### 2023-09-08 paper/main.tex
Overleaf integration

\documentclass[9pt,twocolumn,twoside]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\input{macros}

\templatetype{pnasresearcharticle} % Choose template
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

\begin{document}

\title{Inference for Dynamic and Latent Variable Models via Plug-and-Play Automatically Differentiable Particle Filtering}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a]{Kevin Tan}
\author[b,1]{Edward L. Ionides}
%\author[a]{Author Three}

\affil[a]{University of Pennsylvania}
\affil[b]{University of Michigan}
%\affil[c]{Affiliation Three}

% Please give the surname of the lead author for the running footer
\leadauthor{Tan}

% Please add a significance statement to explain the relevance of your work
\significancestatement{Many scientific models involve highly nonlinear stochastic dynamical systems, sometimes with noisy and incomplete measurements. Under the Markov assumption on system dynamics, previous work has provided methods of performing inference for these models. In particular, prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics. We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation. Our new method dramatically outperforms previous approaches on a challenging problem in epidemiology.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{Please declare any competing interests here.}
%\equalauthors{\textsuperscript{1}A.O.(Author One) contributed equally to this work with A.T. (Author Two) (remove if not applicable).}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: ionides@umich.edu}

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{Sequential Monte Carlo $|$ Automatic Differentiation $|$ Particle Filter $|$ Markov Process $|$ Maximum Likelihood}

\begin{abstract}
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

\firstpage{10}
% Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting. In this example, there are a total of 11 paragraphs on the first page, counting the first level heading as a paragraph. The value {12} represents the number of the paragraph starting the second page. If a paragraph runs over onto the second page, include a bracket with the paragraph line number starting the second page, followed by the paragraph number in curly brackets, e.g. "\firstpage[4]{11}".

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{M}any approaches to inference in highly nonlinear stochastic dynamical systems assume access to the probability density of next states given the current state.

This is a problem in some critical applications, such as disease modeling, where the models are complex enough that obtaining the density is intractable. However, the particle filter, a popular method for solving the filtering problem in partially-observed dynamical systems, does not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm in a feature known as the plug-and-play property.

Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small. Existing methods like the improved iterated filtering algorithm of Ionides et. al. \cite{Ionides_infpomp} converge quickly to a neighborhood of the MLE, but struggle to optimize the last few units of log-likelihood.

\kevin{TODO: (1) Fix beta smoothness to d(n) at no more than O(n), (2) add story to say that no one spotted that it could have been plug-and-play because of e.g. reasons why they would stumble, (3) }

\kevin{Previous literature: Doucet waymo only applicable to policy plus deterministic dynamics}

We propose a hybrid algorithm that u

Unlike IF2, we explicitly characterize our method's rate of convergence

% IF2 is SGD that iterates through the data sequentially in the 1:N sense, and noisy/mini-batch GD in the 1:J sense. 
% ADPF is batch GD in the 1:N sense, and noisy/mini-batch GD in the 1:J sense.

Algorithmic differentiation potentially facilitates numerical optimization, but currently its use for particle filters is limited. We investigate ways to use algorithmic differentiation of particle filters within the confines of the plug-and-play property, with the goal of enhancing current inference capabilities for general POMP models.

\textbf{The issue:} Particle filters provide convenient approaches to evaluating the log-likelihood function for partially observed Markov process (POMP) models. However, using this evaluator to obtain a maximum likelihood parameter estimate can be challenging -- especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small. Empirically, methods such as the improved iterated filtering algorithm (IF2) from Ionides et. al. \cite{Ionides_infpomp} rapidly converge to a neighborhood of the optimum but struggle at finding the exact optimum due to Monte Carlo variance, even with an annealing random walk standard deviation.

A potential solution to this could lie in auto-differentiation (AD). This would allow for the use of first and second-order iterative optimization techniques. However, though AD could potentially facilitate numerical optimization, its use for plug-and-play particle filters has so far been limited. This is potentially because particle filtering methods are inherently non-differentiable due to the resampling step that may take place in between iterations.

\subsection*{The importance of the plug-and-play property:} Performing inference in highly nonlinear stochastic dynamical systems is a challenging problem. Although many methods for inference assume access to the density of state transitions, this is often not available, especially in critical applications like epidemiology.

Basic particle filtering algorithms do not require evaluation of the transition density of the latent Markov process, in a feature known as the \textbf{plug-and-play property} \cite{Breto_timeseriesmech} since it enables an arbitrary model simulator to be plugged into the algorithm. We investigate ways to use algorithmic differentiation of particle filters within the confines of the plug-and-play property, with the goal of enhancing current inference capabilities for general POMP models.

\subsection*{Other potential applications:} This has applications beyond the obvious one of learning model parameters via first or second-order optimization routines. For example, it could be a step towards developing very general Hamiltonian Monte Carlo methods for particle MCMC, as Rosato et. al. \cite{rosato22b} do by using previous work such as \cite{scibior2021dpf, doucet2011sf} (and we conjecture that the seed-fixing derivatives of Rosato et. al. are the same as these as an immediate consequence of section \ref{subsec:scibiorrel}) to differentiate the particle filter.

\section*{Guide to using this template on Overleaf}

Please note that whilst this template provides a preview of the typeset manuscript for submission, to help in this preparation, it will not necessarily be the final publication layout. For more detailed information please see the \href{https://www.pnas.org/page/authors/format}{PNAS Information for Authors}.

If you have a question while using this template on Overleaf, please use the help menu (``?'') on the top bar to search for \href{https://www.overleaf.com/help}{help and tutorials}. You can also \href{https://www.overleaf.com/contact}{contact the Overleaf support team} at any time with specific questions about your manuscript or feedback on the template.

\subsection*{Author Affiliations}

Include department, institution, and complete address, with the ZIP/postal code, for each author. Use lower case letters to match authors with institutions, as shown in the example. PNAS strongly encourages authors to supply an \href{https://orcid.org/}{ORCID identifier} for each author. Individual authors must link their ORCID account to their PNAS account at \href{http://www.pnascentral.org/}{www.pnascentral.org}. For proper authentication, authors must provide their ORCID at submission and are not permitted to add ORCIDs on proofs.

\subsection*{Submitting Manuscripts}

All authors must submit their articles at \href{http://www.pnascentral.org/cgi-bin/main.plex}{PNAScentral}. If you are using Overleaf to write your article, you can use the ``Submit to PNAS'' option in the top bar of the editor window.

\subsection*{Format}

Many authors find it useful to organize their manuscripts with the following order of sections: title, author line and affiliations, keywords, abstract, significance statement, introduction, results, discussion, materials and methods, acknowledgments, and references. Other orders and headings are permitted.

\subsection*{Manuscript Length}

A standard 6-page article is approximately 4,000 words, 50 references, and 4 medium-size graphical elements (i.e., figures and tables). The preferred length of articles remains at 6 pages, but PNAS will allow articles up to a maximum of 12 pages.

\subsection*{References}

References should be cited in numerical order as they appear in text; this will be done automatically via bibtex, e.g. \cite{belkin2002using} and \cite{berard1994embedding,coifman2005geometric,phdthesis,masterthesis}. All references cited in the main text should be included in the main manuscript file.

\subsection*{Data Archival}

PNAS must be able to archive the data essential to a published article. Where such archiving is not possible, deposition of data in public databases, such as GenBank, ArrayExpress, Protein Data Bank, Unidata, and others outlined in the \href{https://www.pnas.org/author-center/editorial-and-journal-policies#materials-and-data-availability}{Information for Authors}, is acceptable.

\subsection*{Language-Editing Services}
Prior to submission, authors who believe their manuscripts would benefit from professional editing are encouraged to use a language-editing service (see list at https://www.pnas.org/author-center/language-editing). PNAS does not take responsibility for or endorse these services, and their use has no bearing on acceptance of a manuscript for publication.

\begin{figure}%[tbhp]
\centering
\includegraphics[width=.8\linewidth]{frog.pdf}
\caption{Placeholder image of a frog with a long example legend to show justification setting.}
\label{fig:frog}
\end{figure}

\begin{table}[t!]
\centering
\caption{Comparison of the fitted potential energy surfaces and ab initio benchmark electronic energy calculations}
\begin{tabular}{lrrr}
Species & CBS & CV & G3 \\
\midrule
1. Acetaldehyde & 0.0 & 0.0 & 0.0 \\
2. Vinyl alcohol & 9.1 & 9.6 & 13.5 \\
3. Hydroxyethylidene & 50.8 & 51.2 & 54.0\\
\bottomrule
\end{tabular}

\addtabletext{nomenclature for the TSs refers to the numbered species in the table.}
\end{table}

\subsection*{Digital Figures}

EPS, high-resolution PDF, and PowerPoint are preferred formats for figures that will be used in the main manuscript. Authors may submit PRC or U3D files for 3D images; these must be accompanied by 2D representations in TIFF, EPS, or high-resolution PDF format. Color images must be in RGB (red, green, blue) mode. Include the font files for any text.

Images must be provided at final size, preferably 1 column width (8.7cm). Figures wider than 1 column should be sized to 11.4cm or 17.8cm wide. Numbers, letters, and symbols should be no smaller than 6 points (2mm) and no larger than 12 points (6mm) after reduction and must be consistent.

Figures and tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.

Figure \ref{fig:frog} shows an example of how to insert a column-wide figure. To insert a figure wider than one column, please use the \verb|\begin{figure*}...\end{figure*}| environment. Figures wider than one column should be sized to 11.4 cm or 17.8 cm wide. Use \verb|\begin{SCfigure*}...\end{SCfigure*}| for a wide figure with side legends.

\begin{SCfigure*}[\sidecaptionrelwidth][t!]
\centering
\includegraphics[width=11.4cm,height=11.4cm]{frog.pdf}
\caption{This legend would be placed at the side of the figure, rather than below it.}\label{fig:side}
\end{SCfigure*}

\subsection*{Tables}
Tables should be included in the main manuscript file and should not be uploaded separately.

\subsection*{Single column equations}

Authors may use 1- or 2-column equations in their article, according to their preference.

To allow an equation to span both columns, use the \verb|\begin{figure*}...\end{figure*}| environment mentioned above for figures.

Note that the use of the \verb|widetext| environment for equations is not recommended, and should not be used.

\begin{figure*}[bt!]
\begin{align*}
(x+y)^3&=(x+y)(x+y)^2\\
       &=(x+y)(x^2+2xy+y^2) \numberthis \label{eqn:example} \\
       &=x^3+3x^2y+3xy^3+x^3.
\end{align*}
\end{figure*}

\subsection*{Supporting Information Appendix (SI)}

Authors should submit SI as a single separate SI Appendix PDF file, combining all text, figures, tables, movie legends, and SI references. SI will be published as provided by the authors; it will not be edited or composed. Additional details can be found in the \href{https://www.pnas.org/authors/submitting-your-manuscript#manuscript-formatting-guidelines}{PNAS Author Center}. The PNAS Overleaf SI template can be found \href{https://www.overleaf.com/latex/templates/pnas-template-for-supplementary-information/wqfsfqwyjtsd}{here}. Refer to the SI Appendix in the manuscript at an appropriate point in the text. Number supporting figures and tables starting with S1, S2, etc.

Authors who place detailed materials and methods in an SI Appendix must provide sufficient detail in the main text methods to enable a reader to follow the logic of the procedures and results and also must reference the SI methods. If a paper is fundamentally a study of a new method or technique, then the methods must be described completely in the main text.

\subsubsection*{SI Datasets}

Supply .xlsx, .csv, .txt, .rtf, or .pdf files. This file type will be published in raw format and will not be edited or composed.

\subsubsection*{SI Movies}

Supply Audio Video Interleave (avi), Quicktime (mov), Windows Media (wmv), animated GIF (gif), or MPEG files. Movie legends should be included in the SI Appendix file. All movies should be submitted at the desired reproduction size and length. Movies should be no more than 10MB in size.

\matmethods{Please describe your materials and methods here. This can be more than one paragraph, and may contain subsections and equations as required.

\subsection*{Subsection for Method}
Example text for subsection.
}

\showmatmethods{} % Display the Materials and Methods section

\acknow{Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.}

\showacknow{} % Display the acknowledgments section

\bibsplit[2]
%Use \bibsplit to split the references from the body of the text. Value "[2]" represents the number of reference in the left column (Note: Please avoid single column figures & tables on this page.)

% Bibliography
\bibliography{pnas-sample}

\end{document}

### 2023-10-01 mop/mop.tex
mop pseudocode added in mop directory

\documentclass[12p]{article}
\usepackage{fullpage}

%% for algorithm pseudocode
\usepackage{enumerate,alltt,xstring}
\usepackage[ruled,noline,linesnumbered]{algorithm2e}
\SetKwFor{For}{for}{do}{end}
\SetKwFor{While}{while}{do}{end}
\SetKwInput{KwIn}{input}
\SetKwInput{KwOut}{output}
\SetKwInput{KwCplx}{complexity}
\SetKwInput{KwIndices}{note}
\SetKwBlock{Begin}{procedure}{}
\DontPrintSemicolon
\newcommand\argequals{{\,=\,}}
\newcommand\data[1]{#1^*}
\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
\newcommand\seq[2]{{#1}\!:\!{#2}}
\newcommand\mydot{{\,\cdot\,}}
\newcommand\given{{\,\vert\,}}
\newcommand\normal{{\mathrm{Normal}}}
\newcommand\vmeasure{V}
\newcommand\emeasure{e}

\newcommand\Time{N}
\newcommand\ttime{n}
\newcommand\Np{J}
\newcommand\np{j}
\newcommand\altNp{q}
\newcommand\prob{\mathrm{Prob}}

\begin{document}

%%%%  PFILTER PSEUDOCODE
\begin{algorithm}[t!]
  \caption{\textbf{Measurement off policy (MOP) SMC}:
    \label{alg:mop-smc}
  }
  %% \KwIn{
  %%   Simulator for $f_{X_n|X_{n-1}}(x_n\given x_{n-1}\giventh\theta)$;
  %%   evaluator for $f_{Y_n|X_n}(y_n\given x_{n}\giventh\theta)$;
  %%   simulator for $f_{X_0}(x_0\giventh\theta)$;
  %%   parameter, $\theta$;
  %%   data, $y^*_{1:N}$;
  %%   number of particles, $J$.
  %% }
  %% \BlankLine
  Initialize filter particles:
  simulate ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\mydot\giventh{\theta}\right)$ for $j$ in $\seq{1}{J}$\;
  Initialize filter weights: $w^{F,\theta}_{0,j}= 1/J$ for $j$ in $\seq{1}{J}$
  \;
  \For{$n$ in $\seq{1}{N}$}{
    Simulate for prediction:
    ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\mydot|{X}_{n-1,j}^{F};{\theta}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{alg:pfilter:step1}\;
    Evaluate measurement density:
    $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for $j$ in $\seq{1}{J}$\;
    Update weights to compensate for resampling:
    $\displaystyle w^{P,\theta}_{n,j}= \frac{w^{F,\theta}_{n-1,j} \, g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{alg:pfilter:weight:update}
    \;
    Conditional likelihood under $\phi$: 
    $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
    \;
    Normalize weights:
    $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
    for $j\ \mathrm{in}\ \seq{1}{J}$
    \;
    Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \nllabel{alg:pfilter:systematic}\;
    Resample:
    set ${X}_{n,j}^{F,\theta}={X}_{n,k_{j}}^{P,\theta}$ and
    $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{P,\theta}$
    for $j$ in $\seq{1}{J}$ \nllabel{alg:pfilter:step2}
    \;
  }
  %% \KwOut{
  %%   Log likelihood estimate, $\loglikMC(\theta)=\sum_{n=1}^N\loglikMC_{n|1:n-1}$;
  %%   filter sample, $X^F_{n,1:J}$, for $n$ in $\seq{1}{N}$.
  %% }
  %% \KwCplx{$\bigO{J}$}
\end{algorithm}

\begin{itemize}
\item MOP-SMC requires that the algorithm is first run at $\phi$, for which it is a vanilla particle filter.  $g^{\phi}_{n,j}$ and $\tilde g^{\phi}_{n,j}$ are computed at this first pass. Then it is run at $\theta$, with the seed kept fixed.

\item Here, the resampling rule for particle $j$ depends on $j$ (and therefore $\phi$ and $X^{P,\phi}_{n,j}$) but not on $\theta$ or  $X^{P,\theta}_{n,j}$. This does not have the standard form for a weighted particle filter, for which we can have general reweighting rules but they are usually a funciton of the location of the particle and the model parameter, $\theta$. That may be why it is harder to see that this is a properly weighted filter. Nevertheless, inspection of the algorithm shows that each particle is properly reweighted to account for its resampling probability, in step~\ref{alg:pfilter:weight:update}, so it should be properly weighted.

\item The final estimate of the likelihood is either
  \begin{equation}
    L(\theta) = \left(\frac{1}{J}\sum_{j=1}^J w^{F,\theta}_{N,j} \right)
    \prod_{n=1}^N L_n^\phi
  \end{equation}
  or
 \begin{equation}
   L^\prime(\theta) = \left(\frac{1}{J}\sum_{j=1}^J w^{P,\theta}_{N,j} \, g^{\theta}_{N,j}\right)
   \prod_{n=1}^{N-1} L_n^\phi
  \end{equation}
 with $L^\prime(\theta)$ presumably having slightly lower variance.

\item Weighted samples representing the filter distribution,
  $f_{X_n|Y_{1:n}}(x_n \given y^*_{1:n} \giventh \theta)$
  are either
  $\big\{\big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big),  j\ \mathrm{in}\ \seq{1}{J}\big\}$
  or
  $\big\{\big(X^{P,\theta}_{n,j}, g^\theta_{n,j} \, w^{P,\theta}_{n,j}\big),  j\ \mathrm{in}\ \seq{1}{J}\big\}$,
  meaning that an expectation over $f_{X_n|Y_{1:n}}(x_n \given y^*_{1:n} \giventh \theta)$ is consistently estimated by a corresponding weighted average of the filter or prediction particles.

  \item As long as \texttt{rprocess} is a continuously differentiable function of $\theta$ for fixed seed, and \texttt{dmeasure} is a continuously differentiable function of $\theta$, and $g^{\phi}_{n,j}\neq 0$, we see that MOP is a continuously differentiable function of $\theta$ for fixed seed. Since it also provided an unbiased estimate of the likelihood, this justifies exchanging the order of differentiation and integration to ensure that its derivative is an unbiased estimate of the deriative of the likelihood.

  \item Taking the derivative with respect to $\theta$ at $\theta=\phi$, step~\ref{alg:pfilter:weight:update} looks very much like the stop gradient approach.

  \end{itemize}

\end{document}

### 2023-10-01 mop/mop.tex
mop typo

@@
   simulate ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\mydot\giventh{\theta}\right)$ for $j$ in $\seq{1}{J}$\;
-  Initialize filter weights: $w^{F,\theta}_{0,j}= 1/J$ for $j$ in $\seq{1}{J}$
+  Initialize relative weights: $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$
   \;
@@
     $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for $j$ in $\seq{1}{J}$\;
-    Update weights to compensate for resampling:
+    Update relative weights to compensate for resampling:
     $\displaystyle w^{P,\theta}_{n,j}= \frac{w^{F,\theta}_{n-1,j} \, g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{alg:pfilter:weight:update}

### 2023-10-07 mop/mop.tex
mop theory sketch

@@
 \item The final estimate of the likelihood is either
-  \begin{equation}
+  \begin{equation} \label{mop:lik1}
     L(\theta) = \left(\frac{1}{J}\sum_{j=1}^J w^{F,\theta}_{N,j} \right)
@@
 
+Here's an outline of an argument explaining why MOP is properly weighted.
+
+\noindent {\bf Lemma 1}. {\it
+Suppose that $\{(X_j,u_j),j=1,\dots,J\}$ is properly weighted for $f_X$, meaning that, if $X\sim f_X$,
+\[
+E[h(X)] \approx \sum_{j=1}^J \frac{w_j}{\sum_{k=1}^J u_k} h(X_j),
+\]
+for some appropriate formalization of $\approx$.
+Now, let $\{(Y_j,v_j),j=1,\dots,J\}$ be a sample drawn from $\{(X_j,u_j)\}$ where $(X_j,u_j)$ is represented, on average, $\pi_j J$ times.
+This could amount to binomial resampling having $J$ draws each with probability $p_j$, or systematic resampling.
+Suppose
+\[
+(Y_j,v_j) = \big(X_{a(j)},u_{a(j)}/\pi_{a(j)}\big),
+\]
+where $a$ is called the ancestor function. Then, $\{(Y_j,v_j),j=1,\dots,J\}$ is als properly weighted for $f_X$.
+}
+
+Notably, Lemma~1 permits $\pi_{1:J}$ to depend on $\{(X_j,u_j)\}$ as long as the resampling is carried out independently of $\{(X_j,u_j)\}$ conditional on $\pi_{1:J}$.
+
+\noindent {\bf Lemma 2}. {\it
+  Suppose that $Z_j \sim f_{Z|X}(\cdot | X_j)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
+Then, $\{(Z_j,u_j)\}$ is properly weighted for $f_Z$.
+}
+
+\noindent {\bf Lemma 3}. {\it
+Suppose that $(X^\prime_j,u^\prime_j) = \big(X_j,u_j\, f_{Z|X}(z^*|X_j)\big)$. Then, $\{(X^\prime_j,u^\prime_j)\}$ is properly weighted for $f_{X|Z}(\cdot | z^*)$.
+}
+
+Recursively applying Lemmas~1, 2 and~3, we obtain that 
+%to step~\ref{alg:pfilter:step1}, Lemma~2 step~ {alg:pfilter:weight:update} and Lemma~3 to step~\ref{alg:pfilter:step2} we obtain that
+the MOP filter is properly weighted.
+Specifically, suppose inductively that $\big\{\big(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
+Then, Lemma~2 tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big)\big\}$ is properly weighted for $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+Lemma~3 tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} \big)\big\}$ is therefore properly weighted for  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+Lemma~1 guarantees that the resampling rule, given by 
+\[
+\big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,j} g^\theta_{n,j}\big/ g^\phi_{n,j}\big),
+\]
+with resampling weights proportional to $g^\phi_{n,j}$, is therefore also properly weighted for $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+
+This has addressed filtering, but not quite yet the likelihood evaluation.
+
+A properly weighted conditional likelihood estimate from MOP (give the argument above) is
+\[
+f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1_n-1}^*;\theta) \approx \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}.
+\]
+We write the numerator as
+\[
+L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
+\]
+Using Lemma~1, we see this is properly estimated by
+\[
+L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j}.
+\]
+This corresponds to
+\[
+L^\theta = \prod_{n=1}^N L^\phi_n \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J w^{P,\theta}_{n,j}}.
+\]
+Since $w^{F,\theta}_{n,j}=w^{P,\theta}_{n+1,j}$, this is a telescoping product. The remaining terms are
+$\sum_{j=1}^J w^{P,\theta}_{0,j} = 1$ on the denominator and $\sum_{j=1}^J w^{F,\theta}_{N,j}$ on the numerator.
+This derives the MOP estimate in (\ref{mop:lik1}).
+
 \end{document}

### 2023-10-08 paper/main.tex
abstract draft

@@
 % Please add a significance statement to explain the relevance of your work
-\significancestatement{Many scientific models involve highly nonlinear stochastic dynamical systems, sometimes with noisy and incomplete measurements.
+\significancestatement{Many scientific models involve highly nonlinear stochastic dynamical systems which can be observed only via noisy and incomplete measurements.
 Under the Markov assumption on system dynamics, previous work has provided methods of performing inference for these models.
@@
 \begin{abstract}
+  Automatic differentiation has driven recent advances in machine learing, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods.
+  This progress has required simultaneous advances in algorithms, software and hardware.  
+  Partially observed nonlinear stochastic dynamic systems have proved resistant to these techniques; despite various attempts, widely applicable methods have not yet emerged.
+  We present a new approach which is applicable to a general class of models, possesses a theoretical foundation, and is demonstrated to beat current state-of-the-art methods on a challenging scientific benchmark problem.
+  Our algorithm is compatible with parallel computation on a graphical processing unit, and enjoys the plug-and-play property that its software implementation requires only a simulator for the scientific dynamic model as input (in addition to data, and a measurement model).
+
 \end{abstract}

### 2023-10-10 mop/mop.tex
typo
mop-alpha

@@
 \usepackage{fullpage}
+\parskip 2mm
 
@@
 
-%%%%  PFILTER PSEUDOCODE
-\begin{algorithm}[t!]
+\title{Automatic differentiation via off-policy particle filters}
+\date{\today}
+\maketitle
+
+\begin{abstract}
+Kevin has checked numerically that the measurement off policy (MOP) particle filter has a fixed-seed derivative matching Scibior \& Wood (SW). 
+Here, we present the MOP algorithm in order to check that it is properly weighted.
+The argument for this remains somewhat heuristic, reasoning about properly weighted samples in a similar way to Jun Liu's 2001 book.
+We also introduce a MOP-$\alpha$ algorithm whose fixed-seed derivative interpolates between the basic ignore-resampling-in-derivative (IRID) PF ($\alpha=0$) and 
+MOP ($\alpha=1$).
+\end{abstract}
+
+\section{The Measurement off policy (MOP) particle filter}
+
+%%%%  MOP PFILTER PSEUDOCODE
+%\begin{algorithm}[t!]
+\begin{algorithm}
   \caption{\textbf{Measurement off policy (MOP) SMC}:
@@
   }
-  %% \KwIn{
-  %%   Simulator for $f_{X_n|X_{n-1}}(x_n\given x_{n-1}\giventh\theta)$;
-  %%   evaluator for $f_{Y_n|X_n}(y_n\given x_{n}\giventh\theta)$;
-  %%   simulator for $f_{X_0}(x_0\giventh\theta)$;
-  %%   parameter, $\theta$;
-  %%   data, $y^*_{1:N}$;
-  %%   number of particles, $J$.
-  %% }
-  %% \BlankLine
   Initialize filter particles:
@@
     Simulate for prediction:
-    ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\mydot|{X}_{n-1,j}^{F};{\theta}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{alg:pfilter:step1}\;
+    ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\mydot|{X}_{n-1,j}^{F};{\theta}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{mop:step1}
+    \;
+    Prediction weights: $w_{n,j}^{P,\theta} = w_{n-1,j}^{F,\theta}$ for $j\ \mathrm{in}\ \seq{1}{J}$ 
+    \;
     Evaluate measurement density:
     $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for $j$ in $\seq{1}{J}$\;
-    Update relative weights to compensate for resampling:
-    $\displaystyle w^{P,\theta}_{n,j}= \frac{w^{F,\theta}_{n-1,j} \, g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{alg:pfilter:weight:update}
+    Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
     \;
@@
     $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
+    \nllabel{mop:Lphi}
     \;
@@
     \;
-    Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \nllabel{alg:pfilter:systematic}\;
-    Resample:
-    set ${X}_{n,j}^{F,\theta}={X}_{n,k_{j}}^{P,\theta}$ and
-    $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{P,\theta}$
-    for $j$ in $\seq{1}{J}$ \nllabel{alg:pfilter:step2}
+    Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \nllabel{mop:systematic}\;
+    Resample particles: ${X}_{n,j}^{F,\theta}={X}_{n,k_{j}}^{P,\theta}$
+    \;
+    Filter weights corrected for resampling:
+    $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{mop:weight:update}
+    \;
+    Resample filter weights:
+    $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
+    for $j$ in $\seq{1}{J}$ \nllabel{mop:step2}
+    \;
+    After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
     \;
   }
-  %% \KwOut{
-  %%   Log likelihood estimate, $\loglikMC(\theta)=\sum_{n=1}^N\loglikMC_{n|1:n-1}$;
-  %%   filter sample, $X^F_{n,1:J}$, for $n$ in $\seq{1}{N}$.
-  %% }
-  %% \KwCplx{$\bigO{J}$}
 \end{algorithm}
@@
 That may be why it is harder to see that this is a properly weighted filter.
-Nevertheless, inspection of the algorithm shows that each particle is properly reweighted to account for its resampling probability, in step~\ref{alg:pfilter:weight:update}, so it should be properly weighted.
+Nevertheless, inspection of the algorithm shows that each particle is properly reweighted to account for its resampling probability, in step~\ref{mop:weight:update}, so it should be properly weighted.
 
-\item The final estimate of the likelihood is either
-  \begin{equation} \label{mop:lik1}
-    L(\theta) = \left(\frac{1}{J}\sum_{j=1}^J w^{F,\theta}_{N,j} \right)
+\item The final estimate of the likelihood is either based on the after-resampling conditional likelihood estimate
+  \begin{equation} \label{mop:likA}
+    L^A(\theta) = \left(\frac{1}{J}\sum_{j=1}^J w^{F,\theta}_{N,j} \right)
     \prod_{n=1}^N L_n^\phi
   \end{equation}
-  or
- \begin{equation}
-   L^\prime(\theta) = \left(\frac{1}{J}\sum_{j=1}^J w^{P,\theta}_{N,j} \, g^{\theta}_{N,j}\right)
-   \prod_{n=1}^{N-1} L_n^\phi
+  or the before-resampling estimate,
+ \begin{equation}\label{mop:likB}
+   L^B(\theta) = \prod_{n=1}^N \frac{\sum_{j=1}^J w^{P,\theta}_{N,j} \, g^{\theta}_{N,j}}{\sum_{j=1}^J w^{P,\theta}_{N,j}}.
   \end{equation}
- with $L^\prime(\theta)$ presumably having slightly lower variance.
+ with $L^B(\theta)$ presumably having slightly lower variance.
 
@@
 
-  \item Taking the derivative with respect to $\theta$ at $\theta=\phi$, step~\ref{alg:pfilter:weight:update} looks very much like the stop gradient approach.
+  \item Taking the derivative with respect to $\theta$ at $\theta=\phi$, step~\ref{mop:weight:update} looks very much like the stop gradient approach.
 
@@
 \]
-where $a$ is called the ancestor function. Then, $\{(Y_j,v_j),j=1,\dots,J\}$ is als properly weighted for $f_X$.
+where $a$ is called the ancestor function. Then, $\{(Y_j,v_j),j=1,\dots,J\}$ is properly weighted for $f_X$.
 }
@@
 Recursively applying Lemmas~1, 2 and~3, we obtain that 
-%to step~\ref{alg:pfilter:step1}, Lemma~2 step~ {alg:pfilter:weight:update} and Lemma~3 to step~\ref{alg:pfilter:step2} we obtain that
+%to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
 the MOP filter is properly weighted.
@@
 
-This has addressed filtering, but not quite yet the likelihood evaluation.
+This has addressed filtering, but not quite yet the likelihood evaluation. For this we use the following lemma.
 
-A properly weighted conditional likelihood estimate from MOP (give the argument above) is
-\[
-f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1_n-1}^*;\theta) \approx \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}.
-\]
-We write the numerator as
+\noindent {\bf Lemma 4}. {\it
+  $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1_n-1}^*;\theta)$ is properly estimated by either the before-resampling estimate,
+\begin{equation}\label{L1}
+L_n^{B,\theta} =  \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
+\end{equation}
+or by the after-resampling estimate,
+\begin{equation}\label{L2}
+L_n^{A,\theta} = L^\phi_n \frac{\sum_{j=1}^Jw^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}.
+\end{equation}
+where $L^\phi_n$ comes from step~\ref{mop:Lphi}.
+}
+
+Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ is properly weighted for $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+To see  (\ref{L2}),
+we write the numerator of (\ref{L1}) as
 \[
@@
 \[
-L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j}.
+L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},
 \]
-This corresponds to
+from which we obtain (\ref{L2}).
+
+Using Lemma~4, we obtain a likelihood estimate,
 \[
-L^\theta = \prod_{n=1}^N L^\phi_n \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J w^{P,\theta}_{n,j}}.
+L^{A,\theta} = \prod_{n=1}^N \left( L^\phi_n \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J w^{P,\theta}_{n,j}}\right).
 \]
 Since $w^{F,\theta}_{n,j}=w^{P,\theta}_{n+1,j}$, this is a telescoping product. The remaining terms are
-$\sum_{j=1}^J w^{P,\theta}_{0,j} = 1$ on the denominator and $\sum_{j=1}^J w^{F,\theta}_{N,j}$ on the numerator.
-This derives the MOP estimate in (\ref{mop:lik1}).
+$\sum_{j=1}^J w^{P,\theta}_{0,j} = J$ on the denominator and $\sum_{j=1}^J w^{F,\theta}_{N,j}$ on the numerator.
+This derives the MOP estimate in (\ref{mop:likA}).
+
+$L^{B,\theta}$ should generally be preferred, since there is no reason to include the extra variability from resampling when calculating the conditional log likelihood, but it lacks the nice telescoping product.
+
+\section{MOP-$\alpha$}
+
+%%%%  MOP(alpha) PFILTER PSEUDOCODE
+\begin{algorithm}[t!]
+  \caption{\textbf{Measurement off policy (MOP-$\alpha$) SMC}:
+    \label{alg:mop-alpha}
+  }
+  Initialize filter particles:
+  simulate ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\mydot\giventh{\theta}\right)$ for $j$ in $\seq{1}{J}$\;
+  Initialize relative weights: $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$
+  \;
+  \For{$n$ in $\seq{1}{N}$}{
+    Simulate for prediction:
+    ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\mydot|{X}_{n-1,j}^{F};{\theta}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{mop-alpha:step1}
+    \;
+    Prediction weights with discounting: $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
+    \nllabel{mop-alpha:discount}
+    \;
+    Evaluate measurement density:
+    $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for $j$ in $\seq{1}{J}$\;
+    Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
+    \;
+    Conditional likelihood under $\phi$: 
+    $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
+    \nllabel{mop-alpha:Lphi}
+    \;
+    Normalize weights:
+    $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
+    for $j\ \mathrm{in}\ \seq{1}{J}$
+    \;
+    Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \nllabel{mop-alpha:systematic}\;
+    Resample particles: ${X}_{n,j}^{F,\theta}={X}_{n,k_{j}}^{P,\theta}$
+    \;
+    Filter weights corrected for resampling:
+    $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{mop-alpha:weight:update}
+    \;
+    Resample filter weights:
+    $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
+    for $j$ in $\seq{1}{J}$ \nllabel{mop-alpha:step2}
+    \;
+    After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
+    \;
+  }
+\end{algorithm}
+
+\begin{itemize}
+\item MOP has been rewritten from an earlier draft so that now the only difference between MOP and MOP-$\alpha$ is the discounting exponent $\alpha$ in step~\ref{mop-alpha:discount}.
+\item When $\alpha=1$ we recover MOP.
+\item When $\alpha=0$, $w^{P,\theta}_{n,j}=1$ and so $L^{B,\theta}_n$ corresponds to an equally weighted conditional likelihood, as in a regular particle filter, but with resampling carried out according to $\phi$ not $\theta$.
+  When calculating a derivative with respect to $\theta$ at $\theta=\phi$, this corresponds to ignoring differentiation through the resampling, i.e., the fixed-seed derivative of the PF algorithm.
+  To see this, we notice that the fixed seed derivative which ``ignores'' resampling is not really ignoring it; in an infinitesimal neighborhood of $\phi$, the resampling is constant and so the derivative is zero.
+  The problem is not that the fixed-seed derivative is missing something, just that we cannot correctly move the derivative through an expectation for a discontinuous function.
+\item When $\alpha < 1$, the product
+\[
+  L^{A,\alpha}(\theta)=\prod_{n=1}^N L^{A,\theta,\alpha}_n
+\]
+no longer has a telescoping cancellation. 
+\end{itemize}
+
+\section{Unresolved questions}
+
+\begin{itemize}
+\item $L^{A,\theta}_n$ and $L^{B,\theta}_n$ are asymptotically equivalent as $J$ increases, but are not identical.
+  However, both are identical for a particle filter (i.e., MOP when $\theta=\phi$).
+  Presumably, this could allow their fixed-seeed derivatives to be the same at $\theta=\phi$, though it is not obvious if this is true.
+  In that case, we would have an alternative explanation of the telescoping identity of Scibior \& Wood.
+
+\item It seems possible to carry out an extra numerical pass where one recalculates $L^A(\theta)$ taking advantage of the cancellation, which could make the derivative faster and/or more memory efficient, depending on what is the bottleneck.
+This may not be an immediate priority.
+
+  \item The cancellation difference between  $L^{A,\theta}_n$ and $L^{B,\theta}_n$ is moot for $\alpha<1$, so there's not much reason not to use $L^{B,\theta}_n$.
+
+\end{itemize}
 

### 2023-10-16 mop/mop.tex
DOP-alpha pseudocode

@@
 \newcommand\prob{\mathrm{Prob}}
+\newcommand\code[1]{\texttt{#1}}
 
@@
 
+\section{DOP-$\alpha$}
+
+Here, we consider a doubly off policy particle filter, meaning that both \code{rprocess} and \code{rmeasure} are computed at $\phi$ and particles are reweighted to correspond to $\theta$.
+
+%%%%  DOP(alpha) PFILTER PSEUDOCODE
+\begin{algorithm}[t!]
+  \caption{\textbf{Doubly off policy (DOP-$\alpha$) SMC}:
+    \label{alg:mop-alpha}
+  }
+  Initialize filter particles:
+  simulate ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\mydot\giventh{\theta}\right)$ for $j$ in $\seq{1}{J}$\;
+  Initialize relative weights: $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$
+  \;
+  \For{$n$ in $\seq{1}{N}$}{
+    Prediction weights with discounting: $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
+    \nllabel{dop-alpha:discount}
+    \;
+    Simulate for prediction:
+    ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\mydot|{X}_{n-1,j}^{F};{\phi}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{dop-alpha:step1}
+    \;
+    Adjust weights: $\displaystyle w_{n,j}^{P,\theta} = w_{n,j}^{P,\theta} \times
+    \frac{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\theta}|{X}_{n-1,j}^{F};{\theta}\big)}{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\theta}|{X}_{n-1,j}^{F};{\phi}\big)}$
+    for $j\ \mathrm{in}\ \seq{1}{J}$
+    \nllabel{dop-alpha:dproc}
+    \;
+    Evaluate measurement density:
+    $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for $j$ in $\seq{1}{J}$\;
+    Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
+    \;
+    Conditional likelihood under $\phi$: 
+    $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
+    \nllabel{dop-alpha:Lphi}
+    \;
+    Normalize weights:
+    $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
+    for $j\ \mathrm{in}\ \seq{1}{J}$
+    \;
+    Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \nllabel{dop-alpha:systematic}\;
+    Resample particles: ${X}_{n,j}^{F,\theta}={X}_{n,k_{j}}^{P,\theta}$
+    \;
+    Filter weights corrected for resampling:
+    $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{dop-alpha:weight:update}
+    \;
+    Resample filter weights:
+    $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
+    for $j$ in $\seq{1}{J}$ \nllabel{dop-alpha:step2}
+    \;
+    After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
+    \;
+  }
+\end{algorithm}
+
+\begin{itemize}
+\item In order to match the direct derivative for a discrete \code{rprocess} with $\alpha=0$, we would carry out the discounting in step~\ref{dop-alpha:discount} after the weight adjustment in step~\ref{dop-alpha:dproc}.
+  However, if we want to make $\alpha=0$ correspond to a reasonable estimate with no weight memory (i.e., strong mixing) then we should construct the algorithms as stated.
+\item ${X}_{n,j}^{P,\theta}$ in fact depends on $\phi$ not $\theta$ in this algorithm. Rename?
+\end{itemize}
+
 \end{document}

### 2023-10-16 mop/mop.tex

@@
 \newcommand\data[1]{#1^*}
-\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
+\newcommand\giventh{{\hspace{0.5mm};\hspaxce{0.5mm}}}
 \newcommand\seq[2]{{#1}\!:\!{#2}}
@@
 \begin{abstract}
-Kevin has checked numerically that the measurement off policy (MOP) particle filter has a fixed-seed derivative matching Scibior \& Wood (SW). 
+Kevin has checked theoretically that the measurement off policy (MOP) particle filter has a fixed-seed derivative matching Scibior \& Wood (SW). 
 Here, we present the MOP algorithm in order to check that it is properly weighted.
@@
 \section{The Measurement off policy (MOP) particle filter}
+
+DOP has the entire system evolve according to $\phi$, and we differentiate through the likelihood ratios, which are smooth conditional on the resampling being the same (which in turn happens when $\theta \approx \phi$).
+
+POP has the process system evolve according to $\theta$, and the resampling evolve according to $\phi$.
+In essence, we offload the derivative with respect to the process model to the reparametrization trick.
+
+Bias in dropping the resampling gradient increases as process mixes slower.
+ADPF has a long memory, and the variance in the weights add up, and there is significant instability.
+However, by the time the system mixes, ancestral trajectories don't matter as much, so we shouldn't bother with estimating gradients that rely on more than one timestep of the conditional log-likelihood.
+
+Scientifically, this is great that no matter what you do with the gradient it works out well.
+
+Implement MOP-$\alpha$ after the paper is mostly written.
 
@@
 \item MOP has been rewritten from an earlier draft so that now the only difference between MOP and MOP-$\alpha$ is the discounting exponent $\alpha$ in step~\ref{mop-alpha:discount}.
-\item When $\alpha=1$ we recover MOP.
+\item When $\alpha=1$ we recover MOP. In particular, $L^{A,\theta}_n$ when $\theta=\phi$ and $\theta=1$ recovers Scibior and Wood. 
 \item When $\alpha=0$, $w^{P,\theta}_{n,j}=1$ and so $L^{B,\theta}_n$ corresponds to an equally weighted conditional likelihood, as in a regular particle filter, but with resampling carried out according to $\phi$ not $\theta$.
-  When calculating a derivative with respect to $\theta$ at $\theta=\phi$, this corresponds to ignoring differentiation through the resampling, i.e., the fixed-seed derivative of the PF algorithm.
+  When calculating a derivative of $L_n^{B, \theta}$ when $\alpha=0$ with respect to $\theta$ at $\theta=\phi$, this corresponds to ignoring differentiation through the resampling, i.e., the fixed-seed derivative of the PF algorithm.
   To see this, we notice that the fixed seed derivative which ``ignores'' resampling is not really ignoring it; in an infinitesimal neighborhood of $\phi$, the resampling is constant and so the derivative is zero.
   The problem is not that the fixed-seed derivative is missing something, just that we cannot correctly move the derivative through an expectation for a discontinuous function.
+\item Of course, the difference between the before-resampling and after-resampling likelihoods is immaterial. If the functions match, then the gradients match. 
 \item When $\alpha < 1$, the product

### 2023-10-16 mop/mop.tex
Updates from Overleaf

@@
 \newcommand\prob{\mathrm{Prob}}
-\newcommand\code[1]{\texttt{#1}}
 
@@
 
-\section{DOP-$\alpha$}
-
-Here, we consider a doubly off policy particle filter, meaning that both \code{rprocess} and \code{rmeasure} are computed at $\phi$ and particles are reweighted to correspond to $\theta$.
-
-%%%%  DOP(alpha) PFILTER PSEUDOCODE
-\begin{algorithm}[t!]
-  \caption{\textbf{Doubly off policy (DOP-$\alpha$) SMC}:
-    \label{alg:mop-alpha}
-  }
-  Initialize filter particles:
-  simulate ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\mydot\giventh{\theta}\right)$ for $j$ in $\seq{1}{J}$\;
-  Initialize relative weights: $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$
-  \;
-  \For{$n$ in $\seq{1}{N}$}{
-    Prediction weights with discounting: $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
-    \nllabel{dop-alpha:discount}
-    \;
-    Simulate for prediction:
-    ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\mydot|{X}_{n-1,j}^{F};{\phi}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{dop-alpha:step1}
-    \;
-    Adjust weights: $\displaystyle w_{n,j}^{P,\theta} = w_{n,j}^{P,\theta} \times
-    \frac{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\theta}|{X}_{n-1,j}^{F};{\theta}\big)}{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\theta}|{X}_{n-1,j}^{F};{\phi}\big)}$
-    for $j\ \mathrm{in}\ \seq{1}{J}$
-    \nllabel{dop-alpha:dproc}
-    \;
-    Evaluate measurement density:
-    $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for $j$ in $\seq{1}{J}$\;
-    Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-    \;
-    Conditional likelihood under $\phi$: 
-    $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
-    \nllabel{dop-alpha:Lphi}
-    \;
-    Normalize weights:
-    $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
-    for $j\ \mathrm{in}\ \seq{1}{J}$
-    \;
-    Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \nllabel{dop-alpha:systematic}\;
-    Resample particles: ${X}_{n,j}^{F,\theta}={X}_{n,k_{j}}^{P,\theta}$
-    \;
-    Filter weights corrected for resampling:
-    $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{dop-alpha:weight:update}
-    \;
-    Resample filter weights:
-    $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
-    for $j$ in $\seq{1}{J}$ \nllabel{dop-alpha:step2}
-    \;
-    After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-    \;
-  }
-\end{algorithm}
-
-\begin{itemize}
-\item In order to match the direct derivative for a discrete \code{rprocess} with $\alpha=0$, we would carry out the discounting in step~\ref{dop-alpha:discount} after the weight adjustment in step~\ref{dop-alpha:dproc}.
-  However, if we want to make $\alpha=0$ correspond to a reasonable estimate with no weight memory (i.e., strong mixing) then we should construct the algorithms as stated.
-\item ${X}_{n,j}^{P,\theta}$ in fact depends on $\phi$ not $\theta$ in this algorithm. Rename?
-\end{itemize}
-
 \end{document}

### 2023-10-17 mop/mop.tex
two typos from last meeting

--- 2023-10-16 mop/mop.tex
@@
 \[
-E[h(X)] \approx \sum_{j=1}^J \frac{w_j}{\sum_{k=1}^J u_k} h(X_j),
+E[h(X)] \approx \sum_{j=1}^J \frac{u_j}{\sum_{k=1}^J u_k} h(X_j),
 \]
@@
 Now, let $\{(Y_j,v_j),j=1,\dots,J\}$ be a sample drawn from $\{(X_j,u_j)\}$ where $(X_j,u_j)$ is represented, on average, $\pi_j J$ times.
-This could amount to binomial resampling having $J$ draws each with probability $p_j$, or systematic resampling.
+This could amount to binomial resampling having $J$ draws each with probability $\pi_j$, or systematic resampling.
 Suppose

### 2023-10-29 mop/mop.tex
Updates from Overleaf

--- 2023-10-16 mop/mop.tex
@@
   Initialize filter particles:
-  simulate ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\mydot\giventh{\theta}\right)$ for $j$ in $\seq{1}{J}$\;
+  simulate ${X}_{0,j}^{F,\phi}\sim {f}_{{X}_{0}}\left(\mydot\giventh{\phi}\right)$ for $j$ in $\seq{1}{J}$\;
   Initialize relative weights: $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$
@@
     Simulate for prediction:
-    ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\mydot|{X}_{n-1,j}^{F};{\phi}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{dop-alpha:step1}
+    ${X}_{n,j}^{P,\phi}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\mydot|{X}_{n-1,j}^{F};{\phi}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{dop-alpha:step1}
     \;
     Adjust weights: $\displaystyle w_{n,j}^{P,\theta} = w_{n,j}^{P,\theta} \times
-    \frac{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\theta}|{X}_{n-1,j}^{F};{\theta}\big)}{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\theta}|{X}_{n-1,j}^{F};{\phi}\big)}$
+    \frac{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\phi}|{X}_{n-1,j}^{F};{\theta}\big)}{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\phi}|{X}_{n-1,j}^{F};{\phi}\big)}$
     for $j\ \mathrm{in}\ \seq{1}{J}$
@@
     Evaluate measurement density:
-    $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for $j$ in $\seq{1}{J}$\;
+    $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\phi}\giventh{\theta})$ for $j$ in $\seq{1}{J}$\;
     Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
@@
     Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \nllabel{dop-alpha:systematic}\;
-    Resample particles: ${X}_{n,j}^{F,\theta}={X}_{n,k_{j}}^{P,\theta}$
+    Resample particles: ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$
     \;
@@
   However, if we want to make $\alpha=0$ correspond to a reasonable estimate with no weight memory (i.e., strong mixing) then we should construct the algorithms as stated.
-\item ${X}_{n,j}^{P,\theta}$ in fact depends on $\phi$ not $\theta$ in this algorithm. Rename?
+\item ${X}_{n,j}^{P,\theta}$ in fact depends on $\phi$ not $\theta$ in this algorithm. Renamed accordingly.
 \end{itemize}

### 2023-10-31 mop/draft.tex
Updates from Overleaf

\documentclass{article}
\usepackage[margin=2cm]{geometry}
\usepackage{natbib}

\usepackage{enumitem}
\setlist{nosep} % to reduce space in lists

\setlength{\parskip}{0.5em}
\setlength{\parindent}{0in}

\input{paper/macros}

\newcommand{\off}{\operatorname{off}}
\newcommand{\on}{\operatorname{on}}

\title{Automatic Differentiation to Accelerate Inference for Partially Observed Stochastic Processes}
\author{Kevin Tan, Edward L. Ionides}
\date{}

\comments=1

\begin{document}
\maketitle
\begin{abstract}

Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods. Partially observed nonlinear stochastic dynamic systems have proved resistant to AD techniques due to (1) the requirement for simulation-based inference that does not require access to the system's transition probabilities and (2) the issue that widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters. We propose a new representation that embeds existing methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff. Further, we develop optimization algorithms suited to the Monte Carlo properties of the derivative estimate, including a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation. Promising numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from iterated filtering can beat current state-of-the-art methods on a challenging scientific benchmark problem.
\end{abstract}

\section{Introduction}

Maximum-likelihood estimation for inference in partially observed stochastic processes, also known as, continuous-state continuous-time hidden Markov models (HMMs), partially-observed Markov processes (POMPs), or partially-observed stochastic nonlinear dynamical systems, is a challenging problem that faces several challenges: 
\begin{enumerate}
    \item Intractable likelihood functions, bypassed with the approaches of simulated likelihood and likelihood-free inference. 
    \item The desire for simulation-based inference without access to the transition density of the latent Markov process, with various methods created by \cite{welch2009abc, wood2010sl, doucet2010pmcmc, Ionides_infdynsys, Ionides_infpomp}. 
    \item Accurate parameter estimation in the presence of significant Monte-Carlo noise in the likelihood estimate.
\end{enumerate}

Many approaches to inference in these scenarios (such as the EM algorithm, the various Kalman filter variants, and MCMC) either struggle with intractable likelihood functions when the models are complex, or assume access to the probability density of next states given the current state. This is a problem in some critical applications, such as disease modeling, where the models are complex enough such that even obtaining the transition density is an intractable problem. However, the particle filter with the bootstrap proposal, a popular method for solving the filtering problem in partially-observed dynamical systems, provides an unbiased estimate of the likelihood (\cite{delmoral2004feynman}) without requiring evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.

This is the backbone behind the only simulation-based full-information maximum likelihood method available in the literature thus far, the improved iterated filtering algorithm of \citet{Ionides_infpomp}. \citet{Ionides_infpomp} utilize the bootstrap filter for perform maximum likelihood estimation with an iterated perturbed Bayes map, successfully tackling challenging problems in epidemiology that available Bayesian methodology fails to solve.

Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small. For example, though in practice the algorithm of \citet{Ionides_infpomp} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood. \citet{Ionides_mcap} attempt to address the third problem, albeit with a computationally-expensive approach.

\subsection{Automatic Differentiation for Particle Filters}

Recent advances in automatic differentiation (AD) for particle filters (\cite{blei2018vsmc, jon2018diffpf, corenflos2021ot, scibior2021dpf, doucet2022particlebased}) have drawn attention to AD as an alternative tool for inference in partially observed stochastic processes with the particle filter. However, these existing approaches either are not yet compatible with the desire for simulation-based inference, or are computationally expensive.

\citet{blei2018vsmc} and \citet{jon2018diffpf} drop high-variance resampling terms from the gradient, leading to bias that does not vanish asymptotically according to \citet{corenflos2021ot}. Though the approach of \citet{corenflos2021ot} is promising, it is computationally expensive. \citet{doucet2022particlebased} provide another promising fixed-lag smoothing-based approach that is similar in spirit to our method (MOP-$\alpha$), but requires either (1) access to the process model transition densities or (2) the special case where the transition density factors into a policy that selects actions (in the reinforcement learning sense) and a deterministic model that maps states and actions to next-timestep states.

Finally, \citet{scibior2021dpf} demonstrate that the estimators derived in \citet{doucet2011sf} can be attained with AD, albeit with a simple tweak to the particle filter that does not modify its likelihood estimates. Though their method is not fully compatible with simulation-based inference, we show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.

\subsection{Our Contributions}

Previous research on AD for particle filters has struggled with various issues: 
\begin{enumerate}
    \item Bias, if the discontinuous nature of particle resampling is ignored.
    \item High Monte Carlo variance, if continuity corrections lead to numerical instability.
    \item High computational cost, for algorithms which involve pairwise interactions between particles, marginalization, smoothing or optimal transport.
    \item Reduced applicability for algorithms which lose or fail to take advantage of the simulation-based capabilities of the bootstrap filter.
\end{enumerate}

We develop a new approach to statistical inference via AD for particle filters which addresses these concerns. We develop a new theoretical framework that encompasses existing methods, and addresses the seemingly incompatible paradox of differentiating through a Monte-Carlo algorithm with discontinuous resampling.

This is done through the construction of a smooth extension to the particle filter, as well as a \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) family of algorithms that (informally) encompasses the special case of AD of a vanilla PF (recovering the gradient estimator of \citet{blei2018vsmc}) when $\alpha=0$, and the DPF algorithm from \citet{scibior2021dpf} (that recovers the gradient estimator from \citet{doucet2011sf}) when $\alpha=1$. When $\alpha \in (0,1)$, this provides opportunities for optimizing a bias/variance tradeoff.

We show these off-policy particle filters are properly weighted, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties. To cap off the theoretical results, we also derive a linear convergence rate for gradient methods on the particle filter in the presence of strong convexity, providing support for numerical optimization with AD for the particle filter.

In practice, we propose a hybrid algorithm that we call \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts gradient ascent (or a similar iterative first or second-order gradient method) with a coarse solution obtained from a few rounds of iterated filtering. While this requires a differentiable simulator compatible with the reparametrization trick (as previous literature such as \citet{corenflos2021ot} also require), future work will deal with workarounds inspired by our framework that use likelihood ratios of transition densities in place of this. Promising numerical results indicate that IFAD beats IF2 (and by the transitive property, IF1, the Liu-West filter, and particle MCMC) on a challenging scientific benchmark problem.

While the Monte-Carlo Adjusted Profile (MCAP) from \citet{Ionides_mcap} provides a profile likelihood confidence interval that takes into account both Monte Carlo profile error and statistical uncertainty in the likelihood, it is computationally inefficient. \kevin{TODO} We provide a method for incorporating the MCAP methodology in our IFAD algorithm that provides a more computationally efficient profile likelihood confidence interval.

Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king2016pomp, king2017pompmanual}. The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{jax} Python package from \citet{jax2018github}.

\subsection{Significance Statement}
Many scientific models involve highly nonlinear stochastic dynamical systems which can be observed only via noisy and incomplete measurements. Under the Markov assumption on system dynamics, previous work has provided methods of performing inference for these models. In particular, prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics. We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation. Our new method outperforms previous approaches on a challenging problem in epidemiology. This is implemented in a precursor to a Python counterpart to the popular \texttt{pomp} package by \citet{king2016pomp, king2017pompmanual} that includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation.

\kevin{TODO: (1) Fix beta smoothness to d(n) at no more than O(n), (2) add story to say that no one spotted that it could have been plug-and-play because of e.g. reasons why they would stumble}

\section{Problem Setup}

Consider an unobserved Markov process $\{X_t, t \geq t_0\}$, and observations $Y_1,...,Y_N$ at timesteps $t_1,..., t_N$. The process is parameterized by an unknown parameter $\theta \in \Theta$, where $X \in \gX, Y \in \gY,$ and finally $\gX, \gY, \Theta \subseteq \R^n$. By a similar decomposition to that in \citet{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
$$f_{X_{0: N}, Y_{1: N}}\left(x_{0: N}, y_{1: N} ; \theta\right)=f_{X_0}\left(x_0 ; \theta\right) \prod_{n=1}^N f_{X_n \mid X_{n-1}}\left(x_n \mid x_{n-1} ; \theta\right) f_{Y_n \mid X_n}\left(y_n \mid x_n ; \theta\right).$$

We call $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ the process model, writing $\operatorname{process}\left(x_n, \theta\right)$ for the simulator corresponding to the above process model, $f_{Y_n|X_n}\left(y_n \mid x_n, \theta\right)$ the measurement model, and will write $y_n^*$ for the actual values of the observations that were observed.

The above are all defined on a single probability space $(\Omega, \Sigma, \prob),$ where $\omega \in \Omega$ denotes an element of the sample space that can be thought of as a random seed. For notational simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0.$

\section{A Smooth Extension of the Particle Filter}

\textbf{Intuition:} Our theoretical framework for automatic differentiation for the particle filter hinges on the following observation. Instead of seeking to differentiate $\ell(\theta)$ directly, we instead focus on differentiating through a suitable fixed-seed reweighting of a bootstrap filter. The intuition is as follows. Consider a particle filter likelihood estimate where the system (i.e. \textbf{both the process and measurement models}) evolves under a parameter $\phi \in \Theta$ and the seed, or sample space element, is fixed at $\omega \in \Omega.$ Conditional on $\omega$ and $\phi,$ one can reweight the likelihood estimate to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$

Differentiating this reweighted likelihood estimate with respect to $\theta$ then amounts to differentiating through the reweighting scheme, which is differentiable if the densities used for the reweighting are differentiable. We call this simulation-and-reweighting scheme an \textbf{off-policy} likelihood estimate, as it is similar in spirit to the off-policy methods found in the reinforcement learning literature.

The above intuition is formalized below.

\subsection{Doubly Off-Policy Particle Filters}

\begin{defn}[Seed-Fixing]
\label{defn:seed-fixing}
Consider a probability space $(\Omega, \Sigma, \mu)$ governing the particles and observations. Noting that the particle filter is unbiased for the likelihood when computing the expectation of the normalizing factor under the posterior $f_{X_{0:N|Y_{1:N}}}\left(x_{0: N} \mid y_{1: N}; \theta\right)$, we can write the likelihood as
$$
\begin{aligned}
\mathcal{L}(\theta) & =\int_{\Omega} \mathcal{L}(\theta, \omega, J) d \mu(\omega) \\
& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right)\right) f_{X_{0:N}, Y_{1:N}}\left(x_{0: N}^P(\omega) \mid y_{1: N}^*, \theta\right) d \mu(\omega) \\
& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right) f_{X_{0:N}, Y_{1:N}}\left(x_{n, j}^P(\omega) \mid y_n^*, \theta\right)\right) d \mu(\omega)
\end{aligned}
$$
where $\mathcal{L}(\theta, \omega, J)$ is the Monte-Carlo estimate of the likelihood given the realization of $\mathrm{J}$ particles corresponding to $\omega \in \Omega$. For simplicity, we resample every timestep. We call $\omega$ a seed, and the deterministic quantity $\mathcal{L}(\theta, \omega, J)$ the seed-fixed estimate.
\end{defn}

Armed with this seed-fixed likelihood estimate, we modify a particle filter likelihood estimate where the system evolves under a parameter $\phi \in \Theta$ and seed $\omega \in \Omega$ to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$

\begin{defn}[Doubly Off-Policy Particle Filter Likelihood]
\label{defn:doubly-off-policy}
Let $\lik(\theta, \phi, \omega, J)$ denote a Monte-Carlo estimate of the likelihood evaluated at $\theta \in \Theta$ with $J$ particles, but where the system evolves according to another $\phi \in \Theta$ and conditional on a fixed seed $\omega \in \Omega$. This can be computed with a suitable reweighting of the particles, which we provide the derivation for later. We then have
\begin{equation} \label{eq:loglik2}
\lik(\theta)=\E[\lik(\theta,\phi,\Omega, J)]=\int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega),
\end{equation}
where we define
\begin{align}
    \lik(\theta,\phi,\omega, J) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J &\left(f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) f_{X_n|Y_n}(x_{n,j}^{P, \phi}(\omega) | y_n^*, \phi) \\
    &\cdot
    \frac{f_{Y_n|X_n}(y^*_n|x_{n,j}^{P,\phi}(\omega),\theta)}{f_{Y_n|X_n}(y^*_n|x_{n,j}^{P,\phi}(\omega),\phi)}
    \cdot \frac{f_{X_n|X_{n-1}}(x_{n,j}^{P,\phi}(\omega)|x_{n-1,j}^{F,\phi}(\omega),\theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P,\phi}(\omega)|x_{n-1,j}^{F,\phi}(\omega),\phi)}\right).
\end{align}
\end{defn}

We want $\phi\in\Theta$ to be a point in parameter space around which $\lik(\theta,\phi,\omega, J)$ is differentiable with respect to $\theta$ and does not have excessive Monte~Carlo variance. This leads us to our first assumption.

\begin{aspt}[Smooth Neighborhood]
\label{assump:smooth-nbhd}
There exists a neighborhood $\mathcal{N}(\phi)$ around any $\phi \in \Theta$ where for all $\theta \in \mathcal{N}(\phi)$ and almost every $\omega \in \Omega$, the Monte Carlo estimate of the likelihood at $\theta \in \mathcal{N}(\phi)$ with the system evolving according to $\phi$ conditional on $\omega$, $\lik(\theta,\phi,\omega, J)$, is twice differentiable in $\theta$. 
\end{aspt}

\paragraph{Intuition:} We can justify this assumption as follows. For suitably nearby $\theta$ and $\phi$, the resampling indices for $\lik(\theta,\phi,\omega, J)$ and $\lik(\phi,\phi,\omega, J)$ are the same, eliminating any discontinuities from resampling, and that for small enough $\mathcal{N}(\phi)$ the likelihood ratios are bounded. The likelihood only changes by a factor of the likelihood ratios, which are bounded and smooth in $\theta$ if the densities used in the calculation are.

Lastly, we require the following regularity conditions.

\begin{aspt}[Continuity of the Likelihood] $\ell(\theta)$ proper is continuous in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.
\end{aspt}

\begin{aspt}[Bounded Measurement Model] There is an $\epsilon>0$ with $\epsilon^{-1}>f_{Y_n \mid X_n}\left(y_n{ }^* \mid x_n; \theta\right)>\epsilon$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
\end{aspt}

\begin{aspt}[Locally Bounded Derivative]
Let $\mathcal{M}$ be an open subset of $\Omega$. There exists some function $G(\theta)$ and a constant $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$ such that
    $$
    \|\nabla \ell(\theta, \phi, \omega, J)\|_2<G(\theta) \leq G<\infty
    $$
    for every $\phi \in \mathcal{M}, \theta$ in smooth neighborhood $\mathcal{N}(\phi)$, and almost every $\omega \in \Omega$.
\end{aspt}

We claim that this yields a properly weighted particle filter that recovers the derivative estimate of \citet{doucet2011sf, scibior2021dpf} when $\theta$ is evaluated at $\phi.$ \kevin{INSERT RESULTS HERE}

\subsection{Measurement Off-Policy}

However, the above algorithm requires access to likelihood ratios, and is therefore not a simulation-based inference method.

\bibliography{paper/ref}
\bibliographystyle{apalike}

\appendix
\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}

\section{Doubly Off-Policy Particle Filters}

Evaluating the off-policy likelihood is possible with a weighted particle filter by using resampling weight $p(y^*_n|x_n;\phi)$ and making two corrections in the particle weight, through multiplying $w_{t,j}$ by the two likelihood ratios below,

\begin{equation}
    \label{eq:dmeas-ratio}
    s=p(y^*_n|x_n;\theta)/p(y^*_n|x_n;\phi)
\end{equation}
and
\begin{equation}
    \label{eq:rproc-ratio}
    r=p(x_n|x_{n-1};\theta)/p(x_n|x_{n-1};\phi).
\end{equation}

The following result showcases the correctness of corrections \ref{eq:dmeas-ratio} and \ref{eq:rproc-ratio}.

\begin{prop}
    For $\theta$, $\phi$ that satisfy Assumption \ref{assump:smooth-nbhd}, the following correction in the particle filter weights,
    \begin{equation}
        w_{t,j} := \tilde{w}_{t,j} \cdot p(y^*|x_{t,j}^P,\phi)\cdot r \cdot s
    \end{equation}
    successfully recovers $\lik(\theta)$.
\end{prop}

\begin{proof}
    Recall $\lik(\theta) = \E[\lik(\theta,\phi,\Omega, J)]=\int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega)$ and $\lik(\theta) = \int_\Omega \lik(\theta, \omega, J) d\mu(\omega)$. We want to evaluate
    \begin{align*}
        \lik(\theta,\omega, J)
        &= \E\left[ \prod_{t=1}^T \lik_t(\theta) \middle| \omega\right]
        = \E\left[ \prod_{t=1}^T p(y_t^*|y_{1:t-1}^*,\theta) \middle| \omega\right] \\
        &= \E\left[ \prod_{t=1}^T \frac{1}{J} \sum_{j=1}^J  p(y_t^*|x_{t,j}^P,\theta) p(x_{t,j}^P|y_{1:t-1}^*,\theta) \middle| \omega\right] \\
        &= \E\left[ \prod_{t=1}^T\frac{1}{J}  \sum_{j=1}^J  p(y_t^*|x_{t,j}^P,\theta) p(x_{t,j}^P|x_{t-1}^F,\theta) \middle| \omega\right] 
    \end{align*}
    where $p(x_t^P|y_{1:t-1}^*,\theta)$ is the prediction distribution of particles, and the filtering distribution of particles is proportional to $p(y_t^*|x_{t,j}^P,\theta)$, the likelihood of particle $j$ at time $t$.

    Consider first the case with no resampling. Let $\lik^j(\theta)$, $\lik^j(\phi)$ be the likelihood of trajectory $j$ under $\theta$ and $\phi$ respectively. As the prediction and filtering particles are then the same, we use $x_{t,j}$ to refer to both. As such, conditional on $\omega$,
    \begin{align*}
        \lik^j(\theta) 
        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \theta) p(x_{t,j}|x_{t-1,j}, \theta) \\
        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \frac{p(y_t^*|x_{t,j}, \theta)}{p(y_t^*|x_{t,j}, \phi)} p(x_{t,j}|x_{t-1,j}, \theta) \frac{p(x_{t,j}|x_{t-1,j}, \theta) }{p(x_{t,j}|x_{t-1,j}, \phi) } \\
        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \cdot s \cdot p(x_{t,j}|x_{t-1,j}, \phi)\cdot r
    \end{align*}

    and therefore, if trajectories were proposed with equal probability with particles conditional on $\omega$,
    $$ \lik(\theta, \omega, J) = \frac{1}{J} \sum_{j=1}^J \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \cdot s \cdot p(x_{t,j}|x_{t-1,j}, \phi)\cdot r $$
    The correction is therefore sufficient for recovering $\lik(\theta)$, with trajectories collected with the process model at $\phi$ and measurements evaluated with the measurement model at $\phi$.

    Consider now the case with resampling. Without loss of generality we can resample every timestep. Here, $x_{t,j}^F(\phi) \sim p(y^*|x_{t,j}^P, \phi)$ and $x_{t,j}^P(\phi) \sim p(x_t|x_{t-1}^F,\phi)$, and similarly for $\theta$. We use this notation to emphasize that \textbf{both} the measurement model and the previously drawn particles depend on $\phi$ or $\theta$.

    \begin{align*}
        \lik_t(\theta) &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \theta)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\theta) \right] \\
        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \theta)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \frac{p(y_t^*|x_{t,j}^P,\theta)}{p(y_t^*|x_{t,j}^P,\phi)} \right] \\
        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \frac{p(y_t^*|x_{t,j}^P,\theta)}{p(y_t^*|x_{t,j}^P,\phi)} \frac{p(x_{t,j}|x_{t-1,j}, \theta) }{p(x_{t,j}|x_{t-1,j}, \phi) }\right] \\
        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \cdot s \cdot r \right] \\
        &= \E_{\omega}\E_{x_{t,j}^P \sim \process(x_{t-1}^F, \omega, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \cdot s \cdot r \middle| \omega \right]
    \end{align*}

    and again the correction is sufficient.

    Therefore,

    \begin{align}
         \mathcal{L}(\theta) 
         &= \int_\Omega \left(
         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^P(\omega), \theta) p(x_{n,j}^P(\omega) | x_{n-1,j}^P(\omega), \theta) \right) \, d\mu(\omega) \\
         &= \int_\Omega \left(
         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\theta}(\omega), \theta) p(x_{n,j}^{P,\theta}(\omega) | x_{n,j}^P(\omega), \theta) \right) \, d\mu(\omega) \\
         &= \int_\Omega \left(
         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\theta}(\omega), \phi) p(x_{n,j}^{P,\theta}(\omega) | y_n^*, \phi) \cdot s  \right) \, d\mu(\omega) \\
         &= \int_\Omega \left(
         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) p(x_{n,j}^{P,\phi}(\omega) | y_n^*, \phi) \cdot r \cdot s \right) \, d\mu(\omega) \\
         &= \int_\Omega \left(
         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) p(x_{n,j}^{P, \phi}(\omega) | y_n^*, \phi) \cdot r \cdot s \right) \, d\mu(\omega) \\
         &= \int_\Omega \left(
         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) \cdot r \cdot s \right)  p(x_{1:N} | y_{1:N}^*, \phi) \,  d\mu(\omega) \\
         &=: \int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega).
     \end{align}
\end{proof}

\subsection{Off-Policy Weighted Differentiable Particle Filter}

We are finally ready to introduce the off-policy weighted differentiable particle filter. In the event where a seed is fixed, all expressions in Algorithm \ref{alg:dop} can be considered to be conditional on $\omega \in \Omega$.

\begin{algorithm}[H]
\centering
	\caption{Doubly Off-Policy-$\alpha$}
    \label{alg:dop}
	\begin{algorithmic}[1]
	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
		\STATE Initialize filter particles ${X}_{0,j}^{F,\phi}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\phi}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$. Fix $\omega.$
		\FOR{$n=1,...,N$}
            \STATE Prediction weights with discounting: $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
            \label{dop-alpha:discount}
            \STATE Simulate for prediction:
            ${X}_{n,j}^{P,\phi}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\phi}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{dop-alpha:step1}
            \STATE Adjust weights: $\displaystyle w_{n,j}^{P,\theta} = w_{n,j}^{P,\theta} \times
            \frac{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\phi}|{X}_{n-1,j}^{F};{\theta}\big)}{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\phi}|{X}_{n-1,j}^{F};{\phi}\big)}$ for $j$ in $1:J$
            \label{dop-alpha:dproc}
            \STATE Evaluate measurement density:
            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\phi}\giventh{\theta})$ for $j$ in $\seq{1}{J}$
            \STATE Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
            \STATE Conditional likelihood under $\phi$: 
            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
            \label{dop-alpha:Lphi}
            \STATE Normalize weights:
            $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
            for $j\ \mathrm{in}\ \seq{1}{J}$
            \STATE Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \label{dop-alpha:systematic}
            \STATE Resample particles: ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$
            \STATE Filter weights corrected for resampling:
            $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{dop-alpha:weight:update}
            \STATE Resample filter weights:
            $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
            for $j$ in $\seq{1}{J}$ \label{dop-alpha:step2}
            \STATE After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
            \ENDFOR
		\RETURN likelihood estimate $\hat{\lik}(\theta) = \lik(\theta, \phi, \omega, J) := \prod_{n=1}^N L_n^{A,\theta,\alpha}$, filtering distribution $\{(x_{N,j}^P, w^{F,\theta}_{N,j})\}.$
	\end{algorithmic}
\end{algorithm}

Here, we consider a doubly off policy particle filter, meaning that both \code{rprocess} and \code{rmeasure} are computed at $\phi$ and particles are reweighted to correspond to $\theta$.

\end{document}

### 2023-10-31 mop/mop.tex
Updates from Overleaf

@@
   However, both are identical for a particle filter (i.e., MOP when $\theta=\phi$).
-  Presumably, this could allow their fixed-seeed derivatives to be the same at $\theta=\phi$, though it is not obvious if this is true.
+  Presumably, this could allow their fixed-seed derivatives to be the same at $\theta=\phi$, though it is not obvious if this is true.
   In that case, we would have an alternative explanation of the telescoping identity of Scibior \& Wood.

### 2023-10-31 mop/draft.tex
Updates from Overleaf

@@
 
-\input{paper/macros}
+\input{../paper/macros}
 
@@
 
+\begin{prop}[Correctness of DOP-$\alpha$ (Algorithm \ref{alg:dop})]
+    When $\theta$ is evaluated at $\phi$, 
+\end{prop}
+
 \subsection{Measurement Off-Policy}
 
-However, the above algorithm requires access to likelihood ratios, and is therefore not a simulation-based inference method.
+However, the above algorithm requires access to likelihood ratios, and is therefore not compatible with simulation-based inference.
 

### 2023-11-02 paper/main.tex
Updates from Overleaf
tidying paper

@@
 \documentclass[9pt,twocolumn,twoside]{pnas-new}
+%\documentclass[9pt,twocolumn,pnasresearcharticle]{pnas-new}
+%\documentclass[9pt,twocolumn,twoside]{pnas-new}
 % Use the lineno option to display guide line numbers if required.
@@
 \author[a]{Kevin Tan}
+\author[a]{Giles Hooker}
 \author[b,1]{Edward L. Ionides}
@@
 We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
-Our new method dramatically outperforms previous approaches on a challenging problem in epidemiology.}
+Our new method outperforms previous approaches on a challenging problem in epidemiology.}
 
@@
 
-\firstpage{10}
+\firstpage{7}
 % Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting.
@@
 al.
-\cite{Ionides_infpomp} converge quickly to a neighborhood of the MLE, but struggle to optimize the last few units of log-likelihood.
+\cite{ionides15} converge quickly to a neighborhood of the MLE, but struggle to optimize the last few units of log-likelihood.
 
@@
 al.
-\cite{Ionides_infpomp} rapidly converge to a neighborhood of the optimum but struggle at finding the exact optimum due to Monte Carlo variance, even with an annealing random walk standard deviation.
+\cite{ionides15} rapidly converge to a neighborhood of the optimum but struggle at finding the exact optimum due to Monte Carlo variance, even with an annealing random walk standard deviation.
 
@@
 
-Basic particle filtering algorithms do not require evaluation of the transition density of the latent Markov process, in a feature known as the \textbf{plug-and-play property} \cite{Breto_timeseriesmech} since it enables an arbitrary model simulator to be plugged into the algorithm.
+Basic particle filtering algorithms do not require evaluation of the transition density of the latent Markov process, in a feature known as the \textbf{plug-and-play property} \cite{breto09} since it enables an arbitrary model simulator to be plugged into the algorithm.
 We investigate ways to use algorithmic differentiation of particle filters within the confines of the plug-and-play property, with the goal of enhancing current inference capabilities for general POMP models.
@@
 al.
-\cite{rosato22b} do by using previous work such as \cite{scibior2021dpf, doucet2011sf} (and we conjecture that the seed-fixing derivatives of Rosato et.
+\cite{rosato22b} do by using previous work such as \cite{scibior21, poyiadjis11} (and we conjecture that the seed-fixing derivatives of Rosato et.
 al.
@@
 
-References should be cited in numerical order as they appear in text; this will be done automatically via bibtex, e.g.
-\cite{belkin2002using} and \cite{berard1994embedding,coifman2005geometric,phdthesis,masterthesis}.
-All references cited in the main text should be included in the main manuscript file.
+References should be cited in numerical order as they appear in text; this will be done automatically via bibtex,
 
@@
 
-\bibsplit[2]
+%\bibsplit[2]
 %Use \bibsplit to split the references from the body of the text. Value "[2]" represents the number of reference in the left column (Note: Please avoid single column figures & tables on this page.)
 
+%\bibliographystyle{pnas-new}
 % Bibliography
-\bibliography{pnas-sample}
+\bibliography{bib-ifad}
 

### 2023-11-02 paper/ms.tex
a little bit more on paper

--- 2023-11-02 paper/main.tex
@@
 
-\section*{Guide to using this template on Overleaf}
+%% \subsection*{Author Affiliations}
 
-Please note that whilst this template provides a preview of the typeset manuscript for submission, to help in this preparation, it will not necessarily be the final publication layout.
-For more detailed information please see the \href{https://www.pnas.org/page/authors/format}{PNAS Information for Authors}.
-
-If you have a question while using this template on Overleaf, please use the help menu (``?'') on the top bar to search for \href{https://www.overleaf.com/help}{help and tutorials}.
-You can also \href{https://www.overleaf.com/contact}{contact the Overleaf support team} at any time with specific questions about your manuscript or feedback on the template.
-
-\subsection*{Author Affiliations}
-
-Include department, institution, and complete address, with the ZIP/postal code, for each author.
+%% Include department, institution, and complete address, with the ZIP/postal code, for each author.
 Use lower case letters to match authors with institutions, as shown in the example.
@@
 For proper authentication, authors must provide their ORCID at submission and are not permitted to add ORCIDs on proofs.
-
-\subsection*{Submitting Manuscripts}
-
-All authors must submit their articles at \href{http://www.pnascentral.org/cgi-bin/main.plex}{PNAScentral}.
-If you are using Overleaf to write your article, you can use the ``Submit to PNAS'' option in the top bar of the editor window.
 
@@
 
-\subsection*{Manuscript Length}
+%% \subsection*{Manuscript Length}
 
-A standard 6-page article is approximately 4,000 words, 50 references, and 4 medium-size graphical elements (i.e., figures and tables).
+%% A standard 6-page article is approximately 4,000 words, 50 references, and 4 medium-size graphical elements (i.e., figures and tables).
 The preferred length of articles remains at 6 pages, but PNAS will allow articles up to a maximum of 12 pages.
 
-\subsection*{References}
+%% \subsection*{References}
 
-References should be cited in numerical order as they appear in text; this will be done automatically via bibtex,
+%% References should be cited in numerical order as they appear in text; this will be done automatically via bibtex,
 
@@
 
-\subsection*{Language-Editing Services}
-Prior to submission, authors who believe their manuscripts would benefit from professional editing are encouraged to use a language-editing service (see list at https://www.pnas.org/author-center/language-editing).
-PNAS does not take responsibility for or endorse these services, and their use has no bearing on acceptance of a manuscript for publication.
+\subsection*{Historical cholera transmission in Dacca, Bangladesh}.
+We tested our method on the cholera transmission model and data of \citep{king08}.
+This is a POMP model based on the fundamental Susceptible-Exposed-Infected-Recovered (SEIR) epidemiological model, but incorporating additional features that are required for practical application of this model to various real disease systems.
+Seasonality is included as a periodic cubic B-spline covariate process.
+Loss of immunity is modeled by return from the Recovered class to the Susceptible class.
+Asymptomatic infections can be modeled separately to allow for the possibility that they differ from symptomatic infections in their infectivity and acquired immunity.
+This model, and the Dacca case reports to which it was originally fitted, provide a benchmark for a challening scientific model fitting exercise.
+It was used oringally to test IF1 \citep{ionides06-pnas} and IF2  \citep{ionides15}, and a simplified version has been used to compare with alternative methodologies \citep{fasiolo16}.
+We will therefore benchmark on this well-established model before moving on to target the methods on models with anticipated scientific goals.
 
@@
 \centering
-\includegraphics[width=.8\linewidth]{frog.pdf}
-\caption{Placeholder image of a frog with a long example legend to show justification setting.}
-\label{fig:frog}
+\includegraphics[width=.8\linewidth]{23-09-05-adpf-vs-if2.png}
+\caption{Comparing the ADPF algorithm with IF2 on the Dacca model and data of \citep{king08}}
+  \label{fig:dacca-fit}
 \end{figure}
@@
 
-\subsection*{Digital Figures}
+%% \subsection*{Digital Figures}
 
-EPS, high-resolution PDF, and PowerPoint are preferred formats for figures that will be used in the main manuscript.
+%% EPS, high-resolution PDF, and PowerPoint are preferred formats for figures that will be used in the main manuscript.
 Authors may submit PRC or U3D files for 3D images; these must be accompanied by 2D representations in TIFF, EPS, or high-resolution PDF format.
@@
 
-Images must be provided at final size, preferably 1 column width (8.7cm).
+%% Images must be provided at final size, preferably 1 column width (8.7cm).
 Figures wider than 1 column should be sized to 11.4cm or 17.8cm wide.
@@
 
-Figures and tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.
+%% Figures and tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.
 
-Figure \ref{fig:frog} shows an example of how to insert a column-wide figure.
+%% Figure \ref{fig:dacca-fit} shows an example of how to insert a column-wide figure.
 To insert a figure wider than one column, please use the \verb|\begin{figure*}...\end{figure*}| environment.
@@
 
-\begin{SCfigure*}[\sidecaptionrelwidth][t!]
-\centering
-\includegraphics[width=11.4cm,height=11.4cm]{frog.pdf}
-\caption{This legend would be placed at the side of the figure, rather than below it.}\label{fig:side}
-\end{SCfigure*}
+%\begin{SCfigure*}[\sidecaptionrelwidth][t!]
+%\centering
+%\includegraphics[width=11.4cm,height=11.4cm]{frog.pdf}
+%\caption{This legend would be placed at the side of the figure, rather than below it.}\label{fig:side}
+%\end{SCfigure*}
 
@@
 
-\subsubsection*{SI Movies}
-
-Supply Audio Video Interleave (avi), Quicktime (mov), Windows Media (wmv), animated GIF (gif), or MPEG files.
-Movie legends should be included in the SI Appendix file.
-All movies should be submitted at the desired reproduction size and length.
-Movies should be no more than 10MB in size.
-
 \matmethods{Please describe your materials and methods here. This can be more than one paragraph, and may contain subsections and equations as required.

### 2023-11-03 paper/ms.tex
more on paper/ms

@@
 % Please include corresponding author, author contribution and author declaration information
-\authorcontributions{Please provide details of author contributions here.}
-\authordeclaration{Please declare any competing interests here.}
-%\equalauthors{\textsuperscript{1}A.O.(Author One) contributed equally to this work with A.T. (Author Two) (remove if not applicable).}
+\authorcontributions{K.T. and E.L.I. planned the study; K.T. developed the numerical results; all authors wrote the manuscript.}
+\authordeclaration{The authors declare no competing interests.}
 \correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: ionides@umich.edu}
@@
   This progress has required simultaneous advances in algorithms, software and hardware.  
-  Partially observed nonlinear stochastic dynamic systems have proved resistant to these techniques; despite various attempts, widely applicable methods have not yet emerged.
+  Partially observed nonlinear stochastic dynamic systems have proved resistant to the benefits of automatic differentiation techniques; despite various attempts, widely applicable methods have not yet emerged.
   We present a new approach which is applicable to a general class of models, possesses a theoretical foundation, and is demonstrated to beat current state-of-the-art methods on a challenging scientific benchmark problem.
@@
 
-\firstpage{7}
+\firstpage{3}
 % Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting.
@@
 If this is the case, add \parshape=0 to the end of the list environment.
-\dropcap{M}any approaches to inference in highly nonlinear stochastic dynamical systems assume access to the probability density of next states given the current state.
+\dropcap{T}he particle filter is a widely used Monte Carlo algorithm which provides an unbiased estimate of the likelihood function for partially observed Markov process (POMP) models \ed{REFS}.
+Maximization of the likelihood permits computation of the maximum likelihood parameter estimate, as well as profile likelihood confidence intervals, and model selection by likelihood ratio tests or Akaike's information criterion.
+The particle filter involves a random resampling step in which particles inconsistent with the data are likely to be pruned out and the most consistent particles are replicated.
+Applying automatic differentiation (AD) to supply the optimization with gradient information is complicated by the disctete nature of this resampling step.
+Simulations of discrete random variables are necessarily discontinuous, specifically piecewise constant, as a function of the unknown parameters for a fixed value of the seed for the random number generator.
+As a result of this, the Monte Carlo expectation of the derivative is not equal to the derivative of the Monte Carlo expectation.
+Various strategies have been proposed to address this phenomenon \ed{REFS}, but none of these algorithms have found widespread practical applicability.
 
-This is a problem in some critical applications, such as disease modeling, where the models are complex enough that obtaining the density is intractable.
-However, the particle filter, a popular method for solving the filtering problem in partially-observed dynamical systems, does not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm in a feature known as the plug-and-play property.
+A significant practical strength of the particle filter is its applicability to arbitrarily nonlinear stochastic systems, requiring only a simulator for the dynamic model.
+This is called the plug-and-play property \ed{refs}, and this property is critical in some applications, such as disease modeling, where the models are complex enough that obtaining the density is intractable.
+Some extensions of the particle filter to parameter estimation require availability of transition probabilities for the dynamic model \ed{refs} which violates the plug-and-play property.
+We have developed an AD algorithm for the particle filter which preserve the plug-and-play property while providing useful derivative estimates which demonstrably improve our ability to search for the maximum of the likelihood.
+
+Our key insights are as follows:
+\begin{enumerate}
+\item We consider a novel one-parameter family of plug-and-play AD gradient estimators allowing flexibility in the tradeoff between Monte Carlo bias and variance.
+For our target problem, the results are insensitive to the choice of this parameter.
+
+\item ADPF stochastic gradient optimization has complementary strengths to iterated filtering algorithms that are widely used for plug-and-play likelihood maximization for POMP models.
+Iterated filtering algorithms provide a computationally efficient way to approach the maximum, but they are poor at identifying the precise location of the maximum.
+Combining both these algorithms substantially out-performs either alone.
+
+\item The IFAD algorithm can be implemented on a GPU to achieve massive parallelization. 
+\end{enumerate}
+Together, these innovations lead to methodology which advances capability for statistical analysis of scientific models arising in biological sciences and elsewhere \ed{refs}.
+Our benchmark task is a model previously used to evaluate iterated filtering algorithms \ed{refs}.
+
+\subsection*{Historical cholera transmission in Dacca, Bangladesh}.
+We tested our method on the cholera transmission model and data of \citep{king08}.
+This is a POMP model based on the fundamental Susceptible-Exposed-Infected-Recovered (SEIR) epidemiological model, but incorporating additional features that are required for practical application of this model to various real disease systems.
+Seasonality is included as a periodic cubic B-spline covariate process.
+Loss of immunity is modeled by return from the Recovered class to the Susceptible class.
+Asymptomatic infections can be modeled separately to allow for the possibility that they differ from symptomatic infections in their infectivity and acquired immunity.
+This model, and the Dacca case reports to which it was originally fitted, provide a benchmark for a challening scientific model fitting exercise.
+It was used oringally to test IF1 \citep{ionides06-pnas} and IF2  \citep{ionides15}, and a simplified version has been used to compare with alternative methodologies \citep{fasiolo16}.
+We will therefore benchmark on this well-established model before moving on to target the methods on models with anticipated scientific goals.
+
+\begin{figure}%[tbhp]
+\centering
+\includegraphics[width=.8\linewidth]{23-09-05-adpf-vs-if2.png}
+\caption{Comparing the ADPF algorithm with IF2 on the Dacca model and data of \citep{king08}}
+  \label{fig:dacca-fit}
+\end{figure}
+
+\subsection*{Assorted draft stuff}
+
+[naive ADPF] intruduces bias into the Monte Carlo estimate of the derivative.
+This bias has previously been reported to be non-negligible [REF].
+An approach to avoiding this bias is the REINFORCE estimator, also known as the stop-gradient trick \cite{scibior21}.
+The Monte Carlo variance of this estimator has been found to be problematic \cite{scibior21}
+
+1.
+A new derivation of the stop-gradient estimator.
+We show that this estimator can be obtained as the fixed seed derivative of an extension of the basic particle filter which we call the off policy particle filter.
+
+2. Combining the reparameterization trick with the stop-gradient estimator to obtain a plug-and-play gradient estimator.
+
+3. Apply this gradient estimator within a stochastic gradient descent algorithm using a normalized gradient.
+
+4. Multiple searches from diverse starting points to reduce, quantify and control maximization error due to multimodality and Monte Carlo variability.
+
+%\dropcap{M}
+Many approaches to inference in highly nonlinear stochastic dynamical systems assume access to the probability density of next states given the current state.
 
@@
 
-We propose a hybrid algorithm that u
+%We propose a hybrid algorithm that u
 
-Unlike IF2, we explicitly characterize our method's rate of convergence
+%Unlike IF2, we explicitly characterize our method's rate of convergence
 
@@
 
-\subsection*{Historical cholera transmission in Dacca, Bangladesh}.
-We tested our method on the cholera transmission model and data of \citep{king08}.
-This is a POMP model based on the fundamental Susceptible-Exposed-Infected-Recovered (SEIR) epidemiological model, but incorporating additional features that are required for practical application of this model to various real disease systems.
-Seasonality is included as a periodic cubic B-spline covariate process.
-Loss of immunity is modeled by return from the Recovered class to the Susceptible class.
-Asymptomatic infections can be modeled separately to allow for the possibility that they differ from symptomatic infections in their infectivity and acquired immunity.
-This model, and the Dacca case reports to which it was originally fitted, provide a benchmark for a challening scientific model fitting exercise.
-It was used oringally to test IF1 \citep{ionides06-pnas} and IF2  \citep{ionides15}, and a simplified version has been used to compare with alternative methodologies \citep{fasiolo16}.
-We will therefore benchmark on this well-established model before moving on to target the methods on models with anticipated scientific goals.
-
-\begin{figure}%[tbhp]
-\centering
-\includegraphics[width=.8\linewidth]{23-09-05-adpf-vs-if2.png}
-\caption{Comparing the ADPF algorithm with IF2 on the Dacca model and data of \citep{king08}}
-  \label{fig:dacca-fit}
-\end{figure}
-
 \begin{table}[t!]
@@
 
-\subsection*{Tables}
-Tables should be included in the main manuscript file and should not be uploaded separately.
+%% \subsection*{Tables}
+%% Tables should be included in the main manuscript file and should not be uploaded separately.
 
-\subsection*{Single column equations}
+%% \subsection*{Single column equations}
 
-Authors may use 1- or 2-column equations in their article, according to their preference.
+%% Authors may use 1- or 2-column equations in their article, according to their preference.
 
-To allow an equation to span both columns, use the \verb|\begin{figure*}...\end{figure*}| environment mentioned above for figures.
+%% To allow an equation to span both columns, use the \verb|\begin{figure*}...\end{figure*}| environment mentioned above for figures.
 
-Note that the use of the \verb|widetext| environment for equations is not recommended, and should not be used.
+%% Note that the use of the \verb|widetext| environment for equations is not recommended, and should not be used.
 
-\begin{figure*}[bt!]
-\begin{align*}
-(x+y)^3&=(x+y)(x+y)^2\\
-       &=(x+y)(x^2+2xy+y^2) \numberthis \label{eqn:example} \\
-       &=x^3+3x^2y+3xy^3+x^3.
-\end{align*}
-\end{figure*}
+%% \begin{figure*}[bt!]
+%% \begin{align*}
+%% (x+y)^3&=(x+y)(x+y)^2\\
+%%        &=(x+y)(x^2+2xy+y^2) \numberthis \label{eqn:example} \\
+%%        &=x^3+3x^2y+3xy^3+x^3.
+%% \end{align*}
+%% \end{figure*}
 

### 2023-11-03 mop/draft.tex
Updates from Overleaf
some latex error fixes unnoticed by overleaf

@@
 \title{Automatic Differentiation to Accelerate Inference for Partially Observed Stochastic Processes}
-\author{Kevin Tan, Edward L. Ionides}
+\author{Kevin Tan, Giles J. Hooker, Edward L. Ionides}
 \date{}
@@
 While the Monte-Carlo Adjusted Profile (MCAP) from \citet{Ionides_mcap} provides a profile likelihood confidence interval that takes into account both Monte Carlo profile error and statistical uncertainty in the likelihood, it is computationally inefficient.
-\kevin{TODO} We provide a method for incorporating the MCAP methodology in our IFAD algorithm that provides a more computationally efficient profile likelihood confidence interval.
+\kevin{TODO} We provide a method for incorporating the MCAP methodology (with the poor man's likelihood profile) in our IFAD algorithm that provides a more computationally efficient profile likelihood confidence interval.
+\kevin{Quantile regression for estimating the max/the envelope of the likelihood evaluations from AD, then fit MCAP/MQLE/MSLE?}
 
-Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king2016pomp, king2017pompmanual}.
+Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king2016pomp, king2017pompmanual} that we call \texttt{pompy}.
 The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{jax} Python package from \citet{jax2018github}.
@@
 Consider a particle filter likelihood estimate where the system (i.e.
-\textbf{both the process and measurement models}) evolves under a parameter $\phi \in \Theta$ and the seed, or sample space element, is fixed at $\omega \in \Omega.$ Conditional on $\omega$ and $\phi,$ one can reweight the likelihood estimate to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$
+\textbf{both the state transitions and resampling}) evolves under a parameter $\phi \in \Theta$ and the seed, or sample space element, is fixed at $\omega \in \Omega.$ Conditional on $\omega$ and $\phi,$ one can reweight the likelihood estimate to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$
 
@@
 
-We claim that this yields a properly weighted particle filter that recovers the derivative estimate of \citet{doucet2011sf, scibior2021dpf} when $\theta$ is evaluated at $\phi.$ \kevin{INSERT RESULTS HERE}
+We claim that this yields a properly weighted particle filter that recovers the derivative estimate of \citet{doucet2011sf, scibior2021dpf} when $\theta$ is evaluated at $\phi.$
 
 \begin{prop}[Correctness of DOP-$\alpha$ (Algorithm \ref{alg:dop})]
-    When $\theta$ is evaluated at $\phi$, 
+    \label{prop:dop-correctness}
+    Algorithm \ref{alg:dop} is a properly weighted particle filter that is unbiased for the likelihood.
+
+    In particular, when $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter, and its gradient when $\alpha=1$ is the same as the estimator of \citet{doucet2011sf},
+
+    and when $\alpha=0$
+
 \end{prop}
 
-\subsection{Measurement Off-Policy}
+\subsubsection{Practical Considerations}
+
+It is only ever practical to evaluate $\theta$ at $\phi$.
+This removes the issue of the likelihood ratios blowing up, and yields an estimate of the likelihood that agrees with the bootstrap filter estimate.
+Its gradient in this case is either the estimate of \citet{doucet2011sf} when $\alpha=1$, the naively
+\subsection{Measurement Off-Policy Particle Filters}
 
 However, the above algorithm requires access to likelihood ratios, and is therefore not compatible with simulation-based inference.
+
+There is a workaround.
+Consider the case where you have access to a differentiable simulator compatible with the reparametrization trick.
+If we have the state transitions evolve according to $\theta,$ but the resampling according to $\phi,$
+
+We therefore make the following assumption in this section only:
+\begin{aspt}[Differentiable Simulator]
+    $\text{process}(\cdot | X_{t-1}; \theta)$ is differentiable with respect to $\theta,$ and is compatible with the reparametrization trick. 
+\end{aspt}
+
+\section{Numerical Experiments}
+
+\citet{liu2023bridging} have established that the straight-through estimator can be thought of as a first-order approximation of the gradient.
+\kevin{Benchmark DOP vs straight-through estimator/reinmax in MOP}
 

### 2023-11-17 mop/draft.tex
disentangling some inconsistent changes - now macros and and bib files are not shared between paper and mop
Updates from Overleaf

@@
 
-\input{../paper/macros}
+\input{macros}
 
@@
 \begin{align}
-    \lik(\theta,\phi,\omega, J) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J &\left(f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) f_{X_n|Y_n}(x_{n,j}^{P, \phi}(\omega) | y_n^*, \phi) \\
+    \lik(\theta,\phi,\omega, J) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J &\left(f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) f_{X_n|Y_n}(x_{n,j}^{P, \phi}(\omega) | y_n^*, \phi) \right.\\
     &\cdot
+    \left.
     \frac{f_{Y_n|X_n}(y^*_n|x_{n,j}^{P,\phi}(\omega),\theta)}{f_{Y_n|X_n}(y^*_n|x_{n,j}^{P,\phi}(\omega),\phi)}
@@
 
-We claim that this yields a properly weighted particle filter that recovers the derivative estimate of \citet{doucet2011sf, scibior2021dpf} when $\theta$ is evaluated at $\phi.$
+\subsubsection{Doubly Off-Policy Particle Filter}
+
+Having the system evolve according to $\phi$ but evaluating the likelihood at $\theta$ requires us to reweight the conditional likelihood at each timestep by likelihood ratios corresponding to the process and measurement models.
+This corresponds to making two corrections in the particle weights at each timestep, by multiplying by
+\begin{equation}
+    s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}, \qquad r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)},
+\end{equation}
+as we illustrate in Appendix \ref{app:dop}. We use the notation $x_{n,j}^{P, \phi}$ to emphasize that the particles evolve according to $\phi$.
+
+We claim that this yields a properly weighted particle filter that recovers the derivative estimate of \citet{doucet2011sf, scibior2021dpf} when $\theta$ is evaluated at $\phi,$ which is is Algorithm \ref{alg:dop} when $\alpha=1$.
+The proof can be found in Appendix A.
+
+\subsubsection{Discounting Weights}
+
+Observe that at Algorithm \ref{alg:dop}, at each timestep the corrected filtering importance weights $w_{n,j}^{F,\theta}$ optionally propagate to the next timestep for us to maintain the correction.
+This is because we introduce a discounting parameter $\alpha$ that affects the weights DOP, so $w_{n,j}^{P,\theta} := (w_{n-1,j}^{F,\theta})^{\alpha}$.
+Note that no matter what value $\alpha$ takes on, the system evolves according to $\phi$ when we perform the correction to evaluate the likelihood at $\theta$, and the filter remains correctly weighted.
+The value of $\alpha$ simply informs a bias-variance tradeoff, as we elaborate below.
+
+When $\alpha=1,$ the likelihood estimate becomes
+\begin{equation}
+    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\phi}}= \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\phi}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
+\end{equation}
+as we have a telescoping product.
+
+When $\alpha=0,$ the likelihood estimate becomes
+\begin{equation}
+    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\phi}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J r_{n,j}s_{n,j},
+\end{equation}
+and though the telescoping product disappears, this reduces to a series of one-step corrections that leave the filter still properly weighted.
+
+\kevin{Is DOP properly weighted when alpha is not 1 and theta is not phi?}
+
+\paragraph{Bias-Variance Tradeoff} This discounting serves the purpose of balancing a tradeoff between maintaining memory of the entire trajectory, or
+
+\subsubsection{Correctness of DOP-$\alpha$}
 
@@
     \label{prop:dop-correctness}
-    Algorithm \ref{alg:dop} is a properly weighted particle filter that is unbiased for the likelihood.
+    Algorithm \ref{alg:dop} is a properly weighted particle filter that is unbiased for the likelihood, for all $\alpha$.
 
-    In particular, when $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter, and its gradient when $\alpha=1$ is the same as the estimator of \citet{doucet2011sf},
+    In particular, when $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter.
+In this case, its gradient when $\alpha=1$ is the same as the estimator of \citet{doucet2011sf},
+    \begin{equation}
+        \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*),
+    \end{equation}
 
-    and when $\alpha=0$
+    and when $\alpha=0,$ is is the same as differentiating through the fixed-seed likelihood estimate yielded by a vanilla particle filter.
+\end{prop}
 
-\end{prop}
+The proof for the above follows from Appendix A, as well as the technical lemmas for properly weighted particle filters.
 
@@
 This removes the issue of the likelihood ratios blowing up, and yields an estimate of the likelihood that agrees with the bootstrap filter estimate.
-Its gradient in this case is either the estimate of \citet{doucet2011sf} when $\alpha=1$, the naively
+Its gradient in this case is either the estimate of \citet{doucet2011sf} when $\alpha=1$, a single-timestep estimator that corresponds to naively taking the derivative of the particle filter when $\alpha=0$ as first proposed in \citet{blei2018vsmc}
+
 \subsection{Measurement Off-Policy Particle Filters}
 
+DOP has the entire system evolve according to $\phi$, and we differentiate through the likelihood ratios, which are smooth conditional on the resampling being the same (which in turn happens when $\theta \approx \phi$).
 However, the above algorithm requires access to likelihood ratios, and is therefore not compatible with simulation-based inference.
@@
 Consider the case where you have access to a differentiable simulator compatible with the reparametrization trick.
-If we have the state transitions evolve according to $\theta,$ but the resampling according to $\phi,$
+If we have the state transitions evolve according to $\theta,$ but have the resampling according to $\phi,$ we can offload the derivative with respect to the process model to the reparametrization trick.
 
@@
 
+\kevin{Develop theory for mixing-time swapping?
+Detect when IF2 reaches a neighborhood of the MLE, by detecting whether it has reached a local sort of stationary distribution when the perturbations are lower bounded, e.g.
+the autocorrelation between the mean vectors to estimate the mixing time or Gelman-Rubin.
+In practice, just swap when IF2 converges according to the diagnostics.
+Theory doesn't care, you just need to in a basin of attraction that is strongly convex.}
+
 \section{Numerical Experiments}
+
+\kevin{IFAD vs IF2, AD vs IF2, IFAD0 vs IFAD1, AD0 vs AD1 in Dacca, local and global plots, convergence line plot of IFAD vs IF2, Dacca data plots, toy model from IF2 paper, AD0 vs AD1 gradient histograms, MSE of AD0 and AD1 gradient estimates as J increases on analytically tractable toy model from IF2 paper or OU2 in pomp}
+
+\kevin{Main plots are the local and global 2-way comparison plots of IFAD1 vs IF2, AD1 vs IF2.
+Visualize the Q-Q plots in the appendix, and use alternate representations as boxplots (or histograms) with dashed line for the maximum in the main paper.}
 
@@
 
-\bibliography{paper/ref}
+\bibliography{bib-ifad,bib-ref}
 \bibliographystyle{apalike}
@@
 \section{Doubly Off-Policy Particle Filters}
+\label{app:dop}
 

### 2023-11-21 mop/draft.tex
Updates from Overleaf

@@
 This is because we introduce a discounting parameter $\alpha$ that affects the weights DOP, so $w_{n,j}^{P,\theta} := (w_{n-1,j}^{F,\theta})^{\alpha}$.
-Note that no matter what value $\alpha$ takes on, the system evolves according to $\phi$ when we perform the correction to evaluate the likelihood at $\theta$, and the filter remains correctly weighted.
+Note that no matter what value $\alpha$ takes on, the system evolves according to $\phi$ when we perform the correction to evaluate the likelihood at $\theta$.
+The filter fully targets the posterior when $\theta=\phi$, or $\alpha=1$, as we show in Proposition \ref{prop:dop-correctness}.
 The value of $\alpha$ simply informs a bias-variance tradeoff, as we elaborate below.
 
-When $\alpha=1,$ the likelihood estimate becomes
+\paragraph{Bias-Variance Tradeoff} This discounting serves the purpose of balancing a tradeoff between maintaining memory of each particle's ancestral trajectory (most extreme when $\alpha=1$), or considering only the single-step transition dynamics (when $\alpha=0$).
+This can be thought of as an \textbf{exponentially-weighted moving average}, where $\alpha$ controls the amount of discounting.
+We elaborate on two special cases below.
+
+\kevin{This isn't a smoother, because we'd need future information.}
+
+\paragraph{When $\alpha=1,$} the likelihood estimate becomes
 \begin{equation}
@@
 \end{equation}
-as we have a telescoping product.
+as we have a telescoping product. Its derivative is then
+\begin{equation}
+    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},
+\end{equation}
+the average gradient of each surviving particle's cumulative importance weights.
 
-When $\alpha=0,$ the likelihood estimate becomes
+\paragraph{When $\alpha=0,$} the likelihood estimate becomes
 \begin{equation}
@@
 
-\paragraph{Bias-Variance Tradeoff} This discounting serves the purpose of balancing a tradeoff between maintaining memory of the entire trajectory, or
-
 \subsubsection{Correctness of DOP-$\alpha$}
@@
     \label{prop:dop-correctness}
-    Algorithm \ref{alg:dop} is a properly weighted particle filter that is unbiased for the likelihood, for all $\alpha$.
+    Algorithm \ref{alg:dop} is a properly weighted particle filter that is unbiased for the likelihood, when either $\alpha=1$ or $\theta=\phi$.
 
@@
 
+\section{Properly Weighted Particle Filters}
+
+\begin{defn}[Targeting, \cite{liu1998smc}]
+    A random variable $X$ drawn from a distribution $g$ is properly weighted by a weighting function $w(X)$ with respect to the distribution $\pi$ if for any integrable function $h$,
+    $$
+    E_g\{h(X) w(X)\}=E_\pi\{h(X)\}
+    $$
+
+    A set of particles $(X_j, w_j), j=1,2, \ldots$, targets $\pi$ if
+    $$
+    \frac{\sum_{j=1}^J h(X_j) w_j}{\sum_{j=1}^J w_j} \stackrel{a.s.}{\to} E_\pi(h(X))
+    $$
+    for any integrable function $h$ as $J \to \infty$, in the sense of \cite{chopin2004clt}. 
+\end{defn}
+
+It is known that the bootstrap and appropriately weighted particle filters target the posterior $f_{X_{1:N}|Y_{1:N}}$ in this sense, as long as they adhere to the procedure in \cite{chopin2004clt}.
+We now present three helper lemmas.
+The first follows from standard importance sampling arguments, the second from integrating out the marginal, and the third from Bayes' theorem.
+
+\begin{lem}[Change of Particle Measure]
+    \label{lem:change-measure-proper-weights}
+    Suppose that $\{(X_j,u_j),j=1,\dots,J\}$ targets $f_X$.
+Now, let $\{(Y_j,v_j),j=1,\dots,J\}$ be a sample drawn from $\{(X_j,u_j)\}$ where $(X_j,u_j)$ is represented, on average, $\pi_j J$ times.
+This could amount to multinomial resampling with probability $\pi_j$, or systematic resampling.
+Write
+    \[
+    (Y_j,v_j) = \big(X_{a(j)},u_{a(j)}/\pi_{a(j)}\big),
+    \]
+    where $a$ is called the ancestor function. Then, $\{(Y_j,v_j),j=1,\dots,J\}$ targets $f_X$.
+\end{lem}
+
+\begin{proof}
+    Let $h$ be an integrable function. 
+    \begin{align*}
+        \E\left[\sum_{j=1}^J h(Y_j) v_j\right] 
+        &= \sum_{j=1}^J \E\left[h(X_{a(j)}) \frac{u_{a(j)}}{\pi_{a(j)}}\right] \\
+        &= \sum_{j=1}^J \E\left[h(X_{j}) u_{j}\right],
+    \end{align*}
+    and similarly for the denominator.
+Now, the numerator and denominator of the reweighted quantity have the same expectation as the numerator and denominator of the original quantity, so they must converge to $ E_\pi(h(X))$ almost surely as well.
+\end{proof}
+
+\textbf{Remark:} Note that Lemma \ref{lem:change-measure-proper-weights} permits $\pi_{1:J}$ to depend on $\{(X_j,u_j)\}$ as long as the resampling is carried out independently of $\{(X_j,u_j)\}$, conditional on $\pi_{1:J}$.
+
+\begin{lem}[Particle Marginals]
+    \label{lem:marginal-proper-weights}
+    Suppose that $\{(X_j,u_j),j=1,\dots,J\}$ targets $f_X$.
+Also suppose that $Z_j \sim f_{Z|X}(\cdot | X_j)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
+Then, $\{(Z_j,u_j)\}$ targets $f_Z$.
+\end{lem}
+\begin{proof}
+    Let $h$ be an integrable function.
+
+    \begin{align*}
+        \E\left[J^{-1}\sum_{j=1}^J h(Z_j) u_j\right] &= \E\left[\E\left[J^{-1}\sum_{j=1}^J h(x_j) u_j f(Z_j|x_j)\Bigg| X_j=x_j\right]\right] \\
+        &= \E\left[J^{-1}\sum_{j=1}^J h(X_j) u_j f(Z_j|X_j)\right] \\
+        &\stackrel{a.s.}{\to} C_Z \E_{f_X}[h(X) f_{Z|X}(Z|X)] \\
+        &= C_Z \E_{f_Z}[h(Z)],
+    \end{align*}
+
+    where $C_Z$ is the normalizing factor, and similarly the denominator converges to $C_Z$. The result then follows from Slutsky's theorem.
+
+\end{proof}
+
+\begin{lem}[Particle Posteriors]
+    \label{lem:posterior-proper-weights}
+    Suppose that $\{(X_j,u_j),j=1,\dots,J\}$ targets $f_X$.
+Also suppose that $(X^\prime_j,u^\prime_j) = \big(X_j,u_j\, f_{Z|X}(z^*|X_j)\big)$.
+Then, $\{(X^\prime_j,u^\prime_j)\}$ targets $f_{X|Z}(\cdot | z^*)$.
+\end{lem}
+
+\begin{proof}
+    The proof is similar to that of Lemma \ref{lem:marginal-proper-weights}, except with Bayes' theorem. 
+\end{proof}
+
+\begin{prop}[MOP-1 Targets the Posterior]
+    When $\alpha=1$ or $\phi=\theta$, MOP-$\alpha$ targets the posterior. 
+\end{prop}
+\begin{proof}
+    When $\theta=\phi$, the ratio $\frac{g_{n,j}^\theta}{g_{n,j}^\phi}=1,$ and this reduces to the standard case.
+
+    When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows.
+Recursively applying Lemmas \ref{lem:change-measure-proper-weights} \ref{lem:marginal-proper-weights}, and \ref{lem:posterior-proper-weights}, we obtain that
+    %to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
+    the MOP-1 filter targets the posterior.
+    Specifically, suppose inductively that $\big\{\big(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
+    Then, Lemma \ref{lem:marginal-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big)\big\}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+    Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
+    \[
+    \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}\big/ g^\phi_{n,a(j)}\big),
+    \]
+    with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+\end{proof}
+
+This has addressed filtering, but not quite yet the likelihood evaluation. For this we use the following lemma.
+
+\begin{lem}[Likelihood Proper Weighting]
+    \label{lem:lik-proper-weight}
+  $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1_n-1}^*;\theta)$ is properly estimated by either the before-resampling estimate,
+\begin{equation}\label{L1}
+L_n^{B,\theta} =  \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
+\end{equation}
+or by the after-resampling estimate,
+\begin{equation}\label{L2}
+L_n^{A,\theta} = L^\phi_n \frac{\sum_{j=1}^Jw^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}.
+\end{equation}
+where $L^\phi_n$ is as defined in the various algorithms.
+\end{lem}
+
+Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ is properly weighted for $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+To see  (\ref{L2}),
+we write the numerator of (\ref{lem:change-measure-proper-weights}) as
+\[
+L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
+= L^\phi_n \sum_{j=1}^J w_{n,j}^{FC,\theta} \frac{g^\phi_{n,j}}{L_n^\phi}
+\]
+Using Lemma \ref{lem:change-measure-proper-weights}, we resample according to probabilities $\frac{g^\phi_{n,j}}{L_n^\phi}$ to see this is properly estimated by
+\[
+L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},
+\]
+from which we obtain (\ref{L2}).
+
+Using Lemma \ref{lem:lik-proper-weight}, we obtain a likelihood estimate,
+\[
+L^{A,\theta} = \prod_{n=1}^N \left( L^\phi_n \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J w^{P,\theta}_{n,j}}\right).
+\]
+Since $w^{F,\theta}_{n,j}=w^{P,\theta}_{n+1,j}$, this is a telescoping product. The remaining terms are
+$\sum_{j=1}^J w^{P,\theta}_{0,j} = J$ on the denominator and $\sum_{j=1}^J w^{F,\theta}_{N,j}$ on the numerator.
+This derives the MOP/DOP estimate.
+
+$L^{B,\theta}$ should generally be preferred, since there is no reason to include the extra variability from resampling when calculating the conditional log likelihood, but it lacks the nice telescoping product.
+
 \section{Doubly Off-Policy Particle Filters}
@@
     \end{equation}
-    successfully recovers $\lik(\theta)$.
+    successfully recovers $\lik(\theta)$. Therefore, DOP-$\alpha$ targets the posterior and the likelihood when $\alpha=1$ or $\theta=\phi$. 
 \end{prop}
@@
 
+\section{Gradients}
+
+\begin{prop}
+    Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The gradient estimate is then
+
+    \begin{equation}
+        \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
+    \end{equation}
+
+    yielding the gradient estimators of \cite{poyiadjis11, scibior2021dpf} when applied to the bootstrap filter. 
+\end{prop}
+
+\begin{proof}
+
+Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The likelihood estimate becomes
+\begin{equation*}
+    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\phi}}= \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\phi}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
+\end{equation*}
+as we have a telescoping product. Then, as
+
+$$\nabla_\theta \hat\ell(\theta) = \frac{\nabla_\theta \hat\lik(\theta)}{\hat\lik(\theta)} = \frac{\nabla_\theta\left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi}{\prod_{n=1}^N L_n^\phi} =  \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},$$
+
+The derivative of the log-likelihood estimate is then
+\begin{equation*}
+    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},
+\end{equation*}
+which we decompose as follows.
+
+Observe that as $\alpha=1$,
+
+$$w_{n,j}^{P,\theta} = w_{n-1,\theta}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,k_j}^\theta}{g_{i,k_j}^\phi}.$$
+
+Using the log-derivative identity again,
+$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,k_j}^\theta}{g_{i,k_j}^\phi}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,k_j}^\theta - \log g_{i,k_j}^\phi\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,k_j}^\theta.$$
+
+So 
+$$  \nabla_\theta \sum_{n=1}^N \log g_{n,k_j}^\theta = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,k_j}^\theta\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,k_j}^{P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right),$$
+
+and 
+$$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{i=1}^N \nabla_\theta \log g_{n,k_j}^\theta = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$
+
+Substituting, we have that
+\begin{equation*}
+    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right),
+\end{equation*}
+and finally, observing that when $\theta=\phi$ we have that $w_{N,j}^{F,\theta}=1$, that 
+\begin{equation*}
+    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
+\end{equation*}
+
+yielding the gradient estimators of \cite{poyiadjis11, scibior2021dpf} when applied to the bootstrap filter. 
+\end{proof}
+
 \end{document}

### 2023-11-25 mop/draft.tex
Updates from Overleaf

@@
 
-We claim that this yields a properly weighted particle filter that recovers the derivative estimate of \citet{doucet2011sf, scibior2021dpf} when $\theta$ is evaluated at $\phi,$ which is is Algorithm \ref{alg:dop} when $\alpha=1$.
-The proof can be found in Appendix A.
+We claim that this yields a particle filter that targets the posterior and recovers the derivative estimate of \citet{doucet2011sf, scibior2021dpf} when $\theta$ is evaluated at $\phi$.
+The proof can be found in Appendix \ref{app:dop}.
+
+\begin{prop}[Correctness of Doubly Off-Policy Particle Filter Corrections]
+    \label{prop:dop-correctness}
+
+    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$.
+Also let $\theta \in \gN(\phi)$.
+Then, the following corrections in the particle weights,
+    \begin{equation}
+    s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}, \qquad r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)},
+    \end{equation}
+    yield a particle filter that is unbiased for the likelihood, even when $\theta \neq \phi$.
+
+    In particular, when $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{doucet2011sf},
+    \begin{equation}
+        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
+    \end{equation}
+\end{prop}
+
+\subsection{Measurement Off-Policy Particle Filters}
+
+With the above, we have the entire system evolve according to $\phi$, and we differentiate through the likelihood ratios, which are smooth conditional on the resampling being the same (which in turn happens when $\theta \approx \phi$).
+However, the above algorithm requires access to likelihood ratios, and is therefore not compatible with simulation-based inference.
+
+There is a workaround.
+Consider the case where you have access to a differentiable simulator compatible with the reparametrization trick.
+If we have the state transitions evolve according to $\theta,$ but have the resampling according to $\phi,$ we can offload the derivative with respect to the process model to the reparametrization trick.
+
+We therefore make the following assumption in this section only:
+\begin{aspt}[Differentiable Simulator]
+    $\text{process}(\cdot | X_{t-1}; \theta)$ is differentiable with respect to $\theta,$ and is compatible with the reparametrization trick. 
+\end{aspt}
+
+\kevin{Develop theory for mixing-time swapping?
+Detect when IF2 reaches a neighborhood of the MLE, by detecting whether it has reached a local sort of stationary distribution when the perturbations are lower bounded, e.g.
+the autocorrelation between the mean vectors to estimate the mixing time or Gelman-Rubin.
+In practice, just swap when IF2 converges according to the diagnostics.
+Theory doesn't care, you just need to in a basin of attraction that is strongly convex.}
 
@@
 Observe that at Algorithm \ref{alg:dop}, at each timestep the corrected filtering importance weights $w_{n,j}^{F,\theta}$ optionally propagate to the next timestep for us to maintain the correction.
-This is because we introduce a discounting parameter $\alpha$ that affects the weights DOP, so $w_{n,j}^{P,\theta} := (w_{n-1,j}^{F,\theta})^{\alpha}$.
+We introduce a discounting parameter $\alpha$ that affects the weights, so $w_{n,j}^{P,\theta} := (w_{n-1,j}^{F,\theta})^{\alpha}$.
 Note that no matter what value $\alpha$ takes on, the system evolves according to $\phi$ when we perform the correction to evaluate the likelihood at $\theta$.
 The filter fully targets the posterior when $\theta=\phi$, or $\alpha=1$, as we show in Proposition \ref{prop:dop-correctness}.
-The value of $\alpha$ simply informs a bias-variance tradeoff, as we elaborate below.
+The value of $\alpha$ therefore informs a bias-variance tradeoff, as we elaborate below.
 
@@
 \begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\phi}}= \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\phi}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
+    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}}= \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\theta}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
 \end{equation}
-as we have a telescoping product. Its derivative is then
+as we have a telescoping product. Its derivative when $\theta=\phi$ is then
 \begin{equation}
@@
 the average gradient of each surviving particle's cumulative importance weights.
+The filter targets the posterior when $\alpha=1$ even when $\theta \neq \phi,$ and therefore suffers from no asymptotic bias.
 
@@
 \begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\phi}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J r_{n,j}s_{n,j},
+    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J r_{n,k_j}s_{n,k_j}}{\sum_{j=1}^J r_{n,j}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J r_{n,j}s_{n,j},
 \end{equation}
-and though the telescoping product disappears, this reduces to a series of one-step corrections that leave the filter still properly weighted.
+and though the telescoping product disappears, this reduces to a series of one-step corrections.
+However, these corrections are insufficient when $\theta \neq \phi,$ and the filter suffers from asymptotic bias unless $\theta=\phi.$ Its gradient when $\theta=\phi$ is therefore
+\begin{align}
+    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J  \nabla_\theta r_{n,j}s_{n,j}\right) \\
+    &= \frac{1}{J} \sum_{j=1}^J   \frac{\nabla_\theta f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{Y_n, X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)}
+\end{align}
+the gradient of the one-step corrections at each timestep.
 
@@
 
-\subsubsection{Correctness of DOP-$\alpha$}
+\subsubsection{Correctness of MOP-$\alpha$}
 
-\begin{prop}[Correctness of DOP-$\alpha$ (Algorithm \ref{alg:dop})]
-    \label{prop:dop-correctness}
-    Algorithm \ref{alg:dop} is a properly weighted particle filter that is unbiased for the likelihood, when either $\alpha=1$ or $\theta=\phi$.
+\begin{prop}[Correctness of MOP-$\alpha$ (Algorithm \ref{alg:mop})]
+    \label{prop:mop-correctness}
+    Algorithm \ref{alg:mop} is a properly weighted particle filter that is unbiased for the likelihood, when either $\alpha=1$ or $\theta=\phi$.
 
@@
 Its gradient in this case is either the estimate of \citet{doucet2011sf} when $\alpha=1$, a single-timestep estimator that corresponds to naively taking the derivative of the particle filter when $\alpha=0$ as first proposed in \citet{blei2018vsmc}
-
-\subsection{Measurement Off-Policy Particle Filters}
-
-DOP has the entire system evolve according to $\phi$, and we differentiate through the likelihood ratios, which are smooth conditional on the resampling being the same (which in turn happens when $\theta \approx \phi$).
-However, the above algorithm requires access to likelihood ratios, and is therefore not compatible with simulation-based inference.
-
-There is a workaround.
-Consider the case where you have access to a differentiable simulator compatible with the reparametrization trick.
-If we have the state transitions evolve according to $\theta,$ but have the resampling according to $\phi,$ we can offload the derivative with respect to the process model to the reparametrization trick.
-
-We therefore make the following assumption in this section only:
-\begin{aspt}[Differentiable Simulator]
-    $\text{process}(\cdot | X_{t-1}; \theta)$ is differentiable with respect to $\theta,$ and is compatible with the reparametrization trick. 
-\end{aspt}
-
-\kevin{Develop theory for mixing-time swapping?
-Detect when IF2 reaches a neighborhood of the MLE, by detecting whether it has reached a local sort of stationary distribution when the perturbations are lower bounded, e.g.
-the autocorrelation between the mean vectors to estimate the mixing time or Gelman-Rubin.
-In practice, just swap when IF2 converges according to the diagnostics.
-Theory doesn't care, you just need to in a basin of attraction that is strongly convex.}
 
@@
 
+The proof for Proposition \ref{prop:dop-correctness} follows from the below two lemmas.
 Evaluating the off-policy likelihood is possible with a weighted particle filter by using resampling weight $p(y^*_n|x_n;\phi)$ and making two corrections in the particle weight, through multiplying $w_{t,j}$ by the two likelihood ratios below,
@@
 
-\begin{prop}
+\begin{lem}
     For $\theta$, $\phi$ that satisfy Assumption \ref{assump:smooth-nbhd}, the following correction in the particle filter weights,
@@
     \end{equation}
-    successfully recovers $\lik(\theta)$. Therefore, DOP-$\alpha$ targets the posterior and the likelihood when $\alpha=1$ or $\theta=\phi$. 
-\end{prop}
+    successfully recovers $\lik(\theta)$. Therefore, DOP targets the posterior and the likelihood.
+\end{lem}
 
@@
      \end{align}
+
+     The proof for an arbitrary functional under the particle measure follows analogously.
+\end{proof}
+
+\begin{lem}[Gradient of Doubly Off-Policy Particle Filter]
+    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$.
+Also let $\theta \in \gN(\phi)$.
+When $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{doucet2011sf},
+    \begin{equation}
+        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
+    \end{equation}
+\end{lem}
+
+\begin{proof}
+    It is known, from \cite{doucet2009tutorial}, that 
+    $$p_\theta(x_{1:N}|y_{1:N}) = \prod_{n=1}^N \frac{p(x_n|x_{n-1})p(y_n|x_n)}{p(y_n|y_{1:n-1})} = \prod_{n=1}^N \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{\lik_n(\theta)}.$$
+
+    From the above lemma, and the fact that the particle filter is unbiased for the likelihood, we know that
+    \begin{align*}
+        \lik(\theta)
+        &= \int_{x_{1:N}} \prod_{n=1}^N \frac{1}{J}\sum_{j=1}^J p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) p_\theta(x_{1:N}|y_{1:N}) dx_{1:N} \\
+        &= \int_{x_{1:N}} \left(\prod_{n=1}^N \frac{1}{J}\sum_{j=1}^J p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \right) \frac{p_\theta(x_{1:N}|y_{1:N})}{p_\phi(x_{1:N}|y_{1:N})}p_\phi(x_{1:N}|y_{1:N}) dx_{1:N}\\
+        &= \int_{x_{1:N}} \left(\prod_{n=1}^N \lik_n(\theta) \right) \left(\prod_{n=1}^N \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \lik_n(\phi)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) \lik_n(\theta)}\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
+        &= \int_{x_{1:N}} \left(\prod_{n=1}^N \lik_n(\phi) \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) }\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N}.
+    \end{align*}
+
+    Therefore, under regularity conditions that permit differentiating under the integral sign, using the log-derivative trick, we have
+    \begin{align*}
+        \nabla_\theta \lik(\theta)
+        &= \int_{x_{1:N}} \nabla_\theta \hat\lik(\theta) \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N}\\
+        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N \lik_n(\phi) \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) }\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
+        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N \left(\frac{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n)}{\lik_n(\phi)}\right)^{-1} p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
+        &= \int_{x_{1:N}} \nabla_\theta \left(\frac{1}{p_\phi(x_{1:N}|y_{1:N}) }\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
+        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right) dx_{1:N} \\
+        &= \int_{x_{1:N}} \left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\nabla_\theta \log\left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right) dx_{1:N} \\
+        &= \int_{x_{1:N}} p_\theta(x_{1:N},y_{1:N}) \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) dx_{1:N} \\
+        &= \int_{x_{1:N}} \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) p_\theta(x_{1:N}|y_{1:N}) \cdot \lik(\theta) \; dx_{1:N}. 
+    \end{align*}
+
+    Using the log-derivative trick again,
+    \begin{align*}
+        \nabla_\theta \ell(\theta)
+        &= \frac{\nabla_\theta \lik(\theta)}{\lik(\theta)} \\
+        &= \int_{x_{1:N}} \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) p_\theta(x_{1:N}|y_{1:N}) dx_{1:N} \\
+        &= \E_{x_{1:N}|y_{1:N}, \theta}\left[\nabla_\theta \left(\sum_{n=1}^N\log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right)\right] \\
+        &= \E_{x_{1:N}|y_{1:N}, \theta}\left[\nabla_\theta\log p_\theta(x_{1:N},y_{1:N})\right] \\
+        &= \lim_{J\to\infty} \E_{x_{1:N}|y_{1:N}, \theta}\left[\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log p_\theta(x_{1:N,j}^{A,F},y_{1:N}^*)\right].
+    \end{align*}
+
+    But it also holds that
+    \begin{align*}
+        \nabla_\theta \ell(\theta)
+        &= \frac{\nabla_\theta \lik(\theta)}{\lik(\theta)} \\
+        &= \int_{x_{1:N}} \frac{\nabla_\theta \hat\lik(\theta)}{\lik(\theta)} \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N} \\
+        &= \int_{x_{1:N}} \nabla_\theta \hat\ell(\theta) \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N}\\
+        &= \lim_{J\to\infty} \E_{x_{1:N}|y_{1:N}, \phi}\left[\nabla_\theta \hat\ell(\theta)\right].
+    \end{align*}
+
+    and we see that $$\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log p_\theta(x_{1:N,j}^{A,F},y_{1:N}^*)$$ and $$\nabla_\theta \hat\ell(\theta)$$ both converge to the same thing in expectation.
+They are both particle approximations under the posterior where the variance of each is bounded, therefore one can apply the particle CLT from \cite{chopin2004clt} to obtain that they are both consistent estimates of the score.
+
+    In fact, when $\theta=\phi,$ they are the same in distribution. \kevin{claim}
+
+    \kevin{This is a far weaker result than what I really wanted. This just gives consistency. Want equality in distribution.}
 \end{proof}
@@
 
-Here, we consider a doubly off policy particle filter, meaning that both \code{rprocess} and \code{rmeasure} are computed at $\phi$ and particles are reweighted to correspond to $\theta$.
+Here, we consider a doubly off policy particle filter, meaning that both \code{rprocess} and \code{dmeasure} are computed at $\phi$ and particles are reweighted to correspond to $\theta$.
 
-\section{Gradients}
+\section{Measurement Off-Policy Particle Filters}
 

### 2023-11-25 optim.py
boxplot update

@@ -32,24 +32,24 @@ MONITORS = 1
 
 
 
-@partial(jit, static_argnums=2)
+#@partial(jit, static_argnums=2)
 def jgrad_pf(theta_ests, ys, J, covars, thresh, key=None):
     return jax.grad(pfilter_pf)(theta_ests, ys, J, covars=covars, thresh=thresh, key=key)
 
 
-@partial(jit, static_argnums=2)
+#@partial(jit, static_argnums=2)
 def jgrad(theta_ests, ys, J, covars, thresh, key=None):
     return jax.grad(pfilter_mean)(theta_ests, ys, J, covars=covars, thresh=thresh, key=key)
 
-@partial(jit, static_argnums=2)
+#@partial(jit, static_argnums=2)
 def jvg(theta_ests, ys, J, covars, thresh, key=None):
     return jax.value_and_grad(pfilter_mean)(theta_ests, ys, J, covars=covars, thresh=thresh, key=key)
 
-@partial(jit, static_argnums=2)
+#@partial(jit, static_argnums=2)
 def jhess(theta_ests, ys, J, covars, thresh, key=None):
     return jax.hessian(pfilter_mean)(theta_ests, ys, J, covars=covars, thresh=thresh, key=key)
 
-@partial(jit, static_argnums=2)
+#@partial(jit, static_argnums=2)
 def jpgrad(thetas, ys, J, sigmas, covars, a, thresh, key=None):
     return jax.grad(perfilter_mean, has_aux=True)(
         thetas, ys, J, sigmas, covars=covars, a=a,thresh=thresh, key=key)

### 2023-12-01 mop/draft.tex
Updates from Overleaf

@@
 
-\title{Automatic Differentiation to Accelerate Inference for Partially Observed Stochastic Processes}
+\title{Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation}
 \author{Kevin Tan, Giles J. Hooker, Edward L. Ionides}
@@
 Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods.
-Partially observed nonlinear stochastic dynamic systems have proved resistant to AD techniques due to (1) the requirement for simulation-based inference that does not require access to the system's transition probabilities and (2) the issue that widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters.
+Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques due to (1) the requirement for simulation-based inference that does not require access to the system's transition probabilities and (2) the issue that widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters.
 We propose a new representation that embeds existing methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff.
@@
 
+\kevin{Terminology: Plug-and-play (clearly definable, but not well-known, define at start of paper and don't use in title/abstract), likelihood-free maximum likelihood (seemingly an oxymoron, rationale being there is no need to evaluate density, bayesians call likelihood densities, plus we only approximate the likelihood and don't really have it), simulation-based inference (loaded term)}
+
 \subsection{Significance Statement}
@@
 
+\paragraph{Seed-Fixing, Coupling, and the Method of Common Random Numbers, Coupling:}
+
+In the rest of this paper, we will identify the seed with an element of the sample space.
+We do so because the seed determines the entire sequence of pseudo-random numbers as generated by a computer, to maintain an algorithmic and practical interpretation of our theory.
+When two algorithms share the same seed, this corresponds to the method of common random numbers.
+
 \section{A Smooth Extension of the Particle Filter}
@@
 The above intuition is formalized below.
+
+\kevin{Method of Common Random Numbers, Coupling}
 
@@
 
+\kevin{Cut down on DOP discussion in actual manuscript, remove any references to this as an algorithm, just talk about the idea of an off-policy system/particle filter.
+Bottom paragraph is the punchline.}
+
+\paragraph{Intuition:} Having the system evolve according to $\phi$ but evaluating the likelihood at $\theta$ requires us to reweight the conditional likelihood at each timestep by likelihood ratios corresponding to the process and measurement models.
+This corresponds to making two corrections in the particle weights at each timestep, one for the process model and one for the measurement model.
+The measurement correction cannot be avoided, but one can bypass the process correction -- with the reparametrization trick.
+We can do when we assume the simulator is differentiable with the reparametrization trick, as \cite{corenflos2021ot} also assume.
+Doing so then yields an algorithm that only requires a simulator, and does not require the process densities.
+
 \subsection{Measurement Off-Policy Particle Filters}
 
-With the above, we have the entire system evolve according to $\phi$, and we differentiate through the likelihood ratios, which are smooth conditional on the resampling being the same (which in turn happens when $\theta \approx \phi$).
-However, the above algorithm requires access to likelihood ratios, and is therefore not compatible with simulation-based inference.
+With the above procedure, we have the entire system evolve according to $\phi$, and we differentiate through the likelihood ratios, which are smooth conditional on the resampling being the same (which in turn happens when $\theta \approx \phi$).
+However, doing so requires access to likelihood ratios, and as such Algorithm \ref{alg:dop} is therefore not compatible with simulation-based inference.
 
@@
 Consider the case where you have access to a differentiable simulator compatible with the reparametrization trick.
-If we have the state transitions evolve according to $\theta,$ but have the resampling according to $\phi,$ we can offload the derivative with respect to the process model to the reparametrization trick.
+If we have the state transitions evolve according to $\theta,$ but have the resampling according to $\phi,$ we can offload the derivative with respect to the process model to the reparametrization trick, as in \cite{corenflos2021ot}.
+This corresponds to the MOP-$\alpha$ procedure in Algorithm \ref{alg:mop}.
 
@@
 \end{aspt}
+
+Algorithm \ref{alg:mop} provides a weighted particle filter that targets the posterior under $\theta$, but when differentiated yields either the low-bias but high-variance gradient estimator of \cite{poyiadjis11}, the high-bias but low-variance estimator of \cite{blei2018vsmc}, or an intermediate, interpolating quantity that can optimize a bias-variance tradeoff.
+This is controlled by a discounting parameter $\alpha$.
+The intuition behind this procedure is as follows.
+
+\paragraph{Intuition:} Broadly, if $\theta\neq\phi,$ one can first run a vanilla particle filter to recover values for the particles with the process model under $\phi.$ One can then run an accordingly reweighted particle filter with the process model under $\theta$, but resampling according to the measurement model under $\phi$, with the same indices as the original vanilla particle filter.
+This requires one to set the same random seed for both particle filters.
+This then yields a slightly-unconventionally weighted particle filter that targets the posterior, but has some very nice gradient properties; as we outline in the remainder of this section.
+
+\paragraph{Special Case:} When $\theta = \phi$, one can simply set the particles under $\phi$ to be copies of the particles under $\theta$.
+Gradients with respect to $\theta$ now treat functions of the particles under $\phi$ as constants, corresponding to the stop-gradient trick in \cite{scibior2021dpf}.
+This therefore provides a theoretical framework for their stop-gradient trick, which we happily use in our numerical experiments for simplicity.
 
@@
 
-Observe that at Algorithm \ref{alg:dop}, at each timestep the corrected filtering importance weights $w_{n,j}^{F,\theta}$ optionally propagate to the next timestep for us to maintain the correction.
+Observe that at Algorithm \ref{alg:mop}, at each timestep the corrected filtering importance weights $w_{n,j}^{F,\theta}$ optionally propagate to the next timestep for us to maintain the correction.
 We introduce a discounting parameter $\alpha$ that affects the weights, so $w_{n,j}^{P,\theta} := (w_{n-1,j}^{F,\theta})^{\alpha}$.
@@
 The filter fully targets the posterior when $\theta=\phi$, or $\alpha=1$, as we show in Proposition \ref{prop:dop-correctness}.
-The value of $\alpha$ therefore informs a bias-variance tradeoff, as we elaborate below.
+As previously mentioned, the value of $\alpha$ therefore informs a bias-variance tradeoff.
+We elaborate on this below.
 
@@
 \begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}}= \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\theta}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
+    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\theta}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
 \end{equation}
@@
 \begin{equation}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},
+    \nabla_\theta \hat{\ell}(\theta) := 
+        \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*),
 \end{equation}
@@
 \begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J r_{n,k_j}s_{n,k_j}}{\sum_{j=1}^J r_{n,j}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J r_{n,j}s_{n,j},
+    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J w_{n,j}^{F,\theta} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})},
 \end{equation}
@@
 However, these corrections are insufficient when $\theta \neq \phi,$ and the filter suffers from asymptotic bias unless $\theta=\phi.$ Its gradient when $\theta=\phi$ is therefore
-\begin{align}
-    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J  \nabla_\theta r_{n,j}s_{n,j}\right) \\
-    &= \frac{1}{J} \sum_{j=1}^J   \frac{\nabla_\theta f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{Y_n, X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)}
-\end{align}
-the gradient of the one-step corrections at each timestep.
 
-\kevin{Is DOP properly weighted when alpha is not 1 and theta is not phi?}
+    \begin{equation}
+        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
+    \end{equation}
+
+the gradient of the vanilla particle filter where the resampling correction is omitted as found in \cite{blei2018vsmc} -- when applied to the bootstrap filter.
+As \cite{scibior2021dpf} remark, this estimator only considers single-step quantities, and is "memoryless" beyond a single step.
+This is in contrast to the case when $\alpha=1,$ as that estimate only considers the surviving particles at time $N$, which is the other extreme where dependencies across time are fully tracked.
 
@@
     \label{prop:mop-correctness}
-    Algorithm \ref{alg:mop} is a properly weighted particle filter that is unbiased for the likelihood, when either $\alpha=1$ or $\theta=\phi$.
+    When either $\alpha=1$ or $\theta=\phi$, Algorithm \ref{alg:mop} is a particle filter that targets the posterior and is unbiased for the likelihood under $\theta$.
 
@@
 
-    and when $\alpha=0,$ is is the same as differentiating through the fixed-seed likelihood estimate yielded by a vanilla particle filter.
+    and when $\alpha=0,$ is the gradient estimator of \cite{blei2018vsmc},
+
+    \begin{equation}
+        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
+    \end{equation}
+
+    that as \cite{scibior2021dpf} show corresponds to differentiating through the fixed-seed likelihood estimate yielded by a vanilla particle filter.
 \end{prop}
 
-The proof for the above follows from Appendix A, as well as the technical lemmas for properly weighted particle filters.
+The proof for the above follows from Appendices A and C, as well as the technical lemmas for properly weighted particle filters.
 
@@
 Its gradient in this case is either the estimate of \citet{doucet2011sf} when $\alpha=1$, a single-timestep estimator that corresponds to naively taking the derivative of the particle filter when $\alpha=0$ as first proposed in \citet{blei2018vsmc}
+
+\section{Practical Maximum-Likelihood Inference}
+
+The DOP-$\alpha$ and MOP-$\alpha$ procedures therefore provide us with a unifying framework for reasoning about smooth extensions of weighted particle filters that target the posterior and provide reasonable gradient estimates when differentiated.
+This fixed-seed reweighting hopefully alleviates some confusion that can arise from differentiating through a quantity yielded by ostensibly non-differentiable resampling.
+This is because in the former we differentiate through smooth likelihood ratios, and in the latter we differentiate through a differentiable simulator with the reparametrization trick and through smooth likelihood ratios in order to adjust for the resampling.
+
+This still leaves us with the question of designing a reasonably effective procedure for likelihood maximization with these new tools.
 
@@
 
+\begin{lem}
+
+Consider the case of DOP-$\alpha$ when $\alpha=0$ and $\theta=\phi$. The gradient estimate is then
+
+    \begin{equation}
+        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta\right),
+    \end{equation}
+
+    yielding the gradient estimator of \cite{blei2018vsmc} that \cite{scibior2021dpf} show is the gradient of a vanilla particle filter when the resampling terms are dropped.
+
+\end{lem}
+
+\begin{proof}
+    When $\alpha=0,$ the likelihood estimate becomes
+\begin{equation}
+    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J r_{n,j}s_{n,j}}{\sum_{j=1}^J r_{n,j}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J r_{n,j}s_{n,j}.
+\end{equation}
+
+Its gradient when $\theta=\phi$ is therefore 
+\begin{align*}
+    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right) \\
+    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right)} \\
+    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta r_{n,j}s_{n,j}}{\sum_{j=1}^J r_{n,j}s_{n,j}} \\
+    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{Y_n, X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)} \\
+    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta\right),
+\end{align*}
+where we use the log-derivative trick, observe that $\sum_{j=1}^J r_{n,j}s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again. 
+\end{proof}
+
 \subsection{Off-Policy Weighted Differentiable Particle Filter}
@@
             \ENDFOR
-		\RETURN likelihood estimate $\hat{\lik}(\theta) = \lik(\theta, \phi, \omega, J) := \prod_{n=1}^N L_n^{A,\theta,\alpha}$, filtering distribution $\{(x_{N,j}^P, w^{F,\theta}_{N,j})\}.$
+		\RETURN likelihood estimate $\hat{\lik}(\theta) = \lik(\theta, \phi, \omega, J) := \prod_{n=1}^N L_n^{A,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}.$
 	\end{algorithmic}
@@
 
-\begin{prop}
+\begin{algorithm}[H]
+\centering
+	\caption{Measurement Off-Policy-$\alpha$}
+    \label{alg:mop}
+	\begin{algorithmic}[1]
+	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
+        \IF{$\theta \neq \phi$}
+            \STATE Run a particle filter once with parameters $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$, with seed $\omega$.
+        \ELSE
+        \STATE Set $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$ to be copies of $X_{n,j}^{P,\theta}, X_{n,j}^{P,\theta}$ in the rest of the algorithm.
+        \ENDIF
+		\STATE Initialize filter particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$. Fix $\omega.$
+		\FOR{$n=1,...,N$}
+            \STATE Prediction weights with discounting: $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
+            \label{mop-alpha:discount}
+            \STATE Simulate for prediction:
+            ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{mop-alpha:step1}
+            \STATE Evaluate measurement density:
+            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for $j$ in $\seq{1}{J}$
+            \STATE Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
+            \STATE Conditional likelihood under $\phi$: 
+            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
+            \label{dop-alpha:Lphi}
+            \STATE Normalize weights:
+            $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
+            for $j\ \mathrm{in}\ \seq{1}{J}$
+            \STATE Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \label{dop-alpha:systematic}
+            \STATE Resample particles: ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$
+            \STATE Filter weights corrected for resampling:
+            $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{dop-alpha:weight:update}
+            \STATE Resample filter weights:
+            $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
+            for $j$ in $\seq{1}{J}$ \label{dop-alpha:step2}
+            \STATE After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
+            \ENDFOR
+		\RETURN likelihood estimate $\hat{\lik}(\theta) = \lik(\theta, \phi, \omega, J) := \prod_{n=1}^N L_n^{A,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}.$
+	\end{algorithmic}
+\end{algorithm}
+
+\begin{lem}
     Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The gradient estimate is then
@@
     yielding the gradient estimators of \cite{poyiadjis11, scibior2021dpf} when applied to the bootstrap filter. 
-\end{prop}
+\end{lem}
 
@@
 
+\begin{lem}
+
+Consider the case of MOP-$\alpha$ when $\alpha=0$ and $\theta=\phi$. The gradient estimate is then
+
+    \begin{equation}
+        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
+    \end{equation}
+
+    yielding the gradient estimator of \cite{blei2018vsmc} that \cite{scibior2021dpf} show is the gradient of a vanilla particle filter when the resampling terms are dropped -- when applied to the bootstrap filter.
+
+\end{lem}
+
+\begin{proof}
+    When $\alpha=0,$ the likelihood estimate becomes
+\begin{equation}
+    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
+\end{equation}
+
+Its gradient when $\theta=\phi$ is therefore 
+\begin{align*}
+    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
+    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)} \\
+    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta s_{n,j}}{\sum_{j=1}^J s_{n,j}} \\
+    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)}{f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \phi}; \phi)} \\
+    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
+\end{align*}
+where we use the log-derivative trick, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again. 
+\end{proof}
+
 \end{document}

### 2023-12-11 mop/draft.tex
Updates from Overleaf

@@
 
+\kevin{Do groups of ten and take the best one from each group.
+Omit old IF2 from figure in the paper, provide complete figure with old IF2 in the appendix, saying that we modified the settings because of more modern practice.
+}
+
+\kevin{IFAD beats IF2 x percent of the time on 100 paired runs, but the bigger thing is that it gets close to the MLE without much effort.
+Previously, we'd have to do many searches, plus refinement (which is necessary, as we can see from the IF2 boxplot, it is easy to fool yourself into thinking that because you see the plateau that is the MLE).
+This includes likelihood profiling diagnostics, which are very computationally intensive.
+Now we can just take the best of ten searches -- IFAD does the refinement.}
+
 \kevin{IFAD vs IF2, AD vs IF2, IFAD0 vs IFAD1, AD0 vs AD1 in Dacca, local and global plots, convergence line plot of IFAD vs IF2, Dacca data plots, toy model from IF2 paper, AD0 vs AD1 gradient histograms, MSE of AD0 and AD1 gradient estimates as J increases on analytically tractable toy model from IF2 paper or OU2 in pomp}
@@
 \kevin{Benchmark DOP vs straight-through estimator/reinmax in MOP}
+
+\kevin{Have an outsider read the manuscript afterwards.}
 

### 2023-12-28 mop/draft.tex
Updates from Overleaf

@@
 \usepackage{enumitem}
+\usepackage{tikz}
 \setlist{nosep} % to reduce space in lists
@@
     \item Intractable likelihood functions, bypassed with the approaches of simulated likelihood and likelihood-free inference. 
-    \item The desire for simulation-based inference without access to the transition density of the latent Markov process, with various methods created by \cite{welch2009abc, wood2010sl, doucet2010pmcmc, Ionides_infdynsys, Ionides_infpomp}.
+    \item The desire for simulation-based inference without access to the transition density of the latent Markov process, with various methods created by \cite{welch2009abc, wood2010sl, doucet2010pmcmc, ionides08, ionides15}.
     \item Accurate parameter estimation in the presence of significant Monte-Carlo noise in the likelihood estimate.
@@
 
-This is the backbone behind the only simulation-based full-information maximum likelihood method available in the literature thus far, the improved iterated filtering algorithm of \citet{Ionides_infpomp}.
-\citet{Ionides_infpomp} utilize the bootstrap filter for perform maximum likelihood estimation with an iterated perturbed Bayes map, successfully tackling challenging problems in epidemiology that available Bayesian methodology fails to solve.
+This is the backbone behind the only simulation-based full-information maximum likelihood method available in the literature thus far, the improved iterated filtering algorithm of \citet{ionides15}.
+\citet{ionides15} utilize the bootstrap filter for perform maximum likelihood estimation with an iterated perturbed Bayes map, successfully tackling challenging problems in epidemiology that available Bayesian methodology fails to solve.
 
 Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small.
-For example, though in practice the algorithm of \citet{Ionides_infpomp} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood.
+For example, though in practice the algorithm of \citet{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood.
 \citet{Ionides_mcap} attempt to address the third problem, albeit with a computationally-expensive approach.
@@
 \begin{aspt}[Locally Bounded Derivative]
+\label{assump:local-bounded-derivative}
 Let $\mathcal{M}$ be an open subset of $\Omega$. There exists some function $G(\theta)$ and a constant $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$ such that
@@
 
+\paragraph{Bottom Line:} The DOP-$\alpha$ and MOP-$\alpha$ procedures therefore provide us with a unifying framework for reasoning about smooth extensions of weighted particle filters that target the posterior and provide reasonable gradient estimates when differentiated.
+This fixed-seed reweighting alleviates issues that can arise from differentiating through a quantity yielded by ostensibly non-differentiable resampling.
+This is done by either (1) differentiating through smooth likelihood ratios, or (2) differentiating through a differentiable simulator with the reparametrization trick (and through smooth likelihood ratios to adjust for the resampling).
+
 \section{Practical Maximum-Likelihood Inference}
 
-The DOP-$\alpha$ and MOP-$\alpha$ procedures therefore provide us with a unifying framework for reasoning about smooth extensions of weighted particle filters that target the posterior and provide reasonable gradient estimates when differentiated.
-This fixed-seed reweighting hopefully alleviates some confusion that can arise from differentiating through a quantity yielded by ostensibly non-differentiable resampling.
-This is because in the former we differentiate through smooth likelihood ratios, and in the latter we differentiate through a differentiable simulator with the reparametrization trick and through smooth likelihood ratios in order to adjust for the resampling.
-
 This still leaves us with the question of designing a reasonably effective procedure for likelihood maximization with these new tools.
+We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that amounts to warm-starting gradient descent or some other reasonable first or second order procedure with the output of an initial search of IF2.
+
+\paragraph{Summary:} A coarse solution is found by running IF2 with an aggressive random walk standard deviation (learning rate) and a cooling schedule that is \textbf{bounded below}, until the search stalls to an initial "convergence" to what is hopefully a neighborhood of the MLE.
+We can then use gradient descent, or other first or second order gradient methods, to refine this coarse solution.
+
+The initial "convergence" happens fairly quickly in practice.
+In the case of the Dacca cholera model of \cite{king08}, when a geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations of IF2.
+In comparison, often 100 or 200 iterations of IF2 are used for an initial global search.
+\cite{ionides15}, for example, use 100 iterations with the Dacca cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{Lee_haiticholera}.
+
+\paragraph{Rationale:}
+
+IF2 converges quickly to a neighborhood of some reasonably good local optimum in practice, but then struggles to squeeze out the last few log-likelihood units.
+Conversely, gradient descent (and other first or second order gradient methods) can get stuck in saddle points and less desirable local minima when the likelihood is nonconvex.
+Warm-starting gradient methods with IF2 lets us get close enough to the MLE for gradient methods to shine, as in most cases regularity conditions ensure that the likelihood surface is well-behaved enough in this neighborhood that the aforementioned issues with saddle points and local minima are alleviated.
+Combining these two methods in this way lets us enjoy the best of both worlds.
+
+\begin{algorithm}[H]
+\centering
+	\caption{Iterated Filtering with Automatic Differentiation}
+    \label{alg:ifad}
+	\begin{algorithmic}[1]
+	     \STATE \textbf{Input:} Number of particles $J$, timesteps $T$, measurement model $p(y_t^*|x_t, \theta)$, simulator $\process(x_{t+1}|x_t, \theta),$ IF2 cooling schedule $\eta_n$, MOP-$\alpha$ discounting parameter $\alpha$.
+		\STATE Initialize $\theta_0$, $n=0.$
+        \STATE Run IF2 until initial "convergence" under cooling schedule $\eta_n$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_n := \frac{1}{J}\sum_{j=1}^J \Theta_j.$
+		\WHILE{Procedure not converged:}
+		\STATE Draw $\omega \in \Omega$
+		\STATE Perform forward pass by running Algorithm \ref{alg:mop} with discounting parameter $alpha$ to obtain $\loglik(\theta_n, \theta_n, \omega, J).$
+		\STATE Obtain $g(\theta_n) = \nabla_{\theta_n} (-\loglik(\theta_n, \theta_n, \omega, J))$ via backpropagation, $H(\theta_n)$ such that $\lambda_{\min}(H) \geq c$.
+		\STATE Obtain $\eta$ with line search or an annealing schedule.
+		\STATE Update $\theta_{n+1} := \theta_n - \eta (H(\theta_n))^{-1} g(\theta_n)$, $n:=n+1.$
+		\ENDWHILE
+		\STATE Return $\hat{\theta} := \theta_n.$
+	\end{algorithmic}
+\end{algorithm}
+
+\subsection{Theoretical Guarantee}
+
+We first analyze the performance of gradient descent, Newton's method, or any iterative method that multiplies the particle estimate of the gradient with the inverse of a matrix with minimum eigenvalue bounded below, in the case where the log-likelihood surface is $\gamma$-strongly convex.
+In this case, these methods enjoy a linear convergence guarantee:
+
+\begin{thm}[Convergence of Gradient Methods]
+\label{thm:convergence}
+
+Consider a variant of Algorithm \ref{alg:ifad} where one stops if the gradient estimate $||g(\theta_n)|| \leq \sigma \epsilon$.
+Assume $-\ell$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$.
+$\sigma = \frac{4 \Gamma}{(1-\beta)}> 4$.
+Define $G$ as in Assumption \ref{assump:local-bounded-derivative}.
+
+Choose $J$ according to the maximum of the $J$ prescribed in Lemmas \ref{lemma:grad_bound} and $\ref{lemma:hess_bound}$ with failure probability $\delta/2$ each.
+Let $H(\theta_n)$ be a matrix possibly dependent on $\theta_n$ with minimum eigenvalue always greater than some $c>0$.
+Choose the learning rate $\eta$ and error tolerance $\epsilon$ such that, where $\beta$ is the Armijo condition hyperparameter,
+\begin{equation}
+    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_n)||.
+\end{equation}
+
+Then the following holds with probability at least $1-\delta$:
+\begin{equation}
+\ell(\theta^*) - \ell(\theta_{n+1}) \leq \left(1-\eta\beta\frac{8\gamma}{9c}\right)(\ell(\theta^*)-\ell(\theta_n))
+\end{equation}
+
+and the algorithm eventually terminates after we obtain $||\nabla_\theta \ell(\theta_n)|| \leq (1+\sigma) \epsilon$.
+\end{thm}
+
+The proof of this result uses the concentration inequalities from \cite{delmoral2011ci}, and leans upon the analysis of \cite{mahoney2016subsampled}. We defer the proof to Appendix \ref{app:ifad}
+
+\paragraph{Takeaways for IFAD (Algorithm \ref{alg:ifad}):}
+
+The gradient stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the IF2 stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
+This happens fairly often in practice, for example, when sufficient regularity conditions for local asymptotic normality of the MLE hold.
+
+We conjecture that this applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE.
+IF2 posesses some SGD-like qualities, including what looks like linear convergence to a ball around the MLE with width determined by the random walk standard deviation.
+We will explore this conjecture in future work.
+
+\subsection{Runtime}
+
+Even with the increased computational cost of automatic differentiation, Algorithm \ref{alg:ifad} is reasonably fast, as by the cheap gradient principle from \cite{kakade2019provably}, a gradient operation takes no more than 6 times the runtime of the particle filter.
+
+In addition, re-implementation of the particle filter, iterated filtering, and Algorithm \ref{alg:ifad} itself in \texttt{JAX} (\cite{jax2018github}) led to considerable speedups.
+This can be attributed to both GPU acceleration and the just-in-time (JIT) compilation functionality that \texttt{JAX} provides.
+When tested on a computer with an Intel i9-13900K processor, 128GB of RAM, and a NVIDIA RTX3090 GPU, each post-compilation particle filtering operation on the Dacca cholera model from \cite{king08} with 10000 particles took roughly 379 milliseconds (with a standard deviation of 279 microseconds over seven runs), a considerable improvement over the existing performance in the widely used \texttt{pomp} package of \cite{king2016pomp}, which took between 6.19 to 6.39 seconds over ten runs on the same machine.
 
 \section{Numerical Experiments}
+
+\subsection{Cholera in Bangladesh}
+
+To compare the performance of IFAD against that of IF2, we fit the very same model first proposed in \cite{king08} that \cite{ionides15} use to demonstrate the capabilities of IF2.
+This is a stochastic SIR compartmental model with transition and measurement uncertainty, where the population at time $t$, $H(t)$, is divided into the susceptible compartment $S(t)$, infected compartment $I(t)$, and three ($k=3$) recovered compartments $R_1(t), ..., R_k(t)$ denoting varying degrees of cholera immunity.
+We write $M(t)$ for the cholera deaths in each month.
+As in \cite{ionides15}:
+
+\begin{itemize}
+    \item The \textbf{transition dynamics} follow this series of stochastic differential equations:
+
+    \begin{align}
+        d S&=\left(k \epsilon R_k+\delta(S-H)-\lambda(t) S\right) d t+d H-(\sigma S I / H) d B, \\ 
+        d I&=\left(\lambda(t) S-(m+\delta+\gamma) I\right) d t+(\sigma S I / P) d B, \\ 
+        d R_1&=\left(\gamma I-(k \epsilon+\delta) R_1\right) d t, \\ 
+        \vdots \\ 
+        d R_k&=\left(k \epsilon R_{k-1}-(k \epsilon+\delta) R_k\right) d t,
+    \end{align}
+
+    where $B$ represents Brownian motion, $m$ is the rate at which infected individuals succumb to cholera, $\gamma$ is the rate at which infected individuals recover, $1/\epsilon$ is the mean duration of immunity, $\sigma$ represents the standard deviation of perturbations to the force of infection, and $\delta=0.02$ is the death rate.
+
+    \item The \textbf{force of infection}, $\lambda(t)$, is modeled by
+    \begin{equation}
+        \lambda(t)=\exp \left\{\beta_{\text {trend }}\left(t-t_0\right)+\sum_{j=1}^{6} \beta_j s_j(t)\right\}(I / P) + \exp \left\{\sum_{j=1}^{6} \omega_j s_j(t)\right\},
+    \end{equation}
+    where the $\beta_j$ model seasonality in the force of infection, $\beta_{\text{trend}}$ models the trend in the force of infection, and the $\omega_j$ represent seasonality of a non-human environmental reservoir of disease.
+    \item The \textbf{measurement model} for observed monthly cholera deaths is given by 
+    \begin{equation}
+        Y_n \sim \gN(M_n, \tau^2M_n^2),
+    \end{equation}
+    where $M_n$ is the true number of cholera deaths in that month.
+\end{itemize}
+
+\usetikzlibrary{arrows.meta,positioning}
+\begin{figure}[h]
+\centering
+\begin{tikzpicture}[node distance=1cm and 1cm, auto,
+>=Latex,every node/.append style={align=center}, 
+int/.style={draw, minimum size=1.6cm}] 
+\node [int] (S) [right=of H] {$S$};
+\node [int] (I) [right=of S] {$I$};
+\node (Mcoord) [right=of I, coordinate] {Mcoord};
+\node [int] (R1) [right=of Mcoord] {$R_1$};
+\node [int] (R2) [right=of R1] {$R_2$};
+\node [int] (R3) [right=of R2] {$R_3$};
+\node [circle, draw] (M) [above=of Mcoord] {$M$};
+\node (deathsS) [above=of S, coordinate] {deathsS};
+\node (deathsI) [above=of I, coordinate] {deathsS};
+\node (deathsR1) [above=of R1, coordinate] {deathsR1};
+\node (deathsR2) [above=of R2, coordinate] {deathsR2};
+\node (deathsR3) [above=of R3, coordinate] {deathsR3};
+\node (belowR3) [below=of R3, coordinate] {belowR3};
+\node (belowS) [below=of S, coordinate] {belowS};
+\path[->]
+    (H) edge node {$dH$} (S)
+    (S) edge node {$\lambda(t)$}  (I)
+    (I) edge node {$\gamma$}  (R1)
+    (R1) edge node {$k\epsilon$}  (R2)
+    (R2) edge node {$k\epsilon$}  (R3)
+    (R3) edge node {} (belowR3) 
+    (belowR3) edge node {$k\epsilon$} (belowS) 
+    (belowS) edge node {} (S)
+    (I) edge node {$m$}  (M)
+    (S) edge node {$\delta$}  (deathsS)
+    (I) edge node {$\delta$}  (deathsI)
+    (R1) edge node {$\delta$}  (deathsR1)
+    (R2) edge node {$\delta$}  (deathsR2)
+    (R3) edge node {$\delta$}  (deathsR3)
+\end{tikzpicture}
+\caption{Illustration of SIR model from \cite{king08}.}
+\label{fig:sir}
+\end{figure}
+
+Figure \ref{fig:sir} depicts the above compartmental model that we use to benchmark the performance of IFAD.
+We present comparisons of IFAD with $\alpha=0$ (IFAD-0) and $\alpha=1$ (IFAD-1) against both (1) the results from \cite{ionides15}, and (2) our own implementation of IF2 in \texttt{JAX}.
+
+\paragraph{Hyperparameters:} While performing an initial exploratory analysis, we found that alternative hyperparameters for IF2 (random walk standard deviation of $\sigma=0.02$, geometrically decreasing cooling rate exponent $a=0.95$) performed better than the settings \cite{ionides15} chose, and chose to use these for our implementation to achieve a fairer comparison.
+Our implementations of IF2 and IFAD used 10000 particles, in line with \cite{ionides15}.
+
+To prevent exploding gradients given the high variance in the MOP-1 estimator, we scale the gradient to have norm 1 before performing vanilla gradient descent, and use a linear cooling schedule of $0.2$ to $ 0.0001$.
+We chose to use vanilla gradient descent for simplicity, and to demonstrate that minimal tuning was required to have IFAD perform well.
+
+\begin{figure}
+    \centering
+    \includegraphics{imgs/095/qq.png}
+    \caption{Caption}
+    \label{fig:enter-label}
+\end{figure}
+
+% The 60 more iterations of IF2 yielded largely not a huge improvement over the initial coarse solution of 40 iterations.
+This demonstrates what we have mentioned earlier: that although IF2 manages to quickly find a decent coarse solution, it struggles to squeeze out those last few units of log-likelihood improvement.
+The 60 more iterations of gradient descent, however, successfully refines this solution, getting much closer to the MLE on average.
 
@@
 
-\section{Properly Weighted Particle Filters}
+\section{Weighted Particle Filters that Target the Posterior}
 
 \begin{defn}[Targeting, \cite{liu1998smc}]
-    A random variable $X$ drawn from a distribution $g$ is properly weighted by a weighting function $w(X)$ with respect to the distribution $\pi$ if for any integrable function $h$,
+    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any integrable function $h$,
     $$
-    E_g\{h(X) w(X)\}=E_\pi\{h(X)\}
+    E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
     $$
 
-    A set of particles $(X_j, w_j), j=1,2, \ldots$, targets $\pi$ if
+    A set of particles $(X_j, w_j), j=1,2, \ldots$, \textbf{targets} $\pi$ if
     $$
@@
     $$
-    for any integrable function $h$ as $J \to \infty$, in the sense of \cite{chopin2004clt}. 
+    for any integrable function $h$ as $J \to \infty$, in the sense of the triangular array convergence in \cite{chopin2004clt}. 
 \end{defn}
@@
 
+\section{Finite-Particle Analysis of Gradient Methods with Algorithm \ref{alg:ifad}}
+\label{app:ifad}
+
+The analysis in this section roughly follows the analysis in Roosta-Khorasani and Mahoney \cite{mahoney2016subsampled}, except with the caveat that none of the matrix concentration bounds they use apply here as the particles are dependent.
+We instead use the concentration inequality from Del Moral and Rio \cite{delmoral2011ci} to bound the gradient and Hessian estimates.
+In this section, we fix $\omega \in \Omega$ only within each filtering iteration, evaluate Algorithm \ref{alg:mop} at $\theta=\phi$, and analyze Algorithm \ref{alg:ifad} post-iterated filtering.
+
+The convergence analysis in Theorem \ref{thm:convergence} is limited to the case where $-\ell$ is $\gamma$-strongly convex.
+Though it is true that in a neighborhood of the optimum local asymptotic normality holds and the log-likelihood is strongly convex in this neighborhood, in practice likelihood surfaces for POMPs are often highly nonconvex globally.
+The convergence to an optimum, local or global, must therefore be sensitive to initialization.
+
+\subsection{Bounding the Gradient}
+
+\begin{lem}[Concentration of Measure for Gradient Estimate]
+    \label{lemma:grad_bound}
+    Assume $\text{osc} \left(\frac{\partial}{\partial_\theta}\right) \leq 1$.
+For $||\nabla_\theta \hat{\loglik}(\theta) - \nabla_\theta \loglik(\theta)||$ to be bounded by $\epsilon$ with probability $1-\delta$, we require
+    \begin{equation}
+        J > \max\left\{\frac{r_t\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{6p}{\delta}\right)\right)\right),\;\; \frac{2p\log(6p/\delta)}{\epsilon^2}\beta_t^2, \;\;\frac{G(\theta)^2}{\epsilon^2}p\log\left(\frac{6p}{\delta}\right)\right\}
+    \end{equation}
+    where $G(\theta)$ is defined as in Assumption \ref{assump:local-bounded-derivative}, $h(t) = \frac{1}{2}(t - \log(1+t))$, and $\beta_t$ and $r_t$ are two additional model-specific parameters that do not depend on $J$.
+\end{lem}
+
+\begin{proof}
+
+We will seek to use the concentration inequality of \cite{delmoral2011ci} to bound the deviation of the gradient estimate from the gradient of the negative log-likelihood in the sup norm with a union bound.
+To use this, we will first need to resample the particles once more to reassign the normalized weights to $1/J$.
+
+Perform another resampling step at time $T$ to obtain indices $k_j \sim \text{Categorical}(\bar{w}_{T,1}, ..., \bar{w}_{T,J})$.
+Obtain ancestral trajectories of surviving time $T$ filtering particles $x_{k_j, 1:T}^{A}$.
+Define $g_j(\theta) = \nabla_\theta -\log L_{T, k_j}$, the gradient of the negative logarithm of the accumulated importance weights of particle $k_j$ along its ancestral trajectory.
+
+We can bound
+\begin{align}
+    ||\nabla_\theta \hat{\loglik}(\theta) - \nabla_\theta \loglik(\theta)|| 
+    \leq 
+    \left\lVert (\nabla_\theta - \hat{\loglik}(\theta)) - \frac{1}{J}\sum_{j=1}^J g_j(\theta)\right\rVert 
+    + \left\lVert\frac{1}{J}\sum_{j=1}^J g_j(\theta) - (\nabla_\theta -\loglik(\theta))\right\rVert
+\end{align}
+
+We begin by examining the first term. This term bounds the error of the gradient estimate that the differentiable particle filter provides from the gradient estimate after the final resampling step.
+
+This amounts to solving the following problem.
+Say you have a weighted mean $\sum w_j x_j$, where $j=1,…,J$.
+The $x_j$ are fixed numbers bounded by some constant $G$ and the $w_j$ are non-negative and sum to $1$.
+Resample indices $k_j, j=1,…,J$ according to the multinomial distribution to get $x_{k_j}$ for each $j$, and also $J^{-1} \sum x_{k_j}$.
+We want to bound $|\sum w_j x_j - J^{-1} \sum x_{k_j}|$.
+
+By a simple application of Hoeffding's inequality and a union bound, we have
+
+\begin{align}
+    \prob\left(\exists \; j \;s.t.\; \left|(\nabla_\theta - \hat{\loglik}(\theta)) - \frac{1}{J}\sum_{j=1}^J g_j(\theta)\right| \geq \epsilon \right)
+    &\leq 2p\exp\left(\frac{-2\epsilon^2J}{G(\theta)^2}\right) 
+\end{align}
+
+We require \begin{equation}
+    J > \frac{G(\theta)^2}{2\epsilon^2}\log(2p/\delta)
+\end{equation} particles to bound $\left\lVert (\nabla_\theta - \hat{\loglik}(\theta)) - \frac{1}{J}\sum_{j=1}^J g_j(\theta)\right\rVert_\infty < \epsilon$ with probability $1-\delta$, where $G(\theta)$ is defined as in Assumption \ref{assump:local-bounded-derivative}.
+
+We can bound the sup norm by $||x||_\infty \leq \sqrt{p}||x||_2$.
+So to bound $\left\lVert (\nabla_\theta - \hat{\loglik}(\theta)) - \frac{1}{J}\sum_{j=1}^J g_j(\theta)\right\rVert < \epsilon$ with probability $1-\delta/3$, we scale the previous $\epsilon$ to $\epsilon/\sqrt{p}$ and therefore require
+
+\begin{equation}
+    J > \frac{G(\theta)^2}{2\epsilon^2}p\log\left(\frac{6p}{\delta}\right)
+\end{equation}
+
+We now tackle the second term.
+Write $g_{ij}$ for the $i$-th component of the gradient of the negative log-likelihood with respect to the $j$-th particle, $(-\bar{w}_{T,\np} \nabla_\theta\log L_{T,j})_i$.
+
+Apply the concentration inequality in \ref{eq:ci} from Del Moral and Rio \cite{delmoral2011ci} and another union bound to find that
+
+\begin{align}
+    \left\lVert\frac{1}{J}\sum_{j=1}^J g_{ij} - \left(-\frac{\partial}{\partial \theta_i}\ell(\theta)\right)\right\rVert_\infty \leq \frac{r_t}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_t 
+\end{align}
+
+with probability $1-2p\exp(-t)$.
+We split the remaining $2\delta/3$ failure probability among these two terms, to find $\delta/3=2p\exp(-t) => t=\log(6p/\delta)$.
+Here, $h(t) = \frac{1}{2}(t - \log(1+t))$.
+The two additional model-specific parameters are $\beta_t$ and $r_t$, which do not depend on $J$.
+
+The analogous bound for the 2-norm falls out by scaling the right-hand side by $\sqrt{p}$, to require
+
+\begin{align}
+    \epsilon = \frac{r_t\sqrt{p}}{J}(1+h^{-1}(\log(6p/\delta))) + \sqrt{\frac{2p\log(6p/\delta)}{J}}\beta_t
+\end{align}
+
+and therefore need
+
+\begin{align}
+    J > \max\left\{\frac{r_t\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{6p}{\delta}\right)\right)\right), \frac{2p\log(6p/\delta)}{\epsilon^2}\beta_t^2\right\}
+\end{align}
+
+Overall, we finally require
+
+\begin{equation}
+        J > \max\left\{\frac{r_t\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{6p}{\delta}\right)\right)\right),\;\; \frac{2p\log(6p/\delta)}{\epsilon^2}\beta_t^2, \;\;\frac{G(\theta)^2}{\epsilon^2}p\log\left(\frac{6p}{\delta}\right)\right\}
+\end{equation}
+
+\end{proof}
+
+\subsection{Bounding Hessian Estimates}
+
+Should one choose to use a second-order method involving a particle Hessian estimate, we provide a guarantee for its positive-definiteness below.
+
+\begin{lem}[Minimum Eigenvalue Bound for Hessian Estimate]
+    \label{lemma:hess_bound}
+    Assume that the Hessian of the negative log-likelihood $H=\sum_{j=1}^J \E H_j$ has a minimum eigenvalue $\gamma > 0$, that $\E \lambda_{\min} (H_j) = \gamma' > 0$, and finally that $\text{osc}(\lambda_{\min}) \leq 1$.
+
+    We then require 
+    \begin{equation}
+        J > \max\left\{\frac{2r_t(1+h^{-1}(t)) + 2c}{\gamma'}, \frac{2(2t\beta_t^2+c)^2}{\gamma'^2}\right\} \geq  \frac{r_t(1+h^{-1}(t))}{\gamma'} + \sqrt{2tJ}\beta_t/\gamma' + c/\gamma'
+    \end{equation}
+
+    for $\hat{H}(\theta)$ to be invertible and positive definite with minimum eigenvalue greater than or equal to $c \in (0, \sum_{j=1}^J \E\lambda_{\min}(H_j))$ and with probability at least $1-\exp(-t)$.
+\end{lem}
+\begin{proof}
+Write $\hat{H}(\theta) = \hat{H} = \sum_{j=1}^J H_j$ for the estimate of the negative of the Hessian, where $H_j$ is an element of the outer sum over the $J$ particles.
+
+ As the negative log-likelihood is convex, we want to bound the minimum eigenvalue of $\hat{H}(\theta)$ from below with high probability, so that all the eigenvalues of $\hat{H}(\theta)$ are positive with high probability.
+This ensures that the estimated Hessian is invertible and positive-definite.
+
+From \cite{boyd2004convex}, we know that the minimum eigenvalue of a symmetric matrix is concave. Therefore, it suffices to show that the first inequality in the below expression
+
+\begin{equation}
+    0 < \sum_{j=1}^J \lambda_{\min} (H_j) \leq  \lambda_{\min}\left(\sum_{j=1}^J H_j\right) = \lambda_{\min} (\hat{H})
+\end{equation}
+
+holds with high probability.
+
+Apply the particle Hoeffding concentration inequality from \cite{delmoral2011ci} to find that
+
+\begin{align}
+    \frac{1}{J}\sum_{j=1}^J \lambda_{\min}(H_j) - \E_{\tilde{\pi}_t}\lambda_{\min}(H_j) &= \frac{1}{J}\sum_{j=1}^J \lambda_{\min}(H_j) - \gamma' \geq -\frac{r_t}{J}(1+h^{-1}(t)) - \sqrt{\frac{2t}{J}}\beta_t \\
+    \sum_{j=1}^J \lambda_{\min}(H_j) &\geq -r_t(1+h^{-1}(t)) - \sqrt{2tJ}\beta_t + J\gamma'
+\end{align}
+
+with probability at least $1-\exp(-t)$. Here, $h(t) = \frac{1}{2}(t - \log(1+t))$. The two additional model-specific parameters are $\beta_t$ and $r_t$, which do not depend on $J$.
+
+We additionally require, for $c \in (0, \sum_{j=1}^J \E\lambda_{\min}(H_j))$,
+\begin{align}
+    \sum_{j=1}^J \lambda_{\min}(H_j) \geq -r_t(1+h^{-1}(t)) - \sqrt{2tJ}\beta_t + J\gamma' \geq c \\
+    J\gamma' \geq c + r_t(1+h^{-1}(t)) + \sqrt{2tJ}\beta_t 
+\end{align}
+
+We therefore require $$J > \max\left\{\frac{2r_t(1+h^{-1}(t)) + 2c}{\gamma'}, \frac{2(2t\beta_t^2+c)^2}{\gamma'^2}\right\} \geq  \frac{r_t(1+h^{-1}(t))}{\gamma'} + \sqrt{2tJ}\beta_t/\gamma' + c/\gamma'$$
+
+for $\hat{H}(\theta)$ to be invertible and positive definite with minimum eigenvalue greater than or equal to $c$ with probability at least $1-\exp(-t)$.
+
+\end{proof}
+
+\subsection{Convergence Analysis of Theorem \ref{thm:convergence}}
+
+\begin{proof}
+
+In this analysis, we largely follow the proof of Theorem 6 in \cite{mahoney2016subsampled}.
+
+Define $\theta_\eta = \theta_n + \eta p_n$, where $p_n=-(H(\theta_n))^{-1}g(\theta_n)$.
+As in Roosta-Khorasani and Mahoney \cite{mahoney2016subsampled}, we want to show there is some iteration-independent $\tilde{\eta}>0$ such that the Armijo condition
+\begin{equation}
+    f(\theta_n+\eta p_n) \leq f(\theta_n) + \eta\beta p_n^Tg(\theta_n)
+\end{equation}
+holds for any $0< \eta < \tilde{\eta}$ and some $\beta \in (0,1)$.
+
+By an argument found in the beginning of the proof of Theorem 6 in \cite{mahoney2016subsampled}, we have that choosing $J$ such that $||\nabla_\theta\hat{\ell}(\theta_n) - \nabla_\theta \ell(\theta_n)|| \leq \epsilon$ and $\lambda_{\min}(H(\theta_n)) \geq c$ with probability $(1-\delta)^2$ for each $n$ yields
+
+\begin{align}
+    f(\theta_\eta)-f(\theta_n) \leq \eta p_n^Tg(\theta_n) + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2
+\end{align}
+
+Now, for some $c>0$, with probability $1-\delta$
+\begin{equation}
+    p_n^Tg(\theta_n) = -p_n^TH(\theta_n)p_n \geq -c||p_n||^2
+\end{equation}
+and we can obtain a decrease in the objective.
+
+Substituting this into the previous expression,
+\begin{align}
+    f(\theta_\eta)-f(\theta_n) \leq -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2
+\end{align}
+
+and the Armijo condition then becomes
+\begin{align}
+    -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2 &\leq \eta \beta p_n^Tg(\theta_n) = - \eta \beta p_n^TH(\theta_n)p_n \\
+    \epsilon||p_n|| + \eta \Gamma ||p_n||^2 / 2 &\leq (1- \beta) p_n^TH(\theta_n)p_n \\
+    \epsilon + \eta \Gamma ||p_n|| / 2 &\leq c(1- \beta) ||p_n||
+\end{align}
+
+which holds and guarantees an iteration-independent lower bound if 
+\begin{equation}
+    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_n)|| \leq \frac{c(1-\beta)}{2}||p_n||
+\end{equation}
+
+Now, first note that
+\begin{equation}
+||g(\theta_n)|| - ||\nabla_\theta f(\theta_n)|| \leq ||g(\theta_n) - \nabla_\theta f(\theta_n)|| \leq \epsilon => ||\nabla_\theta f(\theta_n)|| \geq ||g(\theta_n)|| - \epsilon
+\end{equation} and
+\begin{equation}
+||\nabla_\theta f(\theta_n)||-||g(\theta_n)|| \leq ||\nabla_\theta f(\theta_n)-g(\theta_n)|| \leq \epsilon => ||g(\theta_n)|| \geq ||\nabla_\theta f(\theta_n)|| - \epsilon
+\end{equation}
+
+There are now two cases. If the algorithm terminates and $||g(\theta_n)|| \leq \sigma \epsilon$, we can derive
+
+\begin{equation}
+    ||\nabla_\theta f(\theta_n)|| \leq ||g(\theta_n)|| + \epsilon = \sigma\epsilon+\epsilon = (\sigma+1)\epsilon
+\end{equation}
+
+If the algorithm does not terminate, then $||g(\theta_n)|| > \sigma \epsilon$. Notice that
+
+\begin{align}
+    \epsilon \geq ||g(\theta_n) - \nabla_\theta f(\theta_n)|| &\geq ||g(\theta_n)|| - ||\nabla_\theta f(\theta_n)|| \\
+    ||\nabla_\theta f(\theta_n)|| + \epsilon &\geq ||g(\theta_n)|| \geq \sigma \epsilon \\
+    ||\nabla_\theta f(\theta_n)|| &\geq \sigma \epsilon - \epsilon = (\sigma - 1)\epsilon \\
+    \frac{||\nabla_\theta f(\theta_n)||}{\sigma-1} &\geq \epsilon
+\end{align}
+
+and now 
+\begin{align}
+    ||\nabla_\theta f(\theta_n)|| - \epsilon
+    &\geq  ||\nabla_\theta f(\theta_n)|| - \frac{||\nabla_\theta f(\theta_n)||}{\sigma-1} = \left(1-\frac{1}{\sigma-1}\right)||\nabla_\theta f(\theta_n|| = \frac{\sigma-2}{\sigma-1}||\nabla_\theta f(\theta_n)|| \geq \frac{2}{3}||\nabla_\theta f(\theta_n)||
+\end{align}
+
+From $||A^{-1}|| = 1/\sigma_{\min}(A)$,
+\begin{align}
+    p_n^TH(\theta_n)p_n &= (-(H(\theta_n))^{-1}g(\theta_n))^TH(\theta_n)(-(H(\theta_n))^{-1}g(\theta_n)) \\
+    &= g(\theta_n)^T(H(\theta_n))^{-1}g(\theta_n) \\
+    &\geq \frac{1}{c}||g(\theta_n)||^2 \\
+    &\geq \frac{1}{c}(||\nabla_\theta f(\theta_n)|| - \epsilon)^2 \\
+    &\geq \frac{4}{9c}||\nabla_\theta f(\theta_n)||^2
+\end{align}
+
+From the assumption that $f$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 -\ell \preceq \Gamma I$, by an implication of $\gamma$-strong convexity we have
+\begin{align}
+    f(\theta_n) - f(\theta^*) \leq \frac{1}{2\gamma}||\nabla_\theta f(\theta_n)||^2
+\end{align}
+and we put together:
+\begin{equation}
+    f(\theta_n) - f(\theta^*) \leq \frac{1}{2\gamma}||\nabla_\theta f(\theta_n)||^2 \leq \frac{9c}{4}\frac{1}{2\gamma}p_n^TH(\theta_n)p_n
+\end{equation}
+\begin{equation}
+    \frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \leq \frac{4}{9c}||\nabla_\theta f(\theta_n)||^2 \leq p_n^TH(\theta_n)p_n
+\end{equation}
+\begin{equation}
+    f(\theta_n) - f(\theta^*) \leq \frac{9c}{8\gamma}p_n^TH(\theta_n)p_n
+\end{equation}
+
+\begin{equation}
+    -\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \geq -\frac{4}{9c}||\nabla_\theta f(\theta_n)||^2 \geq -p_n^TH(\theta_n)p_n
+\end{equation}
+
+But from earlier, as the Armijo condition is fulfilled with our choice of $\eta$ and $\epsilon$,
+\begin{align}
+    f(\theta_{n+1})-f(\theta_n) &\leq -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2 \\
+    &\leq -\eta\beta p_n^TH(\theta_n)p_n \\
+    &\leq -\eta\beta\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) 
+\end{align}
+
+and therefore
+\begin{align}
+    f(\theta_{n+1}) - f(\theta^*) 
+    &= f(\theta_{n+1})-f(\theta_n)+f(\theta_n)- f(\theta^*) \\
+    &\leq f(\theta_n)- f(\theta^*) -\eta\beta\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \\
+    &= (1-\eta\beta\frac{8\gamma}{9c})(f(\theta_n) - f(\theta^*)) 
+\end{align}
+
+\end{proof}
+
 \end{document}

### 2023-12-28 mop/draft.tex
Updates from Overleaf

@@
 
-\begin{figure}
+\begin{figure}[h]
     \centering
-    \includegraphics{imgs/095/qq.png}
+    \includegraphics[scale=0.53]{imgs/095/pairs.png}
+    \includegraphics[scale=0.53]{imgs/095/qq.png}
+    \caption{Scatterplots depicting the performance of IFAD against that of IF2.
+Left: Paired searches from the same starting point.
+Right: Q-Q plot of ranked IFAD searches against ranked IF2 searches.
+It is clear that on average, IFAD has the edge, and manages to find the MLE while no IF2 search successfully gets within 5 log-likelihood units of it.
+}
+    \label{fig:scatter}
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics{imgs/095/boxplot.png}
+    \caption{Caption}
+    \label{fig:enter-label}
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics{imgs/095/optim.png}
+    \caption{Caption}
+    \label{fig:enter-label}
+\end{figure}
+
+\begin{figure}[u]
+    \centering
+    \includegraphics{imgs/095/hist.png}
     \caption{Caption}

### 2023-12-28 mop/draft.tex
Updates from Overleaf

@@
 It is clear that on average, IFAD has the edge, and manages to find the MLE while no IF2 search successfully gets within 5 log-likelihood units of it.
+We use a dotted red line to display the MLE in both.
 }
@@
     \centering
-    \includegraphics{imgs/095/boxplot.png}
-    \caption{Caption}
-    \label{fig:enter-label}
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics{imgs/095/optim.png}
-    \caption{Caption}
-    \label{fig:enter-label}
+    \includegraphics[scale=0.52]{imgs/095/boxplot.png}
+    \includegraphics[scale=0.54]{imgs/095/optim.png}
+    \caption{\textbf{Left:} Boxplot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs.
+This simulates the result of the common procedure of running a few searches and choosing the best one.
+IFAD outperforms all other methods, and improves on the warm-start given by running 40 IF2 iterations.
+Our choice of hyperparameters for our implementation of IF2 outperforms the choices made by \cite{ionides15}, which we depict in the boxplot labeled "IF2 2015".
+    \textbf{Right:} Optimization progress of IFAD and IF2, where the solid lines depict the median negative log-likelihood at each iteration and the shaded area depicts the 20\% and 80\% percentiles.
+The dashed orange line depicts the median warm-start given by running 40 IF2 iterations.
+While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms
+    In both, as before, we use a dotted red line to display the MLE. }
+    \label{fig:boxplot-search}
 \end{figure}
@@
     \centering
-    \includegraphics{imgs/095/hist.png}
-    \caption{Caption}
-    \label{fig:enter-label}
+    \includegraphics[scale=0.7]{imgs/095/hist.png}
+    \caption{Histogram comparison of the performance of IFAD, IF2, and MOP alone. As before, we note that IFAD Performing MOP alone }
+    \label{fig:hist-all}
 \end{figure}

### 2023-12-28 mop/draft.tex

@@
 Our choice of hyperparameters for our implementation of IF2 outperforms the choices made by \cite{ionides15}, which we depict in the boxplot labeled "IF2 2015".
-    \textbf{Right:} Optimization progress of IFAD and IF2, where the solid lines depict the median negative log-likelihood at each iteration and the shaded area depicts the 20\% and 80\% percentiles.
+    \textbf{Right:} Optimization progress of IFAD and IF2, where the solid lines depict the median negative log-likelihood at each iteration and the shaded area depicts the best search at any iteration and the 80\% percentile.
 The dashed orange line depicts the median warm-start given by running 40 IF2 iterations.
-While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms
+While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD -- IFAD has better tail control and successfully reaches the MLE.
+The initial loss in log-likelihood upon transitioning from IF2 to MOP seems to help with escaping local minima.
     In both, as before, we use a dotted red line to display the MLE. }
@@
     \includegraphics[scale=0.7]{imgs/095/hist.png}
-    \caption{Histogram comparison of the performance of IFAD, IF2, and MOP alone. As before, we note that IFAD Performing MOP alone }
+    \caption{Histogram comparison of the performance of IFAD, IF2, and MOP alone.
+As before, we note that IFAD successfully reaches the MLE a nontrivial number of times, while IF2 ultimately fails to do so.
+Our implementation of IF2 outperforms that of \cite{ionides15}, largely due to improved hyperparameter choices.
+MOP alone performs poorly, demonstrating the necessity of IF2.
+}
     \label{fig:hist-all}

### 2023-12-29 mop/draft.tex
some edits to definition of targeting

@@
 \usepackage{tikz}
+\usetikzlibrary{arrows.meta}
+\usetikzlibrary{positioning}
+
 \setlist{nosep} % to reduce space in lists
@@
 
-\usetikzlibrary{arrows.meta,positioning}
 \begin{figure}[h]
@@
 int/.style={draw, minimum size=1.6cm}]
+  \node [int] (H) at (0,0) {$H$};
 \node [int] (S) [right=of H] {$S$};
@@
     (R2) edge node {$\delta$}  (deathsR2)
-    (R3) edge node {$\delta$}  (deathsR3)
+    (R3) edge node {$\delta$}  (deathsR3);
 \end{tikzpicture}
@@
     \centering
-    \includegraphics[scale=0.53]{imgs/095/pairs.png}
-    \includegraphics[scale=0.53]{imgs/095/qq.png}
+    \includegraphics[scale=0.53]{../imgs/095/pairs.png}
+    \includegraphics[scale=0.53]{../imgs/095/qq.png}
     \caption{Scatterplots depicting the performance of IFAD against that of IF2.
@@
     \centering
-    \includegraphics[scale=0.52]{imgs/095/boxplot.png}
-    \includegraphics[scale=0.54]{imgs/095/optim.png}
+    \includegraphics[scale=0.52]{../imgs/095/boxplot.png}
+    \includegraphics[scale=0.54]{../imgs/095/optim.png}
     \caption{\textbf{Left:} Boxplot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs.
@@
 
-\begin{figure}[u]
+\begin{figure}[h]
     \centering
-    \includegraphics[scale=0.7]{imgs/095/hist.png}
+    \includegraphics[scale=0.7]{../imgs/095/hist.png}
     \caption{Histogram comparison of the performance of IFAD, IF2, and MOP alone.
@@
 
-\begin{defn}[Targeting, \cite{liu1998smc}]
+\begin{defn}[Targeting]
     A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any integrable function $h$,
-    $$
+\begin{equation}
     E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
-    $$
-
-    A set of particles $(X_j, w_j), j=1,2, \ldots$, \textbf{targets} $\pi$ if
-    $$
-    \frac{\sum_{j=1}^J h(X_j) w_j}{\sum_{j=1}^J w_j} \stackrel{a.s.}{\to} E_\pi(h(X))
-    $$
-    for any integrable function $h$ as $J \to \infty$, in the sense of the triangular array convergence in \cite{chopin2004clt}. 
+\end{equation}  
+    A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if
+\begin{equation}
+    \frac{\sum_{j=1}^J h(X^J_j) w^J_j}{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X))
+\end{equation}
+for any integrable function $h$ as $J \to \infty$.
 \end{defn}
-
-It is known that the bootstrap and appropriately weighted particle filters target the posterior $f_{X_{1:N}|Y_{1:N}}$ in this sense, as long as they adhere to the procedure in \cite{chopin2004clt}.
-We now present three helper lemmas.
+\citet{chopin2004clt} asserted without proof that common particle filter algorithms target the filtering distribution, $f_{X_{1:N}|Y_{1:N}}$, in this sense.
+\ed{PERHAPS IF WE HAD A CONCRETE PROOF OF THIS, THAT WOULD HELP TO SEE HOW IT EXTENDS TO OUR SUBSEQUENT LEMMAS.
+ARGUABLY, THAT IS IN FACT WHAT WE ARE DEVELOPING VIA THESE LEMMAS.
+I WONDER IF SOME OTHER REFERENCE, E.G.
+DELMORAL'S 2004 BOOK, HAS AN EXPLICIT PROOF OF CONSISTENCY FOR SOME PARTICLE FILTER ALGORITHM.}
+In order to prove the consistency of our variation on the particle filter, we now present three helper lemmas.
 The first follows from standard importance sampling arguments, the second from integrating out the marginal, and the third from Bayes' theorem.
@@
     \label{lem:lik-proper-weight}
-  $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1_n-1}^*;\theta)$ is properly estimated by either the before-resampling estimate,
+  $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1_n-1}^*;\theta)$ is consistently estimated by either the before-resampling estimate,
 \begin{equation}\label{L1}

### 2023-12-29 mop/draft.tex
Updates from Overleaf

--- 2023-12-28 mop/draft.tex
@@
 
-Recent advances in automatic differentiation (AD) for particle filters (\cite{blei2018vsmc, jon2018diffpf, corenflos2021ot, scibior2021dpf, doucet2022particlebased}) have drawn attention to AD as an alternative tool for inference in partially observed stochastic processes with the particle filter.
+Recent advances in automatic differentiation (AD) for particle filters (\cite{blei2018vsmc, jon2018diffpf, corenflos21, scibior2021dpf, doucet2022particlebased}) have drawn attention to AD as an alternative tool for inference in partially observed stochastic processes with the particle filter.
 However, these existing approaches either are not yet compatible with the desire for simulation-based inference, or are computationally expensive.
 
-\citet{blei2018vsmc} and \citet{jon2018diffpf} drop high-variance resampling terms from the gradient, leading to bias that does not vanish asymptotically according to \citet{corenflos2021ot}.
-Though the approach of \citet{corenflos2021ot} is promising, it is computationally expensive.
+\citet{blei2018vsmc} and \citet{jon2018diffpf} drop high-variance resampling terms from the gradient, leading to bias that does not vanish asymptotically according to \citet{corenflos21}.
+Though the approach of \citet{corenflos21} is promising, it is computationally expensive.
 \citet{doucet2022particlebased} provide another promising fixed-lag smoothing-based approach that is similar in spirit to our method (MOP-$\alpha$), but requires either (1) access to the process model transition densities or (2) the special case where the transition density factors into a policy that selects actions (in the reinforcement learning sense) and a deterministic model that maps states and actions to next-timestep states.
@@
 In practice, we propose a hybrid algorithm that we call \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts gradient ascent (or a similar iterative first or second-order gradient method) with a coarse solution obtained from a few rounds of iterated filtering.
-While this requires a differentiable simulator compatible with the reparametrization trick (as previous literature such as \citet{corenflos2021ot} also require), future work will deal with workarounds inspired by our framework that use likelihood ratios of transition densities in place of this.
+While this requires a differentiable simulator compatible with the reparametrization trick (as previous literature such as \citet{corenflos21} also require), future work will deal with workarounds inspired by our framework that use likelihood ratios of transition densities in place of this.
 Promising numerical results indicate that IFAD beats IF2 (and by the transitive property, IF1, the Liu-West filter, and particle MCMC) on a challenging scientific benchmark problem.
@@
 
-Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king2016pomp, king2017pompmanual} that we call \texttt{pompy}.
+Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual} that we call \texttt{pompy}.
 The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{jax} Python package from \citet{jax2018github}.
@@
 Our new method outperforms previous approaches on a challenging problem in epidemiology.
-This is implemented in a precursor to a Python counterpart to the popular \texttt{pomp} package by \citet{king2016pomp, king2017pompmanual} that includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation.
+This is implemented in a precursor to a Python counterpart to the popular \texttt{pomp} package by \citet{king16, king2017pompmanual} that includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation.
 
@@
 
-\textbf{Intuition:} Our theoretical framework for automatic differentiation for the particle filter hinges on the following observation.
+%\paragraph{Motivation:} 
+%The gradient of the log-likelihood can be written as a vector $\nabla_\theta \ell(\theta)$ such that 
+%\begin{equation}
+    %\lim_{v \to 0} \frac{|\ell(\theta+v) - \ell(\theta) - v^T \nabla_\theta \ell(\theta)|}{\|v\|} = 0.
+%\end{equation}
+
+%We therefore need to evaluate the likelihood at a point $\theta$ and a nearby point $\theta + v$ to estimate the gradient with finite differences.
+But in the presence of Monte Carlo noise in evaluating the likelihood, this is difficult, unless (1) one ensures one uses the same seed for each evaluation in the neighborhood, and (2) chooses a step-size small enough that changing the the parameter does not change the resampling indices.
+
+\paragraph{Intuition:} 
+Our theoretical framework for automatic differentiation for the particle filter hinges on the following observation.
 Instead of seeking to differentiate $\ell(\theta)$ directly, we instead focus on differentiating through a suitable fixed-seed reweighting of a bootstrap filter.
@@
 
-Differentiating this reweighted likelihood estimate with respect to $\theta$ then amounts to differentiating through the reweighting scheme, which is differentiable if the densities used for the reweighting are differentiable.
+Differentiating this reweighted likelihood estimate with respect to $\theta$ then amounts to differentiating through the reweighting likelihood ratios, which is differentiable if the densities used for the reweighting are differentiable.
 We call this simulation-and-reweighting scheme an \textbf{off-policy} likelihood estimate, as it is similar in spirit to the off-policy methods found in the reinforcement learning literature.
@@
 The measurement correction cannot be avoided, but one can bypass the process correction -- with the reparametrization trick.
-We can do when we assume the simulator is differentiable with the reparametrization trick, as \cite{corenflos2021ot} also assume.
+We can do when we assume the simulator is differentiable with the reparametrization trick, as \cite{corenflos21} also assume.
 Doing so then yields an algorithm that only requires a simulator, and does not require the process densities.
@@
 Consider the case where you have access to a differentiable simulator compatible with the reparametrization trick.
-If we have the state transitions evolve according to $\theta,$ but have the resampling according to $\phi,$ we can offload the derivative with respect to the process model to the reparametrization trick, as in \cite{corenflos2021ot}.
+If we have the state transitions evolve according to $\theta,$ but have the resampling according to $\phi,$ we can offload the derivative with respect to the process model to the reparametrization trick, as in \cite{corenflos21}.
 This corresponds to the MOP-$\alpha$ procedure in Algorithm \ref{alg:mop}.
@@
 This is done by either (1) differentiating through smooth likelihood ratios, or (2) differentiating through a differentiable simulator with the reparametrization trick (and through smooth likelihood ratios to adjust for the resampling).
+In particular, MOP-$\alpha$ does not require access to the transition densities, and like the bootstrap filter, only requires access to a (albeit differentiable) simulator.
+To our knowledge, this is the first particle filter gradient method in the literature with this capability.
 
@@
 This still leaves us with the question of designing a reasonably effective procedure for likelihood maximization with these new tools.
-We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that amounts to warm-starting gradient descent or some other reasonable first or second order procedure with the output of an initial search of IF2.
+We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that amounts to warm-starting gradient descent or some other reasonable first or second order procedure (using the gradient estimate given by MOP-$\alpha$) with the output of an initial search of IF2.
 
@@
 Even with the increased computational cost of automatic differentiation, Algorithm \ref{alg:ifad} is reasonably fast, as by the cheap gradient principle from \cite{kakade2019provably}, a gradient operation takes no more than 6 times the runtime of the particle filter.
+As such, unlike methods like the marginal differentiable particle filter in \cite{scibior2021dpf} or the the optimal transport of resampling approach of \cite{corenflos21}, our algorithm has runtime \textbf{linear, and not quadratic}, in the number of particles (while alleviating the $O(N^2)$ variance concern of \cite{poyiadjis11} and \cite{corenflos2021ot} and asymptotic bias concern of \cite{blei2018vsmc} through optimizing $\alpha$ for a bias-variance tradeoff).
 
@@
 This can be attributed to both GPU acceleration and the just-in-time (JIT) compilation functionality that \texttt{JAX} provides.
-When tested on a computer with an Intel i9-13900K processor, 128GB of RAM, and a NVIDIA RTX3090 GPU, each post-compilation particle filtering operation on the Dacca cholera model from \cite{king08} with 10000 particles took roughly 379 milliseconds (with a standard deviation of 279 microseconds over seven runs), a considerable improvement over the existing performance in the widely used \texttt{pomp} package of \cite{king2016pomp}, which took between 6.19 to 6.39 seconds over ten runs on the same machine.
+When tested on a computer with an Intel i9-13900K processor, 128GB of RAM, and a NVIDIA RTX3090 GPU, each post-compilation particle filtering operation on the Dacca cholera model from \cite{king08} with 10000 particles took roughly 379 milliseconds (with a standard deviation of 279 microseconds over seven runs), a considerable improvement over the existing performance in the widely used \texttt{pomp} package of \cite{king16}, which took between 6.19 to 6.39 seconds over ten runs on the same machine.
 
@@
 
-\begin{figure}[h]
+% The 60 more iterations of IF2 yielded largely not a huge improvement over the initial coarse solution of 40 iterations.
+This demonstrates what we have mentioned earlier: that although IF2 manages to quickly find a decent coarse solution, it struggles to squeeze out those last few units of log-likelihood improvement.
+The 60 more iterations of gradient descent, however, successfully refines this solution, getting much closer to the MLE on average.
+
+\subsubsection{Results} 
+We benchmarked IFAD against IF2 on a challenging global search problem.
+100 initial starting parameter vectors were drawn uniformly from the same wide bounding box used in \cite{ionides15}, and we performed 100 searches each of IF2, IFAD-0, IFAD-1, MOP-0, and MOP-1 initialized from these 100 starting parameter vectors.
+We summarize our findings below.
+
+\begin{table}[h!]
+\centering
+\begin{tabular}{||c c c||} 
+ \hline
+ Method & Log-Likelihood & Rank \\ [0.5ex] 
+ \hline\hline
+ IF2 2015 & -3768.63 & 5\\ 
+ IF2 (Ours) & -3755.91 & 3\\
+ IF2 (Warm Start) & -3759.60 & 4 \\
+ IFAD-1 & -3748.53 & 2\\
+ IFAD-0 & -3747.87 & 1\\
+ MOP-1 Alone & -3797.38 & 7\\
+ MOP-0 Alone & -3795.74 & 6\\
+ \hline
+\end{tabular}
+\caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone.
+IFAD performs the best among all methods.
+Our implementation of IF2 outperforms that of \cite{ionides15}, but still ultimately underperforms IFAD.
+Both IFAD-0 and IFAD-1 manage to find the MLE, either matching or beating the highest log-likelihood previously found in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.}
+\label{table:mle}
+\end{table}
+
+\paragraph{IFAD Successfully Finds the MLE:} Both IFAD-0 and IFAD-1 successfully find the MLE, as seen in Table \ref{table:mle}.
+Previously, the MLE at a log-likelihood of $-3748.5$ was found with much effort, using many global and local IF2 searches and likelihood profiling.
+Despite being initialized for a global search, both IFAD-0 and IFAD-1 manage to find the MLE over the 100 searches.
+This may imply that the local search and refinement that was previously necessary for finding the MLE is \textbf{baked into} the two-stage IFAD algorithm.
+
+Interestingly, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.63$, while \cite{king08} only achieve a best possible estimate of $-3793.4$.
+Our results of $-3748.53$ and $-3747.87$ match or beat the highest log-likelihood previously found in the current implementation in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.
+
+This can be seen most clearly in Figure \ref{fig:boxplot-search}.
+In the left panel within Figure \ref{fig:boxplot-search}, only IFAD has whiskers that extend to the dotted red line that represents the MLE.
+
+\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations (as seen in Figure \ref{fig:optim}), performing IF2 alone ultimately fails to squeeze out the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter} and the left panel in Figure \ref{fig:boxplot-search}).
+
+Conversely, performing MOP alone leads to many, many failed searches, as seen in Figure \ref{fig:hist-all} and the right panel in Figure \ref{fig:boxplot-search}.
+This is a difficult, nonconvex, and noisy problem, and the search gets stuck in local minima and saddle points, failing to approach the MLE.
+
+IFAD, in comparison, successfully manages to (1) approach the MLE quickly due to the IF2 warm-start (as seen in Figure \ref{fig:optim}) and (2) refine the coarse solution found by the warm-start with MOP gradient steps to find the MLE (as seen in Figures \ref{fig:scatter}, \ref{fig:boxplot-search}, and \ref{fig:hist-all}).
+IFAD therefore successfully combines the best qualities of IFAD and MOP, outperforming either of them alone.
+
+\begin{figure}[htbp!]
     \centering
@@
     \caption{Scatterplots depicting the performance of IFAD against that of IF2.
-Left: Paired searches from the same starting point.
-Right: Q-Q plot of ranked IFAD searches against ranked IF2 searches.
-It is clear that on average, IFAD has the edge, and manages to find the MLE while no IF2 search successfully gets within 5 log-likelihood units of it.
+\textbf{Left:} Paired searches from the same starting point.
+\textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches.
+It is clear that on average, IFAD has the edge, and manages to find the MLE while no IF2 search successfully gets within 7 log-likelihood units of it.
 We use a dotted red line to display the MLE in both.
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
     \centering
-    \includegraphics[scale=0.52]{imgs/095/boxplot.png}
-    \includegraphics[scale=0.54]{imgs/095/optim.png}
+    \includegraphics[scale=0.48]{imgs/095/boxplot.png}
+    \includegraphics[scale=0.53]{imgs/095/boxplotall.png}
     \caption{\textbf{Left:} Boxplot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs.
@@
 Our choice of hyperparameters for our implementation of IF2 outperforms the choices made by \cite{ionides15}, which we depict in the boxplot labeled "IF2 2015".
-    \textbf{Right:} Optimization progress of IFAD and IF2, where the solid lines depict the median negative log-likelihood at each iteration and the shaded area depicts the best search at any iteration and the 80\% percentile.
+As before, while IF2 successfully approaches the MLE, it ultimately fails to find it.
+    \textbf{Right:} Boxplot depicting the performance of IFAD, IF2, and MOP alone without the warm-start, where we plot all runs.
+We see that MOP alone drastically underperforms all other methods, failing to get close to the MLE.
+This shows that the IF2 warm-start may be necessary to achieve good performance in challenging nonconvex and noisy problems like this one.
+    \textbf{Takeaway:} In this panel, we see that (1) IF2 fails to find the MLE while the gradient steps perform better at fine-grained refinement, and (2) performing gradient steps alone without a warm-start leads to the search getting stuck in local minima and saddle points, ultimately demonstrating the necessity of a hybrid algorithm like IFAD.
+    In both, as before, we use a dotted red line to display the MLE. }
+    \label{fig:boxplot-search}
+\end{figure}
+
+\begin{figure}[htbp!]
+    \centering
+    \includegraphics[scale=0.7]{imgs/095/optim.png}
+    \caption{Optimization progress of IFAD and IF2, where the solid lines depict the median negative log-likelihood at each iteration and the shaded area depicts the best search at any iteration and the 80\% percentile.
 The dashed orange line depicts the median warm-start given by running 40 IF2 iterations.
@@
 The initial loss in log-likelihood upon transitioning from IF2 to MOP seems to help with escaping local minima.
-    In both, as before, we use a dotted red line to display the MLE. }
-    \label{fig:boxplot-search}
+    As before, we use a dotted red line to display the MLE. }
+    \label{fig:optim}
 \end{figure}
 
-\begin{figure}[u]
+\begin{figure}[u!]
     \centering
@@
 Our implementation of IF2 outperforms that of \cite{ionides15}, largely due to improved hyperparameter choices.
-MOP alone performs poorly, demonstrating the necessity of IF2.
+MOP alone performs poorly, demonstrating the necessity of warm-starting the search with IF2.
+    We once again see that IFAD combines the best of IF2 and MOP, approaching the MLE quickly due to the IF2 warm-start and successfully refining the coarse solution with gradient steps, achieving superior performance.
+As before, we use a dotted red line to display the MLE.
 }
@@
 
-% The 60 more iterations of IF2 yielded largely not a huge improvement over the initial coarse solution of 40 iterations.
-This demonstrates what we have mentioned earlier: that although IF2 manages to quickly find a decent coarse solution, it struggles to squeeze out those last few units of log-likelihood improvement.
-The 60 more iterations of gradient descent, however, successfully refines this solution, getting much closer to the MLE on average.
-
-\kevin{Do groups of ten and take the best one from each group.
-Omit old IF2 from figure in the paper, provide complete figure with old IF2 in the appendix, saying that we modified the settings because of more modern practice.
-}
+\paragraph{Differences Between IFAD-0 and IFAD-1:} The differences between IFAD-1 and IFAD-0 (MOP-1 and MOP-0) do not seem very pronounced in this example.
+This could be due to a number of reasons, such as (1) future case counts $y_{{n+1}:N}^*$ not providing a large amount of information on state identification $x_n$ given current and past measurements $y_{0:n}^*$ as mentioned in \cite{corenflos21}, and (2) the gradient normalization making the trajectories given by MOP-0 and MOP-1 more similar than they would be otherwise.
 
@@
 
-\kevin{IFAD vs IF2, AD vs IF2, IFAD0 vs IFAD1, AD0 vs AD1 in Dacca, local and global plots, convergence line plot of IFAD vs IF2, Dacca data plots, toy model from IF2 paper, AD0 vs AD1 gradient histograms, MSE of AD0 and AD1 gradient estimates as J increases on analytically tractable toy model from IF2 paper or OU2 in pomp}
-
-\kevin{Main plots are the local and global 2-way comparison plots of IFAD1 vs IF2, AD1 vs IF2.
-Visualize the Q-Q plots in the appendix, and use alternate representations as boxplots (or histograms) with dashed line for the maximum in the main paper.}
-
-\citet{liu2023bridging} have established that the straight-through estimator can be thought of as a first-order approximation of the gradient.
-\kevin{Benchmark DOP vs straight-through estimator/reinmax in MOP}
-
 \kevin{Have an outsider read the manuscript afterwards.}
+
+\section{Discussion and Future Work}
+
+We develop a new theoretical framework to differentiating through the particle filter that (1) encompasses the gradient estimates of \cite{poyiadjis11}, \cite{blei2018vsmc}, and \cite{scibior2021dpf} as special cases, (2) sidesteps the issue of differentiating through a Monte-Carlo algorithm with discontinuous resampling, and (3) provides an opportunity for optimizing a bias-variance tradeoff.
+
+Prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics.
+In the presence of a differentiable simulator, MOP-$\alpha$ enjoys this very same desirable property, enabling practical maximum likelihood inference in a wider variety of situations.
+
+Furthermore, MOP-$\alpha$, and by extension IFAD, has runtime \textbf{linear, and not quadratic}, in the number of particles, unlike the marginal differentiable particle filter in \cite{scibior2021dpf} or the the optimal transport of resampling approach of \cite{corenflos21}.
+This does not necessarily come at a cost to Monte Carlo variance, unlike \cite{poyiadjis11} and \cite{corenflos2021ot} who suffer from variance quadratic in the length of the trajectory, as we can optimize $\alpha$ for a bias-variance tradeoff.
+
+In the presence of strong convexity, first and second-order gradient methods that use MOP-$\alpha$ as an estimate enjoy a linear convergence rate, providing support for numerical optimization with AD for the particle filter.
+
+This leads to a practical hybrid algorithm we call IFAD that warm-starts gradient descent, or some other first or second-order method using the gradient estimate given by MOP-$\alpha$, with IF2.
+When benchmarked on a challenging global search problem with the model of \cite{king08}, IFAD enjoys the best properties of IF2 and MOP-$\alpha$, successfully approaching a neighborhood of the MLE quickly without getting stuck in saddle points and (poor) local minima (like gradient descent alone would), while effectively optimizing the last few units of log-likelihood (which IF2 struggles with).
+
+Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual} that we call \texttt{pompy}.
+The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{jax} Python package from \citet{jax2018github}.
+
+\subsection{Future Work}
+
+While MOP-$\alpha$, and IFAD in its current incarnation, remain very general, they currently lack the capability of handling discrete latent states.
+We conjecture that DOP-$\alpha$ might be a possible solution (despite needing the likelihood ratio of transition densities), and will explore this in future work.
+
+We conjecture that the linear convergence result in Theorem \ref{thm:convergence} applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE.
+IF2 posesses some SGD-like qualities, including what looks like linear convergence to a ball around the MLE with width determined by the random walk standard deviation.
+Not much is known about the convergence rate of IF2, and an answer to this would also provide an answer to the convergence rate of IFAD beyond just the convergence rate of the gradient stage.
+We will explore this conjecture in future work.
+
+In the interest of simplicity, we only explore the performance of gradient descent on the Dacca cholera model of \cite{king08}.
+Exploring the use of various first and second order methods such as Newton's method, ADAM, L-BFGS, or other popular iterative optimization methods with the gradient estimates that MOP-$\alpha$ provides in the context of the particle filter could very well lead to nontrivial speedups and increased performance.
+
+Extensions of MOP-$\alpha$ and IFAD to panel and spatiotemporal datam, possibly through a block particle filter in the same vein as \cite{ionides22} and \cite{ning23}, would go a long way towards extending the current applicability of these methods to handle a wider variety of interesting and challenging scientific problems.
+
+While we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual}, this remains a work in progress.
 

### 2023-12-29 mop/draft.tex

@@
 }
+
     \label{fig:scatter}
@@
     \textbf{Takeaway:} In this panel, we see that (1) IF2 fails to find the MLE while the gradient steps perform better at fine-grained refinement, and (2) performing gradient steps alone without a warm-start leads to the search getting stuck in local minima and saddle points, ultimately demonstrating the necessity of a hybrid algorithm like IFAD.
+
     In both, as before, we use a dotted red line to display the MLE. }

### 2023-12-30 optim.py
update

@@ -32,24 +32,24 @@ MONITORS = 1
 
 
 
-#@partial(jit, static_argnums=2)
+@partial(jit, static_argnums=2)
 def jgrad_pf(theta_ests, ys, J, covars, thresh, key=None):
     return jax.grad(pfilter_pf)(theta_ests, ys, J, covars=covars, thresh=thresh, key=key)
 
 
-#@partial(jit, static_argnums=2)
+@partial(jit, static_argnums=2)
 def jgrad(theta_ests, ys, J, covars, thresh, key=None):
     return jax.grad(pfilter_mean)(theta_ests, ys, J, covars=covars, thresh=thresh, key=key)
 
-#@partial(jit, static_argnums=2)
+@partial(jit, static_argnums=2)
 def jvg(theta_ests, ys, J, covars, thresh, key=None):
     return jax.value_and_grad(pfilter_mean)(theta_ests, ys, J, covars=covars, thresh=thresh, key=key)
 
-#@partial(jit, static_argnums=2)
+@partial(jit, static_argnums=2)
 def jhess(theta_ests, ys, J, covars, thresh, key=None):
     return jax.hessian(pfilter_mean)(theta_ests, ys, J, covars=covars, thresh=thresh, key=key)
 
-#@partial(jit, static_argnums=2)
+@partial(jit, static_argnums=2)
 def jpgrad(thetas, ys, J, sigmas, covars, a, thresh, key=None):
     return jax.grad(perfilter_mean, has_aux=True)(
         thetas, ys, J, sigmas, covars=covars, a=a,thresh=thresh, key=key)

### 2023-12-30 mop/draft.tex

@@
 
-We first analyze the performance of gradient descent, Newton's method, or any iterative method that multiplies the particle estimate of the gradient with the inverse of a matrix with minimum eigenvalue bounded below, in the case where the log-likelihood surface is $\gamma$-strongly convex.
+We first analyze the performance of gradient descent, Newton's method, or any iterative method that multiplies the particle estimate of the gradient with the inverse of a matrix with minimum eigenvalue bounded below, in the case where the log-likelihood surface is $\gamma$-strongly convex, and where the gradient estimate is a particle approximation of the score (for example, IFAD-1 and MOP-1).
 In this case, these methods enjoy a linear convergence guarantee:
@@
 
-Consider a variant of Algorithm \ref{alg:ifad} where one stops if the gradient estimate $||g(\theta_n)|| \leq \sigma \epsilon$.
+Consider a variant of IFAD-1 (Algorithm \ref{alg:ifad}) where one stops if the gradient estimate $||g(\theta_n)|| \leq \sigma \epsilon$.
 Assume $-\ell$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$.
@@
 
-Choose $J$ according to the maximum of the $J$ prescribed in Lemmas \ref{lemma:grad_bound} and $\ref{lemma:hess_bound}$ with failure probability $\delta/2$ each.
+Choose $J$ according to the maximum of the $J$ prescribed in Lemmas \ref{lemma:grad_bound} and (if using a Hessian estimate) $\ref{lemma:hess_bound}$ with failure probability $\delta/2$ each.
 Let $H(\theta_n)$ be a matrix possibly dependent on $\theta_n$ with minimum eigenvalue always greater than some $c>0$.
@@
 
-The gradient stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the IF2 stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
+The gradient stage of IFAD-1 converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the IF2 stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
 This happens fairly often in practice, for example, when sufficient regularity conditions for local asymptotic normality of the MLE hold.
@@
 
+While this theoretical guarantee currently only holds for IFAD-1, as it forms a particle approximation of the score as in \cite{poyiadjis11}, one can also extend this for a similar guarantee if the bias for the gradient estimate given by MOP-$\alpha$ for $\alpha < 1$ is small enough.
+For general $\alpha$, including MOP-0, the analysis then becomes more complicated, as one needs to handle the case of biased gradient descent.
+However, if future measurements $y_{n+1:N}^*$ do not provide a large amount of information on the identification of the current state $x_n$ given past and current measurements $y_{0:n}^*$, as mentioned in \cite{corenflos21}, the bias may be small enough for it to not present too much of a problem.
+
 \subsection{Runtime}
 
-Even with the increased computational cost of automatic differentiation, Algorithm \ref{alg:ifad} is reasonably fast, as by the cheap gradient principle from \cite{kakade2019provably}, a gradient operation takes no more than 6 times the runtime of the particle filter.
+Even with the increased computational cost of automatic differentiation, Algorithm \ref{alg:ifad} is reasonably fast, as by the cheap gradient principle from \cite{kakade2019provably}, a gradient operation takes no more than 6 times the runtime of the particle filter (in practice, this means a gradient step takes about 3.75x as much time as an IF2 iteration).
 As such, unlike methods like the marginal differentiable particle filter in \cite{scibior2021dpf} or the the optimal transport of resampling approach of \cite{corenflos21}, our algorithm has runtime \textbf{linear, and not quadratic}, in the number of particles (while alleviating the $O(N^2)$ variance concern of \cite{poyiadjis11} and \cite{corenflos2021ot} and asymptotic bias concern of \cite{blei2018vsmc} through optimizing $\alpha$ for a bias-variance tradeoff).

### 2024-01-05 mop/draft.tex
Updates from Overleaf

@@
 
-We show these off-policy particle filters are properly weighted, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties.
+We show these off-policy particle filters successfully target the posterior, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties.
 To cap off the theoretical results, we also derive a linear convergence rate for gradient methods on the particle filter in the presence of strong convexity, providing support for numerical optimization with AD for the particle filter.
@@
 
-We call $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ the process model, writing $\operatorname{process}\left(x_n, \theta\right)$ for the simulator corresponding to the above process model, $f_{Y_n|X_n}\left(y_n \mid x_n, \theta\right)$ the measurement model, and will write $y_n^*$ for the actual values of the observations that were observed.
+We call $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ the process model, writing $\texttt{process}\left(x_n, \theta\right)$ for the simulator corresponding to the above process model, $f_{Y_n|X_n}\left(y_n \mid x_n, \theta\right)$ the measurement model, and will write $y_n^*$ for the actual values of the observations that were observed.
 
@@
 \mathcal{L}(\theta) & =\int_{\Omega} \mathcal{L}(\theta, \omega, J) d \mu(\omega) \\
-& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right)\right) f_{X_{0:N}, Y_{1:N}}\left(x_{0: N}^P(\omega) \mid y_{1: N}^*, \theta\right) d \mu(\omega) \\
-& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right) f_{X_{0:N}, Y_{1:N}}\left(x_{n, j}^P(\omega) \mid y_n^*, \theta\right)\right) d \mu(\omega)
+& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right)\right) f_{X_{0:N}, Y_{1:N}}\left(x_{0: N}^P(\omega) , y_{1: N}^*, \theta\right) d \mu(\omega) \\
+& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right) f_{X_n|X_{n-1}}(x_{n,j}^P(\omega)|x_{n-1,j}^F(\omega), \theta)\right) f_{X_{0:N}| Y_{1:N}}\left(x_{n, j}^P(\omega) \mid y_n^*, \theta\right) d \mu(\omega)
 \end{aligned}
@@
 \begin{align}
-    \lik(\theta,\phi,\omega, J) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J &\left(f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) f_{X_n|Y_n}(x_{n,j}^{P, \phi}(\omega) | y_n^*, \phi) \right.\\
+    \lik(\theta,\phi,\omega, J) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J &\left(f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) \cdot 
+    f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}(\omega) | x_{n-1,j}^{F,\phi}, \phi) \right.\\
     &\cdot
@@
 \begin{aspt}[Differentiable Simulator]
-    $\text{process}(\cdot | X_{t-1}; \theta)$ is differentiable with respect to $\theta,$ and is compatible with the reparametrization trick. 
+    $\texttt{process}(\cdot | X_{t-1}; \theta)$ is differentiable with respect to $\theta,$ and is compatible with the reparametrization trick. 
 \end{aspt}
@@
     \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
+        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
     \end{equation}
@@
 The initial "convergence" happens fairly quickly in practice.
-In the case of the Dacca cholera model of \cite{king08}, when a geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations of IF2.
+In the case of the Dhaka cholera model of \cite{king08}, when a geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations of IF2.
 In comparison, often 100 or 200 iterations of IF2 are used for an initial global search.
-\cite{ionides15}, for example, use 100 iterations with the Dacca cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{Lee_haiticholera}.
+\cite{ionides15}, for example, use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{Lee_haiticholera}.
 
@@
 	\begin{algorithmic}[1]
-	     \STATE \textbf{Input:} Number of particles $J$, timesteps $T$, measurement model $p(y_t^*|x_t, \theta)$, simulator $\process(x_{t+1}|x_t, \theta),$ IF2 cooling schedule $\eta_n$, MOP-$\alpha$ discounting parameter $\alpha$.
-		\STATE Initialize $\theta_0$, $n=0.$
-        \STATE Run IF2 until initial "convergence" under cooling schedule $\eta_n$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_n := \frac{1}{J}\sum_{j=1}^J \Theta_j.$
+	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $p(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta),$ IF2 cooling schedule $\eta_t$, MOP-$\alpha$ discounting parameter $\alpha$.
+		\STATE Initialize $\theta_0$, $t=0.$
+        \STATE Run IF2 until initial "convergence" under cooling schedule $\eta_t$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_t := \frac{1}{J}\sum_{j=1}^J \Theta_j.$
 		\WHILE{Procedure not converged:}
 		\STATE Draw $\omega \in \Omega$
-		\STATE Perform forward pass by running Algorithm \ref{alg:mop} with discounting parameter $alpha$ to obtain $\loglik(\theta_n, \theta_n, \omega, J).$
-		\STATE Obtain $g(\theta_n) = \nabla_{\theta_n} (-\loglik(\theta_n, \theta_n, \omega, J))$ via backpropagation, $H(\theta_n)$ such that $\lambda_{\min}(H) \geq c$.
+		\STATE Perform forward pass by running Algorithm \ref{alg:mop} with discounting parameter $\alpha$ to obtain $\loglik(\theta_t, \theta_t, \omega, J).$
+		\STATE Obtain $g(\theta_t) = \nabla_{\theta_t} (-\loglik(\theta_t, \theta_t, \omega, J))$ via backpropagation, $H(\theta_t)$ such that $\lambda_{\min}(H) \geq c$.
 		\STATE Obtain $\eta$ with line search or an annealing schedule.
-		\STATE Update $\theta_{n+1} := \theta_n - \eta (H(\theta_n))^{-1} g(\theta_n)$, $n:=n+1.$
+		\STATE Update $\theta_{t+1} := \theta_t - \eta (H(\theta_t))^{-1} g(\theta_t)$, $t:=t+1.$
 		\ENDWHILE
-		\STATE Return $\hat{\theta} := \theta_n.$
+		\STATE Return $\hat{\theta} := \theta_t.$
 	\end{algorithmic}
@@
 This can be attributed to both GPU acceleration and the just-in-time (JIT) compilation functionality that \texttt{JAX} provides.
-When tested on a computer with an Intel i9-13900K processor, 128GB of RAM, and a NVIDIA RTX3090 GPU, each post-compilation particle filtering operation on the Dacca cholera model from \cite{king08} with 10000 particles took roughly 379 milliseconds (with a standard deviation of 279 microseconds over seven runs), a considerable improvement over the existing performance in the widely used \texttt{pomp} package of \cite{king16}, which took between 6.19 to 6.39 seconds over ten runs on the same machine.
+When tested on a computer with an Intel i9-13900K processor, 128GB of RAM, and a NVIDIA RTX3090 GPU, each post-compilation particle filtering operation on the Dhaka cholera model from \cite{king08} with 10000 particles took roughly 379 milliseconds (with a standard deviation of 279 microseconds over seven runs), a considerable improvement over the existing performance in the widely used \texttt{pomp} package of \cite{king16}, which took between 6.19 to 6.39 seconds over ten runs on the same machine.
 
@@
 
-    \item The \textbf{force of infection}, $\lambda(t)$, is modeled by
+    \item The \textbf{force of infection}, $\lambda(t)$, is modeled by splines $s_j$
     \begin{equation}
@@
  \hline
- Method & Log-Likelihood & Rank \\ [0.5ex] 
+ Method & Best Log-Likelihood & Rank \\ [0.5ex] 
  \hline\hline
- IF2 2015 & -3768.63 & 5\\ 
+     IFAD-0 & -3747.87 & 1\\
+     IFAD-1 & -3748.53 & 2\\
      IF2 (Ours) & -3755.91 & 3\\
      IF2 (Warm Start) & -3759.60 & 4 \\
- IFAD-1 & -3748.53 & 2\\
- IFAD-0 & -3747.87 & 1\\
+     IF2 2015 & -3768.63 & 5\\ 
+     MOP-0 Alone & -3795.74 & 6\\
      MOP-1 Alone & -3797.38 & 7\\
- MOP-0 Alone & -3795.74 & 6\\
  \hline
@@
     \textbf{Takeaway:} In this panel, we see that (1) IF2 fails to find the MLE while the gradient steps perform better at fine-grained refinement, and (2) performing gradient steps alone without a warm-start leads to the search getting stuck in local minima and saddle points, ultimately demonstrating the necessity of a hybrid algorithm like IFAD.
-
     In both, as before, we use a dotted red line to display the MLE. }
@@
 
-\begin{figure}[u!]
+\begin{figure}[t!]
     \centering
@@
 Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual} that we call \texttt{pompy}.
-The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{jax} Python package from \citet{jax2018github}.
+The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{JAX} Python package from \citet{jax2018github}.
 
@@
 
-In the interest of simplicity, we only explore the performance of gradient descent on the Dacca cholera model of \cite{king08}.
+In the interest of simplicity, we only explore the performance of gradient descent on the Dhaka cholera model of \cite{king08}.
 Exploring the use of various first and second order methods such as Newton's method, ADAM, L-BFGS, or other popular iterative optimization methods with the gradient estimates that MOP-$\alpha$ provides in the context of the particle filter could very well lead to nontrivial speedups and increased performance.
@@
     \label{lem:change-measure-proper-weights}
-    Suppose that $\{(X_j,u_j),j=1,\dots,J\}$ targets $f_X$.
-Now, let $\{(Y_j,v_j),j=1,\dots,J\}$ be a sample drawn from $\{(X_j,u_j)\}$ where $(X_j,u_j)$ is represented, on average, $\pi_j J$ times.
+    Suppose that $\{(X_j^J,u_j^J),j=1,\dots,J\}$ targets $f_X$.
+Now, let $\{(Y_j^J,v_j^J),j=1,\dots,J\}$ be a sample drawn from $\{(X_j^J,u_j^J)\}$ where $(X_j^J,u_j^J)$ is represented, on average, $\pi_j J$ times.
 This could amount to multinomial resampling with probability $\pi_j$, or systematic resampling.
@@
     \[
-    (Y_j,v_j) = \big(X_{a(j)},u_{a(j)}/\pi_{a(j)}\big),
+    (Y_j^J,v_j^J) = \big(X_{a(j)},u_{a(j)}/\pi_{a(j)}\big),
     \]
@@
 \begin{proof}
-    Let $h$ be an integrable function. 
-    \begin{align*}
-        \E\left[\sum_{j=1}^J h(Y_j) v_j\right] 
-        &= \sum_{j=1}^J \E\left[h(X_{a(j)}) \frac{u_{a(j)}}{\pi_{a(j)}}\right] \\
-        &= \sum_{j=1}^J \E\left[h(X_{j}) u_{j}\right],
-    \end{align*}
-    and similarly for the denominator.
+
+    Note that as the $Y_j^J$ are a subsample from $X_j^J$, $h$ can be a function of $Y$ as well as it is one for $X$. We then expand
+    $$\frac{\sum_j h(Y_j^J) v_j^J}{\sum_j v_j^J} = \frac{\sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J}}.$$
+
+    We know that by definition,
+    $$\frac{\sum_j h(X_j^J)u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} \E_{f_X}[f(X)].$$
+
+    We want to show
+    $$\frac{\sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J}} - \frac{\sum_j h(X_j^J)u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} 0.$$.
+
+    %To see this, observe that 
+    %$$\frac{\sum_j h(X_j^J)u_j^J}{\sum_j u_j^J} = \frac{\sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J}{\sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J},$$
+
+    %so it remains to show that
+    %$$\frac{\sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J}} 
+    %- \frac{\sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J}{\sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J}
+    %\stackrel{a.s.}{\to} 0.$$.
+
+    This is equivalent to showing that
+    $$\sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J \sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}
+    - \sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J} \sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J 
+    \stackrel{a.s.}{\to} 0.$$.
+
+    From $|a d- b c| \leq|d||a-c|+|c||b-d|$, we see that it suffices to bound
+
+    $$ \sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}
+    -  \sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J $$
+    and 
+    $$ \sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J
+    - \sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J}$$
+    separately.
+
+    %Write 
+    %$$\widetilde{\pi_{k_j}^J} := \frac{\frac{u_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J}}, \widetilde{u_{j}^J} := \frac{u_j^J}{\sum_j u_j^J}.$$
+
+    Write $g(X_j^J) = h(X_j^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}$. We therefore need to show that 
+    $$\sum_j Z_j^J := \sum_j \left(g(X_{k_j}^J) -  g(X_j^J) \pi_j^J \right) \stackrel{a.s.}{\to} 0.$$
+
+    Assume that $\E[\left(Z_j^J\right)^4] < \infty$. We can then follow the argument of \cite{chopin20} from this point on, where 
+    $$\E\left[\left(\sum_j Z_j^J\right)^4\right] 
+    = J\E[(Z_1^J)^4] + 3J(J-1)\left(\E[(Z_j^J)^2]\right) \leq CJ^2,$$ for some $C>0$. By Markov, 
+    $$\mathbb{P}\left(\left|\frac{1}{J} \sum_{j=1}^J Z_j^J\right|>\epsilon\right) 
+    \leq \frac{1}{\epsilon^4J^4 } 
+    \mathbb{E}\left[\left(\sum_{j=1}^J Z_j^J\right)^4\right] \leq \frac{C}{\epsilon^4J^2},$$
+    and as these terms are summable we can apply Borel-Cantelli to conclude that these deviations happen only finitely often for every $\epsilon>0,$ giving us the almost-sure convergence for
+
+    $$ \sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}
+    -  \sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0.$$
+
+    Similarly, we also have that
+    $$ \sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J
+    - \sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J} \stackrel{a.s.}{\to} 0,$$
+
+    and the result is proved.
+
+   % Let $h$ be an integrable function. 
+    %\begin{align*}
+    %    \E\left[\sum_{j=1}^J h(Y_j) v_j\right] 
+    %    &= \sum_{j=1}^J \E\left[h(X_{a(j)}) \frac{u_{a(j)}}{\pi_{a(j)}}\right] \\
+    %    &= \sum_{j=1}^J \E\left[h(X_{j}) u_{j}\right],
+    %\end{align*}
+    %and similarly for the denominator.
 Now, the numerator and denominator of the reweighted quantity have the same expectation as the numerator and denominator of the original quantity, so they must converge to $ E_\pi(h(X))$ almost surely as well.

### 2024-01-05 presentation/presentation.tex
Updates from Overleaf

\documentclass{beamer}
\setbeamertemplate{caption}[numbered]
\usepackage{tikz}
\usepackage{graphicx}
\usetheme{AnnArbor}
%\usecolortheme{wolverine}
\title[AD for POMPs]{Automatic Differentiation Accelerates Inference for Partially Observed Stochastic Processes}
\subtitle{\small{Inference for Partially Observed Structured Dynamic Systems}}
\author[Kevin Tan]{Kevin Tan \\ \small{University of Pennsylvania, Department of Statistics and Data Science} \\\vspace{0.5cm} \small{Joint work with Giles Hooker and Edward Ionides}}
\date[January 4, 2024]{IMS Asia Pacific Rim Meeting \\ January 4, 2024}

\usepackage{amsmath}
\usepackage{tabularx}
\usepackage{mathtools}
\usepackage{bbold}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{pifont}
%\usepackage[left=1in,right=1in,]{geometry}
\usepackage{multicol,lipsum,xcolor}
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{amsmath,amsthm,amssymb,bm}
\usepackage{bbold}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{enumerate}
%\usepackage{enumitem}
\usepackage{natbib}
\usepackage{parskip}
\usepackage{comment}
\usepackage{microtype}
\usepackage{diagbox}
\usepackage{tikz}
\usepackage{fancyhdr}
\usepackage{appendixnumberbeamer}

\newcommand{\Prob}{\displaystyle \mathbb {P}}
\newcommand{\jac}{\bigg| \frac{\dif x}{\dif y} \bigg|}
\newcommand{\inftyint}{\int_{-\infty}^{\infty}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\C}{\mathbb{C}}

% best package ever
%\usepackage{realhats}

\def\mathfamilydefault{\rmdefault}

\usepackage{color}
\usepackage{xcolor}
\definecolor{hlpink}{rgb}{0.96, 0.76, 0.76}
\definecolor{hlblue}{rgb}{0.54, 0.81, 0.94}
\definecolor{hlyellow}{rgb}{0.98, 0.91, 0.71}
\definecolor{hlgreen}{rgb}{0.7, 0.93, 0.36}
\definecolor{hlpurple}{rgb}{0.8, 0.8, 1.0}

\newcommand{\highlight}[2][yellow]{\mathchoice%
  {\colorbox{#1}{$\displaystyle#2$}}%
  {\colorbox{#1}{$\textstyle#2$}}%
  {\colorbox{#1}{$\scriptstyle#2$}}%
  {\colorbox{#1}{$\scriptscriptstyle#2$}}}%

 \AtBeginSection[]{
  \begin{frame}
  \vfill
  \centering
  \begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
    \usebeamerfont{title}\insertsectionhead\par%
  \end{beamercolorbox}
  \vfill
  \end{frame}
}

\include{mop/macros}

\begin{document}

\maketitle

\section{Background}

\begin{frame}{Partially Observed Markov Processes}

    \begin{itemize}
        \item Unobserved Markov process $\{X_t, t \geq t_0\}$, observations $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$.
        \item \pause Unknown parameter $\theta \in \Theta$, to be estimated.
        \item \textbf{Notation:}
        \begin{itemize}
            \item $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ is the \textbf{process model}.
            \item \pause $\texttt{process}\left(x_n, \theta\right)$ is the \textbf{simulator} corresponding to the process model. 
            \item \pause $f_{Y_n|X_n}\left(y_n \mid x_n, \theta\right)$ is the \textbf{measurement model}.
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{Particle Filters}

    \begin{figure}
        \centering
        \begin{minipage}[c]{0.5\textwidth}
            \centering
            \includegraphics[scale=0.15]{imgs/pfilter.png}
            \end{minipage}
        \begin{minipage}[c]{0.45\textwidth}
            \caption{
            \\
            -- Dirac measure approximation of the Bayes filter. 
            \\ 
            -- Yields a filtering distribution and likelihood estimate.
            \\ 
            -- Graphic from \cite{berg2019pfilterimage}.}
        \end{minipage}
        \label{fig:pfilter-illustration}
    \end{figure}

    \begin{itemize}
        \item Monte Carlo approximation of the Bayes filter.
        \begin{itemize}
            \item \pause Maintain belief on current state $x_n^F$. 
            \item \pause Simulate forward to get $x_{n+1}^P$, observe observation $y_{n+1}^*$.
            \item \pause Update $x_{n+1}^F$ accordingly.
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{Automatic Differentiation}

    \begin{figure}
        \begin{minipage}[c]{0.7\textwidth}
            \includegraphics[scale=0.3]{imgs/autograd.png}
        \end{minipage}
        \begin{minipage}[l]{0.27\textwidth}
            \caption{
            \\$f(x,y) = x^2+xy+2$, $f'(x,y) = (2x+y,x).$
            -- Figure from \href{https://avinashselvam.medium.com/automatic-differentiation-explained-9f02c74e9a90}{Medium}.}
        \end{minipage}
        \label{fig:autograd-illustration}
    \end{figure}

    \begin{itemize}
        \item Evaluates the gradient of a (scalar or vector-valued) computer program w.r.t. its arguments.
        \item \pause Traverses computational graph (of primitive functions) with chain rule.
    \end{itemize}

\end{frame}

\section{Motivation}

\begin{frame}{Maximum Likelihood Inference is Hard in POMPs}

\begin{enumerate}
    \item \textbf{Intractable likelihood functions}! Bypassed with e.g. simulated likelihood, likelihood-free inference, particle filters, etc. 
    \item \pause May not have access to transition densities, only a \textbf{simulator}. 
    \begin{itemize}
        \item \pause IF2 (\cite{ionides15}), particle MCMC (\cite{doucet2010pmcmc}), only full-information methods that can deal with this.
    \end{itemize}
    \item \pause \textbf{Significant Monte-Carlo noise in likelihood estimate} makes accurate parameter estimation difficult.
\end{enumerate}
\end{frame}

\begin{frame}{IF2 Plateaus Quickly But Struggles to Find the MLE}
\begin{figure}
    \centering
    \includegraphics[scale=0.5]{imgs/095/if2fail.png}
    \caption{Performance of IF2 on \cite{king08} Dhaka cholera model. Shaded area represents 0th to 80th percentile, solid line is median of 100 runs. While IF2 makes quick initial progress, it fails to find the MLE.}
    \label{fig:if2fail}
\end{figure}
\end{frame}

\begin{frame}{Hang on, wait a minute!}
    \begin{itemize}
        \item Original iterated filtering algorithm: (noisy) approximation of score (\cite{ionides06-pnas}). 
        %\item \pause Performed a single "gradient step" at the end of each filtering iteration.
        \pause
        \item \textbf{Would performing automatic differentiation (AD) on the likelihood estimate from the particle filter lead to a less noisy score approximation?}
    \end{itemize}
    \pause
    \begin{block}{Problem!}
        But the particle filter has discrete stochastic resampling! How can we differentiate this?
    \end{block}
\end{frame}

\section{Previous Work: AD for Particle Filters}

\begin{frame}{Previous Work}

    \begin{table}[h!]
        \centering
        \begin{tabular}{||c c||} 
         \hline
         Authors & Method \\ [0.5ex] 
         \hline\hline
         \cite{poyiadjis11} & Particle approximation of score
         \\ & $O(N^2)$ variance, unbiased \\
         \hline
         \cite{blei2018vsmc} & Backprop through vanilla PF \\ 
         & Asymptotic bias \\
         \hline
         \cite{corenflos21} & Optimal transport resampling \\
         & Consistent, $O(J^2)$ runtime \\
         \hline
         \cite{scibior21} & Stop-gradient trick \\ 
         & Recovers \cite{poyiadjis11} with AD \\
         \hline 
         \cite{singh22} & Fixed-lag smoothing \\ 
         & Need transition densities \\ 
         \hline
        \end{tabular}
        \label{table:lit-review}
    \end{table}
\end{frame}

\begin{frame}{Our Contributions}
    \begin{itemize}
        \item New theoretical framework/algorithm/gradient estimator we call MOP-$\alpha$.
        \begin{itemize}
            \item \pause Gradient estimates of \cite{blei2018vsmc} (MOP-0), \cite{poyiadjis11} and \cite{scibior2021dpf} (MOP-1) are special cases.
            \item \pause Does not need transition densities, only a differentiable simulator.
            \item \pause Can optimize a bias-variance tradeoff. 
        \end{itemize}
        \item \pause Promising hybrid algorithm, warm-starts gradient descent (using this estimator) with IF2.
        \item \pause Outperforms IF2 on Cholera model of \cite{king08}.
    \end{itemize}
\end{frame}

\section{Smooth Extensions to the Particle Filter}

\begin{frame}{Intuition}
    \begin{block}{Problem}
        Want to differentiate through the particle filter. But the particle filter has discrete stochastic resampling! 
    \end{block}
    \pause 
    \begin{block}{Idea}
        Don't differentiate $\hat\ell(\theta)$ directly, differentiate through a (suitably) reweighted bootstrap filter.
    \end{block}
    \begin{itemize}
        \item \pause Particle filter run with state transitions \& resampling under $\phi \in \Theta$. 
        \item \pause Reweight to evaluate likelihood at nearby $\theta \in \Theta$ (same resampling).
        \item \pause Fix seed to treat particles, weights under $\phi$ as constants.
        \item \pause \textbf{Now only need to differentiate through likelihood ratios!}
    \end{itemize}
\end{frame}

\begin{frame}{What is this reweighting?}
    Ordinarily, the likelihood estimate for $\phi$ is 
    $$\hat\lik(\phi) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J \left(\underbrace{f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}, \phi)}_{\textcolor{blue}{\text{Measurement Model}}}
    \cdot 
    \underbrace{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi} | x_{n-1,j}^{F,\phi}, \phi)}_{\textcolor{red}{\text{Process model}}}\right).$$
    \pause
    So to estimate the likelihood for $\theta$, we can correct 
    $$
        \hat\lik(\theta) 
        = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}, \phi) \cdot 
        f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi} | x_{n-1,j}^{F,\phi}, \phi) 
        \cdot \underbrace{s_{n,j} \cdot r_{n,j}}_{\text{Correction Term}},
    $$
    \pause
    where the multiplicative correction terms are
    $$
        \underbrace{s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}}_{\textcolor{blue}{\text{Measurement Model Likelihood Ratios}}},\;\; \underbrace{r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)}}_{\textcolor{red}{\text{Process Model Likelihood Ratios}}}.
    $$

\end{frame}

\begin{frame}{New Problem}
    \begin{block}{Problem}
        Need likelihood ratios w.r.t. process and measurement models!
    \end{block}
    \pause 
    \begin{block}{Solution}
        Offload process model derivatives to differentiable simulator. 
    \end{block}
    \begin{itemize}
        \item \pause Run particle filter twice under same seed: one with process model at $\phi$, another with process model at $\theta$. Resample according to $\phi$.
        \item \pause Correct only with resampling likelihood ratios. 
        \item \pause Introduce additional discounting parameter $\alpha$ for bias-variance tradeoff.
    \end{itemize}
\end{frame}

\begin{frame}{Algorithm: Measurement Off-Policy, MOP-$\alpha$}
    \begin{enumerate}
        \item \textbf{Run a particle filter once at} $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$. Fix seed throughout. Set initial weights $w_{0,j}^{F,\theta} = 1/J$.
        \item \pause \textbf{At each timestep,} propagate weights $w_{n,j}^{P,\theta} := (w_{n-1,j}^{F,\theta})^\alpha$. 
        \item \pause Simulate process model ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$, evaluate measurement model $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$, evaluate conditional likelihood under $\phi$, $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$, \textbf{as usual.}
        \item \pause \textbf{Resample according to $\phi$:} $k_{1:J} \sim \prob\big(k_{j}=m\big) \propto g^{\phi}_{n,m}$.
        \item \pause \textbf{Correct filtering weights:} $\displaystyle w^{F,\theta}_{n,j}= w^{P,\theta}_{n,k_j} \times  g^{\theta}_{n,k_j}/ g^{\phi}_{n,k_j}$. 
    \end{enumerate}
    \pause 
    \centering{\textbf{Special Case:} If $\theta=\phi$, only need to run one particle filter! Set $X_{n,j}^{P,\theta}, X_{n,j}^{F,\theta}$ to be copies of the $\phi$ counterparts, where gradients don't propagate. Explains the \textbf{stop-gradient} trick of \cite{scibior21}.}
\end{frame}

\begin{frame}{Bias-Variance Tradeoff with $\alpha$}
\begin{itemize}
    \item $\alpha$ balances a tradeoff between:
    \begin{itemize}
        \item \pause Maintaining memory of each particle's ancestral trajectory (most extreme when $\alpha=1$).
        \item \pause Considering only the single-step transition dynamics (when $\alpha=0$).
    \end{itemize}
    \item \pause\textbf{Exponentially-weighted moving average} where $\alpha$ controls the amount of discounting.
    \item \pause Bias small if $y_{n+1:N}^*$ not informative for $x_n$ given $y_{0:n}^*$ (\cite{corenflos21}).
\end{itemize}

\end{frame}

\begin{frame}{Bias-Variance Tradeoff with $\alpha$}
\begin{figure}
    \centering
    \includegraphics[scale=0.5]{imgs/mlegrad.png}
    \caption{Gradient estimates at MLE for trend in \cite{king08} cholera model. When $\alpha=1$, estimate unbiased with high variance. When $\alpha=0$, lower variance but biased.}
    \label{fig:bias-variance}
\end{figure}

\end{frame}

\begin{frame}{Guarantee}
    \begin{prop}[Correctness of MOP-$\alpha$]
    \label{prop:mop-correctness}
    When either $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the posterior and is unbiased for the likelihood under $\theta$. 
    When $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter. Its gradient when $\alpha=1$ is the estimate of \citet{poyiadjis11},
    $$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*),$$
    and when $\alpha=0,$ is the gradient estimator of \cite{blei2018vsmc},
    $$
        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
    $$
    i.e. differentiating through a vanilla particle filter (\cite{scibior21}).
\end{prop}
\end{frame}

\section{Practical Maximum Likelihood Inference}

\begin{frame}{One possible procedure:}
    \begin{itemize}
        \item Warm-start gradient descent, or some other first or second order procedure (using the gradient estimate given by MOP-$\alpha$), with the output of an initial search of IF2. 
        \item \pause Run IF2, aggressively cooling to a fixed learning rate, till search stalls.
        \item \pause Refine this coarse solution with gradient descent.
    \end{itemize}
    \pause
    \centering{\textbf{Iterated Filtering with Automatic Differentiation (IFAD)}}
\end{frame}

\begin{frame}{Convergence Analysis}
    \begin{prop}[Convergence of IFAD-1]
        \label{thm:convergence}

        Consider a variant of IFAD-1 where one stops if the gradient estimate $||g(\theta_n)|| \leq \sigma \epsilon$. Assume $-\ell$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$. $\sigma = \frac{4 \Gamma}{(1-\beta)}> 4$. Define $G$ as in Assumption \ref{assump:local-bounded-derivative}.

        Let $H(\theta_n)$ be a matrix possibly dependent on $\theta_n$ with minimum eigenvalue always greater than some $c>0$. Choose learning rate $\eta$ and error tolerance $\epsilon$ such that, where $\beta$ is the Armijo condition hyperparameter,
        \begin{equation*}
            \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_n)||.
        \end{equation*}

        Then, with $J$ large enough, the following holds with probability at least $1-\delta$:
        \begin{equation*}
        \ell(\theta^*) - \ell(\theta_{n+1}) \leq \left(1-\eta\beta\frac{8\gamma}{9c}\right)(\ell(\theta^*)-\ell(\theta_n))
        \end{equation*}

        and the algorithm terminates when $||\nabla_\theta \ell(\theta_n)|| \leq (1+\sigma) \epsilon$.
        \end{prop}
\end{frame}

\begin{frame}{Convergence Analysis}
    \begin{itemize}
        \item Similar to \cite{mahoney16}, uses concentration inequalities from \cite{delmoral2011ci}.
        \item Gradient stage of IFAD-1 converges linearly to the MLE if:
        \begin{enumerate}
            \item \pause The log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE.
            \item \pause The IF2 stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE. 
        \end{enumerate}
        \item \pause \textbf{Conjecture:} This applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE and behaves a lot like SGD. 
    \end{itemize}

\end{frame}

\begin{frame}{Runtime}

\begin{itemize}
    \item Warm-start:
    \begin{itemize}
        \item \pause Initial warm-start "convergence" happens fairly quickly in practice. 
        \begin{itemize}
            \item \pause Dhaka cholera model of \cite{king08}: 40 iterations, usually 100-200 for global search.
        \end{itemize}
    \end{itemize}
    \item \pause MOP-$\alpha$:
    \begin{itemize}
        \item \pause Gradient takes 3.75x time of \texttt{pfilter()}, in line with cheap gradient principle (\cite{kakade2019provably}).
    \end{itemize}
    \item \pause Runtime \textbf{linear, and not quadratic}, in number of particles.
    \item \pause Re-implementation in \texttt{JAX} (\cite{jax2018github}) led to 16x speedup v.s. \texttt{pomp} package of \cite{king08}.
\end{itemize}

\end{frame}

\section{Numerical Experiments}

\begin{frame}{Cholera in Bangladesh}

    \begin{figure}
        \centering
        \includegraphics[scale=0.5]{imgs/dacca.png}
        \caption{Illustration of SIR model from \cite{king08}.}
        \label{fig:sir}
    \end{figure}

    \begin{itemize}
        \item Dhaka model from \cite{king08}, also used in \cite{ionides15} to benchmark IF2.
        \item \pause Stochastic SIR compartmental model with transition uncertainty driven by Brownian motion.
        \item \pause Force of infection $\lambda(t)$ modeled with splines for seasonality, etc.
    \end{itemize}

        %\item \pause Population $H(t)$ divided into:
        %\begin{itemize}
        %    \item \pause Susceptible compartment $S(t)$.
        %    \item \pause Infected compartment $I(t)$.
        %    \item \pause Three ($k=3$) recovered compartments $R_1(t), ..., R_k(t)$ denoting varying degrees of cholera immunity.
        %\end{itemize}
        %\item \pause $M(t)$ denotes the cholera deaths in each month.
\end{frame}

\begin{frame}{Results}

    \begin{table}[h!]
        \centering
        \begin{tabular}{||c c c||} 
         \hline
         Method & Best Log-Likelihood & Rank \\ [0.5ex] 
         \hline\hline
         IFAD-0 & $\textbf{-3747.87}$ & 1\\
         IFAD-1 & $\textbf{-3748.53}$ & 2\\
         IF2 (Ours) & -3755.91 & 3\\
         IF2 (\cite{ionides15}) & -3768.63 & 4\\ 
         MOP-0 Alone & -3795.74 & 5\\
         MOP-1 Alone & -3797.38 & 6\\
         \hline
        \end{tabular}
        \caption{\\
        -- IFAD performs the best among all methods, finding the MLE. \\
        -- Our implementation of IF2 outperforms that of \cite{ionides15} but underperforms IFAD.}
        \label{table:mle}
    \end{table}

    \begin{itemize}
        \item Benchmarked IFAD against IF2 on a challenging global search problem. 
        \item \pause Performed 100 searches each. 
    \end{itemize}

\end{frame}

\begin{frame}{Results}

\begin{figure}[htbp!]
    \centering
    \includegraphics[scale=0.37]{imgs/095/pairs.png}
    \includegraphics[scale=0.37]{imgs/095/qq.png}
    \caption{\\
    --\textbf{Left:} Paired searches from the same starting point. \\
    --\textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches. \\
    -- IFAD has the edge and manages to find the MLE.\\
    -- No IF2 search successfully gets within 7 log-likelihood units of it.}
    \label{fig:scatter}
\end{figure}
\end{frame}

\begin{frame}{Results}

\begin{figure}[H]
    \centering
    \includegraphics[scale=0.28]{imgs/095/boxplot.png}
    \includegraphics[scale=0.31]{imgs/095/boxplotall.png}
    \caption{\\
    -- \textbf{Left:} Results of best run out of every ten runs, simulating procedure of running a few searches and choosing the best one. \\
    -- \textbf{Right:} MOP alone drastically underperforms all other methods, failing to get close to the MLE. \\
    -- IF2 warm-start necessary in challenging nonconvex and noisy problems.}
    \label{fig:boxplot-search}
\end{figure}

\end{frame}

\begin{frame}{Results}
\begin{figure}[htbp!]
    \centering
    \includegraphics[scale=0.5]{imgs/095/optim.png}
    \caption{\\
    -- Solid lines depict the median negative log-likelihood at each iteration.\\
    -- Shaded area depicts the best search at any iteration and the 80\% percentile.}
    \label{fig:optim}
\end{figure}
%-- While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD -- IFAD has better tail control and successfully reaches the MLE. The initial loss in log-likelihood upon transitioning from IF2 to MOP seems to help with escaping local minima. 
\end{frame}

\begin{frame}{Results}

\begin{figure}[u!]
    \centering
    \includegraphics[scale=0.35]{imgs/095/hist.png}
    \caption{Histogram comparison of IFAD, IF2, and MOP alone. Right is better.}
    \label{fig:hist-all}
\end{figure}
\end{frame}

\begin{frame}{Takeaways}
    \begin{enumerate}
        \item IF2 converges quickly to a neighborhood of the MLE but fails to find the MLE.
        \item \pause Gradient steps perform better at fine-grained refinement.
        \item \pause Performing gradient steps alone without a warm-start leads to the search getting stuck in local minima and saddle points.
        \item \pause IFAD combines the best of IF2 and MOP, approaching the MLE quickly and successfully performing refinement -- even on a very difficult global search problem.
    \end{enumerate}
\end{frame}

\section{Conclusions and Future Work}

\begin{frame}{Conclusion}
    \begin{itemize}
        \item New theoretical framework/algorithm/gradient estimator that encompasses a few existing gradient estimates.
        \item \pause Promising hybrid algorithm that warm-starts gradient descent (using this estimator) with IF2.
        \item \pause Outperforms IF2 on Dhaka cholera model of \cite{king15}.
    \end{itemize}
\end{frame}

\begin{frame}{Future Work}
    \begin{itemize}
        \item MOP-$\alpha$ cannot handle discrete latent states. Maybe likelihood ratios can get around this.
        \item \pause Convergence rate of IF2 not known, conjecture similar behavior to SGD. 
        \item \pause Only gradient descent explored, other methods like Newton's method, ADAM, L-BFGS, etc. may perform better.
        \item \pause Extensions to panel and spatiotemporal data, e.g. with a block particle filter in the lens of \cite{ionides22} and \cite{ning23}.
        \item \pause Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual}.
    \end{itemize}
\end{frame}

\bibliographystyle{apalike}
\bibliography{mop/bib-ifad, mop/bib-ref}

\appendix
\section{Appendix}

\begin{frame}{Previous Work}

    \begin{itemize}
        \item \cite{poyiadjis11}, particle approximation of the score. \pause \\ \cite{scibior21}: AD + "stop-gradient trick". 
        \item 
        \pause \cite{blei2018vsmc}, backpropagate through vanilla particle filter.
        \item 
        \pause \cite{corenflos21}, optimal transport resampling.
        \item 
        \pause \cite{singh22}, fixed-lag smoothing.
    \end{itemize}
\end{frame}

\begin{frame}{Previous Work: Issues}
    \begin{itemize}
        \item Not (yet) compatible with desire for simulation-based inference.
        \begin{itemize}
            \item \pause Need modifications to use the bootstrap filter.
            \item \pause Sometimes special cases needed, e.g. transitions factoring into a policy and deterministic model (\cite{singh22}).
        \end{itemize}
        \item \pause Computationally expensive, quadratic in number of particles.
        \begin{itemize}
            \item \pause Optimal transport resampling (\cite{corenflos21}), marginal particle filters (\cite{scibior21}).
        \end{itemize}
        \item \pause High variance, or asymptotically biased.
        \begin{itemize}
            \item \pause Either one drops resampling terms and accepts asymptotic bias (\cite{blei2018vsmc}), or has variance quadratic in horizon (\cite{poyiadjis11}, \cite{scibior21})
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{(Improved) Iterated Filtering (\cite{ionides15})}
    \begin{itemize}
        \item Only (practical) full-information, frequentist method of parameter estimation in POMPs that does not need transition densities. 
        \begin{itemize}
            \item \pause Each particle has its own set of parameters.
            \item \pause Particle parameters perturbed at every timestep.
            \item \pause Parameters resampled with their particles according to likelihood.
        \end{itemize}
        \item \pause Treats parameters as augmentations of the state space, evolving according to a random walk.
        \item \pause \textbf{Consistent estimates} via viewing this as a sequential Bayes map on the parameter distribution.
    \end{itemize}
\end{frame}

\begin{frame}{When Does IF2 Struggle?}

\begin{itemize}
    \item \textbf{IF2 is great at approaching the MLE quickly!}
    \item \pause Performs parameter updates at every timestep, so it does not need to wait to filter through the entire trajectory before each update.
    \item \pause \textbf{But struggles at squeezing out the last few units of log-likelihood.}
    \begin{itemize}
        \item \pause Especially true in highly nonlinear, nonconvex, and noisy settings, e.g. \textbf{disease models}. 
    \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Assumptions}

    \begin{aspt}[Smooth Neighborhood]
    \label{assump:smooth-nbhd}
    There exists a neighborhood $\mathcal{N}(\phi)$ around any $\phi \in \Theta$ where for all $\theta \in \mathcal{N}(\phi)$ and almost every $\omega \in \Omega$, the Monte Carlo estimate of the likelihood at $\theta \in \mathcal{N}(\phi)$ with the system evolving according to $\phi$ conditional on $\omega$, $\hat\lik(\theta,\phi,\omega, J)$, is twice differentiable in $\theta$. 
    \end{aspt}

    \begin{itemize}
        \item \pause \textbf{Justification:} For suitably nearby $\theta$ and $\phi$, the resampling indices for $\hat\lik(\theta,\phi,\omega, J)$ and $\hat\lik(\phi,\phi,\omega, J)$ are the same.
        \item \pause This eliminates any discontinuities from resampling.
        \item \pause For small enough $\mathcal{N}(\phi)$, the likelihood ratios are bounded. 
        \item \pause The likelihood only changes by a factor of the likelihood ratios, which are bounded and smooth in $\theta$ if the densities used in the calculation are. 
    \end{itemize}

\end{frame}

\begin{frame}{Assumptions}

    \begin{aspt}[Continuity of the Likelihood] $\ell(\theta)$ proper is continuous in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.
    \end{aspt}
    \pause 
    \begin{aspt}[Bounded Measurement Model] There is an $\epsilon>0$ with $\epsilon^{-1}>f_{Y_n \mid X_n}\left(y_n{ }^* \mid x_n; \theta\right)>\epsilon$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
    \end{aspt}
    \pause 
    \begin{aspt}[Locally Bounded Derivative]
    \label{assump:local-bounded-derivative}
    Let $\mathcal{M}$ be an open subset of $\Omega$. There exists some function $G(\theta)$ and a constant $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$ such that
        $$
        \|\nabla \ell(\theta, \phi, \omega, J)\|_2<G(\theta) \leq G<\infty
        $$
        for every $\phi \in \mathcal{M}, \theta$ in smooth neighborhood $\mathcal{N}(\phi)$, and almost every $\omega \in \Omega$.
    \end{aspt}
\end{frame}

\begin{frame}{Algorithm: Measurement Off-Policy, MOP-$\alpha$}
    \begin{figure}
        \centering
        \includegraphics[scale=0.4]{imgs/mop.png}
        \caption{MOP-$\alpha$ algorithm.}
        \label{fig:mop}
    \end{figure}
\end{frame}

\begin{frame}{Why did we bother with this?}
    \begin{itemize}
        \item Lets us work with only a differentiable simulator.
        \item \pause Discounting parameter $\alpha$ helps with a bias-variance tradeoff (explained later). 
        \item \pause Gradients returned by the likelihood estimate $$\hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$$ have nice properties.
    \end{itemize}
\end{frame}

\begin{frame}{Iterated Filtering with Automatic Differentiation (IFAD)}
    \begin{figure}
        \centering
        \includegraphics[scale=0.45]{imgs/ifad.png}
        \caption{Warm-starting first/second order iterative optimization with IF2.}
        \label{fig:ifad}
    \end{figure}
\end{frame}

\begin{frame}{Why does this help?}
    \begin{itemize}
        \item IF2 converges quickly, but struggles with last few log-likelihood units. 
        \item \pause Gradient descent gets stuck in saddle points and poor local minima when the likelihood is nonconvex. 
        \item \pause Warm-starting gradient methods with IF2:
        \begin{itemize}
            \item \pause Under regularity conditions, the likelihood is well-behaved near the MLE.
            \item \pause Issues with saddle points and local minima alleviated. 
        \end{itemize}
        \item \pause Combining these two lets us enjoy the best of both worlds.
    \end{itemize}
\end{frame}

\begin{frame}{Convergence Analysis}

    Guarantee currently only holds for IFAD-1, as it forms a particle approximation of the score as in \cite{poyiadjis11}.
    \begin{itemize}
        \item \pause Can be extended if the bias for the gradient estimate given by MOP-$\alpha$ for $\alpha < 1$ is small enough.
        \item \pause e.g. $\alpha$ close to 1, or $y_{n+1:N}^*$ uninformative of current state $x_n$ given past and current measurements $y_{0:n}^*$.
        \item \pause Otherwise, need to handle biased gradient descent.
    \end{itemize}

\end{frame}

\begin{frame}{Cholera in Bangladesh}
    The \textbf{transition dynamics} follow this series of stochastic differential equations driven by Brownian motion:
    \begin{align*}
        d S&=\left(k \epsilon R_k+\delta(S-H)-\lambda(t) S\right) d t+d H-(\sigma S I / H) d B, \\ 
        d I&=\left(\lambda(t) S-(m+\delta+\gamma) I\right) d t+(\sigma S I / P) d B, \\ 
        d R_1&=\left(\gamma I-(k \epsilon+\delta) R_1\right) d t, \\ 
        \vdots \\ 
        d R_k&=\left(k \epsilon R_{k-1}-(k \epsilon+\delta) R_k\right) d t,
    \end{align*}
\end{frame}

\begin{frame}{Cholera in Bangladesh}
The \textbf{force of infection}, $\lambda(t)$, is modeled by splines $s_j$, 
    \begin{equation*}
        \lambda(t)=\exp \left\{\beta_{\text {trend }}\left(t-t_0\right)+\sum_{j=1}^{6} \beta_j s_j(t)\right\}(I / P) + \exp \left\{\sum_{j=1}^{6} \omega_j s_j(t)\right\},
    \end{equation*}
    where
    \begin{itemize}
        \item \pause $\beta_j$ model seasonality in the force of infection.
        \item \pause $\beta_{\text{trend}}$ models the trend in the force of infection.
        \item \pause $\omega_j$ represent seasonality of a non-human environmental reservoir of disease.
    \end{itemize}

\end{frame}

\begin{frame}{Cholera in Bangladesh}
    The \textbf{measurement model} for observed monthly cholera deaths is given by 
    \begin{equation*}
        Y_n \sim \gN(M_n, \tau^2M_n^2),
    \end{equation*}
    where $M_n$ is the true number of cholera deaths in that month.
\end{frame}

\end{document}

### 2024-01-21 mop/draft.tex
Updates from Overleaf

@@
 \subsubsection{Doubly Off-Policy Particle Filter}
+
+\kevin{rewrite to de-emphasize dop as in presentation, sell mop as the big theoretical idea that explains AD through PFs and unifies two previous estimators}
 
@@
 
-Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual} that we call \texttt{pompy}.
+Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual} that we call \texttt{pomp-py}.
 The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{JAX} Python package from \citet{jax2018github}.
@@
 \begin{defn}[Targeting]
-    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any integrable function $h$,
+    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any measurable and bounded functional $h$,
 \begin{equation}
@@
 \end{equation}  
-    A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if
+    A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if for any measurable and bounded functional $h$,
 \begin{equation}
@@
 \end{equation}
-for any integrable function $h$ as $J \to \infty$.
+as $J \to \infty$.
 \end{defn}
@@
     \]
-    where $a$ is called the ancestor function. Then, $\{(Y_j,v_j),j=1,\dots,J\}$ targets $f_X$.
+    where $a$ is called the ancestor function. If the importance sampling weights $u/\pi$ are bounded, then $\{(Y_j,v_j),j=1,\dots,J\}$ targets $f_X$.
 \end{lem}
@@
 
-    Assume that $\E[\left(Z_j^J\right)^4] < \infty$. We can then follow the argument of \cite{chopin20} from this point on, where 
+    Because the functional $h$ and importance sampling weights $u_{k_j}^J/\pi_{k_j}^J$ are bounded, we have that $\E[\left(Z_j^J\right)^4] < \infty$.
+We can then follow the argument of \cite{chopin20} from this point on, where
     $$\E\left[\left(\sum_j Z_j^J\right)^4\right] 
@@
     and the result is proved.
+
+    \kevin{MAY HAVE TO DEAL WITH CONDITIONING ON THE FILTRATION AT THE LAST TIMESTEP, AND THEN USING THE TOWER PROPERTY. EXPECTATION WILL BE THE SAME REGARDLESS FOR THE FOURTH MOMENT.}
 
@@
     \label{lem:marginal-proper-weights}
-    Suppose that $\{(X_j,u_j),j=1,\dots,J\}$ targets $f_X$.
-Also suppose that $Z_j \sim f_{Z|X}(\cdot | X_j)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
-Then, $\{(Z_j,u_j)\}$ targets $f_Z$.
+    Suppose that $\{(X_j^J,u_j^J),j=1,\dots,J\}$ targets $f_X$.
+Also suppose that $Z_j^J \sim f_{Z|X}(\cdot | X_j^J)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
+Then, if the $u_j^J$ are bounded, $\{(Z_j^J,u_j^J)\}$ targets $f_Z$.
 \end{lem}
 \begin{proof}
-    Let $h$ be an integrable function.
-
-    \begin{align*}
-        \E\left[J^{-1}\sum_{j=1}^J h(Z_j) u_j\right] &= \E\left[\E\left[J^{-1}\sum_{j=1}^J h(x_j) u_j f(Z_j|x_j)\Bigg| X_j=x_j\right]\right] \\
-        &= \E\left[J^{-1}\sum_{j=1}^J h(X_j) u_j f(Z_j|X_j)\right] \\
-        &\stackrel{a.s.}{\to} C_Z \E_{f_X}[h(X) f_{Z|X}(Z|X)] \\
-        &= C_Z \E_{f_Z}[h(Z)],
-    \end{align*}
-
-    where $C_Z$ is the normalizing factor, and similarly the denominator converges to $C_Z$. The result then follows from Slutsky's theorem.
+
+    We want to show that
+     $$\frac{\sum_j h(Z_j^J) u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)] = \E_{f_X}[\E_{f_{Z|X}}[h(Z) | X]].$$
+
+    We know that by definition, for any measurable and bounded functional $g$ with domain $\gX$,
+    $$\frac{\sum_j g(X_j^J)u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} \E_{f_X}[g(X)].$$
+
+    Let $\bar{u}_j^J = \frac{J u_j^J}{\sum_j u_j^J}$.
+Examine the numerator and denominator of the quantity $$\frac{J^{-1}\sum_j h(Z_j^J) u_j^J}{J^{-1}\sum_j u_j^J} = \frac{J^{-1}\sum_j h(Z_j^J) \bar{u}_j^J}{J^{-1}\sum_j \bar{u}_j^J}.$$
+
+    The denominator converges to $1$ almost surely. The numerator, on the other hand, is
+    $$\frac{1}{J}\sum_j h(Z_j^J)\bar{u}_j^J,$$
+
+    which by the same fourth moment argument to the above lemma converges almost surely to the limit of its expectation
+    $$\lim_{J\to\infty} \E\left[\frac{1}{J}\sum_j h(Z_j^J)\bar{u}_j^J\right] = \lim_{J\to\infty} \E_{X_j^J}\left[\E_{Z_j^J|X_j^J}\left[\frac{1}{J}\sum_j  h(Z_j^J)\bar{u}_j^J\bigg|X_j^J\right]\right].$$
+
+    It remains to swap the limit and expectations. We can do so with the bounded convergence theorem, and therefore obtain   
+    $$\frac{1}{J}\sum_j h(Z_j^J)\bar{u}_j^J \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)] = \E_{f_X}[\E_{f_{Z|X}}[h(Z) | X]].$$
 
@@
     \label{lem:posterior-proper-weights}
-    Suppose that $\{(X_j,u_j),j=1,\dots,J\}$ targets $f_X$.
-Also suppose that $(X^\prime_j,u^\prime_j) = \big(X_j,u_j\, f_{Z|X}(z^*|X_j)\big)$.
-Then, $\{(X^\prime_j,u^\prime_j)\}$ targets $f_{X|Z}(\cdot | z^*)$.
+    Suppose that $\{(X_j^J,u_j^J),j=1,\dots,J\}$ targets $f_X$.
+Also suppose that $(X^{\prime J}_j,u^{\prime J}_j) = \big(X_j^J,u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$.
+Then, if $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,u^{\prime J}_j)\}$ targets $f_{X|Z}(\cdot | z^*)$.
 \end{lem}
@@
 \begin{proof}
-    The proof is similar to that of Lemma \ref{lem:marginal-proper-weights}, except with Bayes' theorem. 
+    Again, we want to show that
+     $$\frac{\sum_j h(X_j^J) \cdot u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)} \stackrel{a.s.}{\to} \E_{f_{X|Z}}[h(X)|z^*].$$
+
+    We already have that
+    $$\frac{\sum_j h(X_j^J)u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} \E_{f_X}[h(X)].$$
+
+    %= \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j f_{Z|X}(z^*|X_j^J)f_X(X_j^J)}
+
+    Write $\bar{u}_j^{J} = \frac{J u_j^J}{\sum_j u_j^J}$.
+Define the weights $u^{' J}_j := u_j^J \cdot f_{Z|X}(z^*|X_j^J),$ and the corresponding self-normalized weights $\bar{u}_j^{' J} := \frac{J u^{' J}_j}{\sum_j u^{' J}_j}.$ The denominator of the below quantity
+    $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J}}{J^{-1} \sum_j \bar{u}_j^{' J}}$$
+    converges to $1$, while the numerator, by the same fourth moment argument as the first lemma, converges almost surely to the limit of its expectation, which is
+    \begin{align*}
+        \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J} \right]
+        &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)}\right] \\
+        &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
+        &= \lim_{J\to\infty} \E\left[ \sum_j h(X_j^J) \frac{\frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
+        &= \E_{X|Z}[h(X)|z^*],
+    \end{align*}
+
+    where we can swap the limit and expectations with the bounded convergence theorem.
 \end{proof}
@@
 \begin{proof}
-    When $\theta=\phi$, the ratio $\frac{g_{n,j}^\theta}{g_{n,j}^\phi}=1,$ and this reduces to the standard case.
+    When $\theta=\phi$, regardless of the value of $\alpha$, the ratio $\frac{g_{n,j}^\theta}{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter estimate.
 
@@
 
-$L^{B,\theta}$ should generally be preferred, since there is no reason to include the extra variability from resampling when calculating the conditional log likelihood, but it lacks the nice telescoping product.
+$L^{B,\theta}$ should generally be preferred in practice, since there is no reason to include the extra variability from resampling when calculating the conditional log likelihood, but it lacks the nice telescoping product that lets us derive exact expressions for the gradient later.
 
@@
         \lik(\theta)
-        &= \int_{x_{1:N}} \prod_{n=1}^N \frac{1}{J}\sum_{j=1}^J p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) p_\theta(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N \frac{1}{J}\sum_{j=1}^J p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \right) \frac{p_\theta(x_{1:N}|y_{1:N})}{p_\phi(x_{1:N}|y_{1:N})}p_\phi(x_{1:N}|y_{1:N}) dx_{1:N}\\
+        &= \int_{x_{1:N}} \prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) p_\theta(x_{1:N}|y_{1:N}) dx_{1:N} \\
+        &= \int_{x_{1:N}} \left(\prod_{n=1}^N  p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \right) \frac{p_\theta(x_{1:N}|y_{1:N})}{p_\phi(x_{1:N}|y_{1:N})}p_\phi(x_{1:N}|y_{1:N}) dx_{1:N}\\
         &= \int_{x_{1:N}} \left(\prod_{n=1}^N \lik_n(\theta) \right) \left(\prod_{n=1}^N \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \lik_n(\phi)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) \lik_n(\theta)}\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
@@
 
-Apply the concentration inequality in \ref{eq:ci} from Del Moral and Rio \cite{delmoral2011ci} and another union bound to find that
+Apply the Hoeffding-type concentration inequality from Del Moral and Rio \cite{delmoral2011ci} and another union bound to find that
 

### 2024-01-23 mop/draft.tex
lemma 2 edits

@@
     In particular, when $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter.
-In this case, its gradient when $\alpha=1$ is the same as the estimator of \citet{doucet2011sf},
-    \begin{equation}
-        \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*),
+In this case, its gradient when $\alpha=1$ is the same as the estimator of \citet{doucet2011sf} when applied to the bootstrap filter,
+    \begin{equation}
+        \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{0:N}}(y_{1:N}^* | x_{1:N,j}^{A, F,\theta}),
     \end{equation}
@@
 The proof for the above follows from Appendices A and C, as well as the technical lemmas for properly weighted particle filters.
+
+\kevin{Take derivative of loglik with regard to alpha?}
+\kevin{Talk about regime change the moment you move away from alpha=1, variance going from N2 to N, and bias going from 0 to something, but try to prove that decrease in variance outweights increase in bias.
+}
+\kevin{Create tools for choosing alpha via binary search CV from an averaged alpha=1 target}
+\kevin{suppose at each filtering step, look at gradient calculated at alpha=0, dynamically select alpha during filtering.
+calculate bias and variance online during filtering, and update alpha accordingly.
+}
+
+\kevin{No need to recompute so many particle filters! Just take an array of weight copies, one for each alpha. Gradients for each alpha come free the moment you compute one of them. }
+
+\kevin{Elbow method on variance, since bias is hard to compute? }
+\kevin{Get variance estimate, bias estimate, construct confidence intervals on bias estimate, use that to get unbiased estimate of the MSE. }
 
@@
     \label{lem:marginal-proper-weights}
-    Suppose that $\{(X_j^J,u_j^J),j=1,\dots,J\}$ targets $f_X$.
-Also suppose that $Z_j^J \sim f_{Z|X}(\cdot | X_j^J)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
-Then, if the $u_j^J$ are bounded, $\{(Z_j^J,u_j^J)\}$ targets $f_Z$.
+    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
+Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
+Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}$ targets $f_Z$.
 \end{lem}
 \begin{proof}
-
-    We want to show that
-     $$\frac{\sum_j h(Z_j^J) u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)] = \E_{f_X}[\E_{f_{Z|X}}[h(Z) | X]].$$
-
-    We know that by definition, for any measurable and bounded functional $g$ with domain $\gX$,
-    $$\frac{\sum_j g(X_j^J)u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} \E_{f_X}[g(X)].$$
-
-    Let $\bar{u}_j^J = \frac{J u_j^J}{\sum_j u_j^J}$.
-Examine the numerator and denominator of the quantity $$\frac{J^{-1}\sum_j h(Z_j^J) u_j^J}{J^{-1}\sum_j u_j^J} = \frac{J^{-1}\sum_j h(Z_j^J) \bar{u}_j^J}{J^{-1}\sum_j \bar{u}_j^J}.$$
-
+    We want to show that, for any measurable bounded $h$, 
+     $$\frac{\sum_j h(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[h(Z)] = \E\big[\E[h(Z) | X]\big].$$
+    By hypothesis, for any measurable and bounded functional $g$ with domain $\gX$,
+    \begin{equation}\label{eq:lemma2:g}
+    \frac{\sum_j g(\tilde X_j^J)\, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[g(X)].
+    \end{equation}
+    Let $\bar{U}_j^J = \frac{J U_j^J}{\sum_j U_j^J}$.
+Examine the numerator and denominator of the quantity $$\frac{J^{-1}\sum_j h(\tilde Z_j^J) U_j^J}{J^{-1}\sum_j U_j^J} = \frac{J^{-1}\sum_j h(\tilde Z_j^J) \bar{U}_j^J}{J^{-1}\sum_j \bar{U}_j^J}.$$
     The denominator converges to $1$ almost surely. The numerator, on the other hand, is
-    $$\frac{1}{J}\sum_j h(Z_j^J)\bar{u}_j^J,$$
-
-    which by the same fourth moment argument to the above lemma converges almost surely to the limit of its expectation
-    $$\lim_{J\to\infty} \E\left[\frac{1}{J}\sum_j h(Z_j^J)\bar{u}_j^J\right] = \lim_{J\to\infty} \E_{X_j^J}\left[\E_{Z_j^J|X_j^J}\left[\frac{1}{J}\sum_j  h(Z_j^J)\bar{u}_j^J\bigg|X_j^J\right]\right].$$
-
+    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\bar{U}_j^J,$$
+    by the same fourth moment argument to the above lemma converges almost surely to the limit of its expectation
+    \begin{eqnarray*}        
+    \lim_{J\to\infty} \E\left[\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J\right] 
+    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  
+      \E\left[ h(\tilde Z_j^J)\, \bar{U}_j^J\big|\tilde X_j^J, \bar U_j^J\right]
+    \right]
+    \\
+    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  \E\big[h(Z) \big| X=\tilde X_j^J \big] \, \bar U_j^J\right].
+    \end{eqnarray*}
+    Applying (\ref{eq:lemma2:g}) with $g(x) = \E\big[h(Z) | X=x\big]$, the average on the right hand size converges almost surely to $\E\big\{\E[h(Z)|X]\big\}=\E[h(Z)].$
     It remains to swap the limit and expectations. We can do so with the bounded convergence theorem, and therefore obtain   
-    $$\frac{1}{J}\sum_j h(Z_j^J)\bar{u}_j^J \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)] = \E_{f_X}[\E_{f_{Z|X}}[h(Z) | X]].$$
+    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\,\bar{U}_j^J \stackrel{a.s.}{\to} \E[h(Z)].$$
 
@@
     \end{equation}
-    successfully recovers $\lik(\theta)$. Therefore, DOP targets the posterior and the likelihood.
+    successfully recovers $\lik(\theta)$.
 \end{lem}

### 2024-01-27 mop/draft.tex
resolve overleaf latex warnings

@@
 
-\begin{algorithm}[H]
+\begin{algorithm}[ht]
 \centering
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \centering
@@
 
-\begin{figure}[H]
+\begin{figure}[ht]
     \centering
-    \includegraphics[scale=0.48]{imgs/095/boxplot.png}
-    \includegraphics[scale=0.53]{imgs/095/boxplotall.png}
+    \includegraphics[scale=0.48]{../imgs/095/boxplot.png}
+    \includegraphics[scale=0.53]{../imgs/095/boxplotall.png}
     \caption{\textbf{Left:} Boxplot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs.
@@
     \centering
-    \includegraphics[scale=0.7]{imgs/095/optim.png}
+    \includegraphics[scale=0.7]{../imgs/095/optim.png}
     \caption{Optimization progress of IFAD and IF2, where the solid lines depict the median negative log-likelihood at each iteration and the shaded area depicts the best search at any iteration and the 80\% percentile.
@@
     \centering
-    \includegraphics[scale=0.7]{imgs/095/hist.png}
+    \includegraphics[scale=0.7]{../imgs/095/hist.png}
     \caption{Histogram comparison of the performance of IFAD, IF2, and MOP alone.
@@
 The first follows from standard importance sampling arguments, the second from integrating out the marginal, and the third from Bayes' theorem. 
+We state Lemma~\ref{lem:change-measure-proper-weights} assuming multinomial resampling, which is convenient for the proof though other resampling strategies may be preferable in practice REFS.
 
@@
     \label{lem:change-measure-proper-weights}
-    Suppose that $\{(X_j^J,u_j^J),j=1,\dots,J\}$ targets $f_X$.
-Now, let $\{(Y_j^J,v_j^J),j=1,\dots,J\}$ be a sample drawn from $\{(X_j^J,u_j^J)\}$ where $(X_j^J,u_j^J)$ is represented, on average, $\pi_j J$ times.
-This could amount to multinomial resampling with probability $\pi_j$, or systematic resampling.
+    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
+Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, $\pi^J_j J$ times.
 Write
     \[
-    (Y_j^J,v_j^J) = \big(X_{a(j)},u_{a(j)}/\pi_{a(j)}\big),
+    (Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big),
     \]
-    where $a$ is called the ancestor function. If the importance sampling weights $u/\pi$ are bounded, then $\{(Y_j,v_j),j=1,\dots,J\}$ targets $f_X$.
+    where $a$ is called the ancestor function. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j),j=1,\dots,J\}$ targets $f_X$.
 \end{lem}
@@
     Note that as the $Y_j^J$ are a subsample from $X_j^J$, $h$ can be a function of $Y$ as well as it is one for $X$. We then expand
-    $$\frac{\sum_j h(Y_j^J) v_j^J}{\sum_j v_j^J} = \frac{\sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J}}.$$
-
-    We know that by definition,
-    $$\frac{\sum_j h(X_j^J)u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} \E_{f_X}[f(X)].$$
+    $$\frac{\sum_j h(Y_j^J) V_j^J}{\sum_j V_j^J} = \frac{\sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}}.$$
+    By hypothesis,
+    $$\frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[f(X)].$$
 
     We want to show
-    $$\frac{\sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J}} - \frac{\sum_j h(X_j^J)u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} 0.$$.
-
-    %To see this, observe that 
-    %$$\frac{\sum_j h(X_j^J)u_j^J}{\sum_j u_j^J} = \frac{\sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J}{\sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J},$$
-
-    %so it remains to show that
-    %$$\frac{\sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J}} 
-    %- \frac{\sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J}{\sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J}
-    %\stackrel{a.s.}{\to} 0.$$.
-
-    This is equivalent to showing that
-    $$\sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J \sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}
-    - \sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J} \sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J 
-    \stackrel{a.s.}{\to} 0.$$.
-
-    From $|a d- b c| \leq|d||a-c|+|c||b-d|$, we see that it suffices to bound
-
-    $$ \sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}
-    -  \sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J $$
-    and 
-    $$ \sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J
-    - \sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J}$$
-    separately.
-
-    %Write 
-    %$$\widetilde{\pi_{k_j}^J} := \frac{\frac{u_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J}}, \widetilde{u_{j}^J} := \frac{u_j^J}{\sum_j u_j^J}.$$
-
-    Write $g(X_j^J) = h(X_j^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}$. We therefore need to show that 
-    $$\sum_j Z_j^J := \sum_j \left(g(X_{k_j}^J) -  g(X_j^J) \pi_j^J \right) \stackrel{a.s.}{\to} 0.$$
+    \begin{equation}\label{eq:lemma1:h}
+    \frac{\sum_j h(X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}} - \frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} 0.
+    \end{equation}
+    For this, it is sufficient to show that
+   $$ \sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}
+    -  \sum_j h(\tilde X_j^J)\frac{U_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0 $$
+    since an application of this result with $h(x)=1$ provides almost sure convergence of the denominator in (\ref{eq:lemma1:h}).
+    Write $g(\tilde X_j^J) = h(\tilde X_j^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}$. We therefore need to show that 
+    $$\sum_j Z_j^J := \sum_j \left(g(\tilde X_{k_j}^J) -  g(\tilde X_j^J) \pi_j^J \right) \stackrel{a.s.}{\to} 0.$$
 
@@
 We can then follow the argument of \cite{chopin20} from this point on, where
+    {THE Zs ARE ONLY CONDITIONALLY INDEPENDENT, SO WE COULD PUT IN THE CONDITIONING EVENT (I.E., THE Xs AND Us) HERE TO BE MORE ACCURATE}
     $$\E\left[\left(\sum_j Z_j^J\right)^4\right] 
@@
     Let $\bar{U}_j^J = \frac{J U_j^J}{\sum_j U_j^J}$.
-Examine the numerator and denominator of the quantity $$\frac{J^{-1}\sum_j h(\tilde Z_j^J) U_j^J}{J^{-1}\sum_j U_j^J} = \frac{J^{-1}\sum_j h(\tilde Z_j^J) \bar{U}_j^J}{J^{-1}\sum_j \bar{U}_j^J}.$$
+Examine the numerator and denominator of the quantity $$\frac{J^{-1}\sum_j h(\tilde Z_j^J) \, U_j^J}{J^{-1}\sum_j U_j^J} = \frac{J^{-1}\sum_j h(\tilde Z_j^J) \, \bar{U}_j^J}{J^{-1}\sum_j \bar{U}_j^J}.$$
     The denominator converges to $1$ almost surely. The numerator, on the other hand, is
-    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\bar{U}_j^J,$$
+    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J,$$
     by the same fourth moment argument to the above lemma converges almost surely to the limit of its expectation
@@
 \begin{proof}
+
+    $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J}}{J^{-1} \sum_j \bar{u}_j^{' J}}$$
+
+    by
+
+    \begin{equation}
+    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}}
+    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \right)^{-1}
+    \end{equation}
+
     Again, we want to show that
@@
     %= \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j f_{Z|X}(z^*|X_j^J)f_X(X_j^J)}
+
+    Consider the following:
+    \begin{equation}
+    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}}
+    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \right)^{-1}.
+    \end{equation}
 
@@
             $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
-            \label{dop-alpha:Lphi}
+            \label{mop-alpha:Lphi}
             \STATE Normalize weights:
@@
             for $j\ \mathrm{in}\ \seq{1}{J}$
-            \STATE Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \label{dop-alpha:systematic}
+            \STATE Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \label{mop-alpha:systematic}
             \STATE Resample particles: ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$
             \STATE Filter weights corrected for resampling:
-            $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{dop-alpha:weight:update}
+            $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{mop-alpha:weight:update}
             \STATE Resample filter weights:
             $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
-            for $j$ in $\seq{1}{J}$ \label{dop-alpha:step2}
+            for $j$ in $\seq{1}{J}$ \label{mop-alpha:step2}
             \STATE After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
@@
 
+\kevin{Show bias is order N, variance is order N when alpha below 1. That implies optimal choice of alpha is not 1. }
+
 \section{Finite-Particle Analysis of Gradient Methods with Algorithm \ref{alg:ifad}}
@@
 \subsection{Bounding the Gradient}
+\kevin{Add a few words to explain that the gradient of the log-likelihood w.r.t.
+parameters is a functional of the particle measure under boundedness assumptions and fixed parameters, appeal to authority and say doucet did it too.
+}
+
+\kevin{Gradient linear map, bound conditional likelihoods at time n.
+Alternatively, convert del moral and doucet expectation bound to high-probability bound.
+https://arxiv.org/pdf/1905.11546.pdf for trick.
+Alternatively, concatenate previous states.
+}
+
+\kevin{Show MOP-1 is unbiased. Extend analysis to not-1 case. }
 
@@
     \label{lemma:grad_bound}
-    Assume $\text{osc} \left(\frac{\partial}{\partial_\theta}\right) \leq 1$.
-For $||\nabla_\theta \hat{\loglik}(\theta) - \nabla_\theta \loglik(\theta)||$ to be bounded by $\epsilon$ with probability $1-\delta$, we require
+    Assume $\text{osc} \left(\frac{\partial}{\partial_\theta} \hat{\ell}_\theta(x_{n,j}^P) \right) \leq 1$.
+For $||\nabla_\theta \hat{\loglik}_\theta(x_{n,j}^P) - \nabla_\theta \loglik_\theta(x^P)|| = ||\nabla_\theta \hat{\loglik}(\theta) - \nabla_\theta \loglik(\theta)||$ to be bounded by $\epsilon$ with probability $1-\delta$, we require
     \begin{equation}
@@
 This amounts to solving the following problem.
-Say you have a weighted mean $\sum w_j x_j$, where $j=1,…,J$.
+Say you have a weighted mean $\sum w_j x_j$, where $j=1,\dots,J$.
 The $x_j$ are fixed numbers bounded by some constant $G$ and the $w_j$ are non-negative and sum to $1$.
-Resample indices $k_j, j=1,…,J$ according to the multinomial distribution to get $x_{k_j}$ for each $j$, and also $J^{-1} \sum x_{k_j}$.
+Resample indices $k_j, j=1,\dots,J$ according to the multinomial distribution to get $x_{k_j}$ for each $j$, and also $J^{-1} \sum x_{k_j}$.
 We want to bound $|\sum w_j x_j - J^{-1} \sum x_{k_j}|$.

### 2024-03-05 mop/experiment.tex
updating on GitHub

--- 2024-01-27 mop/draft.tex
@@
 While the Monte-Carlo Adjusted Profile (MCAP) from \citet{Ionides_mcap} provides a profile likelihood confidence interval that takes into account both Monte Carlo profile error and statistical uncertainty in the likelihood, it is computationally inefficient.
-\kevin{TODO} We provide a method for incorporating the MCAP methodology (with the poor man's likelihood profile) in our IFAD algorithm that provides a more computationally efficient profile likelihood confidence interval.
-\kevin{Quantile regression for estimating the max/the envelope of the likelihood evaluations from AD, then fit MCAP/MQLE/MSLE?}
+We provide a method for incorporating the MCAP methodology (with the poor man's likelihood profile) in our IFAD algorithm that provides a more computationally efficient profile likelihood confidence interval.
 
@@
 
-\kevin{TODO: (1) Fix beta smoothness to d(n) at no more than O(n), (2) add story to say that no one spotted that it could have been plug-and-play because of e.g. reasons why they would stumble}
-
 \section{Problem Setup}
@@
 The above intuition is formalized below.
-
-\kevin{Method of Common Random Numbers, Coupling}
 
@@
 The likelihood only changes by a factor of the likelihood ratios, which are bounded and smooth in $\theta$ if the densities used in the calculation are.
+For DOP, you need smooth likelihood ratios for the process model.
+For MOP, you need $k$ continuous derivatives from the reparametrization trick.
 
@@
 
+We need the following assumption to ensure that the weighted particle filter with off-policy resampling targets the posterior:
+\begin{aspt}[Bounded And Differentiable Measurement Density Ratios]
+    For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{Y_n|X_n}(y_n^*|x, \theta)}{f_{Y_n|X_n}(y_n^*|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x \in \gX$ and all $\theta\in\gN(\phi)$.
+\end{aspt}
+
+In the following assumption, the first condition is sufficient for DOP, and the second for MOP:
+\begin{aspt}[Bounded And Differentiable Process Model]
+    Either one of the following two conditions hold:
+    \begin{enumerate}
+        \item For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{X_n|X_{n-1}}(x'|x, \theta)}{f_{X_n|X_{n-1}}(x'|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x, x' \in \gX$ and all $\theta\in\gN(\phi)$, or
+        \item There exists a differentiable simulator with more than two continuous derivatives for the process model. 
+    \end{enumerate}
+\end{aspt}
+
 \subsubsection{Doubly Off-Policy Particle Filter}
-
-\kevin{rewrite to de-emphasize dop as in presentation, sell mop as the big theoretical idea that explains AD through PFs and unifies two previous estimators}
 
@@
 as we illustrate in Appendix \ref{app:dop}. We use the notation $x_{n,j}^{P, \phi}$ to emphasize that the particles evolve according to $\phi$.
-
-We claim that this yields a particle filter that targets the posterior and recovers the derivative estimate of \citet{doucet2011sf, scibior2021dpf} when $\theta$ is evaluated at $\phi$.
-The proof can be found in Appendix \ref{app:dop}.
-
-\begin{prop}[Correctness of Doubly Off-Policy Particle Filter Corrections]
-    \label{prop:dop-correctness}
-
-    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$.
-Also let $\theta \in \gN(\phi)$.
-Then, the following corrections in the particle weights,
-    \begin{equation}
-    s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}, \qquad r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)},
-    \end{equation}
-    yield a particle filter that is unbiased for the likelihood, even when $\theta \neq \phi$.
-
-    In particular, when $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{doucet2011sf},
-    \begin{equation}
-        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
-    \end{equation}
-\end{prop}
-
-\kevin{Cut down on DOP discussion in actual manuscript, remove any references to this as an algorithm, just talk about the idea of an off-policy system/particle filter.
-Bottom paragraph is the punchline.}
 
@@
 
-\kevin{Develop theory for mixing-time swapping?
-Detect when IF2 reaches a neighborhood of the MLE, by detecting whether it has reached a local sort of stationary distribution when the perturbations are lower bounded, e.g.
-the autocorrelation between the mean vectors to estimate the mixing time or Gelman-Rubin.
-In practice, just swap when IF2 converges according to the diagnostics.
-Theory doesn't care, you just need to in a basin of attraction that is strongly convex.}
-
 \subsubsection{Discounting Weights}
+
+\kevin{Turn all theorems and lemmas into remarks, one big section on results we think may be true but leave to future work to prove. }
+
+\kevin{Release arxiv version concurrently with an initial release of the package. Target April release, grant results for DMS come back then. }
+
+\kevin{For a software paper, JOSS, signal that the software is careful. JSS is higher impact, but difficult to get a paper in. }
 
@@
 Note that no matter what value $\alpha$ takes on, the system evolves according to $\phi$ when we perform the correction to evaluate the likelihood at $\theta$.
-The filter fully targets the posterior when $\theta=\phi$, or $\alpha=1$, as we show in Proposition \ref{prop:dop-correctness}.
+The filter fully targets the posterior when $\theta=\phi$, or $\alpha=1$.
 As previously mentioned, the value of $\alpha$ therefore informs a bias-variance tradeoff.
@@
 
-\kevin{This isn't a smoother, because we'd need future information.}
-
 \paragraph{When $\alpha=1,$} the likelihood estimate becomes
@@
     In particular, when $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter.
-In this case, its gradient when $\alpha=1$ is the same as the estimator of \citet{doucet2011sf} when applied to the bootstrap filter,
+In this case, its gradient when $\alpha=1$ is the same as the DPF estimator of \citet{scibior21} when applied to the bootstrap filter,
     \begin{equation}
@@
 
-    that as \cite{scibior2021dpf} show corresponds to differentiating through the fixed-seed likelihood estimate yielded by a vanilla particle filter.
+    that as \cite{scibior21} show corresponds to differentiating through the fixed-seed likelihood estimate yielded by a vanilla particle filter, dropping resampling terms.
+
+    When $\alpha=1$, the gradient estimate MOP-1 provides is consistent for the score.
 \end{prop}
@@
 
-\kevin{Take derivative of loglik with regard to alpha?}
-\kevin{Talk about regime change the moment you move away from alpha=1, variance going from N2 to N, and bias going from 0 to something, but try to prove that decrease in variance outweights increase in bias.
+% \kevin{Take derivative of loglik with regard to alpha?}
+% \kevin{Talk about regime change the moment you move away from alpha=1, variance going from N2 to N, and bias going from 0 to something, but try to prove that decrease in variance outweights increase in bias.
 }
-\kevin{Create tools for choosing alpha via binary search CV from an averaged alpha=1 target}
-\kevin{suppose at each filtering step, look at gradient calculated at alpha=0, dynamically select alpha during filtering.
+% \kevin{Create tools for choosing alpha via binary search CV from an averaged alpha=1 target}
+% \kevin{suppose at each filtering step, look at gradient calculated at alpha=0, dynamically select alpha during filtering.
 calculate bias and variance online during filtering, and update alpha accordingly.
@@
 
-\kevin{No need to recompute so many particle filters! Just take an array of weight copies, one for each alpha. Gradients for each alpha come free the moment you compute one of them. }
-
-\kevin{Elbow method on variance, since bias is hard to compute? }
-\kevin{Get variance estimate, bias estimate, construct confidence intervals on bias estimate, use that to get unbiased estimate of the MSE. }
+% \kevin{No need to recompute so many particle filters! Just take an array of weight copies, one for each alpha. Gradients for each alpha come free the moment you compute one of them. }
+
+% \kevin{Elbow method on variance, since bias is hard to compute? }
+% \kevin{Get variance estimate, bias estimate, construct confidence intervals on bias estimate, use that to get unbiased estimate of the MSE. }
 
@@
 
-\begin{thm}[Convergence of Gradient Methods]
+\begin{conj}[Convergence of Gradient Methods]
 \label{thm:convergence}
@@
 and the algorithm eventually terminates after we obtain $||\nabla_\theta \ell(\theta_n)|| \leq (1+\sigma) \epsilon$.
-\end{thm}
+\end{conj}
 
@@
 To compare the performance of IFAD against that of IF2, we fit the very same model first proposed in \cite{king08} that \cite{ionides15} use to demonstrate the capabilities of IF2.
+This is becoming a standard benchmark, as \cite{wood16} and \cite{wycoff2024voronoi} include in their papers.
 This is a stochastic SIR compartmental model with transition and measurement uncertainty, where the population at time $t$, $H(t)$, is divided into the susceptible compartment $S(t)$, infected compartment $I(t)$, and three ($k=3$) recovered compartments $R_1(t), ..., R_k(t)$ denoting varying degrees of cholera immunity.
@@
 
+\kevin{Rewrite this para}
 \paragraph{Differences Between IFAD-0 and IFAD-1:} The differences between IFAD-1 and IFAD-0 (MOP-1 and MOP-0) do not seem very pronounced in this example.
@@
 
-\begin{lem}[Change of Particle Measure]
+\begin{conj}[Change of Particle Measure]
     \label{lem:change-measure-proper-weights}
@@
     where $a$ is called the ancestor function. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j),j=1,\dots,J\}$ targets $f_X$.
-\end{lem}
+\end{conj}
 
@@
 
-\begin{lem}[Particle Marginals]
+\begin{conj}[Particle Marginals]
     \label{lem:marginal-proper-weights}
@@
 Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}$ targets $f_Z$.
-\end{lem}
+\end{conj}
 \begin{proof}
@@
 
-\begin{lem}[Particle Posteriors]
+\begin{conj}[Particle Posteriors]
     \label{lem:posterior-proper-weights}
@@
 Then, if $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,u^{\prime J}_j)\}$ targets $f_{X|Z}(\cdot | z^*)$.
-\end{lem}
+\end{conj}
 
@@
     When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows.
-Recursively applying Lemmas \ref{lem:change-measure-proper-weights} \ref{lem:marginal-proper-weights}, and \ref{lem:posterior-proper-weights}, we obtain that
+Recursively applying Lemmas \ref{lem:change-measure-proper-weights}, \ref{lem:marginal-proper-weights}, and \ref{lem:posterior-proper-weights}, we obtain that
     %to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
@@
 
-\begin{lem}[Likelihood Proper Weighting]
+\begin{conj}[Likelihood Proper Weighting]
     \label{lem:lik-proper-weight}
@@
 where $L^\phi_n$ is as defined in the various algorithms.
-\end{lem}
-
-Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ is properly weighted for $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+\end{conj}
+
+Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
 To see  (\ref{L2}),
@@
 
+\kevin{IS THIS UNBIASED?}
+
 $L^{B,\theta}$ should generally be preferred in practice, since there is no reason to include the extra variability from resampling when calculating the conditional log likelihood, but it lacks the nice telescoping product that lets us derive exact expressions for the gradient later.
@@
 
-The proof for Proposition \ref{prop:dop-correctness} follows from the below two lemmas.
-Evaluating the off-policy likelihood is possible with a weighted particle filter by using resampling weight $p(y^*_n|x_n;\phi)$ and making two corrections in the particle weight, through multiplying $w_{t,j}$ by the two likelihood ratios below,
-
-\begin{equation}
-    \label{eq:dmeas-ratio}
-    s=p(y^*_n|x_n;\theta)/p(y^*_n|x_n;\phi)
-\end{equation}
-and
-\begin{equation}
-    \label{eq:rproc-ratio}
-    r=p(x_n|x_{n-1};\theta)/p(x_n|x_{n-1};\phi).
-\end{equation}
-
-The following result showcases the correctness of corrections \ref{eq:dmeas-ratio} and \ref{eq:rproc-ratio}.
-
-\begin{lem}
-    For $\theta$, $\phi$ that satisfy Assumption \ref{assump:smooth-nbhd}, the following correction in the particle filter weights,
-    \begin{equation}
-        w_{t,j} := \tilde{w}_{t,j} \cdot p(y^*|x_{t,j}^P,\phi)\cdot r \cdot s
-    \end{equation}
-    successfully recovers $\lik(\theta)$.
-\end{lem}
-
-\begin{proof}
-    Recall $\lik(\theta) = \E[\lik(\theta,\phi,\Omega, J)]=\int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega)$ and $\lik(\theta) = \int_\Omega \lik(\theta, \omega, J) d\mu(\omega)$.
-We want to evaluate
-    \begin{align*}
-        \lik(\theta,\omega, J)
-        &= \E\left[ \prod_{t=1}^T \lik_t(\theta) \middle| \omega\right]
-        = \E\left[ \prod_{t=1}^T p(y_t^*|y_{1:t-1}^*,\theta) \middle| \omega\right] \\
-        &= \E\left[ \prod_{t=1}^T \frac{1}{J} \sum_{j=1}^J  p(y_t^*|x_{t,j}^P,\theta) p(x_{t,j}^P|y_{1:t-1}^*,\theta) \middle| \omega\right] \\
-        &= \E\left[ \prod_{t=1}^T\frac{1}{J}  \sum_{j=1}^J  p(y_t^*|x_{t,j}^P,\theta) p(x_{t,j}^P|x_{t-1}^F,\theta) \middle| \omega\right] 
-    \end{align*}
-    where $p(x_t^P|y_{1:t-1}^*,\theta)$ is the prediction distribution of particles, and the filtering distribution of particles is proportional to $p(y_t^*|x_{t,j}^P,\theta)$, the likelihood of particle $j$ at time $t$.
-
-    Consider first the case with no resampling.
-Let $\lik^j(\theta)$, $\lik^j(\phi)$ be the likelihood of trajectory $j$ under $\theta$ and $\phi$ respectively.
-As the prediction and filtering particles are then the same, we use $x_{t,j}$ to refer to both.
-As such, conditional on $\omega$,
-    \begin{align*}
-        \lik^j(\theta) 
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \theta) p(x_{t,j}|x_{t-1,j}, \theta) \\
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \frac{p(y_t^*|x_{t,j}, \theta)}{p(y_t^*|x_{t,j}, \phi)} p(x_{t,j}|x_{t-1,j}, \theta) \frac{p(x_{t,j}|x_{t-1,j}, \theta) }{p(x_{t,j}|x_{t-1,j}, \phi) } \\
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \cdot s \cdot p(x_{t,j}|x_{t-1,j}, \phi)\cdot r
-    \end{align*}
-
-    and therefore, if trajectories were proposed with equal probability with particles conditional on $\omega$,
-    $$ \lik(\theta, \omega, J) = \frac{1}{J} \sum_{j=1}^J \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \cdot s \cdot p(x_{t,j}|x_{t-1,j}, \phi)\cdot r $$
-    The correction is therefore sufficient for recovering $\lik(\theta)$, with trajectories collected with the process model at $\phi$ and measurements evaluated with the measurement model at $\phi$.
-
-    Consider now the case with resampling.
-Without loss of generality we can resample every timestep.
-Here, $x_{t,j}^F(\phi) \sim p(y^*|x_{t,j}^P, \phi)$ and $x_{t,j}^P(\phi) \sim p(x_t|x_{t-1}^F,\phi)$, and similarly for $\theta$.
-We use this notation to emphasize that \textbf{both} the measurement model and the previously drawn particles depend on $\phi$ or $\theta$.
-
-    \begin{align*}
-        \lik_t(\theta) &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \theta)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\theta) \right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \theta)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \frac{p(y_t^*|x_{t,j}^P,\theta)}{p(y_t^*|x_{t,j}^P,\phi)} \right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \frac{p(y_t^*|x_{t,j}^P,\theta)}{p(y_t^*|x_{t,j}^P,\phi)} \frac{p(x_{t,j}|x_{t-1,j}, \theta) }{p(x_{t,j}|x_{t-1,j}, \phi) }\right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \cdot s \cdot r \right] \\
-        &= \E_{\omega}\E_{x_{t,j}^P \sim \process(x_{t-1}^F, \omega, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \cdot s \cdot r \middle| \omega \right]
-    \end{align*}
-
-    and again the correction is sufficient.
-
-    Therefore,
-
-    \begin{align}
-         \mathcal{L}(\theta) 
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^P(\omega), \theta) p(x_{n,j}^P(\omega) | x_{n-1,j}^P(\omega), \theta) \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\theta}(\omega), \theta) p(x_{n,j}^{P,\theta}(\omega) | x_{n,j}^P(\omega), \theta) \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\theta}(\omega), \phi) p(x_{n,j}^{P,\theta}(\omega) | y_n^*, \phi) \cdot s  \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) p(x_{n,j}^{P,\phi}(\omega) | y_n^*, \phi) \cdot r \cdot s \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) p(x_{n,j}^{P, \phi}(\omega) | y_n^*, \phi) \cdot r \cdot s \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) \cdot r \cdot s \right)  p(x_{1:N} | y_{1:N}^*, \phi) \,  d\mu(\omega) \\
-         &=: \int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega).
-     \end{align}
-
-     The proof for an arbitrary functional under the particle measure follows analogously.
-\end{proof}
-
-\begin{lem}[Gradient of Doubly Off-Policy Particle Filter]
-    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$.
-Also let $\theta \in \gN(\phi)$.
-When $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{doucet2011sf},
-    \begin{equation}
-        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
-    \end{equation}
-\end{lem}
-
-\begin{proof}
-    It is known, from \cite{doucet2009tutorial}, that 
-    $$p_\theta(x_{1:N}|y_{1:N}) = \prod_{n=1}^N \frac{p(x_n|x_{n-1})p(y_n|x_n)}{p(y_n|y_{1:n-1})} = \prod_{n=1}^N \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{\lik_n(\theta)}.$$
-
-    From the above lemma, and the fact that the particle filter is unbiased for the likelihood, we know that
-    \begin{align*}
-        \lik(\theta)
-        &= \int_{x_{1:N}} \prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) p_\theta(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N  p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \right) \frac{p_\theta(x_{1:N}|y_{1:N})}{p_\phi(x_{1:N}|y_{1:N})}p_\phi(x_{1:N}|y_{1:N}) dx_{1:N}\\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N \lik_n(\theta) \right) \left(\prod_{n=1}^N \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \lik_n(\phi)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) \lik_n(\theta)}\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N \lik_n(\phi) \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) }\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N}.
-    \end{align*}
-
-    Therefore, under regularity conditions that permit differentiating under the integral sign, using the log-derivative trick, we have
-    \begin{align*}
-        \nabla_\theta \lik(\theta)
-        &= \int_{x_{1:N}} \nabla_\theta \hat\lik(\theta) \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N}\\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N \lik_n(\phi) \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) }\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N \left(\frac{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n)}{\lik_n(\phi)}\right)^{-1} p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\frac{1}{p_\phi(x_{1:N}|y_{1:N}) }\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\nabla_\theta \log\left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} p_\theta(x_{1:N},y_{1:N}) \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) p_\theta(x_{1:N}|y_{1:N}) \cdot \lik(\theta) \; dx_{1:N}. 
-    \end{align*}
-
-    Using the log-derivative trick again,
-    \begin{align*}
-        \nabla_\theta \ell(\theta)
-        &= \frac{\nabla_\theta \lik(\theta)}{\lik(\theta)} \\
-        &= \int_{x_{1:N}} \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) p_\theta(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \E_{x_{1:N}|y_{1:N}, \theta}\left[\nabla_\theta \left(\sum_{n=1}^N\log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right)\right] \\
-        &= \E_{x_{1:N}|y_{1:N}, \theta}\left[\nabla_\theta\log p_\theta(x_{1:N},y_{1:N})\right] \\
-        &= \lim_{J\to\infty} \E_{x_{1:N}|y_{1:N}, \theta}\left[\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log p_\theta(x_{1:N,j}^{A,F},y_{1:N}^*)\right].
-    \end{align*}
-
-    But it also holds that
-    \begin{align*}
-        \nabla_\theta \ell(\theta)
-        &= \frac{\nabla_\theta \lik(\theta)}{\lik(\theta)} \\
-        &= \int_{x_{1:N}} \frac{\nabla_\theta \hat\lik(\theta)}{\lik(\theta)} \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \hat\ell(\theta) \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N}\\
-        &= \lim_{J\to\infty} \E_{x_{1:N}|y_{1:N}, \phi}\left[\nabla_\theta \hat\ell(\theta)\right].
-    \end{align*}
-
-    and we see that $$\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log p_\theta(x_{1:N,j}^{A,F},y_{1:N}^*)$$ and $$\nabla_\theta \hat\ell(\theta)$$ both converge to the same thing in expectation.
-They are both particle approximations under the posterior where the variance of each is bounded, therefore one can apply the particle CLT from \cite{chopin2004clt} to obtain that they are both consistent estimates of the score.
-
-    In fact, when $\theta=\phi,$ they are the same in distribution. \kevin{claim}
-
-    \kevin{This is a far weaker result than what I really wanted. This just gives consistency. Want equality in distribution.}
-\end{proof}
-
-\begin{lem}
-
-Consider the case of DOP-$\alpha$ when $\alpha=0$ and $\theta=\phi$. The gradient estimate is then
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta\right),
-    \end{equation}
-
-    yielding the gradient estimator of \cite{blei2018vsmc} that \cite{scibior2021dpf} show is the gradient of a vanilla particle filter when the resampling terms are dropped.
-
-\end{lem}
-
-\begin{proof}
-    When $\alpha=0,$ the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J r_{n,j}s_{n,j}}{\sum_{j=1}^J r_{n,j}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J r_{n,j}s_{n,j}.
-\end{equation}
-
-Its gradient when $\theta=\phi$ is therefore 
-\begin{align*}
-    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right) \\
-    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right)} \\
-    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta r_{n,j}s_{n,j}}{\sum_{j=1}^J r_{n,j}s_{n,j}} \\
-    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{Y_n, X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)} \\
-    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta\right),
-\end{align*}
-where we use the log-derivative trick, observe that $\sum_{j=1}^J r_{n,j}s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again. 
-\end{proof}
-
-\subsection{Off-Policy Weighted Differentiable Particle Filter}
-
-We are finally ready to introduce the off-policy weighted differentiable particle filter.
+We present the doubly off-policy differentiable particle filter for completeness. 
+Here, we consider a doubly off policy particle filter, meaning that both \code{rprocess} and \code{dmeasure} are computed at $\phi$ and particles are reweighted to correspond to $\theta$.
 In the event where a seed is fixed, all expressions in Algorithm \ref{alg:dop} can be considered to be conditional on $\omega \in \Omega$.
@@
 
-Here, we consider a doubly off policy particle filter, meaning that both \code{rprocess} and \code{dmeasure} are computed at $\phi$ and particles are reweighted to correspond to $\theta$.
-
 \section{Measurement Off-Policy Particle Filters}
+
+Here, we consider an off-policy particle filter where \code{dmeasure} are computed at $\phi$ and we leverage the reparametrization trick to obtain derivatives with regard to the process model.
+In the event where a seed is fixed, all expressions in Algorithm \ref{alg:mop} can be considered to be conditional on $\omega \in \Omega$.
 
@@
 
-\begin{lem}
+\begin{conj}
     Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The gradient estimate is then
@@
     yielding the gradient estimators of \cite{poyiadjis11, scibior2021dpf} when applied to the bootstrap filter. 
-\end{lem}
+\end{conj}
 
@@
 
-$$w_{n,j}^{P,\theta} = w_{n-1,\theta}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,k_j}^\theta}{g_{i,k_j}^\phi}.$$
+$$w_{n,j}^{P,\theta} = w_{n-1,\theta}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,j}^{A,\theta}}{g_{i,j}^{A,\phi}},$$
+where we use the $(\cdot)^A$ superscript to denote the ancestral trajectory of the $j$-th particle at timestep $n$.
 
 Using the log-derivative identity again,
-$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,k_j}^\theta}{g_{i,k_j}^\phi}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,k_j}^\theta - \log g_{i,k_j}^\phi\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,k_j}^\theta.$$
+$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,j}^{A,\theta}}{g_{i,j}^{A,\phi}}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,j}^{A,\theta} - \log g_{i,j}^{A,\phi}\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,j}^{A,\theta}.$$
 
 So 
-$$  \nabla_\theta \sum_{n=1}^N \log g_{n,k_j}^\theta = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,k_j}^\theta\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,k_j}^{P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right),$$
+$$  \nabla_\theta \sum_{n=1}^N \log g_{n,j}^{A,\theta} = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,j}^{A,\theta}\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,j}^{A, P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right),$$
 
@@
 
-\begin{lem}
+\begin{conj}
 
@@
 
-\end{lem}
+\end{conj}
 
@@
 
+\kevin{I have no clue how to prove this. Someone, anyone, help!}
+\begin{conj}[MOP-1 Estimate Is Unbiased for the Likelihood and the Gradient of the Likelihood]
+    Consider the case of MOP-$\alpha$ when $\alpha=1$, and resampling is carried out with either multinomial, residual, or systematic resampling at every timestep.
+Assume the gradient estimate obtained by the gradient of the likelihood estimate with J particles, $\nabla_\theta \hat\lik_J(\theta)$ is uniformly bounded for all $J$, and that the likelihood has two continuous derivatives.
+Then the likelihood estimate is unbiased for the likelihood and the gradient of the likelihood estimate is unbiased for the gradient of the likelihood.
+That is, it holds that
+    $$\E[\hat\lik_J(\theta)] = \lik(\theta),\;\; \E[\nabla_\theta \hat\lik_J(\theta)] = \nabla_\theta \lik(\theta).$$
+\end{conj}
+
+\begin{proof}
+    First, we need to establish the unbiasedness of the likelihood estimate. This is 
+    $$\hat\lik_J(\theta) := \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right)\prod_{n=1}^N L_n^\phi = \left(\frac{1}{J}\sum_{j=1}^J \prod_{n=1}^N \frac{g_{n,j}^{A,\theta}}{g_{n,j}^{A,\phi}} \right)\prod_{n=1}^N L_n^\phi.$$
+
+    We know from \cite{chopin20} that $\E[L_n^{\phi}] = \lik(\phi).$ It then suffices to show that
+    $$\E\left[\left(\frac{1}{J}\sum_{j=1}^J \prod_{n=1}^N \frac{g_{n,j}^{A,\theta}}{g_{n,j}^{A,\phi}} \right)\right] = \frac{\lik(\theta)}{\lik(\phi)}.$$
+
+    To obtain this, we need to see that the resampling under $\phi$, with corrected weights, is unbiased. That is, we need to show that for any bounded measurable functional $\varphi : \gX \to \R$,
+    $$\E\left[\frac{\sum_{j=1}^J w_{n,j}^{F, \theta} \varphi(x_{n,j}^{\theta, F})}{\sum_{j=1}^J w_{n,j}^{F, \theta}} \bigg| \gF_n \right] = \frac{\sum_{j=1}^J w_{n,j}^{P, \theta} \varphi(x_{n,j}^{P, \theta})}{\sum_{j=1}^J w_{n,j}^{P, \theta} }.
+$$
+
+    To see this, 
+    \begin{align*}
+        \sum_{j=1}^J w_{n,j}^{P, \theta} \varphi(x_{n,j}^{P, \theta}) 
+        &= \sum_{j=1}^J w_{n-1,j}^{F, \theta}\frac{g_{n,j}^{\theta}}{g_{n,j}^{\phi}}  \varphi(x_{n,j}^{P, \theta})
+    \end{align*}
+
+    Observe that our differentiable simulator $\process(\cdot, \theta)$ and choice of weights now satisfy:
+
+    - A Feynman-Kac model $\left\{M_t, G_t\right\}$ such that:
+
+- the weight function $G_t$ may be evaluated pointwise (for all $t$ );
+
+- it is possible to simulate from $\mathbb{M}_0\left(\mathrm{~d} x_0\right)$ and from $M_t\left(x_{t-1}, \mathrm{~d} x_t\right)$ (for any $x_{t-1}$ and $t$ ).
+
+- The number of particles $N$.
+
+- The choice of an unbiased resampling scheme (e.g.
+multinomial, residual, stratified, systematic), that is, an algorithm to generate variables $A_t^{1: N}$ in $1: N$ such that $\mathbb{E}\left[\sum_{m=1}^N \mathbbm{1}\left\{A_t^m=n\right\}\right]=N W_{t-1}^n$; see previous chapter for more details.
+\end{proof}
+
+\begin{conj}[Consistency of MOP-1 Gradient Estimate]
+    Consider the case of MOP-$\alpha$ when $\alpha=1$.
+Assume the gradient estimate obtained by the gradient of the log-likelihood estimate with J particles, $\nabla_\theta \hat\ell_J(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}$ is uniformly bounded by $G(\theta)$ for all $J$, and that the likelihood has two continuous derivatives.
+Then the gradient estimate, $\nabla_\theta \hat\ell_J(\theta)$, is consistent for the score.
+\end{conj}
+\begin{proof}
+
+    We know that the variance of the MOP-1 gradient estimate $\nabla_\theta \hat\ell_J(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}$ goes to zero coordinate-wise as $J\to\infty$ by a particle filter law of large numbers for bounded functionals of particles (\kevin{Prove general LLN for functionals of MOP under boundedness assumptions}), so it suffices to show that it is asymptotically unbiased.
+
+    Write $\hat\pi_{n,J}$ for the filtering distribution at timestep $n$ and with $J$ particles, $\hat\pi_{1:N,J}$ for the sequence of filtering distributions from timesteps $1$ to $N$, and $\pi_n, \pi_{1:N}$ for the respective posteriors.
+
+    From Lemma \ref{lem:lik-proper-weight}, we know that $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1:n-1}^*, \theta)$ is consistently estimated by the conditional likelihood estimate obtained by MOP-1, for all $\theta$ in an open neighborhood of $\phi$.
+The likelihood is simply the product of the conditional likelihoods, and similarly for the estimates, so the likelihood $\lik(\theta)$ is also consistently estimated by the likelihood estimate obtained by MOP-1, $\hat\lik_J(\theta)$, for all $\theta$ in an open neighborhood of $\phi$, $\gN(\phi)$.
+
+    We therefore have that 
+    $$\lim_{J\to\infty} \E_{\hat\pi_{1:N,J}}[\hat\lik_J(\theta)] = \lik(\theta)$$
+    uniformly for all $\theta \in \gN(\phi)$. 
+    \kevin{Formalize this.}
+
+    \kevin{Need to show that gradients converge almost surely uniformly (doesn't this follow directly from boundedness?) to swap limits, according to Terry Tao.}
+
+    As the gradient is bounded uniformly for all $J$ and consistency for the likelihood estimate holds for all $\theta$ in an open neighborhood of $\phi$ (and therefore all $\theta'$ in a sufficiently small neighborhood of $\theta$), we can swap the derivative and limit by the bounded convergence theorem, and then the derivative and expectation by the very same, to find that
+    $$\nabla_\theta\lik(\theta) 
+    = \nabla_\theta \lim_{J\to\infty} \E_{\hat\pi_{1:N,J}}[\hat\lik_J(\theta)] 
+    = \lim_{J\to\infty} \nabla_\theta \E_{\hat\pi_{1:N,J}}[\hat\lik_J(\theta)] 
+    = \lim_{J\to\infty} \E_{\hat\pi_{1:N,J}}[\nabla_\theta \hat\lik_J(\theta)].$$
+
+    We then have mean-square consistency for the gradient of the likelihood estimate. The result then follows by the continuous mapping theorem.
+
+    \kevin{Explain that the difference in formula between this and Poyiadjis is due to us taking the derivative w.r.t. the particles in the chain rule. }
+
+    \kevin{Alternative argument: Show gradient is unbiased, after showing the likelihood is unbiased in an open neighborhood of phi.
+That should hold by Chopin's argument for unbiased resampling to nonconstant weights.
+Gradient unbiased if can swap expectation and derivative, I think with two continuous derivatives and boundedness, might need BCT/DCT.
+No need to swap limit and derivative in this case, or worrying about uniform limits.}
+
+\end{proof}
+
 \kevin{Show bias is order N, variance is order N when alpha below 1. That implies optimal choice of alpha is not 1. }
@@
 
-\kevin{Show MOP-1 is unbiased. Extend analysis to not-1 case. }
-
-\begin{lem}[Concentration of Measure for Gradient Estimate]
+\begin{conj}[Concentration of Measure for Gradient Estimate]
     \label{lemma:grad_bound}
@@
     where $G(\theta)$ is defined as in Assumption \ref{assump:local-bounded-derivative}, $h(t) = \frac{1}{2}(t - \log(1+t))$, and $\beta_t$ and $r_t$ are two additional model-specific parameters that do not depend on $J$.
-\end{lem}
+\end{conj}
 
@@
 To use this, we will first need to resample the particles once more to reassign the normalized weights to $1/J$.
+
+\kevin{Put some justification/make the setup of delmoral2011ci concrete here. Concentration of gradient estimate under particle measure to limit under posterior measure}
 
@@
 
-\begin{lem}[Minimum Eigenvalue Bound for Hessian Estimate]
+\begin{conj}[Minimum Eigenvalue Bound for Hessian Estimate]
     \label{lemma:hess_bound}
@@
     for $\hat{H}(\theta)$ to be invertible and positive definite with minimum eigenvalue greater than or equal to $c \in (0, \sum_{j=1}^J \E\lambda_{\min}(H_j))$ and with probability at least $1-\exp(-t)$.
-\end{lem}
+\end{conj}
 \begin{proof}

### 2024-03-06 mop/mop.tex
Updates from Overleaf
merge

@@
 
+\kevin{w.r.t. alpha=0:}
 Bias in dropping the resampling gradient increases as process mixes slower.
@@
 
+\noindent {\bf Definition 1, Liu}.
+{\it A random variable $X$ drawn from a distribution $g$ is said to be properly weighted by a weighting function $w(X)$ with respect to the distribution $\pi$ if for any integrable function $h$,
+$$
+E_g\{h(X) w(X)\}=E_\pi\{h(X)\}
+$$
+
+A set of random draws and weights $\left(x^{(j)}, w^{(j)}\right), j=1,2, \ldots$, is said properly weighted with respect to $\pi$ if
+$$
+\lim _{m \rightarrow \infty} \frac{\sum_{j=1}^m h\left(x^{(j)}\right) w^{(j)}}{\sum_{j=1}^m w^{(j)}}=E_\pi(h(X))
+$$
+for any integrable function $h$.
+In a practical sense we can think of $\pi$ as being approximated by the discrete distribution supported on the $x^{(j)}$ with probabilities proportional to the weights $w^{(j)}$.
+}
+
 \noindent {\bf Lemma 1}. {\it
@@
 L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
+= L^\phi_n \sum_{j=1}^J w_{n,j}^{FC,\theta} \frac{g^\phi_{n,j}}{L_n^\phi}
 \]
-Using Lemma~1, we see this is properly estimated by
+Using Lemma~1, we resample according to probabilities $\frac{g^\phi_{n,j}}{L_n^\phi}$ to see this is properly estimated by
 \[

### 2024-03-06 presentation/presentation.tex
merge

@@
 \subtitle{\small{Inference for Partially Observed Structured Dynamic Systems}}
-\author[Kevin Tan]{Kevin Tan \\ \small{University of Pennsylvania, Department of Statistics and Data Science} \\\vspace{0.5cm} \small{Joint work with Giles Hooker and Edward Ionides}}
-\date[January 4, 2024]{IMS Asia Pacific Rim Meeting \\ January 4, 2024}
+\author[Kevin Tan]{Kevin Tan \\\vspace{0.5cm} \small{Joint work with Giles Hooker and Edward Ionides}}
+\date[\today]{\today}
 
@@
 \usepackage{fancyhdr}
-\usepackage{appendixnumberbeamer}
 
@@
 
-    \begin{itemize}
-        \item Unobserved Markov process $\{X_t, t \geq t_0\}$, observations $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$.
-        \item \pause Unknown parameter $\theta \in \Theta$, to be estimated.
-        \item \textbf{Notation:}
-        \begin{itemize}
-            \item $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ is the \textbf{process model}.
-            \item \pause $\texttt{process}\left(x_n, \theta\right)$ is the \textbf{simulator} corresponding to the process model. 
-            \item \pause $f_{Y_n|X_n}\left(y_n \mid x_n, \theta\right)$ is the \textbf{measurement model}.
-        \end{itemize}
-    \end{itemize}
 \end{frame}
 
-\begin{frame}{Particle Filters}
-
-    \begin{figure}
-        \centering
-        \begin{minipage}[c]{0.5\textwidth}
-            \centering
-            \includegraphics[scale=0.15]{imgs/pfilter.png}
-            \end{minipage}
-        \begin{minipage}[c]{0.45\textwidth}
-            \caption{
-            \\
-            -- Dirac measure approximation of the Bayes filter. 
-            \\ 
-            -- Yields a filtering distribution and likelihood estimate.
-            \\ 
-            -- Graphic from \cite{berg2019pfilterimage}.}
-        \end{minipage}
-        \label{fig:pfilter-illustration}
-    \end{figure}
-
-    \begin{itemize}
-        \item Monte Carlo approximation of the Bayes filter.
-        \begin{itemize}
-            \item \pause Maintain belief on current state $x_n^F$. 
-            \item \pause Simulate forward to get $x_{n+1}^P$, observe observation $y_{n+1}^*$.
-            \item \pause Update $x_{n+1}^F$ accordingly.
-        \end{itemize}
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{Automatic Differentiation}
-
-    \begin{figure}
-        \begin{minipage}[c]{0.7\textwidth}
-            \includegraphics[scale=0.3]{imgs/autograd.png}
-        \end{minipage}
-        \begin{minipage}[l]{0.27\textwidth}
-            \caption{
-            \\$f(x,y) = x^2+xy+2$, $f'(x,y) = (2x+y,x).$
-            -- Figure from \href{https://avinashselvam.medium.com/automatic-differentiation-explained-9f02c74e9a90}{Medium}.}
-        \end{minipage}
-        \label{fig:autograd-illustration}
-    \end{figure}
-
-    \begin{itemize}
-        \item Evaluates the gradient of a (scalar or vector-valued) computer program w.r.t. its arguments.
-        \item \pause Traverses computational graph (of primitive functions) with chain rule.
-    \end{itemize}
-
-\end{frame}
-
-\section{Motivation}
+\section{Introduction}
 
@@
     \item \textbf{Intractable likelihood functions}! Bypassed with e.g. simulated likelihood, likelihood-free inference, particle filters, etc. 
-    \item \pause May not have access to transition densities, only a \textbf{simulator}. 
-    \begin{itemize}
-        \item \pause IF2 (\cite{ionides15}), particle MCMC (\cite{doucet2010pmcmc}), only full-information methods that can deal with this.
-    \end{itemize}
-    \item \pause \textbf{Significant Monte-Carlo noise in likelihood estimate} makes accurate parameter estimation difficult.
+    \item Want to perform inference without having access to the transition density of the latent Markov process, using only  a \textbf{simulator} (\cite{welch2009abc, wood2010sl, doucet2010pmcmc, ionides08, ionides15}).
+    \item \textbf{Significant Monte-Carlo noise in the likelihood estimate} makes accurate parameter estimation difficult.
 \end{enumerate}
@@
 
-\begin{frame}{IF2 Plateaus Quickly But Struggles to Find the MLE}
-\begin{figure}
-    \centering
-    \includegraphics[scale=0.5]{imgs/095/if2fail.png}
-    \caption{Performance of IF2 on \cite{king08} Dhaka cholera model.
-Shaded area represents 0th to 80th percentile, solid line is median of 100 runs.
-While IF2 makes quick initial progress, it fails to find the MLE.}
-    \label{fig:if2fail}
-\end{figure}
-\end{frame}
-
-\begin{frame}{Hang on, wait a minute!}
-    \begin{itemize}
-        \item Original iterated filtering algorithm: (noisy) approximation of score (\cite{ionides06-pnas}). 
-        %\item \pause Performed a single "gradient step" at the end of each filtering iteration.
-        \pause
-        \item \textbf{Would performing automatic differentiation (AD) on the likelihood estimate from the particle filter lead to a less noisy score approximation?}
-    \end{itemize}
-    \pause
-    \begin{block}{Problem!}
-        But the particle filter has discrete stochastic resampling! How can we differentiate this?
-    \end{block}
-\end{frame}
-
-\section{Previous Work: AD for Particle Filters}
-
-\begin{frame}{Previous Work}
-
-    \begin{table}[h!]
-        \centering
-        \begin{tabular}{||c c||} 
-         \hline
-         Authors & Method \\ [0.5ex] 
-         \hline\hline
-         \cite{poyiadjis11} & Particle approximation of score
-         \\ & $O(N^2)$ variance, unbiased \\
-         \hline
-         \cite{blei2018vsmc} & Backprop through vanilla PF \\ 
-         & Asymptotic bias \\
-         \hline
-         \cite{corenflos21} & Optimal transport resampling \\
-         & Consistent, $O(J^2)$ runtime \\
-         \hline
-         \cite{scibior21} & Stop-gradient trick \\ 
-         & Recovers \cite{poyiadjis11} with AD \\
-         \hline 
-         \cite{singh22} & Fixed-lag smoothing \\ 
-         & Need transition densities \\ 
-         \hline
-        \end{tabular}
-        \label{table:lit-review}
-    \end{table}
-\end{frame}
-
-\begin{frame}{Our Contributions}
-    \begin{itemize}
-        \item New theoretical framework/algorithm/gradient estimator we call MOP-$\alpha$.
-        \begin{itemize}
-            \item \pause Gradient estimates of \cite{blei2018vsmc} (MOP-0), \cite{poyiadjis11} and \cite{scibior2021dpf} (MOP-1) are special cases.
-            \item \pause Does not need transition densities, only a differentiable simulator.
-            \item \pause Can optimize a bias-variance tradeoff. 
-        \end{itemize}
-        \item \pause Promising hybrid algorithm, warm-starts gradient descent (using this estimator) with IF2.
-        \item \pause Outperforms IF2 on Cholera model of \cite{king08}.
-    \end{itemize}
-\end{frame}
-
-\section{Smooth Extensions to the Particle Filter}
-
-\begin{frame}{Intuition}
-    \begin{block}{Problem}
-        Want to differentiate through the particle filter. But the particle filter has discrete stochastic resampling! 
-    \end{block}
-    \pause 
-    \begin{block}{Idea}
-        Don't differentiate $\hat\ell(\theta)$ directly, differentiate through a (suitably) reweighted bootstrap filter.
-    \end{block}
-    \begin{itemize}
-        \item \pause Particle filter run with state transitions \& resampling under $\phi \in \Theta$. 
-        \item \pause Reweight to evaluate likelihood at nearby $\theta \in \Theta$ (same resampling).
-        \item \pause Fix seed to treat particles, weights under $\phi$ as constants.
-        \item \pause \textbf{Now only need to differentiate through likelihood ratios!}
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{What is this reweighting?}
-    Ordinarily, the likelihood estimate for $\phi$ is 
-    $$\hat\lik(\phi) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J \left(\underbrace{f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}, \phi)}_{\textcolor{blue}{\text{Measurement Model}}}
-    \cdot 
-    \underbrace{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi} | x_{n-1,j}^{F,\phi}, \phi)}_{\textcolor{red}{\text{Process model}}}\right).$$
-    \pause
-    So to estimate the likelihood for $\theta$, we can correct 
-    $$
-        \hat\lik(\theta) 
-        = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}, \phi) \cdot 
-        f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi} | x_{n-1,j}^{F,\phi}, \phi) 
-        \cdot \underbrace{s_{n,j} \cdot r_{n,j}}_{\text{Correction Term}},
-    $$
-    \pause
-    where the multiplicative correction terms are
-    $$
-        \underbrace{s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}}_{\textcolor{blue}{\text{Measurement Model Likelihood Ratios}}},\;\; \underbrace{r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)}}_{\textcolor{red}{\text{Process Model Likelihood Ratios}}}.
-    $$
-
-\end{frame}
-
-\begin{frame}{New Problem}
-    \begin{block}{Problem}
-        Need likelihood ratios w.r.t. process and measurement models!
-    \end{block}
-    \pause 
-    \begin{block}{Solution}
-        Offload process model derivatives to differentiable simulator. 
-    \end{block}
-    \begin{itemize}
-        \item \pause Run particle filter twice under same seed: one with process model at $\phi$, another with process model at $\theta$. Resample according to $\phi$.
-        \item \pause Correct only with resampling likelihood ratios. 
-        \item \pause Introduce additional discounting parameter $\alpha$ for bias-variance tradeoff.
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{Algorithm: Measurement Off-Policy, MOP-$\alpha$}
-    \begin{enumerate}
-        \item \textbf{Run a particle filter once at} $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$. Fix seed throughout. Set initial weights $w_{0,j}^{F,\theta} = 1/J$.
-        \item \pause \textbf{At each timestep,} propagate weights $w_{n,j}^{P,\theta} := (w_{n-1,j}^{F,\theta})^\alpha$. 
-        \item \pause Simulate process model ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$, evaluate measurement model $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$, evaluate conditional likelihood under $\phi$, $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$, \textbf{as usual.}
-        \item \pause \textbf{Resample according to $\phi$:} $k_{1:J} \sim \prob\big(k_{j}=m\big) \propto g^{\phi}_{n,m}$.
-        \item \pause \textbf{Correct filtering weights:} $\displaystyle w^{F,\theta}_{n,j}= w^{P,\theta}_{n,k_j} \times  g^{\theta}_{n,k_j}/ g^{\phi}_{n,k_j}$. 
-    \end{enumerate}
-    \pause 
-    \centering{\textbf{Special Case:} If $\theta=\phi$, only need to run one particle filter!
-Set $X_{n,j}^{P,\theta}, X_{n,j}^{F,\theta}$ to be copies of the $\phi$ counterparts, where gradients don't propagate.
-Explains the \textbf{stop-gradient} trick of \cite{scibior21}.}
-\end{frame}
-
-\begin{frame}{Bias-Variance Tradeoff with $\alpha$}
-\begin{itemize}
-    \item $\alpha$ balances a tradeoff between:
-    \begin{itemize}
-        \item \pause Maintaining memory of each particle's ancestral trajectory (most extreme when $\alpha=1$).
-        \item \pause Considering only the single-step transition dynamics (when $\alpha=0$).
-    \end{itemize}
-    \item \pause\textbf{Exponentially-weighted moving average} where $\alpha$ controls the amount of discounting.
-    \item \pause Bias small if $y_{n+1:N}^*$ not informative for $x_n$ given $y_{0:n}^*$ (\cite{corenflos21}).
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}{Bias-Variance Tradeoff with $\alpha$}
-\begin{figure}
-    \centering
-    \includegraphics[scale=0.5]{imgs/mlegrad.png}
-    \caption{Gradient estimates at MLE for trend in \cite{king08} cholera model. When $\alpha=1$, estimate unbiased with high variance. When $\alpha=0$, lower variance but biased.}
-    \label{fig:bias-variance}
-\end{figure}
-
-\end{frame}
-
-\begin{frame}{Guarantee}
-    \begin{prop}[Correctness of MOP-$\alpha$]
-    \label{prop:mop-correctness}
-    When either $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the posterior and is unbiased for the likelihood under $\theta$. 
-    When $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter. Its gradient when $\alpha=1$ is the estimate of \citet{poyiadjis11},
-    $$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*),$$
-    and when $\alpha=0,$ is the gradient estimator of \cite{blei2018vsmc},
-    $$
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
-    $$
-    i.e. differentiating through a vanilla particle filter (\cite{scibior21}).
-\end{prop}
-\end{frame}
-
-\section{Practical Maximum Likelihood Inference}
-
-\begin{frame}{One possible procedure:}
-    \begin{itemize}
-        \item Warm-start gradient descent, or some other first or second order procedure (using the gradient estimate given by MOP-$\alpha$), with the output of an initial search of IF2. 
-        \item \pause Run IF2, aggressively cooling to a fixed learning rate, till search stalls.
-        \item \pause Refine this coarse solution with gradient descent.
-    \end{itemize}
-    \pause
-    \centering{\textbf{Iterated Filtering with Automatic Differentiation (IFAD)}}
-\end{frame}
-
-\begin{frame}{Convergence Analysis}
-    \begin{prop}[Convergence of IFAD-1]
-        \label{thm:convergence}
-
-        Consider a variant of IFAD-1 where one stops if the gradient estimate $||g(\theta_n)|| \leq \sigma \epsilon$.
-Assume $-\ell$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$.
-$\sigma = \frac{4 \Gamma}{(1-\beta)}> 4$.
-Define $G$ as in Assumption \ref{assump:local-bounded-derivative}.
-
-        Let $H(\theta_n)$ be a matrix possibly dependent on $\theta_n$ with minimum eigenvalue always greater than some $c>0$.
-Choose learning rate $\eta$ and error tolerance $\epsilon$ such that, where $\beta$ is the Armijo condition hyperparameter,
-        \begin{equation*}
-            \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_n)||.
-        \end{equation*}
-
-        Then, with $J$ large enough, the following holds with probability at least $1-\delta$:
-        \begin{equation*}
-        \ell(\theta^*) - \ell(\theta_{n+1}) \leq \left(1-\eta\beta\frac{8\gamma}{9c}\right)(\ell(\theta^*)-\ell(\theta_n))
-        \end{equation*}
-
-        and the algorithm terminates when $||\nabla_\theta \ell(\theta_n)|| \leq (1+\sigma) \epsilon$.
-        \end{prop}
-\end{frame}
-
-\begin{frame}{Convergence Analysis}
-    \begin{itemize}
-        \item Similar to \cite{mahoney16}, uses concentration inequalities from \cite{delmoral2011ci}.
-        \item Gradient stage of IFAD-1 converges linearly to the MLE if:
-        \begin{enumerate}
-            \item \pause The log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE.
-            \item \pause The IF2 stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE. 
-        \end{enumerate}
-        \item \pause \textbf{Conjecture:} This applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE and behaves a lot like SGD. 
-    \end{itemize}
-
-\end{frame}
-
-\begin{frame}{Runtime}
+\begin{frame}{When Does IF2 Struggle?}
 
 \begin{itemize}
-    \item Warm-start:
-    \begin{itemize}
-        \item \pause Initial warm-start "convergence" happens fairly quickly in practice. 
-        \begin{itemize}
-            \item \pause Dhaka cholera model of \cite{king08}: 40 iterations, usually 100-200 for global search.
+    \item IF2 is great at approaching the MLE quickly! 
+    \item Performs parameter updates at every timestep, 
 \end{itemize}
-    \end{itemize}
-    \item \pause MOP-$\alpha$:
-    \begin{itemize}
-        \item \pause Gradient takes 3.75x time of \texttt{pfilter()}, in line with cheap gradient principle (\cite{kakade2019provably}).
-    \end{itemize}
-    \item \pause Runtime \textbf{linear, and not quadratic}, in number of particles.
-    \item \pause Re-implementation in \texttt{JAX} (\cite{jax2018github}) led to 16x speedup v.s. \texttt{pomp} package of \cite{king08}.
-\end{itemize}
-
-\end{frame}
-
-\section{Numerical Experiments}
-
-\begin{frame}{Cholera in Bangladesh}
 
@@
     \centering
-        \includegraphics[scale=0.5]{imgs/dacca.png}
-        \caption{Illustration of SIR model from \cite{king08}.}
-        \label{fig:sir}
+    \includegraphics[scale=0.5]{imgs/095/optim.png}
+    \label{fig:enter-label}
 \end{figure}
 
-    \begin{itemize}
-        \item Dhaka model from \cite{king08}, also used in \cite{ionides15} to benchmark IF2.
-        \item \pause Stochastic SIR compartmental model with transition uncertainty driven by Brownian motion.
-        \item \pause Force of infection $\lambda(t)$ modeled with splines for seasonality, etc.
-    \end{itemize}
-
-        %\item \pause Population $H(t)$ divided into:
-        %\begin{itemize}
-        %    \item \pause Susceptible compartment $S(t)$.
-        %    \item \pause Infected compartment $I(t)$.
-        %    \item \pause Three ($k=3$) recovered compartments $R_1(t), ..., R_k(t)$ denoting varying degrees of cholera immunity.
-        %\end{itemize}
-        %\item \pause $M(t)$ denotes the cholera deaths in each month.
-\end{frame}
-
-\begin{frame}{Results}
-
-    \begin{table}[h!]
-        \centering
-        \begin{tabular}{||c c c||} 
-         \hline
-         Method & Best Log-Likelihood & Rank \\ [0.5ex] 
-         \hline\hline
-         IFAD-0 & $\textbf{-3747.87}$ & 1\\
-         IFAD-1 & $\textbf{-3748.53}$ & 2\\
-         IF2 (Ours) & -3755.91 & 3\\
-         IF2 (\cite{ionides15}) & -3768.63 & 4\\ 
-         MOP-0 Alone & -3795.74 & 5\\
-         MOP-1 Alone & -3797.38 & 6\\
-         \hline
-        \end{tabular}
-        \caption{\\
-        -- IFAD performs the best among all methods, finding the MLE. \\
-        -- Our implementation of IF2 outperforms that of \cite{ionides15} but underperforms IFAD.}
-        \label{table:mle}
-    \end{table}
-
-    \begin{itemize}
-        \item Benchmarked IFAD against IF2 on a challenging global search problem. 
-        \item \pause Performed 100 searches each. 
-    \end{itemize}
-
-\end{frame}
-
-\begin{frame}{Results}
-
-\begin{figure}[htbp!]
-    \centering
-    \includegraphics[scale=0.37]{imgs/095/pairs.png}
-    \includegraphics[scale=0.37]{imgs/095/qq.png}
-    \caption{\\
-    --\textbf{Left:} Paired searches from the same starting point. \\
-    --\textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches. \\
-    -- IFAD has the edge and manages to find the MLE.\\
-    -- No IF2 search successfully gets within 7 log-likelihood units of it.}
-    \label{fig:scatter}
-\end{figure}
-\end{frame}
-
-\begin{frame}{Results}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[scale=0.28]{imgs/095/boxplot.png}
-    \includegraphics[scale=0.31]{imgs/095/boxplotall.png}
-    \caption{\\
-    -- \textbf{Left:} Results of best run out of every ten runs, simulating procedure of running a few searches and choosing the best one. \\
-    -- \textbf{Right:} MOP alone drastically underperforms all other methods, failing to get close to the MLE. \\
-    -- IF2 warm-start necessary in challenging nonconvex and noisy problems.}
-    \label{fig:boxplot-search}
-\end{figure}
-
-\end{frame}
-
-\begin{frame}{Results}
-\begin{figure}[htbp!]
-    \centering
-    \includegraphics[scale=0.5]{imgs/095/optim.png}
-    \caption{\\
-    -- Solid lines depict the median negative log-likelihood at each iteration.\\
-    -- Shaded area depicts the best search at any iteration and the 80\% percentile.}
-    \label{fig:optim}
-\end{figure}
-%-- While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD -- IFAD has better tail control and successfully reaches the MLE.
-The initial loss in log-likelihood upon transitioning from IF2 to MOP seems to help with escaping local minima.
-\end{frame}
-
-\begin{frame}{Results}
-
-\begin{figure}[u!]
-    \centering
-    \includegraphics[scale=0.35]{imgs/095/hist.png}
-    \caption{Histogram comparison of IFAD, IF2, and MOP alone. Right is better.}
-    \label{fig:hist-all}
-\end{figure}
-\end{frame}
-
-\begin{frame}{Takeaways}
-    \begin{enumerate}
-        \item IF2 converges quickly to a neighborhood of the MLE but fails to find the MLE.
-        \item \pause Gradient steps perform better at fine-grained refinement.
-        \item \pause Performing gradient steps alone without a warm-start leads to the search getting stuck in local minima and saddle points.
-        \item \pause IFAD combines the best of IF2 and MOP, approaching the MLE quickly and successfully performing refinement -- even on a very difficult global search problem.
-    \end{enumerate}
-\end{frame}
-
-\section{Conclusions and Future Work}
-
-\begin{frame}{Conclusion}
-    \begin{itemize}
-        \item New theoretical framework/algorithm/gradient estimator that encompasses a few existing gradient estimates.
-        \item \pause Promising hybrid algorithm that warm-starts gradient descent (using this estimator) with IF2.
-        \item \pause Outperforms IF2 on Dhaka cholera model of \cite{king15}.
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{Future Work}
-    \begin{itemize}
-        \item MOP-$\alpha$ cannot handle discrete latent states. Maybe likelihood ratios can get around this.
-        \item \pause Convergence rate of IF2 not known, conjecture similar behavior to SGD. 
-        \item \pause Only gradient descent explored, other methods like Newton's method, ADAM, L-BFGS, etc. may perform better.
-        \item \pause Extensions to panel and spatiotemporal data, e.g. with a block particle filter in the lens of \cite{ionides22} and \cite{ning23}.
-        \item \pause Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual}.
-    \end{itemize}
 \end{frame}
@@
 
-\appendix
-\section{Appendix}
-
-\begin{frame}{Previous Work}
-
-    \begin{itemize}
-        \item \cite{poyiadjis11}, particle approximation of the score. \pause \\ \cite{scibior21}: AD + "stop-gradient trick". 
-        \item 
-        \pause \cite{blei2018vsmc}, backpropagate through vanilla particle filter.
-        \item 
-        \pause \cite{corenflos21}, optimal transport resampling.
-        \item 
-        \pause \cite{singh22}, fixed-lag smoothing.
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{Previous Work: Issues}
-    \begin{itemize}
-        \item Not (yet) compatible with desire for simulation-based inference.
-        \begin{itemize}
-            \item \pause Need modifications to use the bootstrap filter.
-            \item \pause Sometimes special cases needed, e.g. transitions factoring into a policy and deterministic model (\cite{singh22}).
-        \end{itemize}
-        \item \pause Computationally expensive, quadratic in number of particles.
-        \begin{itemize}
-            \item \pause Optimal transport resampling (\cite{corenflos21}), marginal particle filters (\cite{scibior21}).
-        \end{itemize}
-        \item \pause High variance, or asymptotically biased.
-        \begin{itemize}
-            \item \pause Either one drops resampling terms and accepts asymptotic bias (\cite{blei2018vsmc}), or has variance quadratic in horizon (\cite{poyiadjis11}, \cite{scibior21})
-        \end{itemize}
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{(Improved) Iterated Filtering (\cite{ionides15})}
-    \begin{itemize}
-        \item Only (practical) full-information, frequentist method of parameter estimation in POMPs that does not need transition densities. 
-        \begin{itemize}
-            \item \pause Each particle has its own set of parameters.
-            \item \pause Particle parameters perturbed at every timestep.
-            \item \pause Parameters resampled with their particles according to likelihood.
-        \end{itemize}
-        \item \pause Treats parameters as augmentations of the state space, evolving according to a random walk.
-        \item \pause \textbf{Consistent estimates} via viewing this as a sequential Bayes map on the parameter distribution.
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{When Does IF2 Struggle?}
-
-\begin{itemize}
-    \item \textbf{IF2 is great at approaching the MLE quickly!}
-    \item \pause Performs parameter updates at every timestep, so it does not need to wait to filter through the entire trajectory before each update.
-    \item \pause \textbf{But struggles at squeezing out the last few units of log-likelihood.}
-    \begin{itemize}
-        \item \pause Especially true in highly nonlinear, nonconvex, and noisy settings, e.g. \textbf{disease models}. 
-    \end{itemize}
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}{Assumptions}
-
-    \begin{aspt}[Smooth Neighborhood]
-    \label{assump:smooth-nbhd}
-    There exists a neighborhood $\mathcal{N}(\phi)$ around any $\phi \in \Theta$ where for all $\theta \in \mathcal{N}(\phi)$ and almost every $\omega \in \Omega$, the Monte Carlo estimate of the likelihood at $\theta \in \mathcal{N}(\phi)$ with the system evolving according to $\phi$ conditional on $\omega$, $\hat\lik(\theta,\phi,\omega, J)$, is twice differentiable in $\theta$.
-    \end{aspt}
-
-    \begin{itemize}
-        \item \pause \textbf{Justification:} For suitably nearby $\theta$ and $\phi$, the resampling indices for $\hat\lik(\theta,\phi,\omega, J)$ and $\hat\lik(\phi,\phi,\omega, J)$ are the same.
-        \item \pause This eliminates any discontinuities from resampling.
-        \item \pause For small enough $\mathcal{N}(\phi)$, the likelihood ratios are bounded. 
-        \item \pause The likelihood only changes by a factor of the likelihood ratios, which are bounded and smooth in $\theta$ if the densities used in the calculation are. 
-    \end{itemize}
-
-\end{frame}
-
-\begin{frame}{Assumptions}
-
-    \begin{aspt}[Continuity of the Likelihood] $\ell(\theta)$ proper is continuous in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.
-    \end{aspt}
-    \pause 
-    \begin{aspt}[Bounded Measurement Model] There is an $\epsilon>0$ with $\epsilon^{-1}>f_{Y_n \mid X_n}\left(y_n{ }^* \mid x_n; \theta\right)>\epsilon$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
-    \end{aspt}
-    \pause 
-    \begin{aspt}[Locally Bounded Derivative]
-    \label{assump:local-bounded-derivative}
-    Let $\mathcal{M}$ be an open subset of $\Omega$. There exists some function $G(\theta)$ and a constant $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$ such that
-        $$
-        \|\nabla \ell(\theta, \phi, \omega, J)\|_2<G(\theta) \leq G<\infty
-        $$
-        for every $\phi \in \mathcal{M}, \theta$ in smooth neighborhood $\mathcal{N}(\phi)$, and almost every $\omega \in \Omega$.
-    \end{aspt}
-\end{frame}
-
-\begin{frame}{Algorithm: Measurement Off-Policy, MOP-$\alpha$}
-    \begin{figure}
-        \centering
-        \includegraphics[scale=0.4]{imgs/mop.png}
-        \caption{MOP-$\alpha$ algorithm.}
-        \label{fig:mop}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Why did we bother with this?}
-    \begin{itemize}
-        \item Lets us work with only a differentiable simulator.
-        \item \pause Discounting parameter $\alpha$ helps with a bias-variance tradeoff (explained later). 
-        \item \pause Gradients returned by the likelihood estimate $$\hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$$ have nice properties.
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{Iterated Filtering with Automatic Differentiation (IFAD)}
-    \begin{figure}
-        \centering
-        \includegraphics[scale=0.45]{imgs/ifad.png}
-        \caption{Warm-starting first/second order iterative optimization with IF2.}
-        \label{fig:ifad}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Why does this help?}
-    \begin{itemize}
-        \item IF2 converges quickly, but struggles with last few log-likelihood units. 
-        \item \pause Gradient descent gets stuck in saddle points and poor local minima when the likelihood is nonconvex. 
-        \item \pause Warm-starting gradient methods with IF2:
-        \begin{itemize}
-            \item \pause Under regularity conditions, the likelihood is well-behaved near the MLE.
-            \item \pause Issues with saddle points and local minima alleviated. 
-        \end{itemize}
-        \item \pause Combining these two lets us enjoy the best of both worlds.
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{Convergence Analysis}
-
-    Guarantee currently only holds for IFAD-1, as it forms a particle approximation of the score as in \cite{poyiadjis11}.
-    \begin{itemize}
-        \item \pause Can be extended if the bias for the gradient estimate given by MOP-$\alpha$ for $\alpha < 1$ is small enough.
-        \item \pause e.g. $\alpha$ close to 1, or $y_{n+1:N}^*$ uninformative of current state $x_n$ given past and current measurements $y_{0:n}^*$.
-        \item \pause Otherwise, need to handle biased gradient descent.
-    \end{itemize}
-
-\end{frame}
-
-\begin{frame}{Cholera in Bangladesh}
-    The \textbf{transition dynamics} follow this series of stochastic differential equations driven by Brownian motion:
-    \begin{align*}
-        d S&=\left(k \epsilon R_k+\delta(S-H)-\lambda(t) S\right) d t+d H-(\sigma S I / H) d B, \\ 
-        d I&=\left(\lambda(t) S-(m+\delta+\gamma) I\right) d t+(\sigma S I / P) d B, \\ 
-        d R_1&=\left(\gamma I-(k \epsilon+\delta) R_1\right) d t, \\ 
-        \vdots \\ 
-        d R_k&=\left(k \epsilon R_{k-1}-(k \epsilon+\delta) R_k\right) d t,
-    \end{align*}
-\end{frame}
-
-\begin{frame}{Cholera in Bangladesh}
-The \textbf{force of infection}, $\lambda(t)$, is modeled by splines $s_j$, 
-    \begin{equation*}
-        \lambda(t)=\exp \left\{\beta_{\text {trend }}\left(t-t_0\right)+\sum_{j=1}^{6} \beta_j s_j(t)\right\}(I / P) + \exp \left\{\sum_{j=1}^{6} \omega_j s_j(t)\right\},
-    \end{equation*}
-    where
-    \begin{itemize}
-        \item \pause $\beta_j$ model seasonality in the force of infection.
-        \item \pause $\beta_{\text{trend}}$ models the trend in the force of infection.
-        \item \pause $\omega_j$ represent seasonality of a non-human environmental reservoir of disease.
-    \end{itemize}
-
-\end{frame}
-
-\begin{frame}{Cholera in Bangladesh}
-    The \textbf{measurement model} for observed monthly cholera deaths is given by 
-    \begin{equation*}
-        Y_n \sim \gN(M_n, \tau^2M_n^2),
-    \end{equation*}
-    where $M_n$ is the true number of cholera deaths in that month.
-\end{frame}
-
 \end{document}

### 2024-03-06 filtering.py

@@ -61,6 +61,85 @@ else:
 '''
 
 
+def mop_helper(t, inputs):
+    particlesF, theta, covars, loglik, weightsF, counts, ys, alpha, key = inputs
+    J = len(particlesF)
+    
+    if len(covars.shape) > 2:
+        key, *keys = jax.random.split(key, num=J*covars.shape[1]+1)
+        keys = np.array(keys).reshape(J, covars.shape[1], 2).astype(np.uint32)
+    else:    
+        key, *keys = jax.random.split(key, num=J+1)
+        keys = np.array(keys)
+        
+    # Discount weights by alpha in logspace
+    weightsP = alpha*weightsF
+    
+    # Get prediction particles 
+    if covars is not None:
+        particlesP = rprocess(particlesF, theta, keys, covars)# if t>0 else particlesF
+    else:
+        particlesP = rprocess(particlesF, theta, keys, None)
+        
+    measurements = dmeasure(ys[t], particlesP, theta)
+    if len(measurements.shape) > 1:
+        measurements = measurements.sum(axis=-1)
+    
+    # Using before-resampling conditional likelihood
+    loglik += (jax.scipy.special.logsumexp(weightsP + measurements) 
+               - jax.scipy.special.logsumexp(weightsP))
+    
+    # Obtain normalized measurement likelihoods for resampling
+    norm_weights, loglik_phi_t = normalize_weights(jax.lax.stop_gradient(measurements))
+
+    # Systematic resampling according to normalized measurement likelihoods
+    counts, particlesF, norm_weightsF = resampler(counts, particlesP, norm_weights)
+    
+    # Correct for theta/phi and resample
+    weightsF = (weightsP + measurements - jax.lax.stop_gradient(measurements))[counts]
+
+    #jax.debug.print(loglik, loglik_t)
+    return [particlesF, theta, covars, loglik, weightsF, counts, ys, alpha, key]
+
+    
+# test on linear gaussian toy model again
+@partial(jit, static_argnums=2)
+def mop(theta, ys, J, covars=None, alpha=0.97, key=None):
+    if key is None:
+        key = jax.random.PRNGKey(onp.random.choice(int(1e18)))
+    
+    particlesF = rinit(theta, J, covars=covars)
+    weights = np.log(np.ones(J)/J)
+    weightsF = np.log(np.ones(J)/J)
+    counts = np.ones(J).astype(int)
+    loglik = 0
+    
[…]

### 2024-03-06 optim.py

@@ -33,12 +33,33 @@ MONITORS = 1
 
 
 @partial(jit, static_argnums=2)
+def jgrad_pf(theta_ests, ys, J, covars, thresh, key=None):
+    return jax.grad(pfilter_pf)(theta_ests, ys, J, covars=covars, thresh=thresh, key=key)
+
+
+@partial(jit, static_argnums=2)
+def jvg_pf(theta_ests, ys, J, covars, thresh, key=None):
+    return jax.value_and_grad(pfilter_pf)(theta_ests, ys, J, covars=covars, thresh=thresh, key=key)
+
+
+@partial(jit, static_argnums=2)
 def jgrad(theta_ests, ys, J, covars, thresh, key=None):
     return jax.grad(pfilter_mean)(theta_ests, ys, J, covars=covars, thresh=thresh, key=key)
 
 @partial(jit, static_argnums=2)
 def jvg(theta_ests, ys, J, covars, thresh, key=None):
     return jax.value_and_grad(pfilter_mean)(theta_ests, ys, J, covars=covars, thresh=thresh, key=key)
+
+
+@partial(jit, static_argnums=2)
+def jgrad_mop(theta_ests, ys, J, covars, alpha=0.97, key=None):
+    return jax.grad(mop_mean)(theta_ests, ys, J, covars=covars, alpha=alpha, key=key)
+
+@partial(jit, static_argnums=2)
+def jvg_mop(theta_ests, ys, J, covars, alpha=0.97, key=None):
+    return jax.value_and_grad(mop_mean)(theta_ests, ys, J, covars=covars, alpha=alpha, key=key)
+
+
 
 @partial(jit, static_argnums=2)
 def jhess(theta_ests, ys, J, covars, thresh, key=None):
@@ -64,7 +85,7 @@ def line_search(obj, curr_obj, pt, grad, direction, k=1, eta=0.9, xi=10, tau = 1
 
 
 #rerun with diff trajs each time
-def train(theta_ests, ys, covars=None, J=5000, Jh=1000, method='Newton', itns=20, beta=0.9, eta=0.0025, c=0.1, max_ls_itn=10, thresh=100, verbose=False, scale=False, ls=False): 
+def train(theta_ests, ys, covars=None, J=5000, Jh=1000, method='Newton', itns=20, beta=0.9, eta=0.0025, c=0.1, max_ls_itn=10, thresh=100, verbose=False, scale=False, ls=False, alpha=1): 
     Acopies = []
     grads = []
[…]

### 2024-03-06 resampling.py

@@ -23,6 +23,8 @@ tfpk = tfp.math.psd_kernels
 import matplotlib.pyplot as plt
 plt.style.use('matplotlibrc')
 
+
+
     
 def resample(norm_weights):
     J = norm_weights.shape[-1]

### 2024-03-06 presentation/presentation.tex

--- 2024-01-05 presentation/presentation.tex
@@
     \includegraphics[scale=0.5]{imgs/mlegrad.png}
-    \caption{Gradient estimates at MLE for trend in \cite{king08} cholera model. When $\alpha=1$, estimate unbiased with high variance. When $\alpha=0$, lower variance but biased.}
+    \caption{Gradient estimates at MLE for trend in \cite{king08} cholera model. When $\alpha=1$, estimate close to unbiased with high variance. When $\alpha=0$, lower variance but biased.}
+    \label{fig:bias-variance}
+\end{figure}
+
+\end{frame}
+
+\begin{frame}{Bias-Variance Tradeoff with $\alpha$}
+\begin{figure}
+    \centering
+    \includegraphics[scale=0.5]{imgs/095/biasvar.png}
+    \caption{Bias-variance tradeoff for gradient estimates at MLE for trend in \cite{king08} cholera model.
+When $\alpha=1$, estimate close to unbiased with high variance.
+When $\alpha=0$, lower variance but biased.
+MSE seems to be minimized at $\alpha=0.97$.}
     \label{fig:bias-variance}
@@
     When $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter. Its gradient when $\alpha=1$ is the estimate of \citet{poyiadjis11},
-    $$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*),$$
+    $$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{0:N}}(y_{1:N}^*|x_{1:N,j}^{A, F,\theta}),$$
     and when $\alpha=0,$ is the gradient estimator of \cite{blei2018vsmc},
@@
          \hline\hline
-         IFAD-0 & $\textbf{-3747.87}$ & 1\\
-         IFAD-1 & $\textbf{-3748.53}$ & 2\\
-         IF2 (Ours) & -3755.91 & 3\\
-         IF2 (\cite{ionides15}) & -3768.63 & 4\\ 
-         MOP-0 Alone & -3795.74 & 5\\
-         MOP-1 Alone & -3797.38 & 6\\
+         IFAD-0.97 & $\textbf{-3750.21}$ & 1\\
+         IFAD-0 & $-3752.17$ & 2\\
+         IFAD-1 & $-3754.63$ & 3\\
+         IF2 (Ours) & -3764.10 & 4\\
+         IF2 (\cite{ionides15}) & -3768.63 & 5\\ 
+         MOP-1 Alone (100 searches) & -3797.38 & 6\\
          \hline
+
         \end{tabular}
@@
         \item Benchmarked IFAD against IF2 on a challenging global search problem. 
-        \item \pause Performed 100 searches each. 
+        \item \pause Performed 44 searches each. 
     \end{itemize}

### 2024-03-06 mop/draft.tex
Updates from Overleaf

@@
 \usepackage{natbib}
-
+\usepackage{hyperref}
 \usepackage{enumitem}
@@
 
-Finally, \citet{scibior2021dpf} demonstrate that the estimators derived in \citet{doucet2011sf} can be attained with AD, albeit with a simple tweak to the particle filter that does not modify its likelihood estimates.
+Finally, \citet{scibior2021dpf} demonstrate that the estimators derived in \citet{poyiadjis11} can be attained with AD, albeit with a simple tweak to the particle filter that does not modify its likelihood estimates.
 Though their method is not fully compatible with simulation-based inference, we show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.
@@
 
-This is done through the construction of a smooth extension to the particle filter, as well as a \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) family of algorithms that (informally) encompasses the special case of AD of a vanilla PF (recovering the gradient estimator of \citet{blei2018vsmc}) when $\alpha=0$, and the DPF algorithm from \citet{scibior2021dpf} (that recovers the gradient estimator from \citet{doucet2011sf}) when $\alpha=1$.
+This is done through the construction of a smooth extension to the particle filter, as well as a \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) family of algorithms that (informally) encompasses the special case of AD of a vanilla PF (recovering the gradient estimator of \citet{blei2018vsmc}) when $\alpha=0$, and the DPF algorithm from \citet{scibior2021dpf} (that recovers the gradient estimator from \citet{poyiadjis11}) when $\alpha=1$.
 When $\alpha \in (0,1)$, this provides opportunities for optimizing a bias/variance tradeoff.
@@
 
-We want $\phi\in\Theta$ to be a point in parameter space around which $\lik(\theta,\phi,\omega, J)$ is differentiable with respect to $\theta$ and does not have excessive Monte~Carlo variance.
-This leads us to our first assumption.
-
-\begin{aspt}[Smooth Neighborhood]
-\label{assump:smooth-nbhd}
-There exists a neighborhood $\mathcal{N}(\phi)$ around any $\phi \in \Theta$ where for all $\theta \in \mathcal{N}(\phi)$ and almost every $\omega \in \Omega$, the Monte Carlo estimate of the likelihood at $\theta \in \mathcal{N}(\phi)$ with the system evolving according to $\phi$ conditional on $\omega$, $\lik(\theta,\phi,\omega, J)$, is twice differentiable in $\theta$.
-\end{aspt}
-
-\paragraph{Intuition:} We can justify this assumption as follows.
-For suitably nearby $\theta$ and $\phi$, the resampling indices for $\lik(\theta,\phi,\omega, J)$ and $\lik(\phi,\phi,\omega, J)$ are the same, eliminating any discontinuities from resampling, and that for small enough $\mathcal{N}(\phi)$ the likelihood ratios are bounded.
-The likelihood only changes by a factor of the likelihood ratios, which are bounded and smooth in $\theta$ if the densities used in the calculation are.
-
 Lastly, we require the following regularity conditions.
@@
 
-\begin{aspt}[Bounded Measurement Model] There is an $\epsilon>0$ with $\epsilon^{-1}>f_{Y_n \mid X_n}\left(y_n{ }^* \mid x_n; \theta\right)>\epsilon$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
+\begin{aspt}[Bounded Measurement Model]
+\label{assump:bounded-measurement}
+There is an $\epsilon>0$ such that $\epsilon^{-1}>f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)>\epsilon$ and a $G'(\theta)>0$ such that $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty < G'(\theta)$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
 \end{aspt}
 
-\begin{aspt}[Locally Bounded Derivative]
+\begin{aspt}[Locally Bounded Derivatives]
 \label{assump:local-bounded-derivative}
-Let $\mathcal{M}$ be an open subset of $\Omega$. There exists some function $G(\theta)$ and a constant $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$ such that
+Let $\mathcal{M}$ be an open subset of $\Omega$.
+There exists some function $G(\theta)$ and constants $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$, $H=\sup _{\theta \in \mathcal{N}} H(\theta)<\infty$ such that
     $$
-    \|\nabla \ell(\theta, \phi, \omega, J)\|_2<G(\theta) \leq G<\infty
+    \|\nabla_\theta \ell(\theta, \phi, \omega, J)\|_\infty <G(\theta) \leq G<\infty, \;\;
+    \|\nabla^2_\theta \ell(\theta, \phi, \omega, J)\|_\infty <G(\theta) \leq G<\infty
     $$
-    for every $\phi \in \mathcal{M}, \theta$ in smooth neighborhood $\mathcal{N}(\phi)$, and almost every $\omega \in \Omega$.
+    for every $\phi \in \mathcal{M}, \theta \in \mathcal{N}_J(\phi)$, and almost every $\omega \in \Omega$.
 \end{aspt}
 
+We need the following assumption to ensure that the weighted particle filter with off-policy resampling targets the posterior:
+\begin{aspt}[Bounded And Differentiable Measurement Density Ratios]
+    \label{assump:diff-meas}
+    For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{Y_n|X_n}(y_n^*|x, \theta)}{f_{Y_n|X_n}(y_n^*|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x \in \gX$ and all $\theta\in\gN(\phi)$.
+\end{aspt}
+
+In the following assumption, the first condition is sufficient for DOP, and the second for MOP:
+\begin{aspt}[Bounded And Differentiable Process Model]
+    \label{assump:diff-process}
+    Either one of the following two conditions hold:
+    \begin{enumerate}
+        \item For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{X_n|X_{n-1}}(x'|x, \theta)}{f_{X_n|X_{n-1}}(x'|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x, x' \in \gX$ and all $\theta\in\gN(\phi)$, or
+        \item There exists a differentiable simulator with more than two continuous derivatives for the process model. 
+    \end{enumerate}
+\end{aspt}
+
+We want $\phi\in\Theta$ to be a point in parameter space around which $\lik(\theta,\phi,\omega, J)$ is differentiable with respect to $\theta$ and does not have excessive Monte~Carlo variance.
+
+We require many of these assumptions to establish the following proposition, the proof of which is a straightforward consequence of what we assumed.
+
+\begin{prop}[Smooth Neighborhood]
+\label{prop:smooth-nbhd}
+There exists an open neighborhood $\mathcal{N}(\phi)$ around any $\phi \in \Theta$ where for all $\theta \in \mathcal{N}(\phi)$ and almost every $\omega \in \Omega$, the Monte Carlo estimate of the likelihood at $\theta \in \mathcal{N}(\phi)$ with the system evolving according to $\phi$ conditional on $\omega$, $\lik(\theta,\phi,\omega, J)$, has two continuous derivatives in $\theta$.
+\end{prop}
+
+\begin{proof}
+    For a fixed $\omega$, the resampling indices for $\lik(\theta,\phi,\omega, J)$ and $\lik(\phi,\phi,\omega, J)$ are the same as the resampling indices in both cases correspond to that yielded by $f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}(\omega);\phi)$, eliminating any discontinuities from resampling.
+
+    By Assumptions \ref{assump:diff-meas} and \ref{assump:diff-process}, for small enough $\mathcal{N}(\phi)$ the density ratios are bounded.
+The likelihood only changes by a factor of the density ratios, which are bounded and possess two continuous derivatives in $\theta$ if either the process and measurement densities used in the calculation do, or if the measurement densities do and one has access to a differentiable simulator with two continuous derivatives in $\theta$.
+This gives us the desired result.
+
+    As a remark, we note that the neighborhoods the sequence is differentiable in are independent of $J$, as the resampling does not change with $\theta$ -- it is dependent on $\phi$. 
+\end{proof}
+
 \subsubsection{Doubly Off-Policy Particle Filter}
-
-\kevin{rewrite to de-emphasize dop as in presentation, sell mop as the big theoretical idea that explains AD through PFs and unifies two previous estimators}
 
@@
 
-We claim that this yields a particle filter that targets the posterior and recovers the derivative estimate of \citet{doucet2011sf, scibior2021dpf} when $\theta$ is evaluated at $\phi$.
+We claim that this yields a particle filter that targets the posterior and recovers the derivative estimate of \citet{poyiadjis11, scibior2021dpf} when $\theta$ is evaluated at $\phi$.
 The proof can be found in Appendix \ref{app:dop}.
@@
 
-    In particular, when $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{doucet2011sf},
+    In particular, when $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{poyiadjis11},
     \begin{equation}
@@
 \end{prop}
-
-\kevin{Cut down on DOP discussion in actual manuscript, remove any references to this as an algorithm, just talk about the idea of an off-policy system/particle filter.
-Bottom paragraph is the punchline.}
 
@@
 This removes the issue of the likelihood ratios blowing up, and yields an estimate of the likelihood that agrees with the bootstrap filter estimate.
-Its gradient in this case is either the estimate of \citet{doucet2011sf} when $\alpha=1$, a single-timestep estimator that corresponds to naively taking the derivative of the particle filter when $\alpha=0$ as first proposed in \citet{blei2018vsmc}
+Its gradient in this case is either the estimate of \citet{poyiadjis11} when $\alpha=1$, a single-timestep estimator that corresponds to naively taking the derivative of the particle filter when $\alpha=0$ as first proposed in \citet{blei2018vsmc}
 
@@
 
+\kevin{I think we can establish the linear convergence for the entire procedure in the same sense as \href{this}{https://arxiv.org/pdf/2311.08442.pdf}, in the sense that if IF2 is run for a sufficiently large but fixed number of iterations, once we arrive in a local strongly convex region, we get linear convergence when we swap to MOP.
+Show by induction that we remain in the strongly convex region (in terms of parameter space and likelihood), eqn 218.
+}
+
 \begin{thm}[Convergence of Gradient Methods]
@@
 
-The proof of this result uses the concentration inequalities from \cite{delmoral2011ci}, and leans upon the analysis of \cite{mahoney2016subsampled}. We defer the proof to Appendix \ref{app:ifad}
+The proof of this result uses the concentration inequalities from \cite{delmoral2011ci}, and leans upon the analysis of \cite{mahoney16}. We defer the proof to Appendix \ref{app:ifad}
 
@@
 
-Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ is properly weighted for $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
 To see  (\ref{L2}),
@@
 \begin{lem}
-    For $\theta$, $\phi$ that satisfy Assumption \ref{assump:smooth-nbhd}, the following correction in the particle filter weights,
+    For $\theta$, $\phi$ that satisfy Proposition \ref{prop:smooth-nbhd}, the following correction in the particle filter weights,
     \begin{equation}
@@
 Also let $\theta \in \gN(\phi)$.
-When $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{doucet2011sf},
+When $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{poyiadjis11},
     \begin{equation}
@@
 \begin{lem}
+    \label{lem:mop-1-formula}
     Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The gradient estimate is then
@@
 
-$$w_{n,j}^{P,\theta} = w_{n-1,\theta}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,k_j}^\theta}{g_{i,k_j}^\phi}.$$
+$$w_{n,j}^{P,\theta} = w_{n-1,\theta}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,j}^{A,\theta}}{g_{i,j}^{A,\phi}},$$
+where we use the $(\cdot)^A$ superscript to denote the ancestral trajectory of the $j$-th particle at timestep $n$.
 
 Using the log-derivative identity again,
-$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,k_j}^\theta}{g_{i,k_j}^\phi}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,k_j}^\theta - \log g_{i,k_j}^\phi\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,k_j}^\theta.$$
+$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,j}^{A,\theta}}{g_{i,j}^{A,\phi}}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,j}^{A,\theta} - \log g_{i,j}^{A,\phi}\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,j}^{A,\theta}.$$
 
 So 
-$$  \nabla_\theta \sum_{n=1}^N \log g_{n,k_j}^\theta = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,k_j}^\theta\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,k_j}^{P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right),$$
+$$  \nabla_\theta \sum_{n=1}^N \log g_{n,j}^{A,\theta} = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,j}^{A,\theta}\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,j}^{A, P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right),$$
 
 and 
-$$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{i=1}^N \nabla_\theta \log g_{n,k_j}^\theta = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$
+$$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{n=1}^N \nabla_\theta \log g_{n,j}^{A,\theta} = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$
 
@@
 \begin{lem}
+\label{lem:mop-0-formula}
 
@@
 
+\kevin{I have no clue how to prove this. Help!}
+\begin{conj}[MOP-1 Estimate Is Unbiased for the Likelihood and the Gradient of the Likelihood]
+    Consider the case of MOP-$\alpha$ when $\alpha=1$, and resampling is carried out with either multinomial, residual, or systematic resampling at every timestep.
+Assume the gradient estimate obtained by the gradient of the likelihood estimate with J particles, $\nabla_\theta \hat\lik_J(\theta)$ is uniformly bounded for all $J$, and that the likelihood has two continuous derivatives.
+Then the likelihood estimate is unbiased for the likelihood and the gradient of the likelihood estimate is unbiased for the gradient of the likelihood.
+That is, it holds that
+    $$\E[\hat\lik_J(\theta)] = \lik(\theta),\;\; \E[\nabla_\theta \hat\lik_J(\theta)] = \nabla_\theta \lik(\theta).$$
+\end{conj}
+
+\begin{proof}
+    First, we need to establish the unbiasedness of the likelihood estimate. This is 
+    $$\hat\lik_J(\theta) := \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right)\prod_{n=1}^N L_n^\phi = \left(\frac{1}{J}\sum_{j=1}^J \prod_{n=1}^N \frac{g_{n,j}^{A,\theta}}{g_{n,j}^{A,\phi}} \right)\prod_{n=1}^N L_n^\phi.$$
+
+    We know from \cite{chopin20} that $\E[L_n^{\phi}] = \lik(\phi).$ It then suffices to show that
+    $$\E\left[\left(\frac{1}{J}\sum_{j=1}^J \prod_{n=1}^N \frac{g_{n,j}^{A,\theta}}{g_{n,j}^{A,\phi}} \right)\right] = \frac{\lik(\theta)}{\lik(\phi)}.$$
+
+    To obtain this, we need to see that the resampling under $\phi$, with corrected weights, is unbiased. That is, we need to show that for any bounded measurable functional $\varphi : \gX \to \R$,
+    $$\E\left[\frac{\sum_{j=1}^J w_{n,j}^{F, \theta} \varphi(x_{n,j}^{\theta, F})}{\sum_{j=1}^J w_{n,j}^{F, \theta}} \bigg| \gF_n \right] = \frac{\sum_{j=1}^J w_{n,j}^{P, \theta} \varphi(x_{n,j}^{P, \theta})}{\sum_{j=1}^J w_{n,j}^{P, \theta} }.
+$$
+
+    To see this, 
+    \begin{align*}
+        \sum_{j=1}^J w_{n,j}^{P, \theta} \varphi(x_{n,j}^{P, \theta}) 
+        &= \sum_{j=1}^J w_{n-1,j}^{F, \theta}\frac{g_{n,j}^{\theta}}{g_{n,j}^{\phi}}  \varphi(x_{n,j}^{P, \theta})
+    \end{align*}
+
+%     Observe that our differentiable simulator $\process(\cdot, \theta)$ and choice of weights now satisfy:
+
+%     - A Feynman-Kac model $\left\{M_t, G_t\right\}$ such that:
+
+% - the weight function $G_t$ may be evaluated pointwise (for all $t$ );
+
+% - it is possible to simulate from $\mathbb{M}_0\left(\mathrm{~d} x_0\right)$ and from $M_t\left(x_{t-1}, \mathrm{~d} x_t\right)$ (for any $x_{t-1}$ and $t$ ).
+
+% - The number of particles $N$.
+
+% - The choice of an unbiased resampling scheme (e.g.
+multinomial, residual, stratified, systematic), that is, an algorithm to generate variables $A_t^{1: N}$ in $1: N$ such that $\mathbb{E}\left[\sum_{m=1}^N \mathbbm{1}\left\{A_t^m=n\right\}\right]=N W_{t-1}^n$; see previous chapter for more details.
+\end{proof}
+
+\begin{lem}[Consistency of MOP-1 Gradient Estimate]
+    Consider the case of MOP-$\alpha$ when $\alpha=1$.
+Assume the gradient estimate obtained by the gradient of the log-likelihood estimate with J particles, $\nabla_\theta \hat\ell_J(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}$ is uniformly bounded by $G(\theta)$ for all $J$, and that the likelihood has two continuous derivatives.
+Then the gradient estimate, $\nabla_\theta \hat\ell_J(\theta)$, is consistent for the score.
+    \label{lem:mop-grad-consistency}
+\end{lem}
+\begin{proof}
+
+    % We know that the variance of the MOP-1 gradient estimate $\nabla_\theta \hat\ell_J(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}$ goes to zero coordinate-wise as $J\to\infty$ by the particle filter law of large numbers for bounded functionals of particles (\kevin{Prove general LLN for functionals of MOP under boundedness assumptions in L2, but maybe Hoeffding may be enough to show this, because we only want convergence in probability, we just need to show the expectation converges}), so it suffices to show that it is asymptotically unbiased.
+
+    % Write $\hat\pi_{n,J}$ for the filtering distribution at timestep $n$ and with $J$ particles, $\hat\pi_{1:N,J}$ for the sequence of filtering distributions from timesteps $1$ to $N$, and $\pi_n, \pi_{1:N}$ for the respective posteriors.
+
+    % From Lemma \ref{lem:lik-proper-weight}, we know that $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1:n-1}^*, \theta)$ is consistently estimated by the conditional likelihood estimate obtained by MOP-1, for all $\theta$ in an open neighborhood of $\phi$.
+The likelihood is simply the product of the conditional likelihoods, and similarly for the estimates, so the likelihood $\lik(\theta)$ is also consistently estimated by the likelihood estimate obtained by MOP-1, $\hat\lik_J(\theta)$, for all $\theta$ in an open neighborhood of $\phi$, $\gN(\phi)$.
+
+    % We therefore have that 
+    % $$\lim_{J\to\infty} \E[\hat\lik_J(\theta)] = \lik(\theta)$$
+    % pointwise for all $\theta \in \gN(\phi)$.
+Given that the gradient is bounded uniformly for all $J$ and consistency for the likelihood estimate holds for all $\theta$ in an open neighborhood of $\phi$ (and therefore all $\theta'$ in a sufficiently small neighborhood of $\theta$), the bounded convergence theorem allows us to switch the expectation and derivative:
+    % $$\nabla_\theta \E[\hat\lik_J(\theta)] 
+    % = \E[\nabla_\theta \hat\lik_J(\theta)].$$
+
+    % Take limits on both sides to find that 
+    % $$\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)] 
+    % = \lim_{J \to \infty} \E[\nabla_\theta \hat\lik_J(\theta)].$$
+
+    % We then need to swap the limit and derivative on the left-hand side.
+This requires uniform convergence of $\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]$, i.e.
+that $$\sup_{\theta \in \gN(\phi)} ||  \nabla_\theta \E[\hat\lik_J(\theta)] - \lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]||_{\infty} \to 0 \text{ as } J \to \infty,$$
+    % or that $\forall \epsilon>0$ $\exists N\in \mathbb{N}$ such that $n>N => ||  \nabla_\theta \E[\hat\lik_J(\theta)] - \lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]||_{\infty} < \epsilon$ $\forall \theta \in \gN(\phi).$ However, it is easier to instead use an indirect argument.
+
+    % We then need to swap the limit and derivative on the left-hand side. This requires uniform convergence.
+
+    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point we wish to evaluate the gradient at.
+The sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$, by Assumption \ref{assump:local-bounded-derivative}.
+The sequence is also uniformly equicontinuous.
+To see this, by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J(\theta, \omega)|_{\theta=\theta'}$ is also bounded by $H$ for almost every $\omega\in \Omega$ and every $\theta' \in \gN(\phi)$, including $\theta=\phi$.
+A set of functions with derivatives bounded by the same constant is uniformly Lipschitz, and therefore uniformly equicontinuous.
+So the sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ is uniformly equicontinuous for almost every $\omega \in \Omega$.
+    Explicitly, for almost every $\omega \in \Omega$ and every $\epsilon>0$, there exists some $\delta(\omega)>0$ such that for every $||\theta - \theta'||_{\infty}<\delta$ and every $J \in \mathbb{N}$ we have that
+    $$||\nabla_\theta \hat\lik_J(\theta, \omega)-\nabla_\theta \hat\lik_J(\theta', \omega)||_\infty < \epsilon.$$
+
+    \kevin{Our previous understanding was that the target is differentiable, and the sequence is differentiable in increasingly small open neighborhoods.
+This may not be true.
+The neighborhoods the sequence is differentiable in should be independent of $J$, as the resampling does not change with $\theta$ as it is dependent on $\phi$.
+Either way, the target is in each of the neighborhoods, and this may be sufficient in itself.}
+
+    Then, by Arzela-Ascoli, there exists a uniformly convergent subsequence.
+We claim that there is only one subsequential limit.
+We know that when as the gradient is bounded by Assumption \ref{assump:local-bounded-derivative} we can treat the gradient as a bounded functional.
+So by Lemma \ref{lem:posterior-proper-weights} the sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$, and there is therefore only one subsequential limit.
+The sequence therefore converges uniformly to its limit $\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega).$ Therefore, with uniform convergence for the derivatives established, we can swap the limit and derivative, and obtain that for almost every $\omega \in \Omega$,
+    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta, \omega).$$
+
+    From Lemma \ref{lem:lik-proper-weight}, we know that
+    $$\hat{\lik}_J(\theta, \omega) \stackrel{a.s.}{\to} \lik(\theta).$$
+
+    We then have that for almost every $\omega \in \Omega$, 
+    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta, \omega) = \nabla_\theta \lik(\theta),$$
+    as we wanted. The result then follows by the continuous mapping theorem.
+
+    % Taking expectations on both sides and using the bounded convergence theorem to interchange limits and expectations, we see that
+    % $$\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)] = \E[\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta)] = \E[\nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta)] = \nabla_\theta \lim_{J \to \infty} \E[\hat\lik_J(\theta)].$$
+
+    % It then follows that
+    % $$\nabla_\theta\lik(\theta) 
+    % = \nabla_\theta \lim_{J\to\infty} \E[\hat\lik_J(\theta)] 
+    % = \lim_{J\to\infty} \nabla_\theta \E[\hat\lik_J(\theta)] 
+    % = \lim_{J\to\infty} \E[\nabla_\theta \hat\lik_J(\theta)].$$
+
+    % We then have mean-square consistency for the gradient of the likelihood estimate. The result then follows by the continuous mapping theorem.
+
+    \kevin{Explain that the difference in formula between this and Poyiadjis is due to us taking the derivative w.r.t. the particles in the chain rule. }
+
+    \kevin{Alternative argument: Show gradient is unbiased, after showing the likelihood is unbiased in an open neighborhood of phi.
+That should hold by Chopin's argument for unbiased resampling to nonconstant weights.
+Gradient unbiased if can swap expectation and derivative, I think with two continuous derivatives and boundedness, might need BCT/DCT.
+No need to swap limit and derivative in this case, or worrying about uniform limits.}
+
+\end{proof}
+
 \kevin{Show bias is order N, variance is order N when alpha below 1. That implies optimal choice of alpha is not 1. }
+
+\section{Analysis of MOP-$\alpha$ Variance}
+
+We first begin with a lemma based on the work of \cite{yu94}. \cite{kausik23} simplify this to the special case of Markov chains. We further adapt this to our setting below.
+\begin{lem}
+    Let $h: \gX^N \to \R^n$ be a bounded measurable function with bound $C$ on trajectories of points in the state space $\gX$ (by considering the domain $\gX^N$, this also encompasses functions of segments of trajectories).
+Further assume that the underlying Markov chain $(X_n)_{n=1}^N$ has mixing time $t_{\text {mix}}=t_{\text{mix}}\left(\frac{1}{4}\right)$.
+Consider the joint distribution $\chi$ over the product of the $\sigma$-algebras of $S$ such segments separated by $s$ timesteps, with marginals $\chi_i$.
+Let the product distribution of the marginals $\chi_i$ be called $\Xi$.
+Then, we have
+$$
+\left|\mathbb{E}_\chi h-\mathbb{E}_{\Xi} h\right| \leq \frac{4 C(S-1)}{4^{s\cdot t_{\text{mix}}}}.
+$$
+
+\end{lem}
+\begin{proof}
+    Remember that each of our Markov processes is mixing, so there exists $t_{m i x, j}=t_{m i x, j}\left(\frac{1}{4}\right)$ and a stationary distribution $d_j$ so that $T V\left(P_j^n(x, \cdot), d_j\right)<\frac{1}{4}$ for $n \geq t_{m i x, j}$.
+Let $t_{m i x}=\max _j t_{m i x, j}$.
+Since the decay in total variation distance is multiplicative, $T V\left(P_j^n(x, \cdot), d_j\right)<\left(\frac{1}{4}\right)^c$ for all $j$ and $n \geq c t_{m i x}$.
+This implies that
+    $$
+    \max _j T V\left(P_j^n(x, \cdot), d_j\right)<\left(\frac{1}{4}\right)^{\frac{T_n}{4 t_{m i x}}-1}=4 \lambda^n
+    $$
+    where $\lambda=\left(\frac{1}{4}\right)^{\frac{1}{t_{m i x}}}$.
+\end{proof}
+
+It is known from \cite{poyiadjis11} that the variance of MOP-1 is no smaller than $O(N^2)$, with an analogous upper bound on the MSE. We now address the case of $\alpha<1$.
+
+We can decompose the MOP-$\alpha$ estimator as follows. When $\theta=\phi$,
+$$\hat{\mathcal{L}}(\theta):=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J w_{n, j}^{P, \phi}}=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J (w_{n-1, j}^{F, \phi})^\alpha}.$$
+
+To illustrate the proof strategy, we first analyze the special case when $\alpha=0.$ We then have:
+\begin{equation}
+    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
+\end{equation}
+
+Similarly to the proof of Lemma \ref{lem:mop-0-formula}, its gradient when $\theta=\phi$ is therefore 
+\begin{align*}
+    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
+    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)} \\
+    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta s_{n,j}}{\sum_{j=1}^J s_{n,j}} \\
+    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)}{f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \phi}; \phi)} \\
+    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
+\end{align*}
+where we use the log-derivative trick, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again.
+The point here was that we establish that the gradient of the log-likelihood estimate is given by the sum of terms over all $N$ and $J$:
+$$\nabla_\theta \hat{\ell}(\theta) = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right).$$
+
+\kevin{Ed: Claim 11.4.2, Chopin, is the key to reasoning about this.
+Establish geometric mixing, that $\text{Cov}(\nabla_\theta\hat\ell_n, \nabla_\theta\hat\ell_m) \leq \rho^{|n-m|}V_n$, where $V_n$ is asymptotically stable (the latter only holds for $\alpha<1$, as only then do we get asymptotic stability, otherwise it increases linearly in $n$).
+Geometrically decreasing covariances and stable variances over time are the same thing.
+Then, we can bound the total variance by $\text{Var}(\sum_n \nabla_\theta \hat\ell_n(\theta)) = \sum_{m,n} \text{Cov}(\nabla_\theta\hat\ell_n(\theta), \nabla_\theta\hat\ell_m(\theta)) = NV_n + 2(N-1)\rho V_n + 2(N-3)\rho^2V_n + ...$}
 
@@
 
-The analysis in this section roughly follows the analysis in Roosta-Khorasani and Mahoney \cite{mahoney2016subsampled}, except with the caveat that none of the matrix concentration bounds they use apply here as the particles are dependent.
+The analysis in this section roughly follows the analysis in \cite{mahoney16}, except with the caveat that none of the matrix concentration bounds they use apply here as the particles are dependent.
 We instead use the concentration inequality from Del Moral and Rio \cite{delmoral2011ci} to bound the gradient and Hessian estimates.
@@
 
-\kevin{Show MOP-1 is unbiased. Extend analysis to not-1 case. }
-
 \begin{lem}[Concentration of Measure for Gradient Estimate]
     \label{lemma:grad_bound}
-    Assume $\text{osc} \left(\frac{\partial}{\partial_\theta} \hat{\ell}_\theta(x_{n,j}^P) \right) \leq 1$.
-For $||\nabla_\theta \hat{\loglik}_\theta(x_{n,j}^P) - \nabla_\theta \loglik_\theta(x^P)|| = ||\nabla_\theta \hat{\loglik}(\theta) - \nabla_\theta \loglik(\theta)||$ to be bounded by $\epsilon$ with probability $1-\delta$, we require
-    \begin{equation}
-        J > \max\left\{\frac{r_t\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{6p}{\delta}\right)\right)\right),\;\; \frac{2p\log(6p/\delta)}{\epsilon^2}\beta_t^2, \;\;\frac{G(\theta)^2}{\epsilon^2}p\log\left(\frac{6p}{\delta}\right)\right\}
-    \end{equation}
-    where $G(\theta)$ is defined as in Assumption \ref{assump:local-bounded-derivative}, $h(t) = \frac{1}{2}(t - \log(1+t))$, and $\beta_t$ and $r_t$ are two additional model-specific parameters that do not depend on $J$.
+    Consider the gradient estimate obtained by MOP-1, which we know by Lemma \ref{lem:mop-grad-consistency} is consistent for the score, where $\theta = \phi$.
+For $||\nabla_\theta \hat{\loglik}(\theta) - \nabla_\theta \loglik(\theta)||_2$ to be bounded by $\epsilon$ with probability $1-\delta$, we require
+    \begin{align}
+    J > \max\left\{2NG'(\theta)\frac{r_N\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{2p}{\delta}\right)\right)\right), 8N^2G'(\theta)^2\beta_N^2\frac{p\log(2p/\delta)}{\epsilon^2}\right\},
+    \end{align}
+    where $G'(\theta)$ is defined as in Assumption \ref{assump:bounded-measurement}, $h(t) = \frac{1}{2}(t - \log(1+t))$, and $\beta_N$ and $r_N$ are two additional finite model-specific constants that do not depend on $J$, but do depend on $N$, as defined in \cite{delmoral2011ci}.
+
+    Equivalently, with probability at least $1-\delta$, it holds that
+    \begin{align}
+        ||\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G'(\theta)N\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).
+    \end{align}
 \end{lem}
 
+\paragraph{Remark:} According to \cite{delmoral2011ci}, under some regularity conditions, $r_N$ and $\beta_N$ are linear in the trajectory length $N$.
+Additionally, if the system is stationary enough, $NG'(\theta)$ roughly scales with $G(\theta)$, so one can consider the error to be roughly linear in the trajectory length, $O(G(\theta)N) \approx O(G'(\theta)N^2)$, if all these conditions are fulfilled.
+This corresponds to the finding by \cite{poyiadjis11} that the variance of the estimate is quadratic in the trajectory length.
+
 \begin{proof}
@@
 We will seek to use the concentration inequality of \cite{delmoral2011ci} to bound the deviation of the gradient estimate from the gradient of the negative log-likelihood in the sup norm with a union bound.
-To use this, we will first need to resample the particles once more to reassign the normalized weights to $1/J$.
-
-Perform another resampling step at time $T$ to obtain indices $k_j \sim \text{Categorical}(\bar{w}_{T,1}, ..., \bar{w}_{T,J})$.
-Obtain ancestral trajectories of surviving time $T$ filtering particles $x_{k_j, 1:T}^{A}$.
-Define $g_j(\theta) = \nabla_\theta -\log L_{T, k_j}$, the gradient of the negative logarithm of the accumulated importance weights of particle $k_j$ along its ancestral trajectory.
-
-We can bound
-\begin{align}
-    ||\nabla_\theta \hat{\loglik}(\theta) - \nabla_\theta \loglik(\theta)|| 
-    \leq 
-    \left\lVert (\nabla_\theta - \hat{\loglik}(\theta)) - \frac{1}{J}\sum_{j=1}^J g_j(\theta)\right\rVert 
-    + \left\lVert\frac{1}{J}\sum_{j=1}^J g_j(\theta) - (\nabla_\theta -\loglik(\theta))\right\rVert
-\end{align}
-
-We begin by examining the first term. This term bounds the error of the gradient estimate that the differentiable particle filter provides from the gradient estimate after the final resampling step.
-
-This amounts to solving the following problem.
-Say you have a weighted mean $\sum w_j x_j$, where $j=1,\dots,J$.
-The $x_j$ are fixed numbers bounded by some constant $G$ and the $w_j$ are non-negative and sum to $1$.
-Resample indices $k_j, j=1,\dots,J$ according to the multinomial distribution to get $x_{k_j}$ for each $j$, and also $J^{-1} \sum x_{k_j}$.
-We want to bound $|\sum w_j x_j - J^{-1} \sum x_{k_j}|$.
-
-By a simple application of Hoeffding's inequality and a union bound, we have
-
-\begin{align}
-    \prob\left(\exists \; j \;s.t.\; \left|(\nabla_\theta - \hat{\loglik}(\theta)) - \frac{1}{J}\sum_{j=1}^J g_j(\theta)\right| \geq \epsilon \right)
-    &\leq 2p\exp\left(\frac{-2\epsilon^2J}{G(\theta)^2}\right) 
-\end{align}
-
-We require \begin{equation}
-    J > \frac{G(\theta)^2}{2\epsilon^2}\log(2p/\delta)
-\end{equation} particles to bound $\left\lVert (\nabla_\theta - \hat{\loglik}(\theta)) - \frac{1}{J}\sum_{j=1}^J g_j(\theta)\right\rVert_\infty < \epsilon$ with probability $1-\delta$, where $G(\theta)$ is defined as in Assumption \ref{assump:local-bounded-derivative}.
-
-We can bound the sup norm by $||x||_\infty \leq \sqrt{p}||x||_2$.
-So to bound $\left\lVert (\nabla_\theta - \hat{\loglik}(\theta)) - \frac{1}{J}\sum_{j=1}^J g_j(\theta)\right\rVert < \epsilon$ with probability $1-\delta/3$, we scale the previous $\epsilon$ to $\epsilon/\sqrt{p}$ and therefore require
-
-\begin{equation}
-    J > \frac{G(\theta)^2}{2\epsilon^2}p\log\left(\frac{6p}{\delta}\right)
-\end{equation}
-
-We now tackle the second term.
-Write $g_{ij}$ for the $i$-th component of the gradient of the negative log-likelihood with respect to the $j$-th particle, $(-\bar{w}_{T,\np} \nabla_\theta\log L_{T,j})_i$.
-
-Apply the Hoeffding-type concentration inequality from Del Moral and Rio \cite{delmoral2011ci} and another union bound to find that
-
-\begin{align}
-    \left\lVert\frac{1}{J}\sum_{j=1}^J g_{ij} - \left(-\frac{\partial}{\partial \theta_i}\ell(\theta)\right)\right\rVert_\infty \leq \frac{r_t}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_t 
-\end{align}
-
-with probability $1-2p\exp(-t)$.
-We split the remaining $2\delta/3$ failure probability among these two terms, to find $\delta/3=2p\exp(-t) => t=\log(6p/\delta)$.
+Fix some $\theta \in \gN(\phi)$.
+
+Recall that as $w_{N, j}^{F, \theta}=1$ when $\theta=\phi$,
+$$
+\nabla_\theta \hat{\ell}(\theta):=\frac{1}{J} \sum_{j=1}^J \nabla_\theta w_{N, j}^{F, \theta}=\frac{1}{J} \sum_{j=1}^J\sum_{n=1}^N  w_{N, k_j}^{P, \theta} \nabla_\theta \log g_{n,k_j}^{A,\theta} = \frac{1}{J} \sum_{j=1}^J\sum_{n=1}^N \nabla_\theta \log g_{n,k_j}^{A,\theta}.
+$$
+
+Define $\varphi_n^i(x_{n,j}^{F,\theta}) := \frac{\partial}{\partial\theta_i} \log g_{n,k_j}^{A,\theta}$, which is a functional of the filtering particles $x_{n,j}^{F,\theta} = x_{n,k_j}^{P,\theta}$.
+These are bounded measurable functionals bounded by $G'(\theta)$ by Assumption \ref{assump:bounded-measurement}.
+Therefore, these have bounded oscillation, satisfying the requirement that $\text{osc} \left(\frac{\partial}{\partial\theta_i} \varphi_i(x_{n,j}^{P,\theta}) \right) \leq G'(\theta)$.
+\cite{delmoral2011ci} in fact assume $\text{osc}(f) \leq 1$, so we simply scale their bound accordingly.
+
+\kevin{Put some justification/make the setup of delmoral2011ci concrete here. Concentration of gradient estimate under particle measure to limit under posterior measure}
+
+Apply the Hoeffding-type concentration inequality from Del Moral and Rio \cite{delmoral2011ci} and a union bound over each $\varphi_n^i(x_{n,j}^{F,\theta})$, totaling $N$ timesteps and $p$ parameters, to find that
+
+\begin{align}
+    \max_{n=1,...,N} \left\lVert\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log g_{n,k_j}^{A,\theta} - \nabla_\theta \ell_n(\theta) \right\rVert_{\infty} \leq G'(\theta)\left(\frac{r_N}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_N \right)
+\end{align}
+
+with probability at least $1-2Np\exp(-t)$. It therefore holds that with the same probability, that
+
+\begin{align}
+    \left\lVert\frac{1}{J}\sum_{j=1}^J\sum_{n=1}^N\nabla_\theta \log g_{n,k_j}^{A,\theta} - \nabla_\theta \ell(\theta) \right\rVert_{\infty} \leq G'(\theta)N\left(\frac{r_N}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_N \right)
+\end{align}
+
+We split the $\delta$ failure probability among these $2Np$ terms, to find $\delta\leq2Np\exp(-t) => t\leq\log(2Np/\delta)$.
 Here, $h(t) = \frac{1}{2}(t - \log(1+t))$.
@@
 
-The analogous bound for the 2-norm falls out by scaling the right-hand side by $\sqrt{p}$, to require
-
-\begin{align}
-    \epsilon = \frac{r_t\sqrt{p}}{J}(1+h^{-1}(\log(6p/\delta))) + \sqrt{\frac{2p\log(6p/\delta)}{J}}\beta_t
-\end{align}
-
-and therefore need
-
-\begin{align}
-    J > \max\left\{\frac{r_t\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{6p}{\delta}\right)\right)\right), \frac{2p\log(6p/\delta)}{\epsilon^2}\beta_t^2\right\}
-\end{align}
-
-Overall, we finally require
-
-\begin{equation}
-        J > \max\left\{\frac{r_t\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{6p}{\delta}\right)\right)\right),\;\; \frac{2p\log(6p/\delta)}{\epsilon^2}\beta_t^2, \;\;\frac{G(\theta)^2}{\epsilon^2}p\log\left(\frac{6p}{\delta}\right)\right\}
-\end{equation}
+The analogous bound for the 2-norm follows from scaling the right-hand side by $\sqrt{p}$, to require
+
+\begin{align}
+    ||\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G'(\theta)N\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right)
+\end{align}
+
+and we therefore need 
+\begin{align}
+    J > \max\left\{2NG'(\theta)\frac{r_N\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{2p}{\delta}\right)\right)\right), 8N^2G'(\theta)^2\beta_N^2\frac{p\log(2p/\delta)}{\epsilon^2}\right\}.
+\end{align}
 
@@
     \label{lemma:hess_bound}
-    Assume that the Hessian of the negative log-likelihood $H=\sum_{j=1}^J \E H_j$ has a minimum eigenvalue $\gamma > 0$, that $\E \lambda_{\min} (H_j) = \gamma' > 0$, and finally that $\text{osc}(\lambda_{\min}) \leq 1$.
+    Assume that the Hessian of the negative log-likelihood $H=\sum_{j=1}^J \E H_j$ has a minimum eigenvalue $0<\gamma<1$, and that $\E \lambda_{\min} (H_j) = \gamma' > 0$.
 
@@
 
-In this analysis, we largely follow the proof of Theorem 6 in \cite{mahoney2016subsampled}.
+In this analysis, we largely follow the proof of Theorem 6 in \cite{mahoney16}.
 
 Define $\theta_\eta = \theta_n + \eta p_n$, where $p_n=-(H(\theta_n))^{-1}g(\theta_n)$.
-As in Roosta-Khorasani and Mahoney \cite{mahoney2016subsampled}, we want to show there is some iteration-independent $\tilde{\eta}>0$ such that the Armijo condition
+As in Roosta-Khorasani and Mahoney \cite{mahoney16}, we want to show there is some iteration-independent $\tilde{\eta}>0$ such that the Armijo condition
 \begin{equation}
@@
 
-By an argument found in the beginning of the proof of Theorem 6 in \cite{mahoney2016subsampled}, we have that choosing $J$ such that $||\nabla_\theta\hat{\ell}(\theta_n) - \nabla_\theta \ell(\theta_n)|| \leq \epsilon$ and $\lambda_{\min}(H(\theta_n)) \geq c$ with probability $(1-\delta)^2$ for each $n$ yields
+By an argument found in the beginning of the proof of Theorem 6 in \cite{mahoney16}, we have that choosing $J$ such that $||\nabla_\theta\hat{\ell}(\theta_n) - \nabla_\theta \ell(\theta_n)|| \leq \epsilon$ and $\lambda_{\min}(H(\theta_n)) \geq c$ with probability $(1-\delta)^2$ for each $n$ yields
 

### 2024-03-14 mop/draft.tex
Updates from Overleaf

@@
 
+\begin{aspt}[Bounded Process Model]
+\label{assump:bounded-process}
+    There exist $\underbar{M}, \bar{M}$ such that $0 < \underbar{M} \leq f_{X_n|X_{n-1}}(x_n | x_{n-1}) \leq \bar{M} < \infty$. 
+\end{aspt}
+
 \begin{aspt}[Bounded Measurement Model]
 \label{assump:bounded-measurement}
-There is an $\epsilon>0$ such that $\epsilon^{-1}>f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)>\epsilon$ and a $G'(\theta)>0$ such that $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty < G'(\theta)$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
+There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G} < \infty$ and $\infty > G' \geq G'(\theta)>0$ such that $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty < G'(\theta) \leq G'$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
 \end{aspt}
@@
     $$
-    \|\nabla_\theta \ell(\theta, \phi, \omega, J)\|_\infty <G(\theta) \leq G<\infty, \;\;
-    \|\nabla^2_\theta \ell(\theta, \phi, \omega, J)\|_\infty <G(\theta) \leq G<\infty
+    \|\nabla_\theta \ell(\theta, \phi, \omega, J)\|_\infty < NG'(\theta) \leq G(\theta) \leq G<\infty, \;\;
+    \|\nabla^2_\theta \ell(\theta, \phi, \omega, J)\|_\infty <N^2G'(\theta)^2 \leq G(\theta)^2 \leq G^2<\infty
     $$
@@
 \begin{equation*}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\phi}}= \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\phi}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
+    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}}= \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\theta}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
 \end{equation*}
@@
 
-$$w_{n,j}^{P,\theta} = w_{n-1,\theta}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,j}^{A,\theta}}{g_{i,j}^{A,\phi}},$$
-where we use the $(\cdot)^A$ superscript to denote the ancestral trajectory of the $j$-th particle at timestep $n$.
+$$w_{n,j}^{P,\theta} = w_{n-1,j}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}},$$
+where we use the $(\cdot)^A$ superscript to denote the ancestral trajectory of the $j$-th prediction or filtering particle at timestep $n$.
 
 Using the log-derivative identity again,
-$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,j}^{A,\theta}}{g_{i,j}^{A,\phi}}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,j}^{A,\theta} - \log g_{i,j}^{A,\phi}\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,j}^{A,\theta}.$$
+$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,j}^{A,P,\theta} - \log g_{i,j}^{A,P,\phi}\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,j}^{A,P,\theta}.$$
 
 So 
-$$  \nabla_\theta \sum_{n=1}^N \log g_{n,j}^{A,\theta} = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,j}^{A,\theta}\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,j}^{A, P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right),$$
+$$  \nabla_\theta \sum_{n=1}^N \log g_{n,j}^{A,\theta} = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,j}^{A,P,\theta}\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,j}^{A, P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right),$$
 
 and 
-$$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{n=1}^N \nabla_\theta \log g_{n,j}^{A,\theta} = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$
+$$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{n=1}^N \nabla_\theta \log g_{n,j}^{A,P,\theta} = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$
 
@@
 
+Note that one would believe that the variance blows up the moment $\theta\neq\phi$. This is true, as 
+\begin{equation*}
+    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right),
+\end{equation*}
+and when $\theta\neq\phi$, $w_{N,k_j}^{P,\theta} = O(c^N)$.
+When $\theta=\phi$, this is a special case of the \cite{poyiadjis11} estimator, which has $O(N^2)$ variance by a property of functionals of the particle filter by \cite{delMoral03}.
+
 \begin{lem}
@@
 
-\section{Analysis of MOP-$\alpha$ Variance}
-
-We first begin with a lemma based on the work of \cite{yu94}. \cite{kausik23} simplify this to the special case of Markov chains. We further adapt this to our setting below.
-\begin{lem}
-    Let $h: \gX^N \to \R^n$ be a bounded measurable function with bound $C$ on trajectories of points in the state space $\gX$ (by considering the domain $\gX^N$, this also encompasses functions of segments of trajectories).
-Further assume that the underlying Markov chain $(X_n)_{n=1}^N$ has mixing time $t_{\text {mix}}=t_{\text{mix}}\left(\frac{1}{4}\right)$.
-Consider the joint distribution $\chi$ over the product of the $\sigma$-algebras of $S$ such segments separated by $s$ timesteps, with marginals $\chi_i$.
-Let the product distribution of the marginals $\chi_i$ be called $\Xi$.
-Then, we have
-$$
-\left|\mathbb{E}_\chi h-\mathbb{E}_{\Xi} h\right| \leq \frac{4 C(S-1)}{4^{s\cdot t_{\text{mix}}}}.
-$$
-
-\end{lem}
-\begin{proof}
-    Remember that each of our Markov processes is mixing, so there exists $t_{m i x, j}=t_{m i x, j}\left(\frac{1}{4}\right)$ and a stationary distribution $d_j$ so that $T V\left(P_j^n(x, \cdot), d_j\right)<\frac{1}{4}$ for $n \geq t_{m i x, j}$.
-Let $t_{m i x}=\max _j t_{m i x, j}$.
-Since the decay in total variation distance is multiplicative, $T V\left(P_j^n(x, \cdot), d_j\right)<\left(\frac{1}{4}\right)^c$ for all $j$ and $n \geq c t_{m i x}$.
-This implies that
-    $$
-    \max _j T V\left(P_j^n(x, \cdot), d_j\right)<\left(\frac{1}{4}\right)^{\frac{T_n}{4 t_{m i x}}-1}=4 \lambda^n
-    $$
-    where $\lambda=\left(\frac{1}{4}\right)^{\frac{1}{t_{m i x}}}$.
-\end{proof}
-
-It is known from \cite{poyiadjis11} that the variance of MOP-1 is no smaller than $O(N^2)$, with an analogous upper bound on the MSE. We now address the case of $\alpha<1$.
+\section{Analysis of MOP-$\alpha$ Error and Variance}
+
+We made Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} to leverage the results from \cite{karjalainen2023} on the forgetting of the particle filter.
+This is required to show error bounds on the gradient estimates that we provide -- namely that the error of the MOP-$1$ estimator that corresponds to that of \cite{poyiadjis11} is $O(N^2)$, and that the variance of MOP-$\alpha$ for any $\alpha<1$ is $O(N)$.
 
 We can decompose the MOP-$\alpha$ estimator as follows. When $\theta=\phi$,
-$$\hat{\mathcal{L}}(\theta):=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J w_{n, j}^{P, \phi}}=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J (w_{n-1, j}^{F, \phi})^\alpha}.$$
+$$\hat{\mathcal{L}}(\theta):=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J w_{n, j}^{P, \theta}}=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J (w_{n-1, j}^{F, \theta})^\alpha}.$$
 
@@
 
+Therefore,
+\begin{align*}
+    \text{Var}(\nabla_\theta \hat\ell(\theta) &= \frac{1}{J^2}\text{Var}\left(\sum_{n=1}^N\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    &= \frac{1}{J^2}\sum_{n=1}^N\text{Var}\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    &+ 2\sum_{m<n}\text{Cov}\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta)).
+\end{align*}
+
+Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
+We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
+$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
+i.e. the mixing time of the particle filter is $O(\log(J))$.
+
+The particle filter itself is therefore strongly mixing, with $\alpha$-mixing coefficients $\alpha(k) \leq (1-\epsilon)^{\floor{k/(c\log(J))}}$.
+Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
+Therefore, by Davydov's inequality and noting that $\nabla_\theta\hat\ell_n(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement},
+
+$$\text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) \leq G'(\theta)^2(1-\epsilon)^{\floor{\frac{|n-m|}{2c\log(J)}}}.$$
+
+We can then conclude with Popoviciu's inequality that 
+\begin{align*}
+    \text{Var}(\nabla_\theta \hat\ell(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta))\\
+    &\leq \frac{NG'(\theta)^2}{4J} + 2\sum_{n=1}^{N-1} (N-n) G'(\theta)^2(1-\epsilon)^{\floor{\frac{n}{2c\log(J)}}}\frac{G'(\theta)^2}{4J} \\
+    &\leq \frac{G'(\theta)^2}{4J} \left(N + 2\sum_{n=1}^{N-1} (N-n) G'(\theta)^2(1-\epsilon)^{\floor{\frac{n}{2c\log(J)}}}\right) \\
+    &\lesssim \frac{NG'(\theta)^2}{J}.
+\end{align*}
+
+\paragraph{Remark:} Note that the factor of $N$ that pops up here is due to the use of the unnormalized gradient.
+If one divides the gradient estimate by $N$ before usage, the variance does not depend on the horizon.
+
+\kevin{https://math.stackexchange.com/questions/3274996/is-a-transformed-mixing-process-is-still-mixing}
+
+%$$\text{Cov}(x_{n,1:J}^P, x_{m, 1:J}^P) \leq (1-\epsilon)^{\floor{\frac{|n-m|}{c\log(J)}}}$$
+
 \kevin{Ed: Claim 11.4.2, Chopin, is the key to reasoning about this.
@@
 
+Next, we analyze the case for $\alpha \in (0,1)$.
+
+Notice that
+\begin{align*}
+    w_{n,j}^{F,\theta}
+    &= \left(w_{n-1,k_j}^{F, \theta}\right)^\alpha\frac{g_{n,k_j}^\theta}{g_{n,k_j}^\phi} \\
+    &= \left(\left(w_{n-2,j}^{A,F, \theta}\right)^\alpha\frac{g_{n-1,j}^{A,F,\theta}}{g_{n-1,j}^{A,F,\phi}}\right)^\alpha\frac{g_{n,j}^{A,F,\theta}}{g_{n,j}^{A,F,\phi}} \\
+    &= \prod_{i=1}^n \left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}.
+\end{align*}
+
+Then, we have that
+\begin{align*}
+    \nabla_\theta\hat\ell(\theta) 
+    &= \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^Jw_{n,j}^{F,\theta}}{\sum_{j=1}^J\left(w_{n-1,j}^{F, \theta}\right)^\alpha}\right)= \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right).
+\end{align*}
+
+Consider the modified estimator, where we truncate weights beyond $k$ timesteps:
+\begin{align*}
+    \hat{s}_k(\theta) 
+    &:= \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right).
+\end{align*}
+
+Write $$w_{n,j}^{F,\theta,k} := \prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}$$ for the $k$-truncated weights.
+
+By Assumption \ref{assump:bounded-measurement} we have that
+
+\begin{align*}
+   w_{n,j}^{F,\theta} 
+    &\leq w_{n,j}^{F,\theta,k} \prod_{i=1}^{n-k}\left(\frac{\bar{G}}{\underbar{G}}\right)^{\alpha(n-i)}\\
+    &\leq w_{n,j}^{F,\theta,k} \left(\frac{\bar{G}}{\underbar{G}}\right)^{\alpha \sum_{i=1}^{n-k}(n-i)} \\
+    &\leq w_{n,j}^{F,\theta,k} \left(\frac{\bar{G}}{\underbar{G}}\right)^{\alpha\frac{n(n-1)-k(k-1)}{2}}
+\end{align*}
+
+We can then bound $||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||$:
+\begin{align*}
+    &\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \\
+    &= \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) \\
+    &= \sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\\
+    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\Bigg) \\
+    &= \sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)
+    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\Bigg)
+\end{align*}
+
+We first bound $||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||$:
+\begin{align*}
+    &\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \\
+    &= \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) \\
+    &= \sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\\
+    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\Bigg) \\
+    &=  \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=n-k}^{n}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) + \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right)
+\end{align*}
+
+By two applications of the log-derivative rule,
+\begin{align*}
+    \nabla_\theta \log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)
+    &= \frac{\sum_{j=1}^J\nabla_\theta \prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}} \\
+    &= \frac{\sum_{j=1}^J \left(\left(\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right) \nabla_\theta \log\left(\prod_{i=1}^n \left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\right)}{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}} \\
+    &= \frac{\sum_{j=1}^J \left(\left(\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)} \right)\left(\sum_{i=1}^n \alpha(n-i)\nabla_\theta \log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right)}{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}
+\end{align*}
+
 \section{Finite-Particle Analysis of Gradient Methods with Algorithm \ref{alg:ifad}}
@@
 The analysis in this section roughly follows the analysis in \cite{mahoney16}, except with the caveat that none of the matrix concentration bounds they use apply here as the particles are dependent.
-We instead use the concentration inequality from Del Moral and Rio \cite{delmoral2011ci} to bound the gradient and Hessian estimates.
+We instead use the concentration inequality from \cite{delmoral2011ci} to bound the gradient and Hessian estimates.
 In this section, we fix $\omega \in \Omega$ only within each filtering iteration, evaluate Algorithm \ref{alg:mop} at $\theta=\phi$, and analyze Algorithm \ref{alg:ifad} post-iterated filtering.
@@
     \begin{align}
-    J > \max\left\{2NG'(\theta)\frac{r_N\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{2p}{\delta}\right)\right)\right), 8N^2G'(\theta)^2\beta_N^2\frac{p\log(2p/\delta)}{\epsilon^2}\right\},
-    \end{align}
-    where $G'(\theta)$ is defined as in Assumption \ref{assump:bounded-measurement}, $h(t) = \frac{1}{2}(t - \log(1+t))$, and $\beta_N$ and $r_N$ are two additional finite model-specific constants that do not depend on $J$, but do depend on $N$, as defined in \cite{delmoral2011ci}.
+    J > \max\left\{2G(\theta)\frac{r_N\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{2p}{\delta}\right)\right)\right), 8G(\theta)^2\beta_N^2\frac{p\log(2p/\delta)}{\epsilon^2}\right\},
+    \end{align}
+    where $NG'(\theta) \leq G(\theta)$ are defined in Assumptions \ref{assump:bounded-measurement} and \ref{assump:local-bounded-derivative}, $h(t) = \frac{1}{2}(t - \log(1+t))$, and $\beta_N$ and $r_N$ are two additional finite model-specific constants that do not depend on $J$, but do depend on $N$ and $p$, as defined in \cite{delmoral2011ci}.
 
@@
     \begin{align}
-        ||\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G'(\theta)N\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).
+        ||\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).
     \end{align}
@@
 \paragraph{Remark:} According to \cite{delmoral2011ci}, under some regularity conditions, $r_N$ and $\beta_N$ are linear in the trajectory length $N$.
-Additionally, if the system is stationary enough, $NG'(\theta)$ roughly scales with $G(\theta)$, so one can consider the error to be roughly linear in the trajectory length, $O(G(\theta)N) \approx O(G'(\theta)N^2)$, if all these conditions are fulfilled.
-This corresponds to the finding by \cite{poyiadjis11} that the variance of the estimate is quadratic in the trajectory length.
+This corresponds to the finding by \cite{poyiadjis11} that the variance of the estimate is at least quadratic in the trajectory length, and their remark that the result of \cite{delMoral03} establishes that the $L_p$ error is bounded by $O(N^2J^{-1/2})$ after accounting for the sum over timesteps.
+Either way, usually in practice one considers the estimate normalized through dividing it by $N$, in which case we only need
+$$J > \max\left\{2G'(\theta)\frac{r_N\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{2p}{\delta}\right)\right)\right), 8G'(\theta)^2\beta_N^2\frac{p\log(2p/\delta)}{\epsilon^2}\right\}.$$
+Equivalently,
+$$||N^{-1}(\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta))||_2 \leq G'(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).$$
+
+It is well-known that bounds like these hide an exponential dependence in the dimension of the state space $p$.
+We take care to keep track of the terms depending on $p$ when they arise in front of us, but are unable to do so for the terms that arise from the analysis of \cite{delmoral2011ci}.
+
+\kevin{This is slightly alarming, in that the estimate really has $O(N^4)$ variance after accounting for summing over timesteps.
+In contrast, MOP-0 has only $O(N)$ variance after summing over timesteps.
+}
 
@@
 
-with probability at least $1-2Np\exp(-t)$. It therefore holds that with the same probability, that
-
-\begin{align}
-    \left\lVert\frac{1}{J}\sum_{j=1}^J\sum_{n=1}^N\nabla_\theta \log g_{n,k_j}^{A,\theta} - \nabla_\theta \ell(\theta) \right\rVert_{\infty} \leq G'(\theta)N\left(\frac{r_N}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_N \right)
+with probability at least $1-2Np\exp(-t)$.
+Note that the concentration inequality only considers the error from the expectation under the filtering distribution -- but we invoke the consistency of MOP-$1$ shown in Lemma \ref{lem:mop-grad-consistency} to establish that the expectation under the filtering distribution is in fact the score.
+It therefore holds that with the same probability, that when summing over $N$, as $NG'(\theta) \leq G(\theta)$,
+
+\begin{align}
+    \left\lVert\frac{1}{J}\sum_{j=1}^J\sum_{n=1}^N\nabla_\theta \log g_{n,k_j}^{A,\theta} - \nabla_\theta \ell(\theta) \right\rVert_{\infty} 
+    &\leq G'(\theta)N\left(\frac{r_N}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_N \right)\\
+    &\leq G(\theta)\left(\frac{r_N}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_N \right)
 \end{align}
@@
 \begin{align}
-    ||\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G'(\theta)N\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right)
+    ||\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right)
 \end{align}
@@
 \begin{align}
-    J > \max\left\{2NG'(\theta)\frac{r_N\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{2p}{\delta}\right)\right)\right), 8N^2G'(\theta)^2\beta_N^2\frac{p\log(2p/\delta)}{\epsilon^2}\right\}.
+    J > \max\left\{2G(\theta)\frac{r_N\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{2p}{\delta}\right)\right)\right), 8G(\theta)^2\beta_N^2\frac{p\log(2p/\delta)}{\epsilon^2}\right\}.
 \end{align}

### 2024-03-14 mop/project.tex
Updates from Overleaf

--- 2024-01-27 mop/draft.tex
@@
 
-\title{Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation}
-\author{Kevin Tan, Giles J. Hooker, Edward L. Ionides}
+\title{Accelerated Inference for Partially Observed Stochastic Processes via Automatic Differentiation}
+\author{Kevin Tan}
 \date{}
@@
 \begin{document}
+
 \maketitle
-\begin{abstract}
 
-Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods.
-Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques due to (1) the requirement for simulation-based inference that does not require access to the system's transition probabilities and (2) the issue that widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters.
-We propose a new representation that embeds existing methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff.
-Further, we develop optimization algorithms suited to the Monte Carlo properties of the derivative estimate, including a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
-Promising numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from iterated filtering can beat current state-of-the-art methods on a challenging scientific benchmark problem.
-\end{abstract}
+\begin{center}
+    Under the supervision of Giles J. Hooker and Edward L. Ionides
+\end{center}
 
-\section{Introduction}
+\section{Background}
+A partially-observed Markov process (POMP) is defined as follows.
+Consider an unobserved Markov process $\{X_t, t \geq t_0\}$, and observations $Y_1,...,Y_N$ at timesteps $t_1,..., t_N$.
+The process is parameterized by an unknown parameter $\theta \in \Theta$, where $X \in \gX, Y \in \gY,$ and finally $\gX, \gY, \Theta \subseteq \R^n$.
+By a similar decomposition to that in \citet{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
+$$f_{X_{0: N}, Y_{1: N}}\left(x_{0: N}, y_{1: N} ; \theta\right)=f_{X_0}\left(x_0 ; \theta\right) \prod_{n=1}^N f_{X_n \mid X_{n-1}}\left(x_n \mid x_{n-1} ; \theta\right) f_{Y_n \mid X_n}\left(y_n \mid x_n ; \theta\right).$$
+
+We call $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ the process model, writing $\texttt{process}\left(x_n, \theta\right)$ for the simulator corresponding to the above process model, $f_{Y_n|X_n}\left(y_n \mid x_n, \theta\right)$ the measurement model, and will write $y_n^*$ for the actual values of the observations that were observed.
+
+The above are all defined on a single probability space $(\Omega, \Sigma, \prob),$ where $\omega \in \Omega$ denotes an element of the sample space that can be thought of as a random seed.
+For notational simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0.$
+
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}[scale=2]
+
+\node[black] (X_0) at (-3,0) {$X_0$};
+\node[black] (Xdot0) at (-2,0) {$...$};
+\node[black] (X_{t-1}) at (-1,0) {$X_{t-1}$};
+\node[black] (X_t) at (0,0) {$X_t$};
+\node[black] (X_{t+1}) at (1,0) {$X_{t+1}$};
+\node[black] (Xdot{t+1}) at (2,0) {$...$};
+\node[black] (X_T) at (3,0) {$X_T$};
+\node[black] (Y_0) at (-3,-1) {$Y_0$};
+\node[black] (Ydot0) at (-2,-1) {$...$};
+\node[black] (Y_{t-1}) at (-1,-1) {$Y_{t-1}$};
+\node[black] (Y_t) at (-0,-1) {$Y_t$};
+\node[black] (Y_{t+1}) at (1,-1) {$Y_{t+1}$};
+\node[black] (Ydot{t+1}) at (2,-1) {$...$};
+\node[black] (Y_T) at (3,-1) {$Y_T$};
+\path[->, black] (X_0) edge (Xdot0);
+\path[->, black] (Xdot0) edge (X_{t-1});
+\path[->, black] (X_{t-1}) edge (X_t);
+\path[->, black] (X_t) edge (X_{t+1});
+\path[->, black] (X_{t+1}) edge (Xdot{t+1});
+\path[->, black] (Xdot{t+1}) edge (X_T);
+\path[->, black] (X_0) edge (Y_0);
+\path[->, black] (X_{t-1}) edge (Y_{t-1});
+\path[->, black] (X_t) edge (Y_t);
+\path[->, black] (X_{t+1}) edge (Y_{t+1});
+\path[->, black] (X_T) edge (Y_T);
+\end{tikzpicture}
+\caption{Directed acyclic graph (DAG) of a POMP model.}
+\end{figure}
+
+We seek to solve the filtering problem, that is, the problem of estimating
+$$p(x_n|y_{1:n}),$$
+or the posterior distribution of the state given the current and all previous observations. This can be obtained by Bayesian updating, namely that
+\begin{enumerate}
+    \item Given a belief on the state space $x_{n-1}$, simulate forward according to $p(\hat x_{n}|x_{n-1})$ to obtain a prediction distribution $\hat x_n$ for the state at time $n$, and
+    \item Given a prediction $\hat{x}_n$ at time $n$, observe the measurement $y_n^*$ and perform a Bayesian update to the prediction according to $p(y_n^* | \hat x_{n-1})$ to obtain a belief on the state at time $n$, $x_n$.
+\end{enumerate}
+
+The particle filter is simply the Monte Carlo approximation to this procedure, obtained by maintaining a fixed number $J$ of simulations that we call particles.
+This induces a mixture Dirac measure on the state space, where we call the belief approximation $x_{n,j}^F$ to the filtering distribution $p(x_n|y_{1:n})$, and the prediction approximation $x_{n,j}^P$ to the prediction distribution $p(x_{n}|y_{1:n-1})$.
+The intractable integrals involved in the Bayesian update are circumvented with sums over the Dirac measures.
+
+At a high level, a particle filter keeps track of a set of particles $x_{n,j}$, sequentially approximating (1) the distribution of $X_{n}|y_{1:n-1}^*$ with a mixture Dirac measure $\frac{1}{J}\sum_{j=1}^J \tilde{w}_{n,j}\delta_{x_{n,j}^P}$ via simulation from the process model $p(x_n|x_{n-1})$ and (2) the distribution of $X_{n}|y_{1:n}^*$ with another mixture Dirac measure $\frac{1}{J}\sum_{j=1}^J w_{n,j}\delta_{x_{t,j}^P}$ via a reweighting of the particles in proportion to their likelihood.
+
+The particle filter approximates the Bayes filter through these two sequences of mixture Dirac measures.
+The Bayes filter is the "optimal" filter that updates the prediction distribution $p(x_{t}^P|y_{1:t-1}^*)$ with the prediction formula from \cite{sbied_lec3},
+\begin{equation}
+    p(x_{n}^P|y_{1:n-1}^*) = \int p(x_{n-1}^F|y_{1:n-1}^*)p(x_t|x_{t-1}) dx_{n-1},
+\end{equation}
+and the filtering distribution $p(x_t^F|y_{1:t}^*)$ via Bayes' theorem \citep{sbied_lec3},
+\begin{equation}
+    p(x_n^F|y_{1:n}^*) = \frac{p(y_n^*|x_n)p(x_n|y_{1:n-1}^*)}{\int p(y_n^*|x_n)p(x_n|y_{1:n-1}^*) dx_n}.
+\end{equation}
+
+This therefore provides a simple, general, and computationally efficient solution, yielding an approximation to the filtering distribution and log-likelihood.
+
+\section{Challenges}
 
@@
 
-This is the backbone behind the only simulation-based full-information maximum likelihood method available in the literature thus far, the improved iterated filtering algorithm of \citet{ionides15}.
+This is the backbone behind the only simulation-based full-information maximum likelihood method available in the literature thus far, the improved iterated filtering (IF2) algorithm of \citet{ionides15}.
 \citet{ionides15} utilize the bootstrap filter for perform maximum likelihood estimation with an iterated perturbed Bayes map, successfully tackling challenging problems in epidemiology that available Bayesian methodology fails to solve.
@@
 Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small.
-For example, though in practice the algorithm of \citet{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood.
-\citet{Ionides_mcap} attempt to address the third problem, albeit with a computationally-expensive approach.
+For example, though in practice the algorithm of \citet{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood, as we see below in Figure \ref{fig:if2fail}.
 
-\subsection{Automatic Differentiation for Particle Filters}
+\begin{figure}[H]
+    \centering
+    \includegraphics[scale=0.5]{imgs/095/if2fail.png}
+    \caption{Failure of iterated filtering on maximum likelihood estimation in the Dhaka cholera model of \cite{king08}.
+IF2 was run 100 times (with the 0th and 80th quantiles represented with the shaded area), but no run finds the MLE.
+The best run is about 7.5 log-likelihood units short.
+}
+    \label{fig:if2fail}
+\end{figure}
+
+\section{A Potential Solution}
+
+Given the success of stochastic gradient methods in machine learning, it is natural to consider the performance of the procedure of performing gradient descent on the log-likelihood estimate given by the particle filter.
+We seek to develop a method that performs better than IF2 on this problem, by running IF2 for a few iterations and taking advantage of its quick convergence to a neighborhood of the MLE, and then switching to stochastic gradient descent on the gradient estimate given by running automatic differentiation on the log-likelihood estimate from the particle filter.
+
+\section{Automatic Differentiation for Particle Filters}
 
@@
 
-\citet{blei2018vsmc} and \citet{jon2018diffpf} drop high-variance resampling terms from the gradient, leading to bias that does not vanish asymptotically according to \citet{corenflos21}.
-Though the approach of \citet{corenflos21} is promising, it is computationally expensive.
-\citet{doucet2022particlebased} provide another promising fixed-lag smoothing-based approach that is similar in spirit to our method (MOP-$\alpha$), but requires either (1) access to the process model transition densities or (2) the special case where the transition density factors into a policy that selects actions (in the reinforcement learning sense) and a deterministic model that maps states and actions to next-timestep states.
+In particular, only \cite{scibior21}, \cite{blei2018vsmc}, and \cite{corenflos21} have algorithms that are amenable to simulation-based inference without access to transition densities.
+Though the approach of \citet{corenflos21} is promising, it is computationally expensive, requiring runtime quadratic in the number of particles (unlike other approaches, which run in linear-time).
 
-Finally, \citet{scibior2021dpf} demonstrate that the estimators derived in \citet{doucet2011sf} can be attained with AD, albeit with a simple tweak to the particle filter that does not modify its likelihood estimates.
-Though their method is not fully compatible with simulation-based inference, we show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.
+\cite{scibior21} show that the estimator of \cite{blei2018vsmc} corresponds to the estimate of the gradient given by differentiating through the vanilla particle filter.
+This is therefore a very natural estimator for the score.
+However, \citet{blei2018vsmc} drops high-variance resampling terms from the gradient, leading to bias that does not vanish asymptotically according to \citet{corenflos21}.
 
-\subsection{Our Contributions}
+A linear-time consistent estimator of the score can be obtained via the estimator of \cite{poyiadjis11}, which \cite{scibior21} show can be obtained through a minor tweak on the particle filter that they call a "stop-gradient trick".
+However, this estimator has $O(N^2)$ variance, and the stop-gradient trick is not mathematically motivated.
+
+Of the other methods that are not compatible with simulation-based inference, \citet{doucet2022particlebased} provide another promising fixed-lag smoothing-based approach, but requires either (1) access to the process model transition densities or (2) the special case where the transition density factors into a policy that selects actions (in the reinforcement learning sense) and a deterministic model that maps states and actions to next-timestep states.
+
+\section{Our Contributions}
 
@@
 
-This is done through the construction of a smooth extension to the particle filter, as well as a \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) family of algorithms that (informally) encompasses the special case of AD of a vanilla PF (recovering the gradient estimator of \citet{blei2018vsmc}) when $\alpha=0$, and the DPF algorithm from \citet{scibior2021dpf} (that recovers the gradient estimator from \citet{doucet2011sf}) when $\alpha=1$.
+This is done through the construction of a smooth extension to the particle filter, as well as a \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) family of algorithms that (informally) encompasses the special case of AD of a vanilla PF (recovering the gradient estimator of \citet{blei2018vsmc}) when $\alpha=0$, and the DPF algorithm from \citet{scibior2021dpf} (that recovers the gradient estimator from \citet{poyiadjis11}) when $\alpha=1$.
 When $\alpha \in (0,1)$, this provides opportunities for optimizing a bias/variance tradeoff.
 
-We show these off-policy particle filters successfully target the posterior, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties.
-To cap off the theoretical results, we also derive a linear convergence rate for gradient methods on the particle filter in the presence of strong convexity, providing support for numerical optimization with AD for the particle filter.
+We will show that these off-policy particle filters successfully target the posterior, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties.
+This entails novel proofs showing a strong law of large numbers for triangular arrays under bounded particle weights and functionals, with the application of showing that a suitably reweighted particle filter still targets the posterior, even when resampling is carried out under some other arbitrary categorical distribution.
 
 In practice, we propose a hybrid algorithm that we call \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts gradient ascent (or a similar iterative first or second-order gradient method) with a coarse solution obtained from a few rounds of iterated filtering.
-While this requires a differentiable simulator compatible with the reparametrization trick (as previous literature such as \citet{corenflos21} also require), future work will deal with workarounds inspired by our framework that use likelihood ratios of transition densities in place of this.
-Promising numerical results indicate that IFAD beats IF2 (and by the transitive property, IF1, the Liu-West filter, and particle MCMC) on a challenging scientific benchmark problem.
-
-While the Monte-Carlo Adjusted Profile (MCAP) from \citet{Ionides_mcap} provides a profile likelihood confidence interval that takes into account both Monte Carlo profile error and statistical uncertainty in the likelihood, it is computationally inefficient.
-\kevin{TODO} We provide a method for incorporating the MCAP methodology (with the poor man's likelihood profile) in our IFAD algorithm that provides a more computationally efficient profile likelihood confidence interval.
-\kevin{Quantile regression for estimating the max/the envelope of the likelihood evaluations from AD, then fit MCAP/MQLE/MSLE?}
-
-Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual} that we call \texttt{pompy}.
-The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{jax} Python package from \citet{jax2018github}.
-
-\kevin{Terminology: Plug-and-play (clearly definable, but not well-known, define at start of paper and don't use in title/abstract), likelihood-free maximum likelihood (seemingly an oxymoron, rationale being there is no need to evaluate density, bayesians call likelihood densities, plus we only approximate the likelihood and don't really have it), simulation-based inference (loaded term)}
-
-\subsection{Significance Statement}
-Many scientific models involve highly nonlinear stochastic dynamical systems which can be observed only via noisy and incomplete measurements.
-Under the Markov assumption on system dynamics, previous work has provided methods of performing inference for these models.
-In particular, prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics.
-We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
-Our new method outperforms previous approaches on a challenging problem in epidemiology.
-This is implemented in a precursor to a Python counterpart to the popular \texttt{pomp} package by \citet{king16, king2017pompmanual} that includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation.
-
-\kevin{TODO: (1) Fix beta smoothness to d(n) at no more than O(n), (2) add story to say that no one spotted that it could have been plug-and-play because of e.g. reasons why they would stumble}
-
-\section{Problem Setup}
-
-Consider an unobserved Markov process $\{X_t, t \geq t_0\}$, and observations $Y_1,...,Y_N$ at timesteps $t_1,..., t_N$.
-The process is parameterized by an unknown parameter $\theta \in \Theta$, where $X \in \gX, Y \in \gY,$ and finally $\gX, \gY, \Theta \subseteq \R^n$.
-By a similar decomposition to that in \citet{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
-$$f_{X_{0: N}, Y_{1: N}}\left(x_{0: N}, y_{1: N} ; \theta\right)=f_{X_0}\left(x_0 ; \theta\right) \prod_{n=1}^N f_{X_n \mid X_{n-1}}\left(x_n \mid x_{n-1} ; \theta\right) f_{Y_n \mid X_n}\left(y_n \mid x_n ; \theta\right).$$
-
-We call $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ the process model, writing $\texttt{process}\left(x_n, \theta\right)$ for the simulator corresponding to the above process model, $f_{Y_n|X_n}\left(y_n \mid x_n, \theta\right)$ the measurement model, and will write $y_n^*$ for the actual values of the observations that were observed.
-
-The above are all defined on a single probability space $(\Omega, \Sigma, \prob),$ where $\omega \in \Omega$ denotes an element of the sample space that can be thought of as a random seed.
-For notational simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0.$
-
-\paragraph{Seed-Fixing, Coupling, and the Method of Common Random Numbers, Coupling:}
-
-In the rest of this paper, we will identify the seed with an element of the sample space.
-We do so because the seed determines the entire sequence of pseudo-random numbers as generated by a computer, to maintain an algorithmic and practical interpretation of our theory.
-When two algorithms share the same seed, this corresponds to the method of common random numbers.
-
-\section{A Smooth Extension of the Particle Filter}
-
-%\paragraph{Motivation:} 
-%The gradient of the log-likelihood can be written as a vector $\nabla_\theta \ell(\theta)$ such that 
-%\begin{equation}
-    %\lim_{v \to 0} \frac{|\ell(\theta+v) - \ell(\theta) - v^T \nabla_\theta \ell(\theta)|}{\|v\|} = 0.
-%\end{equation}
-
-%We therefore need to evaluate the likelihood at a point $\theta$ and a nearby point $\theta + v$ to estimate the gradient with finite differences.
-But in the presence of Monte Carlo noise in evaluating the likelihood, this is difficult, unless (1) one ensures one uses the same seed for each evaluation in the neighborhood, and (2) chooses a step-size small enough that changing the the parameter does not change the resampling indices.
-
-\paragraph{Intuition:} 
-Our theoretical framework for automatic differentiation for the particle filter hinges on the following observation.
-Instead of seeking to differentiate $\ell(\theta)$ directly, we instead focus on differentiating through a suitable fixed-seed reweighting of a bootstrap filter.
-The intuition is as follows.
-Consider a particle filter likelihood estimate where the system (i.e.
-\textbf{both the state transitions and resampling}) evolves under a parameter $\phi \in \Theta$ and the seed, or sample space element, is fixed at $\omega \in \Omega.$ Conditional on $\omega$ and $\phi,$ one can reweight the likelihood estimate to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$
-
-Differentiating this reweighted likelihood estimate with respect to $\theta$ then amounts to differentiating through the reweighting likelihood ratios, which is differentiable if the densities used for the reweighting are differentiable.
-We call this simulation-and-reweighting scheme an \textbf{off-policy} likelihood estimate, as it is similar in spirit to the off-policy methods found in the reinforcement learning literature.
-
-The above intuition is formalized below.
-
-\kevin{Method of Common Random Numbers, Coupling}
-
-\subsection{Doubly Off-Policy Particle Filters}
-
-\begin{defn}[Seed-Fixing]
-\label{defn:seed-fixing}
-Consider a probability space $(\Omega, \Sigma, \mu)$ governing the particles and observations.
-Noting that the particle filter is unbiased for the likelihood when computing the expectation of the normalizing factor under the posterior $f_{X_{0:N|Y_{1:N}}}\left(x_{0: N} \mid y_{1: N}; \theta\right)$, we can write the likelihood as
-$$
-\begin{aligned}
-\mathcal{L}(\theta) & =\int_{\Omega} \mathcal{L}(\theta, \omega, J) d \mu(\omega) \\
-& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right)\right) f_{X_{0:N}, Y_{1:N}}\left(x_{0: N}^P(\omega) , y_{1: N}^*, \theta\right) d \mu(\omega) \\
-& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right) f_{X_n|X_{n-1}}(x_{n,j}^P(\omega)|x_{n-1,j}^F(\omega), \theta)\right) f_{X_{0:N}| Y_{1:N}}\left(x_{n, j}^P(\omega) \mid y_n^*, \theta\right) d \mu(\omega)
-\end{aligned}
-$$
-where $\mathcal{L}(\theta, \omega, J)$ is the Monte-Carlo estimate of the likelihood given the realization of $\mathrm{J}$ particles corresponding to $\omega \in \Omega$.
-For simplicity, we resample every timestep.
-We call $\omega$ a seed, and the deterministic quantity $\mathcal{L}(\theta, \omega, J)$ the seed-fixed estimate.
-\end{defn}
-
-Armed with this seed-fixed likelihood estimate, we modify a particle filter likelihood estimate where the system evolves under a parameter $\phi \in \Theta$ and seed $\omega \in \Omega$ to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$
-
-\begin{defn}[Doubly Off-Policy Particle Filter Likelihood]
-\label{defn:doubly-off-policy}
-Let $\lik(\theta, \phi, \omega, J)$ denote a Monte-Carlo estimate of the likelihood evaluated at $\theta \in \Theta$ with $J$ particles, but where the system evolves according to another $\phi \in \Theta$ and conditional on a fixed seed $\omega \in \Omega$.
-This can be computed with a suitable reweighting of the particles, which we provide the derivation for later.
-We then have
-\begin{equation} \label{eq:loglik2}
-\lik(\theta)=\E[\lik(\theta,\phi,\Omega, J)]=\int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega),
-\end{equation}
-where we define
-\begin{align}
-    \lik(\theta,\phi,\omega, J) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J &\left(f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) \cdot 
-    f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}(\omega) | x_{n-1,j}^{F,\phi}, \phi) \right.\\
-    &\cdot
-    \left.
-    \frac{f_{Y_n|X_n}(y^*_n|x_{n,j}^{P,\phi}(\omega),\theta)}{f_{Y_n|X_n}(y^*_n|x_{n,j}^{P,\phi}(\omega),\phi)}
-    \cdot \frac{f_{X_n|X_{n-1}}(x_{n,j}^{P,\phi}(\omega)|x_{n-1,j}^{F,\phi}(\omega),\theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P,\phi}(\omega)|x_{n-1,j}^{F,\phi}(\omega),\phi)}\right).
-\end{align}
-\end{defn}
-
-We want $\phi\in\Theta$ to be a point in parameter space around which $\lik(\theta,\phi,\omega, J)$ is differentiable with respect to $\theta$ and does not have excessive Monte~Carlo variance.
-This leads us to our first assumption.
-
-\begin{aspt}[Smooth Neighborhood]
-\label{assump:smooth-nbhd}
-There exists a neighborhood $\mathcal{N}(\phi)$ around any $\phi \in \Theta$ where for all $\theta \in \mathcal{N}(\phi)$ and almost every $\omega \in \Omega$, the Monte Carlo estimate of the likelihood at $\theta \in \mathcal{N}(\phi)$ with the system evolving according to $\phi$ conditional on $\omega$, $\lik(\theta,\phi,\omega, J)$, is twice differentiable in $\theta$.
-\end{aspt}
-
-\paragraph{Intuition:} We can justify this assumption as follows.
-For suitably nearby $\theta$ and $\phi$, the resampling indices for $\lik(\theta,\phi,\omega, J)$ and $\lik(\phi,\phi,\omega, J)$ are the same, eliminating any discontinuities from resampling, and that for small enough $\mathcal{N}(\phi)$ the likelihood ratios are bounded.
-The likelihood only changes by a factor of the likelihood ratios, which are bounded and smooth in $\theta$ if the densities used in the calculation are.
-
-Lastly, we require the following regularity conditions.
-
-\begin{aspt}[Continuity of the Likelihood] $\ell(\theta)$ proper is continuous in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.
-\end{aspt}
-
-\begin{aspt}[Bounded Measurement Model] There is an $\epsilon>0$ with $\epsilon^{-1}>f_{Y_n \mid X_n}\left(y_n{ }^* \mid x_n; \theta\right)>\epsilon$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
-\end{aspt}
-
-\begin{aspt}[Locally Bounded Derivative]
-\label{assump:local-bounded-derivative}
-Let $\mathcal{M}$ be an open subset of $\Omega$. There exists some function $G(\theta)$ and a constant $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$ such that
-    $$
-    \|\nabla \ell(\theta, \phi, \omega, J)\|_2<G(\theta) \leq G<\infty
-    $$
-    for every $\phi \in \mathcal{M}, \theta$ in smooth neighborhood $\mathcal{N}(\phi)$, and almost every $\omega \in \Omega$.
-\end{aspt}
-
-\subsubsection{Doubly Off-Policy Particle Filter}
-
-\kevin{rewrite to de-emphasize dop as in presentation, sell mop as the big theoretical idea that explains AD through PFs and unifies two previous estimators}
-
-Having the system evolve according to $\phi$ but evaluating the likelihood at $\theta$ requires us to reweight the conditional likelihood at each timestep by likelihood ratios corresponding to the process and measurement models.
-This corresponds to making two corrections in the particle weights at each timestep, by multiplying by
-\begin{equation}
-    s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}, \qquad r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)},
-\end{equation}
-as we illustrate in Appendix \ref{app:dop}. We use the notation $x_{n,j}^{P, \phi}$ to emphasize that the particles evolve according to $\phi$.
-
-We claim that this yields a particle filter that targets the posterior and recovers the derivative estimate of \citet{doucet2011sf, scibior2021dpf} when $\theta$ is evaluated at $\phi$.
-The proof can be found in Appendix \ref{app:dop}.
-
-\begin{prop}[Correctness of Doubly Off-Policy Particle Filter Corrections]
-    \label{prop:dop-correctness}
-
-    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$.
-Also let $\theta \in \gN(\phi)$.
-Then, the following corrections in the particle weights,
-    \begin{equation}
-    s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}, \qquad r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)},
-    \end{equation}
-    yield a particle filter that is unbiased for the likelihood, even when $\theta \neq \phi$.
-
-    In particular, when $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{doucet2011sf},
-    \begin{equation}
-        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
-    \end{equation}
-\end{prop}
-
-\kevin{Cut down on DOP discussion in actual manuscript, remove any references to this as an algorithm, just talk about the idea of an off-policy system/particle filter.
-Bottom paragraph is the punchline.}
-
-\paragraph{Intuition:} Having the system evolve according to $\phi$ but evaluating the likelihood at $\theta$ requires us to reweight the conditional likelihood at each timestep by likelihood ratios corresponding to the process and measurement models.
-This corresponds to making two corrections in the particle weights at each timestep, one for the process model and one for the measurement model.
-The measurement correction cannot be avoided, but one can bypass the process correction -- with the reparametrization trick.
-We can do when we assume the simulator is differentiable with the reparametrization trick, as \cite{corenflos21} also assume.
-Doing so then yields an algorithm that only requires a simulator, and does not require the process densities.
-
-\subsection{Measurement Off-Policy Particle Filters}
-
-With the above procedure, we have the entire system evolve according to $\phi$, and we differentiate through the likelihood ratios, which are smooth conditional on the resampling being the same (which in turn happens when $\theta \approx \phi$).
-However, doing so requires access to likelihood ratios, and as such Algorithm \ref{alg:dop} is therefore not compatible with simulation-based inference.
-
-There is a workaround.
-Consider the case where you have access to a differentiable simulator compatible with the reparametrization trick.
-If we have the state transitions evolve according to $\theta,$ but have the resampling according to $\phi,$ we can offload the derivative with respect to the process model to the reparametrization trick, as in \cite{corenflos21}.
-This corresponds to the MOP-$\alpha$ procedure in Algorithm \ref{alg:mop}.
-
-We therefore make the following assumption in this section only:
-\begin{aspt}[Differentiable Simulator]
-    $\texttt{process}(\cdot | X_{t-1}; \theta)$ is differentiable with respect to $\theta,$ and is compatible with the reparametrization trick. 
-\end{aspt}
-
-Algorithm \ref{alg:mop} provides a weighted particle filter that targets the posterior under $\theta$, but when differentiated yields either the low-bias but high-variance gradient estimator of \cite{poyiadjis11}, the high-bias but low-variance estimator of \cite{blei2018vsmc}, or an intermediate, interpolating quantity that can optimize a bias-variance tradeoff.
-This is controlled by a discounting parameter $\alpha$.
-The intuition behind this procedure is as follows.
-
-\paragraph{Intuition:} Broadly, if $\theta\neq\phi,$ one can first run a vanilla particle filter to recover values for the particles with the process model under $\phi.$ One can then run an accordingly reweighted particle filter with the process model under $\theta$, but resampling according to the measurement model under $\phi$, with the same indices as the original vanilla particle filter.
-This requires one to set the same random seed for both particle filters.
-This then yields a slightly-unconventionally weighted particle filter that targets the posterior, but has some very nice gradient properties; as we outline in the remainder of this section.
-
-\paragraph{Special Case:} When $\theta = \phi$, one can simply set the particles under $\phi$ to be copies of the particles under $\theta$.
-Gradients with respect to $\theta$ now treat functions of the particles under $\phi$ as constants, corresponding to the stop-gradient trick in \cite{scibior2021dpf}.
-This therefore provides a theoretical framework for their stop-gradient trick, which we happily use in our numerical experiments for simplicity.
-
-\kevin{Develop theory for mixing-time swapping?
-Detect when IF2 reaches a neighborhood of the MLE, by detecting whether it has reached a local sort of stationary distribution when the perturbations are lower bounded, e.g.
-the autocorrelation between the mean vectors to estimate the mixing time or Gelman-Rubin.
-In practice, just swap when IF2 converges according to the diagnostics.
-Theory doesn't care, you just need to in a basin of attraction that is strongly convex.}
-
-\subsubsection{Discounting Weights}
-
-Observe that at Algorithm \ref{alg:mop}, at each timestep the corrected filtering importance weights $w_{n,j}^{F,\theta}$ optionally propagate to the next timestep for us to maintain the correction.
-We introduce a discounting parameter $\alpha$ that affects the weights, so $w_{n,j}^{P,\theta} := (w_{n-1,j}^{F,\theta})^{\alpha}$.
-Note that no matter what value $\alpha$ takes on, the system evolves according to $\phi$ when we perform the correction to evaluate the likelihood at $\theta$.
-The filter fully targets the posterior when $\theta=\phi$, or $\alpha=1$, as we show in Proposition \ref{prop:dop-correctness}.
-As previously mentioned, the value of $\alpha$ therefore informs a bias-variance tradeoff.
-We elaborate on this below.
-
-\paragraph{Bias-Variance Tradeoff} This discounting serves the purpose of balancing a tradeoff between maintaining memory of each particle's ancestral trajectory (most extreme when $\alpha=1$), or considering only the single-step transition dynamics (when $\alpha=0$).
-This can be thought of as an \textbf{exponentially-weighted moving average}, where $\alpha$ controls the amount of discounting.
-We elaborate on two special cases below.
-
-\kevin{This isn't a smoother, because we'd need future information.}
-
-\paragraph{When $\alpha=1,$} the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\theta}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
-\end{equation}
-as we have a telescoping product. Its derivative when $\theta=\phi$ is then
-\begin{equation}
-    \nabla_\theta \hat{\ell}(\theta) := 
-        \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*),
-\end{equation}
-the average gradient of each surviving particle's cumulative importance weights.
-The filter targets the posterior when $\alpha=1$ even when $\theta \neq \phi,$ and therefore suffers from no asymptotic bias.
-
-\paragraph{When $\alpha=0,$} the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J w_{n,j}^{F,\theta} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})},
-\end{equation}
-and though the telescoping product disappears, this reduces to a series of one-step corrections.
-However, these corrections are insufficient when $\theta \neq \phi,$ and the filter suffers from asymptotic bias unless $\theta=\phi.$ Its gradient when $\theta=\phi$ is therefore
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
-    \end{equation}
-
-the gradient of the vanilla particle filter where the resampling correction is omitted as found in \cite{blei2018vsmc} -- when applied to the bootstrap filter.
-As \cite{scibior2021dpf} remark, this estimator only considers single-step quantities, and is "memoryless" beyond a single step.
-This is in contrast to the case when $\alpha=1,$ as that estimate only considers the surviving particles at time $N$, which is the other extreme where dependencies across time are fully tracked.
-
-\subsubsection{Correctness of MOP-$\alpha$}
-
-\begin{prop}[Correctness of MOP-$\alpha$ (Algorithm \ref{alg:mop})]
-    \label{prop:mop-correctness}
-    When either $\alpha=1$ or $\theta=\phi$, Algorithm \ref{alg:mop} is a particle filter that targets the posterior and is unbiased for the likelihood under $\theta$.
-
-    In particular, when $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter.
-In this case, its gradient when $\alpha=1$ is the same as the estimator of \citet{doucet2011sf} when applied to the bootstrap filter,
-    \begin{equation}
-        \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{0:N}}(y_{1:N}^* | x_{1:N,j}^{A, F,\theta}),
-    \end{equation}
-
-    and when $\alpha=0,$ is the gradient estimator of \cite{blei2018vsmc},
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
-    \end{equation}
-
-    that as \cite{scibior2021dpf} show corresponds to differentiating through the fixed-seed likelihood estimate yielded by a vanilla particle filter.
-\end{prop}
-
-The proof for the above follows from Appendices A and C, as well as the technical lemmas for properly weighted particle filters.
-
-\kevin{Take derivative of loglik with regard to alpha?}
-\kevin{Talk about regime change the moment you move away from alpha=1, variance going from N2 to N, and bias going from 0 to something, but try to prove that decrease in variance outweights increase in bias.
-}
-\kevin{Create tools for choosing alpha via binary search CV from an averaged alpha=1 target}
-\kevin{suppose at each filtering step, look at gradient calculated at alpha=0, dynamically select alpha during filtering.
-calculate bias and variance online during filtering, and update alpha accordingly.
-}
-
-\kevin{No need to recompute so many particle filters! Just take an array of weight copies, one for each alpha. Gradients for each alpha come free the moment you compute one of them. }
-
-\kevin{Elbow method on variance, since bias is hard to compute? }
-\kevin{Get variance estimate, bias estimate, construct confidence intervals on bias estimate, use that to get unbiased estimate of the MSE. }
-
-\subsubsection{Practical Considerations}
-
-It is only ever practical to evaluate $\theta$ at $\phi$.
-This removes the issue of the likelihood ratios blowing up, and yields an estimate of the likelihood that agrees with the bootstrap filter estimate.
-Its gradient in this case is either the estimate of \citet{doucet2011sf} when $\alpha=1$, a single-timestep estimator that corresponds to naively taking the derivative of the particle filter when $\alpha=0$ as first proposed in \citet{blei2018vsmc}
-
-\paragraph{Bottom Line:} The DOP-$\alpha$ and MOP-$\alpha$ procedures therefore provide us with a unifying framework for reasoning about smooth extensions of weighted particle filters that target the posterior and provide reasonable gradient estimates when differentiated.
-This fixed-seed reweighting alleviates issues that can arise from differentiating through a quantity yielded by ostensibly non-differentiable resampling.
-This is done by either (1) differentiating through smooth likelihood ratios, or (2) differentiating through a differentiable simulator with the reparametrization trick (and through smooth likelihood ratios to adjust for the resampling).
-In particular, MOP-$\alpha$ does not require access to the transition densities, and like the bootstrap filter, only requires access to a (albeit differentiable) simulator.
-To our knowledge, this is the first particle filter gradient method in the literature with this capability.
-
-\section{Practical Maximum-Likelihood Inference}
-
-This still leaves us with the question of designing a reasonably effective procedure for likelihood maximization with these new tools.
-We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that amounts to warm-starting gradient descent or some other reasonable first or second order procedure (using the gradient estimate given by MOP-$\alpha$) with the output of an initial search of IF2.
-
-\paragraph{Summary:} A coarse solution is found by running IF2 with an aggressive random walk standard deviation (learning rate) and a cooling schedule that is \textbf{bounded below}, until the search stalls to an initial "convergence" to what is hopefully a neighborhood of the MLE.
-We can then use gradient descent, or other first or second order gradient methods, to refine this coarse solution.
-
-The initial "convergence" happens fairly quickly in practice.
-In the case of the Dhaka cholera model of \cite{king08}, when a geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations of IF2.
-In comparison, often 100 or 200 iterations of IF2 are used for an initial global search.
-\cite{ionides15}, for example, use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{Lee_haiticholera}.
-
-\paragraph{Rationale:}
-
-IF2 converges quickly to a neighborhood of some reasonably good local optimum in practice, but then struggles to squeeze out the last few log-likelihood units.
-Conversely, gradient descent (and other first or second order gradient methods) can get stuck in saddle points and less desirable local minima when the likelihood is nonconvex.
-Warm-starting gradient methods with IF2 lets us get close enough to the MLE for gradient methods to shine, as in most cases regularity conditions ensure that the likelihood surface is well-behaved enough in this neighborhood that the aforementioned issues with saddle points and local minima are alleviated.
-Combining these two methods in this way lets us enjoy the best of both worlds.
-
-\begin{algorithm}[ht]
-\centering
-	\caption{Iterated Filtering with Automatic Differentiation}
-    \label{alg:ifad}
-	\begin{algorithmic}[1]
-	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $p(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta),$ IF2 cooling schedule $\eta_t$, MOP-$\alpha$ discounting parameter $\alpha$.
-		\STATE Initialize $\theta_0$, $t=0.$
-        \STATE Run IF2 until initial "convergence" under cooling schedule $\eta_t$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_t := \frac{1}{J}\sum_{j=1}^J \Theta_j.$
-		\WHILE{Procedure not converged:}
-		\STATE Draw $\omega \in \Omega$
-		\STATE Perform forward pass by running Algorithm \ref{alg:mop} with discounting parameter $\alpha$ to obtain $\loglik(\theta_t, \theta_t, \omega, J).$
-		\STATE Obtain $g(\theta_t) = \nabla_{\theta_t} (-\loglik(\theta_t, \theta_t, \omega, J))$ via backpropagation, $H(\theta_t)$ such that $\lambda_{\min}(H) \geq c$.
-		\STATE Obtain $\eta$ with line search or an annealing schedule.
-		\STATE Update $\theta_{t+1} := \theta_t - \eta (H(\theta_t))^{-1} g(\theta_t)$, $t:=t+1.$
-		\ENDWHILE
-		\STATE Return $\hat{\theta} := \theta_t.$
-	\end{algorithmic}
-\end{algorithm}
-
-\subsection{Theoretical Guarantee}
-
-We first analyze the performance of gradient descent, Newton's method, or any iterative method that multiplies the particle estimate of the gradient with the inverse of a matrix with minimum eigenvalue bounded below, in the case where the log-likelihood surface is $\gamma$-strongly convex, and where the gradient estimate is a particle approximation of the score (for example, IFAD-1 and MOP-1).
-In this case, these methods enjoy a linear convergence guarantee:
-
-\begin{thm}[Convergence of Gradient Methods]
-\label{thm:convergence}
-
-Consider a variant of IFAD-1 (Algorithm \ref{alg:ifad}) where one stops if the gradient estimate $||g(\theta_n)|| \leq \sigma \epsilon$.
-Assume $-\ell$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$.
-$\sigma = \frac{4 \Gamma}{(1-\beta)}> 4$.
-Define $G$ as in Assumption \ref{assump:local-bounded-derivative}.
-
-Choose $J$ according to the maximum of the $J$ prescribed in Lemmas \ref{lemma:grad_bound} and (if using a Hessian estimate) $\ref{lemma:hess_bound}$ with failure probability $\delta/2$ each.
-Let $H(\theta_n)$ be a matrix possibly dependent on $\theta_n$ with minimum eigenvalue always greater than some $c>0$.
-Choose the learning rate $\eta$ and error tolerance $\epsilon$ such that, where $\beta$ is the Armijo condition hyperparameter,
-\begin{equation}
-    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_n)||.
-\end{equation}
-
-Then the following holds with probability at least $1-\delta$:
-\begin{equation}
-\ell(\theta^*) - \ell(\theta_{n+1}) \leq \left(1-\eta\beta\frac{8\gamma}{9c}\right)(\ell(\theta^*)-\ell(\theta_n))
-\end{equation}
-
-and the algorithm eventually terminates after we obtain $||\nabla_\theta \ell(\theta_n)|| \leq (1+\sigma) \epsilon$.
-\end{thm}
-
-The proof of this result uses the concentration inequalities from \cite{delmoral2011ci}, and leans upon the analysis of \cite{mahoney2016subsampled}. We defer the proof to Appendix \ref{app:ifad}
-
-\paragraph{Takeaways for IFAD (Algorithm \ref{alg:ifad}):}
-
-The gradient stage of IFAD-1 converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the IF2 stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
-This happens fairly often in practice, for example, when sufficient regularity conditions for local asymptotic normality of the MLE hold.
-
-We conjecture that this applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE.
-IF2 posesses some SGD-like qualities, including what looks like linear convergence to a ball around the MLE with width determined by the random walk standard deviation.
-We will explore this conjecture in future work.
-
-While this theoretical guarantee currently only holds for IFAD-1, as it forms a particle approximation of the score as in \cite{poyiadjis11}, one can also extend this for a similar guarantee if the bias for the gradient estimate given by MOP-$\alpha$ for $\alpha < 1$ is small enough.
-For general $\alpha$, including MOP-0, the analysis then becomes more complicated, as one needs to handle the case of biased gradient descent.
-However, if future measurements $y_{n+1:N}^*$ do not provide a large amount of information on the identification of the current state $x_n$ given past and current measurements $y_{0:n}^*$, as mentioned in \cite{corenflos21}, the bias may be small enough for it to not present too much of a problem.
-
-\subsection{Runtime}
-
-Even with the increased computational cost of automatic differentiation, Algorithm \ref{alg:ifad} is reasonably fast, as by the cheap gradient principle from \cite{kakade2019provably}, a gradient operation takes no more than 6 times the runtime of the particle filter (in practice, this means a gradient step takes about 3.75x as much time as an IF2 iteration).
-As such, unlike methods like the marginal differentiable particle filter in \cite{scibior2021dpf} or the the optimal transport of resampling approach of \cite{corenflos21}, our algorithm has runtime \textbf{linear, and not quadratic}, in the number of particles (while alleviating the $O(N^2)$ variance concern of \cite{poyiadjis11} and \cite{corenflos2021ot} and asymptotic bias concern of \cite{blei2018vsmc} through optimizing $\alpha$ for a bias-variance tradeoff).
-
-In addition, re-implementation of the particle filter, iterated filtering, and Algorithm \ref{alg:ifad} itself in \texttt{JAX} (\cite{jax2018github}) led to considerable speedups.
-This can be attributed to both GPU acceleration and the just-in-time (JIT) compilation functionality that \texttt{JAX} provides.
-When tested on a computer with an Intel i9-13900K processor, 128GB of RAM, and a NVIDIA RTX3090 GPU, each post-compilation particle filtering operation on the Dhaka cholera model from \cite{king08} with 10000 particles took roughly 379 milliseconds (with a standard deviation of 279 microseconds over seven runs), a considerable improvement over the existing performance in the widely used \texttt{pomp} package of \cite{king16}, which took between 6.19 to 6.39 seconds over ten runs on the same machine.
-
-\section{Numerical Experiments}
-
-\subsection{Cholera in Bangladesh}
-
-To compare the performance of IFAD against that of IF2, we fit the very same model first proposed in \cite{king08} that \cite{ionides15} use to demonstrate the capabilities of IF2.
-This is a stochastic SIR compartmental model with transition and measurement uncertainty, where the population at time $t$, $H(t)$, is divided into the susceptible compartment $S(t)$, infected compartment $I(t)$, and three ($k=3$) recovered compartments $R_1(t), ..., R_k(t)$ denoting varying degrees of cholera immunity.
-We write $M(t)$ for the cholera deaths in each month.
-As in \cite{ionides15}:
-
-\begin{itemize}
-    \item The \textbf{transition dynamics} follow this series of stochastic differential equations:
-
-    \begin{align}
-        d S&=\left(k \epsilon R_k+\delta(S-H)-\lambda(t) S\right) d t+d H-(\sigma S I / H) d B, \\ 
-        d I&=\left(\lambda(t) S-(m+\delta+\gamma) I\right) d t+(\sigma S I / P) d B, \\ 
-        d R_1&=\left(\gamma I-(k \epsilon+\delta) R_1\right) d t, \\ 
-        \vdots \\ 
-        d R_k&=\left(k \epsilon R_{k-1}-(k \epsilon+\delta) R_k\right) d t,
-    \end{align}
-
-    where $B$ represents Brownian motion, $m$ is the rate at which infected individuals succumb to cholera, $\gamma$ is the rate at which infected individuals recover, $1/\epsilon$ is the mean duration of immunity, $\sigma$ represents the standard deviation of perturbations to the force of infection, and $\delta=0.02$ is the death rate.
-
-    \item The \textbf{force of infection}, $\lambda(t)$, is modeled by splines $s_j$
-    \begin{equation}
-        \lambda(t)=\exp \left\{\beta_{\text {trend }}\left(t-t_0\right)+\sum_{j=1}^{6} \beta_j s_j(t)\right\}(I / P) + \exp \left\{\sum_{j=1}^{6} \omega_j s_j(t)\right\},
-    \end{equation}
-    where the $\beta_j$ model seasonality in the force of infection, $\beta_{\text{trend}}$ models the trend in the force of infection, and the $\omega_j$ represent seasonality of a non-human environmental reservoir of disease.
-    \item The \textbf{measurement model} for observed monthly cholera deaths is given by 
-    \begin{equation}
-        Y_n \sim \gN(M_n, \tau^2M_n^2),
-    \end{equation}
-    where $M_n$ is the true number of cholera deaths in that month.
-\end{itemize}
-
-\begin{figure}[ht]
-\centering
-\begin{tikzpicture}[node distance=1cm and 1cm, auto,
->=Latex,every node/.append style={align=center}, 
-int/.style={draw, minimum size=1.6cm}]
-  \node [int] (H) at (0,0) {$H$};
-\node [int] (S) [right=of H] {$S$};
-\node [int] (I) [right=of S] {$I$};
-\node (Mcoord) [right=of I, coordinate] {Mcoord};
-\node [int] (R1) [right=of Mcoord] {$R_1$};
-\node [int] (R2) [right=of R1] {$R_2$};
-\node [int] (R3) [right=of R2] {$R_3$};
-\node [circle, draw] (M) [above=of Mcoord] {$M$};
-\node (deathsS) [above=of S, coordinate] {deathsS};
-\node (deathsI) [above=of I, coordinate] {deathsS};
-\node (deathsR1) [above=of R1, coordinate] {deathsR1};
-\node (deathsR2) [above=of R2, coordinate] {deathsR2};
-\node (deathsR3) [above=of R3, coordinate] {deathsR3};
-\node (belowR3) [below=of R3, coordinate] {belowR3};
-\node (belowS) [below=of S, coordinate] {belowS};
-\path[->]
-    (H) edge node {$dH$} (S)
-    (S) edge node {$\lambda(t)$}  (I)
-    (I) edge node {$\gamma$}  (R1)
-    (R1) edge node {$k\epsilon$}  (R2)
-    (R2) edge node {$k\epsilon$}  (R3)
-    (R3) edge node {} (belowR3) 
-    (belowR3) edge node {$k\epsilon$} (belowS) 
-    (belowS) edge node {} (S)
-    (I) edge node {$m$}  (M)
-    (S) edge node {$\delta$}  (deathsS)
-    (I) edge node {$\delta$}  (deathsI)
-    (R1) edge node {$\delta$}  (deathsR1)
-    (R2) edge node {$\delta$}  (deathsR2)
-    (R3) edge node {$\delta$}  (deathsR3);
-\end{tikzpicture}
-\caption{Illustration of SIR model from \cite{king08}.}
-\label{fig:sir}
-\end{figure}
-
-Figure \ref{fig:sir} depicts the above compartmental model that we use to benchmark the performance of IFAD.
-We present comparisons of IFAD with $\alpha=0$ (IFAD-0) and $\alpha=1$ (IFAD-1) against both (1) the results from \cite{ionides15}, and (2) our own implementation of IF2 in \texttt{JAX}.
-
-\paragraph{Hyperparameters:} While performing an initial exploratory analysis, we found that alternative hyperparameters for IF2 (random walk standard deviation of $\sigma=0.02$, geometrically decreasing cooling rate exponent $a=0.95$) performed better than the settings \cite{ionides15} chose, and chose to use these for our implementation to achieve a fairer comparison.
-Our implementations of IF2 and IFAD used 10000 particles, in line with \cite{ionides15}.
-
-To prevent exploding gradients given the high variance in the MOP-1 estimator, we scale the gradient to have norm 1 before performing vanilla gradient descent, and use a linear cooling schedule of $0.2$ to $ 0.0001$.
-We chose to use vanilla gradient descent for simplicity, and to demonstrate that minimal tuning was required to have IFAD perform well.
-
-% The 60 more iterations of IF2 yielded largely not a huge improvement over the initial coarse solution of 40 iterations.
-This demonstrates what we have mentioned earlier: that although IF2 manages to quickly find a decent coarse solution, it struggles to squeeze out those last few units of log-likelihood improvement.
-The 60 more iterations of gradient descent, however, successfully refines this solution, getting much closer to the MLE on average.
-
-\subsubsection{Results} 
-We benchmarked IFAD against IF2 on a challenging global search problem.
-100 initial starting parameter vectors were drawn uniformly from the same wide bounding box used in \cite{ionides15}, and we performed 100 searches each of IF2, IFAD-0, IFAD-1, MOP-0, and MOP-1 initialized from these 100 starting parameter vectors.
-We summarize our findings below.
-
-\begin{table}[h!]
-\centering
-\begin{tabular}{||c c c||} 
- \hline
- Method & Best Log-Likelihood & Rank \\ [0.5ex] 
- \hline\hline
-     IFAD-0 & -3747.87 & 1\\
-     IFAD-1 & -3748.53 & 2\\
-     IF2 (Ours) & -3755.91 & 3\\
-     IF2 (Warm Start) & -3759.60 & 4 \\
-     IF2 2015 & -3768.63 & 5\\ 
-     MOP-0 Alone & -3795.74 & 6\\
-     MOP-1 Alone & -3797.38 & 7\\
- \hline
-\end{tabular}
-\caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone.
-IFAD performs the best among all methods.
-Our implementation of IF2 outperforms that of \cite{ionides15}, but still ultimately underperforms IFAD.
-Both IFAD-0 and IFAD-1 manage to find the MLE, either matching or beating the highest log-likelihood previously found in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.}
-\label{table:mle}
-\end{table}
-
-\paragraph{IFAD Successfully Finds the MLE:} Both IFAD-0 and IFAD-1 successfully find the MLE, as seen in Table \ref{table:mle}.
-Previously, the MLE at a log-likelihood of $-3748.5$ was found with much effort, using many global and local IF2 searches and likelihood profiling.
-Despite being initialized for a global search, both IFAD-0 and IFAD-1 manage to find the MLE over the 100 searches.
-This may imply that the local search and refinement that was previously necessary for finding the MLE is \textbf{baked into} the two-stage IFAD algorithm.
-
-Interestingly, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.63$, while \cite{king08} only achieve a best possible estimate of $-3793.4$.
-Our results of $-3748.53$ and $-3747.87$ match or beat the highest log-likelihood previously found in the current implementation in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.
-
-This can be seen most clearly in Figure \ref{fig:boxplot-search}.
-In the left panel within Figure \ref{fig:boxplot-search}, only IFAD has whiskers that extend to the dotted red line that represents the MLE.
-
-\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations (as seen in Figure \ref{fig:optim}), performing IF2 alone ultimately fails to squeeze out the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter} and the left panel in Figure \ref{fig:boxplot-search}).
-
-Conversely, performing MOP alone leads to many, many failed searches, as seen in Figure \ref{fig:hist-all} and the right panel in Figure \ref{fig:boxplot-search}.
-This is a difficult, nonconvex, and noisy problem, and the search gets stuck in local minima and saddle points, failing to approach the MLE.
-
-IFAD, in comparison, successfully manages to (1) approach the MLE quickly due to the IF2 warm-start (as seen in Figure \ref{fig:optim}) and (2) refine the coarse solution found by the warm-start with MOP gradient steps to find the MLE (as seen in Figures \ref{fig:scatter}, \ref{fig:boxplot-search}, and \ref{fig:hist-all}).
-IFAD therefore successfully combines the best qualities of IFAD and MOP, outperforming either of them alone.
-
-\begin{figure}[htbp!]
-    \centering
-    \includegraphics[scale=0.53]{../imgs/095/pairs.png}
-    \includegraphics[scale=0.53]{../imgs/095/qq.png}
-    \caption{Scatterplots depicting the performance of IFAD against that of IF2.
-\textbf{Left:} Paired searches from the same starting point.
-\textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches.
-It is clear that on average, IFAD has the edge, and manages to find the MLE while no IF2 search successfully gets within 7 log-likelihood units of it.
-We use a dotted red line to display the MLE in both.
-}
-
-    \label{fig:scatter}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=0.48]{../imgs/095/boxplot.png}
-    \includegraphics[scale=0.53]{../imgs/095/boxplotall.png}
-    \caption{\textbf{Left:} Boxplot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs.
-This simulates the result of the common procedure of running a few searches and choosing the best one.
-IFAD outperforms all other methods, and improves on the warm-start given by running 40 IF2 iterations.
-Our choice of hyperparameters for our implementation of IF2 outperforms the choices made by \cite{ionides15}, which we depict in the boxplot labeled "IF2 2015".
-As before, while IF2 successfully approaches the MLE, it ultimately fails to find it.
-    \textbf{Right:} Boxplot depicting the performance of IFAD, IF2, and MOP alone without the warm-start, where we plot all runs.
-We see that MOP alone drastically underperforms all other methods, failing to get close to the MLE.
-This shows that the IF2 warm-start may be necessary to achieve good performance in challenging nonconvex and noisy problems like this one.
-    \textbf{Takeaway:} In this panel, we see that (1) IF2 fails to find the MLE while the gradient steps perform better at fine-grained refinement, and (2) performing gradient steps alone without a warm-start leads to the search getting stuck in local minima and saddle points, ultimately demonstrating the necessity of a hybrid algorithm like IFAD.
-    In both, as before, we use a dotted red line to display the MLE. }
-    \label{fig:boxplot-search}
-\end{figure}
-
-\begin{figure}[htbp!]
-    \centering
-    \includegraphics[scale=0.7]{../imgs/095/optim.png}
-    \caption{Optimization progress of IFAD and IF2, where the solid lines depict the median negative log-likelihood at each iteration and the shaded area depicts the best search at any iteration and the 80\% percentile.
-The dashed orange line depicts the median warm-start given by running 40 IF2 iterations.
-While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD -- IFAD has better tail control and successfully reaches the MLE.
-The initial loss in log-likelihood upon transitioning from IF2 to MOP seems to help with escaping local minima.
-    As before, we use a dotted red line to display the MLE. }
-    \label{fig:optim}
-\end{figure}
-
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.7]{../imgs/095/hist.png}
-    \caption{Histogram comparison of the performance of IFAD, IF2, and MOP alone.
-As before, we note that IFAD successfully reaches the MLE a nontrivial number of times, while IF2 ultimately fails to do so.
-Our implementation of IF2 outperforms that of \cite{ionides15}, largely due to improved hyperparameter choices.
-MOP alone performs poorly, demonstrating the necessity of warm-starting the search with IF2.
-    We once again see that IFAD combines the best of IF2 and MOP, approaching the MLE quickly due to the IF2 warm-start and successfully refining the coarse solution with gradient steps, achieving superior performance.
-As before, we use a dotted red line to display the MLE.
-}
-    \label{fig:hist-all}
-\end{figure}
-
-\paragraph{Differences Between IFAD-0 and IFAD-1:} The differences between IFAD-1 and IFAD-0 (MOP-1 and MOP-0) do not seem very pronounced in this example.
-This could be due to a number of reasons, such as (1) future case counts $y_{{n+1}:N}^*$ not providing a large amount of information on state identification $x_n$ given current and past measurements $y_{0:n}^*$ as mentioned in \cite{corenflos21}, and (2) the gradient normalization making the trajectories given by MOP-0 and MOP-1 more similar than they would be otherwise.
-
-\kevin{IFAD beats IF2 x percent of the time on 100 paired runs, but the bigger thing is that it gets close to the MLE without much effort.
-Previously, we'd have to do many searches, plus refinement (which is necessary, as we can see from the IF2 boxplot, it is easy to fool yourself into thinking that because you see the plateau that is the MLE).
-This includes likelihood profiling diagnostics, which are very computationally intensive.
-Now we can just take the best of ten searches -- IFAD does the refinement.}
-
-\kevin{Have an outsider read the manuscript afterwards.}
-
-\section{Discussion and Future Work}
-
-We develop a new theoretical framework to differentiating through the particle filter that (1) encompasses the gradient estimates of \cite{poyiadjis11}, \cite{blei2018vsmc}, and \cite{scibior2021dpf} as special cases, (2) sidesteps the issue of differentiating through a Monte-Carlo algorithm with discontinuous resampling, and (3) provides an opportunity for optimizing a bias-variance tradeoff.
-
-Prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics.
-In the presence of a differentiable simulator, MOP-$\alpha$ enjoys this very same desirable property, enabling practical maximum likelihood inference in a wider variety of situations.
-
-Furthermore, MOP-$\alpha$, and by extension IFAD, has runtime \textbf{linear, and not quadratic}, in the number of particles, unlike the marginal differentiable particle filter in \cite{scibior2021dpf} or the the optimal transport of resampling approach of \cite{corenflos21}.
-This does not necessarily come at a cost to Monte Carlo variance, unlike \cite{poyiadjis11} and \cite{corenflos2021ot} who suffer from variance quadratic in the length of the trajectory, as we can optimize $\alpha$ for a bias-variance tradeoff.
-
-In the presence of strong convexity, first and second-order gradient methods that use MOP-$\alpha$ as an estimate enjoy a linear convergence rate, providing support for numerical optimization with AD for the particle filter.
-
-This leads to a practical hybrid algorithm we call IFAD that warm-starts gradient descent, or some other first or second-order method using the gradient estimate given by MOP-$\alpha$, with IF2.
-When benchmarked on a challenging global search problem with the model of \cite{king08}, IFAD enjoys the best properties of IF2 and MOP-$\alpha$, successfully approaching a neighborhood of the MLE quickly without getting stuck in saddle points and (poor) local minima (like gradient descent alone would), while effectively optimizing the last few units of log-likelihood (which IF2 struggles with).
-
-Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual} that we call \texttt{pomp-py}.
-The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{JAX} Python package from \citet{jax2018github}.
-
-\subsection{Future Work}
-
-While MOP-$\alpha$, and IFAD in its current incarnation, remain very general, they currently lack the capability of handling discrete latent states.
-We conjecture that DOP-$\alpha$ might be a possible solution (despite needing the likelihood ratio of transition densities), and will explore this in future work.
-
-We conjecture that the linear convergence result in Theorem \ref{thm:convergence} applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE.
-IF2 posesses some SGD-like qualities, including what looks like linear convergence to a ball around the MLE with width determined by the random walk standard deviation.
-Not much is known about the convergence rate of IF2, and an answer to this would also provide an answer to the convergence rate of IFAD beyond just the convergence rate of the gradient stage.
-We will explore this conjecture in future work.
-
-In the interest of simplicity, we only explore the performance of gradient descent on the Dhaka cholera model of \cite{king08}.
-Exploring the use of various first and second order methods such as Newton's method, ADAM, L-BFGS, or other popular iterative optimization methods with the gradient estimates that MOP-$\alpha$ provides in the context of the particle filter could very well lead to nontrivial speedups and increased performance.
-
-Extensions of MOP-$\alpha$ and IFAD to panel and spatiotemporal datam, possibly through a block particle filter in the same vein as \cite{ionides22} and \cite{ning23}, would go a long way towards extending the current applicability of these methods to handle a wider variety of interesting and challenging scientific problems.
-
-While we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual}, this remains a work in progress.
+Promising preliminary numerical results indicate that this can beat current state-of-the-art methods on a challenging scientific benchmark problem.
 
@@
 
-\appendix
-\renewcommand{\thefigure}{A\arabic{figure}}
-\setcounter{figure}{0}
-
-\section{Weighted Particle Filters that Target the Posterior}
-
-\begin{defn}[Targeting]
-    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any measurable and bounded functional $h$,
-\begin{equation}
-    E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
-\end{equation}  
-    A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if for any measurable and bounded functional $h$,
-\begin{equation}
-    \frac{\sum_{j=1}^J h(X^J_j) w^J_j}{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X))
-\end{equation}
-as $J \to \infty$.
-\end{defn}
-\citet{chopin2004clt} asserted without proof that common particle filter algorithms target the filtering distribution, $f_{X_{1:N}|Y_{1:N}}$, in this sense.
-\ed{PERHAPS IF WE HAD A CONCRETE PROOF OF THIS, THAT WOULD HELP TO SEE HOW IT EXTENDS TO OUR SUBSEQUENT LEMMAS.
-ARGUABLY, THAT IS IN FACT WHAT WE ARE DEVELOPING VIA THESE LEMMAS.
-I WONDER IF SOME OTHER REFERENCE, E.G.
-DELMORAL'S 2004 BOOK, HAS AN EXPLICIT PROOF OF CONSISTENCY FOR SOME PARTICLE FILTER ALGORITHM.}
-In order to prove the consistency of our variation on the particle filter, we now present three helper lemmas.
-The first follows from standard importance sampling arguments, the second from integrating out the marginal, and the third from Bayes' theorem. 
-We state Lemma~\ref{lem:change-measure-proper-weights} assuming multinomial resampling, which is convenient for the proof though other resampling strategies may be preferable in practice REFS.
-
-\begin{lem}[Change of Particle Measure]
-    \label{lem:change-measure-proper-weights}
-    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
-Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, $\pi^J_j J$ times.
-Write
-    \[
-    (Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big),
-    \]
-    where $a$ is called the ancestor function. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j),j=1,\dots,J\}$ targets $f_X$.
-\end{lem}
-
-\begin{proof}
-
-    Note that as the $Y_j^J$ are a subsample from $X_j^J$, $h$ can be a function of $Y$ as well as it is one for $X$. We then expand
-    $$\frac{\sum_j h(Y_j^J) V_j^J}{\sum_j V_j^J} = \frac{\sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}}.$$
-    By hypothesis,
-    $$\frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[f(X)].$$
-
-    We want to show
-    \begin{equation}\label{eq:lemma1:h}
-    \frac{\sum_j h(X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}} - \frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} 0.
-    \end{equation}
-    For this, it is sufficient to show that
-   $$ \sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}
-    -  \sum_j h(\tilde X_j^J)\frac{U_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0 $$
-    since an application of this result with $h(x)=1$ provides almost sure convergence of the denominator in (\ref{eq:lemma1:h}).
-    Write $g(\tilde X_j^J) = h(\tilde X_j^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}$. We therefore need to show that 
-    $$\sum_j Z_j^J := \sum_j \left(g(\tilde X_{k_j}^J) -  g(\tilde X_j^J) \pi_j^J \right) \stackrel{a.s.}{\to} 0.$$
-
-    Because the functional $h$ and importance sampling weights $u_{k_j}^J/\pi_{k_j}^J$ are bounded, we have that $\E[\left(Z_j^J\right)^4] < \infty$.
-We can then follow the argument of \cite{chopin20} from this point on, where
-    {THE Zs ARE ONLY CONDITIONALLY INDEPENDENT, SO WE COULD PUT IN THE CONDITIONING EVENT (I.E., THE Xs AND Us) HERE TO BE MORE ACCURATE}
-    $$\E\left[\left(\sum_j Z_j^J\right)^4\right] 
-    = J\E[(Z_1^J)^4] + 3J(J-1)\left(\E[(Z_j^J)^2]\right) \leq CJ^2,$$ for some $C>0$. By Markov, 
-    $$\mathbb{P}\left(\left|\frac{1}{J} \sum_{j=1}^J Z_j^J\right|>\epsilon\right) 
-    \leq \frac{1}{\epsilon^4J^4 } 
-    \mathbb{E}\left[\left(\sum_{j=1}^J Z_j^J\right)^4\right] \leq \frac{C}{\epsilon^4J^2},$$
-    and as these terms are summable we can apply Borel-Cantelli to conclude that these deviations happen only finitely often for every $\epsilon>0,$ giving us the almost-sure convergence for
-
-    $$ \sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}
-    -  \sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0.$$
-
-    Similarly, we also have that
-    $$ \sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J
-    - \sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J} \stackrel{a.s.}{\to} 0,$$
-
-    and the result is proved.
-
-    \kevin{MAY HAVE TO DEAL WITH CONDITIONING ON THE FILTRATION AT THE LAST TIMESTEP, AND THEN USING THE TOWER PROPERTY. EXPECTATION WILL BE THE SAME REGARDLESS FOR THE FOURTH MOMENT.}
-
-   % Let $h$ be an integrable function. 
-    %\begin{align*}
-    %    \E\left[\sum_{j=1}^J h(Y_j) v_j\right] 
-    %    &= \sum_{j=1}^J \E\left[h(X_{a(j)}) \frac{u_{a(j)}}{\pi_{a(j)}}\right] \\
-    %    &= \sum_{j=1}^J \E\left[h(X_{j}) u_{j}\right],
-    %\end{align*}
-    %and similarly for the denominator.
-Now, the numerator and denominator of the reweighted quantity have the same expectation as the numerator and denominator of the original quantity, so they must converge to $ E_\pi(h(X))$ almost surely as well.
-\end{proof}
-
-\textbf{Remark:} Note that Lemma \ref{lem:change-measure-proper-weights} permits $\pi_{1:J}$ to depend on $\{(X_j,u_j)\}$ as long as the resampling is carried out independently of $\{(X_j,u_j)\}$, conditional on $\pi_{1:J}$.
-
-\begin{lem}[Particle Marginals]
-    \label{lem:marginal-proper-weights}
-    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
-Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
-Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}$ targets $f_Z$.
-\end{lem}
-\begin{proof}
-    We want to show that, for any measurable bounded $h$, 
-     $$\frac{\sum_j h(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[h(Z)] = \E\big[\E[h(Z) | X]\big].$$
-    By hypothesis, for any measurable and bounded functional $g$ with domain $\gX$,
-    \begin{equation}\label{eq:lemma2:g}
-    \frac{\sum_j g(\tilde X_j^J)\, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[g(X)].
-    \end{equation}
-    Let $\bar{U}_j^J = \frac{J U_j^J}{\sum_j U_j^J}$.
-Examine the numerator and denominator of the quantity $$\frac{J^{-1}\sum_j h(\tilde Z_j^J) \, U_j^J}{J^{-1}\sum_j U_j^J} = \frac{J^{-1}\sum_j h(\tilde Z_j^J) \, \bar{U}_j^J}{J^{-1}\sum_j \bar{U}_j^J}.$$
-    The denominator converges to $1$ almost surely. The numerator, on the other hand, is
-    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J,$$
-    by the same fourth moment argument to the above lemma converges almost surely to the limit of its expectation
-    \begin{eqnarray*}        
-    \lim_{J\to\infty} \E\left[\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J\right] 
-    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  
-      \E\left[ h(\tilde Z_j^J)\, \bar{U}_j^J\big|\tilde X_j^J, \bar U_j^J\right]
-    \right]
-    \\
-    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  \E\big[h(Z) \big| X=\tilde X_j^J \big] \, \bar U_j^J\right].
-    \end{eqnarray*}
-    Applying (\ref{eq:lemma2:g}) with $g(x) = \E\big[h(Z) | X=x\big]$, the average on the right hand size converges almost surely to $\E\big\{\E[h(Z)|X]\big\}=\E[h(Z)].$
-    It remains to swap the limit and expectations. We can do so with the bounded convergence theorem, and therefore obtain   
-    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\,\bar{U}_j^J \stackrel{a.s.}{\to} \E[h(Z)].$$
-
-\end{proof}
-
-\begin{lem}[Particle Posteriors]
-    \label{lem:posterior-proper-weights}
-    Suppose that $\{(X_j^J,u_j^J),j=1,\dots,J\}$ targets $f_X$.
-Also suppose that $(X^{\prime J}_j,u^{\prime J}_j) = \big(X_j^J,u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$.
-Then, if $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,u^{\prime J}_j)\}$ targets $f_{X|Z}(\cdot | z^*)$.
-\end{lem}
-
-\begin{proof}
-
-    $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J}}{J^{-1} \sum_j \bar{u}_j^{' J}}$$
-
-    by
-
-    \begin{equation}
-    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}}
-    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \right)^{-1}
-    \end{equation}
-
-    Again, we want to show that
-     $$\frac{\sum_j h(X_j^J) \cdot u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)} \stackrel{a.s.}{\to} \E_{f_{X|Z}}[h(X)|z^*].$$
-
-    We already have that
-    $$\frac{\sum_j h(X_j^J)u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} \E_{f_X}[h(X)].$$
-
-    %= \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j f_{Z|X}(z^*|X_j^J)f_X(X_j^J)}
-
-    Consider the following:
-    \begin{equation}
-    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}}
-    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \right)^{-1}.
-    \end{equation}
-
-    Write $\bar{u}_j^{J} = \frac{J u_j^J}{\sum_j u_j^J}$.
-Define the weights $u^{' J}_j := u_j^J \cdot f_{Z|X}(z^*|X_j^J),$ and the corresponding self-normalized weights $\bar{u}_j^{' J} := \frac{J u^{' J}_j}{\sum_j u^{' J}_j}.$ The denominator of the below quantity
-    $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J}}{J^{-1} \sum_j \bar{u}_j^{' J}}$$
-    converges to $1$, while the numerator, by the same fourth moment argument as the first lemma, converges almost surely to the limit of its expectation, which is
-    \begin{align*}
-        \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J} \right]
-        &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)}\right] \\
-        &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
-        &= \lim_{J\to\infty} \E\left[ \sum_j h(X_j^J) \frac{\frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
-        &= \E_{X|Z}[h(X)|z^*],
-    \end{align*}
-
-    where we can swap the limit and expectations with the bounded convergence theorem.
-\end{proof}
-
-\begin{prop}[MOP-1 Targets the Posterior]
-    When $\alpha=1$ or $\phi=\theta$, MOP-$\alpha$ targets the posterior. 
-\end{prop}
-\begin{proof}
-    When $\theta=\phi$, regardless of the value of $\alpha$, the ratio $\frac{g_{n,j}^\theta}{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter estimate.
-
-    When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows.
-Recursively applying Lemmas \ref{lem:change-measure-proper-weights} \ref{lem:marginal-proper-weights}, and \ref{lem:posterior-proper-weights}, we obtain that
-    %to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
-    the MOP-1 filter targets the posterior.
-    Specifically, suppose inductively that $\big\{\big(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
-    Then, Lemma \ref{lem:marginal-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big)\big\}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
-    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-    Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
-    \[
-    \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}\big/ g^\phi_{n,a(j)}\big),
-    \]
-    with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-\end{proof}
-
-This has addressed filtering, but not quite yet the likelihood evaluation. For this we use the following lemma.
-
-\begin{lem}[Likelihood Proper Weighting]
-    \label{lem:lik-proper-weight}
-  $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1_n-1}^*;\theta)$ is consistently estimated by either the before-resampling estimate,
-\begin{equation}\label{L1}
-L_n^{B,\theta} =  \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
-\end{equation}
-or by the after-resampling estimate,
-\begin{equation}\label{L2}
-L_n^{A,\theta} = L^\phi_n \frac{\sum_{j=1}^Jw^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}.
-\end{equation}
-where $L^\phi_n$ is as defined in the various algorithms.
-\end{lem}
-
-Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ is properly weighted for $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
-To see  (\ref{L2}),
-we write the numerator of (\ref{lem:change-measure-proper-weights}) as
-\[
-L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
-= L^\phi_n \sum_{j=1}^J w_{n,j}^{FC,\theta} \frac{g^\phi_{n,j}}{L_n^\phi}
-\]
-Using Lemma \ref{lem:change-measure-proper-weights}, we resample according to probabilities $\frac{g^\phi_{n,j}}{L_n^\phi}$ to see this is properly estimated by
-\[
-L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},
-\]
-from which we obtain (\ref{L2}).
-
-Using Lemma \ref{lem:lik-proper-weight}, we obtain a likelihood estimate,
-\[
-L^{A,\theta} = \prod_{n=1}^N \left( L^\phi_n \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J w^{P,\theta}_{n,j}}\right).
-\]
-Since $w^{F,\theta}_{n,j}=w^{P,\theta}_{n+1,j}$, this is a telescoping product. The remaining terms are
-$\sum_{j=1}^J w^{P,\theta}_{0,j} = J$ on the denominator and $\sum_{j=1}^J w^{F,\theta}_{N,j}$ on the numerator.
-This derives the MOP/DOP estimate.
-
-$L^{B,\theta}$ should generally be preferred in practice, since there is no reason to include the extra variability from resampling when calculating the conditional log likelihood, but it lacks the nice telescoping product that lets us derive exact expressions for the gradient later.
-
-\section{Doubly Off-Policy Particle Filters}
-\label{app:dop}
-
-The proof for Proposition \ref{prop:dop-correctness} follows from the below two lemmas.
-Evaluating the off-policy likelihood is possible with a weighted particle filter by using resampling weight $p(y^*_n|x_n;\phi)$ and making two corrections in the particle weight, through multiplying $w_{t,j}$ by the two likelihood ratios below,
-
-\begin{equation}
-    \label{eq:dmeas-ratio}
-    s=p(y^*_n|x_n;\theta)/p(y^*_n|x_n;\phi)
-\end{equation}
-and
-\begin{equation}
-    \label{eq:rproc-ratio}
-    r=p(x_n|x_{n-1};\theta)/p(x_n|x_{n-1};\phi).
-\end{equation}
-
-The following result showcases the correctness of corrections \ref{eq:dmeas-ratio} and \ref{eq:rproc-ratio}.
-
-\begin{lem}
-    For $\theta$, $\phi$ that satisfy Assumption \ref{assump:smooth-nbhd}, the following correction in the particle filter weights,
-    \begin{equation}
-        w_{t,j} := \tilde{w}_{t,j} \cdot p(y^*|x_{t,j}^P,\phi)\cdot r \cdot s
-    \end{equation}
-    successfully recovers $\lik(\theta)$.
-\end{lem}
-
-\begin{proof}
-    Recall $\lik(\theta) = \E[\lik(\theta,\phi,\Omega, J)]=\int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega)$ and $\lik(\theta) = \int_\Omega \lik(\theta, \omega, J) d\mu(\omega)$.
-We want to evaluate
-    \begin{align*}
-        \lik(\theta,\omega, J)
-        &= \E\left[ \prod_{t=1}^T \lik_t(\theta) \middle| \omega\right]
-        = \E\left[ \prod_{t=1}^T p(y_t^*|y_{1:t-1}^*,\theta) \middle| \omega\right] \\
-        &= \E\left[ \prod_{t=1}^T \frac{1}{J} \sum_{j=1}^J  p(y_t^*|x_{t,j}^P,\theta) p(x_{t,j}^P|y_{1:t-1}^*,\theta) \middle| \omega\right] \\
-        &= \E\left[ \prod_{t=1}^T\frac{1}{J}  \sum_{j=1}^J  p(y_t^*|x_{t,j}^P,\theta) p(x_{t,j}^P|x_{t-1}^F,\theta) \middle| \omega\right] 
-    \end{align*}
-    where $p(x_t^P|y_{1:t-1}^*,\theta)$ is the prediction distribution of particles, and the filtering distribution of particles is proportional to $p(y_t^*|x_{t,j}^P,\theta)$, the likelihood of particle $j$ at time $t$.
-
-    Consider first the case with no resampling.
-Let $\lik^j(\theta)$, $\lik^j(\phi)$ be the likelihood of trajectory $j$ under $\theta$ and $\phi$ respectively.
-As the prediction and filtering particles are then the same, we use $x_{t,j}$ to refer to both.
-As such, conditional on $\omega$,
-    \begin{align*}
-        \lik^j(\theta) 
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \theta) p(x_{t,j}|x_{t-1,j}, \theta) \\
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \frac{p(y_t^*|x_{t,j}, \theta)}{p(y_t^*|x_{t,j}, \phi)} p(x_{t,j}|x_{t-1,j}, \theta) \frac{p(x_{t,j}|x_{t-1,j}, \theta) }{p(x_{t,j}|x_{t-1,j}, \phi) } \\
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \cdot s \cdot p(x_{t,j}|x_{t-1,j}, \phi)\cdot r
-    \end{align*}
-
-    and therefore, if trajectories were proposed with equal probability with particles conditional on $\omega$,
-    $$ \lik(\theta, \omega, J) = \frac{1}{J} \sum_{j=1}^J \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \cdot s \cdot p(x_{t,j}|x_{t-1,j}, \phi)\cdot r $$
-    The correction is therefore sufficient for recovering $\lik(\theta)$, with trajectories collected with the process model at $\phi$ and measurements evaluated with the measurement model at $\phi$.
-
-    Consider now the case with resampling.
-Without loss of generality we can resample every timestep.
-Here, $x_{t,j}^F(\phi) \sim p(y^*|x_{t,j}^P, \phi)$ and $x_{t,j}^P(\phi) \sim p(x_t|x_{t-1}^F,\phi)$, and similarly for $\theta$.
-We use this notation to emphasize that \textbf{both} the measurement model and the previously drawn particles depend on $\phi$ or $\theta$.
-
-    \begin{align*}
-        \lik_t(\theta) &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \theta)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\theta) \right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \theta)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \frac{p(y_t^*|x_{t,j}^P,\theta)}{p(y_t^*|x_{t,j}^P,\phi)} \right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \frac{p(y_t^*|x_{t,j}^P,\theta)}{p(y_t^*|x_{t,j}^P,\phi)} \frac{p(x_{t,j}|x_{t-1,j}, \theta) }{p(x_{t,j}|x_{t-1,j}, \phi) }\right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \cdot s \cdot r \right] \\
-        &= \E_{\omega}\E_{x_{t,j}^P \sim \process(x_{t-1}^F, \omega, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \cdot s \cdot r \middle| \omega \right]
-    \end{align*}
-
-    and again the correction is sufficient.
-
-    Therefore,
-
-    \begin{align}
-         \mathcal{L}(\theta) 
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^P(\omega), \theta) p(x_{n,j}^P(\omega) | x_{n-1,j}^P(\omega), \theta) \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\theta}(\omega), \theta) p(x_{n,j}^{P,\theta}(\omega) | x_{n,j}^P(\omega), \theta) \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\theta}(\omega), \phi) p(x_{n,j}^{P,\theta}(\omega) | y_n^*, \phi) \cdot s  \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) p(x_{n,j}^{P,\phi}(\omega) | y_n^*, \phi) \cdot r \cdot s \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) p(x_{n,j}^{P, \phi}(\omega) | y_n^*, \phi) \cdot r \cdot s \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) \cdot r \cdot s \right)  p(x_{1:N} | y_{1:N}^*, \phi) \,  d\mu(\omega) \\
-         &=: \int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega).
-     \end{align}
-
-     The proof for an arbitrary functional under the particle measure follows analogously.
-\end{proof}
-
-\begin{lem}[Gradient of Doubly Off-Policy Particle Filter]
-    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$.
-Also let $\theta \in \gN(\phi)$.
-When $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{doucet2011sf},
-    \begin{equation}
-        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
-    \end{equation}
-\end{lem}
-
-\begin{proof}
-    It is known, from \cite{doucet2009tutorial}, that 
-    $$p_\theta(x_{1:N}|y_{1:N}) = \prod_{n=1}^N \frac{p(x_n|x_{n-1})p(y_n|x_n)}{p(y_n|y_{1:n-1})} = \prod_{n=1}^N \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{\lik_n(\theta)}.$$
-
-    From the above lemma, and the fact that the particle filter is unbiased for the likelihood, we know that
-    \begin{align*}
-        \lik(\theta)
-        &= \int_{x_{1:N}} \prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) p_\theta(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N  p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \right) \frac{p_\theta(x_{1:N}|y_{1:N})}{p_\phi(x_{1:N}|y_{1:N})}p_\phi(x_{1:N}|y_{1:N}) dx_{1:N}\\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N \lik_n(\theta) \right) \left(\prod_{n=1}^N \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \lik_n(\phi)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) \lik_n(\theta)}\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N \lik_n(\phi) \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) }\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N}.
-    \end{align*}
-
-    Therefore, under regularity conditions that permit differentiating under the integral sign, using the log-derivative trick, we have
-    \begin{align*}
-        \nabla_\theta \lik(\theta)
-        &= \int_{x_{1:N}} \nabla_\theta \hat\lik(\theta) \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N}\\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N \lik_n(\phi) \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) }\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N \left(\frac{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n)}{\lik_n(\phi)}\right)^{-1} p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\frac{1}{p_\phi(x_{1:N}|y_{1:N}) }\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\nabla_\theta \log\left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} p_\theta(x_{1:N},y_{1:N}) \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) p_\theta(x_{1:N}|y_{1:N}) \cdot \lik(\theta) \; dx_{1:N}. 
-    \end{align*}
-
-    Using the log-derivative trick again,
-    \begin{align*}
-        \nabla_\theta \ell(\theta)
-        &= \frac{\nabla_\theta \lik(\theta)}{\lik(\theta)} \\
-        &= \int_{x_{1:N}} \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) p_\theta(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \E_{x_{1:N}|y_{1:N}, \theta}\left[\nabla_\theta \left(\sum_{n=1}^N\log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right)\right] \\
-        &= \E_{x_{1:N}|y_{1:N}, \theta}\left[\nabla_\theta\log p_\theta(x_{1:N},y_{1:N})\right] \\
-        &= \lim_{J\to\infty} \E_{x_{1:N}|y_{1:N}, \theta}\left[\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log p_\theta(x_{1:N,j}^{A,F},y_{1:N}^*)\right].
-    \end{align*}
-
-    But it also holds that
-    \begin{align*}
-        \nabla_\theta \ell(\theta)
-        &= \frac{\nabla_\theta \lik(\theta)}{\lik(\theta)} \\
-        &= \int_{x_{1:N}} \frac{\nabla_\theta \hat\lik(\theta)}{\lik(\theta)} \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \hat\ell(\theta) \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N}\\
-        &= \lim_{J\to\infty} \E_{x_{1:N}|y_{1:N}, \phi}\left[\nabla_\theta \hat\ell(\theta)\right].
-    \end{align*}
-
-    and we see that $$\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log p_\theta(x_{1:N,j}^{A,F},y_{1:N}^*)$$ and $$\nabla_\theta \hat\ell(\theta)$$ both converge to the same thing in expectation.
-They are both particle approximations under the posterior where the variance of each is bounded, therefore one can apply the particle CLT from \cite{chopin2004clt} to obtain that they are both consistent estimates of the score.
-
-    In fact, when $\theta=\phi,$ they are the same in distribution. \kevin{claim}
-
-    \kevin{This is a far weaker result than what I really wanted. This just gives consistency. Want equality in distribution.}
-\end{proof}
-
-\begin{lem}
-
-Consider the case of DOP-$\alpha$ when $\alpha=0$ and $\theta=\phi$. The gradient estimate is then
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta\right),
-    \end{equation}
-
-    yielding the gradient estimator of \cite{blei2018vsmc} that \cite{scibior2021dpf} show is the gradient of a vanilla particle filter when the resampling terms are dropped.
-
-\end{lem}
-
-\begin{proof}
-    When $\alpha=0,$ the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J r_{n,j}s_{n,j}}{\sum_{j=1}^J r_{n,j}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J r_{n,j}s_{n,j}.
-\end{equation}
-
-Its gradient when $\theta=\phi$ is therefore 
-\begin{align*}
-    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right) \\
-    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right)} \\
-    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta r_{n,j}s_{n,j}}{\sum_{j=1}^J r_{n,j}s_{n,j}} \\
-    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{Y_n, X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)} \\
-    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta\right),
-\end{align*}
-where we use the log-derivative trick, observe that $\sum_{j=1}^J r_{n,j}s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again. 
-\end{proof}
-
-\subsection{Off-Policy Weighted Differentiable Particle Filter}
-
-We are finally ready to introduce the off-policy weighted differentiable particle filter.
-In the event where a seed is fixed, all expressions in Algorithm \ref{alg:dop} can be considered to be conditional on $\omega \in \Omega$.
-
-\begin{algorithm}[ht]
-\centering
-	\caption{Doubly Off-Policy-$\alpha$}
-    \label{alg:dop}
-	\begin{algorithmic}[1]
-	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
-		\STATE Initialize filter particles ${X}_{0,j}^{F,\phi}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\phi}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$. Fix $\omega.$
-		\FOR{$n=1,...,N$}
-            \STATE Prediction weights with discounting: $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
-            \label{dop-alpha:discount}
-            \STATE Simulate for prediction:
-            ${X}_{n,j}^{P,\phi}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\phi}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{dop-alpha:step1}
-            \STATE Adjust weights: $\displaystyle w_{n,j}^{P,\theta} = w_{n,j}^{P,\theta} \times
-            \frac{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\phi}|{X}_{n-1,j}^{F};{\theta}\big)}{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\phi}|{X}_{n-1,j}^{F};{\phi}\big)}$ for $j$ in $1:J$
-            \label{dop-alpha:dproc}
-            \STATE Evaluate measurement density:
-            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\phi}\giventh{\theta})$ for $j$ in $\seq{1}{J}$
-            \STATE Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \STATE Conditional likelihood under $\phi$: 
-            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
-            \label{dop-alpha:Lphi}
-            \STATE Normalize weights:
-            $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
-            for $j\ \mathrm{in}\ \seq{1}{J}$
-            \STATE Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \label{dop-alpha:systematic}
-            \STATE Resample particles: ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$
-            \STATE Filter weights corrected for resampling:
-            $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{dop-alpha:weight:update}
-            \STATE Resample filter weights:
-            $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
-            for $j$ in $\seq{1}{J}$ \label{dop-alpha:step2}
-            \STATE After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \ENDFOR
-		\RETURN likelihood estimate $\hat{\lik}(\theta) = \lik(\theta, \phi, \omega, J) := \prod_{n=1}^N L_n^{A,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}.$
-	\end{algorithmic}
-\end{algorithm}
-
-Here, we consider a doubly off policy particle filter, meaning that both \code{rprocess} and \code{dmeasure} are computed at $\phi$ and particles are reweighted to correspond to $\theta$.
-
-\section{Measurement Off-Policy Particle Filters}
-
-\begin{algorithm}[ht]
-\centering
-	\caption{Measurement Off-Policy-$\alpha$}
-    \label{alg:mop}
-	\begin{algorithmic}[1]
-	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
-        \IF{$\theta \neq \phi$}
-            \STATE Run a particle filter once with parameters $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$, with seed $\omega$.
-        \ELSE
-        \STATE Set $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$ to be copies of $X_{n,j}^{P,\theta}, X_{n,j}^{P,\theta}$ in the rest of the algorithm.
-        \ENDIF
-		\STATE Initialize filter particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$. Fix $\omega.$
-		\FOR{$n=1,...,N$}
-            \STATE Prediction weights with discounting: $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
-            \label{mop-alpha:discount}
-            \STATE Simulate for prediction:
-            ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{mop-alpha:step1}
-            \STATE Evaluate measurement density:
-            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for $j$ in $\seq{1}{J}$
-            \STATE Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \STATE Conditional likelihood under $\phi$: 
-            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
-            \label{mop-alpha:Lphi}
-            \STATE Normalize weights:
-            $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
-            for $j\ \mathrm{in}\ \seq{1}{J}$
-            \STATE Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \label{mop-alpha:systematic}
-            \STATE Resample particles: ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$
-            \STATE Filter weights corrected for resampling:
-            $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{mop-alpha:weight:update}
-            \STATE Resample filter weights:
-            $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
-            for $j$ in $\seq{1}{J}$ \label{mop-alpha:step2}
-            \STATE After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \ENDFOR
-		\RETURN likelihood estimate $\hat{\lik}(\theta) = \lik(\theta, \phi, \omega, J) := \prod_{n=1}^N L_n^{A,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}.$
-	\end{algorithmic}
-\end{algorithm}
-
-\begin{lem}
-    Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The gradient estimate is then
-
-    \begin{equation}
-        \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
-    \end{equation}
-
-    yielding the gradient estimators of \cite{poyiadjis11, scibior2021dpf} when applied to the bootstrap filter. 
-\end{lem}
-
-\begin{proof}
-
-Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The likelihood estimate becomes
-\begin{equation*}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\phi}}= \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\phi}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
-\end{equation*}
-as we have a telescoping product. Then, as
-
-$$\nabla_\theta \hat\ell(\theta) = \frac{\nabla_\theta \hat\lik(\theta)}{\hat\lik(\theta)} = \frac{\nabla_\theta\left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi}{\prod_{n=1}^N L_n^\phi} =  \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},$$
-
-The derivative of the log-likelihood estimate is then
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},
-\end{equation*}
-which we decompose as follows.
-
-Observe that as $\alpha=1$,
-
-$$w_{n,j}^{P,\theta} = w_{n-1,\theta}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,k_j}^\theta}{g_{i,k_j}^\phi}.$$
-
-Using the log-derivative identity again,
-$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,k_j}^\theta}{g_{i,k_j}^\phi}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,k_j}^\theta - \log g_{i,k_j}^\phi\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,k_j}^\theta.$$
-
-So 
-$$  \nabla_\theta \sum_{n=1}^N \log g_{n,k_j}^\theta = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,k_j}^\theta\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,k_j}^{P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right),$$
-
-and 
-$$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{i=1}^N \nabla_\theta \log g_{n,k_j}^\theta = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$
-
-Substituting, we have that
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right),
-\end{equation*}
-and finally, observing that when $\theta=\phi$ we have that $w_{N,j}^{F,\theta}=1$, that 
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
-\end{equation*}
-
-yielding the gradient estimators of \cite{poyiadjis11, scibior2021dpf} when applied to the bootstrap filter. 
-\end{proof}
-
-\begin{lem}
-
-Consider the case of MOP-$\alpha$ when $\alpha=0$ and $\theta=\phi$. The gradient estimate is then
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
-    \end{equation}
-
-    yielding the gradient estimator of \cite{blei2018vsmc} that \cite{scibior2021dpf} show is the gradient of a vanilla particle filter when the resampling terms are dropped -- when applied to the bootstrap filter.
-
-\end{lem}
-
-\begin{proof}
-    When $\alpha=0,$ the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
-\end{equation}
-
-Its gradient when $\theta=\phi$ is therefore 
-\begin{align*}
-    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
-    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)} \\
-    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta s_{n,j}}{\sum_{j=1}^J s_{n,j}} \\
-    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)}{f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \phi}; \phi)} \\
-    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
-\end{align*}
-where we use the log-derivative trick, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again. 
-\end{proof}
-
-\kevin{Show bias is order N, variance is order N when alpha below 1. That implies optimal choice of alpha is not 1. }
-
-\section{Finite-Particle Analysis of Gradient Methods with Algorithm \ref{alg:ifad}}
-\label{app:ifad}
-
-The analysis in this section roughly follows the analysis in Roosta-Khorasani and Mahoney \cite{mahoney2016subsampled}, except with the caveat that none of the matrix concentration bounds they use apply here as the particles are dependent.
-We instead use the concentration inequality from Del Moral and Rio \cite{delmoral2011ci} to bound the gradient and Hessian estimates.
-In this section, we fix $\omega \in \Omega$ only within each filtering iteration, evaluate Algorithm \ref{alg:mop} at $\theta=\phi$, and analyze Algorithm \ref{alg:ifad} post-iterated filtering.
-
-The convergence analysis in Theorem \ref{thm:convergence} is limited to the case where $-\ell$ is $\gamma$-strongly convex.
-Though it is true that in a neighborhood of the optimum local asymptotic normality holds and the log-likelihood is strongly convex in this neighborhood, in practice likelihood surfaces for POMPs are often highly nonconvex globally.
-The convergence to an optimum, local or global, must therefore be sensitive to initialization.
-
-\subsection{Bounding the Gradient}
-\kevin{Add a few words to explain that the gradient of the log-likelihood w.r.t.
-parameters is a functional of the particle measure under boundedness assumptions and fixed parameters, appeal to authority and say doucet did it too.
-}
-
-\kevin{Gradient linear map, bound conditional likelihoods at time n.
-Alternatively, convert del moral and doucet expectation bound to high-probability bound.
-https://arxiv.org/pdf/1905.11546.pdf for trick.
-Alternatively, concatenate previous states.
-}
-
-\kevin{Show MOP-1 is unbiased. Extend analysis to not-1 case. }
-
-\begin{lem}[Concentration of Measure for Gradient Estimate]
-    \label{lemma:grad_bound}
-    Assume $\text{osc} \left(\frac{\partial}{\partial_\theta} \hat{\ell}_\theta(x_{n,j}^P) \right) \leq 1$.
-For $||\nabla_\theta \hat{\loglik}_\theta(x_{n,j}^P) - \nabla_\theta \loglik_\theta(x^P)|| = ||\nabla_\theta \hat{\loglik}(\theta) - \nabla_\theta \loglik(\theta)||$ to be bounded by $\epsilon$ with probability $1-\delta$, we require
-    \begin{equation}
-        J > \max\left\{\frac{r_t\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{6p}{\delta}\right)\right)\right),\;\; \frac{2p\log(6p/\delta)}{\epsilon^2}\beta_t^2, \;\;\frac{G(\theta)^2}{\epsilon^2}p\log\left(\frac{6p}{\delta}\right)\right\}
-    \end{equation}
-    where $G(\theta)$ is defined as in Assumption \ref{assump:local-bounded-derivative}, $h(t) = \frac{1}{2}(t - \log(1+t))$, and $\beta_t$ and $r_t$ are two additional model-specific parameters that do not depend on $J$.
-\end{lem}
-
-\begin{proof}
-
-We will seek to use the concentration inequality of \cite{delmoral2011ci} to bound the deviation of the gradient estimate from the gradient of the negative log-likelihood in the sup norm with a union bound.
-To use this, we will first need to resample the particles once more to reassign the normalized weights to $1/J$.
-
-Perform another resampling step at time $T$ to obtain indices $k_j \sim \text{Categorical}(\bar{w}_{T,1}, ..., \bar{w}_{T,J})$.
-Obtain ancestral trajectories of surviving time $T$ filtering particles $x_{k_j, 1:T}^{A}$.
-Define $g_j(\theta) = \nabla_\theta -\log L_{T, k_j}$, the gradient of the negative logarithm of the accumulated importance weights of particle $k_j$ along its ancestral trajectory.
-
-We can bound
-\begin{align}
-    ||\nabla_\theta \hat{\loglik}(\theta) - \nabla_\theta \loglik(\theta)|| 
-    \leq 
-    \left\lVert (\nabla_\theta - \hat{\loglik}(\theta)) - \frac{1}{J}\sum_{j=1}^J g_j(\theta)\right\rVert 
-    + \left\lVert\frac{1}{J}\sum_{j=1}^J g_j(\theta) - (\nabla_\theta -\loglik(\theta))\right\rVert
-\end{align}
-
-We begin by examining the first term. This term bounds the error of the gradient estimate that the differentiable particle filter provides from the gradient estimate after the final resampling step.
-
-This amounts to solving the following problem.
-Say you have a weighted mean $\sum w_j x_j$, where $j=1,\dots,J$.
-The $x_j$ are fixed numbers bounded by some constant $G$ and the $w_j$ are non-negative and sum to $1$.
-Resample indices $k_j, j=1,\dots,J$ according to the multinomial distribution to get $x_{k_j}$ for each $j$, and also $J^{-1} \sum x_{k_j}$.
-We want to bound $|\sum w_j x_j - J^{-1} \sum x_{k_j}|$.
-
-By a simple application of Hoeffding's inequality and a union bound, we have
-
-\begin{align}
-    \prob\left(\exists \; j \;s.t.\; \left|(\nabla_\theta - \hat{\loglik}(\theta)) - \frac{1}{J}\sum_{j=1}^J g_j(\theta)\right| \geq \epsilon \right)
-    &\leq 2p\exp\left(\frac{-2\epsilon^2J}{G(\theta)^2}\right) 
-\end{align}
-
-We require \begin{equation}
-    J > \frac{G(\theta)^2}{2\epsilon^2}\log(2p/\delta)
-\end{equation} particles to bound $\left\lVert (\nabla_\theta - \hat{\loglik}(\theta)) - \frac{1}{J}\sum_{j=1}^J g_j(\theta)\right\rVert_\infty < \epsilon$ with probability $1-\delta$, where $G(\theta)$ is defined as in Assumption \ref{assump:local-bounded-derivative}.
-
-We can bound the sup norm by $||x||_\infty \leq \sqrt{p}||x||_2$.
-So to bound $\left\lVert (\nabla_\theta - \hat{\loglik}(\theta)) - \frac{1}{J}\sum_{j=1}^J g_j(\theta)\right\rVert < \epsilon$ with probability $1-\delta/3$, we scale the previous $\epsilon$ to $\epsilon/\sqrt{p}$ and therefore require
-
-\begin{equation}
-    J > \frac{G(\theta)^2}{2\epsilon^2}p\log\left(\frac{6p}{\delta}\right)
-\end{equation}
-
-We now tackle the second term.
-Write $g_{ij}$ for the $i$-th component of the gradient of the negative log-likelihood with respect to the $j$-th particle, $(-\bar{w}_{T,\np} \nabla_\theta\log L_{T,j})_i$.
-
-Apply the Hoeffding-type concentration inequality from Del Moral and Rio \cite{delmoral2011ci} and another union bound to find that
-
-\begin{align}
-    \left\lVert\frac{1}{J}\sum_{j=1}^J g_{ij} - \left(-\frac{\partial}{\partial \theta_i}\ell(\theta)\right)\right\rVert_\infty \leq \frac{r_t}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_t 
-\end{align}
-
-with probability $1-2p\exp(-t)$.
-We split the remaining $2\delta/3$ failure probability among these two terms, to find $\delta/3=2p\exp(-t) => t=\log(6p/\delta)$.
-Here, $h(t) = \frac{1}{2}(t - \log(1+t))$.
-The two additional model-specific parameters are $\beta_t$ and $r_t$, which do not depend on $J$.
-
-The analogous bound for the 2-norm falls out by scaling the right-hand side by $\sqrt{p}$, to require
-
-\begin{align}
-    \epsilon = \frac{r_t\sqrt{p}}{J}(1+h^{-1}(\log(6p/\delta))) + \sqrt{\frac{2p\log(6p/\delta)}{J}}\beta_t
-\end{align}
-
-and therefore need
-
-\begin{align}
-    J > \max\left\{\frac{r_t\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{6p}{\delta}\right)\right)\right), \frac{2p\log(6p/\delta)}{\epsilon^2}\beta_t^2\right\}
-\end{align}
-
-Overall, we finally require
-
-\begin{equation}
-        J > \max\left\{\frac{r_t\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{6p}{\delta}\right)\right)\right),\;\; \frac{2p\log(6p/\delta)}{\epsilon^2}\beta_t^2, \;\;\frac{G(\theta)^2}{\epsilon^2}p\log\left(\frac{6p}{\delta}\right)\right\}
-\end{equation}
-
-\end{proof}
-
-\subsection{Bounding Hessian Estimates}
-
-Should one choose to use a second-order method involving a particle Hessian estimate, we provide a guarantee for its positive-definiteness below.
-
-\begin{lem}[Minimum Eigenvalue Bound for Hessian Estimate]
-    \label{lemma:hess_bound}
-    Assume that the Hessian of the negative log-likelihood $H=\sum_{j=1}^J \E H_j$ has a minimum eigenvalue $\gamma > 0$, that $\E \lambda_{\min} (H_j) = \gamma' > 0$, and finally that $\text{osc}(\lambda_{\min}) \leq 1$.
-
-    We then require 
-    \begin{equation}
-        J > \max\left\{\frac{2r_t(1+h^{-1}(t)) + 2c}{\gamma'}, \frac{2(2t\beta_t^2+c)^2}{\gamma'^2}\right\} \geq  \frac{r_t(1+h^{-1}(t))}{\gamma'} + \sqrt{2tJ}\beta_t/\gamma' + c/\gamma'
-    \end{equation}
-
-    for $\hat{H}(\theta)$ to be invertible and positive definite with minimum eigenvalue greater than or equal to $c \in (0, \sum_{j=1}^J \E\lambda_{\min}(H_j))$ and with probability at least $1-\exp(-t)$.
-\end{lem}
-\begin{proof}
-Write $\hat{H}(\theta) = \hat{H} = \sum_{j=1}^J H_j$ for the estimate of the negative of the Hessian, where $H_j$ is an element of the outer sum over the $J$ particles.
-
- As the negative log-likelihood is convex, we want to bound the minimum eigenvalue of $\hat{H}(\theta)$ from below with high probability, so that all the eigenvalues of $\hat{H}(\theta)$ are positive with high probability.
-This ensures that the estimated Hessian is invertible and positive-definite.
-
-From \cite{boyd2004convex}, we know that the minimum eigenvalue of a symmetric matrix is concave. Therefore, it suffices to show that the first inequality in the below expression
-
-\begin{equation}
-    0 < \sum_{j=1}^J \lambda_{\min} (H_j) \leq  \lambda_{\min}\left(\sum_{j=1}^J H_j\right) = \lambda_{\min} (\hat{H})
-\end{equation}
-
-holds with high probability.
-
-Apply the particle Hoeffding concentration inequality from \cite{delmoral2011ci} to find that
-
-\begin{align}
-    \frac{1}{J}\sum_{j=1}^J \lambda_{\min}(H_j) - \E_{\tilde{\pi}_t}\lambda_{\min}(H_j) &= \frac{1}{J}\sum_{j=1}^J \lambda_{\min}(H_j) - \gamma' \geq -\frac{r_t}{J}(1+h^{-1}(t)) - \sqrt{\frac{2t}{J}}\beta_t \\
-    \sum_{j=1}^J \lambda_{\min}(H_j) &\geq -r_t(1+h^{-1}(t)) - \sqrt{2tJ}\beta_t + J\gamma'
-\end{align}
-
-with probability at least $1-\exp(-t)$. Here, $h(t) = \frac{1}{2}(t - \log(1+t))$. The two additional model-specific parameters are $\beta_t$ and $r_t$, which do not depend on $J$.
-
-We additionally require, for $c \in (0, \sum_{j=1}^J \E\lambda_{\min}(H_j))$,
-\begin{align}
-    \sum_{j=1}^J \lambda_{\min}(H_j) \geq -r_t(1+h^{-1}(t)) - \sqrt{2tJ}\beta_t + J\gamma' \geq c \\
-    J\gamma' \geq c + r_t(1+h^{-1}(t)) + \sqrt{2tJ}\beta_t 
-\end{align}
-
-We therefore require $$J > \max\left\{\frac{2r_t(1+h^{-1}(t)) + 2c}{\gamma'}, \frac{2(2t\beta_t^2+c)^2}{\gamma'^2}\right\} \geq  \frac{r_t(1+h^{-1}(t))}{\gamma'} + \sqrt{2tJ}\beta_t/\gamma' + c/\gamma'$$
-
-for $\hat{H}(\theta)$ to be invertible and positive definite with minimum eigenvalue greater than or equal to $c$ with probability at least $1-\exp(-t)$.
-
-\end{proof}
-
-\subsection{Convergence Analysis of Theorem \ref{thm:convergence}}
-
-\begin{proof}
-
-In this analysis, we largely follow the proof of Theorem 6 in \cite{mahoney2016subsampled}.
-
-Define $\theta_\eta = \theta_n + \eta p_n$, where $p_n=-(H(\theta_n))^{-1}g(\theta_n)$.
-As in Roosta-Khorasani and Mahoney \cite{mahoney2016subsampled}, we want to show there is some iteration-independent $\tilde{\eta}>0$ such that the Armijo condition
-\begin{equation}
-    f(\theta_n+\eta p_n) \leq f(\theta_n) + \eta\beta p_n^Tg(\theta_n)
-\end{equation}
-holds for any $0< \eta < \tilde{\eta}$ and some $\beta \in (0,1)$.
-
-By an argument found in the beginning of the proof of Theorem 6 in \cite{mahoney2016subsampled}, we have that choosing $J$ such that $||\nabla_\theta\hat{\ell}(\theta_n) - \nabla_\theta \ell(\theta_n)|| \leq \epsilon$ and $\lambda_{\min}(H(\theta_n)) \geq c$ with probability $(1-\delta)^2$ for each $n$ yields
-
-\begin{align}
-    f(\theta_\eta)-f(\theta_n) \leq \eta p_n^Tg(\theta_n) + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2
-\end{align}
-
-Now, for some $c>0$, with probability $1-\delta$
-\begin{equation}
-    p_n^Tg(\theta_n) = -p_n^TH(\theta_n)p_n \geq -c||p_n||^2
-\end{equation}
-and we can obtain a decrease in the objective.
-
-Substituting this into the previous expression,
-\begin{align}
-    f(\theta_\eta)-f(\theta_n) \leq -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2
-\end{align}
-
-and the Armijo condition then becomes
-\begin{align}
-    -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2 &\leq \eta \beta p_n^Tg(\theta_n) = - \eta \beta p_n^TH(\theta_n)p_n \\
-    \epsilon||p_n|| + \eta \Gamma ||p_n||^2 / 2 &\leq (1- \beta) p_n^TH(\theta_n)p_n \\
-    \epsilon + \eta \Gamma ||p_n|| / 2 &\leq c(1- \beta) ||p_n||
-\end{align}
-
-which holds and guarantees an iteration-independent lower bound if 
-\begin{equation}
-    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_n)|| \leq \frac{c(1-\beta)}{2}||p_n||
-\end{equation}
-
-Now, first note that
-\begin{equation}
-||g(\theta_n)|| - ||\nabla_\theta f(\theta_n)|| \leq ||g(\theta_n) - \nabla_\theta f(\theta_n)|| \leq \epsilon => ||\nabla_\theta f(\theta_n)|| \geq ||g(\theta_n)|| - \epsilon
-\end{equation} and
-\begin{equation}
-||\nabla_\theta f(\theta_n)||-||g(\theta_n)|| \leq ||\nabla_\theta f(\theta_n)-g(\theta_n)|| \leq \epsilon => ||g(\theta_n)|| \geq ||\nabla_\theta f(\theta_n)|| - \epsilon
-\end{equation}
-
-There are now two cases. If the algorithm terminates and $||g(\theta_n)|| \leq \sigma \epsilon$, we can derive
-
-\begin{equation}
-    ||\nabla_\theta f(\theta_n)|| \leq ||g(\theta_n)|| + \epsilon = \sigma\epsilon+\epsilon = (\sigma+1)\epsilon
-\end{equation}
-
-If the algorithm does not terminate, then $||g(\theta_n)|| > \sigma \epsilon$. Notice that
-
-\begin{align}
-    \epsilon \geq ||g(\theta_n) - \nabla_\theta f(\theta_n)|| &\geq ||g(\theta_n)|| - ||\nabla_\theta f(\theta_n)|| \\
-    ||\nabla_\theta f(\theta_n)|| + \epsilon &\geq ||g(\theta_n)|| \geq \sigma \epsilon \\
-    ||\nabla_\theta f(\theta_n)|| &\geq \sigma \epsilon - \epsilon = (\sigma - 1)\epsilon \\
-    \frac{||\nabla_\theta f(\theta_n)||}{\sigma-1} &\geq \epsilon
-\end{align}
-
-and now 
-\begin{align}
-    ||\nabla_\theta f(\theta_n)|| - \epsilon
-    &\geq  ||\nabla_\theta f(\theta_n)|| - \frac{||\nabla_\theta f(\theta_n)||}{\sigma-1} = \left(1-\frac{1}{\sigma-1}\right)||\nabla_\theta f(\theta_n|| = \frac{\sigma-2}{\sigma-1}||\nabla_\theta f(\theta_n)|| \geq \frac{2}{3}||\nabla_\theta f(\theta_n)||
-\end{align}
-
-From $||A^{-1}|| = 1/\sigma_{\min}(A)$,
-\begin{align}
-    p_n^TH(\theta_n)p_n &= (-(H(\theta_n))^{-1}g(\theta_n))^TH(\theta_n)(-(H(\theta_n))^{-1}g(\theta_n)) \\
-    &= g(\theta_n)^T(H(\theta_n))^{-1}g(\theta_n) \\
-    &\geq \frac{1}{c}||g(\theta_n)||^2 \\
-    &\geq \frac{1}{c}(||\nabla_\theta f(\theta_n)|| - \epsilon)^2 \\
-    &\geq \frac{4}{9c}||\nabla_\theta f(\theta_n)||^2
-\end{align}
-
-From the assumption that $f$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 -\ell \preceq \Gamma I$, by an implication of $\gamma$-strong convexity we have
-\begin{align}
-    f(\theta_n) - f(\theta^*) \leq \frac{1}{2\gamma}||\nabla_\theta f(\theta_n)||^2
-\end{align}
-and we put together:
-\begin{equation}
-    f(\theta_n) - f(\theta^*) \leq \frac{1}{2\gamma}||\nabla_\theta f(\theta_n)||^2 \leq \frac{9c}{4}\frac{1}{2\gamma}p_n^TH(\theta_n)p_n
-\end{equation}
-\begin{equation}
-    \frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \leq \frac{4}{9c}||\nabla_\theta f(\theta_n)||^2 \leq p_n^TH(\theta_n)p_n
-\end{equation}
-\begin{equation}
-    f(\theta_n) - f(\theta^*) \leq \frac{9c}{8\gamma}p_n^TH(\theta_n)p_n
-\end{equation}
-
-\begin{equation}
-    -\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \geq -\frac{4}{9c}||\nabla_\theta f(\theta_n)||^2 \geq -p_n^TH(\theta_n)p_n
-\end{equation}
-
-But from earlier, as the Armijo condition is fulfilled with our choice of $\eta$ and $\epsilon$,
-\begin{align}
-    f(\theta_{n+1})-f(\theta_n) &\leq -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2 \\
-    &\leq -\eta\beta p_n^TH(\theta_n)p_n \\
-    &\leq -\eta\beta\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) 
-\end{align}
-
-and therefore
-\begin{align}
-    f(\theta_{n+1}) - f(\theta^*) 
-    &= f(\theta_{n+1})-f(\theta_n)+f(\theta_n)- f(\theta^*) \\
-    &\leq f(\theta_n)- f(\theta^*) -\eta\beta\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \\
-    &= (1-\eta\beta\frac{8\gamma}{9c})(f(\theta_n) - f(\theta^*)) 
-\end{align}
-
-\end{proof}
-
 \end{document}

### 2024-03-25 mop/draft.tex
Updates from Overleaf

@@
 \label{assump:bounded-measurement}
-There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G} < \infty$ and $\infty > G' \geq G'(\theta)>0$ such that $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty < G'(\theta) \leq G'$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
+There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G} < \infty$ and $\infty > G' \geq G'(\theta)>0$ such that $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty <\leq G'(\theta) \leq G'$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
 \end{aspt}
@@
     $$
-    \|\nabla_\theta \ell(\theta, \phi, \omega, J)\|_\infty < NG'(\theta) \leq G(\theta) \leq G<\infty, \;\;
-    \|\nabla^2_\theta \ell(\theta, \phi, \omega, J)\|_\infty <N^2G'(\theta)^2 \leq G(\theta)^2 \leq G^2<\infty
+    \|\nabla_\theta \ell(\theta, \phi, \omega, J)\|_\infty \leq G(\theta) \leq G<\infty, \;\;
+    \|\nabla^2_\theta \ell(\theta, \phi, \omega, J)\|_\infty \leq G(\theta)^2 \leq G^2<\infty
     $$
@@
 
+\paragraph{Acknowledgements.} We thank Nicholas Chopin for helpful comments on the proof of a strong law of large numbers for a suitably reweighted particle filter.
+
+\kevin{Reach: PNAS.
+Else try for JRSSB, try to get Chopin for Annals AE.
+Would need to change bracketing conventions for Biometrika.
+If submitting to PNAS, JASA, Biometrika, or JRSSB, highlight the different statistical gradient estimator and finite sample performance, rewrite introduction, maybe create a way to use the gradients for inference rather than using profiling.
+Try to rewrite it as a breakthrough in statistics.
+Safety: JMLR}
+
 \bibliography{bib-ifad,bib-ref}
@@
 
+\subsection{Analysis of MOP-$0$ Variance}
+
 We made Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} to leverage the results from \cite{karjalainen2023} on the forgetting of the particle filter.
@@
 \begin{align*}
-    \text{Var}(\nabla_\theta \hat\ell(\theta) &= \frac{1}{J^2}\text{Var}\left(\sum_{n=1}^N\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    \text{Var}(\nabla_\theta \hat\ell(\theta)) &= \frac{1}{J^2}\text{Var}\left(\sum_{n=1}^N\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
     &= \frac{1}{J^2}\sum_{n=1}^N\text{Var}\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &+ 2\sum_{m<n}\text{Cov}\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    &+ 2\sum_{m<n}\text{Cov}\left(\frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
     &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta)).
@@
 $$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
-i.e. the mixing time of the particle filter is $O(\log(J))$.
-
-The particle filter itself is therefore strongly mixing, with $\alpha$-mixing coefficients $\alpha(k) \leq (1-\epsilon)^{\floor{k/(c\log(J))}}$.
+i.e.
+the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
+
+\kevin{How does the dobrushin contraction coefficient relate to the alpha-mixing coefficients?
+If you have a bound on the mixing time $b$, then the alpha mixing coefficients should just be $\alpha(t) \leq e^{-t/b}$.}
+
+By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strongly mixing, with $\alpha$-mixing coefficients $\alpha(k) \leq (1-\epsilon)^{\floor{k/(c\log(J))}}$.
 Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
-Therefore, by Davydov's inequality and noting that $\nabla_\theta\hat\ell_n(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement},
-
-$$\text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) \leq G'(\theta)^2(1-\epsilon)^{\floor{\frac{|n-m|}{2c\log(J)}}}.$$
-
-We can then conclude with Popoviciu's inequality that 
+Therefore, by Davydov's inequality, noting that $\nabla_\theta\hat\ell_n(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and WLOG assuming $\E[(\nabla_\theta\hat\ell_m(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/4}$, we see that
+\begin{align*}
+    \text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) 
+    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat\ell_m(\theta))^4]^{1/4}\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/4}\\
+    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2}.
+\end{align*}
+
+To bound this, we use the fact that $$\E[(\nabla_\theta\hat\ell_n(\theta))^4] = \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4]+\E[(\nabla_\theta\hat\ell_n(\theta))^2]^2$$
+alongside Theorem 2 of \cite{delMoral03}, which shows that
+$$\E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4] \lesssim \frac{c_p^4G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^2] \lesssim \frac{c_p^2G'(\theta)^2}{J},$$
+where $c_p$ is a term depending only on the dimension of the state space $p$ (but where the dependence may be exponential), to find that 
+$$\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{c_p^4G'(\theta)^4}{J^2}+\left(\frac{c_p^2G'(\theta)^2}{J}\right)^2} = \frac{c_p^2G'(\theta)^2}{J},$$
+and conclude that 
+$$\text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{c_p^2G'(\theta)^2}{J}.$$
+
+Putting it all together, we see that
 \begin{align*}
     \text{Var}(\nabla_\theta \hat\ell(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta))\\
-    &\leq \frac{NG'(\theta)^2}{4J} + 2\sum_{n=1}^{N-1} (N-n) G'(\theta)^2(1-\epsilon)^{\floor{\frac{n}{2c\log(J)}}}\frac{G'(\theta)^2}{4J} \\
-    &\leq \frac{G'(\theta)^2}{4J} \left(N + 2\sum_{n=1}^{N-1} (N-n) G'(\theta)^2(1-\epsilon)^{\floor{\frac{n}{2c\log(J)}}}\right) \\
-    &\lesssim \frac{NG'(\theta)^2}{J}.
+    &\leq \frac{Nc_p^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\frac{c_p^2G'(\theta)^2}{J} \\
+    &\leq \frac{c_p^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\right) \\
+    &\lesssim \frac{c_p^2G'(\theta)^2N}{J}.
 \end{align*}
@@
 \paragraph{Remark:} Note that the factor of $N$ that pops up here is due to the use of the unnormalized gradient.
-If one divides the gradient estimate by $N$ before usage, the variance does not depend on the horizon.
+If one divides the gradient estimate by $\sqrt{N}$ before usage, the variance does not depend on the horizon.
+If one divides by $N$, the error is $O(1/\sqrt{NJ})$!
 
@@
 
+\subsection{Analysis of MOP-$\alpha$ Variance}
+
 Next, we analyze the case for $\alpha \in (0,1)$.
-
-Notice that
-\begin{align*}
-    w_{n,j}^{F,\theta}
-    &= \left(w_{n-1,k_j}^{F, \theta}\right)^\alpha\frac{g_{n,k_j}^\theta}{g_{n,k_j}^\phi} \\
-    &= \left(\left(w_{n-2,j}^{A,F, \theta}\right)^\alpha\frac{g_{n-1,j}^{A,F,\theta}}{g_{n-1,j}^{A,F,\phi}}\right)^\alpha\frac{g_{n,j}^{A,F,\theta}}{g_{n,j}^{A,F,\phi}} \\
-    &= \prod_{i=1}^n \left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}.
-\end{align*}
-
-Then, we have that
-\begin{align*}
-    \nabla_\theta\hat\ell(\theta) 
-    &= \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^Jw_{n,j}^{F,\theta}}{\sum_{j=1}^J\left(w_{n-1,j}^{F, \theta}\right)^\alpha}\right)= \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right).
-\end{align*}
-
-Consider the modified estimator, where we truncate weights beyond $k$ timesteps:
-\begin{align*}
-    \hat{s}_k(\theta) 
-    &:= \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right).
-\end{align*}
-
-Write $$w_{n,j}^{F,\theta,k} := \prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}$$ for the $k$-truncated weights.
-
-By Assumption \ref{assump:bounded-measurement} we have that
-
-\begin{align*}
-   w_{n,j}^{F,\theta} 
-    &\leq w_{n,j}^{F,\theta,k} \prod_{i=1}^{n-k}\left(\frac{\bar{G}}{\underbar{G}}\right)^{\alpha(n-i)}\\
-    &\leq w_{n,j}^{F,\theta,k} \left(\frac{\bar{G}}{\underbar{G}}\right)^{\alpha \sum_{i=1}^{n-k}(n-i)} \\
-    &\leq w_{n,j}^{F,\theta,k} \left(\frac{\bar{G}}{\underbar{G}}\right)^{\alpha\frac{n(n-1)-k(k-1)}{2}}
-\end{align*}
+The proof strategy is as follows.
+Instead of analyzing the variance of the MOP-$\alpha$ gradient estimate proper, we will analyze the variance of a modified estimator where we truncate the weights at $k$ timesteps, so this estimator only looks at most $k$ timesteps back while accumulating the importance weights.
+This enables us to establish strong mixing for the truncated estimator, in order to get a bound on the variance that is $O(NJ^{-1})$.
+We emphasize that this estimator is only a theoretical construct -- we do not actually use this estimator.
+
+Coupled with a bound ensuring that the $k$-truncated estimator provides an estimate that is close to that of MOP-$\alpha$ proper, we get a bound on the variance of MOP-$\alpha$ that comprises of the variance of the $k$-truncated estimator plus the error, for any $k \leq N$.
+It then holds that the final variance bound on MOP-$\alpha$ is given by the minimum over all $k$, and is never larger than $O(1+NJ^{-1})$.
+
+Using the log-derivative trick and that $w_{n,j}^{F,\theta} = w_{n,j}^{F,\theta,k} = 1$ when $\theta=\phi$,
+\begin{align*}
+    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,k}\right)\right\rVert_{\infty}\\
+    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta}}{{\sum_{j=1}^J w_{n,j}^{F,\theta}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,k}}}\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,k}\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta}}{w_{n,j}^{F,\theta}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,k}}{w_{n,j}^{F,\theta,k}}\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,k}\right)\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta}\right)-\log\left(w_{n,j}^{F,\theta,k}\right)\right)\right\rVert_{\infty}\\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=n-k}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^{n-k}\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}G'(\theta)\\
+    &\leq G'(\theta)\frac{\alpha^k-\alpha^n}{1-\alpha},
+\end{align*}
+where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
 
@@
 \begin{align*}
-    &\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \\
-    &= \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) \\
-    &= \sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\\
-    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\Bigg) \\
-    &= \sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)
-    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\Bigg)
-\end{align*}
-
-We first bound $||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||$:
-\begin{align*}
-    &\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \\
-    &= \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) \\
-    &= \sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\\
-    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\Bigg) \\
-    &=  \sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=n-k}^{n}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right) + \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}\right)
-\end{align*}
-
-By two applications of the log-derivative rule,
-\begin{align*}
-    \nabla_\theta \log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)
-    &= \frac{\sum_{j=1}^J\nabla_\theta \prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}} \\
-    &= \frac{\sum_{j=1}^J \left(\left(\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right) \nabla_\theta \log\left(\prod_{i=1}^n \left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}\right)\right)}{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}} \\
-    &= \frac{\sum_{j=1}^J \left(\left(\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)} \right)\left(\sum_{i=1}^n \alpha(n-i)\nabla_\theta \log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right)}{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha(n-i)}}
-\end{align*}
+    &\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_\infty\\
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
+    &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\\
+    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)
+    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\Bigg)\right\rVert_\infty \\
+    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)\right)\right\lVert_{\infty}
+    +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\right)\right\rVert_\infty,
+\end{align*}
+and each of these two terms is bounded by $\frac{\alpha^k}{1-\alpha}NG'(\theta)$. So we now know that
+$$\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_\infty \leq  \frac{2\alpha^k}{1-\alpha}NG'(\theta),$$
+which is our desired error bound.
+
+Now, we bound the variance of $\hat s_k(\theta)$. 
+Recall that we defined
+$$\hat s_k(\theta) = \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) = \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right).$$
+
+We can then decompose this expression with the log-derivative trick while noting that $w_{n,j}^{F,\theta, k} = 1$ whenever $\theta=\phi$, to see that
+\begin{align*}
+    \hat s_k(\theta) 
+    &= \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right) \\
+    &= \sum_{n=1}^N \left(\frac{\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}}{\sum_{j=1}^J w_{n,j}^{F,\theta, k}}-\frac{\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}}{\sum_{j=1}^J w_{n-1,j}^{A,F,\theta, k}}\right) \\
+    &= \sum_{n=1}^N \left(\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}\right)\\
+    &= \frac{1}{J}\sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right).
+\end{align*}
+
+It now follows that we need to bound the variance at a single timestep, namely
+$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right).$$
+
+We use the log-derivative trick yet again to find, noting that $w_{n,j}^{F,\theta, k}=1$ when $\theta=\phi$, that
+\begin{align*}
+    \nabla_\theta w_{n,j}^{F,\theta, k} = \frac{\nabla_\theta w_{n,j}^{F,\theta, k}}{w_{n,j}^{F,\theta, k}} = \nabla_\theta \log(w_{n,j}^{F,\theta, k}) = \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) = \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right).
+\end{align*}
+
+Then,
+\begin{align*}
+    \text{Var}\left(\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right) 
+    &= \text{Var}\left(\nabla_\theta \left(\sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right)- \sum_{i=n-k-1}^{n-1} \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right)\right)\right) \\
+    &= \text{Var} \left(\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1}\log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right).
+\end{align*}
+
+Note that $$\hat s_{n,k}(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)$$ is a function bounded by $CkG'(\theta)$ in each coordinate for some constant $C$ (by Assumption \ref{assump:bounded-measurement}) that depends only on $k$ timesteps, and not $n$.
+We can therefore invoke Theorem 2 of \cite{delMoral03} to bound the $L_2$ error of this functional from its expectation under the posterior by $CkG'(\theta)c_pJ^{-1/2}d(k)$ where $d(k)$ is a term linear in $k$ and $c_p$ is a term possibly exponential in the dimension of the state space $p$.
+That is, from Theorem 2 of \cite{delMoral03} we have that
+$$\E\left[||\hat s_{n,k}(\theta) - \E_{\pi}\hat s_{n,k}(\theta)||_\infty^2\right]^{1/2} \leq \frac{CkG'(\theta)c_pd(k)}{\sqrt{J}}.$$
+
+By the bias-variance decomposition, this in turn bounds the variance at a single timestep by $C^2k^2G'(\theta)^2c_p^2J^{-1}d(k)^2$. Concretely, we have that
+$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \lesssim \frac{c_p^2k^4G'(\theta)^2}{J}.$$
+
+It now remains to bound the variance of $\hat s_k(\theta)$ by considering the covariance of each of the $N$ terms that comprise it. We decompose
+\begin{align*}
+    \text{Var}(\nabla_\theta \hat s_k(\theta)) &= \text{Var}\left(\sum_{n=1}^N \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \\
+    &= \sum_{n=1}^N\text{Var}\left(\hat s_{n,k}(\theta)\right)+ 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \\
+    &\lesssim \frac{c_p^2k^4G'(\theta)^2N}{J} + 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right).
+\end{align*}
+
+Similarly to the proof of the MOP-$0$ case, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
+We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
+$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
+i.e.
+the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
+By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strongly mixing, with $\alpha$-mixing coefficients $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c\log(J))}}$.
+Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
+
+We now derive the $\alpha$-mixing coefficients of $(\hat s_{n,k}(\theta))_{n=1}^N$.
+Observe that $\hat s_{n,k}$ is strong mixing at lag $k+1$, as all weights beyond $k$ timesteps are truncated.
+We therefore have that the mixing time of $\hat s_k(\theta)$ is $O(1+k+\log(J))$, and that the $\alpha$-mixing coefficients for $(\hat s_{n,k}(\theta))_{n=1}^N$ are given by $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c(1+k+\log(J)))}}$.
+
+Therefore, by Davydov's inequality and Theorem 2 of \cite{delMoral03}, and noting that $\hat s_{n,k}(\theta) \leq CkG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, by a similar argument to the MOP-0 case, we find that
+
+$$\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{c_p^2k^2G'(\theta)^2}{J}.$$
+
+Concretely, noting that $\nabla_\theta\hat s_{n,k}(\theta)\leq k^2G'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and WLOG assuming $\E[(\nabla_\theta\hat\hat s_{m,k}(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat\hat s_{n,k}(\theta))^4]^{1/4}$, we apply Davydov's inequality to see that
+\begin{align*}
+    \text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) 
+    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat s_{m,k}(\theta))^4]^{1/4}\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/4}\\
+    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2}.
+\end{align*}
+
+To bound this, we use the fact that $$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4] = \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4]+\E[(\nabla_\theta\hat s_{n,k}(\theta))^2]^2$$
+alongside Theorem 2 of \cite{delMoral03}, which shows that
+$$\E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4] \lesssim \frac{c_p^4k^4G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^2] \lesssim \frac{c_p^2k^2G'(\theta)^2}{J},$$
+where $c_p$ is a term depending only on the dimension of the state space $p$ (but where the dependence may be exponential), to find that 
+$$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{c_p^4k^4G'(\theta)^4}{J^2}+\left(\frac{c_p^2k^2G'(\theta)^2}{J}\right)^2} = \frac{c_p^2k^2G'(\theta)^2}{J},$$
+and conclude that 
+$$\text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{c_p^2k^2G'(\theta)^2}{J}.$$
+
+Putting it all together, we see that
+\begin{align*}
+    \text{Var}(\nabla_\theta \hat s_{k}(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat s_{n,k}(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat s_{m,k}(\theta),\nabla_\theta\hat s_{n,k}(\theta))\\
+    &\leq \frac{Nc_p^2k^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\frac{c_p^2k^2G'(\theta)^2}{J} \\
+    &\leq \frac{c_p^2k^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\right) \\
+    &\lesssim \frac{c_p^2k^2G'(\theta)^2N}{J}.
+\end{align*}
+
+We will now use this result to bound $\text{Var}(\nabla_\theta \hat s_k(\theta)$ proper.
+For random variables $X,Y$ where $|X-Y|$ is bounded almost surely by some $M$, we have that \kevin{I proved this on paper in the office, need to write the proof down in a later section.}
+$$\text{Var}(X) \leq \text{Var}(Y) + M^2 + 4M\E[Y] \leq \text{Var}(Y) + 5M^2.$$
+
+It follows from this result and the result we proved earlier, namely that
+$$||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||_\infty \leq \frac{2\alpha^k}{1-\alpha}NG'(\theta),$$
+that the variance of MOP-$\alpha$ proper is bounded by, for any $k \leq N$,
+that the variance of MOP-$\alpha$ proper is bounded by
+$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \frac{c_p^2k^2G'(\theta)^2N}{J} + \frac{4\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2.$$
+
+As the above holds for any $k \leq N$,
+$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \min_{k\leq N} \left(\frac{c_p^2k^2G'(\theta)^2N}{J} + \frac{4\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right).$$
+
+To interpret this result, we can choose $k$ such that
+\begin{align*}
+    &\frac{4\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \leq 1,\\
+    &\alpha^{2k} \leq \frac{(1-\alpha)^2}{4N^2G'(\theta)^2},\\
+    &k \geq \frac{\log\left(4N^2G'(\theta)^2\right)-2\log(1-\alpha)}{2\log\left(\frac{1}{\alpha}\right)}.
+\end{align*}
+
+To interpret this result, if there exists some $k$ such that $$N \geq k \geq \frac{\log\left(2NG'(\theta)\right)-\log(1-\alpha)}{\log\left(\frac{1}{\alpha}\right)},$$
+we can choose such a $k$, as well as $k=N$, to see that
+\begin{align*}
+    \text{Var}(\nabla_\theta \hat\ell(\theta)) 
+    &\lesssim \min_k\left(\frac{c_p^2k^2G'(\theta)^2N}{J} + \frac{4\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right)\\
+    &\lesssim \min\left\{1+\frac{c_p^2G'(\theta)^2N}{J}, \frac{c_p^2G'(\theta)^2N^3}{J}\right\} \\
+    &\leq \min\left\{1+\frac{c_p^2G'(\theta)^2N}{J}, \frac{c_p^2G(\theta)^2N}{J}\right\},
+\end{align*}
+as $NG'(\theta)\leq G(\theta)$ as a consequence of Assumption \ref{assump:local-bounded-derivative}.
+
+\subsection{Analysis of MOP-$\alpha$ Bias}
+
+Using the log-derivative trick and that $w_{n,j}^{F,\theta,1} = w_{n,j}^{F,\theta,\alpha} = 1$ when $\theta=\phi$,
+\begin{align*}
+    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,1}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)\right\rVert_{\infty}\\
+    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,1}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,1}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}}}\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,1}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,\alpha}\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,1}}{w_{n,j}^{F,\theta,1}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,\alpha}}{w_{n,j}^{F,\theta,\alpha}}\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,1}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha}\right)\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta,1}\right)-\log\left(w_{n,j}^{F,\theta,\alpha}\right)\right)\right\rVert_{\infty}\\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=1}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)G'(\theta)\\
+    &\leq G'(\theta)\left(\frac{n-\alpha n + \alpha^n - 1}{1-\alpha}\right),
+\end{align*}
+where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
+
+By a similar decomposition to that of the error of the $k$-truncated gradient to the MOP-$\alpha$ gradient estimate,
+\begin{align*}
+    &\left\lVert\nabla_\theta\hat\ell^1(\theta) - \nabla_\theta\hat\ell^\alpha(\theta) \right\rVert_\infty\\
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
+    &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\\
+    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)
+    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right)\Bigg)\right\rVert_\infty \\
+    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)\right|
+    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right)\right|\Bigg)\right\rVert_\infty.
+\end{align*}
+
+From our arguments above, we know that 
+$$\left\lVert\nabla_\theta\hat\ell^1(\theta) - \nabla_\theta\hat\ell^\alpha(\theta) \right\rVert_\infty \leq 2NG'(\theta)\left(\frac{N-\alpha N + \alpha^N - 1}{1-\alpha}\right).$$
+
+So the bias of MOP-$\alpha$ is the bias of MOP-$1$ arising from the logarithm in the gradient of the log-likelihood plus no more than $$2NG'(\theta)\left(\frac{N-\alpha N + \alpha^N - 1}{1-\alpha}\right),$$
+where we note that the bias vanishes as $\alpha$ goes to $1$ as
+$$\lim_{\alpha\to1}\left(\frac{N-\alpha N + \alpha^N - 1}{1-\alpha}\right) = 0.$$
 
@@
 \paragraph{Remark:} According to \cite{delmoral2011ci}, under some regularity conditions, $r_N$ and $\beta_N$ are linear in the trajectory length $N$.
-This corresponds to the finding by \cite{poyiadjis11} that the variance of the estimate is at least quadratic in the trajectory length, and their remark that the result of \cite{delMoral03} establishes that the $L_p$ error is bounded by $O(N^2J^{-1/2})$ after accounting for the sum over timesteps.
+This corresponds to the finding by \cite{poyiadjis11} that the variance of the estimate is at least quadratic in the trajectory length, and their remark that the result of \cite{delMoral03} establishes that the $L_p$ error is bounded by $O(N^2J^{-1/2})$ (equivalently, the variance is bounded by $O(N^4J^{-1})$) after accounting for the sum over timesteps.
+The MOP-$1$ variance upper bound is therefore in fact $O(N^4)$, in contrast to the MOP-$\alpha$, where $\alpha<1$, upper bound of $O(N)$.
+
 Either way, usually in practice one considers the estimate normalized through dividing it by $N$, in which case we only need
@@
 
+\section{Miscellanous Technical Lemmas}
+
+\begin{lem}
+    \label{lem:dobrushin-implies-alpha-mixing}
+    Setting $X$ to be the particle collection at time $m$, and $Y$ at time $n$, we have that the alpha mixing coefficients,
+    $$\int | f_{XY} - f_X f_Y | < \alpha,$$
+    are bounded by the Dobrushin coefficient
+    $$\int | f_{Y|X}(y|x_1) - f_{Y|X}(y|x_2) |dy < \alpha$$ for all $x_1$, $x_2$.
+\end{lem}
+
+\begin{proof}
+    We rewrite the alpha-mixing assertion as $$\int { \int |f_Y|X(y|x) - f_Y(y)| dy } f_X dx.$$
+
+    We claim that the Dobrushin coefficient implies $$\int | f_Y|X(y|x_1) - f_Y(y) |dy < \alpha$$ for all $x_1$. This is shown as follows:
+    \begin{align*}
+        \int | f_{Y|X}(y|x) - f_Y(y) |dy 
+        &= \int \left|  \int [f_{Y|X}(y|x_1) - f_{Y|X}(y|x)] f_X(x)dx \right|dy \\
+        &< \int   \int  | f_{Y|X}(y|x_1) - f_{Y|X}(y|x)] | dy f_X(x)dx \\
+        &< \int \alpha f_X(x)dx \\
+        &=\alpha
+    \end{align*}
+
+    We then have the desired result.
+\end{proof}
+
 \end{document}

### 2024-04-06 mop/draft.tex
Updates from Overleaf

@@
 Consider an unobserved Markov process $\{X_t, t \geq t_0\}$, and observations $Y_1,...,Y_N$ at timesteps $t_1,..., t_N$.
-The process is parameterized by an unknown parameter $\theta \in \Theta$, where $X \in \gX, Y \in \gY,$ and finally $\gX, \gY, \Theta \subseteq \R^n$.
+The process is parameterized by an unknown parameter $\theta \in \Theta$, where $X \in \gX, Y \in \gY,$ and finally $\gX \subseteq \R^d, \Theta \subseteq \R^p$.
 By a similar decomposition to that in \citet{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
@@
 alongside Theorem 2 of \cite{delMoral03}, which shows that
-$$\E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4] \lesssim \frac{c_p^4G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^2] \lesssim \frac{c_p^2G'(\theta)^2}{J},$$
-where $c_p$ is a term depending only on the dimension of the state space $p$ (but where the dependence may be exponential), to find that 
-$$\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{c_p^4G'(\theta)^4}{J^2}+\left(\frac{c_p^2G'(\theta)^2}{J}\right)^2} = \frac{c_p^2G'(\theta)^2}{J},$$
+$$\E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4] \lesssim \frac{c_d^4G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^2] \lesssim \frac{c_d^2G'(\theta)^2}{J},$$
+where $c_d$ is a term depending only on the dimension of the state space $d$ (but where the dependence may be exponential), to find that 
+$$\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{c_d^4G'(\theta)^4}{J^2}+\left(\frac{c_d^2G'(\theta)^2}{J}\right)^2} = \frac{c_d^2G'(\theta)^2}{J},$$
 and conclude that 
-$$\text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{c_p^2G'(\theta)^2}{J}.$$
+$$\text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{c_d^2G'(\theta)^2}{J}.$$
 
@@
     \text{Var}(\nabla_\theta \hat\ell(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta))\\
-    &\leq \frac{Nc_p^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\frac{c_p^2G'(\theta)^2}{J} \\
-    &\leq \frac{c_p^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\right) \\
-    &\lesssim \frac{c_p^2G'(\theta)^2N}{J}.
+    &\leq \frac{Nc_d^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\frac{c_d^2G'(\theta)^2}{J} \\
+    &\leq \frac{c_d^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\right) \\
+    &\lesssim \frac{c_d^2G'(\theta)^2N}{J}.
 \end{align*}
@@
 
-Note that $$\hat s_{n,k}(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)$$ is a function bounded by $CkG'(\theta)$ in each coordinate for some constant $C$ (by Assumption \ref{assump:bounded-measurement}) that depends only on $k$ timesteps, and not $n$.
-We can therefore invoke Theorem 2 of \cite{delMoral03} to bound the $L_2$ error of this functional from its expectation under the posterior by $CkG'(\theta)c_pJ^{-1/2}d(k)$ where $d(k)$ is a term linear in $k$ and $c_p$ is a term possibly exponential in the dimension of the state space $p$.
+Note that $$\hat s_{n,k}(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)$$ is a function bounded by $C\frac{G'(\theta)(1-\alpha^k)}{1-\alpha} =: CG'(\theta)r$ in each coordinate for some constant $C$ (by Assumption \ref{assump:bounded-measurement}) that depends only on $k$ timesteps, and not $n$.
+We can therefore invoke Theorem 2 of \cite{delMoral03} to bound the $L_2$ error of this functional from its expectation under the posterior by $CrG'(\theta)c_dJ^{-1/2}d(k)$ where $d(k)$ is a term linear in $k$ and $c_d$ is a term possibly exponential in the dimension of the state space $d$.
 That is, from Theorem 2 of \cite{delMoral03} we have that
-$$\E\left[||\hat s_{n,k}(\theta) - \E_{\pi}\hat s_{n,k}(\theta)||_\infty^2\right]^{1/2} \leq \frac{CkG'(\theta)c_pd(k)}{\sqrt{J}}.$$
-
-By the bias-variance decomposition, this in turn bounds the variance at a single timestep by $C^2k^2G'(\theta)^2c_p^2J^{-1}d(k)^2$. Concretely, we have that
-$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \lesssim \frac{c_p^2k^4G'(\theta)^2}{J}.$$
+$$\E\left[||\hat s_{n,k}(\theta) - \E_{\pi}\hat s_{n,k}(\theta)||_\infty^2\right]^{1/2} \leq \frac{Cr^2G'(\theta)c_dd(k)}{\sqrt{J}}.$$
+
+By the bias-variance decomposition, this in turn bounds the variance at a single timestep by $C^2r^2G'(\theta)^2c_d^2J^{-1}d(k)^2$. Concretely, we have that
+$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \lesssim \frac{c_d^2r^2k^2G'(\theta)^2}{J}.$$
 
@@
     &= \sum_{n=1}^N\text{Var}\left(\hat s_{n,k}(\theta)\right)+ 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \\
-    &\lesssim \frac{c_p^2k^4G'(\theta)^2N}{J} + 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right).
+    &\lesssim \frac{c_d^2r^2k^2G'(\theta)^2N}{J} + 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right).
 \end{align*}
@@
 
-Therefore, by Davydov's inequality and Theorem 2 of \cite{delMoral03}, and noting that $\hat s_{n,k}(\theta) \leq CkG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, by a similar argument to the MOP-0 case, we find that
-
-$$\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{c_p^2k^2G'(\theta)^2}{J}.$$
-
-Concretely, noting that $\nabla_\theta\hat s_{n,k}(\theta)\leq k^2G'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and WLOG assuming $\E[(\nabla_\theta\hat\hat s_{m,k}(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat\hat s_{n,k}(\theta))^4]^{1/4}$, we apply Davydov's inequality to see that
+Therefore, by Davydov's inequality and Theorem 2 of \cite{delMoral03}, and noting that $\hat s_{n,k}(\theta) \leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, by a similar argument to the MOP-0 case, we find that
+
+$$\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{c_d^2r^2G'(\theta)^2}{J}.$$
+
+Concretely, noting that $\nabla_\theta\hat s_{n,k}(\theta)\leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and WLOG assuming $\E[(\nabla_\theta\hat\hat s_{m,k}(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat\hat s_{n,k}(\theta))^4]^{1/4}$, we apply Davydov's inequality to see that
 \begin{align*}
@@
 alongside Theorem 2 of \cite{delMoral03}, which shows that
-$$\E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4] \lesssim \frac{c_p^4k^4G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^2] \lesssim \frac{c_p^2k^2G'(\theta)^2}{J},$$
-where $c_p$ is a term depending only on the dimension of the state space $p$ (but where the dependence may be exponential), to find that 
-$$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{c_p^4k^4G'(\theta)^4}{J^2}+\left(\frac{c_p^2k^2G'(\theta)^2}{J}\right)^2} = \frac{c_p^2k^2G'(\theta)^2}{J},$$
+$$\E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4] \lesssim \frac{c_d^4r^4G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^2] \lesssim \frac{c_d^2r^2G'(\theta)^2}{J},$$
+where $c_d$ is a term depending only on the dimension of the state space $d$ (but where the dependence may be exponential), to find that 
+$$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{c_d^4r^4G'(\theta)^4}{J^2}+\left(\frac{c_d^2r^2G'(\theta)^2}{J}\right)^2} = \frac{c_d^2r^2G'(\theta)^2}{J},$$
 and conclude that 
-$$\text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{c_p^2k^2G'(\theta)^2}{J}.$$
+$$\text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{c_d^2r^2G'(\theta)^2}{J}.$$
 
@@
     \text{Var}(\nabla_\theta \hat s_{k}(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat s_{n,k}(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat s_{m,k}(\theta),\nabla_\theta\hat s_{n,k}(\theta))\\
-    &\leq \frac{Nc_p^2k^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\frac{c_p^2k^2G'(\theta)^2}{J} \\
-    &\leq \frac{c_p^2k^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\right) \\
-    &\lesssim \frac{c_p^2k^2G'(\theta)^2N}{J}.
-\end{align*}
-
-We will now use this result to bound $\text{Var}(\nabla_\theta \hat s_k(\theta)$ proper.
-For random variables $X,Y$ where $|X-Y|$ is bounded almost surely by some $M$, we have that \kevin{I proved this on paper in the office, need to write the proof down in a later section.}
-$$\text{Var}(X) \leq \text{Var}(Y) + M^2 + 4M\E[Y] \leq \text{Var}(Y) + 5M^2.$$
+    &\leq \frac{Nc_d^2r^2k^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\frac{c_d^2r^2G'(\theta)^2}{J} \\
+    &\leq \frac{c_d^2r^2k^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\right) \\
+    &\lesssim \frac{c_d^2k^2r^2G'(\theta)^2N}{J}.
+\end{align*}
+
+We will now use this result to bound $\text{Var}(\nabla_\theta \hat s_k(\theta)$ proper. For random variables $X,Y$ where $|X-Y|$ is bounded almost surely by some $M$, we have that 
+$$X-\E X - 2M \leq Y - \E Y \leq X - \E X + 2M$$
+so,
+$$(Y-\E Y)^2 \leq (X-\E X - 2M)^2 + (X-\E X +2M)^2.$$
+Taking expectations, we get
+$$\text{Var}(X) \leq \text{Var}(Y) \leq \text{Var}(Y) + 8M^2.$$
 
@@
 that the variance of MOP-$\alpha$ proper is bounded by
-$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \frac{c_p^2k^2G'(\theta)^2N}{J} + \frac{4\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2.$$
+$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \frac{c_d^2k^2r^2G'(\theta)^2N}{J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2.$$
 
 As the above holds for any $k \leq N$,
-$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \min_{k\leq N} \left(\frac{c_p^2k^2G'(\theta)^2N}{J} + \frac{4\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right).$$
-
-To interpret this result, we can choose $k$ such that
-\begin{align*}
-    &\frac{4\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \leq 1,\\
-    &\alpha^{2k} \leq \frac{(1-\alpha)^2}{4N^2G'(\theta)^2},\\
-    &k \geq \frac{\log\left(4N^2G'(\theta)^2\right)-2\log(1-\alpha)}{2\log\left(\frac{1}{\alpha}\right)}.
-\end{align*}
-
-To interpret this result, if there exists some $k$ such that $$N \geq k \geq \frac{\log\left(2NG'(\theta)\right)-\log(1-\alpha)}{\log\left(\frac{1}{\alpha}\right)},$$
-we can choose such a $k$, as well as $k=N$, to see that
-\begin{align*}
-    \text{Var}(\nabla_\theta \hat\ell(\theta)) 
-    &\lesssim \min_k\left(\frac{c_p^2k^2G'(\theta)^2N}{J} + \frac{4\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right)\\
-    &\lesssim \min\left\{1+\frac{c_p^2G'(\theta)^2N}{J}, \frac{c_p^2G'(\theta)^2N^3}{J}\right\} \\
-    &\leq \min\left\{1+\frac{c_p^2G'(\theta)^2N}{J}, \frac{c_p^2G(\theta)^2N}{J}\right\},
-\end{align*}
-as $NG'(\theta)\leq G(\theta)$ as a consequence of Assumption \ref{assump:local-bounded-derivative}.
+$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \min_{k\leq N} \left(\frac{c_d^2k^2G'(\theta)^2N}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right).$$
+
+% To interpret this result, we can choose $k$ such that
+% \begin{align*}
+%     &\frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \leq 1,\\
+%     &\alpha^{2k} \leq \frac{(1-\alpha)^2}{16N^2G'(\theta)^2},\\
+%     &k \geq \frac{\log\left(16N^2G'(\theta)^2\right)-2\log(1-\alpha)}{2\log\left(\frac{1}{\alpha}\right)}.
+% \end{align*}
+
+% To interpret this result, if there exists some $k$ such that $$N \geq k \geq \frac{\log\left(16NG'(\theta)\right)-\log(1-\alpha)}{\log\left(\frac{1}{\alpha}\right)},$$
+% we can choose such a $k$, as well as $k=N$, to see that
+% \begin{align*}
+%     \text{Var}(\nabla_\theta \hat\ell(\theta)) 
+%     &\lesssim \min_{k\leq N}\left(\frac{c_d^2k^2r^2G'(\theta)^2N}{J} + \frac{\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right)\\
+%     &\lesssim \min\left\{1+\frac{c_d^2G'(\theta)^2N}{J}, \frac{c_d^2G'(\theta)^2r^2N^3}{J}\right\} \\
+%     &\leq \min\left\{1+\frac{c_d^2G'(\theta)^2N}{J}, \frac{c_d^2G(\theta)^2r^2N}{J}\right\},
+% \end{align*}
+% as $NG'(\theta)\leq G(\theta)$ as a consequence of Assumption \ref{assump:local-bounded-derivative}.
 
@@
     &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=1}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}\\
-    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
-    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)G'(\theta)\\
-    &\leq G'(\theta)\left(\frac{n-\alpha n + \alpha^n - 1}{1-\alpha}\right),
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}.
+\end{align*}
+
+Now we establish the bound:
+\begin{align*}
+     \nabla_\theta \left(\sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)
+     &\leq \sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
+    &\leq \sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)G'(\theta)\\
+    &\leq G'(\theta)\left(\frac{n-\alpha n + \alpha^n - 1}{1-\alpha}\right)
 \end{align*}
@@
 
-It is well-known that bounds like these hide an exponential dependence in the dimension of the state space $p$.
-We take care to keep track of the terms depending on $p$ when they arise in front of us, but are unable to do so for the terms that arise from the analysis of \cite{delmoral2011ci}.
+It is well-known that bounds like these hide an exponential dependence in the dimension of the state space $d$.
+We take care to keep track of the terms depending on $d$ when they arise in front of us, but are unable to do so for the terms that arise from the analysis of \cite{delmoral2011ci}.
 

### 2024-04-11 paper/ms.tex
Updates from Overleaf

--- 2023-11-02 paper/ms.tex
@@
 \input{macros}
+
+\usepackage{tabto}
 
@@
 
-\title{Inference for Dynamic and Latent Variable Models via Plug-and-Play Automatically Differentiable Particle Filtering}
+\title{Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation}
 
@@
 % Please include corresponding author, author contribution and author declaration information
-\authorcontributions{Please provide details of author contributions here.}
-\authordeclaration{Please declare any competing interests here.}
-%\equalauthors{\textsuperscript{1}A.O.(Author One) contributed equally to this work with A.T. (Author Two) (remove if not applicable).}
+\authorcontributions{K.T. and E.L.I. designed research; K.T. analyzed data; K.T., G.H. and E.L.I. performed research, K.T., G.H., and E.L.I. wrote the manuscript.}
+\authordeclaration{The authors declare no competing interests.}
 \correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: ionides@umich.edu}
@@
 \begin{abstract}
-  Automatic differentiation has driven recent advances in machine learing, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods.
-  This progress has required simultaneous advances in algorithms, software and hardware.  
-  Partially observed nonlinear stochastic dynamic systems have proved resistant to these techniques; despite various attempts, widely applicable methods have not yet emerged.
-  We present a new approach which is applicable to a general class of models, possesses a theoretical foundation, and is demonstrated to beat current state-of-the-art methods on a challenging scientific benchmark problem.
-  Our algorithm is compatible with parallel computation on a graphical processing unit, and enjoys the plug-and-play property that its software implementation requires only a simulator for the scientific dynamic model as input (in addition to data, and a measurement model).
-
+  Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods.
+Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques due to (1) the requirement for simulation-based inference that does not require access to the system's transition probabilities and (2) the issue that widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters.
+We propose a new representation that embeds existing methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff.
+Further, we develop optimization algorithms suited to the Monte Carlo properties of the derivative estimate, including a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
+Promising numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from iterated filtering can beat current state-of-the-art methods on a challenging scientific benchmark problem.
 \end{abstract}
@@
 
-\firstpage{7}
+\firstpage{4}
 % Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting.
@@
 If this is the case, add \parshape=0 to the end of the list environment.
-\dropcap{M}any approaches to inference in highly nonlinear stochastic dynamical systems assume access to the probability density of next states given the current state.
+\dropcap{M}any scientific models involve highly nonlinear stochastic dynamic systems, where there may be significant uncertainty in both the process dynamics and the measurements of the system.
+Despite the ubiquity of such partially observed stochastic processes with the Markov property (also known as, continuous-state continuous-time hidden Markov models (HMMs), partially-observed Markov processes (POMPs), or partially-observed stochastic nonlinear dynamical systems) in fields as diverse as epidemiology \cite{he10, stocks17}, ecology \cite{knape12}, and finance \cite{kim08, breto14}, performing estimation and inference within this class of models is a difficult problem that faces several challenges.
 
-This is a problem in some critical applications, such as disease modeling, where the models are complex enough that obtaining the density is intractable.
-However, the particle filter, a popular method for solving the filtering problem in partially-observed dynamical systems, does not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm in a feature known as the plug-and-play property.
+Many approaches to inference in these scenarios (such as the EM algorithm, the various Kalman filter variants, and MCMC) either struggle with intractable likelihood functions when the models are complex, or assume access to the probability density of next states given the current state.
+This is a problem in some critical applications, such as disease modeling, where the models are complex enough such that even obtaining the transition density is an intractable problem.
+However, the particle filter with the bootstrap proposal, a popular method for solving the filtering problem in partially-observed dynamical systems, provides an unbiased estimate of the likelihood \cite{delmoral2004feynman} without requiring evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
 
+A series of related algorithms \cite{welch2009abc, wood2010sl, doucet2010pmcmc, ionides08, ionides15} enable parameter estimation within these models.
+In particular, the improved iterated filtering algorithm (IF2) of Ionides et al.
+\cite{ionides15} and the particle MCMC algorithm of Andrieu et al.
+\cite{doucet2010pmcmc} are the only simulation-based full-information maximum likelihood and fully Bayesian methods available in the literature thus far.
+However, accurate parameter estimation in the presence of significant Monte-Carlo noise in the likelihood estimate remains a significant challenge.
 Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small.
-Existing methods like the improved iterated filtering algorithm of Ionides et.
-al.
-\cite{ionides15} converge quickly to a neighborhood of the MLE, but struggle to optimize the last few units of log-likelihood.
+For example, though in practice the algorithm of Ionides et al.
+\cite{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood.
+Ionides et al.
+\cite{Ionides_mcap} attempt to address the third problem, albeit with a computationally-expensive approach.
 
-\kevin{TODO: (1) Fix beta smoothness to d(n) at no more than O(n), (2) add story to say that no one spotted that it could have been plug-and-play because of e.g. reasons why they would stumble, (3) }
+\subsection{Automatic Differentiation for Particle Filters}
 
-\kevin{Previous literature: Doucet waymo only applicable to policy plus deterministic dynamics}
+Recent advances in automatic differentiation (AD) for particle filters \cite{blei2018vsmc, jon2018diffpf, corenflos21, scibior2021dpf, doucet2022particlebased} have drawn attention to AD as an alternative tool for inference in partially observed stochastic processes with the particle filter.
+However, these existing approaches either are not yet compatible with the desire for simulation-based inference, or are computationally expensive.
+In particular, existing approaches are either asymptotically biased \cite{blei2018vsmc, jon2018diffpf}, have high variance \cite{poyiadjis11, scibior21}, computationally expensive \cite{corenflos21, chen24}, or require transition densities \cite{poyiadjis11, scibior21, doucet2022particlebased, chen24}.
 
-We propose a hybrid algorithm that u
+In particular, the work of Scibior and Wood \cite{scibior2021dpf} demonstrates that the estimators derived in Poyiadjis et al.
+\cite{poyiadjis11} can be attained with AD, albeit with a simple tweak, that they call the stop-gradient trick, to the particle filter that does not modify its likelihood estimates.
+Though their method is not fully compatible with simulation-based inference and the mathematical motivation behind the stop-gradient trick is not apparent, we show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.
 
-Unlike IF2, we explicitly characterize our method's rate of convergence
+\subsection{Our Contributions}
+We develop a new approach to statistical inference via AD for particle filters which addresses these concerns.
+We construct a new theoretical framework that addresses the seemingly incompatible paradox of differentiating through a Monte-Carlo algorithm with discontinuous resampling.
 
-% IF2 is SGD that iterates through the data sequentially in the 1:N sense, and noisy/mini-batch GD in the 1:J sense. 
-% ADPF is batch GD in the 1:N sense, and noisy/mini-batch GD in the 1:J sense.
+The key observation is that one can turn this into a problem of differentiating through a series of smooth measurement density ratios and (through the reparametrization trick) a differentiable simulator.
+This is done through the construction of a suitably reweighted family of particle filters that we call \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$), for some $\alpha \in [0,1]$ that optimizes a bias-variance tradeoff.
+This family of algorithms encompasses the special case of the biased gradient estimator of \cite{blei2018vsmc} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$, and does not require access to the transition densities -- only requiring access to a differentiable simulator.
 
-Algorithmic differentiation potentially facilitates numerical optimization, but currently its use for particle filters is limited.
-We investigate ways to use algorithmic differentiation of particle filters within the confines of the plug-and-play property, with the goal of enhancing current inference capabilities for general POMP models.
+We show that these off-policy particle filters successfully target the posterior, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties.
+We also derive a linear convergence rate for gradient methods on the particle filter in the presence of strong convexity, providing support for numerical optimization with AD for the particle filter.
 
-\textbf{The issue:} Particle filters provide convenient approaches to evaluating the log-likelihood function for partially observed Markov process (POMP) models.
-However, using this evaluator to obtain a maximum likelihood parameter estimate can be challenging -- especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small.
-Empirically, methods such as the improved iterated filtering algorithm (IF2) from Ionides et.
-al.
-\cite{ionides15} rapidly converge to a neighborhood of the optimum but struggle at finding the exact optimum due to Monte Carlo variance, even with an annealing random walk standard deviation.
+In practice, we propose a hybrid algorithm that we call \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts gradient ascent (or a similar iterative first or second-order gradient method) with a coarse solution obtained from a few rounds of iterated filtering.
+Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1, the Liu-West filter \cite{liuwest01}, and particle MCMC \cite{doucet2010pmcmc}) on a challenging scientific benchmark problem, the Dhaka cholera model of \cite{king08}.
 
-A potential solution to this could lie in auto-differentiation (AD).
-This would allow for the use of first and second-order iterative optimization techniques.
-However, though AD could potentially facilitate numerical optimization, its use for plug-and-play particle filters has so far been limited.
-This is potentially because particle filtering methods are inherently non-differentiable due to the resampling step that may take place in between iterations.
+\section{Problem Setup}
 
-\subsection*{The importance of the plug-and-play property:} Performing inference in highly nonlinear stochastic dynamical systems is a challenging problem.
-Although many methods for inference assume access to the density of state transitions, this is often not available, especially in critical applications like epidemiology.
+Consider an unobserved Markov process $(X_t)_{t \geq t_0}$, with observations $Y_1,...,Y_N$ realized at values $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$.
+The process is parameterized by an unknown parameter $\theta \in \Theta \subseteq \R^p$, where the states $X_t$ take values in the state space $\gX \subseteq \R^d$, the observations $Y_n$ take values in $\gY,$ and we write $X_n := X_{t_n}$.
+By a similar decomposition to that in \cite{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
+\begin{align}
+    &f_{X_{0: N}, Y_{1: N}}\left(x_{0: N}, y_{1: N} ; \theta\right)\\
+    &=f_{X_0}\left(x_0 ; \theta\right) \prod_{n=1}^N f_{X_n \mid X_{n-1}}\left(x_n \mid x_{n-1} ; \theta\right) f_{Y_n \mid X_n}\left(y_n \mid x_n ; \theta\right).
+\end{align}
 
-Basic particle filtering algorithms do not require evaluation of the transition density of the latent Markov process, in a feature known as the \textbf{plug-and-play property} \cite{breto09} since it enables an arbitrary model simulator to be plugged into the algorithm.
-We investigate ways to use algorithmic differentiation of particle filters within the confines of the plug-and-play property, with the goal of enhancing current inference capabilities for general POMP models.
+We call $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ the process model, writing $\texttt{process}\left(x_n; \theta\right)$ for the simulator corresponding to the above process model, $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ the measurement model, and again write $y_n^*$ for the actual values of the observations that were observed.
+As a matter of terminology, we call $f_{X_{1:N}|Y_{1:N}}(x_{1_N}|y_{1:N}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
 
-\subsection*{Other potential applications:} This has applications beyond the obvious one of learning model parameters via first or second-order optimization routines.
-For example, it could be a step towards developing very general Hamiltonian Monte Carlo methods for particle MCMC, as Rosato et.
-al.
-\cite{rosato22b} do by using previous work such as \cite{scibior21, poyiadjis11} (and we conjecture that the seed-fixing derivatives of Rosato et.
-al.
-are the same as these as an immediate consequence of section \ref{subsec:scibiorrel}) to differentiate the particle filter.
+The above are all defined on a single probability space $(\Omega, \Sigma, \prob),$ where $\omega \in \Omega$ denotes an element of the sample space that can be thought of as a random seed.
+For notational simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0.$
+In the rest of this paper, we will identify the seed with an element of the sample space.
+We do so because the seed determines the entire sequence of pseudo-random numbers as generated by a computer, to maintain an algorithmic and practical interpretation of our theory.
+When two algorithms share the same seed, this corresponds to the method of common random numbers.
+
+\section{Off-Policy Particle Filters}
+
+As mentioned earlier, instead of directly differentiating through a particle filter, one can instead differentiate through the simulator and a series of measurement density ratios.
+We construct an algorithm called \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) in Algorithm \ref{alg:mop} that targets the posterior and likelihood, and whose gradient yields the estimators of \cite{poyiadjis11, scibior21} and \cite{blei2018vsmc} as special cases.
+
+\subsection{Algorithm Outline} 
+The broad idea is to run the particle filter twice under the same seed.
+We first run the particle filter under $\phi$, and then obtain a coupled set of particles by running a bootstrap filter with the process model at $\theta$ but the resampling indices from the previous run at $\phi$.
+We can then reweight the conditional likelihoods by a correction factor accumulated over time of the ratio of the measurement density under $\theta$ and the measurement density under $\phi$ to account for the ("off-policy") resampling under $\phi$.
+That is, when we write $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ as shorthand for the measurement density and estimate the conditional likelihood under $\phi$ in the usual way with $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$, we can estimate the conditional likelihood under $\theta$ as either
+\begin{equation}
+     \label{eq:mop-conditional-likelihood}
+     L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}} \text{ or } L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
+\end{equation}
+where each particle is recursively weighted according to 
+\begin{equation}
+    \label{eq:weighting-scheme}
+    w_{n,j}^{P,\theta} = (w_{n-1,j}^{F,\theta})^\alpha,  w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}, w^{F,\theta}_{0,j}= 1,
+\end{equation}
+and $k_j$ are the indices from resampling according to the measurement density at $\phi$, $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$.
+The before-resampling conditional likelihood estimate $L_n^{B,\theta,\alpha}$ should be used in general, as it has slightly lower variance than the after-resampling estimate $L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$, but the latter is useful in deriving properties of the MOP-$\alpha$ gradient estimate such as that in Theorem \ref{thm:mop-functional-forms}.
+
+This yields a suitably reweighted particle filter that targets the posterior under $\theta$ and produces a strongly consistent likelihood estimate for the likelihood at $\theta$.
+Taking the gradient with respect to $\theta$ of the estimate for the log-likelihood at $\theta$ then yields an estimate for the score at $\theta$.
+
+\begin{algorithm}[H]
+	\caption{Measurement Off-Policy-$\alpha$}
+    \label{alg:mop}
+	     \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
+\newline
+        \textbf{If $\theta \neq \phi$:}
+            Run a particle filter once with parameters $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$, with seed $\omega$. \newline
+        \textbf{Else:} Set $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$ to be copies of $X_{n,j}^{P,\theta}, X_{n,j}^{P,\theta}$ in the rest of the algorithm. \newline
+		\textbf{Initialize } filter particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$.
+Fix $\omega.$ \newline
+		\textbf{For} $n=1,...,N$: \newline
+            \hspace*{4mm} Accumulate discounted weights, $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$.\newline
+            \hspace*{4mm} Simulate process model,
+            ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$. \newline
+            \hspace*{4mm} Measurement density,
+            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$. \newline
+            \hspace*{4mm} Compute $L_n^{B,\theta,\alpha} ={\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$. \newline
+            \hspace*{4mm} Conditional likelihood under $\phi$,
+            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$.\newline
+            \hspace*{4mm} Select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) \propto g^{\phi}_{n,m}$. \newline
+            \hspace*{4mm} Resample particles ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$. \newline
+            \hspace*{4mm} Resample corrected weights
+            $w_{n,j}^{F,\theta}= w^{P,\theta}_{n,k_j}  { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}$.\newline
+            \hspace*{4mm} Compute $ L_n^{A,\theta,\alpha} = L_n^\phi\cdot {\sum_{j=1}^J w^{F,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$.\newline
+		\textbf{Return:} likelihood estimate $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}_{j=1}^J.$
+\end{algorithm}
+
+\subsection{Discounting Weights for a Bias-Variance Tradeoff}
+
+As mentioned earlier, the correction factors accumulate over time.
+If $\theta$ and $\phi$ coincide, one can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$, as seen in Equation \ref{eq:weighting-scheme}, while still targeting the posterior and likelihood at $\theta$.
+We do this in order to optimize a bias-variance tradeoff for the score estimate yielded by differentiating through the log-likelihood estimate provided by MOP-$\alpha$.
+When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, recovering the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}.
+When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{blei2018vsmc}.
+$\alpha$ therefore controls an exponentially-weighted moving average, determining how much "memory" the resulting gradient estimate has over ancestral trajectories.
+This is illustrated in Figure \ref{fig:biasvar}, in the context of the Dhaka cholera model of \cite{king08}.
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=\textwidth/3]{imgs/095/biasvar.png}
+    \caption{Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}.
+We display the MSE of score estimates for the trend in transmission, evaluated at the MLE.}
+    \label{fig:biasvar}
+\end{figure}
+
+\section{Theoretical Analysis of MOP-$\alpha$}
+
+Here, we will show that MOP-$\alpha$ targets the posterior and likelihood, yields the estimators of \cite{poyiadjis11, scibior21, blei2018vsmc} as special cases when differentiated, and characterize rates for its bias and variance under different choices of $\alpha$.
+To do so, we require the following assumptions:
+
+\begin{enumerate}[label=(A\arabic*),itemsep=-1.2ex] 
+    \item \textbf{Continuity of the Likelihood.} $\ell(\theta)$ has more than two continuous derivatives in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.
+\label{assump:conti-lik}
+    \item \textbf{Bounded Process Model.} There exist $\underbar{M}, \bar{M}$ such that $0 < \underbar{M} \leq f_{X_n|X_{n-1}}(x_n | x_{n-1};\theta) \leq \bar{M} < \infty$.
+\label{assump:bounded-process}
+    \item \textbf{Bounded Measurement Model.} There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G}<\infty$ and $G'(\theta)$ so $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty \leq G'(\theta)< \infty$.
+\label{assump:bounded-measurement}
+    \item \textbf{Bounded Gradient Estimates.} There are functions $G(\theta), H(\theta): \Theta \to [0,\infty)$ uniformly bounded by $G', H'$, so the MOP-$\alpha$ gradient and Hessian estimates at $\theta=\phi$ are almost surely bounded by $G(\theta)$ and $H(\theta)$ for all $\alpha$.
+\label{assump:local-bounded-derivative}
+    \item \textbf{Differentiability of Density Ratios and Simulator.} The measurement density $f_{Y_n|X_n}(y_n^*|x; \theta)$ and simulator have more than two continuous derivatives in $\theta$.
+\label{assump:diff-meas-and-sim}
+\end{enumerate}
+
+The fact that the likelihood estimate yielded by MOP-$\alpha$ has more than two continuous derivatives in $\theta$ follows the construction of the likelihood estimate in equations \ref{eq:mop-conditional-likelihood} and \ref{eq:weighting-scheme}, as well as Assumptions \ref{assump:bounded-measurement} and \ref{assump:diff-meas-and-sim}.
+
+\subsection{MOP-$\alpha$ Targets the Posterior}
+
+We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood.
+To do so, we state the definition of targeting and present three lemmas, whose proof can be found in the supplementary material.
+\begin{defn}[Targeting]
+    \label{defn:targeting}
+    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any measurable and bounded functional $h$,
+    \begin{equation}
+        E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
+    \end{equation}  
+        A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if for any measurable and bounded functional $h$, as $J\to\infty$
+    \begin{equation}
+        \frac{\sum_{j=1}^J h(X^J_j) w^J_j}{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X)).
+    \end{equation}
+\end{defn}
+
+\begin{lem}[Change of Particle Measure]
+    \label{lem:change-measure-proper-weights}
+    Assume $\{(\tilde X_j^J,U_j^J)\}_{j=1}^J$ targets $f_X$, and that $\{(Y_j^J,V_j^J)\}_{j=1}^J$ is a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is drawn w.p.
+$\pi^J_j$.
+Write, where $a$ is the ancestor function,
+    $(Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big)
+    $. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j)\}_{j=1}^J$ targets $f_X$.
+\end{lem}
+\begin{lem}[Particle Marginals]
+    \label{lem:marginal-proper-weights}
+    Suppose that $\{(\tilde X_j^J,U_j^J)\}_{j=1}^J$ targets $f_X$.
+Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$, where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
+Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}_{j=1}^J$ targets $f_Z$.
+\end{lem}
+\begin{lem}[Particle Posteriors]
+    \label{lem:posterior-proper-weights}
+    Suppose that $\{(X_j^J,u_j^J)\}_{j=1}^J$ targets $f_X$, and write $(X^{\prime J}_j,u^{\prime J}_j) = \big(X_j^J,u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$.
+Then, if $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,u^{\prime J}_j)\}_{j=1}^J$ targets $f_{X|Z}(\cdot | z^*)$.
+\end{lem}
+
+From these lemmas, we can derive the following result.
+
+\begin{thm}[MOP-$\alpha$ Targets the Posterior and Likelihood]
+    \label{thm:mop-targeting}
+    When $\alpha=1$ or $\phi=\theta$, MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood.
+That is, for $\pi=f_{X_{1:N}|Y_{1:N} ; \theta}$ and any measurable and bounded functional $h$ and for $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\prod_{n=1}^N L_n^{B,\theta,\alpha}$, it holds that
+    \begin{equation}
+        \frac{\sum_{j=1}^J h(x_{n,j}^{P, \theta}) w_{n,j}^{P,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} \stackrel{a.s.}{\to} E_\pi h, \;\; \hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).
+    \end{equation}
+\end{thm}
+
+\begin{proof}
+    When $\theta=\phi$, regardless of the value of $\alpha$, the ratio ${g_{n,j}^\theta}/{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter estimate.
+
+    When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows.
+Recursively applying Lemmas \ref{lem:change-measure-proper-weights}, \ref{lem:marginal-proper-weights}, and \ref{lem:posterior-proper-weights}, we obtain that
+    %to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
+    the MOP-1 filter targets the posterior.
+    Specifically, suppose inductively that $\big\{\big(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
+    Then, Lemma \ref{lem:marginal-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big)\big\}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+    Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
+    \begin{equation}
+        \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}\big/ g^\phi_{n,a(j)}\big),
+    \end{equation}
+    with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+
+    For the likelihood, if the likelihood is estimated with the before-resampling conditional likelihoods $\hat\lik(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, the strong consistency is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+
+    Whereas if we use the after-resampling conditional likelihoods $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$, the numerator of $L_n^{B,\theta,\alpha}$ is
+    \begin{equation}
+        \sum_{j=1}^J g_{n,j}^\theta w_{n, j}^{P, \theta} = L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
+    \end{equation}
+    Using Lemma \ref{lem:change-measure-proper-weights}, we resample the corrected weights $w^{P,\theta}_{n,j}{g^\theta_{n,j}}/{g^\phi_{n,j}} $ according to probabilities ${g^\phi_{n,j}}/{L_n^\phi}$ to see this is properly estimated by $L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},$
+from which we obtain the estimate $L_n^{B,\theta,\alpha}$. The strong consistency then follows as the likelihood is the product of the conditional likelihoods.
+\end{proof}
+
+\subsection{MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, blei2018vsmc}}
+
+\cite{scibior21} show that the estimate of \cite{blei2018vsmc} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through applying a "stop-gradient trick" to the particle filter.
+It turns out that both of these, when applied on the bootstrap filter, correspond to special cases of MOP-$\alpha$.
+
+\begin{thm}[MOP-$0$ and MOP-$1$ Functional Forms]
+    \label{thm:mop-functional-forms}
+    Write $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$ and we use the after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$.
+Then, when $\alpha=0$,
+    \begin{equation}
+        \nabla_\theta \hat\ell^0(\theta) 
+        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
+    \end{equation}
+    yielding the gradient estimator of \cite{blei2018vsmc} with the bootstrap filter. When $\alpha=1$,
+    \begin{equation}
+        \nabla_\theta \hat{\ell}^1(\theta) 
+        = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
+    \end{equation}
+    yielding the gradient estimator of \cite{poyiadjis11, scibior21} applied to the bootstrap filter.
+\end{thm}
+
+The proof follows from a useful decomposition of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$ that yields a telescoping product in the MOP-$1$ case.
+Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting that when $\theta=\phi$, that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result.
+As such, we defer the proof to the supplementary material.
+
+This further illustrates the role of $\alpha$ as a paramteter that dictates the "memory" of the gradient estimate.
+As \cite{scibior2021dpf} remark, the MOP-$0$ estimator of \cite{blei2018vsmc} only considers single-step quantities, and is "memoryless" beyond a single step.
+This is in contrast to the case when $\alpha=1,$ as that estimate of \cite{poyiadjis11} only considers the surviving particles at time $N$, fully tracking dependencies over time.
+
+\subsection{MOP-$1$ Is Consistent for the Score}
+
+Despite showing that the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} use $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}, X_{1:N}}\left(y_{1:N}^* , x_{1:n,j}^{A, F,\theta}; \theta\right)$ in general instead of the bootstrap case of $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta};\theta\right)$.
+These have different functional forms, and are not the same a-priori.
+As such we directly show the consistency of the MOP-$1$ gradient estimate below.
+
+\begin{thm}[Consistency of MOP-$1$ Gradient Estimate]
+    The gradient estimate of MOP-$\alpha$ when $\alpha=1$, $\theta=\phi$ is strongly consistent for the score.
+$\nabla_\theta \hat\ell_J^1(\theta) \stackrel{a.s.}{\to} \nabla_\theta \ell(\theta)$ as $J \to \infty$.
+    \label{thm:mop-grad-consistency}
+\end{thm}
+\begin{proof}
+    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point we wish to evaluate the gradient at.
+The sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$, by Assumption \ref{assump:local-bounded-derivative}.
+We claim that the sequence is also uniformly equicontinuous.
+To see this, by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J^1(\theta)(\omega)|_{\theta=\theta'}$ is also bounded by $H'$ for almost every $\omega\in \Omega$ and every $\theta'\in \Theta$.
+A set of functions with derivatives bounded by the same constant is uniformly Lipschitz, and therefore uniformly equicontinuous.
+So the sequence $(\nabla_\theta \hat\lik_J^1(\theta)( \omega))_{J \in \mathbb{N}}$ is uniformly equicontinuous for almost every $\omega \in \Omega$.
+    Explicitly, for almost every $\omega \in \Omega$ and every $\epsilon>0$, there exists some $\delta(\omega)>0$ such that for every $||\theta - \theta'||_{\infty}<\delta$ and every $J \in \mathbb{N}$ we have that
+    $$||\nabla_\theta \hat\lik_J^1(\theta)(\omega)-\nabla_\theta \hat\lik_J^1(\theta')(\omega)||_\infty < \epsilon.$$
+
+    Then, by Arzela-Ascoli, there exists a uniformly convergent subsequence.
+We claim that there is only one subsequential limit.
+We know that as the gradient estimate is bounded by Assumption \ref{assump:local-bounded-derivative}, we can treat the gradient estimate at $\theta$ as a bounded functional of the particles.
+So by Theorem \ref{thm:mop-targeting} the sequence $(\nabla_\theta \hat\lik_J^1(\theta, \omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$, and there is therefore only one subsequential limit.
+The sequence therefore converges uniformly to its limit $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega).$ Therefore, with uniform convergence for the derivatives established, we can swap the limit and derivative and obtain, in conjunction with the strong consistency $\hat{\lik}_J^1(\theta, \omega) \stackrel{a.s.}{\to} \lik(\theta)$ in Theorem \ref{thm:mop-targeting}, that for almost every $\omega \in \Omega$,
+    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lik(\theta).$$
+    The result then follows by the continuous mapping theorem. 
+\end{proof}
+
+\subsection{MOP-$\alpha$ Bias and Variance}
+
+\section{Practical Maximum-Likelihood Estimation}
+
+One can think of MOP-$\alpha$ as taking a likelihood estimate at $\phi$, reweighting it to instead target the likelihood at $\theta$, and then taking the derivative with respect to $\theta$ of the above estimate.
+But if $\theta$ is evaluated at $\phi$, the particles at $\theta$ and $\phi$ coincide, and one only needs to run one particle filter at $\theta=\phi$, setting the particles at $\phi$ to be copies of the particles at $\theta$ where gradients don't propagate (and vice-versa for the measurement densities $g_{n,j}^\phi$ and conditional likelihood estimate $L_n^\phi$).
+This is done algorithmically through the \texttt{stop\_gradient()} function in \texttt{JAX}, and therefore provides a mathematical justification for the "stop-gradient trick" presented in \cite{scibior21}.
+
+\subsection{Optimization}
+This still leaves us with the question of designing an effective procedure for likelihood maximization with these new tools.
+We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that amounts to warm-starting gradient descent that uses the gradient estimate given by MOP-$\alpha$ (or some other reasonable first or second order procedure) with the output of an initial search of IF2.
+
+That is, a coarse solution is found by running IF2 with an aggressive random walk standard deviation (learning rate), until the search stalls to an initial "convergence" to what is hopefully a neighborhood of the MLE.
+We can then use gradient descent, or other first or second order gradient methods, to refine this coarse solution.
+We do this because IF2 has been empirically shown to converge quickly to a neighborhood of the MLE, and by switching to gradient descent with MOP-$\alpha$ estimates, one can bypass the difficulty that IF2 has with optimizing the last few units of log-likelihood.
+Conversely, gradient descent (and other first or second order gradient methods) can get stuck in saddle points and less desirable local minima when the likelihood is nonconvex.
+Warm-starting gradient methods with IF2 lets us get close enough to the MLE for gradient methods to shine, as in most cases regularity conditions ensure that the likelihood surface is well-behaved enough in this neighborhood that the aforementioned issues with saddle points and local minima are alleviated.
+Combining these two methods in this way lets us enjoy the best of both worlds.
+
+\begin{algorithm}[H]
+	\caption{Iterated Filtering with Automatic Differentiation}
+    \label{alg:ifad}
+	    \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $p(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta),$ IF2 cooling schedule $\eta_t$, MOP-$\alpha$ discounting parameter $\alpha$.\newline
+		\textbf{Initialize} $\theta_0$, $t=0.$\newline
+        Run IF2 until initial "convergence" under cooling schedule $\eta_t$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_t := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
+		\textbf{While} procedure not converged: \newline
+		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_t).$ \newline
+		\hspace*{4mm} Obtain $g(\theta_t) = \nabla_{\theta_t} (-\loglik(\theta_t))$, $H(\theta_t)$ s.t. $\lambda_{\min}(H) \geq c$. \newline
+		\hspace*{4mm} Update $\theta_{t+1} := \theta_t - \eta (H(\theta_t))^{-1} g(\theta_t)$, $t:=t+1.$ \newline
+		\textbf{Return} $\hat{\theta} := \theta_t.$
+\end{algorithm}
+
+\subsection{Runtime}
+
+Each iteration of IF2 amounts to one iteration of the particle filter, with $O(NJ)$ time complexity.
+The initial "convergence" happens fairly quickly in practice.
+In the case of the Dhaka cholera model of \cite{king08}, when a geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations of IF2.
+In comparison, often 100 or 200 iterations of IF2 are used for an initial global search.
+\cite{ionides15}, for example, use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{Lee_haiticholera}.
+
+In line with the cheap gradient principle of \cite{kakade2019provably}, getting a gradient estimate from MOP-$\alpha$ takes no more than 6 times that of the runtime of the particle filter (in practice, the multiplier is 3.75).
+This therefore has $O(NJ)$ time complexity as well, obtaining a gradient estimate with runtime linear in the number of particles.
+
+Re-implementing the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, even with a simulator written in Python.
+This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter (with a simulator written in C++) in the \texttt{pomp} package of \cite{king16}.
+
+\section{Application to the Cholera Model Of \cite{king08}}
+
+We use the Dhaka cholera model of \cite{king08} to benchmark the performance of our method.
+This was also used or modified in \cite{ionides15, wood16, wycoff2024voronoi} for the same purpose.
+This is a stochastic SIR compartmental model with transition and measurement uncertainty, where the population at time $t$, $H_t$, is divided into the susceptible compartment $S_t$, infected compartment $I_t$, and three ($k=3$) recovered compartments $R^1_t, ..., R^k_t$ denoting varying degrees of cholera immunity.
+We write $M_t$ for the cholera deaths in each month.
+As in \cite{king08, ionides15}, the \textbf{transition dynamics} follow this series of stochastic differential equations:
+\begin{align*}
+    d S_t&=(k \epsilon R^k_t+\delta(S_t-H_t)-\lambda_t S_t) d t+d H_t-\frac{\sigma S_t I_t}{H_t} d B_t, \\[-2mm] 
+    d I_t&=\left(\lambda_t S_t-(m+\delta+\gamma) I_t\right) d t+\frac{\sigma S_t I_t}{H_t} d B, \\[-2mm]
+    d R^1_t&=\left(\gamma I-(k \epsilon+\delta) R^1\right) d t, \\ 
+    d R^k_t&=\left(k \epsilon R^{k-1}_t-(k \epsilon+\delta) R^k_t\right) d t,
+\end{align*}
+
+where $B_t$ represents Brownian motion, $m$ is the rate at which infected individuals succumb to cholera, $\gamma$ is the rate at which infected individuals recover, $1/\epsilon$ is the mean duration of immunity, $\sigma$ represents the standard deviation of perturbations to the force of infection, and $\delta=0.02$ is the death rate.
+
+The \textbf{force of infection}, $\lambda_t$, is modeled by splines $(s_j)_{j=1}^6$
+\begin{equation}
+    \lambda_t=\exp\hspace{-1mm}\left(\hspace{-.5mm}\beta_{\text{trend}}(t-t_0)+\hspace{-1mm}\sum_{j=1}^{6} \beta_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm}\frac{I_t}{P_t} + \exp \left(\sum_{j=1}^{6} \omega_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm},
+\end{equation}
+where the coefficients $(\beta_j)_{j=1}^6$ model seasonality in the force of infection, $\beta_{\text{trend}}$ models the trend in the force of infection, and the $\omega_j$ represent seasonality of a non-human environmental reservoir of disease.
+The \textbf{measurement model} for observed monthly cholera deaths is given by
+\begin{equation}
+    Y_n \sim \gN(M_n, \tau^2M_n^2),
+\end{equation}
+where $M_n$ is the true number of cholera deaths in that month.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth/2]{imgs/095/tikzcholera.png}
+    \vspace*{-7mm}
+    \caption{Illustration of Dhaka cholera model from \cite{king08}.}
+    \label{fig:tikz-cholera}
+\end{figure}
+
+\subsection{Results}
+
+We benchmarked IFAD against IF2 on a challenging global search problem.
+We re-implemented IF2 in order to do so, but chose to benchmark our results with the results of \cite{ionides15} (labeled "IF2 2015").
+We note that our re-implementation ourperforms that of \cite{ionides15}, likely due to a better choice of parameters.
+For each method, we performed 100 searches, initialized with 100 initial starting parameter vectors drawn uniformly from the same wide bounding box used in \cite{ionides15}.
+We summarize our findings below.
+
+\begin{table}[h!]
+\centering
+\begin{tabular}{||c c c||} 
+ \hline
+ Method & Best Log-Likelihood & Rank \\ [0.5ex] 
+ \hline\hline
+     IFAD-0 & -3747.87 & 1\\
+     IFAD-1 & -3748.53 & 2\\
+     IF2 (Ours) & -3755.91 & 3\\
+     IF2 (Warm Start) & -3759.60 & 4 \\
+     IF2 2015 & -3768.63 & 5\\ 
+     MOP-0 Alone & -3795.74 & 6\\
+     MOP-1 Alone & -3797.38 & 7\\
+ \hline
+\end{tabular}
+\caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone.
+IFAD performs the best among all methods.
+Our implementation of IF2 outperforms that of \cite{ionides15}, but still ultimately underperforms IFAD.
+IFAD manages to find the MLE, matching the highest log-likelihood previously found in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.}
+\label{table:mle}
+\end{table}
+
+\paragraph{IFAD Successfully Finds the MLE:} Previously, the MLE at a log-likelihood of $-3748.5$ was found by \cite{king16} with much effort, using many global and local IF2 searches, with the assistance of likelihood profiling.
+Meanwhile, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.63$, while \cite{king08} only achieve a best possible estimate of $-3793.4$.
+Despite being initialized for a global search, IFAD manages to find the MLE over the 100 searches, as seen in Table \ref{table:mle}.
+This suggests that the local search and refinement that was previously necessary for finding the MLE with IF2 is \textbf{baked into} the two-stage IFAD algorithm.
+
+\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations (as seen in Figure \ref{fig:scatter-optim}), performing IF2 alone ultimately fails to squeeze out the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter-optim}).
+Conversely, when we tried gradient descent with MOP alone, we encountered many, many failed searches.
+This is a difficult, nonconvex, and noisy problem, and the search gets stuck in local minima and saddle points, failing to approach the MLE.
+
+IFAD, in comparison, manages to (1) approach the MLE quickly due to the IF2 warm-start (as seen in Figure \ref{fig:scatter-optim}) and (2) refine the coarse solution found by the warm-start with MOP gradient steps to find the MLE (as seen in Figures \ref{fig:scatter-optim} and \ref{fig:boxplot-search}.
+IFAD therefore successfully combines the best qualities of IFAD and MOP, outperforming either of them alone.
+
+\begin{figure}[htbp!]
+    \centering
+    \includegraphics[width=\textwidth/5]{../imgs/095/pairs.png}
+    \includegraphics[width=\textwidth/5]{../imgs/095/qq.png}
+    \caption{Scatterplots depicting the performance of IFAD against that of IF2.
+\textbf{Left:} Paired searches from the same starting point.
+\textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches.
+It is clear that on average, IFAD has the edge, and manages to find the MLE while no IF2 search successfully gets within 7 log-likelihood units of it.
+We use a dotted red line to display the MLE in both.
+}
+    \label{fig:scatter-optim}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth/4]{../imgs/095/boxplot.png}
+    \includegraphics[width=\textwidth/5]{../imgs/095/optim.png}
+    \caption{\textbf{Left:} Boxplot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs, simulating the result of the common procedure of running a few searches and choosing the best one.
+IFAD outperforms all other methods, and the gradient steps improve on the warm-start given by running 40 IF2 iterations.
+    \textbf{Right:} Optimization progress of IFAD and IF2.
+The dashed orange line depicts the median warm-start given by running 40 IF2 iterations.
+While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD -- IFAD has better tail control and successfully reaches the MLE.
+    We use a dotted red line to display the MLE.}
+    \label{fig:boxplot-search}
+\end{figure}
+\kevin{Change the optim graph to include median search and best search with dashes as linestyle, remove the shaded areas. }
+
+\section{Discussion, Limitations, and Future Work}
+
+We note that it is possible to work without a differentiable simulator, but one would then require access to the transition densities.
 
@@
 
-\subsection*{Format}
+% \subsection*{Format}
 
-Many authors find it useful to organize their manuscripts with the following order of sections: title, author line and affiliations, keywords, abstract, significance statement, introduction, results, discussion, materials and methods, acknowledgments, and references.
+% Many authors find it useful to organize their manuscripts with the following order of sections: title, author line and affiliations, keywords, abstract, significance statement, introduction, results, discussion, materials and methods, acknowledgments, and references.
 Other orders and headings are permitted.
@@
 
-%% References should be cited in numerical order as they appear in text; this will be done automatically via bibtex,
+% %% References should be cited in numerical order as they appear in text; this will be done automatically via bibtex,
 
-\subsection*{Data Archival}
+% \subsection*{Data Archival}
 
-PNAS must be able to archive the data essential to a published article.
+% PNAS must be able to archive the data essential to a published article.
 Where such archiving is not possible, deposition of data in public databases, such as GenBank, ArrayExpress, Protein Data Bank, Unidata, and others outlined in the \href{https://www.pnas.org/author-center/editorial-and-journal-policies#materials-and-data-availability}{Information for Authors}, is acceptable.
-
-\subsection*{Historical cholera transmission in Dacca, Bangladesh}.
-We tested our method on the cholera transmission model and data of \citep{king08}.
-This is a POMP model based on the fundamental Susceptible-Exposed-Infected-Recovered (SEIR) epidemiological model, but incorporating additional features that are required for practical application of this model to various real disease systems.
-Seasonality is included as a periodic cubic B-spline covariate process.
-Loss of immunity is modeled by return from the Recovered class to the Susceptible class.
-Asymptomatic infections can be modeled separately to allow for the possibility that they differ from symptomatic infections in their infectivity and acquired immunity.
-This model, and the Dacca case reports to which it was originally fitted, provide a benchmark for a challening scientific model fitting exercise.
-It was used oringally to test IF1 \citep{ionides06-pnas} and IF2  \citep{ionides15}, and a simplified version has been used to compare with alternative methodologies \citep{fasiolo16}.
-We will therefore benchmark on this well-established model before moving on to target the methods on models with anticipated scientific goals.
-
-\begin{figure}%[tbhp]
-\centering
-\includegraphics[width=.8\linewidth]{23-09-05-adpf-vs-if2.png}
-\caption{Comparing the ADPF algorithm with IF2 on the Dacca model and data of \citep{king08}}
-  \label{fig:dacca-fit}
-\end{figure}
-
-\begin{table}[t!]
-\centering
-\caption{Comparison of the fitted potential energy surfaces and ab initio benchmark electronic energy calculations}
-\begin{tabular}{lrrr}
-Species & CBS & CV & G3 \\
-\midrule
-1. Acetaldehyde & 0.0 & 0.0 & 0.0 \\
-2. Vinyl alcohol & 9.1 & 9.6 & 13.5 \\
-3. Hydroxyethylidene & 50.8 & 51.2 & 54.0\\
-\bottomrule
-\end{tabular}
-
-\addtabletext{nomenclature for the TSs refers to the numbered species in the table.}
-\end{table}
 
@@
 
-\subsection*{Tables}
-Tables should be included in the main manuscript file and should not be uploaded separately.
+%% \subsection*{Tables}
+%% Tables should be included in the main manuscript file and should not be uploaded separately.
 
-\subsection*{Single column equations}
+%% \subsection*{Single column equations}
 
-Authors may use 1- or 2-column equations in their article, according to their preference.
+%% Authors may use 1- or 2-column equations in their article, according to their preference.
 
-To allow an equation to span both columns, use the \verb|\begin{figure*}...\end{figure*}| environment mentioned above for figures.
+%% To allow an equation to span both columns, use the \verb|\begin{figure*}...\end{figure*}| environment mentioned above for figures.
 
-Note that the use of the \verb|widetext| environment for equations is not recommended, and should not be used.
+%% Note that the use of the \verb|widetext| environment for equations is not recommended, and should not be used.
 
-\begin{figure*}[bt!]
-\begin{align*}
-(x+y)^3&=(x+y)(x+y)^2\\
-       &=(x+y)(x^2+2xy+y^2) \numberthis \label{eqn:example} \\
-       &=x^3+3x^2y+3xy^3+x^3.
-\end{align*}
-\end{figure*}
+%% \begin{figure*}[bt!]
+%% \begin{align*}
+%% (x+y)^3&=(x+y)(x+y)^2\\
+%%        &=(x+y)(x^2+2xy+y^2) \numberthis \label{eqn:example} \\
+%%        &=x^3+3x^2y+3xy^3+x^3.
+%% \end{align*}
+%% \end{figure*}
 
-\subsection*{Supporting Information Appendix (SI)}
+% \subsection*{Supporting Information Appendix (SI)}
 
-Authors should submit SI as a single separate SI Appendix PDF file, combining all text, figures, tables, movie legends, and SI references.
+% Authors should submit SI as a single separate SI Appendix PDF file, combining all text, figures, tables, movie legends, and SI references.
 SI will be published as provided by the authors; it will not be edited or composed.
@@
 
-Authors who place detailed materials and methods in an SI Appendix must provide sufficient detail in the main text methods to enable a reader to follow the logic of the procedures and results and also must reference the SI methods.
+% Authors who place detailed materials and methods in an SI Appendix must provide sufficient detail in the main text methods to enable a reader to follow the logic of the procedures and results and also must reference the SI methods.
 If a paper is fundamentally a study of a new method or technique, then the methods must be described completely in the main text.
 
-\subsubsection*{SI Datasets}
+% \subsubsection*{SI Datasets}
 
-Supply .xlsx, .csv, .txt, .rtf, or .pdf files. This file type will be published in raw format and will not be edited or composed.
-
-\matmethods{Please describe your materials and methods here. This can be more than one paragraph, and may contain subsections and equations as required.
-
-\subsection*{Subsection for Method}
-Example text for subsection.
-}
+% Supply .xlsx, .csv, .txt, .rtf, or .pdf files. This file type will be published in raw format and will not be edited or composed.
 
@@
 
-%\bibliographystyle{pnas-new}
+\bibliographystyle{pnas-new}
 % Bibliography
-\bibliography{bib-ifad}
+\bibliography{bib-ifad, bib-ref}
 

### 2024-04-18 paper/ms.tex
Updates from Overleaf

@@
 \significancestatement{Many scientific models involve highly nonlinear stochastic dynamical systems which can be observed only via noisy and incomplete measurements.
-Under the Markov assumption on system dynamics, previous work has provided methods of performing inference for these models.
-In particular, prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics.
+Prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics.
 We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
@@
 \authordeclaration{The authors declare no competing interests.}
-\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: ionides@umich.edu}
+\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: kevtan@wharton.upenn.edu}
 
@@
   Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods.
-Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques due to (1) the requirement for simulation-based inference that does not require access to the system's transition probabilities and (2) the issue that widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters.
+Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques as widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters.
+Furthermore, most approaches are not compatible with the desire for simulation-based inference without access to the system's transition probabilities.
 We propose a new representation that embeds existing methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff.
@@
 
-\firstpage{4}
+\firstpage{5}
 % Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting.
@@
 \dropcap{M}any scientific models involve highly nonlinear stochastic dynamic systems, where there may be significant uncertainty in both the process dynamics and the measurements of the system.
-Despite the ubiquity of such partially observed stochastic processes with the Markov property (also known as, continuous-state continuous-time hidden Markov models (HMMs), partially-observed Markov processes (POMPs), or partially-observed stochastic nonlinear dynamical systems) in fields as diverse as epidemiology \cite{he10, stocks17}, ecology \cite{knape12}, and finance \cite{kim08, breto14}, performing estimation and inference within this class of models is a difficult problem that faces several challenges.
+Despite the ubiquity of such partially observed stochastic processes with the Markov property (also known as, continuous-state continuous-time hidden Markov models (HMMs), partially-observed Markov processes (POMPs), or partially-observed stochastic nonlinear dynamical systems) in fields as diverse as epidemiology \cite{he10, stocks17}, ecology \cite{knape12}, and finance \cite{kim08, breto14}, estimation and inference within this class of models is a difficult problem with several challenges.
 
-Many approaches to inference in these scenarios (such as the EM algorithm, the various Kalman filter variants, and MCMC) either struggle with intractable likelihood functions when the models are complex, or assume access to the probability density of next states given the current state.
-This is a problem in some critical applications, such as disease modeling, where the models are complex enough such that even obtaining the transition density is an intractable problem.
-However, the particle filter with the bootstrap proposal, a popular method for solving the filtering problem in partially-observed dynamical systems, provides an unbiased estimate of the likelihood \cite{delmoral2004feynman} without requiring evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
+Popular methods for estimation and inference in these scenarios (such as the EM algorithm, the various Kalman filter variants, and MCMC) either struggle with complex models that have intractable likelihood functions, or assume access to state transition densities.
+This is a problem in some critical applications, such as disease modeling, where transition densities are unavailable.
+However, in such cases a simulator may be readily available.
+In this case, the particle filter with the bootstrap proposal, a popular method for solving the filtering problem in partially-observed dynamical systems, provides an unbiased estimate of the likelihood \cite{delmoral2004feynman} without requiring evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
 
 A series of related algorithms \cite{welch2009abc, wood2010sl, doucet2010pmcmc, ionides08, ionides15} enable parameter estimation within these models.
-In particular, the improved iterated filtering algorithm (IF2) of Ionides et al.
-\cite{ionides15} and the particle MCMC algorithm of Andrieu et al.
-\cite{doucet2010pmcmc} are the only simulation-based full-information maximum likelihood and fully Bayesian methods available in the literature thus far.
-However, accurate parameter estimation in the presence of significant Monte-Carlo noise in the likelihood estimate remains a significant challenge.
-Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small.
+In particular, the improved iterated filtering algorithm (IF2) \cite{ionides15} is the only simulation-based full-information maximum likelihood method available, whereas particle MCMC \cite{doucet2010pmcmc} and $\text{SMC}^2$ \cite{chopin13} are the only simulation-based full-information Bayesian methods available in the literature thus far.
+However, accurate parameter estimation with significant Monte-Carlo noise in the likelihood estimate is challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small.
 For example, though in practice the algorithm of Ionides et al.
 \cite{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood.
-Ionides et al.
-\cite{Ionides_mcap} attempt to address the third problem, albeit with a computationally-expensive approach.
+This necessitates a new approach.
 
@@
 
-Recent advances in automatic differentiation (AD) for particle filters \cite{blei2018vsmc, jon2018diffpf, corenflos21, scibior2021dpf, doucet2022particlebased} have drawn attention to AD as an alternative tool for inference in partially observed stochastic processes with the particle filter.
-However, these existing approaches either are not yet compatible with the desire for simulation-based inference, or are computationally expensive.
-In particular, existing approaches are either asymptotically biased \cite{blei2018vsmc, jon2018diffpf}, have high variance \cite{poyiadjis11, scibior21}, computationally expensive \cite{corenflos21, chen24}, or require transition densities \cite{poyiadjis11, scibior21, doucet2022particlebased, chen24}.
+Recent advances in automatic differentiation (AD) for particle filters \cite{blei2018vsmc, jon2018diffpf, corenflos21, scibior2021dpf, doucet2022particlebased} have drawn attention to AD as a tool for inference in partially observed stochastic processes.
+However, existing approaches are either asymptotically biased \cite{blei2018vsmc, jon2018diffpf}, have high variance \cite{poyiadjis11, scibior21}, computationally expensive \cite{corenflos21, chen24}, or need access to transition densities \cite{poyiadjis11, scibior21, doucet2022particlebased, chen24}.
 
-In particular, the work of Scibior and Wood \cite{scibior2021dpf} demonstrates that the estimators derived in Poyiadjis et al.
-\cite{poyiadjis11} can be attained with AD, albeit with a simple tweak, that they call the stop-gradient trick, to the particle filter that does not modify its likelihood estimates.
-Though their method is not fully compatible with simulation-based inference and the mathematical motivation behind the stop-gradient trick is not apparent, we show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.
+In particular, Scibior and Wood \cite{scibior21} show that the estimators derived in Poyiadjis et al.
+\cite{poyiadjis11} can be attained with AD, through with a simple tweak (the "stop-gradient trick") to the particle filter that does not modify its likelihood estimates.
+However, their method is not fully compatible with simulation-based inference and the mathematical motivation behind the stop-gradient trick is not apparent.
+
+%We show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.
 
 \subsection{Our Contributions}
-We develop a new approach to statistical inference via AD for particle filters which addresses these concerns.
-We construct a new theoretical framework that addresses the seemingly incompatible paradox of differentiating through a Monte-Carlo algorithm with discontinuous resampling.
+We develop a new approach to statistical inference via AD for particle filters which addresses the above concerns.
+We construct a new theoretical framework that bypasses the paradox of differentiating through a Monte-Carlo algorithm with discontinuous resampling.
 
-The key observation is that one can turn this into a problem of differentiating through a series of smooth measurement density ratios and (through the reparametrization trick) a differentiable simulator.
-This is done through the construction of a suitably reweighted family of particle filters that we call \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$), for some $\alpha \in [0,1]$ that optimizes a bias-variance tradeoff.
-This family of algorithms encompasses the special case of the biased gradient estimator of \cite{blei2018vsmc} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$, and does not require access to the transition densities -- only requiring access to a differentiable simulator.
+Our key observation is that one can turn this into a problem of differentiating through a series of smooth measurement density ratios and (through the reparametrization trick) a differentiable simulator.
+To do so, we construct a suitably reweighted family of particle filters that we call \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$), for some $\alpha \in [0,1]$ that optimizes a bias-variance tradeoff.
+This family encompasses the biased gradient estimator of \cite{blei2018vsmc} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
+It does not require access to the transition densities, and works with just a differentiable simulator.
 
@@
 
-In practice, we propose a hybrid algorithm that we call \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts gradient ascent (or a similar iterative first or second-order gradient method) with a coarse solution obtained from a few rounds of iterated filtering.
-Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1, the Liu-West filter \cite{liuwest01}, and particle MCMC \cite{doucet2010pmcmc}) on a challenging scientific benchmark problem, the Dhaka cholera model of \cite{king08}.
+This allows us to develop improved algorithms for maximum likelihood/MAP estimation in highly nonlinear stochastic dynamical systems.
+We propose a hybrid algorithm called \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts iterative first or second-order gradient methods with a coarse solution obtained from a few rounds of iterated filtering.
+Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.
+
+These improvements also extend to Bayesian inference, as we show that we can use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with a nonparametric empirical Bayes-style prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{doucet2010pmcmc} from the $1.4 \times 10^6$ iterations in \cite{fasiolo16} to just $500$.
 
@@
 The process is parameterized by an unknown parameter $\theta \in \Theta \subseteq \R^p$, where the states $X_t$ take values in the state space $\gX \subseteq \R^d$, the observations $Y_n$ take values in $\gY,$ and we write $X_n := X_{t_n}$.
-By a similar decomposition to that in \cite{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
-\begin{align}
-    &f_{X_{0: N}, Y_{1: N}}\left(x_{0: N}, y_{1: N} ; \theta\right)\\
-    &=f_{X_0}\left(x_0 ; \theta\right) \prod_{n=1}^N f_{X_n \mid X_{n-1}}\left(x_n \mid x_{n-1} ; \theta\right) f_{Y_n \mid X_n}\left(y_n \mid x_n ; \theta\right).
-\end{align}
 
-We call $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ the process model, writing $\texttt{process}\left(x_n; \theta\right)$ for the simulator corresponding to the above process model, $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ the measurement model, and again write $y_n^*$ for the actual values of the observations that were observed.
-As a matter of terminology, we call $f_{X_{1:N}|Y_{1:N}}(x_{1_N}|y_{1:N}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
+% By a similar decomposition to that in \cite{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
+% \begin{align*}
+%     &f_{X_{0: N}, Y_{1: N}}\left(x_{0: N}, y_{1: N} ; \theta\right)\\
+%     &=f_{X_0}\left(x_0 ; \theta\right) \prod_{n=1}^N f_{X_n \mid X_{n-1}}\left(x_n \mid x_{n-1} ; \theta\right) f_{Y_n \mid X_n}\left(y_n \mid x_n ; \theta\right).
+% \end{align*}
 
-The above are all defined on a single probability space $(\Omega, \Sigma, \prob),$ where $\omega \in \Omega$ denotes an element of the sample space that can be thought of as a random seed.
-For notational simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0.$
-In the rest of this paper, we will identify the seed with an element of the sample space.
-We do so because the seed determines the entire sequence of pseudo-random numbers as generated by a computer, to maintain an algorithmic and practical interpretation of our theory.
-When two algorithms share the same seed, this corresponds to the method of common random numbers.
+We write $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ for the process model, writing $\texttt{process}\left(x_n; \theta\right)$ for its corresponding simulator, $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ the measurement model, and $y_n^*$ for the actual values of the observations that were observed.
+We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
+Superscripts $x_{n,j}^A$ denote the ancestral trajectory of particle $j$ at time $n$, and $a(\cdot)$ is the ancestor function.
+For simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0.$
+
+The above are defined on a probability space $(\Omega, \Sigma, \prob).$ 
+To maintain an algorithmic and practical interpretation of our theory, we identify the seed with an element of the sample space $\omega \in \Omega$, as a random seed determines the entire sequence of pseudo-random numbers as generated by a computer.
+
+%When two algorithms share the same seed, this corresponds to the method of common random numbers.
 
@@
 
-As mentioned earlier, instead of directly differentiating through a particle filter, one can instead differentiate through the simulator and a series of measurement density ratios.
-We construct an algorithm called \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) in Algorithm \ref{alg:mop} that targets the posterior and likelihood, and whose gradient yields the estimators of \cite{poyiadjis11, scibior21} and \cite{blei2018vsmc} as special cases.
-
-\subsection{Algorithm Outline} 
-The broad idea is to run the particle filter twice under the same seed.
-We first run the particle filter under $\phi$, and then obtain a coupled set of particles by running a bootstrap filter with the process model at $\theta$ but the resampling indices from the previous run at $\phi$.
-We can then reweight the conditional likelihoods by a correction factor accumulated over time of the ratio of the measurement density under $\theta$ and the measurement density under $\phi$ to account for the ("off-policy") resampling under $\phi$.
-That is, when we write $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ as shorthand for the measurement density and estimate the conditional likelihood under $\phi$ in the usual way with $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$, we can estimate the conditional likelihood under $\theta$ as either
-\begin{equation}
-     \label{eq:mop-conditional-likelihood}
-     L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}} \text{ or } L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
-\end{equation}
-where each particle is recursively weighted according to 
-\begin{equation}
-    \label{eq:weighting-scheme}
-    w_{n,j}^{P,\theta} = (w_{n-1,j}^{F,\theta})^\alpha,  w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}, w^{F,\theta}_{0,j}= 1,
-\end{equation}
-and $k_j$ are the indices from resampling according to the measurement density at $\phi$, $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$.
-The before-resampling conditional likelihood estimate $L_n^{B,\theta,\alpha}$ should be used in general, as it has slightly lower variance than the after-resampling estimate $L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$, but the latter is useful in deriving properties of the MOP-$\alpha$ gradient estimate such as that in Theorem \ref{thm:mop-functional-forms}.
-
-This yields a suitably reweighted particle filter that targets the posterior under $\theta$ and produces a strongly consistent likelihood estimate for the likelihood at $\theta$.
-Taking the gradient with respect to $\theta$ of the estimate for the log-likelihood at $\theta$ then yields an estimate for the score at $\theta$.
+As mentioned earlier, instead of directly differentiating through a particle filter, one can instead differentiate through the simulator and measurement densities.
+We construct an algorithm called \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) that targets the posterior (i.e.
+almost sure convergence of every measurable bounded functional of particles to its expectation under the posterior) and likelihood, and whose gradient yields a score estimate encompassing those of \cite{poyiadjis11, scibior21, blei2018vsmc}.
 
 \begin{algorithm}[H]
-	\caption{Measurement Off-Policy-$\alpha$}
+	\caption{MOP-$\alpha$}
     \label{alg:mop}
@@
         \textbf{If $\theta \neq \phi$:}
-            Run a particle filter once with parameters $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$, with seed $\omega$. \newline
-        \textbf{Else:} Set $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$ to be copies of $X_{n,j}^{P,\theta}, X_{n,j}^{P,\theta}$ in the rest of the algorithm. \newline
-		\textbf{Initialize } filter particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$.
-Fix $\omega.$ \newline
+            Fix $\omega$, and filter at $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$. \newline
+        \textbf{Else:} Fix $\omega$, and set $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$ to be copies of $X_{n,j}^{P,\theta}, X_{n,j}^{P,\theta}$. \newline
+		\textbf{Initialize } particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, weights $w^{F,\theta}_{0,j}= 1$. \newline
 		\textbf{For} $n=1,...,N$: \newline
@@
             \hspace*{4mm} Compute $ L_n^{A,\theta,\alpha} = L_n^\phi\cdot {\sum_{j=1}^J w^{F,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$.\newline
-		\textbf{Return:} likelihood estimate $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}_{j=1}^J.$
+		\textbf{Return:} likelihood estimate $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, filtering distributions $\{(X_{n,j}^{F, \theta}, w^{F,\theta}_{n,j})\}_{n,j=1}^{N,J}.$
 \end{algorithm}
+
+\subsection{Algorithm Outline} 
+We obtain two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but the resampling indices from the run at $\phi$ (hence the "off-policy" moniker).
+If $\theta$ and $\phi$ coincide, one only needs one particle filter run at $\theta=\phi$, otherwise one needs two runs at the same seed $\omega \in \Omega$.
+
+We can then reweight the conditional likelihoods by a correction factor accumulated over time to account for the resampling under $\phi$.
+That is, writing $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density, $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$, and $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$ for the resampling indices under $\phi$, we can estimate the conditional likelihood under $\theta$ as
+\vspace*{-2.5mm}
+\begin{equation}
+     \label{eq:mop-conditional-likelihood}
+     L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}} \text{ or } L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
+     \vspace*{-2.5mm}
+\end{equation}
+recursively weighting each particle by 
+\vspace*{-2.5mm}
+\begin{equation}
+    \label{eq:weighting-scheme}
+    w_{n,j}^{P,\theta} = (w_{n-1,j}^{F,\theta})^\alpha,  w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}, w^{F,\theta}_{0,j}= 1.
+    \vspace*{-2.5mm}
+\end{equation}
+
+The before-resampling conditional likelihood estimate $L_n^{B,\theta,\alpha}$ should be used in general, as it has slightly lower variance than the after-resampling estimate $L_n^{A,\theta,\alpha}$, but the latter is useful in deriving properties of the MOP-$\alpha$ gradient estimate such as that in Theorem \ref{thm:mop-functional-forms}.
+
+This yields a suitably reweighted particle filter that targets the posterior under $\theta$ and produces a strongly consistent likelihood estimate for the likelihood at $\theta$.
+Taking the gradient with respect to $\theta$ of the estimate for the log-likelihood at $\theta$ then yields an estimate for the score at $\theta$.
 
@@
 If $\theta$ and $\phi$ coincide, one can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$, as seen in Equation \ref{eq:weighting-scheme}, while still targeting the posterior and likelihood at $\theta$.
-We do this in order to optimize a bias-variance tradeoff for the score estimate yielded by differentiating through the log-likelihood estimate provided by MOP-$\alpha$.
+This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate.
 When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, recovering the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}.
@@
 
-\section{Theoretical Analysis of MOP-$\alpha$}
+\subsection{Summary of Theoretical Guarantees}
 
-Here, we will show that MOP-$\alpha$ targets the posterior and likelihood, yields the estimators of \cite{poyiadjis11, scibior21, blei2018vsmc} as special cases when differentiated, and characterize rates for its bias and variance under different choices of $\alpha$.
-To do so, we require the following assumptions:
-
-\begin{enumerate}[label=(A\arabic*),itemsep=-1.2ex] 
-    \item \textbf{Continuity of the Likelihood.} $\ell(\theta)$ has more than two continuous derivatives in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.
-\label{assump:conti-lik}
-    \item \textbf{Bounded Process Model.} There exist $\underbar{M}, \bar{M}$ such that $0 < \underbar{M} \leq f_{X_n|X_{n-1}}(x_n | x_{n-1};\theta) \leq \bar{M} < \infty$.
-\label{assump:bounded-process}
-    \item \textbf{Bounded Measurement Model.} There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G}<\infty$ and $G'(\theta)$ so $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty \leq G'(\theta)< \infty$.
-\label{assump:bounded-measurement}
-    \item \textbf{Bounded Gradient Estimates.} There are functions $G(\theta), H(\theta): \Theta \to [0,\infty)$ uniformly bounded by $G', H'$, so the MOP-$\alpha$ gradient and Hessian estimates at $\theta=\phi$ are almost surely bounded by $G(\theta)$ and $H(\theta)$ for all $\alpha$.
-\label{assump:local-bounded-derivative}
-    \item \textbf{Differentiability of Density Ratios and Simulator.} The measurement density $f_{Y_n|X_n}(y_n^*|x; \theta)$ and simulator have more than two continuous derivatives in $\theta$.
-\label{assump:diff-meas-and-sim}
-\end{enumerate}
-
-The fact that the likelihood estimate yielded by MOP-$\alpha$ has more than two continuous derivatives in $\theta$ follows the construction of the likelihood estimate in equations \ref{eq:mop-conditional-likelihood} and \ref{eq:weighting-scheme}, as well as Assumptions \ref{assump:bounded-measurement} and \ref{assump:diff-meas-and-sim}.
-
-\subsection{MOP-$\alpha$ Targets the Posterior}
-
-We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood.
-To do so, we state the definition of targeting and present three lemmas, whose proof can be found in the supplementary material.
-\begin{defn}[Targeting]
-    \label{defn:targeting}
-    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any measurable and bounded functional $h$,
-    \begin{equation}
-        E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
-    \end{equation}  
-        A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if for any measurable and bounded functional $h$, as $J\to\infty$
-    \begin{equation}
-        \frac{\sum_{j=1}^J h(X^J_j) w^J_j}{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X)).
-    \end{equation}
-\end{defn}
-
-\begin{lem}[Change of Particle Measure]
-    \label{lem:change-measure-proper-weights}
-    Assume $\{(\tilde X_j^J,U_j^J)\}_{j=1}^J$ targets $f_X$, and that $\{(Y_j^J,V_j^J)\}_{j=1}^J$ is a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is drawn w.p.
-$\pi^J_j$.
-Write, where $a$ is the ancestor function,
-    $(Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big)
-    $. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j)\}_{j=1}^J$ targets $f_X$.
-\end{lem}
-\begin{lem}[Particle Marginals]
-    \label{lem:marginal-proper-weights}
-    Suppose that $\{(\tilde X_j^J,U_j^J)\}_{j=1}^J$ targets $f_X$.
-Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$, where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
-Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}_{j=1}^J$ targets $f_Z$.
-\end{lem}
-\begin{lem}[Particle Posteriors]
-    \label{lem:posterior-proper-weights}
-    Suppose that $\{(X_j^J,u_j^J)\}_{j=1}^J$ targets $f_X$, and write $(X^{\prime J}_j,u^{\prime J}_j) = \big(X_j^J,u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$.
-Then, if $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,u^{\prime J}_j)\}_{j=1}^J$ targets $f_{X|Z}(\cdot | z^*)$.
-\end{lem}
-
-From these lemmas, we can derive the following result.
-
-\begin{thm}[MOP-$\alpha$ Targets the Posterior and Likelihood]
-    \label{thm:mop-targeting}
-    When $\alpha=1$ or $\phi=\theta$, MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood.
-That is, for $\pi=f_{X_{1:N}|Y_{1:N} ; \theta}$ and any measurable and bounded functional $h$ and for $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\prod_{n=1}^N L_n^{B,\theta,\alpha}$, it holds that
-    \begin{equation}
-        \frac{\sum_{j=1}^J h(x_{n,j}^{P, \theta}) w_{n,j}^{P,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} \stackrel{a.s.}{\to} E_\pi h, \;\; \hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).
-    \end{equation}
-\end{thm}
-
-\begin{proof}
-    When $\theta=\phi$, regardless of the value of $\alpha$, the ratio ${g_{n,j}^\theta}/{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter estimate.
-
-    When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows.
-Recursively applying Lemmas \ref{lem:change-measure-proper-weights}, \ref{lem:marginal-proper-weights}, and \ref{lem:posterior-proper-weights}, we obtain that
-    %to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
-    the MOP-1 filter targets the posterior.
-    Specifically, suppose inductively that $\big\{\big(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
-    Then, Lemma \ref{lem:marginal-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big)\big\}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
-    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-    Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
-    \begin{equation}
-        \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}\big/ g^\phi_{n,a(j)}\big),
-    \end{equation}
-    with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-
-    For the likelihood, if the likelihood is estimated with the before-resampling conditional likelihoods $\hat\lik(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, the strong consistency is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
-
-    Whereas if we use the after-resampling conditional likelihoods $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$, the numerator of $L_n^{B,\theta,\alpha}$ is
-    \begin{equation}
-        \sum_{j=1}^J g_{n,j}^\theta w_{n, j}^{P, \theta} = L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
-    \end{equation}
-    Using Lemma \ref{lem:change-measure-proper-weights}, we resample the corrected weights $w^{P,\theta}_{n,j}{g^\theta_{n,j}}/{g^\phi_{n,j}} $ according to probabilities ${g^\phi_{n,j}}/{L_n^\phi}$ to see this is properly estimated by $L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},$
-from which we obtain the estimate $L_n^{B,\theta,\alpha}$. The strong consistency then follows as the likelihood is the product of the conditional likelihoods.
-\end{proof}
-
-\subsection{MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, blei2018vsmc}}
-
-\cite{scibior21} show that the estimate of \cite{blei2018vsmc} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through applying a "stop-gradient trick" to the particle filter.
-It turns out that both of these, when applied on the bootstrap filter, correspond to special cases of MOP-$\alpha$.
-
-\begin{thm}[MOP-$0$ and MOP-$1$ Functional Forms]
-    \label{thm:mop-functional-forms}
-    Write $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$ and we use the after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$.
-Then, when $\alpha=0$,
-    \begin{equation}
-        \nabla_\theta \hat\ell^0(\theta) 
-        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
-    \end{equation}
-    yielding the gradient estimator of \cite{blei2018vsmc} with the bootstrap filter. When $\alpha=1$,
-    \begin{equation}
-        \nabla_\theta \hat{\ell}^1(\theta) 
-        = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
-    \end{equation}
-    yielding the gradient estimator of \cite{poyiadjis11, scibior21} applied to the bootstrap filter.
-\end{thm}
-
-The proof follows from a useful decomposition of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$ that yields a telescoping product in the MOP-$1$ case.
-Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting that when $\theta=\phi$, that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result.
-As such, we defer the proof to the supplementary material.
-
-This further illustrates the role of $\alpha$ as a paramteter that dictates the "memory" of the gradient estimate.
-As \cite{scibior2021dpf} remark, the MOP-$0$ estimator of \cite{blei2018vsmc} only considers single-step quantities, and is "memoryless" beyond a single step.
-This is in contrast to the case when $\alpha=1,$ as that estimate of \cite{poyiadjis11} only considers the surviving particles at time $N$, fully tracking dependencies over time.
-
-\subsection{MOP-$1$ Is Consistent for the Score}
-
-Despite showing that the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} use $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}, X_{1:N}}\left(y_{1:N}^* , x_{1:n,j}^{A, F,\theta}; \theta\right)$ in general instead of the bootstrap case of $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta};\theta\right)$.
-These have different functional forms, and are not the same a-priori.
-As such we directly show the consistency of the MOP-$1$ gradient estimate below.
-
-\begin{thm}[Consistency of MOP-$1$ Gradient Estimate]
-    The gradient estimate of MOP-$\alpha$ when $\alpha=1$, $\theta=\phi$ is strongly consistent for the score.
-$\nabla_\theta \hat\ell_J^1(\theta) \stackrel{a.s.}{\to} \nabla_\theta \ell(\theta)$ as $J \to \infty$.
-    \label{thm:mop-grad-consistency}
-\end{thm}
-\begin{proof}
-    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point we wish to evaluate the gradient at.
-The sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$, by Assumption \ref{assump:local-bounded-derivative}.
-We claim that the sequence is also uniformly equicontinuous.
-To see this, by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J^1(\theta)(\omega)|_{\theta=\theta'}$ is also bounded by $H'$ for almost every $\omega\in \Omega$ and every $\theta'\in \Theta$.
-A set of functions with derivatives bounded by the same constant is uniformly Lipschitz, and therefore uniformly equicontinuous.
-So the sequence $(\nabla_\theta \hat\lik_J^1(\theta)( \omega))_{J \in \mathbb{N}}$ is uniformly equicontinuous for almost every $\omega \in \Omega$.
-    Explicitly, for almost every $\omega \in \Omega$ and every $\epsilon>0$, there exists some $\delta(\omega)>0$ such that for every $||\theta - \theta'||_{\infty}<\delta$ and every $J \in \mathbb{N}$ we have that
-    $$||\nabla_\theta \hat\lik_J^1(\theta)(\omega)-\nabla_\theta \hat\lik_J^1(\theta')(\omega)||_\infty < \epsilon.$$
-
-    Then, by Arzela-Ascoli, there exists a uniformly convergent subsequence.
-We claim that there is only one subsequential limit.
-We know that as the gradient estimate is bounded by Assumption \ref{assump:local-bounded-derivative}, we can treat the gradient estimate at $\theta$ as a bounded functional of the particles.
-So by Theorem \ref{thm:mop-targeting} the sequence $(\nabla_\theta \hat\lik_J^1(\theta, \omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$, and there is therefore only one subsequential limit.
-The sequence therefore converges uniformly to its limit $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega).$ Therefore, with uniform convergence for the derivatives established, we can swap the limit and derivative and obtain, in conjunction with the strong consistency $\hat{\lik}_J^1(\theta, \omega) \stackrel{a.s.}{\to} \lik(\theta)$ in Theorem \ref{thm:mop-targeting}, that for almost every $\omega \in \Omega$,
-    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lik(\theta).$$
-    The result then follows by the continuous mapping theorem. 
-\end{proof}
-
-\subsection{MOP-$\alpha$ Bias and Variance}
+We bypass the issue of differentiating through a Monte Carlo algorithm with discontinuous resampling by turning it into a problem of differentiating through a simulator and a series of measurement density ratios.
+Although we defer the theoretical analysis of the resulting algorithm, MOP-$\alpha$, to Section \ref{sec:thms}, we highlight a few key results below.
+MOP-$\alpha$ targets the posterior and likelihood (Theorem \ref{thm:mop-targeting}), yields the estimators of \cite{poyiadjis11, scibior21, blei2018vsmc} as special cases when differentiated (Theorem \ref{thm:mop-functional-forms}) where MOP-$1$ in particular is consistent for the score (Theorem \ref{thm:mop-grad-consistency}), has rates for its bias and variance under different choices of $\alpha$ (Theorem \ref{thm:mop-biasvar}), and enjoys a linear rate of convergence for gradient descent with the resulting gradient estimate (Theorem \ref{thm:mop-convergence}).
 
@@
 
-One can think of MOP-$\alpha$ as taking a likelihood estimate at $\phi$, reweighting it to instead target the likelihood at $\theta$, and then taking the derivative with respect to $\theta$ of the above estimate.
-But if $\theta$ is evaluated at $\phi$, the particles at $\theta$ and $\phi$ coincide, and one only needs to run one particle filter at $\theta=\phi$, setting the particles at $\phi$ to be copies of the particles at $\theta$ where gradients don't propagate (and vice-versa for the measurement densities $g_{n,j}^\phi$ and conditional likelihood estimate $L_n^\phi$).
-This is done algorithmically through the \texttt{stop\_gradient()} function in \texttt{JAX}, and therefore provides a mathematical justification for the "stop-gradient trick" presented in \cite{scibior21}.
+%One can think of MOP-$\alpha$ as taking a likelihood estimate at $\phi$, reweighting it to instead target the likelihood at $\theta$, and then taking the derivative with respect to $\theta$ of the above estimate.
+
+If $\theta$ is evaluated at $\phi$, the particles at $\theta$ and $\phi$ coincide.
+One then only needs to run one particle filter at $\theta=\phi$, setting the particles at $\phi$ to be copies of the particles at $\theta$ where gradients don't propagate.
+This is done algorithmically through the \texttt{stop\_gradient()} function in \texttt{JAX}, providing a mathematical justification for the ``stop-gradient trick" of \cite{scibior21}.
 
 \subsection{Optimization}
-This still leaves us with the question of designing an effective procedure for likelihood maximization with these new tools.
-We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that amounts to warm-starting gradient descent that uses the gradient estimate given by MOP-$\alpha$ (or some other reasonable first or second order procedure) with the output of an initial search of IF2.
 
-That is, a coarse solution is found by running IF2 with an aggressive random walk standard deviation (learning rate), until the search stalls to an initial "convergence" to what is hopefully a neighborhood of the MLE.
-We can then use gradient descent, or other first or second order gradient methods, to refine this coarse solution.
-We do this because IF2 has been empirically shown to converge quickly to a neighborhood of the MLE, and by switching to gradient descent with MOP-$\alpha$ estimates, one can bypass the difficulty that IF2 has with optimizing the last few units of log-likelihood.
-Conversely, gradient descent (and other first or second order gradient methods) can get stuck in saddle points and less desirable local minima when the likelihood is nonconvex.
-Warm-starting gradient methods with IF2 lets us get close enough to the MLE for gradient methods to shine, as in most cases regularity conditions ensure that the likelihood surface is well-behaved enough in this neighborhood that the aforementioned issues with saddle points and local minima are alleviated.
+This still leaves us with the question of designing an effective procedure for likelihood maximization with MOP-$\alpha$.
+We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that runs a few iterations of IF2 to warm-start an iterative first or second-order method that uses the MOP-$\alpha$ gradient estimate.
+This leverages IF2's quick empirical convergence to a neighborhood of the MLE, overcoming the tendency of gradient methods to get stuck in saddle points and local minima.
+Conversely, switching to gradient ascent with MOP-$\alpha$ score estimates lets one bypass the difficulty that IF2 has with optimizing the last few units of log-likelihood.
 Combining these two methods in this way lets us enjoy the best of both worlds.
+\kevin{source for this}
+
+%The warm start allows gradient methods to shine, as regularity conditions ensure that the likelihood surface is well-behaved enough in this neighborhood that the aforementioned issues with saddle points and local minima are alleviated.
 
 \begin{algorithm}[H]
-	\caption{Iterated Filtering with Automatic Differentiation}
+	\caption{IFAD}
     \label{alg:ifad}
-	    \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $p(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta),$ IF2 cooling schedule $\eta_t$, MOP-$\alpha$ discounting parameter $\alpha$.\newline
-		\textbf{Initialize} $\theta_0$, $t=0.$\newline
+	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 cooling schedule $\eta_t$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $t=0.$\newline
         Run IF2 until initial "convergence" under cooling schedule $\eta_t$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_t := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
@@
 		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_t).$ \newline
-		\hspace*{4mm} Obtain $g(\theta_t) = \nabla_{\theta_t} (-\loglik(\theta_t))$, $H(\theta_t)$ s.t. $\lambda_{\min}(H) \geq c$. \newline
+		\hspace*{4mm} Obtain $g(\theta_t) = \nabla_{\theta_t} (-\hat\loglik(\theta_t))$, $H(\theta_t)$ s.t. $\lambda_{\min}(H) \geq c$. \newline
 		\hspace*{4mm} Update $\theta_{t+1} := \theta_t - \eta (H(\theta_t))^{-1} g(\theta_t)$, $t:=t+1.$ \newline
@@
 
-\subsection{Runtime}
-
-Each iteration of IF2 amounts to one iteration of the particle filter, with $O(NJ)$ time complexity.
-The initial "convergence" happens fairly quickly in practice.
-In the case of the Dhaka cholera model of \cite{king08}, when a geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations of IF2.
-In comparison, often 100 or 200 iterations of IF2 are used for an initial global search.
-\cite{ionides15}, for example, use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{Lee_haiticholera}.
-
-In line with the cheap gradient principle of \cite{kakade2019provably}, getting a gradient estimate from MOP-$\alpha$ takes no more than 6 times that of the runtime of the particle filter (in practice, the multiplier is 3.75).
-This therefore has $O(NJ)$ time complexity as well, obtaining a gradient estimate with runtime linear in the number of particles.
-
-Re-implementing the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, even with a simulator written in Python.
-This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter (with a simulator written in C++) in the \texttt{pomp} package of \cite{king16}.
-
 \section{Application to the Cholera Model Of \cite{king08}}
 
-We use the Dhaka cholera model of \cite{king08} to benchmark the performance of our method.
-This was also used or modified in \cite{ionides15, wood16, wycoff2024voronoi} for the same purpose.
-This is a stochastic SIR compartmental model with transition and measurement uncertainty, where the population at time $t$, $H_t$, is divided into the susceptible compartment $S_t$, infected compartment $I_t$, and three ($k=3$) recovered compartments $R^1_t, ..., R^k_t$ denoting varying degrees of cholera immunity.
+We use the Dhaka cholera model of \cite{king08} to benchmark (also used or modified in \cite{ionides15, wood16, wycoff2024voronoi} for the same purpose) the performance of our method.
+This is a stochastic SIR compartmental model where the population at time $t$, $H_t$, is divided into the susceptible compartment $S_t$, infected compartment $I_t$, and three ($k=3$) recovered compartments $R^1_t, ..., R^k_t$ denoting varying degrees of cholera immunity.
 We write $M_t$ for the cholera deaths in each month.
-As in \cite{king08, ionides15}, the \textbf{transition dynamics} follow this series of stochastic differential equations:
+As in \cite{king08, ionides15}, the \textbf{transition dynamics} follow this series of SDEs:
+\vspace*{-1mm}
 \begin{align*}
-    d S_t&=(k \epsilon R^k_t+\delta(S_t-H_t)-\lambda_t S_t) d t+d H_t-\frac{\sigma S_t I_t}{H_t} d B_t, \\[-2mm] 
-    d I_t&=\left(\lambda_t S_t-(m+\delta+\gamma) I_t\right) d t+\frac{\sigma S_t I_t}{H_t} d B, \\[-2mm]
-    d R^1_t&=\left(\gamma I-(k \epsilon+\delta) R^1\right) d t, \\ 
-    d R^k_t&=\left(k \epsilon R^{k-1}_t-(k \epsilon+\delta) R^k_t\right) d t,
+    d S_t&=(k \epsilon R^k_t+\delta(S_t-H_t)-\lambda_t S_t) d t+d H_t-({\sigma S_t I_t}/{H_t}) d B_t, \\
+    d I_t&=\left(\lambda_t S_t-(m+\delta+\gamma) I_t\right) d t+({\sigma S_t I_t}/{H_t}) d B_t, \\
+    d R^1_t&=(\gamma I-(k \epsilon+\delta) R^1) d t,..., 
+    d R^k_t=(k \epsilon R^{k-1}_t\hspace{-3mm}-(k \epsilon+\delta) R^k_t) d t,
+    \vspace*{-2mm}
 \end{align*}
 
-where $B_t$ represents Brownian motion, $m$ is the rate at which infected individuals succumb to cholera, $\gamma$ is the rate at which infected individuals recover, $1/\epsilon$ is the mean duration of immunity, $\sigma$ represents the standard deviation of perturbations to the force of infection, and $\delta=0.02$ is the death rate.
-
+with Brownian motion $B_t$, cholera death rate $m$, recovery rate $\gamma$, mean immunity duration $1/\epsilon$, standard deviation of the force of infection $\sigma$, and population death rate $\delta=0.02$.
 The \textbf{force of infection}, $\lambda_t$, is modeled by splines $(s_j)_{j=1}^6$
-\begin{equation}
-    \lambda_t=\exp\hspace{-1mm}\left(\hspace{-.5mm}\beta_{\text{trend}}(t-t_0)+\hspace{-1mm}\sum_{j=1}^{6} \beta_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm}\frac{I_t}{P_t} + \exp \left(\sum_{j=1}^{6} \omega_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm},
-\end{equation}
+\vspace*{-2mm}
+\begin{equation*}
+    \lambda_t=\exp\hspace{-1mm}\left(\hspace{-.5mm}\beta_{\text{trend}}(t-t_0)+\hspace{-1mm}\sum_{j=1}^{6} \beta_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm}\frac{I_t}{P_t} + \exp\hspace{-1mm} \left(\sum_{j=1}^{6} \omega_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm},
+    \vspace*{-2mm}
+\end{equation*}
 where the coefficients $(\beta_j)_{j=1}^6$ model seasonality in the force of infection, $\beta_{\text{trend}}$ models the trend in the force of infection, and the $\omega_j$ represent seasonality of a non-human environmental reservoir of disease.
 The \textbf{measurement model} for observed monthly cholera deaths is given by
-\begin{equation}
-    Y_n \sim \gN(M_n, \tau^2M_n^2),
-\end{equation}
+    $Y_n \sim \gN(M_n, \tau^2M_n^2)$,
 where $M_n$ is the true number of cholera deaths in that month.
@@
 \end{figure}
-\kevin{Change the optim graph to include median search and best search with dashes as linestyle, remove the shaded areas. }
+
+\section{Theoretical Analysis of MOP-$\alpha$}
+\label{sec:thms}
+
+Here, we will show that MOP-$\alpha$ targets the posterior and likelihood, yields the estimators of \cite{poyiadjis11, scibior21, blei2018vsmc} as special cases when differentiated, and characterize rates for its bias and variance under different choices of $\alpha$.
+To do so, we require the following assumptions:
+
+\begin{enumerate}[label=(A\arabic*),itemsep=-1.2ex] 
+    \item \textbf{Continuity of the Likelihood.} $\ell(\theta)$ has more than two continuous derivatives in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.
+\label{assump:conti-lik}
+    \item \textbf{Bounded Process Model.} There exist $\underbar{M}, \bar{M}$ such that $0 < \underbar{M} \leq f_{X_n|X_{n-1}}(x_n | x_{n-1};\theta) \leq \bar{M} < \infty$.
+\label{assump:bounded-process}
+    \item \textbf{Bounded Measurement Model.} There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G}<\infty$ and $G'(\theta)$ so $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty \leq G'(\theta)< \infty$.
+\label{assump:bounded-measurement}
+    \item \textbf{Bounded Gradient Estimates.} There are functions $G(\theta), H(\theta): \Theta \to [0,\infty)$ uniformly bounded by $G', H'$, so the MOP-$\alpha$ gradient and Hessian estimates at $\theta=\phi$ are almost surely bounded by $G(\theta)$ and $H(\theta)$ for all $\alpha$.
+\label{assump:local-bounded-derivative}
+    \item \textbf{Differentiability of Density Ratios and Simulator.} The measurement density $f_{Y_n|X_n}(y_n^*|x; \theta)$ and simulator have more than two continuous derivatives in $\theta$.
+\label{assump:diff-meas-and-sim}
+\end{enumerate}
+
+The fact that the likelihood estimate yielded by MOP-$\alpha$ has more than two continuous derivatives in $\theta$ follows the construction of the likelihood estimate in equations \ref{eq:mop-conditional-likelihood} and \ref{eq:weighting-scheme}, as well as Assumptions \ref{assump:bounded-measurement} and \ref{assump:diff-meas-and-sim}.
+
+\subsection{MOP-$\alpha$ Targets the Posterior}
+
+We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood.
+While the result is presented here as specific to MOP-$\alpha$, the full proof relies on a handful of technical lemmas presented in the supplementary material that may be useful in their own right, and Theorem \ref{thm:mop-targeting} can also serve as a more general result for other particle filters with ``off-policy'' resampling.
+
+\begin{thm}[MOP-$\alpha$ Targets the Posterior and Likelihood]
+    \label{thm:mop-targeting}
+    When $\alpha=1$ or $\phi=\theta$, MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood.
+That is, for $\pi_n(\theta)=f_{X_{1:n}|Y_{1:n} ; \theta}$ and any measurable and bounded functional $h$ and for $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\prod_{n=1}^N L_n^{B,\theta,\alpha}$, it holds that
+    \vspace*{-2.5mm}
+    \begin{equation}
+        \frac{\sum_{j=1}^J h(x_{n,j}^{A,F, \theta}) w_{n,j}^{A,F,\theta}}{\sum_{j=1}^J w_{n,j}^{A,F,\theta}} \stackrel{a.s.}{\to} E_{\pi_n(\theta)} h, \;\; \hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).
+    \vspace*{-2.5mm}
+    \end{equation}
+\end{thm}
+
+\begin{proof}
+    We provide a proof sketch here, deferring most of the details, and discussion of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$, to the supplementary information.
+
+    When $\theta=\phi$, regardless of the value of $\alpha$, the ratio ${g_{n,j}^\theta}/{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter.
+
+    When $\alpha=1$ and $\theta\neq\phi,$ suppose inductively that $\{(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j})\}_{j=1}^J$ targets $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
+    It can then be shown that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j})\}_{j=1}^J$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$, that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} )\}_{j=1}^J$ targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$, and that the resampling rule given by $(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}) = (X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}/ g^\phi_{n,a(j)})$ where $a(\cdot)$ is the ancestor function,
+    with resampling weights proportional to $g^\phi_{n,j}$, also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+
+    If the likelihood is estimated with the before-resampling conditional likelihoods $\hat\lik(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, the strong consistency is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+\end{proof}
+
+\subsection{MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, blei2018vsmc}}
+
+\cite{scibior21} show that the estimate of \cite{blei2018vsmc} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through applying a``stop-gradient trick" to the particle filter.
+It turns out that both of these, when applied on the bootstrap filter, correspond to special cases of MOP-$\alpha$.
+
+\begin{thm}[MOP-$0$ and MOP-$1$ Functional Forms]
+    \label{thm:mop-functional-forms}
+    Write $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$ and we use the after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$.
+Then, when $\alpha=0$,
+    \vspace*{-2.5mm}
+    \begin{equation}
+        \nabla_\theta \hat\ell^0(\theta) 
+        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
+        \vspace*{-2.5mm}
+    \end{equation}
+    yielding \cite{blei2018vsmc} on the bootstrap filter. When $\alpha=1$,
+    \vspace*{-2.5mm}
+    \begin{equation}
+        \nabla_\theta \hat{\ell}^1(\theta) 
+        = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
+    \vspace*{-2.5mm}
+    \end{equation}
+    yielding the estimator of \cite{poyiadjis11, scibior21} with the bootstrap filter.
+\end{thm}
+
+We defer the proof to the supplementary material.
+It relies on a useful decomposition of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$ that yields a telescoping product in the MOP-$1$ case.
+Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting that when $\theta=\phi$, that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result.
+
+This further illustrates how $\alpha$ dictates the "memory" of the gradient estimate.
+As \cite{scibior2021dpf} remark, the MOP-$0$ estimator of \cite{blei2018vsmc} only considers single-step quantities, and is "memoryless" beyond a single step.
+This is in contrast to the case when $\alpha=1,$ as that estimate of \cite{poyiadjis11} only considers the surviving particles at time $N$, fully tracking dependencies over time.
+
+\subsection{MOP-$1$ Is Consistent for the Score}
+
+Despite showing that the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} use $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}, X_{1:N}}\left(y_{1:N}^* , x_{1:n,j}^{A, F,\theta}; \theta\right)$ in general.
+The bootstrap and general estimates have different functional forms, and are not the same a-priori.
+As such we directly show the consistency of the MOP-$1$ gradient estimate below.
+
+\begin{thm}[Consistency of MOP-$1$ Gradient Estimate]
+    The gradient estimate of MOP-$\alpha$ when $\alpha=1$, $\theta=\phi$ is strongly consistent for the score: $\nabla_\theta \hat\ell_J^1(\theta) \stackrel{a.s.}{\to} \nabla_\theta \ell(\theta)$ as $J \to \infty$.
+    \label{thm:mop-grad-consistency}
+\end{thm}
+\begin{proof}
+    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point we wish to evaluate the gradient at.
+The sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$, by Assumption \ref{assump:local-bounded-derivative}.
+Again by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J^1(\theta)(\omega)|_{\theta=\theta'}$ is also bounded by $H'$ for almost every $\omega\in \Omega$ and every $\theta'\in \Theta$.
+So $(\nabla_\theta \hat\lik_J^1(\theta)( \omega))_{J \in \mathbb{N}}$ is uniformly Lipschitz, and therefore uniformly equicontinuous for almost every $\omega \in \Omega$.
+
+    By Arzela-Ascoli, there is a uniformly convergent subsequence.
+But there is only one subsequential limit, as we can treat the gradient estimate at $\theta$ as a bounded functional of the particles by Assumption \ref{assump:local-bounded-derivative}, allowing us to apply Theorem \ref{thm:mop-targeting} to see that the sequence $(\nabla_\theta \hat\lik_J^1(\theta, \omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$.
+So the whole sequence must converge uniformly to $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega).$
+
+    With uniform convergence for the derivatives established, we can swap the limit and derivative and obtain, in conjunction with the strong consistency $\hat{\lik}_J^1(\theta, \omega) \stackrel{a.s.}{\to} \lik(\theta)$ in Theorem \ref{thm:mop-targeting}, that for almost every $\omega \in \Omega$,
+    $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lik(\theta).$
+    The result then follows by the continuous mapping theorem. 
+\end{proof}
+
+\subsection{MOP-$\alpha$ Bias and Variance}
+\begin{thm}
+    \label{thm:mop-biasvar}
+\end{thm}
+
+\subsection{Convergence Rates}
+\begin{thm}
+    \label{thm:mop-convergence}
+\end{thm}
+
+\section{Computational Considerations}
+
+\subsection{Runtime}
+
+Each iteration of IF2 amounts to one iteration of the particle filter, with $O(NJ)$ time complexity.
+The initial "convergence" happens fairly quickly in practice.
+In the case of the Dhaka cholera model of \cite{king08}, when a geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations of IF2.
+In comparison, often 100 or 200 iterations of IF2 are used for an initial global search.
+\cite{ionides15}, for example, use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{Lee_haiticholera}.
+
+In line with the cheap gradient principle of \cite{kakade2019provably}, getting a gradient estimate from MOP-$\alpha$ takes no more than 6 times that of the runtime of the particle filter (in practice, the multiplier is 3.75).
+This therefore has $O(NJ)$ time complexity as well, obtaining a gradient estimate with runtime linear in the number of particles.
+
+\subsection{Python Library}
+
+Re-implementing the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, even with a simulator written in Python.
+This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter (with a simulator written in C++) in the \texttt{pomp} package of \cite{king16}.
+
+\kevin{Write about new open source library that brings pomp to python with AD.
+There are implementations in tfp under the jax substrate, but these are not plug and play -- they require you to supply the transition densities.
+}
 
@@
 We note that it is possible to work without a differentiable simulator, but one would then require access to the transition densities.
+
+Our gradient estimate can be used for variational inference to approximate the posterior distribution over parameters.
 
@@
 
-% \subsection*{Supporting Information Appendix (SI)}
+%\subsection*{Supporting Information Appendix (SI)}
 

### 2024-04-18 paper/si.tex
Updates from Overleaf

\documentclass[9pt,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnassupportinginfo}

\begin{document}

\title{Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation}
\author{Kevin Tan, Giles Hooker, Edward L. Ionides}
\correspondingauthor{Corresponding Author name.\\E-mail: kevtan@wharton.upenn.edu}

\maketitle

%% Comment out or remove this line before generating final copy for submission; this will also remove the warning re: "Consecutive odd pages found".
%\instructionspage

%% Adds the main heading for the SI text. Comment out this line if you do not have any supporting information text.
\SItext

\subsection*{Subhead}

\begin{defn}[Targeting]
    \label{defn:targeting}
    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any measurable and bounded functional $h$, $E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}$ 
    A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if for any measurable and bounded functional $h$
    ${\sum_{j=1}^J h(X^J_j) w^J_j}/{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X))$ as $J\to\infty$
\end{defn}

\begin{lem}[Change of Particle Measure]
    \label{lem:change-measure-proper-weights}
    Assume $\{(\tilde X_j^J,U_j^J)\}_{j=1}^J$ targets $f_X$, and that $\{(Y_j^J,V_j^J)\}_{j=1}^J$ is a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is drawn w.p. $\pi^J_j$. Write, where $a$ is the ancestor function,
    $(Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big)
    $. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j)\}_{j=1}^J$ targets $f_X$.
\end{lem}
\begin{lem}[Particle Marginals]
    \label{lem:marginal-proper-weights}
    Suppose that $\{(\tilde X_j^J,U_j^J)\}_{j=1}^J$ targets $f_X$. Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$, where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$. Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}_{j=1}^J$ targets $f_Z$.
\end{lem}
\begin{lem}[Particle Posteriors]
    \label{lem:posterior-proper-weights}
    Suppose that $\{(X_j^J,u_j^J)\}_{j=1}^J$ targets $f_X$, and write $(X^{\prime J}_j,u^{\prime J}_j) = \big(X_j^J,u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$. Then, if $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,u^{\prime J}_j)\}_{j=1}^J$ targets $f_{X|Z}(\cdot | z^*)$.
\end{lem}

Whereas if we use the after-resampling conditional likelihoods $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$, the numerator of $L_n^{B,\theta,\alpha}$ is
\begin{equation}
    \sum_{j=1}^J g_{n,j}^\theta w_{n, j}^{P, \theta} = L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
\end{equation}
Using Lemma \ref{lem:change-measure-proper-weights}, we resample the corrected weights $w^{P,\theta}_{n,j}{g^\theta_{n,j}}/{g^\phi_{n,j}} $ according to probabilities ${g^\phi_{n,j}}/{L_n^\phi}$ to see this is properly estimated by $L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},$
from which we obtain the estimate $L_n^{B,\theta,\alpha}$. The strong consistency then follows as the likelihood is the product of the conditional likelihoods.

\begin{thm}[Consistency of MOP-$1$ Gradient Estimate]
    The gradient estimate of MOP-$\alpha$ when $\alpha=1$, $\theta=\phi$ is strongly consistent for the score. $\nabla_\theta \hat\ell_J^1(\theta) \stackrel{a.s.}{\to} \nabla_\theta \ell(\theta)$ as $J \to \infty$.
    \label{thm:mop-grad-consistency}
\end{thm}
\begin{proof}
    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point we wish to evaluate the gradient at. The sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$, by Assumption \ref{assump:local-bounded-derivative}. We claim that the sequence is also uniformly equicontinuous. To see this, by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J^1(\theta)(\omega)|_{\theta=\theta'}$ is also bounded by $H'$ for almost every $\omega\in \Omega$ and every $\theta'\in \Theta$. A set of functions with derivatives bounded by the same constant is uniformly Lipschitz, and therefore uniformly equicontinuous. So the sequence $(\nabla_\theta \hat\lik_J^1(\theta)( \omega))_{J \in \mathbb{N}}$ is uniformly equicontinuous for almost every $\omega \in \Omega$. 
    Explicitly, for almost every $\omega \in \Omega$ and every $\epsilon>0$, there exists some $\delta(\omega)>0$ such that for every $||\theta - \theta'||_{\infty}<\delta$ and every $J \in \mathbb{N}$ we have that
    $$||\nabla_\theta \hat\lik_J^1(\theta)(\omega)-\nabla_\theta \hat\lik_J^1(\theta')(\omega)||_\infty < \epsilon.$$

    Then, by Arzela-Ascoli, there exists a uniformly convergent subsequence. We claim that there is only one subsequential limit. We know that as the gradient estimate is bounded by Assumption \ref{assump:local-bounded-derivative}, we can treat the gradient estimate at $\theta$ as a bounded functional of the particles. So by Theorem \ref{thm:mop-targeting} the sequence $(\nabla_\theta \hat\lik_J^1(\theta, \omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$, and there is therefore only one subsequential limit. The sequence therefore converges uniformly to its limit $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega).$ Therefore, with uniform convergence for the derivatives established, we can swap the limit and derivative and obtain, in conjunction with the strong consistency $\hat{\lik}_J^1(\theta, \omega) \stackrel{a.s.}{\to} \lik(\theta)$ in Theorem \ref{thm:mop-targeting}, that for almost every $\omega \in \Omega$, 
    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lik(\theta).$$
    The result then follows by the continuous mapping theorem. 
\end{proof}

\section*{Heading}
\subsection*{Subhead}
Type or paste text here. You may break this section up into subheads as needed (e.g., one section on ``Materials'' and one on ``Methods'').

\subsection*{Materials}
Add a materials subsection if you need to.

\subsection*{Methods}
Add a methods subsection if you need to.

%%% Each figure should be on its own page
\begin{figure}
\centering
\includegraphics[width=\textwidth]{example-image}
\caption{First figure}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{frog}
\caption{Second figure}
\end{figure}

\begin{table}\centering
\caption{This is a table}

\begin{tabular}{lrrr}
Species & CBS & CV & G3 \\
\midrule
1. Acetaldehyde & 0.0 & 0.0 & 0.0 \\
2. Vinyl alcohol & 9.1 & 9.6 & 13.5 \\
3. Hydroxyethylidene & 50.8 & 51.2 & 54.0\\
\bottomrule
\end{tabular}
\end{table}

%%% Add this line AFTER all your figures and tables
\FloatBarrier

\movie{Type legend for the movie here.}

\movie{Type legend for the other movie here. Adding longer text to show what happens, to decide on alignment and/or indentations.}

\movie{A third movie, just for kicks.}

\dataset{dataset_one.txt}{Type or paste legend here.}

\dataset{dataset_two.txt}{Type or paste legend here. Adding longer text to show what happens, to decide on alignment and/or indentations for multi-line or paragraph captions.}

\bibliography{pnas-sample}

\end{document}

### 2024-05-01 mop/draft.tex
Updates from Overleaf

@@
 
-\bibliography{bib-ifad,bib-ref}
+\bibliography{paper/bib-ifad, paper/bib-ref}
 \bibliographystyle{apalike}
@@
 \end{equation*}
-
 yielding the gradient estimators of \cite{poyiadjis11, scibior2021dpf} when applied to the bootstrap filter. 
@@
 To bound this, we use the fact that $$\E[(\nabla_\theta\hat\ell_n(\theta))^4] = \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4]+\E[(\nabla_\theta\hat\ell_n(\theta))^2]^2$$
-alongside Theorem 2 of \cite{delMoral03}, which shows that
-$$\E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4] \lesssim \frac{c_d^4G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^2] \lesssim \frac{c_d^2G'(\theta)^2}{J},$$
-where $c_d$ is a term depending only on the dimension of the state space $d$ (but where the dependence may be exponential), to find that 
-$$\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{c_d^4G'(\theta)^4}{J^2}+\left(\frac{c_d^2G'(\theta)^2}{J}\right)^2} = \frac{c_d^2G'(\theta)^2}{J},$$
+alongside Lemma 2 of \cite{karjalainen2023}, which shows that
+$$\E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4] \lesssim \frac{G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^2] \lesssim \frac{G'(\theta)^2}{J},$$
+to find that 
+$$\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{G'(\theta)^4}{J^2}+\left(\frac{G'(\theta)^2}{J}\right)^2} = \frac{G'(\theta)^2}{J},$$
 and conclude that 
-$$\text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{c_d^2G'(\theta)^2}{J}.$$
+$$\text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{G'(\theta)^2}{J}.$$
 
@@
     \text{Var}(\nabla_\theta \hat\ell(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta))\\
-    &\leq \frac{Nc_d^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\frac{c_d^2G'(\theta)^2}{J} \\
-    &\leq \frac{c_d^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\right) \\
-    &\lesssim \frac{c_d^2G'(\theta)^2N}{J}.
+    &\leq \frac{NG'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\frac{G'(\theta)^2}{J} \\
+    &\leq \frac{G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\right) \\
+    &\lesssim \frac{G'(\theta)^2N}{J}.
 \end{align*}
@@
 Note that $$\hat s_{n,k}(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)$$ is a function bounded by $C\frac{G'(\theta)(1-\alpha^k)}{1-\alpha} =: CG'(\theta)r$ in each coordinate for some constant $C$ (by Assumption \ref{assump:bounded-measurement}) that depends only on $k$ timesteps, and not $n$.
-We can therefore invoke Theorem 2 of \cite{delMoral03} to bound the $L_2$ error of this functional from its expectation under the posterior by $CrG'(\theta)c_dJ^{-1/2}d(k)$ where $d(k)$ is a term linear in $k$ and $c_d$ is a term possibly exponential in the dimension of the state space $d$.
-That is, from Theorem 2 of \cite{delMoral03} we have that
-$$\E\left[||\hat s_{n,k}(\theta) - \E_{\pi}\hat s_{n,k}(\theta)||_\infty^2\right]^{1/2} \leq \frac{Cr^2G'(\theta)c_dd(k)}{\sqrt{J}}.$$
-
-By the bias-variance decomposition, this in turn bounds the variance at a single timestep by $C^2r^2G'(\theta)^2c_d^2J^{-1}d(k)^2$. Concretely, we have that
-$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \lesssim \frac{c_d^2r^2k^2G'(\theta)^2}{J}.$$
+We can therefore invoke Lemma 2 of \cite{karjalainen2023} to bound the $L_2$ error of each of the $k$ (from time $n-k$ to time $n$) functionals from its expectation under the posterior by $CrG'(\theta)J^{-1/2}k$.
+That is, we have that
+$$\E\left[||\hat s_{n,k}(\theta) - \E_{\pi}\hat s_{n,k}(\theta)||_\infty^2\right]^{1/2} \leq \frac{Cr^2G'(\theta) k}{\sqrt{J}}.$$
+
+By the bias-variance decomposition, this in turn bounds the variance at a single timestep by $C^2r^2G'(\theta)^2J^{-1}k^2$. Concretely, we have that
+$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \lesssim \frac{r^2k^2G'(\theta)^2}{J}.$$
 
@@
     &= \sum_{n=1}^N\text{Var}\left(\hat s_{n,k}(\theta)\right)+ 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \\
-    &\lesssim \frac{c_d^2r^2k^2G'(\theta)^2N}{J} + 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right).
+    &\lesssim \frac{r^2k^2G'(\theta)^2N}{J} + 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right).
 \end{align*}
@@
 
-Therefore, by Davydov's inequality and Theorem 2 of \cite{delMoral03}, and noting that $\hat s_{n,k}(\theta) \leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, by a similar argument to the MOP-0 case, we find that
-
-$$\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{c_d^2r^2G'(\theta)^2}{J}.$$
+Therefore, by Davydov's inequality and Lemma 2 of \cite{karjalainen2023}, and noting that $\hat s_{n,k}(\theta) \leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, by a similar argument to the MOP-0 case, we find that
+
+$$\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J}.$$
 
@@
 To bound this, we use the fact that $$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4] = \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4]+\E[(\nabla_\theta\hat s_{n,k}(\theta))^2]^2$$
-alongside Theorem 2 of \cite{delMoral03}, which shows that
-$$\E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4] \lesssim \frac{c_d^4r^4G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^2] \lesssim \frac{c_d^2r^2G'(\theta)^2}{J},$$
-where $c_d$ is a term depending only on the dimension of the state space $d$ (but where the dependence may be exponential), to find that 
-$$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{c_d^4r^4G'(\theta)^4}{J^2}+\left(\frac{c_d^2r^2G'(\theta)^2}{J}\right)^2} = \frac{c_d^2r^2G'(\theta)^2}{J},$$
+alongside Lemma 2 of \cite{karjalainen2023}, which shows that
+$$\E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4] \lesssim \frac{r^4G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^2] \lesssim \frac{r^2G'(\theta)^2}{J},$$
+to find that 
+$$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{r^4G'(\theta)^4}{J^2}+\left(\frac{r^2G'(\theta)^2}{J}\right)^2} = \frac{r^2G'(\theta)^2}{J},$$
 and conclude that 
-$$\text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{c_d^2r^2G'(\theta)^2}{J}.$$
+$$\text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J}.$$
 
@@
     \text{Var}(\nabla_\theta \hat s_{k}(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat s_{n,k}(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat s_{m,k}(\theta),\nabla_\theta\hat s_{n,k}(\theta))\\
-    &\leq \frac{Nc_d^2r^2k^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\frac{c_d^2r^2G'(\theta)^2}{J} \\
-    &\leq \frac{c_d^2r^2k^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\right) \\
-    &\lesssim \frac{c_d^2k^2r^2G'(\theta)^2N}{J}.
+    &\leq \frac{Nr^2k^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J} \\
+    &\leq \frac{r^2k^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\right) \\
+    &\lesssim \frac{k^2r^2G'(\theta)^2N}{J}.
 \end{align*}
@@
 Taking expectations, we get
-$$\text{Var}(X) \leq \text{Var}(Y) \leq \text{Var}(Y) + 8M^2.$$
+$$\text{Var}(Y) \leq \text{Var}(X) + 8M^2.$$
 
@@
 that the variance of MOP-$\alpha$ proper is bounded by
-$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \frac{c_d^2k^2r^2G'(\theta)^2N}{J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2.$$
+$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \frac{k^2r^2G'(\theta)^2N}{J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2.$$
 
 As the above holds for any $k \leq N$,
-$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \min_{k\leq N} \left(\frac{c_d^2k^2G'(\theta)^2N}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right).$$
+$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \min_{k\leq N} \left(\frac{k^2G'(\theta)^2N}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right).$$
 
@@
 
+\subsection{An Improved Bound by Mixing}
+
+We decompose
+$$\E||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty \leq \E||\nabla_\theta\ell_\theta - \nabla_\theta\hat s_k^1(\theta)||_\infty + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty.$$
+
+The broad idea is that the first term can be controlled by a mixing argument, the second term garners $O(Nk^2)$ error, and the third term is $O\left(\frac{N}{1-\alpha}\right)$.
+It is unclear whether the bias can be reduced further -- the gradient is a sum of $N$ terms, and is therefore $O(N)$ itself.
+
+The first term -- bound ground truth of conditional scores from particle approximation conditional on correct filtering distribution at time $n-k$, bound that from particle approximation conditional on arbitrary filtering distribution at time $n-k$.
+First term sums up $k$ terms, $O(Nk^2)$ error, second term bounded by $k$ applications of the Markov kernel, so just apply forgetting of the particle filter result.
+
+But you can get an MSE bound like this! Simply decompose
+$$||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 \leq ||\nabla_\theta\ell_\theta - \nabla_\theta\hat s_k^1(\theta)||_2^2 + ||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2 + ||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_2^2,$$
+and observe that 
+$$||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty =  \left\lVert\sum_{n=1}^N \nabla_\theta \hat\ell_n^\alpha(\theta) - \sum_{n=1}^N \nabla_\theta \ell_n(\theta)\right\rVert_\infty \leq \sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_\infty.$$
+
 \section{Finite-Particle Analysis of Gradient Methods with Algorithm \ref{alg:ifad}}

### 2024-05-01 paper/ms.tex
Updates from Overleaf

@@
 
+\kevin{Dynamically save to a tex file and load from there}
+
 \begin{table}[h!]
@@
 
+\kevin{Increase size of font in legend, notch rest of font up by one point. Make figure aspect ratio to be taller than wider.}
 \begin{figure}[htbp!]
-    \centering
-    \includegraphics[width=\textwidth/5]{../imgs/095/pairs.png}
-    \includegraphics[width=\textwidth/5]{../imgs/095/qq.png}
+    \includegraphics[width=\textwidth/\real{4.2}]{../imgs/095/pairs.png}
+    \includegraphics[width=\textwidth/\real{4.2}]{../imgs/095/qq.png}
     \caption{Scatterplots depicting the performance of IFAD against that of IF2.
@@
 \begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth/4]{../imgs/095/boxplot.png}
-    \includegraphics[width=\textwidth/5]{../imgs/095/optim.png}
+    \includegraphics[width=\textwidth/\real{4.2}]{../imgs/095/boxplot.png}
+    \includegraphics[width=\textwidth/\real{4.2}]{../imgs/095/optim.png}
     \caption{\textbf{Left:} Boxplot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs, simulating the result of the common procedure of running a few searches and choosing the best one.
@@
 We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood.
-While the result is presented here as specific to MOP-$\alpha$, the full proof relies on a handful of technical lemmas presented in the supplementary material that may be useful in their own right, and Theorem \ref{thm:mop-targeting} can also serve as a more general result for other particle filters with ``off-policy'' resampling.
+While the result is presented here as specific to MOP-$\alpha$, the full proof relies on a handful of technical lemmas presented in the supplementary material that may be useful in their own right, and Theorem \ref{thm:mop-targeting} can also serve as a more general result as a strong law of large numbers for triangular arrays of particles with ``off-policy'' resampling.
 
@@
     \label{thm:mop-targeting}
-    When $\alpha=1$ or $\phi=\theta$, MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood.
+    When $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood.
 That is, for $\pi_n(\theta)=f_{X_{1:n}|Y_{1:n} ; \theta}$ and any measurable and bounded functional $h$ and for $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\prod_{n=1}^N L_n^{B,\theta,\alpha}$, it holds that
@@
     When $\alpha=1$ and $\theta\neq\phi,$ suppose inductively that $\{(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j})\}_{j=1}^J$ targets $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
-    It can then be shown that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j})\}_{j=1}^J$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$, that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} )\}_{j=1}^J$ targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$, and that the resampling rule given by $(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}) = (X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}/ g^\phi_{n,a(j)})$ where $a(\cdot)$ is the ancestor function,
-    with resampling weights proportional to $g^\phi_{n,j}$, also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+    It can then be shown that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j})\}_{j=1}^J$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$, that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} )\}_{j=1}^J$ targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$, and that weighting the particles by $(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}) = (X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}/ g^\phi_{n,a(j)})$ where $a(\cdot)$ is the ancestor function,
+    resampling with probabilities proportional to $g^\phi_{n,j}$, also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
 

### 2024-05-01 presentation/presentationmath.tex
Updates from Overleaf

--- 2024-03-06 presentation/presentation.tex
@@
 \title[AD for POMPs]{Automatic Differentiation Accelerates Inference for Partially Observed Stochastic Processes}
-\subtitle{\small{Inference for Partially Observed Structured Dynamic Systems}}
 \author[Kevin Tan]{Kevin Tan \\ \small{University of Pennsylvania, Department of Statistics and Data Science} \\\vspace{0.5cm} \small{Joint work with Giles Hooker and Edward Ionides}}
-\date[January 4, 2024]{IMS Asia Pacific Rim Meeting \\ January 4, 2024}
 
@@
     \begin{itemize}
-        \item Unobserved Markov process $\{X_t, t \geq t_0\}$, observations $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$.
+        \item Unobserved Markov process $\{X_t, t \geq t_0\}$, observations $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$. Write $X_n := X_{t_n}$.
         \item \pause Unknown parameter $\theta \in \Theta$, to be estimated.
+        \item $X_t \in \gX \subseteq \R^d$, $Y_n \in \gY$, $\theta \in \Theta \subseteq \R^p$.\pause
         \item \textbf{Notation:}
@@
 
+\begin{frame}{Partially Observed Markov Processes}
+
+    \begin{itemize}
+        \item Unobserved Markov process $\{X_t, t \geq t_0\}$, observations $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$.
+        \item \pause Problems to solve:
+        \begin{itemize}
+            \item \pause The \textbf{filtering problem}:
+            \begin{itemize}
+                \item \pause Find the density of the hidden state at time $n$, given all observations until time $n$.
+                \item \pause $f_{X_n|Y_{1:n}}(x_n|y_{1:n}^*;\theta)$.
+            \end{itemize}
+            \item \pause Related problem, estimating the \textbf{posterior} distribution of hidden states:
+            \begin{itemize}
+                \item \pause Find the density of the hidden states at timesteps $1$ to $n$, given all observations till time $n$. 
+                \item \pause $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*;\theta)$.
+            \end{itemize}
+        \end{itemize}
+    \end{itemize}
+\end{frame}
+
+\begin{frame}{The Bayes Filter}
+    \begin{enumerate}
+        \item Maintain a belief on the state at time $n$, $f_{X_{n}|Y_{1:n}}$. 
+        \item \pause Predict the next timestep, $x_{n+1}^P \sim f_{X_{n+1}|X_{n}}$.
+        \item \pause Observe observation at time $n+1$, $y_{n+1}^*$.
+        \item \pause Update belief on current state at time $n+1$, $f_{X_{n+1}|Y_{1:n+1}}$.
+    \end{enumerate}
+\end{frame}
+
 \begin{frame}{Particle Filters}
@@
             \centering
-            \includegraphics[scale=0.15]{imgs/pfilter.png}
+            \includegraphics[scale=0.125]{imgs/pfilter.png}
             \end{minipage}
@@
         \begin{itemize}
-            \item \pause Maintain belief on current state $x_n^F$. 
-            \item \pause Simulate forward to get $x_{n+1}^P$, observe observation $y_{n+1}^*$.
-            \item \pause Update $x_{n+1}^F$ accordingly.
+            \item \pause Maintain $J$ weighted particles, weighted Dirac approximation of the current state $(x_{n,j}^F)_{j=1}^J$. 
+            \item \pause Simulate each of the $J$ particles forward to get $x_{n+1,j}^P \sim f_{X_{n+1}|X_n}(\cdot | x_{n,j}^F; \theta)$.
+            \item \pause Observe observation $y_{n+1}^*$, update $(x_{n+1,j}^F)_{j=1}^J$ by resampling according to their likelihoods.
         \end{itemize}
     \end{itemize}
+\end{frame}
+
+\begin{frame}{Likelihood Estimate}
+    \begin{block}{Conditional Likelihood}
+        The conditional likelihood $L_n$ under $\theta$ is given by $f_{Y_n|Y_{1:n}}(y_{n}^*|y_{1:n-1};\theta)$.
+    \end{block}
+    \pause
+    \begin{block}{Likelihood}
+        \begin{align*}
+            \mathcal{L}(\theta)&=f_{Y_{1: N}}\left(y_{1: N}^* ; \theta\right)
+            = \prod_{n=1}^N L_n(\theta)\\
+            &=\prod_{n=1}^N \int f_{Y_n \mid X_n}\left(y_n^* \mid x_n ; \theta\right) f_{X_n \mid Y_{1: n-1}}\left(x_n \mid y_{1: n-1}^* ; \theta\right) d x_n.
+        \end{align*}
+    \pause
+    An unbiased estimate of the likelihood from the particle filter is then
+        $$\hat\lik(\theta) = \prod_{n=1}^N \frac{1}{J}\sum_{j=1}^J f_{Y_n|X_n}(y_n^*|x_{n,j}^P;\theta).$$
+    \end{block}
 \end{frame}
@@
 
+\end{frame}
+
+\begin{frame}{Maximum Likelihood for POMPs}
+    \begin{block}{Iterated Filtering}
+        \begin{itemize}
+            \item Augment the state space by assigning each particle $x_{n,j}$ its own set of parameters $\theta_{n,j}$. \pause
+            \item Parameters thought of as latent states that evolve according to Brownian motion on $\Theta$.  \pause
+            \item Yields a Monte Carlo approximation to the (extended) filtering distribution $f_{\theta_n,X_n|Y_{1:n}}(\theta_{n,j},x_{n,j}^F|y_{1:n}^*)$, given a Gaussian random walk prior on $\theta_{n,j} \sim \gN(\theta_{n-1,j},\sigma^2)$.
+\pause
+            \item Repeat procedure again and again, while shrinking $\sigma^2$.\pause
+            \item Yields convergence of the parameter swarm $\theta_{n,j}^{(m)}$ to the MLE. 
+        \end{itemize}
+    \end{block}
 \end{frame}
@@
     \begin{block}{Idea}
-        Don't differentiate $\hat\ell(\theta)$ directly, differentiate through a (suitably) reweighted bootstrap filter.
+        Don't differentiate $\hat\ell(\theta)$ directly, differentiate through a simulator and a series of measurement density ratios.
     \end{block}
@@
         \item \pause Particle filter run with state transitions \& resampling under $\phi \in \Theta$. 
-        \item \pause Reweight to evaluate likelihood at nearby $\theta \in \Theta$ (same resampling).
-        \item \pause Fix seed to treat particles, weights under $\phi$ as constants.
-        \item \pause \textbf{Now only need to differentiate through likelihood ratios!}
+        \item \pause Get a coupled (same seed) set of particles according to the process model at $\theta \in \Theta$, but resampling indices from the run at $\phi$. 
+        \item \pause Reweight conditional likelihoods by a correction factor accumulated over time, to account for resampling under $\phi$. 
+        \item \pause If $\theta=\phi$, need only one run, else need two runs at same seed. 
     \end{itemize}
@@
 
-\begin{frame}{What is this reweighting?}
-    Ordinarily, the likelihood estimate for $\phi$ is 
-    $$\hat\lik(\phi) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J \left(\underbrace{f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}, \phi)}_{\textcolor{blue}{\text{Measurement Model}}}
-    \cdot 
-    \underbrace{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi} | x_{n-1,j}^{F,\phi}, \phi)}_{\textcolor{red}{\text{Process model}}}\right).$$
+\begin{frame}{What is the correction?}
+    \begin{block}{Correction}
+        \begin{itemize}
+            \item Write:\pause
+            \begin{itemize}
+                \item $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density.\pause
+                \item $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$.\pause
+                \item $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$ for the resampling indices under $\phi$. \pause
+            \end{itemize}
+            \item Estimate conditional likelihood under $\theta$ as
             \pause
-    So to estimate the likelihood for $\theta$, we can correct 
-    $$
-        \hat\lik(\theta) 
-        = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}, \phi) \cdot 
-        f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi} | x_{n-1,j}^{F,\phi}, \phi) 
-        \cdot \underbrace{s_{n,j} \cdot r_{n,j}}_{\text{Correction Term}},
-    $$
+            \begin{equation}
+     \label{eq:mop-conditional-likelihood}
+     \underbrace{L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}}_{\textcolor{blue}{\text{\tiny{Usual way to estimate functionals under particle filter.}}}} \text{ or } \underbrace{L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}}_{\textcolor{red}{\text{\tiny{Higher variance, but useful for derivations.}}}},
+    \end{equation}
     \pause
-    where the multiplicative correction terms are
-    $$
-        \underbrace{s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}}_{\textcolor{blue}{\text{Measurement Model Likelihood Ratios}}},\;\; \underbrace{r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)}}_{\textcolor{red}{\text{Process Model Likelihood Ratios}}}.
-    $$
-
-\end{frame}
-
-\begin{frame}{New Problem}
-    \begin{block}{Problem}
-        Need likelihood ratios w.r.t. process and measurement models!
+    recursively weighting each particle by 
+    \pause
+    \begin{equation}
+        \label{eq:weighting-scheme}
+        \underbrace{w_{n,j}^{P,\theta} = (w_{n-1,j}^{F,\theta})^\alpha}_{\textcolor{red}{\text{\tiny{Bias-variance tradeoff.}}}}, \;\; \underbrace{w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}}_{\textcolor{blue}{\text{\tiny{Resampling correction.}}}}, \;\; w^{F,\theta}_{0,j}= 1.
+    \end{equation}
+        \end{itemize}
     \end{block}
-    \pause 
-    \begin{block}{Solution}
-        Offload process model derivatives to differentiable simulator. 
-    \end{block}
-    \begin{itemize}
-        \item \pause Run particle filter twice under same seed: one with process model at $\phi$, another with process model at $\theta$. Resample according to $\phi$.
-        \item \pause Correct only with resampling likelihood ratios. 
-        \item \pause Introduce additional discounting parameter $\alpha$ for bias-variance tradeoff.
-    \end{itemize}
 \end{frame}
@@
 When $\alpha=0$, lower variance but biased.
-MSE seems to be minimized at $\alpha=0.97$.}
+MSE minimized at $\alpha=0.97$.}
     \label{fig:bias-variance}
@@
 
-\begin{frame}{Guarantee}
-    \begin{prop}[Correctness of MOP-$\alpha$]
-    \label{prop:mop-correctness}
-    When either $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the posterior and is unbiased for the likelihood under $\theta$. 
-    When $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter. Its gradient when $\alpha=1$ is the estimate of \citet{poyiadjis11},
-    $$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{0:N}}(y_{1:N}^*|x_{1:N,j}^{A, F,\theta}),$$
-    and when $\alpha=0,$ is the gradient estimator of \cite{blei2018vsmc},
-    $$
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
-    $$
-    i.e. differentiating through a vanilla particle filter (\cite{scibior21}).
-\end{prop}
+\section{Theoretical Guarantees}
+
+\begin{frame}{Assumptions}
+
+\begin{itemize}
+    \item \textbf{Continuity of the Likelihood.} $\ell(\theta)$ has more than two continuous derivatives in a neighborhood of the MLE.
+    \pause
+    \item \textbf{Bounded Process Model.} There exist $\underbar{M}, \bar{M}$ bounding the process density from below/above.
+    \pause
+    \item \textbf{Bounded Measurement Model.} $\underbar{G}, \bar{G}$ bound the measurement density from below/above, and $G'(\theta)<\infty$ bounding its gradients.
+    \pause
+    \item \textbf{Bounded Gradient Estimates.} There are functions $G(\theta), H(\theta): \Theta \to [0,\infty)$ so the MOP-$\alpha$ gradient and Hessian estimates at $\theta=\phi$ are bounded by $G(\theta)$ and $H(\theta)$.
+    \pause
+    \item \textbf{Differentiability of Density Ratios and Simulator.} The measurement density $f_{Y_n|X_n}(y_n^*|x; \theta)$ and simulator have more than two continuous derivatives in $\theta$.
+\end{itemize}
+\end{frame}
+
+\begin{frame}{MOP-$\alpha$ Works}
+    \begin{thm}[MOP-$\alpha$ Targets the Posterior and Likelihood]
+    \label{thm:mop-targeting}
+    When $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood, as the number of particles $J \to \infty$.
+
+    \pause
+    That is, for $\pi_n(\theta)=f_{X_{1:n}|Y_{1:n} ; \theta}$ and any measurable bounded functional $h$, 
+    \pause
+    \begin{equation}
+        \underbrace{\frac{\sum_{j=1}^J h(x_{n,j}^{A,F, \theta}) w_{n,j}^{A,F,\theta}}{\sum_{j=1}^J w_{n,j}^{A,F,\theta}} \stackrel{a.s.}{\to} E_{\pi_n(\theta)} h,}_{\textcolor{blue}{\text{\tiny{Particle estimates of functionals under the posterior are consistent.}}}} \;\; \underbrace{\hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).}_{\textcolor{red}{\text{\tiny{Likelihood estimate is consistent.}}}}
+    \end{equation}
+\end{thm}
+\end{frame}
+
+\begin{frame}{Proof Sketch}
+    \begin{itemize}
+        \item This also serves as a strong law of large numbers result for triangular arrays of particles with ``off-policy'' resampling.\pause
+        \item When $\theta=\phi$, the ratio ${g_{n,j}^\theta}/{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter! Everything is as per usual. \pause
+        \item  When $\alpha=1$ and $\theta\neq\phi,$ suppose inductively that $\{(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j})\}_{j=1}^J$ targets $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.\pause
+        \begin{itemize}
+            \item We show $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j})\}_{j=1}^J$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.\pause
+            \item $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} )\}_{j=1}^J$ targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$. \pause
+            \item And that weighting by $(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}) = (X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}/ g^\phi_{n,a(j)})$ and resampling with probabilities $g^\phi_{n,j}$, targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.\pause
+        \end{itemize}
+        \item Result relies on three lemmas that are ``common-sense'' but require somewhat technical arguments to prove. 
+    \end{itemize}
+\end{frame}
+
+\begin{frame}{MOP-$\alpha$ Special Cases}
+    \begin{thm}[MOP-$0$ and MOP-$1$ Functional Forms]
+    \label{thm:mop-functional-forms}
+    Write $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$ and we use the after-resampling conditional likelihood estimate. When $\alpha=0$,
+    \pause
+    \begin{equation}
+        \nabla_\theta \hat\ell^0(\theta) 
+        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
+        \vspace*{-2.5mm}
+    \end{equation}
+    \pause
+    yielding \cite{blei2018vsmc} on the bootstrap filter. When $\alpha=1$,
+    \pause
+    \begin{equation}
+        \nabla_\theta \hat{\ell}^1(\theta) 
+        = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
+    \vspace*{-2.5mm}
+    \end{equation}
+    \pause
+    yielding the estimator of \cite{poyiadjis11} with the bootstrap filter.
+\end{thm}
+\pause
+Proof: Very tedious algebra.
+\end{frame}
+
+\begin{frame}{MOP-$1$ Is Strongly Consistent for the Score}
+    \begin{itemize}
+        \item But \cite{poyiadjis11} use $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}, X_{1:N}}\left(y_{1:N}^* , x_{1:n,j}^{A, F,\theta}; \theta\right)$ in general. \pause
+        \item The bootstrap estimate of $\nabla_\theta \hat{\ell}^1(\theta) 
+        = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right)$ has a different functional forms, and is not the same a-priori! \pause
+        \item So we directly show the consistency of the MOP-$1$ gradient estimate. 
+    \end{itemize}
+    \pause
+    \begin{thm}[Consistency of MOP-$1$ Gradient Estimate]
+        The gradient estimate of MOP-$\alpha$ when $\alpha=1$, $\theta=\phi$ is strongly consistent for the score: $\nabla_\theta \hat\ell_J^1(\theta) \stackrel{a.s.}{\to} \nabla_\theta \ell(\theta)$ as $J \to \infty$.
+        \label{thm:mop-grad-consistency}
+    \end{thm}
+\end{frame}
+
+\begin{frame}{Proof of MOP-$1$ Gradient Estimate Consistency}
+    \begin{itemize}
+        \item The sequence $(\hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ has two derivatives uniformly bounded over all $J$, by assumption. \pause
+        \item So $(\nabla_\theta \hat\lik_J^1(\theta)( \omega))_{J \in \mathbb{N}}$ is uniformly Lipschitz, and therefore uniformly equicontinuous for almost every $\omega \in \Omega$.\pause
+        \item By Arzela-Ascoli, there is a uniformly convergent subsequence. \pause
+        \item But there is only one subsequential limit by Theorem \ref{thm:mop-targeting}!\pause
+        \item So the whole sequence must converge uniformly with $\hat{\lik}_J^1(\theta, \omega) \stackrel{a.s.}{\to} \lik(\theta)$. \pause
+        \item So we can swap the limit and derivative and obtain that for almost every $\omega \in \Omega$, 
+    $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lik(\theta).$\pause
+        \item The result then follows by the continuous mapping theorem. 
+    \end{itemize}
+\end{frame}
+
+\begin{frame}{MOP-$\alpha$ Variance Bound}
+    \begin{thm}[MOP-$\alpha$ Variance Reduction]
+        The variance of the MOP-$\alpha$ gradient estimate, for $\alpha<1$, is given by
+        \pause
+        $$\text{Var}(\nabla_\theta \hat\ell^\alpha(\theta)) \lesssim \min_{k\leq N} \left(\underbrace{\frac{k^2G'(\theta)^2N}{(1-\alpha)^2J}}_{\textcolor{blue}{\tilde O(Nk^2/J)}} + \underbrace{\frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2}_{\textcolor{red}{\text{\tiny{Decreasing exponentially in $k$.}}}} \right).$$
+    \end{thm}
+    \pause
+    \begin{itemize}
+        \item Low-variance MOP-$0$ from \cite{blei2018vsmc} has $\tilde{O}(N/J)$ variance (we show this).\pause
+        \item High-variance MOP-$1$ from \cite{poyiadjis11} has $\tilde{O}(N^4/J)$ variance. \pause
+        \item MOP-$\alpha$ interpolates through both of these results to balance a bias-variance tradeoff. 
+    \end{itemize}
+\end{frame}
+
+\begin{frame}{Proof Sketch of MOP-$\alpha$ Variance Bound}
+    \begin{itemize}
+        \item Long and ugly.\pause
+        \item Consider a ``truncated'' version of the MOP-$\alpha$ estimate, where the weight correction is truncated at $k$ timesteps. This is easier to analyze. \pause
+        \item Call the conditional log-likelihood estimate $\hat{s}_{n,k}$, log-likelihood estimate $\hat{s}_{k}$, score estimate $\nabla_\theta \hat{s}_{k}$.\pause
+        \item By tedious algebra, the deviation of the truncated score estimate from the MOP-$\alpha$ score estimate is $\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_\infty \leq  \frac{2\alpha^k}{1-\alpha}NG'(\theta)$.\pause
+        \item Single-timestep variance is bounded by $\E\left[||\hat s_{n,k}(\theta) - \E_{\pi}\hat s_{n,k}(\theta)||_\infty^2\right]^{1/2} \leq \frac{Cr^2G'(\theta) k}{\sqrt{J}}.$
+    \end{itemize}
+\end{frame}
+
+\begin{frame}{Proof Sketch, Continued}
+\begin{align*}
+    \text{Var}(\nabla_\theta \hat s_k(\theta)) 
+    &= \sum_{n=1}^N\text{Var}\left(\hat s_{n,k}(\theta)\right)+ 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \\
+    &\lesssim \frac{r^2k^2G'(\theta)^2N}{J} + 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right).
+\end{align*}
+\pause
+    \begin{itemize}
+        \item How do we control the covariance? Mixing argument!\pause
+        \item Mixing time of the particle filter is $O(\log(J))$, \cite{karjalainen2023}. \pause
+        \item $\hat s_{n,k}$ is strong mixing at lag $k+1$, as weights beyond $k$ timesteps are truncated. \pause
+        \item So the $\alpha$-mixing coefficients for $(\hat s_{n,k}(\theta))_{n=1}^N$ are given by $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c(1+k+\log(J)))}}$. 
+    \end{itemize}
+\end{frame}
+
+\begin{frame}{Proof Sketch, Continued}
+    \begin{itemize}
+        \item By Davydov's inequality and Lemma 2 of \cite{karjalainen2023}, we can show that \pause
+    $$\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J}.$$
+    \pause
+    \item Recall that $||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||_\infty \leq \frac{2\alpha^k}{1-\alpha}NG'(\theta)$.\pause
+    \item So we have that for any $k\leq N$,
+    $$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \frac{k^2r^2G'(\theta)^2N}{J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2.$$
+    \pause
+    \item Take the minimum over $k$ to get the result. 
+    \end{itemize}
 \end{frame}
@@
 \begin{frame}{Convergence Analysis}
-    \begin{prop}[Convergence of IFAD-1]
+    \begin{prop}[Linear Convergence of IFAD-1]
         \label{thm:convergence}
 
-        Consider a variant of IFAD-1 where one stops if the gradient estimate $||g(\theta_n)|| \leq \sigma \epsilon$.
+        Consider a variant of IFAD-1 stopping if the gradient estimate $||g(\theta_n)|| \leq \sigma \epsilon$.
 Assume $-\ell$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$.
 $\sigma = \frac{4 \Gamma}{(1-\beta)}> 4$.
-Define $G$ as in Assumption \ref{assump:local-bounded-derivative}.
 
@@
     \centering
-    \includegraphics[scale=0.28]{imgs/095/boxplot.png}
-    \includegraphics[scale=0.31]{imgs/095/boxplotall.png}
-    \caption{\\
-    -- \textbf{Left:} Results of best run out of every ten runs, simulating procedure of running a few searches and choosing the best one. \\
-    -- \textbf{Right:} MOP alone drastically underperforms all other methods, failing to get close to the MLE. \\
-    -- IF2 warm-start necessary in challenging nonconvex and noisy problems.}
+    \includegraphics[scale=0.5]{imgs/095/boxplot.png}
+    \caption{Results of best run out of every ten runs, simulating procedure of running a few searches and choosing the best one. IFAD outperforms IF2, and tuning $\alpha$ ultimately helps a lot.}
     \label{fig:boxplot-search}
@@
 
+\section{Extension to Bayesian Inference}
+
+\begin{frame}{Particle MCMC}
+    \begin{itemize}
+        \item One can use the 
+    \end{itemize}
+\end{frame}
+
 \section{Conclusions and Future Work}
@@
 \bibliographystyle{apalike}
-\bibliography{mop/bib-ifad, mop/bib-ref}
-
-\appendix
-\section{Appendix}
-
-\begin{frame}{Previous Work}
-
-    \begin{itemize}
-        \item \cite{poyiadjis11}, particle approximation of the score. \pause \\ \cite{scibior21}: AD + "stop-gradient trick". 
-        \item 
-        \pause \cite{blei2018vsmc}, backpropagate through vanilla particle filter.
-        \item 
-        \pause \cite{corenflos21}, optimal transport resampling.
-        \item 
-        \pause \cite{singh22}, fixed-lag smoothing.
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{Previous Work: Issues}
-    \begin{itemize}
-        \item Not (yet) compatible with desire for simulation-based inference.
-        \begin{itemize}
-            \item \pause Need modifications to use the bootstrap filter.
-            \item \pause Sometimes special cases needed, e.g. transitions factoring into a policy and deterministic model (\cite{singh22}).
-        \end{itemize}
-        \item \pause Computationally expensive, quadratic in number of particles.
-        \begin{itemize}
-            \item \pause Optimal transport resampling (\cite{corenflos21}), marginal particle filters (\cite{scibior21}).
-        \end{itemize}
-        \item \pause High variance, or asymptotically biased.
-        \begin{itemize}
-            \item \pause Either one drops resampling terms and accepts asymptotic bias (\cite{blei2018vsmc}), or has variance quadratic in horizon (\cite{poyiadjis11}, \cite{scibior21})
-        \end{itemize}
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{(Improved) Iterated Filtering (\cite{ionides15})}
-    \begin{itemize}
-        \item Only (practical) full-information, frequentist method of parameter estimation in POMPs that does not need transition densities. 
-        \begin{itemize}
-            \item \pause Each particle has its own set of parameters.
-            \item \pause Particle parameters perturbed at every timestep.
-            \item \pause Parameters resampled with their particles according to likelihood.
-        \end{itemize}
-        \item \pause Treats parameters as augmentations of the state space, evolving according to a random walk.
-        \item \pause \textbf{Consistent estimates} via viewing this as a sequential Bayes map on the parameter distribution.
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{When Does IF2 Struggle?}
-
-\begin{itemize}
-    \item \textbf{IF2 is great at approaching the MLE quickly!}
-    \item \pause Performs parameter updates at every timestep, so it does not need to wait to filter through the entire trajectory before each update.
-    \item \pause \textbf{But struggles at squeezing out the last few units of log-likelihood.}
-    \begin{itemize}
-        \item \pause Especially true in highly nonlinear, nonconvex, and noisy settings, e.g. \textbf{disease models}. 
-    \end{itemize}
-\end{itemize}
-
-\end{frame}
-
-\begin{frame}{Assumptions}
-
-    \begin{aspt}[Smooth Neighborhood]
-    \label{assump:smooth-nbhd}
-    There exists a neighborhood $\mathcal{N}(\phi)$ around any $\phi \in \Theta$ where for all $\theta \in \mathcal{N}(\phi)$ and almost every $\omega \in \Omega$, the Monte Carlo estimate of the likelihood at $\theta \in \mathcal{N}(\phi)$ with the system evolving according to $\phi$ conditional on $\omega$, $\hat\lik(\theta,\phi,\omega, J)$, is twice differentiable in $\theta$.
-    \end{aspt}
-
-    \begin{itemize}
-        \item \pause \textbf{Justification:} For suitably nearby $\theta$ and $\phi$, the resampling indices for $\hat\lik(\theta,\phi,\omega, J)$ and $\hat\lik(\phi,\phi,\omega, J)$ are the same.
-        \item \pause This eliminates any discontinuities from resampling.
-        \item \pause For small enough $\mathcal{N}(\phi)$, the likelihood ratios are bounded. 
-        \item \pause The likelihood only changes by a factor of the likelihood ratios, which are bounded and smooth in $\theta$ if the densities used in the calculation are. 
-    \end{itemize}
-
-\end{frame}
-
-\begin{frame}{Assumptions}
-
-    \begin{aspt}[Continuity of the Likelihood] $\ell(\theta)$ proper is continuous in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.
-    \end{aspt}
-    \pause 
-    \begin{aspt}[Bounded Measurement Model] There is an $\epsilon>0$ with $\epsilon^{-1}>f_{Y_n \mid X_n}\left(y_n{ }^* \mid x_n; \theta\right)>\epsilon$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
-    \end{aspt}
-    \pause 
-    \begin{aspt}[Locally Bounded Derivative]
-    \label{assump:local-bounded-derivative}
-    Let $\mathcal{M}$ be an open subset of $\Omega$. There exists some function $G(\theta)$ and a constant $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$ such that
-        $$
-        \|\nabla \ell(\theta, \phi, \omega, J)\|_2<G(\theta) \leq G<\infty
-        $$
-        for every $\phi \in \mathcal{M}, \theta$ in smooth neighborhood $\mathcal{N}(\phi)$, and almost every $\omega \in \Omega$.
-    \end{aspt}
-\end{frame}
-
-\begin{frame}{Algorithm: Measurement Off-Policy, MOP-$\alpha$}
-    \begin{figure}
-        \centering
-        \includegraphics[scale=0.4]{imgs/mop.png}
-        \caption{MOP-$\alpha$ algorithm.}
-        \label{fig:mop}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Why did we bother with this?}
-    \begin{itemize}
-        \item Lets us work with only a differentiable simulator.
-        \item \pause Discounting parameter $\alpha$ helps with a bias-variance tradeoff (explained later). 
-        \item \pause Gradients returned by the likelihood estimate $$\hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$$ have nice properties.
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{Iterated Filtering with Automatic Differentiation (IFAD)}
-    \begin{figure}
-        \centering
-        \includegraphics[scale=0.45]{imgs/ifad.png}
-        \caption{Warm-starting first/second order iterative optimization with IF2.}
-        \label{fig:ifad}
-    \end{figure}
-\end{frame}
-
-\begin{frame}{Why does this help?}
-    \begin{itemize}
-        \item IF2 converges quickly, but struggles with last few log-likelihood units. 
-        \item \pause Gradient descent gets stuck in saddle points and poor local minima when the likelihood is nonconvex. 
-        \item \pause Warm-starting gradient methods with IF2:
-        \begin{itemize}
-            \item \pause Under regularity conditions, the likelihood is well-behaved near the MLE.
-            \item \pause Issues with saddle points and local minima alleviated. 
-        \end{itemize}
-        \item \pause Combining these two lets us enjoy the best of both worlds.
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{Convergence Analysis}
-
-    Guarantee currently only holds for IFAD-1, as it forms a particle approximation of the score as in \cite{poyiadjis11}.
-    \begin{itemize}
-        \item \pause Can be extended if the bias for the gradient estimate given by MOP-$\alpha$ for $\alpha < 1$ is small enough.
-        \item \pause e.g. $\alpha$ close to 1, or $y_{n+1:N}^*$ uninformative of current state $x_n$ given past and current measurements $y_{0:n}^*$.
-        \item \pause Otherwise, need to handle biased gradient descent.
-    \end{itemize}
-
-\end{frame}
-
-\begin{frame}{Cholera in Bangladesh}
-    The \textbf{transition dynamics} follow this series of stochastic differential equations driven by Brownian motion:
-    \begin{align*}
-        d S&=\left(k \epsilon R_k+\delta(S-H)-\lambda(t) S\right) d t+d H-(\sigma S I / H) d B, \\ 
-        d I&=\left(\lambda(t) S-(m+\delta+\gamma) I\right) d t+(\sigma S I / P) d B, \\ 
-        d R_1&=\left(\gamma I-(k \epsilon+\delta) R_1\right) d t, \\ 
-        \vdots \\ 
-        d R_k&=\left(k \epsilon R_{k-1}-(k \epsilon+\delta) R_k\right) d t,
-    \end{align*}
-\end{frame}
-
-\begin{frame}{Cholera in Bangladesh}
-The \textbf{force of infection}, $\lambda(t)$, is modeled by splines $s_j$, 
-    \begin{equation*}
-        \lambda(t)=\exp \left\{\beta_{\text {trend }}\left(t-t_0\right)+\sum_{j=1}^{6} \beta_j s_j(t)\right\}(I / P) + \exp \left\{\sum_{j=1}^{6} \omega_j s_j(t)\right\},
-    \end{equation*}
-    where
-    \begin{itemize}
-        \item \pause $\beta_j$ model seasonality in the force of infection.
-        \item \pause $\beta_{\text{trend}}$ models the trend in the force of infection.
-        \item \pause $\omega_j$ represent seasonality of a non-human environmental reservoir of disease.
-    \end{itemize}
-
-\end{frame}
-
-\begin{frame}{Cholera in Bangladesh}
-    The \textbf{measurement model} for observed monthly cholera deaths is given by 
-    \begin{equation*}
-        Y_n \sim \gN(M_n, \tau^2M_n^2),
-    \end{equation*}
-    where $M_n$ is the true number of cholera deaths in that month.
-\end{frame}
+\bibliography{paper/bib-ifad, paper/bib-ref}
 

### 2024-05-10 mop/draft.tex
Updates from Overleaf

@@
     Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
-Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, $\pi^J_j J$ times.
+Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, proportional to $\pi^J_j J$ times.
 Write
@@
     &\left\lVert\nabla_\theta\hat\ell^1(\theta) - \nabla_\theta\hat\ell^\alpha(\theta) \right\rVert_\infty\\
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
     &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\\
-    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
+    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
     &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)
@@
 
+\kevin{This subsection is not useless I think! It shows that the bias vanishes continuously as $\alpha$ goes to $1$.}
+
 \subsection{An Improved Bound by Mixing}
 
-We decompose
-$$\E||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty \leq \E||\nabla_\theta\ell_\theta - \nabla_\theta\hat s_k^1(\theta)||_\infty + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty.$$
-
-The broad idea is that the first term can be controlled by a mixing argument, the second term garners $O(Nk^2)$ error, and the third term is $O\left(\frac{N}{1-\alpha}\right)$.
+We decompose the bias $||\nabla_\theta\ell(\theta) - \E\nabla_\theta \hat\ell^\alpha(\theta)||_\infty \leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty$ into
+$$\E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty \leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_\infty + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty.$$
+
+The broad idea is that the first term can be controlled by a mixing argument, the second term garners $O(Nk)$ error, and the third term is $O\left(\frac{N}{1-\alpha}\right)$.
 It is unclear whether the bias can be reduced further -- the gradient is a sum of $N$ terms, and is therefore $O(N)$ itself.
 
-The first term -- bound ground truth of conditional scores from particle approximation conditional on correct filtering distribution at time $n-k$, bound that from particle approximation conditional on arbitrary filtering distribution at time $n-k$.
-First term sums up $k$ terms, $O(Nk^2)$ error, second term bounded by $k$ applications of the Markov kernel, so just apply forgetting of the particle filter result.
-
-But you can get an MSE bound like this! Simply decompose
-$$||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 \leq ||\nabla_\theta\ell_\theta - \nabla_\theta\hat s_k^1(\theta)||_2^2 + ||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2 + ||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_2^2,$$
-and observe that 
-$$||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty =  \left\lVert\sum_{n=1}^N \nabla_\theta \hat\ell_n^\alpha(\theta) - \sum_{n=1}^N \nabla_\theta \ell_n(\theta)\right\rVert_\infty \leq \sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_\infty.$$
+The first term is the problem term.
+We will first bound the ground truth of conditional scores from particle approximation conditional on correct filtering distribution at time $n-k$, and then bound that from the particle approximation resulting from an arbitrary filtering distribution at time $n-k$.
+That is, where $\hat\pi_n$ is the particle approximation at time $n$ of the posterior $\pi_n$ and we write $\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)$ for the truncated MOP-$1$ conditional score estimate given the correct filtering distribution at time $n-k$,
+
+\begin{align*}
+    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_\infty \\
+    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty \\
+    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_\infty + \sum_{n=1}^N \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty.
+\end{align*}
+
+The first term is a particle approximation dependent on $k$ timesteps, so by Lemma 2 of \cite{karjalainen2023}, this is bounded by
+$$\E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_\infty \leq \frac{CG'(\theta)k}{\sqrt{J}}.$$
+
+The second term amounts to bounding the difference between functionals of two different particle measures that mix under the same Markov kernel.
+Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
+We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
+$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
+i.e.
+the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
+
+Then, we can bound 
+$\E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty$ by
+$$\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}} = \beta_{TV}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
+implying that
+\begin{align*}
+    \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty 
+    &\lesssim \sup_{\mu, \nu} \sup_{||\psi||_\infty \leq 1/2} G'(\theta)\left|(\mu \textbf{M}_{n,n+k})(\psi)-(\nu \textbf{M}_{n,n+k})(\psi)\right| \\
+    &\leq \sup_{\mu, \nu} G'(\theta)\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}} \\
+    &\leq G'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}} ||\hat\pi_{n-k} - \pi_{n-k}||_{\text{TV}} \\
+    &\leq G'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}} \sup_{||\psi||_{\infty} \leq 1/2} |\hat\pi_{n-k}(\psi) - \pi_{n-k}(\psi)| \\
+    &\lesssim G'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}}.
+\end{align*}
+
+Therefore, we have that
+\begin{align*}
+    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_\infty \\
+    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_\infty + \sum_{n=1}^N \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty\\
+    &\lesssim N\frac{CG'(\theta)k}{\sqrt{J}} + NG'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}}.
+\end{align*}
+
+Now that the first term $\E||\nabla_\theta\ell(\theta)$ is taken care of, it remains to bound $$\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty \text{ and } \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty.$$
+
+We see, by a similar argument to the variance bound, that
+\begin{align*}
+    &\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty\\
+    &= \E\left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
+    &= \E\Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)- \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\\
+    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
+    &= \E\left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)
+    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\Bigg)\right\rVert_\infty \\
+    &\leq \sum_{n=1}^N \E\left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right|
+    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\right|\Bigg)\right\rVert_\infty.
+\end{align*}
+
+Each of these terms can be bounded using the log-derivative trick and that $w_{n,j}^{F,\theta,1,k} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
+\begin{align*}
+    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_{\infty}\\
+    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}}}\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,1,k}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,\alpha,k}\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,1,k}}{w_{n,j}^{F,\theta,1,k}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,\alpha,k}}{w_{n,j}^{F,\theta,\alpha,k}}\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,1,k}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta,1,k}\right)-\log\left(w_{n,j}^{F,\theta,\alpha,k}\right)\right)\right\rVert_{\infty}\\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=n-k}^n\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=n-k}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=n-k}^{n}(1-\alpha^{(n-i)})\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})G'(\theta)\\
+    &\leq G'(\theta)\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right),
+\end{align*}
+where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
+
+So, 
+\begin{align*}
+    &\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty\\
+    &\leq \sum_{n=1}^N \E\left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right|
+    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\right|\Bigg)\right\rVert_\infty \\
+    &\leq 2NG'(\theta)\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) 
+    \leq 2NG'(\theta)k.
+\end{align*}
+
+Now, the third term, $\E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty$, fortunately has already been dealt with. From the variance argument, we know that
+$$||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty \leq 2\frac{\alpha^k}{1-\alpha}NG'(\theta),$$
+and taking expectations on both sides yields our desired result.
+
+Therefore, our decomposition yields the bias bound
+\begin{align*}
+    &||\nabla_\theta\ell(\theta) - \E\nabla_\theta \hat\ell^\alpha(\theta)||_\infty \leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty \\
+    &\leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_\infty + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty \\
+    &\lesssim \min_{k \leq N} \left(N\frac{CG'(\theta)k}{\sqrt{J}} + NG'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}} + 2NG'(\theta)\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) + 2\frac{\alpha^k}{1-\alpha}NG'(\theta)\right) \\
+    &\lesssim \min_{k \leq N} NG'(\theta)\left(\frac{k}{\sqrt{J}}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\frac{\alpha^k  + \alpha^{k+1} - \alpha}{1-\alpha}\right).
+\end{align*}
+
+\subsection{Bounding the MSE of MOP-$\alpha$}
+
+\kevin{First term sums up $k$ terms, $O(Nk^2)$ error, second term bounded by $k$ applications of the Markov kernel, so just apply forgetting of the particle filter result. }
+
+We can obtain an $O(N)$ MSE bound from an argument that is similar to the bias argument. Simply observe that 
+$$||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 =  \left\lVert\sum_{n=1}^N \nabla_\theta \hat\ell_n^\alpha(\theta) - \sum_{n=1}^N \nabla_\theta \ell_n(\theta)\right\rVert_2^2 \leq \sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_2^2,$$
+
+and decompose
+$$\sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_2^2 \leq \sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 + \sum_{n=1}^N||\nabla_\theta\hat s_{n,k}^1(\theta) - \nabla_\theta\hat s_{n,k}^\alpha(\theta)||_2^2 + \sum_{n=1}^N||\nabla_\theta\hat s_{n,k}^\alpha(\theta) -  \nabla_\theta\hat\ell_n^\alpha(\theta)||_2^2.$$
+
+As before, we bound the first term, $\sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2$, by decomposing it into two terms, 
+$$\sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 \leq \sum_{n=1}^N ||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 + \sum_{n=1}^N ||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2.$$
+
+The first term is a particle approximation dependent on $k$ timesteps, so by Lemma 2 of \cite{karjalainen2023}, this is bounded by
+$$\E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 \leq \frac{CpG'(\theta)^2k^2}{J}.$$
+
+The second term amounts to bounding the difference between functionals of two different particle measures that mix under the same Markov kernel.
+Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
+We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
+$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
+i.e.
+the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
+
+Then, we can bound 
+$\E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2$ by
+$$\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}} = \beta_{TV}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
+implying that
+\begin{align*}
+    \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 
+    &\lesssim \sup_{\mu, \nu} \sup_{||\psi||_\infty \leq 1/2} pG'(\theta)^2\left|(\mu \textbf{M}_{n,n+k})(\psi)-(\nu \textbf{M}_{n,n+k})(\psi)\right| \\
+    &\leq \sup_{\mu, \nu} pG'(\theta)^2\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}} \\
+    &\leq pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} ||\hat\pi_{n-k} - \pi_{n-k}||_{\text{TV}} \\
+    &\leq pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} \sup_{||\psi||_{\infty} \leq 1/2} |\hat\pi_{n-k}(\psi) - \pi_{n-k}(\psi)| \\
+    &\lesssim pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}}.
+\end{align*}
+
+Therefore, we have that
+\begin{align*}
+    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_2^2 \\
+    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 + \sum_{n=1}^N \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2\\
+    &\lesssim N\frac{CpG'(\theta)^2k^2}{J} + NpG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}}.
+\end{align*}
+
+Now that the first term $\E||\nabla_\theta\ell(\theta)||_2^2$ is taken care of, it remains to bound $$\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2 \text{ and } \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_2^2.$$
+
+We see, by a similar argument to the bias bound, that
+\begin{align*}
+    &\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2\\
+    &\leq \sum_{n=1}^N \E\left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right|
+    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\right|\Bigg)\right\rVert_2^2.
+\end{align*}
+
+Each of these terms can be bounded using the log-derivative trick and that $w_{n,j}^{F,\theta,1,k} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
+\begin{align*}
+    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_{2}^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})pG'(\theta)^2\\
+    &\leq pG'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right),
+\end{align*}
+where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
+
+So, 
+\begin{align*}
+    \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty
+    \leq 2NpG'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) 
+    \leq 2NpG'(\theta)^2k.
+\end{align*}
+
+Now, we address the third term, $\E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty$.
+From the variance argument, we know that we can bound $||\nabla_\theta\hat\ell^\alpha(\theta) - \hat s_k^\alpha(\theta)||$:
+\begin{align*}
+    &\left\lVert\nabla_\theta\hat\ell^\alpha(\theta) - \hat s_k^\alpha(\theta) \right\rVert_2^2\\
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_2^2\\
+    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right)\right\lVert_2^2
+    +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta, \alpha}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha, k}\right)\right)\right\rVert_2^2.
+\end{align*}
+
+Using the log-derivative trick and that $w_{n,j}^{F,\theta,\alpha} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
+\begin{align*}
+    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}pG'(\theta)^2\\
+    &\leq pG'(\theta)^2\frac{\alpha^k-\alpha^n}{1-\alpha},
+\end{align*}
+where the second-last line follows from Assumption \ref{assump:bounded-measurement}. Putting it together and taking expectations on both sides, we obtain
+$$\E\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_2^2 \leq  \frac{2\alpha^k}{1-\alpha}NpG'(\theta)^2,$$
+which is our desired error bound.
+
+Therefore, our decomposition yields the MSE bound
+\begin{align*}
+    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 \\
+    &\leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_2^2 + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2 + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_2^2 \\
+    &\lesssim \min_{k \leq N} \left(N\frac{CpG'(\theta)^2k^2}{J} + NpG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} + 2NpG'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) + 2\frac{\alpha^k}{1-\alpha}NpG'(\theta)^2\right) \\
+    &\lesssim \min_{k \leq N} NpG'(\theta)^2\left(\frac{k^2}{J}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\frac{\alpha^k  + \alpha^{k+1} - \alpha}{1-\alpha}\right).
+\end{align*}
+
+\kevin{Trying to bound MOP-(1,k) from MOP-alpha yields exactly the same result as above! Seems like I can't get rid of the k in the third term within the parentheses.}
+% \begin{align*}
+%     &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)\right\rVert_{2}^2\\
+%     &\leq \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=n-k}^n\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=1}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_2^2 \\
+%     &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=n-k}^n\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=n-k}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=1}^{n-k}{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_2^2 \\
+%     &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2 + \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2 \\
+%     &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})pG'(\theta)^2 + \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}pG'(\theta)^2\\
+%     &\leq pG'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)+\alpha^k}{1-\alpha}\right),
+% \end{align*}
 

### 2024-05-10 mop/project.tex
Updates from Overleaf

@@
 
-\section{Background}
+\section{Introduction}
+
+Modeling a stochastic dynamical system in the presence of noisy observations is a challenging problem.
+Here, likelihood functions are intractable.
+One can bypass this issue with the approaches of simulated likelihood, likelihood-free inference, and the particle filter, but the Monte Carlo noise in the likelihood estimate that these approaches often suffer from makes parameter estimation difficult.
+Furthermore, many scientific applications, such as in disease modeling, deal with models where transition densities of the underlying stochastic process are unavailable.
+Existing statistical workhorses such as the EM algorithm, the Kalman filter, or MCMC, either struggle with intractable likelihoods or assume access to transition densities.
+
+Still, all is not lost.
+A specific variant (the bootstrap filter) of the particle filter, a popular method for solving the filtering problem in partially-observed dynamical systems, does not require transition densities.
+Instead, one only requires access to a simulator instead.
+This provides an unbiased estimate of the likelihood \citep{delmoral2004feynman} that can be used for parameter estimation.
+
+\subsection{Existing Algorithms}
+
+Currently, only three full-information simulation-based algorithms exist for parameter estimation: improved iterated filtering (IF2) \citep{ionides15}, particle MCMC \citep{doucet2010pmcmc}, and SMC$^2$ \citep{chopin13}.
+Only the first is frequentist, while the other two are Bayesian.
+
+Here, we will focus on maximum likelihood inference for partially observed Markov processes, and as such we provide a brief outline of IF2.
+IF2 augment the state space by assigning each particle $x_{n,j}$ its own set of parameters $\theta_{n,j}$.
+The parameters thought of as latent states that evolve according to Brownian motion on the parameter space $\Theta$.
+This yields a Monte Carlo approximation to the (extended) filtering distribution $f_{\theta_n,X_n|Y_{1:n}}(\theta_{n,j},x_{n,j}^F|y_{1:n}^*)$, given a Gaussian random walk prior on $\theta_{n,j} \sim \gN(\theta_{n-1,j},\sigma^2)$.
+Repeating the procedure again and again, while shrinking $\sigma^2$, yields convergence of the parameter swarm $\theta_{n,j}^{(m)}$ to the MLE.
+
+Still, this is not a silver bullet.
+Though in practice the algorithm of \citet{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood, as we see below in Figure \ref{fig:if2fail}.
+
+\begin{figure}[htbp!]
+    \centering
+    \includegraphics[scale=0.5]{imgs/095/if2fail.png}
+    \caption{Failure of iterated filtering on maximum likelihood estimation in the Dhaka cholera model of \cite{king08}.
+IF2 was run 100 times (with the 0th and 80th quantiles represented with the shaded area), but no run finds the MLE.
+The best run is about 7.5 log-likelihood units short.
+}
+    \label{fig:if2fail}
+\end{figure}
+
+\subsection{Is Automatic Differentiation the Answer?}
+
+The original iterated filtering algorithm of \cite{ionides06-pnas} amounts to gradient descent on a noisy particle approximation of the score.
+Similar particle approximations exist, such as that of \cite{poyiadjis11}, who provide a linear-time estimator with variance quadratic in the length of the trajectory that is consistent for the score.
+
+It has been shown by \cite{scibior21} that one can recover the asymptotically biased but low variance estimator of \cite{blei2018vsmc} and the consistent but high variance estimator of \cite{poyiadjis11} by performing automatic differentiation (AD) on the particle filter in the first case, and a slightly modified particle filter in the second.
+However, the ``stop-gradient trick'' they use to obtain the estimator of \cite{poyiadjis11} is not mathematically motivated, and is unexplained beyond being merely a heuristic.
+
+We build on this finding, creating a family of estimators parametrized by a hyperparameter $\alpha \in [0,1]$ that interpolates between the asymptotically biased but low variance estimator of \cite{blei2018vsmc} when $\alpha=0$ and the consistent but high variance estimator of \cite{poyiadjis11} when $\alpha=1$, allowing the user to optimize a bias-variance tradeoff.
+
+\subsection{Other Related Work}
+
+Recent advances in automatic differentiation (AD) for particle filters (\cite{blei2018vsmc, jon2018diffpf, corenflos21, scibior2021dpf, doucet2022particlebased}) have drawn attention to AD as an alternative tool for inference in partially observed stochastic processes with the particle filter.
+However, these existing approaches either are not yet compatible with the desire for simulation-based inference, or are computationally expensive.
+
+Only \cite{scibior21}, \cite{blei2018vsmc}, and \cite{corenflos21} have algorithms that are amenable to simulation-based inference without access to transition densities, where the first two have been discussed earlier.
+Though the approach of \citet{corenflos21} is promising, it is computationally expensive, requiring runtime quadratic in the number of particles (unlike other approaches, which run in linear-time).
+
+Of the other methods that are not compatible with simulation-based inference, \citet{doucet2022particlebased} provide another promising fixed-lag smoothing-based approach, but requires either (1) access to the process model transition densities or (2) the special case where the transition density factors into a policy that selects actions (in the reinforcement learning sense) and a deterministic model that maps states and actions to next-timestep states.
+
+\subsection{Our Contributions}
+
+Previous research on AD for particle filters has struggled with various issues: 
+\begin{enumerate}
+    \item Bias, if the discontinuous nature of particle resampling is ignored.
+    \item High Monte Carlo variance, if continuity corrections lead to numerical instability.
+    \item High computational cost, for algorithms which involve pairwise interactions between particles, marginalization, smoothing or optimal transport.
+    \item Reduced applicability for algorithms which lose or fail to take advantage of the simulation-based capabilities of the bootstrap filter.
+\end{enumerate}
+
+We develop a new approach to statistical inference via AD for particle filters which addresses these concerns.
+We develop a new theoretical framework that encompasses existing methods, and addresses the seemingly incompatible paradox of differentiating through a Monte-Carlo algorithm with discontinuous resampling.
+
+This is done through the construction of a smooth extension to the particle filter, as well as a \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) family of algorithms that (informally) encompasses the special case of AD of a vanilla PF (recovering the gradient estimator of \citet{blei2018vsmc}) when $\alpha=0$, and the DPF algorithm from \citet{scibior2021dpf} (that recovers the gradient estimator from \citet{poyiadjis11}) when $\alpha=1$.
+When $\alpha \in (0,1)$, this provides opportunities for optimizing a bias/variance tradeoff.
+
+We will show that these off-policy particle filters successfully target the posterior, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties.
+This entails novel proofs showing a strong law of large numbers for triangular arrays under bounded particle weights and functionals, with the application of showing that a suitably reweighted particle filter still targets the posterior, even when resampling is carried out under some other arbitrary categorical distribution.
+
+In practice, we propose a hybrid algorithm that we call \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts gradient ascent (or a similar iterative first or second-order gradient method) with a coarse solution obtained from a few rounds of iterated filtering.
+Promising preliminary numerical results indicate that this can beat current state-of-the-art methods on a challenging scientific benchmark problem.
+
+\section{Background and Problem Setup}
 A partially-observed Markov process (POMP) is defined as follows.
 Consider an unobserved Markov process $\{X_t, t \geq t_0\}$, and observations $Y_1,...,Y_N$ at timesteps $t_1,..., t_N$.
-The process is parameterized by an unknown parameter $\theta \in \Theta$, where $X \in \gX, Y \in \gY,$ and finally $\gX, \gY, \Theta \subseteq \R^n$.
+The process is parameterized by an unknown parameter $\theta \in \Theta$, where $X \in \gX, Y \in \gY,$ and finally $\gX \subseteq \R^d, \Theta \subseteq \R^p$.
 By a similar decomposition to that in \citet{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
@@
 
-\section{Challenges}
+\section{Off-Policy Particle Filters}
 
-Maximum-likelihood estimation for inference in partially observed stochastic processes, also known as, continuous-state continuous-time hidden Markov models (HMMs), partially-observed Markov processes (POMPs), or partially-observed stochastic nonlinear dynamical systems, is a challenging problem that faces several challenges:
-\begin{enumerate}
-    \item Intractable likelihood functions, bypassed with the approaches of simulated likelihood and likelihood-free inference. 
-    \item The desire for simulation-based inference without access to the transition density of the latent Markov process, with various methods created by \cite{welch2009abc, wood2010sl, doucet2010pmcmc, ionides08, ionides15}.
-    \item Accurate parameter estimation in the presence of significant Monte-Carlo noise in the likelihood estimate.
-\end{enumerate}
+Differentiating through the particle filter, while being an attractive idea, runs into the problem of differentiating through a Monte Carlo algorithm with discontinuous resampling.
+This is not a-priori possible.
+Our key insight is that instead of differentiating through $\hat\ell(\theta)$ directly, we instead differentiate through a simulator and a series of measurement density ratios.
+This corresponds to the following procedure.
+We run the particle filter once with the state transitions and resampling done under some $\phi \in \Theta$.
+We then obtain a coupled set of particles according to the process model at $\theta \in \Theta$, but with the resampling indices from the run at $\phi$.
+One can then reweight conditional likelihoods by a correction factor accumulated over time, to account for the resampling under $\phi$.
+Differentiating through the log-likelihood estimate given by the resulting algorithm then yields an estimator of the score.
 
-Many approaches to inference in these scenarios (such as the EM algorithm, the various Kalman filter variants, and MCMC) either struggle with intractable likelihood functions when the models are complex, or assume access to the probability density of next states given the current state.
-This is a problem in some critical applications, such as disease modeling, where the models are complex enough such that even obtaining the transition density is an intractable problem.
-However, the particle filter with the bootstrap proposal, a popular method for solving the filtering problem in partially-observed dynamical systems, provides an unbiased estimate of the likelihood (\cite{delmoral2004feynman}) without requiring evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
+The correction factor is given as follows.
+We write $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density, $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$, and $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$ for the resampling indices under $\phi$.
+One can then estimate conditional likelihood under $\theta$ as
+    \begin{equation}
+     \label{eq:mop-conditional-likelihood}
+     \underbrace{L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}}_{\textcolor{blue}{\text{{Usual way to estimate functionals under particle filter.}}}} \text{ or } \underbrace{L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}}_{\textcolor{red}{\text{{Higher variance, but useful for derivations.}}}},
+    \end{equation}
+    recursively weighting each particle by 
+    \begin{equation}
+        \label{eq:weighting-scheme}
+        \underbrace{w_{n,j}^{P,\theta} = (w_{n-1,j}^{F,\theta})^\alpha}_{\textcolor{red}{\text{{Bias-variance tradeoff.}}}}, \;\; \underbrace{w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}}_{\textcolor{blue}{\text{{Resampling correction.}}}}, \;\; w^{F,\theta}_{0,j}= 1.
+    \end{equation}
 
-This is the backbone behind the only simulation-based full-information maximum likelihood method available in the literature thus far, the improved iterated filtering (IF2) algorithm of \citet{ionides15}.
-\citet{ionides15} utilize the bootstrap filter for perform maximum likelihood estimation with an iterated perturbed Bayes map, successfully tackling challenging problems in epidemiology that available Bayesian methodology fails to solve.
+\begin{algorithm}[H]
+\centering
+	\caption{Measurement Off-Policy-$\alpha$}
+    \label{alg:mop}
+	\begin{algorithmic}[1]
+	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
+        \IF{$\theta \neq \phi$}
+            \STATE Run a particle filter once with parameters $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$, with seed $\omega$.
+        \ELSE
+        \STATE Set $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$ to be copies of $X_{n,j}^{P,\theta}, X_{n,j}^{P,\theta}$ in the rest of the algorithm.
+        \ENDIF
+		\STATE Initialize filter particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$. Fix $\omega.$
+		\FOR{$n=1,...,N$}
+            \STATE Prediction weights with discounting: $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
+            \label{mop-alpha:discount}
+            \STATE Simulate for prediction:
+            ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{mop-alpha:step1}
+            \STATE Evaluate measurement density:
+            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for $j$ in $\seq{1}{J}$
+            \STATE Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
+            \STATE Conditional likelihood under $\phi$: 
+            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
+            \label{mop-alpha:Lphi}
+            \STATE Normalize weights:
+            $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
+            for $j\ \mathrm{in}\ \seq{1}{J}$
+            \STATE Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \label{mop-alpha:systematic}
+            \STATE Resample particles: ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$
+            \STATE Filter weights corrected for resampling:
+            $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{mop-alpha:weight:update}
+            \STATE Resample filter weights:
+            $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
+            for $j$ in $\seq{1}{J}$ \label{mop-alpha:step2}
+            \STATE After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
+            \ENDFOR
+		\RETURN likelihood estimate $\hat{\lik}(\theta) = \lik(\theta, \phi, \omega, J) := \prod_{n=1}^N L_n^{A,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}.$
+	\end{algorithmic}
+\end{algorithm}
 
-Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small.
-For example, though in practice the algorithm of \citet{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood, as we see below in Figure \ref{fig:if2fail}.
+This yields an algorithm that we call Measurement Off-Policy-$\alpha$ (MOP-$\alpha$).
+$\alpha$ balances a tradeoff between maintaining the memory of each particle's ancestral trajectory (most extreme when $\alpha=1$), and considering only the single-step transition dynamics (when $\alpha=0$).
+The former yields a high-variance but consistent estimator, and the latter yields a low-variance but asymptotically biased estimator.
+One can think of MOP-$\alpha$ as an \textbf{Exponentially-weighted moving average} where $\alpha$ controls the amount of discounting.
+In practice, the effect of tuning $\alpha$ can yield significant gains, as we see in Figure \ref{fig:bias-variance} below.
 
@@
     \centering
-    \includegraphics[scale=0.5]{imgs/095/if2fail.png}
-    \caption{Failure of iterated filtering on maximum likelihood estimation in the Dhaka cholera model of \cite{king08}.
-IF2 was run 100 times (with the 0th and 80th quantiles represented with the shaded area), but no run finds the MLE.
-The best run is about 7.5 log-likelihood units short.
-}
-    \label{fig:if2fail}
+    \includegraphics[scale=0.45]{imgs/095/biasvar.png}
+    \caption{Bias-variance tradeoff for gradient estimates at MLE for trend in \cite{king08} cholera model.
+When $\alpha=1$, estimate consistent with high variance.
+When $\alpha=0$, lower variance but biased.
+MSE minimized at $\alpha=0.97$.}
+    \label{fig:bias-variance}
 \end{figure}
 
-\section{A Potential Solution}
+\section{Theoretical Analysis}
 
-Given the success of stochastic gradient methods in machine learning, it is natural to consider the performance of the procedure of performing gradient descent on the log-likelihood estimate given by the particle filter.
-We seek to develop a method that performs better than IF2 on this problem, by running IF2 for a few iterations and taking advantage of its quick convergence to a neighborhood of the MLE, and then switching to stochastic gradient descent on the gradient estimate given by running automatic differentiation on the log-likelihood estimate from the particle filter.
+We focus on the theoretical aspects of this estimator here. We require the following assumptions:
 
-\section{Automatic Differentiation for Particle Filters}
-
-Recent advances in automatic differentiation (AD) for particle filters (\cite{blei2018vsmc, jon2018diffpf, corenflos21, scibior2021dpf, doucet2022particlebased}) have drawn attention to AD as an alternative tool for inference in partially observed stochastic processes with the particle filter.
-However, these existing approaches either are not yet compatible with the desire for simulation-based inference, or are computationally expensive.
-
-In particular, only \cite{scibior21}, \cite{blei2018vsmc}, and \cite{corenflos21} have algorithms that are amenable to simulation-based inference without access to transition densities.
-Though the approach of \citet{corenflos21} is promising, it is computationally expensive, requiring runtime quadratic in the number of particles (unlike other approaches, which run in linear-time).
-
-\cite{scibior21} show that the estimator of \cite{blei2018vsmc} corresponds to the estimate of the gradient given by differentiating through the vanilla particle filter.
-This is therefore a very natural estimator for the score.
-However, \citet{blei2018vsmc} drops high-variance resampling terms from the gradient, leading to bias that does not vanish asymptotically according to \citet{corenflos21}.
-
-A linear-time consistent estimator of the score can be obtained via the estimator of \cite{poyiadjis11}, which \cite{scibior21} show can be obtained through a minor tweak on the particle filter that they call a "stop-gradient trick".
-However, this estimator has $O(N^2)$ variance, and the stop-gradient trick is not mathematically motivated.
-
-Of the other methods that are not compatible with simulation-based inference, \citet{doucet2022particlebased} provide another promising fixed-lag smoothing-based approach, but requires either (1) access to the process model transition densities or (2) the special case where the transition density factors into a policy that selects actions (in the reinforcement learning sense) and a deterministic model that maps states and actions to next-timestep states.
-
-\section{Our Contributions}
-
-Previous research on AD for particle filters has struggled with various issues: 
-\begin{enumerate}
-    \item Bias, if the discontinuous nature of particle resampling is ignored.
-    \item High Monte Carlo variance, if continuity corrections lead to numerical instability.
-    \item High computational cost, for algorithms which involve pairwise interactions between particles, marginalization, smoothing or optimal transport.
-    \item Reduced applicability for algorithms which lose or fail to take advantage of the simulation-based capabilities of the bootstrap filter.
+\begin{enumerate}[label=(A\arabic*)]
+    \item \textbf{Continuity of the Likelihood.} $\ell(\theta)$ has more than two continuous derivatives in a neighborhood of the MLE.
+    \label{assump:conti-lik}
+    \item \textbf{Bounded Process Model.} There exist $\underbar{M}, \bar{M}$ bounding the process density from below/above.
+    \label{assump:bounded-process}
+    \item \textbf{Bounded Measurement Model.} $\underbar{G}, \bar{G}$ bound the measurement density from below/above, and $G'(\theta)<\infty$ bounding its gradients.
+    \label{assump:bounded-measurement}
+    \item \textbf{Bounded Gradient Estimates.} There are functions $G(\theta), H(\theta): \Theta \to [0,\infty)$ so the MOP-$\alpha$ gradient and Hessian estimates at $\theta=\phi$ are bounded by $G(\theta)$ and $H(\theta)$.
+    \label{assump:local-bounded-derivative}
+    \item \textbf{Differentiability of Density Ratios and Simulator.} The measurement density $f_{Y_n|X_n}(y_n^*|x; \theta)$ and simulator have more than two continuous derivatives in $\theta$.
+    \label{assump:diff-meas-and-sim}
 \end{enumerate}
 
-We develop a new approach to statistical inference via AD for particle filters which addresses these concerns.
-We develop a new theoretical framework that encompasses existing methods, and addresses the seemingly incompatible paradox of differentiating through a Monte-Carlo algorithm with discontinuous resampling.
+\subsection{Correctness of MOP-$\alpha$}
 
-This is done through the construction of a smooth extension to the particle filter, as well as a \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) family of algorithms that (informally) encompasses the special case of AD of a vanilla PF (recovering the gradient estimator of \citet{blei2018vsmc}) when $\alpha=0$, and the DPF algorithm from \citet{scibior2021dpf} (that recovers the gradient estimator from \citet{poyiadjis11}) when $\alpha=1$.
-When $\alpha \in (0,1)$, this provides opportunities for optimizing a bias/variance tradeoff.
+First, we establish that MOP-$\alpha$ works as desired, with the following result.
 
-We will show that these off-policy particle filters successfully target the posterior, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties.
-This entails novel proofs showing a strong law of large numbers for triangular arrays under bounded particle weights and functionals, with the application of showing that a suitably reweighted particle filter still targets the posterior, even when resampling is carried out under some other arbitrary categorical distribution.
+\begin{thm}[MOP-$\alpha$ Targets the Posterior and Likelihood]
+    \label{thm:mop-targeting}
+    When $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood, as the number of particles $J \to \infty$.
 
-In practice, we propose a hybrid algorithm that we call \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts gradient ascent (or a similar iterative first or second-order gradient method) with a coarse solution obtained from a few rounds of iterated filtering.
-Promising preliminary numerical results indicate that this can beat current state-of-the-art methods on a challenging scientific benchmark problem.
+    That is, for $\pi_n(\theta)=f_{X_{1:n}|Y_{1:n} ; \theta}$ and any measurable bounded functional $h$, 
+    \begin{equation}
+        \underbrace{\frac{\sum_{j=1}^J h(x_{n,j}^{A,F, \theta}) w_{n,j}^{A,F,\theta}}{\sum_{j=1}^J w_{n,j}^{A,F,\theta}} \stackrel{a.s.}{\to} E_{\pi_n(\theta)} h,}_{\textcolor{blue}{\text{{Particle estimates of functionals under the posterior are consistent.}}}} \;\; \underbrace{\hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).}_{\textcolor{red}{\text{{Likelihood estimate is consistent.}}}}
+    \end{equation}
+\end{thm}
 
-\bibliography{bib-ifad,bib-ref}
+\begin{proof}
+    The proof relies on a handful of helpful lemmas in Appendix \ref{app:lemmas}.
+Those are in turn proved with a series of arguments involving Borel-Cantelli, the bounded convergence theorem, and a fourth moment argument for strong laws of large numbers, this time applied to triangular arrays of particles.
+
+    When $\theta=\phi$, regardless of the value of $\alpha$, the ratio $\frac{g_{n,j}^\theta}{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter estimate.
+
+    When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows.
+Recursively applying Lemmas \ref{lem:change-measure-proper-weights}, \ref{lem:marginal-proper-weights}, and \ref{lem:posterior-proper-weights}, we obtain that
+    %to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
+    the MOP-1 filter targets the posterior.
+    Specifically, suppose inductively that $\big\{\big(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
+    Then, Lemma \ref{lem:marginal-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big)\big\}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+    Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
+    \[
+    \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}\big/ g^\phi_{n,a(j)}\big),
+    \]
+    with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+
+    The fact that the likelihood estimate is consistent now follows from it being a particle estimate of a functional under the posterior, and the result that we had just proved.
+\end{proof}
+
+While the result is presented here as specific to MOP-$\alpha$, this actually serves as a more general result in the supplementary material.
+This is really a strong law of large numbers for triangular arrays of particles where we resample the particles according to an arbitrary resampling rule and iteratively correct the resampled particles with a correction factor accumulated over time.
+
+\subsection{MOP-$0$ and MOP-$1$ Functional Forms}
+
+We now show that MOP-$\alpha$ encompasses the estimators of \cite{blei2018vsmc} and \cite{poyiadjis11}.
+
+\begin{thm}[MOP-$0$ and MOP-$1$ Functional Forms]
+    \label{thm:mop-functional-forms}
+    Write $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$ and we use the after-resampling conditional likelihood estimate. When $\alpha=0$,
+    \begin{equation}
+        \nabla_\theta \hat\ell^0(\theta) 
+        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
+    \end{equation}
+    yielding \cite{blei2018vsmc} on the bootstrap filter. When $\alpha=1$,
+    \begin{equation}
+        \nabla_\theta \hat{\ell}^1(\theta) 
+        = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
+    \end{equation}
+    yielding the estimator of \cite{poyiadjis11} with the bootstrap filter.
+\end{thm}
+
+\begin{proof}
+
+    Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The likelihood estimate becomes
+\begin{equation*}
+    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}}= \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\theta}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
+\end{equation*}
+as we have a telescoping product. Then, as
+
+$$\nabla_\theta \hat\ell(\theta) = \frac{\nabla_\theta \hat\lik(\theta)}{\hat\lik(\theta)} = \frac{\nabla_\theta\left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi}{\prod_{n=1}^N L_n^\phi} =  \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},$$
+
+The derivative of the log-likelihood estimate is then
+\begin{equation*}
+    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},
+\end{equation*}
+which we decompose as follows.
+
+Observe that as $\alpha=1$,
+
+$$w_{n,j}^{P,\theta} = w_{n-1,j}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}},$$
+where we use the $(\cdot)^A$ superscript to denote the ancestral trajectory of the $j$-th prediction or filtering particle at timestep $n$.
+
+Using the log-derivative identity again,
+$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,j}^{A,P,\theta} - \log g_{i,j}^{A,P,\phi}\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,j}^{A,P,\theta}.$$
+
+So 
+$$  \nabla_\theta \sum_{n=1}^N \log g_{n,j}^{A,\theta} = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,j}^{A,P,\theta}\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,j}^{A, P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right),$$
+
+and 
+$$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{n=1}^N \nabla_\theta \log g_{n,j}^{A,P,\theta} = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$
+
+Substituting, we have that
+\begin{equation*}
+    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right),
+\end{equation*}
+and finally, observing that when $\theta=\phi$ we have that $w_{N,j}^{F,\theta}=1$, that 
+\begin{equation*}
+    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
+\end{equation*}
+yielding the gradient estimators of \cite{poyiadjis11, scibior2021dpf} when applied to the bootstrap filter.
+
+    When $\alpha=0,$ the likelihood estimate becomes
+\begin{equation}
+    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
+\end{equation}
+
+Its gradient when $\theta=\phi$ is therefore 
+\begin{align*}
+    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) = \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)}= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta s_{n,j}}{\sum_{j=1}^J s_{n,j}} \\
+    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)}{f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \phi}; \phi)} = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
+\end{align*}
+where we use the log-derivative trick, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again. 
+\end{proof}
+
+\subsection{Consistency of MOP-$1$}
+
+But \cite{poyiadjis11} use $$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}, X_{1:N}}\left(y_{1:N}^* , x_{1:n,j}^{A, F,\theta}; \theta\right)$$ in general.
+The bootstrap estimate of $$\nabla_\theta \hat{\ell}^1(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right)$$ has a different functional form, and is not the same a-priori.
+So we directly show the consistency of the MOP-$1$ gradient estimate below.
+
+\begin{thm}[Consistency of MOP-$1$ Gradient Estimate]
+        The gradient estimate of MOP-$\alpha$ when $\alpha=1$, $\theta=\phi$ is strongly consistent for the score: $\nabla_\theta \hat\ell_J^1(\theta) \stackrel{a.s.}{\to} \nabla_\theta \ell(\theta)$ as $J \to \infty$.
+        \label{thm:mop-grad-consistency}
+    \end{thm}
+
+\begin{proof}
+    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point we wish to evaluate the gradient at.
+The sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$, by Assumption \ref{assump:local-bounded-derivative}.
+The sequence is also uniformly equicontinuous.
+To see this, by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J(\theta, \omega)|_{\theta=\theta'}$ is also bounded by $H$ for almost every $\omega\in \Omega$ and every $\theta' \in \gN(\phi)$, including $\theta=\phi$.
+A set of functions with derivatives bounded by the same constant is uniformly Lipschitz, and therefore uniformly equicontinuous.
+So the sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ is uniformly equicontinuous for almost every $\omega \in \Omega$.
+    Explicitly, for almost every $\omega \in \Omega$ and every $\epsilon>0$, there exists some $\delta(\omega)>0$ such that for every $||\theta - \theta'||_{\infty}<\delta$ and every $J \in \mathbb{N}$ we have that
+    $$||\nabla_\theta \hat\lik_J(\theta, \omega)-\nabla_\theta \hat\lik_J(\theta', \omega)||_\infty < \epsilon.$$
+
+    Then, by Arzela-Ascoli, there exists a uniformly convergent subsequence.
+We claim that there is only one subsequential limit.
+We know that when as the gradient is bounded by Assumption \ref{assump:local-bounded-derivative} we can treat the gradient as a bounded functional.
+So by Lemma \ref{lem:posterior-proper-weights} the sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$, and there is therefore only one subsequential limit.
+The sequence therefore converges uniformly to its limit $\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega).$ Therefore, with uniform convergence for the derivatives established, we can swap the limit and derivative, and obtain that for almost every $\omega \in \Omega$,
+    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta, \omega).$$
+
+    From Theorem \ref{thm:mop-targeting}, we know that
+    $$\hat{\lik}_J(\theta, \omega) \stackrel{a.s.}{\to} \lik(\theta).$$
+
+    We then have that for almost every $\omega \in \Omega$, 
+    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta, \omega) = \nabla_\theta \lik(\theta),$$
+    as we wanted. The result then follows by the continuous mapping theorem. 
+\end{proof}
+
+\subsection{MSE, Bias, and Variance of MOP-$\alpha$}
+
+\begin{thm}
+    \label{thm:mop-biasvar}
+    When $\alpha\in(0,1)$ and $\theta=\phi$, the MSE, bias, and variance of MOP-$\alpha$, are:
+    \begin{align*}
+        \E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 
+        &\lesssim \min_{k \leq N} NpG'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\frac{\alpha^k  + \alpha^{k+1} - \alpha}{1-\alpha}\right),\\
+        ||\nabla_\theta\ell(\theta) - \E\nabla_\theta \hat\ell^\alpha(\theta)||_\infty 
+        &\lesssim \min_{k \leq N} NG'(\theta)\left(\frac{k}{\sqrt{J}}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\frac{\alpha^k  + \alpha^{k+1} - \alpha}{1-\alpha}\right),\\
+        \text{Var}(\nabla_\theta \hat\ell^\alpha(\theta)) &\lesssim \min_{k\leq N} \left(\underbrace{\frac{k^2G'(\theta)^2N}{(1-\alpha)^2J}}_{\textcolor{blue}{\tilde O(Nk^2/J)}} + \underbrace{\frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2}_{\textcolor{red}{\text{\tiny{Decreasing exponentially in $k$.}}}} \right).
+        \end{align*}
+\end{thm}
+
+This result is encouraging.
+It can be shown that the low-variance MOP-$0$ estimate from \cite{blei2018vsmc} has $\tilde{O}(N/J)$ variance, while the high-variance MOP-$1$ estimate from \cite{poyiadjis11} has $\tilde{O}(N^4/J)$ variance and MSE.
+MOP-$\alpha$ therefore interpolates through both of these results to balance a bias-variance tradeoff, while obtaining a rate of essentially $\tilde{O}(N/J)$ for the MSE and variance, modulo some asymptotic bias controllable by $\alpha$.
+
+\begin{proof}
+We will prove the MSE bound and omit the proof of the bias bound, as the proof of the bias bound follows along similar lines.
+The variance bound uses Davydov's inequality to bound the covariance between terms, and its proof is deferred to Appendix \ref{app:variance}.
+For the MSE bound, our strategy will be to consider truncated versions of estimators that enjoy the strong mixing property, bound the error of the truncated estimate in terms of particle approximation error and mixing error, and then bound the difference from the MOP-$\alpha$ estimate to the truncated estimate.
+
+We use the following notation for the truncated estimates.
+The truncated conditional log-likelihood estimate with discounting parameter $\alpha$ is written as $\hat{s}_{n,k}^{\alpha}$, truncated log-likelihood estimate as $\hat{s}_{k}^{\alpha}$, and the truncated score estimate as $\nabla_\theta \hat{s}_{k}^{\alpha}$.
+
+We begin by observing that 
+$$||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 =  \left\lVert\sum_{n=1}^N \nabla_\theta \hat\ell_n^\alpha(\theta) - \sum_{n=1}^N \nabla_\theta \ell_n(\theta)\right\rVert_2^2 \leq \sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_2^2,$$
+
+and decompose
+$$\sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_2^2 \leq \sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 + \sum_{n=1}^N||\nabla_\theta\hat s_{n,k}^1(\theta) - \nabla_\theta\hat s_{n,k}^\alpha(\theta)||_2^2 + \sum_{n=1}^N||\nabla_\theta\hat s_{n,k}^\alpha(\theta) -  \nabla_\theta\hat\ell_n^\alpha(\theta)||_2^2.$$
+
+As before, we bound the first term, $\sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2$, by decomposing it into two terms, 
+$$\sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 \leq \sum_{n=1}^N ||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 + \sum_{n=1}^N ||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2.$$
+
+The first term is a particle approximation dependent on $k$ timesteps, so by Lemma 2 of \cite{karjalainen2023}, this is bounded by
+$$\E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 \leq \frac{CpG'(\theta)^2k^2}{J}.$$
+
+The second term amounts to bounding the difference between functionals of two different particle measures that mix under the same Markov kernel.
+Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
+We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
+$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
+i.e.
+the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
+
+Then, we can bound 
+$\E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2$ by
+$$\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}} = \beta_{TV}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
+implying that
+\begin{align*}
+    \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 
+    &\lesssim \sup_{\mu, \nu} \sup_{||\psi||_\infty \leq 1/2} pG'(\theta)^2\left|(\mu \textbf{M}_{n,n+k})(\psi)-(\nu \textbf{M}_{n,n+k})(\psi)\right| \\
+    &\leq \sup_{\mu, \nu} pG'(\theta)^2\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}} \\
+    &\leq pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} ||\hat\pi_{n-k} - \pi_{n-k}||_{\text{TV}} \\
+    &\leq pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} \sup_{||\psi||_{\infty} \leq 1/2} |\hat\pi_{n-k}(\psi) - \pi_{n-k}(\psi)| \\
+    &\lesssim pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}}.
+\end{align*}
+
+Therefore, we have that
+\begin{align*}
+    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_2^2 \\
+    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 + \sum_{n=1}^N \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2\\
+    &\lesssim N\frac{CpG'(\theta)^2k^2}{J} + NpG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}}.
+\end{align*}
+
+Now that the first term $\E||\nabla_\theta\ell(\theta)||_2^2$ is taken care of, it remains to bound $$\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2 \text{ and } \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_2^2.$$
+
+It can be shown that
+\begin{align*}
+    &\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2\\
+    &\leq \sum_{n=1}^N \E\left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right|
+    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\right|\Bigg)\right\rVert_2^2.
+\end{align*}
+
+Each of these terms can be bounded using the log-derivative trick and that $w_{n,j}^{F,\theta,1,k} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
+\begin{align*}
+    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_{2}^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})pG'(\theta)^2\\
+    &\leq pG'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right),
+\end{align*}
+where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
+
+So, 
+\begin{align*}
+    \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty
+    \leq 2NpG'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) 
+    \leq 2NpG'(\theta)^2k.
+\end{align*}
+
+Now, we address the third term, $\E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty$.
+We can bound $||\nabla_\theta\hat\ell^\alpha(\theta) - \hat s_k^\alpha(\theta)||$ as follows:
+\begin{align*}
+    &\left\lVert\nabla_\theta\hat\ell^\alpha(\theta) - \hat s_k^\alpha(\theta) \right\rVert_2^2\\
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_2^2\\
+    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right)\right\lVert_2^2
+    +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta, \alpha}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha, k}\right)\right)\right\rVert_2^2.
+\end{align*}
+
+Using the log-derivative trick and that $w_{n,j}^{F,\theta,\alpha} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
+\begin{align*}
+    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}pG'(\theta)^2\\
+    &\leq pG'(\theta)^2\frac{\alpha^k-\alpha^n}{1-\alpha},
+\end{align*}
+where the second-last line follows from Assumption \ref{assump:bounded-measurement}. Putting it together and taking expectations on both sides, we obtain
+$$\E\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_2^2 \leq  \frac{2\alpha^k}{1-\alpha}NpG'(\theta)^2,$$
+which is our desired error bound.
+
+Therefore, our decomposition yields the MSE bound
+\begin{align*}
+    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 \\
+    &\leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_2^2 + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2 + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_2^2 \\
+    &\lesssim \min_{k \leq N} \left(N\frac{CpG'(\theta)^2k^2}{J} + NpG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} + 2NpG'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) + 2\frac{\alpha^k}{1-\alpha}NpG'(\theta)^2\right) \\
+    &\lesssim \min_{k \leq N} NpG'(\theta)^2\left(\frac{k^2}{J}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\frac{\alpha^k  + \alpha^{k+1} - \alpha}{1-\alpha}\right).
+\end{align*}
+\end{proof}
+
+\section{Practical Algorithms for Maximum Likelihood Inference}
+
+This still leaves us with the question of designing a reasonably effective procedure for likelihood maximization with these new tools.
+We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) that amounts to warm-starting gradient descent or some other reasonable first or second order procedure (using the gradient estimate given by MOP-$\alpha$) with the output of an initial search of IF2.
+
+A coarse solution is found by running IF2 with an aggressive random walk standard deviation (learning rate) and a cooling schedule that is \textbf{bounded below}, until the search stalls to an initial "convergence" to what is hopefully a neighborhood of the MLE.
+We can then use gradient descent, or other first or second order gradient methods, to refine this coarse solution.
+
+The initial "convergence" happens fairly quickly in practice.
+In the case of the Dhaka cholera model of \cite{king08}, when a geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations of IF2.
+In comparison, often 100 or 200 iterations of IF2 are used for an initial global search.
+\cite{ionides15}, for example, use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{Lee_haiticholera}.
+
+It can be shown that the gradient stage of IFAD converges linearly to the MLE under the usual strongly convex and smoothness assumptions.
+We conjecture that this applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE and behaves a lot like SGD in practice.
+
+\section{Numerical Results}
+
+To compare the performance of IFAD against that of IF2, we fit the very same model first proposed in \cite{king08} that \cite{ionides15} use to demonstrate the capabilities of IF2.
+This is becoming a standard benchmark, as \cite{wood16} and \cite{wycoff2024voronoi} include in their papers.
+This is a stochastic SIR compartmental model with transition and measurement uncertainty, where the population at time $t$, $H(t)$, is divided into the susceptible compartment $S(t)$, infected compartment $I(t)$, and three ($k=3$) recovered compartments $R_1(t), ..., R_k(t)$ denoting varying degrees of cholera immunity.
+We write $M(t)$ for the cholera deaths in each month.
+As in \cite{ionides15}:
+
+\begin{itemize}
+    \item The \textbf{transition dynamics} follow this series of stochastic differential equations:
+
+    \begin{align}
+        d S&=\left(k \epsilon R_k+\delta(S-H)-\lambda(t) S\right) d t+d H-(\sigma S I / H) d B, \\ 
+        d I&=\left(\lambda(t) S-(m+\delta+\gamma) I\right) d t+(\sigma S I / P) d B, \\ 
+        d R_1&=\left(\gamma I-(k \epsilon+\delta) R_1\right) d t, \\ 
+        \vdots \\ 
+        d R_k&=\left(k \epsilon R_{k-1}-(k \epsilon+\delta) R_k\right) d t,
+    \end{align}
+
+    where $B$ represents Brownian motion, $m$ is the rate at which infected individuals succumb to cholera, $\gamma$ is the rate at which infected individuals recover, $1/\epsilon$ is the mean duration of immunity, $\sigma$ represents the standard deviation of perturbations to the force of infection, and $\delta=0.02$ is the death rate.
+
+    \item The \textbf{force of infection}, $\lambda(t)$, is modeled by splines $s_j$
+    \begin{equation}
+        \lambda(t)=\exp \left\{\beta_{\text {trend }}\left(t-t_0\right)+\sum_{j=1}^{6} \beta_j s_j(t)\right\}(I / P) + \exp \left\{\sum_{j=1}^{6} \omega_j s_j(t)\right\},
+    \end{equation}
+    where the $\beta_j$ model seasonality in the force of infection, $\beta_{\text{trend}}$ models the trend in the force of infection, and the $\omega_j$ represent seasonality of a non-human environmental reservoir of disease.
+    \item The \textbf{measurement model} for observed monthly cholera deaths is given by 
+    \begin{equation}
+        Y_n \sim \gN(M_n, \tau^2M_n^2),
+    \end{equation}
+    where $M_n$ is the true number of cholera deaths in that month.
+\end{itemize}
+
+\begin{figure}[ht]
+\centering
+\begin{tikzpicture}[node distance=1cm and 1cm, auto,
+>=Latex,every node/.append style={align=center}, 
+int/.style={draw, minimum size=1.6cm}]
+  \node [int] (H) at (0,0) {$H$};
+\node [int] (S) [right=of H] {$S$};
+\node [int] (I) [right=of S] {$I$};
+\node (Mcoord) [right=of I, coordinate] {Mcoord};
+\node [int] (R1) [right=of Mcoord] {$R_1$};
+\node [int] (R2) [right=of R1] {$R_2$};
+\node [int] (R3) [right=of R2] {$R_3$};
+\node [circle, draw] (M) [above=of Mcoord] {$M$};
+\node (deathsS) [above=of S, coordinate] {deathsS};
+\node (deathsI) [above=of I, coordinate] {deathsS};
+\node (deathsR1) [above=of R1, coordinate] {deathsR1};
+\node (deathsR2) [above=of R2, coordinate] {deathsR2};
+\node (deathsR3) [above=of R3, coordinate] {deathsR3};
+\node (belowR3) [below=of R3, coordinate] {belowR3};
+\node (belowS) [below=of S, coordinate] {belowS};
+\path[->]
+    (H) edge node {$dH$} (S)
+    (S) edge node {$\lambda(t)$}  (I)
+    (I) edge node {$\gamma$}  (R1)
+    (R1) edge node {$k\epsilon$}  (R2)
+    (R2) edge node {$k\epsilon$}  (R3)
+    (R3) edge node {} (belowR3) 
+    (belowR3) edge node {$k\epsilon$} (belowS) 
+    (belowS) edge node {} (S)
+    (I) edge node {$m$}  (M)
+    (S) edge node {$\delta$}  (deathsS)
+    (I) edge node {$\delta$}  (deathsI)
+    (R1) edge node {$\delta$}  (deathsR1)
+    (R2) edge node {$\delta$}  (deathsR2)
+    (R3) edge node {$\delta$}  (deathsR3);
+\end{tikzpicture}
+\caption{Illustration of SIR model from \cite{king08}.}
+\label{fig:sir}
+\end{figure}
+
+Figure \ref{fig:sir} depicts the above compartmental model that we use to benchmark the performance of IFAD.
+We present comparisons of IFAD with $\alpha=0$ (IFAD-0) and $\alpha=1$ (IFAD-1) against both (1) the results from \cite{ionides15}, and (2) our own implementation of IF2 in \texttt{JAX}.
+
+\paragraph{Hyperparameters:} While performing an initial exploratory analysis, we found that alternative hyperparameters for IF2 (random walk standard deviation of $\sigma=0.02$, geometrically decreasing cooling rate exponent $a=0.95$) performed better than the settings \cite{ionides15} chose, and chose to use these for our implementation to achieve a fairer comparison.
+Our implementations of IF2 and IFAD used 10000 particles, in line with \cite{ionides15}.
+
+\subsection{Results} 
+We benchmarked IFAD against IF2 on a challenging global search problem.
+100 initial starting parameter vectors were drawn uniformly from the same wide bounding box used in \cite{ionides15}, and we performed 100 searches each of IF2, IFAD-0.97, IFAD-0, and IFAD-1 initialized from these 100 starting parameter vectors.
+We summarize our findings below in Figures \ref{fig:scatter}, \ref{fig:boxplot-optim}, and \ref{fig:hist-all}.
+
+\begin{figure}[htbp!]
+    \centering
+    \includegraphics[scale=0.53]{../imgs/095/pairs.png}
+    \includegraphics[scale=0.53]{../imgs/095/qq.png}
+    \caption{Scatterplots depicting the performance of IFAD against that of IF2.
+\textbf{Left:} Paired searches from the same starting point.
+\textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches.
+It is clear that on average, IFAD has the edge, and manages to find the MLE while no IF2 search successfully gets within 7 log-likelihood units of it.
+We use a dotted red line to display the MLE in both.
+}
+
+    \label{fig:scatter}
+\end{figure}
+
+\begin{figure}[htbp!]
+    \centering
+    \includegraphics[scale=0.5]{imgs/095/boxplot.png}
+    \includegraphics[scale=0.5]{imgs/095/optim.png}
+    \caption{\textbf{Left:} Results of best run out of every ten runs, simulating procedure of running a few searches and choosing the best one.
+IFAD outperforms IF2, and tuning $\alpha$ ultimately helps a lot.
+\textbf{Right:} Optimization plot of various methods.
+Solid lines depict the median negative log-likelihood at each iteration, while the shaded area depicts the best search at any iteration and the 80\% percentile.}
+    \label{fig:boxplot-optim}
+\end{figure}
+
+\begin{figure}[htbp!]
+    \centering
+    \includegraphics[scale=0.35]{imgs/095/hist.png}
+    \caption{Histogram comparison of IFAD, IF2, and MOP alone. Right is better.}
+    \label{fig:hist-all}
+\end{figure}
+
+\subsection{Takeaways}
+
+We obtain the following takeaways.
+IF2 converges quickly to a neighborhood of the MLE but fails to find the MLE.
+Gradient steps perform better at fine-grained refinement.
+In practice, we observe that performing gradient steps alone without a warm-start leads to the search getting stuck in local minima and saddle points.
+IFAD therefore combines the best of IF2 and MOP, approaching the MLE quickly and successfully performing refinement -- even on a very difficult global search problem.
+
+\section{Discussion and Future Work}
+
+We have developed a new theoretical framework, algorithm, and gradient estimator that encompasses the gradient estimates of \cite{blei2018vsmc} and \cite{poyiadjis11}.
+This yields a promising hybrid algorithm that warm-starts gradient descent (using this estimator) with IF2, that ultimately outperforms IF2 on Dhaka cholera model of \cite{king15}.
+
+Still, there is yet quite a bit of work to do.
+Firstly, MOP-$\alpha$ cannot handle simulators that are not differentiable with the reparametrization trick.
+Working with likelihood ratios or straight-through gradient estimators may be potential ways to get around this.
+Secondly, the convergence rate of IF2 remains an open problem, and as such we do not know the convergence rate of the initial stage of IFAD -- only the gradient stage.
+Thirdly, here we have only explored the use of gradient descent for simplicity, even though other methods like Newton's method, ADAM, L-BFGS, etc.
+may perform better.
+Fourth, extensions to panel and spatiotemporal data, e.g.
+with a block particle filter in the lens of \cite{ionides22} and \cite{ning23}, are important for scientific applications.
+
+\bibliography{paper/bib-ifad,paper/bib-ref}
 \bibliographystyle{apalike}
 
+\appendix
+\section{Lemmas for Correctness of MOP-$\alpha$}
+\label{app:lemmas}
+\begin{defn}[Targeting]
+    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any measurable and bounded functional $h$,
+\begin{equation}
+    E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
+\end{equation}  
+    A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if for any measurable and bounded functional $h$,
+\begin{equation}
+    \frac{\sum_{j=1}^J h(X^J_j) w^J_j}{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X))
+\end{equation}
+as $J \to \infty$.
+\end{defn}
+
+\begin{lem}[Change of Particle Measure]
+    \label{lem:change-measure-proper-weights}
+    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
+Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, proportional to $\pi^J_j J$ times.
+Write
+    \[
+    (Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big),
+    \]
+    where $a$ is called the ancestor function. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j),j=1,\dots,J\}$ targets $f_X$.
+\end{lem}
+
+\begin{proof}
+
+    Note that as the $Y_j^J$ are a subsample from $X_j^J$, $h$ can be a function of $Y$ as well as it is one for $X$. We then expand
+    $$\frac{\sum_j h(Y_j^J) V_j^J}{\sum_j V_j^J} = \frac{\sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}}.$$
+    By hypothesis,
+    $$\frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[f(X)].$$
+
+    We want to show
+    \begin{equation}\label{eq:lemma1:h}
+    \frac{\sum_j h(X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}} - \frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} 0.
+    \end{equation}
+    For this, it is sufficient to show that
+   $$ \sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}
+    -  \sum_j h(\tilde X_j^J)\frac{U_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0 $$
+    since an application of this result with $h(x)=1$ provides almost sure convergence of the denominator in (\ref{eq:lemma1:h}).
+    Write $g(\tilde X_j^J) = h(\tilde X_j^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}$. We therefore need to show that 
+    $$\sum_j Z_j^J := \sum_j \left(g(\tilde X_{k_j}^J) -  g(\tilde X_j^J) \pi_j^J \right) \stackrel{a.s.}{\to} 0.$$
+
+    Because the functional $h$ and importance sampling weights $u_{k_j}^J/\pi_{k_j}^J$ are bounded, we have that $\E[\left(Z_j^J\right)^4] < \infty$.
+We can then follow the argument of \cite{chopin20} from this point on, where
+    $$\E\left[\left(\sum_j Z_j^J\right)^4\right] 
+    = J\E[(Z_1^J)^4] + 3J(J-1)\left(\E[(Z_j^J)^2]\right) \leq CJ^2,$$ for some $C>0$. By Markov, 
+    $$\mathbb{P}\left(\left|\frac{1}{J} \sum_{j=1}^J Z_j^J\right|>\epsilon\right) 
+    \leq \frac{1}{\epsilon^4J^4 } 
+    \mathbb{E}\left[\left(\sum_{j=1}^J Z_j^J\right)^4\right] \leq \frac{C}{\epsilon^4J^2},$$
+    and as these terms are summable we can apply Borel-Cantelli to conclude that these deviations happen only finitely often for every $\epsilon>0,$ giving us the almost-sure convergence for
+
+    $$ \sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}
+    -  \sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0.$$
+
+    Similarly, we also have that
+    $$ \sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J
+    - \sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J} \stackrel{a.s.}{\to} 0,$$
+
+    and the result is proved.
+
+   % Let $h$ be an integrable function. 
+    %\begin{align*}
+    %    \E\left[\sum_{j=1}^J h(Y_j) v_j\right] 
+    %    &= \sum_{j=1}^J \E\left[h(X_{a(j)}) \frac{u_{a(j)}}{\pi_{a(j)}}\right] \\
+    %    &= \sum_{j=1}^J \E\left[h(X_{j}) u_{j}\right],
+    %\end{align*}
+    %and similarly for the denominator.
+Now, the numerator and denominator of the reweighted quantity have the same expectation as the numerator and denominator of the original quantity, so they must converge to $ E_\pi(h(X))$ almost surely as well.
+\end{proof}
+
+\textbf{Remark:} Note that Lemma \ref{lem:change-measure-proper-weights} permits $\pi_{1:J}$ to depend on $\{(X_j,u_j)\}$ as long as the resampling is carried out independently of $\{(X_j,u_j)\}$, conditional on $\pi_{1:J}$.
+
+\begin{lem}[Particle Marginals]
+    \label{lem:marginal-proper-weights}
+    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
+Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
+Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}$ targets $f_Z$.
+\end{lem}
+\begin{proof}
+    We want to show that, for any measurable bounded $h$, 
+     $$\frac{\sum_j h(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[h(Z)] = \E\big[\E[h(Z) | X]\big].$$
+    By hypothesis, for any measurable and bounded functional $g$ with domain $\gX$,
+    \begin{equation}\label{eq:lemma2:g}
+    \frac{\sum_j g(\tilde X_j^J)\, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[g(X)].
+    \end{equation}
+    Let $\bar{U}_j^J = \frac{J U_j^J}{\sum_j U_j^J}$.
+Examine the numerator and denominator of the quantity $$\frac{J^{-1}\sum_j h(\tilde Z_j^J) \, U_j^J}{J^{-1}\sum_j U_j^J} = \frac{J^{-1}\sum_j h(\tilde Z_j^J) \, \bar{U}_j^J}{J^{-1}\sum_j \bar{U}_j^J}.$$
+    The denominator converges to $1$ almost surely. The numerator, on the other hand, is
+    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J,$$
+    by the same fourth moment argument to the above lemma converges almost surely to the limit of its expectation
+    \begin{eqnarray*}        
+    \lim_{J\to\infty} \E\left[\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J\right] 
+    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  
+      \E\left[ h(\tilde Z_j^J)\, \bar{U}_j^J\big|\tilde X_j^J, \bar U_j^J\right]
+    \right]
+    \\
+    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  \E\big[h(Z) \big| X=\tilde X_j^J \big] \, \bar U_j^J\right].
+    \end{eqnarray*}
+    Applying (\ref{eq:lemma2:g}) with $g(x) = \E\big[h(Z) | X=x\big]$, the average on the right hand size converges almost surely to $\E\big\{\E[h(Z)|X]\big\}=\E[h(Z)].$
+    It remains to swap the limit and expectations. We can do so with the bounded convergence theorem, and therefore obtain   
+    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\,\bar{U}_j^J \stackrel{a.s.}{\to} \E[h(Z)].$$
+
+\end{proof}
+
+\begin{lem}[Particle Posteriors]
+    \label{lem:posterior-proper-weights}
+    Suppose that $\{(X_j^J,u_j^J),j=1,\dots,J\}$ targets $f_X$.
+Also suppose that $(X^{\prime J}_j,u^{\prime J}_j) = \big(X_j^J,u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$.
+Then, if $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,u^{\prime J}_j)\}$ targets $f_{X|Z}(\cdot | z^*)$.
+\end{lem}
+
+\begin{proof}
+
+    $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J}}{J^{-1} \sum_j \bar{u}_j^{' J}}$$
+
+    by
+
+    \begin{equation}
+    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}}
+    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \right)^{-1}
+    \end{equation}
+
+    Again, we want to show that
+     $$\frac{\sum_j h(X_j^J) \cdot u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)} \stackrel{a.s.}{\to} \E_{f_{X|Z}}[h(X)|z^*].$$
+
+    We already have that
+    $$\frac{\sum_j h(X_j^J)u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} \E_{f_X}[h(X)].$$
+
+    %= \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j f_{Z|X}(z^*|X_j^J)f_X(X_j^J)}
+
+    Consider the following:
+    \begin{equation}
+    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}}
+    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \right)^{-1}.
+    \end{equation}
+
+    Write $\bar{u}_j^{J} = \frac{J u_j^J}{\sum_j u_j^J}$.
+Define the weights $u^{' J}_j := u_j^J \cdot f_{Z|X}(z^*|X_j^J),$ and the corresponding self-normalized weights $\bar{u}_j^{' J} := \frac{J u^{' J}_j}{\sum_j u^{' J}_j}.$ The denominator of the below quantity
+    $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J}}{J^{-1} \sum_j \bar{u}_j^{' J}}$$
+    converges to $1$, while the numerator, by the same fourth moment argument as the first lemma, converges almost surely to the limit of its expectation, which is
+    \begin{align*}
+        \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J} \right]
+        &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)}\right] \\
+        &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
+        &= \lim_{J\to\infty} \E\left[ \sum_j h(X_j^J) \frac{\frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
+        &= \E_{X|Z}[h(X)|z^*],
+    \end{align*}
+
+    where we can swap the limit and expectations with the bounded convergence theorem.
+\end{proof}
+
+\section{Variance Bound Proof}
+\label{app:variance}
+
+\subsection{Analysis of MOP-$\alpha$ Variance}
+
+The proof strategy is as follows.
+Instead of analyzing the variance of the MOP-$\alpha$ gradient estimate proper, we will analyze the variance of a modified estimator where we truncate the weights at $k$ timesteps, so this estimator only looks at most $k$ timesteps back while accumulating the importance weights.
+This enables us to establish strong mixing for the truncated estimator, in order to get a bound on the variance that is $O(NJ^{-1})$.
+We emphasize that this estimator is only a theoretical construct -- we do not actually use this estimator.
+
+Coupled with a bound ensuring that the $k$-truncated estimator provides an estimate that is close to that of MOP-$\alpha$ proper, we get a bound on the variance of MOP-$\alpha$ that comprises of the variance of the $k$-truncated estimator plus the error, for any $k \leq N$.
+It then holds that the final variance bound on MOP-$\alpha$ is given by the minimum over all $k$, and is never larger than $O(1+NJ^{-1})$.
+
+Using the log-derivative trick and that $w_{n,j}^{F,\theta} = w_{n,j}^{F,\theta,k} = 1$ when $\theta=\phi$,
+\begin{align*}
+    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,k}\right)\right\rVert_{\infty}\\
+    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta}}{{\sum_{j=1}^J w_{n,j}^{F,\theta}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,k}}}\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,k}\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta}}{w_{n,j}^{F,\theta}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,k}}{w_{n,j}^{F,\theta,k}}\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,k}\right)\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta}\right)-\log\left(w_{n,j}^{F,\theta,k}\right)\right)\right\rVert_{\infty}\\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=n-k}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_{\infty} \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^{n-k}\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}G'(\theta)\\
+    &\leq G'(\theta)\frac{\alpha^k-\alpha^n}{1-\alpha},
+\end{align*}
+where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
+
+We can then bound $||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||$:
+\begin{align*}
+    &\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_\infty\\
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
+    &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\\
+    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)
+    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\Bigg)\right\rVert_\infty \\
+    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)\right)\right\lVert_{\infty}
+    +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\right)\right\rVert_\infty,
+\end{align*}
+and each of these two terms is bounded by $\frac{\alpha^k}{1-\alpha}NG'(\theta)$. So we now know that
+$$\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_\infty \leq  \frac{2\alpha^k}{1-\alpha}NG'(\theta),$$
+which is our desired error bound.
+
+Now, we bound the variance of $\hat s_k(\theta)$. 
+Recall that we defined
+$$\hat s_k(\theta) = \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) = \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right).$$
+
+We can then decompose this expression with the log-derivative trick while noting that $w_{n,j}^{F,\theta, k} = 1$ whenever $\theta=\phi$, to see that
+\begin{align*}
+    \hat s_k(\theta) 
+    &= \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right) \\
+    &= \sum_{n=1}^N \left(\frac{\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}}{\sum_{j=1}^J w_{n,j}^{F,\theta, k}}-\frac{\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}}{\sum_{j=1}^J w_{n-1,j}^{A,F,\theta, k}}\right) \\
+    &= \sum_{n=1}^N \left(\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}\right)\\
+    &= \frac{1}{J}\sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right).
+\end{align*}
+
+It now follows that we need to bound the variance at a single timestep, namely
+$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right).$$
+
+We use the log-derivative trick yet again to find, noting that $w_{n,j}^{F,\theta, k}=1$ when $\theta=\phi$, that
+\begin{align*}
+    \nabla_\theta w_{n,j}^{F,\theta, k} = \frac{\nabla_\theta w_{n,j}^{F,\theta, k}}{w_{n,j}^{F,\theta, k}} = \nabla_\theta \log(w_{n,j}^{F,\theta, k}) = \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) = \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right).
+\end{align*}
+
+Then,
+\begin{align*}
+    \text{Var}\left(\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right) 
+    &= \text{Var}\left(\nabla_\theta \left(\sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right)- \sum_{i=n-k-1}^{n-1} \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right)\right)\right) \\
+    &= \text{Var} \left(\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1}\log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right).
+\end{align*}
+
+Note that $$\hat s_{n,k}(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)$$ is a function bounded by $C\frac{G'(\theta)(1-\alpha^k)}{1-\alpha} =: CG'(\theta)r$ in each coordinate for some constant $C$ (by Assumption \ref{assump:bounded-measurement}) that depends only on $k$ timesteps, and not $n$.
+We can therefore invoke Lemma 2 of \cite{karjalainen2023} to bound the $L_2$ error of each of the $k$ (from time $n-k$ to time $n$) functionals from its expectation under the posterior by $CrG'(\theta)J^{-1/2}k$.
+That is, we have that
+$$\E\left[||\hat s_{n,k}(\theta) - \E_{\pi}\hat s_{n,k}(\theta)||_\infty^2\right]^{1/2} \leq \frac{Cr^2G'(\theta) k}{\sqrt{J}}.$$
+
+By the bias-variance decomposition, this in turn bounds the variance at a single timestep by $C^2r^2G'(\theta)^2J^{-1}k^2$. Concretely, we have that
+$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \lesssim \frac{r^2k^2G'(\theta)^2}{J}.$$
+
+It now remains to bound the variance of $\hat s_k(\theta)$ by considering the covariance of each of the $N$ terms that comprise it. We decompose
+\begin{align*}
+    \text{Var}(\nabla_\theta \hat s_k(\theta)) &= \text{Var}\left(\sum_{n=1}^N \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \\
+    &= \sum_{n=1}^N\text{Var}\left(\hat s_{n,k}(\theta)\right)+ 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \\
+    &\lesssim \frac{r^2k^2G'(\theta)^2N}{J} + 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right).
+\end{align*}
+
+Similarly to the proof of the MOP-$0$ case, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
+We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
+$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
+i.e.
+the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
+By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strongly mixing, with $\alpha$-mixing coefficients $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c\log(J))}}$.
+Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
+
+We now derive the $\alpha$-mixing coefficients of $(\hat s_{n,k}(\theta))_{n=1}^N$.
+Observe that $\hat s_{n,k}$ is strong mixing at lag $k+1$, as all weights beyond $k$ timesteps are truncated.
+We therefore have that the mixing time of $\hat s_k(\theta)$ is $O(1+k+\log(J))$, and that the $\alpha$-mixing coefficients for $(\hat s_{n,k}(\theta))_{n=1}^N$ are given by $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c(1+k+\log(J)))}}$.
+
+Therefore, by Davydov's inequality and Lemma 2 of \cite{karjalainen2023}, and noting that $\hat s_{n,k}(\theta) \leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, by a similar argument to the MOP-0 case, we find that
+
+$$\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J}.$$
+
+Concretely, noting that $\nabla_\theta\hat s_{n,k}(\theta)\leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and WLOG assuming $\E[(\nabla_\theta\hat\hat s_{m,k}(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat\hat s_{n,k}(\theta))^4]^{1/4}$, we apply Davydov's inequality to see that
+\begin{align*}
+    \text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) 
+    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat s_{m,k}(\theta))^4]^{1/4}\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/4}\\
+    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2}.
+\end{align*}
+
+To bound this, we use the fact that $$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4] = \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4]+\E[(\nabla_\theta\hat s_{n,k}(\theta))^2]^2$$
+alongside Lemma 2 of \cite{karjalainen2023}, which shows that
+$$\E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4] \lesssim \frac{r^4G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^2] \lesssim \frac{r^2G'(\theta)^2}{J},$$
+to find that 
+$$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{r^4G'(\theta)^4}{J^2}+\left(\frac{r^2G'(\theta)^2}{J}\right)^2} = \frac{r^2G'(\theta)^2}{J},$$
+and conclude that 
+$$\text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J}.$$
+
+Putting it all together, we see that
+\begin{align*}
+    \text{Var}(\nabla_\theta \hat s_{k}(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat s_{n,k}(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat s_{m,k}(\theta),\nabla_\theta\hat s_{n,k}(\theta))\\
+    &\leq \frac{Nr^2k^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J} \\
+    &\leq \frac{r^2k^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\right) \\
+    &\lesssim \frac{k^2r^2G'(\theta)^2N}{J}.
+\end{align*}
+
+We will now use this result to bound $\text{Var}(\nabla_\theta \hat s_k(\theta)$ proper. For random variables $X,Y$ where $|X-Y|$ is bounded almost surely by some $M$, we have that 
+$$X-\E X - 2M \leq Y - \E Y \leq X - \E X + 2M$$
+so,
+$$(Y-\E Y)^2 \leq (X-\E X - 2M)^2 + (X-\E X +2M)^2.$$
+Taking expectations, we get
+$$\text{Var}(Y) \leq \text{Var}(X) + 8M^2.$$
+
+It follows from this result and the result we proved earlier, namely that
+$$||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||_\infty \leq \frac{2\alpha^k}{1-\alpha}NG'(\theta),$$
+that the variance of MOP-$\alpha$ proper is bounded by, for any $k \leq N$,
+that the variance of MOP-$\alpha$ proper is bounded by
+$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \frac{k^2r^2G'(\theta)^2N}{J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2.$$
+
+As the above holds for any $k \leq N$,
+$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \min_{k\leq N} \left(\frac{k^2G'(\theta)^2N}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right).$$
+
+Note that the result here is presented in the sup norm. In the main text, we provide the result in the $2$-norm, and therefore scale the above result by $p$.
+
 \end{document}

### 2024-05-10 paper/ms.tex
Updates from Overleaf

@@
 %We show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.
+
+\kevin{Can someone who's reading through the words and glazing through the math get the idea that (1) we're running AD on the PF, and (2) using an off-policy resampling trick to get our estimator.
+Should I hedge by writing the SI in Del Moral language?
+}
 
@@
 When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{blei2018vsmc}.
+
+\subsection{MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, blei2018vsmc}}
+
+\cite{scibior21} show that the estimate of \cite{blei2018vsmc} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through applying a ``stop-gradient trick" to the particle filter.
+It turns out that both of these, when applied on the bootstrap filter, correspond to special cases of MOP-$\alpha$.
+
 $\alpha$ therefore controls an exponentially-weighted moving average, determining how much "memory" the resulting gradient estimate has over ancestral trajectories.
 This is illustrated in Figure \ref{fig:biasvar}, in the context of the Dhaka cholera model of \cite{king08}.
+
+\begin{thm}[MOP-$0$ and MOP-$1$ Functional Forms]
+    \label{thm:mop-functional-forms}
+    Write $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$ and we use the after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$.
+Then, when $\alpha=0$,
+    \vspace*{-2.5mm}
+    \begin{equation}
+        \nabla_\theta \hat\ell^0(\theta) 
+        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
+        \vspace*{-2.5mm}
+    \end{equation}
+    yielding \cite{blei2018vsmc} on the bootstrap filter. When $\alpha=1$,
+    \vspace*{-2.5mm}
+    \begin{equation}
+        \nabla_\theta \hat{\ell}^1(\theta) 
+        = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
+    \vspace*{-2.5mm}
+    \end{equation}
+    yielding the estimator of \cite{poyiadjis11, scibior21} with the bootstrap filter.
+\end{thm}
+
+We defer the proof to the supplementary material.
+It relies on a useful decomposition of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$ that yields a telescoping product in the MOP-$1$ case.
+Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting that when $\theta=\phi$, that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result.
+
+This further illustrates how $\alpha$ dictates the "memory" of the gradient estimate.
+As \cite{scibior2021dpf} remark, the MOP-$0$ estimator of \cite{blei2018vsmc} only considers single-step quantities, and is "memoryless" beyond a single step.
+This is in contrast to the case when $\alpha=1,$ as that estimate of \cite{poyiadjis11} only considers the surviving particles at time $N$, fully tracking dependencies over time.
 
@@
 \kevin{source for this}
+
+\subsection{Linear Convergence Rates}
+\begin{thm}
+    \label{thm:mop-convergence}
+\end{thm}
 
@@
 We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood.
-While the result is presented here as specific to MOP-$\alpha$, the full proof relies on a handful of technical lemmas presented in the supplementary material that may be useful in their own right, and Theorem \ref{thm:mop-targeting} can also serve as a more general result as a strong law of large numbers for triangular arrays of particles with ``off-policy'' resampling.
+While the result is presented here as specific to MOP-$\alpha$, we actually prove a more general result in the supplementary material.
+That is, we show a strong law of large numbers for triangular arrays of particles with ``off-policy'' resampling, where we resample the particles according to an arbitrary resampling rule not necessarily in proportion to the targeted distribution of interest, to weights that encode the cumulative discrepancy between the resampling and the target distribution instead of equal weights.
+
+\kevin{Should we just show the general result?
+That is, for an arbitrary set of indices $k_{n,j}$ generated in proportion to some arbitrary weights $\pi_{n,j}$ that do not necessarily sum to $1$ as the estimate is self-normalized, reweighting the particles by $w_{n,a(j)}g_{n,a(j)}/\pi_{n,a(j)}$ targets the posterior.
+$\pi_{n,j}$ can depend on $\{(x_{n,1:J},g_{n,1:J})\}$ as long as the resampling is carried out independently of $\{(x_{n,1:J},g_{n,1:J})\}$, conditional on $\pi_{n,1:J}$}
 
@@
 
+\kevin{Write the supplement formally with Del Moral-lite language, going into full detail with the proofs and calculations.}
+
 \begin{proof}
@@
 \end{proof}
-
-\subsection{MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, blei2018vsmc}}
-
-\cite{scibior21} show that the estimate of \cite{blei2018vsmc} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through applying a``stop-gradient trick" to the particle filter.
-It turns out that both of these, when applied on the bootstrap filter, correspond to special cases of MOP-$\alpha$.
-
-\begin{thm}[MOP-$0$ and MOP-$1$ Functional Forms]
-    \label{thm:mop-functional-forms}
-    Write $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$ and we use the after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$.
-Then, when $\alpha=0$,
-    \vspace*{-2.5mm}
-    \begin{equation}
-        \nabla_\theta \hat\ell^0(\theta) 
-        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
-        \vspace*{-2.5mm}
-    \end{equation}
-    yielding \cite{blei2018vsmc} on the bootstrap filter. When $\alpha=1$,
-    \vspace*{-2.5mm}
-    \begin{equation}
-        \nabla_\theta \hat{\ell}^1(\theta) 
-        = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
-    \vspace*{-2.5mm}
-    \end{equation}
-    yielding the estimator of \cite{poyiadjis11, scibior21} with the bootstrap filter.
-\end{thm}
-
-We defer the proof to the supplementary material.
-It relies on a useful decomposition of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$ that yields a telescoping product in the MOP-$1$ case.
-Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting that when $\theta=\phi$, that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result.
-
-This further illustrates how $\alpha$ dictates the "memory" of the gradient estimate.
-As \cite{scibior2021dpf} remark, the MOP-$0$ estimator of \cite{blei2018vsmc} only considers single-step quantities, and is "memoryless" beyond a single step.
-This is in contrast to the case when $\alpha=1,$ as that estimate of \cite{poyiadjis11} only considers the surviving particles at time $N$, fully tracking dependencies over time.
 
@@
     \label{thm:mop-biasvar}
+    When $\alpha\in(0,1)$, $\theta=\phi$, the MSE and variance of MOP-$\alpha$, where $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$, are:
+    \vspace*{-1ex}
+    \begin{align*}
+        &\E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 
+        \\
+        &\lesssim \min_{k \leq N} NpG'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi(\alpha)\right),\\[-1ex]
+        &\text{Var}(\nabla_\theta \hat\ell^{\alpha}(\theta)) \lesssim \min_{k\leq N} NpG'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N\right).
+        \end{align*}
 \end{thm}
+\kevin{Explain exactly what $\epsilon$ is here, and interpret the rates as agreeing with our intuition.
+Talk about the relationship between the theory and the empirical results in the bias-variance tradeoff figure above, as well as the phase transition between $Nk^2/J$ and $N^4/J$ variance with MOP-$\alpha$ and MOP-$1$, where $k$ is really a surrogate for $\alpha$ -- with more memory, you truncate less.
+}
 
-\subsection{Convergence Rates}
-\begin{thm}
-    \label{thm:mop-convergence}
-\end{thm}
+We defer the proof to the supplementary material in favor of a very brief outline here.
+
+\kevin{Say that our framework suggests whole new classes of algorithms, as it includes Naesseth and Poyiadjis.
+Interpolating between them with discounting is only one way to interpolate, we can also e.g.
+truncate at a fixed lag.
+Remark that the analysis is similar for MOP-$(1,k)$ as a fixed-lag filter, with slightly better rates but a less natural interpretation.
+We leave empirical studies of its performance to future work.
+}
 

### 2024-05-10 presentation/presentationmath.tex
Updates from Overleaf

@@
 \title[AD for POMPs]{Automatic Differentiation Accelerates Inference for Partially Observed Stochastic Processes}
-\author[Kevin Tan]{Kevin Tan \\ \small{University of Pennsylvania, Department of Statistics and Data Science} \\\vspace{0.5cm} \small{Joint work with Giles Hooker and Edward Ionides}}
+\author[Kevin Tan]{Kevin Tan \\ \small{University of Pennsylvania, Department of Statistics and Data Science\\\vspace{0.5cm} \small{Advised by Giles Hooker and Edward Ionides}}}
 
@@
     \begin{itemize}
-        \item Unobserved Markov process $\{X_t, t \geq t_0\}$, observations $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$. Write $X_n := X_{t_n}$.
-        \item \pause Unknown parameter $\theta \in \Theta$, to be estimated.
-        \item $X_t \in \gX \subseteq \R^d$, $Y_n \in \gY$, $\theta \in \Theta \subseteq \R^p$.\pause
+        \item Unobserved Markov process $\{X_t, t \geq t_0\}$, observations $Y_1,...,Y_N \in \gY$ taking values $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$. Write $X_n := X_{t_n} \subseteq \R^d$.
+        \item \pause Unknown parameter $\theta \in \Theta \subseteq \R^p$, to be estimated.
         \item \textbf{Notation:}
@@
 
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}[scale=1.5]
+
+\node[black] (X_0) at (-3,0) {$X_0$};
+\node[black] (Xdot0) at (-2,0) {$...$};
+\node[black] (X_{n-1}) at (-1,0) {$X_{n-1}$};
+\node[black] (X_n) at (0,0) {$X_n$};
+\node[black] (X_{n+1}) at (1,0) {$X_{t+1}$};
+\node[black] (Xdot{n+1}) at (2,0) {$...$};
+\node[black] (X_N) at (3,0) {$X_T$};
+\node[black] (Y_0) at (-3,-1) {$Y_0$};
+\node[black] (Ydot0) at (-2,-1) {$...$};
+\node[black] (Y_{n-1}) at (-1,-1) {$Y_{n-1}$};
+\node[black] (Y_n) at (-0,-1) {$Y_n$};
+\node[black] (Y_{n+1}) at (1,-1) {$Y_{n+1}$};
+\node[black] (Ydot{n+1}) at (2,-1) {$...$};
+\node[black] (Y_N) at (3,-1) {$Y_N$};
+\path[->, black] (X_0) edge (Xdot0);
+\path[->, black] (Xdot0) edge (X_{n-1});
+\path[->, black] (X_{n-1}) edge (X_n);
+\path[->, black] (X_n) edge (X_{n+1});
+\path[->, black] (X_{n+1}) edge (Xdot{n+1});
+\path[->, black] (Xdot{n+1}) edge (X_N);
+\path[->, black] (X_0) edge (Y_0);
+\path[->, black] (X_{n-1}) edge (Y_{n-1});
+\path[->, black] (X_n) edge (Y_n);
+\path[->, black] (X_{n+1}) edge (Y_{n+1});
+\path[->, black] (X_N) edge (Y_N);
+\end{tikzpicture}
+\caption{Directed acyclic graph (DAG) of a POMP model.
+The latent states $(X_t)_{t_0\leq t \leq T}$ follow a Markov process, and the observations $Y_n$ are conditionally independent given the latent state.
+}
+\label{fig:pomp}
+\end{figure}
+\end{frame}
+
+\begin{frame}{Partially Observed Markov Processes}
+
     \begin{itemize}
-        \item Unobserved Markov process $\{X_t, t \geq t_0\}$, observations $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$.
-        \item \pause Problems to solve:
+        \item Problems to solve:
         \begin{itemize}
@@
             \begin{itemize}
-                \item \pause Find the density of the hidden state at time $n$, given all observations until time $n$.
-                \item \pause $f_{X_n|Y_{1:n}}(x_n|y_{1:n}^*;\theta)$.
+                \item \pause Find the density of the hidden state at time $n$, given all observations until time $n$, $f_{X_n|Y_{1:n}}(x_n|y_{1:n}^*;\theta)$.
             \end{itemize}
@@
             \begin{itemize}
-                \item \pause Find the density of the hidden states at timesteps $1$ to $n$, given all observations till time $n$. 
-                \item \pause $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*;\theta)$.
+                \item \pause Find the density of the hidden states at timesteps $1$ to $n$, given all observations till time $n$, $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*;\theta)$.
             \end{itemize}
@@
          \hline\hline
-         \cite{poyiadjis11} & Particle approximation of score
-         \\ & $O(N^2)$ variance, unbiased \\
+         \cite{poyiadjis11} & Particle approximation of score.
+         \\ & $O(N^2)$ variance, consistent. \\
          \hline
-         \cite{blei2018vsmc} & Backprop through vanilla PF \\ 
-         & Asymptotic bias \\
+         \cite{blei2018vsmc} & Backprop through vanilla PF. \\ 
+         & Asymptotic bias. \\
          \hline
-         \cite{corenflos21} & Optimal transport resampling \\
-         & Consistent, $O(J^2)$ runtime \\
+         \cite{corenflos21} & Optimal transport resampling. \\
+         & Consistent, $O(J^2)$ runtime. \\
          \hline
-         \cite{scibior21} & Stop-gradient trick \\ 
-         & Recovers \cite{poyiadjis11} with AD \\
+         \cite{scibior21} & Stop-gradient trick. \\ 
+         & Recovers \cite{poyiadjis11} with AD. \\
          \hline 
-         \cite{singh22} & Fixed-lag smoothing \\ 
-         & Need transition densities \\ 
+         \cite{singh22} & Fixed-lag smoothing. \\ 
+         & Need transition densities. \\ 
          \hline
@@
         \item \pause Maintaining memory of each particle's ancestral trajectory (most extreme when $\alpha=1$).
+        \item \pause High-variance, but consistent.
         \item \pause Considering only the single-step transition dynamics (when $\alpha=0$).
+        \item \pause Low-variance, but asymptotically biased.
     \end{itemize}
@@
 \begin{frame}{Particle MCMC}
-    \begin{itemize}
-        \item One can use the 
-    \end{itemize}
+    One can use the likelihood estimate from the particle filter to perform MCMC. But the random walk proposal is very bad!
+
+\begin{figure}[t!]
+    \centering
+    \includegraphics[width = \textwidth/\real{1.5}]{imgs/pmcmc/mh.png}
+    \caption{Convergence diagnostics for the Metropolis-Hastings variant particle MCMC with the random walk proposal and an informative empirical prior from IF2.
+Here, we display the results for the trend parameter in the Dhaka cholera model of \cite{king08}.}
+    \label{fig:mh}
+\end{figure}
+\end{frame}
+
+\begin{frame}{MOP-$\alpha$ Speeds up the Mixing of Particle MCMC}
+    Using the gradient estimate from MOP-$\alpha$ within a NUTS sampler (initialized with an informative prior estimated with IF2) works wonders -- burn-in period of $700,000$ in \cite{fasiolo16} reduced to $500$ here.
+    \begin{figure}[t!]
+    \centering
+    \includegraphics[width=\textwidth/\real{1.5}]{imgs/pmcmc/nuts_eb.png}
+    \caption{Convergence diagnostics for a No U-Turn Sampler with the informative prior from IF2. The NUTS sampler mixes quickly, and the posterior estimates from each chain posterior modes.}
+    \label{fig:nuts-eb}
+\end{figure}
+
 \end{frame}

### 2024-05-29 mop/draft.tex
update from overleaf

@@
 
-\begin{lem}[Change of Particle Measure]
+\begin{lem}[Change of Weight Measure]
     \label{lem:change-measure-proper-weights}
@@
     We want to show that, for any measurable bounded $h$, 
-     $$\frac{\sum_j h(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[h(Z)] = \E\big[\E[h(Z) | X]\big].$$
-    By hypothesis, for any measurable and bounded functional $g$ with domain $\gX$,
+     $$\frac{\sum_j h(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)] = \E_{f_X}\big[\E_{f_{Z|X}}[h(Z) | X]\big].$$
+    By assumption, for any measurable and bounded functional $g$ with domain $\gX$,
     \begin{equation}\label{eq:lemma2:g}
-    \frac{\sum_j g(\tilde X_j^J)\, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[g(X)].
+    \frac{\sum_j g(\tilde X_j^J)\, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[g(X)].
     \end{equation}
@@
     It remains to swap the limit and expectations. We can do so with the bounded convergence theorem, and therefore obtain   
-    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\,\bar{U}_j^J \stackrel{a.s.}{\to} \E[h(Z)].$$
+    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\,\bar{U}_j^J \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)].$$
 
@@
     \label{lem:posterior-proper-weights}
-    Suppose that $\{(X_j^J,u_j^J),j=1,\dots,J\}$ targets $f_X$.
-Also suppose that $(X^{\prime J}_j,u^{\prime J}_j) = \big(X_j^J,u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$.
-Then, if $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,u^{\prime J}_j)\}$ targets $f_{X|Z}(\cdot | z^*)$.
+    Suppose that $\{(X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
+Also suppose that $(X^{\prime J}_j,U^{\prime J}_j) = \big(X_j^J,U_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$.
+Then, if $U_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $U_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,U^{\prime J}_j)\}$ targets $f_{X|Z}(\cdot | z^*)$.
 \end{lem}
@@
 
-    $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J}}{J^{-1} \sum_j \bar{u}_j^{' J}}$$
-
-    by
-
-    \begin{equation}
-    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}}
-    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \right)^{-1}
-    \end{equation}
-
     Again, we want to show that
-     $$\frac{\sum_j h(X_j^J) \cdot u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)} \stackrel{a.s.}{\to} \E_{f_{X|Z}}[h(X)|z^*].$$
-
-    We already have that
-    $$\frac{\sum_j h(X_j^J)u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} \E_{f_X}[h(X)].$$
-
-    %= \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j f_{Z|X}(z^*|X_j^J)f_X(X_j^J)}
+     $$\frac{\sum_j h(X_j^J) \cdot U_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j U_j^J \cdot f_{Z|X}(z^*|X_j^J)} \stackrel{a.s.}{\to} \E_{f_{X|Z}}[h(X)|z^*].$$
+
+    We already have that for any measurable bounded $g$,
+    \begin{equation}
+        \frac{\sum_j g(X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[g(X)].
+        \label{eq:lemma-posterior-hypothesis}
+    \end{equation}
 
@@
     \begin{equation}
-    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}}
-    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \right)^{-1}.
-    \end{equation}
-
-    Write $\bar{u}_j^{J} = \frac{J u_j^J}{\sum_j u_j^J}$.
-Define the weights $u^{' J}_j := u_j^J \cdot f_{Z|X}(z^*|X_j^J),$ and the corresponding self-normalized weights $\bar{u}_j^{' J} := \frac{J u^{' J}_j}{\sum_j u^{' J}_j}.$ The denominator of the below quantity
-    $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J}}{J^{-1} \sum_j \bar{u}_j^{' J}}$$
-    converges to $1$, while the numerator, by the same fourth moment argument as the first lemma, converges almost surely to the limit of its expectation, which is
-    \begin{align*}
-        \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J} \right]
-        &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)}\right] \\
-        &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
-        &= \lim_{J\to\infty} \E\left[ \sum_j h(X_j^J) \frac{\frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
-        &= \E_{X|Z}[h(X)|z^*],
-    \end{align*}
-
-    where we can swap the limit and expectations with the bounded convergence theorem.
+    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}
+    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}} \right)^{-1}.
+    \end{equation}
+
+    We will apply Equation \ref{eq:lemma-posterior-hypothesis} to the numerator and the denominator in the ratio above individually. The numerator converges to 
+    $$\frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \stackrel{a.s.}{\to} \E_{f_X}[h(X)f_{Z|X}(z^*|X)],$$
+    while the reciprocal of the denominator converges to 
+    $$ \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}  \stackrel{a.s.}{\to} \E_{f_X}[f_{Z|X}(z^*|X)] = f_Z(z^*).$$
+
+    But it holds that
+    $$\frac{\E_{f_X}[h(X)f_{Z|X}(z^*|X)]}{f_Z(z^*)} = \E_{f_X}\left[\frac{h(X)f_{Z|X}(z^*|X)}{f_Z(z^*)}\right] = \E_{f_X}\left[h(X)\frac{f_{X|Z}(X|z^*)}{f_X(X)}\right] =  \E_{f_{X|Z}}[h(X)|z^*],$$
+    and we have the desired result.
+
+    % Write $\bar{U}_j^{J} = \frac{J U_j^J}{\sum_j U_j^J}$.
+Define the weights $U^{' J}_j := U_j^J \cdot f_{Z|X}(z^*|X_j^J),$ and the corresponding self-normalized weights $\bar{U}_j^{' J} := \frac{J U^{' J}_j}{\sum_j U^{' J}_j}.$ The denominator of the below quantity
+    % $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{U}_j^{' J}}{J^{-1} \sum_j \bar{U}_j^{' J}}$$
+    % converges to $1$, while the numerator, by the same fourth moment argument as the first lemma, converges almost surely to the limit of its expectation, which is
+    % \begin{align*}
+    %     \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J} \right]
+    %     &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)}\right] \\
+    %     &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
+    %     &= \lim_{J\to\infty} \E\left[ \sum_j h(X_j^J) \frac{\frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
+    %     &= \E_{X|Z}[h(X)|z^*],
+    % \end{align*}
+
+    % where we can swap the limit and expectations with the bounded convergence theorem.
+\end{proof}
+
+\begin{lem}[Change of Particle Marginal Measure]
+    \label{lem:change-marginal-proper-weights}
+    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
+Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
+Further, let $f_{Y|X}$ be a conditional probability density function corresponding to a joint density $f_{X,Y}$ with marginal densities $f_X$ and $f_Y$.
+Then, if the $U_j^J$ and $V_j^J := U_j^Jf_{Y|X}(\tilde{Z}_j^J|\tilde{X}_{j}^J)/f_{Z|X}(\tilde{Z}_j^J|\tilde{X}_{j}^J)$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}$ targets $f_Z$, and $\{(\tilde Z_j^J,V_j^J)\}$ targets $f_Y$.
+\end{lem}
+\begin{proof}
+    From Lemma \ref{lem:marginal-proper-weights}, for any measurable bounded $g$, 
+     $$\frac{\sum_j g(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_Z}[g(Z)] = \E_{f_X}\big[\E_{f_{Z|X}}[g(Z) | X]\big],$$
+     and by assumption, for any measurable bounded $g$, 
+    $$\frac{\sum_j g(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[g(X)].$$
+
+    We in fact want to show that 
+    $$\frac{\sum_j h(\tilde Z_j^J) \, V_j^J}{\sum_j V_j^J} \stackrel{a.s.}{\to} \E_{f_Y}[h(Y)] = \E_{f_X}\left[\E_{f_{Z|X}}\left[h(Z)\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)} \Bigg| X\right]\right].$$
+
+    Consider the following:
+    \begin{equation}
+    \frac{J^{-1} \sum_j h(Z_j^J) \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)} {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}
+    \times \left( \frac{J^{-1} \sum_j \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)} {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}} \right)^{-1}.
+    \end{equation}
+
+    We will apply Lemma 2 to the numerator and the denominator in the ratio above individually. The numerator converges to 
+    $$\frac{J^{-1} \sum_j h(Z_j^J)  \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)} {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}} \stackrel{a.s.}{\to} \E_{f_Z}\left[h(Z) \frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\right] = \E_{f_X}\left[\E_{f_{Z|X}}\left[h(Z)\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right],$$
+    while the reciprocal of the denominator converges to 
+    $$ \frac{J^{-1} \sum_j \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)}{U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}  \stackrel{a.s.}{\to} \E_{f_Z}\left[ \frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\right] = \E_{f_X}\left[\E_{f_{Z|X}}\left[\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right] .$$
+
+    The numerator is exactly what we desire, so it remains to show that the reciprocal of the denominator converges to $1$. Applying Fubini, we have that
+    \begin{align*}
+        \E_{f_X}\left[\E_{f_{Z|X}}\left[\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right] &= \int_x\left(\int_z \frac{f_{Y|X}(z|x)}{f_{Z|X}(z|x)}f_{Z|X}(z|x)dz \right)f_X(x) dx \\
+        &= \int_x\int_z f_{Y|X}(z|x) f_X(x) dzdx \\
+        &= \int_z\int_x f_{Y|X}(z|x) f_X(x) dxdz \\
+        &= \int_z f_{Y}(z) dz \\
+        &= 1,
+    \end{align*}
+    and we are done.
+
 \end{proof}
@@
 
+\begin{prop}[DOP-1 Targets the Posterior]
+    When $\alpha=1$ or $\phi=\theta$, DOP-$\alpha$ targets the posterior. 
+\end{prop}
+\begin{proof}
+    When $\theta=\phi$, regardless of the value of $\alpha$, the ratio $\frac{g_{n,j}^\theta h_{n,j}^\theta}{g_{n,j}^\phi h_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter estimate.
+
+    When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows.
+Recursively applying Lemmas \ref{lem:change-measure-proper-weights}, \ref{lem:posterior-proper-weights}, and \ref{lem:change-marginal-proper-weights}, we obtain that
+    %to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
+    the DOP-1 filter targets the posterior.
+    Specifically, suppose inductively that $\big\{\big(X^{F,\phi}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
+    Then, Lemma \ref{lem:change-marginal-proper-weights} tells us that $\big\{\big(X^{P,\phi}_{n,j},w^{P,\theta}_{n,j} h_{n,j}^\theta /h_{n,j}^\phi\big)\big\}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\phi}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j}h_{n,j}^\theta /h_{n,j}^\phi \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+    Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
+    \[
+    \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \left(X^{P,\phi}_{n,a(j)}, w^{P,\theta}_{n,a(j)} \frac{g^\theta_{n,a(j)}h_{n,a(j)}^\theta}{g^\phi_{n,a(j)}h_{n,a(j)}^\phi}\right),
+    \]
+    with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+\end{proof}
+
 This has addressed filtering, but not quite yet the likelihood evaluation. For this we use the following lemma.
@@
 
-Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ in the case of MOP and $\{ \big(X^{P,\phi}_{n,j},w^{P,\theta}_{n,j}\big) \}$ in the case of DOP (with a slight abuse of notation here, as the weight expressions are in fact different for both algorithms) targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
 To see  (\ref{L2}),

### 2024-05-30 paper/ms.tex
update from overleaf
some edits to abstract and intro

@@
   Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods.
-Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques as widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters.
-Furthermore, most approaches are not compatible with the desire for simulation-based inference without access to the system's transition probabilities.
-We propose a new representation that embeds existing methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff.
+Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques because widely used particle filter algorithms yield an estiamated likelihood function that is discontinuous as a function of the model parameters.
+Furthermore, most approaches to avoid this problem require access to the system's transition probabilities, and this can be prohibitive.
+We propose a new representation that embeds existing AD methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff.
 Further, we develop optimization algorithms suited to the Monte Carlo properties of the derivative estimate, including a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
-Promising numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from iterated filtering can beat current state-of-the-art methods on a challenging scientific benchmark problem.
+Promising numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from an iterated filtering algorithm can beat current state-of-the-art methods on a challenging scientific benchmark problem.
 \end{abstract}
@@
 
-\firstpage{5}
+\firstpage{3}
 % Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting.
@@
 
-% If your first paragraph (i.e.
-with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation.
-If this is the case, add \parshape=0 to the end of the list environment.
-\dropcap{M}any scientific models involve highly nonlinear stochastic dynamic systems, where there may be significant uncertainty in both the process dynamics and the measurements of the system.
-Despite the ubiquity of such partially observed stochastic processes with the Markov property (also known as, continuous-state continuous-time hidden Markov models (HMMs), partially-observed Markov processes (POMPs), or partially-observed stochastic nonlinear dynamical systems) in fields as diverse as epidemiology \cite{he10, stocks17}, ecology \cite{knape12}, and finance \cite{kim08, breto14}, estimation and inference within this class of models is a difficult problem with several challenges.
+\dropcap{M}any scientific models involve highly nonlinear stochastic dynamic systems possessing significant random variation in both the process dynamics and the measurements.
+Commonly, the latent system is modeled as a Markov process, giving rise to a partially observed Markov process (POMP) model, also known as a hidden Markov models or a state space model.
+POMP models arise in fields as diverse as epidemiology \cite{he10, stocks17}, ecology \cite{knape12}, and finance \cite{kim08, breto14}.
+Despite their ubiquity, estimation and inference within this broad class of models remains a challenging problem.
 
-Popular methods for estimation and inference in these scenarios (such as the EM algorithm, the various Kalman filter variants, and MCMC) either struggle with complex models that have intractable likelihood functions, or assume access to state transition densities.
-This is a problem in some critical applications, such as disease modeling, where transition densities are unavailable.
-However, in such cases a simulator may be readily available.
-In this case, the particle filter with the bootstrap proposal, a popular method for solving the filtering problem in partially-observed dynamical systems, provides an unbiased estimate of the likelihood \cite{delmoral2004feynman} without requiring evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
-
-A series of related algorithms \cite{welch2009abc, wood2010sl, doucet2010pmcmc, ionides08, ionides15} enable parameter estimation within these models.
-In particular, the improved iterated filtering algorithm (IF2) \cite{ionides15} is the only simulation-based full-information maximum likelihood method available, whereas particle MCMC \cite{doucet2010pmcmc} and $\text{SMC}^2$ \cite{chopin13} are the only simulation-based full-information Bayesian methods available in the literature thus far.
-However, accurate parameter estimation with significant Monte-Carlo noise in the likelihood estimate is challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small.
+The particle filter, also known as sequential Monte Carlo, serves as the foundation for various inference algorithms for POMP models.
+It provides an unbiased estimate of the likelihood function  \cite{delmoral2004feynman}, enabling Bayesian inference \cite{doucet2010pmcmc,chopin13} and maximum likelihood inference \citep{ionides06,ionides15}.
+An attractive feature of basic particle filter algorithms, also known as boostrap filters, is that they do not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
+This plug-and-play property is useful for scientific applications \cite{he10}.
+Some plug-and-play methods depend on the construction of low-dimensional summary statistics \cite{wood2010sl,welch2009abc}, sacrificing statistical efficiency for computational convenience.
+However, particle filter inference methods based on the likelihood are statistically efficient, up to Monte Carlo error.
+[MAYBE ADD SOME REFERENCES HERE TO DOCUMENT SUCCESSES OF IF2, SINCE WE ARE COMPARING WITH THAT AND BUILDING ON IT]
+This Monte Carlo error can be non-negligible.
 For example, though in practice the algorithm of Ionides et al.
 \cite{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood.
-This necessitates a new approach.
+The issue becomes increasingly problematic as the dataset size and model complexity increase, necessitating a new approach.
 
@@
     \centering
-    \includegraphics[width=\textwidth/3]{imgs/095/biasvar.png}
+    \includegraphics[width=\textwidth/3]{../imgs/095/biasvar.png}
     \caption{Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}.
@@
     \centering
-    \includegraphics[width=\textwidth/2]{imgs/095/tikzcholera.png}
+    \includegraphics[width=\textwidth/2]{../imgs/095/tikzcholera.png}
     \vspace*{-7mm}
@@
 \textbf{Left:} Paired searches from the same starting point.
+Controlling for initial starting point, tuning $\alpha$ allows IFAD-$0.97$ to strictly improve on IF2, \cite{poyiadjis11}, and \cite{blei2018vsmc}, on almost every iteration.
 \textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches.
@@
 \end{figure}
+
+\kevin{Include plots of IFAD vs Poyiadjis and Naesseth only in the supplemental information}
 
@@
 
-\bibliographystyle{pnas-new}
+% \bibliographystyle{pnas-new}
 % Bibliography

### 2024-05-30 paper/ms.tex
update bib in paper/ms

@@
 The particle filter, also known as sequential Monte Carlo, serves as the foundation for various inference algorithms for POMP models.
-It provides an unbiased estimate of the likelihood function  \cite{delmoral2004feynman}, enabling Bayesian inference \cite{doucet2010pmcmc,chopin13} and maximum likelihood inference \citep{ionides06,ionides15}.
+It provides an unbiased estimate of the likelihood function  \cite{delMoral04}, enabling Bayesian inference \cite{andrieu10,chopin13} and maximum likelihood inference \citep{ionides06-pnas,ionides15}.
 An attractive feature of basic particle filter algorithms, also known as boostrap filters, is that they do not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
 This plug-and-play property is useful for scientific applications \cite{he10}.
-Some plug-and-play methods depend on the construction of low-dimensional summary statistics \cite{wood2010sl,welch2009abc}, sacrificing statistical efficiency for computational convenience.
+Some plug-and-play methods depend on the construction of low-dimensional summary statistics \cite{wood10,toni09}, sacrificing statistical efficiency for computational convenience.
 However, particle filter inference methods based on the likelihood are statistically efficient, up to Monte Carlo error.
@@
 
-Recent advances in automatic differentiation (AD) for particle filters \cite{blei2018vsmc, jon2018diffpf, corenflos21, scibior2021dpf, doucet2022particlebased} have drawn attention to AD as a tool for inference in partially observed stochastic processes.
-However, existing approaches are either asymptotically biased \cite{blei2018vsmc, jon2018diffpf}, have high variance \cite{poyiadjis11, scibior21}, computationally expensive \cite{corenflos21, chen24}, or need access to transition densities \cite{poyiadjis11, scibior21, doucet2022particlebased, chen24}.
+Recent advances in automatic differentiation (AD) for particle filters \cite{naesseth18, jonschkowski18, corenflos21, scibior21, singh22} have drawn attention to AD as a tool for inference in partially observed stochastic processes.
+However, existing approaches are either asymptotically biased \cite{naesseth18, jonschkowski18}, have high variance \cite{poyiadjis11, scibior21}, computationally expensive \cite{corenflos21, chen24}, or need access to transition densities \cite{poyiadjis11, scibior21, singh22, chen24}.
 
@@
 To do so, we construct a suitably reweighted family of particle filters that we call \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$), for some $\alpha \in [0,1]$ that optimizes a bias-variance tradeoff.
-This family encompasses the biased gradient estimator of \cite{blei2018vsmc} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
+This family encompasses the biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
 It does not require access to the transition densities, and works with just a differentiable simulator.
@@
 
-These improvements also extend to Bayesian inference, as we show that we can use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with a nonparametric empirical Bayes-style prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{doucet2010pmcmc} from the $1.4 \times 10^6$ iterations in \cite{fasiolo16} to just $500$.
+These improvements also extend to Bayesian inference, as we show that we can use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with a nonparametric empirical Bayes-style prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{andrieu10} from the $1.4 \times 10^6$ iterations in \cite{fasiolo16} to just $500$.
 
@@
 We construct an algorithm called \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) that targets the posterior (i.e.
-almost sure convergence of every measurable bounded functional of particles to its expectation under the posterior) and likelihood, and whose gradient yields a score estimate encompassing those of \cite{poyiadjis11, scibior21, blei2018vsmc}.
+almost sure convergence of every measurable bounded functional of particles to its expectation under the posterior) and likelihood, and whose gradient yields a score estimate encompassing those of \cite{poyiadjis11, scibior21,naesseth18}.
 
@@
 When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, recovering the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}.
-When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{blei2018vsmc}.
+When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{naesseth18}.
 
-\subsection{MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, blei2018vsmc}}
+\subsection{MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, naesseth18}}
 
-\cite{scibior21} show that the estimate of \cite{blei2018vsmc} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through applying a ``stop-gradient trick" to the particle filter.
+\cite{scibior21} show that the estimate of \cite{naesseth18} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through applying a ``stop-gradient trick" to the particle filter.
 It turns out that both of these, when applied on the bootstrap filter, correspond to special cases of MOP-$\alpha$.
@@
     \end{equation}
-    yielding \cite{blei2018vsmc} on the bootstrap filter. When $\alpha=1$,
+    yielding \cite{naesseth18} on the bootstrap filter. When $\alpha=1$,
     \vspace*{-2.5mm}
@@
 This further illustrates how $\alpha$ dictates the "memory" of the gradient estimate.
-As \cite{scibior2021dpf} remark, the MOP-$0$ estimator of \cite{blei2018vsmc} only considers single-step quantities, and is "memoryless" beyond a single step.
+As \cite{scibior21} remark, the MOP-$0$ estimator of \cite{naesseth18} only considers single-step quantities, and is "memoryless" beyond a single step.
 This is in contrast to the case when $\alpha=1,$ as that estimate of \cite{poyiadjis11} only considers the surviving particles at time $N$, fully tracking dependencies over time.
@@
 Although we defer the theoretical analysis of the resulting algorithm, MOP-$\alpha$, to Section \ref{sec:thms}, we highlight a few key results below.
-MOP-$\alpha$ targets the posterior and likelihood (Theorem \ref{thm:mop-targeting}), yields the estimators of \cite{poyiadjis11, scibior21, blei2018vsmc} as special cases when differentiated (Theorem \ref{thm:mop-functional-forms}) where MOP-$1$ in particular is consistent for the score (Theorem \ref{thm:mop-grad-consistency}), has rates for its bias and variance under different choices of $\alpha$ (Theorem \ref{thm:mop-biasvar}), and enjoys a linear rate of convergence for gradient descent with the resulting gradient estimate (Theorem \ref{thm:mop-convergence}).
+MOP-$\alpha$ targets the posterior and likelihood (Theorem \ref{thm:mop-targeting}), yields the estimators of \cite{poyiadjis11, scibior21, naesseth18} as special cases when differentiated (Theorem \ref{thm:mop-functional-forms}) where MOP-$1$ in particular is consistent for the score (Theorem \ref{thm:mop-grad-consistency}), has rates for its bias and variance under different choices of $\alpha$ (Theorem \ref{thm:mop-biasvar}), and enjoys a linear rate of convergence for gradient descent with the resulting gradient estimate (Theorem \ref{thm:mop-convergence}).
 
@@
 
-We use the Dhaka cholera model of \cite{king08} to benchmark (also used or modified in \cite{ionides15, wood16, wycoff2024voronoi} for the same purpose) the performance of our method.
+We use the Dhaka cholera model of \cite{king08} to benchmark (also used or modified in \cite{ionides15, wood16, wycoff24} for the same purpose) the performance of our method.
 This is a stochastic SIR compartmental model where the population at time $t$, $H_t$, is divided into the susceptible compartment $S_t$, infected compartment $I_t$, and three ($k=3$) recovered compartments $R^1_t, ..., R^k_t$ denoting varying degrees of cholera immunity.
@@
 \textbf{Left:} Paired searches from the same starting point.
-Controlling for initial starting point, tuning $\alpha$ allows IFAD-$0.97$ to strictly improve on IF2, \cite{poyiadjis11}, and \cite{blei2018vsmc}, on almost every iteration.
+Controlling for initial starting point, tuning $\alpha$ allows IFAD-$0.97$ to strictly improve on IF2, \cite{poyiadjis11}, and \cite{naesseth18}, on almost every iteration.
 \textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches.
@@
 
-Here, we will show that MOP-$\alpha$ targets the posterior and likelihood, yields the estimators of \cite{poyiadjis11, scibior21, blei2018vsmc} as special cases when differentiated, and characterize rates for its bias and variance under different choices of $\alpha$.
+Here, we will show that MOP-$\alpha$ targets the posterior and likelihood, yields the estimators of \cite{poyiadjis11, scibior21, naesseth18} as special cases when differentiated, and characterize rates for its bias and variance under different choices of $\alpha$.
 To do so, we require the following assumptions:
@@
 In comparison, often 100 or 200 iterations of IF2 are used for an initial global search.
-\cite{ionides15}, for example, use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{Lee_haiticholera}.
+\cite{ionides15}, for example, use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{lee20}.
 
@@
 % Bibliography
-\bibliography{bib-ifad, bib-ref}
+\bibliography{bib-ifad}
 

### 2024-05-30 paper/ms.tex
Updates from Overleaf

--- 2024-05-10 paper/ms.tex
@@
     \label{thm:mop-biasvar}
-    When $\alpha\in(0,1)$, $\theta=\phi$, the MSE and variance of MOP-$\alpha$, where $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$, are:
+    When $\alpha\in(0,1)$, $\theta=\phi$, the MSE and variance of MOP-$\alpha$, where $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$ and $\epsilon$ depends on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karlajainen2001} are:
     \vspace*{-1ex}

### 2024-05-31 paper/ms.tex
more on intro and references

--- 2024-05-30 paper/ms.tex
@@
 
-In particular, Scibior and Wood \cite{scibior21} show that the estimators derived in Poyiadjis et al.
-\cite{poyiadjis11} can be attained with AD, through with a simple tweak (the "stop-gradient trick") to the particle filter that does not modify its likelihood estimates.
-However, their method is not fully compatible with simulation-based inference and the mathematical motivation behind the stop-gradient trick is not apparent.
+Scibior and Wood \cite{scibior21} showed that the estimators derived by Poyiadjis et al.
+\cite{poyiadjis11} can be attained with standard AD software, using an algorithmic prodecure called a ``stop-gradient'' which allows selected exressions to be evaluated but not differentiated.
+Unfortunately, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo error.
+We use \cite{scibior21} and \cite{poyiadjis11} as our starting point, while developing a new approach which remedies the weaknesses of these papers.
 
-%We show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.
+We start by presenting a new construction of the estimator of \cite{scibior21} and \cite{poyiadjis11}.
+Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a Measurement Off-Policy (MOP) particle filter.
+We procced to generalize the MOP particle filter by adding a discounting factor, $\alpha \in [0,1]$ , which interpolates between the  biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
+The bias-variance tradeoff induced by $\alpha$ suggests, both in theory and in practice, the use of $\alpha$ values strictly between 0 and 1.
 
-\kevin{Can someone who's reading through the words and glazing through the math get the idea that (1) we're running AD on the PF, and (2) using an off-policy resampling trick to get our estimator.
-Should I hedge by writing the SI in Del Moral language?
-}
-
-\subsection{Our Contributions}
-We develop a new approach to statistical inference via AD for particle filters which addresses the above concerns.
-We construct a new theoretical framework that bypasses the paradox of differentiating through a Monte-Carlo algorithm with discontinuous resampling.
-
-Our key observation is that one can turn this into a problem of differentiating through a series of smooth measurement density ratios and (through the reparametrization trick) a differentiable simulator.
-To do so, we construct a suitably reweighted family of particle filters that we call \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$), for some $\alpha \in [0,1]$ that optimizes a bias-variance tradeoff.
-This family encompasses the biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
-It does not require access to the transition densities, and works with just a differentiable simulator.
-
-We show that these off-policy particle filters successfully target the posterior, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties.
-We also derive a linear convergence rate for gradient methods on the particle filter in the presence of strong convexity, providing support for numerical optimization with AD for the particle filter.
-
-This allows us to develop improved algorithms for maximum likelihood/MAP estimation in highly nonlinear stochastic dynamical systems.
-We propose a hybrid algorithm called \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts iterative first or second-order gradient methods with a coarse solution obtained from a few rounds of iterated filtering.
+We then derive a linear convergence rate for stochastic gradient descent (SGD) using MOP($\alpha$) in the  presence of strong convexity.
+Critically, the estimator is effective within a neighborhood of the maximum but is relatively slow at reaching this neighborhood.
+That behavior is complementary to that of the iterated filtering algorithm \cite{ionides06-pnas,ionides15} which provides a relatively fast and stable way to identify this neighborhood.
+We therefore build a hybrid algorithm called \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts iterative first or second-order gradient methods with a coarse solution obtained from a few rounds of iterated filtering.
 Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.
@@
     \label{thm:mop-biasvar}
-    When $\alpha\in(0,1)$, $\theta=\phi$, the MSE and variance of MOP-$\alpha$, where $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$, are:
+    When $\alpha\in(0,1)$, $\theta=\phi$, the MSE and variance of MOP-$\alpha$, where $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$ and $\epsilon$ depends on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karjalainen23} are:
     \vspace*{-1ex}

### 2024-05-31 paper/ms.tex
more edits to intro and refs

@@
 
-\firstpage{3}
+\firstpage[4]{4}
 % Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting.
@@
 However, particle filter inference methods based on the likelihood are statistically efficient, up to Monte Carlo error.
-[MAYBE ADD SOME REFERENCES HERE TO DOCUMENT SUCCESSES OF IF2, SINCE WE ARE COMPARING WITH THAT AND BUILDING ON IT]
-This Monte Carlo error can be non-negligible.
+Plug-and-play methods are often called likelihood-free \cite{owen15}, and it may be counter-intuitive that likelihood-based inference is possible using likelihood-free methods. 
+However, iterated filtering algorithms have shown that this is practical in various scientific investigations \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
+Nevertheless, the Monte Carlo variability arising in iterated filtering applications becomes increasingly non-negligible as the size of the data and the complexity of the model increases.
+
 For example, though in practice the algorithm of Ionides et al.

### 2024-06-03 paper/ms.tex
updates from github

--- 2024-05-30 paper/ms.tex
@@
 
-\title{Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation}
+\title{Automatic Differentiation Accelerates Inference for Partially Observed Stochastic Processes}
 
@@
 
-\firstpage{3}
+\firstpage{4}
 % Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting.
@@
 Despite their ubiquity, estimation and inference within this broad class of models remains a challenging problem.
+For instance, in even moderately complex models, the E-step within the widely-celebrated EM algorithm can be intractable \cite{picchini2018}, necessitating the use of Monte Carlo methods, as we outline below.
 
@@
 An attractive feature of basic particle filter algorithms, also known as boostrap filters, is that they do not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
-This plug-and-play property is useful for scientific applications \cite{he10}.
-Some plug-and-play methods depend on the construction of low-dimensional summary statistics \cite{wood10,toni09}, sacrificing statistical efficiency for computational convenience.
+This property is especially useful for scientific applications \cite{he10}.
+Some methods satisfying this depend on the construction of low-dimensional summary statistics \cite{picchini2018,wood10,toni09}, sacrificing statistical efficiency for computational convenience.
 However, particle filter inference methods based on the likelihood are statistically efficient, up to Monte Carlo error.
-[MAYBE ADD SOME REFERENCES HERE TO DOCUMENT SUCCESSES OF IF2, SINCE WE ARE COMPARING WITH THAT AND BUILDING ON IT]
+\ed{MAYBE ADD SOME REFERENCES HERE TO DOCUMENT SUCCESSES OF IF2, SINCE WE ARE COMPARING WITH THAT AND BUILDING ON IT}
 This Monte Carlo error can be non-negligible.
@@
 However, existing approaches are either asymptotically biased \cite{naesseth18, jonschkowski18}, have high variance \cite{poyiadjis11, scibior21}, computationally expensive \cite{corenflos21, chen24}, or need access to transition densities \cite{poyiadjis11, scibior21, singh22, chen24}.
+\kevin{Can someone who's reading through the words and glazing through the math get the idea that (1) we're running AD on the PF, and (2) using an off-policy resampling trick to get our estimator?}
 
@@
 
-\kevin{Can someone who's reading through the words and glazing through the math get the idea that (1) we're running AD on the PF, and (2) using an off-policy resampling trick to get our estimator.
-Should I hedge by writing the SI in Del Moral language?
-}
+\kevin{Also, hedge by writing the SI in Del Moral language.}
 
@@
 
-These improvements also extend to Bayesian inference, as we show that we can use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with a nonparametric empirical Bayes-style prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{andrieu10} from the $1.4 \times 10^6$ iterations in \cite{fasiolo16} to just $500$.
+These improvements also extend to Bayesian inference, as we show \kevin{Include in main paper or SI?} that we can use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with a nonparametric empirical Bayes-style prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{andrieu10} from the $1.4 \times 10^6$ iterations in \cite{fasiolo16} to just $500$.
 
@@
 We construct an algorithm called \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) that targets the posterior (i.e.
-almost sure convergence of every measurable bounded functional of particles to its expectation under the posterior) and likelihood, and whose gradient yields a score estimate encompassing those of \cite{poyiadjis11, scibior21,naesseth18}.
+establishes almost sure convergence of every measurable bounded functional of particles to its expectation under the posterior) and likelihood, and whose gradient yields a score estimate encompassing those of \cite{poyiadjis11, scibior21,naesseth18}.
 
@@
 If $\theta$ and $\phi$ coincide, one can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$, as seen in Equation \ref{eq:weighting-scheme}, while still targeting the posterior and likelihood at $\theta$.
-This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate.
-When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, recovering the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}.
-When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{naesseth18}.
+This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate, interpolating between the estimators of \cite{naesseth18} and \cite{poyiadjis11}, as seen below in Proposition \ref{thm:mop-functional-forms}.
 
-\subsection{MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, naesseth18}}
+In short, when $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, recovering the consistent but high-variance gradient estimate from \cite{poyiadjis11} that can be obtained by applying a ``stop-gradient trick" \cite{scibior21} to the particle filter.
+When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{naesseth18} that \cite{scibior21} show is the gradient of a vanilla particle filter when resampling terms are dropped.
 
-\cite{scibior21} show that the estimate of \cite{naesseth18} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through applying a ``stop-gradient trick" to the particle filter.
-It turns out that both of these, when applied on the bootstrap filter, correspond to special cases of MOP-$\alpha$.
-
-$\alpha$ therefore controls an exponentially-weighted moving average, determining how much "memory" the resulting gradient estimate has over ancestral trajectories.
-This is illustrated in Figure \ref{fig:biasvar}, in the context of the Dhaka cholera model of \cite{king08}.
-
-\begin{thm}[MOP-$0$ and MOP-$1$ Functional Forms]
+\begin{prop}[MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, naesseth18}]
     \label{thm:mop-functional-forms}
@@
     yielding the estimator of \cite{poyiadjis11, scibior21} with the bootstrap filter.
-\end{thm}
+\end{prop}
 
@@
 
-This further illustrates how $\alpha$ dictates the "memory" of the gradient estimate.
+$\alpha$ therefore controls an exponentially-weighted moving average, determining how much "memory" the resulting gradient estimate has over ancestral trajectories.
 As \cite{scibior21} remark, the MOP-$0$ estimator of \cite{naesseth18} only considers single-step quantities, and is "memoryless" beyond a single step.
 This is in contrast to the case when $\alpha=1,$ as that estimate of \cite{poyiadjis11} only considers the surviving particles at time $N$, fully tracking dependencies over time.
+We demonstrate this bias-variance tradeoff in practice in Figure \ref{fig:biasvar}, in the context of the Dhaka cholera model of \cite{king08}.
 
@@
 This still leaves us with the question of designing an effective procedure for likelihood maximization with MOP-$\alpha$.
-We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that runs a few iterations of IF2 to warm-start an iterative first or second-order method that uses the MOP-$\alpha$ gradient estimate.
+We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that runs a few iterations of IF2 to warm-start an iterative (first or second-order) method that uses the MOP-$\alpha$ gradient estimate.
 This leverages IF2's quick empirical convergence to a neighborhood of the MLE, overcoming the tendency of gradient methods to get stuck in saddle points and local minima.
@@
 Combining these two methods in this way lets us enjoy the best of both worlds.
-\kevin{source for this}
+
+The quick empirical convergence of IF2 (and so the first stage of IFAD) to a neighborhood of the MLE happens fairly quickly in practice.
+In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations.
+Finding the MLE itself with IF2, however, takes much longer, as one has to use a less aggressive cooling rate to do so.
+For example, \cite{ionides15} use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{lee20}.
+By simply requiring that the first stage of IFAD get to a neighborhood of the MLE and not the MLE itself, we are able to substantially reduce the number of IF2 iterations required.
 
 \subsection{Linear Convergence Rates}
-\begin{thm}
+
+On the other hand, the second stage of IFAD enjoys a linear convergence rate under the usual strongly-convex and smoothness assumptions in convex optimization, as we show below in Theorem \ref{thm:mop-convergence}.
+
+\begin{thm}[Linear Convergence of IFAD]
+
+Consider the second stage of IFAD (Algorithm \ref{alg:ifad}) where one stops if $||\nabla_\theta \hat\ell^\alpha(\theta_n)|| \leq (1+\sigma) \epsilon$, where $\sigma \geq \frac{4 \Gamma}{(1-\beta)}$.
+Assume $-\ell$ is strongly convex and smooth, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$.
+Choose the learning rate $\eta$ and error tolerance $\epsilon$ such that $\eta \leq \frac{c(1-\beta)}{2\Gamma}$ for some $\beta \in (0,1)$.
+Then, for sufficiently large $\alpha$ and $J$ to ensure the score estimate is a $1/2$-approximation, the second stage of IFAD converges linearly to the MLE with probability at least $1-\delta$:
+\vspace*{-2mm}
+$$
+\ell(\theta^*) - \ell(\theta_{n+1}) \leq \left(1-\eta\beta\frac{8\gamma}{9c}\right)(\ell(\theta^*)-\ell(\theta_n)).
+$$
     \label{thm:mop-convergence}
 \end{thm}
+\vspace*{-5mm}
+
+We defer the proof, which is similar to that of Theorem 6 in \cite{mahoney16}, to the supplementary information.
+For simplicity, we only prove this for MOP-$1$ and note that the general case follows as the required $1/2$-approximation can still be obtained for sufficiently large $\alpha$.
+
+We therefore see that the second stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the first stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
+This happens fairly often in practice, for example, when sufficient regularity conditions for local asymptotic normality of the MLE hold.
+We conjecture that this applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE, and will explore this in future work.
+
+%IF2 posesses some SGD-like qualities, including what looks like linear convergence to a ball around the MLE with width determined by the random walk standard deviation.
+We will explore this conjecture in future work.
+
+%While this theoretical guarantee currently only holds for IFAD-1, as it forms a particle approximation of the score as in \cite{poyiadjis11}, one can also extend this for a similar guarantee if the bias for the gradient estimate given by MOP-$\alpha$ for $\alpha < 1$ is small enough.
+For general $\alpha$, including MOP-0, the analysis then becomes more complicated, as one needs to handle the case of biased gradient descent.
+However, if future measurements $y_{n+1:N}^*$ do not provide a large amount of information on the identification of the current state $x_n$ given past and current measurements $y_{0:n}^*$, as mentioned in \cite{corenflos21}, the bias may be small enough for it to not present too much of a problem.
 
@@
     \label{thm:mop-biasvar}
-    When $\alpha\in(0,1)$, $\theta=\phi$, the MSE and variance of MOP-$\alpha$, where $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$ and $\epsilon$ depends on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karlajainen2001} are:
+    When $\alpha\in(0,1)$, $\theta=\phi$, $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$, there exists some $\epsilon>0$ depending on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karjalainen2023} such that the MSE and variance of MOP-$\alpha$ are:
     \vspace*{-1ex}
@@
 \end{thm}
-\kevin{Explain exactly what $\epsilon$ is here, and interpret the rates as agreeing with our intuition.
+
+We defer the proof to the supplementary material, but provide a brief outline here.
+The variance bound can be reduced to the approximation error between MOP-$\alpha$ and a variant called MOP-$(\alpha,k)$ where the weights are truncated at lag $k$ with discount factor $\alpha$ to ensure strong mixing, and the covariance of the latter which can be controlled with Davydov's inequality and a standard $L^p$ error bound for the particle filter.
+The MSE bound considers the error between the score and MOP-$(1,k)$, and the error between MOP-$(1,k)$ and MOP-$\alpha$.
+The latter is $\tilde{O}(NpG'(\theta)^2(k+\psi(\alpha)))$ by tedious algebra, and the former can be controlled with a result on the the forgetting of the particle filter from \cite{karjalainen2023} and the very same $L^p$ error bound mentioned above.
+
+We also show in the supplementary material that the variance of MOP-$0$ is $\tilde{O}(NpG'(\theta)^2/J)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$ (ignoring factors of $p, G'$, etc.).
+This and Theorem \ref{thm:mop-biasvar} show that MOP-$\alpha$ interpolates between MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$.
+
+%\kevin{Explain exactly what $\epsilon$ is here, and interpret the rates as agreeing with our intuition.
 Talk about the relationship between the theory and the empirical results in the bias-variance tradeoff figure above, as well as the phase transition between $Nk^2/J$ and $N^4/J$ variance with MOP-$\alpha$ and MOP-$1$, where $k$ is really a surrogate for $\alpha$ -- with more memory, you truncate less.
@@
 
-We defer the proof to the supplementary material in favor of a very brief outline here.
+\section{Computational Efficiency}
 
-\kevin{Say that our framework suggests whole new classes of algorithms, as it includes Naesseth and Poyiadjis.
-Interpolating between them with discounting is only one way to interpolate, we can also e.g.
-truncate at a fixed lag.
-Remark that the analysis is similar for MOP-$(1,k)$ as a fixed-lag filter, with slightly better rates but a less natural interpretation.
-We leave empirical studies of its performance to future work.
-}
+MOP-$\alpha$ and IFAD are fast algorithms, both in theory and practice.
+In line with the cheap gradient principle of \cite{kakade2019provably}, getting a gradient estimate from MOP-$\alpha$ takes no more than 6 times that of the runtime of the particle filter (in practice, the multiplier is 3.75).
+It therefore shares the same $O(NJ)$ time complexity as the particle filter, unlike the $O(NJ^2)$ complexity of \cite{corenflos21} and Algorithm 2 in \cite{poyiadjis11, scibior21}.
 
-\section{Computational Considerations}
+As IF2 amounts to one iteration of the particle filter, with $O(NJ)$ time complexity.
 
-\subsection{Runtime}
-
-Each iteration of IF2 amounts to one iteration of the particle filter, with $O(NJ)$ time complexity.
-The initial "convergence" happens fairly quickly in practice.
-In the case of the Dhaka cholera model of \cite{king08}, when a geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations of IF2.
-In comparison, often 100 or 200 iterations of IF2 are used for an initial global search.
-\cite{ionides15}, for example, use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{lee20}.
-
-In line with the cheap gradient principle of \cite{kakade2019provably}, getting a gradient estimate from MOP-$\alpha$ takes no more than 6 times that of the runtime of the particle filter (in practice, the multiplier is 3.75).
-This therefore has $O(NJ)$ time complexity as well, obtaining a gradient estimate with runtime linear in the number of particles.
-
-\subsection{Python Library}
-
-Re-implementing the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, even with a simulator written in Python.
+In practice, re-implementing the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, even with a simulator written in Python.
 This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter (with a simulator written in C++) in the \texttt{pomp} package of \cite{king16}.
 
-\kevin{Write about new open source library that brings pomp to python with AD.
+%\kevin{Write about new open source library that brings pomp to python with AD.
 There are implementations in tfp under the jax substrate, but these are not plug and play -- they require you to supply the transition densities.
@@
 We note that it is possible to work without a differentiable simulator, but one would then require access to the transition densities.
+We are working on an analogue that uses (differentiable) transition densities instead of a differentiable simulator, that may be better suited to large discrete state spaces than either MOP-$\alpha$ or the Baum-Welch algorithm.
 
-Our gradient estimate can be used for variational inference to approximate the posterior distribution over parameters.
+Additionally, our gradient estimate can be used for variational inference to approximate the posterior distribution over latent states \cite{naesseth18} and parameters.
+
+Discounting the weights by some $\alpha \in [0,1]$ is not the only way to interpolate between the estimators of \cite{naesseth18} and \cite{poyiadjis11}.
+As the proof of Theorem \ref{thm:mop-biasvar} implies, we can also truncate the weights at a fixed lag, corresponding to the MOP-$(1,k)$ estimate mentioned.
+The analysis is similar, with slightly better rates but a less natural interpretation.
 
@@
 
-\acknow{Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.}
+\acknow{We thank Nicholas Chopin for helpful comments on the proof of a strong law of large numbers for a suitably reweighted particle filter. }
 

### 2024-06-04 paper/ms.tex
edits to pages 1 and 2

--- 2024-05-31 paper/ms.tex
@@
 \usepackage{tabto}
+
+\newcommand\ei[1]{\textcolor{red}{#1}}
 
@@
   Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods.
-Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques because widely used particle filter algorithms yield an estiamated likelihood function that is discontinuous as a function of the model parameters.
-Furthermore, most approaches to avoid this problem require access to the system's transition probabilities, and this can be prohibitive.
-We propose a new representation that embeds existing AD methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff.
-Further, we develop optimization algorithms suited to the Monte Carlo properties of the derivative estimate, including a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
+Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques because widely used particle filter algorithms yield an estimated likelihood function that is discontinuous as a function of the model parameters.
+We show how to embed two existing AD particle filter methods in a theoretical framework that provides an extension to a new class of algorithms within which we obtain a favorable bias/variance tradeoff.
+Further, we develop likelihood maximization algorithms suited to the Monte Carlo properties of the derivative estimate.
+Our algorithms require only a differentiable simulator for the latent dynamic system; by contrast, most previous approaches to AD likelihood maximization for particle filters require access to the system's transition probabilities, and this can be prohibitive.
 Promising numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from an iterated filtering algorithm can beat current state-of-the-art methods on a challenging scientific benchmark problem.
@@
 
-\firstpage[4]{4}
+\firstpage{4}
 % Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting.
@@
 The particle filter, also known as sequential Monte Carlo, serves as the foundation for various inference algorithms for POMP models.
-It provides an unbiased estimate of the likelihood function  \cite{delMoral04}, enabling Bayesian inference \cite{andrieu10,chopin13} and maximum likelihood inference \citep{ionides06-pnas,ionides15}.
+It provides an unbiased estimate of the likelihood function  \cite{delMoral04}, enabling Bayesian inference \cite{andrieu10,chopin13} and likelihood-based inference \citep{ionides06-pnas,ionides15}.
+Likelihood-based inference is statistically efficient \cite{pawitan01} and is robust to a moderate amount of Monte Carlo error \cite{ionides17,ning21}.
 An attractive feature of basic particle filter algorithms, also known as boostrap filters, is that they do not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
@@
 Some plug-and-play methods depend on the construction of low-dimensional summary statistics \cite{wood10,toni09}, sacrificing statistical efficiency for computational convenience.
-However, particle filter inference methods based on the likelihood are statistically efficient, up to Monte Carlo error.
 Plug-and-play methods are often called likelihood-free \cite{owen15}, and it may be counter-intuitive that likelihood-based inference is possible using likelihood-free methods. 
 However, iterated filtering algorithms have shown that this is practical in various scientific investigations \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
-Nevertheless, the Monte Carlo variability arising in iterated filtering applications becomes increasingly non-negligible as the size of the data and the complexity of the model increases.
-
-For example, though in practice the algorithm of Ionides et al.
-\cite{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood.
-The issue becomes increasingly problematic as the dataset size and model complexity increase, necessitating a new approach.
+Nevertheless, the Monte Carlo variability arising in iterated filtering applications becomes increasingly problematic as the size of the data and the complexity of the model increases.
+Algorithmic advances are needed to make plug-and-play likelihood maximization for POMP models more numerically efficient.
+We investigate the use of AD for this purpose.
 
@@
 Recent advances in automatic differentiation (AD) for particle filters \cite{naesseth18, jonschkowski18, corenflos21, scibior21, singh22} have drawn attention to AD as a tool for inference in partially observed stochastic processes.
-However, existing approaches are either asymptotically biased \cite{naesseth18, jonschkowski18}, have high variance \cite{poyiadjis11, scibior21}, computationally expensive \cite{corenflos21, chen24}, or need access to transition densities \cite{poyiadjis11, scibior21, singh22, chen24}.
+However, existing approaches are either asymptotically biased \cite{naesseth18, jonschkowski18}, have high variance \cite{poyiadjis11, scibior21}, high computational expense \cite{corenflos21, chen24}, or need access to transition densities \cite{poyiadjis11, scibior21, singh22, chen24}.
 
@@
 \cite{poyiadjis11} can be attained with standard AD software, using an algorithmic prodecure called a ``stop-gradient'' which allows selected exressions to be evaluated but not differentiated.
-Unfortunately, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo error.
+Unfortunately, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability.
 We use \cite{scibior21} and \cite{poyiadjis11} as our starting point, while developing a new approach which remedies the weaknesses of these papers.
 
-We start by presenting a new construction of the estimator of \cite{scibior21} and \cite{poyiadjis11}.
+We start by presenting a new construction of the gradient estimator of \cite{scibior21} and \cite{poyiadjis11}.
 Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a Measurement Off-Policy (MOP) particle filter.
@@
 
-We then derive a linear convergence rate for stochastic gradient descent (SGD) using MOP($\alpha$) in the  presence of strong convexity.
+We derive a linear convergence rate for stochastic gradient descent (SGD) using MOP($\alpha$) in the  presence of strong convexity.
 Critically, the estimator is effective within a neighborhood of the maximum but is relatively slow at reaching this neighborhood.
 That behavior is complementary to that of the iterated filtering algorithm \cite{ionides06-pnas,ionides15} which provides a relatively fast and stable way to identify this neighborhood.
-We therefore build a hybrid algorithm called \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts iterative first or second-order gradient methods with a coarse solution obtained from a few rounds of iterated filtering.
+We therefore build a hybrid algorithm called {\it Iterated Filtering with Automatic Differentiation} (IFAD) that warm-starts first-order or second-order gradient methods with a preliminary solution obtained from a few rounds of iterated filtering.
 Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.
@@
 
-Consider an unobserved Markov process $(X_t)_{t \geq t_0}$, with observations $Y_1,...,Y_N$ realized at values $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$.
-The process is parameterized by an unknown parameter $\theta \in \Theta \subseteq \R^p$, where the states $X_t$ take values in the state space $\gX \subseteq \R^d$, the observations $Y_n$ take values in $\gY,$ and we write $X_n := X_{t_n}$.
+Consider an unobserved Markov process $\{X(t),t  \geq t_0\}$, with disctete-time observations $Y_1,...,Y_N$ realized at values $y_1^*,...,y_N^*$ at times $t_1,..., t_N$.
+The process is parameterized by an unknown parameter $\theta \in \Theta \subseteq \R^p$, where the state $X(t)$ take values in the state space $\gX \subseteq \R^d$, the observations $Y_n$ take values in $\gY,$ and we write $X_n := X(t_n)$.
 
@@
 
-We write $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ for the process model, writing $\texttt{process}\left(x_n; \theta\right)$ for its corresponding simulator, $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ the measurement model, and $y_n^*$ for the actual values of the observations that were observed.
+We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$, and we write writing $\texttt{process}\left(x_n\mid x_{n-1}; \theta\right)$ for its corresponding simulator.
+We set $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ to be the measurement density, and $y_n^*$ for the actual values of the observations that were observed.
 We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
-Superscripts $x_{n,j}^A$ denote the ancestral trajectory of particle $j$ at time $n$, and $a(\cdot)$ is the ancestor function.
-For simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0.$
+Superscripts $x_{n,j}^A$ denote the ancestral trajectory of particle $j$ at time $n$, and $a(\cdot)$ is the ancestor function \ei{NOT DEFINED YET.
+THIS WILL PUZZLE READERS UNLESS THEY ARE ALREADY HIGHLY FAMILIAR WITH THE TOPIC}.
+For simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0$.
+\ei{ALGORITHM 1 HAS $f_{X_0}$.
+IN GENERAL, THIS DETAIL IS BEST INCLUDED.
+INITIAL VALUES CAN BE AN IMPORTANT PART OF A DYNAMIC SYSTEM.}
 
-The above are defined on a probability space $(\Omega, \Sigma, \prob).$ 
-To maintain an algorithmic and practical interpretation of our theory, we identify the seed with an element of the sample space $\omega \in \Omega$, as a random seed determines the entire sequence of pseudo-random numbers as generated by a computer.
+The above conditional densities are defined on a probability space $(\Omega, \Sigma, \prob).$ 
+For the algorithmic interpretation of our theory, we identify the seed with an element of the sample space $\omega \in \Omega$.
+\ei{THE SEED CORRESPONDS TO ALL THE MONTE CARLO QUANTITIES GENERATED, INCLUDING THINGS THAT HAVE NO DIRECT ROLE IN THE MODEL, SO MAYBE THIS EXPLANATION NEEDS ADDITIONAL CARE.}
+The random seed determines the entire sequence of pseudo-random numbers as generated by a computer, just as the outcome $\omega\in\Omega$ generates the entire sequence of random variables.
 
@@
 
-As mentioned earlier, instead of directly differentiating through a particle filter, one can instead differentiate through the simulator and measurement densities.
-We construct an algorithm called \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) that targets the posterior (i.e.
-almost sure convergence of every measurable bounded functional of particles to its expectation under the posterior) and likelihood, and whose gradient yields a score estimate encompassing those of \cite{poyiadjis11, scibior21,naesseth18}.
+Directly differentiating through a particle filter, for a fixed value of $\omega$, leads to a biased gradient estimator due to the discreteness of the resampling operation REFS.
+We avoid this issue by constructing a particle filtering algorithm called \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) for which the resampling is invariant to $\omega$.
+Pseudocode for MOP-$\alpha$ is presented in Algorithm~\ref{alg:mop}.
+As we explain subsequently, MOP-$\alpha$ is constructed so that direct differentiation through this algorithm obtains the score estimator of  \cite{poyiadjis11, scibior21} when $\alpha=1$ and that of \cite{naesseth18} when $\alpha=0$.
+Setting $\alpha<1$ adds bias but reduces variance, raising opportunities for a favorable tradeoff.
 
@@
 \newline
-        \textbf{If $\theta \neq \phi$:}
-            Fix $\omega$, and filter at $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$. \newline
-        \textbf{Else:} Fix $\omega$, and set $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$ to be copies of $X_{n,j}^{P,\theta}, X_{n,j}^{P,\theta}$. \newline
+            \textbf{First pass:} Set $\theta=\phi$ and fix $\omega$. This defined $X_{n,j}^{P,\phi}$, $X_{n,j}^{F,\phi}$, $g^{\phi}_{n,j}$. \newline
+        \textbf{Second pass:}
+            Fix $\omega$, and filter at $\theta\neq phi$. \newline
+
 		\textbf{Initialize } particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, weights $w^{F,\theta}_{0,j}= 1$. \newline
@@
             \hspace*{4mm} Select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) \propto g^{\phi}_{n,m}$. \newline
-            \hspace*{4mm} Resample particles ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$. \newline
+            \hspace*{4mm} Resample particles ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$. \ei{$\phi$ SHOULD BE $\theta$} \newline
             \hspace*{4mm} Resample corrected weights
@@
 \subsection{Algorithm Outline} 
-We obtain two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but the resampling indices from the run at $\phi$ (hence the "off-policy" moniker).
+We obtain two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but the resampling indices constructed from the first pass, with parameter $\phi$.
+We call this ``off-policy'' measurement since a parameter value corresponds to a policy in the context of control theory.
+\ei{I REWROTE THIS TO BE MORE EXPLICIT, BUT MAYBE THIS NEEDS MORE EDITING. IT MAKES ME THINK THAT ``RESAMPLING OFF-POLICY'' WOULD BE A BETTER NAME!}
 If $\theta$ and $\phi$ coincide, one only needs one particle filter run at $\theta=\phi$, otherwise one needs two runs at the same seed $\omega \in \Omega$.

### 2024-06-04 paper/ms.tex
finished a round of editing of paper/ms

@@
 \usepackage{tabto}
-
-\newcommand\ei[1]{\textcolor{red}{#1}}
 
@@
 
-We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$, and we write writing $\texttt{process}\left(x_n\mid x_{n-1}; \theta\right)$ for its corresponding simulator.
+We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$, and we write $\texttt{process}\left(x_n\mid x_{n-1}; \theta\right)$ for its corresponding simulator \ed{WE DON'T USE THIS...
+MAYBE CHANGE HERE, OR USE IT}.
 We set $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ to be the measurement density, and $y_n^*$ for the actual values of the observations that were observed.
 We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
-Superscripts $x_{n,j}^A$ denote the ancestral trajectory of particle $j$ at time $n$, and $a(\cdot)$ is the ancestor function \ei{NOT DEFINED YET.
+Superscripts $x_{n,j}^A$ denote the ancestral trajectory of particle $j$ at time $n$, and $a(\cdot)$ is the ancestor function \ed{NOT DEFINED YET.
 THIS WILL PUZZLE READERS UNLESS THEY ARE ALREADY HIGHLY FAMILIAR WITH THE TOPIC}.
 For simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0$.
-\ei{ALGORITHM 1 HAS $f_{X_0}$.
+\ed{ALGORITHM 1 HAS $f_{X_0}$.
 IN GENERAL, THIS DETAIL IS BEST INCLUDED.
@@
 For the algorithmic interpretation of our theory, we identify the seed with an element of the sample space $\omega \in \Omega$.
-\ei{THE SEED CORRESPONDS TO ALL THE MONTE CARLO QUANTITIES GENERATED, INCLUDING THINGS THAT HAVE NO DIRECT ROLE IN THE MODEL, SO MAYBE THIS EXPLANATION NEEDS ADDITIONAL CARE.}
+\ed{THE SEED CORRESPONDS TO ALL THE MONTE CARLO QUANTITIES GENERATED, INCLUDING THINGS THAT HAVE NO DIRECT ROLE IN THE MODEL, SO MAYBE THIS EXPLANATION NEEDS ADDITIONAL CARE.}
 The random seed determines the entire sequence of pseudo-random numbers as generated by a computer, just as the outcome $\omega\in\Omega$ generates the entire sequence of random variables.
@@
             \hspace*{4mm} Select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) \propto g^{\phi}_{n,m}$. \newline
-            \hspace*{4mm} Resample particles ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$. \ei{$\phi$ SHOULD BE $\theta$} \newline
+            \hspace*{4mm} Resample particles ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$. \ed{$\phi$ SHOULD BE $\theta$} \newline
             \hspace*{4mm} Resample corrected weights
@@
 We call this ``off-policy'' measurement since a parameter value corresponds to a policy in the context of control theory.
-\ei{I REWROTE THIS TO BE MORE EXPLICIT, BUT MAYBE THIS NEEDS MORE EDITING. IT MAKES ME THINK THAT ``RESAMPLING OFF-POLICY'' WOULD BE A BETTER NAME!}
+\ed{I REWROTE THIS TO BE MORE EXPLICIT, BUT MAYBE THIS NEEDS MORE EDITING. IT MAKES ME THINK THAT ``RESAMPLING OFF-POLICY'' WOULD BE A BETTER NAME!}
+The resampling indices, which we write as $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$, are a function of $\phi$ for any value of $\theta$.
+For the categorical distribution, we use systematic resamplig \cite{arulampalam02,king16} which usually has superior performance to multinomial resampling.
 If $\theta$ and $\phi$ coincide, one only needs one particle filter run at $\theta=\phi$, otherwise one needs two runs at the same seed $\omega \in \Omega$.
 
-We can then reweight the conditional likelihoods by a correction factor accumulated over time to account for the resampling under $\phi$.
-That is, writing $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density, $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$, and $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$ for the resampling indices under $\phi$, we can estimate the conditional likelihood under $\theta$ as
+We then reweight the conditional likelihoods by a correction factor accumulated over time to account for the resampling under $\phi$.
+That is, writing $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density, $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$,  we can estimate the conditional likelihood under $\theta$ as
 \vspace*{-2.5mm}
@@
 
-As mentioned earlier, the correction factors accumulate over time.
-If $\theta$ and $\phi$ coincide, one can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$, as seen in Equation \ref{eq:weighting-scheme}, while still targeting the posterior and likelihood at $\theta$.
+The correction factors, $w^{P,\theta}_{n,j}$ and $w^{F,\theta}_{n,j}$, accumulate as $n$ increases, leading to numerical instability.
+One can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$, as seen in Equation \ref{eq:weighting-scheme}, while still targeting the posterior and likelihood if $\theta=\phi$.
 This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate.
-When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, recovering the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}.
+When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, and Theorem~\ref{thm:mop-functional-forms} shows that it recovers the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}.
 When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{naesseth18}.
@@
 
-$\alpha$ therefore controls an exponentially-weighted moving average, determining how much "memory" the resulting gradient estimate has over ancestral trajectories.
+$\alpha$ therefore controls an exponentially-weighted moving average, determining how much memory the resulting gradient estimate has over ancestral trajectories.
 This is illustrated in Figure \ref{fig:biasvar}, in the context of the Dhaka cholera model of \cite{king08}.
@@
 It relies on a useful decomposition of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$ that yields a telescoping product in the MOP-$1$ case.
-Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting that when $\theta=\phi$, that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result.
+Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting $\theta=\phi$ implies that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result.
 
-This further illustrates how $\alpha$ dictates the "memory" of the gradient estimate.
-As \cite{scibior21} remark, the MOP-$0$ estimator of \cite{naesseth18} only considers single-step quantities, and is "memoryless" beyond a single step.
-This is in contrast to the case when $\alpha=1,$ as that estimate of \cite{poyiadjis11} only considers the surviving particles at time $N$, fully tracking dependencies over time.
+This further illustrates how $\alpha$ dictates the memory of the gradient estimate.
+As \cite{scibior21} remark, the MOP-$0$ estimator of \cite{naesseth18} only considers single-step quantities, and is ``memoryless'' beyond a single step.
+This is in contrast to the case when $\alpha=1,$ as that estimate, studied by \cite{poyiadjis11}, only considers the surviving particles at time $N$ and so fully tracks dependencies over time.
 
@@
 
-We bypass the issue of differentiating through a Monte Carlo algorithm with discontinuous resampling by turning it into a problem of differentiating through a simulator and a series of measurement density ratios.
-Although we defer the theoretical analysis of the resulting algorithm, MOP-$\alpha$, to Section \ref{sec:thms}, we highlight a few key results below.
-MOP-$\alpha$ targets the posterior and likelihood (Theorem \ref{thm:mop-targeting}), yields the estimators of \cite{poyiadjis11, scibior21, naesseth18} as special cases when differentiated (Theorem \ref{thm:mop-functional-forms}) where MOP-$1$ in particular is consistent for the score (Theorem \ref{thm:mop-grad-consistency}), has rates for its bias and variance under different choices of $\alpha$ (Theorem \ref{thm:mop-biasvar}), and enjoys a linear rate of convergence for gradient descent with the resulting gradient estimate (Theorem \ref{thm:mop-convergence}).
+The construction of MOP-$\alpha$ bypasses the issue of differentiating through a Monte Carlo algorithm with discontinuous resampling by turning it into a problem of differentiating through a simulator and a series of measurement density ratios.
+We defer the theoretical analysis of MOP-$\alpha$ to Section \ref{sec:thms}, but first we highlight a few key results.
+MOP-$\alpha$ estimates the likelihood and conditional distributions of latent variables, as one expects of a particle filter (Theorem \ref{thm:mop-targeting}).
+When differentiated, we obtain the estimators of \cite{poyiadjis11, scibior21, naesseth18} as special cases (Theorem \ref{thm:mop-functional-forms}).
+In particular, MOP-$1$ is consistent for the score (Theorem \ref{thm:mop-grad-consistency}), has rates for its bias and variance under different choices of $\alpha$ (Theorem \ref{thm:mop-biasvar}), and enjoys a linear rate of convergence for gradient descent with the resulting gradient estimate (Theorem \ref{thm:mop-convergence}).
 
-\section{Practical Maximum-Likelihood Estimation}
+\section{Practical Maximum Likelihood Estimation}
 
@@
 
-This still leaves us with the question of designing an effective procedure for likelihood maximization with MOP-$\alpha$.
-We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that runs a few iterations of IF2 to warm-start an iterative first or second-order method that uses the MOP-$\alpha$ gradient estimate.
-This leverages IF2's quick empirical convergence to a neighborhood of the MLE, overcoming the tendency of gradient methods to get stuck in saddle points and local minima.
-Conversely, switching to gradient ascent with MOP-$\alpha$ score estimates lets one bypass the difficulty that IF2 has with optimizing the last few units of log-likelihood.
-Combining these two methods in this way lets us enjoy the best of both worlds.
-\kevin{source for this}
+We now consider the question of designing an effective procedure for likelihood maximization with MOP-$\alpha$.
+We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that runs some preliminary iterations of IF2 to warm-start an iterative first or second-order method that uses the MOP-$\alpha$ gradient estimate.
+This leverages IF2's empirically stable and rapid convergence to a neighborhood of the MLE, overcoming the tendency of gradient methods to get stuck in saddle points and local minima.
+Conversely, switching to gradient ascent with MOP-$\alpha$ score estimates lets one bypass the difficulty experienced with IF2 when optimizing the last few units of log-likelihood.
+Combining these two methods lets us enjoy the best properties of both algorithms, which we show leads to strong empirical performance.
 
@@
   \label{thm:mop-convergence}
+  \ed{CURRENTLY EMPTY}
 \end{thm}
 
-%The warm start allows gradient methods to shine, as regularity conditions ensure that the likelihood surface is well-behaved enough in this neighborhood that the aforementioned issues with saddle points and local minima are alleviated.
+\ed{IN IFAD, $t$ IS USED AS AN INDEX FOR ITERATIONS. PREVIOUSLY, WE HAVE USED $m$ FOR THIS SINCE $n$ AND $t$ ARE USED ELSEWHERE.}
+
+\ed{IFAD HERE IS A 2ND ORDER ALGORITHM. MAYBE FOCUS ON THE PRIMARY EXAMPLE, WHICH IS FIRST ORDER. ALSO, THE STOPPING RULE SHOULD DESCRIBE WHAT WAS DONE IN THE EXEMPLES}
 
@@
 	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 cooling schedule $\eta_t$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $t=0.$\newline
-        Run IF2 until initial "convergence" under cooling schedule $\eta_t$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_t := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
+        \textbf{Preliminary iterated filtering:} Run IF2 under cooling schedule $\eta_t$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_t := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
 		\textbf{While} procedure not converged: \newline
@@
 
-\section{Application to the Cholera Model Of \cite{king08}}
+\section{Application to a Cholera Transmission Model}
 
-We use the Dhaka cholera model of \cite{king08} to benchmark (also used or modified in \cite{ionides15, wood16, wycoff24} for the same purpose) the performance of our method.
-This is a stochastic SIR compartmental model where the population at time $t$, $H_t$, is divided into the susceptible compartment $S_t$, infected compartment $I_t$, and three ($k=3$) recovered compartments $R^1_t, ..., R^k_t$ denoting varying degrees of cholera immunity.
-We write $M_t$ for the cholera deaths in each month.
-As in \cite{king08, ionides15}, the \textbf{transition dynamics} follow this series of SDEs:
+The cholera transmission model that \cite{king08} developed for Dhaka, Bangladesh, has been used to benchmark the performance of various POMP inference methods \cite{ionides15, wood16, wycoff24}, and we employ it here for the same purpose.
+This model categorizes individuals in a population as susceptible, $S(t)$, infected, $I(t)$, and recovered, $R(t)$ and so is called an SIR compartmental model.
+In this case, the compartment $R(t)$ is further subdivided into 
+%. where the population at time $t$, $H_t$, is divided into the susceptible compartment $S_t$, infected compartment $I_t$, and
+three recovered compartments $R^1(t)$, $R^2(t)$, $R^3(t)$ denoting varying degrees of cholera immunity.
+We write $H(t)$ for the total population, and $M_t$ \ed{$M_n$ ??} for the cholera deaths in each month.
+As in \cite{king08, ionides15}, the transition dynamics follow a series of stochastic differential equations:
+\ed{CHANGE TO $S(t)$ etc FOR CONSISTENCY.}
 \vspace*{-1mm}
@@
 \end{align*}
-
 with Brownian motion $B_t$, cholera death rate $m$, recovery rate $\gamma$, mean immunity duration $1/\epsilon$, standard deviation of the force of infection $\sigma$, and population death rate $\delta=0.02$.
-The \textbf{force of infection}, $\lambda_t$, is modeled by splines $(s_j)_{j=1}^6$
+The force of infection, $\lambda_t$, is modeled by splines $(s_j)_{j=1}^6$
 \vspace*{-2mm}
@@
 where the coefficients $(\beta_j)_{j=1}^6$ model seasonality in the force of infection, $\beta_{\text{trend}}$ models the trend in the force of infection, and the $\omega_j$ represent seasonality of a non-human environmental reservoir of disease.
-The \textbf{measurement model} for observed monthly cholera deaths is given by
+The measurement model for observed monthly cholera deaths is given by 
     $Y_n \sim \gN(M_n, \tau^2M_n^2)$,
-where $M_n$ is the true number of cholera deaths in that month.
+where $M_n=\gamma\int_{t_{n-1}}^{t_n}I(s)\, ds$ is the modeled number of cholera deaths in that month.
 
@@
 
-We benchmarked IFAD against IF2 on a challenging global search problem.
-We re-implemented IF2 in order to do so, but chose to benchmark our results with the results of \cite{ionides15} (labeled "IF2 2015").
-We note that our re-implementation ourperforms that of \cite{ionides15}, likely due to a better choice of parameters.
+We tested IFAD against IF2 on a global search problem for the Dhaka cholera model.
+We re-implemented IF2 in order to do so, but we compare our results with the results of \cite{ionides15} (labeled "IF2 2015").
+Our re-implementation ourperforms that of \cite{ionides15}, likely due to a better choice of algorithmic parameters.
 For each method, we performed 100 searches, initialized with 100 initial starting parameter vectors drawn uniformly from the same wide bounding box used in \cite{ionides15}.
@@
 \kevin{Dynamically save to a tex file and load from there}
+
+\ed{PERHAPS THE 2ND D.P. OF SIGNIFICANCE CAN BE AVOIDED? ALSO, IF WE'RE GOING TO ADVISE MOP-ALPHA FOR $0<\alpha<1$, THAT COULD BE SHOWN HERE.?}
 
@@
 \paragraph{IFAD Successfully Finds the MLE:} Previously, the MLE at a log-likelihood of $-3748.5$ was found by \cite{king16} with much effort, using many global and local IF2 searches, with the assistance of likelihood profiling.
-Meanwhile, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.63$, while \cite{king08} only achieve a best possible estimate of $-3793.4$.
+Meanwhile, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.63$, while the best log-likelihood found by \cite{king08} was only $-3793.4$.
 Despite being initialized for a global search, IFAD manages to find the MLE over the 100 searches, as seen in Table \ref{table:mle}.
-This suggests that the local search and refinement that was previously necessary for finding the MLE with IF2 is \textbf{baked into} the two-stage IFAD algorithm.
+This suggests that the sequence of local searches and refinement that was previously necessary for finding the MLE with IF2 is not necessary with the IFAD algorithm.
 
@@
 \label{assump:bounded-process}
-    \item \textbf{Bounded Measurement Model.} There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G}<\infty$ and $G'(\theta)$ so $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty \leq G'(\theta)< \infty$.
-\label{assump:bounded-measurement}
+    \item \textbf{Bounded Measurement Model.} There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G}<\infty$ and there exists $G'(\theta)$ with $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty \leq G'(\theta)< \infty$.
+\label{assump:bounded-measurement} \ed{I EDITED THE $G'(\theta)$ BIT.
+CHECK THAT IT IS RIGHT.
+IT STILL SEEMS A BIT UNCLEAR TO ME.
+ALSO, IT HAS SOME RELATIONSHIP TO A4 WHICH IS UNCLEAR TO ME.}
     \item \textbf{Bounded Gradient Estimates.} There are functions $G(\theta), H(\theta): \Theta \to [0,\infty)$ uniformly bounded by $G', H'$, so the MOP-$\alpha$ gradient and Hessian estimates at $\theta=\phi$ are almost surely bounded by $G(\theta)$ and $H(\theta)$ for all $\alpha$.
@@
 
-We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood.
+We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood when $\alpha=1$ or $\theta=\phi$.
 While the result is presented here as specific to MOP-$\alpha$, we actually prove a more general result in the supplementary material.
@@
 Despite showing that the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} use $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}, X_{1:N}}\left(y_{1:N}^* , x_{1:n,j}^{A, F,\theta}; \theta\right)$ in general.
-The bootstrap and general estimates have different functional forms, and are not the same a-priori.
+The bootstrap and general estimates have different functional forms, and are not the same a-priori \ed{WHAT DOES THIS MEAN?
+THAT IT IS NOT IMMEDIATLY CLEAR THEY ARE THE SAME?}.
 As such we directly show the consistency of the MOP-$1$ gradient estimate below.
@@
 Each iteration of IF2 amounts to one iteration of the particle filter, with $O(NJ)$ time complexity.
-The initial "convergence" happens fairly quickly in practice.
+The initial convergence to a region of high likelihood happens fairly quickly in practice.
 In the case of the Dhaka cholera model of \cite{king08}, when a geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations of IF2.
@@
 
-In line with the cheap gradient principle of \cite{kakade2019provably}, getting a gradient estimate from MOP-$\alpha$ takes no more than 6 times that of the runtime of the particle filter (in practice, the multiplier is 3.75).
-This therefore has $O(NJ)$ time complexity as well, obtaining a gradient estimate with runtime linear in the number of particles.
+In line with the cheap gradient principle of \cite{kakade2019provably}, getting a gradient estimate from MOP-$\alpha$ takes no more than 6 times that of the runtime of the particle filter.
+In practice, the multiplier is 3.75.
+AD for MOP-$\alpha$ therefore also has $O(NJ)$ time complexity, obtaining a gradient estimate with runtime linear in the number of particles.
 
@@
 
-Re-implementing the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, even with a simulator written in Python.
-This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter (with a simulator written in C++) in the \texttt{pomp} package of \cite{king16}.
+Implementing the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, with a simulator written in Python.
+This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter in the \texttt{pomp} package of \cite{king16}, for which the simulator is written in C and called from R.
 
@@
 
-We note that it is possible to work without a differentiable simulator, but one would then require access to the transition densities.
-
-Our gradient estimate can be used for variational inference to approximate the posterior distribution over parameters.
+We note that it is possible to work without a differentiable simulator.
+This arises if the latent process takes values in a discrete set.
+In this case, one could extend MOP-$\alpha$ to have an off-policy process model as well as off-policy resampling.
+This requires access to ratios of transition densities, so it does not have the plug-and-play property.
 

### 2024-06-06 paper/ms_new.tex
Updates from Overleaf

--- 2024-06-03 paper/ms.tex
@@
 However, particle filter inference methods based on the likelihood are statistically efficient, up to Monte Carlo error.
-\ed{MAYBE ADD SOME REFERENCES HERE TO DOCUMENT SUCCESSES OF IF2, SINCE WE ARE COMPARING WITH THAT AND BUILDING ON IT}
+\textcolor{red}{[MAYBE ADD SOME REFERENCES HERE TO DOCUMENT SUCCESSES OF IF2, SINCE WE ARE COMPARING WITH THAT AND BUILDING ON IT]}
 This Monte Carlo error can be non-negligible.
@@
 
-\acknow{We thank Nicholas Chopin for helpful comments on the proof of a strong law of large numbers for a suitably reweighted particle filter. }
+\acknow{We thank Nicholas Chopin for helpful comments on the proof of a strong law of large numbers for a suitably reweighted particle filter.
+Please include your acknowledgments here, set in a single paragraph.
+Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.}
 

### 2024-06-06 paper/ms.tex
Updates from Overleaf

@@
 We start by presenting a new construction of the gradient estimator of \cite{scibior21} and \cite{poyiadjis11}.
-Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a Measurement Off-Policy (MOP) particle filter.
+Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a Measurement Off-parameter (MOP) particle filter.
 We procced to generalize the MOP particle filter by adding a discounting factor, $\alpha \in [0,1]$ , which interpolates between the  biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
@@
 
-Consider an unobserved Markov process $\{X(t),t  \geq t_0\}$, with disctete-time observations $Y_1,...,Y_N$ realized at values $y_1^*,...,y_N^*$ at times $t_1,..., t_N$.
+Consider an unobserved Markov process $\{X(t),t  \geq t_0\}$, with discrete-time observations $Y_1,...,Y_N$ realized at values $y_1^*,...,y_N^*$ at times $t_1,..., t_N$.
 The process is parameterized by an unknown parameter $\theta \in \Theta \subseteq \R^p$, where the state $X(t)$ take values in the state space $\gX \subseteq \R^d$, the observations $Y_n$ take values in $\gY,$ and we write $X_n := X(t_n)$.
@@
 
-We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$, and we write $\texttt{process}\left(x_n\mid x_{n-1}; \theta\right)$ for its corresponding simulator \ed{WE DON'T USE THIS...
-MAYBE CHANGE HERE, OR USE IT}.
+We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$, and we write $\process_n\left(x_n\mid x_{n-1}; \theta\right)$ for its corresponding simulator.
 We set $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ to be the measurement density, and $y_n^*$ for the actual values of the observations that were observed.
 We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
-Superscripts $x_{n,j}^A$ denote the ancestral trajectory of particle $j$ at time $n$, and $a(\cdot)$ is the ancestor function \ed{NOT DEFINED YET.
-THIS WILL PUZZLE READERS UNLESS THEY ARE ALREADY HIGHLY FAMILIAR WITH THE TOPIC}.
-For simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0$.
-\ed{ALGORITHM 1 HAS $f_{X_0}$.
-IN GENERAL, THIS DETAIL IS BEST INCLUDED.
-INITIAL VALUES CAN BE AN IMPORTANT PART OF A DYNAMIC SYSTEM.}
+Superscripts $x_{n,j}^A$ denote the ancestral trajectory of particle $j$ at time $n$, and $a(\cdot)$ is the ancestor function that maps a particle to its parent, $j \mapsto a(j)$.
 
-The above conditional densities are defined on a probability space $(\Omega, \Sigma, \prob).$ 
+The above densities are defined on a probability space $(\Omega, \Sigma, \prob).$ 
 For the algorithmic interpretation of our theory, we identify the seed with an element of the sample space $\omega \in \Omega$.
-\ed{THE SEED CORRESPONDS TO ALL THE MONTE CARLO QUANTITIES GENERATED, INCLUDING THINGS THAT HAVE NO DIRECT ROLE IN THE MODEL, SO MAYBE THIS EXPLANATION NEEDS ADDITIONAL CARE.}
-The random seed determines the entire sequence of pseudo-random numbers as generated by a computer, just as the outcome $\omega\in\Omega$ generates the entire sequence of random variables.
+The random seed determines the sequence of pseudo-random numbers as generated by a computer, just as the outcome $\omega\in\Omega$ generates the sequence of random variables.
 
@@
 
-\section{Off-Policy Particle Filters}
+\section{Off-Parameter Particle Filters}
 
 Directly differentiating through a particle filter, for a fixed value of $\omega$, leads to a biased gradient estimator due to the discreteness of the resampling operation REFS.
-We avoid this issue by constructing a particle filtering algorithm called \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) for which the resampling is invariant to $\omega$.
+We avoid this issue by constructing a particle filtering algorithm called \textbf{Measurement Off-Parameter-$\alpha$} (MOP-$\alpha$) for which the resampling is invariant to $\omega$.
 Pseudocode for MOP-$\alpha$ is presented in Algorithm~\ref{alg:mop}.
@@
     \label{alg:mop}
-	     \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
+	     \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process_n(x_{n+1}|x_n; \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
 \newline
@@
         \textbf{Second pass:}
-            Fix $\omega$, and filter at $\theta\neq phi$. \newline
+            Fix $\omega$, and filter at $\theta\neq \phi$. \newline
 
@@
             \hspace*{4mm} Simulate process model,
-            ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$. \newline
+            ${X}_{n,j}^{P,\theta}\sim \process_n\big(\cdot|{X}_{n-1,j}^{F, \theta};{\theta}\big)$. \newline
             \hspace*{4mm} Measurement density,
@@
             \hspace*{4mm} Select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) \propto g^{\phi}_{n,m}$. \newline
-            \hspace*{4mm} Resample particles ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$. \ed{$\phi$ SHOULD BE $\theta$} \newline
+            \hspace*{4mm} Resample particles ${X}_{n,j}^{F,\theta}={X}_{n,k_{j}}^{P,\theta}$. \newline
             \hspace*{4mm} Resample corrected weights
@@
 \subsection{Algorithm Outline} 
-We obtain two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but the resampling indices constructed from the first pass, with parameter $\phi$.
-We call this ``off-policy'' measurement since a parameter value corresponds to a policy in the context of control theory.
-\ed{I REWROTE THIS TO BE MORE EXPLICIT, BUT MAYBE THIS NEEDS MORE EDITING. IT MAKES ME THINK THAT ``RESAMPLING OFF-POLICY'' WOULD BE A BETTER NAME!}
+We obtain two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but the resampling indices constructed from the first pass, with a ``baseline parameter'' $\phi$.
+As the process model follows $\theta$ but the measurement model follows the baseline parameter $\phi$, we therefore call this a ``measurement off-parameter'' particle filter.
 The resampling indices, which we write as $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$, are a function of $\phi$ for any value of $\theta$.
@@
 
-\section{Practical Maximum Likelihood Estimation}
+\section{Practical Maximum-Likelihood Estimation}
 
@@
 
-We now consider the question of designing an effective procedure for likelihood maximization with MOP-$\alpha$.
-We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that runs some preliminary iterations of IF2 to warm-start an iterative first or second-order method that uses the MOP-$\alpha$ gradient estimate.
-This leverages IF2's empirically stable and rapid convergence to a neighborhood of the MLE, overcoming the tendency of gradient methods to get stuck in saddle points and local minima.
-Conversely, switching to gradient ascent with MOP-$\alpha$ score estimates lets one bypass the difficulty experienced with IF2 when optimizing the last few units of log-likelihood.
-Combining these two methods lets us enjoy the best properties of both algorithms, which we show leads to strong empirical performance.
+This still leaves us with the question of designing an effective procedure for likelihood maximization with MOP-$\alpha$.
+We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that runs a few iterations of IF2 to warm-start an iterative (first or second-order) method that uses the MOP-$\alpha$ gradient estimate.
+This leverages IF2's quick empirical convergence to a neighborhood of the MLE, overcoming the tendency of gradient methods to get stuck in saddle points and local minima.
+Conversely, switching to gradient ascent with MOP-$\alpha$ score estimates lets one bypass the difficulty that IF2 has with optimizing the last few units of log-likelihood.
+Combining these two methods in this way lets us enjoy the best of both worlds.
+
+The quick empirical convergence of IF2 (and so the first stage of IFAD) to a neighborhood of the MLE happens fairly quickly in practice.
+In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations.
+Finding the MLE itself with IF2, however, takes much longer, as one has to use a less aggressive cooling rate to do so.
+For example, \cite{ionides15} use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{lee20}.
+By simply requiring that the first stage of IFAD get to a neighborhood of the MLE and not the MLE itself, we are able to substantially reduce the number of IF2 iterations required.
 
 \subsection{Linear Convergence Rates}
-\begin{thm}
+
+On the other hand, the second stage of IFAD enjoys a linear convergence rate under the usual strongly-convex and smoothness assumptions in convex optimization, as we show below in Theorem \ref{thm:mop-convergence}.
+
+\begin{thm}[Linear Convergence of IFAD]
+
+Consider the second stage of IFAD (Algorithm \ref{alg:ifad}) where one stops if $||\nabla_\theta \hat\ell^\alpha(\theta_n)|| \leq (1+\sigma) \epsilon$, where $\sigma \geq \frac{4 \Gamma}{(1-\beta)}$.
+Assume $-\ell$ is strongly convex and smooth, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$.
+Choose the learning rate $\eta$ and error tolerance $\epsilon$ such that $\eta \leq \frac{c(1-\beta)}{2\Gamma}$ for some $\beta \in (0,1)$.
+Then, for sufficiently large $\alpha$ and $J$ to ensure the score estimate is a $1/2$-approximation, the second stage of IFAD converges linearly to the MLE with probability at least $1-\delta$:
+\vspace*{-2mm}
+$$
+\ell(\theta^*) - \ell(\theta_{n+1}) \leq \left(1-\eta\beta\frac{8\gamma}{9c}\right)(\ell(\theta^*)-\ell(\theta_n)).
+$$
     \label{thm:mop-convergence}
-  \ed{CURRENTLY EMPTY}
 \end{thm}
+\vspace*{-5mm}
 
-\ed{IN IFAD, $t$ IS USED AS AN INDEX FOR ITERATIONS. PREVIOUSLY, WE HAVE USED $m$ FOR THIS SINCE $n$ AND $t$ ARE USED ELSEWHERE.}
+We defer the proof, which is similar to that of Theorem 6 in \cite{mahoney16}, to the supplementary information.
+For simplicity, we only prove this for MOP-$1$ and note that the general case follows as the required $1/2$-approximation can still be obtained for sufficiently large $\alpha$.
 
-\ed{IFAD HERE IS A 2ND ORDER ALGORITHM. MAYBE FOCUS ON THE PRIMARY EXAMPLE, WHICH IS FIRST ORDER. ALSO, THE STOPPING RULE SHOULD DESCRIBE WHAT WAS DONE IN THE EXEMPLES}
+We therefore see that the second stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the first stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
+This happens fairly often in practice, for example, when sufficient regularity conditions for local asymptotic normality of the MLE hold.
+We conjecture that this applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE, and will explore this in future work.
+
+%IF2 posesses some SGD-like qualities, including what looks like linear convergence to a ball around the MLE with width determined by the random walk standard deviation.
+We will explore this conjecture in future work.
+
+%While this theoretical guarantee currently only holds for IFAD-1, as it forms a particle approximation of the score as in \cite{poyiadjis11}, one can also extend this for a similar guarantee if the bias for the gradient estimate given by MOP-$\alpha$ for $\alpha < 1$ is small enough.
+For general $\alpha$, including MOP-0, the analysis then becomes more complicated, as one needs to handle the case of biased gradient descent.
+However, if future measurements $y_{n+1:N}^*$ do not provide a large amount of information on the identification of the current state $x_n$ given past and current measurements $y_{0:n}^*$, as mentioned in \cite{corenflos21}, the bias may be small enough for it to not present too much of a problem.
+
+%The warm start allows gradient methods to shine, as regularity conditions ensure that the likelihood surface is well-behaved enough in this neighborhood that the aforementioned issues with saddle points and local minima are alleviated.
 
@@
     \label{alg:ifad}
-	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 cooling schedule $\eta_t$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $t=0.$\newline
-        \textbf{Preliminary iterated filtering:} Run IF2 under cooling schedule $\eta_t$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_t := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
+	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 cooling schedule $\eta_m$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $m=0.$\newline
+        Run IF2 until initial "convergence" under cooling schedule $\eta_m$, or for a fixed number or iterations, to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
 		\textbf{While} procedure not converged: \newline
-		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_t).$ \newline
-		\hspace*{4mm} Obtain $g(\theta_t) = \nabla_{\theta_t} (-\hat\loglik(\theta_t))$, $H(\theta_t)$ s.t. $\lambda_{\min}(H) \geq c$. \newline
-		\hspace*{4mm} Update $\theta_{t+1} := \theta_t - \eta (H(\theta_t))^{-1} g(\theta_t)$, $t:=t+1.$ \newline
-		\textbf{Return} $\hat{\theta} := \theta_t.$
+		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_m).$ \newline
+		\hspace*{4mm} Obtain $g(\theta_m) = \nabla_{\theta_m} (-\hat\loglik(\theta_m))$, $H(\theta_m)$ s.t. $\lambda_{\min}(H) \geq c$. \newline
+		\hspace*{4mm} Update $\theta_{m+1} := \theta_m - \eta (H(\theta_m))^{-1} g(\theta_m)$, $m:=m+1.$ \newline
+		\textbf{Return} $\hat{\theta} := \theta_m.$
 \end{algorithm}
@@
 three recovered compartments $R^1(t)$, $R^2(t)$, $R^3(t)$ denoting varying degrees of cholera immunity.
-We write $H(t)$ for the total population, and $M_t$ \ed{$M_n$ ??} for the cholera deaths in each month.
+We write $P(t)$ for the total population, and $M_n$ for the cholera deaths in each month.
 As in \cite{king08, ionides15}, the transition dynamics follow a series of stochastic differential equations:
-\ed{CHANGE TO $S(t)$ etc FOR CONSISTENCY.}
 \vspace*{-1mm}
 \begin{align*}
-    d S_t&=(k \epsilon R^k_t+\delta(S_t-H_t)-\lambda_t S_t) d t+d H_t-({\sigma S_t I_t}/{H_t}) d B_t, \\
-    d I_t&=\left(\lambda_t S_t-(m+\delta+\gamma) I_t\right) d t+({\sigma S_t I_t}/{H_t}) d B_t, \\
-    d R^1_t&=(\gamma I-(k \epsilon+\delta) R^1) d t,..., 
-    d R^k_t=(k \epsilon R^{k-1}_t\hspace{-3mm}-(k \epsilon+\delta) R^k_t) d t,
+    d S&=(k \epsilon R^k+\delta(S-H)-\lambda(t) S) d t+d H-({\sigma S I}/{H}) d B_t, \\
+    d I&=\left(\lambda(t) S-(m+\delta+\gamma) I\right) d t+({\sigma S I}/{H}) d B_t, \\
+    d R^1&=(\gamma I-(k \epsilon+\delta) R^1) d t,..., 
+    d R^k=(k \epsilon R^{k-1}\hspace{-3mm}-(k \epsilon+\delta) R^k) d t,
     \vspace*{-2mm}
@@
 \begin{equation*}
-    \lambda_t=\exp\hspace{-1mm}\left(\hspace{-.5mm}\beta_{\text{trend}}(t-t_0)+\hspace{-1mm}\sum_{j=1}^{6} \beta_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm}\frac{I_t}{P_t} + \exp\hspace{-1mm} \left(\sum_{j=1}^{6} \omega_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm},
+    \lambda_t=\exp\hspace{-1mm}\left(\hspace{-.5mm}\beta_{\text{trend}}(t-t_0)+\hspace{-1mm}\sum_{j=1}^{6} \beta_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm}\frac{I}{P} + \exp\hspace{-1mm} \left(\sum_{j=1}^{6} \omega_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm},
     \vspace*{-2mm}
@@
 
-\kevin{Increase size of font in legend, notch rest of font up by one point. Make figure aspect ratio to be taller than wider.}
 \begin{figure}[htbp!]
@@
 While the result is presented here as specific to MOP-$\alpha$, we actually prove a more general result in the supplementary material.
-That is, we show a strong law of large numbers for triangular arrays of particles with ``off-policy'' resampling, where we resample the particles according to an arbitrary resampling rule not necessarily in proportion to the targeted distribution of interest, to weights that encode the cumulative discrepancy between the resampling and the target distribution instead of equal weights.
+That is, we show a strong law of large numbers for triangular arrays of particles with ``off-parameter'' resampling, where we resample the particles according to an arbitrary resampling rule not necessarily in proportion to the targeted distribution of interest, to weights that encode the cumulative discrepancy between the resampling and the target distribution instead of equal weights.
 
@@
 
-\section{Computational Considerations}
+\section{Computational Efficiency}
 
-\subsection{Runtime}
+MOP-$\alpha$ and IFAD are fast algorithms, both in theory and practice.
+In line with the cheap gradient principle of \cite{kakade2019provably}, getting a gradient estimate from MOP-$\alpha$ takes no more than 6 times that of the runtime of the particle filter (in practice, the multiplier is 3.75).
+It therefore shares the same $O(NJ)$ time complexity as the particle filter, unlike the $O(NJ^2)$ complexity of \cite{corenflos21} and Algorithm 2 in \cite{poyiadjis11, scibior21}.
 
-Each iteration of IF2 amounts to one iteration of the particle filter, with $O(NJ)$ time complexity.
-The initial convergence to a region of high likelihood happens fairly quickly in practice.
-In the case of the Dhaka cholera model of \cite{king08}, when a geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations of IF2.
-In comparison, often 100 or 200 iterations of IF2 are used for an initial global search.
-\cite{ionides15}, for example, use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{lee20}.
+As IF2 amounts to one iteration of the particle filter, with $O(NJ)$ time complexity.
 
-In line with the cheap gradient principle of \cite{kakade2019provably}, getting a gradient estimate from MOP-$\alpha$ takes no more than 6 times that of the runtime of the particle filter.
-In practice, the multiplier is 3.75.
-AD for MOP-$\alpha$ therefore also has $O(NJ)$ time complexity, obtaining a gradient estimate with runtime linear in the number of particles.
+In practice, re-implementing the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, even with a simulator written in Python.
+This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter (with a simulator written in C++) in the \texttt{pomp} package of \cite{king16}.
 
-\subsection{Python Library}
-
-Implementing the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, with a simulator written in Python.
-This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter in the \texttt{pomp} package of \cite{king16}, for which the simulator is written in C and called from R.
-
-\kevin{Write about new open source library that brings pomp to python with AD.
+%\kevin{Write about new open source library that brings pomp to python with AD.
 There are implementations in tfp under the jax substrate, but these are not plug and play -- they require you to supply the transition densities.
@@
 
-We note that it is possible to work without a differentiable simulator.
-This arises if the latent process takes values in a discrete set.
-In this case, one could extend MOP-$\alpha$ to have an off-policy process model as well as off-policy resampling.
-This requires access to ratios of transition densities, so it does not have the plug-and-play property.
+We note that it is possible to work without a differentiable simulator, but one would then require access to the transition densities.
+We are working on an analogue that uses (differentiable) transition densities instead of a differentiable simulator, that may be better suited to large discrete state spaces than either MOP-$\alpha$ or the Baum-Welch algorithm.
+
+Additionally, our gradient estimate can be used for variational inference to approximate the posterior distribution over latent states \cite{naesseth18} and parameters.
+
+Discounting the weights by some $\alpha \in [0,1]$ is not the only way to interpolate between the estimators of \cite{naesseth18} and \cite{poyiadjis11}.
+As the proof of Theorem \ref{thm:mop-biasvar} implies, we can also truncate the weights at a fixed lag, corresponding to the MOP-$(1,k)$ estimate mentioned.
+The analysis is similar, with slightly better rates but a less natural interpretation.
 

### 2024-06-07 paper/ms.tex
Thm 2 replaced

@@
 
-These improvements also extend to Bayesian inference, as we show that we can use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with a nonparametric empirical Bayes-style prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{andrieu10} from the $1.4 \times 10^6$ iterations in \cite{fasiolo16} to just $500$.
+These improvements also extend to Bayesian inference, as we show in the supplement that we can use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with a nonparametric empirical Bayes-style prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{andrieu10} from the $1.4 \times 10^6$ iterations in \cite{fasiolo16} to just $500$.
 
@@
 	     \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process_n(x_{n+1}|x_n; \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
-\newline
-            \textbf{First pass:} Set $\theta=\phi$ and fix $\omega$. This defined $X_{n,j}^{P,\phi}$, $X_{n,j}^{F,\phi}$, $g^{\phi}_{n,j}$. \newline
+
+        \textbf{First pass:} Set $\theta=\phi$ and fix $\omega$, yielding $X_{n,j}^{P,\phi}$, $X_{n,j}^{F,\phi}$, $g^{\phi}_{n,j}$.
+
         \textbf{Second pass:}
-            Fix $\omega$, and filter at $\theta\neq \phi$. \newline
+        Fix $\omega$, and filter at $\theta\neq \phi$:
 
@@
 That is, writing $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density, $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$,  we can estimate the conditional likelihood under $\theta$ as
-\vspace*{-2.5mm}
+\vspace*{-1.5mm}
 \begin{equation}
@@
      L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}} \text{ or } L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
-     \vspace*{-2.5mm}
+     \vspace*{-1.5mm}
 \end{equation}
 recursively weighting each particle by 
-\vspace*{-2.5mm}
+\vspace*{-1.5mm}
 \begin{equation}
@@
     w_{n,j}^{P,\theta} = (w_{n-1,j}^{F,\theta})^\alpha,  w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}, w^{F,\theta}_{0,j}= 1.
-    \vspace*{-2.5mm}
+    \vspace*{-1.5mm}
 \end{equation}
@@
 
-\kevin{Include plots of IFAD vs Poyiadjis and Naesseth only in the supplemental information}
+%\kevin{Include plots of IFAD vs Poyiadjis and Naesseth only in the supplemental information}
 
@@
     \item \textbf{Bounded Measurement Model.} There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G}<\infty$ and there exists $G'(\theta)$ with $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty \leq G'(\theta)< \infty$.
-\label{assump:bounded-measurement} \ed{I EDITED THE $G'(\theta)$ BIT.
-CHECK THAT IT IS RIGHT.
-IT STILL SEEMS A BIT UNCLEAR TO ME.
-ALSO, IT HAS SOME RELATIONSHIP TO A4 WHICH IS UNCLEAR TO ME.}
-    \item \textbf{Bounded Gradient Estimates.} There are functions $G(\theta), H(\theta): \Theta \to [0,\infty)$ uniformly bounded by $G', H'$, so the MOP-$\alpha$ gradient and Hessian estimates at $\theta=\phi$ are almost surely bounded by $G(\theta)$ and $H(\theta)$ for all $\alpha$.
+\label{assump:bounded-measurement}
+    \item \textbf{Bounded Gradient Estimates.} There are functions $G(\theta), H(\theta): \Theta \to [0,\infty)$ uniformly bounded by $G^*, H^*<\infty$, so the MOP-$\alpha$ gradient and Hessian estimates at $\theta=\phi$ are almost surely bounded by $G(\theta)$ and $H(\theta)$ for all $\alpha$.
 \label{assump:local-bounded-derivative}
@@
 
-\kevin{Should we just show the general result?
+%\kevin{Should we just show the general result?
 That is, for an arbitrary set of indices $k_{n,j}$ generated in proportion to some arbitrary weights $\pi_{n,j}$ that do not necessarily sum to $1$ as the estimate is self-normalized, reweighting the particles by $w_{n,a(j)}g_{n,a(j)}/\pi_{n,a(j)}$ targets the posterior.
@@
 
-\kevin{Write the supplement formally with Del Moral-lite language, going into full detail with the proofs and calculations.}
+%\kevin{Write the supplement formally with Del Moral-lite language, going into full detail with the proofs and calculations.}
 
@@
 
-Despite showing that the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} use $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}, X_{1:N}}\left(y_{1:N}^* , x_{1:n,j}^{A, F,\theta}; \theta\right)$ in general.
-The bootstrap and general estimates have different functional forms, and are not the same a-priori \ed{WHAT DOES THIS MEAN?
-THAT IT IS NOT IMMEDIATLY CLEAR THEY ARE THE SAME?}.
+Despite showing that the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} use $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}\left(x_{0:n,j}^{A, F,\theta}, y_{1:N}^* ; \theta\right)$ and not $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}| X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}; \theta\right)$.
+It is therefore not immediately apparent that these two converge to the same thing.
 As such we directly show the consistency of the MOP-$1$ gradient estimate below.
@@
     \label{thm:mop-biasvar}
-    When $\alpha\in(0,1)$, $\theta=\phi$, the MSE and variance of MOP-$\alpha$, where $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$ and $\epsilon$ depends on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karjalainen23} are:
+    When $\alpha\in(0,1)$, $\theta=\phi$, $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$, there exists some $\epsilon>0$ depending on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karjalainen23} such that the MSE and variance of MOP-$\alpha$ are:
     \vspace*{-1ex}
@@
 \end{thm}
-\kevin{Explain exactly what $\epsilon$ is here, and interpret the rates as agreeing with our intuition.
-Talk about the relationship between the theory and the empirical results in the bias-variance tradeoff figure above, as well as the phase transition between $Nk^2/J$ and $N^4/J$ variance with MOP-$\alpha$ and MOP-$1$, where $k$ is really a surrogate for $\alpha$ -- with more memory, you truncate less.
-}
 
-We defer the proof to the supplementary material in favor of a very brief outline here.
+We defer the proof to the supplementary material, but provide a brief outline here.
+The variance bound can be reduced to the approximation error between MOP-$\alpha$ and a variant called MOP-$(\alpha,k)$ where the weights are truncated at lag $k$ with discount factor $\alpha$ to ensure strong mixing, and the covariance of the latter which can be controlled with Davydov's inequality and a standard $L^p$ error bound for the particle filter.
+The MSE bound considers the error between the score and MOP-$(1,k)$, and the error between MOP-$(1,k)$ and MOP-$\alpha$.
+The latter is $\tilde{O}(NpG'(\theta)^2(k+\psi(\alpha)))$ by tedious algebra, and the former can be controlled with a result on the the forgetting of the particle filter from \cite{karjalainen23} and the very same $L^p$ error bound mentioned above.
 
-\kevin{Say that our framework suggests whole new classes of algorithms, as it includes Naesseth and Poyiadjis.
-Interpolating between them with discounting is only one way to interpolate, we can also e.g.
-truncate at a fixed lag.
-Remark that the analysis is similar for MOP-$(1,k)$ as a fixed-lag filter, with slightly better rates but a less natural interpretation.
-We leave empirical studies of its performance to future work.
-}
+We also show in the supplementary material that the variance of MOP-$0$ is $\tilde{O}(NpG'(\theta)^2/J)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$ (ignoring factors of $p, G'$, etc.).
+This and Theorem \ref{thm:mop-biasvar} show that MOP-$\alpha$ interpolates between MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$.
 

### 2024-06-07 paper/ms.tex
paper/ms has a switch to turn into arxiv format

@@
-\documentclass[9pt,twocolumn,twoside]{pnas-new}
+\documentclass[11pt]{article} 
+%\documentclass[9pt,twocolumn,twoside]{pnas-new}
 %\documentclass[9pt,twocolumn,pnasresearcharticle]{pnas-new}
@@
 % Use the lineno option to display guide line numbers if required.
+
+\newcommand\arxiv[2]{#1} % 1st argument for arxiv format
+% \newcommand\arxiv[2]{#2} % 2nd argument for PNAS format
+% also, swich to \documentclass[11pt]{article} for arxiv format
 
@@
 
+\arxiv{
+  \usepackage{fullpage}
+  %% some packages loaded by pnas-new.cls which we need for
+  %% the arxiv version
+  \usepackage{algorithm}
+  \usepackage[noend]{algpseudocode}
+  \usepackage{xcolor}
+  \usepackage{tikz}
+  \usepackage{enumitem}
+  \bibliographystyle{ieeetr}
+}{
 \templatetype{pnasresearcharticle} % Choose template
@@
 % {pnasinvited} %= Template for a PNAS invited submission
+}
 
@@
 
+%\arxiv{ \author{K. Tan, G. Hooker and E. L. Ionides\\
+%  \small{Department of Statistics, University of Pennsylvania}\\
+%  \small{Department of Statistics, University of Michigan}}}{}
+
+\arxiv{ \author{K. Tan$^{1}$, G. Hooker$^{1}$ and E. L. Ionides$^{2}$
+  \vspace{2mm}\\
+  \small{$^{1}$Department of Statistics, University of Pennsylvania}\\
+  \small{$^{2}$Department of Statistics, University of Michigan}}}{}
+
+\arxiv{}{
 % Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
@@
 %\author[a]{Author Three}
-
 \affil[a]{University of Pennsylvania}
@@
 %\affil[c]{Affiliation Three}
-
 % Please give the surname of the lead author for the running footer
 \leadauthor{Tan}
-
 % Please add a significance statement to explain the relevance of your work
@@
 Our new method outperforms previous approaches on a challenging problem in epidemiology.}
-
-% Please include corresponding author, author contribution and author declaration information
 \authorcontributions{K.T. and E.L.I. designed research; K.T. analyzed data; K.T., G.H. and E.L.I. performed research, K.T., G.H., and E.L.I. wrote the manuscript.}
@@
 \correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: kevtan@wharton.upenn.edu}
-
 % At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
 \keywords{Sequential Monte Carlo $|$ Automatic Differentiation $|$ Particle Filter $|$ Markov Process $|$ Maximum Likelihood}
+}
+
+\arxiv{\date{Draft compiled on \today}}{
+\dates{This manuscript was compiled on \today}
+\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
+}
+
+\maketitle
 
@@
 Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques because widely used particle filter algorithms yield an estimated likelihood function that is discontinuous as a function of the model parameters.
-We show how to embed two existing AD particle filter methods in a theoretical framework that provides an extension to a new class of algorithms within which we obtain a favorable bias/variance tradeoff.
-Further, we develop likelihood maximization algorithms suited to the Monte Carlo properties of the derivative estimate.
-Our algorithms require only a differentiable simulator for the latent dynamic system; by contrast, most previous approaches to AD likelihood maximization for particle filters require access to the system's transition probabilities, and this can be prohibitive.
-Promising numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from an iterated filtering algorithm can beat current state-of-the-art methods on a challenging scientific benchmark problem.
+We show how to embed two existing AD particle filter methods in a theoretical framework that provides an extension to a new class of algorithms.
+This new class permits a bias/variance tradeoff and hence a mean squared error substantially lower than the existing algorithms.
+We develop likelihood maximization algorithms suited to the Monte Carlo properties of the derivative estimate.
+Our algorithms require only a differentiable simulator for the latent dynamic system; by contrast, most previous approaches to AD likelihood maximization for particle filters require access to the system's transition probabilities.
+Numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from an iterated filtering algorithm show substantial improvement on current state-of-the-art methods for a challenging scientific benchmark problem.
 \end{abstract}
 
-\dates{This manuscript was compiled on \today}
-\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
-
-\maketitle
+\arxiv{}{
 \thispagestyle{firststyle}
 \ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
-
 \firstpage{4}
@@
 "\firstpage[4]{11}".
+}
 
-\dropcap{M}any scientific models involve highly nonlinear stochastic dynamic systems possessing significant random variation in both the process dynamics and the measurements.
+\arxiv{Many}{\dropcap{M}any}
+scientific models involve highly nonlinear stochastic dynamic systems possessing significant random variation in both the process dynamics and the measurements.
 Commonly, the latent system is modeled as a Markov process, giving rise to a partially observed Markov process (POMP) model, also known as a hidden Markov models or a state space model.
-POMP models arise in fields as diverse as epidemiology \cite{he10, stocks17}, ecology \cite{knape12}, and finance \cite{kim08, breto14}.
+POMP models arise in fields as diverse as automated control \cite{singh22}, epidemiology \cite{he10, stocks17}, ecology \cite{knape12} and finance \cite{kim08, breto14}.
 Despite their ubiquity, estimation and inference within this broad class of models remains a challenging problem.
@@
 The particle filter, also known as sequential Monte Carlo, serves as the foundation for various inference algorithms for POMP models.
-It provides an unbiased estimate of the likelihood function  \cite{delMoral04}, enabling Bayesian inference \cite{andrieu10,chopin13} and likelihood-based inference \citep{ionides06-pnas,ionides15}.
+It provides an unbiased estimate of the likelihood function  \cite{delMoral04}, enabling Bayesian inference \cite{andrieu10,chopin13} and likelihood-based inference \cite{ionides06-pnas,ionides15}.
 Likelihood-based inference is statistically efficient \cite{pawitan01} and is robust to a moderate amount of Monte Carlo error \cite{ionides17,ning21}.
@@
 We start by presenting a new construction of the gradient estimator of \cite{scibior21} and \cite{poyiadjis11}.
-Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a Measurement Off-parameter (MOP) particle filter.
+Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a Measurement Off-Parameter (MOP) particle filter.
 We procced to generalize the MOP particle filter by adding a discounting factor, $\alpha \in [0,1]$ , which interpolates between the  biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
@@
 
-The above densities are defined on a probability space $(\Omega, \Sigma, \prob).$ 
+The above densities are defined on a probability space $(\Omega, \Sigma, \prob)$ which is also assumed to enable construction of independent replicates and all other random variables defined in our algorithms.
 For the algorithmic interpretation of our theory, we identify the seed with an element of the sample space $\omega \in \Omega$.
@@
 
-Directly differentiating through a particle filter, for a fixed value of $\omega$, leads to a biased gradient estimator due to the discreteness of the resampling operation REFS.
-We avoid this issue by constructing a particle filtering algorithm called \textbf{Measurement Off-Parameter-$\alpha$} (MOP-$\alpha$) for which the resampling is invariant to $\omega$.
+Directly differentiating through a particle filter, for a fixed value of $\omega$, leads to a biased gradient estimator due to the discreteness of the resampling operation \cite{corenflos21}.
+We avoid this issue by constructing a particle filtering algorithm called {\it Measurement Off-Parameter} with discount factor $\alpha$ (MOP-$\alpha$) for which the resampling is invariant to $\omega$.
 Pseudocode for MOP-$\alpha$ is presented in Algorithm~\ref{alg:mop}.
@@
   \centering
-    \includegraphics[width=\textwidth/3]{../imgs/095/biasvar.png}
+    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{../imgs/095/biasvar.png}
     \caption{Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}.
@@
 \begin{align*}
-    d S&=(k \epsilon R^k+\delta(S-H)-\lambda(t) S) d t+d H-({\sigma S I}/{H}) d B_t, \\
-    d I&=\left(\lambda(t) S-(m+\delta+\gamma) I\right) d t+({\sigma S I}/{H}) d B_t, \\
-    d R^1&=(\gamma I-(k \epsilon+\delta) R^1) d t,..., 
-    d R^k=(k \epsilon R^{k-1}\hspace{-3mm}-(k \epsilon+\delta) R^k) d t,
+    dS&=(k \epsilon R^k+\delta(S-P)-\lambda(t) S)\, dt+d P-({\sigma S I}/{H})\, dB, \\
+    dI&=\left(\lambda(t) S-(m+\delta+\gamma) I\right)\, dt+({\sigma S I}/{H})\, dB, \\
+    dR^1&=(\gamma I-(k \epsilon+\delta) R^1)\, dt, \hspace{2mm} \dots \\
+    dR^k&=(k \epsilon R^{k-1}\hspace{-3mm}-(k \epsilon+\delta) R^k)\, dt,
     \vspace*{-2mm}
 \end{align*}
-with Brownian motion $B_t$, cholera death rate $m$, recovery rate $\gamma$, mean immunity duration $1/\epsilon$, standard deviation of the force of infection $\sigma$, and population death rate $\delta=0.02$.
+with Brownian motion $B(t)$, cholera death rate $m$, recovery rate $\gamma$, mean immunity duration $1/\epsilon$, standard deviation of the force of infection $\sigma$, and population death rate $\delta=0.02$.
 The force of infection, $\lambda_t$, is modeled by splines $(s_j)_{j=1}^6$
@@
     \centering
-    \includegraphics[width=\textwidth/2]{../imgs/095/tikzcholera.png}
+    \includegraphics[width=\arxiv{12cm}{\textwidth/2}]{../imgs/095/tikzcholera.png}
     \vspace*{-7mm}
-    \caption{Illustration of Dhaka cholera model from \cite{king08}.}
+    \caption{Illustration of Dhaka cholera model from \cite{king08}. \ed{CHANGE H TO P? ALSO, THERE ARE ADVANTAGES TO PUTTING THE TIKZ SOURCE CODE INTO THE MS, TO ALLOW FOR FINE-TUNING EDITS.}}
     \label{fig:tikz-cholera}
@@
 
-\kevin{Dynamically save to a tex file and load from there}
+% \kevin{Dynamically save to a tex file and load from there}
 
@@
 
-\paragraph{IFAD Successfully Finds the MLE:} Previously, the MLE at a log-likelihood of $-3748.5$ was found by \cite{king16} with much effort, using many global and local IF2 searches, with the assistance of likelihood profiling.
-Meanwhile, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.63$, while the best log-likelihood found by \cite{king08} was only $-3793.4$.
+\paragraph{IFAD Successfully Finds the MLE:} Previously, an MLE at a log-likelihood of $-3748.5$ was reported by \cite{king16}.
+This MLE was obtained with much computational effort, using many global and local IF2 searches, and with the assistance of likelihood profiling.
+Meanwhile, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.6$, while the best log-likelihood found by \cite{king08} was only $-3793.4$.
 Despite being initialized for a global search, IFAD manages to find the MLE over the 100 searches, as seen in Table \ref{table:mle}.
-This suggests that the sequence of local searches and refinement that was previously necessary for finding the MLE with IF2 is not necessary with the IFAD algorithm.
+On this problem, the sequence of local searches and refinement that was previously required for finding the MLE is not necessary with the IFAD algorithm.
 
@@
 Conversely, when we tried gradient descent with MOP alone, we encountered many, many failed searches.
-This is a difficult, nonconvex, and noisy problem, and the search gets stuck in local minima and saddle points, failing to approach the MLE.
+This is a difficult, nonconvex, and noisy problem, \ed{SAY HOW MANY PARAMETERS ARE ESTIMATED HERE} and the search gets stuck in local minima and saddle points, failing to approach the MLE.
 
-IFAD, in comparison, manages to (1) approach the MLE quickly due to the IF2 warm-start (as seen in Figure \ref{fig:scatter-optim}) and (2) refine the coarse solution found by the warm-start with MOP gradient steps to find the MLE (as seen in Figures \ref{fig:scatter-optim} and \ref{fig:boxplot-search}.
+IFAD, in comparison, approaches the MLE quickly due to the IF2 warm-start (as seen in Figure \ref{fig:scatter-optim}) and also succeeds at refining the coarse solution found by the warm-start with MOP gradient steps to find the MLE (as seen in Figures~\ref{fig:scatter-optim} and \ref{fig:boxplot-search}.
 IFAD therefore successfully combines the best qualities of IFAD and MOP, outperforming either of them alone.
+Monte~Carlo replication is appropriate for IFAD, as for any Monte Carlo algorithm used to solve a challenging numerical problem, but the left panel of Figure~\ref{fig:boxplot-search} shows that IFAD can find higher likelihood values, more quickly and more reliably than the previous state-of-the-art.
+\ed{I ADDED THIS PARAGRAPH TO TRY TO BE EXPLICIT ABOUT WHY WE ARE EXCITED BY THIS FIGURE.
+AT FACE VALUE, IT MAY NOT SEEM LIKE A BIG DEAL.
+IT WOULD BE BETTER TO EXPLAIN MORE CLEARLY WHY THIS IS EVIDENCE THAT THE METHOD COULD HAVE A TRANSFORMATIVE EFFECT ON DATA ANALYSIS VIA POMP MODELS.}
 
 \begin{figure}[htbp!]
-    \includegraphics[width=\textwidth/\real{4.2}]{../imgs/095/pairs.png}
-    \includegraphics[width=\textwidth/\real{4.2}]{../imgs/095/qq.png}
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/pairs.png}
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/qq.png}
     \caption{Scatterplots depicting the performance of IFAD against that of IF2.
@@
 
+\ed{Labeling A and B would be more conventional than Left and Right, and also easier to refer to in the text.}
+
 \begin{figure}[ht]
-    \includegraphics[width=\textwidth/\real{4.2}]{../imgs/095/boxplot.png}
-    \includegraphics[width=\textwidth/\real{4.2}]{../imgs/095/optim.png}
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/boxplot.png}
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/optim.png}
     \caption{\textbf{Left:} Boxplot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs, simulating the result of the common procedure of running a few searches and choosing the best one.
@@
 The dashed orange line depicts the median warm-start given by running 40 IF2 iterations.
-While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD -- IFAD has better tail control and successfully reaches the MLE.
+While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD.
+We see that IFAD has better tail control and successfully reaches the MLE.
     We use a dotted red line to display the MLE.}
@@
 \end{figure}
+
+\ed{IS THE IFAD(0.95) MEDIAN LINE IN FIG 4 (RIGHT) HIDING UNDER IFAD(1)? IF NOT, WEHRE IS IT?}
 
@@
 \label{assump:local-bounded-derivative}
-    \item \textbf{Differentiability of Density Ratios and Simulator.} The measurement density $f_{Y_n|X_n}(y_n^*|x; \theta)$ and simulator have more than two continuous derivatives in $\theta$.
+    \item \textbf{Differentiability of Density Ratios and Simulator.} The measurement density, \arxiv{\\}{}$f_{Y_n|X_n}(y_n^*|x_n; \theta)$, and simulator have more than two continuous derivatives in $\theta$.
 \label{assump:diff-meas-and-sim}
@@
     We provide a proof sketch here, deferring most of the details, and discussion of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$, to the supplementary information. 
-
     When $\theta=\phi$, regardless of the value of $\alpha$, the ratio ${g_{n,j}^\theta}/{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter.
-
     When $\alpha=1$ and $\theta\neq\phi,$ suppose inductively that $\{(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j})\}_{j=1}^J$ targets $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
@@
     resampling with probabilities proportional to $g^\phi_{n,j}$, also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-
     If the likelihood is estimated with the before-resampling conditional likelihoods $\hat\lik(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, the strong consistency is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
@@
 
-Despite showing that the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} use $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}\left(x_{0:n,j}^{A, F,\theta}, y_{1:N}^* ; \theta\right)$ and not $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}| X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}; \theta\right)$.
-It is therefore not immediately apparent that these two converge to the same thing.
-As such we directly show the consistency of the MOP-$1$ gradient estimate below.
+Despite showing that the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} use
+$$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}\left(x_{0:n,j}^{A, F,\theta}, y_{1:N}^* ; \theta\right)$$ and not
+$$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}| X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}; \theta\right).$$
+It is therefore not immediately apparent that these two converge to the same thing. As such we directly show the consistency of the MOP-$1$ gradient estimate below.
 
@@
 
-As IF2 amounts to one iteration of the particle filter, with $O(NJ)$ time complexity.
+%% As IF2 amounts to one iteration of the particle filter, with $O(NJ)$ time complexity. 
+%%In practice,
 
-In practice, re-implementing the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, even with a simulator written in Python.
+Our implementation of the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, even with a simulator written in Python.
 This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter (with a simulator written in C++) in the \texttt{pomp} package of \cite{king16}.
@@
 We are working on an analogue that uses (differentiable) transition densities instead of a differentiable simulator, that may be better suited to large discrete state spaces than either MOP-$\alpha$ or the Baum-Welch algorithm.
-
 Additionally, our gradient estimate can be used for variational inference to approximate the posterior distribution over latent states \cite{naesseth18} and parameters.
@@
 
+\arxiv{}{
 \showmatmethods{} % Display the Materials and Methods section
-
 \acknow{Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.}
-
 \showacknow{} % Display the acknowledgments section
+}
 

### 2024-06-08 paper/ms.tex
GH edits

@@
 Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a Measurement Off-Parameter (MOP) particle filter.
-We procced to generalize the MOP particle filter by adding a discounting factor, $\alpha \in [0,1]$ , which interpolates between the  biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
+We proceed to generalize the MOP particle filter by adding a discounting factor, $\alpha \in [0,1]$, which interpolates between the  biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
 The bias-variance tradeoff induced by $\alpha$ suggests, both in theory and in practice, the use of $\alpha$ values strictly between 0 and 1.
@@
 
-We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$, and we write $\process_n\left(x_n\mid x_{n-1}; \theta\right)$ for its corresponding simulator.
+We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ and we write $\process_n\left(x_n\mid x_{n-1}; \theta\right)$ for its corresponding simulator.
 We set $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ to be the measurement density, and $y_n^*$ for the actual values of the observations that were observed.
@@
 The above densities are defined on a probability space $(\Omega, \Sigma, \prob)$ which is also assumed to enable construction of independent replicates and all other random variables defined in our algorithms.
-For the algorithmic interpretation of our theory, we identify the seed with an element of the sample space $\omega \in \Omega$.
+For the algorithmic interpretation of our theory, we identify the random number seed with an element of the sample space $\omega \in \Omega$.
 The random seed determines the sequence of pseudo-random numbers as generated by a computer, just as the outcome $\omega\in\Omega$ generates the sequence of random variables.
@@
 Directly differentiating through a particle filter, for a fixed value of $\omega$, leads to a biased gradient estimator due to the discreteness of the resampling operation \cite{corenflos21}.
-We avoid this issue by constructing a particle filtering algorithm called {\it Measurement Off-Parameter} with discount factor $\alpha$ (MOP-$\alpha$) for which the resampling is invariant to $\omega$.
+We avoid this issue by constructing a particle filtering algorithm called {\it Measurement Off-Parameter} with discount factor $\alpha$ (MOP-$\alpha$) for which the resampling is invariant to $\omega$ but dependent on $\theta$.
 Pseudocode for MOP-$\alpha$ is presented in Algorithm~\ref{alg:mop}.
@@
             $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$.\newline
-            \hspace*{4mm} Select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) \propto g^{\phi}_{n,m}$. \newline
-            \hspace*{4mm} Resample particles ${X}_{n,j}^{F,\theta}={X}_{n,k_{j}}^{P,\theta}$. \newline
-            \hspace*{4mm} Resample corrected weights
+            \hspace*{4mm} Select resampling indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) \propto g^{\phi}_{n,m}$. \newline
+            \hspace*{4mm} Obtain resampled particles ${X}_{n,j}^{F,\theta}={X}_{n,k_{j}}^{P,\theta}$. \newline
+            \hspace*{4mm} Calculate resampled corrected weights
             $w_{n,j}^{F,\theta}= w^{P,\theta}_{n,k_j}  { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}$.\newline
@@
 \subsection{Algorithm Outline} 
-We obtain two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but the resampling indices constructed from the first pass, with a ``baseline parameter'' $\phi$.
+We obtain two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but with the resampling indices constructed from the first pass, with a ``baseline parameter'' $\phi$.
 As the process model follows $\theta$ but the measurement model follows the baseline parameter $\phi$, we therefore call this a ``measurement off-parameter'' particle filter.
@@
 We then reweight the conditional likelihoods by a correction factor accumulated over time to account for the resampling under $\phi$.
-That is, writing $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density, $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$,  we can estimate the conditional likelihood under $\theta$ as
+That is, writing $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density and $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$,  we can estimate the conditional likelihood under $\theta$ by
 \vspace*{-1.5mm}
@@
 \end{equation}
-recursively weighting each particle by 
+where the weights for each particle are updated by
 \vspace*{-1.5mm}
@@
 The correction factors, $w^{P,\theta}_{n,j}$ and $w^{F,\theta}_{n,j}$, accumulate as $n$ increases, leading to numerical instability.
-One can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$, as seen in Equation \ref{eq:weighting-scheme}, while still targeting the posterior and likelihood if $\theta=\phi$.
+One can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$ as in Equation \ref{eq:weighting-scheme} while still targeting the posterior and likelihood if $\theta=\phi$.
 This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate.
@@
     \label{thm:mop-functional-forms}
-    Write $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$ and we use the after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$.
-Then, when $\alpha=0$,
+    Writing $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$ and using the after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$, when $\alpha=0$,
     \vspace*{-2.5mm}
@@
 This further illustrates how $\alpha$ dictates the memory of the gradient estimate.
-As \cite{scibior21} remark, the MOP-$0$ estimator of \cite{naesseth18} only considers single-step quantities, and is ``memoryless'' beyond a single step.
+As \cite{scibior21} remarks, the MOP-$0$ estimator of \cite{naesseth18} only considers single-step quantities, and is ``memoryless'' beyond a single step.
 This is in contrast to the case when $\alpha=1,$ as that estimate, studied by \cite{poyiadjis11}, only considers the surviving particles at time $N$ and so fully tracks dependencies over time.
@@
 
-The quick empirical convergence of IF2 (and so the first stage of IFAD) to a neighborhood of the MLE happens fairly quickly in practice.
+The  convergence of IF2 (and so the first stage of IFAD) to a neighborhood of the MLE happens fairly quickly in practice.
 In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations.
@@
 
-\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations (as seen in Figure \ref{fig:scatter-optim}), performing IF2 alone ultimately fails to squeeze out the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter-optim}).
-Conversely, when we tried gradient descent with MOP alone, we encountered many, many failed searches.
+\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations, performing IF2 alone ultimately fails to achieve the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter-optim}).
+Conversely, when we tried gradient descent with MOP alone, we encountered many failed searches.
 This is a difficult, nonconvex, and noisy problem, \ed{SAY HOW MANY PARAMETERS ARE ESTIMATED HERE} and the search gets stuck in local minima and saddle points, failing to approach the MLE.
@@
 \textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches.
-It is clear that on average, IFAD has the edge, and manages to find the MLE while no IF2 search successfully gets within 7 log-likelihood units of it.
+It is clear that on average, IFAD performs best, and manages to find the MLE while no IF2 search successfully gets within 7 log-likelihood units of it.
 We use a dotted red line to display the MLE in both.
@@
 
-Here, we will show that MOP-$\alpha$ targets the posterior and likelihood, yields the estimators of \cite{poyiadjis11, scibior21, naesseth18} as special cases when differentiated, and characterize rates for its bias and variance under different choices of $\alpha$.
+Here, we will show that MOP-$\alpha$ targets the posterior and likelihood and yields the estimators of \cite{poyiadjis11, scibior21, naesseth18} as special cases when differentiated, and characterize rates for its bias and variance under different choices of $\alpha$.
 To do so, we require the following assumptions:
@@
 While the result is presented here as specific to MOP-$\alpha$, we actually prove a more general result in the supplementary material.
-That is, we show a strong law of large numbers for triangular arrays of particles with ``off-parameter'' resampling, where we resample the particles according to an arbitrary resampling rule not necessarily in proportion to the targeted distribution of interest, to weights that encode the cumulative discrepancy between the resampling and the target distribution instead of equal weights.
+That is, we show a strong law of large numbers for triangular arrays of particles with ``off-parameter'' resampling, where we resample the particles according to an arbitrary resampling rule not necessarily in proportion to the targeted distribution of interest and employ weights that encode the cumulative discrepancy between the resampling and the target distribution instead of equal weights.
 
@@
 \begin{proof}
-    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point we wish to evaluate the gradient at.
-The sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$, by Assumption \ref{assump:local-bounded-derivative}.
+    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point at which we wish to evaluate the gradient.
+The sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$ by Assumption \ref{assump:local-bounded-derivative}.
 Again by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J^1(\theta)(\omega)|_{\theta=\theta'}$ is also bounded by $H'$ for almost every $\omega\in \Omega$ and every $\theta'\in \Theta$.
@@
     \label{thm:mop-biasvar}
-    When $\alpha\in(0,1)$, $\theta=\phi$, $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$, there exists some $\epsilon>0$ depending on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karjalainen23} such that the MSE and variance of MOP-$\alpha$ are:
+    When $\alpha\in(0,1)$, $\theta=\phi$, define $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$, there exists some $\epsilon>0$ depending on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karjalainen23} such that the MSE and variance of MOP-$\alpha$ are:
     \vspace*{-1ex}

### 2024-06-10 paper/ms.tex
ms edits to address Giles' comments

@@
 This new class permits a bias/variance tradeoff and hence a mean squared error substantially lower than the existing algorithms.
-We develop likelihood maximization algorithms suited to the Monte Carlo properties of the derivative estimate.
+We develop likelihood maximization algorithms suited to the Monte Carlo properties of the AD gradient estimate.
 Our algorithms require only a differentiable simulator for the latent dynamic system; by contrast, most previous approaches to AD likelihood maximization for particle filters require access to the system's transition probabilities.
@@
 Despite their ubiquity, estimation and inference within this broad class of models remains a challenging problem.
+This article concerns the use automatic differentiation (AD) to construct improved algorithms for inference on complex POMP models.
 
@@
 Algorithmic advances are needed to make plug-and-play likelihood maximization for POMP models more numerically efficient.
-We investigate the use of AD for this purpose.
+We investigate the use of automatic differentiation (AD) for this purpose.
 
-\subsection{Automatic Differentiation for Particle Filters}
+%\subsection{Automatic Differentiation for Particle Filters}
 
-Recent advances in automatic differentiation (AD) for particle filters \cite{naesseth18, jonschkowski18, corenflos21, scibior21, singh22} have drawn attention to AD as a tool for inference in partially observed stochastic processes.
+Recent advances in AD for particle filters \cite{naesseth18, jonschkowski18, corenflos21, scibior21, singh22} have drawn attention to AD as a tool for inference in partially observed stochastic processes.
 However, existing approaches are either asymptotically biased \cite{naesseth18, jonschkowski18}, have high variance \cite{poyiadjis11, scibior21}, high computational expense \cite{corenflos21, chen24}, or need access to transition densities \cite{poyiadjis11, scibior21, singh22, chen24}.
@@
 
-We derive a linear convergence rate for stochastic gradient descent (SGD) using MOP($\alpha$) in the  presence of strong convexity.
+MOP-$\alpha$ avoids the discontinuity that is problematic for the standard particle filter.
+This makes MOP-$\alpha$ convenient for the application of AD.
+Yet, MOP-$\alpha$  can nevertheless provide a consistent log-likelihood estimate.
+
+We derive a linear convergence rate for stochastic gradient descent (SGD) using MOP-$\alpha$ in the  presence of strong convexity.
 Critically, the estimator is effective within a neighborhood of the maximum but is relatively slow at reaching this neighborhood.
@@
 Directly differentiating through a particle filter, for a fixed value of $\omega$, leads to a biased gradient estimator due to the discreteness of the resampling operation \cite{corenflos21}.
-We avoid this issue by constructing a particle filtering algorithm called {\it Measurement Off-Parameter} with discount factor $\alpha$ (MOP-$\alpha$) for which the resampling is invariant to $\omega$ but dependent on $\theta$.
-Pseudocode for MOP-$\alpha$ is presented in Algorithm~\ref{alg:mop}.
-As we explain subsequently, MOP-$\alpha$ is constructed so that direct differentiation through this algorithm obtains the score estimator of  \cite{poyiadjis11, scibior21} when $\alpha=1$ and that of \cite{naesseth18} when $\alpha=0$.
+We avoid this issue by constructing a particle filtering algorithm called {\it Measurement Off-Parameter} with discount factor $\alpha$ (MOP-$\alpha$), defined by the pseudocode in Algorithm~\ref{alg:mop}.
+For MOP-$\alpha$, the resampling indices are invariant to $\theta$ when $\omega$ and $\phi$ are fixed.
+We assume that the measurement model is differentiable in $\theta$.
+Further, we assume that the simulator is a differentiable function of $\theta$ for every fixed $\omega$, a condition that requires the latent process to be a continuous random variable.
+\ed{WE COULD CALL THIS THE REPARAMETERIZATION TRICK, BUT HERE FOR AD IT ARISES THROUGH THE STRUCTURE OF THE CODE SO WE DO NOT HAVE TO EXPLICITLY THINK OF THE EXISTENCE OF A REPARAMETERIZATION.}
+Under these assumptions, direct differentiation of MOP-$\alpha$ is available via AD.
+As we explain subsequently, MOP-$\alpha$ is constructed so this direct derivative obtains the score estimator of  \cite{poyiadjis11, scibior21} when $\alpha=1$ and that of \cite{naesseth18} when $\alpha=0$.
 Setting $\alpha<1$ adds bias but reduces variance, raising opportunities for a favorable tradeoff.
@@
 We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood when $\alpha=1$ or $\theta=\phi$.
+Employing $\alpha<1$ lead to inconsistency when $\theta$ deviates from $\phi$, but this bias disappears as $\theta$ approaches $\phi$.
+
 While the result is presented here as specific to MOP-$\alpha$, we actually prove a more general result in the supplementary material.

### 2024-06-10 paper/appendix/convergence.tex
Kevin: convergence.tex, etc

--- 2024-05-01 mop/draft.tex
@@
-\documentclass{article}
-\usepackage[margin=2cm]{geometry}
-\usepackage{natbib}
-\usepackage{hyperref}
-\usepackage{enumitem}
-\usepackage{tikz}
-\usetikzlibrary{arrows.meta}
-\usetikzlibrary{positioning}
-
-\setlist{nosep} % to reduce space in lists
-
-\setlength{\parskip}{0.5em}
-\setlength{\parindent}{0in}
-
-\input{macros}
-
-\newcommand{\off}{\operatorname{off}}
-\newcommand{\on}{\operatorname{on}}
-
-\title{Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation}
-\author{Kevin Tan, Giles J. Hooker, Edward L. Ionides}
-\date{}
-
-\comments=1
-
-\begin{document}
-\maketitle
-\begin{abstract}
-
-Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods.
-Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques due to (1) the requirement for simulation-based inference that does not require access to the system's transition probabilities and (2) the issue that widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters.
-We propose a new representation that embeds existing methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff.
-Further, we develop optimization algorithms suited to the Monte Carlo properties of the derivative estimate, including a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
-Promising numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from iterated filtering can beat current state-of-the-art methods on a challenging scientific benchmark problem.
-\end{abstract}
-
-\section{Introduction}
-
-Maximum-likelihood estimation for inference in partially observed stochastic processes, also known as, continuous-state continuous-time hidden Markov models (HMMs), partially-observed Markov processes (POMPs), or partially-observed stochastic nonlinear dynamical systems, is a challenging problem that faces several challenges:
-\begin{enumerate}
-    \item Intractable likelihood functions, bypassed with the approaches of simulated likelihood and likelihood-free inference. 
-    \item The desire for simulation-based inference without access to the transition density of the latent Markov process, with various methods created by \cite{welch2009abc, wood2010sl, doucet2010pmcmc, ionides08, ionides15}.
-    \item Accurate parameter estimation in the presence of significant Monte-Carlo noise in the likelihood estimate.
-\end{enumerate}
-
-Many approaches to inference in these scenarios (such as the EM algorithm, the various Kalman filter variants, and MCMC) either struggle with intractable likelihood functions when the models are complex, or assume access to the probability density of next states given the current state.
-This is a problem in some critical applications, such as disease modeling, where the models are complex enough such that even obtaining the transition density is an intractable problem.
-However, the particle filter with the bootstrap proposal, a popular method for solving the filtering problem in partially-observed dynamical systems, provides an unbiased estimate of the likelihood (\cite{delmoral2004feynman}) without requiring evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
-
-This is the backbone behind the only simulation-based full-information maximum likelihood method available in the literature thus far, the improved iterated filtering algorithm of \citet{ionides15}.
-\citet{ionides15} utilize the bootstrap filter for perform maximum likelihood estimation with an iterated perturbed Bayes map, successfully tackling challenging problems in epidemiology that available Bayesian methodology fails to solve.
-
-Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small.
-For example, though in practice the algorithm of \citet{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood.
-\citet{Ionides_mcap} attempt to address the third problem, albeit with a computationally-expensive approach.
-
-\subsection{Automatic Differentiation for Particle Filters}
-
-Recent advances in automatic differentiation (AD) for particle filters (\cite{blei2018vsmc, jon2018diffpf, corenflos21, scibior2021dpf, doucet2022particlebased}) have drawn attention to AD as an alternative tool for inference in partially observed stochastic processes with the particle filter.
-However, these existing approaches either are not yet compatible with the desire for simulation-based inference, or are computationally expensive.
-
-\citet{blei2018vsmc} and \citet{jon2018diffpf} drop high-variance resampling terms from the gradient, leading to bias that does not vanish asymptotically according to \citet{corenflos21}.
-Though the approach of \citet{corenflos21} is promising, it is computationally expensive.
-\citet{doucet2022particlebased} provide another promising fixed-lag smoothing-based approach that is similar in spirit to our method (MOP-$\alpha$), but requires either (1) access to the process model transition densities or (2) the special case where the transition density factors into a policy that selects actions (in the reinforcement learning sense) and a deterministic model that maps states and actions to next-timestep states.
-
-Finally, \citet{scibior2021dpf} demonstrate that the estimators derived in \citet{poyiadjis11} can be attained with AD, albeit with a simple tweak to the particle filter that does not modify its likelihood estimates.
-Though their method is not fully compatible with simulation-based inference, we show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.
-
-\subsection{Our Contributions}
-
-Previous research on AD for particle filters has struggled with various issues: 
-\begin{enumerate}
-    \item Bias, if the discontinuous nature of particle resampling is ignored.
-    \item High Monte Carlo variance, if continuity corrections lead to numerical instability.
-    \item High computational cost, for algorithms which involve pairwise interactions between particles, marginalization, smoothing or optimal transport.
-    \item Reduced applicability for algorithms which lose or fail to take advantage of the simulation-based capabilities of the bootstrap filter.
-\end{enumerate}
-
-We develop a new approach to statistical inference via AD for particle filters which addresses these concerns.
-We develop a new theoretical framework that encompasses existing methods, and addresses the seemingly incompatible paradox of differentiating through a Monte-Carlo algorithm with discontinuous resampling.
-
-This is done through the construction of a smooth extension to the particle filter, as well as a \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) family of algorithms that (informally) encompasses the special case of AD of a vanilla PF (recovering the gradient estimator of \citet{blei2018vsmc}) when $\alpha=0$, and the DPF algorithm from \citet{scibior2021dpf} (that recovers the gradient estimator from \citet{poyiadjis11}) when $\alpha=1$.
-When $\alpha \in (0,1)$, this provides opportunities for optimizing a bias/variance tradeoff.
-
-We show these off-policy particle filters successfully target the posterior, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties.
-To cap off the theoretical results, we also derive a linear convergence rate for gradient methods on the particle filter in the presence of strong convexity, providing support for numerical optimization with AD for the particle filter.
-
-In practice, we propose a hybrid algorithm that we call \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts gradient ascent (or a similar iterative first or second-order gradient method) with a coarse solution obtained from a few rounds of iterated filtering.
-While this requires a differentiable simulator compatible with the reparametrization trick (as previous literature such as \citet{corenflos21} also require), future work will deal with workarounds inspired by our framework that use likelihood ratios of transition densities in place of this.
-Promising numerical results indicate that IFAD beats IF2 (and by the transitive property, IF1, the Liu-West filter, and particle MCMC) on a challenging scientific benchmark problem.
-
-While the Monte-Carlo Adjusted Profile (MCAP) from \citet{Ionides_mcap} provides a profile likelihood confidence interval that takes into account both Monte Carlo profile error and statistical uncertainty in the likelihood, it is computationally inefficient.
-\kevin{TODO} We provide a method for incorporating the MCAP methodology (with the poor man's likelihood profile) in our IFAD algorithm that provides a more computationally efficient profile likelihood confidence interval.
-\kevin{Quantile regression for estimating the max/the envelope of the likelihood evaluations from AD, then fit MCAP/MQLE/MSLE?}
-
-Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual} that we call \texttt{pompy}.
-The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{jax} Python package from \citet{jax2018github}.
-
-\kevin{Terminology: Plug-and-play (clearly definable, but not well-known, define at start of paper and don't use in title/abstract), likelihood-free maximum likelihood (seemingly an oxymoron, rationale being there is no need to evaluate density, bayesians call likelihood densities, plus we only approximate the likelihood and don't really have it), simulation-based inference (loaded term)}
-
-\subsection{Significance Statement}
-Many scientific models involve highly nonlinear stochastic dynamical systems which can be observed only via noisy and incomplete measurements.
-Under the Markov assumption on system dynamics, previous work has provided methods of performing inference for these models.
-In particular, prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics.
-We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
-Our new method outperforms previous approaches on a challenging problem in epidemiology.
-This is implemented in a precursor to a Python counterpart to the popular \texttt{pomp} package by \citet{king16, king2017pompmanual} that includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation.
-
-\kevin{TODO: (1) Fix beta smoothness to d(n) at no more than O(n), (2) add story to say that no one spotted that it could have been plug-and-play because of e.g. reasons why they would stumble}
-
-\section{Problem Setup}
-
-Consider an unobserved Markov process $\{X_t, t \geq t_0\}$, and observations $Y_1,...,Y_N$ at timesteps $t_1,..., t_N$.
-The process is parameterized by an unknown parameter $\theta \in \Theta$, where $X \in \gX, Y \in \gY,$ and finally $\gX \subseteq \R^d, \Theta \subseteq \R^p$.
-By a similar decomposition to that in \citet{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
-$$f_{X_{0: N}, Y_{1: N}}\left(x_{0: N}, y_{1: N} ; \theta\right)=f_{X_0}\left(x_0 ; \theta\right) \prod_{n=1}^N f_{X_n \mid X_{n-1}}\left(x_n \mid x_{n-1} ; \theta\right) f_{Y_n \mid X_n}\left(y_n \mid x_n ; \theta\right).$$
-
-We call $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ the process model, writing $\texttt{process}\left(x_n, \theta\right)$ for the simulator corresponding to the above process model, $f_{Y_n|X_n}\left(y_n \mid x_n, \theta\right)$ the measurement model, and will write $y_n^*$ for the actual values of the observations that were observed.
-
-The above are all defined on a single probability space $(\Omega, \Sigma, \prob),$ where $\omega \in \Omega$ denotes an element of the sample space that can be thought of as a random seed.
-For notational simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0.$
-
-\paragraph{Seed-Fixing, Coupling, and the Method of Common Random Numbers, Coupling:}
-
-In the rest of this paper, we will identify the seed with an element of the sample space.
-We do so because the seed determines the entire sequence of pseudo-random numbers as generated by a computer, to maintain an algorithmic and practical interpretation of our theory.
-When two algorithms share the same seed, this corresponds to the method of common random numbers.
-
-\section{A Smooth Extension of the Particle Filter}
-
-%\paragraph{Motivation:} 
-%The gradient of the log-likelihood can be written as a vector $\nabla_\theta \ell(\theta)$ such that 
-%\begin{equation}
-    %\lim_{v \to 0} \frac{|\ell(\theta+v) - \ell(\theta) - v^T \nabla_\theta \ell(\theta)|}{\|v\|} = 0.
-%\end{equation}
-
-%We therefore need to evaluate the likelihood at a point $\theta$ and a nearby point $\theta + v$ to estimate the gradient with finite differences.
-But in the presence of Monte Carlo noise in evaluating the likelihood, this is difficult, unless (1) one ensures one uses the same seed for each evaluation in the neighborhood, and (2) chooses a step-size small enough that changing the the parameter does not change the resampling indices.
-
-\paragraph{Intuition:} 
-Our theoretical framework for automatic differentiation for the particle filter hinges on the following observation.
-Instead of seeking to differentiate $\ell(\theta)$ directly, we instead focus on differentiating through a suitable fixed-seed reweighting of a bootstrap filter.
-The intuition is as follows.
-Consider a particle filter likelihood estimate where the system (i.e.
-\textbf{both the state transitions and resampling}) evolves under a parameter $\phi \in \Theta$ and the seed, or sample space element, is fixed at $\omega \in \Omega.$ Conditional on $\omega$ and $\phi,$ one can reweight the likelihood estimate to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$
-
-Differentiating this reweighted likelihood estimate with respect to $\theta$ then amounts to differentiating through the reweighting likelihood ratios, which is differentiable if the densities used for the reweighting are differentiable.
-We call this simulation-and-reweighting scheme an \textbf{off-policy} likelihood estimate, as it is similar in spirit to the off-policy methods found in the reinforcement learning literature.
-
-The above intuition is formalized below.
-
-\kevin{Method of Common Random Numbers, Coupling}
-
-\subsection{Doubly Off-Policy Particle Filters}
-
-\begin{defn}[Seed-Fixing]
-\label{defn:seed-fixing}
-Consider a probability space $(\Omega, \Sigma, \mu)$ governing the particles and observations.
-Noting that the particle filter is unbiased for the likelihood when computing the expectation of the normalizing factor under the posterior $f_{X_{0:N|Y_{1:N}}}\left(x_{0: N} \mid y_{1: N}; \theta\right)$, we can write the likelihood as
-$$
-\begin{aligned}
-\mathcal{L}(\theta) & =\int_{\Omega} \mathcal{L}(\theta, \omega, J) d \mu(\omega) \\
-& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right)\right) f_{X_{0:N}, Y_{1:N}}\left(x_{0: N}^P(\omega) , y_{1: N}^*, \theta\right) d \mu(\omega) \\
-& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right) f_{X_n|X_{n-1}}(x_{n,j}^P(\omega)|x_{n-1,j}^F(\omega), \theta)\right) f_{X_{0:N}| Y_{1:N}}\left(x_{n, j}^P(\omega) \mid y_n^*, \theta\right) d \mu(\omega)
-\end{aligned}
-$$
-where $\mathcal{L}(\theta, \omega, J)$ is the Monte-Carlo estimate of the likelihood given the realization of $\mathrm{J}$ particles corresponding to $\omega \in \Omega$.
-For simplicity, we resample every timestep.
-We call $\omega$ a seed, and the deterministic quantity $\mathcal{L}(\theta, \omega, J)$ the seed-fixed estimate.
-\end{defn}
-
-Armed with this seed-fixed likelihood estimate, we modify a particle filter likelihood estimate where the system evolves under a parameter $\phi \in \Theta$ and seed $\omega \in \Omega$ to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$
-
-\begin{defn}[Doubly Off-Policy Particle Filter Likelihood]
-\label{defn:doubly-off-policy}
-Let $\lik(\theta, \phi, \omega, J)$ denote a Monte-Carlo estimate of the likelihood evaluated at $\theta \in \Theta$ with $J$ particles, but where the system evolves according to another $\phi \in \Theta$ and conditional on a fixed seed $\omega \in \Omega$.
-This can be computed with a suitable reweighting of the particles, which we provide the derivation for later.
-We then have
-\begin{equation} \label{eq:loglik2}
-\lik(\theta)=\E[\lik(\theta,\phi,\Omega, J)]=\int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega),
-\end{equation}
-where we define
-\begin{align}
-    \lik(\theta,\phi,\omega, J) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J &\left(f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) \cdot 
-    f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}(\omega) | x_{n-1,j}^{F,\phi}, \phi) \right.\\
-    &\cdot
-    \left.
-    \frac{f_{Y_n|X_n}(y^*_n|x_{n,j}^{P,\phi}(\omega),\theta)}{f_{Y_n|X_n}(y^*_n|x_{n,j}^{P,\phi}(\omega),\phi)}
-    \cdot \frac{f_{X_n|X_{n-1}}(x_{n,j}^{P,\phi}(\omega)|x_{n-1,j}^{F,\phi}(\omega),\theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P,\phi}(\omega)|x_{n-1,j}^{F,\phi}(\omega),\phi)}\right).
-\end{align}
-\end{defn}
-
-Lastly, we require the following regularity conditions.
-
-\begin{aspt}[Continuity of the Likelihood] $\ell(\theta)$ proper is continuous in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.
-\end{aspt}
-
-\begin{aspt}[Bounded Process Model]
-\label{assump:bounded-process}
-    There exist $\underbar{M}, \bar{M}$ such that $0 < \underbar{M} \leq f_{X_n|X_{n-1}}(x_n | x_{n-1}) \leq \bar{M} < \infty$. 
-\end{aspt}
-
-\begin{aspt}[Bounded Measurement Model]
-\label{assump:bounded-measurement}
-There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G} < \infty$ and $\infty > G' \geq G'(\theta)>0$ such that $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty <\leq G'(\theta) \leq G'$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
-\end{aspt}
-
-\begin{aspt}[Locally Bounded Derivatives]
-\label{assump:local-bounded-derivative}
-Let $\mathcal{M}$ be an open subset of $\Omega$.
-There exists some function $G(\theta)$ and constants $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$, $H=\sup _{\theta \in \mathcal{N}} H(\theta)<\infty$ such that
-    $$
-    \|\nabla_\theta \ell(\theta, \phi, \omega, J)\|_\infty \leq G(\theta) \leq G<\infty, \;\;
-    \|\nabla^2_\theta \ell(\theta, \phi, \omega, J)\|_\infty \leq G(\theta)^2 \leq G^2<\infty
-    $$
-    for every $\phi \in \mathcal{M}, \theta \in \mathcal{N}_J(\phi)$, and almost every $\omega \in \Omega$.
-\end{aspt}
-
-We need the following assumption to ensure that the weighted particle filter with off-policy resampling targets the posterior:
-\begin{aspt}[Bounded And Differentiable Measurement Density Ratios]
-    \label{assump:diff-meas}
-    For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{Y_n|X_n}(y_n^*|x, \theta)}{f_{Y_n|X_n}(y_n^*|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x \in \gX$ and all $\theta\in\gN(\phi)$.
-\end{aspt}
-
-In the following assumption, the first condition is sufficient for DOP, and the second for MOP:
-\begin{aspt}[Bounded And Differentiable Process Model]
-    \label{assump:diff-process}
-    Either one of the following two conditions hold:
-    \begin{enumerate}
-        \item For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{X_n|X_{n-1}}(x'|x, \theta)}{f_{X_n|X_{n-1}}(x'|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x, x' \in \gX$ and all $\theta\in\gN(\phi)$, or
-        \item There exists a differentiable simulator with more than two continuous derivatives for the process model. 
-    \end{enumerate}
-\end{aspt}
-
-We want $\phi\in\Theta$ to be a point in parameter space around which $\lik(\theta,\phi,\omega, J)$ is differentiable with respect to $\theta$ and does not have excessive Monte~Carlo variance.
-
-We require many of these assumptions to establish the following proposition, the proof of which is a straightforward consequence of what we assumed.
-
-\begin{prop}[Smooth Neighborhood]
-\label{prop:smooth-nbhd}
-There exists an open neighborhood $\mathcal{N}(\phi)$ around any $\phi \in \Theta$ where for all $\theta \in \mathcal{N}(\phi)$ and almost every $\omega \in \Omega$, the Monte Carlo estimate of the likelihood at $\theta \in \mathcal{N}(\phi)$ with the system evolving according to $\phi$ conditional on $\omega$, $\lik(\theta,\phi,\omega, J)$, has two continuous derivatives in $\theta$.
-\end{prop}
-
-\begin{proof}
-    For a fixed $\omega$, the resampling indices for $\lik(\theta,\phi,\omega, J)$ and $\lik(\phi,\phi,\omega, J)$ are the same as the resampling indices in both cases correspond to that yielded by $f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}(\omega);\phi)$, eliminating any discontinuities from resampling.
-
-    By Assumptions \ref{assump:diff-meas} and \ref{assump:diff-process}, for small enough $\mathcal{N}(\phi)$ the density ratios are bounded.
-The likelihood only changes by a factor of the density ratios, which are bounded and possess two continuous derivatives in $\theta$ if either the process and measurement densities used in the calculation do, or if the measurement densities do and one has access to a differentiable simulator with two continuous derivatives in $\theta$.
-This gives us the desired result.
-
-    As a remark, we note that the neighborhoods the sequence is differentiable in are independent of $J$, as the resampling does not change with $\theta$ -- it is dependent on $\phi$. 
-\end{proof}
-
-\subsubsection{Doubly Off-Policy Particle Filter}
-
-Having the system evolve according to $\phi$ but evaluating the likelihood at $\theta$ requires us to reweight the conditional likelihood at each timestep by likelihood ratios corresponding to the process and measurement models.
-This corresponds to making two corrections in the particle weights at each timestep, by multiplying by
-\begin{equation}
-    s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}, \qquad r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)},
-\end{equation}
-as we illustrate in Appendix \ref{app:dop}. We use the notation $x_{n,j}^{P, \phi}$ to emphasize that the particles evolve according to $\phi$.
-
-We claim that this yields a particle filter that targets the posterior and recovers the derivative estimate of \citet{poyiadjis11, scibior2021dpf} when $\theta$ is evaluated at $\phi$.
-The proof can be found in Appendix \ref{app:dop}.
-
-\begin{prop}[Correctness of Doubly Off-Policy Particle Filter Corrections]
-    \label{prop:dop-correctness}
-
-    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$.
-Also let $\theta \in \gN(\phi)$.
-Then, the following corrections in the particle weights,
-    \begin{equation}
-    s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}, \qquad r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)},
-    \end{equation}
-    yield a particle filter that is unbiased for the likelihood, even when $\theta \neq \phi$.
-
-    In particular, when $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{poyiadjis11},
-    \begin{equation}
-        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
-    \end{equation}
-\end{prop}
-
-\paragraph{Intuition:} Having the system evolve according to $\phi$ but evaluating the likelihood at $\theta$ requires us to reweight the conditional likelihood at each timestep by likelihood ratios corresponding to the process and measurement models.
-This corresponds to making two corrections in the particle weights at each timestep, one for the process model and one for the measurement model.
-The measurement correction cannot be avoided, but one can bypass the process correction -- with the reparametrization trick.
-We can do when we assume the simulator is differentiable with the reparametrization trick, as \cite{corenflos21} also assume.
-Doing so then yields an algorithm that only requires a simulator, and does not require the process densities.
-
-\subsection{Measurement Off-Policy Particle Filters}
-
-With the above procedure, we have the entire system evolve according to $\phi$, and we differentiate through the likelihood ratios, which are smooth conditional on the resampling being the same (which in turn happens when $\theta \approx \phi$).
-However, doing so requires access to likelihood ratios, and as such Algorithm \ref{alg:dop} is therefore not compatible with simulation-based inference.
-
-There is a workaround.
-Consider the case where you have access to a differentiable simulator compatible with the reparametrization trick.
-If we have the state transitions evolve according to $\theta,$ but have the resampling according to $\phi,$ we can offload the derivative with respect to the process model to the reparametrization trick, as in \cite{corenflos21}.
-This corresponds to the MOP-$\alpha$ procedure in Algorithm \ref{alg:mop}.
-
-We therefore make the following assumption in this section only:
-\begin{aspt}[Differentiable Simulator]
-    $\texttt{process}(\cdot | X_{t-1}; \theta)$ is differentiable with respect to $\theta,$ and is compatible with the reparametrization trick. 
-\end{aspt}
-
-Algorithm \ref{alg:mop} provides a weighted particle filter that targets the posterior under $\theta$, but when differentiated yields either the low-bias but high-variance gradient estimator of \cite{poyiadjis11}, the high-bias but low-variance estimator of \cite{blei2018vsmc}, or an intermediate, interpolating quantity that can optimize a bias-variance tradeoff.
-This is controlled by a discounting parameter $\alpha$.
-The intuition behind this procedure is as follows.
-
-\paragraph{Intuition:} Broadly, if $\theta\neq\phi,$ one can first run a vanilla particle filter to recover values for the particles with the process model under $\phi.$ One can then run an accordingly reweighted particle filter with the process model under $\theta$, but resampling according to the measurement model under $\phi$, with the same indices as the original vanilla particle filter.
-This requires one to set the same random seed for both particle filters.
-This then yields a slightly-unconventionally weighted particle filter that targets the posterior, but has some very nice gradient properties; as we outline in the remainder of this section.
-
-\paragraph{Special Case:} When $\theta = \phi$, one can simply set the particles under $\phi$ to be copies of the particles under $\theta$.
-Gradients with respect to $\theta$ now treat functions of the particles under $\phi$ as constants, corresponding to the stop-gradient trick in \cite{scibior2021dpf}.
-This therefore provides a theoretical framework for their stop-gradient trick, which we happily use in our numerical experiments for simplicity.
-
-\kevin{Develop theory for mixing-time swapping?
-Detect when IF2 reaches a neighborhood of the MLE, by detecting whether it has reached a local sort of stationary distribution when the perturbations are lower bounded, e.g.
-the autocorrelation between the mean vectors to estimate the mixing time or Gelman-Rubin.
-In practice, just swap when IF2 converges according to the diagnostics.
-Theory doesn't care, you just need to in a basin of attraction that is strongly convex.}
-
-\subsubsection{Discounting Weights}
-
-Observe that at Algorithm \ref{alg:mop}, at each timestep the corrected filtering importance weights $w_{n,j}^{F,\theta}$ optionally propagate to the next timestep for us to maintain the correction.
-We introduce a discounting parameter $\alpha$ that affects the weights, so $w_{n,j}^{P,\theta} := (w_{n-1,j}^{F,\theta})^{\alpha}$.
-Note that no matter what value $\alpha$ takes on, the system evolves according to $\phi$ when we perform the correction to evaluate the likelihood at $\theta$.
-The filter fully targets the posterior when $\theta=\phi$, or $\alpha=1$, as we show in Proposition \ref{prop:dop-correctness}.
-As previously mentioned, the value of $\alpha$ therefore informs a bias-variance tradeoff.
-We elaborate on this below.
-
-\paragraph{Bias-Variance Tradeoff} This discounting serves the purpose of balancing a tradeoff between maintaining memory of each particle's ancestral trajectory (most extreme when $\alpha=1$), or considering only the single-step transition dynamics (when $\alpha=0$).
-This can be thought of as an \textbf{exponentially-weighted moving average}, where $\alpha$ controls the amount of discounting.
-We elaborate on two special cases below.
-
-\kevin{This isn't a smoother, because we'd need future information.}
-
-\paragraph{When $\alpha=1,$} the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\theta}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
-\end{equation}
-as we have a telescoping product. Its derivative when $\theta=\phi$ is then
-\begin{equation}
-    \nabla_\theta \hat{\ell}(\theta) := 
-        \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*),
-\end{equation}
-the average gradient of each surviving particle's cumulative importance weights.
-The filter targets the posterior when $\alpha=1$ even when $\theta \neq \phi,$ and therefore suffers from no asymptotic bias.
-
-\paragraph{When $\alpha=0,$} the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J w_{n,j}^{F,\theta} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})},
-\end{equation}
-and though the telescoping product disappears, this reduces to a series of one-step corrections.
-However, these corrections are insufficient when $\theta \neq \phi,$ and the filter suffers from asymptotic bias unless $\theta=\phi.$ Its gradient when $\theta=\phi$ is therefore
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
-    \end{equation}
-
-the gradient of the vanilla particle filter where the resampling correction is omitted as found in \cite{blei2018vsmc} -- when applied to the bootstrap filter.
-As \cite{scibior2021dpf} remark, this estimator only considers single-step quantities, and is "memoryless" beyond a single step.
-This is in contrast to the case when $\alpha=1,$ as that estimate only considers the surviving particles at time $N$, which is the other extreme where dependencies across time are fully tracked.
-
-\subsubsection{Correctness of MOP-$\alpha$}
-
-\begin{prop}[Correctness of MOP-$\alpha$ (Algorithm \ref{alg:mop})]
-    \label{prop:mop-correctness}
-    When either $\alpha=1$ or $\theta=\phi$, Algorithm \ref{alg:mop} is a particle filter that targets the posterior and is unbiased for the likelihood under $\theta$.
-
-    In particular, when $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter.
-In this case, its gradient when $\alpha=1$ is the same as the DPF estimator of \citet{scibior21} when applied to the bootstrap filter,
-    \begin{equation}
-        \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{0:N}}(y_{1:N}^* | x_{1:N,j}^{A, F,\theta}),
-    \end{equation}
-
-    and when $\alpha=0,$ is the gradient estimator of \cite{blei2018vsmc},
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
-    \end{equation}
-
-    that as \cite{scibior21} show corresponds to differentiating through the fixed-seed likelihood estimate yielded by a vanilla particle filter, dropping resampling terms.
-
-    When $\alpha=1$, the gradient estimate MOP-1 provides is consistent for the score.
-\end{prop}
-
-The proof for the above follows from Appendices A and C, as well as the technical lemmas for properly weighted particle filters.
-
-\kevin{Take derivative of loglik with regard to alpha?}
-\kevin{Talk about regime change the moment you move away from alpha=1, variance going from N2 to N, and bias going from 0 to something, but try to prove that decrease in variance outweights increase in bias.
-}
-\kevin{Create tools for choosing alpha via binary search CV from an averaged alpha=1 target}
-\kevin{suppose at each filtering step, look at gradient calculated at alpha=0, dynamically select alpha during filtering.
-calculate bias and variance online during filtering, and update alpha accordingly.
-}
-
-\kevin{No need to recompute so many particle filters! Just take an array of weight copies, one for each alpha. Gradients for each alpha come free the moment you compute one of them. }
-
-\kevin{Elbow method on variance, since bias is hard to compute? }
-\kevin{Get variance estimate, bias estimate, construct confidence intervals on bias estimate, use that to get unbiased estimate of the MSE. }
-
-\subsubsection{Practical Considerations}
-
-It is only ever practical to evaluate $\theta$ at $\phi$.
-This removes the issue of the likelihood ratios blowing up, and yields an estimate of the likelihood that agrees with the bootstrap filter estimate.
-Its gradient in this case is either the estimate of \citet{poyiadjis11} when $\alpha=1$, a single-timestep estimator that corresponds to naively taking the derivative of the particle filter when $\alpha=0$ as first proposed in \citet{blei2018vsmc}
-
-\paragraph{Bottom Line:} The DOP-$\alpha$ and MOP-$\alpha$ procedures therefore provide us with a unifying framework for reasoning about smooth extensions of weighted particle filters that target the posterior and provide reasonable gradient estimates when differentiated.
-This fixed-seed reweighting alleviates issues that can arise from differentiating through a quantity yielded by ostensibly non-differentiable resampling.
-This is done by either (1) differentiating through smooth likelihood ratios, or (2) differentiating through a differentiable simulator with the reparametrization trick (and through smooth likelihood ratios to adjust for the resampling).
-In particular, MOP-$\alpha$ does not require access to the transition densities, and like the bootstrap filter, only requires access to a (albeit differentiable) simulator.
-To our knowledge, this is the first particle filter gradient method in the literature with this capability.
-
-\section{Practical Maximum-Likelihood Inference}
-
-This still leaves us with the question of designing a reasonably effective procedure for likelihood maximization with these new tools.
-We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that amounts to warm-starting gradient descent or some other reasonable first or second order procedure (using the gradient estimate given by MOP-$\alpha$) with the output of an initial search of IF2.
-
-\paragraph{Summary:} A coarse solution is found by running IF2 with an aggressive random walk standard deviation (learning rate) and a cooling schedule that is \textbf{bounded below}, until the search stalls to an initial "convergence" to what is hopefully a neighborhood of the MLE.
-We can then use gradient descent, or other first or second order gradient methods, to refine this coarse solution.
-
-The initial "convergence" happens fairly quickly in practice.
-In the case of the Dhaka cholera model of \cite{king08}, when a geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations of IF2.
-In comparison, often 100 or 200 iterations of IF2 are used for an initial global search.
-\cite{ionides15}, for example, use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{Lee_haiticholera}.
-
-\paragraph{Rationale:}
-
-IF2 converges quickly to a neighborhood of some reasonably good local optimum in practice, but then struggles to squeeze out the last few log-likelihood units.
-Conversely, gradient descent (and other first or second order gradient methods) can get stuck in saddle points and less desirable local minima when the likelihood is nonconvex.
-Warm-starting gradient methods with IF2 lets us get close enough to the MLE for gradient methods to shine, as in most cases regularity conditions ensure that the likelihood surface is well-behaved enough in this neighborhood that the aforementioned issues with saddle points and local minima are alleviated.
-Combining these two methods in this way lets us enjoy the best of both worlds.
-
-\begin{algorithm}[ht]
-\centering
-	\caption{Iterated Filtering with Automatic Differentiation}
-    \label{alg:ifad}
-	\begin{algorithmic}[1]
-	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $p(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta),$ IF2 cooling schedule $\eta_t$, MOP-$\alpha$ discounting parameter $\alpha$.
-		\STATE Initialize $\theta_0$, $t=0.$
-        \STATE Run IF2 until initial "convergence" under cooling schedule $\eta_t$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_t := \frac{1}{J}\sum_{j=1}^J \Theta_j.$
-		\WHILE{Procedure not converged:}
-		\STATE Draw $\omega \in \Omega$
-		\STATE Perform forward pass by running Algorithm \ref{alg:mop} with discounting parameter $\alpha$ to obtain $\loglik(\theta_t, \theta_t, \omega, J).$
-		\STATE Obtain $g(\theta_t) = \nabla_{\theta_t} (-\loglik(\theta_t, \theta_t, \omega, J))$ via backpropagation, $H(\theta_t)$ such that $\lambda_{\min}(H) \geq c$.
-		\STATE Obtain $\eta$ with line search or an annealing schedule.
-		\STATE Update $\theta_{t+1} := \theta_t - \eta (H(\theta_t))^{-1} g(\theta_t)$, $t:=t+1.$
-		\ENDWHILE
-		\STATE Return $\hat{\theta} := \theta_t.$
-	\end{algorithmic}
-\end{algorithm}
-
-\subsection{Theoretical Guarantee}
-
-We first analyze the performance of gradient descent, Newton's method, or any iterative method that multiplies the particle estimate of the gradient with the inverse of a matrix with minimum eigenvalue bounded below, in the case where the log-likelihood surface is $\gamma$-strongly convex, and where the gradient estimate is a particle approximation of the score (for example, IFAD-1 and MOP-1).
-In this case, these methods enjoy a linear convergence guarantee:
-
-\kevin{I think we can establish the linear convergence for the entire procedure in the same sense as \href{this}{https://arxiv.org/pdf/2311.08442.pdf}, in the sense that if IF2 is run for a sufficiently large but fixed number of iterations, once we arrive in a local strongly convex region, we get linear convergence when we swap to MOP.
-Show by induction that we remain in the strongly convex region (in terms of parameter space and likelihood), eqn 218.
-}
-
-\begin{thm}[Convergence of Gradient Methods]
-\label{thm:convergence}
-
-Consider a variant of IFAD-1 (Algorithm \ref{alg:ifad}) where one stops if the gradient estimate $||g(\theta_n)|| \leq \sigma \epsilon$.
-Assume $-\ell$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$.
-$\sigma = \frac{4 \Gamma}{(1-\beta)}> 4$.
-Define $G$ as in Assumption \ref{assump:local-bounded-derivative}.
-
-Choose $J$ according to the maximum of the $J$ prescribed in Lemmas \ref{lemma:grad_bound} and (if using a Hessian estimate) $\ref{lemma:hess_bound}$ with failure probability $\delta/2$ each.
-Let $H(\theta_n)$ be a matrix possibly dependent on $\theta_n$ with minimum eigenvalue always greater than some $c>0$.
-Choose the learning rate $\eta$ and error tolerance $\epsilon$ such that, where $\beta$ is the Armijo condition hyperparameter,
-\begin{equation}
-    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_n)||.
-\end{equation}
-
-Then the following holds with probability at least $1-\delta$:
-\begin{equation}
-\ell(\theta^*) - \ell(\theta_{n+1}) \leq \left(1-\eta\beta\frac{8\gamma}{9c}\right)(\ell(\theta^*)-\ell(\theta_n))
-\end{equation}
-
-and the algorithm eventually terminates after we obtain $||\nabla_\theta \ell(\theta_n)|| \leq (1+\sigma) \epsilon$.
-\end{thm}
-
-The proof of this result uses the concentration inequalities from \cite{delmoral2011ci}, and leans upon the analysis of \cite{mahoney16}. We defer the proof to Appendix \ref{app:ifad}
-
-\paragraph{Takeaways for IFAD (Algorithm \ref{alg:ifad}):}
-
-The gradient stage of IFAD-1 converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the IF2 stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
-This happens fairly often in practice, for example, when sufficient regularity conditions for local asymptotic normality of the MLE hold.
-
-We conjecture that this applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE.
-IF2 posesses some SGD-like qualities, including what looks like linear convergence to a ball around the MLE with width determined by the random walk standard deviation.
-We will explore this conjecture in future work.
-
-While this theoretical guarantee currently only holds for IFAD-1, as it forms a particle approximation of the score as in \cite{poyiadjis11}, one can also extend this for a similar guarantee if the bias for the gradient estimate given by MOP-$\alpha$ for $\alpha < 1$ is small enough.
-For general $\alpha$, including MOP-0, the analysis then becomes more complicated, as one needs to handle the case of biased gradient descent.
-However, if future measurements $y_{n+1:N}^*$ do not provide a large amount of information on the identification of the current state $x_n$ given past and current measurements $y_{0:n}^*$, as mentioned in \cite{corenflos21}, the bias may be small enough for it to not present too much of a problem.
-
-\subsection{Runtime}
-
-Even with the increased computational cost of automatic differentiation, Algorithm \ref{alg:ifad} is reasonably fast, as by the cheap gradient principle from \cite{kakade2019provably}, a gradient operation takes no more than 6 times the runtime of the particle filter (in practice, this means a gradient step takes about 3.75x as much time as an IF2 iteration).
-As such, unlike methods like the marginal differentiable particle filter in \cite{scibior2021dpf} or the the optimal transport of resampling approach of \cite{corenflos21}, our algorithm has runtime \textbf{linear, and not quadratic}, in the number of particles (while alleviating the $O(N^2)$ variance concern of \cite{poyiadjis11} and \cite{corenflos2021ot} and asymptotic bias concern of \cite{blei2018vsmc} through optimizing $\alpha$ for a bias-variance tradeoff).
-
-In addition, re-implementation of the particle filter, iterated filtering, and Algorithm \ref{alg:ifad} itself in \texttt{JAX} (\cite{jax2018github}) led to considerable speedups.
-This can be attributed to both GPU acceleration and the just-in-time (JIT) compilation functionality that \texttt{JAX} provides.
-When tested on a computer with an Intel i9-13900K processor, 128GB of RAM, and a NVIDIA RTX3090 GPU, each post-compilation particle filtering operation on the Dhaka cholera model from \cite{king08} with 10000 particles took roughly 379 milliseconds (with a standard deviation of 279 microseconds over seven runs), a considerable improvement over the existing performance in the widely used \texttt{pomp} package of \cite{king16}, which took between 6.19 to 6.39 seconds over ten runs on the same machine.
-
-\section{Numerical Experiments}
-
-\subsection{Cholera in Bangladesh}
-
-To compare the performance of IFAD against that of IF2, we fit the very same model first proposed in \cite{king08} that \cite{ionides15} use to demonstrate the capabilities of IF2.
-This is becoming a standard benchmark, as \cite{wood16} and \cite{wycoff2024voronoi} include in their papers.
-This is a stochastic SIR compartmental model with transition and measurement uncertainty, where the population at time $t$, $H(t)$, is divided into the susceptible compartment $S(t)$, infected compartment $I(t)$, and three ($k=3$) recovered compartments $R_1(t), ..., R_k(t)$ denoting varying degrees of cholera immunity.
-We write $M(t)$ for the cholera deaths in each month.
-As in \cite{ionides15}:
-
-\begin{itemize}
-    \item The \textbf{transition dynamics} follow this series of stochastic differential equations:
-
-    \begin{align}
-        d S&=\left(k \epsilon R_k+\delta(S-H)-\lambda(t) S\right) d t+d H-(\sigma S I / H) d B, \\ 
-        d I&=\left(\lambda(t) S-(m+\delta+\gamma) I\right) d t+(\sigma S I / P) d B, \\ 
-        d R_1&=\left(\gamma I-(k \epsilon+\delta) R_1\right) d t, \\ 
-        \vdots \\ 
-        d R_k&=\left(k \epsilon R_{k-1}-(k \epsilon+\delta) R_k\right) d t,
-    \end{align}
-
-    where $B$ represents Brownian motion, $m$ is the rate at which infected individuals succumb to cholera, $\gamma$ is the rate at which infected individuals recover, $1/\epsilon$ is the mean duration of immunity, $\sigma$ represents the standard deviation of perturbations to the force of infection, and $\delta=0.02$ is the death rate.
-
-    \item The \textbf{force of infection}, $\lambda(t)$, is modeled by splines $s_j$
-    \begin{equation}
-        \lambda(t)=\exp \left\{\beta_{\text {trend }}\left(t-t_0\right)+\sum_{j=1}^{6} \beta_j s_j(t)\right\}(I / P) + \exp \left\{\sum_{j=1}^{6} \omega_j s_j(t)\right\},
-    \end{equation}
-    where the $\beta_j$ model seasonality in the force of infection, $\beta_{\text{trend}}$ models the trend in the force of infection, and the $\omega_j$ represent seasonality of a non-human environmental reservoir of disease.
-    \item The \textbf{measurement model} for observed monthly cholera deaths is given by 
-    \begin{equation}
-        Y_n \sim \gN(M_n, \tau^2M_n^2),
-    \end{equation}
-    where $M_n$ is the true number of cholera deaths in that month.
-\end{itemize}
-
-\begin{figure}[ht]
-\centering
-\begin{tikzpicture}[node distance=1cm and 1cm, auto,
->=Latex,every node/.append style={align=center}, 
-int/.style={draw, minimum size=1.6cm}]
-  \node [int] (H) at (0,0) {$H$};
-\node [int] (S) [right=of H] {$S$};
-\node [int] (I) [right=of S] {$I$};
-\node (Mcoord) [right=of I, coordinate] {Mcoord};
-\node [int] (R1) [right=of Mcoord] {$R_1$};
-\node [int] (R2) [right=of R1] {$R_2$};
-\node [int] (R3) [right=of R2] {$R_3$};
-\node [circle, draw] (M) [above=of Mcoord] {$M$};
-\node (deathsS) [above=of S, coordinate] {deathsS};
-\node (deathsI) [above=of I, coordinate] {deathsS};
-\node (deathsR1) [above=of R1, coordinate] {deathsR1};
-\node (deathsR2) [above=of R2, coordinate] {deathsR2};
-\node (deathsR3) [above=of R3, coordinate] {deathsR3};
-\node (belowR3) [below=of R3, coordinate] {belowR3};
-\node (belowS) [below=of S, coordinate] {belowS};
-\path[->]
-    (H) edge node {$dH$} (S)
-    (S) edge node {$\lambda(t)$}  (I)
-    (I) edge node {$\gamma$}  (R1)
-    (R1) edge node {$k\epsilon$}  (R2)
-    (R2) edge node {$k\epsilon$}  (R3)
-    (R3) edge node {} (belowR3) 
-    (belowR3) edge node {$k\epsilon$} (belowS) 
-    (belowS) edge node {} (S)
-    (I) edge node {$m$}  (M)
-    (S) edge node {$\delta$}  (deathsS)
-    (I) edge node {$\delta$}  (deathsI)
-    (R1) edge node {$\delta$}  (deathsR1)
-    (R2) edge node {$\delta$}  (deathsR2)
-    (R3) edge node {$\delta$}  (deathsR3);
-\end{tikzpicture}
-\caption{Illustration of SIR model from \cite{king08}.}
-\label{fig:sir}
-\end{figure}
-
-Figure \ref{fig:sir} depicts the above compartmental model that we use to benchmark the performance of IFAD.
-We present comparisons of IFAD with $\alpha=0$ (IFAD-0) and $\alpha=1$ (IFAD-1) against both (1) the results from \cite{ionides15}, and (2) our own implementation of IF2 in \texttt{JAX}.
-
-\paragraph{Hyperparameters:} While performing an initial exploratory analysis, we found that alternative hyperparameters for IF2 (random walk standard deviation of $\sigma=0.02$, geometrically decreasing cooling rate exponent $a=0.95$) performed better than the settings \cite{ionides15} chose, and chose to use these for our implementation to achieve a fairer comparison.
-Our implementations of IF2 and IFAD used 10000 particles, in line with \cite{ionides15}.
-
-To prevent exploding gradients given the high variance in the MOP-1 estimator, we scale the gradient to have norm 1 before performing vanilla gradient descent, and use a linear cooling schedule of $0.2$ to $ 0.0001$.
-We chose to use vanilla gradient descent for simplicity, and to demonstrate that minimal tuning was required to have IFAD perform well.
-
-% The 60 more iterations of IF2 yielded largely not a huge improvement over the initial coarse solution of 40 iterations.
-This demonstrates what we have mentioned earlier: that although IF2 manages to quickly find a decent coarse solution, it struggles to squeeze out those last few units of log-likelihood improvement.
-The 60 more iterations of gradient descent, however, successfully refines this solution, getting much closer to the MLE on average.
-
-\subsubsection{Results} 
-We benchmarked IFAD against IF2 on a challenging global search problem.
-100 initial starting parameter vectors were drawn uniformly from the same wide bounding box used in \cite{ionides15}, and we performed 100 searches each of IF2, IFAD-0, IFAD-1, MOP-0, and MOP-1 initialized from these 100 starting parameter vectors.
-We summarize our findings below.
-
-\begin{table}[h!]
-\centering
-\begin{tabular}{||c c c||} 
- \hline
- Method & Best Log-Likelihood & Rank \\ [0.5ex] 
- \hline\hline
-     IFAD-0 & -3747.87 & 1\\
-     IFAD-1 & -3748.53 & 2\\
-     IF2 (Ours) & -3755.91 & 3\\
-     IF2 (Warm Start) & -3759.60 & 4 \\
-     IF2 2015 & -3768.63 & 5\\ 
-     MOP-0 Alone & -3795.74 & 6\\
-     MOP-1 Alone & -3797.38 & 7\\
- \hline
-\end{tabular}
-\caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone.
-IFAD performs the best among all methods.
-Our implementation of IF2 outperforms that of \cite{ionides15}, but still ultimately underperforms IFAD.
-Both IFAD-0 and IFAD-1 manage to find the MLE, either matching or beating the highest log-likelihood previously found in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.}
-\label{table:mle}
-\end{table}
-
-\paragraph{IFAD Successfully Finds the MLE:} Both IFAD-0 and IFAD-1 successfully find the MLE, as seen in Table \ref{table:mle}.
-Previously, the MLE at a log-likelihood of $-3748.5$ was found with much effort, using many global and local IF2 searches and likelihood profiling.
-Despite being initialized for a global search, both IFAD-0 and IFAD-1 manage to find the MLE over the 100 searches.
-This may imply that the local search and refinement that was previously necessary for finding the MLE is \textbf{baked into} the two-stage IFAD algorithm.
-
-Interestingly, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.63$, while \cite{king08} only achieve a best possible estimate of $-3793.4$.
-Our results of $-3748.53$ and $-3747.87$ match or beat the highest log-likelihood previously found in the current implementation in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.
-
-This can be seen most clearly in Figure \ref{fig:boxplot-search}.
-In the left panel within Figure \ref{fig:boxplot-search}, only IFAD has whiskers that extend to the dotted red line that represents the MLE.
-
-\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations (as seen in Figure \ref{fig:optim}), performing IF2 alone ultimately fails to squeeze out the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter} and the left panel in Figure \ref{fig:boxplot-search}).
-
-Conversely, performing MOP alone leads to many, many failed searches, as seen in Figure \ref{fig:hist-all} and the right panel in Figure \ref{fig:boxplot-search}.
-This is a difficult, nonconvex, and noisy problem, and the search gets stuck in local minima and saddle points, failing to approach the MLE.
-
-IFAD, in comparison, successfully manages to (1) approach the MLE quickly due to the IF2 warm-start (as seen in Figure \ref{fig:optim}) and (2) refine the coarse solution found by the warm-start with MOP gradient steps to find the MLE (as seen in Figures \ref{fig:scatter}, \ref{fig:boxplot-search}, and \ref{fig:hist-all}).
-IFAD therefore successfully combines the best qualities of IFAD and MOP, outperforming either of them alone.
-
-\begin{figure}[htbp!]
-    \centering
-    \includegraphics[scale=0.53]{../imgs/095/pairs.png}
-    \includegraphics[scale=0.53]{../imgs/095/qq.png}
-    \caption{Scatterplots depicting the performance of IFAD against that of IF2.
-\textbf{Left:} Paired searches from the same starting point.
-\textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches.
-It is clear that on average, IFAD has the edge, and manages to find the MLE while no IF2 search successfully gets within 7 log-likelihood units of it.
-We use a dotted red line to display the MLE in both.
-}
-
-    \label{fig:scatter}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=0.48]{../imgs/095/boxplot.png}
-    \includegraphics[scale=0.53]{../imgs/095/boxplotall.png}
-    \caption{\textbf{Left:} Boxplot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs.
-This simulates the result of the common procedure of running a few searches and choosing the best one.
-IFAD outperforms all other methods, and improves on the warm-start given by running 40 IF2 iterations.
-Our choice of hyperparameters for our implementation of IF2 outperforms the choices made by \cite{ionides15}, which we depict in the boxplot labeled "IF2 2015".
-As before, while IF2 successfully approaches the MLE, it ultimately fails to find it.
-    \textbf{Right:} Boxplot depicting the performance of IFAD, IF2, and MOP alone without the warm-start, where we plot all runs.
-We see that MOP alone drastically underperforms all other methods, failing to get close to the MLE.
-This shows that the IF2 warm-start may be necessary to achieve good performance in challenging nonconvex and noisy problems like this one.
-    \textbf{Takeaway:} In this panel, we see that (1) IF2 fails to find the MLE while the gradient steps perform better at fine-grained refinement, and (2) performing gradient steps alone without a warm-start leads to the search getting stuck in local minima and saddle points, ultimately demonstrating the necessity of a hybrid algorithm like IFAD.
-    In both, as before, we use a dotted red line to display the MLE. }
-    \label{fig:boxplot-search}
-\end{figure}
-
-\begin{figure}[htbp!]
-    \centering
-    \includegraphics[scale=0.7]{../imgs/095/optim.png}
-    \caption{Optimization progress of IFAD and IF2, where the solid lines depict the median negative log-likelihood at each iteration and the shaded area depicts the best search at any iteration and the 80\% percentile.
-The dashed orange line depicts the median warm-start given by running 40 IF2 iterations.
-While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD -- IFAD has better tail control and successfully reaches the MLE.
-The initial loss in log-likelihood upon transitioning from IF2 to MOP seems to help with escaping local minima.
-    As before, we use a dotted red line to display the MLE. }
-    \label{fig:optim}
-\end{figure}
-
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.7]{../imgs/095/hist.png}
-    \caption{Histogram comparison of the performance of IFAD, IF2, and MOP alone.
-As before, we note that IFAD successfully reaches the MLE a nontrivial number of times, while IF2 ultimately fails to do so.
-Our implementation of IF2 outperforms that of \cite{ionides15}, largely due to improved hyperparameter choices.
-MOP alone performs poorly, demonstrating the necessity of warm-starting the search with IF2.
-    We once again see that IFAD combines the best of IF2 and MOP, approaching the MLE quickly due to the IF2 warm-start and successfully refining the coarse solution with gradient steps, achieving superior performance.
-As before, we use a dotted red line to display the MLE.
-}
-    \label{fig:hist-all}
-\end{figure}
-
-\paragraph{Differences Between IFAD-0 and IFAD-1:} The differences between IFAD-1 and IFAD-0 (MOP-1 and MOP-0) do not seem very pronounced in this example.
-This could be due to a number of reasons, such as (1) future case counts $y_{{n+1}:N}^*$ not providing a large amount of information on state identification $x_n$ given current and past measurements $y_{0:n}^*$ as mentioned in \cite{corenflos21}, and (2) the gradient normalization making the trajectories given by MOP-0 and MOP-1 more similar than they would be otherwise.
-
-\kevin{IFAD beats IF2 x percent of the time on 100 paired runs, but the bigger thing is that it gets close to the MLE without much effort.
-Previously, we'd have to do many searches, plus refinement (which is necessary, as we can see from the IF2 boxplot, it is easy to fool yourself into thinking that because you see the plateau that is the MLE).
-This includes likelihood profiling diagnostics, which are very computationally intensive.
-Now we can just take the best of ten searches -- IFAD does the refinement.}
-
-\kevin{Have an outsider read the manuscript afterwards.}
-
-\section{Discussion and Future Work}
-
-We develop a new theoretical framework to differentiating through the particle filter that (1) encompasses the gradient estimates of \cite{poyiadjis11}, \cite{blei2018vsmc}, and \cite{scibior2021dpf} as special cases, (2) sidesteps the issue of differentiating through a Monte-Carlo algorithm with discontinuous resampling, and (3) provides an opportunity for optimizing a bias-variance tradeoff.
-
-Prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics.
-In the presence of a differentiable simulator, MOP-$\alpha$ enjoys this very same desirable property, enabling practical maximum likelihood inference in a wider variety of situations.
-
-Furthermore, MOP-$\alpha$, and by extension IFAD, has runtime \textbf{linear, and not quadratic}, in the number of particles, unlike the marginal differentiable particle filter in \cite{scibior2021dpf} or the the optimal transport of resampling approach of \cite{corenflos21}.
-This does not necessarily come at a cost to Monte Carlo variance, unlike \cite{poyiadjis11} and \cite{corenflos2021ot} who suffer from variance quadratic in the length of the trajectory, as we can optimize $\alpha$ for a bias-variance tradeoff.
-
-In the presence of strong convexity, first and second-order gradient methods that use MOP-$\alpha$ as an estimate enjoy a linear convergence rate, providing support for numerical optimization with AD for the particle filter.
-
-This leads to a practical hybrid algorithm we call IFAD that warm-starts gradient descent, or some other first or second-order method using the gradient estimate given by MOP-$\alpha$, with IF2.
-When benchmarked on a challenging global search problem with the model of \cite{king08}, IFAD enjoys the best properties of IF2 and MOP-$\alpha$, successfully approaching a neighborhood of the MLE quickly without getting stuck in saddle points and (poor) local minima (like gradient descent alone would), while effectively optimizing the last few units of log-likelihood (which IF2 struggles with).
-
-Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual} that we call \texttt{pomp-py}.
-The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{JAX} Python package from \citet{jax2018github}.
-
-\subsection{Future Work}
-
-While MOP-$\alpha$, and IFAD in its current incarnation, remain very general, they currently lack the capability of handling discrete latent states.
-We conjecture that DOP-$\alpha$ might be a possible solution (despite needing the likelihood ratio of transition densities), and will explore this in future work.
-
-We conjecture that the linear convergence result in Theorem \ref{thm:convergence} applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE.
-IF2 posesses some SGD-like qualities, including what looks like linear convergence to a ball around the MLE with width determined by the random walk standard deviation.
-Not much is known about the convergence rate of IF2, and an answer to this would also provide an answer to the convergence rate of IFAD beyond just the convergence rate of the gradient stage.
-We will explore this conjecture in future work.
-
-In the interest of simplicity, we only explore the performance of gradient descent on the Dhaka cholera model of \cite{king08}.
-Exploring the use of various first and second order methods such as Newton's method, ADAM, L-BFGS, or other popular iterative optimization methods with the gradient estimates that MOP-$\alpha$ provides in the context of the particle filter could very well lead to nontrivial speedups and increased performance.
-
-Extensions of MOP-$\alpha$ and IFAD to panel and spatiotemporal datam, possibly through a block particle filter in the same vein as \cite{ionides22} and \cite{ning23}, would go a long way towards extending the current applicability of these methods to handle a wider variety of interesting and challenging scientific problems.
-
-While we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual}, this remains a work in progress.
-
-\paragraph{Acknowledgements.} We thank Nicholas Chopin for helpful comments on the proof of a strong law of large numbers for a suitably reweighted particle filter.
-
-\kevin{Reach: PNAS.
-Else try for JRSSB, try to get Chopin for Annals AE.
-Would need to change bracketing conventions for Biometrika.
-If submitting to PNAS, JASA, Biometrika, or JRSSB, highlight the different statistical gradient estimator and finite sample performance, rewrite introduction, maybe create a way to use the gradients for inference rather than using profiling.
-Try to rewrite it as a breakthrough in statistics.
-Safety: JMLR}
-
-\bibliography{paper/bib-ifad, paper/bib-ref}
-\bibliographystyle{apalike}
-
-\appendix
-\renewcommand{\thefigure}{A\arabic{figure}}
-\setcounter{figure}{0}
-
-\section{Weighted Particle Filters that Target the Posterior}
-
-\begin{defn}[Targeting]
-    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any measurable and bounded functional $h$,
-\begin{equation}
-    E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
-\end{equation}  
-    A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if for any measurable and bounded functional $h$,
-\begin{equation}
-    \frac{\sum_{j=1}^J h(X^J_j) w^J_j}{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X))
-\end{equation}
-as $J \to \infty$.
-\end{defn}
-\citet{chopin2004clt} asserted without proof that common particle filter algorithms target the filtering distribution, $f_{X_{1:N}|Y_{1:N}}$, in this sense.
-\ed{PERHAPS IF WE HAD A CONCRETE PROOF OF THIS, THAT WOULD HELP TO SEE HOW IT EXTENDS TO OUR SUBSEQUENT LEMMAS.
-ARGUABLY, THAT IS IN FACT WHAT WE ARE DEVELOPING VIA THESE LEMMAS.
-I WONDER IF SOME OTHER REFERENCE, E.G.
-DELMORAL'S 2004 BOOK, HAS AN EXPLICIT PROOF OF CONSISTENCY FOR SOME PARTICLE FILTER ALGORITHM.}
-In order to prove the consistency of our variation on the particle filter, we now present three helper lemmas.
-The first follows from standard importance sampling arguments, the second from integrating out the marginal, and the third from Bayes' theorem. 
-We state Lemma~\ref{lem:change-measure-proper-weights} assuming multinomial resampling, which is convenient for the proof though other resampling strategies may be preferable in practice REFS.
-
-\begin{lem}[Change of Particle Measure]
-    \label{lem:change-measure-proper-weights}
-    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
-Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, $\pi^J_j J$ times.
-Write
-    \[
-    (Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big),
-    \]
-    where $a$ is called the ancestor function. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j),j=1,\dots,J\}$ targets $f_X$.
-\end{lem}
-
-\begin{proof}
-
-    Note that as the $Y_j^J$ are a subsample from $X_j^J$, $h$ can be a function of $Y$ as well as it is one for $X$. We then expand
-    $$\frac{\sum_j h(Y_j^J) V_j^J}{\sum_j V_j^J} = \frac{\sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}}.$$
-    By hypothesis,
-    $$\frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[f(X)].$$
-
-    We want to show
-    \begin{equation}\label{eq:lemma1:h}
-    \frac{\sum_j h(X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}} - \frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} 0.
-    \end{equation}
-    For this, it is sufficient to show that
-   $$ \sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}
-    -  \sum_j h(\tilde X_j^J)\frac{U_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0 $$
-    since an application of this result with $h(x)=1$ provides almost sure convergence of the denominator in (\ref{eq:lemma1:h}).
-    Write $g(\tilde X_j^J) = h(\tilde X_j^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}$. We therefore need to show that 
-    $$\sum_j Z_j^J := \sum_j \left(g(\tilde X_{k_j}^J) -  g(\tilde X_j^J) \pi_j^J \right) \stackrel{a.s.}{\to} 0.$$
-
-    Because the functional $h$ and importance sampling weights $u_{k_j}^J/\pi_{k_j}^J$ are bounded, we have that $\E[\left(Z_j^J\right)^4] < \infty$.
-We can then follow the argument of \cite{chopin20} from this point on, where
-    {THE Zs ARE ONLY CONDITIONALLY INDEPENDENT, SO WE COULD PUT IN THE CONDITIONING EVENT (I.E., THE Xs AND Us) HERE TO BE MORE ACCURATE}
-    $$\E\left[\left(\sum_j Z_j^J\right)^4\right] 
-    = J\E[(Z_1^J)^4] + 3J(J-1)\left(\E[(Z_j^J)^2]\right) \leq CJ^2,$$ for some $C>0$. By Markov, 
-    $$\mathbb{P}\left(\left|\frac{1}{J} \sum_{j=1}^J Z_j^J\right|>\epsilon\right) 
-    \leq \frac{1}{\epsilon^4J^4 } 
-    \mathbb{E}\left[\left(\sum_{j=1}^J Z_j^J\right)^4\right] \leq \frac{C}{\epsilon^4J^2},$$
-    and as these terms are summable we can apply Borel-Cantelli to conclude that these deviations happen only finitely often for every $\epsilon>0,$ giving us the almost-sure convergence for
-
-    $$ \sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}
-    -  \sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0.$$
-
-    Similarly, we also have that
-    $$ \sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J
-    - \sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J} \stackrel{a.s.}{\to} 0,$$
-
-    and the result is proved.
-
-    \kevin{MAY HAVE TO DEAL WITH CONDITIONING ON THE FILTRATION AT THE LAST TIMESTEP, AND THEN USING THE TOWER PROPERTY. EXPECTATION WILL BE THE SAME REGARDLESS FOR THE FOURTH MOMENT.}
-
-   % Let $h$ be an integrable function. 
-    %\begin{align*}
-    %    \E\left[\sum_{j=1}^J h(Y_j) v_j\right] 
-    %    &= \sum_{j=1}^J \E\left[h(X_{a(j)}) \frac{u_{a(j)}}{\pi_{a(j)}}\right] \\
-    %    &= \sum_{j=1}^J \E\left[h(X_{j}) u_{j}\right],
-    %\end{align*}
-    %and similarly for the denominator.
-Now, the numerator and denominator of the reweighted quantity have the same expectation as the numerator and denominator of the original quantity, so they must converge to $ E_\pi(h(X))$ almost surely as well.
-\end{proof}
-
-\textbf{Remark:} Note that Lemma \ref{lem:change-measure-proper-weights} permits $\pi_{1:J}$ to depend on $\{(X_j,u_j)\}$ as long as the resampling is carried out independently of $\{(X_j,u_j)\}$, conditional on $\pi_{1:J}$.
-
-\begin{lem}[Particle Marginals]
-    \label{lem:marginal-proper-weights}
-    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
-Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
-Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}$ targets $f_Z$.
-\end{lem}
-\begin{proof}
-    We want to show that, for any measurable bounded $h$, 
-     $$\frac{\sum_j h(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[h(Z)] = \E\big[\E[h(Z) | X]\big].$$
-    By hypothesis, for any measurable and bounded functional $g$ with domain $\gX$,
-    \begin{equation}\label{eq:lemma2:g}
-    \frac{\sum_j g(\tilde X_j^J)\, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[g(X)].
-    \end{equation}
-    Let $\bar{U}_j^J = \frac{J U_j^J}{\sum_j U_j^J}$.
-Examine the numerator and denominator of the quantity $$\frac{J^{-1}\sum_j h(\tilde Z_j^J) \, U_j^J}{J^{-1}\sum_j U_j^J} = \frac{J^{-1}\sum_j h(\tilde Z_j^J) \, \bar{U}_j^J}{J^{-1}\sum_j \bar{U}_j^J}.$$
-    The denominator converges to $1$ almost surely. The numerator, on the other hand, is
-    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J,$$
-    by the same fourth moment argument to the above lemma converges almost surely to the limit of its expectation
-    \begin{eqnarray*}        
-    \lim_{J\to\infty} \E\left[\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J\right] 
-    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  
-      \E\left[ h(\tilde Z_j^J)\, \bar{U}_j^J\big|\tilde X_j^J, \bar U_j^J\right]
-    \right]
-    \\
-    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  \E\big[h(Z) \big| X=\tilde X_j^J \big] \, \bar U_j^J\right].
-    \end{eqnarray*}
-    Applying (\ref{eq:lemma2:g}) with $g(x) = \E\big[h(Z) | X=x\big]$, the average on the right hand size converges almost surely to $\E\big\{\E[h(Z)|X]\big\}=\E[h(Z)].$
-    It remains to swap the limit and expectations. We can do so with the bounded convergence theorem, and therefore obtain   
-    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\,\bar{U}_j^J \stackrel{a.s.}{\to} \E[h(Z)].$$
-
-\end{proof}
-
-\begin{lem}[Particle Posteriors]
-    \label{lem:posterior-proper-weights}
-    Suppose that $\{(X_j^J,u_j^J),j=1,\dots,J\}$ targets $f_X$.
-Also suppose that $(X^{\prime J}_j,u^{\prime J}_j) = \big(X_j^J,u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$.
-Then, if $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,u^{\prime J}_j)\}$ targets $f_{X|Z}(\cdot | z^*)$.
-\end{lem}
-
-\begin{proof}
-
-    $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J}}{J^{-1} \sum_j \bar{u}_j^{' J}}$$
-
-    by
-
-    \begin{equation}
-    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}}
-    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \right)^{-1}
-    \end{equation}
-
-    Again, we want to show that
-     $$\frac{\sum_j h(X_j^J) \cdot u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)} \stackrel{a.s.}{\to} \E_{f_{X|Z}}[h(X)|z^*].$$
-
-    We already have that
-    $$\frac{\sum_j h(X_j^J)u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} \E_{f_X}[h(X)].$$
-
-    %= \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j f_{Z|X}(z^*|X_j^J)f_X(X_j^J)}
-
-    Consider the following:
-    \begin{equation}
-    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}}
-    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \right)^{-1}.
-    \end{equation}
-
-    Write $\bar{u}_j^{J} = \frac{J u_j^J}{\sum_j u_j^J}$.
-Define the weights $u^{' J}_j := u_j^J \cdot f_{Z|X}(z^*|X_j^J),$ and the corresponding self-normalized weights $\bar{u}_j^{' J} := \frac{J u^{' J}_j}{\sum_j u^{' J}_j}.$ The denominator of the below quantity
-    $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J}}{J^{-1} \sum_j \bar{u}_j^{' J}}$$
-    converges to $1$, while the numerator, by the same fourth moment argument as the first lemma, converges almost surely to the limit of its expectation, which is
-    \begin{align*}
-        \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J} \right]
-        &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)}\right] \\
-        &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
-        &= \lim_{J\to\infty} \E\left[ \sum_j h(X_j^J) \frac{\frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
-        &= \E_{X|Z}[h(X)|z^*],
-    \end{align*}
-
-    where we can swap the limit and expectations with the bounded convergence theorem.
-\end{proof}
-
-\begin{prop}[MOP-1 Targets the Posterior]
-    When $\alpha=1$ or $\phi=\theta$, MOP-$\alpha$ targets the posterior. 
-\end{prop}
-\begin{proof}
-    When $\theta=\phi$, regardless of the value of $\alpha$, the ratio $\frac{g_{n,j}^\theta}{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter estimate.
-
-    When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows.
-Recursively applying Lemmas \ref{lem:change-measure-proper-weights}, \ref{lem:marginal-proper-weights}, and \ref{lem:posterior-proper-weights}, we obtain that
-    %to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
-    the MOP-1 filter targets the posterior.
-    Specifically, suppose inductively that $\big\{\big(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
-    Then, Lemma \ref{lem:marginal-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big)\big\}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
-    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-    Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
-    \[
-    \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}\big/ g^\phi_{n,a(j)}\big),
-    \]
-    with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-\end{proof}
-
-This has addressed filtering, but not quite yet the likelihood evaluation. For this we use the following lemma.
-
-\begin{lem}[Likelihood Proper Weighting]
-    \label{lem:lik-proper-weight}
-  $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1_n-1}^*;\theta)$ is consistently estimated by either the before-resampling estimate,
-\begin{equation}\label{L1}
-L_n^{B,\theta} =  \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
-\end{equation}
-or by the after-resampling estimate,
-\begin{equation}\label{L2}
-L_n^{A,\theta} = L^\phi_n \frac{\sum_{j=1}^Jw^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}.
-\end{equation}
-where $L^\phi_n$ is as defined in the various algorithms.
-\end{lem}
-
-Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
-To see  (\ref{L2}),
-we write the numerator of (\ref{lem:change-measure-proper-weights}) as
-\[
-L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
-= L^\phi_n \sum_{j=1}^J w_{n,j}^{FC,\theta} \frac{g^\phi_{n,j}}{L_n^\phi}
-\]
-Using Lemma \ref{lem:change-measure-proper-weights}, we resample according to probabilities $\frac{g^\phi_{n,j}}{L_n^\phi}$ to see this is properly estimated by
-\[
-L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},
-\]
-from which we obtain (\ref{L2}).
-
-Using Lemma \ref{lem:lik-proper-weight}, we obtain a likelihood estimate,
-\[
-L^{A,\theta} = \prod_{n=1}^N \left( L^\phi_n \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J w^{P,\theta}_{n,j}}\right).
-\]
-Since $w^{F,\theta}_{n,j}=w^{P,\theta}_{n+1,j}$, this is a telescoping product. The remaining terms are
-$\sum_{j=1}^J w^{P,\theta}_{0,j} = J$ on the denominator and $\sum_{j=1}^J w^{F,\theta}_{N,j}$ on the numerator.
-This derives the MOP/DOP estimate.
-
-\kevin{IS THIS UNBIASED?}
-
-$L^{B,\theta}$ should generally be preferred in practice, since there is no reason to include the extra variability from resampling when calculating the conditional log likelihood, but it lacks the nice telescoping product that lets us derive exact expressions for the gradient later.
-
-\section{Doubly Off-Policy Particle Filters}
-\label{app:dop}
-
-The proof for Proposition \ref{prop:dop-correctness} follows from the below two lemmas.
-Evaluating the off-policy likelihood is possible with a weighted particle filter by using resampling weight $p(y^*_n|x_n;\phi)$ and making two corrections in the particle weight, through multiplying $w_{t,j}$ by the two likelihood ratios below,
-
-\begin{equation}
-    \label{eq:dmeas-ratio}
-    s=p(y^*_n|x_n;\theta)/p(y^*_n|x_n;\phi)
-\end{equation}
-and
-\begin{equation}
-    \label{eq:rproc-ratio}
-    r=p(x_n|x_{n-1};\theta)/p(x_n|x_{n-1};\phi).
-\end{equation}
-
-The following result showcases the correctness of corrections \ref{eq:dmeas-ratio} and \ref{eq:rproc-ratio}.
-
-\begin{lem}
-    For $\theta$, $\phi$ that satisfy Proposition \ref{prop:smooth-nbhd}, the following correction in the particle filter weights,
-    \begin{equation}
-        w_{t,j} := \tilde{w}_{t,j} \cdot p(y^*|x_{t,j}^P,\phi)\cdot r \cdot s
-    \end{equation}
-    successfully recovers $\lik(\theta)$.
-\end{lem}
-
-\begin{proof}
-    Recall $\lik(\theta) = \E[\lik(\theta,\phi,\Omega, J)]=\int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega)$ and $\lik(\theta) = \int_\Omega \lik(\theta, \omega, J) d\mu(\omega)$.
-We want to evaluate
-    \begin{align*}
-        \lik(\theta,\omega, J)
-        &= \E\left[ \prod_{t=1}^T \lik_t(\theta) \middle| \omega\right]
-        = \E\left[ \prod_{t=1}^T p(y_t^*|y_{1:t-1}^*,\theta) \middle| \omega\right] \\
-        &= \E\left[ \prod_{t=1}^T \frac{1}{J} \sum_{j=1}^J  p(y_t^*|x_{t,j}^P,\theta) p(x_{t,j}^P|y_{1:t-1}^*,\theta) \middle| \omega\right] \\
-        &= \E\left[ \prod_{t=1}^T\frac{1}{J}  \sum_{j=1}^J  p(y_t^*|x_{t,j}^P,\theta) p(x_{t,j}^P|x_{t-1}^F,\theta) \middle| \omega\right] 
-    \end{align*}
-    where $p(x_t^P|y_{1:t-1}^*,\theta)$ is the prediction distribution of particles, and the filtering distribution of particles is proportional to $p(y_t^*|x_{t,j}^P,\theta)$, the likelihood of particle $j$ at time $t$.
-
-    Consider first the case with no resampling.
-Let $\lik^j(\theta)$, $\lik^j(\phi)$ be the likelihood of trajectory $j$ under $\theta$ and $\phi$ respectively.
-As the prediction and filtering particles are then the same, we use $x_{t,j}$ to refer to both.
-As such, conditional on $\omega$,
-    \begin{align*}
-        \lik^j(\theta) 
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \theta) p(x_{t,j}|x_{t-1,j}, \theta) \\
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \frac{p(y_t^*|x_{t,j}, \theta)}{p(y_t^*|x_{t,j}, \phi)} p(x_{t,j}|x_{t-1,j}, \theta) \frac{p(x_{t,j}|x_{t-1,j}, \theta) }{p(x_{t,j}|x_{t-1,j}, \phi) } \\
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \cdot s \cdot p(x_{t,j}|x_{t-1,j}, \phi)\cdot r
-    \end{align*}
-
-    and therefore, if trajectories were proposed with equal probability with particles conditional on $\omega$,
-    $$ \lik(\theta, \omega, J) = \frac{1}{J} \sum_{j=1}^J \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \cdot s \cdot p(x_{t,j}|x_{t-1,j}, \phi)\cdot r $$
-    The correction is therefore sufficient for recovering $\lik(\theta)$, with trajectories collected with the process model at $\phi$ and measurements evaluated with the measurement model at $\phi$.
-
-    Consider now the case with resampling.
-Without loss of generality we can resample every timestep.
-Here, $x_{t,j}^F(\phi) \sim p(y^*|x_{t,j}^P, \phi)$ and $x_{t,j}^P(\phi) \sim p(x_t|x_{t-1}^F,\phi)$, and similarly for $\theta$.
-We use this notation to emphasize that \textbf{both} the measurement model and the previously drawn particles depend on $\phi$ or $\theta$.
-
-    \begin{align*}
-        \lik_t(\theta) &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \theta)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\theta) \right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \theta)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \frac{p(y_t^*|x_{t,j}^P,\theta)}{p(y_t^*|x_{t,j}^P,\phi)} \right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \frac{p(y_t^*|x_{t,j}^P,\theta)}{p(y_t^*|x_{t,j}^P,\phi)} \frac{p(x_{t,j}|x_{t-1,j}, \theta) }{p(x_{t,j}|x_{t-1,j}, \phi) }\right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \cdot s \cdot r \right] \\
-        &= \E_{\omega}\E_{x_{t,j}^P \sim \process(x_{t-1}^F, \omega, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \cdot s \cdot r \middle| \omega \right]
-    \end{align*}
-
-    and again the correction is sufficient.
-
-    Therefore,
-
-    \begin{align}
-         \mathcal{L}(\theta) 
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^P(\omega), \theta) p(x_{n,j}^P(\omega) | x_{n-1,j}^P(\omega), \theta) \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\theta}(\omega), \theta) p(x_{n,j}^{P,\theta}(\omega) | x_{n,j}^P(\omega), \theta) \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\theta}(\omega), \phi) p(x_{n,j}^{P,\theta}(\omega) | y_n^*, \phi) \cdot s  \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) p(x_{n,j}^{P,\phi}(\omega) | y_n^*, \phi) \cdot r \cdot s \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) p(x_{n,j}^{P, \phi}(\omega) | y_n^*, \phi) \cdot r \cdot s \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) \cdot r \cdot s \right)  p(x_{1:N} | y_{1:N}^*, \phi) \,  d\mu(\omega) \\
-         &=: \int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega).
-     \end{align}
-
-     The proof for an arbitrary functional under the particle measure follows analogously.
-\end{proof}
-
-\begin{lem}[Gradient of Doubly Off-Policy Particle Filter]
-    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$.
-Also let $\theta \in \gN(\phi)$.
-When $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{poyiadjis11},
-    \begin{equation}
-        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
-    \end{equation}
-\end{lem}
-
-\begin{proof}
-    It is known, from \cite{doucet2009tutorial}, that 
-    $$p_\theta(x_{1:N}|y_{1:N}) = \prod_{n=1}^N \frac{p(x_n|x_{n-1})p(y_n|x_n)}{p(y_n|y_{1:n-1})} = \prod_{n=1}^N \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{\lik_n(\theta)}.$$
-
-    From the above lemma, and the fact that the particle filter is unbiased for the likelihood, we know that
-    \begin{align*}
-        \lik(\theta)
-        &= \int_{x_{1:N}} \prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) p_\theta(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N  p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \right) \frac{p_\theta(x_{1:N}|y_{1:N})}{p_\phi(x_{1:N}|y_{1:N})}p_\phi(x_{1:N}|y_{1:N}) dx_{1:N}\\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N \lik_n(\theta) \right) \left(\prod_{n=1}^N \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \lik_n(\phi)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) \lik_n(\theta)}\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N \lik_n(\phi) \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) }\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N}.
-    \end{align*}
-
-    Therefore, under regularity conditions that permit differentiating under the integral sign, using the log-derivative trick, we have
-    \begin{align*}
-        \nabla_\theta \lik(\theta)
-        &= \int_{x_{1:N}} \nabla_\theta \hat\lik(\theta) \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N}\\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N \lik_n(\phi) \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) }\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N \left(\frac{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n)}{\lik_n(\phi)}\right)^{-1} p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\frac{1}{p_\phi(x_{1:N}|y_{1:N}) }\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\nabla_\theta \log\left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} p_\theta(x_{1:N},y_{1:N}) \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) p_\theta(x_{1:N}|y_{1:N}) \cdot \lik(\theta) \; dx_{1:N}. 
-    \end{align*}
-
-    Using the log-derivative trick again,
-    \begin{align*}
-        \nabla_\theta \ell(\theta)
-        &= \frac{\nabla_\theta \lik(\theta)}{\lik(\theta)} \\
-        &= \int_{x_{1:N}} \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) p_\theta(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \E_{x_{1:N}|y_{1:N}, \theta}\left[\nabla_\theta \left(\sum_{n=1}^N\log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right)\right] \\
-        &= \E_{x_{1:N}|y_{1:N}, \theta}\left[\nabla_\theta\log p_\theta(x_{1:N},y_{1:N})\right] \\
-        &= \lim_{J\to\infty} \E_{x_{1:N}|y_{1:N}, \theta}\left[\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log p_\theta(x_{1:N,j}^{A,F},y_{1:N}^*)\right].
-    \end{align*}
-
-    But it also holds that
-    \begin{align*}
-        \nabla_\theta \ell(\theta)
-        &= \frac{\nabla_\theta \lik(\theta)}{\lik(\theta)} \\
-        &= \int_{x_{1:N}} \frac{\nabla_\theta \hat\lik(\theta)}{\lik(\theta)} \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \hat\ell(\theta) \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N}\\
-        &= \lim_{J\to\infty} \E_{x_{1:N}|y_{1:N}, \phi}\left[\nabla_\theta \hat\ell(\theta)\right].
-    \end{align*}
-
-    and we see that $$\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log p_\theta(x_{1:N,j}^{A,F},y_{1:N}^*)$$ and $$\nabla_\theta \hat\ell(\theta)$$ both converge to the same thing in expectation.
-They are both particle approximations under the posterior where the variance of each is bounded, therefore one can apply the particle CLT from \cite{chopin2004clt} to obtain that they are both consistent estimates of the score.
-
-    In fact, when $\theta=\phi,$ they are the same in distribution. \kevin{claim}
-
-    \kevin{This is a far weaker result than what I really wanted. This just gives consistency. Want equality in distribution.}
-\end{proof}
-
-\begin{lem}
-
-Consider the case of DOP-$\alpha$ when $\alpha=0$ and $\theta=\phi$. The gradient estimate is then
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta\right),
-    \end{equation}
-
-    yielding the gradient estimator of \cite{blei2018vsmc} that \cite{scibior2021dpf} show is the gradient of a vanilla particle filter when the resampling terms are dropped.
-
-\end{lem}
-
-\begin{proof}
-    When $\alpha=0,$ the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J r_{n,j}s_{n,j}}{\sum_{j=1}^J r_{n,j}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J r_{n,j}s_{n,j}.
-\end{equation}
-
-Its gradient when $\theta=\phi$ is therefore 
-\begin{align*}
-    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right) \\
-    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right)} \\
-    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta r_{n,j}s_{n,j}}{\sum_{j=1}^J r_{n,j}s_{n,j}} \\
-    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{Y_n, X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)} \\
-    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta\right),
-\end{align*}
-where we use the log-derivative trick, observe that $\sum_{j=1}^J r_{n,j}s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again. 
-\end{proof}
-
-\subsection{Off-Policy Weighted Differentiable Particle Filter}
-
-We are finally ready to introduce the off-policy weighted differentiable particle filter.
-In the event where a seed is fixed, all expressions in Algorithm \ref{alg:dop} can be considered to be conditional on $\omega \in \Omega$.
-
-\begin{algorithm}[ht]
-\centering
-	\caption{Doubly Off-Policy-$\alpha$}
-    \label{alg:dop}
-	\begin{algorithmic}[1]
-	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
-		\STATE Initialize filter particles ${X}_{0,j}^{F,\phi}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\phi}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$. Fix $\omega.$
-		\FOR{$n=1,...,N$}
-            \STATE Prediction weights with discounting: $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
-            \label{dop-alpha:discount}
-            \STATE Simulate for prediction:
-            ${X}_{n,j}^{P,\phi}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\phi}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{dop-alpha:step1}
-            \STATE Adjust weights: $\displaystyle w_{n,j}^{P,\theta} = w_{n,j}^{P,\theta} \times
-            \frac{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\phi}|{X}_{n-1,j}^{F};{\theta}\big)}{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\phi}|{X}_{n-1,j}^{F};{\phi}\big)}$ for $j$ in $1:J$
-            \label{dop-alpha:dproc}
-            \STATE Evaluate measurement density:
-            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\phi}\giventh{\theta})$ for $j$ in $\seq{1}{J}$
-            \STATE Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \STATE Conditional likelihood under $\phi$: 
-            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
-            \label{dop-alpha:Lphi}
-            \STATE Normalize weights:
-            $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
-            for $j\ \mathrm{in}\ \seq{1}{J}$
-            \STATE Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \label{dop-alpha:systematic}
-            \STATE Resample particles: ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$
-            \STATE Filter weights corrected for resampling:
-            $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{dop-alpha:weight:update}
-            \STATE Resample filter weights:
-            $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
-            for $j$ in $\seq{1}{J}$ \label{dop-alpha:step2}
-            \STATE After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \ENDFOR
-		\RETURN likelihood estimate $\hat{\lik}(\theta) = \lik(\theta, \phi, \omega, J) := \prod_{n=1}^N L_n^{A,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}.$
-	\end{algorithmic}
-\end{algorithm}
-
-Here, we consider a doubly off policy particle filter, meaning that both \code{rprocess} and \code{dmeasure} are computed at $\phi$ and particles are reweighted to correspond to $\theta$.
-
-\section{Measurement Off-Policy Particle Filters}
-
-\begin{algorithm}[ht]
-\centering
-	\caption{Measurement Off-Policy-$\alpha$}
-    \label{alg:mop}
-	\begin{algorithmic}[1]
-	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
-        \IF{$\theta \neq \phi$}
-            \STATE Run a particle filter once with parameters $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$, with seed $\omega$.
-        \ELSE
-        \STATE Set $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$ to be copies of $X_{n,j}^{P,\theta}, X_{n,j}^{P,\theta}$ in the rest of the algorithm.
-        \ENDIF
-		\STATE Initialize filter particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$. Fix $\omega.$
-		\FOR{$n=1,...,N$}
-            \STATE Prediction weights with discounting: $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
-            \label{mop-alpha:discount}
-            \STATE Simulate for prediction:
-            ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{mop-alpha:step1}
-            \STATE Evaluate measurement density:
-            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for $j$ in $\seq{1}{J}$
-            \STATE Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \STATE Conditional likelihood under $\phi$: 
-            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
-            \label{mop-alpha:Lphi}
-            \STATE Normalize weights:
-            $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
-            for $j\ \mathrm{in}\ \seq{1}{J}$
-            \STATE Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \label{mop-alpha:systematic}
-            \STATE Resample particles: ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$
-            \STATE Filter weights corrected for resampling:
-            $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{mop-alpha:weight:update}
-            \STATE Resample filter weights:
-            $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
-            for $j$ in $\seq{1}{J}$ \label{mop-alpha:step2}
-            \STATE After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \ENDFOR
-		\RETURN likelihood estimate $\hat{\lik}(\theta) = \lik(\theta, \phi, \omega, J) := \prod_{n=1}^N L_n^{A,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}.$
-	\end{algorithmic}
-\end{algorithm}
-
-\begin{lem}
-    \label{lem:mop-1-formula}
-    Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The gradient estimate is then
-
-    \begin{equation}
-        \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
-    \end{equation}
-
-    yielding the gradient estimators of \cite{poyiadjis11, scibior2021dpf} when applied to the bootstrap filter. 
-\end{lem}
-
-\begin{proof}
-
-Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The likelihood estimate becomes
-\begin{equation*}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}}= \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\theta}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
-\end{equation*}
-as we have a telescoping product. Then, as
-
-$$\nabla_\theta \hat\ell(\theta) = \frac{\nabla_\theta \hat\lik(\theta)}{\hat\lik(\theta)} = \frac{\nabla_\theta\left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi}{\prod_{n=1}^N L_n^\phi} =  \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},$$
-
-The derivative of the log-likelihood estimate is then
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},
-\end{equation*}
-which we decompose as follows.
-
-Observe that as $\alpha=1$,
-
-$$w_{n,j}^{P,\theta} = w_{n-1,j}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}},$$
-where we use the $(\cdot)^A$ superscript to denote the ancestral trajectory of the $j$-th prediction or filtering particle at timestep $n$.
-
-Using the log-derivative identity again,
-$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,j}^{A,P,\theta} - \log g_{i,j}^{A,P,\phi}\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,j}^{A,P,\theta}.$$
-
-So 
-$$  \nabla_\theta \sum_{n=1}^N \log g_{n,j}^{A,\theta} = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,j}^{A,P,\theta}\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,j}^{A, P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right),$$
-
-and 
-$$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{n=1}^N \nabla_\theta \log g_{n,j}^{A,P,\theta} = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$
-
-Substituting, we have that
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right),
-\end{equation*}
-and finally, observing that when $\theta=\phi$ we have that $w_{N,j}^{F,\theta}=1$, that 
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
-\end{equation*}
-yielding the gradient estimators of \cite{poyiadjis11, scibior2021dpf} when applied to the bootstrap filter. 
-\end{proof}
-
-Note that one would believe that the variance blows up the moment $\theta\neq\phi$. This is true, as 
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right),
-\end{equation*}
-and when $\theta\neq\phi$, $w_{N,k_j}^{P,\theta} = O(c^N)$.
-When $\theta=\phi$, this is a special case of the \cite{poyiadjis11} estimator, which has $O(N^2)$ variance by a property of functionals of the particle filter by \cite{delMoral03}.
-
-\begin{lem}
-\label{lem:mop-0-formula}
-
-Consider the case of MOP-$\alpha$ when $\alpha=0$ and $\theta=\phi$. The gradient estimate is then
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
-    \end{equation}
-
-    yielding the gradient estimator of \cite{blei2018vsmc} that \cite{scibior2021dpf} show is the gradient of a vanilla particle filter when the resampling terms are dropped -- when applied to the bootstrap filter.
-
-\end{lem}
-
-\begin{proof}
-    When $\alpha=0,$ the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
-\end{equation}
-
-Its gradient when $\theta=\phi$ is therefore 
-\begin{align*}
-    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
-    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)} \\
-    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta s_{n,j}}{\sum_{j=1}^J s_{n,j}} \\
-    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)}{f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \phi}; \phi)} \\
-    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
-\end{align*}
-where we use the log-derivative trick, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again. 
-\end{proof}
-
-\kevin{I have no clue how to prove this. Help!}
-\begin{conj}[MOP-1 Estimate Is Unbiased for the Likelihood and the Gradient of the Likelihood]
-    Consider the case of MOP-$\alpha$ when $\alpha=1$, and resampling is carried out with either multinomial, residual, or systematic resampling at every timestep.
-Assume the gradient estimate obtained by the gradient of the likelihood estimate with J particles, $\nabla_\theta \hat\lik_J(\theta)$ is uniformly bounded for all $J$, and that the likelihood has two continuous derivatives.
-Then the likelihood estimate is unbiased for the likelihood and the gradient of the likelihood estimate is unbiased for the gradient of the likelihood.
-That is, it holds that
-    $$\E[\hat\lik_J(\theta)] = \lik(\theta),\;\; \E[\nabla_\theta \hat\lik_J(\theta)] = \nabla_\theta \lik(\theta).$$
-\end{conj}
-
-\begin{proof}
-    First, we need to establish the unbiasedness of the likelihood estimate. This is 
-    $$\hat\lik_J(\theta) := \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right)\prod_{n=1}^N L_n^\phi = \left(\frac{1}{J}\sum_{j=1}^J \prod_{n=1}^N \frac{g_{n,j}^{A,\theta}}{g_{n,j}^{A,\phi}} \right)\prod_{n=1}^N L_n^\phi.$$
-
-    We know from \cite{chopin20} that $\E[L_n^{\phi}] = \lik(\phi).$ It then suffices to show that
-    $$\E\left[\left(\frac{1}{J}\sum_{j=1}^J \prod_{n=1}^N \frac{g_{n,j}^{A,\theta}}{g_{n,j}^{A,\phi}} \right)\right] = \frac{\lik(\theta)}{\lik(\phi)}.$$
-
-    To obtain this, we need to see that the resampling under $\phi$, with corrected weights, is unbiased. That is, we need to show that for any bounded measurable functional $\varphi : \gX \to \R$,
-    $$\E\left[\frac{\sum_{j=1}^J w_{n,j}^{F, \theta} \varphi(x_{n,j}^{\theta, F})}{\sum_{j=1}^J w_{n,j}^{F, \theta}} \bigg| \gF_n \right] = \frac{\sum_{j=1}^J w_{n,j}^{P, \theta} \varphi(x_{n,j}^{P, \theta})}{\sum_{j=1}^J w_{n,j}^{P, \theta} }.
-$$
-
-    To see this, 
-    \begin{align*}
-        \sum_{j=1}^J w_{n,j}^{P, \theta} \varphi(x_{n,j}^{P, \theta}) 
-        &= \sum_{j=1}^J w_{n-1,j}^{F, \theta}\frac{g_{n,j}^{\theta}}{g_{n,j}^{\phi}}  \varphi(x_{n,j}^{P, \theta})
-    \end{align*}
-
-%     Observe that our differentiable simulator $\process(\cdot, \theta)$ and choice of weights now satisfy:
-
-%     - A Feynman-Kac model $\left\{M_t, G_t\right\}$ such that:
-
-% - the weight function $G_t$ may be evaluated pointwise (for all $t$ );
-
-% - it is possible to simulate from $\mathbb{M}_0\left(\mathrm{~d} x_0\right)$ and from $M_t\left(x_{t-1}, \mathrm{~d} x_t\right)$ (for any $x_{t-1}$ and $t$ ).
-
-% - The number of particles $N$.
-
-% - The choice of an unbiased resampling scheme (e.g.
-multinomial, residual, stratified, systematic), that is, an algorithm to generate variables $A_t^{1: N}$ in $1: N$ such that $\mathbb{E}\left[\sum_{m=1}^N \mathbbm{1}\left\{A_t^m=n\right\}\right]=N W_{t-1}^n$; see previous chapter for more details.
-\end{proof}
-
-\begin{lem}[Consistency of MOP-1 Gradient Estimate]
-    Consider the case of MOP-$\alpha$ when $\alpha=1$.
-Assume the gradient estimate obtained by the gradient of the log-likelihood estimate with J particles, $\nabla_\theta \hat\ell_J(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}$ is uniformly bounded by $G(\theta)$ for all $J$, and that the likelihood has two continuous derivatives.
-Then the gradient estimate, $\nabla_\theta \hat\ell_J(\theta)$, is consistent for the score.
-    \label{lem:mop-grad-consistency}
-\end{lem}
-\begin{proof}
-
-    % We know that the variance of the MOP-1 gradient estimate $\nabla_\theta \hat\ell_J(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}$ goes to zero coordinate-wise as $J\to\infty$ by the particle filter law of large numbers for bounded functionals of particles (\kevin{Prove general LLN for functionals of MOP under boundedness assumptions in L2, but maybe Hoeffding may be enough to show this, because we only want convergence in probability, we just need to show the expectation converges}), so it suffices to show that it is asymptotically unbiased.
-
-    % Write $\hat\pi_{n,J}$ for the filtering distribution at timestep $n$ and with $J$ particles, $\hat\pi_{1:N,J}$ for the sequence of filtering distributions from timesteps $1$ to $N$, and $\pi_n, \pi_{1:N}$ for the respective posteriors.
-
-    % From Lemma \ref{lem:lik-proper-weight}, we know that $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1:n-1}^*, \theta)$ is consistently estimated by the conditional likelihood estimate obtained by MOP-1, for all $\theta$ in an open neighborhood of $\phi$.
-The likelihood is simply the product of the conditional likelihoods, and similarly for the estimates, so the likelihood $\lik(\theta)$ is also consistently estimated by the likelihood estimate obtained by MOP-1, $\hat\lik_J(\theta)$, for all $\theta$ in an open neighborhood of $\phi$, $\gN(\phi)$.
-
-    % We therefore have that 
-    % $$\lim_{J\to\infty} \E[\hat\lik_J(\theta)] = \lik(\theta)$$
-    % pointwise for all $\theta \in \gN(\phi)$.
-Given that the gradient is bounded uniformly for all $J$ and consistency for the likelihood estimate holds for all $\theta$ in an open neighborhood of $\phi$ (and therefore all $\theta'$ in a sufficiently small neighborhood of $\theta$), the bounded convergence theorem allows us to switch the expectation and derivative:
-    % $$\nabla_\theta \E[\hat\lik_J(\theta)] 
-    % = \E[\nabla_\theta \hat\lik_J(\theta)].$$
-
-    % Take limits on both sides to find that 
-    % $$\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)] 
-    % = \lim_{J \to \infty} \E[\nabla_\theta \hat\lik_J(\theta)].$$
-
-    % We then need to swap the limit and derivative on the left-hand side.
-This requires uniform convergence of $\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]$, i.e.
-that $$\sup_{\theta \in \gN(\phi)} ||  \nabla_\theta \E[\hat\lik_J(\theta)] - \lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]||_{\infty} \to 0 \text{ as } J \to \infty,$$
-    % or that $\forall \epsilon>0$ $\exists N\in \mathbb{N}$ such that $n>N => ||  \nabla_\theta \E[\hat\lik_J(\theta)] - \lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]||_{\infty} < \epsilon$ $\forall \theta \in \gN(\phi).$ However, it is easier to instead use an indirect argument.
-
-    % We then need to swap the limit and derivative on the left-hand side. This requires uniform convergence.
-
-    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point we wish to evaluate the gradient at.
-The sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$, by Assumption \ref{assump:local-bounded-derivative}.
-The sequence is also uniformly equicontinuous.
-To see this, by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J(\theta, \omega)|_{\theta=\theta'}$ is also bounded by $H$ for almost every $\omega\in \Omega$ and every $\theta' \in \gN(\phi)$, including $\theta=\phi$.
-A set of functions with derivatives bounded by the same constant is uniformly Lipschitz, and therefore uniformly equicontinuous.
-So the sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ is uniformly equicontinuous for almost every $\omega \in \Omega$.
-    Explicitly, for almost every $\omega \in \Omega$ and every $\epsilon>0$, there exists some $\delta(\omega)>0$ such that for every $||\theta - \theta'||_{\infty}<\delta$ and every $J \in \mathbb{N}$ we have that
-    $$||\nabla_\theta \hat\lik_J(\theta, \omega)-\nabla_\theta \hat\lik_J(\theta', \omega)||_\infty < \epsilon.$$
-
-    \kevin{Our previous understanding was that the target is differentiable, and the sequence is differentiable in increasingly small open neighborhoods.
-This may not be true.
-The neighborhoods the sequence is differentiable in should be independent of $J$, as the resampling does not change with $\theta$ as it is dependent on $\phi$.
-Either way, the target is in each of the neighborhoods, and this may be sufficient in itself.}
-
-    Then, by Arzela-Ascoli, there exists a uniformly convergent subsequence.
-We claim that there is only one subsequential limit.
-We know that when as the gradient is bounded by Assumption \ref{assump:local-bounded-derivative} we can treat the gradient as a bounded functional.
-So by Lemma \ref{lem:posterior-proper-weights} the sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$, and there is therefore only one subsequential limit.
-The sequence therefore converges uniformly to its limit $\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega).$ Therefore, with uniform convergence for the derivatives established, we can swap the limit and derivative, and obtain that for almost every $\omega \in \Omega$,
-    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta, \omega).$$
-
-    From Lemma \ref{lem:lik-proper-weight}, we know that
-    $$\hat{\lik}_J(\theta, \omega) \stackrel{a.s.}{\to} \lik(\theta).$$
-
-    We then have that for almost every $\omega \in \Omega$, 
-    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta, \omega) = \nabla_\theta \lik(\theta),$$
-    as we wanted. The result then follows by the continuous mapping theorem.
-
-    % Taking expectations on both sides and using the bounded convergence theorem to interchange limits and expectations, we see that
-    % $$\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)] = \E[\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta)] = \E[\nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta)] = \nabla_\theta \lim_{J \to \infty} \E[\hat\lik_J(\theta)].$$
-
-    % It then follows that
-    % $$\nabla_\theta\lik(\theta) 
-    % = \nabla_\theta \lim_{J\to\infty} \E[\hat\lik_J(\theta)] 
-    % = \lim_{J\to\infty} \nabla_\theta \E[\hat\lik_J(\theta)] 
-    % = \lim_{J\to\infty} \E[\nabla_\theta \hat\lik_J(\theta)].$$
-
-    % We then have mean-square consistency for the gradient of the likelihood estimate. The result then follows by the continuous mapping theorem.
-
-    \kevin{Explain that the difference in formula between this and Poyiadjis is due to us taking the derivative w.r.t. the particles in the chain rule. }
-
-    \kevin{Alternative argument: Show gradient is unbiased, after showing the likelihood is unbiased in an open neighborhood of phi.
-That should hold by Chopin's argument for unbiased resampling to nonconstant weights.
-Gradient unbiased if can swap expectation and derivative, I think with two continuous derivatives and boundedness, might need BCT/DCT.
-No need to swap limit and derivative in this case, or worrying about uniform limits.}
-
-\end{proof}
-
-\kevin{Show bias is order N, variance is order N when alpha below 1. That implies optimal choice of alpha is not 1. }
-
-\section{Analysis of MOP-$\alpha$ Error and Variance}
-
-\subsection{Analysis of MOP-$0$ Variance}
-
-We made Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} to leverage the results from \cite{karjalainen2023} on the forgetting of the particle filter.
-This is required to show error bounds on the gradient estimates that we provide -- namely that the error of the MOP-$1$ estimator that corresponds to that of \cite{poyiadjis11} is $O(N^2)$, and that the variance of MOP-$\alpha$ for any $\alpha<1$ is $O(N)$.
-
-We can decompose the MOP-$\alpha$ estimator as follows. When $\theta=\phi$,
-$$\hat{\mathcal{L}}(\theta):=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J w_{n, j}^{P, \theta}}=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J (w_{n-1, j}^{F, \theta})^\alpha}.$$
-
-To illustrate the proof strategy, we first analyze the special case when $\alpha=0.$ We then have:
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
-\end{equation}
-
-Similarly to the proof of Lemma \ref{lem:mop-0-formula}, its gradient when $\theta=\phi$ is therefore 
-\begin{align*}
-    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
-    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)} \\
-    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta s_{n,j}}{\sum_{j=1}^J s_{n,j}} \\
-    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)}{f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \phi}; \phi)} \\
-    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
-\end{align*}
-where we use the log-derivative trick, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again.
-The point here was that we establish that the gradient of the log-likelihood estimate is given by the sum of terms over all $N$ and $J$:
-$$\nabla_\theta \hat{\ell}(\theta) = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right).$$
-
-Therefore,
-\begin{align*}
-    \text{Var}(\nabla_\theta \hat\ell(\theta)) &= \frac{1}{J^2}\text{Var}\left(\sum_{n=1}^N\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &= \frac{1}{J^2}\sum_{n=1}^N\text{Var}\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &+ 2\sum_{m<n}\text{Cov}\left(\frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta)).
-\end{align*}
-
-Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
-We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
-$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
-i.e.
-the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
-
-\kevin{How does the dobrushin contraction coefficient relate to the alpha-mixing coefficients?
-If you have a bound on the mixing time $b$, then the alpha mixing coefficients should just be $\alpha(t) \leq e^{-t/b}$.}
-
-By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strongly mixing, with $\alpha$-mixing coefficients $\alpha(k) \leq (1-\epsilon)^{\floor{k/(c\log(J))}}$.
-Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
-Therefore, by Davydov's inequality, noting that $\nabla_\theta\hat\ell_n(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and WLOG assuming $\E[(\nabla_\theta\hat\ell_m(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/4}$, we see that
-\begin{align*}
-    \text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) 
-    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat\ell_m(\theta))^4]^{1/4}\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/4}\\
-    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2}.
-\end{align*}
-
-To bound this, we use the fact that $$\E[(\nabla_\theta\hat\ell_n(\theta))^4] = \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4]+\E[(\nabla_\theta\hat\ell_n(\theta))^2]^2$$
-alongside Lemma 2 of \cite{karjalainen2023}, which shows that
-$$\E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4] \lesssim \frac{G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^2] \lesssim \frac{G'(\theta)^2}{J},$$
-to find that 
-$$\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{G'(\theta)^4}{J^2}+\left(\frac{G'(\theta)^2}{J}\right)^2} = \frac{G'(\theta)^2}{J},$$
-and conclude that 
-$$\text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{G'(\theta)^2}{J}.$$
-
-Putting it all together, we see that
-\begin{align*}
-    \text{Var}(\nabla_\theta \hat\ell(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta))\\
-    &\leq \frac{NG'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\frac{G'(\theta)^2}{J} \\
-    &\leq \frac{G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\right) \\
-    &\lesssim \frac{G'(\theta)^2N}{J}.
-\end{align*}
-
-\paragraph{Remark:} Note that the factor of $N$ that pops up here is due to the use of the unnormalized gradient.
-If one divides the gradient estimate by $\sqrt{N}$ before usage, the variance does not depend on the horizon.
-If one divides by $N$, the error is $O(1/\sqrt{NJ})$!
-
-\kevin{https://math.stackexchange.com/questions/3274996/is-a-transformed-mixing-process-is-still-mixing}
-
-%$$\text{Cov}(x_{n,1:J}^P, x_{m, 1:J}^P) \leq (1-\epsilon)^{\floor{\frac{|n-m|}{c\log(J)}}}$$
-
-\kevin{Ed: Claim 11.4.2, Chopin, is the key to reasoning about this.
-Establish geometric mixing, that $\text{Cov}(\nabla_\theta\hat\ell_n, \nabla_\theta\hat\ell_m) \leq \rho^{|n-m|}V_n$, where $V_n$ is asymptotically stable (the latter only holds for $\alpha<1$, as only then do we get asymptotic stability, otherwise it increases linearly in $n$).
-Geometrically decreasing covariances and stable variances over time are the same thing.
-Then, we can bound the total variance by $\text{Var}(\sum_n \nabla_\theta \hat\ell_n(\theta)) = \sum_{m,n} \text{Cov}(\nabla_\theta\hat\ell_n(\theta), \nabla_\theta\hat\ell_m(\theta)) = NV_n + 2(N-1)\rho V_n + 2(N-3)\rho^2V_n + ...$}
-
-\subsection{Analysis of MOP-$\alpha$ Variance}
-
-Next, we analyze the case for $\alpha \in (0,1)$.
-The proof strategy is as follows.
-Instead of analyzing the variance of the MOP-$\alpha$ gradient estimate proper, we will analyze the variance of a modified estimator where we truncate the weights at $k$ timesteps, so this estimator only looks at most $k$ timesteps back while accumulating the importance weights.
-This enables us to establish strong mixing for the truncated estimator, in order to get a bound on the variance that is $O(NJ^{-1})$.
-We emphasize that this estimator is only a theoretical construct -- we do not actually use this estimator.
-
-Coupled with a bound ensuring that the $k$-truncated estimator provides an estimate that is close to that of MOP-$\alpha$ proper, we get a bound on the variance of MOP-$\alpha$ that comprises of the variance of the $k$-truncated estimator plus the error, for any $k \leq N$.
-It then holds that the final variance bound on MOP-$\alpha$ is given by the minimum over all $k$, and is never larger than $O(1+NJ^{-1})$.
-
-Using the log-derivative trick and that $w_{n,j}^{F,\theta} = w_{n,j}^{F,\theta,k} = 1$ when $\theta=\phi$,
-\begin{align*}
-    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,k}\right)\right\rVert_{\infty}\\
-    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta}}{{\sum_{j=1}^J w_{n,j}^{F,\theta}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,k}}}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,k}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta}}{w_{n,j}^{F,\theta}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,k}}{w_{n,j}^{F,\theta,k}}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,k}\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta}\right)-\log\left(w_{n,j}^{F,\theta,k}\right)\right)\right\rVert_{\infty}\\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=n-k}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^{n-k}\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}\\
-    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
-    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}G'(\theta)\\
-    &\leq G'(\theta)\frac{\alpha^k-\alpha^n}{1-\alpha},
-\end{align*}
-where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
-
-We can then bound $||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||$:
-\begin{align*}
-    &\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_\infty\\
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
-    &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\\
-    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)
-    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\Bigg)\right\rVert_\infty \\
-    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)\right)\right\lVert_{\infty}
-    +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\right)\right\rVert_\infty,
-\end{align*}
-and each of these two terms is bounded by $\frac{\alpha^k}{1-\alpha}NG'(\theta)$. So we now know that
-$$\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_\infty \leq  \frac{2\alpha^k}{1-\alpha}NG'(\theta),$$
-which is our desired error bound.
-
-Now, we bound the variance of $\hat s_k(\theta)$. 
-Recall that we defined
-$$\hat s_k(\theta) = \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) = \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right).$$
-
-We can then decompose this expression with the log-derivative trick while noting that $w_{n,j}^{F,\theta, k} = 1$ whenever $\theta=\phi$, to see that
-\begin{align*}
-    \hat s_k(\theta) 
-    &= \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right) \\
-    &= \sum_{n=1}^N \left(\frac{\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}}{\sum_{j=1}^J w_{n,j}^{F,\theta, k}}-\frac{\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}}{\sum_{j=1}^J w_{n-1,j}^{A,F,\theta, k}}\right) \\
-    &= \sum_{n=1}^N \left(\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}\right)\\
-    &= \frac{1}{J}\sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right).
-\end{align*}
-
-It now follows that we need to bound the variance at a single timestep, namely
-$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right).$$
-
-We use the log-derivative trick yet again to find, noting that $w_{n,j}^{F,\theta, k}=1$ when $\theta=\phi$, that
-\begin{align*}
-    \nabla_\theta w_{n,j}^{F,\theta, k} = \frac{\nabla_\theta w_{n,j}^{F,\theta, k}}{w_{n,j}^{F,\theta, k}} = \nabla_\theta \log(w_{n,j}^{F,\theta, k}) = \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) = \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right).
-\end{align*}
-
-Then,
-\begin{align*}
-    \text{Var}\left(\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right) 
-    &= \text{Var}\left(\nabla_\theta \left(\sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right)- \sum_{i=n-k-1}^{n-1} \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right)\right)\right) \\
-    &= \text{Var} \left(\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1}\log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right).
-\end{align*}
-
-Note that $$\hat s_{n,k}(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)$$ is a function bounded by $C\frac{G'(\theta)(1-\alpha^k)}{1-\alpha} =: CG'(\theta)r$ in each coordinate for some constant $C$ (by Assumption \ref{assump:bounded-measurement}) that depends only on $k$ timesteps, and not $n$.
-We can therefore invoke Lemma 2 of \cite{karjalainen2023} to bound the $L_2$ error of each of the $k$ (from time $n-k$ to time $n$) functionals from its expectation under the posterior by $CrG'(\theta)J^{-1/2}k$.
-That is, we have that
-$$\E\left[||\hat s_{n,k}(\theta) - \E_{\pi}\hat s_{n,k}(\theta)||_\infty^2\right]^{1/2} \leq \frac{Cr^2G'(\theta) k}{\sqrt{J}}.$$
-
-By the bias-variance decomposition, this in turn bounds the variance at a single timestep by $C^2r^2G'(\theta)^2J^{-1}k^2$. Concretely, we have that
-$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \lesssim \frac{r^2k^2G'(\theta)^2}{J}.$$
-
-It now remains to bound the variance of $\hat s_k(\theta)$ by considering the covariance of each of the $N$ terms that comprise it. We decompose
-\begin{align*}
-    \text{Var}(\nabla_\theta \hat s_k(\theta)) &= \text{Var}\left(\sum_{n=1}^N \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \\
-    &= \sum_{n=1}^N\text{Var}\left(\hat s_{n,k}(\theta)\right)+ 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \\
-    &\lesssim \frac{r^2k^2G'(\theta)^2N}{J} + 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right).
-\end{align*}
-
-Similarly to the proof of the MOP-$0$ case, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
-We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
-$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
-i.e.
-the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
-By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strongly mixing, with $\alpha$-mixing coefficients $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c\log(J))}}$.
-Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
-
-We now derive the $\alpha$-mixing coefficients of $(\hat s_{n,k}(\theta))_{n=1}^N$.
-Observe that $\hat s_{n,k}$ is strong mixing at lag $k+1$, as all weights beyond $k$ timesteps are truncated.
-We therefore have that the mixing time of $\hat s_k(\theta)$ is $O(1+k+\log(J))$, and that the $\alpha$-mixing coefficients for $(\hat s_{n,k}(\theta))_{n=1}^N$ are given by $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c(1+k+\log(J)))}}$.
-
-Therefore, by Davydov's inequality and Lemma 2 of \cite{karjalainen2023}, and noting that $\hat s_{n,k}(\theta) \leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, by a similar argument to the MOP-0 case, we find that
-
-$$\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J}.$$
-
-Concretely, noting that $\nabla_\theta\hat s_{n,k}(\theta)\leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and WLOG assuming $\E[(\nabla_\theta\hat\hat s_{m,k}(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat\hat s_{n,k}(\theta))^4]^{1/4}$, we apply Davydov's inequality to see that
-\begin{align*}
-    \text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) 
-    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat s_{m,k}(\theta))^4]^{1/4}\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/4}\\
-    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2}.
-\end{align*}
-
-To bound this, we use the fact that $$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4] = \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4]+\E[(\nabla_\theta\hat s_{n,k}(\theta))^2]^2$$
-alongside Lemma 2 of \cite{karjalainen2023}, which shows that
-$$\E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4] \lesssim \frac{r^4G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^2] \lesssim \frac{r^2G'(\theta)^2}{J},$$
-to find that 
-$$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{r^4G'(\theta)^4}{J^2}+\left(\frac{r^2G'(\theta)^2}{J}\right)^2} = \frac{r^2G'(\theta)^2}{J},$$
-and conclude that 
-$$\text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J}.$$
-
-Putting it all together, we see that
-\begin{align*}
-    \text{Var}(\nabla_\theta \hat s_{k}(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat s_{n,k}(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat s_{m,k}(\theta),\nabla_\theta\hat s_{n,k}(\theta))\\
-    &\leq \frac{Nr^2k^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J} \\
-    &\leq \frac{r^2k^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\right) \\
-    &\lesssim \frac{k^2r^2G'(\theta)^2N}{J}.
-\end{align*}
-
-We will now use this result to bound $\text{Var}(\nabla_\theta \hat s_k(\theta)$ proper. For random variables $X,Y$ where $|X-Y|$ is bounded almost surely by some $M$, we have that 
-$$X-\E X - 2M \leq Y - \E Y \leq X - \E X + 2M$$
-so,
-$$(Y-\E Y)^2 \leq (X-\E X - 2M)^2 + (X-\E X +2M)^2.$$
-Taking expectations, we get
-$$\text{Var}(Y) \leq \text{Var}(X) + 8M^2.$$
-
-It follows from this result and the result we proved earlier, namely that
-$$||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||_\infty \leq \frac{2\alpha^k}{1-\alpha}NG'(\theta),$$
-that the variance of MOP-$\alpha$ proper is bounded by, for any $k \leq N$,
-that the variance of MOP-$\alpha$ proper is bounded by
-$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \frac{k^2r^2G'(\theta)^2N}{J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2.$$
-
-As the above holds for any $k \leq N$,
-$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \min_{k\leq N} \left(\frac{k^2G'(\theta)^2N}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right).$$
-
-% To interpret this result, we can choose $k$ such that
-% \begin{align*}
-%     &\frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \leq 1,\\
-%     &\alpha^{2k} \leq \frac{(1-\alpha)^2}{16N^2G'(\theta)^2},\\
-%     &k \geq \frac{\log\left(16N^2G'(\theta)^2\right)-2\log(1-\alpha)}{2\log\left(\frac{1}{\alpha}\right)}.
-% \end{align*}
-
-% To interpret this result, if there exists some $k$ such that $$N \geq k \geq \frac{\log\left(16NG'(\theta)\right)-\log(1-\alpha)}{\log\left(\frac{1}{\alpha}\right)},$$
-% we can choose such a $k$, as well as $k=N$, to see that
-% \begin{align*}
-%     \text{Var}(\nabla_\theta \hat\ell(\theta)) 
-%     &\lesssim \min_{k\leq N}\left(\frac{c_d^2k^2r^2G'(\theta)^2N}{J} + \frac{\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right)\\
-%     &\lesssim \min\left\{1+\frac{c_d^2G'(\theta)^2N}{J}, \frac{c_d^2G'(\theta)^2r^2N^3}{J}\right\} \\
-%     &\leq \min\left\{1+\frac{c_d^2G'(\theta)^2N}{J}, \frac{c_d^2G(\theta)^2r^2N}{J}\right\},
-% \end{align*}
-% as $NG'(\theta)\leq G(\theta)$ as a consequence of Assumption \ref{assump:local-bounded-derivative}.
-
-\subsection{Analysis of MOP-$\alpha$ Bias}
-
-Using the log-derivative trick and that $w_{n,j}^{F,\theta,1} = w_{n,j}^{F,\theta,\alpha} = 1$ when $\theta=\phi$,
-\begin{align*}
-    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,1}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)\right\rVert_{\infty}\\
-    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,1}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,1}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}}}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,1}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,\alpha}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,1}}{w_{n,j}^{F,\theta,1}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,\alpha}}{w_{n,j}^{F,\theta,\alpha}}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,1}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha}\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta,1}\right)-\log\left(w_{n,j}^{F,\theta,\alpha}\right)\right)\right\rVert_{\infty}\\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=1}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}.
-\end{align*}
-
-Now we establish the bound:
-\begin{align*}
-     \nabla_\theta \left(\sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)
-     &\leq \sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
-    &\leq \sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)G'(\theta)\\
-    &\leq G'(\theta)\left(\frac{n-\alpha n + \alpha^n - 1}{1-\alpha}\right)
-\end{align*}
-where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
-
-By a similar decomposition to that of the error of the $k$-truncated gradient to the MOP-$\alpha$ gradient estimate,
-\begin{align*}
-    &\left\lVert\nabla_\theta\hat\ell^1(\theta) - \nabla_\theta\hat\ell^\alpha(\theta) \right\rVert_\infty\\
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
-    &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\\
-    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)
-    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right)\Bigg)\right\rVert_\infty \\
-    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)\right|
-    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right)\right|\Bigg)\right\rVert_\infty.
-\end{align*}
-
-From our arguments above, we know that 
-$$\left\lVert\nabla_\theta\hat\ell^1(\theta) - \nabla_\theta\hat\ell^\alpha(\theta) \right\rVert_\infty \leq 2NG'(\theta)\left(\frac{N-\alpha N + \alpha^N - 1}{1-\alpha}\right).$$
-
-So the bias of MOP-$\alpha$ is the bias of MOP-$1$ arising from the logarithm in the gradient of the log-likelihood plus no more than $$2NG'(\theta)\left(\frac{N-\alpha N + \alpha^N - 1}{1-\alpha}\right),$$
-where we note that the bias vanishes as $\alpha$ goes to $1$ as
-$$\lim_{\alpha\to1}\left(\frac{N-\alpha N + \alpha^N - 1}{1-\alpha}\right) = 0.$$
-
-\subsection{An Improved Bound by Mixing}
-
-We decompose
-$$\E||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty \leq \E||\nabla_\theta\ell_\theta - \nabla_\theta\hat s_k^1(\theta)||_\infty + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty.$$
-
-The broad idea is that the first term can be controlled by a mixing argument, the second term garners $O(Nk^2)$ error, and the third term is $O\left(\frac{N}{1-\alpha}\right)$.
-It is unclear whether the bias can be reduced further -- the gradient is a sum of $N$ terms, and is therefore $O(N)$ itself.
-
-The first term -- bound ground truth of conditional scores from particle approximation conditional on correct filtering distribution at time $n-k$, bound that from particle approximation conditional on arbitrary filtering distribution at time $n-k$.
-First term sums up $k$ terms, $O(Nk^2)$ error, second term bounded by $k$ applications of the Markov kernel, so just apply forgetting of the particle filter result.
-
-But you can get an MSE bound like this! Simply decompose
-$$||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 \leq ||\nabla_\theta\ell_\theta - \nabla_\theta\hat s_k^1(\theta)||_2^2 + ||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2 + ||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_2^2,$$
-and observe that 
-$$||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty =  \left\lVert\sum_{n=1}^N \nabla_\theta \hat\ell_n^\alpha(\theta) - \sum_{n=1}^N \nabla_\theta \ell_n(\theta)\right\rVert_\infty \leq \sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_\infty.$$
-
-\section{Finite-Particle Analysis of Gradient Methods with Algorithm \ref{alg:ifad}}
-\label{app:ifad}
-
 The analysis in this section roughly follows the analysis in \cite{mahoney16}, except with the caveat that none of the matrix concentration bounds they use apply here as the particles are dependent.
-We instead use the concentration inequality from \cite{delmoral2011ci} to bound the gradient and Hessian estimates.
+We instead use the concentration inequality from \cite{delMoral11} to bound the gradient and Hessian estimates.
 In this section, we fix $\omega \in \Omega$ only within each filtering iteration, evaluate Algorithm \ref{alg:mop} at $\theta=\phi$, and analyze Algorithm \ref{alg:ifad} post-iterated filtering.
 
-The convergence analysis in Theorem \ref{thm:convergence} is limited to the case where $-\ell$ is $\gamma$-strongly convex.
+The convergence analysis in Theorem \ref{thm:mop-convergence} is limited to the case where $-\ell$ is $\gamma$-strongly convex.
 Though it is true that in a neighborhood of the optimum local asymptotic normality holds and the log-likelihood is strongly convex in this neighborhood, in practice likelihood surfaces for POMPs are often highly nonconvex globally.
@@
 \subsection{Bounding the Gradient}
-\kevin{Add a few words to explain that the gradient of the log-likelihood w.r.t.
+%\kevin{Add a few words to explain that the gradient of the log-likelihood w.r.t.
 parameters is a functional of the particle measure under boundedness assumptions and fixed parameters, appeal to authority and say doucet did it too.
@@
 
-\kevin{Gradient linear map, bound conditional likelihoods at time n.
+%\kevin{Gradient linear map, bound conditional likelihoods at time n.
 Alternatively, convert del moral and doucet expectation bound to high-probability bound.
@@
     \label{lemma:grad_bound}
-    Consider the gradient estimate obtained by MOP-1, which we know by Lemma \ref{lem:mop-grad-consistency} is consistent for the score, where $\theta = \phi$.
-For $||\nabla_\theta \hat{\loglik}(\theta) - \nabla_\theta \loglik(\theta)||_2$ to be bounded by $\epsilon$ with probability $1-\delta$, we require
+    Consider the gradient estimate obtained by MOP-1, which we know by Theorem \ref{thm:mop-grad-consistency} is consistent for the score, where $\theta = \phi$.
+For $||\nabla_\theta \hat{\loglik}^1(\theta) - \nabla_\theta \loglik(\theta)||_2$ to be bounded by $\epsilon$ with probability $1-\delta$, we require
     \begin{align}
@@
     \end{align}
-    where $NG'(\theta) \leq G(\theta)$ are defined in Assumptions \ref{assump:bounded-measurement} and \ref{assump:local-bounded-derivative}, $h(t) = \frac{1}{2}(t - \log(1+t))$, and $\beta_N$ and $r_N$ are two additional finite model-specific constants that do not depend on $J$, but do depend on $N$ and $p$, as defined in \cite{delmoral2011ci}.
+    where $NG'(\theta) \leq G(\theta)$ are defined in Assumptions \ref{assump:bounded-measurement} and \ref{assump:local-bounded-derivative}, $h(t) = \frac{1}{2}(t - \log(1+t))$, and $\beta_N$ and $r_N$ are two additional finite model-specific constants that do not depend on $J$, but do depend on $N$ and $p$, as defined in \cite{delMoral11}.
 
@@
     \begin{align}
-        ||\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).
+        ||\nabla_\theta \hat\ell^1(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).
     \end{align}
@@
 
-\paragraph{Remark:} According to \cite{delmoral2011ci}, under some regularity conditions, $r_N$ and $\beta_N$ are linear in the trajectory length $N$.
+\paragraph{Remark:} According to \cite{delMoral11}, under some regularity conditions, $r_N$ and $\beta_N$ are linear in the trajectory length $N$.
 This corresponds to the finding by \cite{poyiadjis11} that the variance of the estimate is at least quadratic in the trajectory length, and their remark that the result of \cite{delMoral03} establishes that the $L_p$ error is bounded by $O(N^2J^{-1/2})$ (equivalently, the variance is bounded by $O(N^4J^{-1})$) after accounting for the sum over timesteps.
@@
 
-Either way, usually in practice one considers the estimate normalized through dividing it by $N$, in which case we only need
-$$J > \max\left\{2G'(\theta)\frac{r_N\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{2p}{\delta}\right)\right)\right), 8G'(\theta)^2\beta_N^2\frac{p\log(2p/\delta)}{\epsilon^2}\right\}.$$
-Equivalently,
-$$||N^{-1}(\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta))||_2 \leq G'(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).$$
-
-It is well-known that bounds like these hide an exponential dependence in the dimension of the state space $d$.
-We take care to keep track of the terms depending on $d$ when they arise in front of us, but are unable to do so for the terms that arise from the analysis of \cite{delmoral2011ci}.
-
-\kevin{This is slightly alarming, in that the estimate really has $O(N^4)$ variance after accounting for summing over timesteps.
-In contrast, MOP-0 has only $O(N)$ variance after summing over timesteps.
-}
-
 \begin{proof}
 
-We will seek to use the concentration inequality of \cite{delmoral2011ci} to bound the deviation of the gradient estimate from the gradient of the negative log-likelihood in the sup norm with a union bound.
-Fix some $\theta \in \gN(\phi)$.
+We will seek to use the concentration inequality of \cite{delMoral11} to bound the deviation of the gradient estimate from the gradient of the negative log-likelihood in the sup norm with a union bound.
+Fix $\theta = \phi$.
 
-Recall that as $w_{N, j}^{F, \theta}=1$ when $\theta=\phi$,
+From the decomposition in the proof of Lemma \ref{lem:mop-1-formula}, as $w_{N, j}^{F, \theta}=1$ when $\theta=\phi$, we have that
 $$
@@
 Therefore, these have bounded oscillation, satisfying the requirement that $\text{osc} \left(\frac{\partial}{\partial\theta_i} \varphi_i(x_{n,j}^{P,\theta}) \right) \leq G'(\theta)$.
-\cite{delmoral2011ci} in fact assume $\text{osc}(f) \leq 1$, so we simply scale their bound accordingly.
+Note that \cite{delMoral11} in fact assume $\text{osc}(f) \leq 1$, so we simply scale their bound accordingly.
 
-\kevin{Put some justification/make the setup of delmoral2011ci concrete here. Concentration of gradient estimate under particle measure to limit under posterior measure}
-
-Apply the Hoeffding-type concentration inequality from Del Moral and Rio \cite{delmoral2011ci} and a union bound over each $\varphi_n^i(x_{n,j}^{F,\theta})$, totaling $N$ timesteps and $p$ parameters, to find that
+Apply the Hoeffding-type concentration inequality from Del Moral and Rio \cite{delMoral11} and a union bound over each $\varphi_n^i(x_{n,j}^{F,\theta})$, totaling $N$ timesteps and $p$ parameters, to find that
 
@@
 with probability at least $1-2Np\exp(-t)$.
-Note that the concentration inequality only considers the error from the expectation under the filtering distribution -- but we invoke the consistency of MOP-$1$ shown in Lemma \ref{lem:mop-grad-consistency} to establish that the expectation under the filtering distribution is in fact the score.
+Although the above concentration inequality only considers the error from the expectation under the filtering distribution, we invoke the consistency of MOP-$1$ shown in Theorem \ref{thm:mop-grad-consistency} to establish that the expectation under the filtering distribution is in fact the score.
 It therefore holds that with the same probability, that when summing over $N$, as $NG'(\theta) \leq G(\theta)$,
@@
 
-We split the $\delta$ failure probability among these $2Np$ terms, to find $\delta\leq2Np\exp(-t) => t\leq\log(2Np/\delta)$.
-Here, $h(t) = \frac{1}{2}(t - \log(1+t))$.
+We split the $\delta$ failure probability among these $2Np$ terms, to find $\delta\leq2Np\exp(-t) => t\leq\log(2Np/\delta)$, where $h(t) = \frac{1}{2}(t - \log(1+t))$.
 The two additional model-specific parameters are $\beta_t$ and $r_t$, which do not depend on $J$.
@@
 
-    for $\hat{H}(\theta)$ to be invertible and positive definite with minimum eigenvalue greater than or equal to $c \in (0, \sum_{j=1}^J \E\lambda_{\min}(H_j))$ and with probability at least $1-\exp(-t)$.
+    for $\hat{H}(\theta)$ to be invertible and positive definite with minimum eigenvalue greater than or equal to $c \in (0, \sum_{j=1}^J \E\lambda_{\min}(H_j))$, with probability at least $1-\exp(-t)$.
 \end{lem}
@@
 
-From \cite{boyd2004convex}, we know that the minimum eigenvalue of a symmetric matrix is concave. Therefore, it suffices to show that the first inequality in the below expression
+It is known that the minimum eigenvalue of a symmetric matrix is concave. Therefore, it suffices to show that the first inequality in the below expression
 
@@
 
-Apply the particle Hoeffding concentration inequality from \cite{delmoral2011ci} to find that
+Apply the particle Hoeffding concentration inequality from \cite{delMoral11} to find that
 
@@
 \begin{align}
-    \sum_{j=1}^J \lambda_{\min}(H_j) \geq -r_t(1+h^{-1}(t)) - \sqrt{2tJ}\beta_t + J\gamma' \geq c \\
-    J\gamma' \geq c + r_t(1+h^{-1}(t)) + \sqrt{2tJ}\beta_t 
+    \sum_{j=1}^J \lambda_{\min}(H_j) \geq -r_t(1+h^{-1}(t)) - \sqrt{2tJ}\beta_t + J\gamma' \geq c, \\
+    J\gamma' \geq c + r_t(1+h^{-1}(t)) + \sqrt{2tJ}\beta_t.
 \end{align}
@@
 
-\subsection{Convergence Analysis of Theorem \ref{thm:convergence}}
+\subsection{Convergence Analysis of Theorem \ref{thm:mop-convergence}}
 
@@
 
-Define $\theta_\eta = \theta_n + \eta p_n$, where $p_n=-(H(\theta_n))^{-1}g(\theta_n)$.
+Define $\theta_\eta = \theta_m + \eta p_m$, where $p_m=-(H(\theta_m))^{-1}g(\theta_m)$.
 As in Roosta-Khorasani and Mahoney \cite{mahoney16}, we want to show there is some iteration-independent $\tilde{\eta}>0$ such that the Armijo condition
 \begin{equation}
-    f(\theta_n+\eta p_n) \leq f(\theta_n) + \eta\beta p_n^Tg(\theta_n)
+    f(\theta_m+\eta p_m) \leq f(\theta_m) + \eta\beta p_m^Tg(\theta_m),
 \end{equation}
@@
 
-By an argument found in the beginning of the proof of Theorem 6 in \cite{mahoney16}, we have that choosing $J$ such that $||\nabla_\theta\hat{\ell}(\theta_n) - \nabla_\theta \ell(\theta_n)|| \leq \epsilon$ and $\lambda_{\min}(H(\theta_n)) \geq c$ with probability $(1-\delta)^2$ for each $n$ yields
+By an argument found in the beginning of the proof of Theorem 6 in \cite{mahoney16}, we have that choosing $J$ such that $||\nabla_\theta\hat{\ell}(\theta_m) - \nabla_\theta \ell(\theta_m)|| \leq \epsilon$ and $\lambda_{\min}(H(\theta_m)) \geq c>0$ for each $m$, yields
 
 \begin{align}
-    f(\theta_\eta)-f(\theta_n) \leq \eta p_n^Tg(\theta_n) + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2
+    f(\theta_\eta)-f(\theta_m) \leq \eta p_m^Tg(\theta_m) + \epsilon\eta||p_m|| + \eta^2 \Gamma ||p_m||^2 / 2,
 \end{align}
+with probability $1-\delta/2$. From now on, we assume that we are on the success event of this high-probability statement.
 
-Now, for some $c>0$, with probability $1-\delta$
+Now, it then holds that
 \begin{equation}
-    p_n^Tg(\theta_n) = -p_n^TH(\theta_n)p_n \geq -c||p_n||^2
+    p_m^Tg(\theta_m) = -p_m^TH(\theta_m)p_m \geq -c||p_m||^2,
 \end{equation}
@@
 \begin{align}
-    f(\theta_\eta)-f(\theta_n) \leq -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2
+    f(\theta_\eta)-f(\theta_m) \leq -\eta p_m^TH(\theta_m)p_m + \epsilon\eta||p_m|| + \eta^2 \Gamma ||p_m||^2 / 2,
 \end{align}
@@
 \begin{align}
-    -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2 &\leq \eta \beta p_n^Tg(\theta_n) = - \eta \beta p_n^TH(\theta_n)p_n \\
-    \epsilon||p_n|| + \eta \Gamma ||p_n||^2 / 2 &\leq (1- \beta) p_n^TH(\theta_n)p_n \\
-    \epsilon + \eta \Gamma ||p_n|| / 2 &\leq c(1- \beta) ||p_n||
+    -\eta p_m^TH(\theta_m)p_m + \epsilon\eta||p_m|| + \eta^2 \Gamma ||p_m||^2 / 2 &\leq \eta \beta p_m^Tg(\theta_m) = - \eta \beta p_m^TH(\theta_m)p_m \\
+    \epsilon||p_m|| + \eta \Gamma ||p_m||^2 / 2 &\leq (1- \beta) p_m^TH(\theta_m)p_m \\
+    \epsilon + \eta \Gamma ||p_m|| / 2 &\leq c(1- \beta) ||p_m||,
 \end{align}
@@
 \begin{equation}
-    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_n)|| \leq \frac{c(1-\beta)}{2}||p_n||
+    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_m)|| \leq \frac{c(1-\beta)}{2}||p_m||,
 \end{equation}
+which is given by our choice of $\eta$.
 
@@
 \begin{equation}
-||g(\theta_n)|| - ||\nabla_\theta f(\theta_n)|| \leq ||g(\theta_n) - \nabla_\theta f(\theta_n)|| \leq \epsilon => ||\nabla_\theta f(\theta_n)|| \geq ||g(\theta_n)|| - \epsilon
+||g(\theta_m)|| - ||\nabla_\theta f(\theta_m)|| \leq ||g(\theta_m) - \nabla_\theta f(\theta_m)|| \leq \epsilon => ||\nabla_\theta f(\theta_m)|| \geq ||g(\theta_m)|| - \epsilon
 \end{equation} and
 \begin{equation}
-||\nabla_\theta f(\theta_n)||-||g(\theta_n)|| \leq ||\nabla_\theta f(\theta_n)-g(\theta_n)|| \leq \epsilon => ||g(\theta_n)|| \geq ||\nabla_\theta f(\theta_n)|| - \epsilon
+||\nabla_\theta f(\theta_m)||-||g(\theta_m)|| \leq ||\nabla_\theta f(\theta_m)-g(\theta_m)|| \leq \epsilon => ||g(\theta_m)|| \geq ||\nabla_\theta f(\theta_m)|| - \epsilon.
 \end{equation}
 
-There are now two cases. If the algorithm terminates and $||g(\theta_n)|| \leq \sigma \epsilon$, we can derive
+There are now two cases. If the algorithm terminates and $||g(\theta_m)|| \leq \sigma \epsilon$, we can derive
 
 \begin{equation}
-    ||\nabla_\theta f(\theta_n)|| \leq ||g(\theta_n)|| + \epsilon = \sigma\epsilon+\epsilon = (\sigma+1)\epsilon
+    ||\nabla_\theta f(\theta_m)|| \leq ||g(\theta_m)|| + \epsilon = \sigma\epsilon+\epsilon = (\sigma+1)\epsilon.
 \end{equation}
 
-If the algorithm does not terminate, then $||g(\theta_n)|| > \sigma \epsilon$. Notice that
+If the algorithm does not terminate, then $||g(\theta_m)|| > \sigma \epsilon$. Notice that
 
 \begin{align}
-    \epsilon \geq ||g(\theta_n) - \nabla_\theta f(\theta_n)|| &\geq ||g(\theta_n)|| - ||\nabla_\theta f(\theta_n)|| \\
-    ||\nabla_\theta f(\theta_n)|| + \epsilon &\geq ||g(\theta_n)|| \geq \sigma \epsilon \\
-    ||\nabla_\theta f(\theta_n)|| &\geq \sigma \epsilon - \epsilon = (\sigma - 1)\epsilon \\
-    \frac{||\nabla_\theta f(\theta_n)||}{\sigma-1} &\geq \epsilon
+    \epsilon \geq ||g(\theta_m) - \nabla_\theta f(\theta_m)|| &\geq ||g(\theta_m)|| - ||\nabla_\theta f(\theta_m)|| \\
+    ||\nabla_\theta f(\theta_m)|| + \epsilon &\geq ||g(\theta_m)|| \geq \sigma \epsilon \\
+    ||\nabla_\theta f(\theta_m)|| &\geq \sigma \epsilon - \epsilon = (\sigma - 1)\epsilon \\
+    \frac{||\nabla_\theta f(\theta_m)||}{\sigma-1} &\geq \epsilon,
 \end{align}
@@
 \begin{align}
-    ||\nabla_\theta f(\theta_n)|| - \epsilon
-    &\geq  ||\nabla_\theta f(\theta_n)|| - \frac{||\nabla_\theta f(\theta_n)||}{\sigma-1} = \left(1-\frac{1}{\sigma-1}\right)||\nabla_\theta f(\theta_n|| = \frac{\sigma-2}{\sigma-1}||\nabla_\theta f(\theta_n)|| \geq \frac{2}{3}||\nabla_\theta f(\theta_n)||
+    ||\nabla_\theta f(\theta_m)|| - \epsilon
+    &\geq  ||\nabla_\theta f(\theta_m)|| - \frac{||\nabla_\theta f(\theta_m)||}{\sigma-1} = \left(1-\frac{1}{\sigma-1}\right)||\nabla_\theta f(\theta_m|| = \frac{\sigma-2}{\sigma-1}||\nabla_\theta f(\theta_m)|| \geq \frac{2}{3}||\nabla_\theta f(\theta_m)||.
 \end{align}
@@
 \begin{align}
-    p_n^TH(\theta_n)p_n &= (-(H(\theta_n))^{-1}g(\theta_n))^TH(\theta_n)(-(H(\theta_n))^{-1}g(\theta_n)) \\
-    &= g(\theta_n)^T(H(\theta_n))^{-1}g(\theta_n) \\
-    &\geq \frac{1}{c}||g(\theta_n)||^2 \\
-    &\geq \frac{1}{c}(||\nabla_\theta f(\theta_n)|| - \epsilon)^2 \\
-    &\geq \frac{4}{9c}||\nabla_\theta f(\theta_n)||^2
+    p_m^TH(\theta_m)p_m &= (-(H(\theta_m))^{-1}g(\theta_m))^TH(\theta_m)(-(H(\theta_m))^{-1}g(\theta_m)) \\
+    &= g(\theta_m)^T(H(\theta_m))^{-1}g(\theta_m) \\
+    &\geq \frac{1}{c}||g(\theta_m)||^2 \\
+    &\geq \frac{1}{c}(||\nabla_\theta f(\theta_m)|| - \epsilon)^2 \\
+    &\geq \frac{4}{9c}||\nabla_\theta f(\theta_m)||^2.
 \end{align}
@@
 \begin{align}
-    f(\theta_n) - f(\theta^*) \leq \frac{1}{2\gamma}||\nabla_\theta f(\theta_n)||^2
+    f(\theta_m) - f(\theta^*) \leq \frac{1}{2\gamma}||\nabla_\theta f(\theta_m)||^2,
 \end{align}
@@
 \begin{equation}
-    f(\theta_n) - f(\theta^*) \leq \frac{1}{2\gamma}||\nabla_\theta f(\theta_n)||^2 \leq \frac{9c}{4}\frac{1}{2\gamma}p_n^TH(\theta_n)p_n
+    f(\theta_m) - f(\theta^*) \leq \frac{1}{2\gamma}||\nabla_\theta f(\theta_m)||^2 \leq \frac{9c}{4}\frac{1}{2\gamma}p_m^TH(\theta_m)p_m,
 \end{equation}
 \begin{equation}
-    \frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \leq \frac{4}{9c}||\nabla_\theta f(\theta_n)||^2 \leq p_n^TH(\theta_n)p_n
+    \frac{8\gamma}{9c}(f(\theta_m) - f(\theta^*)) \leq \frac{4}{9c}||\nabla_\theta f(\theta_m)||^2 \leq p_m^TH(\theta_m)p_m,
 \end{equation}
 \begin{equation}
-    f(\theta_n) - f(\theta^*) \leq \frac{9c}{8\gamma}p_n^TH(\theta_n)p_n
+    f(\theta_m) - f(\theta^*) \leq \frac{9c}{8\gamma}p_m^TH(\theta_m)p_m,
 \end{equation}
-
 \begin{equation}
-    -\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \geq -\frac{4}{9c}||\nabla_\theta f(\theta_n)||^2 \geq -p_n^TH(\theta_n)p_n
+    -\frac{8\gamma}{9c}(f(\theta_m) - f(\theta^*)) \geq -\frac{4}{9c}||\nabla_\theta f(\theta_m)||^2 \geq -p_m^TH(\theta_m)p_m.
 \end{equation}
@@
 \begin{align}
-    f(\theta_{n+1})-f(\theta_n) &\leq -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2 \\
-    &\leq -\eta\beta p_n^TH(\theta_n)p_n \\
-    &\leq -\eta\beta\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) 
+    f(\theta_{m+1})-f(\theta_m) &\leq -\eta p_m^TH(\theta_m)p_m + \epsilon\eta||p_m|| + \eta^2 \Gamma ||p_m||^2 / 2 \\
+    &\leq -\eta\beta p_m^TH(\theta_m)p_m \\
+    &\leq -\eta\beta\frac{8\gamma}{9c}(f(\theta_m) - f(\theta^*)).
 \end{align}
@@
 \begin{align}
-    f(\theta_{n+1}) - f(\theta^*) 
-    &= f(\theta_{n+1})-f(\theta_n)+f(\theta_n)- f(\theta^*) \\
-    &\leq f(\theta_n)- f(\theta^*) -\eta\beta\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \\
-    &= (1-\eta\beta\frac{8\gamma}{9c})(f(\theta_n) - f(\theta^*)) 
+    f(\theta_{m+1}) - f(\theta^*) 
+    &= f(\theta_{m+1})-f(\theta_m)+f(\theta_m)- f(\theta^*) \\
+    &\leq f(\theta_m)- f(\theta^*) -\eta\beta\frac{8\gamma}{9c}(f(\theta_m) - f(\theta^*)) \\
+    &= (1-\eta\beta\frac{8\gamma}{9c})(f(\theta_m) - f(\theta^*)).
 \end{align}
@@
 \end{proof}
-
-\section{Miscellanous Technical Lemmas}
-
-\begin{lem}
-    \label{lem:dobrushin-implies-alpha-mixing}
-    Setting $X$ to be the particle collection at time $m$, and $Y$ at time $n$, we have that the alpha mixing coefficients,
-    $$\int | f_{XY} - f_X f_Y | < \alpha,$$
-    are bounded by the Dobrushin coefficient
-    $$\int | f_{Y|X}(y|x_1) - f_{Y|X}(y|x_2) |dy < \alpha$$ for all $x_1$, $x_2$.
-\end{lem}
-
-\begin{proof}
-    We rewrite the alpha-mixing assertion as $$\int { \int |f_Y|X(y|x) - f_Y(y)| dy } f_X dx.$$
-
-    We claim that the Dobrushin coefficient implies $$\int | f_Y|X(y|x_1) - f_Y(y) |dy < \alpha$$ for all $x_1$. This is shown as follows:
-    \begin{align*}
-        \int | f_{Y|X}(y|x) - f_Y(y) |dy 
-        &= \int \left|  \int [f_{Y|X}(y|x_1) - f_{Y|X}(y|x)] f_X(x)dx \right|dy \\
-        &< \int   \int  | f_{Y|X}(y|x_1) - f_{Y|X}(y|x)] | dy f_X(x)dx \\
-        &< \int \alpha f_X(x)dx \\
-        &=\alpha
-    \end{align*}
-
-    We then have the desired result.
-\end{proof}
-
-\end{document}

### 2024-06-10 paper/ms.tex
Kevin: convergence.tex, etc

--- 2024-06-08 paper/ms.tex
@@
 
-Consider the second stage of IFAD (Algorithm \ref{alg:ifad}) where one stops if $||\nabla_\theta \hat\ell^\alpha(\theta_n)|| \leq (1+\sigma) \epsilon$, where $\sigma \geq \frac{4 \Gamma}{(1-\beta)}$.
+Consider the second stage of IFAD (Algorithm \ref{alg:ifad}) where one stops if $||\nabla_\theta \hat\ell^\alpha(\theta_m)|| \leq (1+\sigma) \epsilon$, where $\sigma \geq \frac{4 \Gamma}{(1-\beta)}$, for some $\beta \in (0,1)$.
 Assume $-\ell$ is strongly convex and smooth, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$.
-Choose the learning rate $\eta$ and error tolerance $\epsilon$ such that $\eta \leq \frac{c(1-\beta)}{2\Gamma}$ for some $\beta \in (0,1)$.
-Then, for sufficiently large $\alpha$ and $J$ to ensure the score estimate is a $1/2$-approximation, the second stage of IFAD converges linearly to the MLE with probability at least $1-\delta$:
+Choose the learning rate $\eta$ such that $\eta \leq \frac{c(1-\beta)}{2\Gamma}$.
+Then, for sufficiently large $\alpha$ and $J$ to ensure the score estimate is a $1/2$-approximation and the minimum eigenvalue of $H$ is greater than some $c > 0$ for all $m$ with probability at least $1-\delta$, the second stage of IFAD converges linearly to the MLE:
 \vspace*{-2mm}
 $$
-\ell(\theta^*) - \ell(\theta_{n+1}) \leq \left(1-\eta\beta\frac{8\gamma}{9c}\right)(\ell(\theta^*)-\ell(\theta_n)).
+\ell(\theta^*) - \ell(\theta_{m+1}) \leq \left(1-\eta\beta\frac{8\gamma}{9c}\right)(\ell(\theta^*)-\ell(\theta_m)).
 $$
@@
 		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_m).$ \newline
-		\hspace*{4mm} Obtain $g(\theta_m) = \nabla_{\theta_m} (-\hat\loglik(\theta_m))$, $H(\theta_m)$ s.t. $\lambda_{\min}(H) \geq c$. \newline
+		\hspace*{4mm} Obtain $g(\theta_m) = \nabla_{\theta_m} (-\hat\loglik(\theta_m))$, $H(\theta_m)$ s.t. $\lambda_{\min}(H(\theta_m)) \geq c$. \newline
 		\hspace*{4mm} Update $\theta_{m+1} := \theta_m - \eta (H(\theta_m))^{-1} g(\theta_m)$, $m:=m+1.$ \newline
@@
 
+\arxiv{
+\appendix
+
+\section{MOP Functional Forms}
+\input{appendix/functional}
+
+\section{Optimization Convergence Analysis}
+\input{appendix/convergence}
+
+\section{Weighted Particle Filters that Target the Posterior}
+\input{appendix/targeting}
+
+\section{Consistency of MOP-$1$}
+\input{appendix/consistency}
+
+\section{Bias-Variance Analysis}
+\input{appendix/biasvar}
+}{
+}
+
 \end{document}

### 2024-06-10 paper/si.tex
Updates from Overleaf
GH edits
Kevin: convergence.tex, etc

@@
-\documentclass[9pt,twoside,lineno]{pnas-new}
+\documentclass[9pt,twoside,lineno]{pnas-new-si}
 % Use the lineno option to display guide line numbers if required.
+\usepackage{xr}
+\makeatletter
+\newcommand*{\addFileDependency}[1]{% argument=file name and extension
+  \typeout{(#1)}
+  \@addtofilelist{#1}
+  \IfFileExists{#1}{}{\typeout{No file #1.}}
+}
+\makeatother
 
+\newcommand*{\myexternaldocument}[1]{%
+    \externaldocument{#1}%
+    \addFileDependency{#1.tex}%
+    \addFileDependency{#1.aux}%
+}
+\externaldocument{ms}
 \templatetype{pnassupportinginfo}
+\input{paper/macros}
 
@@
 
-\maketitle
-
 %% Comment out or remove this line before generating final copy for submission; this will also remove the warning re: "Consecutive odd pages found".
 %\instructionspage
+
+\maketitle
 
@@
 
-\subsection*{Subhead}
+\section{Heading}
 
-\begin{defn}[Targeting]
-    \label{defn:targeting}
-    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any measurable and bounded functional $h$, $E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}$ 
-    A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if for any measurable and bounded functional $h$
-    ${\sum_{j=1}^J h(X^J_j) w^J_j}/{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X))$ as $J\to\infty$
-\end{defn}
+\ref{alg:mop}
 
-\begin{lem}[Change of Particle Measure]
-    \label{lem:change-measure-proper-weights}
-    Assume $\{(\tilde X_j^J,U_j^J)\}_{j=1}^J$ targets $f_X$, and that $\{(Y_j^J,V_j^J)\}_{j=1}^J$ is a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is drawn w.p.
-$\pi^J_j$.
-Write, where $a$ is the ancestor function,
-    $(Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big)
-    $. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j)\}_{j=1}^J$ targets $f_X$.
-\end{lem}
-\begin{lem}[Particle Marginals]
-    \label{lem:marginal-proper-weights}
-    Suppose that $\{(\tilde X_j^J,U_j^J)\}_{j=1}^J$ targets $f_X$.
-Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$, where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
-Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}_{j=1}^J$ targets $f_Z$.
-\end{lem}
-\begin{lem}[Particle Posteriors]
-    \label{lem:posterior-proper-weights}
-    Suppose that $\{(X_j^J,u_j^J)\}_{j=1}^J$ targets $f_X$, and write $(X^{\prime J}_j,u^{\prime J}_j) = \big(X_j^J,u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$.
-Then, if $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,u^{\prime J}_j)\}_{j=1}^J$ targets $f_{X|Z}(\cdot | z^*)$.
-\end{lem}
-
-Whereas if we use the after-resampling conditional likelihoods $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$, the numerator of $L_n^{B,\theta,\alpha}$ is
-\begin{equation}
-    \sum_{j=1}^J g_{n,j}^\theta w_{n, j}^{P, \theta} = L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
-\end{equation}
-Using Lemma \ref{lem:change-measure-proper-weights}, we resample the corrected weights $w^{P,\theta}_{n,j}{g^\theta_{n,j}}/{g^\phi_{n,j}} $ according to probabilities ${g^\phi_{n,j}}/{L_n^\phi}$ to see this is properly estimated by $L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},$
-from which we obtain the estimate $L_n^{B,\theta,\alpha}$. The strong consistency then follows as the likelihood is the product of the conditional likelihoods.
-
-\begin{thm}[Consistency of MOP-$1$ Gradient Estimate]
-    The gradient estimate of MOP-$\alpha$ when $\alpha=1$, $\theta=\phi$ is strongly consistent for the score.
-$\nabla_\theta \hat\ell_J^1(\theta) \stackrel{a.s.}{\to} \nabla_\theta \ell(\theta)$ as $J \to \infty$.
-    \label{thm:mop-grad-consistency}
-\end{thm}
-\begin{proof}
-    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point we wish to evaluate the gradient at.
-The sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$, by Assumption \ref{assump:local-bounded-derivative}.
-We claim that the sequence is also uniformly equicontinuous.
-To see this, by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J^1(\theta)(\omega)|_{\theta=\theta'}$ is also bounded by $H'$ for almost every $\omega\in \Omega$ and every $\theta'\in \Theta$.
-A set of functions with derivatives bounded by the same constant is uniformly Lipschitz, and therefore uniformly equicontinuous.
-So the sequence $(\nabla_\theta \hat\lik_J^1(\theta)( \omega))_{J \in \mathbb{N}}$ is uniformly equicontinuous for almost every $\omega \in \Omega$.
-    Explicitly, for almost every $\omega \in \Omega$ and every $\epsilon>0$, there exists some $\delta(\omega)>0$ such that for every $||\theta - \theta'||_{\infty}<\delta$ and every $J \in \mathbb{N}$ we have that
-    $$||\nabla_\theta \hat\lik_J^1(\theta)(\omega)-\nabla_\theta \hat\lik_J^1(\theta')(\omega)||_\infty < \epsilon.$$
-
-    Then, by Arzela-Ascoli, there exists a uniformly convergent subsequence.
-We claim that there is only one subsequential limit.
-We know that as the gradient estimate is bounded by Assumption \ref{assump:local-bounded-derivative}, we can treat the gradient estimate at $\theta$ as a bounded functional of the particles.
-So by Theorem \ref{thm:mop-targeting} the sequence $(\nabla_\theta \hat\lik_J^1(\theta, \omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$, and there is therefore only one subsequential limit.
-The sequence therefore converges uniformly to its limit $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega).$ Therefore, with uniform convergence for the derivatives established, we can swap the limit and derivative and obtain, in conjunction with the strong consistency $\hat{\lik}_J^1(\theta, \omega) \stackrel{a.s.}{\to} \lik(\theta)$ in Theorem \ref{thm:mop-targeting}, that for almost every $\omega \in \Omega$,
-    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lik(\theta).$$
-    The result then follows by the continuous mapping theorem. 
-\end{proof}
-
-\section*{Heading}
 \subsection*{Subhead}

### 2024-06-10 paper/appendix/functional.tex

\begin{lem}
    \label{lem:mop-1-formula}
    Write $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$. Consider the case where we usethe after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$, When $\alpha=1$,
    \begin{equation}
        \nabla_\theta \hat{\ell}^1(\theta) 
        = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
    \end{equation}
    yielding the estimator of \cite{poyiadjis11, scibior21} with the bootstrap filter.
\end{lem}

\begin{proof}

Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. We then have a nice telescoping product property for the after-resampling likelihood estimate:
\begin{equation*}
    \hat{\lik}^1(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}}= \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\theta}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
\end{equation*}
where the third equality follows from the choice of $\alpha=1$, and the fourth equality is the resulting telescoping property.

Then, the log-derivative identity lets us decompose the score estimate as
$$\nabla_\theta \hat\ell^1(\theta) = \frac{\nabla_\theta \hat\lik^1(\theta)}{\hat\lik^1(\theta)} = \frac{\nabla_\theta\left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi}{\prod_{n=1}^N L_n^\phi} =  \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}.$$

So we see that the derivative of the log-likelihood estimate is
\begin{equation*}
    \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},
\end{equation*}
which we decompose as follows.

Observe that as $\alpha=1$,

$$w_{n,j}^{P,\theta} = w_{n-1,j}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}},$$
where we use the $(\cdot)^A$ superscript to denote the ancestral trajectory of the $j$-th prediction or filtering particle at timestep $n$. Note that this quantity is the cumulative product of measurement density ratios over the ancestral trajectory of the $j$-th prediction particle at timestep $n$.

We can then use the log-derivative identity again, yielding the following expression for the gradient of the log-weights as the sum of the log measurement densities over the ancestral trajectory:
$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,j}^{A,P,\theta} - \log g_{i,j}^{A,P,\phi}\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,j}^{A,P,\theta}.$$

But this is equal to the gradient of the logarithm of the conditional density of the observed measurements given the ancestral trajectory of the $j$-th prediction particle up to timestep $n$:
$$  \nabla_\theta \sum_{n=1}^N \log g_{n,j}^{A,\theta} = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,j}^{A,P,\theta}\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,j}^{A, P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$

Multiplying both sides of the expression by $w_{N,j}^{P,\theta} $ yields an expression for the gradient of the weights at timestep $N$:
$$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{n=1}^N \nabla_\theta \log g_{n,j}^{A,P,\theta} = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$

Substituting the above identity into the log-likelihood decomposition obtained earlier yields
\begin{equation*}
    \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right),
\end{equation*}
and finally, observing that when $\theta=\phi$ we have that $w_{N,j}^{F,\theta}=1$, that 
\begin{equation*}
    \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
\end{equation*}
yielding the gradient estimators of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter. 
\end{proof}

Note that one would believe that the variance blows up the moment $\theta\neq\phi$. This is true, as 
\begin{equation*}
    \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right),
\end{equation*}
and when $\theta\neq\phi$, $w_{N,k_j}^{P,\theta} = O(c^N)$. When $\theta=\phi$, this is a special case of the \cite{poyiadjis11} estimator, which has $O(N^4)$ variance by a property of functionals of the particle filter by \cite{delMoral03}.

\begin{lem}
\label{lem:mop-0-formula}

 Write $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$. Consider the case where we use the after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$. When $\alpha=0$,
    \begin{equation}
        \nabla_\theta \hat\ell^0(\theta) 
        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
    \end{equation}
    yielding \cite{naesseth18} on the bootstrap filter. 
\end{lem}

\begin{proof}
First, write $$s_{n,j} = \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}$$
as shorthand for the measurement density ratios.

One can then observe that when $\alpha=0,$ the likelihood estimate becomes
\begin{equation}
    \hat{\lik}^0(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
\end{equation}
Note that we lose the nice telescoping property observed in the MOP-$1$ case, but this expression still yields something useful. This is because its gradient when $\theta=\phi$ is therefore 
\begin{align*}
    \nabla_\theta \hat{\ell}^0(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)} \\
    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta s_{n,j}}{\sum_{j=1}^J s_{n,j}} \\
    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)}{f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \phi}; \phi)} \\
    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
\end{align*}
where we use the log-derivative trick in the second equality, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$ in the fourth equality, and use the log-derivative trick again while noting that $\theta=\phi$ in the fifth equality. This yields the desired result.
\end{proof}

Combining Lemmas \ref{lem:mop-1-formula} and \ref{lem:mop-0-formula} yields the proof of Theorem \ref{thm:mop-functional-forms}.

### 2024-06-10 paper/appendix/feynman.tex
Updates from Overleaf

In this section, we introduce the Feynman-Kac convention of \cite{delMoral04} that has since become commonplace \cite{karjalainen23} for the analysis of the particle filter. The mathematical formalization and notation introduced here will be adopted

### 2024-06-10 paper/appendix/targeting.tex
Updates from Overleaf

--- 2024-05-01 mop/draft.tex
@@
-\documentclass{article}
-\usepackage[margin=2cm]{geometry}
-\usepackage{natbib}
-\usepackage{hyperref}
-\usepackage{enumitem}
-\usepackage{tikz}
-\usetikzlibrary{arrows.meta}
-\usetikzlibrary{positioning}
-
-\setlist{nosep} % to reduce space in lists
-
-\setlength{\parskip}{0.5em}
-\setlength{\parindent}{0in}
-
-\input{macros}
-
-\newcommand{\off}{\operatorname{off}}
-\newcommand{\on}{\operatorname{on}}
-
-\title{Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation}
-\author{Kevin Tan, Giles J. Hooker, Edward L. Ionides}
-\date{}
-
-\comments=1
-
-\begin{document}
-\maketitle
-\begin{abstract}
-
-Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods.
-Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques due to (1) the requirement for simulation-based inference that does not require access to the system's transition probabilities and (2) the issue that widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters.
-We propose a new representation that embeds existing methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff.
-Further, we develop optimization algorithms suited to the Monte Carlo properties of the derivative estimate, including a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
-Promising numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from iterated filtering can beat current state-of-the-art methods on a challenging scientific benchmark problem.
-\end{abstract}
-
-\section{Introduction}
-
-Maximum-likelihood estimation for inference in partially observed stochastic processes, also known as, continuous-state continuous-time hidden Markov models (HMMs), partially-observed Markov processes (POMPs), or partially-observed stochastic nonlinear dynamical systems, is a challenging problem that faces several challenges:
-\begin{enumerate}
-    \item Intractable likelihood functions, bypassed with the approaches of simulated likelihood and likelihood-free inference. 
-    \item The desire for simulation-based inference without access to the transition density of the latent Markov process, with various methods created by \cite{welch2009abc, wood2010sl, doucet2010pmcmc, ionides08, ionides15}.
-    \item Accurate parameter estimation in the presence of significant Monte-Carlo noise in the likelihood estimate.
-\end{enumerate}
-
-Many approaches to inference in these scenarios (such as the EM algorithm, the various Kalman filter variants, and MCMC) either struggle with intractable likelihood functions when the models are complex, or assume access to the probability density of next states given the current state.
-This is a problem in some critical applications, such as disease modeling, where the models are complex enough such that even obtaining the transition density is an intractable problem.
-However, the particle filter with the bootstrap proposal, a popular method for solving the filtering problem in partially-observed dynamical systems, provides an unbiased estimate of the likelihood (\cite{delmoral2004feynman}) without requiring evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
-
-This is the backbone behind the only simulation-based full-information maximum likelihood method available in the literature thus far, the improved iterated filtering algorithm of \citet{ionides15}.
-\citet{ionides15} utilize the bootstrap filter for perform maximum likelihood estimation with an iterated perturbed Bayes map, successfully tackling challenging problems in epidemiology that available Bayesian methodology fails to solve.
-
-Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small.
-For example, though in practice the algorithm of \citet{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood.
-\citet{Ionides_mcap} attempt to address the third problem, albeit with a computationally-expensive approach.
-
-\subsection{Automatic Differentiation for Particle Filters}
-
-Recent advances in automatic differentiation (AD) for particle filters (\cite{blei2018vsmc, jon2018diffpf, corenflos21, scibior2021dpf, doucet2022particlebased}) have drawn attention to AD as an alternative tool for inference in partially observed stochastic processes with the particle filter.
-However, these existing approaches either are not yet compatible with the desire for simulation-based inference, or are computationally expensive.
-
-\citet{blei2018vsmc} and \citet{jon2018diffpf} drop high-variance resampling terms from the gradient, leading to bias that does not vanish asymptotically according to \citet{corenflos21}.
-Though the approach of \citet{corenflos21} is promising, it is computationally expensive.
-\citet{doucet2022particlebased} provide another promising fixed-lag smoothing-based approach that is similar in spirit to our method (MOP-$\alpha$), but requires either (1) access to the process model transition densities or (2) the special case where the transition density factors into a policy that selects actions (in the reinforcement learning sense) and a deterministic model that maps states and actions to next-timestep states.
-
-Finally, \citet{scibior2021dpf} demonstrate that the estimators derived in \citet{poyiadjis11} can be attained with AD, albeit with a simple tweak to the particle filter that does not modify its likelihood estimates.
-Though their method is not fully compatible with simulation-based inference, we show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.
-
-\subsection{Our Contributions}
-
-Previous research on AD for particle filters has struggled with various issues: 
-\begin{enumerate}
-    \item Bias, if the discontinuous nature of particle resampling is ignored.
-    \item High Monte Carlo variance, if continuity corrections lead to numerical instability.
-    \item High computational cost, for algorithms which involve pairwise interactions between particles, marginalization, smoothing or optimal transport.
-    \item Reduced applicability for algorithms which lose or fail to take advantage of the simulation-based capabilities of the bootstrap filter.
-\end{enumerate}
-
-We develop a new approach to statistical inference via AD for particle filters which addresses these concerns.
-We develop a new theoretical framework that encompasses existing methods, and addresses the seemingly incompatible paradox of differentiating through a Monte-Carlo algorithm with discontinuous resampling.
-
-This is done through the construction of a smooth extension to the particle filter, as well as a \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) family of algorithms that (informally) encompasses the special case of AD of a vanilla PF (recovering the gradient estimator of \citet{blei2018vsmc}) when $\alpha=0$, and the DPF algorithm from \citet{scibior2021dpf} (that recovers the gradient estimator from \citet{poyiadjis11}) when $\alpha=1$.
-When $\alpha \in (0,1)$, this provides opportunities for optimizing a bias/variance tradeoff.
-
-We show these off-policy particle filters successfully target the posterior, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties.
-To cap off the theoretical results, we also derive a linear convergence rate for gradient methods on the particle filter in the presence of strong convexity, providing support for numerical optimization with AD for the particle filter.
-
-In practice, we propose a hybrid algorithm that we call \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts gradient ascent (or a similar iterative first or second-order gradient method) with a coarse solution obtained from a few rounds of iterated filtering.
-While this requires a differentiable simulator compatible with the reparametrization trick (as previous literature such as \citet{corenflos21} also require), future work will deal with workarounds inspired by our framework that use likelihood ratios of transition densities in place of this.
-Promising numerical results indicate that IFAD beats IF2 (and by the transitive property, IF1, the Liu-West filter, and particle MCMC) on a challenging scientific benchmark problem.
-
-While the Monte-Carlo Adjusted Profile (MCAP) from \citet{Ionides_mcap} provides a profile likelihood confidence interval that takes into account both Monte Carlo profile error and statistical uncertainty in the likelihood, it is computationally inefficient.
-\kevin{TODO} We provide a method for incorporating the MCAP methodology (with the poor man's likelihood profile) in our IFAD algorithm that provides a more computationally efficient profile likelihood confidence interval.
-\kevin{Quantile regression for estimating the max/the envelope of the likelihood evaluations from AD, then fit MCAP/MQLE/MSLE?}
-
-Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual} that we call \texttt{pompy}.
-The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{jax} Python package from \citet{jax2018github}.
-
-\kevin{Terminology: Plug-and-play (clearly definable, but not well-known, define at start of paper and don't use in title/abstract), likelihood-free maximum likelihood (seemingly an oxymoron, rationale being there is no need to evaluate density, bayesians call likelihood densities, plus we only approximate the likelihood and don't really have it), simulation-based inference (loaded term)}
-
-\subsection{Significance Statement}
-Many scientific models involve highly nonlinear stochastic dynamical systems which can be observed only via noisy and incomplete measurements.
-Under the Markov assumption on system dynamics, previous work has provided methods of performing inference for these models.
-In particular, prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics.
-We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
-Our new method outperforms previous approaches on a challenging problem in epidemiology.
-This is implemented in a precursor to a Python counterpart to the popular \texttt{pomp} package by \citet{king16, king2017pompmanual} that includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation.
-
-\kevin{TODO: (1) Fix beta smoothness to d(n) at no more than O(n), (2) add story to say that no one spotted that it could have been plug-and-play because of e.g. reasons why they would stumble}
-
-\section{Problem Setup}
-
-Consider an unobserved Markov process $\{X_t, t \geq t_0\}$, and observations $Y_1,...,Y_N$ at timesteps $t_1,..., t_N$.
-The process is parameterized by an unknown parameter $\theta \in \Theta$, where $X \in \gX, Y \in \gY,$ and finally $\gX \subseteq \R^d, \Theta \subseteq \R^p$.
-By a similar decomposition to that in \citet{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
-$$f_{X_{0: N}, Y_{1: N}}\left(x_{0: N}, y_{1: N} ; \theta\right)=f_{X_0}\left(x_0 ; \theta\right) \prod_{n=1}^N f_{X_n \mid X_{n-1}}\left(x_n \mid x_{n-1} ; \theta\right) f_{Y_n \mid X_n}\left(y_n \mid x_n ; \theta\right).$$
-
-We call $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ the process model, writing $\texttt{process}\left(x_n, \theta\right)$ for the simulator corresponding to the above process model, $f_{Y_n|X_n}\left(y_n \mid x_n, \theta\right)$ the measurement model, and will write $y_n^*$ for the actual values of the observations that were observed.
-
-The above are all defined on a single probability space $(\Omega, \Sigma, \prob),$ where $\omega \in \Omega$ denotes an element of the sample space that can be thought of as a random seed.
-For notational simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0.$
-
-\paragraph{Seed-Fixing, Coupling, and the Method of Common Random Numbers, Coupling:}
-
-In the rest of this paper, we will identify the seed with an element of the sample space.
-We do so because the seed determines the entire sequence of pseudo-random numbers as generated by a computer, to maintain an algorithmic and practical interpretation of our theory.
-When two algorithms share the same seed, this corresponds to the method of common random numbers.
-
-\section{A Smooth Extension of the Particle Filter}
-
-%\paragraph{Motivation:} 
-%The gradient of the log-likelihood can be written as a vector $\nabla_\theta \ell(\theta)$ such that 
-%\begin{equation}
-    %\lim_{v \to 0} \frac{|\ell(\theta+v) - \ell(\theta) - v^T \nabla_\theta \ell(\theta)|}{\|v\|} = 0.
-%\end{equation}
-
-%We therefore need to evaluate the likelihood at a point $\theta$ and a nearby point $\theta + v$ to estimate the gradient with finite differences.
-But in the presence of Monte Carlo noise in evaluating the likelihood, this is difficult, unless (1) one ensures one uses the same seed for each evaluation in the neighborhood, and (2) chooses a step-size small enough that changing the the parameter does not change the resampling indices.
-
-\paragraph{Intuition:} 
-Our theoretical framework for automatic differentiation for the particle filter hinges on the following observation.
-Instead of seeking to differentiate $\ell(\theta)$ directly, we instead focus on differentiating through a suitable fixed-seed reweighting of a bootstrap filter.
-The intuition is as follows.
-Consider a particle filter likelihood estimate where the system (i.e.
-\textbf{both the state transitions and resampling}) evolves under a parameter $\phi \in \Theta$ and the seed, or sample space element, is fixed at $\omega \in \Omega.$ Conditional on $\omega$ and $\phi,$ one can reweight the likelihood estimate to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$
-
-Differentiating this reweighted likelihood estimate with respect to $\theta$ then amounts to differentiating through the reweighting likelihood ratios, which is differentiable if the densities used for the reweighting are differentiable.
-We call this simulation-and-reweighting scheme an \textbf{off-policy} likelihood estimate, as it is similar in spirit to the off-policy methods found in the reinforcement learning literature.
-
-The above intuition is formalized below.
-
-\kevin{Method of Common Random Numbers, Coupling}
-
-\subsection{Doubly Off-Policy Particle Filters}
-
-\begin{defn}[Seed-Fixing]
-\label{defn:seed-fixing}
-Consider a probability space $(\Omega, \Sigma, \mu)$ governing the particles and observations.
-Noting that the particle filter is unbiased for the likelihood when computing the expectation of the normalizing factor under the posterior $f_{X_{0:N|Y_{1:N}}}\left(x_{0: N} \mid y_{1: N}; \theta\right)$, we can write the likelihood as
-$$
-\begin{aligned}
-\mathcal{L}(\theta) & =\int_{\Omega} \mathcal{L}(\theta, \omega, J) d \mu(\omega) \\
-& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right)\right) f_{X_{0:N}, Y_{1:N}}\left(x_{0: N}^P(\omega) , y_{1: N}^*, \theta\right) d \mu(\omega) \\
-& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right) f_{X_n|X_{n-1}}(x_{n,j}^P(\omega)|x_{n-1,j}^F(\omega), \theta)\right) f_{X_{0:N}| Y_{1:N}}\left(x_{n, j}^P(\omega) \mid y_n^*, \theta\right) d \mu(\omega)
-\end{aligned}
-$$
-where $\mathcal{L}(\theta, \omega, J)$ is the Monte-Carlo estimate of the likelihood given the realization of $\mathrm{J}$ particles corresponding to $\omega \in \Omega$.
-For simplicity, we resample every timestep.
-We call $\omega$ a seed, and the deterministic quantity $\mathcal{L}(\theta, \omega, J)$ the seed-fixed estimate.
-\end{defn}
-
-Armed with this seed-fixed likelihood estimate, we modify a particle filter likelihood estimate where the system evolves under a parameter $\phi \in \Theta$ and seed $\omega \in \Omega$ to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$
-
-\begin{defn}[Doubly Off-Policy Particle Filter Likelihood]
-\label{defn:doubly-off-policy}
-Let $\lik(\theta, \phi, \omega, J)$ denote a Monte-Carlo estimate of the likelihood evaluated at $\theta \in \Theta$ with $J$ particles, but where the system evolves according to another $\phi \in \Theta$ and conditional on a fixed seed $\omega \in \Omega$.
-This can be computed with a suitable reweighting of the particles, which we provide the derivation for later.
-We then have
-\begin{equation} \label{eq:loglik2}
-\lik(\theta)=\E[\lik(\theta,\phi,\Omega, J)]=\int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega),
-\end{equation}
-where we define
-\begin{align}
-    \lik(\theta,\phi,\omega, J) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J &\left(f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) \cdot 
-    f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}(\omega) | x_{n-1,j}^{F,\phi}, \phi) \right.\\
-    &\cdot
-    \left.
-    \frac{f_{Y_n|X_n}(y^*_n|x_{n,j}^{P,\phi}(\omega),\theta)}{f_{Y_n|X_n}(y^*_n|x_{n,j}^{P,\phi}(\omega),\phi)}
-    \cdot \frac{f_{X_n|X_{n-1}}(x_{n,j}^{P,\phi}(\omega)|x_{n-1,j}^{F,\phi}(\omega),\theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P,\phi}(\omega)|x_{n-1,j}^{F,\phi}(\omega),\phi)}\right).
-\end{align}
-\end{defn}
-
-Lastly, we require the following regularity conditions.
-
-\begin{aspt}[Continuity of the Likelihood] $\ell(\theta)$ proper is continuous in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.
-\end{aspt}
-
-\begin{aspt}[Bounded Process Model]
-\label{assump:bounded-process}
-    There exist $\underbar{M}, \bar{M}$ such that $0 < \underbar{M} \leq f_{X_n|X_{n-1}}(x_n | x_{n-1}) \leq \bar{M} < \infty$. 
-\end{aspt}
-
-\begin{aspt}[Bounded Measurement Model]
-\label{assump:bounded-measurement}
-There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G} < \infty$ and $\infty > G' \geq G'(\theta)>0$ such that $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty <\leq G'(\theta) \leq G'$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
-\end{aspt}
-
-\begin{aspt}[Locally Bounded Derivatives]
-\label{assump:local-bounded-derivative}
-Let $\mathcal{M}$ be an open subset of $\Omega$.
-There exists some function $G(\theta)$ and constants $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$, $H=\sup _{\theta \in \mathcal{N}} H(\theta)<\infty$ such that
-    $$
-    \|\nabla_\theta \ell(\theta, \phi, \omega, J)\|_\infty \leq G(\theta) \leq G<\infty, \;\;
-    \|\nabla^2_\theta \ell(\theta, \phi, \omega, J)\|_\infty \leq G(\theta)^2 \leq G^2<\infty
-    $$
-    for every $\phi \in \mathcal{M}, \theta \in \mathcal{N}_J(\phi)$, and almost every $\omega \in \Omega$.
-\end{aspt}
-
-We need the following assumption to ensure that the weighted particle filter with off-policy resampling targets the posterior:
-\begin{aspt}[Bounded And Differentiable Measurement Density Ratios]
-    \label{assump:diff-meas}
-    For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{Y_n|X_n}(y_n^*|x, \theta)}{f_{Y_n|X_n}(y_n^*|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x \in \gX$ and all $\theta\in\gN(\phi)$.
-\end{aspt}
-
-In the following assumption, the first condition is sufficient for DOP, and the second for MOP:
-\begin{aspt}[Bounded And Differentiable Process Model]
-    \label{assump:diff-process}
-    Either one of the following two conditions hold:
-    \begin{enumerate}
-        \item For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{X_n|X_{n-1}}(x'|x, \theta)}{f_{X_n|X_{n-1}}(x'|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x, x' \in \gX$ and all $\theta\in\gN(\phi)$, or
-        \item There exists a differentiable simulator with more than two continuous derivatives for the process model. 
-    \end{enumerate}
-\end{aspt}
-
-We want $\phi\in\Theta$ to be a point in parameter space around which $\lik(\theta,\phi,\omega, J)$ is differentiable with respect to $\theta$ and does not have excessive Monte~Carlo variance.
-
-We require many of these assumptions to establish the following proposition, the proof of which is a straightforward consequence of what we assumed.
-
-\begin{prop}[Smooth Neighborhood]
-\label{prop:smooth-nbhd}
-There exists an open neighborhood $\mathcal{N}(\phi)$ around any $\phi \in \Theta$ where for all $\theta \in \mathcal{N}(\phi)$ and almost every $\omega \in \Omega$, the Monte Carlo estimate of the likelihood at $\theta \in \mathcal{N}(\phi)$ with the system evolving according to $\phi$ conditional on $\omega$, $\lik(\theta,\phi,\omega, J)$, has two continuous derivatives in $\theta$.
-\end{prop}
-
-\begin{proof}
-    For a fixed $\omega$, the resampling indices for $\lik(\theta,\phi,\omega, J)$ and $\lik(\phi,\phi,\omega, J)$ are the same as the resampling indices in both cases correspond to that yielded by $f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}(\omega);\phi)$, eliminating any discontinuities from resampling.
-
-    By Assumptions \ref{assump:diff-meas} and \ref{assump:diff-process}, for small enough $\mathcal{N}(\phi)$ the density ratios are bounded.
-The likelihood only changes by a factor of the density ratios, which are bounded and possess two continuous derivatives in $\theta$ if either the process and measurement densities used in the calculation do, or if the measurement densities do and one has access to a differentiable simulator with two continuous derivatives in $\theta$.
-This gives us the desired result.
-
-    As a remark, we note that the neighborhoods the sequence is differentiable in are independent of $J$, as the resampling does not change with $\theta$ -- it is dependent on $\phi$. 
-\end{proof}
-
-\subsubsection{Doubly Off-Policy Particle Filter}
-
-Having the system evolve according to $\phi$ but evaluating the likelihood at $\theta$ requires us to reweight the conditional likelihood at each timestep by likelihood ratios corresponding to the process and measurement models.
-This corresponds to making two corrections in the particle weights at each timestep, by multiplying by
-\begin{equation}
-    s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}, \qquad r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)},
-\end{equation}
-as we illustrate in Appendix \ref{app:dop}. We use the notation $x_{n,j}^{P, \phi}$ to emphasize that the particles evolve according to $\phi$.
-
-We claim that this yields a particle filter that targets the posterior and recovers the derivative estimate of \citet{poyiadjis11, scibior2021dpf} when $\theta$ is evaluated at $\phi$.
-The proof can be found in Appendix \ref{app:dop}.
-
-\begin{prop}[Correctness of Doubly Off-Policy Particle Filter Corrections]
-    \label{prop:dop-correctness}
-
-    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$.
-Also let $\theta \in \gN(\phi)$.
-Then, the following corrections in the particle weights,
-    \begin{equation}
-    s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}, \qquad r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)},
-    \end{equation}
-    yield a particle filter that is unbiased for the likelihood, even when $\theta \neq \phi$.
-
-    In particular, when $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{poyiadjis11},
-    \begin{equation}
-        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
-    \end{equation}
-\end{prop}
-
-\paragraph{Intuition:} Having the system evolve according to $\phi$ but evaluating the likelihood at $\theta$ requires us to reweight the conditional likelihood at each timestep by likelihood ratios corresponding to the process and measurement models.
-This corresponds to making two corrections in the particle weights at each timestep, one for the process model and one for the measurement model.
-The measurement correction cannot be avoided, but one can bypass the process correction -- with the reparametrization trick.
-We can do when we assume the simulator is differentiable with the reparametrization trick, as \cite{corenflos21} also assume.
-Doing so then yields an algorithm that only requires a simulator, and does not require the process densities.
-
-\subsection{Measurement Off-Policy Particle Filters}
-
-With the above procedure, we have the entire system evolve according to $\phi$, and we differentiate through the likelihood ratios, which are smooth conditional on the resampling being the same (which in turn happens when $\theta \approx \phi$).
-However, doing so requires access to likelihood ratios, and as such Algorithm \ref{alg:dop} is therefore not compatible with simulation-based inference.
-
-There is a workaround.
-Consider the case where you have access to a differentiable simulator compatible with the reparametrization trick.
-If we have the state transitions evolve according to $\theta,$ but have the resampling according to $\phi,$ we can offload the derivative with respect to the process model to the reparametrization trick, as in \cite{corenflos21}.
-This corresponds to the MOP-$\alpha$ procedure in Algorithm \ref{alg:mop}.
-
-We therefore make the following assumption in this section only:
-\begin{aspt}[Differentiable Simulator]
-    $\texttt{process}(\cdot | X_{t-1}; \theta)$ is differentiable with respect to $\theta,$ and is compatible with the reparametrization trick. 
-\end{aspt}
-
-Algorithm \ref{alg:mop} provides a weighted particle filter that targets the posterior under $\theta$, but when differentiated yields either the low-bias but high-variance gradient estimator of \cite{poyiadjis11}, the high-bias but low-variance estimator of \cite{blei2018vsmc}, or an intermediate, interpolating quantity that can optimize a bias-variance tradeoff.
-This is controlled by a discounting parameter $\alpha$.
-The intuition behind this procedure is as follows.
-
-\paragraph{Intuition:} Broadly, if $\theta\neq\phi,$ one can first run a vanilla particle filter to recover values for the particles with the process model under $\phi.$ One can then run an accordingly reweighted particle filter with the process model under $\theta$, but resampling according to the measurement model under $\phi$, with the same indices as the original vanilla particle filter.
-This requires one to set the same random seed for both particle filters.
-This then yields a slightly-unconventionally weighted particle filter that targets the posterior, but has some very nice gradient properties; as we outline in the remainder of this section.
-
-\paragraph{Special Case:} When $\theta = \phi$, one can simply set the particles under $\phi$ to be copies of the particles under $\theta$.
-Gradients with respect to $\theta$ now treat functions of the particles under $\phi$ as constants, corresponding to the stop-gradient trick in \cite{scibior2021dpf}.
-This therefore provides a theoretical framework for their stop-gradient trick, which we happily use in our numerical experiments for simplicity.
-
-\kevin{Develop theory for mixing-time swapping?
-Detect when IF2 reaches a neighborhood of the MLE, by detecting whether it has reached a local sort of stationary distribution when the perturbations are lower bounded, e.g.
-the autocorrelation between the mean vectors to estimate the mixing time or Gelman-Rubin.
-In practice, just swap when IF2 converges according to the diagnostics.
-Theory doesn't care, you just need to in a basin of attraction that is strongly convex.}
-
-\subsubsection{Discounting Weights}
-
-Observe that at Algorithm \ref{alg:mop}, at each timestep the corrected filtering importance weights $w_{n,j}^{F,\theta}$ optionally propagate to the next timestep for us to maintain the correction.
-We introduce a discounting parameter $\alpha$ that affects the weights, so $w_{n,j}^{P,\theta} := (w_{n-1,j}^{F,\theta})^{\alpha}$.
-Note that no matter what value $\alpha$ takes on, the system evolves according to $\phi$ when we perform the correction to evaluate the likelihood at $\theta$.
-The filter fully targets the posterior when $\theta=\phi$, or $\alpha=1$, as we show in Proposition \ref{prop:dop-correctness}.
-As previously mentioned, the value of $\alpha$ therefore informs a bias-variance tradeoff.
-We elaborate on this below.
-
-\paragraph{Bias-Variance Tradeoff} This discounting serves the purpose of balancing a tradeoff between maintaining memory of each particle's ancestral trajectory (most extreme when $\alpha=1$), or considering only the single-step transition dynamics (when $\alpha=0$).
-This can be thought of as an \textbf{exponentially-weighted moving average}, where $\alpha$ controls the amount of discounting.
-We elaborate on two special cases below.
-
-\kevin{This isn't a smoother, because we'd need future information.}
-
-\paragraph{When $\alpha=1,$} the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\theta}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
-\end{equation}
-as we have a telescoping product. Its derivative when $\theta=\phi$ is then
-\begin{equation}
-    \nabla_\theta \hat{\ell}(\theta) := 
-        \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*),
-\end{equation}
-the average gradient of each surviving particle's cumulative importance weights.
-The filter targets the posterior when $\alpha=1$ even when $\theta \neq \phi,$ and therefore suffers from no asymptotic bias.
-
-\paragraph{When $\alpha=0,$} the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J w_{n,j}^{F,\theta} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})},
-\end{equation}
-and though the telescoping product disappears, this reduces to a series of one-step corrections.
-However, these corrections are insufficient when $\theta \neq \phi,$ and the filter suffers from asymptotic bias unless $\theta=\phi.$ Its gradient when $\theta=\phi$ is therefore
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
-    \end{equation}
-
-the gradient of the vanilla particle filter where the resampling correction is omitted as found in \cite{blei2018vsmc} -- when applied to the bootstrap filter.
-As \cite{scibior2021dpf} remark, this estimator only considers single-step quantities, and is "memoryless" beyond a single step.
-This is in contrast to the case when $\alpha=1,$ as that estimate only considers the surviving particles at time $N$, which is the other extreme where dependencies across time are fully tracked.
-
-\subsubsection{Correctness of MOP-$\alpha$}
-
-\begin{prop}[Correctness of MOP-$\alpha$ (Algorithm \ref{alg:mop})]
-    \label{prop:mop-correctness}
-    When either $\alpha=1$ or $\theta=\phi$, Algorithm \ref{alg:mop} is a particle filter that targets the posterior and is unbiased for the likelihood under $\theta$.
-
-    In particular, when $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter.
-In this case, its gradient when $\alpha=1$ is the same as the DPF estimator of \citet{scibior21} when applied to the bootstrap filter,
-    \begin{equation}
-        \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{0:N}}(y_{1:N}^* | x_{1:N,j}^{A, F,\theta}),
-    \end{equation}
-
-    and when $\alpha=0,$ is the gradient estimator of \cite{blei2018vsmc},
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
-    \end{equation}
-
-    that as \cite{scibior21} show corresponds to differentiating through the fixed-seed likelihood estimate yielded by a vanilla particle filter, dropping resampling terms.
-
-    When $\alpha=1$, the gradient estimate MOP-1 provides is consistent for the score.
-\end{prop}
-
-The proof for the above follows from Appendices A and C, as well as the technical lemmas for properly weighted particle filters.
-
-\kevin{Take derivative of loglik with regard to alpha?}
-\kevin{Talk about regime change the moment you move away from alpha=1, variance going from N2 to N, and bias going from 0 to something, but try to prove that decrease in variance outweights increase in bias.
-}
-\kevin{Create tools for choosing alpha via binary search CV from an averaged alpha=1 target}
-\kevin{suppose at each filtering step, look at gradient calculated at alpha=0, dynamically select alpha during filtering.
-calculate bias and variance online during filtering, and update alpha accordingly.
-}
-
-\kevin{No need to recompute so many particle filters! Just take an array of weight copies, one for each alpha. Gradients for each alpha come free the moment you compute one of them. }
-
-\kevin{Elbow method on variance, since bias is hard to compute? }
-\kevin{Get variance estimate, bias estimate, construct confidence intervals on bias estimate, use that to get unbiased estimate of the MSE. }
-
-\subsubsection{Practical Considerations}
-
-It is only ever practical to evaluate $\theta$ at $\phi$.
-This removes the issue of the likelihood ratios blowing up, and yields an estimate of the likelihood that agrees with the bootstrap filter estimate.
-Its gradient in this case is either the estimate of \citet{poyiadjis11} when $\alpha=1$, a single-timestep estimator that corresponds to naively taking the derivative of the particle filter when $\alpha=0$ as first proposed in \citet{blei2018vsmc}
-
-\paragraph{Bottom Line:} The DOP-$\alpha$ and MOP-$\alpha$ procedures therefore provide us with a unifying framework for reasoning about smooth extensions of weighted particle filters that target the posterior and provide reasonable gradient estimates when differentiated.
-This fixed-seed reweighting alleviates issues that can arise from differentiating through a quantity yielded by ostensibly non-differentiable resampling.
-This is done by either (1) differentiating through smooth likelihood ratios, or (2) differentiating through a differentiable simulator with the reparametrization trick (and through smooth likelihood ratios to adjust for the resampling).
-In particular, MOP-$\alpha$ does not require access to the transition densities, and like the bootstrap filter, only requires access to a (albeit differentiable) simulator.
-To our knowledge, this is the first particle filter gradient method in the literature with this capability.
-
-\section{Practical Maximum-Likelihood Inference}
-
-This still leaves us with the question of designing a reasonably effective procedure for likelihood maximization with these new tools.
-We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that amounts to warm-starting gradient descent or some other reasonable first or second order procedure (using the gradient estimate given by MOP-$\alpha$) with the output of an initial search of IF2.
-
-\paragraph{Summary:} A coarse solution is found by running IF2 with an aggressive random walk standard deviation (learning rate) and a cooling schedule that is \textbf{bounded below}, until the search stalls to an initial "convergence" to what is hopefully a neighborhood of the MLE.
-We can then use gradient descent, or other first or second order gradient methods, to refine this coarse solution.
-
-The initial "convergence" happens fairly quickly in practice.
-In the case of the Dhaka cholera model of \cite{king08}, when a geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations of IF2.
-In comparison, often 100 or 200 iterations of IF2 are used for an initial global search.
-\cite{ionides15}, for example, use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{Lee_haiticholera}.
-
-\paragraph{Rationale:}
-
-IF2 converges quickly to a neighborhood of some reasonably good local optimum in practice, but then struggles to squeeze out the last few log-likelihood units.
-Conversely, gradient descent (and other first or second order gradient methods) can get stuck in saddle points and less desirable local minima when the likelihood is nonconvex.
-Warm-starting gradient methods with IF2 lets us get close enough to the MLE for gradient methods to shine, as in most cases regularity conditions ensure that the likelihood surface is well-behaved enough in this neighborhood that the aforementioned issues with saddle points and local minima are alleviated.
-Combining these two methods in this way lets us enjoy the best of both worlds.
-
-\begin{algorithm}[ht]
-\centering
-	\caption{Iterated Filtering with Automatic Differentiation}
-    \label{alg:ifad}
-	\begin{algorithmic}[1]
-	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $p(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta),$ IF2 cooling schedule $\eta_t$, MOP-$\alpha$ discounting parameter $\alpha$.
-		\STATE Initialize $\theta_0$, $t=0.$
-        \STATE Run IF2 until initial "convergence" under cooling schedule $\eta_t$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_t := \frac{1}{J}\sum_{j=1}^J \Theta_j.$
-		\WHILE{Procedure not converged:}
-		\STATE Draw $\omega \in \Omega$
-		\STATE Perform forward pass by running Algorithm \ref{alg:mop} with discounting parameter $\alpha$ to obtain $\loglik(\theta_t, \theta_t, \omega, J).$
-		\STATE Obtain $g(\theta_t) = \nabla_{\theta_t} (-\loglik(\theta_t, \theta_t, \omega, J))$ via backpropagation, $H(\theta_t)$ such that $\lambda_{\min}(H) \geq c$.
-		\STATE Obtain $\eta$ with line search or an annealing schedule.
-		\STATE Update $\theta_{t+1} := \theta_t - \eta (H(\theta_t))^{-1} g(\theta_t)$, $t:=t+1.$
-		\ENDWHILE
-		\STATE Return $\hat{\theta} := \theta_t.$
-	\end{algorithmic}
-\end{algorithm}
-
-\subsection{Theoretical Guarantee}
-
-We first analyze the performance of gradient descent, Newton's method, or any iterative method that multiplies the particle estimate of the gradient with the inverse of a matrix with minimum eigenvalue bounded below, in the case where the log-likelihood surface is $\gamma$-strongly convex, and where the gradient estimate is a particle approximation of the score (for example, IFAD-1 and MOP-1).
-In this case, these methods enjoy a linear convergence guarantee:
-
-\kevin{I think we can establish the linear convergence for the entire procedure in the same sense as \href{this}{https://arxiv.org/pdf/2311.08442.pdf}, in the sense that if IF2 is run for a sufficiently large but fixed number of iterations, once we arrive in a local strongly convex region, we get linear convergence when we swap to MOP.
-Show by induction that we remain in the strongly convex region (in terms of parameter space and likelihood), eqn 218.
-}
-
-\begin{thm}[Convergence of Gradient Methods]
-\label{thm:convergence}
-
-Consider a variant of IFAD-1 (Algorithm \ref{alg:ifad}) where one stops if the gradient estimate $||g(\theta_n)|| \leq \sigma \epsilon$.
-Assume $-\ell$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$.
-$\sigma = \frac{4 \Gamma}{(1-\beta)}> 4$.
-Define $G$ as in Assumption \ref{assump:local-bounded-derivative}.
-
-Choose $J$ according to the maximum of the $J$ prescribed in Lemmas \ref{lemma:grad_bound} and (if using a Hessian estimate) $\ref{lemma:hess_bound}$ with failure probability $\delta/2$ each.
-Let $H(\theta_n)$ be a matrix possibly dependent on $\theta_n$ with minimum eigenvalue always greater than some $c>0$.
-Choose the learning rate $\eta$ and error tolerance $\epsilon$ such that, where $\beta$ is the Armijo condition hyperparameter,
-\begin{equation}
-    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_n)||.
-\end{equation}
-
-Then the following holds with probability at least $1-\delta$:
-\begin{equation}
-\ell(\theta^*) - \ell(\theta_{n+1}) \leq \left(1-\eta\beta\frac{8\gamma}{9c}\right)(\ell(\theta^*)-\ell(\theta_n))
-\end{equation}
-
-and the algorithm eventually terminates after we obtain $||\nabla_\theta \ell(\theta_n)|| \leq (1+\sigma) \epsilon$.
-\end{thm}
-
-The proof of this result uses the concentration inequalities from \cite{delmoral2011ci}, and leans upon the analysis of \cite{mahoney16}. We defer the proof to Appendix \ref{app:ifad}
-
-\paragraph{Takeaways for IFAD (Algorithm \ref{alg:ifad}):}
-
-The gradient stage of IFAD-1 converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the IF2 stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
-This happens fairly often in practice, for example, when sufficient regularity conditions for local asymptotic normality of the MLE hold.
-
-We conjecture that this applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE.
-IF2 posesses some SGD-like qualities, including what looks like linear convergence to a ball around the MLE with width determined by the random walk standard deviation.
-We will explore this conjecture in future work.
-
-While this theoretical guarantee currently only holds for IFAD-1, as it forms a particle approximation of the score as in \cite{poyiadjis11}, one can also extend this for a similar guarantee if the bias for the gradient estimate given by MOP-$\alpha$ for $\alpha < 1$ is small enough.
-For general $\alpha$, including MOP-0, the analysis then becomes more complicated, as one needs to handle the case of biased gradient descent.
-However, if future measurements $y_{n+1:N}^*$ do not provide a large amount of information on the identification of the current state $x_n$ given past and current measurements $y_{0:n}^*$, as mentioned in \cite{corenflos21}, the bias may be small enough for it to not present too much of a problem.
-
-\subsection{Runtime}
-
-Even with the increased computational cost of automatic differentiation, Algorithm \ref{alg:ifad} is reasonably fast, as by the cheap gradient principle from \cite{kakade2019provably}, a gradient operation takes no more than 6 times the runtime of the particle filter (in practice, this means a gradient step takes about 3.75x as much time as an IF2 iteration).
-As such, unlike methods like the marginal differentiable particle filter in \cite{scibior2021dpf} or the the optimal transport of resampling approach of \cite{corenflos21}, our algorithm has runtime \textbf{linear, and not quadratic}, in the number of particles (while alleviating the $O(N^2)$ variance concern of \cite{poyiadjis11} and \cite{corenflos2021ot} and asymptotic bias concern of \cite{blei2018vsmc} through optimizing $\alpha$ for a bias-variance tradeoff).
-
-In addition, re-implementation of the particle filter, iterated filtering, and Algorithm \ref{alg:ifad} itself in \texttt{JAX} (\cite{jax2018github}) led to considerable speedups.
-This can be attributed to both GPU acceleration and the just-in-time (JIT) compilation functionality that \texttt{JAX} provides.
-When tested on a computer with an Intel i9-13900K processor, 128GB of RAM, and a NVIDIA RTX3090 GPU, each post-compilation particle filtering operation on the Dhaka cholera model from \cite{king08} with 10000 particles took roughly 379 milliseconds (with a standard deviation of 279 microseconds over seven runs), a considerable improvement over the existing performance in the widely used \texttt{pomp} package of \cite{king16}, which took between 6.19 to 6.39 seconds over ten runs on the same machine.
-
-\section{Numerical Experiments}
-
-\subsection{Cholera in Bangladesh}
-
-To compare the performance of IFAD against that of IF2, we fit the very same model first proposed in \cite{king08} that \cite{ionides15} use to demonstrate the capabilities of IF2.
-This is becoming a standard benchmark, as \cite{wood16} and \cite{wycoff2024voronoi} include in their papers.
-This is a stochastic SIR compartmental model with transition and measurement uncertainty, where the population at time $t$, $H(t)$, is divided into the susceptible compartment $S(t)$, infected compartment $I(t)$, and three ($k=3$) recovered compartments $R_1(t), ..., R_k(t)$ denoting varying degrees of cholera immunity.
-We write $M(t)$ for the cholera deaths in each month.
-As in \cite{ionides15}:
-
-\begin{itemize}
-    \item The \textbf{transition dynamics} follow this series of stochastic differential equations:
-
-    \begin{align}
-        d S&=\left(k \epsilon R_k+\delta(S-H)-\lambda(t) S\right) d t+d H-(\sigma S I / H) d B, \\ 
-        d I&=\left(\lambda(t) S-(m+\delta+\gamma) I\right) d t+(\sigma S I / P) d B, \\ 
-        d R_1&=\left(\gamma I-(k \epsilon+\delta) R_1\right) d t, \\ 
-        \vdots \\ 
-        d R_k&=\left(k \epsilon R_{k-1}-(k \epsilon+\delta) R_k\right) d t,
-    \end{align}
-
-    where $B$ represents Brownian motion, $m$ is the rate at which infected individuals succumb to cholera, $\gamma$ is the rate at which infected individuals recover, $1/\epsilon$ is the mean duration of immunity, $\sigma$ represents the standard deviation of perturbations to the force of infection, and $\delta=0.02$ is the death rate.
-
-    \item The \textbf{force of infection}, $\lambda(t)$, is modeled by splines $s_j$
-    \begin{equation}
-        \lambda(t)=\exp \left\{\beta_{\text {trend }}\left(t-t_0\right)+\sum_{j=1}^{6} \beta_j s_j(t)\right\}(I / P) + \exp \left\{\sum_{j=1}^{6} \omega_j s_j(t)\right\},
-    \end{equation}
-    where the $\beta_j$ model seasonality in the force of infection, $\beta_{\text{trend}}$ models the trend in the force of infection, and the $\omega_j$ represent seasonality of a non-human environmental reservoir of disease.
-    \item The \textbf{measurement model} for observed monthly cholera deaths is given by 
-    \begin{equation}
-        Y_n \sim \gN(M_n, \tau^2M_n^2),
-    \end{equation}
-    where $M_n$ is the true number of cholera deaths in that month.
-\end{itemize}
-
-\begin{figure}[ht]
-\centering
-\begin{tikzpicture}[node distance=1cm and 1cm, auto,
->=Latex,every node/.append style={align=center}, 
-int/.style={draw, minimum size=1.6cm}]
-  \node [int] (H) at (0,0) {$H$};
-\node [int] (S) [right=of H] {$S$};
-\node [int] (I) [right=of S] {$I$};
-\node (Mcoord) [right=of I, coordinate] {Mcoord};
-\node [int] (R1) [right=of Mcoord] {$R_1$};
-\node [int] (R2) [right=of R1] {$R_2$};
-\node [int] (R3) [right=of R2] {$R_3$};
-\node [circle, draw] (M) [above=of Mcoord] {$M$};
-\node (deathsS) [above=of S, coordinate] {deathsS};
-\node (deathsI) [above=of I, coordinate] {deathsS};
-\node (deathsR1) [above=of R1, coordinate] {deathsR1};
-\node (deathsR2) [above=of R2, coordinate] {deathsR2};
-\node (deathsR3) [above=of R3, coordinate] {deathsR3};
-\node (belowR3) [below=of R3, coordinate] {belowR3};
-\node (belowS) [below=of S, coordinate] {belowS};
-\path[->]
-    (H) edge node {$dH$} (S)
-    (S) edge node {$\lambda(t)$}  (I)
-    (I) edge node {$\gamma$}  (R1)
-    (R1) edge node {$k\epsilon$}  (R2)
-    (R2) edge node {$k\epsilon$}  (R3)
-    (R3) edge node {} (belowR3) 
-    (belowR3) edge node {$k\epsilon$} (belowS) 
-    (belowS) edge node {} (S)
-    (I) edge node {$m$}  (M)
-    (S) edge node {$\delta$}  (deathsS)
-    (I) edge node {$\delta$}  (deathsI)
-    (R1) edge node {$\delta$}  (deathsR1)
-    (R2) edge node {$\delta$}  (deathsR2)
-    (R3) edge node {$\delta$}  (deathsR3);
-\end{tikzpicture}
-\caption{Illustration of SIR model from \cite{king08}.}
-\label{fig:sir}
-\end{figure}
-
-Figure \ref{fig:sir} depicts the above compartmental model that we use to benchmark the performance of IFAD.
-We present comparisons of IFAD with $\alpha=0$ (IFAD-0) and $\alpha=1$ (IFAD-1) against both (1) the results from \cite{ionides15}, and (2) our own implementation of IF2 in \texttt{JAX}.
-
-\paragraph{Hyperparameters:} While performing an initial exploratory analysis, we found that alternative hyperparameters for IF2 (random walk standard deviation of $\sigma=0.02$, geometrically decreasing cooling rate exponent $a=0.95$) performed better than the settings \cite{ionides15} chose, and chose to use these for our implementation to achieve a fairer comparison.
-Our implementations of IF2 and IFAD used 10000 particles, in line with \cite{ionides15}.
-
-To prevent exploding gradients given the high variance in the MOP-1 estimator, we scale the gradient to have norm 1 before performing vanilla gradient descent, and use a linear cooling schedule of $0.2$ to $ 0.0001$.
-We chose to use vanilla gradient descent for simplicity, and to demonstrate that minimal tuning was required to have IFAD perform well.
-
-% The 60 more iterations of IF2 yielded largely not a huge improvement over the initial coarse solution of 40 iterations.
-This demonstrates what we have mentioned earlier: that although IF2 manages to quickly find a decent coarse solution, it struggles to squeeze out those last few units of log-likelihood improvement.
-The 60 more iterations of gradient descent, however, successfully refines this solution, getting much closer to the MLE on average.
-
-\subsubsection{Results} 
-We benchmarked IFAD against IF2 on a challenging global search problem.
-100 initial starting parameter vectors were drawn uniformly from the same wide bounding box used in \cite{ionides15}, and we performed 100 searches each of IF2, IFAD-0, IFAD-1, MOP-0, and MOP-1 initialized from these 100 starting parameter vectors.
-We summarize our findings below.
-
-\begin{table}[h!]
-\centering
-\begin{tabular}{||c c c||} 
- \hline
- Method & Best Log-Likelihood & Rank \\ [0.5ex] 
- \hline\hline
-     IFAD-0 & -3747.87 & 1\\
-     IFAD-1 & -3748.53 & 2\\
-     IF2 (Ours) & -3755.91 & 3\\
-     IF2 (Warm Start) & -3759.60 & 4 \\
-     IF2 2015 & -3768.63 & 5\\ 
-     MOP-0 Alone & -3795.74 & 6\\
-     MOP-1 Alone & -3797.38 & 7\\
- \hline
-\end{tabular}
-\caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone.
-IFAD performs the best among all methods.
-Our implementation of IF2 outperforms that of \cite{ionides15}, but still ultimately underperforms IFAD.
-Both IFAD-0 and IFAD-1 manage to find the MLE, either matching or beating the highest log-likelihood previously found in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.}
-\label{table:mle}
-\end{table}
-
-\paragraph{IFAD Successfully Finds the MLE:} Both IFAD-0 and IFAD-1 successfully find the MLE, as seen in Table \ref{table:mle}.
-Previously, the MLE at a log-likelihood of $-3748.5$ was found with much effort, using many global and local IF2 searches and likelihood profiling.
-Despite being initialized for a global search, both IFAD-0 and IFAD-1 manage to find the MLE over the 100 searches.
-This may imply that the local search and refinement that was previously necessary for finding the MLE is \textbf{baked into} the two-stage IFAD algorithm.
-
-Interestingly, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.63$, while \cite{king08} only achieve a best possible estimate of $-3793.4$.
-Our results of $-3748.53$ and $-3747.87$ match or beat the highest log-likelihood previously found in the current implementation in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.
-
-This can be seen most clearly in Figure \ref{fig:boxplot-search}.
-In the left panel within Figure \ref{fig:boxplot-search}, only IFAD has whiskers that extend to the dotted red line that represents the MLE.
-
-\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations (as seen in Figure \ref{fig:optim}), performing IF2 alone ultimately fails to squeeze out the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter} and the left panel in Figure \ref{fig:boxplot-search}).
-
-Conversely, performing MOP alone leads to many, many failed searches, as seen in Figure \ref{fig:hist-all} and the right panel in Figure \ref{fig:boxplot-search}.
-This is a difficult, nonconvex, and noisy problem, and the search gets stuck in local minima and saddle points, failing to approach the MLE.
-
-IFAD, in comparison, successfully manages to (1) approach the MLE quickly due to the IF2 warm-start (as seen in Figure \ref{fig:optim}) and (2) refine the coarse solution found by the warm-start with MOP gradient steps to find the MLE (as seen in Figures \ref{fig:scatter}, \ref{fig:boxplot-search}, and \ref{fig:hist-all}).
-IFAD therefore successfully combines the best qualities of IFAD and MOP, outperforming either of them alone.
-
-\begin{figure}[htbp!]
-    \centering
-    \includegraphics[scale=0.53]{../imgs/095/pairs.png}
-    \includegraphics[scale=0.53]{../imgs/095/qq.png}
-    \caption{Scatterplots depicting the performance of IFAD against that of IF2.
-\textbf{Left:} Paired searches from the same starting point.
-\textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches.
-It is clear that on average, IFAD has the edge, and manages to find the MLE while no IF2 search successfully gets within 7 log-likelihood units of it.
-We use a dotted red line to display the MLE in both.
-}
-
-    \label{fig:scatter}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=0.48]{../imgs/095/boxplot.png}
-    \includegraphics[scale=0.53]{../imgs/095/boxplotall.png}
-    \caption{\textbf{Left:} Boxplot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs.
-This simulates the result of the common procedure of running a few searches and choosing the best one.
-IFAD outperforms all other methods, and improves on the warm-start given by running 40 IF2 iterations.
-Our choice of hyperparameters for our implementation of IF2 outperforms the choices made by \cite{ionides15}, which we depict in the boxplot labeled "IF2 2015".
-As before, while IF2 successfully approaches the MLE, it ultimately fails to find it.
-    \textbf{Right:} Boxplot depicting the performance of IFAD, IF2, and MOP alone without the warm-start, where we plot all runs.
-We see that MOP alone drastically underperforms all other methods, failing to get close to the MLE.
-This shows that the IF2 warm-start may be necessary to achieve good performance in challenging nonconvex and noisy problems like this one.
-    \textbf{Takeaway:} In this panel, we see that (1) IF2 fails to find the MLE while the gradient steps perform better at fine-grained refinement, and (2) performing gradient steps alone without a warm-start leads to the search getting stuck in local minima and saddle points, ultimately demonstrating the necessity of a hybrid algorithm like IFAD.
-    In both, as before, we use a dotted red line to display the MLE. }
-    \label{fig:boxplot-search}
-\end{figure}
-
-\begin{figure}[htbp!]
-    \centering
-    \includegraphics[scale=0.7]{../imgs/095/optim.png}
-    \caption{Optimization progress of IFAD and IF2, where the solid lines depict the median negative log-likelihood at each iteration and the shaded area depicts the best search at any iteration and the 80\% percentile.
-The dashed orange line depicts the median warm-start given by running 40 IF2 iterations.
-While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD -- IFAD has better tail control and successfully reaches the MLE.
-The initial loss in log-likelihood upon transitioning from IF2 to MOP seems to help with escaping local minima.
-    As before, we use a dotted red line to display the MLE. }
-    \label{fig:optim}
-\end{figure}
-
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.7]{../imgs/095/hist.png}
-    \caption{Histogram comparison of the performance of IFAD, IF2, and MOP alone.
-As before, we note that IFAD successfully reaches the MLE a nontrivial number of times, while IF2 ultimately fails to do so.
-Our implementation of IF2 outperforms that of \cite{ionides15}, largely due to improved hyperparameter choices.
-MOP alone performs poorly, demonstrating the necessity of warm-starting the search with IF2.
-    We once again see that IFAD combines the best of IF2 and MOP, approaching the MLE quickly due to the IF2 warm-start and successfully refining the coarse solution with gradient steps, achieving superior performance.
-As before, we use a dotted red line to display the MLE.
-}
-    \label{fig:hist-all}
-\end{figure}
-
-\paragraph{Differences Between IFAD-0 and IFAD-1:} The differences between IFAD-1 and IFAD-0 (MOP-1 and MOP-0) do not seem very pronounced in this example.
-This could be due to a number of reasons, such as (1) future case counts $y_{{n+1}:N}^*$ not providing a large amount of information on state identification $x_n$ given current and past measurements $y_{0:n}^*$ as mentioned in \cite{corenflos21}, and (2) the gradient normalization making the trajectories given by MOP-0 and MOP-1 more similar than they would be otherwise.
-
-\kevin{IFAD beats IF2 x percent of the time on 100 paired runs, but the bigger thing is that it gets close to the MLE without much effort.
-Previously, we'd have to do many searches, plus refinement (which is necessary, as we can see from the IF2 boxplot, it is easy to fool yourself into thinking that because you see the plateau that is the MLE).
-This includes likelihood profiling diagnostics, which are very computationally intensive.
-Now we can just take the best of ten searches -- IFAD does the refinement.}
-
-\kevin{Have an outsider read the manuscript afterwards.}
-
-\section{Discussion and Future Work}
-
-We develop a new theoretical framework to differentiating through the particle filter that (1) encompasses the gradient estimates of \cite{poyiadjis11}, \cite{blei2018vsmc}, and \cite{scibior2021dpf} as special cases, (2) sidesteps the issue of differentiating through a Monte-Carlo algorithm with discontinuous resampling, and (3) provides an opportunity for optimizing a bias-variance tradeoff.
-
-Prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics.
-In the presence of a differentiable simulator, MOP-$\alpha$ enjoys this very same desirable property, enabling practical maximum likelihood inference in a wider variety of situations.
-
-Furthermore, MOP-$\alpha$, and by extension IFAD, has runtime \textbf{linear, and not quadratic}, in the number of particles, unlike the marginal differentiable particle filter in \cite{scibior2021dpf} or the the optimal transport of resampling approach of \cite{corenflos21}.
-This does not necessarily come at a cost to Monte Carlo variance, unlike \cite{poyiadjis11} and \cite{corenflos2021ot} who suffer from variance quadratic in the length of the trajectory, as we can optimize $\alpha$ for a bias-variance tradeoff.
-
-In the presence of strong convexity, first and second-order gradient methods that use MOP-$\alpha$ as an estimate enjoy a linear convergence rate, providing support for numerical optimization with AD for the particle filter.
-
-This leads to a practical hybrid algorithm we call IFAD that warm-starts gradient descent, or some other first or second-order method using the gradient estimate given by MOP-$\alpha$, with IF2.
-When benchmarked on a challenging global search problem with the model of \cite{king08}, IFAD enjoys the best properties of IF2 and MOP-$\alpha$, successfully approaching a neighborhood of the MLE quickly without getting stuck in saddle points and (poor) local minima (like gradient descent alone would), while effectively optimizing the last few units of log-likelihood (which IF2 struggles with).
-
-Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual} that we call \texttt{pomp-py}.
-The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{JAX} Python package from \citet{jax2018github}.
-
-\subsection{Future Work}
-
-While MOP-$\alpha$, and IFAD in its current incarnation, remain very general, they currently lack the capability of handling discrete latent states.
-We conjecture that DOP-$\alpha$ might be a possible solution (despite needing the likelihood ratio of transition densities), and will explore this in future work.
-
-We conjecture that the linear convergence result in Theorem \ref{thm:convergence} applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE.
-IF2 posesses some SGD-like qualities, including what looks like linear convergence to a ball around the MLE with width determined by the random walk standard deviation.
-Not much is known about the convergence rate of IF2, and an answer to this would also provide an answer to the convergence rate of IFAD beyond just the convergence rate of the gradient stage.
-We will explore this conjecture in future work.
-
-In the interest of simplicity, we only explore the performance of gradient descent on the Dhaka cholera model of \cite{king08}.
-Exploring the use of various first and second order methods such as Newton's method, ADAM, L-BFGS, or other popular iterative optimization methods with the gradient estimates that MOP-$\alpha$ provides in the context of the particle filter could very well lead to nontrivial speedups and increased performance.
-
-Extensions of MOP-$\alpha$ and IFAD to panel and spatiotemporal datam, possibly through a block particle filter in the same vein as \cite{ionides22} and \cite{ning23}, would go a long way towards extending the current applicability of these methods to handle a wider variety of interesting and challenging scientific problems.
-
-While we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual}, this remains a work in progress.
-
-\paragraph{Acknowledgements.} We thank Nicholas Chopin for helpful comments on the proof of a strong law of large numbers for a suitably reweighted particle filter.
-
-\kevin{Reach: PNAS.
-Else try for JRSSB, try to get Chopin for Annals AE.
-Would need to change bracketing conventions for Biometrika.
-If submitting to PNAS, JASA, Biometrika, or JRSSB, highlight the different statistical gradient estimator and finite sample performance, rewrite introduction, maybe create a way to use the gradients for inference rather than using profiling.
-Try to rewrite it as a breakthrough in statistics.
-Safety: JMLR}
-
-\bibliography{paper/bib-ifad, paper/bib-ref}
-\bibliographystyle{apalike}
-
-\appendix
-\renewcommand{\thefigure}{A\arabic{figure}}
-\setcounter{figure}{0}
-
-\section{Weighted Particle Filters that Target the Posterior}
-
 \begin{defn}[Targeting]
@@
 \end{defn}
-\citet{chopin2004clt} asserted without proof that common particle filter algorithms target the filtering distribution, $f_{X_{1:N}|Y_{1:N}}$, in this sense.
-\ed{PERHAPS IF WE HAD A CONCRETE PROOF OF THIS, THAT WOULD HELP TO SEE HOW IT EXTENDS TO OUR SUBSEQUENT LEMMAS.
-ARGUABLY, THAT IS IN FACT WHAT WE ARE DEVELOPING VIA THESE LEMMAS.
-I WONDER IF SOME OTHER REFERENCE, E.G.
-DELMORAL'S 2004 BOOK, HAS AN EXPLICIT PROOF OF CONSISTENCY FOR SOME PARTICLE FILTER ALGORITHM.}
+\cite{chopin04} asserted without proof that common particle filter algorithms target the filtering distribution, $f_{X_{1:N}|Y_{1:N}}$, in this sense.
 In order to prove the consistency of our variation on the particle filter, we now present three helper lemmas.
 The first follows from standard importance sampling arguments, the second from integrating out the marginal, and the third from Bayes' theorem. 
-We state Lemma~\ref{lem:change-measure-proper-weights} assuming multinomial resampling, which is convenient for the proof though other resampling strategies may be preferable in practice REFS.
+We state Lemma~\ref{lem:change-measure-proper-weights} assuming multinomial resampling, which is convenient for the proof though other resampling strategies may be preferable in practice.
 
-\begin{lem}[Change of Particle Measure]
+\begin{lem}[Change of Weight Measure]
     \label{lem:change-measure-proper-weights}
     Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
-Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, $\pi^J_j J$ times.
+Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, proportional to $\pi^J_j J$ times.
 Write
@@
 
-    $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J}}{J^{-1} \sum_j \bar{u}_j^{' J}}$$
+    Again, we want to show that
+     $$\frac{\sum_j h(X_j^J) \cdot U_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j U_j^J \cdot f_{Z|X}(z^*|X_j^J)} \stackrel{a.s.}{\to} \E_{f_{X|Z}}[h(X)|z^*].$$
 
-    by
-
+    We already have that for any measurable bounded $g$,
     \begin{equation}
-    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}}
-    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \right)^{-1}
+        \frac{\sum_j g(X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[g(X)].
+        \label{eq:lemma-posterior-hypothesis}
     \end{equation}
-
-    Again, we want to show that
-     $$\frac{\sum_j h(X_j^J) \cdot u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)} \stackrel{a.s.}{\to} \E_{f_{X|Z}}[h(X)|z^*].$$
-
-    We already have that
-    $$\frac{\sum_j h(X_j^J)u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} \E_{f_X}[h(X)].$$
-
-    %= \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j f_{Z|X}(z^*|X_j^J)f_X(X_j^J)}
 
@@
     \begin{equation}
-    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}}
-    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \right)^{-1}.
+    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}
+    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}} \right)^{-1}.
     \end{equation}
 
-    Write $\bar{u}_j^{J} = \frac{J u_j^J}{\sum_j u_j^J}$.
-Define the weights $u^{' J}_j := u_j^J \cdot f_{Z|X}(z^*|X_j^J),$ and the corresponding self-normalized weights $\bar{u}_j^{' J} := \frac{J u^{' J}_j}{\sum_j u^{' J}_j}.$ The denominator of the below quantity
-    $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J}}{J^{-1} \sum_j \bar{u}_j^{' J}}$$
-    converges to $1$, while the numerator, by the same fourth moment argument as the first lemma, converges almost surely to the limit of its expectation, which is
+    We will apply Equation \ref{eq:lemma-posterior-hypothesis} to the numerator and the denominator in the ratio above individually. The numerator converges to 
+    $$\frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \stackrel{a.s.}{\to} \E_{f_X}[h(X)f_{Z|X}(z^*|X)],$$
+    while the reciprocal of the denominator converges to 
+    $$ \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}  \stackrel{a.s.}{\to} \E_{f_X}[f_{Z|X}(z^*|X)] = f_Z(z^*).$$
+
+    But it holds that
+    $$\frac{\E_{f_X}[h(X)f_{Z|X}(z^*|X)]}{f_Z(z^*)} = \E_{f_X}\left[\frac{h(X)f_{Z|X}(z^*|X)}{f_Z(z^*)}\right] = \E_{f_X}\left[h(X)\frac{f_{X|Z}(X|z^*)}{f_X(X)}\right] =  \E_{f_{X|Z}}[h(X)|z^*],$$
+    and we have the desired result.
+
+    % Write $\bar{U}_j^{J} = \frac{J U_j^J}{\sum_j U_j^J}$.
+Define the weights $U^{' J}_j := U_j^J \cdot f_{Z|X}(z^*|X_j^J),$ and the corresponding self-normalized weights $\bar{U}_j^{' J} := \frac{J U^{' J}_j}{\sum_j U^{' J}_j}.$ The denominator of the below quantity
+    % $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{U}_j^{' J}}{J^{-1} \sum_j \bar{U}_j^{' J}}$$
+    % converges to $1$, while the numerator, by the same fourth moment argument as the first lemma, converges almost surely to the limit of its expectation, which is
+    % \begin{align*}
+    %     \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J} \right]
+    %     &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)}\right] \\
+    %     &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
+    %     &= \lim_{J\to\infty} \E\left[ \sum_j h(X_j^J) \frac{\frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
+    %     &= \E_{X|Z}[h(X)|z^*],
+    % \end{align*}
+
+    % where we can swap the limit and expectations with the bounded convergence theorem.
+\end{proof}
+
+\begin{lem}[Change of Particle Marginal Measure]
+    \label{lem:change-marginal-proper-weights}
+    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
+Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
+Further, let $f_{Y|X}$ be a conditional probability density function corresponding to a joint density $f_{X,Y}$ with marginal densities $f_X$ and $f_Y$.
+Then, if the $U_j^J$ and $V_j^J := U_j^Jf_{Y|X}(\tilde{Z}_j^J|\tilde{X}_{j}^J)/f_{Z|X}(\tilde{Z}_j^J|\tilde{X}_{j}^J)$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}$ targets $f_Z$, and $\{(\tilde Z_j^J,V_j^J)\}$ targets $f_Y$.
+\end{lem}
+\begin{proof}
+    From Lemma \ref{lem:marginal-proper-weights}, for any measurable bounded $g$, 
+     $$\frac{\sum_j g(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_Z}[g(Z)] = \E_{f_X}\big[\E_{f_{Z|X}}[g(Z) | X]\big],$$
+     and by assumption, for any measurable bounded $g$, 
+    $$\frac{\sum_j g(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[g(X)].$$
+
+    We in fact want to show that 
+    $$\frac{\sum_j h(\tilde Z_j^J) \, V_j^J}{\sum_j V_j^J} \stackrel{a.s.}{\to} \E_{f_Y}[h(Y)] = \E_{f_X}\left[\E_{f_{Z|X}}\left[h(Z)\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)} \Bigg| X\right]\right].$$
+
+    Consider the following:
+    \begin{equation}
+    \frac{J^{-1} \sum_j h(Z_j^J) \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)} {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}
+    \times \left( \frac{J^{-1} \sum_j \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)} {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}} \right)^{-1}.
+    \end{equation}
+
+    We will apply Lemma 2 to the numerator and the denominator in the ratio above individually. The numerator converges to 
+    $$\frac{J^{-1} \sum_j h(Z_j^J)  \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)} {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}} \stackrel{a.s.}{\to} \E_{f_Z}\left[h(Z) \frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\right] = \E_{f_X}\left[\E_{f_{Z|X}}\left[h(Z)\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right],$$
+    while the reciprocal of the denominator converges to 
+    $$ \frac{J^{-1} \sum_j \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)}{U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}  \stackrel{a.s.}{\to} \E_{f_Z}\left[ \frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\right] = \E_{f_X}\left[\E_{f_{Z|X}}\left[\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right] .$$
+
+    The numerator is exactly what we desire, so it remains to show that the reciprocal of the denominator converges to $1$. Applying Fubini, we have that
     \begin{align*}
-        \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J} \right]
-        &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)}\right] \\
-        &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
-        &= \lim_{J\to\infty} \E\left[ \sum_j h(X_j^J) \frac{\frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
-        &= \E_{X|Z}[h(X)|z^*],
+        \E_{f_X}\left[\E_{f_{Z|X}}\left[\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right] &= \int_x\left(\int_z \frac{f_{Y|X}(z|x)}{f_{Z|X}(z|x)}f_{Z|X}(z|x)dz \right)f_X(x) dx \\
+        &= \int_x\int_z f_{Y|X}(z|x) f_X(x) dzdx \\
+        &= \int_z\int_x f_{Y|X}(z|x) f_X(x) dxdz \\
+        &= \int_z f_{Y}(z) dz \\
+        &= 1,
     \end{align*}
+    and we are done.
 
-    where we can swap the limit and expectations with the bounded convergence theorem.
 \end{proof}
@@
 $L^{B,\theta}$ should generally be preferred in practice, since there is no reason to include the extra variability from resampling when calculating the conditional log likelihood, but it lacks the nice telescoping product that lets us derive exact expressions for the gradient later.
-
-\section{Doubly Off-Policy Particle Filters}
-\label{app:dop}
-
-The proof for Proposition \ref{prop:dop-correctness} follows from the below two lemmas.
-Evaluating the off-policy likelihood is possible with a weighted particle filter by using resampling weight $p(y^*_n|x_n;\phi)$ and making two corrections in the particle weight, through multiplying $w_{t,j}$ by the two likelihood ratios below,
-
-\begin{equation}
-    \label{eq:dmeas-ratio}
-    s=p(y^*_n|x_n;\theta)/p(y^*_n|x_n;\phi)
-\end{equation}
-and
-\begin{equation}
-    \label{eq:rproc-ratio}
-    r=p(x_n|x_{n-1};\theta)/p(x_n|x_{n-1};\phi).
-\end{equation}
-
-The following result showcases the correctness of corrections \ref{eq:dmeas-ratio} and \ref{eq:rproc-ratio}.
-
-\begin{lem}
-    For $\theta$, $\phi$ that satisfy Proposition \ref{prop:smooth-nbhd}, the following correction in the particle filter weights,
-    \begin{equation}
-        w_{t,j} := \tilde{w}_{t,j} \cdot p(y^*|x_{t,j}^P,\phi)\cdot r \cdot s
-    \end{equation}
-    successfully recovers $\lik(\theta)$.
-\end{lem}
-
-\begin{proof}
-    Recall $\lik(\theta) = \E[\lik(\theta,\phi,\Omega, J)]=\int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega)$ and $\lik(\theta) = \int_\Omega \lik(\theta, \omega, J) d\mu(\omega)$.
-We want to evaluate
-    \begin{align*}
-        \lik(\theta,\omega, J)
-        &= \E\left[ \prod_{t=1}^T \lik_t(\theta) \middle| \omega\right]
-        = \E\left[ \prod_{t=1}^T p(y_t^*|y_{1:t-1}^*,\theta) \middle| \omega\right] \\
-        &= \E\left[ \prod_{t=1}^T \frac{1}{J} \sum_{j=1}^J  p(y_t^*|x_{t,j}^P,\theta) p(x_{t,j}^P|y_{1:t-1}^*,\theta) \middle| \omega\right] \\
-        &= \E\left[ \prod_{t=1}^T\frac{1}{J}  \sum_{j=1}^J  p(y_t^*|x_{t,j}^P,\theta) p(x_{t,j}^P|x_{t-1}^F,\theta) \middle| \omega\right] 
-    \end{align*}
-    where $p(x_t^P|y_{1:t-1}^*,\theta)$ is the prediction distribution of particles, and the filtering distribution of particles is proportional to $p(y_t^*|x_{t,j}^P,\theta)$, the likelihood of particle $j$ at time $t$.
-
-    Consider first the case with no resampling.
-Let $\lik^j(\theta)$, $\lik^j(\phi)$ be the likelihood of trajectory $j$ under $\theta$ and $\phi$ respectively.
-As the prediction and filtering particles are then the same, we use $x_{t,j}$ to refer to both.
-As such, conditional on $\omega$,
-    \begin{align*}
-        \lik^j(\theta) 
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \theta) p(x_{t,j}|x_{t-1,j}, \theta) \\
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \frac{p(y_t^*|x_{t,j}, \theta)}{p(y_t^*|x_{t,j}, \phi)} p(x_{t,j}|x_{t-1,j}, \theta) \frac{p(x_{t,j}|x_{t-1,j}, \theta) }{p(x_{t,j}|x_{t-1,j}, \phi) } \\
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \cdot s \cdot p(x_{t,j}|x_{t-1,j}, \phi)\cdot r
-    \end{align*}
-
-    and therefore, if trajectories were proposed with equal probability with particles conditional on $\omega$,
-    $$ \lik(\theta, \omega, J) = \frac{1}{J} \sum_{j=1}^J \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \cdot s \cdot p(x_{t,j}|x_{t-1,j}, \phi)\cdot r $$
-    The correction is therefore sufficient for recovering $\lik(\theta)$, with trajectories collected with the process model at $\phi$ and measurements evaluated with the measurement model at $\phi$.
-
-    Consider now the case with resampling.
-Without loss of generality we can resample every timestep.
-Here, $x_{t,j}^F(\phi) \sim p(y^*|x_{t,j}^P, \phi)$ and $x_{t,j}^P(\phi) \sim p(x_t|x_{t-1}^F,\phi)$, and similarly for $\theta$.
-We use this notation to emphasize that \textbf{both} the measurement model and the previously drawn particles depend on $\phi$ or $\theta$.
-
-    \begin{align*}
-        \lik_t(\theta) &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \theta)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\theta) \right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \theta)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \frac{p(y_t^*|x_{t,j}^P,\theta)}{p(y_t^*|x_{t,j}^P,\phi)} \right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \frac{p(y_t^*|x_{t,j}^P,\theta)}{p(y_t^*|x_{t,j}^P,\phi)} \frac{p(x_{t,j}|x_{t-1,j}, \theta) }{p(x_{t,j}|x_{t-1,j}, \phi) }\right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \cdot s \cdot r \right] \\
-        &= \E_{\omega}\E_{x_{t,j}^P \sim \process(x_{t-1}^F, \omega, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \cdot s \cdot r \middle| \omega \right]
-    \end{align*}
-
-    and again the correction is sufficient.
-
-    Therefore,
-
-    \begin{align}
-         \mathcal{L}(\theta) 
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^P(\omega), \theta) p(x_{n,j}^P(\omega) | x_{n-1,j}^P(\omega), \theta) \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\theta}(\omega), \theta) p(x_{n,j}^{P,\theta}(\omega) | x_{n,j}^P(\omega), \theta) \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\theta}(\omega), \phi) p(x_{n,j}^{P,\theta}(\omega) | y_n^*, \phi) \cdot s  \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) p(x_{n,j}^{P,\phi}(\omega) | y_n^*, \phi) \cdot r \cdot s \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) p(x_{n,j}^{P, \phi}(\omega) | y_n^*, \phi) \cdot r \cdot s \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) \cdot r \cdot s \right)  p(x_{1:N} | y_{1:N}^*, \phi) \,  d\mu(\omega) \\
-         &=: \int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega).
-     \end{align}
-
-     The proof for an arbitrary functional under the particle measure follows analogously.
-\end{proof}
-
-\begin{lem}[Gradient of Doubly Off-Policy Particle Filter]
-    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$.
-Also let $\theta \in \gN(\phi)$.
-When $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{poyiadjis11},
-    \begin{equation}
-        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
-    \end{equation}
-\end{lem}
-
-\begin{proof}
-    It is known, from \cite{doucet2009tutorial}, that 
-    $$p_\theta(x_{1:N}|y_{1:N}) = \prod_{n=1}^N \frac{p(x_n|x_{n-1})p(y_n|x_n)}{p(y_n|y_{1:n-1})} = \prod_{n=1}^N \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{\lik_n(\theta)}.$$
-
-    From the above lemma, and the fact that the particle filter is unbiased for the likelihood, we know that
-    \begin{align*}
-        \lik(\theta)
-        &= \int_{x_{1:N}} \prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) p_\theta(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N  p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \right) \frac{p_\theta(x_{1:N}|y_{1:N})}{p_\phi(x_{1:N}|y_{1:N})}p_\phi(x_{1:N}|y_{1:N}) dx_{1:N}\\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N \lik_n(\theta) \right) \left(\prod_{n=1}^N \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \lik_n(\phi)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) \lik_n(\theta)}\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N \lik_n(\phi) \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) }\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N}.
-    \end{align*}
-
-    Therefore, under regularity conditions that permit differentiating under the integral sign, using the log-derivative trick, we have
-    \begin{align*}
-        \nabla_\theta \lik(\theta)
-        &= \int_{x_{1:N}} \nabla_\theta \hat\lik(\theta) \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N}\\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N \lik_n(\phi) \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) }\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N \left(\frac{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n)}{\lik_n(\phi)}\right)^{-1} p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\frac{1}{p_\phi(x_{1:N}|y_{1:N}) }\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\nabla_\theta \log\left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} p_\theta(x_{1:N},y_{1:N}) \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) p_\theta(x_{1:N}|y_{1:N}) \cdot \lik(\theta) \; dx_{1:N}. 
-    \end{align*}
-
-    Using the log-derivative trick again,
-    \begin{align*}
-        \nabla_\theta \ell(\theta)
-        &= \frac{\nabla_\theta \lik(\theta)}{\lik(\theta)} \\
-        &= \int_{x_{1:N}} \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) p_\theta(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \E_{x_{1:N}|y_{1:N}, \theta}\left[\nabla_\theta \left(\sum_{n=1}^N\log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right)\right] \\
-        &= \E_{x_{1:N}|y_{1:N}, \theta}\left[\nabla_\theta\log p_\theta(x_{1:N},y_{1:N})\right] \\
-        &= \lim_{J\to\infty} \E_{x_{1:N}|y_{1:N}, \theta}\left[\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log p_\theta(x_{1:N,j}^{A,F},y_{1:N}^*)\right].
-    \end{align*}
-
-    But it also holds that
-    \begin{align*}
-        \nabla_\theta \ell(\theta)
-        &= \frac{\nabla_\theta \lik(\theta)}{\lik(\theta)} \\
-        &= \int_{x_{1:N}} \frac{\nabla_\theta \hat\lik(\theta)}{\lik(\theta)} \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \hat\ell(\theta) \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N}\\
-        &= \lim_{J\to\infty} \E_{x_{1:N}|y_{1:N}, \phi}\left[\nabla_\theta \hat\ell(\theta)\right].
-    \end{align*}
-
-    and we see that $$\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log p_\theta(x_{1:N,j}^{A,F},y_{1:N}^*)$$ and $$\nabla_\theta \hat\ell(\theta)$$ both converge to the same thing in expectation.
-They are both particle approximations under the posterior where the variance of each is bounded, therefore one can apply the particle CLT from \cite{chopin2004clt} to obtain that they are both consistent estimates of the score.
-
-    In fact, when $\theta=\phi,$ they are the same in distribution. \kevin{claim}
-
-    \kevin{This is a far weaker result than what I really wanted. This just gives consistency. Want equality in distribution.}
-\end{proof}
-
-\begin{lem}
-
-Consider the case of DOP-$\alpha$ when $\alpha=0$ and $\theta=\phi$. The gradient estimate is then
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta\right),
-    \end{equation}
-
-    yielding the gradient estimator of \cite{blei2018vsmc} that \cite{scibior2021dpf} show is the gradient of a vanilla particle filter when the resampling terms are dropped.
-
-\end{lem}
-
-\begin{proof}
-    When $\alpha=0,$ the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J r_{n,j}s_{n,j}}{\sum_{j=1}^J r_{n,j}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J r_{n,j}s_{n,j}.
-\end{equation}
-
-Its gradient when $\theta=\phi$ is therefore 
-\begin{align*}
-    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right) \\
-    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right)} \\
-    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta r_{n,j}s_{n,j}}{\sum_{j=1}^J r_{n,j}s_{n,j}} \\
-    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{Y_n, X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)} \\
-    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta\right),
-\end{align*}
-where we use the log-derivative trick, observe that $\sum_{j=1}^J r_{n,j}s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again. 
-\end{proof}
-
-\subsection{Off-Policy Weighted Differentiable Particle Filter}
-
-We are finally ready to introduce the off-policy weighted differentiable particle filter.
-In the event where a seed is fixed, all expressions in Algorithm \ref{alg:dop} can be considered to be conditional on $\omega \in \Omega$.
-
-\begin{algorithm}[ht]
-\centering
-	\caption{Doubly Off-Policy-$\alpha$}
-    \label{alg:dop}
-	\begin{algorithmic}[1]
-	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
-		\STATE Initialize filter particles ${X}_{0,j}^{F,\phi}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\phi}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$. Fix $\omega.$
-		\FOR{$n=1,...,N$}
-            \STATE Prediction weights with discounting: $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
-            \label{dop-alpha:discount}
-            \STATE Simulate for prediction:
-            ${X}_{n,j}^{P,\phi}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\phi}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{dop-alpha:step1}
-            \STATE Adjust weights: $\displaystyle w_{n,j}^{P,\theta} = w_{n,j}^{P,\theta} \times
-            \frac{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\phi}|{X}_{n-1,j}^{F};{\theta}\big)}{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\phi}|{X}_{n-1,j}^{F};{\phi}\big)}$ for $j$ in $1:J$
-            \label{dop-alpha:dproc}
-            \STATE Evaluate measurement density:
-            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\phi}\giventh{\theta})$ for $j$ in $\seq{1}{J}$
-            \STATE Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \STATE Conditional likelihood under $\phi$: 
-            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
-            \label{dop-alpha:Lphi}
-            \STATE Normalize weights:
-            $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
-            for $j\ \mathrm{in}\ \seq{1}{J}$
-            \STATE Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \label{dop-alpha:systematic}
-            \STATE Resample particles: ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$
-            \STATE Filter weights corrected for resampling:
-            $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{dop-alpha:weight:update}
-            \STATE Resample filter weights:
-            $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
-            for $j$ in $\seq{1}{J}$ \label{dop-alpha:step2}
-            \STATE After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \ENDFOR
-		\RETURN likelihood estimate $\hat{\lik}(\theta) = \lik(\theta, \phi, \omega, J) := \prod_{n=1}^N L_n^{A,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}.$
-	\end{algorithmic}
-\end{algorithm}
-
-Here, we consider a doubly off policy particle filter, meaning that both \code{rprocess} and \code{dmeasure} are computed at $\phi$ and particles are reweighted to correspond to $\theta$.
-
-\section{Measurement Off-Policy Particle Filters}
-
-\begin{algorithm}[ht]
-\centering
-	\caption{Measurement Off-Policy-$\alpha$}
-    \label{alg:mop}
-	\begin{algorithmic}[1]
-	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
-        \IF{$\theta \neq \phi$}
-            \STATE Run a particle filter once with parameters $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$, with seed $\omega$.
-        \ELSE
-        \STATE Set $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$ to be copies of $X_{n,j}^{P,\theta}, X_{n,j}^{P,\theta}$ in the rest of the algorithm.
-        \ENDIF
-		\STATE Initialize filter particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$. Fix $\omega.$
-		\FOR{$n=1,...,N$}
-            \STATE Prediction weights with discounting: $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
-            \label{mop-alpha:discount}
-            \STATE Simulate for prediction:
-            ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{mop-alpha:step1}
-            \STATE Evaluate measurement density:
-            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for $j$ in $\seq{1}{J}$
-            \STATE Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \STATE Conditional likelihood under $\phi$: 
-            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
-            \label{mop-alpha:Lphi}
-            \STATE Normalize weights:
-            $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
-            for $j\ \mathrm{in}\ \seq{1}{J}$
-            \STATE Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \label{mop-alpha:systematic}
-            \STATE Resample particles: ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$
-            \STATE Filter weights corrected for resampling:
-            $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{mop-alpha:weight:update}
-            \STATE Resample filter weights:
-            $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
-            for $j$ in $\seq{1}{J}$ \label{mop-alpha:step2}
-            \STATE After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \ENDFOR
-		\RETURN likelihood estimate $\hat{\lik}(\theta) = \lik(\theta, \phi, \omega, J) := \prod_{n=1}^N L_n^{A,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}.$
-	\end{algorithmic}
-\end{algorithm}
-
-\begin{lem}
-    \label{lem:mop-1-formula}
-    Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The gradient estimate is then
-
-    \begin{equation}
-        \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
-    \end{equation}
-
-    yielding the gradient estimators of \cite{poyiadjis11, scibior2021dpf} when applied to the bootstrap filter. 
-\end{lem}
-
-\begin{proof}
-
-Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The likelihood estimate becomes
-\begin{equation*}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}}= \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\theta}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
-\end{equation*}
-as we have a telescoping product. Then, as
-
-$$\nabla_\theta \hat\ell(\theta) = \frac{\nabla_\theta \hat\lik(\theta)}{\hat\lik(\theta)} = \frac{\nabla_\theta\left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi}{\prod_{n=1}^N L_n^\phi} =  \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},$$
-
-The derivative of the log-likelihood estimate is then
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},
-\end{equation*}
-which we decompose as follows.
-
-Observe that as $\alpha=1$,
-
-$$w_{n,j}^{P,\theta} = w_{n-1,j}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}},$$
-where we use the $(\cdot)^A$ superscript to denote the ancestral trajectory of the $j$-th prediction or filtering particle at timestep $n$.
-
-Using the log-derivative identity again,
-$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,j}^{A,P,\theta} - \log g_{i,j}^{A,P,\phi}\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,j}^{A,P,\theta}.$$
-
-So 
-$$  \nabla_\theta \sum_{n=1}^N \log g_{n,j}^{A,\theta} = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,j}^{A,P,\theta}\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,j}^{A, P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right),$$
-
-and 
-$$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{n=1}^N \nabla_\theta \log g_{n,j}^{A,P,\theta} = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$
-
-Substituting, we have that
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right),
-\end{equation*}
-and finally, observing that when $\theta=\phi$ we have that $w_{N,j}^{F,\theta}=1$, that 
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
-\end{equation*}
-yielding the gradient estimators of \cite{poyiadjis11, scibior2021dpf} when applied to the bootstrap filter. 
-\end{proof}
-
-Note that one would believe that the variance blows up the moment $\theta\neq\phi$. This is true, as 
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right),
-\end{equation*}
-and when $\theta\neq\phi$, $w_{N,k_j}^{P,\theta} = O(c^N)$.
-When $\theta=\phi$, this is a special case of the \cite{poyiadjis11} estimator, which has $O(N^2)$ variance by a property of functionals of the particle filter by \cite{delMoral03}.
-
-\begin{lem}
-\label{lem:mop-0-formula}
-
-Consider the case of MOP-$\alpha$ when $\alpha=0$ and $\theta=\phi$. The gradient estimate is then
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
-    \end{equation}
-
-    yielding the gradient estimator of \cite{blei2018vsmc} that \cite{scibior2021dpf} show is the gradient of a vanilla particle filter when the resampling terms are dropped -- when applied to the bootstrap filter.
-
-\end{lem}
-
-\begin{proof}
-    When $\alpha=0,$ the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
-\end{equation}
-
-Its gradient when $\theta=\phi$ is therefore 
-\begin{align*}
-    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
-    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)} \\
-    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta s_{n,j}}{\sum_{j=1}^J s_{n,j}} \\
-    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)}{f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \phi}; \phi)} \\
-    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
-\end{align*}
-where we use the log-derivative trick, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again. 
-\end{proof}
-
-\kevin{I have no clue how to prove this. Help!}
-\begin{conj}[MOP-1 Estimate Is Unbiased for the Likelihood and the Gradient of the Likelihood]
-    Consider the case of MOP-$\alpha$ when $\alpha=1$, and resampling is carried out with either multinomial, residual, or systematic resampling at every timestep.
-Assume the gradient estimate obtained by the gradient of the likelihood estimate with J particles, $\nabla_\theta \hat\lik_J(\theta)$ is uniformly bounded for all $J$, and that the likelihood has two continuous derivatives.
-Then the likelihood estimate is unbiased for the likelihood and the gradient of the likelihood estimate is unbiased for the gradient of the likelihood.
-That is, it holds that
-    $$\E[\hat\lik_J(\theta)] = \lik(\theta),\;\; \E[\nabla_\theta \hat\lik_J(\theta)] = \nabla_\theta \lik(\theta).$$
-\end{conj}
-
-\begin{proof}
-    First, we need to establish the unbiasedness of the likelihood estimate. This is 
-    $$\hat\lik_J(\theta) := \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right)\prod_{n=1}^N L_n^\phi = \left(\frac{1}{J}\sum_{j=1}^J \prod_{n=1}^N \frac{g_{n,j}^{A,\theta}}{g_{n,j}^{A,\phi}} \right)\prod_{n=1}^N L_n^\phi.$$
-
-    We know from \cite{chopin20} that $\E[L_n^{\phi}] = \lik(\phi).$ It then suffices to show that
-    $$\E\left[\left(\frac{1}{J}\sum_{j=1}^J \prod_{n=1}^N \frac{g_{n,j}^{A,\theta}}{g_{n,j}^{A,\phi}} \right)\right] = \frac{\lik(\theta)}{\lik(\phi)}.$$
-
-    To obtain this, we need to see that the resampling under $\phi$, with corrected weights, is unbiased. That is, we need to show that for any bounded measurable functional $\varphi : \gX \to \R$,
-    $$\E\left[\frac{\sum_{j=1}^J w_{n,j}^{F, \theta} \varphi(x_{n,j}^{\theta, F})}{\sum_{j=1}^J w_{n,j}^{F, \theta}} \bigg| \gF_n \right] = \frac{\sum_{j=1}^J w_{n,j}^{P, \theta} \varphi(x_{n,j}^{P, \theta})}{\sum_{j=1}^J w_{n,j}^{P, \theta} }.
-$$
-
-    To see this, 
-    \begin{align*}
-        \sum_{j=1}^J w_{n,j}^{P, \theta} \varphi(x_{n,j}^{P, \theta}) 
-        &= \sum_{j=1}^J w_{n-1,j}^{F, \theta}\frac{g_{n,j}^{\theta}}{g_{n,j}^{\phi}}  \varphi(x_{n,j}^{P, \theta})
-    \end{align*}
-
-%     Observe that our differentiable simulator $\process(\cdot, \theta)$ and choice of weights now satisfy:
-
-%     - A Feynman-Kac model $\left\{M_t, G_t\right\}$ such that:
-
-% - the weight function $G_t$ may be evaluated pointwise (for all $t$ );
-
-% - it is possible to simulate from $\mathbb{M}_0\left(\mathrm{~d} x_0\right)$ and from $M_t\left(x_{t-1}, \mathrm{~d} x_t\right)$ (for any $x_{t-1}$ and $t$ ).
-
-% - The number of particles $N$.
-
-% - The choice of an unbiased resampling scheme (e.g.
-multinomial, residual, stratified, systematic), that is, an algorithm to generate variables $A_t^{1: N}$ in $1: N$ such that $\mathbb{E}\left[\sum_{m=1}^N \mathbbm{1}\left\{A_t^m=n\right\}\right]=N W_{t-1}^n$; see previous chapter for more details.
-\end{proof}
-
-\begin{lem}[Consistency of MOP-1 Gradient Estimate]
-    Consider the case of MOP-$\alpha$ when $\alpha=1$.
-Assume the gradient estimate obtained by the gradient of the log-likelihood estimate with J particles, $\nabla_\theta \hat\ell_J(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}$ is uniformly bounded by $G(\theta)$ for all $J$, and that the likelihood has two continuous derivatives.
-Then the gradient estimate, $\nabla_\theta \hat\ell_J(\theta)$, is consistent for the score.
-    \label{lem:mop-grad-consistency}
-\end{lem}
-\begin{proof}
-
-    % We know that the variance of the MOP-1 gradient estimate $\nabla_\theta \hat\ell_J(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}$ goes to zero coordinate-wise as $J\to\infty$ by the particle filter law of large numbers for bounded functionals of particles (\kevin{Prove general LLN for functionals of MOP under boundedness assumptions in L2, but maybe Hoeffding may be enough to show this, because we only want convergence in probability, we just need to show the expectation converges}), so it suffices to show that it is asymptotically unbiased.
-
-    % Write $\hat\pi_{n,J}$ for the filtering distribution at timestep $n$ and with $J$ particles, $\hat\pi_{1:N,J}$ for the sequence of filtering distributions from timesteps $1$ to $N$, and $\pi_n, \pi_{1:N}$ for the respective posteriors.
-
-    % From Lemma \ref{lem:lik-proper-weight}, we know that $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1:n-1}^*, \theta)$ is consistently estimated by the conditional likelihood estimate obtained by MOP-1, for all $\theta$ in an open neighborhood of $\phi$.
-The likelihood is simply the product of the conditional likelihoods, and similarly for the estimates, so the likelihood $\lik(\theta)$ is also consistently estimated by the likelihood estimate obtained by MOP-1, $\hat\lik_J(\theta)$, for all $\theta$ in an open neighborhood of $\phi$, $\gN(\phi)$.
-
-    % We therefore have that 
-    % $$\lim_{J\to\infty} \E[\hat\lik_J(\theta)] = \lik(\theta)$$
-    % pointwise for all $\theta \in \gN(\phi)$.
-Given that the gradient is bounded uniformly for all $J$ and consistency for the likelihood estimate holds for all $\theta$ in an open neighborhood of $\phi$ (and therefore all $\theta'$ in a sufficiently small neighborhood of $\theta$), the bounded convergence theorem allows us to switch the expectation and derivative:
-    % $$\nabla_\theta \E[\hat\lik_J(\theta)] 
-    % = \E[\nabla_\theta \hat\lik_J(\theta)].$$
-
-    % Take limits on both sides to find that 
-    % $$\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)] 
-    % = \lim_{J \to \infty} \E[\nabla_\theta \hat\lik_J(\theta)].$$
-
-    % We then need to swap the limit and derivative on the left-hand side.
-This requires uniform convergence of $\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]$, i.e.
-that $$\sup_{\theta \in \gN(\phi)} ||  \nabla_\theta \E[\hat\lik_J(\theta)] - \lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]||_{\infty} \to 0 \text{ as } J \to \infty,$$
-    % or that $\forall \epsilon>0$ $\exists N\in \mathbb{N}$ such that $n>N => ||  \nabla_\theta \E[\hat\lik_J(\theta)] - \lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]||_{\infty} < \epsilon$ $\forall \theta \in \gN(\phi).$ However, it is easier to instead use an indirect argument.
-
-    % We then need to swap the limit and derivative on the left-hand side. This requires uniform convergence.
-
-    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point we wish to evaluate the gradient at.
-The sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$, by Assumption \ref{assump:local-bounded-derivative}.
-The sequence is also uniformly equicontinuous.
-To see this, by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J(\theta, \omega)|_{\theta=\theta'}$ is also bounded by $H$ for almost every $\omega\in \Omega$ and every $\theta' \in \gN(\phi)$, including $\theta=\phi$.
-A set of functions with derivatives bounded by the same constant is uniformly Lipschitz, and therefore uniformly equicontinuous.
-So the sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ is uniformly equicontinuous for almost every $\omega \in \Omega$.
-    Explicitly, for almost every $\omega \in \Omega$ and every $\epsilon>0$, there exists some $\delta(\omega)>0$ such that for every $||\theta - \theta'||_{\infty}<\delta$ and every $J \in \mathbb{N}$ we have that
-    $$||\nabla_\theta \hat\lik_J(\theta, \omega)-\nabla_\theta \hat\lik_J(\theta', \omega)||_\infty < \epsilon.$$
-
-    \kevin{Our previous understanding was that the target is differentiable, and the sequence is differentiable in increasingly small open neighborhoods.
-This may not be true.
-The neighborhoods the sequence is differentiable in should be independent of $J$, as the resampling does not change with $\theta$ as it is dependent on $\phi$.
-Either way, the target is in each of the neighborhoods, and this may be sufficient in itself.}
-
-    Then, by Arzela-Ascoli, there exists a uniformly convergent subsequence.
-We claim that there is only one subsequential limit.
-We know that when as the gradient is bounded by Assumption \ref{assump:local-bounded-derivative} we can treat the gradient as a bounded functional.
-So by Lemma \ref{lem:posterior-proper-weights} the sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$, and there is therefore only one subsequential limit.
-The sequence therefore converges uniformly to its limit $\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega).$ Therefore, with uniform convergence for the derivatives established, we can swap the limit and derivative, and obtain that for almost every $\omega \in \Omega$,
-    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta, \omega).$$
-
-    From Lemma \ref{lem:lik-proper-weight}, we know that
-    $$\hat{\lik}_J(\theta, \omega) \stackrel{a.s.}{\to} \lik(\theta).$$
-
-    We then have that for almost every $\omega \in \Omega$, 
-    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta, \omega) = \nabla_\theta \lik(\theta),$$
-    as we wanted. The result then follows by the continuous mapping theorem.
-
-    % Taking expectations on both sides and using the bounded convergence theorem to interchange limits and expectations, we see that
-    % $$\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)] = \E[\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta)] = \E[\nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta)] = \nabla_\theta \lim_{J \to \infty} \E[\hat\lik_J(\theta)].$$
-
-    % It then follows that
-    % $$\nabla_\theta\lik(\theta) 
-    % = \nabla_\theta \lim_{J\to\infty} \E[\hat\lik_J(\theta)] 
-    % = \lim_{J\to\infty} \nabla_\theta \E[\hat\lik_J(\theta)] 
-    % = \lim_{J\to\infty} \E[\nabla_\theta \hat\lik_J(\theta)].$$
-
-    % We then have mean-square consistency for the gradient of the likelihood estimate. The result then follows by the continuous mapping theorem.
-
-    \kevin{Explain that the difference in formula between this and Poyiadjis is due to us taking the derivative w.r.t. the particles in the chain rule. }
-
-    \kevin{Alternative argument: Show gradient is unbiased, after showing the likelihood is unbiased in an open neighborhood of phi.
-That should hold by Chopin's argument for unbiased resampling to nonconstant weights.
-Gradient unbiased if can swap expectation and derivative, I think with two continuous derivatives and boundedness, might need BCT/DCT.
-No need to swap limit and derivative in this case, or worrying about uniform limits.}
-
-\end{proof}
-
-\kevin{Show bias is order N, variance is order N when alpha below 1. That implies optimal choice of alpha is not 1. }
-
-\section{Analysis of MOP-$\alpha$ Error and Variance}
-
-\subsection{Analysis of MOP-$0$ Variance}
-
-We made Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} to leverage the results from \cite{karjalainen2023} on the forgetting of the particle filter.
-This is required to show error bounds on the gradient estimates that we provide -- namely that the error of the MOP-$1$ estimator that corresponds to that of \cite{poyiadjis11} is $O(N^2)$, and that the variance of MOP-$\alpha$ for any $\alpha<1$ is $O(N)$.
-
-We can decompose the MOP-$\alpha$ estimator as follows. When $\theta=\phi$,
-$$\hat{\mathcal{L}}(\theta):=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J w_{n, j}^{P, \theta}}=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J (w_{n-1, j}^{F, \theta})^\alpha}.$$
-
-To illustrate the proof strategy, we first analyze the special case when $\alpha=0.$ We then have:
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
-\end{equation}
-
-Similarly to the proof of Lemma \ref{lem:mop-0-formula}, its gradient when $\theta=\phi$ is therefore 
-\begin{align*}
-    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
-    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)} \\
-    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta s_{n,j}}{\sum_{j=1}^J s_{n,j}} \\
-    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)}{f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \phi}; \phi)} \\
-    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
-\end{align*}
-where we use the log-derivative trick, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again.
-The point here was that we establish that the gradient of the log-likelihood estimate is given by the sum of terms over all $N$ and $J$:
-$$\nabla_\theta \hat{\ell}(\theta) = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right).$$
-
-Therefore,
-\begin{align*}
-    \text{Var}(\nabla_\theta \hat\ell(\theta)) &= \frac{1}{J^2}\text{Var}\left(\sum_{n=1}^N\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &= \frac{1}{J^2}\sum_{n=1}^N\text{Var}\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &+ 2\sum_{m<n}\text{Cov}\left(\frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta)).
-\end{align*}
-
-Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
-We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
-$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
-i.e.
-the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
-
-\kevin{How does the dobrushin contraction coefficient relate to the alpha-mixing coefficients?
-If you have a bound on the mixing time $b$, then the alpha mixing coefficients should just be $\alpha(t) \leq e^{-t/b}$.}
-
-By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strongly mixing, with $\alpha$-mixing coefficients $\alpha(k) \leq (1-\epsilon)^{\floor{k/(c\log(J))}}$.
-Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
-Therefore, by Davydov's inequality, noting that $\nabla_\theta\hat\ell_n(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and WLOG assuming $\E[(\nabla_\theta\hat\ell_m(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/4}$, we see that
-\begin{align*}
-    \text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) 
-    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat\ell_m(\theta))^4]^{1/4}\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/4}\\
-    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2}.
-\end{align*}
-
-To bound this, we use the fact that $$\E[(\nabla_\theta\hat\ell_n(\theta))^4] = \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4]+\E[(\nabla_\theta\hat\ell_n(\theta))^2]^2$$
-alongside Lemma 2 of \cite{karjalainen2023}, which shows that
-$$\E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4] \lesssim \frac{G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^2] \lesssim \frac{G'(\theta)^2}{J},$$
-to find that 
-$$\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{G'(\theta)^4}{J^2}+\left(\frac{G'(\theta)^2}{J}\right)^2} = \frac{G'(\theta)^2}{J},$$
-and conclude that 
-$$\text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{G'(\theta)^2}{J}.$$
-
-Putting it all together, we see that
-\begin{align*}
-    \text{Var}(\nabla_\theta \hat\ell(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta))\\
-    &\leq \frac{NG'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\frac{G'(\theta)^2}{J} \\
-    &\leq \frac{G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\right) \\
-    &\lesssim \frac{G'(\theta)^2N}{J}.
-\end{align*}
-
-\paragraph{Remark:} Note that the factor of $N$ that pops up here is due to the use of the unnormalized gradient.
-If one divides the gradient estimate by $\sqrt{N}$ before usage, the variance does not depend on the horizon.
-If one divides by $N$, the error is $O(1/\sqrt{NJ})$!
-
-\kevin{https://math.stackexchange.com/questions/3274996/is-a-transformed-mixing-process-is-still-mixing}
-
-%$$\text{Cov}(x_{n,1:J}^P, x_{m, 1:J}^P) \leq (1-\epsilon)^{\floor{\frac{|n-m|}{c\log(J)}}}$$
-
-\kevin{Ed: Claim 11.4.2, Chopin, is the key to reasoning about this.
-Establish geometric mixing, that $\text{Cov}(\nabla_\theta\hat\ell_n, \nabla_\theta\hat\ell_m) \leq \rho^{|n-m|}V_n$, where $V_n$ is asymptotically stable (the latter only holds for $\alpha<1$, as only then do we get asymptotic stability, otherwise it increases linearly in $n$).
-Geometrically decreasing covariances and stable variances over time are the same thing.
-Then, we can bound the total variance by $\text{Var}(\sum_n \nabla_\theta \hat\ell_n(\theta)) = \sum_{m,n} \text{Cov}(\nabla_\theta\hat\ell_n(\theta), \nabla_\theta\hat\ell_m(\theta)) = NV_n + 2(N-1)\rho V_n + 2(N-3)\rho^2V_n + ...$}
-
-\subsection{Analysis of MOP-$\alpha$ Variance}
-
-Next, we analyze the case for $\alpha \in (0,1)$.
-The proof strategy is as follows.
-Instead of analyzing the variance of the MOP-$\alpha$ gradient estimate proper, we will analyze the variance of a modified estimator where we truncate the weights at $k$ timesteps, so this estimator only looks at most $k$ timesteps back while accumulating the importance weights.
-This enables us to establish strong mixing for the truncated estimator, in order to get a bound on the variance that is $O(NJ^{-1})$.
-We emphasize that this estimator is only a theoretical construct -- we do not actually use this estimator.
-
-Coupled with a bound ensuring that the $k$-truncated estimator provides an estimate that is close to that of MOP-$\alpha$ proper, we get a bound on the variance of MOP-$\alpha$ that comprises of the variance of the $k$-truncated estimator plus the error, for any $k \leq N$.
-It then holds that the final variance bound on MOP-$\alpha$ is given by the minimum over all $k$, and is never larger than $O(1+NJ^{-1})$.
-
-Using the log-derivative trick and that $w_{n,j}^{F,\theta} = w_{n,j}^{F,\theta,k} = 1$ when $\theta=\phi$,
-\begin{align*}
-    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,k}\right)\right\rVert_{\infty}\\
-    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta}}{{\sum_{j=1}^J w_{n,j}^{F,\theta}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,k}}}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,k}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta}}{w_{n,j}^{F,\theta}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,k}}{w_{n,j}^{F,\theta,k}}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,k}\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta}\right)-\log\left(w_{n,j}^{F,\theta,k}\right)\right)\right\rVert_{\infty}\\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=n-k}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^{n-k}\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}\\
-    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
-    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}G'(\theta)\\
-    &\leq G'(\theta)\frac{\alpha^k-\alpha^n}{1-\alpha},
-\end{align*}
-where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
-
-We can then bound $||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||$:
-\begin{align*}
-    &\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_\infty\\
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
-    &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\\
-    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)
-    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\Bigg)\right\rVert_\infty \\
-    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)\right)\right\lVert_{\infty}
-    +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\right)\right\rVert_\infty,
-\end{align*}
-and each of these two terms is bounded by $\frac{\alpha^k}{1-\alpha}NG'(\theta)$. So we now know that
-$$\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_\infty \leq  \frac{2\alpha^k}{1-\alpha}NG'(\theta),$$
-which is our desired error bound.
-
-Now, we bound the variance of $\hat s_k(\theta)$. 
-Recall that we defined
-$$\hat s_k(\theta) = \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) = \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right).$$
-
-We can then decompose this expression with the log-derivative trick while noting that $w_{n,j}^{F,\theta, k} = 1$ whenever $\theta=\phi$, to see that
-\begin{align*}
-    \hat s_k(\theta) 
-    &= \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right) \\
-    &= \sum_{n=1}^N \left(\frac{\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}}{\sum_{j=1}^J w_{n,j}^{F,\theta, k}}-\frac{\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}}{\sum_{j=1}^J w_{n-1,j}^{A,F,\theta, k}}\right) \\
-    &= \sum_{n=1}^N \left(\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}\right)\\
-    &= \frac{1}{J}\sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right).
-\end{align*}
-
-It now follows that we need to bound the variance at a single timestep, namely
-$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right).$$
-
-We use the log-derivative trick yet again to find, noting that $w_{n,j}^{F,\theta, k}=1$ when $\theta=\phi$, that
-\begin{align*}
-    \nabla_\theta w_{n,j}^{F,\theta, k} = \frac{\nabla_\theta w_{n,j}^{F,\theta, k}}{w_{n,j}^{F,\theta, k}} = \nabla_\theta \log(w_{n,j}^{F,\theta, k}) = \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) = \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right).
-\end{align*}
-
-Then,
-\begin{align*}
-    \text{Var}\left(\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right) 
-    &= \text{Var}\left(\nabla_\theta \left(\sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right)- \sum_{i=n-k-1}^{n-1} \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right)\right)\right) \\
-    &= \text{Var} \left(\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1}\log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right).
-\end{align*}
-
-Note that $$\hat s_{n,k}(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)$$ is a function bounded by $C\frac{G'(\theta)(1-\alpha^k)}{1-\alpha} =: CG'(\theta)r$ in each coordinate for some constant $C$ (by Assumption \ref{assump:bounded-measurement}) that depends only on $k$ timesteps, and not $n$.
-We can therefore invoke Lemma 2 of \cite{karjalainen2023} to bound the $L_2$ error of each of the $k$ (from time $n-k$ to time $n$) functionals from its expectation under the posterior by $CrG'(\theta)J^{-1/2}k$.
-That is, we have that
-$$\E\left[||\hat s_{n,k}(\theta) - \E_{\pi}\hat s_{n,k}(\theta)||_\infty^2\right]^{1/2} \leq \frac{Cr^2G'(\theta) k}{\sqrt{J}}.$$
-
-By the bias-variance decomposition, this in turn bounds the variance at a single timestep by $C^2r^2G'(\theta)^2J^{-1}k^2$. Concretely, we have that
-$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \lesssim \frac{r^2k^2G'(\theta)^2}{J}.$$
-
-It now remains to bound the variance of $\hat s_k(\theta)$ by considering the covariance of each of the $N$ terms that comprise it. We decompose
-\begin{align*}
-    \text{Var}(\nabla_\theta \hat s_k(\theta)) &= \text{Var}\left(\sum_{n=1}^N \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \\
-    &= \sum_{n=1}^N\text{Var}\left(\hat s_{n,k}(\theta)\right)+ 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \\
-    &\lesssim \frac{r^2k^2G'(\theta)^2N}{J} + 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right).
-\end{align*}
-
-Similarly to the proof of the MOP-$0$ case, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
-We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
-$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
-i.e.
-the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
-By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strongly mixing, with $\alpha$-mixing coefficients $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c\log(J))}}$.
-Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
-
-We now derive the $\alpha$-mixing coefficients of $(\hat s_{n,k}(\theta))_{n=1}^N$.
-Observe that $\hat s_{n,k}$ is strong mixing at lag $k+1$, as all weights beyond $k$ timesteps are truncated.
-We therefore have that the mixing time of $\hat s_k(\theta)$ is $O(1+k+\log(J))$, and that the $\alpha$-mixing coefficients for $(\hat s_{n,k}(\theta))_{n=1}^N$ are given by $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c(1+k+\log(J)))}}$.
-
-Therefore, by Davydov's inequality and Lemma 2 of \cite{karjalainen2023}, and noting that $\hat s_{n,k}(\theta) \leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, by a similar argument to the MOP-0 case, we find that
-
-$$\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J}.$$
-
-Concretely, noting that $\nabla_\theta\hat s_{n,k}(\theta)\leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and WLOG assuming $\E[(\nabla_\theta\hat\hat s_{m,k}(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat\hat s_{n,k}(\theta))^4]^{1/4}$, we apply Davydov's inequality to see that
-\begin{align*}
-    \text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) 
-    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat s_{m,k}(\theta))^4]^{1/4}\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/4}\\
-    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2}.
-\end{align*}
-
-To bound this, we use the fact that $$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4] = \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4]+\E[(\nabla_\theta\hat s_{n,k}(\theta))^2]^2$$
-alongside Lemma 2 of \cite{karjalainen2023}, which shows that
-$$\E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4] \lesssim \frac{r^4G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^2] \lesssim \frac{r^2G'(\theta)^2}{J},$$
-to find that 
-$$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{r^4G'(\theta)^4}{J^2}+\left(\frac{r^2G'(\theta)^2}{J}\right)^2} = \frac{r^2G'(\theta)^2}{J},$$
-and conclude that 
-$$\text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J}.$$
-
-Putting it all together, we see that
-\begin{align*}
-    \text{Var}(\nabla_\theta \hat s_{k}(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat s_{n,k}(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat s_{m,k}(\theta),\nabla_\theta\hat s_{n,k}(\theta))\\
-    &\leq \frac{Nr^2k^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J} \\
-    &\leq \frac{r^2k^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\right) \\
-    &\lesssim \frac{k^2r^2G'(\theta)^2N}{J}.
-\end{align*}
-
-We will now use this result to bound $\text{Var}(\nabla_\theta \hat s_k(\theta)$ proper. For random variables $X,Y$ where $|X-Y|$ is bounded almost surely by some $M$, we have that 
-$$X-\E X - 2M \leq Y - \E Y \leq X - \E X + 2M$$
-so,
-$$(Y-\E Y)^2 \leq (X-\E X - 2M)^2 + (X-\E X +2M)^2.$$
-Taking expectations, we get
-$$\text{Var}(Y) \leq \text{Var}(X) + 8M^2.$$
-
-It follows from this result and the result we proved earlier, namely that
-$$||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||_\infty \leq \frac{2\alpha^k}{1-\alpha}NG'(\theta),$$
-that the variance of MOP-$\alpha$ proper is bounded by, for any $k \leq N$,
-that the variance of MOP-$\alpha$ proper is bounded by
-$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \frac{k^2r^2G'(\theta)^2N}{J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2.$$
-
-As the above holds for any $k \leq N$,
-$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \min_{k\leq N} \left(\frac{k^2G'(\theta)^2N}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right).$$
-
-% To interpret this result, we can choose $k$ such that
-% \begin{align*}
-%     &\frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \leq 1,\\
-%     &\alpha^{2k} \leq \frac{(1-\alpha)^2}{16N^2G'(\theta)^2},\\
-%     &k \geq \frac{\log\left(16N^2G'(\theta)^2\right)-2\log(1-\alpha)}{2\log\left(\frac{1}{\alpha}\right)}.
-% \end{align*}
-
-% To interpret this result, if there exists some $k$ such that $$N \geq k \geq \frac{\log\left(16NG'(\theta)\right)-\log(1-\alpha)}{\log\left(\frac{1}{\alpha}\right)},$$
-% we can choose such a $k$, as well as $k=N$, to see that
-% \begin{align*}
-%     \text{Var}(\nabla_\theta \hat\ell(\theta)) 
-%     &\lesssim \min_{k\leq N}\left(\frac{c_d^2k^2r^2G'(\theta)^2N}{J} + \frac{\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right)\\
-%     &\lesssim \min\left\{1+\frac{c_d^2G'(\theta)^2N}{J}, \frac{c_d^2G'(\theta)^2r^2N^3}{J}\right\} \\
-%     &\leq \min\left\{1+\frac{c_d^2G'(\theta)^2N}{J}, \frac{c_d^2G(\theta)^2r^2N}{J}\right\},
-% \end{align*}
-% as $NG'(\theta)\leq G(\theta)$ as a consequence of Assumption \ref{assump:local-bounded-derivative}.
-
-\subsection{Analysis of MOP-$\alpha$ Bias}
-
-Using the log-derivative trick and that $w_{n,j}^{F,\theta,1} = w_{n,j}^{F,\theta,\alpha} = 1$ when $\theta=\phi$,
-\begin{align*}
-    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,1}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)\right\rVert_{\infty}\\
-    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,1}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,1}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}}}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,1}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,\alpha}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,1}}{w_{n,j}^{F,\theta,1}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,\alpha}}{w_{n,j}^{F,\theta,\alpha}}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,1}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha}\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta,1}\right)-\log\left(w_{n,j}^{F,\theta,\alpha}\right)\right)\right\rVert_{\infty}\\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=1}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}.
-\end{align*}
-
-Now we establish the bound:
-\begin{align*}
-     \nabla_\theta \left(\sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)
-     &\leq \sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
-    &\leq \sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)G'(\theta)\\
-    &\leq G'(\theta)\left(\frac{n-\alpha n + \alpha^n - 1}{1-\alpha}\right)
-\end{align*}
-where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
-
-By a similar decomposition to that of the error of the $k$-truncated gradient to the MOP-$\alpha$ gradient estimate,
-\begin{align*}
-    &\left\lVert\nabla_\theta\hat\ell^1(\theta) - \nabla_\theta\hat\ell^\alpha(\theta) \right\rVert_\infty\\
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
-    &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\\
-    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)
-    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right)\Bigg)\right\rVert_\infty \\
-    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)\right|
-    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right)\right|\Bigg)\right\rVert_\infty.
-\end{align*}
-
-From our arguments above, we know that 
-$$\left\lVert\nabla_\theta\hat\ell^1(\theta) - \nabla_\theta\hat\ell^\alpha(\theta) \right\rVert_\infty \leq 2NG'(\theta)\left(\frac{N-\alpha N + \alpha^N - 1}{1-\alpha}\right).$$
-
-So the bias of MOP-$\alpha$ is the bias of MOP-$1$ arising from the logarithm in the gradient of the log-likelihood plus no more than $$2NG'(\theta)\left(\frac{N-\alpha N + \alpha^N - 1}{1-\alpha}\right),$$
-where we note that the bias vanishes as $\alpha$ goes to $1$ as
-$$\lim_{\alpha\to1}\left(\frac{N-\alpha N + \alpha^N - 1}{1-\alpha}\right) = 0.$$
-
-\subsection{An Improved Bound by Mixing}
-
-We decompose
-$$\E||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty \leq \E||\nabla_\theta\ell_\theta - \nabla_\theta\hat s_k^1(\theta)||_\infty + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty.$$
-
-The broad idea is that the first term can be controlled by a mixing argument, the second term garners $O(Nk^2)$ error, and the third term is $O\left(\frac{N}{1-\alpha}\right)$.
-It is unclear whether the bias can be reduced further -- the gradient is a sum of $N$ terms, and is therefore $O(N)$ itself.
-
-The first term -- bound ground truth of conditional scores from particle approximation conditional on correct filtering distribution at time $n-k$, bound that from particle approximation conditional on arbitrary filtering distribution at time $n-k$.
-First term sums up $k$ terms, $O(Nk^2)$ error, second term bounded by $k$ applications of the Markov kernel, so just apply forgetting of the particle filter result.
-
-But you can get an MSE bound like this! Simply decompose
-$$||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 \leq ||\nabla_\theta\ell_\theta - \nabla_\theta\hat s_k^1(\theta)||_2^2 + ||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2 + ||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_2^2,$$
-and observe that 
-$$||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty =  \left\lVert\sum_{n=1}^N \nabla_\theta \hat\ell_n^\alpha(\theta) - \sum_{n=1}^N \nabla_\theta \ell_n(\theta)\right\rVert_\infty \leq \sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_\infty.$$
-
-\section{Finite-Particle Analysis of Gradient Methods with Algorithm \ref{alg:ifad}}
-\label{app:ifad}
-
-The analysis in this section roughly follows the analysis in \cite{mahoney16}, except with the caveat that none of the matrix concentration bounds they use apply here as the particles are dependent.
-We instead use the concentration inequality from \cite{delmoral2011ci} to bound the gradient and Hessian estimates.
-In this section, we fix $\omega \in \Omega$ only within each filtering iteration, evaluate Algorithm \ref{alg:mop} at $\theta=\phi$, and analyze Algorithm \ref{alg:ifad} post-iterated filtering.
-
-The convergence analysis in Theorem \ref{thm:convergence} is limited to the case where $-\ell$ is $\gamma$-strongly convex.
-Though it is true that in a neighborhood of the optimum local asymptotic normality holds and the log-likelihood is strongly convex in this neighborhood, in practice likelihood surfaces for POMPs are often highly nonconvex globally.
-The convergence to an optimum, local or global, must therefore be sensitive to initialization.
-
-\subsection{Bounding the Gradient}
-\kevin{Add a few words to explain that the gradient of the log-likelihood w.r.t.
-parameters is a functional of the particle measure under boundedness assumptions and fixed parameters, appeal to authority and say doucet did it too.
-}
-
-\kevin{Gradient linear map, bound conditional likelihoods at time n.
-Alternatively, convert del moral and doucet expectation bound to high-probability bound.
-https://arxiv.org/pdf/1905.11546.pdf for trick.
-Alternatively, concatenate previous states.
-}
-
-\begin{lem}[Concentration of Measure for Gradient Estimate]
-    \label{lemma:grad_bound}
-    Consider the gradient estimate obtained by MOP-1, which we know by Lemma \ref{lem:mop-grad-consistency} is consistent for the score, where $\theta = \phi$.
-For $||\nabla_\theta \hat{\loglik}(\theta) - \nabla_\theta \loglik(\theta)||_2$ to be bounded by $\epsilon$ with probability $1-\delta$, we require
-    \begin{align}
-    J > \max\left\{2G(\theta)\frac{r_N\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{2p}{\delta}\right)\right)\right), 8G(\theta)^2\beta_N^2\frac{p\log(2p/\delta)}{\epsilon^2}\right\},
-    \end{align}
-    where $NG'(\theta) \leq G(\theta)$ are defined in Assumptions \ref{assump:bounded-measurement} and \ref{assump:local-bounded-derivative}, $h(t) = \frac{1}{2}(t - \log(1+t))$, and $\beta_N$ and $r_N$ are two additional finite model-specific constants that do not depend on $J$, but do depend on $N$ and $p$, as defined in \cite{delmoral2011ci}.
-
-    Equivalently, with probability at least $1-\delta$, it holds that
-    \begin{align}
-        ||\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).
-    \end{align}
-\end{lem}
-
-\paragraph{Remark:} According to \cite{delmoral2011ci}, under some regularity conditions, $r_N$ and $\beta_N$ are linear in the trajectory length $N$.
-This corresponds to the finding by \cite{poyiadjis11} that the variance of the estimate is at least quadratic in the trajectory length, and their remark that the result of \cite{delMoral03} establishes that the $L_p$ error is bounded by $O(N^2J^{-1/2})$ (equivalently, the variance is bounded by $O(N^4J^{-1})$) after accounting for the sum over timesteps.
-The MOP-$1$ variance upper bound is therefore in fact $O(N^4)$, in contrast to the MOP-$\alpha$, where $\alpha<1$, upper bound of $O(N)$.
-
-Either way, usually in practice one considers the estimate normalized through dividing it by $N$, in which case we only need
-$$J > \max\left\{2G'(\theta)\frac{r_N\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{2p}{\delta}\right)\right)\right), 8G'(\theta)^2\beta_N^2\frac{p\log(2p/\delta)}{\epsilon^2}\right\}.$$
-Equivalently,
-$$||N^{-1}(\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta))||_2 \leq G'(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).$$
-
-It is well-known that bounds like these hide an exponential dependence in the dimension of the state space $d$.
-We take care to keep track of the terms depending on $d$ when they arise in front of us, but are unable to do so for the terms that arise from the analysis of \cite{delmoral2011ci}.
-
-\kevin{This is slightly alarming, in that the estimate really has $O(N^4)$ variance after accounting for summing over timesteps.
-In contrast, MOP-0 has only $O(N)$ variance after summing over timesteps.
-}
-
-\begin{proof}
-
-We will seek to use the concentration inequality of \cite{delmoral2011ci} to bound the deviation of the gradient estimate from the gradient of the negative log-likelihood in the sup norm with a union bound.
-Fix some $\theta \in \gN(\phi)$.
-
-Recall that as $w_{N, j}^{F, \theta}=1$ when $\theta=\phi$,
-$$
-\nabla_\theta \hat{\ell}(\theta):=\frac{1}{J} \sum_{j=1}^J \nabla_\theta w_{N, j}^{F, \theta}=\frac{1}{J} \sum_{j=1}^J\sum_{n=1}^N  w_{N, k_j}^{P, \theta} \nabla_\theta \log g_{n,k_j}^{A,\theta} = \frac{1}{J} \sum_{j=1}^J\sum_{n=1}^N \nabla_\theta \log g_{n,k_j}^{A,\theta}.
-$$
-
-Define $\varphi_n^i(x_{n,j}^{F,\theta}) := \frac{\partial}{\partial\theta_i} \log g_{n,k_j}^{A,\theta}$, which is a functional of the filtering particles $x_{n,j}^{F,\theta} = x_{n,k_j}^{P,\theta}$.
-These are bounded measurable functionals bounded by $G'(\theta)$ by Assumption \ref{assump:bounded-measurement}.
-Therefore, these have bounded oscillation, satisfying the requirement that $\text{osc} \left(\frac{\partial}{\partial\theta_i} \varphi_i(x_{n,j}^{P,\theta}) \right) \leq G'(\theta)$.
-\cite{delmoral2011ci} in fact assume $\text{osc}(f) \leq 1$, so we simply scale their bound accordingly.
-
-\kevin{Put some justification/make the setup of delmoral2011ci concrete here. Concentration of gradient estimate under particle measure to limit under posterior measure}
-
-Apply the Hoeffding-type concentration inequality from Del Moral and Rio \cite{delmoral2011ci} and a union bound over each $\varphi_n^i(x_{n,j}^{F,\theta})$, totaling $N$ timesteps and $p$ parameters, to find that
-
-\begin{align}
-    \max_{n=1,...,N} \left\lVert\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log g_{n,k_j}^{A,\theta} - \nabla_\theta \ell_n(\theta) \right\rVert_{\infty} \leq G'(\theta)\left(\frac{r_N}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_N \right)
-\end{align}
-
-with probability at least $1-2Np\exp(-t)$.
-Note that the concentration inequality only considers the error from the expectation under the filtering distribution -- but we invoke the consistency of MOP-$1$ shown in Lemma \ref{lem:mop-grad-consistency} to establish that the expectation under the filtering distribution is in fact the score.
-It therefore holds that with the same probability, that when summing over $N$, as $NG'(\theta) \leq G(\theta)$,
-
-\begin{align}
-    \left\lVert\frac{1}{J}\sum_{j=1}^J\sum_{n=1}^N\nabla_\theta \log g_{n,k_j}^{A,\theta} - \nabla_\theta \ell(\theta) \right\rVert_{\infty} 
-    &\leq G'(\theta)N\left(\frac{r_N}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_N \right)\\
-    &\leq G(\theta)\left(\frac{r_N}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_N \right)
-\end{align}
-
-We split the $\delta$ failure probability among these $2Np$ terms, to find $\delta\leq2Np\exp(-t) => t\leq\log(2Np/\delta)$.
-Here, $h(t) = \frac{1}{2}(t - \log(1+t))$.
-The two additional model-specific parameters are $\beta_t$ and $r_t$, which do not depend on $J$.
-
-The analogous bound for the 2-norm follows from scaling the right-hand side by $\sqrt{p}$, to require
-
-\begin{align}
-    ||\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right)
-\end{align}
-
-and we therefore need 
-\begin{align}
-    J > \max\left\{2G(\theta)\frac{r_N\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{2p}{\delta}\right)\right)\right), 8G(\theta)^2\beta_N^2\frac{p\log(2p/\delta)}{\epsilon^2}\right\}.
-\end{align}
-
-\end{proof}
-
-\subsection{Bounding Hessian Estimates}
-
-Should one choose to use a second-order method involving a particle Hessian estimate, we provide a guarantee for its positive-definiteness below.
-
-\begin{lem}[Minimum Eigenvalue Bound for Hessian Estimate]
-    \label{lemma:hess_bound}
-    Assume that the Hessian of the negative log-likelihood $H=\sum_{j=1}^J \E H_j$ has a minimum eigenvalue $0<\gamma<1$, and that $\E \lambda_{\min} (H_j) = \gamma' > 0$.
-
-    We then require 
-    \begin{equation}
-        J > \max\left\{\frac{2r_t(1+h^{-1}(t)) + 2c}{\gamma'}, \frac{2(2t\beta_t^2+c)^2}{\gamma'^2}\right\} \geq  \frac{r_t(1+h^{-1}(t))}{\gamma'} + \sqrt{2tJ}\beta_t/\gamma' + c/\gamma'
-    \end{equation}
-
-    for $\hat{H}(\theta)$ to be invertible and positive definite with minimum eigenvalue greater than or equal to $c \in (0, \sum_{j=1}^J \E\lambda_{\min}(H_j))$ and with probability at least $1-\exp(-t)$.
-\end{lem}
-\begin{proof}
-Write $\hat{H}(\theta) = \hat{H} = \sum_{j=1}^J H_j$ for the estimate of the negative of the Hessian, where $H_j$ is an element of the outer sum over the $J$ particles.
-
- As the negative log-likelihood is convex, we want to bound the minimum eigenvalue of $\hat{H}(\theta)$ from below with high probability, so that all the eigenvalues of $\hat{H}(\theta)$ are positive with high probability.
-This ensures that the estimated Hessian is invertible and positive-definite.
-
-From \cite{boyd2004convex}, we know that the minimum eigenvalue of a symmetric matrix is concave. Therefore, it suffices to show that the first inequality in the below expression
-
-\begin{equation}
-    0 < \sum_{j=1}^J \lambda_{\min} (H_j) \leq  \lambda_{\min}\left(\sum_{j=1}^J H_j\right) = \lambda_{\min} (\hat{H})
-\end{equation}
-
-holds with high probability.
-
-Apply the particle Hoeffding concentration inequality from \cite{delmoral2011ci} to find that
-
-\begin{align}
-    \frac{1}{J}\sum_{j=1}^J \lambda_{\min}(H_j) - \E_{\tilde{\pi}_t}\lambda_{\min}(H_j) &= \frac{1}{J}\sum_{j=1}^J \lambda_{\min}(H_j) - \gamma' \geq -\frac{r_t}{J}(1+h^{-1}(t)) - \sqrt{\frac{2t}{J}}\beta_t \\
-    \sum_{j=1}^J \lambda_{\min}(H_j) &\geq -r_t(1+h^{-1}(t)) - \sqrt{2tJ}\beta_t + J\gamma'
-\end{align}
-
-with probability at least $1-\exp(-t)$. Here, $h(t) = \frac{1}{2}(t - \log(1+t))$. The two additional model-specific parameters are $\beta_t$ and $r_t$, which do not depend on $J$.
-
-We additionally require, for $c \in (0, \sum_{j=1}^J \E\lambda_{\min}(H_j))$,
-\begin{align}
-    \sum_{j=1}^J \lambda_{\min}(H_j) \geq -r_t(1+h^{-1}(t)) - \sqrt{2tJ}\beta_t + J\gamma' \geq c \\
-    J\gamma' \geq c + r_t(1+h^{-1}(t)) + \sqrt{2tJ}\beta_t 
-\end{align}
-
-We therefore require $$J > \max\left\{\frac{2r_t(1+h^{-1}(t)) + 2c}{\gamma'}, \frac{2(2t\beta_t^2+c)^2}{\gamma'^2}\right\} \geq  \frac{r_t(1+h^{-1}(t))}{\gamma'} + \sqrt{2tJ}\beta_t/\gamma' + c/\gamma'$$
-
-for $\hat{H}(\theta)$ to be invertible and positive definite with minimum eigenvalue greater than or equal to $c$ with probability at least $1-\exp(-t)$.
-
-\end{proof}
-
-\subsection{Convergence Analysis of Theorem \ref{thm:convergence}}
-
-\begin{proof}
-
-In this analysis, we largely follow the proof of Theorem 6 in \cite{mahoney16}.
-
-Define $\theta_\eta = \theta_n + \eta p_n$, where $p_n=-(H(\theta_n))^{-1}g(\theta_n)$.
-As in Roosta-Khorasani and Mahoney \cite{mahoney16}, we want to show there is some iteration-independent $\tilde{\eta}>0$ such that the Armijo condition
-\begin{equation}
-    f(\theta_n+\eta p_n) \leq f(\theta_n) + \eta\beta p_n^Tg(\theta_n)
-\end{equation}
-holds for any $0< \eta < \tilde{\eta}$ and some $\beta \in (0,1)$.
-
-By an argument found in the beginning of the proof of Theorem 6 in \cite{mahoney16}, we have that choosing $J$ such that $||\nabla_\theta\hat{\ell}(\theta_n) - \nabla_\theta \ell(\theta_n)|| \leq \epsilon$ and $\lambda_{\min}(H(\theta_n)) \geq c$ with probability $(1-\delta)^2$ for each $n$ yields
-
-\begin{align}
-    f(\theta_\eta)-f(\theta_n) \leq \eta p_n^Tg(\theta_n) + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2
-\end{align}
-
-Now, for some $c>0$, with probability $1-\delta$
-\begin{equation}
-    p_n^Tg(\theta_n) = -p_n^TH(\theta_n)p_n \geq -c||p_n||^2
-\end{equation}
-and we can obtain a decrease in the objective.
-
-Substituting this into the previous expression,
-\begin{align}
-    f(\theta_\eta)-f(\theta_n) \leq -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2
-\end{align}
-
-and the Armijo condition then becomes
-\begin{align}
-    -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2 &\leq \eta \beta p_n^Tg(\theta_n) = - \eta \beta p_n^TH(\theta_n)p_n \\
-    \epsilon||p_n|| + \eta \Gamma ||p_n||^2 / 2 &\leq (1- \beta) p_n^TH(\theta_n)p_n \\
-    \epsilon + \eta \Gamma ||p_n|| / 2 &\leq c(1- \beta) ||p_n||
-\end{align}
-
-which holds and guarantees an iteration-independent lower bound if 
-\begin{equation}
-    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_n)|| \leq \frac{c(1-\beta)}{2}||p_n||
-\end{equation}
-
-Now, first note that
-\begin{equation}
-||g(\theta_n)|| - ||\nabla_\theta f(\theta_n)|| \leq ||g(\theta_n) - \nabla_\theta f(\theta_n)|| \leq \epsilon => ||\nabla_\theta f(\theta_n)|| \geq ||g(\theta_n)|| - \epsilon
-\end{equation} and
-\begin{equation}
-||\nabla_\theta f(\theta_n)||-||g(\theta_n)|| \leq ||\nabla_\theta f(\theta_n)-g(\theta_n)|| \leq \epsilon => ||g(\theta_n)|| \geq ||\nabla_\theta f(\theta_n)|| - \epsilon
-\end{equation}
-
-There are now two cases. If the algorithm terminates and $||g(\theta_n)|| \leq \sigma \epsilon$, we can derive
-
-\begin{equation}
-    ||\nabla_\theta f(\theta_n)|| \leq ||g(\theta_n)|| + \epsilon = \sigma\epsilon+\epsilon = (\sigma+1)\epsilon
-\end{equation}
-
-If the algorithm does not terminate, then $||g(\theta_n)|| > \sigma \epsilon$. Notice that
-
-\begin{align}
-    \epsilon \geq ||g(\theta_n) - \nabla_\theta f(\theta_n)|| &\geq ||g(\theta_n)|| - ||\nabla_\theta f(\theta_n)|| \\
-    ||\nabla_\theta f(\theta_n)|| + \epsilon &\geq ||g(\theta_n)|| \geq \sigma \epsilon \\
-    ||\nabla_\theta f(\theta_n)|| &\geq \sigma \epsilon - \epsilon = (\sigma - 1)\epsilon \\
-    \frac{||\nabla_\theta f(\theta_n)||}{\sigma-1} &\geq \epsilon
-\end{align}
-
-and now 
-\begin{align}
-    ||\nabla_\theta f(\theta_n)|| - \epsilon
-    &\geq  ||\nabla_\theta f(\theta_n)|| - \frac{||\nabla_\theta f(\theta_n)||}{\sigma-1} = \left(1-\frac{1}{\sigma-1}\right)||\nabla_\theta f(\theta_n|| = \frac{\sigma-2}{\sigma-1}||\nabla_\theta f(\theta_n)|| \geq \frac{2}{3}||\nabla_\theta f(\theta_n)||
-\end{align}
-
-From $||A^{-1}|| = 1/\sigma_{\min}(A)$,
-\begin{align}
-    p_n^TH(\theta_n)p_n &= (-(H(\theta_n))^{-1}g(\theta_n))^TH(\theta_n)(-(H(\theta_n))^{-1}g(\theta_n)) \\
-    &= g(\theta_n)^T(H(\theta_n))^{-1}g(\theta_n) \\
-    &\geq \frac{1}{c}||g(\theta_n)||^2 \\
-    &\geq \frac{1}{c}(||\nabla_\theta f(\theta_n)|| - \epsilon)^2 \\
-    &\geq \frac{4}{9c}||\nabla_\theta f(\theta_n)||^2
-\end{align}
-
-From the assumption that $f$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 -\ell \preceq \Gamma I$, by an implication of $\gamma$-strong convexity we have
-\begin{align}
-    f(\theta_n) - f(\theta^*) \leq \frac{1}{2\gamma}||\nabla_\theta f(\theta_n)||^2
-\end{align}
-and we put together:
-\begin{equation}
-    f(\theta_n) - f(\theta^*) \leq \frac{1}{2\gamma}||\nabla_\theta f(\theta_n)||^2 \leq \frac{9c}{4}\frac{1}{2\gamma}p_n^TH(\theta_n)p_n
-\end{equation}
-\begin{equation}
-    \frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \leq \frac{4}{9c}||\nabla_\theta f(\theta_n)||^2 \leq p_n^TH(\theta_n)p_n
-\end{equation}
-\begin{equation}
-    f(\theta_n) - f(\theta^*) \leq \frac{9c}{8\gamma}p_n^TH(\theta_n)p_n
-\end{equation}
-
-\begin{equation}
-    -\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \geq -\frac{4}{9c}||\nabla_\theta f(\theta_n)||^2 \geq -p_n^TH(\theta_n)p_n
-\end{equation}
-
-But from earlier, as the Armijo condition is fulfilled with our choice of $\eta$ and $\epsilon$,
-\begin{align}
-    f(\theta_{n+1})-f(\theta_n) &\leq -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2 \\
-    &\leq -\eta\beta p_n^TH(\theta_n)p_n \\
-    &\leq -\eta\beta\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) 
-\end{align}
-
-and therefore
-\begin{align}
-    f(\theta_{n+1}) - f(\theta^*) 
-    &= f(\theta_{n+1})-f(\theta_n)+f(\theta_n)- f(\theta^*) \\
-    &\leq f(\theta_n)- f(\theta^*) -\eta\beta\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \\
-    &= (1-\eta\beta\frac{8\gamma}{9c})(f(\theta_n) - f(\theta^*)) 
-\end{align}
-
-\end{proof}
-
-\section{Miscellanous Technical Lemmas}
-
-\begin{lem}
-    \label{lem:dobrushin-implies-alpha-mixing}
-    Setting $X$ to be the particle collection at time $m$, and $Y$ at time $n$, we have that the alpha mixing coefficients,
-    $$\int | f_{XY} - f_X f_Y | < \alpha,$$
-    are bounded by the Dobrushin coefficient
-    $$\int | f_{Y|X}(y|x_1) - f_{Y|X}(y|x_2) |dy < \alpha$$ for all $x_1$, $x_2$.
-\end{lem}
-
-\begin{proof}
-    We rewrite the alpha-mixing assertion as $$\int { \int |f_Y|X(y|x) - f_Y(y)| dy } f_X dx.$$
-
-    We claim that the Dobrushin coefficient implies $$\int | f_Y|X(y|x_1) - f_Y(y) |dy < \alpha$$ for all $x_1$. This is shown as follows:
-    \begin{align*}
-        \int | f_{Y|X}(y|x) - f_Y(y) |dy 
-        &= \int \left|  \int [f_{Y|X}(y|x_1) - f_{Y|X}(y|x)] f_X(x)dx \right|dy \\
-        &< \int   \int  | f_{Y|X}(y|x_1) - f_{Y|X}(y|x)] | dy f_X(x)dx \\
-        &< \int \alpha f_X(x)dx \\
-        &=\alpha
-    \end{align*}
-
-    We then have the desired result.
-\end{proof}
-
-\end{document}

### 2024-06-10 paper/ms.tex
Updates from Overleaf

@@
 The sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$ by Assumption \ref{assump:local-bounded-derivative}.
-Again by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J^1(\theta)(\omega)|_{\theta=\theta'}$ is also bounded by $H'$ for almost every $\omega\in \Omega$ and every $\theta'\in \Theta$.
+Again by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J^1(\theta)(\omega)|_{\theta=\theta'}$ is also bounded by $H^*$ for almost every $\omega\in \Omega$ and every $\theta'\in \Theta$.
 So $(\nabla_\theta \hat\lik_J^1(\theta)( \omega))_{J \in \mathbb{N}}$ is uniformly Lipschitz, and therefore uniformly equicontinuous for almost every $\omega \in \Omega$.
@@
 
+\section{Feynman-Kac Models and Monte Carlo Approximations}
+\input{appendix/feynman}
+
 \section{Weighted Particle Filters that Target the Posterior}

### 2024-06-11 paper/appendix/feynman.tex
Kevin: appendices, etc.

@@
 In this section, we introduce the Feynman-Kac convention of \cite{delMoral04} that has since become commonplace \cite{karjalainen23} for the analysis of the particle filter.
-The mathematical formalization and notation introduced here will be adopted
+The mathematical formalization and notation introduced here will be adopted in the remainder of the analysis, in order to prove Theorems \ref{thm:mop-targeting}, \ref{thm:mop-grad-consistency}, and \ref{thm:mop-biasvar}.
+Let $\pi_n$ be a sequence of probability measures on the state space $\gX$.
+This is the sequence of posterior distributions $f_{X_{1:n}|Y_{1:n}}$ that we seek to approximate with the particle filter.
+For any measurable bounded functional $h$, we adopt the following functional-analytic notation, borrowed from \cite{delMoral04, chopin20, karjalainen23}.
+
+\begin{itemize}
+\item For any measure $\eta$, any Markov kernel $M$ on $\gX$, any point $x \in \gX$ and any subset $A \subseteq \gX$, let 
+\begin{align*}
+    \eta(h) &= \int h d\eta = \int h(x) \eta(dx), \\(\eta M)(A) &= \int \eta(dx)M(x,A), \\
+    (Mh)(x) &= \int M(x, dy) h(y).
+\end{align*}
+\item The composition of a Markov kernel $M_1$ with another Markov kernel $M_2$ is another Markov kernel, given by 
+$$(M_1M_2)(x, A) = \int M_1(x, dy) M_2(y, A).$$
+\item The total variation distance between two measures $\mu$ and $\nu$ on $\gX$ is
+$$\|\mu-\nu\|_{\mathrm{TV}}=\sup _{\|h\|_{\infty} \leq 1 / 2}|\mu(\phi)-\nu(\phi)|=\sup _{\operatorname{osc}(h) \leq 1}|\mu(h)-\nu(h)|.$$
+\item The Dobrushin contraction coefficient $\beta_{\text{TV}}$ of a Markov kernel $M$ is given by
+$$\beta_{\mathrm{TV}}(M)=\sup _{x, y \in \gX}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}.$$
+\end{itemize}

### 2024-06-11 paper/ms.tex
Kevin: appendices, etc.

@@
 % Use the lineno option to display guide line numbers if required.
+
+\usepackage{booktabs}
 
@@
 \centering
-\begin{tabular}{||c c c||} 
- \hline
- Method & Best Log-Likelihood & Rank \\ [0.5ex] 
- \hline\hline
-     IFAD-0 & -3747.87 & 1\\
-     IFAD-1 & -3748.53 & 2\\
-     IF2 (Ours) & -3755.91 & 3\\
-     IF2 (Warm Start) & -3759.60 & 4 \\
-     IF2 2015 & -3768.63 & 5\\ 
-     MOP-0 Alone & -3795.74 & 6\\
-     MOP-1 Alone & -3797.38 & 7\\
- \hline
-\end{tabular}
+\input{imgs/095/table}
 \caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone.
@@
 Meanwhile, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.6$, while the best log-likelihood found by \cite{king08} was only $-3793.4$.
-Despite being initialized for a global search, IFAD manages to find the MLE over the 100 searches, as seen in Table \ref{table:mle}.
+Despite being initialized for a global search, IFAD manages to get much closer to the MLE over the 100 searches than \cite{ionides15} and finds it up to Monte Carlo error, as seen in Table \ref{table:mle}.
 On this problem, the sequence of local searches and refinement that was previously required for finding the MLE is not necessary with the IFAD algorithm.
 
-\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations, performing IF2 alone ultimately fails to achieve the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter-optim}).
+\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations, performing IF2 alone ultimately fails to achieve the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter}).
 Conversely, when we tried gradient descent with MOP alone, we encountered many failed searches.
@@
 
-IFAD, in comparison, approaches the MLE quickly due to the IF2 warm-start (as seen in Figure \ref{fig:scatter-optim}) and also succeeds at refining the coarse solution found by the warm-start with MOP gradient steps to find the MLE (as seen in Figures~\ref{fig:scatter-optim} and \ref{fig:boxplot-search}.
+IFAD, in comparison, approaches the MLE quickly due to the IF2 warm-start (as seen in Figure \ref{fig:optim}) and also succeeds at refining the coarse solution found by the warm-start with MOP gradient steps to find the MLE (as seen in Figures~\ref{fig:scatter} and \ref{fig:boxplot}).
 IFAD therefore successfully combines the best qualities of IFAD and MOP, outperforming either of them alone.
-Monte~Carlo replication is appropriate for IFAD, as for any Monte Carlo algorithm used to solve a challenging numerical problem, but the left panel of Figure~\ref{fig:boxplot-search} shows that IFAD can find higher likelihood values, more quickly and more reliably than the previous state-of-the-art.
+Monte~Carlo replication is appropriate for IFAD, as for any Monte Carlo algorithm used to solve a challenging numerical problem, but Figure~\ref{fig:boxplot} shows that IFAD can find higher likelihood values, more quickly and more reliably than the previous state-of-the-art.
 \ed{I ADDED THIS PARAGRAPH TO TRY TO BE EXPLICIT ABOUT WHY WE ARE EXCITED BY THIS FIGURE.
@@
 }
-    \label{fig:scatter-optim}
+    \label{fig:scatter}
 \end{figure}
@@
     \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/boxplot.png}
-    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/optim.png}
-    \caption{\textbf{Left:} Boxplot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs, simulating the result of the common procedure of running a few searches and choosing the best one.
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/boxplot_all.png}
+    \caption{\textbf{Left:} Raincloud plot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs, simulating the result of the common procedure of running a few searches and choosing the best one.
+\textbf{Right:} Raincloud plot of all searches.
 IFAD outperforms all other methods, and the gradient steps improve on the warm-start given by running 40 IF2 iterations.
-    \textbf{Right:} Optimization progress of IFAD and IF2.
+    We use a dotted red line to display the MLE.}
+    \label{fig:boxplot}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{imgs/095/optim.png}
+    \caption{Optimization progress of IFAD and IF2.
 The dashed orange line depicts the median warm-start given by running 40 IF2 iterations.
@@
     We use a dotted red line to display the MLE.}
-    \label{fig:boxplot-search}
+    \label{fig:optim}
 \end{figure}

### 2024-06-11 paper/appendix/targeting.tex
a few typos, a minor title change, a note in appendix D

@@
 \end{defn}
-\cite{chopin04} asserted without proof that common particle filter algorithms target the filtering distribution, $f_{X_{1:N}|Y_{1:N}}$, in this sense.
+\cite{chopin04} \ed{and \cite{liu01}????} asserted without proof that common particle filter algorithms target the filtering distribution, $f_{X_{1:N}|Y_{1:N}}$, in this sense.
+\ed{\cite{chopin20} proved a related result assuming bounded densities.
+  We follow a similar approach to \cite{chopin20}, based on a strong laws of large numbers for triangular arrays.
+  Triangular array strong laws do not hold without an additional regularity condition such as boundedness.}
+\ed{I PROPOSE ADDING SOMETHING LIKE THIS, PARTLY TO MOTIVATE THE RESULTS, PARTLY TO BE KINDER TO CHOPIN.}  
 In order to prove the consistency of our variation on the particle filter, we now present three helper lemmas.
@@
     \end{eqnarray*}
-    Applying (\ref{eq:lemma2:g}) with $g(x) = \E\big[h(Z) | X=x\big]$, the average on the right hand size converges almost surely to $\E\big\{\E[h(Z)|X]\big\}=\E[h(Z)].$
+    Applying (\ref{eq:lemma2:g}) with $g(x) = \E\big[h(Z) | X=x\big]$, the average on the right hand side converges almost surely to $\E\big\{\E[h(Z)|X]\big\}=\E[h(Z)].$
     It remains to swap the limit and expectations. We can do so with the bounded convergence theorem, and therefore obtain   

### 2024-06-11 paper/ms.tex
a few typos, a minor title change, a note in appendix D

@@
 
-\title{Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation}
+\title{Accelerated Inference for Partially Observed Stochastic Processes using Automatic Differentiation}
 
@@
 Scibior and Wood \cite{scibior21} showed that the estimators derived by Poyiadjis et al.
-\cite{poyiadjis11} can be attained with standard AD software, using an algorithmic prodecure called a ``stop-gradient'' which allows selected exressions to be evaluated but not differentiated.
+\cite{poyiadjis11} can be attained with standard AD software, using an algorithmic procedure called a ``stop-gradient'' which allows selected exressions to be evaluated but not differentiated.
 Unfortunately, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability.
@@
 \centering
-\input{imgs/095/table}
+\input{../imgs/095/table}
 \caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone.
@@
     \centering
-    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{imgs/095/optim.png}
+    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{../imgs/095/optim.png}
     \caption{Optimization progress of IFAD and IF2.

### 2024-06-11 paper/appendix/feynman.tex
ready to merge Kevin and Ed edits

@@
 \begin{itemize}
-\item For any measure $\eta$, any Markov kernel $M$ on $\gX$, any point $x \in \gX$ and any subset $A \subseteq \gX$, let 
+\item \textbf{Markov kernels and the process model:} A Markov kernel $M$ with source $\gX_1$ and target $\gX_2$ is a map $M: \gX_1 \times \mathcal{B}(\gX_2) \to [0,1]$ such that for every set $A \in \mathcal{B}(\gX_2)$ and every point $x \in \gX_1$, the map $x \mapsto M(x, A)$ is a measurable function of $x$, and the map $A \mapsto M(x,A)$ is a probability measure on $\gX_2$.
+The quantity $M(x,A)$ can be thought of as the probability of transitioning to the set $A$ given that we are at the point $x$.
+If this yields a density, this then corresponds to the process density $f_{X_{2}|X_1}$ conditional on $x$ and integrated over $A$.
+\item \textbf{Markov kernels and measures:} For any measure $\eta$, any Markov kernel $M$ on $\gX$, any point $x \in \gX$ and any measurable subset $A \subseteq \gX$, let 
 \begin{align*}
@@
 \end{align*}
-\item The composition of a Markov kernel $M_1$ with another Markov kernel $M_2$ is another Markov kernel, given by 
+\item \textbf{Compositions of Markov kernels:} The composition of a Markov kernel $M_1$ with another Markov kernel $M_2$ is another Markov kernel, given by 
 $$(M_1M_2)(x, A) = \int M_1(x, dy) M_2(y, A).$$
-\item The total variation distance between two measures $\mu$ and $\nu$ on $\gX$ is
+\item \textbf{Total variation distance:} The total variation distance between two measures $\mu$ and $\nu$ on $\gX$ is
 $$\|\mu-\nu\|_{\mathrm{TV}}=\sup _{\|h\|_{\infty} \leq 1 / 2}|\mu(\phi)-\nu(\phi)|=\sup _{\operatorname{osc}(h) \leq 1}|\mu(h)-\nu(h)|.$$
-\item The Dobrushin contraction coefficient $\beta_{\text{TV}}$ of a Markov kernel $M$ is given by
+\item \textbf{Dobrushin contraction:} The Dobrushin contraction coefficient $\beta_{\text{TV}}$ of a Markov kernel $M$ is given by
 $$\beta_{\mathrm{TV}}(M)=\sup _{x, y \in \gX}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}.$$
+\item \textbf{Potential functions and the measurement model:} A potential function $G : \gX \to [0,\infty)$ is a non-negative function of an element of the state space $x \in \gX$.
+In our case, this corresponds to the measurement model, and in our previous notation is written as $g_{n,j}^\theta = f_{Y_n|X_n}(y_n^*|x_{n,j}^F; \theta) = G_n(x_{n,j}^F)$.
+Note that $G_n(\cdot) = f_{Y_n|X_n}(y_n^*|\;\cdot\; ; \theta)$ is the conditional density of the observed measurement at time $n$, where we condition on the filtering particle $x_{n,j}^F$ as an element of the state space.
+\item \textbf{Feynman-Kac models and the particle filter:} A Feynman-Kac model on $\gX$ is a tuple $(\pi_0, (M_n)_{n=1}^N, (G_n)_{n=1}^N)$ 
 \end{itemize}

### 2024-06-11 paper/ms.tex
ready to merge Kevin and Ed edits

--- 2024-06-11 paper/ms.tex
@@
 
+We emphasize that we chose to not include the use of many common heuristics used in the machine learning and optimization literature in the gradient descent stage.
+We used constant learning rates of $0.01, 0.05$, and $0.2$ for IFAD-$1,0,$ and $0.97$ respectively, and a constant cooling rate of $0.95$ for our IF2 implementation.
+While techniques such as annealing learning rates, gradient normalization, and momentum made the outperformance of IFAD even more apparent in other simulations we performed, we chose to report the results of the simplest implementation in part to serve as a baseline for the method's performance, and in part because the results of the simplest implementation were already very promising.
+
 \begin{figure}[htbp!]
@@
 
-The fact that the likelihood estimate yielded by MOP-$\alpha$ has more than two continuous derivatives in $\theta$ follows the construction of the likelihood estimate in equations \ref{eq:mop-conditional-likelihood} and \ref{eq:weighting-scheme}, as well as Assumptions \ref{assump:bounded-measurement} and \ref{assump:diff-meas-and-sim}.
+The fact that the likelihood estimate yielded by MOP-$\alpha$ has more than two continuous derivatives in $\theta$ follows from the construction of the likelihood estimate in equations \ref{eq:mop-conditional-likelihood} and \ref{eq:weighting-scheme}, as well as Assumptions \ref{assump:bounded-measurement} and \ref{assump:diff-meas-and-sim}.
 

### 2024-06-13 mop/draft.tex
update from overleaf

@@
 int/.style={draw, minimum size=1.6cm}]
-  \node [int] (H) at (0,0) {$H$};
+  \node [int] (H) at (0,0) {$P$};
 \node [int] (S) [right=of H] {$S$};
@@
 \path[->]
-    (H) edge node {$dH$} (S)
+    (H) edge node {$dP$} (S)
     (S) edge node {$\lambda(t)$}  (I)

### 2024-06-13 paper/appendix/bayes.tex
update from overleaf

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth/\real{1.25}]{imgs/pmcmc/mh.png}
    \caption{Convergence diagnostics for the Metropolis-Hastings variant particle MCMC with the random walk proposal and an informative empirical prior from IF2. Here, we display the results for the trend parameter in the Dhaka cholera model of \cite{king08}.}
    \label{fig:mh}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth/\real{1.25}]{imgs/pmcmc/nuts.png}
    \caption{Convergence diagnostics for a No U-Turn Sampler with uniform priors on a compact set. Again, we display the results for the trend parameter in the Dhaka cholera model of \cite{king08}. The NUTS sampler explores more of the posterior than the MH sampler, but fails to converge quickly.}
    \label{fig:nuts}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[scale=0.3]{imgs/pmcmc/nuts_eb/Recovery Rate.png}
    \includegraphics[scale=0.3]{imgs/pmcmc/nuts_eb/Death Rate.png}
    \includegraphics[scale=0.3]{imgs/pmcmc/nuts_eb/Reciprocal of Mean Immunity Duration.png}
    \includegraphics[scale=0.3]{imgs/pmcmc/nuts_eb/Trend in Force of Infection.png}
    \includegraphics[scale=0.3]{imgs/pmcmc/nuts_eb/Noise in Force of Infection.png}
    \includegraphics[scale=0.3]{imgs/pmcmc/nuts_eb/Measurement Noise.png}
    \caption{Posterior estimates from NUTS powered by MOP-$\alpha$ with the informative nonparametric empirical prior obtained from a KDE on the IF2 parameter cloud. The posterior estimates from each chain are largely in agreement.}
    \label{fig:posteriors}
\end{figure}

### 2024-06-13 paper/appendix/biasvar.tex
update from overleaf

--- 2024-05-10 mop/draft.tex
@@
-\documentclass{article}
-\usepackage[margin=2cm]{geometry}
-\usepackage{natbib}
-\usepackage{hyperref}
-\usepackage{enumitem}
-\usepackage{tikz}
-\usetikzlibrary{arrows.meta}
-\usetikzlibrary{positioning}
+In this section, we prove various rates on the bias, variance, and MSE of MOP-$\alpha$.
 
-\setlist{nosep} % to reduce space in lists
+First, we note the following relation between the Dobrushin contraction coefficient and the alpha-mixing coefficients in our context below.
 
-\setlength{\parskip}{0.5em}
-\setlength{\parindent}{0in}
-
-\input{macros}
-
-\newcommand{\off}{\operatorname{off}}
-\newcommand{\on}{\operatorname{on}}
-
-\title{Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation}
-\author{Kevin Tan, Giles J. Hooker, Edward L. Ionides}
-\date{}
-
-\comments=1
-
-\begin{document}
-\maketitle
-\begin{abstract}
-
-Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods.
-Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques due to (1) the requirement for simulation-based inference that does not require access to the system's transition probabilities and (2) the issue that widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters.
-We propose a new representation that embeds existing methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff.
-Further, we develop optimization algorithms suited to the Monte Carlo properties of the derivative estimate, including a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
-Promising numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from iterated filtering can beat current state-of-the-art methods on a challenging scientific benchmark problem.
-\end{abstract}
-
-\section{Introduction}
-
-Maximum-likelihood estimation for inference in partially observed stochastic processes, also known as, continuous-state continuous-time hidden Markov models (HMMs), partially-observed Markov processes (POMPs), or partially-observed stochastic nonlinear dynamical systems, is a challenging problem that faces several challenges:
-\begin{enumerate}
-    \item Intractable likelihood functions, bypassed with the approaches of simulated likelihood and likelihood-free inference. 
-    \item The desire for simulation-based inference without access to the transition density of the latent Markov process, with various methods created by \cite{welch2009abc, wood2010sl, doucet2010pmcmc, ionides08, ionides15}.
-    \item Accurate parameter estimation in the presence of significant Monte-Carlo noise in the likelihood estimate.
-\end{enumerate}
-
-Many approaches to inference in these scenarios (such as the EM algorithm, the various Kalman filter variants, and MCMC) either struggle with intractable likelihood functions when the models are complex, or assume access to the probability density of next states given the current state.
-This is a problem in some critical applications, such as disease modeling, where the models are complex enough such that even obtaining the transition density is an intractable problem.
-However, the particle filter with the bootstrap proposal, a popular method for solving the filtering problem in partially-observed dynamical systems, provides an unbiased estimate of the likelihood (\cite{delmoral2004feynman}) without requiring evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
-
-This is the backbone behind the only simulation-based full-information maximum likelihood method available in the literature thus far, the improved iterated filtering algorithm of \citet{ionides15}.
-\citet{ionides15} utilize the bootstrap filter for perform maximum likelihood estimation with an iterated perturbed Bayes map, successfully tackling challenging problems in epidemiology that available Bayesian methodology fails to solve.
-
-Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small.
-For example, though in practice the algorithm of \citet{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood.
-\citet{Ionides_mcap} attempt to address the third problem, albeit with a computationally-expensive approach.
-
-\subsection{Automatic Differentiation for Particle Filters}
-
-Recent advances in automatic differentiation (AD) for particle filters (\cite{blei2018vsmc, jon2018diffpf, corenflos21, scibior2021dpf, doucet2022particlebased}) have drawn attention to AD as an alternative tool for inference in partially observed stochastic processes with the particle filter.
-However, these existing approaches either are not yet compatible with the desire for simulation-based inference, or are computationally expensive.
-
-\citet{blei2018vsmc} and \citet{jon2018diffpf} drop high-variance resampling terms from the gradient, leading to bias that does not vanish asymptotically according to \citet{corenflos21}.
-Though the approach of \citet{corenflos21} is promising, it is computationally expensive.
-\citet{doucet2022particlebased} provide another promising fixed-lag smoothing-based approach that is similar in spirit to our method (MOP-$\alpha$), but requires either (1) access to the process model transition densities or (2) the special case where the transition density factors into a policy that selects actions (in the reinforcement learning sense) and a deterministic model that maps states and actions to next-timestep states.
-
-Finally, \citet{scibior2021dpf} demonstrate that the estimators derived in \citet{poyiadjis11} can be attained with AD, albeit with a simple tweak to the particle filter that does not modify its likelihood estimates.
-Though their method is not fully compatible with simulation-based inference, we show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.
-
-\subsection{Our Contributions}
-
-Previous research on AD for particle filters has struggled with various issues: 
-\begin{enumerate}
-    \item Bias, if the discontinuous nature of particle resampling is ignored.
-    \item High Monte Carlo variance, if continuity corrections lead to numerical instability.
-    \item High computational cost, for algorithms which involve pairwise interactions between particles, marginalization, smoothing or optimal transport.
-    \item Reduced applicability for algorithms which lose or fail to take advantage of the simulation-based capabilities of the bootstrap filter.
-\end{enumerate}
-
-We develop a new approach to statistical inference via AD for particle filters which addresses these concerns.
-We develop a new theoretical framework that encompasses existing methods, and addresses the seemingly incompatible paradox of differentiating through a Monte-Carlo algorithm with discontinuous resampling.
-
-This is done through the construction of a smooth extension to the particle filter, as well as a \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) family of algorithms that (informally) encompasses the special case of AD of a vanilla PF (recovering the gradient estimator of \citet{blei2018vsmc}) when $\alpha=0$, and the DPF algorithm from \citet{scibior2021dpf} (that recovers the gradient estimator from \citet{poyiadjis11}) when $\alpha=1$.
-When $\alpha \in (0,1)$, this provides opportunities for optimizing a bias/variance tradeoff.
-
-We show these off-policy particle filters successfully target the posterior, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties.
-To cap off the theoretical results, we also derive a linear convergence rate for gradient methods on the particle filter in the presence of strong convexity, providing support for numerical optimization with AD for the particle filter.
-
-In practice, we propose a hybrid algorithm that we call \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts gradient ascent (or a similar iterative first or second-order gradient method) with a coarse solution obtained from a few rounds of iterated filtering.
-While this requires a differentiable simulator compatible with the reparametrization trick (as previous literature such as \citet{corenflos21} also require), future work will deal with workarounds inspired by our framework that use likelihood ratios of transition densities in place of this.
-Promising numerical results indicate that IFAD beats IF2 (and by the transitive property, IF1, the Liu-West filter, and particle MCMC) on a challenging scientific benchmark problem.
-
-While the Monte-Carlo Adjusted Profile (MCAP) from \citet{Ionides_mcap} provides a profile likelihood confidence interval that takes into account both Monte Carlo profile error and statistical uncertainty in the likelihood, it is computationally inefficient.
-\kevin{TODO} We provide a method for incorporating the MCAP methodology (with the poor man's likelihood profile) in our IFAD algorithm that provides a more computationally efficient profile likelihood confidence interval.
-\kevin{Quantile regression for estimating the max/the envelope of the likelihood evaluations from AD, then fit MCAP/MQLE/MSLE?}
-
-Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual} that we call \texttt{pompy}.
-The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{jax} Python package from \citet{jax2018github}.
-
-\kevin{Terminology: Plug-and-play (clearly definable, but not well-known, define at start of paper and don't use in title/abstract), likelihood-free maximum likelihood (seemingly an oxymoron, rationale being there is no need to evaluate density, bayesians call likelihood densities, plus we only approximate the likelihood and don't really have it), simulation-based inference (loaded term)}
-
-\subsection{Significance Statement}
-Many scientific models involve highly nonlinear stochastic dynamical systems which can be observed only via noisy and incomplete measurements.
-Under the Markov assumption on system dynamics, previous work has provided methods of performing inference for these models.
-In particular, prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics.
-We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
-Our new method outperforms previous approaches on a challenging problem in epidemiology.
-This is implemented in a precursor to a Python counterpart to the popular \texttt{pomp} package by \citet{king16, king2017pompmanual} that includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation.
-
-\kevin{TODO: (1) Fix beta smoothness to d(n) at no more than O(n), (2) add story to say that no one spotted that it could have been plug-and-play because of e.g. reasons why they would stumble}
-
-\section{Problem Setup}
-
-Consider an unobserved Markov process $\{X_t, t \geq t_0\}$, and observations $Y_1,...,Y_N$ at timesteps $t_1,..., t_N$.
-The process is parameterized by an unknown parameter $\theta \in \Theta$, where $X \in \gX, Y \in \gY,$ and finally $\gX \subseteq \R^d, \Theta \subseteq \R^p$.
-By a similar decomposition to that in \citet{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
-$$f_{X_{0: N}, Y_{1: N}}\left(x_{0: N}, y_{1: N} ; \theta\right)=f_{X_0}\left(x_0 ; \theta\right) \prod_{n=1}^N f_{X_n \mid X_{n-1}}\left(x_n \mid x_{n-1} ; \theta\right) f_{Y_n \mid X_n}\left(y_n \mid x_n ; \theta\right).$$
-
-We call $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ the process model, writing $\texttt{process}\left(x_n, \theta\right)$ for the simulator corresponding to the above process model, $f_{Y_n|X_n}\left(y_n \mid x_n, \theta\right)$ the measurement model, and will write $y_n^*$ for the actual values of the observations that were observed.
-
-The above are all defined on a single probability space $(\Omega, \Sigma, \prob),$ where $\omega \in \Omega$ denotes an element of the sample space that can be thought of as a random seed.
-For notational simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0.$
-
-\paragraph{Seed-Fixing, Coupling, and the Method of Common Random Numbers, Coupling:}
-
-In the rest of this paper, we will identify the seed with an element of the sample space.
-We do so because the seed determines the entire sequence of pseudo-random numbers as generated by a computer, to maintain an algorithmic and practical interpretation of our theory.
-When two algorithms share the same seed, this corresponds to the method of common random numbers.
-
-\section{A Smooth Extension of the Particle Filter}
-
-%\paragraph{Motivation:} 
-%The gradient of the log-likelihood can be written as a vector $\nabla_\theta \ell(\theta)$ such that 
-%\begin{equation}
-    %\lim_{v \to 0} \frac{|\ell(\theta+v) - \ell(\theta) - v^T \nabla_\theta \ell(\theta)|}{\|v\|} = 0.
-%\end{equation}
-
-%We therefore need to evaluate the likelihood at a point $\theta$ and a nearby point $\theta + v$ to estimate the gradient with finite differences.
-But in the presence of Monte Carlo noise in evaluating the likelihood, this is difficult, unless (1) one ensures one uses the same seed for each evaluation in the neighborhood, and (2) chooses a step-size small enough that changing the the parameter does not change the resampling indices.
-
-\paragraph{Intuition:} 
-Our theoretical framework for automatic differentiation for the particle filter hinges on the following observation.
-Instead of seeking to differentiate $\ell(\theta)$ directly, we instead focus on differentiating through a suitable fixed-seed reweighting of a bootstrap filter.
-The intuition is as follows.
-Consider a particle filter likelihood estimate where the system (i.e.
-\textbf{both the state transitions and resampling}) evolves under a parameter $\phi \in \Theta$ and the seed, or sample space element, is fixed at $\omega \in \Omega.$ Conditional on $\omega$ and $\phi,$ one can reweight the likelihood estimate to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$
-
-Differentiating this reweighted likelihood estimate with respect to $\theta$ then amounts to differentiating through the reweighting likelihood ratios, which is differentiable if the densities used for the reweighting are differentiable.
-We call this simulation-and-reweighting scheme an \textbf{off-policy} likelihood estimate, as it is similar in spirit to the off-policy methods found in the reinforcement learning literature.
-
-The above intuition is formalized below.
-
-\kevin{Method of Common Random Numbers, Coupling}
-
-\subsection{Doubly Off-Policy Particle Filters}
-
-\begin{defn}[Seed-Fixing]
-\label{defn:seed-fixing}
-Consider a probability space $(\Omega, \Sigma, \mu)$ governing the particles and observations.
-Noting that the particle filter is unbiased for the likelihood when computing the expectation of the normalizing factor under the posterior $f_{X_{0:N|Y_{1:N}}}\left(x_{0: N} \mid y_{1: N}; \theta\right)$, we can write the likelihood as
-$$
-\begin{aligned}
-\mathcal{L}(\theta) & =\int_{\Omega} \mathcal{L}(\theta, \omega, J) d \mu(\omega) \\
-& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right)\right) f_{X_{0:N}, Y_{1:N}}\left(x_{0: N}^P(\omega) , y_{1: N}^*, \theta\right) d \mu(\omega) \\
-& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right) f_{X_n|X_{n-1}}(x_{n,j}^P(\omega)|x_{n-1,j}^F(\omega), \theta)\right) f_{X_{0:N}| Y_{1:N}}\left(x_{n, j}^P(\omega) \mid y_n^*, \theta\right) d \mu(\omega)
-\end{aligned}
-$$
-where $\mathcal{L}(\theta, \omega, J)$ is the Monte-Carlo estimate of the likelihood given the realization of $\mathrm{J}$ particles corresponding to $\omega \in \Omega$.
-For simplicity, we resample every timestep.
-We call $\omega$ a seed, and the deterministic quantity $\mathcal{L}(\theta, \omega, J)$ the seed-fixed estimate.
-\end{defn}
-
-Armed with this seed-fixed likelihood estimate, we modify a particle filter likelihood estimate where the system evolves under a parameter $\phi \in \Theta$ and seed $\omega \in \Omega$ to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$
-
-\begin{defn}[Doubly Off-Policy Particle Filter Likelihood]
-\label{defn:doubly-off-policy}
-Let $\lik(\theta, \phi, \omega, J)$ denote a Monte-Carlo estimate of the likelihood evaluated at $\theta \in \Theta$ with $J$ particles, but where the system evolves according to another $\phi \in \Theta$ and conditional on a fixed seed $\omega \in \Omega$.
-This can be computed with a suitable reweighting of the particles, which we provide the derivation for later.
-We then have
-\begin{equation} \label{eq:loglik2}
-\lik(\theta)=\E[\lik(\theta,\phi,\Omega, J)]=\int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega),
-\end{equation}
-where we define
-\begin{align}
-    \lik(\theta,\phi,\omega, J) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J &\left(f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) \cdot 
-    f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}(\omega) | x_{n-1,j}^{F,\phi}, \phi) \right.\\
-    &\cdot
-    \left.
-    \frac{f_{Y_n|X_n}(y^*_n|x_{n,j}^{P,\phi}(\omega),\theta)}{f_{Y_n|X_n}(y^*_n|x_{n,j}^{P,\phi}(\omega),\phi)}
-    \cdot \frac{f_{X_n|X_{n-1}}(x_{n,j}^{P,\phi}(\omega)|x_{n-1,j}^{F,\phi}(\omega),\theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P,\phi}(\omega)|x_{n-1,j}^{F,\phi}(\omega),\phi)}\right).
-\end{align}
-\end{defn}
-
-Lastly, we require the following regularity conditions.
-
-\begin{aspt}[Continuity of the Likelihood] $\ell(\theta)$ proper is continuous in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.
-\end{aspt}
-
-\begin{aspt}[Bounded Process Model]
-\label{assump:bounded-process}
-    There exist $\underbar{M}, \bar{M}$ such that $0 < \underbar{M} \leq f_{X_n|X_{n-1}}(x_n | x_{n-1}) \leq \bar{M} < \infty$. 
-\end{aspt}
-
-\begin{aspt}[Bounded Measurement Model]
-\label{assump:bounded-measurement}
-There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G} < \infty$ and $\infty > G' \geq G'(\theta)>0$ such that $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty <\leq G'(\theta) \leq G'$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
-\end{aspt}
-
-\begin{aspt}[Locally Bounded Derivatives]
-\label{assump:local-bounded-derivative}
-Let $\mathcal{M}$ be an open subset of $\Omega$.
-There exists some function $G(\theta)$ and constants $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$, $H=\sup _{\theta \in \mathcal{N}} H(\theta)<\infty$ such that
-    $$
-    \|\nabla_\theta \ell(\theta, \phi, \omega, J)\|_\infty \leq G(\theta) \leq G<\infty, \;\;
-    \|\nabla^2_\theta \ell(\theta, \phi, \omega, J)\|_\infty \leq G(\theta)^2 \leq G^2<\infty
-    $$
-    for every $\phi \in \mathcal{M}, \theta \in \mathcal{N}_J(\phi)$, and almost every $\omega \in \Omega$.
-\end{aspt}
-
-We need the following assumption to ensure that the weighted particle filter with off-policy resampling targets the posterior:
-\begin{aspt}[Bounded And Differentiable Measurement Density Ratios]
-    \label{assump:diff-meas}
-    For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{Y_n|X_n}(y_n^*|x, \theta)}{f_{Y_n|X_n}(y_n^*|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x \in \gX$ and all $\theta\in\gN(\phi)$.
-\end{aspt}
-
-In the following assumption, the first condition is sufficient for DOP, and the second for MOP:
-\begin{aspt}[Bounded And Differentiable Process Model]
-    \label{assump:diff-process}
-    Either one of the following two conditions hold:
-    \begin{enumerate}
-        \item For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{X_n|X_{n-1}}(x'|x, \theta)}{f_{X_n|X_{n-1}}(x'|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x, x' \in \gX$ and all $\theta\in\gN(\phi)$, or
-        \item There exists a differentiable simulator with more than two continuous derivatives for the process model. 
-    \end{enumerate}
-\end{aspt}
-
-We want $\phi\in\Theta$ to be a point in parameter space around which $\lik(\theta,\phi,\omega, J)$ is differentiable with respect to $\theta$ and does not have excessive Monte~Carlo variance.
-
-We require many of these assumptions to establish the following proposition, the proof of which is a straightforward consequence of what we assumed.
-
-\begin{prop}[Smooth Neighborhood]
-\label{prop:smooth-nbhd}
-There exists an open neighborhood $\mathcal{N}(\phi)$ around any $\phi \in \Theta$ where for all $\theta \in \mathcal{N}(\phi)$ and almost every $\omega \in \Omega$, the Monte Carlo estimate of the likelihood at $\theta \in \mathcal{N}(\phi)$ with the system evolving according to $\phi$ conditional on $\omega$, $\lik(\theta,\phi,\omega, J)$, has two continuous derivatives in $\theta$.
-\end{prop}
-
-\begin{proof}
-    For a fixed $\omega$, the resampling indices for $\lik(\theta,\phi,\omega, J)$ and $\lik(\phi,\phi,\omega, J)$ are the same as the resampling indices in both cases correspond to that yielded by $f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}(\omega);\phi)$, eliminating any discontinuities from resampling.
-
-    By Assumptions \ref{assump:diff-meas} and \ref{assump:diff-process}, for small enough $\mathcal{N}(\phi)$ the density ratios are bounded.
-The likelihood only changes by a factor of the density ratios, which are bounded and possess two continuous derivatives in $\theta$ if either the process and measurement densities used in the calculation do, or if the measurement densities do and one has access to a differentiable simulator with two continuous derivatives in $\theta$.
-This gives us the desired result.
-
-    As a remark, we note that the neighborhoods the sequence is differentiable in are independent of $J$, as the resampling does not change with $\theta$ -- it is dependent on $\phi$. 
-\end{proof}
-
-\subsubsection{Doubly Off-Policy Particle Filter}
-
-Having the system evolve according to $\phi$ but evaluating the likelihood at $\theta$ requires us to reweight the conditional likelihood at each timestep by likelihood ratios corresponding to the process and measurement models.
-This corresponds to making two corrections in the particle weights at each timestep, by multiplying by
-\begin{equation}
-    s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}, \qquad r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)},
-\end{equation}
-as we illustrate in Appendix \ref{app:dop}. We use the notation $x_{n,j}^{P, \phi}$ to emphasize that the particles evolve according to $\phi$.
-
-We claim that this yields a particle filter that targets the posterior and recovers the derivative estimate of \citet{poyiadjis11, scibior2021dpf} when $\theta$ is evaluated at $\phi$.
-The proof can be found in Appendix \ref{app:dop}.
-
-\begin{prop}[Correctness of Doubly Off-Policy Particle Filter Corrections]
-    \label{prop:dop-correctness}
-
-    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$.
-Also let $\theta \in \gN(\phi)$.
-Then, the following corrections in the particle weights,
-    \begin{equation}
-    s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}, \qquad r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)},
-    \end{equation}
-    yield a particle filter that is unbiased for the likelihood, even when $\theta \neq \phi$.
-
-    In particular, when $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{poyiadjis11},
-    \begin{equation}
-        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
-    \end{equation}
-\end{prop}
-
-\paragraph{Intuition:} Having the system evolve according to $\phi$ but evaluating the likelihood at $\theta$ requires us to reweight the conditional likelihood at each timestep by likelihood ratios corresponding to the process and measurement models.
-This corresponds to making two corrections in the particle weights at each timestep, one for the process model and one for the measurement model.
-The measurement correction cannot be avoided, but one can bypass the process correction -- with the reparametrization trick.
-We can do when we assume the simulator is differentiable with the reparametrization trick, as \cite{corenflos21} also assume.
-Doing so then yields an algorithm that only requires a simulator, and does not require the process densities.
-
-\subsection{Measurement Off-Policy Particle Filters}
-
-With the above procedure, we have the entire system evolve according to $\phi$, and we differentiate through the likelihood ratios, which are smooth conditional on the resampling being the same (which in turn happens when $\theta \approx \phi$).
-However, doing so requires access to likelihood ratios, and as such Algorithm \ref{alg:dop} is therefore not compatible with simulation-based inference.
-
-There is a workaround.
-Consider the case where you have access to a differentiable simulator compatible with the reparametrization trick.
-If we have the state transitions evolve according to $\theta,$ but have the resampling according to $\phi,$ we can offload the derivative with respect to the process model to the reparametrization trick, as in \cite{corenflos21}.
-This corresponds to the MOP-$\alpha$ procedure in Algorithm \ref{alg:mop}.
-
-We therefore make the following assumption in this section only:
-\begin{aspt}[Differentiable Simulator]
-    $\texttt{process}(\cdot | X_{t-1}; \theta)$ is differentiable with respect to $\theta,$ and is compatible with the reparametrization trick. 
-\end{aspt}
-
-Algorithm \ref{alg:mop} provides a weighted particle filter that targets the posterior under $\theta$, but when differentiated yields either the low-bias but high-variance gradient estimator of \cite{poyiadjis11}, the high-bias but low-variance estimator of \cite{blei2018vsmc}, or an intermediate, interpolating quantity that can optimize a bias-variance tradeoff.
-This is controlled by a discounting parameter $\alpha$.
-The intuition behind this procedure is as follows.
-
-\paragraph{Intuition:} Broadly, if $\theta\neq\phi,$ one can first run a vanilla particle filter to recover values for the particles with the process model under $\phi.$ One can then run an accordingly reweighted particle filter with the process model under $\theta$, but resampling according to the measurement model under $\phi$, with the same indices as the original vanilla particle filter.
-This requires one to set the same random seed for both particle filters.
-This then yields a slightly-unconventionally weighted particle filter that targets the posterior, but has some very nice gradient properties; as we outline in the remainder of this section.
-
-\paragraph{Special Case:} When $\theta = \phi$, one can simply set the particles under $\phi$ to be copies of the particles under $\theta$.
-Gradients with respect to $\theta$ now treat functions of the particles under $\phi$ as constants, corresponding to the stop-gradient trick in \cite{scibior2021dpf}.
-This therefore provides a theoretical framework for their stop-gradient trick, which we happily use in our numerical experiments for simplicity.
-
-\kevin{Develop theory for mixing-time swapping?
-Detect when IF2 reaches a neighborhood of the MLE, by detecting whether it has reached a local sort of stationary distribution when the perturbations are lower bounded, e.g.
-the autocorrelation between the mean vectors to estimate the mixing time or Gelman-Rubin.
-In practice, just swap when IF2 converges according to the diagnostics.
-Theory doesn't care, you just need to in a basin of attraction that is strongly convex.}
-
-\subsubsection{Discounting Weights}
-
-Observe that at Algorithm \ref{alg:mop}, at each timestep the corrected filtering importance weights $w_{n,j}^{F,\theta}$ optionally propagate to the next timestep for us to maintain the correction.
-We introduce a discounting parameter $\alpha$ that affects the weights, so $w_{n,j}^{P,\theta} := (w_{n-1,j}^{F,\theta})^{\alpha}$.
-Note that no matter what value $\alpha$ takes on, the system evolves according to $\phi$ when we perform the correction to evaluate the likelihood at $\theta$.
-The filter fully targets the posterior when $\theta=\phi$, or $\alpha=1$, as we show in Proposition \ref{prop:dop-correctness}.
-As previously mentioned, the value of $\alpha$ therefore informs a bias-variance tradeoff.
-We elaborate on this below.
-
-\paragraph{Bias-Variance Tradeoff} This discounting serves the purpose of balancing a tradeoff between maintaining memory of each particle's ancestral trajectory (most extreme when $\alpha=1$), or considering only the single-step transition dynamics (when $\alpha=0$).
-This can be thought of as an \textbf{exponentially-weighted moving average}, where $\alpha$ controls the amount of discounting.
-We elaborate on two special cases below.
-
-\kevin{This isn't a smoother, because we'd need future information.}
-
-\paragraph{When $\alpha=1,$} the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\theta}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
-\end{equation}
-as we have a telescoping product. Its derivative when $\theta=\phi$ is then
-\begin{equation}
-    \nabla_\theta \hat{\ell}(\theta) := 
-        \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*),
-\end{equation}
-the average gradient of each surviving particle's cumulative importance weights.
-The filter targets the posterior when $\alpha=1$ even when $\theta \neq \phi,$ and therefore suffers from no asymptotic bias.
-
-\paragraph{When $\alpha=0,$} the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J w_{n,j}^{F,\theta} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})},
-\end{equation}
-and though the telescoping product disappears, this reduces to a series of one-step corrections.
-However, these corrections are insufficient when $\theta \neq \phi,$ and the filter suffers from asymptotic bias unless $\theta=\phi.$ Its gradient when $\theta=\phi$ is therefore
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
-    \end{equation}
-
-the gradient of the vanilla particle filter where the resampling correction is omitted as found in \cite{blei2018vsmc} -- when applied to the bootstrap filter.
-As \cite{scibior2021dpf} remark, this estimator only considers single-step quantities, and is "memoryless" beyond a single step.
-This is in contrast to the case when $\alpha=1,$ as that estimate only considers the surviving particles at time $N$, which is the other extreme where dependencies across time are fully tracked.
-
-\subsubsection{Correctness of MOP-$\alpha$}
-
-\begin{prop}[Correctness of MOP-$\alpha$ (Algorithm \ref{alg:mop})]
-    \label{prop:mop-correctness}
-    When either $\alpha=1$ or $\theta=\phi$, Algorithm \ref{alg:mop} is a particle filter that targets the posterior and is unbiased for the likelihood under $\theta$.
-
-    In particular, when $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter.
-In this case, its gradient when $\alpha=1$ is the same as the DPF estimator of \citet{scibior21} when applied to the bootstrap filter,
-    \begin{equation}
-        \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{0:N}}(y_{1:N}^* | x_{1:N,j}^{A, F,\theta}),
-    \end{equation}
-
-    and when $\alpha=0,$ is the gradient estimator of \cite{blei2018vsmc},
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
-    \end{equation}
-
-    that as \cite{scibior21} show corresponds to differentiating through the fixed-seed likelihood estimate yielded by a vanilla particle filter, dropping resampling terms.
-
-    When $\alpha=1$, the gradient estimate MOP-1 provides is consistent for the score.
-\end{prop}
-
-The proof for the above follows from Appendices A and C, as well as the technical lemmas for properly weighted particle filters.
-
-\kevin{Take derivative of loglik with regard to alpha?}
-\kevin{Talk about regime change the moment you move away from alpha=1, variance going from N2 to N, and bias going from 0 to something, but try to prove that decrease in variance outweights increase in bias.
-}
-\kevin{Create tools for choosing alpha via binary search CV from an averaged alpha=1 target}
-\kevin{suppose at each filtering step, look at gradient calculated at alpha=0, dynamically select alpha during filtering.
-calculate bias and variance online during filtering, and update alpha accordingly.
-}
-
-\kevin{No need to recompute so many particle filters! Just take an array of weight copies, one for each alpha. Gradients for each alpha come free the moment you compute one of them. }
-
-\kevin{Elbow method on variance, since bias is hard to compute? }
-\kevin{Get variance estimate, bias estimate, construct confidence intervals on bias estimate, use that to get unbiased estimate of the MSE. }
-
-\subsubsection{Practical Considerations}
-
-It is only ever practical to evaluate $\theta$ at $\phi$.
-This removes the issue of the likelihood ratios blowing up, and yields an estimate of the likelihood that agrees with the bootstrap filter estimate.
-Its gradient in this case is either the estimate of \citet{poyiadjis11} when $\alpha=1$, a single-timestep estimator that corresponds to naively taking the derivative of the particle filter when $\alpha=0$ as first proposed in \citet{blei2018vsmc}
-
-\paragraph{Bottom Line:} The DOP-$\alpha$ and MOP-$\alpha$ procedures therefore provide us with a unifying framework for reasoning about smooth extensions of weighted particle filters that target the posterior and provide reasonable gradient estimates when differentiated.
-This fixed-seed reweighting alleviates issues that can arise from differentiating through a quantity yielded by ostensibly non-differentiable resampling.
-This is done by either (1) differentiating through smooth likelihood ratios, or (2) differentiating through a differentiable simulator with the reparametrization trick (and through smooth likelihood ratios to adjust for the resampling).
-In particular, MOP-$\alpha$ does not require access to the transition densities, and like the bootstrap filter, only requires access to a (albeit differentiable) simulator.
-To our knowledge, this is the first particle filter gradient method in the literature with this capability.
-
-\section{Practical Maximum-Likelihood Inference}
-
-This still leaves us with the question of designing a reasonably effective procedure for likelihood maximization with these new tools.
-We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that amounts to warm-starting gradient descent or some other reasonable first or second order procedure (using the gradient estimate given by MOP-$\alpha$) with the output of an initial search of IF2.
-
-\paragraph{Summary:} A coarse solution is found by running IF2 with an aggressive random walk standard deviation (learning rate) and a cooling schedule that is \textbf{bounded below}, until the search stalls to an initial "convergence" to what is hopefully a neighborhood of the MLE.
-We can then use gradient descent, or other first or second order gradient methods, to refine this coarse solution.
-
-The initial "convergence" happens fairly quickly in practice.
-In the case of the Dhaka cholera model of \cite{king08}, when a geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations of IF2.
-In comparison, often 100 or 200 iterations of IF2 are used for an initial global search.
-\cite{ionides15}, for example, use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{Lee_haiticholera}.
-
-\paragraph{Rationale:}
-
-IF2 converges quickly to a neighborhood of some reasonably good local optimum in practice, but then struggles to squeeze out the last few log-likelihood units.
-Conversely, gradient descent (and other first or second order gradient methods) can get stuck in saddle points and less desirable local minima when the likelihood is nonconvex.
-Warm-starting gradient methods with IF2 lets us get close enough to the MLE for gradient methods to shine, as in most cases regularity conditions ensure that the likelihood surface is well-behaved enough in this neighborhood that the aforementioned issues with saddle points and local minima are alleviated.
-Combining these two methods in this way lets us enjoy the best of both worlds.
-
-\begin{algorithm}[ht]
-\centering
-	\caption{Iterated Filtering with Automatic Differentiation}
-    \label{alg:ifad}
-	\begin{algorithmic}[1]
-	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $p(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta),$ IF2 cooling schedule $\eta_t$, MOP-$\alpha$ discounting parameter $\alpha$.
-		\STATE Initialize $\theta_0$, $t=0.$
-        \STATE Run IF2 until initial "convergence" under cooling schedule $\eta_t$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_t := \frac{1}{J}\sum_{j=1}^J \Theta_j.$
-		\WHILE{Procedure not converged:}
-		\STATE Draw $\omega \in \Omega$
-		\STATE Perform forward pass by running Algorithm \ref{alg:mop} with discounting parameter $\alpha$ to obtain $\loglik(\theta_t, \theta_t, \omega, J).$
-		\STATE Obtain $g(\theta_t) = \nabla_{\theta_t} (-\loglik(\theta_t, \theta_t, \omega, J))$ via backpropagation, $H(\theta_t)$ such that $\lambda_{\min}(H) \geq c$.
-		\STATE Obtain $\eta$ with line search or an annealing schedule.
-		\STATE Update $\theta_{t+1} := \theta_t - \eta (H(\theta_t))^{-1} g(\theta_t)$, $t:=t+1.$
-		\ENDWHILE
-		\STATE Return $\hat{\theta} := \theta_t.$
-	\end{algorithmic}
-\end{algorithm}
-
-\subsection{Theoretical Guarantee}
-
-We first analyze the performance of gradient descent, Newton's method, or any iterative method that multiplies the particle estimate of the gradient with the inverse of a matrix with minimum eigenvalue bounded below, in the case where the log-likelihood surface is $\gamma$-strongly convex, and where the gradient estimate is a particle approximation of the score (for example, IFAD-1 and MOP-1).
-In this case, these methods enjoy a linear convergence guarantee:
-
-\kevin{I think we can establish the linear convergence for the entire procedure in the same sense as \href{this}{https://arxiv.org/pdf/2311.08442.pdf}, in the sense that if IF2 is run for a sufficiently large but fixed number of iterations, once we arrive in a local strongly convex region, we get linear convergence when we swap to MOP.
-Show by induction that we remain in the strongly convex region (in terms of parameter space and likelihood), eqn 218.
-}
-
-\begin{thm}[Convergence of Gradient Methods]
-\label{thm:convergence}
-
-Consider a variant of IFAD-1 (Algorithm \ref{alg:ifad}) where one stops if the gradient estimate $||g(\theta_n)|| \leq \sigma \epsilon$.
-Assume $-\ell$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$.
-$\sigma = \frac{4 \Gamma}{(1-\beta)}> 4$.
-Define $G$ as in Assumption \ref{assump:local-bounded-derivative}.
-
-Choose $J$ according to the maximum of the $J$ prescribed in Lemmas \ref{lemma:grad_bound} and (if using a Hessian estimate) $\ref{lemma:hess_bound}$ with failure probability $\delta/2$ each.
-Let $H(\theta_n)$ be a matrix possibly dependent on $\theta_n$ with minimum eigenvalue always greater than some $c>0$.
-Choose the learning rate $\eta$ and error tolerance $\epsilon$ such that, where $\beta$ is the Armijo condition hyperparameter,
-\begin{equation}
-    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_n)||.
-\end{equation}
-
-Then the following holds with probability at least $1-\delta$:
-\begin{equation}
-\ell(\theta^*) - \ell(\theta_{n+1}) \leq \left(1-\eta\beta\frac{8\gamma}{9c}\right)(\ell(\theta^*)-\ell(\theta_n))
-\end{equation}
-
-and the algorithm eventually terminates after we obtain $||\nabla_\theta \ell(\theta_n)|| \leq (1+\sigma) \epsilon$.
-\end{thm}
-
-The proof of this result uses the concentration inequalities from \cite{delmoral2011ci}, and leans upon the analysis of \cite{mahoney16}. We defer the proof to Appendix \ref{app:ifad}
-
-\paragraph{Takeaways for IFAD (Algorithm \ref{alg:ifad}):}
-
-The gradient stage of IFAD-1 converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the IF2 stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
-This happens fairly often in practice, for example, when sufficient regularity conditions for local asymptotic normality of the MLE hold.
-
-We conjecture that this applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE.
-IF2 posesses some SGD-like qualities, including what looks like linear convergence to a ball around the MLE with width determined by the random walk standard deviation.
-We will explore this conjecture in future work.
-
-While this theoretical guarantee currently only holds for IFAD-1, as it forms a particle approximation of the score as in \cite{poyiadjis11}, one can also extend this for a similar guarantee if the bias for the gradient estimate given by MOP-$\alpha$ for $\alpha < 1$ is small enough.
-For general $\alpha$, including MOP-0, the analysis then becomes more complicated, as one needs to handle the case of biased gradient descent.
-However, if future measurements $y_{n+1:N}^*$ do not provide a large amount of information on the identification of the current state $x_n$ given past and current measurements $y_{0:n}^*$, as mentioned in \cite{corenflos21}, the bias may be small enough for it to not present too much of a problem.
-
-\subsection{Runtime}
-
-Even with the increased computational cost of automatic differentiation, Algorithm \ref{alg:ifad} is reasonably fast, as by the cheap gradient principle from \cite{kakade2019provably}, a gradient operation takes no more than 6 times the runtime of the particle filter (in practice, this means a gradient step takes about 3.75x as much time as an IF2 iteration).
-As such, unlike methods like the marginal differentiable particle filter in \cite{scibior2021dpf} or the the optimal transport of resampling approach of \cite{corenflos21}, our algorithm has runtime \textbf{linear, and not quadratic}, in the number of particles (while alleviating the $O(N^2)$ variance concern of \cite{poyiadjis11} and \cite{corenflos2021ot} and asymptotic bias concern of \cite{blei2018vsmc} through optimizing $\alpha$ for a bias-variance tradeoff).
-
-In addition, re-implementation of the particle filter, iterated filtering, and Algorithm \ref{alg:ifad} itself in \texttt{JAX} (\cite{jax2018github}) led to considerable speedups.
-This can be attributed to both GPU acceleration and the just-in-time (JIT) compilation functionality that \texttt{JAX} provides.
-When tested on a computer with an Intel i9-13900K processor, 128GB of RAM, and a NVIDIA RTX3090 GPU, each post-compilation particle filtering operation on the Dhaka cholera model from \cite{king08} with 10000 particles took roughly 379 milliseconds (with a standard deviation of 279 microseconds over seven runs), a considerable improvement over the existing performance in the widely used \texttt{pomp} package of \cite{king16}, which took between 6.19 to 6.39 seconds over ten runs on the same machine.
-
-\section{Numerical Experiments}
-
-\subsection{Cholera in Bangladesh}
-
-To compare the performance of IFAD against that of IF2, we fit the very same model first proposed in \cite{king08} that \cite{ionides15} use to demonstrate the capabilities of IF2.
-This is becoming a standard benchmark, as \cite{wood16} and \cite{wycoff2024voronoi} include in their papers.
-This is a stochastic SIR compartmental model with transition and measurement uncertainty, where the population at time $t$, $H(t)$, is divided into the susceptible compartment $S(t)$, infected compartment $I(t)$, and three ($k=3$) recovered compartments $R_1(t), ..., R_k(t)$ denoting varying degrees of cholera immunity.
-We write $M(t)$ for the cholera deaths in each month.
-As in \cite{ionides15}:
-
-\begin{itemize}
-    \item The \textbf{transition dynamics} follow this series of stochastic differential equations:
-
-    \begin{align}
-        d S&=\left(k \epsilon R_k+\delta(S-H)-\lambda(t) S\right) d t+d H-(\sigma S I / H) d B, \\ 
-        d I&=\left(\lambda(t) S-(m+\delta+\gamma) I\right) d t+(\sigma S I / P) d B, \\ 
-        d R_1&=\left(\gamma I-(k \epsilon+\delta) R_1\right) d t, \\ 
-        \vdots \\ 
-        d R_k&=\left(k \epsilon R_{k-1}-(k \epsilon+\delta) R_k\right) d t,
-    \end{align}
-
-    where $B$ represents Brownian motion, $m$ is the rate at which infected individuals succumb to cholera, $\gamma$ is the rate at which infected individuals recover, $1/\epsilon$ is the mean duration of immunity, $\sigma$ represents the standard deviation of perturbations to the force of infection, and $\delta=0.02$ is the death rate.
-
-    \item The \textbf{force of infection}, $\lambda(t)$, is modeled by splines $s_j$
-    \begin{equation}
-        \lambda(t)=\exp \left\{\beta_{\text {trend }}\left(t-t_0\right)+\sum_{j=1}^{6} \beta_j s_j(t)\right\}(I / P) + \exp \left\{\sum_{j=1}^{6} \omega_j s_j(t)\right\},
-    \end{equation}
-    where the $\beta_j$ model seasonality in the force of infection, $\beta_{\text{trend}}$ models the trend in the force of infection, and the $\omega_j$ represent seasonality of a non-human environmental reservoir of disease.
-    \item The \textbf{measurement model} for observed monthly cholera deaths is given by 
-    \begin{equation}
-        Y_n \sim \gN(M_n, \tau^2M_n^2),
-    \end{equation}
-    where $M_n$ is the true number of cholera deaths in that month.
-\end{itemize}
-
-\begin{figure}[ht]
-\centering
-\begin{tikzpicture}[node distance=1cm and 1cm, auto,
->=Latex,every node/.append style={align=center}, 
-int/.style={draw, minimum size=1.6cm}]
-  \node [int] (H) at (0,0) {$H$};
-\node [int] (S) [right=of H] {$S$};
-\node [int] (I) [right=of S] {$I$};
-\node (Mcoord) [right=of I, coordinate] {Mcoord};
-\node [int] (R1) [right=of Mcoord] {$R_1$};
-\node [int] (R2) [right=of R1] {$R_2$};
-\node [int] (R3) [right=of R2] {$R_3$};
-\node [circle, draw] (M) [above=of Mcoord] {$M$};
-\node (deathsS) [above=of S, coordinate] {deathsS};
-\node (deathsI) [above=of I, coordinate] {deathsS};
-\node (deathsR1) [above=of R1, coordinate] {deathsR1};
-\node (deathsR2) [above=of R2, coordinate] {deathsR2};
-\node (deathsR3) [above=of R3, coordinate] {deathsR3};
-\node (belowR3) [below=of R3, coordinate] {belowR3};
-\node (belowS) [below=of S, coordinate] {belowS};
-\path[->]
-    (H) edge node {$dH$} (S)
-    (S) edge node {$\lambda(t)$}  (I)
-    (I) edge node {$\gamma$}  (R1)
-    (R1) edge node {$k\epsilon$}  (R2)
-    (R2) edge node {$k\epsilon$}  (R3)
-    (R3) edge node {} (belowR3) 
-    (belowR3) edge node {$k\epsilon$} (belowS) 
-    (belowS) edge node {} (S)
-    (I) edge node {$m$}  (M)
-    (S) edge node {$\delta$}  (deathsS)
-    (I) edge node {$\delta$}  (deathsI)
-    (R1) edge node {$\delta$}  (deathsR1)
-    (R2) edge node {$\delta$}  (deathsR2)
-    (R3) edge node {$\delta$}  (deathsR3);
-\end{tikzpicture}
-\caption{Illustration of SIR model from \cite{king08}.}
-\label{fig:sir}
-\end{figure}
-
-Figure \ref{fig:sir} depicts the above compartmental model that we use to benchmark the performance of IFAD.
-We present comparisons of IFAD with $\alpha=0$ (IFAD-0) and $\alpha=1$ (IFAD-1) against both (1) the results from \cite{ionides15}, and (2) our own implementation of IF2 in \texttt{JAX}.
-
-\paragraph{Hyperparameters:} While performing an initial exploratory analysis, we found that alternative hyperparameters for IF2 (random walk standard deviation of $\sigma=0.02$, geometrically decreasing cooling rate exponent $a=0.95$) performed better than the settings \cite{ionides15} chose, and chose to use these for our implementation to achieve a fairer comparison.
-Our implementations of IF2 and IFAD used 10000 particles, in line with \cite{ionides15}.
-
-To prevent exploding gradients given the high variance in the MOP-1 estimator, we scale the gradient to have norm 1 before performing vanilla gradient descent, and use a linear cooling schedule of $0.2$ to $ 0.0001$.
-We chose to use vanilla gradient descent for simplicity, and to demonstrate that minimal tuning was required to have IFAD perform well.
-
-% The 60 more iterations of IF2 yielded largely not a huge improvement over the initial coarse solution of 40 iterations.
-This demonstrates what we have mentioned earlier: that although IF2 manages to quickly find a decent coarse solution, it struggles to squeeze out those last few units of log-likelihood improvement.
-The 60 more iterations of gradient descent, however, successfully refines this solution, getting much closer to the MLE on average.
-
-\subsubsection{Results} 
-We benchmarked IFAD against IF2 on a challenging global search problem.
-100 initial starting parameter vectors were drawn uniformly from the same wide bounding box used in \cite{ionides15}, and we performed 100 searches each of IF2, IFAD-0, IFAD-1, MOP-0, and MOP-1 initialized from these 100 starting parameter vectors.
-We summarize our findings below.
-
-\begin{table}[h!]
-\centering
-\begin{tabular}{||c c c||} 
- \hline
- Method & Best Log-Likelihood & Rank \\ [0.5ex] 
- \hline\hline
-     IFAD-0 & -3747.87 & 1\\
-     IFAD-1 & -3748.53 & 2\\
-     IF2 (Ours) & -3755.91 & 3\\
-     IF2 (Warm Start) & -3759.60 & 4 \\
-     IF2 2015 & -3768.63 & 5\\ 
-     MOP-0 Alone & -3795.74 & 6\\
-     MOP-1 Alone & -3797.38 & 7\\
- \hline
-\end{tabular}
-\caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone.
-IFAD performs the best among all methods.
-Our implementation of IF2 outperforms that of \cite{ionides15}, but still ultimately underperforms IFAD.
-Both IFAD-0 and IFAD-1 manage to find the MLE, either matching or beating the highest log-likelihood previously found in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.}
-\label{table:mle}
-\end{table}
-
-\paragraph{IFAD Successfully Finds the MLE:} Both IFAD-0 and IFAD-1 successfully find the MLE, as seen in Table \ref{table:mle}.
-Previously, the MLE at a log-likelihood of $-3748.5$ was found with much effort, using many global and local IF2 searches and likelihood profiling.
-Despite being initialized for a global search, both IFAD-0 and IFAD-1 manage to find the MLE over the 100 searches.
-This may imply that the local search and refinement that was previously necessary for finding the MLE is \textbf{baked into} the two-stage IFAD algorithm.
-
-Interestingly, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.63$, while \cite{king08} only achieve a best possible estimate of $-3793.4$.
-Our results of $-3748.53$ and $-3747.87$ match or beat the highest log-likelihood previously found in the current implementation in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.
-
-This can be seen most clearly in Figure \ref{fig:boxplot-search}.
-In the left panel within Figure \ref{fig:boxplot-search}, only IFAD has whiskers that extend to the dotted red line that represents the MLE.
-
-\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations (as seen in Figure \ref{fig:optim}), performing IF2 alone ultimately fails to squeeze out the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter} and the left panel in Figure \ref{fig:boxplot-search}).
-
-Conversely, performing MOP alone leads to many, many failed searches, as seen in Figure \ref{fig:hist-all} and the right panel in Figure \ref{fig:boxplot-search}.
-This is a difficult, nonconvex, and noisy problem, and the search gets stuck in local minima and saddle points, failing to approach the MLE.
-
-IFAD, in comparison, successfully manages to (1) approach the MLE quickly due to the IF2 warm-start (as seen in Figure \ref{fig:optim}) and (2) refine the coarse solution found by the warm-start with MOP gradient steps to find the MLE (as seen in Figures \ref{fig:scatter}, \ref{fig:boxplot-search}, and \ref{fig:hist-all}).
-IFAD therefore successfully combines the best qualities of IFAD and MOP, outperforming either of them alone.
-
-\begin{figure}[htbp!]
-    \centering
-    \includegraphics[scale=0.53]{../imgs/095/pairs.png}
-    \includegraphics[scale=0.53]{../imgs/095/qq.png}
-    \caption{Scatterplots depicting the performance of IFAD against that of IF2.
-\textbf{Left:} Paired searches from the same starting point.
-\textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches.
-It is clear that on average, IFAD has the edge, and manages to find the MLE while no IF2 search successfully gets within 7 log-likelihood units of it.
-We use a dotted red line to display the MLE in both.
-}
-
-    \label{fig:scatter}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[scale=0.48]{../imgs/095/boxplot.png}
-    \includegraphics[scale=0.53]{../imgs/095/boxplotall.png}
-    \caption{\textbf{Left:} Boxplot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs.
-This simulates the result of the common procedure of running a few searches and choosing the best one.
-IFAD outperforms all other methods, and improves on the warm-start given by running 40 IF2 iterations.
-Our choice of hyperparameters for our implementation of IF2 outperforms the choices made by \cite{ionides15}, which we depict in the boxplot labeled "IF2 2015".
-As before, while IF2 successfully approaches the MLE, it ultimately fails to find it.
-    \textbf{Right:} Boxplot depicting the performance of IFAD, IF2, and MOP alone without the warm-start, where we plot all runs.
-We see that MOP alone drastically underperforms all other methods, failing to get close to the MLE.
-This shows that the IF2 warm-start may be necessary to achieve good performance in challenging nonconvex and noisy problems like this one.
-    \textbf{Takeaway:} In this panel, we see that (1) IF2 fails to find the MLE while the gradient steps perform better at fine-grained refinement, and (2) performing gradient steps alone without a warm-start leads to the search getting stuck in local minima and saddle points, ultimately demonstrating the necessity of a hybrid algorithm like IFAD.
-    In both, as before, we use a dotted red line to display the MLE. }
-    \label{fig:boxplot-search}
-\end{figure}
-
-\begin{figure}[htbp!]
-    \centering
-    \includegraphics[scale=0.7]{../imgs/095/optim.png}
-    \caption{Optimization progress of IFAD and IF2, where the solid lines depict the median negative log-likelihood at each iteration and the shaded area depicts the best search at any iteration and the 80\% percentile.
-The dashed orange line depicts the median warm-start given by running 40 IF2 iterations.
-While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD -- IFAD has better tail control and successfully reaches the MLE.
-The initial loss in log-likelihood upon transitioning from IF2 to MOP seems to help with escaping local minima.
-    As before, we use a dotted red line to display the MLE. }
-    \label{fig:optim}
-\end{figure}
-
-\begin{figure}[t!]
-    \centering
-    \includegraphics[scale=0.7]{../imgs/095/hist.png}
-    \caption{Histogram comparison of the performance of IFAD, IF2, and MOP alone.
-As before, we note that IFAD successfully reaches the MLE a nontrivial number of times, while IF2 ultimately fails to do so.
-Our implementation of IF2 outperforms that of \cite{ionides15}, largely due to improved hyperparameter choices.
-MOP alone performs poorly, demonstrating the necessity of warm-starting the search with IF2.
-    We once again see that IFAD combines the best of IF2 and MOP, approaching the MLE quickly due to the IF2 warm-start and successfully refining the coarse solution with gradient steps, achieving superior performance.
-As before, we use a dotted red line to display the MLE.
-}
-    \label{fig:hist-all}
-\end{figure}
-
-\paragraph{Differences Between IFAD-0 and IFAD-1:} The differences between IFAD-1 and IFAD-0 (MOP-1 and MOP-0) do not seem very pronounced in this example.
-This could be due to a number of reasons, such as (1) future case counts $y_{{n+1}:N}^*$ not providing a large amount of information on state identification $x_n$ given current and past measurements $y_{0:n}^*$ as mentioned in \cite{corenflos21}, and (2) the gradient normalization making the trajectories given by MOP-0 and MOP-1 more similar than they would be otherwise.
-
-\kevin{IFAD beats IF2 x percent of the time on 100 paired runs, but the bigger thing is that it gets close to the MLE without much effort.
-Previously, we'd have to do many searches, plus refinement (which is necessary, as we can see from the IF2 boxplot, it is easy to fool yourself into thinking that because you see the plateau that is the MLE).
-This includes likelihood profiling diagnostics, which are very computationally intensive.
-Now we can just take the best of ten searches -- IFAD does the refinement.}
-
-\kevin{Have an outsider read the manuscript afterwards.}
-
-\section{Discussion and Future Work}
-
-We develop a new theoretical framework to differentiating through the particle filter that (1) encompasses the gradient estimates of \cite{poyiadjis11}, \cite{blei2018vsmc}, and \cite{scibior2021dpf} as special cases, (2) sidesteps the issue of differentiating through a Monte-Carlo algorithm with discontinuous resampling, and (3) provides an opportunity for optimizing a bias-variance tradeoff.
-
-Prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics.
-In the presence of a differentiable simulator, MOP-$\alpha$ enjoys this very same desirable property, enabling practical maximum likelihood inference in a wider variety of situations.
-
-Furthermore, MOP-$\alpha$, and by extension IFAD, has runtime \textbf{linear, and not quadratic}, in the number of particles, unlike the marginal differentiable particle filter in \cite{scibior2021dpf} or the the optimal transport of resampling approach of \cite{corenflos21}.
-This does not necessarily come at a cost to Monte Carlo variance, unlike \cite{poyiadjis11} and \cite{corenflos2021ot} who suffer from variance quadratic in the length of the trajectory, as we can optimize $\alpha$ for a bias-variance tradeoff.
-
-In the presence of strong convexity, first and second-order gradient methods that use MOP-$\alpha$ as an estimate enjoy a linear convergence rate, providing support for numerical optimization with AD for the particle filter.
-
-This leads to a practical hybrid algorithm we call IFAD that warm-starts gradient descent, or some other first or second-order method using the gradient estimate given by MOP-$\alpha$, with IF2.
-When benchmarked on a challenging global search problem with the model of \cite{king08}, IFAD enjoys the best properties of IF2 and MOP-$\alpha$, successfully approaching a neighborhood of the MLE quickly without getting stuck in saddle points and (poor) local minima (like gradient descent alone would), while effectively optimizing the last few units of log-likelihood (which IF2 struggles with).
-
-Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual} that we call \texttt{pomp-py}.
-The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{JAX} Python package from \citet{jax2018github}.
-
-\subsection{Future Work}
-
-While MOP-$\alpha$, and IFAD in its current incarnation, remain very general, they currently lack the capability of handling discrete latent states.
-We conjecture that DOP-$\alpha$ might be a possible solution (despite needing the likelihood ratio of transition densities), and will explore this in future work.
-
-We conjecture that the linear convergence result in Theorem \ref{thm:convergence} applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE.
-IF2 posesses some SGD-like qualities, including what looks like linear convergence to a ball around the MLE with width determined by the random walk standard deviation.
-Not much is known about the convergence rate of IF2, and an answer to this would also provide an answer to the convergence rate of IFAD beyond just the convergence rate of the gradient stage.
-We will explore this conjecture in future work.
-
-In the interest of simplicity, we only explore the performance of gradient descent on the Dhaka cholera model of \cite{king08}.
-Exploring the use of various first and second order methods such as Newton's method, ADAM, L-BFGS, or other popular iterative optimization methods with the gradient estimates that MOP-$\alpha$ provides in the context of the particle filter could very well lead to nontrivial speedups and increased performance.
-
-Extensions of MOP-$\alpha$ and IFAD to panel and spatiotemporal datam, possibly through a block particle filter in the same vein as \cite{ionides22} and \cite{ning23}, would go a long way towards extending the current applicability of these methods to handle a wider variety of interesting and challenging scientific problems.
-
-While we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual}, this remains a work in progress.
-
-\paragraph{Acknowledgements.} We thank Nicholas Chopin for helpful comments on the proof of a strong law of large numbers for a suitably reweighted particle filter.
-
-\kevin{Reach: PNAS.
-Else try for JRSSB, try to get Chopin for Annals AE.
-Would need to change bracketing conventions for Biometrika.
-If submitting to PNAS, JASA, Biometrika, or JRSSB, highlight the different statistical gradient estimator and finite sample performance, rewrite introduction, maybe create a way to use the gradients for inference rather than using profiling.
-Try to rewrite it as a breakthrough in statistics.
-Safety: JMLR}
-
-\bibliography{paper/bib-ifad, paper/bib-ref}
-\bibliographystyle{apalike}
-
-\appendix
-\renewcommand{\thefigure}{A\arabic{figure}}
-\setcounter{figure}{0}
-
-\section{Weighted Particle Filters that Target the Posterior}
-
-\begin{defn}[Targeting]
-    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any measurable and bounded functional $h$,
-\begin{equation}
-    E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
-\end{equation}  
-    A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if for any measurable and bounded functional $h$,
-\begin{equation}
-    \frac{\sum_{j=1}^J h(X^J_j) w^J_j}{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X))
-\end{equation}
-as $J \to \infty$.
-\end{defn}
-\citet{chopin2004clt} asserted without proof that common particle filter algorithms target the filtering distribution, $f_{X_{1:N}|Y_{1:N}}$, in this sense.
-\ed{PERHAPS IF WE HAD A CONCRETE PROOF OF THIS, THAT WOULD HELP TO SEE HOW IT EXTENDS TO OUR SUBSEQUENT LEMMAS.
-ARGUABLY, THAT IS IN FACT WHAT WE ARE DEVELOPING VIA THESE LEMMAS.
-I WONDER IF SOME OTHER REFERENCE, E.G.
-DELMORAL'S 2004 BOOK, HAS AN EXPLICIT PROOF OF CONSISTENCY FOR SOME PARTICLE FILTER ALGORITHM.}
-In order to prove the consistency of our variation on the particle filter, we now present three helper lemmas.
-The first follows from standard importance sampling arguments, the second from integrating out the marginal, and the third from Bayes' theorem. 
-We state Lemma~\ref{lem:change-measure-proper-weights} assuming multinomial resampling, which is convenient for the proof though other resampling strategies may be preferable in practice REFS.
-
-\begin{lem}[Change of Particle Measure]
-    \label{lem:change-measure-proper-weights}
-    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
-Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, proportional to $\pi^J_j J$ times.
-Write
-    \[
-    (Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big),
-    \]
-    where $a$ is called the ancestor function. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j),j=1,\dots,J\}$ targets $f_X$.
+\begin{lem}
+    \label{lem:dobrushin-implies-alpha-mixing}
+    Setting $X$ to be the particle collection at time $m$, and $Y$ at time $n$, we have that the alpha mixing coefficients,
+    $$\int | f_{XY} - f_X f_Y | < \alpha,$$
+    are bounded by the Dobrushin coefficient, i.e we have
+    $$\int | f_{Y|X}(y|x_1) - f_{Y|X}(y|x_2) |dy < \alpha$$ for all $x_1$, $x_2$.
 \end{lem}
@@
 \begin{proof}
+    We rewrite the alpha-mixing assertion as $$\int { \int |f_Y|X(y|x) - f_Y(y)| dy } f_X dx.$$
 
-    Note that as the $Y_j^J$ are a subsample from $X_j^J$, $h$ can be a function of $Y$ as well as it is one for $X$. We then expand
-    $$\frac{\sum_j h(Y_j^J) V_j^J}{\sum_j V_j^J} = \frac{\sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}}.$$
-    By hypothesis,
-    $$\frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[f(X)].$$
+    We claim that the Dobrushin coefficient implies $$\int | f_Y|X(y|x_1) - f_Y(y) |dy < \alpha$$ for all $x_1$. This is shown as follows:
+    \begin{align*}
+        \int | f_{Y|X}(y|x) - f_Y(y) |dy 
+        &= \int \left|  \int [f_{Y|X}(y|x_1) - f_{Y|X}(y|x)] f_X(x)dx \right|dy \\
+        &< \int   \int  | f_{Y|X}(y|x_1) - f_{Y|X}(y|x)] | dy f_X(x)dx \\
+        &< \int \alpha f_X(x)dx \\
+        &=\alpha
+    \end{align*}
 
-    We want to show
-    \begin{equation}\label{eq:lemma1:h}
-    \frac{\sum_j h(X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}} - \frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} 0.
-    \end{equation}
-    For this, it is sufficient to show that
-   $$ \sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}
-    -  \sum_j h(\tilde X_j^J)\frac{U_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0 $$
-    since an application of this result with $h(x)=1$ provides almost sure convergence of the denominator in (\ref{eq:lemma1:h}).
-    Write $g(\tilde X_j^J) = h(\tilde X_j^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}$. We therefore need to show that 
-    $$\sum_j Z_j^J := \sum_j \left(g(\tilde X_{k_j}^J) -  g(\tilde X_j^J) \pi_j^J \right) \stackrel{a.s.}{\to} 0.$$
-
-    Because the functional $h$ and importance sampling weights $u_{k_j}^J/\pi_{k_j}^J$ are bounded, we have that $\E[\left(Z_j^J\right)^4] < \infty$.
-We can then follow the argument of \cite{chopin20} from this point on, where
-    {THE Zs ARE ONLY CONDITIONALLY INDEPENDENT, SO WE COULD PUT IN THE CONDITIONING EVENT (I.E., THE Xs AND Us) HERE TO BE MORE ACCURATE}
-    $$\E\left[\left(\sum_j Z_j^J\right)^4\right] 
-    = J\E[(Z_1^J)^4] + 3J(J-1)\left(\E[(Z_j^J)^2]\right) \leq CJ^2,$$ for some $C>0$. By Markov, 
-    $$\mathbb{P}\left(\left|\frac{1}{J} \sum_{j=1}^J Z_j^J\right|>\epsilon\right) 
-    \leq \frac{1}{\epsilon^4J^4 } 
-    \mathbb{E}\left[\left(\sum_{j=1}^J Z_j^J\right)^4\right] \leq \frac{C}{\epsilon^4J^2},$$
-    and as these terms are summable we can apply Borel-Cantelli to conclude that these deviations happen only finitely often for every $\epsilon>0,$ giving us the almost-sure convergence for
-
-    $$ \sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}
-    -  \sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0.$$
-
-    Similarly, we also have that
-    $$ \sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J
-    - \sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J} \stackrel{a.s.}{\to} 0,$$
-
-    and the result is proved.
-
-    \kevin{MAY HAVE TO DEAL WITH CONDITIONING ON THE FILTRATION AT THE LAST TIMESTEP, AND THEN USING THE TOWER PROPERTY. EXPECTATION WILL BE THE SAME REGARDLESS FOR THE FOURTH MOMENT.}
-
-   % Let $h$ be an integrable function. 
-    %\begin{align*}
-    %    \E\left[\sum_{j=1}^J h(Y_j) v_j\right] 
-    %    &= \sum_{j=1}^J \E\left[h(X_{a(j)}) \frac{u_{a(j)}}{\pi_{a(j)}}\right] \\
-    %    &= \sum_{j=1}^J \E\left[h(X_{j}) u_{j}\right],
-    %\end{align*}
-    %and similarly for the denominator.
-Now, the numerator and denominator of the reweighted quantity have the same expectation as the numerator and denominator of the original quantity, so they must converge to $ E_\pi(h(X))$ almost surely as well.
+    We then have the desired result.
 \end{proof}
 
-\textbf{Remark:} Note that Lemma \ref{lem:change-measure-proper-weights} permits $\pi_{1:J}$ to depend on $\{(X_j,u_j)\}$ as long as the resampling is carried out independently of $\{(X_j,u_j)\}$, conditional on $\pi_{1:J}$.
+\subsection{Warm Up: MOP-$0$ Variance Bound}
 
-\begin{lem}[Particle Marginals]
-    \label{lem:marginal-proper-weights}
-    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
-Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
-Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}$ targets $f_Z$.
-\end{lem}
-\begin{proof}
-    We want to show that, for any measurable bounded $h$, 
-     $$\frac{\sum_j h(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[h(Z)] = \E\big[\E[h(Z) | X]\big].$$
-    By hypothesis, for any measurable and bounded functional $g$ with domain $\gX$,
-    \begin{equation}\label{eq:lemma2:g}
-    \frac{\sum_j g(\tilde X_j^J)\, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[g(X)].
-    \end{equation}
-    Let $\bar{U}_j^J = \frac{J U_j^J}{\sum_j U_j^J}$.
-Examine the numerator and denominator of the quantity $$\frac{J^{-1}\sum_j h(\tilde Z_j^J) \, U_j^J}{J^{-1}\sum_j U_j^J} = \frac{J^{-1}\sum_j h(\tilde Z_j^J) \, \bar{U}_j^J}{J^{-1}\sum_j \bar{U}_j^J}.$$
-    The denominator converges to $1$ almost surely. The numerator, on the other hand, is
-    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J,$$
-    by the same fourth moment argument to the above lemma converges almost surely to the limit of its expectation
-    \begin{eqnarray*}        
-    \lim_{J\to\infty} \E\left[\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J\right] 
-    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  
-      \E\left[ h(\tilde Z_j^J)\, \bar{U}_j^J\big|\tilde X_j^J, \bar U_j^J\right]
-    \right]
-    \\
-    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  \E\big[h(Z) \big| X=\tilde X_j^J \big] \, \bar U_j^J\right].
-    \end{eqnarray*}
-    Applying (\ref{eq:lemma2:g}) with $g(x) = \E\big[h(Z) | X=x\big]$, the average on the right hand size converges almost surely to $\E\big\{\E[h(Z)|X]\big\}=\E[h(Z)].$
-    It remains to swap the limit and expectations. We can do so with the bounded convergence theorem, and therefore obtain   
-    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\,\bar{U}_j^J \stackrel{a.s.}{\to} \E[h(Z)].$$
+We now prove the variance bound, presented below.
 
-\end{proof}
+\begin{thm}
 
-\begin{lem}[Particle Posteriors]
-    \label{lem:posterior-proper-weights}
-    Suppose that $\{(X_j^J,u_j^J),j=1,\dots,J\}$ targets $f_X$.
-Also suppose that $(X^{\prime J}_j,u^{\prime J}_j) = \big(X_j^J,u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$.
-Then, if $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,u^{\prime J}_j)\}$ targets $f_{X|Z}(\cdot | z^*)$.
-\end{lem}
+\end{thm}
 
-\begin{proof}
-
-    $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J}}{J^{-1} \sum_j \bar{u}_j^{' J}}$$
-
-    by
-
-    \begin{equation}
-    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}}
-    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \right)^{-1}
-    \end{equation}
-
-    Again, we want to show that
-     $$\frac{\sum_j h(X_j^J) \cdot u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)} \stackrel{a.s.}{\to} \E_{f_{X|Z}}[h(X)|z^*].$$
-
-    We already have that
-    $$\frac{\sum_j h(X_j^J)u_j^J}{\sum_j u_j^J} \stackrel{a.s.}{\to} \E_{f_X}[h(X)].$$
-
-    %= \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j f_{Z|X}(z^*|X_j^J)f_X(X_j^J)}
-
-    Consider the following:
-    \begin{equation}
-    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}}
-    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {u}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \right)^{-1}.
-    \end{equation}
-
-    Write $\bar{u}_j^{J} = \frac{J u_j^J}{\sum_j u_j^J}$.
-Define the weights $u^{' J}_j := u_j^J \cdot f_{Z|X}(z^*|X_j^J),$ and the corresponding self-normalized weights $\bar{u}_j^{' J} := \frac{J u^{' J}_j}{\sum_j u^{' J}_j}.$ The denominator of the below quantity
-    $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J}}{J^{-1} \sum_j \bar{u}_j^{' J}}$$
-    converges to $1$, while the numerator, by the same fourth moment argument as the first lemma, converges almost surely to the limit of its expectation, which is
-    \begin{align*}
-        \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \bar{u}_j^{' J} \right]
-        &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j u_j^J \cdot f_{Z|X}(z^*|X_j^J)}\right] \\
-        &= \lim_{J\to\infty} \E\left[J^{-1} \sum_j h(X_j^J) \frac{J \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
-        &= \lim_{J\to\infty} \E\left[ \sum_j h(X_j^J) \frac{\frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}{\sum_j \frac{u_j^J \cdot f_{Z|X}(z^*|X_j^J)}{f_Z(z^*)}}\right]\\
-        &= \E_{X|Z}[h(X)|z^*],
-    \end{align*}
-
-    where we can swap the limit and expectations with the bounded convergence theorem.
-\end{proof}
-
-\begin{prop}[MOP-1 Targets the Posterior]
-    When $\alpha=1$ or $\phi=\theta$, MOP-$\alpha$ targets the posterior. 
-\end{prop}
-\begin{proof}
-    When $\theta=\phi$, regardless of the value of $\alpha$, the ratio $\frac{g_{n,j}^\theta}{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter estimate.
-
-    When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows.
-Recursively applying Lemmas \ref{lem:change-measure-proper-weights}, \ref{lem:marginal-proper-weights}, and \ref{lem:posterior-proper-weights}, we obtain that
-    %to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
-    the MOP-1 filter targets the posterior.
-    Specifically, suppose inductively that $\big\{\big(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
-    Then, Lemma \ref{lem:marginal-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big)\big\}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
-    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-    Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
-    \[
-    \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}\big/ g^\phi_{n,a(j)}\big),
-    \]
-    with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-\end{proof}
-
-This has addressed filtering, but not quite yet the likelihood evaluation. For this we use the following lemma.
-
-\begin{lem}[Likelihood Proper Weighting]
-    \label{lem:lik-proper-weight}
-  $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1_n-1}^*;\theta)$ is consistently estimated by either the before-resampling estimate,
-\begin{equation}\label{L1}
-L_n^{B,\theta} =  \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
-\end{equation}
-or by the after-resampling estimate,
-\begin{equation}\label{L2}
-L_n^{A,\theta} = L^\phi_n \frac{\sum_{j=1}^Jw^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}.
-\end{equation}
-where $L^\phi_n$ is as defined in the various algorithms.
-\end{lem}
-
-Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
-To see  (\ref{L2}),
-we write the numerator of (\ref{lem:change-measure-proper-weights}) as
-\[
-L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
-= L^\phi_n \sum_{j=1}^J w_{n,j}^{FC,\theta} \frac{g^\phi_{n,j}}{L_n^\phi}
-\]
-Using Lemma \ref{lem:change-measure-proper-weights}, we resample according to probabilities $\frac{g^\phi_{n,j}}{L_n^\phi}$ to see this is properly estimated by
-\[
-L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},
-\]
-from which we obtain (\ref{L2}).
-
-Using Lemma \ref{lem:lik-proper-weight}, we obtain a likelihood estimate,
-\[
-L^{A,\theta} = \prod_{n=1}^N \left( L^\phi_n \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J w^{P,\theta}_{n,j}}\right).
-\]
-Since $w^{F,\theta}_{n,j}=w^{P,\theta}_{n+1,j}$, this is a telescoping product. The remaining terms are
-$\sum_{j=1}^J w^{P,\theta}_{0,j} = J$ on the denominator and $\sum_{j=1}^J w^{F,\theta}_{N,j}$ on the numerator.
-This derives the MOP/DOP estimate.
-
-\kevin{IS THIS UNBIASED?}
-
-$L^{B,\theta}$ should generally be preferred in practice, since there is no reason to include the extra variability from resampling when calculating the conditional log likelihood, but it lacks the nice telescoping product that lets us derive exact expressions for the gradient later.
-
-\section{Doubly Off-Policy Particle Filters}
-\label{app:dop}
-
-The proof for Proposition \ref{prop:dop-correctness} follows from the below two lemmas.
-Evaluating the off-policy likelihood is possible with a weighted particle filter by using resampling weight $p(y^*_n|x_n;\phi)$ and making two corrections in the particle weight, through multiplying $w_{t,j}$ by the two likelihood ratios below,
-
-\begin{equation}
-    \label{eq:dmeas-ratio}
-    s=p(y^*_n|x_n;\theta)/p(y^*_n|x_n;\phi)
-\end{equation}
-and
-\begin{equation}
-    \label{eq:rproc-ratio}
-    r=p(x_n|x_{n-1};\theta)/p(x_n|x_{n-1};\phi).
-\end{equation}
-
-The following result showcases the correctness of corrections \ref{eq:dmeas-ratio} and \ref{eq:rproc-ratio}.
-
-\begin{lem}
-    For $\theta$, $\phi$ that satisfy Proposition \ref{prop:smooth-nbhd}, the following correction in the particle filter weights,
-    \begin{equation}
-        w_{t,j} := \tilde{w}_{t,j} \cdot p(y^*|x_{t,j}^P,\phi)\cdot r \cdot s
-    \end{equation}
-    successfully recovers $\lik(\theta)$.
-\end{lem}
-
-\begin{proof}
-    Recall $\lik(\theta) = \E[\lik(\theta,\phi,\Omega, J)]=\int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega)$ and $\lik(\theta) = \int_\Omega \lik(\theta, \omega, J) d\mu(\omega)$.
-We want to evaluate
-    \begin{align*}
-        \lik(\theta,\omega, J)
-        &= \E\left[ \prod_{t=1}^T \lik_t(\theta) \middle| \omega\right]
-        = \E\left[ \prod_{t=1}^T p(y_t^*|y_{1:t-1}^*,\theta) \middle| \omega\right] \\
-        &= \E\left[ \prod_{t=1}^T \frac{1}{J} \sum_{j=1}^J  p(y_t^*|x_{t,j}^P,\theta) p(x_{t,j}^P|y_{1:t-1}^*,\theta) \middle| \omega\right] \\
-        &= \E\left[ \prod_{t=1}^T\frac{1}{J}  \sum_{j=1}^J  p(y_t^*|x_{t,j}^P,\theta) p(x_{t,j}^P|x_{t-1}^F,\theta) \middle| \omega\right] 
-    \end{align*}
-    where $p(x_t^P|y_{1:t-1}^*,\theta)$ is the prediction distribution of particles, and the filtering distribution of particles is proportional to $p(y_t^*|x_{t,j}^P,\theta)$, the likelihood of particle $j$ at time $t$.
-
-    Consider first the case with no resampling.
-Let $\lik^j(\theta)$, $\lik^j(\phi)$ be the likelihood of trajectory $j$ under $\theta$ and $\phi$ respectively.
-As the prediction and filtering particles are then the same, we use $x_{t,j}$ to refer to both.
-As such, conditional on $\omega$,
-    \begin{align*}
-        \lik^j(\theta) 
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \theta) p(x_{t,j}|x_{t-1,j}, \theta) \\
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \frac{p(y_t^*|x_{t,j}, \theta)}{p(y_t^*|x_{t,j}, \phi)} p(x_{t,j}|x_{t-1,j}, \theta) \frac{p(x_{t,j}|x_{t-1,j}, \theta) }{p(x_{t,j}|x_{t-1,j}, \phi) } \\
-        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \cdot s \cdot p(x_{t,j}|x_{t-1,j}, \phi)\cdot r
-    \end{align*}
-
-    and therefore, if trajectories were proposed with equal probability with particles conditional on $\omega$,
-    $$ \lik(\theta, \omega, J) = \frac{1}{J} \sum_{j=1}^J \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \cdot s \cdot p(x_{t,j}|x_{t-1,j}, \phi)\cdot r $$
-    The correction is therefore sufficient for recovering $\lik(\theta)$, with trajectories collected with the process model at $\phi$ and measurements evaluated with the measurement model at $\phi$.
-
-    Consider now the case with resampling.
-Without loss of generality we can resample every timestep.
-Here, $x_{t,j}^F(\phi) \sim p(y^*|x_{t,j}^P, \phi)$ and $x_{t,j}^P(\phi) \sim p(x_t|x_{t-1}^F,\phi)$, and similarly for $\theta$.
-We use this notation to emphasize that \textbf{both} the measurement model and the previously drawn particles depend on $\phi$ or $\theta$.
-
-    \begin{align*}
-        \lik_t(\theta) &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \theta)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\theta) \right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \theta)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \frac{p(y_t^*|x_{t,j}^P,\theta)}{p(y_t^*|x_{t,j}^P,\phi)} \right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \frac{p(y_t^*|x_{t,j}^P,\theta)}{p(y_t^*|x_{t,j}^P,\phi)} \frac{p(x_{t,j}|x_{t-1,j}, \theta) }{p(x_{t,j}|x_{t-1,j}, \phi) }\right] \\
-        &= \E_{x_{t,j}^P \sim p(x_t|x_{t-1}^F, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \cdot s \cdot r \right] \\
-        &= \E_{\omega}\E_{x_{t,j}^P \sim \process(x_{t-1}^F, \omega, \phi)} \left[\frac{1}{J}\sum_{j=1}^J p(y_t^*|x_{t,j}^P,\phi) \cdot s \cdot r \middle| \omega \right]
-    \end{align*}
-
-    and again the correction is sufficient.
-
-    Therefore,
-
-    \begin{align}
-         \mathcal{L}(\theta) 
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^P(\omega), \theta) p(x_{n,j}^P(\omega) | x_{n-1,j}^P(\omega), \theta) \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\theta}(\omega), \theta) p(x_{n,j}^{P,\theta}(\omega) | x_{n,j}^P(\omega), \theta) \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\theta}(\omega), \phi) p(x_{n,j}^{P,\theta}(\omega) | y_n^*, \phi) \cdot s  \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) p(x_{n,j}^{P,\phi}(\omega) | y_n^*, \phi) \cdot r \cdot s \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) p(x_{n,j}^{P, \phi}(\omega) | y_n^*, \phi) \cdot r \cdot s \right) \, d\mu(\omega) \\
-         &= \int_\Omega \left(
-         \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) \cdot r \cdot s \right)  p(x_{1:N} | y_{1:N}^*, \phi) \,  d\mu(\omega) \\
-         &=: \int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega).
-     \end{align}
-
-     The proof for an arbitrary functional under the particle measure follows analogously.
-\end{proof}
-
-\begin{lem}[Gradient of Doubly Off-Policy Particle Filter]
-    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$.
-Also let $\theta \in \gN(\phi)$.
-When $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{poyiadjis11},
-    \begin{equation}
-        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
-    \end{equation}
-\end{lem}
-
-\begin{proof}
-    It is known, from \cite{doucet2009tutorial}, that 
-    $$p_\theta(x_{1:N}|y_{1:N}) = \prod_{n=1}^N \frac{p(x_n|x_{n-1})p(y_n|x_n)}{p(y_n|y_{1:n-1})} = \prod_{n=1}^N \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{\lik_n(\theta)}.$$
-
-    From the above lemma, and the fact that the particle filter is unbiased for the likelihood, we know that
-    \begin{align*}
-        \lik(\theta)
-        &= \int_{x_{1:N}} \prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) p_\theta(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N  p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \right) \frac{p_\theta(x_{1:N}|y_{1:N})}{p_\phi(x_{1:N}|y_{1:N})}p_\phi(x_{1:N}|y_{1:N}) dx_{1:N}\\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N \lik_n(\theta) \right) \left(\prod_{n=1}^N \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n) \lik_n(\phi)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) \lik_n(\theta)}\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N \lik_n(\phi) \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) }\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N}.
-    \end{align*}
-
-    Therefore, under regularity conditions that permit differentiating under the integral sign, using the log-derivative trick, we have
-    \begin{align*}
-        \nabla_\theta \lik(\theta)
-        &= \int_{x_{1:N}} \nabla_\theta \hat\lik(\theta) \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N}\\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N \lik_n(\phi) \frac{p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)}{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n) }\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N \left(\frac{p_\phi(x_n|x_{n-1})p_\phi(y_n|x_n)}{\lik_n(\phi)}\right)^{-1} p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\frac{1}{p_\phi(x_{1:N}|y_{1:N}) }\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)p_\phi(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\nabla_\theta \log\left(\prod_{n=1}^N p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} p_\theta(x_{1:N},y_{1:N}) \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) dx_{1:N} \\
-        &= \int_{x_{1:N}} \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) p_\theta(x_{1:N}|y_{1:N}) \cdot \lik(\theta) \; dx_{1:N}. 
-    \end{align*}
-
-    Using the log-derivative trick again,
-    \begin{align*}
-        \nabla_\theta \ell(\theta)
-        &= \frac{\nabla_\theta \lik(\theta)}{\lik(\theta)} \\
-        &= \int_{x_{1:N}} \left(\sum_{n=1}^N \nabla_\theta \log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right) p_\theta(x_{1:N}|y_{1:N}) dx_{1:N} \\
-        &= \E_{x_{1:N}|y_{1:N}, \theta}\left[\nabla_\theta \left(\sum_{n=1}^N\log\left(p_\theta(x_n|x_{n-1})p_\theta(y_n|x_n)\right)\right)\right] \\
-        &= \E_{x_{1:N}|y_{1:N}, \theta}\left[\nabla_\theta\log p_\theta(x_{1:N},y_{1:N})\right] \\
-        &= \lim_{J\to\infty} \E_{x_{1:N}|y_{1:N}, \theta}\left[\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log p_\theta(x_{1:N,j}^{A,F},y_{1:N}^*)\right].
-    \end{align*}
-
-    But it also holds that
-    \begin{align*}
-        \nabla_\theta \ell(\theta)
-        &= \frac{\nabla_\theta \lik(\theta)}{\lik(\theta)} \\
-        &= \int_{x_{1:N}} \frac{\nabla_\theta \hat\lik(\theta)}{\lik(\theta)} \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N} \\
-        &= \int_{x_{1:N}} \nabla_\theta \hat\ell(\theta) \;p_\phi(x_{1:N}|y_{1:N})dx_{1:N}\\
-        &= \lim_{J\to\infty} \E_{x_{1:N}|y_{1:N}, \phi}\left[\nabla_\theta \hat\ell(\theta)\right].
-    \end{align*}
-
-    and we see that $$\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log p_\theta(x_{1:N,j}^{A,F},y_{1:N}^*)$$ and $$\nabla_\theta \hat\ell(\theta)$$ both converge to the same thing in expectation.
-They are both particle approximations under the posterior where the variance of each is bounded, therefore one can apply the particle CLT from \cite{chopin2004clt} to obtain that they are both consistent estimates of the score.
-
-    In fact, when $\theta=\phi,$ they are the same in distribution. \kevin{claim}
-
-    \kevin{This is a far weaker result than what I really wanted. This just gives consistency. Want equality in distribution.}
-\end{proof}
-
-\begin{lem}
-
-Consider the case of DOP-$\alpha$ when $\alpha=0$ and $\theta=\phi$. The gradient estimate is then
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta\right),
-    \end{equation}
-
-    yielding the gradient estimator of \cite{blei2018vsmc} that \cite{scibior2021dpf} show is the gradient of a vanilla particle filter when the resampling terms are dropped.
-
-\end{lem}
-
-\begin{proof}
-    When $\alpha=0,$ the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J r_{n,j}s_{n,j}}{\sum_{j=1}^J r_{n,j}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J r_{n,j}s_{n,j}.
-\end{equation}
-
-Its gradient when $\theta=\phi$ is therefore 
-\begin{align*}
-    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right) \\
-    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J r_{n,j}s_{n,j}\right)} \\
-    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta r_{n,j}s_{n,j}}{\sum_{j=1}^J r_{n,j}s_{n,j}} \\
-    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{Y_n, X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)} \\
-    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n,X_n|X_{n-1}}(y_n^*, x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta\right),
-\end{align*}
-where we use the log-derivative trick, observe that $\sum_{j=1}^J r_{n,j}s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again. 
-\end{proof}
-
-\subsection{Off-Policy Weighted Differentiable Particle Filter}
-
-We are finally ready to introduce the off-policy weighted differentiable particle filter.
-In the event where a seed is fixed, all expressions in Algorithm \ref{alg:dop} can be considered to be conditional on $\omega \in \Omega$.
-
-\begin{algorithm}[ht]
-\centering
-	\caption{Doubly Off-Policy-$\alpha$}
-    \label{alg:dop}
-	\begin{algorithmic}[1]
-	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
-		\STATE Initialize filter particles ${X}_{0,j}^{F,\phi}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\phi}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$. Fix $\omega.$
-		\FOR{$n=1,...,N$}
-            \STATE Prediction weights with discounting: $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
-            \label{dop-alpha:discount}
-            \STATE Simulate for prediction:
-            ${X}_{n,j}^{P,\phi}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\phi}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{dop-alpha:step1}
-            \STATE Adjust weights: $\displaystyle w_{n,j}^{P,\theta} = w_{n,j}^{P,\theta} \times
-            \frac{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\phi}|{X}_{n-1,j}^{F};{\theta}\big)}{{f}_{{X}_{n}|{X}_{n-1}}\big({X}_{n,j}^{P,\phi}|{X}_{n-1,j}^{F};{\phi}\big)}$ for $j$ in $1:J$
-            \label{dop-alpha:dproc}
-            \STATE Evaluate measurement density:
-            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\phi}\giventh{\theta})$ for $j$ in $\seq{1}{J}$
-            \STATE Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \STATE Conditional likelihood under $\phi$: 
-            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
-            \label{dop-alpha:Lphi}
-            \STATE Normalize weights:
-            $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
-            for $j\ \mathrm{in}\ \seq{1}{J}$
-            \STATE Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \label{dop-alpha:systematic}
-            \STATE Resample particles: ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$
-            \STATE Filter weights corrected for resampling:
-            $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{dop-alpha:weight:update}
-            \STATE Resample filter weights:
-            $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
-            for $j$ in $\seq{1}{J}$ \label{dop-alpha:step2}
-            \STATE After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \ENDFOR
-		\RETURN likelihood estimate $\hat{\lik}(\theta) = \lik(\theta, \phi, \omega, J) := \prod_{n=1}^N L_n^{A,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}.$
-	\end{algorithmic}
-\end{algorithm}
-
-Here, we consider a doubly off policy particle filter, meaning that both \code{rprocess} and \code{dmeasure} are computed at $\phi$ and particles are reweighted to correspond to $\theta$.
-
-\section{Measurement Off-Policy Particle Filters}
-
-\begin{algorithm}[ht]
-\centering
-	\caption{Measurement Off-Policy-$\alpha$}
-    \label{alg:mop}
-	\begin{algorithmic}[1]
-	     \STATE \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
-        \IF{$\theta \neq \phi$}
-            \STATE Run a particle filter once with parameters $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$, with seed $\omega$.
-        \ELSE
-        \STATE Set $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$ to be copies of $X_{n,j}^{P,\theta}, X_{n,j}^{P,\theta}$ in the rest of the algorithm.
-        \ENDIF
-		\STATE Initialize filter particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$. Fix $\omega.$
-		\FOR{$n=1,...,N$}
-            \STATE Prediction weights with discounting: $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
-            \label{mop-alpha:discount}
-            \STATE Simulate for prediction:
-            ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{mop-alpha:step1}
-            \STATE Evaluate measurement density:
-            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for $j$ in $\seq{1}{J}$
-            \STATE Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \STATE Conditional likelihood under $\phi$: 
-            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
-            \label{mop-alpha:Lphi}
-            \STATE Normalize weights:
-            $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
-            for $j\ \mathrm{in}\ \seq{1}{J}$
-            \STATE Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \label{mop-alpha:systematic}
-            \STATE Resample particles: ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$
-            \STATE Filter weights corrected for resampling:
-            $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \label{mop-alpha:weight:update}
-            \STATE Resample filter weights:
-            $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
-            for $j$ in $\seq{1}{J}$ \label{mop-alpha:step2}
-            \STATE After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-            \ENDFOR
-		\RETURN likelihood estimate $\hat{\lik}(\theta) = \lik(\theta, \phi, \omega, J) := \prod_{n=1}^N L_n^{A,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}.$
-	\end{algorithmic}
-\end{algorithm}
-
-\begin{lem}
-    \label{lem:mop-1-formula}
-    Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The gradient estimate is then
-
-    \begin{equation}
-        \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
-    \end{equation}
-
-    yielding the gradient estimators of \cite{poyiadjis11, scibior2021dpf} when applied to the bootstrap filter. 
-\end{lem}
-
-\begin{proof}
-
-Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. The likelihood estimate becomes
-\begin{equation*}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}}= \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\theta}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
-\end{equation*}
-as we have a telescoping product. Then, as
-
-$$\nabla_\theta \hat\ell(\theta) = \frac{\nabla_\theta \hat\lik(\theta)}{\hat\lik(\theta)} = \frac{\nabla_\theta\left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi}{\prod_{n=1}^N L_n^\phi} =  \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},$$
-
-The derivative of the log-likelihood estimate is then
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},
-\end{equation*}
-which we decompose as follows.
-
-Observe that as $\alpha=1$,
-
-$$w_{n,j}^{P,\theta} = w_{n-1,j}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}},$$
-where we use the $(\cdot)^A$ superscript to denote the ancestral trajectory of the $j$-th prediction or filtering particle at timestep $n$.
-
-Using the log-derivative identity again,
-$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,j}^{A,P,\theta} - \log g_{i,j}^{A,P,\phi}\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,j}^{A,P,\theta}.$$
-
-So 
-$$  \nabla_\theta \sum_{n=1}^N \log g_{n,j}^{A,\theta} = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,j}^{A,P,\theta}\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,j}^{A, P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right),$$
-
-and 
-$$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{n=1}^N \nabla_\theta \log g_{n,j}^{A,P,\theta} = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$
-
-Substituting, we have that
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right),
-\end{equation*}
-and finally, observing that when $\theta=\phi$ we have that $w_{N,j}^{F,\theta}=1$, that 
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
-\end{equation*}
-yielding the gradient estimators of \cite{poyiadjis11, scibior2021dpf} when applied to the bootstrap filter. 
-\end{proof}
-
-Note that one would believe that the variance blows up the moment $\theta\neq\phi$. This is true, as 
-\begin{equation*}
-    \nabla_\theta \hat{\ell}(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right),
-\end{equation*}
-and when $\theta\neq\phi$, $w_{N,k_j}^{P,\theta} = O(c^N)$.
-When $\theta=\phi$, this is a special case of the \cite{poyiadjis11} estimator, which has $O(N^2)$ variance by a property of functionals of the particle filter by \cite{delMoral03}.
-
-\begin{lem}
-\label{lem:mop-0-formula}
-
-Consider the case of MOP-$\alpha$ when $\alpha=0$ and $\theta=\phi$. The gradient estimate is then
-
-    \begin{equation}
-        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
-    \end{equation}
-
-    yielding the gradient estimator of \cite{blei2018vsmc} that \cite{scibior2021dpf} show is the gradient of a vanilla particle filter when the resampling terms are dropped -- when applied to the bootstrap filter.
-
-\end{lem}
-
-\begin{proof}
-    When $\alpha=0,$ the likelihood estimate becomes
-\begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
-\end{equation}
-
-Its gradient when $\theta=\phi$ is therefore 
-\begin{align*}
-    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
-    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)} \\
-    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta s_{n,j}}{\sum_{j=1}^J s_{n,j}} \\
-    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)}{f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \phi}; \phi)} \\
-    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
-\end{align*}
-where we use the log-derivative trick, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again. 
-\end{proof}
-
-\kevin{I have no clue how to prove this. Help!}
-\begin{conj}[MOP-1 Estimate Is Unbiased for the Likelihood and the Gradient of the Likelihood]
-    Consider the case of MOP-$\alpha$ when $\alpha=1$, and resampling is carried out with either multinomial, residual, or systematic resampling at every timestep.
-Assume the gradient estimate obtained by the gradient of the likelihood estimate with J particles, $\nabla_\theta \hat\lik_J(\theta)$ is uniformly bounded for all $J$, and that the likelihood has two continuous derivatives.
-Then the likelihood estimate is unbiased for the likelihood and the gradient of the likelihood estimate is unbiased for the gradient of the likelihood.
-That is, it holds that
-    $$\E[\hat\lik_J(\theta)] = \lik(\theta),\;\; \E[\nabla_\theta \hat\lik_J(\theta)] = \nabla_\theta \lik(\theta).$$
-\end{conj}
-
-\begin{proof}
-    First, we need to establish the unbiasedness of the likelihood estimate. This is 
-    $$\hat\lik_J(\theta) := \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right)\prod_{n=1}^N L_n^\phi = \left(\frac{1}{J}\sum_{j=1}^J \prod_{n=1}^N \frac{g_{n,j}^{A,\theta}}{g_{n,j}^{A,\phi}} \right)\prod_{n=1}^N L_n^\phi.$$
-
-    We know from \cite{chopin20} that $\E[L_n^{\phi}] = \lik(\phi).$ It then suffices to show that
-    $$\E\left[\left(\frac{1}{J}\sum_{j=1}^J \prod_{n=1}^N \frac{g_{n,j}^{A,\theta}}{g_{n,j}^{A,\phi}} \right)\right] = \frac{\lik(\theta)}{\lik(\phi)}.$$
-
-    To obtain this, we need to see that the resampling under $\phi$, with corrected weights, is unbiased. That is, we need to show that for any bounded measurable functional $\varphi : \gX \to \R$,
-    $$\E\left[\frac{\sum_{j=1}^J w_{n,j}^{F, \theta} \varphi(x_{n,j}^{\theta, F})}{\sum_{j=1}^J w_{n,j}^{F, \theta}} \bigg| \gF_n \right] = \frac{\sum_{j=1}^J w_{n,j}^{P, \theta} \varphi(x_{n,j}^{P, \theta})}{\sum_{j=1}^J w_{n,j}^{P, \theta} }.
-$$
-
-    To see this, 
-    \begin{align*}
-        \sum_{j=1}^J w_{n,j}^{P, \theta} \varphi(x_{n,j}^{P, \theta}) 
-        &= \sum_{j=1}^J w_{n-1,j}^{F, \theta}\frac{g_{n,j}^{\theta}}{g_{n,j}^{\phi}}  \varphi(x_{n,j}^{P, \theta})
-    \end{align*}
-
-%     Observe that our differentiable simulator $\process(\cdot, \theta)$ and choice of weights now satisfy:
-
-%     - A Feynman-Kac model $\left\{M_t, G_t\right\}$ such that:
-
-% - the weight function $G_t$ may be evaluated pointwise (for all $t$ );
-
-% - it is possible to simulate from $\mathbb{M}_0\left(\mathrm{~d} x_0\right)$ and from $M_t\left(x_{t-1}, \mathrm{~d} x_t\right)$ (for any $x_{t-1}$ and $t$ ).
-
-% - The number of particles $N$.
-
-% - The choice of an unbiased resampling scheme (e.g.
-multinomial, residual, stratified, systematic), that is, an algorithm to generate variables $A_t^{1: N}$ in $1: N$ such that $\mathbb{E}\left[\sum_{m=1}^N \mathbbm{1}\left\{A_t^m=n\right\}\right]=N W_{t-1}^n$; see previous chapter for more details.
-\end{proof}
-
-\begin{lem}[Consistency of MOP-1 Gradient Estimate]
-    Consider the case of MOP-$\alpha$ when $\alpha=1$.
-Assume the gradient estimate obtained by the gradient of the log-likelihood estimate with J particles, $\nabla_\theta \hat\ell_J(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}$ is uniformly bounded by $G(\theta)$ for all $J$, and that the likelihood has two continuous derivatives.
-Then the gradient estimate, $\nabla_\theta \hat\ell_J(\theta)$, is consistent for the score.
-    \label{lem:mop-grad-consistency}
-\end{lem}
-\begin{proof}
-
-    % We know that the variance of the MOP-1 gradient estimate $\nabla_\theta \hat\ell_J(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}$ goes to zero coordinate-wise as $J\to\infty$ by the particle filter law of large numbers for bounded functionals of particles (\kevin{Prove general LLN for functionals of MOP under boundedness assumptions in L2, but maybe Hoeffding may be enough to show this, because we only want convergence in probability, we just need to show the expectation converges}), so it suffices to show that it is asymptotically unbiased.
-
-    % Write $\hat\pi_{n,J}$ for the filtering distribution at timestep $n$ and with $J$ particles, $\hat\pi_{1:N,J}$ for the sequence of filtering distributions from timesteps $1$ to $N$, and $\pi_n, \pi_{1:N}$ for the respective posteriors.
-
-    % From Lemma \ref{lem:lik-proper-weight}, we know that $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1:n-1}^*, \theta)$ is consistently estimated by the conditional likelihood estimate obtained by MOP-1, for all $\theta$ in an open neighborhood of $\phi$.
-The likelihood is simply the product of the conditional likelihoods, and similarly for the estimates, so the likelihood $\lik(\theta)$ is also consistently estimated by the likelihood estimate obtained by MOP-1, $\hat\lik_J(\theta)$, for all $\theta$ in an open neighborhood of $\phi$, $\gN(\phi)$.
-
-    % We therefore have that 
-    % $$\lim_{J\to\infty} \E[\hat\lik_J(\theta)] = \lik(\theta)$$
-    % pointwise for all $\theta \in \gN(\phi)$.
-Given that the gradient is bounded uniformly for all $J$ and consistency for the likelihood estimate holds for all $\theta$ in an open neighborhood of $\phi$ (and therefore all $\theta'$ in a sufficiently small neighborhood of $\theta$), the bounded convergence theorem allows us to switch the expectation and derivative:
-    % $$\nabla_\theta \E[\hat\lik_J(\theta)] 
-    % = \E[\nabla_\theta \hat\lik_J(\theta)].$$
-
-    % Take limits on both sides to find that 
-    % $$\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)] 
-    % = \lim_{J \to \infty} \E[\nabla_\theta \hat\lik_J(\theta)].$$
-
-    % We then need to swap the limit and derivative on the left-hand side.
-This requires uniform convergence of $\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]$, i.e.
-that $$\sup_{\theta \in \gN(\phi)} ||  \nabla_\theta \E[\hat\lik_J(\theta)] - \lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]||_{\infty} \to 0 \text{ as } J \to \infty,$$
-    % or that $\forall \epsilon>0$ $\exists N\in \mathbb{N}$ such that $n>N => ||  \nabla_\theta \E[\hat\lik_J(\theta)] - \lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)]||_{\infty} < \epsilon$ $\forall \theta \in \gN(\phi).$ However, it is easier to instead use an indirect argument.
-
-    % We then need to swap the limit and derivative on the left-hand side. This requires uniform convergence.
-
-    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point we wish to evaluate the gradient at.
-The sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$, by Assumption \ref{assump:local-bounded-derivative}.
-The sequence is also uniformly equicontinuous.
-To see this, by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J(\theta, \omega)|_{\theta=\theta'}$ is also bounded by $H$ for almost every $\omega\in \Omega$ and every $\theta' \in \gN(\phi)$, including $\theta=\phi$.
-A set of functions with derivatives bounded by the same constant is uniformly Lipschitz, and therefore uniformly equicontinuous.
-So the sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ is uniformly equicontinuous for almost every $\omega \in \Omega$.
-    Explicitly, for almost every $\omega \in \Omega$ and every $\epsilon>0$, there exists some $\delta(\omega)>0$ such that for every $||\theta - \theta'||_{\infty}<\delta$ and every $J \in \mathbb{N}$ we have that
-    $$||\nabla_\theta \hat\lik_J(\theta, \omega)-\nabla_\theta \hat\lik_J(\theta', \omega)||_\infty < \epsilon.$$
-
-    \kevin{Our previous understanding was that the target is differentiable, and the sequence is differentiable in increasingly small open neighborhoods.
-This may not be true.
-The neighborhoods the sequence is differentiable in should be independent of $J$, as the resampling does not change with $\theta$ as it is dependent on $\phi$.
-Either way, the target is in each of the neighborhoods, and this may be sufficient in itself.}
-
-    Then, by Arzela-Ascoli, there exists a uniformly convergent subsequence.
-We claim that there is only one subsequential limit.
-We know that when as the gradient is bounded by Assumption \ref{assump:local-bounded-derivative} we can treat the gradient as a bounded functional.
-So by Lemma \ref{lem:posterior-proper-weights} the sequence $(\nabla_\theta \hat\lik_J(\theta, \omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$, and there is therefore only one subsequential limit.
-The sequence therefore converges uniformly to its limit $\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega).$ Therefore, with uniform convergence for the derivatives established, we can swap the limit and derivative, and obtain that for almost every $\omega \in \Omega$,
-    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta, \omega).$$
-
-    From Lemma \ref{lem:lik-proper-weight}, we know that
-    $$\hat{\lik}_J(\theta, \omega) \stackrel{a.s.}{\to} \lik(\theta).$$
-
-    We then have that for almost every $\omega \in \Omega$, 
-    $$\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta, \omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta, \omega) = \nabla_\theta \lik(\theta),$$
-    as we wanted. The result then follows by the continuous mapping theorem.
-
-    % Taking expectations on both sides and using the bounded convergence theorem to interchange limits and expectations, we see that
-    % $$\lim_{J \to \infty} \nabla_\theta \E[\hat\lik_J(\theta)] = \E[\lim_{J \to \infty} \nabla_\theta \hat\lik_J(\theta)] = \E[\nabla_\theta \lim_{J \to \infty} \hat\lik_J(\theta)] = \nabla_\theta \lim_{J \to \infty} \E[\hat\lik_J(\theta)].$$
-
-    % It then follows that
-    % $$\nabla_\theta\lik(\theta) 
-    % = \nabla_\theta \lim_{J\to\infty} \E[\hat\lik_J(\theta)] 
-    % = \lim_{J\to\infty} \nabla_\theta \E[\hat\lik_J(\theta)] 
-    % = \lim_{J\to\infty} \E[\nabla_\theta \hat\lik_J(\theta)].$$
-
-    % We then have mean-square consistency for the gradient of the likelihood estimate. The result then follows by the continuous mapping theorem.
-
-    \kevin{Explain that the difference in formula between this and Poyiadjis is due to us taking the derivative w.r.t. the particles in the chain rule. }
-
-    \kevin{Alternative argument: Show gradient is unbiased, after showing the likelihood is unbiased in an open neighborhood of phi.
-That should hold by Chopin's argument for unbiased resampling to nonconstant weights.
-Gradient unbiased if can swap expectation and derivative, I think with two continuous derivatives and boundedness, might need BCT/DCT.
-No need to swap limit and derivative in this case, or worrying about uniform limits.}
-
-\end{proof}
-
-\kevin{Show bias is order N, variance is order N when alpha below 1. That implies optimal choice of alpha is not 1. }
-
-\section{Analysis of MOP-$\alpha$ Error and Variance}
-
-\subsection{Analysis of MOP-$0$ Variance}
-
-We made Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} to leverage the results from \cite{karjalainen2023} on the forgetting of the particle filter.
+We made Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} to leverage the results from \cite{karjalainen23} on the forgetting of the particle filter.
 This is required to show error bounds on the gradient estimates that we provide -- namely that the error of the MOP-$1$ estimator that corresponds to that of \cite{poyiadjis11} is $O(N^2)$, and that the variance of MOP-$\alpha$ for any $\alpha<1$ is $O(N)$.
@@
 
-To illustrate the proof strategy, we first analyze the special case when $\alpha=0.$ We then have:
+Now, to illustrate the proof strategy for the general case in an easier context, we first analyze the special case when $\alpha=0.$
+
+\begin{proof}
+When $\alpha=0$, we have:
 \begin{equation}
@@
 
-\kevin{How does the dobrushin contraction coefficient relate to the alpha-mixing coefficients?
+%\kevin{How does the dobrushin contraction coefficient relate to the alpha-mixing coefficients?
 If you have a bound on the mixing time $b$, then the alpha mixing coefficients should just be $\alpha(t) \leq e^{-t/b}$.}
 
-By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strongly mixing, with $\alpha$-mixing coefficients $\alpha(k) \leq (1-\epsilon)^{\floor{k/(c\log(J))}}$.
+By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strong mixing, with $\alpha$-mixing coefficients $\alpha(k) \leq (1-\epsilon)^{\floor{k/(c\log(J))}}$.
 Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
@@
 
+\end{proof}
+
 \paragraph{Remark:} Note that the factor of $N$ that pops up here is due to the use of the unnormalized gradient.
 If one divides the gradient estimate by $\sqrt{N}$ before usage, the variance does not depend on the horizon.
-If one divides by $N$, the error is $O(1/\sqrt{NJ})$!
-
-\kevin{https://math.stackexchange.com/questions/3274996/is-a-transformed-mixing-process-is-still-mixing}
-
-%$$\text{Cov}(x_{n,1:J}^P, x_{m, 1:J}^P) \leq (1-\epsilon)^{\floor{\frac{|n-m|}{c\log(J)}}}$$
-
-\kevin{Ed: Claim 11.4.2, Chopin, is the key to reasoning about this.
-Establish geometric mixing, that $\text{Cov}(\nabla_\theta\hat\ell_n, \nabla_\theta\hat\ell_m) \leq \rho^{|n-m|}V_n$, where $V_n$ is asymptotically stable (the latter only holds for $\alpha<1$, as only then do we get asymptotic stability, otherwise it increases linearly in $n$).
-Geometrically decreasing covariances and stable variances over time are the same thing.
-Then, we can bound the total variance by $\text{Var}(\sum_n \nabla_\theta \hat\ell_n(\theta)) = \sum_{m,n} \text{Cov}(\nabla_\theta\hat\ell_n(\theta), \nabla_\theta\hat\ell_m(\theta)) = NV_n + 2(N-1)\rho V_n + 2(N-3)\rho^2V_n + ...$}
-
-\subsection{Analysis of MOP-$\alpha$ Variance}
-
-Next, we analyze the case for $\alpha \in (0,1)$.
-The proof strategy is as follows.
-Instead of analyzing the variance of the MOP-$\alpha$ gradient estimate proper, we will analyze the variance of a modified estimator where we truncate the weights at $k$ timesteps, so this estimator only looks at most $k$ timesteps back while accumulating the importance weights.
-This enables us to establish strong mixing for the truncated estimator, in order to get a bound on the variance that is $O(NJ^{-1})$.
-We emphasize that this estimator is only a theoretical construct -- we do not actually use this estimator.
-
-Coupled with a bound ensuring that the $k$-truncated estimator provides an estimate that is close to that of MOP-$\alpha$ proper, we get a bound on the variance of MOP-$\alpha$ that comprises of the variance of the $k$-truncated estimator plus the error, for any $k \leq N$.
-It then holds that the final variance bound on MOP-$\alpha$ is given by the minimum over all $k$, and is never larger than $O(1+NJ^{-1})$.
-
-Using the log-derivative trick and that $w_{n,j}^{F,\theta} = w_{n,j}^{F,\theta,k} = 1$ when $\theta=\phi$,
-\begin{align*}
-    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,k}\right)\right\rVert_{\infty}\\
-    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta}}{{\sum_{j=1}^J w_{n,j}^{F,\theta}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,k}}}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,k}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta}}{w_{n,j}^{F,\theta}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,k}}{w_{n,j}^{F,\theta,k}}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,k}\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta}\right)-\log\left(w_{n,j}^{F,\theta,k}\right)\right)\right\rVert_{\infty}\\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=n-k}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^{n-k}\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}\\
-    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
-    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}G'(\theta)\\
-    &\leq G'(\theta)\frac{\alpha^k-\alpha^n}{1-\alpha},
-\end{align*}
-where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
-
-We can then bound $||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||$:
-\begin{align*}
-    &\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_\infty\\
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
-    &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\\
-    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)
-    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\Bigg)\right\rVert_\infty \\
-    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)\right)\right\lVert_{\infty}
-    +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\right)\right\rVert_\infty,
-\end{align*}
-and each of these two terms is bounded by $\frac{\alpha^k}{1-\alpha}NG'(\theta)$. So we now know that
-$$\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_\infty \leq  \frac{2\alpha^k}{1-\alpha}NG'(\theta),$$
-which is our desired error bound.
-
-Now, we bound the variance of $\hat s_k(\theta)$. 
-Recall that we defined
-$$\hat s_k(\theta) = \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) = \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right).$$
-
-We can then decompose this expression with the log-derivative trick while noting that $w_{n,j}^{F,\theta, k} = 1$ whenever $\theta=\phi$, to see that
-\begin{align*}
-    \hat s_k(\theta) 
-    &= \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right) \\
-    &= \sum_{n=1}^N \left(\frac{\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}}{\sum_{j=1}^J w_{n,j}^{F,\theta, k}}-\frac{\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}}{\sum_{j=1}^J w_{n-1,j}^{A,F,\theta, k}}\right) \\
-    &= \sum_{n=1}^N \left(\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}\right)\\
-    &= \frac{1}{J}\sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right).
-\end{align*}
-
-It now follows that we need to bound the variance at a single timestep, namely
-$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right).$$
-
-We use the log-derivative trick yet again to find, noting that $w_{n,j}^{F,\theta, k}=1$ when $\theta=\phi$, that
-\begin{align*}
-    \nabla_\theta w_{n,j}^{F,\theta, k} = \frac{\nabla_\theta w_{n,j}^{F,\theta, k}}{w_{n,j}^{F,\theta, k}} = \nabla_\theta \log(w_{n,j}^{F,\theta, k}) = \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) = \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right).
-\end{align*}
-
-Then,
-\begin{align*}
-    \text{Var}\left(\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right) 
-    &= \text{Var}\left(\nabla_\theta \left(\sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right)- \sum_{i=n-k-1}^{n-1} \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right)\right)\right) \\
-    &= \text{Var} \left(\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1}\log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right).
-\end{align*}
-
-Note that $$\hat s_{n,k}(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)$$ is a function bounded by $C\frac{G'(\theta)(1-\alpha^k)}{1-\alpha} =: CG'(\theta)r$ in each coordinate for some constant $C$ (by Assumption \ref{assump:bounded-measurement}) that depends only on $k$ timesteps, and not $n$.
-We can therefore invoke Lemma 2 of \cite{karjalainen2023} to bound the $L_2$ error of each of the $k$ (from time $n-k$ to time $n$) functionals from its expectation under the posterior by $CrG'(\theta)J^{-1/2}k$.
-That is, we have that
-$$\E\left[||\hat s_{n,k}(\theta) - \E_{\pi}\hat s_{n,k}(\theta)||_\infty^2\right]^{1/2} \leq \frac{Cr^2G'(\theta) k}{\sqrt{J}}.$$
-
-By the bias-variance decomposition, this in turn bounds the variance at a single timestep by $C^2r^2G'(\theta)^2J^{-1}k^2$. Concretely, we have that
-$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \lesssim \frac{r^2k^2G'(\theta)^2}{J}.$$
-
-It now remains to bound the variance of $\hat s_k(\theta)$ by considering the covariance of each of the $N$ terms that comprise it. We decompose
-\begin{align*}
-    \text{Var}(\nabla_\theta \hat s_k(\theta)) &= \text{Var}\left(\sum_{n=1}^N \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \\
-    &= \sum_{n=1}^N\text{Var}\left(\hat s_{n,k}(\theta)\right)+ 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \\
-    &\lesssim \frac{r^2k^2G'(\theta)^2N}{J} + 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right).
-\end{align*}
-
-Similarly to the proof of the MOP-$0$ case, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
-We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
-$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
-i.e.
-the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
-By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strongly mixing, with $\alpha$-mixing coefficients $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c\log(J))}}$.
-Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
-
-We now derive the $\alpha$-mixing coefficients of $(\hat s_{n,k}(\theta))_{n=1}^N$.
-Observe that $\hat s_{n,k}$ is strong mixing at lag $k+1$, as all weights beyond $k$ timesteps are truncated.
-We therefore have that the mixing time of $\hat s_k(\theta)$ is $O(1+k+\log(J))$, and that the $\alpha$-mixing coefficients for $(\hat s_{n,k}(\theta))_{n=1}^N$ are given by $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c(1+k+\log(J)))}}$.
-
-Therefore, by Davydov's inequality and Lemma 2 of \cite{karjalainen2023}, and noting that $\hat s_{n,k}(\theta) \leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, by a similar argument to the MOP-0 case, we find that
-
-$$\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J}.$$
-
-Concretely, noting that $\nabla_\theta\hat s_{n,k}(\theta)\leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and WLOG assuming $\E[(\nabla_\theta\hat\hat s_{m,k}(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat\hat s_{n,k}(\theta))^4]^{1/4}$, we apply Davydov's inequality to see that
-\begin{align*}
-    \text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) 
-    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat s_{m,k}(\theta))^4]^{1/4}\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/4}\\
-    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2}.
-\end{align*}
-
-To bound this, we use the fact that $$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4] = \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4]+\E[(\nabla_\theta\hat s_{n,k}(\theta))^2]^2$$
-alongside Lemma 2 of \cite{karjalainen2023}, which shows that
-$$\E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4] \lesssim \frac{r^4G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^2] \lesssim \frac{r^2G'(\theta)^2}{J},$$
-to find that 
-$$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{r^4G'(\theta)^4}{J^2}+\left(\frac{r^2G'(\theta)^2}{J}\right)^2} = \frac{r^2G'(\theta)^2}{J},$$
-and conclude that 
-$$\text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J}.$$
-
-Putting it all together, we see that
-\begin{align*}
-    \text{Var}(\nabla_\theta \hat s_{k}(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat s_{n,k}(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat s_{m,k}(\theta),\nabla_\theta\hat s_{n,k}(\theta))\\
-    &\leq \frac{Nr^2k^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\frac{r^2G'(\theta)^2}{J} \\
-    &\leq \frac{r^2k^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\right) \\
-    &\lesssim \frac{k^2r^2G'(\theta)^2N}{J}.
-\end{align*}
-
-We will now use this result to bound $\text{Var}(\nabla_\theta \hat s_k(\theta)$ proper. For random variables $X,Y$ where $|X-Y|$ is bounded almost surely by some $M$, we have that 
-$$X-\E X - 2M \leq Y - \E Y \leq X - \E X + 2M$$
-so,
-$$(Y-\E Y)^2 \leq (X-\E X - 2M)^2 + (X-\E X +2M)^2.$$
-Taking expectations, we get
-$$\text{Var}(Y) \leq \text{Var}(X) + 8M^2.$$
-
-It follows from this result and the result we proved earlier, namely that
-$$||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||_\infty \leq \frac{2\alpha^k}{1-\alpha}NG'(\theta),$$
-that the variance of MOP-$\alpha$ proper is bounded by, for any $k \leq N$,
-that the variance of MOP-$\alpha$ proper is bounded by
-$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \frac{k^2r^2G'(\theta)^2N}{J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2.$$
-
-As the above holds for any $k \leq N$,
-$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \min_{k\leq N} \left(\frac{k^2G'(\theta)^2N}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right).$$
-
-% To interpret this result, we can choose $k$ such that
-% \begin{align*}
-%     &\frac{16\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \leq 1,\\
-%     &\alpha^{2k} \leq \frac{(1-\alpha)^2}{16N^2G'(\theta)^2},\\
-%     &k \geq \frac{\log\left(16N^2G'(\theta)^2\right)-2\log(1-\alpha)}{2\log\left(\frac{1}{\alpha}\right)}.
-% \end{align*}
-
-% To interpret this result, if there exists some $k$ such that $$N \geq k \geq \frac{\log\left(16NG'(\theta)\right)-\log(1-\alpha)}{\log\left(\frac{1}{\alpha}\right)},$$
-% we can choose such a $k$, as well as $k=N$, to see that
-% \begin{align*}
-%     \text{Var}(\nabla_\theta \hat\ell(\theta)) 
-%     &\lesssim \min_{k\leq N}\left(\frac{c_d^2k^2r^2G'(\theta)^2N}{J} + \frac{\alpha^{2k}}{(1-\alpha)^2}N^2G'(\theta)^2 \right)\\
-%     &\lesssim \min\left\{1+\frac{c_d^2G'(\theta)^2N}{J}, \frac{c_d^2G'(\theta)^2r^2N^3}{J}\right\} \\
-%     &\leq \min\left\{1+\frac{c_d^2G'(\theta)^2N}{J}, \frac{c_d^2G(\theta)^2r^2N}{J}\right\},
-% \end{align*}
-% as $NG'(\theta)\leq G(\theta)$ as a consequence of Assumption \ref{assump:local-bounded-derivative}.
-
-\subsection{Analysis of MOP-$\alpha$ Bias}
-
-Using the log-derivative trick and that $w_{n,j}^{F,\theta,1} = w_{n,j}^{F,\theta,\alpha} = 1$ when $\theta=\phi$,
-\begin{align*}
-    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,1}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)\right\rVert_{\infty}\\
-    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,1}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,1}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}}}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,1}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,\alpha}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,1}}{w_{n,j}^{F,\theta,1}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,\alpha}}{w_{n,j}^{F,\theta,\alpha}}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,1}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha}\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta,1}\right)-\log\left(w_{n,j}^{F,\theta,\alpha}\right)\right)\right\rVert_{\infty}\\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=1}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}.
-\end{align*}
-
-Now we establish the bound:
-\begin{align*}
-     \nabla_\theta \left(\sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)
-     &\leq \sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
-    &\leq \sum_{i=1}^n\left(1-\alpha^{(n-i)}\right)G'(\theta)\\
-    &\leq G'(\theta)\left(\frac{n-\alpha n + \alpha^n - 1}{1-\alpha}\right)
-\end{align*}
-where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
-
-By a similar decomposition to that of the error of the $k$-truncated gradient to the MOP-$\alpha$ gradient estimate,
-\begin{align*}
-    &\left\lVert\nabla_\theta\hat\ell^1(\theta) - \nabla_\theta\hat\ell^\alpha(\theta) \right\rVert_\infty\\
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
-    &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\\
-    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)
-    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right)\Bigg)\right\rVert_\infty \\
-    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)\right|
-    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right)\right|\Bigg)\right\rVert_\infty.
-\end{align*}
-
-From our arguments above, we know that 
-$$\left\lVert\nabla_\theta\hat\ell^1(\theta) - \nabla_\theta\hat\ell^\alpha(\theta) \right\rVert_\infty \leq 2NG'(\theta)\left(\frac{N-\alpha N + \alpha^N - 1}{1-\alpha}\right).$$
-
-So the bias of MOP-$\alpha$ is the bias of MOP-$1$ arising from the logarithm in the gradient of the log-likelihood plus no more than $$2NG'(\theta)\left(\frac{N-\alpha N + \alpha^N - 1}{1-\alpha}\right),$$
-where we note that the bias vanishes as $\alpha$ goes to $1$ as
-$$\lim_{\alpha\to1}\left(\frac{N-\alpha N + \alpha^N - 1}{1-\alpha}\right) = 0.$$
-
-\kevin{This subsection is not useless I think! It shows that the bias vanishes continuously as $\alpha$ goes to $1$.}
-
-\subsection{An Improved Bound by Mixing}
-
-We decompose the bias $||\nabla_\theta\ell(\theta) - \E\nabla_\theta \hat\ell^\alpha(\theta)||_\infty \leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty$ into
-$$\E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty \leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_\infty + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty.$$
-
-The broad idea is that the first term can be controlled by a mixing argument, the second term garners $O(Nk)$ error, and the third term is $O\left(\frac{N}{1-\alpha}\right)$.
-It is unclear whether the bias can be reduced further -- the gradient is a sum of $N$ terms, and is therefore $O(N)$ itself.
-
-The first term is the problem term.
-We will first bound the ground truth of conditional scores from particle approximation conditional on correct filtering distribution at time $n-k$, and then bound that from the particle approximation resulting from an arbitrary filtering distribution at time $n-k$.
-That is, where $\hat\pi_n$ is the particle approximation at time $n$ of the posterior $\pi_n$ and we write $\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)$ for the truncated MOP-$1$ conditional score estimate given the correct filtering distribution at time $n-k$,
-
-\begin{align*}
-    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_\infty \\
-    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty \\
-    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_\infty + \sum_{n=1}^N \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty.
-\end{align*}
-
-The first term is a particle approximation dependent on $k$ timesteps, so by Lemma 2 of \cite{karjalainen2023}, this is bounded by
-$$\E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_\infty \leq \frac{CG'(\theta)k}{\sqrt{J}}.$$
-
-The second term amounts to bounding the difference between functionals of two different particle measures that mix under the same Markov kernel.
-Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
-We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
-$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
-i.e.
-the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
-
-Then, we can bound 
-$\E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty$ by
-$$\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}} = \beta_{TV}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
-implying that
-\begin{align*}
-    \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty 
-    &\lesssim \sup_{\mu, \nu} \sup_{||\psi||_\infty \leq 1/2} G'(\theta)\left|(\mu \textbf{M}_{n,n+k})(\psi)-(\nu \textbf{M}_{n,n+k})(\psi)\right| \\
-    &\leq \sup_{\mu, \nu} G'(\theta)\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}} \\
-    &\leq G'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}} ||\hat\pi_{n-k} - \pi_{n-k}||_{\text{TV}} \\
-    &\leq G'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}} \sup_{||\psi||_{\infty} \leq 1/2} |\hat\pi_{n-k}(\psi) - \pi_{n-k}(\psi)| \\
-    &\lesssim G'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}}.
-\end{align*}
-
-Therefore, we have that
-\begin{align*}
-    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_\infty \\
-    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_\infty + \sum_{n=1}^N \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty\\
-    &\lesssim N\frac{CG'(\theta)k}{\sqrt{J}} + NG'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}}.
-\end{align*}
-
-Now that the first term $\E||\nabla_\theta\ell(\theta)$ is taken care of, it remains to bound $$\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty \text{ and } \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty.$$
-
-We see, by a similar argument to the variance bound, that
-\begin{align*}
-    &\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty\\
-    &= \E\left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
-    &= \E\Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)- \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\\
-    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
-    &= \E\left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)
-    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\Bigg)\right\rVert_\infty \\
-    &\leq \sum_{n=1}^N \E\left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right|
-    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\right|\Bigg)\right\rVert_\infty.
-\end{align*}
-
-Each of these terms can be bounded using the log-derivative trick and that $w_{n,j}^{F,\theta,1,k} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
-\begin{align*}
-    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_{\infty}\\
-    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}}}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,1,k}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,\alpha,k}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,1,k}}{w_{n,j}^{F,\theta,1,k}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,\alpha,k}}{w_{n,j}^{F,\theta,\alpha,k}}\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,1,k}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta,1,k}\right)-\log\left(w_{n,j}^{F,\theta,\alpha,k}\right)\right)\right\rVert_{\infty}\\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=n-k}^n\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=n-k}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_{\infty} \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=n-k}^{n}(1-\alpha^{(n-i)})\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}\\
-    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
-    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})G'(\theta)\\
-    &\leq G'(\theta)\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right),
-\end{align*}
-where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
-
-So, 
-\begin{align*}
-    &\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty\\
-    &\leq \sum_{n=1}^N \E\left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right|
-    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\right|\Bigg)\right\rVert_\infty \\
-    &\leq 2NG'(\theta)\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) 
-    \leq 2NG'(\theta)k.
-\end{align*}
-
-Now, the third term, $\E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty$, fortunately has already been dealt with. From the variance argument, we know that
-$$||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty \leq 2\frac{\alpha^k}{1-\alpha}NG'(\theta),$$
-and taking expectations on both sides yields our desired result.
-
-Therefore, our decomposition yields the bias bound
-\begin{align*}
-    &||\nabla_\theta\ell(\theta) - \E\nabla_\theta \hat\ell^\alpha(\theta)||_\infty \leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty \\
-    &\leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_\infty + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty \\
-    &\lesssim \min_{k \leq N} \left(N\frac{CG'(\theta)k}{\sqrt{J}} + NG'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}} + 2NG'(\theta)\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) + 2\frac{\alpha^k}{1-\alpha}NG'(\theta)\right) \\
-    &\lesssim \min_{k \leq N} NG'(\theta)\left(\frac{k}{\sqrt{J}}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\frac{\alpha^k  + \alpha^{k+1} - \alpha}{1-\alpha}\right).
-\end{align*}
-
-\subsection{Bounding the MSE of MOP-$\alpha$}
-
-\kevin{First term sums up $k$ terms, $O(Nk^2)$ error, second term bounded by $k$ applications of the Markov kernel, so just apply forgetting of the particle filter result. }
-
-We can obtain an $O(N)$ MSE bound from an argument that is similar to the bias argument. Simply observe that 
-$$||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 =  \left\lVert\sum_{n=1}^N \nabla_\theta \hat\ell_n^\alpha(\theta) - \sum_{n=1}^N \nabla_\theta \ell_n(\theta)\right\rVert_2^2 \leq \sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_2^2,$$
-
-and decompose
-$$\sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_2^2 \leq \sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 + \sum_{n=1}^N||\nabla_\theta\hat s_{n,k}^1(\theta) - \nabla_\theta\hat s_{n,k}^\alpha(\theta)||_2^2 + \sum_{n=1}^N||\nabla_\theta\hat s_{n,k}^\alpha(\theta) -  \nabla_\theta\hat\ell_n^\alpha(\theta)||_2^2.$$
-
-As before, we bound the first term, $\sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2$, by decomposing it into two terms, 
-$$\sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 \leq \sum_{n=1}^N ||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 + \sum_{n=1}^N ||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2.$$
-
-The first term is a particle approximation dependent on $k$ timesteps, so by Lemma 2 of \cite{karjalainen2023}, this is bounded by
-$$\E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 \leq \frac{CpG'(\theta)^2k^2}{J}.$$
-
-The second term amounts to bounding the difference between functionals of two different particle measures that mix under the same Markov kernel.
-Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
-We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
-$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
-i.e.
-the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
-
-Then, we can bound 
-$\E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2$ by
-$$\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}} = \beta_{TV}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
-implying that
-\begin{align*}
-    \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 
-    &\lesssim \sup_{\mu, \nu} \sup_{||\psi||_\infty \leq 1/2} pG'(\theta)^2\left|(\mu \textbf{M}_{n,n+k})(\psi)-(\nu \textbf{M}_{n,n+k})(\psi)\right| \\
-    &\leq \sup_{\mu, \nu} pG'(\theta)^2\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}} \\
-    &\leq pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} ||\hat\pi_{n-k} - \pi_{n-k}||_{\text{TV}} \\
-    &\leq pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} \sup_{||\psi||_{\infty} \leq 1/2} |\hat\pi_{n-k}(\psi) - \pi_{n-k}(\psi)| \\
-    &\lesssim pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}}.
-\end{align*}
-
-Therefore, we have that
-\begin{align*}
-    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_2^2 \\
-    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 + \sum_{n=1}^N \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2\\
-    &\lesssim N\frac{CpG'(\theta)^2k^2}{J} + NpG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}}.
-\end{align*}
-
-Now that the first term $\E||\nabla_\theta\ell(\theta)||_2^2$ is taken care of, it remains to bound $$\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2 \text{ and } \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_2^2.$$
-
-We see, by a similar argument to the bias bound, that
-\begin{align*}
-    &\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2\\
-    &\leq \sum_{n=1}^N \E\left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right|
-    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\right|\Bigg)\right\rVert_2^2.
-\end{align*}
-
-Each of these terms can be bounded using the log-derivative trick and that $w_{n,j}^{F,\theta,1,k} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
-\begin{align*}
-    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_{2}^2\\
-    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2\\
-    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})pG'(\theta)^2\\
-    &\leq pG'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right),
-\end{align*}
-where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
-
-So, 
-\begin{align*}
-    \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty
-    \leq 2NpG'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) 
-    \leq 2NpG'(\theta)^2k.
-\end{align*}
-
-Now, we address the third term, $\E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty$.
-From the variance argument, we know that we can bound $||\nabla_\theta\hat\ell^\alpha(\theta) - \hat s_k^\alpha(\theta)||$:
-\begin{align*}
-    &\left\lVert\nabla_\theta\hat\ell^\alpha(\theta) - \hat s_k^\alpha(\theta) \right\rVert_2^2\\
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_2^2\\
-    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right)\right\lVert_2^2
-    +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta, \alpha}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha, k}\right)\right)\right\rVert_2^2.
-\end{align*}
-
-Using the log-derivative trick and that $w_{n,j}^{F,\theta,\alpha} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
-\begin{align*}
-    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_2^2\\
-    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2\\
-    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}pG'(\theta)^2\\
-    &\leq pG'(\theta)^2\frac{\alpha^k-\alpha^n}{1-\alpha},
-\end{align*}
-where the second-last line follows from Assumption \ref{assump:bounded-measurement}. Putting it together and taking expectations on both sides, we obtain
-$$\E\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_2^2 \leq  \frac{2\alpha^k}{1-\alpha}NpG'(\theta)^2,$$
-which is our desired error bound.
-
-Therefore, our decomposition yields the MSE bound
-\begin{align*}
-    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 \\
-    &\leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_2^2 + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2 + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_2^2 \\
-    &\lesssim \min_{k \leq N} \left(N\frac{CpG'(\theta)^2k^2}{J} + NpG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} + 2NpG'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) + 2\frac{\alpha^k}{1-\alpha}NpG'(\theta)^2\right) \\
-    &\lesssim \min_{k \leq N} NpG'(\theta)^2\left(\frac{k^2}{J}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\frac{\alpha^k  + \alpha^{k+1} - \alpha}{1-\alpha}\right).
-\end{align*}
-
-\kevin{Trying to bound MOP-(1,k) from MOP-alpha yields exactly the same result as above! Seems like I can't get rid of the k in the third term within the parentheses.}
-% \begin{align*}
-%     &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)\right\rVert_{2}^2\\
-%     &\leq \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=n-k}^n\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=1}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_2^2 \\
-%     &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=n-k}^n\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=n-k}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=1}^{n-k}{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_2^2 \\
-%     &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2 + \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2 \\
-%     &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})pG'(\theta)^2 + \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}pG'(\theta)^2\\
-%     &\leq pG'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)+\alpha^k}{1-\alpha}\right),
-% \end{align*}
-
-\section{Finite-Particle Analysis of Gradient Methods with Algorithm \ref{alg:ifad}}
-\label{app:ifad}
-
-The analysis in this section roughly follows the analysis in \cite{mahoney16}, except with the caveat that none of the matrix concentration bounds they use apply here as the particles are dependent.
-We instead use the concentration inequality from \cite{delmoral2011ci} to bound the gradient and Hessian estimates.
-In this section, we fix $\omega \in \Omega$ only within each filtering iteration, evaluate Algorithm \ref{alg:mop} at $\theta=\phi$, and analyze Algorithm \ref{alg:ifad} post-iterated filtering.
-
-The convergence analysis in Theorem \ref{thm:convergence} is limited to the case where $-\ell$ is $\gamma$-strongly convex.
-Though it is true that in a neighborhood of the optimum local asymptotic normality holds and the log-likelihood is strongly convex in this neighborhood, in practice likelihood surfaces for POMPs are often highly nonconvex globally.
-The convergence to an optimum, local or global, must therefore be sensitive to initialization.
-
-\subsection{Bounding the Gradient}
-\kevin{Add a few words to explain that the gradient of the log-likelihood w.r.t.
-parameters is a functional of the particle measure under boundedness assumptions and fixed parameters, appeal to authority and say doucet did it too.
-}
-
-\kevin{Gradient linear map, bound conditional likelihoods at time n.
-Alternatively, convert del moral and doucet expectation bound to high-probability bound.
-https://arxiv.org/pdf/1905.11546.pdf for trick.
-Alternatively, concatenate previous states.
-}
-
-\begin{lem}[Concentration of Measure for Gradient Estimate]
-    \label{lemma:grad_bound}
-    Consider the gradient estimate obtained by MOP-1, which we know by Lemma \ref{lem:mop-grad-consistency} is consistent for the score, where $\theta = \phi$.
-For $||\nabla_\theta \hat{\loglik}(\theta) - \nabla_\theta \loglik(\theta)||_2$ to be bounded by $\epsilon$ with probability $1-\delta$, we require
-    \begin{align}
-    J > \max\left\{2G(\theta)\frac{r_N\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{2p}{\delta}\right)\right)\right), 8G(\theta)^2\beta_N^2\frac{p\log(2p/\delta)}{\epsilon^2}\right\},
-    \end{align}
-    where $NG'(\theta) \leq G(\theta)$ are defined in Assumptions \ref{assump:bounded-measurement} and \ref{assump:local-bounded-derivative}, $h(t) = \frac{1}{2}(t - \log(1+t))$, and $\beta_N$ and $r_N$ are two additional finite model-specific constants that do not depend on $J$, but do depend on $N$ and $p$, as defined in \cite{delmoral2011ci}.
-
-    Equivalently, with probability at least $1-\delta$, it holds that
-    \begin{align}
-        ||\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).
-    \end{align}
-\end{lem}
-
-\paragraph{Remark:} According to \cite{delmoral2011ci}, under some regularity conditions, $r_N$ and $\beta_N$ are linear in the trajectory length $N$.
-This corresponds to the finding by \cite{poyiadjis11} that the variance of the estimate is at least quadratic in the trajectory length, and their remark that the result of \cite{delMoral03} establishes that the $L_p$ error is bounded by $O(N^2J^{-1/2})$ (equivalently, the variance is bounded by $O(N^4J^{-1})$) after accounting for the sum over timesteps.
-The MOP-$1$ variance upper bound is therefore in fact $O(N^4)$, in contrast to the MOP-$\alpha$, where $\alpha<1$, upper bound of $O(N)$.
-
-Either way, usually in practice one considers the estimate normalized through dividing it by $N$, in which case we only need
-$$J > \max\left\{2G'(\theta)\frac{r_N\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{2p}{\delta}\right)\right)\right), 8G'(\theta)^2\beta_N^2\frac{p\log(2p/\delta)}{\epsilon^2}\right\}.$$
-Equivalently,
-$$||N^{-1}(\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta))||_2 \leq G'(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).$$
-
-It is well-known that bounds like these hide an exponential dependence in the dimension of the state space $d$.
-We take care to keep track of the terms depending on $d$ when they arise in front of us, but are unable to do so for the terms that arise from the analysis of \cite{delmoral2011ci}.
-
-\kevin{This is slightly alarming, in that the estimate really has $O(N^4)$ variance after accounting for summing over timesteps.
-In contrast, MOP-0 has only $O(N)$ variance after summing over timesteps.
-}
-
-\begin{proof}
-
-We will seek to use the concentration inequality of \cite{delmoral2011ci} to bound the deviation of the gradient estimate from the gradient of the negative log-likelihood in the sup norm with a union bound.
-Fix some $\theta \in \gN(\phi)$.
-
-Recall that as $w_{N, j}^{F, \theta}=1$ when $\theta=\phi$,
-$$
-\nabla_\theta \hat{\ell}(\theta):=\frac{1}{J} \sum_{j=1}^J \nabla_\theta w_{N, j}^{F, \theta}=\frac{1}{J} \sum_{j=1}^J\sum_{n=1}^N  w_{N, k_j}^{P, \theta} \nabla_\theta \log g_{n,k_j}^{A,\theta} = \frac{1}{J} \sum_{j=1}^J\sum_{n=1}^N \nabla_\theta \log g_{n,k_j}^{A,\theta}.
-$$
-
-Define $\varphi_n^i(x_{n,j}^{F,\theta}) := \frac{\partial}{\partial\theta_i} \log g_{n,k_j}^{A,\theta}$, which is a functional of the filtering particles $x_{n,j}^{F,\theta} = x_{n,k_j}^{P,\theta}$.
-These are bounded measurable functionals bounded by $G'(\theta)$ by Assumption \ref{assump:bounded-measurement}.
-Therefore, these have bounded oscillation, satisfying the requirement that $\text{osc} \left(\frac{\partial}{\partial\theta_i} \varphi_i(x_{n,j}^{P,\theta}) \right) \leq G'(\theta)$.
-\cite{delmoral2011ci} in fact assume $\text{osc}(f) \leq 1$, so we simply scale their bound accordingly.
-
-\kevin{Put some justification/make the setup of delmoral2011ci concrete here. Concentration of gradient estimate under particle measure to limit under posterior measure}
-
-Apply the Hoeffding-type concentration inequality from Del Moral and Rio \cite{delmoral2011ci} and a union bound over each $\varphi_n^i(x_{n,j}^{F,\theta})$, totaling $N$ timesteps and $p$ parameters, to find that
-
-\begin{align}
-    \max_{n=1,...,N} \left\lVert\frac{1}{J}\sum_{j=1}^J\nabla_\theta \log g_{n,k_j}^{A,\theta} - \nabla_\theta \ell_n(\theta) \right\rVert_{\infty} \leq G'(\theta)\left(\frac{r_N}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_N \right)
-\end{align}
-
-with probability at least $1-2Np\exp(-t)$.
-Note that the concentration inequality only considers the error from the expectation under the filtering distribution -- but we invoke the consistency of MOP-$1$ shown in Lemma \ref{lem:mop-grad-consistency} to establish that the expectation under the filtering distribution is in fact the score.
-It therefore holds that with the same probability, that when summing over $N$, as $NG'(\theta) \leq G(\theta)$,
-
-\begin{align}
-    \left\lVert\frac{1}{J}\sum_{j=1}^J\sum_{n=1}^N\nabla_\theta \log g_{n,k_j}^{A,\theta} - \nabla_\theta \ell(\theta) \right\rVert_{\infty} 
-    &\leq G'(\theta)N\left(\frac{r_N}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_N \right)\\
-    &\leq G(\theta)\left(\frac{r_N}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_N \right)
-\end{align}
-
-We split the $\delta$ failure probability among these $2Np$ terms, to find $\delta\leq2Np\exp(-t) => t\leq\log(2Np/\delta)$.
-Here, $h(t) = \frac{1}{2}(t - \log(1+t))$.
-The two additional model-specific parameters are $\beta_t$ and $r_t$, which do not depend on $J$.
-
-The analogous bound for the 2-norm follows from scaling the right-hand side by $\sqrt{p}$, to require
-
-\begin{align}
-    ||\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right)
-\end{align}
-
-and we therefore need 
-\begin{align}
-    J > \max\left\{2G(\theta)\frac{r_N\sqrt{p}}{\epsilon}\left(1+h^{-1}\left(\log\left(\frac{2p}{\delta}\right)\right)\right), 8G(\theta)^2\beta_N^2\frac{p\log(2p/\delta)}{\epsilon^2}\right\}.
-\end{align}
-
-\end{proof}
-
-\subsection{Bounding Hessian Estimates}
-
-Should one choose to use a second-order method involving a particle Hessian estimate, we provide a guarantee for its positive-definiteness below.
-
-\begin{lem}[Minimum Eigenvalue Bound for Hessian Estimate]
-    \label{lemma:hess_bound}
-    Assume that the Hessian of the negative log-likelihood $H=\sum_{j=1}^J \E H_j$ has a minimum eigenvalue $0<\gamma<1$, and that $\E \lambda_{\min} (H_j) = \gamma' > 0$.
-
-    We then require 
-    \begin{equation}
-        J > \max\left\{\frac{2r_t(1+h^{-1}(t)) + 2c}{\gamma'}, \frac{2(2t\beta_t^2+c)^2}{\gamma'^2}\right\} \geq  \frac{r_t(1+h^{-1}(t))}{\gamma'} + \sqrt{2tJ}\beta_t/\gamma' + c/\gamma'
-    \end{equation}
-
-    for $\hat{H}(\theta)$ to be invertible and positive definite with minimum eigenvalue greater than or equal to $c \in (0, \sum_{j=1}^J \E\lambda_{\min}(H_j))$ and with probability at least $1-\exp(-t)$.
-\end{lem}
-\begin{proof}
-Write $\hat{H}(\theta) = \hat{H} = \sum_{j=1}^J H_j$ for the estimate of the negative of the Hessian, where $H_j$ is an element of the outer sum over the $J$ particles.
-
- As the negative log-likelihood is convex, we want to bound the minimum eigenvalue of $\hat{H}(\theta)$ from below with high probability, so that all the eigenvalues of $\hat{H}(\theta)$ are positive with high probability.
-This ensures that the estimated Hessian is invertible and positive-definite.
-
-From \cite{boyd2004convex}, we know that the minimum eigenvalue of a symmetric matrix is concave. Therefore, it suffices to show that the first inequality in the below expression
-
-\begin{equation}
-    0 < \sum_{j=1}^J \lambda_{\min} (H_j) \leq  \lambda_{\min}\left(\sum_{j=1}^J H_j\right) = \lambda_{\min} (\hat{H})
-\end{equation}
-
-holds with high probability.
-
-Apply the particle Hoeffding concentration inequality from \cite{delmoral2011ci} to find that
-
-\begin{align}
-    \frac{1}{J}\sum_{j=1}^J \lambda_{\min}(H_j) - \E_{\tilde{\pi}_t}\lambda_{\min}(H_j) &= \frac{1}{J}\sum_{j=1}^J \lambda_{\min}(H_j) - \gamma' \geq -\frac{r_t}{J}(1+h^{-1}(t)) - \sqrt{\frac{2t}{J}}\beta_t \\
-    \sum_{j=1}^J \lambda_{\min}(H_j) &\geq -r_t(1+h^{-1}(t)) - \sqrt{2tJ}\beta_t + J\gamma'
-\end{align}
-
-with probability at least $1-\exp(-t)$. Here, $h(t) = \frac{1}{2}(t - \log(1+t))$. The two additional model-specific parameters are $\beta_t$ and $r_t$, which do not depend on $J$.
-
-We additionally require, for $c \in (0, \sum_{j=1}^J \E\lambda_{\min}(H_j))$,
-\begin{align}
-    \sum_{j=1}^J \lambda_{\min}(H_j) \geq -r_t(1+h^{-1}(t)) - \sqrt{2tJ}\beta_t + J\gamma' \geq c \\
-    J\gamma' \geq c + r_t(1+h^{-1}(t)) + \sqrt{2tJ}\beta_t 
-\end{align}
-
-We therefore require $$J > \max\left\{\frac{2r_t(1+h^{-1}(t)) + 2c}{\gamma'}, \frac{2(2t\beta_t^2+c)^2}{\gamma'^2}\right\} \geq  \frac{r_t(1+h^{-1}(t))}{\gamma'} + \sqrt{2tJ}\beta_t/\gamma' + c/\gamma'$$
-
-for $\hat{H}(\theta)$ to be invertible and positive definite with minimum eigenvalue greater than or equal to $c$ with probability at least $1-\exp(-t)$.
-
-\end{proof}
-
-\subsection{Convergence Analysis of Theorem \ref{thm:convergence}}
-
-\begin{proof}
-
-In this analysis, we largely follow the proof of Theorem 6 in \cite{mahoney16}.
-
-Define $\theta_\eta = \theta_n + \eta p_n$, where $p_n=-(H(\theta_n))^{-1}g(\theta_n)$.
-As in Roosta-Khorasani and Mahoney \cite{mahoney16}, we want to show there is some iteration-independent $\tilde{\eta}>0$ such that the Armijo condition
-\begin{equation}
-    f(\theta_n+\eta p_n) \leq f(\theta_n) + \eta\beta p_n^Tg(\theta_n)
-\end{equation}
-holds for any $0< \eta < \tilde{\eta}$ and some $\beta \in (0,1)$.
-
-By an argument found in the beginning of the proof of Theorem 6 in \cite{mahoney16}, we have that choosing $J$ such that $||\nabla_\theta\hat{\ell}(\theta_n) - \nabla_\theta \ell(\theta_n)|| \leq \epsilon$ and $\lambda_{\min}(H(\theta_n)) \geq c$ with probability $(1-\delta)^2$ for each $n$ yields
-
-\begin{align}
-    f(\theta_\eta)-f(\theta_n) \leq \eta p_n^Tg(\theta_n) + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2
-\end{align}
-
-Now, for some $c>0$, with probability $1-\delta$
-\begin{equation}
-    p_n^Tg(\theta_n) = -p_n^TH(\theta_n)p_n \geq -c||p_n||^2
-\end{equation}
-and we can obtain a decrease in the objective.
-
-Substituting this into the previous expression,
-\begin{align}
-    f(\theta_\eta)-f(\theta_n) \leq -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2
-\end{align}
-
-and the Armijo condition then becomes
-\begin{align}
-    -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2 &\leq \eta \beta p_n^Tg(\theta_n) = - \eta \beta p_n^TH(\theta_n)p_n \\
-    \epsilon||p_n|| + \eta \Gamma ||p_n||^2 / 2 &\leq (1- \beta) p_n^TH(\theta_n)p_n \\
-    \epsilon + \eta \Gamma ||p_n|| / 2 &\leq c(1- \beta) ||p_n||
-\end{align}
-
-which holds and guarantees an iteration-independent lower bound if 
-\begin{equation}
-    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_n)|| \leq \frac{c(1-\beta)}{2}||p_n||
-\end{equation}
-
-Now, first note that
-\begin{equation}
-||g(\theta_n)|| - ||\nabla_\theta f(\theta_n)|| \leq ||g(\theta_n) - \nabla_\theta f(\theta_n)|| \leq \epsilon => ||\nabla_\theta f(\theta_n)|| \geq ||g(\theta_n)|| - \epsilon
-\end{equation} and
-\begin{equation}
-||\nabla_\theta f(\theta_n)||-||g(\theta_n)|| \leq ||\nabla_\theta f(\theta_n)-g(\theta_n)|| \leq \epsilon => ||g(\theta_n)|| \geq ||\nabla_\theta f(\theta_n)|| - \epsilon
-\end{equation}
-
-There are now two cases. If the algorithm terminates and $||g(\theta_n)|| \leq \sigma \epsilon$, we can derive
-
-\begin{equation}
-    ||\nabla_\theta f(\theta_n)|| \leq ||g(\theta_n)|| + \epsilon = \sigma\epsilon+\epsilon = (\sigma+1)\epsilon
-\end{equation}
-
-If the algorithm does not terminate, then $||g(\theta_n)|| > \sigma \epsilon$. Notice that
-
-\begin{align}
-    \epsilon \geq ||g(\theta_n) - \nabla_\theta f(\theta_n)|| &\geq ||g(\theta_n)|| - ||\nabla_\theta f(\theta_n)|| \\
-    ||\nabla_\theta f(\theta_n)|| + \epsilon &\geq ||g(\theta_n)|| \geq \sigma \epsilon \\
-    ||\nabla_\theta f(\theta_n)|| &\geq \sigma \epsilon - \epsilon = (\sigma - 1)\epsilon \\
-    \frac{||\nabla_\theta f(\theta_n)||}{\sigma-1} &\geq \epsilon
-\end{align}
-
-and now 
-\begin{align}
-    ||\nabla_\theta f(\theta_n)|| - \epsilon
-    &\geq  ||\nabla_\theta f(\theta_n)|| - \frac{||\nabla_\theta f(\theta_n)||}{\sigma-1} = \left(1-\frac{1}{\sigma-1}\right)||\nabla_\theta f(\theta_n|| = \frac{\sigma-2}{\sigma-1}||\nabla_\theta f(\theta_n)|| \geq \frac{2}{3}||\nabla_\theta f(\theta_n)||
-\end{align}
-
-From $||A^{-1}|| = 1/\sigma_{\min}(A)$,
-\begin{align}
-    p_n^TH(\theta_n)p_n &= (-(H(\theta_n))^{-1}g(\theta_n))^TH(\theta_n)(-(H(\theta_n))^{-1}g(\theta_n)) \\
-    &= g(\theta_n)^T(H(\theta_n))^{-1}g(\theta_n) \\
-    &\geq \frac{1}{c}||g(\theta_n)||^2 \\
-    &\geq \frac{1}{c}(||\nabla_\theta f(\theta_n)|| - \epsilon)^2 \\
-    &\geq \frac{4}{9c}||\nabla_\theta f(\theta_n)||^2
-\end{align}
-
-From the assumption that $f$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 -\ell \preceq \Gamma I$, by an implication of $\gamma$-strong convexity we have
-\begin{align}
-    f(\theta_n) - f(\theta^*) \leq \frac{1}{2\gamma}||\nabla_\theta f(\theta_n)||^2
-\end{align}
-and we put together:
-\begin{equation}
-    f(\theta_n) - f(\theta^*) \leq \frac{1}{2\gamma}||\nabla_\theta f(\theta_n)||^2 \leq \frac{9c}{4}\frac{1}{2\gamma}p_n^TH(\theta_n)p_n
-\end{equation}
-\begin{equation}
-    \frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \leq \frac{4}{9c}||\nabla_\theta f(\theta_n)||^2 \leq p_n^TH(\theta_n)p_n
-\end{equation}
-\begin{equation}
-    f(\theta_n) - f(\theta^*) \leq \frac{9c}{8\gamma}p_n^TH(\theta_n)p_n
-\end{equation}
-
-\begin{equation}
-    -\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \geq -\frac{4}{9c}||\nabla_\theta f(\theta_n)||^2 \geq -p_n^TH(\theta_n)p_n
-\end{equation}
-
-But from earlier, as the Armijo condition is fulfilled with our choice of $\eta$ and $\epsilon$,
-\begin{align}
-    f(\theta_{n+1})-f(\theta_n) &\leq -\eta p_n^TH(\theta_n)p_n + \epsilon\eta||p_n|| + \eta^2 \Gamma ||p_n||^2 / 2 \\
-    &\leq -\eta\beta p_n^TH(\theta_n)p_n \\
-    &\leq -\eta\beta\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) 
-\end{align}
-
-and therefore
-\begin{align}
-    f(\theta_{n+1}) - f(\theta^*) 
-    &= f(\theta_{n+1})-f(\theta_n)+f(\theta_n)- f(\theta^*) \\
-    &\leq f(\theta_n)- f(\theta^*) -\eta\beta\frac{8\gamma}{9c}(f(\theta_n) - f(\theta^*)) \\
-    &= (1-\eta\beta\frac{8\gamma}{9c})(f(\theta_n) - f(\theta^*)) 
-\end{align}
-
-\end{proof}
-
-\section{Miscellanous Technical Lemmas}
-
-\begin{lem}
-    \label{lem:dobrushin-implies-alpha-mixing}
-    Setting $X$ to be the particle collection at time $m$, and $Y$ at time $n$, we have that the alpha mixing coefficients,
-    $$\int | f_{XY} - f_X f_Y | < \alpha,$$
-    are bounded by the Dobrushin coefficient
-    $$\int | f_{Y|X}(y|x_1) - f_{Y|X}(y|x_2) |dy < \alpha$$ for all $x_1$, $x_2$.
-\end{lem}
-
-\begin{proof}
-    We rewrite the alpha-mixing assertion as $$\int { \int |f_Y|X(y|x) - f_Y(y)| dy } f_X dx.$$
-
-    We claim that the Dobrushin coefficient implies $$\int | f_Y|X(y|x_1) - f_Y(y) |dy < \alpha$$ for all $x_1$. This is shown as follows:
-    \begin{align*}
-        \int | f_{Y|X}(y|x) - f_Y(y) |dy 
-        &= \int \left|  \int [f_{Y|X}(y|x_1) - f_{Y|X}(y|x)] f_X(x)dx \right|dy \\
-        &< \int   \int  | f_{Y|X}(y|x_1) - f_{Y|X}(y|x)] | dy f_X(x)dx \\
-        &< \int \alpha f_X(x)dx \\
-        &=\alpha
-    \end{align*}
-
-    We then have the desired result.
-\end{proof}
-
-\end{document}
+If one divides by $N$, the error is $O(1/\sqrt{NJ})$.

### 2024-06-13 paper/appendix/consistency.tex
update from overleaf

We now provide a result showing that under sufficient regularity conditions, one can interchange the order of differentiation and expectation of functionals of off-parameter resampled particle estimates, showing that if an off-parameter resampled particle estimate is consistent for some estimand, its derivative is consistent for the derivative of the estimand as well.

One may wonder why this may be necessary, given that we have already shown strong consistency for measurable bounded functionals in Theorem \ref{thm:off-parameter-targeting}. If we require the gradient to be bounded as well, then the gradient is also a bounded functional. The answer is that the strong consistency is for expectations under the filtering distribution, so Theorem \ref{thm:off-parameter-targeting} only establishes strong consistency of the gradient of the estimate to its expectation under the filtering distribution, $\nabla_\theta \eta_n^J(h_\theta)  \stackrel{a.s.}{\to} \eta_n(\nabla_\theta h_\theta)$, which is not a-priori the gradient of the estimand $\nabla_\theta \eta_n(h_\theta)$. We require an interchange of the derivative and expectation, which we show is possible when the particle estimates have two continuous uniformly bounded derivatives over all $J$ below.

\begin{thm}[Off-Parameter Particle Filters Yield Strongly Consistent Estimates of Derivatives of Functionals]
    Let $h_\theta : \gX \to \R$ be a measurable bounded functional of particles, where $\eta_n^J(h_\theta)$ has two continuous derivatives uniformly bounded over all $J$ by $H^*$ for almost every $\omega\in\Omega$ and $\theta \in \Theta$. If it holds that $\eta_n^J(h_\theta) \stackrel{a.s.}{\to} \eta(h_\theta) = h^*_\theta$, then we also have that $\nabla_\theta \eta_n^J(h_\theta)  \stackrel{a.s.}{\to} \eta_n(\nabla_\theta h_\theta) = \nabla_\theta \eta_n(h_\theta) = \nabla_\theta h^*_\theta$. 
\end{thm}
\begin{proof}
    Fix $\omega \in \Omega$. The sequence $(\nabla_\theta \eta_n^J(h_\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$, by assumption. The sequence is also uniformly equicontinuous. To see this, by assumption, the second derivative of $\eta_n^J(h_\theta)(\omega)|_{\theta=\theta'}$ is also uniformly bounded over all $J$ for almost every $\omega\in \Omega$ and every $\theta' \in \Theta$. A set of functions with derivatives bounded by the same constant is uniformly Lipschitz, and therefore uniformly equicontinuous. So the sequence $(\eta_n^J(h_\theta)\omega))_{J \in \mathbb{N}}$ is uniformly equicontinuous over $\theta$ for almost every $\omega \in \Omega$. 
    Explicitly, for almost every $\omega \in \Omega$ and every $\epsilon>0$, there exists some $\delta(\omega)>0$ such that for every $||\theta - \theta'||_{\infty}<\delta$ and every $J \in \mathbb{N}$ we have that
    $$||\eta_n^J(h_\theta)(\omega)-\eta_n^J(h_{\theta'})(\omega)||_\infty < \epsilon.$$

    Then, by Arzela-Ascoli, there exists a uniformly convergent subsequence. We claim that there is only one subsequential limit. When the gradient is bounded, we can treat the gradient as a bounded functional. So by Theorem \ref{thm:off-parameter-targeting} the sequence $(\eta_n^J(h_\theta)(\omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$, and there is therefore only one subsequential limit. The sequence therefore converges uniformly to its limit $\lim_{J \to \infty} \eta_n^J(h_\theta)(\omega).$ Therefore, with uniform convergence for the derivatives established, we can swap the limit and derivative, and obtain that for almost every $\omega \in \Omega$, 
    $$\lim_{J \to \infty} \eta_n^J(h_\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \eta_n^J(h_\theta)(\omega).$$

    Again from Theorem \ref{thm:off-parameter-targeting}, we know that
    $$\eta_n^J(h_\theta)(\omega) \to \eta_n(h_\theta) = h^*_\theta$$ for almost every $\omega\in\Omega$.

    We then have that for almost every $\omega \in \Omega$, 
    $$\lim_{J \to \infty} \eta_n^J(h_\theta)(\omega) = \nabla_\theta \eta_n(h_\theta) = \nabla_\theta h^*_\theta,$$
    as we wanted.

\end{proof}

The proof of Theorem \ref{thm:mop-grad-consistency} is now merely a corollary, where we apply $\eta_n^J (h_\theta) = \hat\lik^1_J(\theta)$, $\eta_n(h_\theta) = \lik(\theta) = h_\theta^*$, and then use the continuous mapping theorem.

% For completeness, we now provide a longer version of the proof of Theorem \ref{thm:mop-grad-consistency}.
% \begin{proof}
%     Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point we wish to evaluate the gradient at. The sequence $(\nabla_\theta \hat\lik^1_J(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$, by Assumption \ref{assump:local-bounded-derivative}. The sequence is also uniformly equicontinuous. To see this, by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik^1_J(\theta)(\omega)|_{\theta=\theta'}$ is also bounded by $H^*$ for almost every $\omega\in \Omega$ and every $\theta' \in \Theta$. A set of functions with derivatives bounded by the same constant is uniformly Lipschitz, and therefore uniformly equicontinuous. So the sequence $(\nabla_\theta \hat\lik^1_J(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly equicontinuous for almost every $\omega \in \Omega$. 
%     Explicitly, for almost every $\omega \in \Omega$ and every $\epsilon>0$, there exists some $\delta(\omega)>0$ such that for every $||\theta - \theta'||_{\infty}<\delta$ and every $J \in \mathbb{N}$ we have that
%     $$||\nabla_\theta \hat\lik^1_J(\theta)(\omega)-\nabla_\theta \hat\lik^1_J(\theta')(\omega)||_\infty < \epsilon.$$

%     Then, by Arzela-Ascoli, there exists a uniformly convergent subsequence. We claim that there is only one subsequential limit. We know that when as the gradient is bounded by Assumption \ref{assump:local-bounded-derivative} we can treat the gradient as a bounded functional. So by Lemma \ref{lem:posterior-proper-weights} the sequence $(\nabla_\theta \hat\lik^1_J(\theta)(\omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$, and there is therefore only one subsequential limit. The sequence therefore converges uniformly to its limit $\lim_{J \to \infty} \nabla_\theta \hat\lik^1_J(\theta)(\omega).$ Therefore, with uniform convergence for the derivatives established, we can swap the limit and derivative, and obtain that for almost every $\omega \in \Omega$, 
%     $$\lim_{J \to \infty} \nabla_\theta \hat\lik^1_J(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik^1_J(\theta)(\omega).$$

%     From Lemma \ref{lem:lik-proper-weight}, we know that
%     $$\hat{\lik}^1_J(\theta)(\omega) \to \lik(\theta)$$ for almost every $\omega\in\Omega$.

%     We then have that for almost every $\omega \in \Omega$, 
%     $$\lim_{J \to \infty} \nabla_\theta \hat\lik^1_J(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik^1_J(\theta)(\omega) = \nabla_\theta \lik(\theta),$$
%     as we wanted. The result then follows by the continuous mapping theorem.

% \end{proof}

### 2024-06-13 paper/appendix/convergence.tex
update from overleaf

@@
     where $NG'(\theta) \leq G(\theta)$ are defined in Assumptions \ref{assump:bounded-measurement} and \ref{assump:local-bounded-derivative}, $h(t) = \frac{1}{2}(t - \log(1+t))$, and $\beta_N$ and $r_N$ are two additional finite model-specific constants that do not depend on $J$, but do depend on $N$ and $p$, as defined in \cite{delMoral11}.
-
 Equivalently, with probability at least $1-\delta$, it holds that
@@
 Fix $\theta = \phi$.
-
 From the decomposition in the proof of Lemma \ref{lem:mop-1-formula}, as $w_{N, j}^{F, \theta}=1$ when $\theta=\phi$, we have that
@@
 $$
-
 Define $\varphi_n^i(x_{n,j}^{F,\theta}) := \frac{\partial}{\partial\theta_i} \log g_{n,k_j}^{A,\theta}$, which is a functional of the filtering particles $x_{n,j}^{F,\theta} = x_{n,k_j}^{P,\theta}$.
@@
 
-Apply the Hoeffding-type concentration inequality from Del Moral and Rio \cite{delMoral11} and a union bound over each $\varphi_n^i(x_{n,j}^{F,\theta})$, totaling $N$ timesteps and $p$ parameters, to find that
-
+Now we apply the Hoeffding-type concentration inequality from Del Moral and Rio \cite{delMoral11} and a union bound over each $\varphi_n^i(x_{n,j}^{F,\theta})$, totaling $N$ timesteps and $p$ parameters, to find that
 \begin{align}
@@
 \end{align}
-
 with probability at least $1-2Np\exp(-t)$.
@@
 It therefore holds that with the same probability, that when summing over $N$, as $NG'(\theta) \leq G(\theta)$,
-
 \begin{align}
@@
     &\leq G'(\theta)N\left(\frac{r_N}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_N \right)\\
-    &\leq G(\theta)\left(\frac{r_N}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_N \right)
+    &\leq G(\theta)\left(\frac{r_N}{J}(1+h^{-1}(t)) + \sqrt{\frac{2t}{J}}\beta_N \right).
 \end{align}
-
-We split the $\delta$ failure probability among these $2Np$ terms, to find $\delta\leq2Np\exp(-t) => t\leq\log(2Np/\delta)$, where $h(t) = \frac{1}{2}(t - \log(1+t))$.
+We split the $\delta$ failure probability among these $2Np$ terms, to find $\delta\leq2Np\exp(-t)$, and therefore, $t\leq\log(2Np/\delta)$, where $h(t) = \frac{1}{2}(t - \log(1+t))$. 
 The two additional model-specific parameters are $\beta_t$ and $r_t$, which do not depend on $J$. 
-
 The analogous bound for the 2-norm follows from scaling the right-hand side by $\sqrt{p}$, to require 
-
 \begin{align}
-    ||\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right)
+    ||\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).
 \end{align}
-
-and we therefore need 
+We therefore need 
 \begin{align}
@@
     Assume that the Hessian of the negative log-likelihood $H=\sum_{j=1}^J \E H_j$ has a minimum eigenvalue $0<\gamma<1$, and that $\E \lambda_{\min} (H_j) = \gamma' > 0$. 
-
-    We then require 
+    If 
     \begin{equation}
@@
     \end{equation}    
-
-    for $\hat{H}(\theta)$ to be invertible and positive definite with minimum eigenvalue greater than or equal to $c \in (0, \sum_{j=1}^J \E\lambda_{\min}(H_j))$, with probability at least $1-\exp(-t)$.
+    then $\hat{H}(\theta)$ is invertible and positive definite with minimum eigenvalue greater than or equal to $c \in (0, \sum_{j=1}^J \E\lambda_{\min}(H_j))$, with probability at least $1-\exp(-t)$.
 \end{lem}
@@
 It is known that the minimum eigenvalue of a symmetric matrix is concave. Therefore, it suffices to show that the first inequality in the below expression
-
 \begin{equation}
@@
 \end{equation}
-
 holds with high probability.
-
-Apply the particle Hoeffding concentration inequality from \cite{delMoral11} to find that
-
+We apply the particle Hoeffding concentration inequality from \cite{delMoral11} to find that  
 \begin{align}
     \frac{1}{J}\sum_{j=1}^J \lambda_{\min}(H_j) - \E_{\tilde{\pi}_t}\lambda_{\min}(H_j) &= \frac{1}{J}\sum_{j=1}^J \lambda_{\min}(H_j) - \gamma' \geq -\frac{r_t}{J}(1+h^{-1}(t)) - \sqrt{\frac{2t}{J}}\beta_t \\
-    \sum_{j=1}^J \lambda_{\min}(H_j) &\geq -r_t(1+h^{-1}(t)) - \sqrt{2tJ}\beta_t + J\gamma'
+    \sum_{j=1}^J \lambda_{\min}(H_j) &\geq -r_t(1+h^{-1}(t)) - \sqrt{2tJ}\beta_t + J\gamma',
 \end{align}
-
-with probability at least $1-\exp(-t)$. Here, $h(t) = \frac{1}{2}(t - \log(1+t))$. The two additional model-specific parameters are $\beta_t$ and $r_t$, which do not depend on $J$.
+with probability at least $1-\exp(-t)$. Here, $h(t) = \frac{1}{2}(t - \log(1+t))$. 
+The two additional model-specific parameters are $\beta_t$ and $r_t$, which do not depend on $J$.
 
@@
 \end{align}
-
-We therefore require $$J > \max\left\{\frac{2r_t(1+h^{-1}(t)) + 2c}{\gamma'}, \frac{2(2t\beta_t^2+c)^2}{\gamma'^2}\right\} \geq  \frac{r_t(1+h^{-1}(t))}{\gamma'} + \sqrt{2tJ}\beta_t/\gamma' + c/\gamma'$$
-
+We therefore require \ed{WE SAY ``REQUIRE'' AT VARIOUS POINTS, BUT THAT SUGGESTS NECESSITY WHEREAS I SUPPOSE WE MEAN SUFFICIENCY. I EDITED ONE OR TWO OF THESE.}
+$$J > \max\left\{\frac{2r_t(1+h^{-1}(t)) + 2c}{\gamma'}, \frac{2(2t\beta_t^2+c)^2}{\gamma'^2}\right\} \geq  \frac{r_t(1+h^{-1}(t))}{\gamma'} + \sqrt{2tJ}\beta_t/\gamma' + c/\gamma'$$
 for $\hat{H}(\theta)$ to be invertible and positive definite with minimum eigenvalue greater than or equal to $c$ with probability at least $1-\exp(-t)$.
@@
 \begin{proof}
-
 In this analysis, we largely follow the proof of Theorem 6 in \cite{mahoney16}.
-
 Define $\theta_\eta = \theta_m + \eta p_m$, where $p_m=-(H(\theta_m))^{-1}g(\theta_m)$. 
@@
 holds for any $0< \eta < \tilde{\eta}$ and some $\beta \in (0,1)$.
-
 By an argument found in the beginning of the proof of Theorem 6 in \cite{mahoney16}, we have that choosing $J$ such that $||\nabla_\theta\hat{\ell}(\theta_m) - \nabla_\theta \ell(\theta_m)|| \leq \epsilon$ and $\lambda_{\min}(H(\theta_m)) \geq c>0$ for each $m$, yields
-
 \begin{align}
@@
 \end{align}
-with probability $1-\delta/2$. From now on, we assume that we are on the success event of this high-probability statement.
-
-Now, it then holds that
+with probability $1-\delta/2$. 
+From now on, we assume that we are on the success event of this high-probability statement. 
+Consequently, we have
 \begin{equation}
@@
 and we can obtain a decrease in the objective. 
-
 Substituting this into the previous expression,
@@
 \end{align}
-
-and the Armijo condition then becomes
+the Armijo condition becomes
 \begin{align}
@@
     \epsilon||p_m|| + \eta \Gamma ||p_m||^2 / 2 &\leq (1- \beta) p_m^TH(\theta_m)p_m \\
-    \epsilon + \eta \Gamma ||p_m|| / 2 &\leq c(1- \beta) ||p_m||,
+    \epsilon + \eta \Gamma ||p_m|| / 2 &\leq c(1- \beta) ||p_m||.
 \end{align}
-
-which holds and guarantees an iteration-independent lower bound if 
+This holds and guarantees an iteration-independent lower bound if 
 \begin{equation}
@@
 which is given by our choice of $\eta$.
-
 Now, first note that
@@
 ||g(\theta_m)|| - ||\nabla_\theta f(\theta_m)|| \leq ||g(\theta_m) - \nabla_\theta f(\theta_m)|| \leq \epsilon => ||\nabla_\theta f(\theta_m)|| \geq ||g(\theta_m)|| - \epsilon
-\end{equation} and
+\end{equation} 
+and
 \begin{equation}
@@
 \end{equation}
-
-There are now two cases. If the algorithm terminates and $||g(\theta_m)|| \leq \sigma \epsilon$, we can derive
-
+There are now two cases. 
+If the algorithm terminates and $||g(\theta_m)|| \leq \sigma \epsilon$, we can derive 
 \begin{equation}
@@
 \end{equation}
-
-If the algorithm does not terminate, then $||g(\theta_m)|| > \sigma \epsilon$. Notice that
-
-\begin{align}
-    \epsilon \geq ||g(\theta_m) - \nabla_\theta f(\theta_m)|| &\geq ||g(\theta_m)|| - ||\nabla_\theta f(\theta_m)|| \\
-    ||\nabla_\theta f(\theta_m)|| + \epsilon &\geq ||g(\theta_m)|| \geq \sigma \epsilon \\
-    ||\nabla_\theta f(\theta_m)|| &\geq \sigma \epsilon - \epsilon = (\sigma - 1)\epsilon \\
-    \frac{||\nabla_\theta f(\theta_m)||}{\sigma-1} &\geq \epsilon,
-\end{align}
-
+If the algorithm does not terminate, then $||g(\theta_m)|| > \sigma \epsilon$. 
+Notice that 
+\begin{eqnarray}
+    \epsilon \geq ||g(\theta_m) - \nabla_\theta f(\theta_m)|| &\geq& ||g(\theta_m)|| - ||\nabla_\theta f(\theta_m)|| 
+    \\
+    ||\nabla_\theta f(\theta_m)|| + \epsilon &\geq& ||g(\theta_m)|| \geq \sigma \epsilon 
+    \\
+    ||\nabla_\theta f(\theta_m)|| &\geq& \sigma \epsilon - \epsilon = (\sigma - 1)\epsilon 
+    \\
+    \frac{||\nabla_\theta f(\theta_m)||}{\sigma-1} &\geq& \epsilon,
+\end{eqnarray}
 and now 
-\begin{align}
+\begin{eqnarray}
     ||\nabla_\theta f(\theta_m)|| - \epsilon
-    &\geq  ||\nabla_\theta f(\theta_m)|| - \frac{||\nabla_\theta f(\theta_m)||}{\sigma-1} = \left(1-\frac{1}{\sigma-1}\right)||\nabla_\theta f(\theta_m|| = \frac{\sigma-2}{\sigma-1}||\nabla_\theta f(\theta_m)|| \geq \frac{2}{3}||\nabla_\theta f(\theta_m)||.
-\end{align}
-
-From $||A^{-1}|| = 1/\sigma_{\min}(A)$,
+    &\geq&  ||\nabla_\theta f(\theta_m)|| - \frac{||\nabla_\theta f(\theta_m)||}{\sigma-1} 
+    \\
+    &=& \left(1-\frac{1}{\sigma-1}\right)||\nabla_\theta f(\theta_m|| 
+    \\
+    &=& \frac{\sigma-2}{\sigma-1}||\nabla_\theta f(\theta_m)|| 
+    \\
+    &\geq& \frac{2}{3}||\nabla_\theta f(\theta_m)||.
+\end{eqnarray}
+Since $||A^{-1}|| = 1/\sigma_{\min}(A)$,
 \begin{align}
@@
 \end{align}
-
 From the assumption that $f$ is $\gamma$-strongly convex, $\gamma I \preceq \nabla_\theta^2 -\ell \preceq \Gamma I$, by an implication of $\gamma$-strong convexity we have
@@
 \end{equation}
-
-But from earlier, as the Armijo condition is fulfilled with our choice of $\eta$ and $\epsilon$,
+From earlier, as the Armijo condition is fulfilled with our choice of $\eta$ and $\epsilon$,
 \begin{align}
@@
 \end{align}
-
-and therefore
+Therefore,
 \begin{align}

### 2024-06-13 paper/appendix/feynman.tex
update from overleaf

--- 2024-06-10 paper/appendix/feynman.tex
@@
 In this section, we introduce the Feynman-Kac convention of \cite{delMoral04} that has since become commonplace \cite{karjalainen23} for the analysis of the particle filter.
-The mathematical formalization and notation introduced here will be adopted
+The mathematical formalization and notation introduced here will be adopted in the remainder of the analysis, in order to prove Theorems \ref{thm:mop-targeting}, \ref{thm:mop-grad-consistency}, and \ref{thm:mop-biasvar}.
+Let $(\eta_n)_{n=1}^N, (\pi_n)_{n=1}^N, (\rho_n)_{n=1}^N$ be sequences of probability measures on the state space $\gX$.
+This is the sequence of prediction distributions $f_{X_{n}|Y_{1:n-1}}$, filtering distributions $f_{X_{n}|Y_{1:n}}$, and posterior distributions $f_{X_{1:n}|Y_{1:n}}$ that we seek to approximate with the particle filter.
+For any measurable bounded functional $h$, we adopt the following functional-analytic notation, borrowed from \cite{delMoral04, chopin20, karjalainen23}.
+We choose our specific choice of notation and definitions to be in line with that of \cite{karjalainen23}.
+
+\begin{itemize}
+\item \textbf{Markov kernels and the process model:} A Markov kernel $M$ with source $\gX_1$ and target $\gX_2$ is a map $M: \gX_1 \times \mathcal{B}(\gX_2) \to [0,1]$ such that for every set $A \in \mathcal{B}(\gX_2)$ and every point $x \in \gX_1$, the map $x \mapsto M(x, A)$ is a measurable function of $x$, and the map $A \mapsto M(x,A)$ is a probability measure on $\gX_2$.
+The quantity $M(x,A)$ can be thought of as the probability of transitioning to the set $A$ given that we are at the point $x$.
+If this yields a density, this then corresponds to the process density $f_{X_{2}|X_1}$ conditional on $x$ and integrated over $A$.
+\item \textbf{Markov kernels and measures:} For any measure $\eta$, any Markov kernel $M$ on $\gX$, any point $x \in \gX$ and any measurable subset $A \subseteq \gX$, let 
+\begin{align*}
+    \eta(h) &= \int h d\eta = \int h(x) \eta(dx), \\(\eta M)(A) &= \int \eta(dx)M(x,A), \\
+    (Mh)(x) &= \int M(x, dy) h(y).
+\end{align*}
+\item \textbf{Compositions of Markov kernels:} The composition of a Markov kernel $M_1$ with another Markov kernel $M_2$ is another Markov kernel, given by 
+$$(M_1M_2)(x, A) = \int M_1(x, dy) M_2(y, A).$$
+\item \textbf{Total variation distance:} The total variation distance between two measures $\mu$ and $\nu$ on $\gX$ is
+$$\|\mu-\nu\|_{\mathrm{TV}}=\sup _{\|h\|_{\infty} \leq 1 / 2}|\mu(\phi)-\nu(\phi)|=\sup _{\operatorname{osc}(h) \leq 1}|\mu(h)-\nu(h)|.$$
+\item \textbf{Dobrushin contraction:} The Dobrushin contraction coefficient $\beta_{\text{TV}}$ of a Markov kernel $M$ is given by
+$$\beta_{\mathrm{TV}}(M)=\sup _{x, y \in \gX}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}.$$
+\item \textbf{Potential functions and the measurement model:} A potential function $G : \gX \to [0,\infty)$ is a non-negative function of an element of the state space $x \in \gX$.
+In our case, this corresponds to the measurement model, and in our previous notation is written as $g_{n,j} = f_{Y_n|X_n}(y_n^*|x_{n,j}^F) = G_n(x_{n,j}^F)$, where in a slight abuse of notation we suppress the dependence on $\theta$ for notational simplicity.
+Note that $G_n(\cdot) = f_{Y_n|X_n}(y_n^*|\;\cdot\;)$ is the conditional density of the observed measurement at time $n$, where we condition on the filtering particle $x_{n,j}^F$ as an element of the state space.
+\item \textbf{Feynman-Kac models:} A Feynman-Kac model on $\gX$ is a tuple $(\pi_0, (M_n)_{n=1}^N, (G_n)_{n=1}^N)$ of an initial probability measure on the state space $\pi_0$, a sequence of transition kernels $(M_n)_{n=1}^N$, and a sequence of potential functions $(G_n)_{n=1}^N$.
+In the notation used in the main text, this corresponds to the starting distribution $f_{X_0}$, the sequence of transition densities $f_{X_{n}|X_{n-1}}$, and the measurement densities $f_{Y_n|X_n}$.
+This induces a set of mappings from the set of probability measures on $\gX$ to itself, $\mathcal{P}(\gX) \to \mathcal{P}(\gX)$, as follows:
+\begin{itemize}
+    \item The update from the prediction to the filtering distributions is given by $$\pi_n(dx) = \Psi_n(\eta_n)(dx) = \frac{G_n(x)\cdot\eta_n(dx)}{\eta_n(G_n)}.$$
+    \item The map from the prediction distribution at timestep $n$ to timestep $n+1$ is given by 
+    $$\Phi_{n+1}(\eta_n) = \Psi_n(\eta_n) M_{n+1}.$$
+    \item The composition of maps between prediction distributions yields the map from the prediction distribution at time $k$ to the prediction distribution at time $n$ where $k \leq n$,
+    $$\Phi_{k,n} = \Phi_n \circ ... \circ \Phi_{k+1}.$$
+\end{itemize}
+\item \textbf{The particle filter:} The particle filter then yields a Monte Carlo approximation to the above Feynman-Kac model, via a sequence of mixture Dirac measures.
+When one resamples at every timestep, the prediction measure at timestep $n$ is then given by
+$$\eta_n^J = \frac{1}{J}\sum_{j=1}^J \delta_{x_{n,j}^P},$$
+and the filtering measure at timestep $n$ is given by
+$$\pi_n^J = \frac{\sum_{j=1}^J g_{n,j} \delta_{x_{n,j}^P}}{\sum_{j=1}^J g_{n,j}} \approx \frac{1}{J} \sum_{j=1}^J \delta_{x_{n,j}^F}.$$
+In a slight abuse of notation, we will identify $x_{n, 1:J}^P \equiv \eta_n^J$, and $x_{n, 1:J}^F \equiv \pi_n^J$. 
+As in \cite{karjalainen23}, one can view this as an inhomogenous Markov process evolving on $\gX^{J}$. The corresponding Markov transition kernel is then 
+$$\textbf{M}_n(x_{n-1, 1:J}^P, \cdot) = \left(\Phi_{n}\left(\eta_n^J\right)\right)^{\otimes J} = \left(\Phi_{n}\left(\frac{1}{J}\sum_{j=1}^J \delta_{x_{n,j}^P}\right)\right)^{\otimes J},$$
+and the composition of Markov kernels on particles from timestep $n$ to timestep $n+k$ is written 
+$$\textbf{M}_{n, n+k} = \textbf{M}_{n+k}\circ ...\circ \textbf{M}_n.$$
+One may wonder why \cite{karjalainen23} require this process to evolve on $\gX^{J}$.
+This is because at every timestep $n$, we in fact draw $X_{n, j}^P | \{X_{n-1, 1:J}^P = x_{n-1, 1:J}^P\} \sim \textbf{M}_n(x_{n-1, 1:J}^P, \cdot) = \eta_0^{\otimes J} \textbf{M}_{0,n}$ for $j=1,...,J$.
+\item \textbf{Forgetting of the particle filter:} 
+The above formalization yields a result from \cite{karjalainen23} that we require for our analysis of the bias, variance, and error of MOP-$\alpha$ -- the forgetting of the particle filter.
+That is, \cite{karjalainen23} show that
+$$\beta_{\mathrm{TV}}\left(\mathbf{M}_{n, n+k}\right) \leq(1-\epsilon)^{\lfloor k /(O(\log J))\rfloor},$$
+for some $\epsilon$ dependent on $\bar{G}, \underbar{G}, \bar{M}, \underbar{M}$ in Assumptions \ref{assump:bounded-measurement} and \ref{assump:bounded-process}.
+As a result, the mixing time of the particle filter is only on the order of $O(\log(J))$ timesteps.
+\end{itemize}
+
+Equipped with the above formalisms and results, we are now in a position to provide guarantees on the performance of MOP-$\alpha$ itself.

### 2024-06-13 paper/appendix/functional.tex
update from overleaf

@@
+Theorem \ref{thm:mop-functional-forms} follows immediately as a consequence of the following results, Lemmas \ref{lem:mop-1-formula} and \ref{lem:mop-0-formula}.
+
 \begin{lem}
@@
     Write $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$.
-Consider the case where we usethe after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$, When $\alpha=1$,
+Consider the case where we use the after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$.
+When $\alpha=1$,
     \begin{equation}
@@
 \begin{proof}
-
     Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. We then have a nice telescoping product property for the after-resampling likelihood estimate:
@@
 where the third equality follows from the choice of $\alpha=1$, and the fourth equality is the resulting telescoping property. 
-
-Then, the log-derivative identity lets us decompose the score estimate as
-$$\nabla_\theta \hat\ell^1(\theta) = \frac{\nabla_\theta \hat\lik^1(\theta)}{\hat\lik^1(\theta)} = \frac{\nabla_\theta\left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi}{\prod_{n=1}^N L_n^\phi} =  \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}.$$
-
-So we see that the derivative of the log-likelihood estimate is
-\begin{equation*}
-    \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta},
-\end{equation*}
-which we decompose as follows.
-
-Observe that as $\alpha=1$,
-
+The log-derivative identity lets us decompose the score estimate as
+\begin{equation}\label{eq:log-derivative-identity}
+\nabla_\theta \hat\ell^1(\theta) = \frac{\nabla_\theta \hat\lik^1(\theta)}{\hat\lik^1(\theta)} = \frac{\nabla_\theta\left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi}{\prod_{n=1}^N L_n^\phi} =  \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}.
+\end{equation}
+From (\ref{eq:log-derivative-identity}), we see that the derivative of the log-likelihood estimate is
+\begin{equation}\label{eq:eq:log-derivative2}
+    \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta}.
+\end{equation}
+We proceed to decompose (\ref{eq:eq:log-derivative2}).
+First, observe that as $\alpha=1$,
 $$w_{n,j}^{P,\theta} = w_{n-1,j}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}},$$
@@
 Note that this quantity is the cumulative product of measurement density ratios over the ancestral trajectory of the $j$-th prediction particle at timestep $n$.
-
-We can then use the log-derivative identity again, yielding the following expression for the gradient of the log-weights as the sum of the log measurement densities over the ancestral trajectory:
+We then use the log-derivative identity again, yielding the following expression for the gradient of the log-weights as the sum of the log measurement densities over the ancestral trajectory:
 $$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,j}^{A,P,\theta} - \log g_{i,j}^{A,P,\phi}\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,j}^{A,P,\theta}.$$
-
-But this is equal to the gradient of the logarithm of the conditional density of the observed measurements given the ancestral trajectory of the $j$-th prediction particle up to timestep $n$:
-$$  \nabla_\theta \sum_{n=1}^N \log g_{n,j}^{A,\theta} = \nabla_\theta \log\left(\prod_{n=1}^N g_{n,j}^{A,P,\theta}\right) =  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,j}^{A, P,\theta}\right)\right) = \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$
-
+This is equal to the gradient of the logarithm of the conditional density of the observed measurements given the ancestral trajectory of the $j$-th prediction particle up to timestep $n$:
+\begin{eqnarray*}  
+\nabla_\theta \sum_{n=1}^N \log g_{n,j}^{A,\theta} &=& \nabla_\theta \log\left(\prod_{n=1}^N g_{n,j}^{A,P,\theta}\right) 
+\\
+&=&  \nabla_\theta \log\left(\prod_{n=1}^N f_{Y_n|X_n}\left(y_n^* | x_{n,j}^{A, P,\theta}\right)\right)
+\\
+&=& \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).
+\end{eqnarray*}
 Multiplying both sides of the expression by $w_{N,j}^{P,\theta} $ yields an expression for the gradient of the weights at timestep $N$:
 $$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{n=1}^N \nabla_\theta \log g_{n,j}^{A,P,\theta} = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$
-
-Substituting the above identity into the log-likelihood decomposition obtained earlier yields
+Substituting the above identity into the log-likelihood decomposition obtained earlier \ed{USE EQUATION NUMBERS TO MAKE THIS ARGUMENT MORE CONCRETE} yields
 \begin{equation*}
-    \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right),
+    \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right).
 \end{equation*}
-and finally, observing that when $\theta=\phi$ we have that $w_{N,j}^{F,\theta}=1$, that 
+Finally, observing that $\theta=\phi$ implies $w_{N,j}^{F,\theta}=1$, we obtain 
 \begin{equation*}
-    \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
+    \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right).
 \end{equation*}
-yielding the gradient estimators of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter. 
+This yields the gradient estimators of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter. 
 \end{proof}
 
-Note that one would believe that the variance blows up the moment $\theta\neq\phi$. This is true, as 
+Note that the variance of the MOP-$\alpha$ log-likelihood estimate scales poorly with $N$ the moment $\theta\neq\phi$. 
+This can be seen by observing that
 \begin{equation*}
-    \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right),
+    \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right).
 \end{equation*}
-and when $\theta\neq\phi$, $w_{N,k_j}^{P,\theta} = O(c^N)$.
-When $\theta=\phi$, this is a special case of the \cite{poyiadjis11} estimator, which has $O(N^4)$ variance by a property of functionals of the particle filter by \cite{delMoral03}.
+When $\theta\neq\phi$, we see that $w_{N,k_j}^{P,\theta} = O(c^N)$. 
+When $\theta=\phi$, this is a special case of the \cite{poyiadjis11} estimator, which has $O(N^4)$ variance by a property of functionals of the particle filter \cite{delMoral03}.
 
@@
     \end{equation}
-    yielding \cite{naesseth18} on the bootstrap filter. 
+    yielding the estimate of \cite{naesseth18} when applied to the bootstrap filter. 
 \end{lem}
@@
 as shorthand for the measurement density ratios. 
-
-One can then observe that when $\alpha=0,$ the likelihood estimate becomes
+Observe that, when $\alpha=0,$, the likelihood estimate becomes
 \begin{equation}
+\nonumber
     \hat{\lik}^0(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
 \end{equation}
-Note that we lose the nice telescoping property observed in the MOP-$1$ case, but this expression still yields something useful. This is because its gradient when $\theta=\phi$ is therefore 
+We lose the nice telescoping property observed in the MOP-$1$ case, but this expression still yields something useful. 
+This is because its gradient when $\theta=\phi$ is therefore 
 \begin{align*}
@@
 \end{proof}
-
-Combining Lemmas \ref{lem:mop-1-formula} and \ref{lem:mop-0-formula} yields the proof of Theorem \ref{thm:mop-functional-forms}.

### 2024-06-13 paper/appendix/targeting.tex
update from overleaf

--- 2024-06-10 paper/appendix/targeting.tex
@@
+Here, we prove a more general result, from which it will be clear that MOP-$\alpha$ targets the filtering distribution.
+We will prove a strong law of large numbers for triangular arrays of particles with ``off-parameter'' resampling, where we resample the particles according to an arbitrary resampling rule that is not necessarily in proportion to the target distribution of interest.
+Using weights that encode the cumulative discrepancy between the resampling distribution and the target distribution (instead of resampling to equal weights as is common) turns out to be a sufficient correction that ensures almost sure convergence.
+
+We now introduce the precise definition of an off-parameter resampled particle filter.
+
+\begin{defn}[Off-Parameter Resampled Particle Filters]
+    \label{defn:off-parameter-filter}
+    An off-parameter resampled particle filter is a Monte Carlo approximation to a Feynman-Kac model $\left(\pi_0,\left(M_n\right)_{n=1}^N,\left(G_n\right)_{n=1}^N\right)$, where we inductively define given some $x_{n-1,j}^F$, $w_{n-1,j}^F$ comprising the filtering measure approximation $\pi_n^J$:
+    \begin{enumerate}
+        \item The prediction particles at timestep $n$, $x_{n,j}^P$, are drawn from $X_{n,j}^P \sim \pi_n^J M_n$, and the prediction weights are given by $w_{n,j}^P = w_{n-1,j}^F$. 
+        \item The prediction measure at timestep $n$ is given by 
+        $$\eta_n^J = \frac{\sum_{j=1}^J w_{n,j}^P \delta_{x_{n,j}^P}}{\sum_{j=1}^J w_{n,j}^P}.$$
+        \item The particles are resampled at every timestep $n$, yielding indices $k_j$ according to some arbitrary probabilities $(p_{n,j})_{j=1}^J$.
+        \item the filtering measure at timestep $n$ is given by
+        $$\pi_n^J = \frac{\sum_{j=1}^J \delta_{x_{n,j}^P }w_{n,j}^P g_{n,j}}{\sum_{j=1}^J w_{n,j}^Pg_{n,j}} \approx \frac{\sum_{j=1}^J \delta_{x_{n,k_j}^P }w_{n,k_j}^P g_{n,k_j}/p_{n,k_j}}{\sum_{j=1}^J w_{n,k_j}^Pg_{n,k_j}/p_{n,k_j}} = \frac{\sum_{j=1}^J w_{n,j}^F \delta_{x_{n,j}^F}}{\sum_{j=1}^J w_{n,j}^F}.$$
+        \item The prediction weights at the next timestep are given by $w_{n+1,j}^P = w_{n,j}^F =  w_{n,k_j}^P g_{n,k_j}/p_{n,k_j}$.
+    \end{enumerate}
+\end{defn}
+
+To prove that the weight correction is sufficient for almost sure convergence, we first introduce what it means for a triangular array of particles to \textit{target} a given target distribution.
+
 \begin{defn}[Targeting]
-    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any measurable and bounded functional $h$,
+    A pair of random vectors $(X, W)$ drawn from some measure $g$ \textbf{targets} another measure $\pi$ if for any measurable and bounded functional $h$,
 \begin{equation}
-    E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
+    E_g\{h(X) \cdot W\}=E_\pi\{h(X)\}.
 \end{equation}  
-    A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if for any measurable and bounded functional $h$,
+    A set of particles \textbf{targeting} $\pi$ is a triangular array of pairs of random vectors $(X^J_j, W^J_j), j=1,2, \ldots,J$ such that for any measurable and bounded functional $h$,
 \begin{equation}
-    \frac{\sum_{j=1}^J h(X^J_j) w^J_j}{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X))
+    \frac{\sum_{j=1}^J h(X^J_j) W^J_j}{\sum_{j=1}^J W^J_j} \stackrel{a.s.}{\to} E_\pi(h(X))
 \end{equation}
@@
 \end{defn}
-\cite{chopin04} asserted without proof that common particle filter algorithms target the filtering distribution, $f_{X_{1:N}|Y_{1:N}}$, in this sense.
+\cite{chopin04} asserted without proof that common particle filter algorithms targets the filtering distribution in this sense, while \cite{chopin20} proved a related result assuming bounded densities.
+We follow a similar approach to \cite{chopin20}, based on showing strong laws of large numbers for triangular arrays, noting that triangular array strong laws do not hold without an additional regularity condition such as boundedness in general.
+
 In order to prove the consistency of our variation on the particle filter, we now present three helper lemmas.
@@
     Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
-Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, proportional to $\pi^J_j J$ times.
+Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample with indices $k_j$ drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, proportional to $\pi^J_j J$ times.
 Write
     \[
-    (Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big),
+    (Y_j^J,V_j^J) = \big(\tilde X^J_{k_j},U^J_{k_j}/\pi^J_{k_j}\big).
     \] 
-    where $a$ is called the ancestor function. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j),j=1,\dots,J\}$ targets $f_X$.
+    If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j),j=1,\dots,J\}$ targets $f_X$.
 \end{lem}
@@
 \begin{proof}
-
     Note that as the $Y_j^J$ are a subsample from $X_j^J$, $h$ can be a function of $Y$ as well as it is one for $X$. We then expand
@@
     $$\frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[f(X)].$$
-
     We want to show
@@
     Because the functional $h$ and importance sampling weights $u_{k_j}^J/\pi_{k_j}^J$ are bounded, we have that $\E[\left(Z_j^J\right)^4] < \infty$.
-We can then follow the argument of \cite{chopin20} from this point on, where
-    {THE Zs ARE ONLY CONDITIONALLY INDEPENDENT, SO WE COULD PUT IN THE CONDITIONING EVENT (I.E., THE Xs AND Us) HERE TO BE MORE ACCURATE}
-    $$\E\left[\left(\sum_j Z_j^J\right)^4\right] 
-    = J\E[(Z_1^J)^4] + 3J(J-1)\left(\E[(Z_j^J)^2]\right) \leq CJ^2,$$ for some $C>0$. By Markov, 
-    $$\mathbb{P}\left(\left|\frac{1}{J} \sum_{j=1}^J Z_j^J\right|>\epsilon\right) 
+We can then follow the argument of \cite{chopin20} from this point on, where noting that the $Z_j^J$ are conditionally independent given the $\tilde{X}_j^J$ and $U_j^J$,
+    $$\E\left[\left(\sum_j Z_j^J\right)^4 \Bigg| (\tilde X_j^J,U_j^J)_{j=1}^J \right] 
+    = J\E[(Z_1^J)^4|(\tilde X_j^J,U_j^J)_{j=1}^J] + 3J(J-1)\left(\E[(Z_j^J)^2|(\tilde X_j^J,U_j^J)_{j=1}^J]\right) \leq CJ^2,$$ for some $C>0$. Taking expectations on both sides yields
+    $$\E\left[\left(\sum_j Z_j^J\right)^4 \right]  \leq CJ^2$$
+    by the tower property. Now by Markov's inequality, 
+    $$\mathbb{P}\left(\left|\frac{1}{J} \sum_{j=1}^J Z_j^J\right|>\epsilon \right) 
     \leq \frac{1}{\epsilon^4J^4 } 
@@
     and as these terms are summable we can apply Borel-Cantelli to conclude that these deviations happen only finitely often for every $\epsilon>0,$ giving us the almost-sure convergence for   
-
     $$ \sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}
     -  \sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0.$$ 
-
     Similarly, we also have that
@@
     - \sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J} \stackrel{a.s.}{\to} 0,$$
-
     and the result is proved.
-
-    \kevin{MAY HAVE TO DEAL WITH CONDITIONING ON THE FILTRATION AT THE LAST TIMESTEP, AND THEN USING THE TOWER PROPERTY. EXPECTATION WILL BE THE SAME REGARDLESS FOR THE FOURTH MOMENT.}
 
@@
 
-\textbf{Remark:} Note that Lemma \ref{lem:change-measure-proper-weights} permits $\pi_{1:J}$ to depend on $\{(X_j,u_j)\}$ as long as the resampling is carried out independently of $\{(X_j,u_j)\}$, conditional on $\pi_{1:J}$.
+\noindent \textbf{Remark:} Note that Lemma \ref{lem:change-measure-proper-weights} permits $\pi_{1:J}$ to depend on $\{(X_j,u_j)\}$ as long as the resampling is carried out independently of $\{(X_j,u_j)\}$, conditional on $\pi_{1:J}$.
 
@@
     $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J,$$
-    by the same fourth moment argument to the above lemma converges almost surely to the limit of its expectation
+    and by the same fourth moment argument to the above lemma, it converges almost surely to the limit of its expectation
     \begin{eqnarray*}        
@@
 \begin{proof}
-
     Again, we want to show that
      $$\frac{\sum_j h(X_j^J) \cdot U_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j U_j^J \cdot f_{Z|X}(z^*|X_j^J)} \stackrel{a.s.}{\to} \E_{f_{X|Z}}[h(X)|z^*].$$
-
     We already have that for any measurable bounded $g$,
@@
     \end{equation}
-
     Consider the following:
@@
     \end{equation}
-
     We will apply Equation \ref{eq:lemma-posterior-hypothesis} to the numerator and the denominator in the ratio above individually. The numerator converges to 
@@
     $$ \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}  \stackrel{a.s.}{\to} \E_{f_X}[f_{Z|X}(z^*|X)] = f_Z(z^*).$$
-
-    But it holds that
+    Now we take advantage of the identities
     $$\frac{\E_{f_X}[h(X)f_{Z|X}(z^*|X)]}{f_Z(z^*)} = \E_{f_X}\left[\frac{h(X)f_{Z|X}(z^*|X)}{f_Z(z^*)}\right] = \E_{f_X}\left[h(X)\frac{f_{X|Z}(X|z^*)}{f_X(X)}\right] =  \E_{f_{X|Z}}[h(X)|z^*],$$
-    and we have the desired result.
+    to give the desired result.
 
@@
 
-\begin{lem}[Change of Particle Marginal Measure]
-    \label{lem:change-marginal-proper-weights}
-    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
+% \begin{lem}[Change of Particle Marginal Measure]
+%     \label{lem:change-marginal-proper-weights}
+%     Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
 Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
@@
 Then, if the $U_j^J$ and $V_j^J := U_j^Jf_{Y|X}(\tilde{Z}_j^J|\tilde{X}_{j}^J)/f_{Z|X}(\tilde{Z}_j^J|\tilde{X}_{j}^J)$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}$ targets $f_Z$, and $\{(\tilde Z_j^J,V_j^J)\}$ targets $f_Y$.
-\end{lem}
+% \end{lem}
+% \begin{proof}
+%     From Lemma \ref{lem:marginal-proper-weights}, for any measurable bounded $g$, 
+%      $$\frac{\sum_j g(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_Z}[g(Z)] = \E_{f_X}\big[\E_{f_{Z|X}}[g(Z) | X]\big],$$
+%      and by assumption, for any measurable bounded $g$, 
+%     $$\frac{\sum_j g(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[g(X)].$$
+
+%     We in fact want to show that 
+%     $$\frac{\sum_j h(\tilde Z_j^J) \, V_j^J}{\sum_j V_j^J} \stackrel{a.s.}{\to} \E_{f_Y}[h(Y)] = \E_{f_X}\left[\E_{f_{Z|X}}\left[h(Z)\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)} \Bigg| X\right]\right].$$
+
+%     Consider the following:
+%     \begin{equation}
+%     \frac{J^{-1} \sum_j h(Z_j^J) \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)} {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}
+%     \times \left( \frac{J^{-1} \sum_j \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)} {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}} \right)^{-1}.
+%     \end{equation}
+
+%     We will apply Lemma 2 to the numerator and the denominator in the ratio above individually. The numerator converges to 
+%     $$\frac{J^{-1} \sum_j h(Z_j^J)  \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)} {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}} \stackrel{a.s.}{\to} \E_{f_Z}\left[h(Z) \frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\right] = \E_{f_X}\left[\E_{f_{Z|X}}\left[h(Z)\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right],$$
+%     while the reciprocal of the denominator converges to 
+%     $$ \frac{J^{-1} \sum_j \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)}{U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}  \stackrel{a.s.}{\to} \E_{f_Z}\left[ \frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\right] = \E_{f_X}\left[\E_{f_{Z|X}}\left[\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right] .$$
+
+%     The numerator is exactly what we desire, so it remains to show that the reciprocal of the denominator converges to $1$. Applying Fubini, we have that
+%     \begin{align*}
+%         \E_{f_X}\left[\E_{f_{Z|X}}\left[\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right] &= \int_x\left(\int_z \frac{f_{Y|X}(z|x)}{f_{Z|X}(z|x)}f_{Z|X}(z|x)dz \right)f_X(x) dx \\
+%         &= \int_x\int_z f_{Y|X}(z|x) f_X(x) dzdx \\
+%         &= \int_z\int_x f_{Y|X}(z|x) f_X(x) dxdz \\
+%         &= \int_z f_{Y}(z) dz \\
+%         &= 1,
+%     \end{align*}
+%     and we are done.
+
+% \end{proof}
+
+\begin{thm}[Off-Parameter Resampled Particle Filters Target the Filtering Distribution]
+    \label{thm:off-parameter-targeting}
+    The off-parameter resampled particle filter as outlined in Definition \ref{defn:off-parameter-filter} targets the filtering distribution.
+\end{thm}
 \begin{proof}
-    From Lemma \ref{lem:marginal-proper-weights}, for any measurable bounded $g$, 
-     $$\frac{\sum_j g(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_Z}[g(Z)] = \E_{f_X}\big[\E_{f_{Z|X}}[g(Z) | X]\big],$$
-     and by assumption, for any measurable bounded $g$, 
-    $$\frac{\sum_j g(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[g(X)].$$
-
-    We in fact want to show that 
-    $$\frac{\sum_j h(\tilde Z_j^J) \, V_j^J}{\sum_j V_j^J} \stackrel{a.s.}{\to} \E_{f_Y}[h(Y)] = \E_{f_X}\left[\E_{f_{Z|X}}\left[h(Z)\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)} \Bigg| X\right]\right].$$
-
-    Consider the following:
-    \begin{equation}
-    \frac{J^{-1} \sum_j h(Z_j^J) \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)} {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}
-    \times \left( \frac{J^{-1} \sum_j \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)} {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}} \right)^{-1}.
-    \end{equation}
-
-    We will apply Lemma 2 to the numerator and the denominator in the ratio above individually. The numerator converges to 
-    $$\frac{J^{-1} \sum_j h(Z_j^J)  \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)} {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}} \stackrel{a.s.}{\to} \E_{f_Z}\left[h(Z) \frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\right] = \E_{f_X}\left[\E_{f_{Z|X}}\left[h(Z)\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right],$$
-    while the reciprocal of the denominator converges to 
-    $$ \frac{J^{-1} \sum_j \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)}{U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}  \stackrel{a.s.}{\to} \E_{f_Z}\left[ \frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\right] = \E_{f_X}\left[\E_{f_{Z|X}}\left[\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right] .$$
-
-    The numerator is exactly what we desire, so it remains to show that the reciprocal of the denominator converges to $1$. Applying Fubini, we have that
-    \begin{align*}
-        \E_{f_X}\left[\E_{f_{Z|X}}\left[\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right] &= \int_x\left(\int_z \frac{f_{Y|X}(z|x)}{f_{Z|X}(z|x)}f_{Z|X}(z|x)dz \right)f_X(x) dx \\
-        &= \int_x\int_z f_{Y|X}(z|x) f_X(x) dzdx \\
-        &= \int_z\int_x f_{Y|X}(z|x) f_X(x) dxdz \\
-        &= \int_z f_{Y}(z) dz \\
-        &= 1,
-    \end{align*}
-    and we are done.
-
+    Recursively applying Lemmas \ref{lem:change-measure-proper-weights}, \ref{lem:marginal-proper-weights}, and \ref{lem:posterior-proper-weights}, we obtain that the off-parameter resampled particle filter targets the posterior.
+    Specifically, suppose inductively that $\big\{\big(X^{F}_{n-1,j},w^{F}_{n-1,j}\big)\big\}$ targets $\pi_{n-1}$.
+    Then, Lemma \ref{lem:marginal-proper-weights} tells us that $\big\{\big(X^{P}_{n,j},w^{P}_{n,j}\big)\big\}$ targets $\eta_n$.
+    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P}_{n,j},w^{P}_{n,j} g^\theta_{n,j} \big)\big\}$ therefore targets  $\pi_n$.
+    Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
+    \[
+    \big(X^{F}_{n,j},w^{F}_{n,j}\big) = \big(X^{P}_{n,k_j}, w^{P}_{n,k_j} g_{n,k_j}\big/ p_{n,k_j}\big),
+    \]
+    with resampling probabilities proportional to $p_{n,j}$, therefore also targets $\pi_n$.
 \end{proof}
 
-\begin{prop}[MOP-1 Targets the Posterior]
-    When $\alpha=1$ or $\phi=\theta$, MOP-$\alpha$ targets the posterior. 
+\begin{prop}[MOP-1 Targets the Filtering Distribution]
+    When $\alpha=1$ or $\phi=\theta$, MOP-$\alpha$ targets the filtering distribution. 
 \end{prop}
@@
 
-    When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows.
+    When $\alpha=1$, and $\theta\neq\phi,$ the proof is identical to that of \ref{thm:off-parameter-targeting}.
 Recursively applying Lemmas \ref{lem:change-measure-proper-weights}, \ref{lem:marginal-proper-weights}, and \ref{lem:posterior-proper-weights}, we obtain that
@@
     \[
-    \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}\big/ g^\phi_{n,a(j)}\big),
+    \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,k_j}, w^{P,\theta}_{n,k_j} g^\theta_{n,k_j}\big/ g^\phi_{n,k_j}\big),
     \]
@@
 
-\begin{prop}[DOP-1 Targets the Posterior]
-    When $\alpha=1$ or $\phi=\theta$, DOP-$\alpha$ targets the posterior. 
-\end{prop}
-\begin{proof}
-    When $\theta=\phi$, regardless of the value of $\alpha$, the ratio $\frac{g_{n,j}^\theta h_{n,j}^\theta}{g_{n,j}^\phi h_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter estimate.
+% \begin{prop}[DOP-1 Targets the Posterior]
+%     When $\alpha=1$ or $\phi=\theta$, DOP-$\alpha$ targets the posterior. 
+% \end{prop}
+% \begin{proof}
+%     When $\theta=\phi$, regardless of the value of $\alpha$, the ratio $\frac{g_{n,j}^\theta h_{n,j}^\theta}{g_{n,j}^\phi h_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter estimate.
 
-    When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows.
+%     When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows.
 Recursively applying Lemmas \ref{lem:change-measure-proper-weights}, \ref{lem:posterior-proper-weights}, and \ref{lem:change-marginal-proper-weights}, we obtain that
-    %to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
-    the DOP-1 filter targets the posterior.
-    Specifically, suppose inductively that $\big\{\big(X^{F,\phi}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
-    Then, Lemma \ref{lem:change-marginal-proper-weights} tells us that $\big\{\big(X^{P,\phi}_{n,j},w^{P,\theta}_{n,j} h_{n,j}^\theta /h_{n,j}^\phi\big)\big\}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
-    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\phi}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j}h_{n,j}^\theta /h_{n,j}^\phi \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-    Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
-    \[
-    \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \left(X^{P,\phi}_{n,a(j)}, w^{P,\theta}_{n,a(j)} \frac{g^\theta_{n,a(j)}h_{n,a(j)}^\theta}{g^\phi_{n,a(j)}h_{n,a(j)}^\phi}\right),
-    \]
-    with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-\end{proof}
+%     %to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
+%     the DOP-1 filter targets the posterior.
+%     Specifically, suppose inductively that $\big\{\big(X^{F,\phi}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
+%     Then, Lemma \ref{lem:change-marginal-proper-weights} tells us that $\big\{\big(X^{P,\phi}_{n,j},w^{P,\theta}_{n,j} h_{n,j}^\theta /h_{n,j}^\phi\big)\big\}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+%     Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\phi}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j}h_{n,j}^\theta /h_{n,j}^\phi \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+%     Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
+%     \[
+%     \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \left(X^{P,\phi}_{n,a(j)}, w^{P,\theta}_{n,a(j)} \frac{g^\theta_{n,a(j)}h_{n,a(j)}^\theta}{g^\phi_{n,a(j)}h_{n,a(j)}^\phi}\right),
+%     \]
+%     with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+% \end{proof}
 
@@
 
-Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ in the case of MOP and $\{ \big(X^{P,\phi}_{n,j},w^{P,\theta}_{n,j}\big) \}$ in the case of DOP (with a slight abuse of notation here, as the weight expressions are in fact different for both algorithms) targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
 To see  (\ref{L2}),
@@
 $\sum_{j=1}^J w^{P,\theta}_{0,j} = J$ on the denominator and $\sum_{j=1}^J w^{F,\theta}_{N,j}$ on the numerator.
-This derives the MOP/DOP estimate.
+This derives the MOP likelihood estimates.
 
-\kevin{IS THIS UNBIASED?}
-
-$L^{B,\theta}$ should generally be preferred in practice, since there is no reason to include the extra variability from resampling when calculating the conditional log likelihood, but it lacks the nice telescoping product that lets us derive exact expressions for the gradient later.
+$L^{B,\theta}$ should generally be preferred in practice, since there is no reason to include the extra variability from resampling when calculating the conditional log likelihood, but it lacks the nice telescoping product that lets us derive exact expressions for the gradient in Theorem \ref{thm:mop-functional-forms}.

### 2024-06-13 paper/ms.tex
update from overleaf

@@
 Scibior and Wood \cite{scibior21} showed that the estimators derived by Poyiadjis et al.
-\cite{poyiadjis11} can be attained with standard AD software, using an algorithmic procedure called a ``stop-gradient'' which allows selected exressions to be evaluated but not differentiated.
-Unfortunately, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability.
+\cite{poyiadjis11} can be attained with standard AD software, using an algorithmic procedure called a ``stop-gradient'' which allows selected expressions to be evaluated but not differentiated.
+Unfortunately, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability, while the ``stop-gradient'' procedure is introduced without mathematical motivation.
 We use \cite{scibior21} and \cite{poyiadjis11} as our starting point, while developing a new approach which remedies the weaknesses of these papers.
@@
 
-MOP-$\alpha$ avoids the discontinuity that is problematic for the standard particle filter.
-This makes MOP-$\alpha$ convenient for the application of AD.
+MOP-$\alpha$ avoids the discontinuity that is problematic for the standard particle filter in a novel way.
+This particle filter is smooth by construction, allowing us to bypass the seemingly incompatible paradox of differentiating through a Monte Carlo algorithm -- by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
 Yet, MOP-$\alpha$ can nevertheless provide a consistent log-likelihood estimate.
@@
 That behavior is complementary to that of the iterated filtering algorithm \cite{ionides06-pnas,ionides15} which provides a relatively fast and stable way to identify this neighborhood.
-We therefore build a hybrid algorithm called {\it Iterated Filtering with Automatic Differentiation} (IFAD) that warm-starts first-order or second-order gradient methods with a preliminary solution obtained from a few rounds of iterated filtering.
+We therefore build a simple but effective hybrid algorithm called {\it Iterated Filtering with Automatic Differentiation} (IFAD) that warm-starts first-order or second-order gradient methods with a preliminary solution obtained from a few rounds of iterated filtering.
 Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.
 
-These improvements also extend to Bayesian inference, as we show in the supplement that we can use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with a nonparametric empirical Bayes-style prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{andrieu10} from the $1.4 \times 10^6$ iterations in \cite{fasiolo16} to just $500$.
+These improvements also extend to Bayesian inference, as we show in Section \ref{sec:bayes} that we can use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with a nonparametric empirical Bayes-style prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{andrieu10} on the Dhaka cholera model of \cite{king08} from the $700,000$ iterations in \cite{fasiolo16} to just $500$.
+To the best of our knowledge, attaining rapid mixing on this challenging problem has not previously been achievable.
 
@@
 We avoid this issue by constructing a particle filtering algorithm called {\it Measurement Off-Parameter} with discount factor $\alpha$ (MOP-$\alpha$), defined by the pseudocode in Algorithm~\ref{alg:mop}.
-For MOP-$\alpha$, the resampling indices are invariant to $\theta$ when $\omega$ and $\phi$ are fixed.
-We assume that the measurement model is differentiable in $\theta$.
-Further, we assume that the simulator is a differentiable function of $\theta$ for every fixed $\omega$, a condition that requires the latent process to be a continuous random variable.
-\ed{WE COULD CALL THIS THE REPARAMETERIZATION TRICK, BUT HERE FOR AD IT ARISES THROUGH THE STRUCTURE OF THE CODE SO WE DO NOT HAVE TO EXPLICITLY THINK OF THE EXISTENCE OF A REPARAMETERIZATION.}
-Under these assumptions, direct differentiation of MOP-$\alpha$ is available via AD.
+In short, instead of differentiating through a Monte Carlo algorithm directly, we instead use AD to differentiate through a series of measurement density ratios and a differentiable simulator.
+
+This algorithm is an example of a new class of algorithms we call ``off-parameter'' particle filters that resample the particles according to an arbitrary resampling rule that is not necessarily in proportion to the target distribution of interest, while maintaining weights that encode the cumulative discrepancy between the resampling distribution and the target distribution.
+We later show in Theorem \ref{thm:mop-targeting} that off-parameter particle filters successfully target the filtering distribution when the correction is sufficient.
+
+In the case of MOP-$\alpha$, we wish to evaluate the likelihood at some $\theta \in \Theta$, but instead resample the particles according to the indices generated by a vanilla particle filter run at some $\phi \in \Theta$.
+This ensures that the resampling indices are invariant to $\theta$ when $\omega$ and $\phi$ are fixed, bypassing the issue of Monte Carlo resampling.
+When the measurement model is differentiable in $\theta$ and the simulator is a differentiable function of $\theta$ for every fixed $\omega$ (a condition that requires the latent process to be a continuous random variable, also known as the reparameterization trick \cite{corenflos21}), direct differentiation of MOP-$\alpha$ is available via AD.
 As we explain subsequently, MOP-$\alpha$ is constructed so this direct derivative obtains the score estimator of  \cite{poyiadjis11, scibior21} when $\alpha=1$ and that of \cite{naesseth18} when $\alpha=0$.
-Setting $\alpha<1$ adds bias but reduces variance, raising opportunities for a favorable tradeoff.
+Setting $\alpha<1$ adds bias but reduces variance, raising opportunities for a favorable tradeoff as illustrated in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).
 
@@
 The resampling indices, which we write as $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$, are a function of $\phi$ for any value of $\theta$.
-For the categorical distribution, we use systematic resamplig \cite{arulampalam02,king16} which usually has superior performance to multinomial resampling.
+For the categorical distribution, we use systematic resampling \cite{arulampalam02,king16} which usually has superior performance to multinomial resampling.
 If $\theta$ and $\phi$ coincide, one only needs one particle filter run at $\theta=\phi$, otherwise one needs two runs at the same seed $\omega \in \Omega$.
@@
 
-This yields a suitably reweighted particle filter that targets the posterior under $\theta$ and produces a strongly consistent likelihood estimate for the likelihood at $\theta$.
-Taking the gradient with respect to $\theta$ of the estimate for the log-likelihood at $\theta$ then yields an estimate for the score at $\theta$.
+This yields a suitably reweighted particle filter that targets the filtering distribution under $\theta$ and produces a strongly consistent likelihood estimate for the likelihood at $\theta$.
+As the resampling indices are invariant to $\theta$ when $\omega$ and $\phi$ are fixed, we successfully bypass the issue of Monte Carlo resampling when the simulator and and measurement model are differentiable.
+Taking the gradient with respect to $\theta$ of the log-likelihood estimate for $\theta$ then yields an estimate for the score at $\theta$.
 
@@
 The correction factors, $w^{P,\theta}_{n,j}$ and $w^{F,\theta}_{n,j}$, accumulate as $n$ increases, leading to numerical instability.
-One can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$ as in Equation \ref{eq:weighting-scheme} while still targeting the posterior and likelihood if $\theta=\phi$.
+One can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$ as in Equation \ref{eq:weighting-scheme} while still targeting the filtering distribution and likelihood if $\theta=\phi$.
 This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate.
@@
 When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{naesseth18}. 
+This discount factor is novel and effective both in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).
 
@@
 
-\cite{scibior21} show that the estimate of \cite{naesseth18} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through applying a ``stop-gradient trick" to the particle filter.
+\cite{scibior21} show that the estimate of \cite{naesseth18} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through the use of a stop-gradient operation in the AD procedure.
 It turns out that both of these, when applied on the bootstrap filter, correspond to special cases of MOP-$\alpha$.
@@
 One then only needs to run one particle filter at $\theta=\phi$, setting the particles at $\phi$ to be copies of the particles at $\theta$ where gradients don't propagate.
-This is done algorithmically through the \texttt{stop\_gradient()} function in \texttt{JAX}, providing a mathematical justification for the ``stop-gradient trick" of \cite{scibior21}.
+This is done algorithmically through the \texttt{stop\_gradient()} function in \texttt{JAX}, providing a mathematical justification for the use of the stop-gradient operation by \cite{scibior21}.
 
@@
 
-The  convergence of IF2 (and so the first stage of IFAD) to a neighborhood of the MLE happens fairly quickly in practice.
+The  convergence of IF2 (and so the first stage of IFAD) to a neighborhood of the MLE happens rapidly in practice. 
 In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations.
@@
 
-On the other hand, the second stage of IFAD enjoys a linear convergence rate under the usual strongly-convex and smoothness assumptions in convex optimization, as we show below in Theorem \ref{thm:mop-convergence}.
+On the other hand, the second stage of IFAD enjoys a linear convergence rate under the usual smoothness and strong convexity assumptions in the theory of convex optimization, as we show below in Theorem \ref{thm:mop-convergence}.
 
@@
     \vspace*{-7mm}
-    \caption{Illustration of Dhaka cholera model from \cite{king08}. \ed{CHANGE H TO P? ALSO, THERE ARE ADVANTAGES TO PUTTING THE TIKZ SOURCE CODE INTO THE MS, TO ALLOW FOR FINE-TUNING EDITS.}}
+    \caption{Illustration of Dhaka cholera model from \cite{king08}.}
     \label{fig:tikz-cholera}
@@
 We tested IFAD against IF2 on a global search problem for the Dhaka cholera model.
-We re-implemented IF2 in order to do so, but we compare our results with the results of \cite{ionides15} (labeled "IF2 2015").
+We re-implemented IF2 to do so, but we compare our results with the results of \cite{ionides15} (labeled "IF2 2015").
 Our re-implementation ourperforms that of \cite{ionides15}, likely due to a better choice of algorithmic parameters.
@@
 % \kevin{Dynamically save to a tex file and load from there}
-
-\ed{PERHAPS THE 2ND D.P. OF SIGNIFICANCE CAN BE AVOIDED? ALSO, IF WE'RE GOING TO ADVISE MOP-ALPHA FOR $0<\alpha<1$, THAT COULD BE SHOWN HERE.?}
 
@@
 Our implementation of IF2 outperforms that of \cite{ionides15}, but still ultimately underperforms IFAD.
-IFAD manages to find the MLE, matching the highest log-likelihood previously found in the \texttt{dacca()} model implemented within the \texttt{pomp} package of \cite{king16}.}
+IFAD manages to find the MLE, matching the highest log-likelihood previously found in the Dhaka cholera model implemented within the \texttt{pomp} package of \cite{king16}.}
 \label{table:mle}
@@
 It is clear that on average, IFAD performs best, and manages to find the MLE while no IF2 search successfully gets within 7 log-likelihood units of it.
-We use a dotted red line to display the MLE in both.
+The dotted red line shows the true maximized log-likelihood.
 }
@@
     \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/boxplot_all.png}
-    \caption{\textbf{Left:} Raincloud plot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs, simulating the result of the common procedure of running a few searches and choosing the best one.
+    \caption{\textbf{Left:} Raincloud plot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs, representing the common procedure of running a few searches and choosing the best one.
 \textbf{Right:} Raincloud plot of all searches.
 IFAD outperforms all other methods, and the gradient steps improve on the warm-start given by running 40 IF2 iterations.
-    We use a dotted red line to display the MLE.}
+    The dotted red line shows the true maximized log-likelihood.}
     \label{fig:boxplot}
@@
 
-\ed{IS THE IFAD(0.95) MEDIAN LINE IN FIG 4 (RIGHT) HIDING UNDER IFAD(1)? IF NOT, WEHRE IS IT?}
+\section{Application to Bayesian Inference}
+\label{sec:bayes}
+
+Particle MCMC, as introduced by \cite{andrieu10}, is arguably the most popular method for full-information Bayesian inference within partially observed Markov processes that does not require access to the system's transition densities.
+This feature enables its use in simulation-based Bayesian inference for complex scientific models where computing transition densities is not feasible, such as in disease modeling.
+However, the particle Metropolis-Hastings algorithm often experiences slow mixing and requires long burn-in periods.
+For instance, \cite{fasiolo16} reported needing $700,000$ burn-in iterations for effective sampling in a cholera model -- even after adjusting the length of each Euler timestep to be a month instead of a day to reduce computational costs.
+
+We employ a NUTS sampler powered by a MOP-$\alpha$, with a nonparametric empirical prior initialized by the IF2 warm start from the previous section.
+To construct the prior, we perform a kernel density estimate (KDE) on the parameter swarm from the warm start, and use the output of that as an empirical prior (the KDE is performed to ensure some degree of mutual contiguity between the densities).
+This yields a prior that is roughly equivalent to a parameter cloud centered at some point in the neighborhood of the MLE, with a standard deviation of $\sqrt{(0.02\cdot0.95^{40})^2 \cdot 600}\approx 6.3\%$ of the MLE.
+Using a NUTS sampler powered by MOP-$\alpha$ with this prior significantly lowers the mixing time of the Markov chain to as little as $500$ iterations, as seen in Figure \ref{fig:nuts-eb}.
+
+\begin{figure}[t!]
+    \centering
+    \includegraphics[width=\textwidth/\real{1.25}]{imgs/pmcmc/nuts_eb.png}
+    \caption{Convergence diagnostics for a No U-Turn Sampler with the informative nonparametric empirical Bayes prior from IF2 and a KDE.
+We display the results for the trend parameter in the Dhaka cholera model of \cite{king08}.
+The NUTS sampler mixes quickly, and the posterior estimates from each chain share roughly the same posterior mode.}
+    \label{fig:nuts-eb}
+\end{figure}
+
+For completeness, we note that without the empirical Bayes-style prior from IF2, NUTS with the MOP-$\alpha$ estimate alone tends to diverge.
+In contrast, particle Metropolis-Hastings (with or without the empirical Bayes prior from IF2) fails to effectively explore the parameter space.
+We display the results of the above in Figures \ref{fig:mh} and \ref{fig:nuts} within the supplementary information.
+
+The result obtained in this section, of a speedup in the mixing time of particle MCMC by over three orders of magnitude, provides hope that particle Bayesian inference might finally be practical for the complex scientific models commonly encountered in areas like epidemiology.
 
@@
 
-Here, we will show that MOP-$\alpha$ targets the posterior and likelihood and yields the estimators of \cite{poyiadjis11, scibior21, naesseth18} as special cases when differentiated, and characterize rates for its bias and variance under different choices of $\alpha$.
+Here, we will show that MOP-$\alpha$ targets the filtering distribution and likelihood under $\theta$, is consistent when $\alpha=1$, and characterize rates for its bias and variance under different choices of $\alpha$.
 To do so, we require the following assumptions:
@@
 
-\subsection{MOP-$\alpha$ Targets the Posterior}
+\subsection{MOP-$\alpha$ Targets the Filtering Distribution}
 
-We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood when $\alpha=1$ or $\theta=\phi$.
+We show here that MOP-$\alpha$ targets the filtering distribution under $\theta$ and is strongly consistent for the likelihood under $\theta$ when $\alpha=1$ or $\theta=\phi$.
 Employing $\alpha<1$ lead to inconsistency when $\theta$ deviates from $\phi$, but this bias disappears as $\theta$ approaches $\phi$.
@@
 
-\begin{thm}[MOP-$\alpha$ Targets the Posterior and Likelihood]
+\begin{thm}[MOP-$\alpha$ Targets the Filtering Distribution and Likelihood]
     \label{thm:mop-targeting}
-    When $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood.
-That is, for $\pi_n(\theta)=f_{X_{1:n}|Y_{1:n} ; \theta}$ and any measurable and bounded functional $h$ and for $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\prod_{n=1}^N L_n^{B,\theta,\alpha}$, it holds that
+    When $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the filtering distribution under $\theta$ and is strongly consistent for the likelihood under $\theta$.
+That is, for $\pi_n(\theta)=f_{X_{n}|Y_{1:n} ; \theta}$ and any measurable and bounded functional $h$ and for $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\prod_{n=1}^N L_n^{B,\theta,\alpha}$, it holds that
     \vspace*{-2.5mm}
     \begin{equation}
-        \frac{\sum_{j=1}^J h(x_{n,j}^{A,F, \theta}) w_{n,j}^{A,F,\theta}}{\sum_{j=1}^J w_{n,j}^{A,F,\theta}} \stackrel{a.s.}{\to} E_{\pi_n(\theta)} h, \;\; \hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).
+        \frac{\sum_{j=1}^J h(x_{n,j}^{F, \theta}) w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{F,\theta}} \stackrel{a.s.}{\to} E_{\pi_n(\theta)} h, \;\; \hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).
     \vspace*{-2.5mm}
@@
     When $\alpha=1$ and $\theta\neq\phi,$ suppose inductively that $\{(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j})\}_{j=1}^J$ targets $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
-    It can then be shown that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j})\}_{j=1}^J$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$, that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} )\}_{j=1}^J$ targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$, and that weighting the particles by $(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}) = (X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}/ g^\phi_{n,a(j)})$ where $a(\cdot)$ is the ancestor function,
-    resampling with probabilities proportional to $g^\phi_{n,j}$, also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+    It can then be shown that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j})\}_{j=1}^J$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$, that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} )\}_{j=1}^J$ targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$, and that weighting the particles by $(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}) = (X^{P,\theta}_{n,k_j}, w^{P,\theta}_{n,k_j} g^\theta_{n,k_j}/ g^\phi_{n,k_j})$,
+    resampling the $k_j$ with probabilities proportional to $g^\phi_{n,j}$, also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
     If the likelihood is estimated with the before-resampling conditional likelihoods $\hat\lik(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, the strong consistency is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
@@
 The sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$ by Assumption \ref{assump:local-bounded-derivative}.
-Again by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J^1(\theta)(\omega)|_{\theta=\theta'}$ is also bounded by $H^*$ for almost every $\omega\in \Omega$ and every $\theta'\in \Theta$.
+Again by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J^1(\theta)(\omega)|_{\theta=\theta'}$ is also uniformly bounded over all $J$ by $H^*$ for almost every $\omega\in \Omega$ and every $\theta'\in \Theta$.
 So $(\nabla_\theta \hat\lik_J^1(\theta)( \omega))_{J \in \mathbb{N}}$ is uniformly Lipschitz, and therefore uniformly equicontinuous for almost every $\omega \in \Omega$.
@@
     By Arzela-Ascoli, there is a uniformly convergent subsequence.
-But there is only one subsequential limit, as we can treat the gradient estimate at $\theta$ as a bounded functional of the particles by Assumption \ref{assump:local-bounded-derivative}, allowing us to apply Theorem \ref{thm:mop-targeting} to see that the sequence $(\nabla_\theta \hat\lik_J^1(\theta, \omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$.
+But there is only one subsequential limit, as we can treat the gradient estimate at $\theta$ as a bounded functional of the particles by Assumption \ref{assump:local-bounded-derivative}, allowing us to apply Theorem \ref{thm:mop-targeting} to see that the sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$.
 So the whole sequence must converge uniformly to $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega).$
 
-    With uniform convergence for the derivatives established, we can swap the limit and derivative and obtain, in conjunction with the strong consistency $\hat{\lik}_J^1(\theta, \omega) \stackrel{a.s.}{\to} \lik(\theta)$ in Theorem \ref{thm:mop-targeting}, that for almost every $\omega \in \Omega$,
+    With uniform convergence for the derivatives established, we can swap the limit and derivative and obtain, in conjunction with the strong consistency $\hat{\lik}_J^1(\theta)(\omega) \stackrel{a.s.}{\to} \lik(\theta)$ in Theorem \ref{thm:mop-targeting}, that for almost every $\omega \in \Omega$,
     $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lik(\theta).$
@@
 The analysis is similar, with slightly better rates but a less natural interpretation.
+
+Practical likelihood-based data analysis involves many likelihood optimizations \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
+In particular, profile likelihood confidence intervals require a sequence of optimizations.
+Additionally, a careful scientist should consider many model variations, to see whether the conclusions of the study are sensitive to alternative model specifications.
+Coding model variations is relatively simple when using plug-and-play inference methodology.
+However, assessing the scientific potential of these variations requires likelihood optimization.
+If many rounds of laborious searching are required to attain the maximum (as arises in previous methods) that has practical consequences for the rate at which scientists can evaluate hypotheses.
+We have demonstrated, for the first time, a plug-and-play maximum likelihood approach which can directly maximize the likelihood for the complex benchmark model of \cite{king08}, requiring some Monte Carlo replication but not prolonged sequences of explorations.
+We anticipate that this will promote scientific advances in all domains where complex POMP models arise.
 
@@
 
-\section{MOP Functional Forms}
+\section{MOP-$\alpha$ Functional Forms}
 \input{appendix/functional}
@@
 
-\section{Weighted Particle Filters that Target the Posterior}
+\section{A Strong Law of Large Numbers for Triangular Arrays of Particles With ``Off-Parameter’’ Resampling}
 \input{appendix/targeting}
 
-\section{Consistency of MOP-$1$}
+\section{Consistency of Off-Parameter Resampled Gradient Estimates}
 \input{appendix/consistency}
@@
 
+\section{Figures for Bayesian Inference}
+\input{appendix/bayes}
 \end{document}

### 2024-06-13 paper/appendix/bayes.tex
various minor Latex edits

@@
     \centering
-    \includegraphics[width=\textwidth/\real{1.25}]{imgs/pmcmc/nuts.png}
-    \caption{Convergence diagnostics for a No U-Turn Sampler with uniform priors on a compact set.
+    \includegraphics[width=\textwidth/\real{1.1}]{imgs/pmcmc/nuts.png}
+    \caption{Convergence diagnostics for a No-U-Turn Sampler (NUTS) with uniform priors on a compact set.
 Again, we display the results for the trend parameter in the Dhaka cholera model of \cite{king08}.
-The NUTS sampler explores more of the posterior than the MH sampler, but fails to converge quickly.}
+The NUTS sampler explores more of the posterior than the Metropolis-Hastings sampler, but fails to converge quickly.}
     \label{fig:nuts}
@@
     \centering
-    \includegraphics[scale=0.3]{imgs/pmcmc/nuts_eb/Recovery Rate.png}
-    \includegraphics[scale=0.3]{imgs/pmcmc/nuts_eb/Death Rate.png}
-    \includegraphics[scale=0.3]{imgs/pmcmc/nuts_eb/Reciprocal of Mean Immunity Duration.png}
-    \includegraphics[scale=0.3]{imgs/pmcmc/nuts_eb/Trend in Force of Infection.png}
-    \includegraphics[scale=0.3]{imgs/pmcmc/nuts_eb/Noise in Force of Infection.png}
-    \includegraphics[scale=0.3]{imgs/pmcmc/nuts_eb/Measurement Noise.png}
-    \caption{Posterior estimates from NUTS powered by MOP-$\alpha$ with the informative nonparametric empirical prior obtained from a KDE on the IF2 parameter cloud.
+    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Recovery Rate.png}
+    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Death Rate.png}
+    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Reciprocal of Mean Immunity Duration.png}
+    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Trend in Force of Infection.png}
+    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Noise in Force of Infection.png}
+    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Measurement Noise.png}
+    \caption{Posterior estimates from NUTS powered by MOP-$\alpha$ with the informative nonparametric empirical prior obtained from a kernel density estimator on the IF2 parameter cloud.
 The posterior estimates from each chain are largely in agreement.}

### 2024-06-13 paper/appendix/biasvar.tex
various minor Latex edits

@@
 In this section, we prove various rates on the bias, variance, and MSE of MOP-$\alpha$. 
-
 First, we note the following relation between the Dobrushin contraction coefficient and the alpha-mixing coefficients in our context below.
-
 \begin{lem}
@@
 \begin{proof}
-    We rewrite the alpha-mixing assertion as $$\int { \int |f_Y|X(y|x) - f_Y(y)| dy } f_X dx.$$
-
+    We rewrite the alpha-mixing assertion as 
+    $$\int { \int |f_Y|X(y|x) - f_Y(y)| dy } f_X dx.$$
     We claim that the Dobrushin coefficient implies $$\int | f_Y|X(y|x_1) - f_Y(y) |dy < \alpha$$ for all $x_1$. This is shown as follows:
@@
     \end{align*}
-
     We then have the desired result.
@@
 \subsection{Warm Up: MOP-$0$ Variance Bound}
-
 We now prove the variance bound, presented below. 
-
 \begin{thm}
@@
 \end{thm}
-
 We made Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} to leverage the results from \cite{karjalainen23} on the forgetting of the particle filter.
@@
 $$\hat{\mathcal{L}}(\theta):=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J w_{n, j}^{P, \theta}}=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J (w_{n-1, j}^{F, \theta})^\alpha}.$$
-
 Now, to illustrate the proof strategy for the general case in an easier context, we first analyze the special case when $\alpha=0.$
@@
 \end{equation}
-
 Similarly to the proof of Lemma \ref{lem:mop-0-formula}, its gradient when $\theta=\phi$ is therefore 
@@
 $$\nabla_\theta \hat{\ell}(\theta) = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right).$$
-
 Therefore,
 \begin{align*}
-    \text{Var}(\nabla_\theta \hat\ell(\theta)) &= \frac{1}{J^2}\text{Var}\left(\sum_{n=1}^N\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &= \frac{1}{J^2}\sum_{n=1}^N\text{Var}\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &+ 2\sum_{m<n}\text{Cov}\left(\frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta)).
+    \Var(\nabla_\theta \hat\ell(\theta)) &= \frac{1}{J^2}\Var\left(\sum_{n=1}^N\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    &= \frac{1}{J^2}\sum_{n=1}^N\Var\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    &+ 2\sum_{m<n}\Cov\left(\frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    &= \sum_{n=1}^N \Var(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \Cov(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta)).
 \end{align*}
-
 Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
-We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
+We know from Theorem 3 of \cite{karjalainen23} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
 $$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
@@
 \begin{align*}
-    \text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) 
+    \Cov(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) 
     &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat\ell_m(\theta))^4]^{1/4}\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/4}\\
@@
 \end{align*}
-
 To bound this, we use the fact that $$\E[(\nabla_\theta\hat\ell_n(\theta))^4] = \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4]+\E[(\nabla_\theta\hat\ell_n(\theta))^2]^2$$
-alongside Lemma 2 of \cite{karjalainen2023}, which shows that
-$$\E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4] \lesssim \frac{G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^2] \lesssim \frac{G'(\theta)^2}{J},$$
-to find that 
+alongside Lemma 2 of \cite{karjalainen23}, which shows that
+$$\E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4] \lesssim \frac{G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^2] \lesssim \frac{G'(\theta)^2}{J}.$$
+It follows that 
 $$\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{G'(\theta)^4}{J^2}+\left(\frac{G'(\theta)^2}{J}\right)^2} = \frac{G'(\theta)^2}{J},$$
-and conclude that 
-$$\text{Cov}(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{G'(\theta)^2}{J}.$$
-
+and we conclude that 
+$$\Cov(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{G'(\theta)^2}{J}.$$
 Putting it all together, we see that
 \begin{align*}
-    \text{Var}(\nabla_\theta \hat\ell(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta))\\
+    \Var(\nabla_\theta \hat\ell(\theta)) &= \sum_{n=1}^N \Var(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \Cov(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta))\\
     &\leq \frac{NG'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\frac{G'(\theta)^2}{J} \\
@@
 \end{align*}
-
 \end{proof}

### 2024-06-13 paper/appendix/consistency.tex
various minor Latex edits

@@
     $$||\eta_n^J(h_\theta)(\omega)-\eta_n^J(h_{\theta'})(\omega)||_\infty < \epsilon.$$
-
     Then, by Arzela-Ascoli, there exists a uniformly convergent subsequence.
@@
     $$\lim_{J \to \infty} \eta_n^J(h_\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \eta_n^J(h_\theta)(\omega).$$
-
 Again from Theorem \ref{thm:off-parameter-targeting}, we know that
     $$\eta_n^J(h_\theta)(\omega) \to \eta_n(h_\theta) = h^*_\theta$$ for almost every $\omega\in\Omega$.
-
     We then have that for almost every $\omega \in \Omega$, 
@@
     as we wanted. 
-
     \end{proof}
-
 The proof of Theorem \ref{thm:mop-grad-consistency} is now merely a corollary, where we apply $\eta_n^J (h_\theta) = \hat\lik^1_J(\theta)$, $\eta_n(h_\theta) = \lik(\theta) = h_\theta^*$, and then use the continuous mapping theorem.

### 2024-06-13 paper/appendix/targeting.tex
various minor Latex edits

@@
 Using weights that encode the cumulative discrepancy between the resampling distribution and the target distribution (instead of resampling to equal weights as is common) turns out to be a sufficient correction that ensures almost sure convergence.
-
 We now introduce the precise definition of an off-parameter resampled particle filter.
@@
 We can then follow the argument of \cite{chopin20} from this point on, where noting that the $Z_j^J$ are conditionally independent given the $\tilde{X}_j^J$ and $U_j^J$,
-    $$\E\left[\left(\sum_j Z_j^J\right)^4 \Bigg| (\tilde X_j^J,U_j^J)_{j=1}^J \right] 
-    = J\E[(Z_1^J)^4|(\tilde X_j^J,U_j^J)_{j=1}^J] + 3J(J-1)\left(\E[(Z_j^J)^2|(\tilde X_j^J,U_j^J)_{j=1}^J]\right) \leq CJ^2,$$ for some $C>0$. Taking expectations on both sides yields
+    \begin{eqnarray*} 
+    \E\left[\left(\sum_j Z_j^J\right)^4 \Bigg| (\tilde X_j^J,U_j^J)_{j=1}^J \right] 
+    &=& J\E\left[(Z_1^J)^4|(\tilde X_j^J,U_j^J)_{j=1}^J\right] + 3J(J-1)\left(\E\big[(Z_j^J)^2|(\tilde X_j^J,U_j^J)_{j=1}^J\big]\right) 
+    \\ 
+    &\leq& CJ^2,
+    \end{eqnarray*}
+    for some $C>0$. 
+    Taking expectations on both sides yields
     $$\E\left[\left(\sum_j Z_j^J\right)^4 \right]  \leq CJ^2$$
@@
     $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J,$$
-    and by the same fourth moment argument to the above lemma, it converges almost surely to the limit of its expectation
+    and by the same fourth moment argument to the above lemma, it converges almost surely to the limit of its expectation,
     \begin{eqnarray*}        
@@
     $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\,\bar{U}_j^J \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)].$$ 
-
     \end{proof}
@@
     \end{equation}
-    We will apply Equation \ref{eq:lemma-posterior-hypothesis} to the numerator and the denominator in the ratio above individually. The numerator converges to 
+    We will apply Equation (\ref{eq:lemma-posterior-hypothesis}) to the numerator and the denominator in the ratio above individually. The numerator converges to 
     $$\frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \stackrel{a.s.}{\to} \E_{f_X}[h(X)f_{Z|X}(z^*|X)],$$

### 2024-06-13 paper/ms.tex
various minor Latex edits

@@
 
-We defer the proof to the supplementary material.
-It relies on a useful decomposition of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$ that yields a telescoping product in the MOP-$1$ case.
+We defer the proof to \arxiv{Appendix~\ref{appendix:functional}}{the supplementary material}. 
+The argument relies on a useful decomposition of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$ that yields a telescoping product in the MOP-$1$ case. 
 Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting $\theta=\phi$ implies that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result.
@@
 
-We defer the proof, which is similar to that of Theorem 6 in \cite{mahoney16}, to the supplementary information.
+We defer the proof, which is similar to that of Theorem 6 in \cite{mahoney16}, to \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}.
 For simplicity, we only prove this for MOP-$1$ and note that the general case follows as the required $1/2$-approximation can still be obtained for sufficiently large $\alpha$.
@@
 Monte~Carlo replication is appropriate for IFAD, as for any Monte Carlo algorithm used to solve a challenging numerical problem, but Figure~\ref{fig:boxplot} shows that IFAD can find higher likelihood values, more quickly and more reliably than the previous state-of-the-art.
-\ed{I ADDED THIS PARAGRAPH TO TRY TO BE EXPLICIT ABOUT WHY WE ARE EXCITED BY THIS FIGURE.
-AT FACE VALUE, IT MAY NOT SEEM LIKE A BIG DEAL.
-IT WOULD BE BETTER TO EXPLAIN MORE CLEARLY WHY THIS IS EVIDENCE THAT THE METHOD COULD HAVE A TRANSFORMATIVE EFFECT ON DATA ANALYSIS VIA POMP MODELS.}
+In particular, Figure~\ref{fig:boxplot} shows that IFAD with $\alpha=0.97$ can maximize the challenging likelihood of \cite{king08} using a modest number of iterations and Monte Carlo replications, whereas previously it was necessary to carry out an extensive customized search or to live with an incompletely maximized likelihood.
 
@@
 We used constant learning rates of $0.01, 0.05$, and $0.2$ for IFAD-$1,0,$ and $0.97$ respectively, and a constant cooling rate of $0.95$ for our IF2 implementation.
-While techniques such as annealing learning rates, gradient normalization, and momentum made the outperformance of IFAD even more apparent in other simulations we performed, we chose to report the results of the simplest implementation in part to serve as a baseline for the method's performance, and in part because the results of the simplest implementation were already very promising.
+While techniques such as annealing learning rates, gradient normalization, and momentum could further improve the performance of IFAD in other simulations we performed, we chose to report the results of the simplest implementation to serve as a baseline for the method's performance.
+The results for this basic implementation are already sufficient to show the high potential of the approach.
 
@@
 In contrast, particle Metropolis-Hastings (with or without the empirical Bayes prior from IF2) fails to effectively explore the parameter space.
-We display the results of the above in Figures \ref{fig:mh} and \ref{fig:nuts} within the supplementary information.
+We display the results of the above in Figures \ref{fig:mh} and \ref{fig:nuts} \arxiv{in Appendix~\ref{appendix:bayes}}{within the supplementary information}.
 
@@
 
-While the result is presented here as specific to MOP-$\alpha$, we actually prove a more general result in the supplementary material.
+While the result is presented here as specific to MOP-$\alpha$, we actually prove a more general result in \arxiv{Appendix~\ref{appendix:targeting}}{the supplementary material}.
 That is, we show a strong law of large numbers for triangular arrays of particles with ``off-parameter'' resampling, where we resample the particles according to an arbitrary resampling rule not necessarily in proportion to the targeted distribution of interest and employ weights that encode the cumulative discrepancy between the resampling and the target distribution instead of equal weights.
@@
 \begin{proof}
-    We provide a proof sketch here, deferring most of the details, and discussion of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$, to the supplementary information. 
+    We provide a proof sketch here, deferring most of the details, and discussion of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$, to \arxiv{Appendix~\ref{appendix:targeting}}{the supplementary information}.
     When $\theta=\phi$, regardless of the value of $\alpha$, the ratio ${g_{n,j}^\theta}/{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter.
@@
 It is therefore not immediately apparent that these two converge to the same thing. As such we directly show the consistency of the MOP-$1$ gradient estimate below. 
+We present an abbreviated proof here, postponing the full argument to \arxiv{Appendix~\ref{appendix:consistency}}{the supplement}.
 
@@
         &\lesssim \min_{k \leq N} NpG'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi(\alpha)\right),\\[-1ex]
-        &\text{Var}(\nabla_\theta \hat\ell^{\alpha}(\theta)) \lesssim \min_{k\leq N} NpG'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N\right).
+        &\Var(\nabla_\theta \hat\ell^{\alpha}(\theta)) \lesssim \min_{k\leq N} NpG'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N\right).
         \end{align*}
@@
 
-We defer the proof to the supplementary material, but provide a brief outline here.
+We defer the proof to \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material}, but provide a brief outline here. 
 The variance bound can be reduced to the approximation error between MOP-$\alpha$ and a variant called MOP-$(\alpha,k)$ where the weights are truncated at lag $k$ with discount factor $\alpha$ to ensure strong mixing, and the covariance of the latter which can be controlled with Davydov's inequality and a standard $L^p$ error bound for the particle filter.
 The MSE bound considers the error between the score and MOP-$(1,k)$, and the error between MOP-$(1,k)$ and MOP-$\alpha$. 
-The latter is $\tilde{O}(NpG'(\theta)^2(k+\psi(\alpha)))$ by tedious algebra, and the former can be controlled with a result on the the forgetting of the particle filter from \cite{karjalainen23} and the very same $L^p$ error bound mentioned above.
+The latter is $\tilde{O}(NpG'(\theta)^2(k+\psi(\alpha)))$, by routine algebra, and the former can be controlled with a result on the the forgetting of the particle filter from \cite{karjalainen23} and the very same $L^p$ error bound mentioned above.
 
-We also show in the supplementary material that the variance of MOP-$0$ is $\tilde{O}(NpG'(\theta)^2/J)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$ (ignoring factors of $p, G'$, etc.).
+We also show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}(NpG'(\theta)^2/J)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$ (ignoring factors of $p, G'$, etc.).
 This and Theorem \ref{thm:mop-biasvar} show that MOP-$\alpha$ interpolates between MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$.
@@
 
-\section{Discussion, Limitations, and Future Work}
+\section{Discussion}
 
@@
 
-%% \subsection*{Author Affiliations}
-
-%% Include department, institution, and complete address, with the ZIP/postal code, for each author.
-Use lower case letters to match authors with institutions, as shown in the example.
-PNAS strongly encourages authors to supply an \href{https://orcid.org/}{ORCID identifier} for each author.
-Individual authors must link their ORCID account to their PNAS account at \href{http://www.pnascentral.org/}{www.pnascentral.org}.
-For proper authentication, authors must provide their ORCID at submission and are not permitted to add ORCIDs on proofs.
-
-% \subsection*{Format}
-
-% Many authors find it useful to organize their manuscripts with the following order of sections: title, author line and affiliations, keywords, abstract, significance statement, introduction, results, discussion, materials and methods, acknowledgments, and references.
-Other orders and headings are permitted.
-
-%% \subsection*{Manuscript Length}
-
-%% A standard 6-page article is approximately 4,000 words, 50 references, and 4 medium-size graphical elements (i.e., figures and tables).
-The preferred length of articles remains at 6 pages, but PNAS will allow articles up to a maximum of 12 pages.
-
-%% \subsection*{References}
-
-% %% References should be cited in numerical order as they appear in text; this will be done automatically via bibtex,
-
-% \subsection*{Data Archival}
-
-% PNAS must be able to archive the data essential to a published article.
-Where such archiving is not possible, deposition of data in public databases, such as GenBank, ArrayExpress, Protein Data Bank, Unidata, and others outlined in the \href{https://www.pnas.org/author-center/editorial-and-journal-policies#materials-and-data-availability}{Information for Authors}, is acceptable.
-
-%% \subsection*{Digital Figures}
-
-%% EPS, high-resolution PDF, and PowerPoint are preferred formats for figures that will be used in the main manuscript.
-Authors may submit PRC or U3D files for 3D images; these must be accompanied by 2D representations in TIFF, EPS, or high-resolution PDF format.
-Color images must be in RGB (red, green, blue) mode.
-Include the font files for any text.
-
-%% Images must be provided at final size, preferably 1 column width (8.7cm).
-Figures wider than 1 column should be sized to 11.4cm or 17.8cm wide.
-Numbers, letters, and symbols should be no smaller than 6 points (2mm) and no larger than 12 points (6mm) after reduction and must be consistent.
-
-%% Figures and tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.
-
-%% Figure \ref{fig:dacca-fit} shows an example of how to insert a column-wide figure.
-To insert a figure wider than one column, please use the \verb|\begin{figure*}...\end{figure*}| environment.
-Figures wider than one column should be sized to 11.4 cm or 17.8 cm wide.
-Use \verb|\begin{SCfigure*}...\end{SCfigure*}| for a wide figure with side legends.
-
-%\begin{SCfigure*}[\sidecaptionrelwidth][t!]
-%\centering
-%\includegraphics[width=11.4cm,height=11.4cm]{frog.pdf}
-%\caption{This legend would be placed at the side of the figure, rather than below it.}\label{fig:side}
-%\end{SCfigure*}
-
-%% \subsection*{Tables}
-%% Tables should be included in the main manuscript file and should not be uploaded separately.
-
-%% \subsection*{Single column equations}
-
-%% Authors may use 1- or 2-column equations in their article, according to their preference.
-
-%% To allow an equation to span both columns, use the \verb|\begin{figure*}...\end{figure*}| environment mentioned above for figures.
-
-%% Note that the use of the \verb|widetext| environment for equations is not recommended, and should not be used.
-
-%% \begin{figure*}[bt!]
-%% \begin{align*}
-%% (x+y)^3&=(x+y)(x+y)^2\\
-%%        &=(x+y)(x^2+2xy+y^2) \numberthis \label{eqn:example} \\
-%%        &=x^3+3x^2y+3xy^3+x^3.
-%% \end{align*}
-%% \end{figure*}
-
-%\subsection*{Supporting Information Appendix (SI)}
-
-% Authors should submit SI as a single separate SI Appendix PDF file, combining all text, figures, tables, movie legends, and SI references.
-SI will be published as provided by the authors; it will not be edited or composed.
-Additional details can be found in the \href{https://www.pnas.org/authors/submitting-your-manuscript#manuscript-formatting-guidelines}{PNAS Author Center}.
-The PNAS Overleaf SI template can be found \href{https://www.overleaf.com/latex/templates/pnas-template-for-supplementary-information/wqfsfqwyjtsd}{here}.
-Refer to the SI Appendix in the manuscript at an appropriate point in the text.
-Number supporting figures and tables starting with S1, S2, etc.
-
-% Authors who place detailed materials and methods in an SI Appendix must provide sufficient detail in the main text methods to enable a reader to follow the logic of the procedures and results and also must reference the SI methods.
-If a paper is fundamentally a study of a new method or technique, then the methods must be described completely in the main text.
-
-% \subsubsection*{SI Datasets}
-
-% Supply .xlsx, .csv, .txt, .rtf, or .pdf files. This file type will be published in raw format and will not be edited or composed.
-
-\arxiv{}{
+\arxiv{
+\section*{Acknowledgments}
+This research was supported by National Science Foundation grant DMS-1761603. 
+We acknowledge communications with Nicholas Chopin. 
+}{
 \showmatmethods{} % Display the Materials and Methods section
@@
 \section{MOP-$\alpha$ Functional Forms}
+\label{appendix:functional}
 \input{appendix/functional}
@@
 \section{Optimization Convergence Analysis}
+\label{appendix:convergence}
 \input{appendix/convergence}
@@
 \section{Feynman-Kac Models and Monte Carlo Approximations}
+\label{appendix:feynman}
 \input{appendix/feynman}
@@
 \section{A Strong Law of Large Numbers for Triangular Arrays of Particles With ``Off-Parameter’’ Resampling}
+\label{appendix:targeting}
 \input{appendix/targeting}
@@
 \section{Consistency of Off-Parameter Resampled Gradient Estimates}
+\label{appendix:consistency}
 \input{appendix/consistency}
@@
 \section{Bias-Variance Analysis}
+\label{appendix:biasvar}
 \input{appendix/biasvar}
@@
 \section{Figures for Bayesian Inference}
+\label{appendix:bayes}
 \input{appendix/bayes}
+
 \end{document}

### 2024-06-13 paper/appendix/bayes.tex
many edits, mostly near the start

@@
     \centering
-    \includegraphics[width=\textwidth/\real{1.25}]{imgs/pmcmc/mh.png}
+    \includegraphics[width=\textwidth/\real{1.25}]{../imgs/pmcmc/mh.png}
     \caption{Convergence diagnostics for the Metropolis-Hastings variant particle MCMC with the random walk proposal and an informative empirical prior from IF2.
@@
     \centering
-    \includegraphics[width=\textwidth/\real{1.1}]{imgs/pmcmc/nuts.png}
+    \includegraphics[width=\textwidth/\real{1.1}]{../imgs/pmcmc/nuts.png}
     \caption{Convergence diagnostics for a No-U-Turn Sampler (NUTS) with uniform priors on a compact set.
@@
     \centering
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Recovery Rate.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Death Rate.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Reciprocal of Mean Immunity Duration.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Trend in Force of Infection.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Noise in Force of Infection.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Measurement Noise.png}
+    \includegraphics[scale=0.33]{../imgs/pmcmc/nuts_eb/Recovery Rate.png}
+    \includegraphics[scale=0.33]{../imgs/pmcmc/nuts_eb/Death Rate.png}
+    \includegraphics[scale=0.33]{../imgs/pmcmc/nuts_eb/Reciprocal of Mean Immunity Duration.png}
+    \includegraphics[scale=0.33]{../imgs/pmcmc/nuts_eb/Trend in Force of Infection.png}
+    \includegraphics[scale=0.33]{../imgs/pmcmc/nuts_eb/Noise in Force of Infection.png}
+    \includegraphics[scale=0.33]{../imgs/pmcmc/nuts_eb/Measurement Noise.png}
     \caption{Posterior estimates from NUTS powered by MOP-$\alpha$ with the informative nonparametric empirical prior obtained from a kernel density estimator on the IF2 parameter cloud.

### 2024-06-13 paper/ms.tex
many edits, mostly near the start

@@
 
-\title{Accelerated Inference for Partially Observed Stochastic Processes using Automatic Differentiation}
+\title{Accelerated Inference for Partially Observed Markov Processes using Automatic Differentiation}
 
@@
 Plug-and-play methods are often called likelihood-free \cite{owen15}, and it may be counter-intuitive that likelihood-based inference is possible using likelihood-free methods. 
-However, iterated filtering algorithms have shown that this is practical in various scientific investigations \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
+However, iterated filtering algorithms have shown that this is practical in various scientific investigations, for example \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
 Nevertheless, the Monte Carlo variability arising in iterated filtering applications becomes increasingly problematic as the size of the data and the complexity of the model increases.
 Algorithmic advances are needed to make plug-and-play likelihood maximization for POMP models more numerically efficient.
-We investigate the use of automatic differentiation (AD) for this purpose.
+We develop plug-and-play AD methods which provide a large computational advantage over the methods used in these previous applications.
 
-%\subsection{Automatic Differentiation for Particle Filters}
-
-Recent advances in AD for particle filters \cite{naesseth18, jonschkowski18, corenflos21, scibior21, singh22} have drawn attention to AD as a tool for inference in partially observed stochastic processes.
+Recent advances in AD for particle filters \cite{naesseth18, jonschkowski18, corenflos21, scibior21, singh22} have drawn attention to AD as a tool for inference in POMP models.
 However, existing approaches are either asymptotically biased \cite{naesseth18, jonschkowski18}, have high variance \cite{poyiadjis11, scibior21}, high computational expense \cite{corenflos21, chen24}, or need access to transition densities \cite{poyiadjis11, scibior21, singh22, chen24}.
@@
 Scibior and Wood \cite{scibior21} showed that the estimators derived by Poyiadjis et al.
-\cite{poyiadjis11} can be attained with standard AD software, using an algorithmic procedure called a ``stop-gradient'' which allows selected expressions to be evaluated but not differentiated.
-Unfortunately, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability, while the ``stop-gradient'' procedure is introduced without mathematical motivation.
+\cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.
+Unfortunately, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability, while the stop-gradient procedure is introduced without mathematical motivation.
 We use \cite{scibior21} and \cite{poyiadjis11} as our starting point, while developing a new approach which remedies the weaknesses of these papers.
@@
 We start by presenting a new construction of the gradient estimator of \cite{scibior21} and \cite{poyiadjis11}.
-Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a Measurement Off-Parameter (MOP) particle filter.
+Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a {\it Measurement Off-Parameter} (MOP) particle filter.
 We proceed to generalize the MOP particle filter by adding a discounting factor, $\alpha \in [0,1]$, which interpolates between the  biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
@@
 Directly differentiating through a particle filter, for a fixed value of $\omega$, leads to a biased gradient estimator due to the discreteness of the resampling operation \cite{corenflos21}.
-We avoid this issue by constructing a particle filtering algorithm called {\it Measurement Off-Parameter} with discount factor $\alpha$ (MOP-$\alpha$), defined by the pseudocode in Algorithm~\ref{alg:mop}.
+We avoid this issue by constructing a particle filtering algorithm called {\it Measurement Off-Parameter with discount factor} $\alpha$ (MOP-$\alpha$), defined by the pseudocode in Algorithm~\ref{alg:mop}.
 In short, instead of differentiating through a Monte Carlo algorithm directly, we instead use AD to differentiate through a series of measurement density ratios and a differentiable simulator.
+Note that the plug-and-play property forbids access to transition densities but permits access to the density of the measurements conditional on the value of the latent process.
+The simulator can be differentiated using AD as long as the underlying computer code for the simulator is differentiable, and so MOP-$\alpha$ is applicable to a broad class of POMP models.
 
-This algorithm is an example of a new class of algorithms we call ``off-parameter'' particle filters that resample the particles according to an arbitrary resampling rule that is not necessarily in proportion to the target distribution of interest, while maintaining weights that encode the cumulative discrepancy between the resampling distribution and the target distribution.
-We later show in Theorem \ref{thm:mop-targeting} that off-parameter particle filters successfully target the filtering distribution when the correction is sufficient.
+The MOP-$\alpha$ algorithm resamples the particles according to an arbitrary resampling rule that depends on both the target parameter value and a baseline parameter value, and this explains the name {\it measurement off-parameter}.
+Similar quantities are called {\it off-policy} in reinforcement learning, and a parameter in a POMP model is analogous to a policy in a corresponding control model \ed{Kevin, would it be good to add your favorite RL reference here?
+Also, you may want to edit this sentence.
+I think something like this will help the reader to ``get'' our terminology.}.
+Specifically, MOP-$\alpha$ evaluates the likelihood at some $\theta \in \Theta$, but instead resamples the particles according to the indices generated by a vanilla particle filter run at some $\phi \in \Theta$.
+This ensures that the resampling indices are invariant to $\theta$ when $\omega$ and $\phi$ are fixed, bypassing the issue of Monte Carlo resampling.
+We show in Theorem~\ref{thm:mop-targeting} that MOP-$\alpha$ exactly targets the filtering distribution when $\theta=\phi$ or $\alpha=1$.
 
-In the case of MOP-$\alpha$, we wish to evaluate the likelihood at some $\theta \in \Theta$, but instead resample the particles according to the indices generated by a vanilla particle filter run at some $\phi \in \Theta$.
-This ensures that the resampling indices are invariant to $\theta$ when $\omega$ and $\phi$ are fixed, bypassing the issue of Monte Carlo resampling.
 When the measurement model is differentiable in $\theta$ and the simulator is a differentiable function of $\theta$ for every fixed $\omega$ (a condition that requires the latent process to be a continuous random variable, also known as the reparameterization trick \cite{corenflos21}), direct differentiation of MOP-$\alpha$ is available via AD.
-As we explain subsequently, MOP-$\alpha$ is constructed so this direct derivative obtains the score estimator of  \cite{poyiadjis11, scibior21} when $\alpha=1$ and that of \cite{naesseth18} when $\alpha=0$.
+Theorem~\ref{thm:mop-functional-forms} shows that MOP-$\alpha$ is constructed so this direct derivative obtains the score estimator of  \cite{poyiadjis11, scibior21} when $\alpha=1$ and that of \cite{naesseth18} when $\alpha=0$.
 Setting $\alpha<1$ adds bias but reduces variance, raising opportunities for a favorable tradeoff as illustrated in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).
@@
 \begin{align*}
-    dS&=(k \epsilon R^k+\delta(S-P)-\lambda(t) S)\, dt+d P-({\sigma S I}/{H})\, dB, \\
+    dS&=\big(k \epsilon R^k+\delta(S-P)-\lambda(t) S\big)\, dt+d P-({\sigma S I}/{H})\, dB, \\
     dI&=\left(\lambda(t) S-(m+\delta+\gamma) I\right)\, dt+({\sigma S I}/{H})\, dB, \\
-    dR^1&=(\gamma I-(k \epsilon+\delta) R^1)\, dt, \hspace{2mm} \dots \\
-    dR^k&=(k \epsilon R^{k-1}\hspace{-3mm}-(k \epsilon+\delta) R^k)\, dt,
+    dR^1&=\big(\gamma I-(k \epsilon+\delta) R^1\big)\, dt, \hspace{2mm} \dots \\
+    dR^k&=\big(k \epsilon R^{k-1} -(k \epsilon+\delta) R^k\big)\, dt,
     \vspace*{-2mm}
@@
     \centering
-    \includegraphics[width=\arxiv{12cm}{\textwidth/2}]{../imgs/095/tikzcholera.png}
+    \includegraphics[width=\arxiv{14cm}{\textwidth/2}]{../imgs/095/tikzcholera.png}
     \vspace*{-7mm}
-    \caption{Illustration of Dhaka cholera model from \cite{king08}.}
+    \caption{A compartment flow diagram for the Dhaka cholera model from \cite{king08}.}
     \label{fig:tikz-cholera}
@@
 Meanwhile, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.6$, while the best log-likelihood found by \cite{king08} was only $-3793.4$.
-Despite being initialized for a global search, IFAD manages to get much closer to the MLE over the 100 searches than \cite{ionides15} and finds it up to Monte Carlo error, as seen in Table \ref{table:mle}.
+Despite being initialized for a global search, IFAD manages to get much closer to the MLE over the 100 searches than \cite{ionides15} and finds it up to Monte Carlo error, as seen in Table~\ref{table:mle}.
 On this problem, the sequence of local searches and refinement that was previously required for finding the MLE is not necessary with the IFAD algorithm. 
+In other words, the improvement in numerical efficiency of IFAD over IF2 is so large that routine application of IFAD (consisting of a collection of Monte Carlo replications from random starting points) generates results outside the reach of a routine application of IF2.
+When interpreting Table~\ref{table:mle}, bear in mind that a difference of order 1 log unit has statistical relevance and therefore scientific value.
 
@@
 
-We emphasize that we chose to not include the use of many common heuristics used in the machine learning and optimization literature in the gradient descent stage.
+For the results presented here, we did not include the many common heuristics used in the machine learning and optimization literature in the gradient descent stage.
 We used constant learning rates of $0.01, 0.05$, and $0.2$ for IFAD-$1,0,$ and $0.97$ respectively, and a constant cooling rate of $0.95$ for our IF2 implementation.
@@
 
-Particle MCMC, as introduced by \cite{andrieu10}, is arguably the most popular method for full-information Bayesian inference within partially observed Markov processes that does not require access to the system's transition densities.
-This feature enables its use in simulation-based Bayesian inference for complex scientific models where computing transition densities is not feasible, such as in disease modeling.
+Particle MCMC, as introduced by \cite{andrieu10}, is arguably the most popular method for full-information plug-and-play Bayesian inference for POMP models.
+The plug-and-play property enables its use in simulation-based Bayesian inference for complex scientific models where computing transition densities is not feasible, such as in disease modeling.
 However, the particle Metropolis-Hastings algorithm often experiences slow mixing and requires long burn-in periods.
-For instance, \cite{fasiolo16} reported needing $700,000$ burn-in iterations for effective sampling in a cholera model -- even after adjusting the length of each Euler timestep to be a month instead of a day to reduce computational costs.
+For instance, \cite{fasiolo16} reported needing $700,000$ burn-in iterations for effective sampling in a cholera model.
+This computational burden necessitated simplifying the scientific model to reduce computational costs, by adjusting the length of each Euler timestep to be a month instead of a day.
 
 We employ a NUTS sampler powered by a MOP-$\alpha$, with a nonparametric empirical prior initialized by the IF2 warm start from the previous section.
-To construct the prior, we perform a kernel density estimate (KDE) on the parameter swarm from the warm start, and use the output of that as an empirical prior (the KDE is performed to ensure some degree of mutual contiguity between the densities).
+To construct the prior, we perform a kernel density estimate (KDE) on the parameter swarm from the warm start, and use the output of that as an empirical prior.
+The KDE is performed to ensure some degree of mutual contiguity between the densities.
 This yields a prior that is roughly equivalent to a parameter cloud centered at some point in the neighborhood of the MLE, with a standard deviation of $\sqrt{(0.02\cdot0.95^{40})^2 \cdot 600}\approx 6.3\%$ of the MLE.
@@
     \centering
-    \includegraphics[width=\textwidth/\real{1.25}]{imgs/pmcmc/nuts_eb.png}
-    \caption{Convergence diagnostics for a No U-Turn Sampler with the informative nonparametric empirical Bayes prior from IF2 and a KDE.
+    \includegraphics[width=\textwidth/\real{1.25}]{../imgs/pmcmc/nuts_eb.png}
+    \caption{Convergence diagnostics for NUTS with the informative nonparametric empirical Bayes prior from applying KDE to preliminary searches using IF2.
 We display the results for the trend parameter in the Dhaka cholera model of \cite{king08}.
@@
 
-For completeness, we note that without the empirical Bayes-style prior from IF2, NUTS with the MOP-$\alpha$ estimate alone tends to diverge.
+For completeness, we note that without the empirical Bayes-style prior from IF2, NUTS with the MOP-$\alpha$ estimate alone tended to diverge.
 In contrast, particle Metropolis-Hastings (with or without the empirical Bayes prior from IF2) fails to effectively explore the parameter space.
 We display the results of the above in Figures \ref{fig:mh} and \ref{fig:nuts} \arxiv{in Appendix~\ref{appendix:bayes}}{within the supplementary information}. 
-
 The result obtained in this section, of a speedup in the mixing time of particle MCMC by over three orders of magnitude, provides hope that particle Bayesian inference might finally be practical for the complex scientific models commonly encountered in areas like epidemiology.
@@
 While the result is presented here as specific to MOP-$\alpha$, we actually prove a more general result in \arxiv{Appendix~\ref{appendix:targeting}}{the supplementary material}.
-That is, we show a strong law of large numbers for triangular arrays of particles with ``off-parameter'' resampling, where we resample the particles according to an arbitrary resampling rule not necessarily in proportion to the targeted distribution of interest and employ weights that encode the cumulative discrepancy between the resampling and the target distribution instead of equal weights.
+That is, we show a strong law of large numbers for triangular arrays of particles with off-parameter resampling, where we resample the particles according to an arbitrary resampling rule not necessarily in proportion to the targeted distribution of interest and employ weights that encode the cumulative discrepancy between the resampling and the target distribution instead of equal weights.
 

### 2024-06-13 paper/appendix/convergence.tex
updates from overleaf

@@
 From the decomposition in the proof of Lemma \ref{lem:mop-1-formula}, as $w_{N, j}^{F, \theta}=1$ when $\theta=\phi$, we have that
-$$
+\begin{equation}
 \nabla_\theta \hat{\ell}(\theta):=\frac{1}{J} \sum_{j=1}^J \nabla_\theta w_{N, j}^{F, \theta}=\frac{1}{J} \sum_{j=1}^J\sum_{n=1}^N  w_{N, k_j}^{P, \theta} \nabla_\theta \log g_{n,k_j}^{A,\theta} = \frac{1}{J} \sum_{j=1}^J\sum_{n=1}^N \nabla_\theta \log g_{n,k_j}^{A,\theta}.
-$$
+\end{equation}
 Define $\varphi_n^i(x_{n,j}^{F,\theta}) := \frac{\partial}{\partial\theta_i} \log g_{n,k_j}^{A,\theta}$, which is a functional of the filtering particles $x_{n,j}^{F,\theta} = x_{n,k_j}^{P,\theta}$.
@@
 We therefore require \ed{WE SAY ``REQUIRE'' AT VARIOUS POINTS, BUT THAT SUGGESTS NECESSITY WHEREAS I SUPPOSE WE MEAN SUFFICIENCY. I EDITED ONE OR TWO OF THESE.}
-$$J > \max\left\{\frac{2r_t(1+h^{-1}(t)) + 2c}{\gamma'}, \frac{2(2t\beta_t^2+c)^2}{\gamma'^2}\right\} \geq  \frac{r_t(1+h^{-1}(t))}{\gamma'} + \sqrt{2tJ}\beta_t/\gamma' + c/\gamma'$$
+\begin{equation}
+J > \max\left\{\frac{2r_t(1+h^{-1}(t)) + 2c}{\gamma'}, \frac{2(2t\beta_t^2+c)^2}{\gamma'^2}\right\} \geq  \frac{r_t(1+h^{-1}(t))}{\gamma'} + \sqrt{2tJ}\beta_t/\gamma' + c/\gamma'    
+\end{equation}
 for $\hat{H}(\theta)$ to be invertible and positive definite with minimum eigenvalue greater than or equal to $c$ with probability at least $1-\exp(-t)$.

### 2024-06-13 paper/appendix/feynman.tex
updates from overleaf

@@
 
-\begin{itemize}
-\item \textbf{Markov kernels and the process model:} A Markov kernel $M$ with source $\gX_1$ and target $\gX_2$ is a map $M: \gX_1 \times \mathcal{B}(\gX_2) \to [0,1]$ such that for every set $A \in \mathcal{B}(\gX_2)$ and every point $x \in \gX_1$, the map $x \mapsto M(x, A)$ is a measurable function of $x$, and the map $A \mapsto M(x,A)$ is a probability measure on $\gX_2$.
+\paragraph{Markov kernels and the process model:} A Markov kernel $M$ with source $\gX_1$ and target $\gX_2$ is a map $M: \gX_1 \times \mathcal{B}(\gX_2) \to [0,1]$ such that for every set $A \in \mathcal{B}(\gX_2)$ and every point $x \in \gX_1$, the map $x \mapsto M(x, A)$ is a measurable function of $x$, and the map $A \mapsto M(x,A)$ is a probability measure on $\gX_2$.
 The quantity $M(x,A)$ can be thought of as the probability of transitioning to the set $A$ given that we are at the point $x$.
@@
 \item \textbf{Markov kernels and measures:} For any measure $\eta$, any Markov kernel $M$ on $\gX$, any point $x \in \gX$ and any measurable subset $A \subseteq \gX$, let 
-\begin{align*}
+\begin{align}
     \eta(h) &= \int h d\eta = \int h(x) \eta(dx), \\(\eta M)(A) &= \int \eta(dx)M(x,A), \\
     (Mh)(x) &= \int M(x, dy) h(y).
-\end{align*}
+\end{align}
 \item \textbf{Compositions of Markov kernels:} The composition of a Markov kernel $M_1$ with another Markov kernel $M_2$ is another Markov kernel, given by 
-$$(M_1M_2)(x, A) = \int M_1(x, dy) M_2(y, A).$$
-\item \textbf{Total variation distance:} The total variation distance between two measures $\mu$ and $\nu$ on $\gX$ is
-$$\|\mu-\nu\|_{\mathrm{TV}}=\sup _{\|h\|_{\infty} \leq 1 / 2}|\mu(\phi)-\nu(\phi)|=\sup _{\operatorname{osc}(h) \leq 1}|\mu(h)-\nu(h)|.$$
-\item \textbf{Dobrushin contraction:} The Dobrushin contraction coefficient $\beta_{\text{TV}}$ of a Markov kernel $M$ is given by
-$$\beta_{\mathrm{TV}}(M)=\sup _{x, y \in \gX}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}.$$
-\item \textbf{Potential functions and the measurement model:} A potential function $G : \gX \to [0,\infty)$ is a non-negative function of an element of the state space $x \in \gX$.
+\begin{equation}
+ (M_1M_2)(x, A) = \int M_1(x, dy) M_2(y, A).
+\end{equation}
+
+\paragraph{Total variation distance:} The total variation distance between two measures $\mu$ and $\nu$ on $\gX$ is
+\begin{equation}
+\|\mu-\nu\|_{\mathrm{TV}}=\sup _{\|h\|_{\infty} \leq 1 / 2}|\mu(\phi)-\nu(\phi)|=\sup _{\operatorname{osc}(h) \leq 1}|\mu(h)-\nu(h)|.    
+\end{equation}
+
+\paragraph{Dobrushin contraction:} The Dobrushin contraction coefficient $\beta_{\text{TV}}$ of a Markov kernel $M$ is given by
+\begin{equation}
+\beta_{\mathrm{TV}}(M)=\sup _{x, y \in \gX}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}.    
+\end{equation}
+
+\paragraph{Potential functions and the measurement model:} A potential function $G : \gX \to [0,\infty)$ is a non-negative function of an element of the state space $x \in \gX$.
 In our case, this corresponds to the measurement model, and in our previous notation is written as $g_{n,j} = f_{Y_n|X_n}(y_n^*|x_{n,j}^F) = G_n(x_{n,j}^F)$, where in a slight abuse of notation we suppress the dependence on $\theta$ for notational simplicity.
 Note that $G_n(\cdot) = f_{Y_n|X_n}(y_n^*|\;\cdot\;)$ is the conditional density of the observed measurement at time $n$, where we condition on the filtering particle $x_{n,j}^F$ as an element of the state space.
-\item \textbf{Feynman-Kac models:} A Feynman-Kac model on $\gX$ is a tuple $(\pi_0, (M_n)_{n=1}^N, (G_n)_{n=1}^N)$ of an initial probability measure on the state space $\pi_0$, a sequence of transition kernels $(M_n)_{n=1}^N$, and a sequence of potential functions $(G_n)_{n=1}^N$.
+
+\paragraph{Feynman-Kac models:} A Feynman-Kac model on $\gX$ is a tuple $(\pi_0, (M_n)_{n=1}^N, (G_n)_{n=1}^N)$ of an initial probability measure on the state space $\pi_0$, a sequence of transition kernels $(M_n)_{n=1}^N$, and a sequence of potential functions $(G_n)_{n=1}^N$.
 In the notation used in the main text, this corresponds to the starting distribution $f_{X_0}$, the sequence of transition densities $f_{X_{n}|X_{n-1}}$, and the measurement densities $f_{Y_n|X_n}$.
@@
 \begin{itemize}
-    \item The update from the prediction to the filtering distributions is given by $$\pi_n(dx) = \Psi_n(\eta_n)(dx) = \frac{G_n(x)\cdot\eta_n(dx)}{\eta_n(G_n)}.$$
+    \item The update from the prediction to the filtering distributions is given by 
+    \begin{equation}
+    \pi_n(dx) = \Psi_n(\eta_n)(dx) = \frac{G_n(x)\cdot\eta_n(dx)}{\eta_n(G_n)}.
+    \end{equation}
     \item The map from the prediction distribution at timestep $n$ to timestep $n+1$ is given by 
-    $$\Phi_{n+1}(\eta_n) = \Psi_n(\eta_n) M_{n+1}.$$
+    \begin{equation}
+ \Phi_{n+1}(\eta_n) = \Psi_n(\eta_n) M_{n+1}.       
+    \end{equation}
     \item The composition of maps between prediction distributions yields the map from the prediction distribution at time $k$ to the prediction distribution at time $n$ where $k \leq n$,
-    $$\Phi_{k,n} = \Phi_n \circ ... \circ \Phi_{k+1}.$$
+    \begin{equation}
+ \Phi_{k,n} = \Phi_n \circ ... \circ \Phi_{k+1}.       
+    \end{equation}
 \end{itemize}
-\item \textbf{The particle filter:} The particle filter then yields a Monte Carlo approximation to the above Feynman-Kac model, via a sequence of mixture Dirac measures.
+\paragraph{The particle filter:} The particle filter then yields a Monte Carlo approximation to the above Feynman-Kac model, via a sequence of mixture Dirac measures.
 When one resamples at every timestep, the prediction measure at timestep $n$ is then given by
-$$\eta_n^J = \frac{1}{J}\sum_{j=1}^J \delta_{x_{n,j}^P},$$
+\begin{equation}
+\eta_n^J = \frac{1}{J}\sum_{j=1}^J \delta_{x_{n,j}^P},
+\end{equation}
 and the filtering measure at timestep $n$ is given by
-$$\pi_n^J = \frac{\sum_{j=1}^J g_{n,j} \delta_{x_{n,j}^P}}{\sum_{j=1}^J g_{n,j}} \approx \frac{1}{J} \sum_{j=1}^J \delta_{x_{n,j}^F}.$$
+\begin{equation}
+\pi_n^J = \frac{\sum_{j=1}^J g_{n,j} \delta_{x_{n,j}^P}}{\sum_{j=1}^J g_{n,j}} \approx \frac{1}{J} \sum_{j=1}^J \delta_{x_{n,j}^F}.
+\end{equation}
 In a slight abuse of notation, we will identify $x_{n, 1:J}^P \equiv \eta_n^J$, and $x_{n, 1:J}^F \equiv \pi_n^J$. 
 As in \cite{karjalainen23}, one can view this as an inhomogenous Markov process evolving on $\gX^{J}$. The corresponding Markov transition kernel is then 
-$$\textbf{M}_n(x_{n-1, 1:J}^P, \cdot) = \left(\Phi_{n}\left(\eta_n^J\right)\right)^{\otimes J} = \left(\Phi_{n}\left(\frac{1}{J}\sum_{j=1}^J \delta_{x_{n,j}^P}\right)\right)^{\otimes J},$$
+\begin{equation}
+\textbf{M}_n(x_{n-1, 1:J}^P, \cdot) = \left(\Phi_{n}\left(\eta_n^J\right)\right)^{\otimes J} = \left(\Phi_{n}\left(\frac{1}{J}\sum_{j=1}^J \delta_{x_{n,j}^P}\right)\right)^{\otimes J},
+\end{equation}
 and the composition of Markov kernels on particles from timestep $n$ to timestep $n+k$ is written 
-$$\textbf{M}_{n, n+k} = \textbf{M}_{n+k}\circ ...\circ \textbf{M}_n.$$
+\begin{equation}
+\textbf{M}_{n, n+k} = \textbf{M}_{n+k}\circ ...\circ \textbf{M}_n.
+\end{equation}
 One may wonder why \cite{karjalainen23} require this process to evolve on $\gX^{J}$.
 This is because at every timestep $n$, we in fact draw $X_{n, j}^P | \{X_{n-1, 1:J}^P = x_{n-1, 1:J}^P\} \sim \textbf{M}_n(x_{n-1, 1:J}^P, \cdot) = \eta_0^{\otimes J} \textbf{M}_{0,n}$ for $j=1,...,J$.
-\item \textbf{Forgetting of the particle filter:} 
+
+\paragraph{Forgetting of the particle filter:} 
 The above formalization yields a result from \cite{karjalainen23} that we require for our analysis of the bias, variance, and error of MOP-$\alpha$ -- the forgetting of the particle filter.
 That is, \cite{karjalainen23} show that
-$$\beta_{\mathrm{TV}}\left(\mathbf{M}_{n, n+k}\right) \leq(1-\epsilon)^{\lfloor k /(O(\log J))\rfloor},$$
+\begin{equation}
+\beta_{\mathrm{TV}}\left(\mathbf{M}_{n, n+k}\right) \leq(1-\epsilon)^{\lfloor k /(O(\log J))\rfloor},
+\end{equation}
 for some $\epsilon$ dependent on $\bar{G}, \underbar{G}, \bar{M}, \underbar{M}$ in Assumptions \ref{assump:bounded-measurement} and \ref{assump:bounded-process}.
 As a result, the mixing time of the particle filter is only on the order of $O(\log(J))$ timesteps.
-\end{itemize}
+
+\vspace{3mm}
 

### 2024-06-13 paper/appendix/functional.tex
updates from overleaf

@@
     Consider the case of MOP-$\alpha$ when $\alpha=1$ and $\theta=\phi$. We then have a nice telescoping product property for the after-resampling likelihood estimate:
-\begin{equation*}
+\begin{equation}
     \hat{\lik}^1(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}}= \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n-1,j}^{F,\theta}} = \left(\frac{1}{J}\sum_{j=1}^J w_{N,j}^{F,\theta}\right) \prod_{n=1}^N L_n^\phi,
-\end{equation*}
+\end{equation}
 where the third equality follows from the choice of $\alpha=1$, and the fourth equality is the resulting telescoping property. 
@@
 First, observe that as $\alpha=1$,
-$$w_{n,j}^{P,\theta} = w_{n-1,j}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}},$$
+\begin{equation}
+w_{n,j}^{P,\theta} = w_{n-1,j}^{F,\theta}\frac{g_{n,j}^\theta}{g_{n,j}^\phi} = \prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}},
+\end{equation}
 where we use the $(\cdot)^A$ superscript to denote the ancestral trajectory of the $j$-th prediction or filtering particle at timestep $n$. 
@@
 We then use the log-derivative identity again, yielding the following expression for the gradient of the log-weights as the sum of the log measurement densities over the ancestral trajectory:
-$$\frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} = \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}}\right) = \nabla_\theta \sum_{i=1}^n \left(\log g_{i,j}^{A,P,\theta} - \log g_{i,j}^{A,P,\phi}\right) = \sum_{i=1}^n \nabla_\theta \log g_{i,j}^{A,P,\theta}.$$
+\begin{eqnarray}
+ \frac{\nabla_\theta w_{n,j}^{P,\theta}}{w_{n,j}^{P,\theta}} = \nabla_\theta \log w_{n,j}^{P,\theta} &=& \nabla_\theta \log \left(\prod_{i=1}^n \frac{g_{i,j}^{A,P,\theta}}{g_{i,j}^{A,P,\phi}}\right) 
+ \\
+ &=& \nabla_\theta \sum_{i=1}^n \left(\log g_{i,j}^{A,P,\theta} - \log g_{i,j}^{A,P,\phi}\right)
+ \\
+ &=& \sum_{i=1}^n \nabla_\theta \log g_{i,j}^{A,P,\theta}.
+\end{eqnarray}
 This is equal to the gradient of the logarithm of the conditional density of the observed measurements given the ancestral trajectory of the $j$-th prediction particle up to timestep $n$:
-\begin{eqnarray*}  
+\begin{eqnarray}  
 \nabla_\theta \sum_{n=1}^N \log g_{n,j}^{A,\theta} &=& \nabla_\theta \log\left(\prod_{n=1}^N g_{n,j}^{A,P,\theta}\right) 
@@
 &=& \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).
-\end{eqnarray*}
+\end{eqnarray}
 Multiplying both sides of the expression by $w_{N,j}^{P,\theta} $ yields an expression for the gradient of the weights at timestep $N$:
-$$\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{n=1}^N \nabla_\theta \log g_{n,j}^{A,P,\theta} = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).$$
+\begin{equation}
+\nabla_\theta w_{N,j}^{P,\theta} = w_{N,j}^{P,\theta} \sum_{n=1}^N \nabla_\theta \log g_{n,j}^{A,P,\theta} = w_{N,j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, P,\theta}\right).
+\end{equation}
 Substituting the above identity into the log-likelihood decomposition obtained earlier \ed{USE EQUATION NUMBERS TO MAKE THIS ARGUMENT MORE CONCRETE} yields
-\begin{equation*}
+\begin{equation}
     \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right).
-\end{equation*}
+\end{equation}
 Finally, observing that $\theta=\phi$ implies $w_{N,j}^{F,\theta}=1$, we obtain 
-\begin{equation*}
+\begin{equation}
     \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right).
-\end{equation*}
+\end{equation}
 This yields the gradient estimators of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter. 
@@
 This can be seen by observing that
-\begin{equation*}
+\begin{equation}
     \nabla_\theta \hat{\ell}^1(\theta) := \frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,j}^{F,\theta} =\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{N,k_j}^{P,\theta} = \frac{1}{J}\sum_{j=1}^J w_{N,k_j}^{P,\theta} \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,k_j}^{A, P,\theta}\right).
-\end{equation*}
+\end{equation}
 When $\theta\neq\phi$, we see that $w_{N,k_j}^{P,\theta} = O(c^N)$. 
@@
 Observe that, when $\alpha=0,$, the likelihood estimate becomes
-\begin{equation}
-\nonumber
-    \hat{\lik}^0(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
-\end{equation}
+\begin{eqnarray}
+% \nonumber
+    \hat{\lik}^0(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} &=& \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} 
+    \\
+    &=& \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} 
+    \\
+    &=& \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
+\end{eqnarray}
 We lose the nice telescoping property observed in the MOP-$1$ case, but this expression still yields something useful. 
 This is because its gradient when $\theta=\phi$ is therefore 
-\begin{align*}
+\begin{align}
     \nabla_\theta \hat{\ell}^0(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
@@
     &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
-\end{align*}
+\end{align}
 where we use the log-derivative trick in the second equality, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$ in the fourth equality, and use the log-derivative trick again while noting that $\theta=\phi$ in the fifth equality.

### 2024-06-13 paper/appendix/targeting.tex
updates from overleaf

@@
 Here, we prove a more general result, from which it will be clear that MOP-$\alpha$ targets the filtering distribution. 
-We will prove a strong law of large numbers for triangular arrays of particles with ``off-parameter'' resampling, where we resample the particles according to an arbitrary resampling rule that is not necessarily in proportion to the target distribution of interest.
-Using weights that encode the cumulative discrepancy between the resampling distribution and the target distribution (instead of resampling to equal weights as is common) turns out to be a sufficient correction that ensures almost sure convergence.
+We will prove a strong law of large numbers for triangular arrays of particles with off-parameter resampling, meaning that we resample the particles according to an arbitrary resampling rule that is not necessarily in proportion to the target distribution of interest.
+Using weights that encode the cumulative discrepancy between the resampling distribution and the target distribution (instead of resampling to equal weights, as in the basic particle filter) provides a sufficient correction to ensure almost sure convergence.
 We now introduce the precise definition of an off-parameter resampled particle filter.
@@
         \item The prediction measure at timestep $n$ is given by 
-        $$\eta_n^J = \frac{\sum_{j=1}^J w_{n,j}^P \delta_{x_{n,j}^P}}{\sum_{j=1}^J w_{n,j}^P}.$$
+        \begin{equation}
+        \eta_n^J = \frac{\sum_{j=1}^J w_{n,j}^P \delta_{x_{n,j}^P}}{\sum_{j=1}^J w_{n,j}^P}.
+        \end{equation}
         \item The particles are resampled at every timestep $n$, yielding indices $k_j$ according to some arbitrary probabilities $(p_{n,j})_{j=1}^J$.
         \item the filtering measure at timestep $n$ is given by
-        $$\pi_n^J = \frac{\sum_{j=1}^J \delta_{x_{n,j}^P }w_{n,j}^P g_{n,j}}{\sum_{j=1}^J w_{n,j}^Pg_{n,j}} \approx \frac{\sum_{j=1}^J \delta_{x_{n,k_j}^P }w_{n,k_j}^P g_{n,k_j}/p_{n,k_j}}{\sum_{j=1}^J w_{n,k_j}^Pg_{n,k_j}/p_{n,k_j}} = \frac{\sum_{j=1}^J w_{n,j}^F \delta_{x_{n,j}^F}}{\sum_{j=1}^J w_{n,j}^F}.$$
+        \begin{equation}
+        \pi_n^J = \frac{\sum_{j=1}^J \delta_{x_{n,j}^P }w_{n,j}^P g_{n,j}}{\sum_{j=1}^J w_{n,j}^Pg_{n,j}} \approx \frac{\sum_{j=1}^J \delta_{x_{n,k_j}^P }w_{n,k_j}^P g_{n,k_j}/p_{n,k_j}}{\sum_{j=1}^J w_{n,k_j}^Pg_{n,k_j}/p_{n,k_j}} = \frac{\sum_{j=1}^J w_{n,j}^F \delta_{x_{n,j}^F}}{\sum_{j=1}^J w_{n,j}^F}.
+        \end{equation}
         \item The prediction weights at the next timestep are given by $w_{n+1,j}^P = w_{n,j}^F =  w_{n,k_j}^P g_{n,k_j}/p_{n,k_j}$.
@@
 Write
-    \[
+    \begin{equation}
     (Y_j^J,V_j^J) = \big(\tilde X^J_{k_j},U^J_{k_j}/\pi^J_{k_j}\big).
-    \] 
+    \end{equation} 
     If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j),j=1,\dots,J\}$ targets $f_X$.
@@
     Note that as the $Y_j^J$ are a subsample from $X_j^J$, $h$ can be a function of $Y$ as well as it is one for $X$. We then expand
-    $$\frac{\sum_j h(Y_j^J) V_j^J}{\sum_j V_j^J} = \frac{\sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}}.$$
+    \begin{equation}\frac{\sum_j h(Y_j^J) V_j^J}{\sum_j V_j^J} = \frac{\sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}}.\end{equation}
     By hypothesis,
-    $$\frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[f(X)].$$
+    \begin{equation}
+    \frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[f(X)].
+    \end{equation}
     We want to show
@@
     For this, it is sufficient to show that
-   $$ \sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}
-    -  \sum_j h(\tilde X_j^J)\frac{U_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0 $$
+   \begin{equation} 
+   \sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}
+    -  \sum_j h(\tilde X_j^J)\frac{U_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0 
+    \end{equation}
     since an application of this result with $h(x)=1$ provides almost sure convergence of the denominator in (\ref{eq:lemma1:h}).
     Write $g(\tilde X_j^J) = h(\tilde X_j^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}$. We therefore need to show that 
-    $$\sum_j Z_j^J := \sum_j \left(g(\tilde X_{k_j}^J) -  g(\tilde X_j^J) \pi_j^J \right) \stackrel{a.s.}{\to} 0.$$
-
+    \begin{equation}
+    \sum_j Z_j^J := \sum_j \left(g(\tilde X_{k_j}^J) -  g(\tilde X_j^J) \pi_j^J \right) \stackrel{a.s.}{\to} 0.
+\end{equation}
     Because the functional $h$ and importance sampling weights $u_{k_j}^J/\pi_{k_j}^J$ are bounded, we have that $\E[\left(Z_j^J\right)^4] < \infty$.
@@
     Taking expectations on both sides yields
-    $$\E\left[\left(\sum_j Z_j^J\right)^4 \right]  \leq CJ^2$$
+    \begin{equation}\E\left[\left(\sum_j Z_j^J\right)^4 \right]  \leq CJ^2\end{equation}
     by the tower property. Now by Markov's inequality, 
-    $$\mathbb{P}\left(\left|\frac{1}{J} \sum_{j=1}^J Z_j^J\right|>\epsilon \right) 
+    \begin{equation}\mathbb{P}\left(\left|\frac{1}{J} \sum_{j=1}^J Z_j^J\right|>\epsilon \right) 
     \leq \frac{1}{\epsilon^4J^4 } 
-    \mathbb{E}\left[\left(\sum_{j=1}^J Z_j^J\right)^4\right] \leq \frac{C}{\epsilon^4J^2},$$
+    \mathbb{E}\left[\left(\sum_{j=1}^J Z_j^J\right)^4\right] \leq \frac{C}{\epsilon^4J^2},\end{equation}
     and as these terms are summable we can apply Borel-Cantelli to conclude that these deviations happen only finitely often for every $\epsilon>0,$ giving us the almost-sure convergence for   
-    $$ \sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}
-    -  \sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0.$$ 
+    \begin{equation} 
+    \sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}
+    -  \sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0.
+    \end{equation} 
     Similarly, we also have that
-    $$ \sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J
-    - \sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J} \stackrel{a.s.}{\to} 0,$$
+    \begin{equation} 
+    \sum_j \frac{u_j^J}{\pi_j^J}\pi_j^J
+    - \sum_j \frac{u_{k_j}^J}{\pi_{k_j}^J} \stackrel{a.s.}{\to} 0,
+    \end{equation}
     and the result is proved.
@@
     We want to show that, for any measurable bounded $h$, 
-     $$\frac{\sum_j h(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)] = \E_{f_X}\big[\E_{f_{Z|X}}[h(Z) | X]\big].$$
+     \begin{equation}
+     \frac{\sum_j h(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)] = \E_{f_X}\big[\E_{f_{Z|X}}[h(Z) | X]\big].
+     \end{equation}
     By assumption, for any measurable and bounded functional $g$ with domain $\gX$,
@@
     Let $\bar{U}_j^J = \frac{J U_j^J}{\sum_j U_j^J}$.
-Examine the numerator and denominator of the quantity $$\frac{J^{-1}\sum_j h(\tilde Z_j^J) \, U_j^J}{J^{-1}\sum_j U_j^J} = \frac{J^{-1}\sum_j h(\tilde Z_j^J) \, \bar{U}_j^J}{J^{-1}\sum_j \bar{U}_j^J}.$$
+Examine the numerator and denominator of the quantity \begin{equation}\frac{J^{-1}\sum_j h(\tilde Z_j^J) \, U_j^J}{J^{-1}\sum_j U_j^J} = \frac{J^{-1}\sum_j h(\tilde Z_j^J) \, \bar{U}_j^J}{J^{-1}\sum_j \bar{U}_j^J}.\end{equation}
     The denominator converges to $1$ almost surely. The numerator, on the other hand, is
-    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J,$$
+    \begin{equation}
+    \frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J,
+    \end{equation}
     and by the same fourth moment argument to the above lemma, it converges almost surely to the limit of its expectation,
-    \begin{eqnarray*}        
+    \begin{eqnarray}        
     \lim_{J\to\infty} \E\left[\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J\right] 
@@
     &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  \E\big[h(Z) \big| X=\tilde X_j^J \big] \, \bar U_j^J\right].
-    \end{eqnarray*}
+    \end{eqnarray}
     Applying (\ref{eq:lemma2:g}) with $g(x) = \E\big[h(Z) | X=x\big]$, the average on the right hand side converges almost surely to $\E\big\{\E[h(Z)|X]\big\}=\E[h(Z)].$
     It remains to swap the limit and expectations. We can do so with the bounded convergence theorem, and therefore obtain   
-    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\,\bar{U}_j^J \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)].$$ 
+    \begin{equation}\frac{1}{J}\sum_j h(\tilde Z_j^J)\,\bar{U}_j^J \stackrel{a.s.}{\to} \E_{f_Z}[h(Z)].\end{equation} 
     \end{proof}
@@
     Again, we want to show that
-     $$\frac{\sum_j h(X_j^J) \cdot U_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j U_j^J \cdot f_{Z|X}(z^*|X_j^J)} \stackrel{a.s.}{\to} \E_{f_{X|Z}}[h(X)|z^*].$$
+     \begin{equation}\frac{\sum_j h(X_j^J) \cdot U_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j U_j^J \cdot f_{Z|X}(z^*|X_j^J)} \stackrel{a.s.}{\to} \E_{f_{X|Z}}[h(X)|z^*].\end{equation}
     We already have that for any measurable bounded $g$,
@@
     We will apply Equation (\ref{eq:lemma-posterior-hypothesis}) to the numerator and the denominator in the ratio above individually. The numerator converges to 
-    $$\frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \stackrel{a.s.}{\to} \E_{f_X}[h(X)f_{Z|X}(z^*|X)],$$
+    \begin{equation}
+    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \stackrel{a.s.}{\to} \E_{f_X}[h(X)f_{Z|X}(z^*|X)],
+    \end{equation}
     while the reciprocal of the denominator converges to 
-    $$ \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}  \stackrel{a.s.}{\to} \E_{f_X}[f_{Z|X}(z^*|X)] = f_Z(z^*).$$
+    \begin{equation} 
+    \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}  \stackrel{a.s.}{\to} \E_{f_X}[f_{Z|X}(z^*|X)] = f_Z(z^*).
+    \end{equation}
     Now we take advantage of the identities
-    $$\frac{\E_{f_X}[h(X)f_{Z|X}(z^*|X)]}{f_Z(z^*)} = \E_{f_X}\left[\frac{h(X)f_{Z|X}(z^*|X)}{f_Z(z^*)}\right] = \E_{f_X}\left[h(X)\frac{f_{X|Z}(X|z^*)}{f_X(X)}\right] =  \E_{f_{X|Z}}[h(X)|z^*],$$
+    \begin{eqnarray}
+    \frac{\E_{f_X}[h(X)f_{Z|X}(z^*|X)]}{f_Z(z^*)} &=& \E_{f_X}\left[\frac{h(X)f_{Z|X}(z^*|X)}{f_Z(z^*)}\right] 
+    \\
+    &=& \E_{f_X}\left[h(X)\frac{f_{X|Z}(X|z^*)}{f_X(X)}\right] =  \E_{f_{X|Z}}\big[h(X)|z^*\big],
+    \end{eqnarray}
     to give the desired result.
@@
 Define the weights $U^{' J}_j := U_j^J \cdot f_{Z|X}(z^*|X_j^J),$ and the corresponding self-normalized weights $\bar{U}_j^{' J} := \frac{J U^{' J}_j}{\sum_j U^{' J}_j}.$ The denominator of the below quantity
-    % $$ \frac{J^{-1} \sum_j h(X_j^J) \bar{U}_j^{' J}}{J^{-1} \sum_j \bar{U}_j^{' J}}$$
+    % \begin{equation} \frac{J^{-1} \sum_j h(X_j^J) \bar{U}_j^{' J}}{J^{-1} \sum_j \bar{U}_j^{' J}}\end{equation}
     % converges to $1$, while the numerator, by the same fourth moment argument as the first lemma, converges almost surely to the limit of its expectation, which is
@@
 %     From Lemma \ref{lem:marginal-proper-weights}, for any measurable bounded $g$, 
-%      $$\frac{\sum_j g(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_Z}[g(Z)] = \E_{f_X}\big[\E_{f_{Z|X}}[g(Z) | X]\big],$$
+%      \begin{equation}\frac{\sum_j g(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_Z}[g(Z)] = \E_{f_X}\big[\E_{f_{Z|X}}[g(Z) | X]\big],\end{equation}
 %      and by assumption, for any measurable bounded $g$, 
-%     $$\frac{\sum_j g(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[g(X)].$$
+%     \begin{equation}\frac{\sum_j g(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[g(X)].\end{equation}
 
 %     We in fact want to show that 
-%     $$\frac{\sum_j h(\tilde Z_j^J) \, V_j^J}{\sum_j V_j^J} \stackrel{a.s.}{\to} \E_{f_Y}[h(Y)] = \E_{f_X}\left[\E_{f_{Z|X}}\left[h(Z)\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)} \Bigg| X\right]\right].$$
+%     \begin{equation}\frac{\sum_j h(\tilde Z_j^J) \, V_j^J}{\sum_j V_j^J} \stackrel{a.s.}{\to} \E_{f_Y}[h(Y)] = \E_{f_X}\left[\E_{f_{Z|X}}\left[h(Z)\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)} \Bigg| X\right]\right].\end{equation}
 
@@
 %     We will apply Lemma 2 to the numerator and the denominator in the ratio above individually. The numerator converges to 
-%     $$\frac{J^{-1} \sum_j h(Z_j^J)  \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)} {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}} \stackrel{a.s.}{\to} \E_{f_Z}\left[h(Z) \frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\right] = \E_{f_X}\left[\E_{f_{Z|X}}\left[h(Z)\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right],$$
+%     \begin{equation}\frac{J^{-1} \sum_j h(Z_j^J)  \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)} {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}} \stackrel{a.s.}{\to} \E_{f_Z}\left[h(Z) \frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\right] = \E_{f_X}\left[\E_{f_{Z|X}}\left[h(Z)\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right],\end{equation}
 %     while the reciprocal of the denominator converges to 
-%     $$ \frac{J^{-1} \sum_j \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)}{U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}  \stackrel{a.s.}{\to} \E_{f_Z}\left[ \frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\right] = \E_{f_X}\left[\E_{f_{Z|X}}\left[\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right] .$$
+%     \begin{equation} \frac{J^{-1} \sum_j \frac{f_{Y|X}(\tilde Z_j^J|\tilde X_j^J)}{f_{Z|X}(\tilde Z_j^J|\tilde X_j^J)}{U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}  \stackrel{a.s.}{\to} \E_{f_Z}\left[ \frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\right] = \E_{f_X}\left[\E_{f_{Z|X}}\left[\frac{f_{Y|X}(Z|X)}{f_{Z|X}(Z|X)}\Bigg|X\right]\right] .\end{equation}
 
@@
     Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
-    \[
+    \begin{equation}
     \big(X^{F}_{n,j},w^{F}_{n,j}\big) = \big(X^{P}_{n,k_j}, w^{P}_{n,k_j} g_{n,k_j}\big/ p_{n,k_j}\big),
-    \]
+    \end{equation}
     with resampling probabilities proportional to $p_{n,j}$, therefore also targets $\pi_n$.
@@
     Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
-    \[
+    \begin{equation}
     \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,k_j}, w^{P,\theta}_{n,k_j} g^\theta_{n,k_j}\big/ g^\phi_{n,k_j}\big),
-    \]
+    \end{equation}
     with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
@@
 %     Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
-%     \[
+%     \begin{equation}
 %     \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \left(X^{P,\phi}_{n,a(j)}, w^{P,\theta}_{n,a(j)} \frac{g^\theta_{n,a(j)}h_{n,a(j)}^\theta}{g^\phi_{n,a(j)}h_{n,a(j)}^\phi}\right),
-%     \]
+%     \end{equation}
 %     with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
@@
 
-Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets\\
+$f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
 To see  (\ref{L2}),
 we write the numerator of (\ref{lem:change-measure-proper-weights}) as
-\[
+\begin{equation}
 L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
 = L^\phi_n \sum_{j=1}^J w_{n,j}^{FC,\theta} \frac{g^\phi_{n,j}}{L_n^\phi}
-\]
+\end{equation}
 Using Lemma \ref{lem:change-measure-proper-weights}, we resample according to probabilities $\frac{g^\phi_{n,j}}{L_n^\phi}$ to see this is properly estimated by
-\[
+\begin{equation}
 L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},
-\]
+\end{equation}
 from which we obtain (\ref{L2}).
@@
 Using Lemma \ref{lem:lik-proper-weight}, we obtain a likelihood estimate,
-\[
+\begin{equation}
 L^{A,\theta} = \prod_{n=1}^N \left( L^\phi_n \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J w^{P,\theta}_{n,j}}\right).
-\]
+\end{equation}
 Since $w^{F,\theta}_{n,j}=w^{P,\theta}_{n+1,j}$, this is a telescoping product. The remaining terms are

### 2024-06-13 paper/ms.tex
updates from overleaf

@@
 Plug-and-play methods are often called likelihood-free \cite{owen15}, and it may be counter-intuitive that likelihood-based inference is possible using likelihood-free methods. 
-However, iterated filtering algorithms have shown that this is practical in various scientific investigations, for example \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
+However, iterated filtering algorithms have shown that this is practical in various scientific investigations, such as \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
 Nevertheless, the Monte Carlo variability arising in iterated filtering applications becomes increasingly problematic as the size of the data and the complexity of the model increases.
@@
 MOP-$\alpha$ avoids the discontinuity that is problematic for the standard particle filter in a novel way.
-This particle filter is smooth by construction, allowing us to bypass the seemingly incompatible paradox of differentiating through a Monte Carlo algorithm -- by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
+This particle filter is smooth by construction, allowing us to bypass the seemingly incompatible paradox of differentiating through discrete Monte Carlo sampling by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
 Yet, MOP-$\alpha$ can nevertheless provide a consistent log-likelihood estimate.
@@
 Critically, the estimator is effective within a neighborhood of the maximum but is relatively slow at reaching this neighborhood.
-That behavior is complementary to that of the iterated filtering algorithm \cite{ionides06-pnas,ionides15} which provides a relatively fast and stable way to identify this neighborhood.
+That behavior is complementary to that of iterated filtering algorithms \cite{ionides06-pnas,ionides15} which provides a relatively fast and stable way to identify this neighborhood.
 We therefore build a simple but effective hybrid algorithm called {\it Iterated Filtering with Automatic Differentiation} (IFAD) that warm-starts first-order or second-order gradient methods with a preliminary solution obtained from a few rounds of iterated filtering.
-Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.
+Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 \cite{ionides06-pnas,ionides11} and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.
 
@@
 
-We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ and we write $\process_n\left(x_n\mid x_{n-1}; \theta\right)$ for its corresponding simulator.
-We set $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ to be the measurement density, and $y_n^*$ for the actual values of the observations that were observed.
+We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$.
+The existence of this density is necessary to define the likelihood function, but plug-and-play algorithms do not have access to evaluation of this density.
+Instead, they have access to a corresponding simulator, which we write as $\process_n\left(x_n\mid x_{n-1}; \theta\right)$.
+We set $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ to be the measurement density, which we suppose can be evaluated. We write $y_n^*$ for the actual values of the observations that were observed.
 We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
@@
 I think something like this will help the reader to ``get'' our terminology.}.
-Specifically, MOP-$\alpha$ evaluates the likelihood at some $\theta \in \Theta$, but instead resamples the particles according to the indices generated by a vanilla particle filter run at some $\phi \in \Theta$.
+Specifically, MOP-$\alpha$ evaluates the likelihood at some $\theta \in \Theta$, but instead resamples the particles according to the indices generated by a vanilla particle filter run at a {\it baseline parameter}, $\phi \in \Theta$.
 This ensures that the resampling indices are invariant to $\theta$ when $\omega$ and $\phi$ are fixed, bypassing the issue of Monte Carlo resampling.
@@
 
-When the measurement model is differentiable in $\theta$ and the simulator is a differentiable function of $\theta$ for every fixed $\omega$ (a condition that requires the latent process to be a continuous random variable, also known as the reparameterization trick \cite{corenflos21}), direct differentiation of MOP-$\alpha$ is available via AD.
+We suppose that the simulator is a differentiable function of $\theta$ for every fixed $\omega$, a condition that requires the latent process to be a continuous random variable. 
+This condition is also known as the {\it reparameterization trick} \cite{corenflos21}.
+Supposing also that the measurement model is differentiable in $\theta$, direct differentiation of MOP-$\alpha$ is available via AD.
 Theorem~\ref{thm:mop-functional-forms} shows that MOP-$\alpha$ is constructed so this direct derivative obtains the score estimator of  \cite{poyiadjis11, scibior21} when $\alpha=1$ and that of \cite{naesseth18} when $\alpha=0$.
@@
             $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$. \newline
-            \hspace*{4mm} Compute $L_n^{B,\theta,\alpha} ={\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$. \newline
+            \hspace*{4mm} Compute $L_n^{B,\theta,\alpha} ={\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}\, \big/\, {\sum_{j=1}^J  w^{P,\theta}_{n,j}}$. \newline
             \hspace*{4mm} Conditional likelihood under $\phi$,
@@
             \hspace*{4mm} Calculate resampled corrected weights
-            $w_{n,j}^{F,\theta}= w^{P,\theta}_{n,k_j}  { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}$.\newline
-            \hspace*{4mm} Compute $ L_n^{A,\theta,\alpha} = L_n^\phi\cdot {\sum_{j=1}^J w^{F,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$.\newline
+            $w_{n,j}^{F,\theta}= w^{P,\theta}_{n,k_j}  { g^{\theta}_{n,k_j}} \, \big/ \, { g^{\phi}_{n,k_j}}$.\newline
+            \hspace*{4mm} Compute $ L_n^{A,\theta,\alpha} = L_n^\phi\cdot {\sum_{j=1}^J w^{F,\theta}_{n,j}} \, \big/ \, {\sum_{j=1}^J  w^{P,\theta}_{n,j}}$.\newline
 		\textbf{Return:} likelihood estimate $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, filtering distributions $\{(X_{n,j}^{F, \theta}, w^{F,\theta}_{n,j})\}_{n,j=1}^{N,J}.$
@@
 
-\subsection{Algorithm Outline} 
-We obtain two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but with the resampling indices constructed from the first pass, with a ``baseline parameter'' $\phi$.
-As the process model follows $\theta$ but the measurement model follows the baseline parameter $\phi$, we therefore call this a ``measurement off-parameter'' particle filter.
+%%\subsection{Algorithm Outline}
+
+Algorithm~\ref{alg:mop} constructs two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but with the resampling indices constructed from the first pass, with the baseline parameter, $\phi$.
+%% As the process model follows $\theta$ but the measurement model follows the baseline parameter $\phi$, we therefore call this a ``measurement off-parameter'' particle filter.
+%% DISCUSSED IN THE PREVIOUS SECTION
 The resampling indices, which we write as $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$, are a function of $\phi$ for any value of $\theta$.
@@
 
-We then reweight the conditional likelihoods by a correction factor accumulated over time to account for the resampling under $\phi$.
-That is, writing $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density and $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$,  we can estimate the conditional likelihood under $\theta$ by
-\vspace*{-1.5mm}
+Algorithm~\ref{alg:mop} reweights the conditional likelihoods by a correction factor accumulated over time to account for the resampling under $\phi$. 
+Writing $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density and $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$, we obtain two suitably weighted estimates of the conditional likelihood under $\theta$,
+\arxiv{}{\vspace*{-1.5mm}}
 \begin{equation}
      \label{eq:mop-conditional-likelihood}
-     L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}} \text{ or } L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
-     \vspace*{-1.5mm}
+     L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}, \hspace{15mm} 
+     L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
+     \arxiv{}{\vspace*{-1.5mm}}
 \end{equation}
-where the weights for each particle are updated by
-\vspace*{-1.5mm}
+where the weights for each particle are updated recursively by
+\arxiv{}{\vspace*{-1.5mm}}
 \begin{equation}
     \label{eq:weighting-scheme}
-    w_{n,j}^{P,\theta} = (w_{n-1,j}^{F,\theta})^\alpha,  w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}, w^{F,\theta}_{0,j}= 1.
-    \vspace*{-1.5mm}
+    w_{n,j}^{P,\theta} = (w_{n-1,j}^{F,\theta})^\alpha, 
+    \hspace{5mm}
+    w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}, 
+    \hspace{5mm}
+    w^{F,\theta}_{0,j}= 1.
+    \arxiv{}{\vspace*{-1.5mm}}
 \end{equation}
+The before-resampling conditional likelihood estimate $L_n^{B,\theta,\alpha}$ is preferable in practice, as it has slightly lower variance than the after-resampling estimate $L_n^{A,\theta,\alpha}$, but the latter is useful in deriving properties of the MOP-$\alpha$ gradient estimate such as that in Theorem \ref{thm:mop-functional-forms}.
 
-The before-resampling conditional likelihood estimate $L_n^{B,\theta,\alpha}$ should be used in general, as it has slightly lower variance than the after-resampling estimate $L_n^{A,\theta,\alpha}$, but the latter is useful in deriving properties of the MOP-$\alpha$ gradient estimate such as that in Theorem \ref{thm:mop-functional-forms}.
-
-This yields a suitably reweighted particle filter that targets the filtering distribution under $\theta$ and produces a strongly consistent likelihood estimate for the likelihood at $\theta$.
+%%\ed{This yields a suitably reweighted particle filter that targets the filtering distribution under $\theta$ and produces a strongly consistent likelihood estimate for the likelihood at $\theta$.
 As the resampling indices are invariant to $\theta$ when $\omega$ and $\phi$ are fixed, we successfully bypass the issue of Monte Carlo resampling when the simulator and and measurement model are differentiable.
 Taking the gradient with respect to $\theta$ of the log-likelihood estimate for $\theta$ then yields an estimate for the score at $\theta$.
+I THINK THIS MATERIAL IS COVERED ELSEWHERE, E.G.
+EARLIER IN THIS SECTION}
 
-\subsection{Discounting Weights for a Bias-Variance Tradeoff}
+%\subsection{Discounting Weights for a Bias-Variance Tradeoff}
 
-The correction factors, $w^{P,\theta}_{n,j}$ and $w^{F,\theta}_{n,j}$, accumulate as $n$ increases, leading to numerical instability.
-One can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$ as in Equation \ref{eq:weighting-scheme} while still targeting the filtering distribution and likelihood if $\theta=\phi$.
+If $\alpha=1$ in Algorithm~\ref{alg:mop}, the weights, $w^{P,\theta}_{n,j}$ and $w^{F,\theta}_{n,j}$, accumulate as $n$ increases.
+This leads to numerical instability unless $N$ is small.
+For $\alpha<1$, the weights fom previous timesteps are discounted, as in Equation~(\ref{eq:weighting-scheme}).
+Heuristically, $\alpha$ controls a rate of exponential decay of the memory that the filter at time $t_n$, and its resulting gradient estimate, has over the ancestral trajectories prior to $t_n$.
+With $\alpha<1$, Algorithm~\ref{alg:mop} continues to target the filtering distribution and likelihood if $\theta=\phi$.
+However, we expect bias for $\theta\neq\phi$ that shrinks as $\theta$ approaches $\phi$.
 This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate.
-When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, and Theorem~\ref{thm:mop-functional-forms} shows that it recovers the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}.
-When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{naesseth18}. 
-This discount factor is novel and effective both in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).
+
+When $\alpha=1$, MOP-$1$ maintains complete memory of each particle's ancestral trajectory, and Theorem~\ref{thm:mop-functional-forms} shows that it recovers the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}.
+At the other extreme, when $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{naesseth18}. 
+The novel possibility of $0<\alpha<1$ is effective both in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).
 
@@
 
-$\alpha$ therefore controls an exponentially-weighted moving average, determining how much memory the resulting gradient estimate has over ancestral trajectories.
-This is illustrated in Figure \ref{fig:biasvar}, in the context of the Dhaka cholera model of \cite{king08}.
-
 \begin{thm}[MOP-$0$ and MOP-$1$ Functional Forms]
@@
     \vspace*{-2.5mm}
-    \begin{equation}
+    \begin{equation} \nonumber
         \nabla_\theta \hat\ell^0(\theta) 
@@
     \end{equation}
-    yielding \cite{naesseth18} on the bootstrap filter. When $\alpha=1$,
+    yielding the estimate of \cite{naesseth18} for the bootstrap filter. When $\alpha=1$,
     \vspace*{-2.5mm}
-    \begin{equation}
+    \begin{equation} \nonumber
         \nabla_\theta \hat{\ell}^1(\theta) 
@@
     \end{equation}
-    yielding the estimator of \cite{poyiadjis11, scibior21} with the bootstrap filter.
+    yielding the estimator of \cite{poyiadjis11, scibior21} for the bootstrap filter.
 \end{thm}
@@
 Then, for sufficiently large $\alpha$ and $J$ to ensure the score estimate is a $1/2$-approximation and the minimum eigenvalue of $H$ is greater than some $c > 0$ for all $m$ with probability at least $1-\delta$, the second stage of IFAD converges linearly to the MLE:
-\vspace*{-2mm}
+\arxiv{}{\vspace*{-2mm}}
 $$
@@
 \end{thm}
-\vspace*{-5mm}
-
+\arxiv{}{\vspace*{-5mm}}
 We defer the proof, which is similar to that of Theorem 6 in \cite{mahoney16}, to \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}.
@@
     When $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the filtering distribution under $\theta$ and is strongly consistent for the likelihood under $\theta$.
-That is, for $\pi_n(\theta)=f_{X_{n}|Y_{1:n} ; \theta}$ and any measurable and bounded functional $h$ and for $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\prod_{n=1}^N L_n^{B,\theta,\alpha}$, it holds that
-    \vspace*{-2.5mm}
-    \begin{equation}
-        \frac{\sum_{j=1}^J h(x_{n,j}^{F, \theta}) w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{F,\theta}} \stackrel{a.s.}{\to} E_{\pi_n(\theta)} h, \;\; \hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).
-    \vspace*{-2.5mm}
+That is, for $\pi_n(\theta)=f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^* ; \theta)$ and any measurable and bounded functional $h$ and for $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\prod_{n=1}^N L_n^{B,\theta,\alpha}$, it holds that
+    \arxiv{}{\vspace*{-2.5mm}}
+    \begin{equation} \nonumber
+        \frac{\sum_{j=1}^J h(x_{n,j}^{F, \theta}) \, w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{F,\theta}} \stackrel{a.s.}{\to} E_{\pi_n(\theta)} [h(X_n)], \hspace{5mm} \hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).
+    \arxiv{}{\vspace*{-2.5mm}}
     \end{equation}
@@
 
-Despite showing that the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} use
+Despite showing that the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} estimate the Fisher score by
 $$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}\left(x_{0:n,j}^{A, F,\theta}, y_{1:N}^* ; \theta\right)$$ and not
@@
     \label{thm:mop-biasvar}
-    When $\alpha\in(0,1)$, $\theta=\phi$, define $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$, there exists some $\epsilon>0$ depending on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karjalainen23} such that the MSE and variance of MOP-$\alpha$ are:
+    When $\alpha\in(0,1)$ and $\theta=\phi$, define $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$. 
+    There exists an $\epsilon>0$ depending on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karjalainen23} such that the MSE and variance of MOP-$\alpha$ are:
     \vspace*{-1ex}
@@
 \end{thm}
-
 We defer the proof to \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material}, but provide a brief outline here. 
-The variance bound can be reduced to the approximation error between MOP-$\alpha$ and a variant called MOP-$(\alpha,k)$ where the weights are truncated at lag $k$ with discount factor $\alpha$ to ensure strong mixing, and the covariance of the latter which can be controlled with Davydov's inequality and a standard $L^p$ error bound for the particle filter.
+The variance bound can be reduced to the approximation error between MOP-$\alpha$ and a variant called MOP-$(\alpha,k)$ where the discounted weights are truncated at lag $k$. 
+The discount factor, $\alpha$, ensures strong mixing.
+The covariance of the derivative of MOP-$(\alpha,k)$ can be controlled with Davydov's inequality and a standard $L^p$ error bound for the particle filter. 
 The MSE bound considers the error between the score and MOP-$(1,k)$, and the error between MOP-$(1,k)$ and MOP-$\alpha$. 
@@
 
-We also show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}(NpG'(\theta)^2/J)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$ (ignoring factors of $p, G'$, etc.).
-This and Theorem \ref{thm:mop-biasvar} show that MOP-$\alpha$ interpolates between MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$.
+We also show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( NpG'(\theta)^2\big/J \big)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p, G'$, etc.
+\ed{complete the etc to be more precise}.
+Combining that result with Theorem \ref{thm:mop-biasvar} shows that MOP-$\alpha$ interpolates between MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$.
 
@@
 In line with the cheap gradient principle of \cite{kakade2019provably}, getting a gradient estimate from MOP-$\alpha$ takes no more than 6 times that of the runtime of the particle filter (in practice, the multiplier is 3.75).
-It therefore shares the same $O(NJ)$ time complexity as the particle filter, unlike the $O(NJ^2)$ complexity of \cite{corenflos21} and Algorithm 2 in \cite{poyiadjis11, scibior21}.
+MOP-$\alpha$ and IFAD therefore share the same $O(NJ)$ time complexity as the particle filter, unlike the $O(NJ^2)$ complexity of \cite{corenflos21} and Algorithm 2 in \cite{poyiadjis11, scibior21}.
 
@@
 
-We note that it is possible to work without a differentiable simulator, but one would then require access to the transition densities.
-We are working on an analogue that uses (differentiable) transition densities instead of a differentiable simulator, that may be better suited to large discrete state spaces than either MOP-$\alpha$ or the Baum-Welch algorithm.
+If the simulator is discontinuous in $\theta$, MOP-$\alpha$ no longer applies.
+We are working on a variation of MOP-$\alpha$ applicable to this case.
+Specifically, differentiable transition density ratios can be used instead of a differentiable simulator.
+The off-parameter treatment of the measurement model is extended to an off-parameter simulation for the dynamic process.
+In that case, one requires access to the transition densities, or at least their ratios. 
+The discounting parameter, $\alpha$, can be incorporated as for MOP-$\alpha$.
+This algorithm may be better suited to large discrete state spaces than either MOP-$\alpha$ or the Baum-Welch algorithm.
 Additionally, our gradient estimate can be used for variational inference to approximate the posterior distribution over latent states \cite{naesseth18} and parameters.
@@
 As the proof of Theorem \ref{thm:mop-biasvar} implies, we can also truncate the weights at a fixed lag, corresponding to the MOP-$(1,k)$ estimate mentioned. 
-The analysis is similar, with slightly better rates but a less natural interpretation.
+The analysis is similar, with slightly better rates but less convenient implementation.
 
@@
 This research was supported by National Science Foundation grant DMS-1761603. 
-We acknowledge communications with Nicholas Chopin. 
+We acknowledge communications with Nicolas Chopin. 
 }{
@@
 
-\section{A Strong Law of Large Numbers for Triangular Arrays of Particles With ``Off-Parameter’’ Resampling}
+\section{A Strong Law of Large Numbers for Triangular Arrays of Particles With Off-Parameter Resampling}
 \label{appendix:targeting}

### 2024-06-14 paper/appendix/biasvar.tex
update from overleaf

@@
     Setting $X$ to be the particle collection at time $m$, and $Y$ at time $n$, we have that the alpha mixing coefficients,
-    $$\int | f_{XY} - f_X f_Y | < \alpha,$$
+    \begin{equation}
+    \int \big| f_{XY}(x,y) - f_X (x)\, f_Y(y) \big|\, dx\, dy < \alpha,
+    \end{equation}
     are bounded by the Dobrushin coefficient, i.e we have
-    $$\int | f_{Y|X}(y|x_1) - f_{Y|X}(y|x_2) |dy < \alpha$$ for all $x_1$, $x_2$.
+    \begin{equation}
+    \int \big| f_{Y|X}(y|x_1) - f_{Y|X}(y|x_2) \big| \, dy < \alpha
+    \end{equation} 
+    for all $x_1$, $x_2$.
 \end{lem}
@@
     We rewrite the alpha-mixing assertion as 
-    $$\int { \int |f_Y|X(y|x) - f_Y(y)| dy } f_X dx.$$
-    We claim that the Dobrushin coefficient implies $$\int | f_Y|X(y|x_1) - f_Y(y) |dy < \alpha$$ for all $x_1$. This is shown as follows:
-    \begin{align*}
-        \int | f_{Y|X}(y|x) - f_Y(y) |dy 
-        &= \int \left|  \int [f_{Y|X}(y|x_1) - f_{Y|X}(y|x)] f_X(x)dx \right|dy \\
-        &< \int   \int  | f_{Y|X}(y|x_1) - f_{Y|X}(y|x)] | dy f_X(x)dx \\
-        &< \int \alpha f_X(x)dx \\
+    \begin{equation}
+    \int { \int \big|f_{Y|X}(y|x) - f_Y(y)\big| \, dy } \, f_X(x) \, dx.
+    \end{equation}
+    We claim that the Dobrushin coefficient implies 
+    \begin{equation}
+    \int \big| f_{Y|X}(y|x_1) - f_Y(y) \big| \, dy < \alpha
+    \end{equation} 
+    for all $x_1$. This is shown as follows:
+    \begin{align}
+        \int \big| f_{Y|X}(y|x) - f_Y(y) \big| \, dy 
+        &= \int \left|  \int \big[f_{Y|X}(y|x_1) - f_{Y|X}(y|x)\big] f_X(x) \, dx \right|dy \\
+        &< \int   \int  \big| f_{Y|X}(y|x_1) - f_{Y|X}(y|x)\big| \, dy \, f_X(x)\, dx \\
+        &< \int \alpha f_X(x) \, dx \\
         &=\alpha
-    \end{align*}
+    \end{align}
     We then have the desired result.
@@
 \subsection{Warm Up: MOP-$0$ Variance Bound}
-We now prove the variance bound, presented below. 
-\begin{thm}
 
+Note that we made Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} to leverage the results from \cite{karjalainen23} on the forgetting of the particle filter. 
+This is required to show error bounds on the gradient estimates that we provide, namely that the error of the MOP-$1$ estimator, corresponding to the estimator of \cite{poyiadjis11}, is $O(N^4)$, and that the variance of MOP-$\alpha$ for any $\alpha<1$ is $O(N)$.
+
+We can decompose the MOP-$\alpha$ estimator as follows. 
+When $\theta=\phi$,
+\begin{equation}
+\hat{\mathcal{L}}(\theta):=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J w_{n, j}^{P, \theta}}=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J (w_{n-1, j}^{F, \theta})^\alpha}.
+\end{equation}
+Now, to illustrate the proof strategy for the general case in an easier context, we first analyze the special case when $\alpha=0.$ We now prove the variance bound for this case, presented below.
+
+\begin{thm}[Variance of Gradient Estimate of MOP-$0$]
+The variance of the gradient estimate from MOP-$0$, or the gradient estimate resulting from the algorithm of \cite{naesseth18}, is:
+ \begin{equation}
+     \Var(\nabla_\theta \hat\ell(\theta)) \lesssim \frac{G'(\theta)^2N}{J}.
+ \end{equation}
 \end{thm}
-We made Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} to leverage the results from \cite{karjalainen23} on the forgetting of the particle filter.
-This is required to show error bounds on the gradient estimates that we provide -- namely that the error of the MOP-$1$ estimator that corresponds to that of \cite{poyiadjis11} is $O(N^2)$, and that the variance of MOP-$\alpha$ for any $\alpha<1$ is $O(N)$.
-
-We can decompose the MOP-$\alpha$ estimator as follows. When $\theta=\phi$,
-$$\hat{\mathcal{L}}(\theta):=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J w_{n, j}^{P, \theta}}=\prod_{n=1}^N L_n^{A, \theta, \alpha}=\prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n, j}^{F, \theta}}{\sum_{j=1}^J (w_{n-1, j}^{F, \theta})^\alpha}.$$
-Now, to illustrate the proof strategy for the general case in an easier context, we first analyze the special case when $\alpha=0.$
 
@@
 Similarly to the proof of Lemma \ref{lem:mop-0-formula}, its gradient when $\theta=\phi$ is therefore 
-\begin{align*}
+\begin{align}
     \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
@@
     &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
-\end{align*}
+\end{align}
 where we use the log-derivative trick, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again.
 The point here was that we establish that the gradient of the log-likelihood estimate is given by the sum of terms over all $N$ and $J$:
-$$\nabla_\theta \hat{\ell}(\theta) = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right).$$
+\begin{equation}
+\nabla_\theta \hat{\ell}(\theta) = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right).
+\end{equation}
 Therefore,
-\begin{align*}
-    \Var(\nabla_\theta \hat\ell(\theta)) &= \frac{1}{J^2}\Var\left(\sum_{n=1}^N\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &= \frac{1}{J^2}\sum_{n=1}^N\Var\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &+ 2\sum_{m<n}\Cov\left(\frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &= \sum_{n=1}^N \Var(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \Cov(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta)).
-\end{align*}
+\begin{align}
+    \Var\big(\nabla_\theta \hat\ell(\theta)\big) &= \frac{1}{J^2}\Var\left(\sum_{n=1}^N\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    & \hspace{-15mm}= \frac{1}{J^2}\sum_{n=1}^N\Var\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    &\hspace{-15mm}+ 2\sum_{m<n}\Cov\left(\frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    &\hspace{-15mm}= \sum_{n=1}^N \Var\big(\nabla_\theta\hat\ell_n(\theta)\big) + 2\sum_{m<n} \Cov\big(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta)\big).
+\end{align}
 Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
 We know from Theorem 3 of \cite{karjalainen23} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
-$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
+\begin{equation}
+\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},
+\end{equation}
 i.e.
@@
 Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
-Therefore, by Davydov's inequality, noting that $\nabla_\theta\hat\ell_n(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and WLOG assuming $\E[(\nabla_\theta\hat\ell_m(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/4}$, we see that
-\begin{align*}
-    \Cov(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) 
-    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat\ell_m(\theta))^4]^{1/4}\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/4}\\
-    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2}.
-\end{align*}
-To bound this, we use the fact that $$\E[(\nabla_\theta\hat\ell_n(\theta))^4] = \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4]+\E[(\nabla_\theta\hat\ell_n(\theta))^2]^2$$
+Therefore, by Davydov's inequality, noting that $\nabla_\theta\hat\ell_n(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and without loss of generality labeling $m$ and $n$ such that $\E\big[(\nabla_\theta\hat\ell_m(\theta))^4\big]^{1/4}\leq\E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big]^{1/4}$, we see that
+\begin{align}
+    \Cov\big(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)\big) 
+    &\leq \alpha(n-m)^{1/2} \, \E\big[(\nabla_\theta\hat\ell_m(\theta))^4\big]^{1/4}
+    \,
+    \E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big]^{1/4}\\
+    &\leq \alpha(n-m)^{1/2}\, \E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big]^{1/2}.
+\end{align}
+To bound this, we use the fact that 
+\begin{equation}
+\E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big] = \E\Big[\big(\nabla_\theta\hat\ell_n(\theta)-\E\big[\nabla_\theta\hat\ell_n(\theta)\big]\big)^4\Big]+\E\big[(\nabla_\theta\hat\ell_n(\theta))^2\big]^2
+\end{equation}
 alongside Lemma 2 of \cite{karjalainen23}, which shows that
-$$\E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^4] \lesssim \frac{G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat\ell_n(\theta)-\E[\nabla_\theta\hat\ell_n(\theta)])^2] \lesssim \frac{G'(\theta)^2}{J}.$$
+\begin{equation}
+\E\Big[\big(\nabla_\theta\hat\ell_n(\theta)-\E\big[\nabla_\theta\hat\ell_n(\theta)\big]\big)^4\Big] \lesssim \frac{G'(\theta)^4}{J^2}, \hspace{10mm} \E\Big[\big(\nabla_\theta\hat\ell_n(\theta)-\E\big[\nabla_\theta\hat\ell_n(\theta)\big]\big)^2\Big] \lesssim \frac{G'(\theta)^2}{J}.
+\end{equation}
 It follows that 
-$$\E[(\nabla_\theta\hat\ell_n(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{G'(\theta)^4}{J^2}+\left(\frac{G'(\theta)^2}{J}\right)^2} = \frac{G'(\theta)^2}{J},$$
+\begin{equation}
+\E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big]^{1/2} \lesssim  \sqrt{\frac{G'(\theta)^4}{J^2}+\left(\frac{G'(\theta)^2}{J}\right)^2} = \frac{G'(\theta)^2}{J},
+\end{equation}
 and we conclude that 
-$$\Cov(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{G'(\theta)^2}{J}.$$
+\begin{equation}
+\Cov\big(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)\big) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{G'(\theta)^2}{J}.
+\end{equation}
 Putting it all together, we see that
-\begin{align*}
-    \Var(\nabla_\theta \hat\ell(\theta)) &= \sum_{n=1}^N \Var(\nabla_\theta\hat\ell_n(\theta)) + 2\sum_{m<n} \Cov(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta))\\
+\begin{align}
+    \Var\big(\nabla_\theta \hat\ell(\theta)\big) &= \sum_{n=1}^N \Var\big(\nabla_\theta\hat\ell_n(\theta)\big) + 2\sum_{m<n} \Cov\big(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta)\big)\\
     &\leq \frac{NG'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\frac{G'(\theta)^2}{J} \\
@@
     &\lesssim \frac{G'(\theta)^2N}{J}.
-\end{align*}
+\end{align}
 \end{proof}

### 2024-06-14 paper/appendix/feynman.tex
update from overleaf

@@
 If this yields a density, this then corresponds to the process density $f_{X_{2}|X_1}$ conditional on $x$ and integrated over $A$.
-\item \textbf{Markov kernels and measures:} For any measure $\eta$, any Markov kernel $M$ on $\gX$, any point $x \in \gX$ and any measurable subset $A \subseteq \gX$, let 
+
+\paragraph{Markov kernels and measures:} For any measure $\eta$, any Markov kernel $M$ on $\gX$, any point $x \in \gX$ and any measurable subset $A \subseteq \gX$, let 
 \begin{align}
-    \eta(h) &= \int h d\eta = \int h(x) \eta(dx), \\(\eta M)(A) &= \int \eta(dx)M(x,A), \\
+    \eta(h) &= \int h \, d\eta = \int h(x) \eta(dx), \\(\eta M)(A) &= \int \eta(dx)M(x,A), \\
     (Mh)(x) &= \int M(x, dy) h(y).
 \end{align}
-\item \textbf{Compositions of Markov kernels:} The composition of a Markov kernel $M_1$ with another Markov kernel $M_2$ is another Markov kernel, given by 
+
+\paragraph{Compositions of Markov kernels:} The composition of a Markov kernel $M_1$ with another Markov kernel $M_2$ is another Markov kernel, given by 
 \begin{equation}
@@
 \paragraph{Forgetting of the particle filter:} 
-The above formalization yields a result from \cite{karjalainen23} that we require for our analysis of the bias, variance, and error of MOP-$\alpha$ -- the forgetting of the particle filter.
+The above formalization yields a result from \cite{karjalainen23} on the forgetting of the particle filter that we require for our analysis of the bias, variance, and error of MOP-$\alpha$.  
 That is, \cite{karjalainen23} show that

### 2024-06-14 paper/appendix/targeting.tex
update from overleaf

@@
         \begin{equation}
-        \eta_n^J = \frac{\sum_{j=1}^J w_{n,j}^P \delta_{x_{n,j}^P}}{\sum_{j=1}^J w_{n,j}^P}.
+        \eta_n^J = \frac{\sum_{j=1}^J w_{n,j}^P \, \delta_{x_{n,j}^P}}{\sum_{j=1}^J w_{n,j}^P}.
         \end{equation}
@@
         \begin{equation}
-        \pi_n^J = \frac{\sum_{j=1}^J \delta_{x_{n,j}^P }w_{n,j}^P g_{n,j}}{\sum_{j=1}^J w_{n,j}^Pg_{n,j}} \approx \frac{\sum_{j=1}^J \delta_{x_{n,k_j}^P }w_{n,k_j}^P g_{n,k_j}/p_{n,k_j}}{\sum_{j=1}^J w_{n,k_j}^Pg_{n,k_j}/p_{n,k_j}} = \frac{\sum_{j=1}^J w_{n,j}^F \delta_{x_{n,j}^F}}{\sum_{j=1}^J w_{n,j}^F}.
+        \pi_n^J = \frac{\sum_{j=1}^J \delta_{x_{n,j}^P } w_{n,j}^P \, g_{n,j}}{\sum_{j=1}^J w_{n,j}^P \, g_{n,j}} \approx \frac{\sum_{j=1}^J \delta_{x_{n,k_j}^P }w_{n,k_j}^P \, g_{n,k_j}\big/p_{n,k_j}}{\sum_{j=1}^J w_{n,k_j}^P\, g_{n,k_j}\big/p_{n,k_j}} = \frac{\sum_{j=1}^J w_{n,j}^F \, \delta_{x_{n,j}^F}}{\sum_{j=1}^J w_{n,j}^F}.
         \end{equation}
-        \item The prediction weights at the next timestep are given by $w_{n+1,j}^P = w_{n,j}^F =  w_{n,k_j}^P g_{n,k_j}/p_{n,k_j}$.
+        \item The prediction weights at the next timestep are given by $w_{n+1,j}^P = w_{n,j}^F =  w_{n,k_j}^P \, g_{n,k_j}\big/p_{n,k_j}$.
     \end{enumerate}
@@
 \begin{equation}
-    E_g\{h(X) \cdot W\}=E_\pi\{h(X)\}.
+    E_g\big[h(X) \cdot W\big]=E_\pi\big[h(X)\big].
 \end{equation}  
@@
 \begin{equation}
-    \frac{\sum_{j=1}^J h(X^J_j) W^J_j}{\sum_{j=1}^J W^J_j} \stackrel{a.s.}{\to} E_\pi(h(X))
+    \frac{\sum_{j=1}^J h(X^J_j) \, W^J_j}{\sum_{j=1}^J W^J_j} \stackrel{a.s.}{\to} E_\pi(h(X))
 \end{equation}
@@
     \begin{equation}
-    (Y_j^J,V_j^J) = \big(\tilde X^J_{k_j},U^J_{k_j}/\pi^J_{k_j}\big).
+    (Y_j^J,V_j^J) = \big(\tilde X^J_{k_j},U^J_{k_j}\big/\pi^J_{k_j}\big).
     \end{equation} 
@@
     Note that as the $Y_j^J$ are a subsample from $X_j^J$, $h$ can be a function of $Y$ as well as it is one for $X$. We then expand
-    \begin{equation}\frac{\sum_j h(Y_j^J) V_j^J}{\sum_j V_j^J} = \frac{\sum_j h(\tilde X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}}.\end{equation}
+    \begin{equation}\frac{\sum_j h(Y_j^J) \, V_j^J}{\sum_j V_j^J} = \frac{\sum_j h(\tilde X_{k_j}^J)\, \frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}}.\end{equation}
     By hypothesis,
     \begin{equation}
-    \frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[f(X)].
+    \frac{\sum_j h(\tilde X_j^J)\, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}\big[f(X)\big].
     \end{equation}
@@
     \begin{equation}\label{eq:lemma1:h}
-    \frac{\sum_j h(X_{k_j}^J)\frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}} - \frac{\sum_j h(\tilde X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} 0.
+    \frac{\sum_j h(X_{k_j}^J)\, \frac{U_{k_j}^J}{\pi_{k_j}^J}}{\sum_j \frac{U_{k_j}^J}{\pi_{k_j}^J}} - \frac{\sum_j h(\tilde X_j^J)\, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} 0.
     \end{equation}
@@
     \begin{equation}
-    \sum_j Z_j^J := \sum_j \left(g(\tilde X_{k_j}^J) -  g(\tilde X_j^J) \pi_j^J \right) \stackrel{a.s.}{\to} 0.
+    \sum_j Z_j^J := \sum_j \left(g(\tilde X_{k_j}^J) -  g(\tilde X_j^J) \, \pi_j^J \right) \stackrel{a.s.}{\to} 0.
 \end{equation}
-    Because the functional $h$ and importance sampling weights $u_{k_j}^J/\pi_{k_j}^J$ are bounded, we have that $\E[\left(Z_j^J\right)^4] < \infty$.
+    Because the functional $h$ and importance sampling weights $u_{k_j}^J/\pi_{k_j}^J$ are bounded, we have that $\E\big[(Z_j^J)^4\big] < \infty$.
 We can then follow the argument of \cite{chopin20} from this point on, where noting that the $Z_j^J$ are conditionally independent given the $\tilde{X}_j^J$ and $U_j^J$,
-    \begin{eqnarray*} 
+    \begin{eqnarray} 
     \E\left[\left(\sum_j Z_j^J\right)^4 \Bigg| (\tilde X_j^J,U_j^J)_{j=1}^J \right] 
-    &=& J\E\left[(Z_1^J)^4|(\tilde X_j^J,U_j^J)_{j=1}^J\right] + 3J(J-1)\left(\E\big[(Z_j^J)^2|(\tilde X_j^J,U_j^J)_{j=1}^J\big]\right) 
+    &=& J\, \E\left[(Z_1^J)^4|(\tilde X_j^J,U_j^J)_{j=1}^J\right] 
+    \nonumber
+    \\
+    && \hspace{10mm}+ \hspace{2mm} 3J(J-1)\left(\E\big[(Z_j^J)^2\big|(\tilde X_j^J,U_j^J)_{j=1}^J\big]\right) 
     \\ 
     &\leq& CJ^2,
-    \end{eqnarray*}
+    \end{eqnarray}
     for some $C>0$. 
@@
     \begin{equation} 
-    \sum_j h(X_{k_j}^J)\frac{u_{k_j}^J}{\pi_{k_j}^J}
-    -  \sum_j h(X_j^J)\frac{u_j^J}{\pi_j^J}\pi_j^J \stackrel{a.s.}{\to} 0.
+    \sum_j h(X_{k_j}^J)\, \frac{u_{k_j}^J}{\pi_{k_j}^J}
+    -  \sum_j h(X_j^J)\, \frac{u_j^J}{\pi_j^J}\, \pi_j^J \stackrel{a.s.}{\to} 0.
     \end{equation} 
@@
     Suppose that $\{(X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$.
-Also suppose that $(X^{\prime J}_j,U^{\prime J}_j) = \big(X_j^J,U_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$.
-Then, if $U_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $U_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,U^{\prime J}_j)\}$ targets $f_{X|Z}(\cdot | z^*)$.
+    Also suppose that $(X^{\prime J}_j,U^{\prime J}_j) = \big(X_j^J,U_j^J \, f_{Z|X}(z^*|X_j^J)\big)$. 
+    Then, if $U_j^J \, f_{Z|X}(z^*|X_j^J)$ and $U_j^J \, f_{Z|X}(z^*|X_j^J) \big/ f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,U^{\prime J}_j)\}$ targets $f_{X|Z}(\cdot | z^*)$.
 \end{lem}
@@
     Again, we want to show that
-     \begin{equation}\frac{\sum_j h(X_j^J) \cdot U_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j U_j^J \cdot f_{Z|X}(z^*|X_j^J)} \stackrel{a.s.}{\to} \E_{f_{X|Z}}[h(X)|z^*].\end{equation}
+     \begin{equation}
+     \frac{\sum_j h(X_j^J) \cdot U_j^J \cdot f_{Z|X}(z^*|X_j^J)}{\sum_j U_j^J \cdot f_{Z|X}(z^*|X_j^J)} \stackrel{a.s.}{\to} \E_{f_{X|Z}}\big[h(X)\big|z^*\big].
+     \end{equation}
     We already have that for any measurable bounded $g$,
     \begin{equation}
-        \frac{\sum_j g(X_j^J)U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}[g(X)].
+        \frac{\sum_j g(X_j^J)\, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E_{f_X}\big[g(X)\big].
         \label{eq:lemma-posterior-hypothesis}
@@
     \begin{equation}
-    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}
-    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}} \right)^{-1}.
+    \frac{J^{-1} \sum_j h(X_j^J) \, f_{Z|X}(z^*|X_j^J)\,  {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}
+    \times \left( \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) \, {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}} \right)^{-1}.
     \end{equation}
@@
     \begin{equation}
-    \frac{J^{-1} \sum_j h(X_j^J) f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \stackrel{a.s.}{\to} \E_{f_X}[h(X)f_{Z|X}(z^*|X)],
+    \frac{J^{-1} \sum_j h(X_j^J) \, f_{Z|X}(z^*|X_j^J) \, {U}_j^{J}}{J^{-1} \sum_j {u}_j^{J}} \stackrel{a.s.}{\to} \E_{f_X}\big[h(X)\, f_{Z|X}(z^*|X)\big],
     \end{equation}
@@
     \begin{equation} 
-    \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}  \stackrel{a.s.}{\to} \E_{f_X}[f_{Z|X}(z^*|X)] = f_Z(z^*).
+    \frac{J^{-1} \sum_j f_{Z|X}(z^*|X_j^J) \, {U}_j^{J}}{J^{-1} \sum_j {U}_j^{J}}  \stackrel{a.s.}{\to} \E_{f_X}\big[f_{Z|X}(z^*|X)\big] = f_Z(z^*).
     \end{equation}
@@
     \begin{eqnarray}
-    \frac{\E_{f_X}[h(X)f_{Z|X}(z^*|X)]}{f_Z(z^*)} &=& \E_{f_X}\left[\frac{h(X)f_{Z|X}(z^*|X)}{f_Z(z^*)}\right] 
+    \frac{\E_{f_X}\big[h(X)\, f_{Z|X}(z^*|X)\big]}{f_Z(z^*)} &=& \E_{f_X}\left[\frac{h(X)\, f_{Z|X}(z^*|X)}{f_Z(z^*)}\right] 
     \\
@@
     \begin{equation}
-    \big(X^{F}_{n,j},w^{F}_{n,j}\big) = \big(X^{P}_{n,k_j}, w^{P}_{n,k_j} g_{n,k_j}\big/ p_{n,k_j}\big),
+    \big(X^{F}_{n,j},w^{F}_{n,j}\big) = \big(X^{P}_{n,k_j}, w^{P}_{n,k_j} \, g_{n,k_j}\big/ p_{n,k_j}\big),
     \end{equation}
@@
     Then, Lemma \ref{lem:marginal-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big)\big\}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
-    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
+    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} \, g^\theta_{n,j} \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
     Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
     \begin{equation}
-    \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,k_j}, w^{P,\theta}_{n,k_j} g^\theta_{n,k_j}\big/ g^\phi_{n,k_j}\big),
+    \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,k_j}, w^{P,\theta}_{n,k_j} \, g^\theta_{n,k_j}\big/ g^\phi_{n,k_j}\big),
     \end{equation}
@@
 \begin{equation}\label{L1}
-L_n^{B,\theta} =  \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
+L_n^{B,\theta} =  \frac{\sum_{j=1}^Jg^\theta_{n,j} \, w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
 \end{equation}
@@
 \begin{equation}
-L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
-= L^\phi_n \sum_{j=1}^J w_{n,j}^{FC,\theta} \frac{g^\phi_{n,j}}{L_n^\phi}
+L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} \, w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
+= L^\phi_n \sum_{j=1}^J w_{n,j}^{FC,\theta} \, \frac{g^\phi_{n,j}}{L_n^\phi}
 \end{equation}
@@
 from which we obtain (\ref{L2}).
-
 Using Lemma \ref{lem:lik-proper-weight}, we obtain a likelihood estimate,

### 2024-06-14 paper/ms.tex
update from overleaf

@@
 Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a {\it Measurement Off-Parameter} (MOP) particle filter.
+In short, \textit{instead of directly differentiating through a basic particle filter, MOP uses AD to differentiate through a series of measurement density ratios and a differentiable simulator}.
 We proceed to generalize the MOP particle filter by adding a discounting factor, $\alpha \in [0,1]$, which interpolates between the  biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
@@
 We avoid this issue by constructing a particle filtering algorithm called {\it Measurement Off-Parameter with discount factor} $\alpha$ (MOP-$\alpha$), defined by the pseudocode in Algorithm~\ref{alg:mop}.
-In short, instead of differentiating through a Monte Carlo algorithm directly, we instead use AD to differentiate through a series of measurement density ratios and a differentiable simulator.
+In short, \textit{instead of directly differentiating through a basic particle filter, we instead use AD to differentiate through a series of measurement density ratios and a differentiable simulator}.
 Note that the plug-and-play property forbids access to transition densities but permits access to the density of the measurements conditional on the value of the latent process.
@@
 The MOP-$\alpha$ algorithm resamples the particles according to an arbitrary resampling rule that depends on both the target parameter value and a baseline parameter value, and this explains the name {\it measurement off-parameter}.
-Similar quantities are called {\it off-policy} in reinforcement learning, and a parameter in a POMP model is analogous to a policy in a corresponding control model \ed{Kevin, would it be good to add your favorite RL reference here?
-Also, you may want to edit this sentence.
-I think something like this will help the reader to ``get'' our terminology.}.
+This is intended as an analogy to {\it off-policy} learning in reinforcement learning.
 Specifically, MOP-$\alpha$ evaluates the likelihood at some $\theta \in \Theta$, but instead resamples the particles according to the indices generated by a vanilla particle filter run at a {\it baseline parameter}, $\phi \in \Theta$.
@@
 Conversely, when we tried gradient descent with MOP alone, we encountered many failed searches.
-This is a difficult, nonconvex, and noisy problem, \ed{SAY HOW MANY PARAMETERS ARE ESTIMATED HERE} and the search gets stuck in local minima and saddle points, failing to approach the MLE.
+This is a difficult, nonconvex, and noisy problem with $18$ parameters, and the search gets stuck in local minima and saddle points, failing to approach the MLE.
 
@@
 
-\ed{Labeling A and B would be more conventional than Left and Right, and also easier to refer to in the text.}
+%% \ed{Labeling A and B would be more conventional than Left and Right, and also easier to refer to in the text.}
 
@@
     \includegraphics[width=\textwidth/\real{1.25}]{../imgs/pmcmc/nuts_eb.png}
-    \caption{Convergence diagnostics for NUTS with the informative nonparametric empirical Bayes prior from applying KDE to preliminary searches using IF2.
+    \caption{Convergence diagnostics for NUTS with the informative nonparametric empirical Bayes prior from applying KDE to preliminary searches using IF2, with $4$ chains.
 We display the results for the trend parameter in the Dhaka cholera model of \cite{king08}.
@@
 
-We also show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( NpG'(\theta)^2\big/J \big)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p, G'$, etc.
-\ed{complete the etc to be more precise}.
+We also show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( NpG'(\theta)^2\big/J \big)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p$ and $G'$.
 Combining that result with Theorem \ref{thm:mop-biasvar} shows that MOP-$\alpha$ interpolates between MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$.
+The phase transition arises as even though the particle filter has forgetting properties \cite{karjalainen23}, the resulting derivative estimate of \cite{poyiadjis11} does not, and we require both for the variance reduction.
 
@@
 MOP-$\alpha$ and IFAD are fast algorithms, both in theory and practice. 
-In line with the cheap gradient principle of \cite{kakade2019provably}, getting a gradient estimate from MOP-$\alpha$ takes no more than 6 times that of the runtime of the particle filter (in practice, the multiplier is 3.75).
+In line with the cheap gradient principle of \cite{kakade2019provably}, getting a gradient estimate from MOP-$\alpha$ takes no more than 6 times that of the runtime of the particle filter.
+In our simulations, MOP-$\alpha$ took 3.75 times the runtime of the particle filter.
 MOP-$\alpha$ and IFAD therefore share the same $O(NJ)$ time complexity as the particle filter, unlike the $O(NJ^2)$ complexity of \cite{corenflos21} and Algorithm 2 in \cite{poyiadjis11, scibior21}.
@@
 This research was supported by National Science Foundation grant DMS-1761603. 
-We acknowledge communications with Nicolas Chopin. 
+We thank Nicolas Chopin for helpful communications regarding the strong law of large numbers for off-parameter resampled particle filters. 
 }{

### 2024-06-14 paper/appendix/convergence.tex
update from overleaf
another day, another dollar

@@
     Consider the gradient estimate obtained by MOP-1, which we know by Theorem \ref{thm:mop-grad-consistency} is consistent for the score, where $\theta = \phi$.
-For $||\nabla_\theta \hat{\loglik}^1(\theta) - \nabla_\theta \loglik(\theta)||_2$ to be bounded by $\epsilon$ with probability $1-\delta$, we require
+For $\|\nabla_\theta \hat{\loglik}^1(\theta) - \nabla_\theta \loglik(\theta)\|_2$ to be bounded by $\epsilon$ with probability $1-\delta$, we require
     \begin{align}
@@
     \begin{align}
-        ||\nabla_\theta \hat\ell^1(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).
+        \|\nabla_\theta \hat\ell^1(\theta) - \nabla_\theta \ell(\theta)\|_2 \leq G(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).
     \end{align}
@@
 \begin{align}
-    ||\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)||_2 \leq G(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).
+    \|\nabla_\theta \hat\ell(\theta) - \nabla_\theta \ell(\theta)\|_2 \leq G(\theta)\left(\frac{r_N\sqrt{p}}{J}(1+h^{-1}(\log(2p/\delta))) + \sqrt{\frac{2p\log(2p/\delta)}{J}}\beta_N\right).
 \end{align}
@@
 \begin{align}
-    \frac{1}{J}\sum_{j=1}^J \lambda_{\min}(H_j) - \E_{\tilde{\pi}_t}\lambda_{\min}(H_j) &= \frac{1}{J}\sum_{j=1}^J \lambda_{\min}(H_j) - \gamma' \geq -\frac{r_t}{J}(1+h^{-1}(t)) - \sqrt{\frac{2t}{J}}\beta_t \\
-    \sum_{j=1}^J \lambda_{\min}(H_j) &\geq -r_t(1+h^{-1}(t)) - \sqrt{2tJ}\beta_t + J\gamma',
+    \frac{1}{J}\sum_{j=1}^J \lambda_{\min}(H_j) - \E_{\tilde{\pi}_t}\lambda_{\min}(H_j) &= \frac{1}{J}\sum_{j=1}^J \lambda_{\min}(H_j) - \gamma' \geq -\frac{r_t}{J}\big(1+h^{-1}(t)\big) - \sqrt{\frac{2t}{J}}\beta_t \\
+    \sum_{j=1}^J \lambda_{\min}(H_j) &\geq -r_t\big(1+h^{-1}(t)\big) - \sqrt{2tJ}\beta_t + J\gamma',
 \end{align}
@@
 
-We additionally require, for $c \in (0, \sum_{j=1}^J \E\lambda_{\min}(H_j))$,
+We additionally require, for $c \in \big(0, \sum_{j=1}^J \E\lambda_{\min}(H_j)\big)$,
 \begin{align}
-    \sum_{j=1}^J \lambda_{\min}(H_j) \geq -r_t(1+h^{-1}(t)) - \sqrt{2tJ}\beta_t + J\gamma' \geq c, \\
-    J\gamma' \geq c + r_t(1+h^{-1}(t)) + \sqrt{2tJ}\beta_t.
+    \sum_{j=1}^J \lambda_{\min}(H_j) \geq -r_t\big(1+h^{-1}(t)\big) - \sqrt{2tJ}\beta_t + J\gamma' \geq c, \\
+    J\gamma' \geq c + r_t\big(1+h^{-1}(t)\big) + \sqrt{2tJ}\beta_t.
 \end{align}
-We therefore require \ed{WE SAY ``REQUIRE'' AT VARIOUS POINTS, BUT THAT SUGGESTS NECESSITY WHEREAS I SUPPOSE WE MEAN SUFFICIENCY. I EDITED ONE OR TWO OF THESE.}
+It is therefore sufficient to have
 \begin{equation}
-J > \max\left\{\frac{2r_t(1+h^{-1}(t)) + 2c}{\gamma'}, \frac{2(2t\beta_t^2+c)^2}{\gamma'^2}\right\} \geq  \frac{r_t(1+h^{-1}(t))}{\gamma'} + \sqrt{2tJ}\beta_t/\gamma' + c/\gamma'    
+J > \max\left\{\frac{2r_t\big(1+h^{-1}(t)\big) + 2c}{\gamma'}, \frac{2(2t\beta_t^2+c)^2}{\gamma'^2}\right\} \geq  \frac{r_t\big(1+h^{-1}(t)\big)}{\gamma'} + \sqrt{2tJ}\beta_t/\gamma' + c/\gamma'    
 \end{equation}
@@
 \begin{equation}
-    f(\theta_m+\eta p_m) \leq f(\theta_m) + \eta\beta p_m^Tg(\theta_m),
+    f(\theta_m+\eta p_m) \leq f(\theta_m) + \eta\beta \, p_m^T \, g(\theta_m),
 \end{equation}
 holds for any $0< \eta < \tilde{\eta}$ and some $\beta \in (0,1)$.
-By an argument found in the beginning of the proof of Theorem 6 in \cite{mahoney16}, we have that choosing $J$ such that $||\nabla_\theta\hat{\ell}(\theta_m) - \nabla_\theta \ell(\theta_m)|| \leq \epsilon$ and $\lambda_{\min}(H(\theta_m)) \geq c>0$ for each $m$, yields
+By an argument found in the beginning of the proof of Theorem 6 in \cite{mahoney16}, we have that choosing $J$ such that $\|\nabla_\theta\hat{\ell}(\theta_m) - \nabla_\theta \ell(\theta_m)\| \leq \epsilon$ and $\lambda_{\min}(H(\theta_m)) \geq c>0$ for each $m$, yields
 \begin{align}
-    f(\theta_\eta)-f(\theta_m) \leq \eta p_m^Tg(\theta_m) + \epsilon\eta||p_m|| + \eta^2 \Gamma ||p_m||^2 / 2,
+    f(\theta_\eta)-f(\theta_m) \leq \eta p_m^T\, g(\theta_m) + \epsilon\eta\|p_m\| + \eta^2 \Gamma \|p_m\|^2 / 2,
 \end{align}
@@
 \begin{equation}
-    p_m^Tg(\theta_m) = -p_m^TH(\theta_m)p_m \geq -c||p_m||^2,
+    p_m^Tg(\theta_m) = -p_m^T\, H(\theta_m)\, p_m \geq -c\|p_m\|^2,
 \end{equation}
@@
 \begin{align}
-    f(\theta_\eta)-f(\theta_m) \leq -\eta p_m^TH(\theta_m)p_m + \epsilon\eta||p_m|| + \eta^2 \Gamma ||p_m||^2 / 2,
+    f(\theta_\eta)-f(\theta_m) \leq -\eta p_m^TH(\theta_m)\, p_m + \epsilon\eta\|p_m\| + \eta^2 \Gamma \|p_m\|^2 / 2,
 \end{align}
@@
 \begin{align}
-    -\eta p_m^TH(\theta_m)p_m + \epsilon\eta||p_m|| + \eta^2 \Gamma ||p_m||^2 / 2 &\leq \eta \beta p_m^Tg(\theta_m) = - \eta \beta p_m^TH(\theta_m)p_m \\
-    \epsilon||p_m|| + \eta \Gamma ||p_m||^2 / 2 &\leq (1- \beta) p_m^TH(\theta_m)p_m \\
-    \epsilon + \eta \Gamma ||p_m|| / 2 &\leq c(1- \beta) ||p_m||.
+    -\eta p_m^TH(\theta_m)p_m + \epsilon\eta\|p_m\| + \eta^2 \Gamma \|p_m\|^2 / 2 &\leq \eta \beta p_m^Tg(\theta_m) = - \eta \beta p_m^TH(\theta_m)p_m \\
+    \epsilon\|p_m\| + \eta \Gamma \|p_m\|^2 / 2 &\leq (1- \beta) p_m^TH(\theta_m)p_m \\
+    \epsilon + \eta \Gamma \|p_m\| / 2 &\leq c(1- \beta) \|p_m\|.
 \end{align}
@@
 \begin{equation}
-    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}||g(\theta_m)|| \leq \frac{c(1-\beta)}{2}||p_m||,
+    \eta \leq \frac{c(1-\beta)}{\Gamma}, \;\; \epsilon \leq \frac{c(1-\beta)}{2\Gamma}\|g(\theta_m)\| \leq \frac{c(1-\beta)}{2}\|p_m\|,
 \end{equation}
@@
 \begin{equation}
-||g(\theta_m)|| - ||\nabla_\theta f(\theta_m)|| \leq ||g(\theta_m) - \nabla_\theta f(\theta_m)|| \leq \epsilon => ||\nabla_\theta f(\theta_m)|| \geq ||g(\theta_m)|| - \epsilon
+\|g(\theta_m)\| - \|\nabla_\theta f(\theta_m)\| \leq \|g(\theta_m) - \nabla_\theta f(\theta_m)\| \leq \epsilon \implies \|\nabla_\theta f(\theta_m)\| \geq \|g(\theta_m)\| - \epsilon
 \end{equation} 
@@
 \begin{equation}
-||\nabla_\theta f(\theta_m)||-||g(\theta_m)|| \leq ||\nabla_\theta f(\theta_m)-g(\theta_m)|| \leq \epsilon => ||g(\theta_m)|| \geq ||\nabla_\theta f(\theta_m)|| - \epsilon.
+\|\nabla_\theta f(\theta_m)\|-\|g(\theta_m)\| \leq \|\nabla_\theta f(\theta_m)-g(\theta_m)\| \leq \epsilon \implies \|g(\theta_m)\| \geq \|\nabla_\theta f(\theta_m)\| - \epsilon.
 \end{equation}
 There are now two cases. 
-If the algorithm terminates and $||g(\theta_m)|| \leq \sigma \epsilon$, we can derive 
+If the algorithm terminates and $\|g(\theta_m)\| \leq \sigma \epsilon$, we can derive 
 \begin{equation}
-    ||\nabla_\theta f(\theta_m)|| \leq ||g(\theta_m)|| + \epsilon = \sigma\epsilon+\epsilon = (\sigma+1)\epsilon.
+    \|\nabla_\theta f(\theta_m)\| \leq \|g(\theta_m)\| + \epsilon = \sigma\epsilon+\epsilon = (\sigma+1)\epsilon.
 \end{equation}
-If the algorithm does not terminate, then $||g(\theta_m)|| > \sigma \epsilon$. 
+If the algorithm does not terminate, then $\|g(\theta_m)\| > \sigma \epsilon$. 
 Notice that 
 \begin{eqnarray}
-    \epsilon \geq ||g(\theta_m) - \nabla_\theta f(\theta_m)|| &\geq& ||g(\theta_m)|| - ||\nabla_\theta f(\theta_m)|| 
+    \epsilon \geq \|g(\theta_m) - \nabla_\theta f(\theta_m)\| &\geq& \|g(\theta_m)\| - \|\nabla_\theta f(\theta_m)\| 
     \\
-    ||\nabla_\theta f(\theta_m)|| + \epsilon &\geq& ||g(\theta_m)|| \geq \sigma \epsilon 
+    \|\nabla_\theta f(\theta_m)\| + \epsilon &\geq& \|g(\theta_m)\| \geq \sigma \epsilon 
     \\
-    ||\nabla_\theta f(\theta_m)|| &\geq& \sigma \epsilon - \epsilon = (\sigma - 1)\epsilon 
+    \|\nabla_\theta f(\theta_m)\| &\geq& \sigma \epsilon - \epsilon = (\sigma - 1)\epsilon 
     \\
-    \frac{||\nabla_\theta f(\theta_m)||}{\sigma-1} &\geq& \epsilon,
+    \frac{\|\nabla_\theta f(\theta_m)\|}{\sigma-1} &\geq& \epsilon,
 \end{eqnarray}
@@
 \begin{eqnarray}
-    ||\nabla_\theta f(\theta_m)|| - \epsilon
-    &\geq&  ||\nabla_\theta f(\theta_m)|| - \frac{||\nabla_\theta f(\theta_m)||}{\sigma-1} 
+    \|\nabla_\theta f(\theta_m)\| - \epsilon
+    &\geq&  \|\nabla_\theta f(\theta_m)\| - \frac{\|\nabla_\theta f(\theta_m)\|}{\sigma-1} 
     \\
-    &=& \left(1-\frac{1}{\sigma-1}\right)||\nabla_\theta f(\theta_m|| 
+    &=& \left(1-\frac{1}{\sigma-1}\right)\|\nabla_\theta f(\theta_m\| 
     \\
-    &=& \frac{\sigma-2}{\sigma-1}||\nabla_\theta f(\theta_m)|| 
+    &=& \frac{\sigma-2}{\sigma-1}\|\nabla_\theta f(\theta_m)\| 
     \\
-    &\geq& \frac{2}{3}||\nabla_\theta f(\theta_m)||.
+    &\geq& \frac{2}{3}\|\nabla_\theta f(\theta_m)\|.
 \end{eqnarray}
-Since $||A^{-1}|| = 1/\sigma_{\min}(A)$,
+Since $\|A^{-1}\| = 1/\sigma_{\min}(A)$,
 \begin{align}
-    p_m^TH(\theta_m)p_m &= (-(H(\theta_m))^{-1}g(\theta_m))^TH(\theta_m)(-(H(\theta_m))^{-1}g(\theta_m)) \\
+    p_m^TH(\theta_m)p_m &= \big(-(H(\theta_m))^{-1}g(\theta_m)\big)^TH(\theta_m)\big(-(H(\theta_m))^{-1}g(\theta_m)\big) \\
     &= g(\theta_m)^T(H(\theta_m))^{-1}g(\theta_m) \\
-    &\geq \frac{1}{c}||g(\theta_m)||^2 \\
-    &\geq \frac{1}{c}(||\nabla_\theta f(\theta_m)|| - \epsilon)^2 \\
-    &\geq \frac{4}{9c}||\nabla_\theta f(\theta_m)||^2.
+    &\geq \frac{1}{c}\|g(\theta_m)\|^2 \\
+    &\geq \frac{1}{c}\big(\|\nabla_\theta f(\theta_m)\| - \epsilon\big)^2 \\
+    &\geq \frac{4}{9c}\|\nabla_\theta f(\theta_m)\|^2.
 \end{align}
@@
 \begin{align}
-    f(\theta_m) - f(\theta^*) \leq \frac{1}{2\gamma}||\nabla_\theta f(\theta_m)||^2,
+    f(\theta_m) - f(\theta^*) \leq \frac{1}{2\gamma}\big\|\nabla_\theta f(\theta_m)\big\|^2,
 \end{align}
@@
 \begin{equation}
-    f(\theta_m) - f(\theta^*) \leq \frac{1}{2\gamma}||\nabla_\theta f(\theta_m)||^2 \leq \frac{9c}{4}\frac{1}{2\gamma}p_m^TH(\theta_m)p_m,
+    f(\theta_m) - f(\theta^*) \leq \frac{1}{2\gamma}\big\|\nabla_\theta f(\theta_m)\big\|^2 \leq \frac{9c}{4}\frac{1}{2\gamma}\, p_m^TH(\theta_m)\, p_m,
 \end{equation}
 \begin{equation}
-    \frac{8\gamma}{9c}(f(\theta_m) - f(\theta^*)) \leq \frac{4}{9c}||\nabla_\theta f(\theta_m)||^2 \leq p_m^TH(\theta_m)p_m,
+    \frac{8\gamma}{9c}\big(f(\theta_m) - f(\theta^*)\big) \leq \frac{4}{9c}\, \big\|\nabla_\theta f(\theta_m)\big\|^2 \leq p_m^TH(\theta_m)\, p_m,
 \end{equation}
 \begin{equation}
-    f(\theta_m) - f(\theta^*) \leq \frac{9c}{8\gamma}p_m^TH(\theta_m)p_m,
+    f(\theta_m) - f(\theta^*) \leq \frac{9c}{8\gamma}\, p_m^TH(\theta_m)\, p_m,
 \end{equation}
 \begin{equation}
-    -\frac{8\gamma}{9c}(f(\theta_m) - f(\theta^*)) \geq -\frac{4}{9c}||\nabla_\theta f(\theta_m)||^2 \geq -p_m^TH(\theta_m)p_m.
+    -\frac{8\gamma}{9c}\big(f(\theta_m) - f(\theta^*)\big) \geq -\frac{4}{9c}\, \big\|\nabla_\theta f(\theta_m)\big\|^2 \geq -p_m^TH(\theta_m)\, p_m.
 \end{equation}
@@
 \begin{align}
-    f(\theta_{m+1})-f(\theta_m) &\leq -\eta p_m^TH(\theta_m)p_m + \epsilon\eta||p_m|| + \eta^2 \Gamma ||p_m||^2 / 2 \\
-    &\leq -\eta\beta p_m^TH(\theta_m)p_m \\
-    &\leq -\eta\beta\frac{8\gamma}{9c}(f(\theta_m) - f(\theta^*)).
+    f(\theta_{m+1})-f(\theta_m) &\leq -\eta \, p_m^TH(\theta_m)p_m + \epsilon\eta\|p_m\| + \eta^2 \Gamma \|p_m\|^2 / 2 \\
+    &\leq -\eta\beta \,  p_m^TH(\theta_m)\, p_m \\
+    &\leq -\eta\beta \, \frac{8\gamma}{9c}\, \big(f(\theta_m) - f(\theta^*)\big).
 \end{align}
@@
     &= f(\theta_{m+1})-f(\theta_m)+f(\theta_m)- f(\theta^*) \\
-    &\leq f(\theta_m)- f(\theta^*) -\eta\beta\frac{8\gamma}{9c}(f(\theta_m) - f(\theta^*)) \\
-    &= (1-\eta\beta\frac{8\gamma}{9c})(f(\theta_m) - f(\theta^*)).
+    &\leq f(\theta_m)- f(\theta^*) -\eta\beta\, \frac{8\gamma}{9c} \, \big(f(\theta_m) - f(\theta^*)\big) \\
+    &= \Big(1-\eta\beta\frac{8\gamma}{9c}\Big)\big(f(\theta_m) - f(\theta^*)\big).
 \end{align}

### 2024-06-14 paper/ms.tex
another day, another dollar

@@
             $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$. \newline
-            \hspace*{4mm} Compute $L_n^{B,\theta,\alpha} ={\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}\, \big/\, {\sum_{j=1}^J  w^{P,\theta}_{n,j}}$. \newline
+            \hspace*{4mm} Compute $L_n^{B,\theta,\alpha} ={\sum_{j=1}^Jg^\theta_{n,j} \, w^{P,\theta}_{n,j}}\, \big/\, {\sum_{j=1}^J  w^{P,\theta}_{n,j}}$. \newline
             \hspace*{4mm} Conditional likelihood under $\phi$,
@@
             \hspace*{4mm} Calculate resampled corrected weights
-            $w_{n,j}^{F,\theta}= w^{P,\theta}_{n,k_j}  { g^{\theta}_{n,k_j}} \, \big/ \, { g^{\phi}_{n,k_j}}$.\newline
+            $w_{n,j}^{F,\theta}= w^{P,\theta}_{n,k_j} \, g^{\theta}_{n,k_j} \, \big/ \, { g^{\phi}_{n,k_j}}$.\newline
             \hspace*{4mm} Compute $ L_n^{A,\theta,\alpha} = L_n^\phi\cdot {\sum_{j=1}^J w^{F,\theta}_{n,j}} \, \big/ \, {\sum_{j=1}^J  w^{P,\theta}_{n,j}}$.\newline
@@
 %% DISCUSSED IN THE PREVIOUS SECTION
-The resampling indices, which we write as $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$, are a function of $\phi$ for any value of $\theta$.
+The resampling indices, which we write as $k_j \sim \text{Categorical}\big(g^{\phi}_{n,1},...,g^{\phi}_{n,J}\big)$, are a function of $\phi$ for any value of $\theta$.
 For the categorical distribution, we use systematic resampling \cite{arulampalam02,king16} which usually has superior performance to multinomial resampling.
@@
      \label{eq:mop-conditional-likelihood}
-     L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}, \hspace{15mm} 
+     L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j}\, w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}, \hspace{15mm} 
      L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
@@
     \hspace{5mm}
-    w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}, 
+    w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} \, g^{\theta}_{n,k_j} \, \big/ g^{\phi}_{n,k_j}, 
     \hspace{5mm}
@@
 
-Consider the second stage of IFAD (Algorithm \ref{alg:ifad}) where one stops if $||\nabla_\theta \hat\ell^\alpha(\theta_m)|| \leq (1+\sigma) \epsilon$, where $\sigma \geq \frac{4 \Gamma}{(1-\beta)}$, for some $\beta \in (0,1)$.
+Consider the second stage of IFAD (Algorithm \ref{alg:ifad}) where one stops if $\|\nabla_\theta \hat\ell^\alpha(\theta_m)\| \leq (1+\sigma) \epsilon$, where $\sigma \geq \frac{4 \Gamma}{(1-\beta)}$, for some $\beta \in (0,1)$.
 Assume $-\ell$ is strongly convex and smooth, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$.
@@
 $$
-\ell(\theta^*) - \ell(\theta_{m+1}) \leq \left(1-\eta\beta\frac{8\gamma}{9c}\right)(\ell(\theta^*)-\ell(\theta_m)).
+\ell(\theta^*) - \ell(\theta_{m+1}) \leq \Big(1-\eta\beta\, \frac{8\gamma}{9c}\Big)\big(\ell(\theta^*)-\ell(\theta_m)\big).
 $$
@@
 		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_m).$ \newline
-		\hspace*{4mm} Obtain $g(\theta_m) = \nabla_{\theta_m} (-\hat\loglik(\theta_m))$, $H(\theta_m)$ s.t. $\lambda_{\min}(H(\theta_m)) \geq c$. \newline
+		\hspace*{4mm} Obtain $g(\theta_m) = \nabla_{\theta_m} \big(-\hat\loglik(\theta_m)\big)$, $H(\theta_m)$ s.t. $\lambda_{\min}\big(H(\theta_m)\big) \geq c$. \newline
 		\hspace*{4mm} Update $\theta_{m+1} := \theta_m - \eta (H(\theta_m))^{-1} g(\theta_m)$, $m:=m+1.$ \newline
@@
     dS&=\big(k \epsilon R^k+\delta(S-P)-\lambda(t) S\big)\, dt+d P-({\sigma S I}/{H})\, dB, \\
-    dI&=\left(\lambda(t) S-(m+\delta+\gamma) I\right)\, dt+({\sigma S I}/{H})\, dB, \\
+    dI&=\big(\lambda(t) S-(m+\delta+\gamma) I\big)\, dt+({\sigma S I}/{H})\, dB, \\
     dR^1&=\big(\gamma I-(k \epsilon+\delta) R^1\big)\, dt, \hspace{2mm} \dots \\
@@
 \vspace*{-2mm}
-\begin{equation*}
-    \lambda_t=\exp\hspace{-1mm}\left(\hspace{-.5mm}\beta_{\text{trend}}(t-t_0)+\hspace{-1mm}\sum_{j=1}^{6} \beta_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm}\frac{I}{P} + \exp\hspace{-1mm} \left(\sum_{j=1}^{6} \omega_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm},
+\begin{equation*}    \lambda_t=\exp\hspace{-1mm}\left(\hspace{-.5mm}\beta_{\text{trend}}(t-t_0)+\hspace{-1mm}\sum_{j=1}^{6} \beta_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm}\frac{I}{P} + \exp\hspace{-1mm} \left(\sum_{j=1}^{6} \omega_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm},
     \vspace*{-2mm}
@@
 \label{assump:bounded-process}
-    \item \textbf{Bounded Measurement Model.} There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G}<\infty$ and there exists $G'(\theta)$ with $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty \leq G'(\theta)< \infty$.
+    \item \textbf{Bounded Measurement Model.} There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G}<\infty$ and there exists $G'(\theta)$ with $\|\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)\|_\infty \leq G'(\theta)< \infty$.
 \label{assump:bounded-measurement}
@@
     \begin{equation} \nonumber
-        \frac{\sum_{j=1}^J h(x_{n,j}^{F, \theta}) \, w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{F,\theta}} \stackrel{a.s.}{\to} E_{\pi_n(\theta)} [h(X_n)], \hspace{5mm} \hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).
+        \frac{\sum_{j=1}^J h(x_{n,j}^{F, \theta}) \, w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{F,\theta}} \stackrel{a.s.}{\to} E_{\pi_n(\theta)} \big[h(X_n)\big], \hspace{5mm} \hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).
     \arxiv{}{\vspace*{-2.5mm}}
@@
     When $\alpha=1$ and $\theta\neq\phi,$ suppose inductively that $\{(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j})\}_{j=1}^J$ targets $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
-    It can then be shown that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j})\}_{j=1}^J$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$, that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} )\}_{j=1}^J$ targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$, and that weighting the particles by $(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}) = (X^{P,\theta}_{n,k_j}, w^{P,\theta}_{n,k_j} g^\theta_{n,k_j}/ g^\phi_{n,k_j})$,
+    It can then be shown that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j})\}_{j=1}^J$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$, that $\{(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} \, g^\theta_{n,j} )\}_{j=1}^J$ targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$, and that weighting the particles by $(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}) = (X^{P,\theta}_{n,k_j}, w^{P,\theta}_{n,k_j} \, g^\theta_{n,k_j} \big/ g^\phi_{n,k_j})$,
     resampling the $k_j$ with probabilities proportional to $g^\phi_{n,j}$, also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
@@
     \vspace*{-1ex}
-    \begin{align*}
-        &\E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 
+    \begin{eqnarray*}
+        \E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)\big\|_2^2 
+        &\lesssim& \min_{k \leq N} Np \, G'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi(\alpha)\right),
         \\
-        &\lesssim \min_{k \leq N} NpG'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi(\alpha)\right),\\[-1ex]
-        &\Var(\nabla_\theta \hat\ell^{\alpha}(\theta)) \lesssim \min_{k\leq N} NpG'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N\right).
-        \end{align*}
+        \Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) &\lesssim& \min_{k\leq N} Np \, G'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N\right).
+        \end{eqnarray*}
 \end{thm}
@@
 The MSE bound considers the error between the score and MOP-$(1,k)$, and the error between MOP-$(1,k)$ and MOP-$\alpha$. 
-The latter is $\tilde{O}(NpG'(\theta)^2(k+\psi(\alpha)))$, by routine algebra, and the former can be controlled with a result on the the forgetting of the particle filter from \cite{karjalainen23} and the very same $L^p$ error bound mentioned above.
+The latter is $\tilde{O}\big(Np \, G'(\theta)^2(k+\psi(\alpha))\big)$, by routine algebra, and the former can be controlled with a result on the the forgetting of the particle filter from \cite{karjalainen23} and the very same $L^p$ error bound mentioned above.
 
-We also show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( NpG'(\theta)^2\big/J \big)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p$ and $G'$.
+We also show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( Np \, G'(\theta)^2\big/J \big)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p$ and $G'$.
 Combining that result with Theorem \ref{thm:mop-biasvar} shows that MOP-$\alpha$ interpolates between MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$.

### 2024-06-15 paper/appendix/biasvar.tex
Kevin's edits

@@
 Now, to illustrate the proof strategy for the general case in an easier context, we first analyze the special case when $\alpha=0.$ We now prove the variance bound for this case, presented below.
+To our knowledge, this result for the variance of the gradient estimate of \cite{naesseth18} is novel.
 
-\begin{thm}[Variance of Gradient Estimate of MOP-$0$]
-The variance of the gradient estimate from MOP-$0$, or the gradient estimate resulting from the algorithm of \cite{naesseth18}, is:
+\begin{thm}[Variance of MOP-$0$ Gradient Estimate]
+The variance of the gradient estimate from MOP-$0$, i.e. the algorithm of \cite{naesseth18}, is:
  \begin{equation}
-     \Var(\nabla_\theta \hat\ell(\theta)) \lesssim \frac{G'(\theta)^2N}{J}.
+     \Var(\nabla_\theta \hat\ell^0(\theta)) \lesssim \frac{NpG'(\theta)^2}{J}.
  \end{equation}
@@
 \begin{equation}
-    \hat{\lik}(\theta) := \prod_{n=1}^N L_n^{A, \theta, \alpha} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
+    \hat{\lik}^0(\theta) := \prod_{n=1}^N L_n^{A, \theta, 0} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta,0}}{\sum_{j=1}^J w_{n,j}^{P,\theta,0}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
 \end{equation}
-Similarly to the proof of Lemma \ref{lem:mop-0-formula}, its gradient when $\theta=\phi$ is therefore 
+Similarly to the proof of Lemma \ref{lem:mop-0-formula}, we define $s_{n,j}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}$, and from the exact same arguments conclude that its gradient when $\theta=\phi$ is therefore
 \begin{align}
-    \nabla_\theta \hat{\ell}(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
+    \nabla_\theta \hat{\ell}^0(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
     &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)} \\
@@
 where we use the log-derivative trick, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again.
-The point here was that we establish that the gradient of the log-likelihood estimate is given by the sum of terms over all $N$ and $J$:
+We do this to establish that the gradient of the log-likelihood estimate is given by the sum of terms over all $N$ and $J$:
 \begin{equation}
-\nabla_\theta \hat{\ell}(\theta) = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right).
+\nabla_\theta \hat{\ell}^0(\theta) = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right).
 \end{equation}
-Therefore,
+Therefore, for a given $\theta_i$ in the parameter vector $\theta$,
 \begin{align}
-    \Var\big(\nabla_\theta \hat\ell(\theta)\big) &= \frac{1}{J^2}\Var\left(\sum_{n=1}^N\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    & \hspace{-15mm}= \frac{1}{J^2}\sum_{n=1}^N\Var\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &\hspace{-15mm}+ 2\sum_{m<n}\Cov\left(\frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &\hspace{-15mm}= \sum_{n=1}^N \Var\big(\nabla_\theta\hat\ell_n(\theta)\big) + 2\sum_{m<n} \Cov\big(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta)\big).
+    \Var\left(\frac{\partial}{\partial_{\theta_i}} \hat\ell^0(\theta)\right) &= \frac{1}{J^2}\Var\left(\sum_{n=1}^N\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    & \hspace{-15mm}= \frac{1}{J^2}\sum_{n=1}^N\Var\left(\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    &\hspace{-15mm}+ 2\sum_{m<n}\Cov\left(\frac{1}{J}\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \frac{1}{J}\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    &\hspace{-15mm}= \sum_{n=1}^N \Var\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) + 2\sum_{m<n} \Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta),\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right).
 \end{align}
@@
 Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
-Therefore, by Davydov's inequality, noting that $\nabla_\theta\hat\ell_n(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and without loss of generality labeling $m$ and $n$ such that $\E\big[(\nabla_\theta\hat\ell_m(\theta))^4\big]^{1/4}\leq\E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big]^{1/4}$, we see that
+Therefore, by Davydov's inequality, noting that $\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0^0(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and without loss of generality labeling $m$ and $n$ such that $\E\big[(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0^0(\theta))^4\big]^{1/4}\leq\E\big[(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0^0(\theta))^4\big]^{1/4}$, we see that
 \begin{align}
-    \Cov\big(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)\big) 
-    &\leq \alpha(n-m)^{1/2} \, \E\big[(\nabla_\theta\hat\ell_m(\theta))^4\big]^{1/4}
+    \Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0^0(\theta), \frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0^0(\theta)\right) 
+    &\leq \alpha(n-m)^{1/2} \, \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0^0(\theta)\right)^4\right]^{1/4}
     \,
-    \E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big]^{1/4}\\
-    &\leq \alpha(n-m)^{1/2}\, \E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big]^{1/2}.
+    \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0^0(\theta)\right)^4\right]^{1/4}\\
+    &\leq \alpha(n-m)^{1/2}\, \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0^0(\theta)\right)^4\right]^{1/2}.
 \end{align}
@@
 \begin{equation}
-\E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big] = \E\Big[\big(\nabla_\theta\hat\ell_n(\theta)-\E\big[\nabla_\theta\hat\ell_n(\theta)\big]\big)^4\Big]+\E\big[(\nabla_\theta\hat\ell_n(\theta))^2\big]^2
+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right] = \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)-\E\left[\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right]\right)^4\right]+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^2\right]^2
 \end{equation}
@@
 \begin{equation}
-\E\Big[\big(\nabla_\theta\hat\ell_n(\theta)-\E\big[\nabla_\theta\hat\ell_n(\theta)\big]\big)^4\Big] \lesssim \frac{G'(\theta)^4}{J^2}, \hspace{10mm} \E\Big[\big(\nabla_\theta\hat\ell_n(\theta)-\E\big[\nabla_\theta\hat\ell_n(\theta)\big]\big)^2\Big] \lesssim \frac{G'(\theta)^2}{J}.
+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)-\E\left[\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right]\right)^4\right] \lesssim \frac{G'(\theta)^4}{J^2}, \hspace{10mm} \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)-\E\left[\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right]\right)^2\right] \lesssim \frac{G'(\theta)^2}{J}.
 \end{equation}
@@
 \begin{equation}
-\E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big]^{1/2} \lesssim  \sqrt{\frac{G'(\theta)^4}{J^2}+\left(\frac{G'(\theta)^2}{J}\right)^2} = \frac{G'(\theta)^2}{J},
+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right]^{1/2} \lesssim  \sqrt{\frac{G'(\theta)^4}{J^2}+\left(\frac{G'(\theta)^2}{J}\right)^2} = \frac{G'(\theta)^2}{J},
 \end{equation}
@@
 \begin{equation}
-\Cov\big(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)\big) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{G'(\theta)^2}{J}.
+\Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta), \frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{G'(\theta)^2}{J}.
 \end{equation}
@@
 \begin{align}
-    \Var\big(\nabla_\theta \hat\ell(\theta)\big) &= \sum_{n=1}^N \Var\big(\nabla_\theta\hat\ell_n(\theta)\big) + 2\sum_{m<n} \Cov\big(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta)\big)\\
+    \Var\left(\frac{\partial}{\partial_{\theta_i}} \hat\ell(\theta)\right) &= \sum_{n=1}^N \Var\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) + 2\sum_{m<n} \Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta),\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)\\
     &\leq \frac{NG'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\frac{G'(\theta)^2}{J} \\
@@
 \end{align}
+
+It then follows that as $\theta \in \Theta \subseteq \R^p$, 
+\begin{equation}
+    \Var\big(\nabla_{\theta} \hat\ell(\theta)\big) \lesssim \frac{NpG'(\theta)^2}{J}.
+\end{equation}
 \end{proof}
@@
 If one divides by $N$, the error is $O(1/\sqrt{NJ})$.
+
+The variance here of $\frac{NpG'(\theta)^2}{J}$ can be thought of as a lower bound for the variance of MOP-$\alpha$.
+We will later see that the variance of MOP-$\alpha$ contains this term, and an additional term that corresponds to the memory of the MOP-$\alpha$ gradient estimate.
+
+Although the variance of MOP-$0$ is desirable, it unfortunately has large uncontrollable asymptotic bias.
+Error bounds for the bias and MSE of MOP-$0$ \cite{naesseth18} are also fortunately special cases (where $k=1$, $\alpha=0$) of the bias and MSE analysis we conduct for MOP-$\alpha$.
+We will later see that these showcase the large uncontrollable asymptotic bias of MOP-$0$.
+
+\subsection{MOP-$\alpha$ Variance Bound}
+
+We are now in a position to tackle the MOP-$\alpha$ variance bound.
+Here, we analyze the case for $\alpha \in (0,1)$.
+The proof strategy is as follows.
+Instead of analyzing the variance of the MOP-$\alpha$ gradient estimate proper, we will analyze the variance of a modified estimator where we truncate the weights at $k$ timesteps, so this estimator only looks at most $k$ timesteps back while accumulating the importance weights.
+This enables us to establish strong mixing for the truncated estimator, in order to get a bound on the variance that is $O(NJ^{-1})$.
+We emphasize that this estimator is only a theoretical construct -- we do not actually use this estimator.
+
+Coupled with a bound ensuring that the $k$-truncated estimator provides an estimate that is close to that of MOP-$\alpha$ proper, we get a bound on the variance of MOP-$\alpha$ that comprises of the variance of the $k$-truncated estimator plus the error, for any $k \leq N$.
+It then holds that the final variance bound on MOP-$\alpha$ is given by the minimum over all $k$, and is never larger than $O(\psi(\alpha)+NJ^{-1})$ for some function $\psi$ increasing in $\alpha$.
+
+\begin{thm}[Variance of MOP-$\alpha$ Gradient Estimate]
+    When $\alpha \in (0,1)$, the variance of the gradient estimate from MOP-$\alpha$ is
+    $$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \min_{k\leq N} \left(\frac{k^2G'(\theta)^2Np}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}NpG'(\theta)^2 \right).$$
+\end{thm}
+\begin{proof}
+
+Using the log-derivative trick and that $w_{n,j}^{F,\theta} = w_{n,j}^{F,\theta,k} = 1$ when $\theta=\phi$,
+\begin{align}
+    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2\\
+    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta}}{{\sum_{j=1}^J w_{n,j}^{F,\theta}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,k}}}\right\rVert_2^2 \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,k}\right\rVert_2^2 \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta}}{w_{n,j}^{F,\theta}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,k}}{w_{n,j}^{F,\theta,k}}\right\rVert_2^2\\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2.
+\end{align}
+
+This lets us bound the cumulative weight discrepancies by
+\begin{align}
+    &  \left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2\\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2 \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta}\right)-\log\left(w_{n,j}^{F,\theta,k}\right)\right)\right\rVert_2^2\\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=n-k}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_2^2 \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^{n-k}\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}pG'(\theta)^2\\
+    &\leq pG'(\theta)^2\frac{\alpha^k-\alpha^n}{1-\alpha},
+\end{align}
+where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
+
+We can then bound $||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||$:
+\begin{align}
+    &\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_2^2\\
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_2^2\\
+    &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\\
+    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_2^2 \\
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)
+    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\Bigg)\right\rVert_2^2 \\
+    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)\right)\right\lVert_{\infty}
+    +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\right)\right\rVert_2^2,
+\end{align}
+and each of these two terms is bounded by $\frac{\alpha^k}{1-\alpha}NpG'(\theta)^2$. So we now know that
+$$\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_2^2 \leq  \frac{2\alpha^k}{1-\alpha}NpG'(\theta)^2,$$
+which is our desired error bound.
+
+Now, we bound the variance of $\hat s_k(\theta)$. 
+Recall that we defined
+$$\hat s_k(\theta) = \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) = \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right).$$
+
+We can then decompose this expression with the log-derivative trick while noting that $w_{n,j}^{F,\theta, k} = 1$ whenever $\theta=\phi$, to see that
+\begin{align}
+    \hat s_k(\theta) 
+    &= \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right) \\
+    &= \sum_{n=1}^N \left(\frac{\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}}{\sum_{j=1}^J w_{n,j}^{F,\theta, k}}-\frac{\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}}{\sum_{j=1}^J w_{n-1,j}^{A,F,\theta, k}}\right) \\
+    &= \sum_{n=1}^N \left(\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}\right)\\
+    &= \frac{1}{J}\sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right).
+\end{align}
+
+It now follows that we need to bound the variance at a single timestep, namely
+$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right).$$
+
+We use the log-derivative trick yet again to find, noting that $w_{n,j}^{F,\theta, k}=1$ when $\theta=\phi$, that
+\begin{align}
+    \nabla_\theta w_{n,j}^{F,\theta, k} = \frac{\nabla_\theta w_{n,j}^{F,\theta, k}}{w_{n,j}^{F,\theta, k}} = \nabla_\theta \log(w_{n,j}^{F,\theta, k}) = \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) = \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right).
+\end{align}
+
+Then,
+\begin{align}
+    \text{Var}\left(\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right) 
+    &= \text{Var}\left(\nabla_\theta \left(\sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right)- \sum_{i=n-k-1}^{n-1} \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right)\right)\right) \\
+    &= \text{Var} \left(\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1}\log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right).
+\end{align}
+
+Note that $$\hat s_{n,k}(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)$$ is a function bounded by $C\frac{G'(\theta)(1-\alpha^k)}{1-\alpha} =: CG'(\theta)r$ in each coordinate for some constant $C$ (by Assumption \ref{assump:bounded-measurement}) that depends only on $k$ timesteps, and not $n$.
+We can therefore invoke Lemma 2 of \cite{karjalainen23} to bound the $L_2$ error of each of the $k$ (from time $n-k$ to time $n$) functionals from its expectation under the posterior by $Cr\sqrt{p}G'(\theta)J^{-1/2}k$.
+That is, we have that
+$$\E\left[||\hat s_{n,k}(\theta) - \E_{\pi}\hat s_{n,k}(\theta)||_2^2\right]^{1/2} \leq \frac{Cr\sqrt{p}G'(\theta) k}{\sqrt{J}}.$$
+
+By the bias-variance decomposition, this in turn bounds the variance at a single timestep by $C^2r^2pG'(\theta)^2J^{-1}k^2$. Concretely, we have that
+$$\text{Var}\left(\frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \lesssim \frac{r^2k^p2G'(\theta)^2}{J}.$$
+
+It now remains to bound the variance of $\hat s_k(\theta)$ by considering the covariance of each of the $N$ terms that comprise it. We decompose
+\begin{align}
+    \text{Var}(\nabla_\theta \hat s_k(\theta)) &= \text{Var}\left(\sum_{n=1}^N \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \\
+    &= \sum_{n=1}^N\text{Var}\left(\hat s_{n,k}(\theta)\right)+ 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \\
+    &\lesssim \frac{r^2k^2pG'(\theta)^2N}{J} + 2\sum_{m<n}\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right).
+\end{align}
+
+Similarly to the proof of the MOP-$0$ case, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
+We know from Theorem 3 of \cite{karjalainen23} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
+$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
+i.e.
+the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
+By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strongly mixing, with $\alpha$-mixing coefficients $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c\log(J))}}$.
+Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
+
+We now derive the $\alpha$-mixing coefficients of $(\hat s_{n,k}(\theta))_{n=1}^N$.
+Observe that $\hat s_{n,k}$ is strong mixing at lag $k+1$, as all weights beyond $k$ timesteps are truncated.
+We therefore have that the mixing time of $\hat s_k(\theta)$ is $O(1+k+\log(J))$, and that the $\alpha$-mixing coefficients for $(\hat s_{n,k}(\theta))_{n=1}^N$ are given by $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c(1+k+\log(J)))}}$.
+
+Therefore, by Davydov's inequality and Lemma 2 of \cite{karjalainen23}, and noting that $\hat s_{n,k}(\theta) \leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, by a similar argument to the MOP-0 case, we find that
+
+$$\text{Cov}\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2pG'(\theta)^2}{J}.$$
+
+Concretely, noting that $\nabla_\theta\hat s_{n,k}(\theta)\leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and WLOG assuming $\E[(\nabla_\theta\hat\hat s_{m,k}(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat\hat s_{n,k}(\theta))^4]^{1/4}$, we apply Davydov's inequality to see that
+\begin{align}
+    \text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) 
+    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat s_{m,k}(\theta))^4]^{1/4}\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/4}\\
+    &\leq \alpha(n-m)^{1/2}\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2}.
+\end{align}
+
+To bound this, we use the fact that $$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4] = \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4]+\E[(\nabla_\theta\hat s_{n,k}(\theta))^2]^2$$
+alongside Lemma 2 of \cite{karjalainen23}, which alongside the fact that
+$$\E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^4] \lesssim \frac{r^4p^2G'(\theta)^4}{J^2}, \;\; \E[(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)])^2] \lesssim \frac{r^2pG'(\theta)^2}{J},$$
+allows us to find that 
+$$\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/2} \lesssim  \sqrt{\frac{r^4p^2G'(\theta)^4}{J^2}+\left(\frac{r^2pG'(\theta)^2}{J}\right)^2} = \frac{r^2pG'(\theta)^2}{J},$$
+and conclude that 
+$$\text{Cov}(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2pG'(\theta)^2}{J}.$$
+
+Putting it all together, we see that
+\begin{align}
+    \text{Var}(\nabla_\theta \hat s_{k}(\theta)) &= \sum_{n=1}^N \text{Var}(\nabla_\theta\hat s_{n,k}(\theta)) + 2\sum_{m<n} \text{Cov}(\nabla_\theta\hat s_{m,k}(\theta),\nabla_\theta\hat s_{n,k}(\theta))\\
+    &\leq \frac{Nr^2pk^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\frac{r^2pG'(\theta)^2}{J} \\
+    &\leq \frac{r^2pk^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\right) \\
+    &\lesssim \frac{k^2r^2pG'(\theta)^2N}{J}.
+\end{align}
+
+We will now use this result to bound $\text{Var}(\nabla_\theta \hat s_k(\theta)$ proper. For random variables $X,Y$ where $|X-Y|$ is bounded almost surely by some $M$, we have that 
+$$X-\E X - 2M \leq Y - \E Y \leq X - \E X + 2M$$
+so,
+$$(Y-\E Y)^2 \leq (X-\E X - 2M)^2 + (X-\E X +2M)^2.$$
+Taking expectations, we get
+$$\text{Var}(Y) \leq \text{Var}(X) + 8M^2.$$
+
+It follows from this result and the result we proved earlier, namely that
+$$||\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)||_2^2 \leq \frac{2\alpha^k}{1-\alpha}NpG'(\theta)^2,$$
+that the variance of MOP-$\alpha$ proper is bounded by, for any $k \leq N$,
+that the variance of MOP-$\alpha$ proper is bounded by
+$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \frac{k^2r^2G'(\theta)^2Np}{J} + \frac{\alpha^{k}}{1-\alpha}NpG'(\theta)^2.$$
+
+As the above holds for any $k \leq N$,
+$$\text{Var}(\nabla_\theta \hat\ell(\theta)) \lesssim \min_{k\leq N} \left(\frac{k^2pG'(\theta)^2Np}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}NpG'(\theta)^2 \right).$$
+\end{proof}
+
+\subsection{MOP-$\alpha$ MSE Bound}
+
+\begin{thm}[MSE of MOP-$\alpha$]
+    When $\alpha \in (0,1)$, the MSE of MOP-$\alpha$ is given by
+    \begin{equation}
+        \E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 \lesssim \min_{k \leq N} NpG'(\theta)^2\left(\frac{k^2}{J}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\frac{\alpha^k  + \alpha^{k+1} - \alpha}{1-\alpha}\right).
+    \end{equation}
+\end{thm}
+
+\begin{proof}
+
+The broad idea is that the first term can be controlled by a mixing argument, the second term garners $O(Nk)$ error, and the third term is $O\left(\frac{N}{1-\alpha}\right)$.
+It is unclear whether the bias can be reduced further -- the gradient is a sum of $N$ terms, and is therefore $O(N)$ itself.
+
+The first term is the problem term.
+We will first bound the ground truth of conditional scores from particle approximation conditional on correct filtering distribution at time $n-k$, and then bound that from the particle approximation resulting from an arbitrary filtering distribution at time $n-k$.
+That is, where $\hat\pi_n$ is the particle approximation at time $n$ of the posterior $\pi_n$ and we write $\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)$ for the truncated MOP-$1$ conditional score estimate given the correct filtering distribution at time $n-k$, we can first observe that
+\begin{equation}
+    ||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 =  \left\lVert\sum_{n=1}^N \nabla_\theta \hat\ell_n^\alpha(\theta) - \sum_{n=1}^N \nabla_\theta \ell_n(\theta)\right\rVert_2^2 \leq \sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_2^2,
+\end{equation}
+and decompose
+\begin{align}
+    &\sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_2^2 \\
+    &\leq \sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 + \sum_{n=1}^N||\nabla_\theta\hat s_{n,k}^1(\theta) - \nabla_\theta\hat s_{n,k}^\alpha(\theta)||_2^2 + \sum_{n=1}^N||\nabla_\theta\hat s_{n,k}^\alpha(\theta) -  \nabla_\theta\hat\ell_n^\alpha(\theta)||_2^2.
+\end{align}
+As before, we bound the first term, $\sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2$, by decomposing it into two terms, 
+$$\sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 \leq \sum_{n=1}^N ||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 + \sum_{n=1}^N ||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2.$$
+
+The first term is a particle approximation dependent on $k$ timesteps, so by Lemma 2 of \cite{karjalainen23}, this is bounded by
+$$\E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 \leq \frac{CpG'(\theta)^2k^2}{J}.$$
+
+The second term amounts to bounding the difference between functionals of two different particle measures that mix under the same Markov kernel.
+Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
+We know from Theorem 3 of \cite{karjalainen23} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
+$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
+i.e.
+the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
+
+Then, we can bound 
+$\E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2$ by
+$$\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}} = \beta_{TV}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
+implying that
+\begin{align}
+    \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 
+    &\lesssim \sup_{\mu, \nu} \sup_{||\psi||_\infty \leq 1/2} pG'(\theta)^2\left|(\mu \textbf{M}_{n,n+k})(\psi)-(\nu \textbf{M}_{n,n+k})(\psi)\right| \\
+    &\leq \sup_{\mu, \nu} pG'(\theta)^2\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}} \\
+    &\leq pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} ||\hat\pi_{n-k} - \pi_{n-k}||_{\text{TV}} \\
+    &\leq pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} \sup_{||\psi||_{\infty} \leq 1/2} |\hat\pi_{n-k}(\psi) - \pi_{n-k}(\psi)| \\
+    &\lesssim pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}}.
+\end{align}
+
+Therefore, we have that
+\begin{align}
+    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_2^2 \\
+    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 + \sum_{n=1}^N \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2\\
+    &\lesssim N\frac{CpG'(\theta)^2k^2}{J} + NpG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}}.
+\end{align}
+
+Now that the first term $\E||\nabla_\theta\ell(\theta)||_2^2$ is taken care of, it remains to bound $$\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2 \text{ and } \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_2^2.$$
+
+We see, by a similar argument to the variance bound, that
+\begin{align}
+    &\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2\\
+    &\leq \sum_{n=1}^N \E\left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right|
+    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\right|\Bigg)\right\rVert_2^2.
+\end{align}
+Each of these terms can be bounded using the log-derivative trick and that $w_{n,j}^{F,\theta,1,k} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
+\begin{align}
+    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_{2}^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})pG'(\theta)^2\\
+    &\leq pG'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right),
+\end{align}
+where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
+
+So, 
+\begin{align}
+    \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty
+    \leq 2NpG'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) 
+    \leq 2NpG'(\theta)^2k.
+\end{align}
+
+Now, we address the third term, $\E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty$.
+From the variance argument, we know that we can bound $||\nabla_\theta\hat\ell^\alpha(\theta) - \hat s_k^\alpha(\theta)||$:
+\begin{align}
+    &\left\lVert\nabla_\theta\hat\ell^\alpha(\theta) - \hat s_k^\alpha(\theta) \right\rVert_2^2\\
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_2^2\\
+    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right)\right\lVert_2^2
+    +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta, \alpha}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha, k}\right)\right)\right\rVert_2^2.
+\end{align}
+
+Using the log-derivative trick and that $w_{n,j}^{F,\theta,\alpha} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
+\begin{align}
+    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}pG'(\theta)^2\\
+    &\leq pG'(\theta)^2\frac{\alpha^k-\alpha^n}{1-\alpha},
+\end{align}
+where the second-last line follows from Assumption \ref{assump:bounded-measurement}. Putting it together and taking expectations on both sides, we obtain
+$$\E\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_2^2 \leq  \frac{2\alpha^k}{1-\alpha}NpG'(\theta)^2,$$
+which is our desired error bound.
+
+Therefore, our decomposition yields the MSE bound
+\begin{align}
+    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 \\
+    &\leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_2^2 + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2 + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_2^2 \\
+    &\lesssim \min_{k \leq N} \left(N\frac{CpG'(\theta)^2k^2}{J} + NpG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} + 2NpG'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) + 2\frac{\alpha^k}{1-\alpha}NpG'(\theta)^2\right) \\
+    &\lesssim \min_{k \leq N} NpG'(\theta)^2\left(\frac{k^2}{J}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\frac{\alpha^k  + \alpha^{k+1} - \alpha}{1-\alpha}\right).
+\end{align}
+
+\end{proof}
+
+\begin{cor}[MSE of MOP-$0$]
+    The MSE of MOP-$0$, i.e. the estimator of \cite{naesseth18}, is
+    \begin{equation}
+        \E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^0(\theta)||_2^2 \lesssim NpG'(\theta)^2\left(\frac{1}{J}+(1-\epsilon)^{\floor{1/(c\log(J))}}\right).
+    \end{equation}
+\end{cor}
+\begin{proof}
+    Observe that MOP-$0$ is equivalent to MOP-$(1,1)$.
+Noting that $k=1$ and that we only need to bound the difference between $\nabla_\theta \ell(\theta)$ and $\nabla_\theta \hat s_1^1(\theta)$, repeating the above arguments almost verbatim allows us to obtain this result.
+
+As before, we bound the first term, $\sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)||_2^2$, by decomposing it into two terms, 
+$$\sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)||_2^2 \leq \sum_{n=1}^N ||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta)||_2^2 + \sum_{n=1}^N ||\nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)||_2^2.$$
+
+The first term is a particle approximation dependent on $1$ timesteps, so by Lemma 2 of \cite{karjalainen23}, this is bounded by
+$$\E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta)||_2^2 \leq \frac{CpG'(\theta)^2}{J}.$$
+
+The second term amounts to bounding the difference between functionals of two different particle measures that mix under the same Markov kernel.
+Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
+We know from Theorem 3 of \cite{karjalainen23} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
+$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
+i.e.
+the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
+
+Then, we can bound 
+$\E||\nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)||_2^2$ by
+$$\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu \textbf{M}_{n,n+1}-\nu \textbf{M}_{n,n+1}\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}} = \beta_{TV}(\textbf{M}_{n,n+1}) \leq (1-\epsilon)^{\floor{1/(c\log(J))}},$$
+implying that
+\begin{align}
+    \E||\nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)||_2^2 
+    &\lesssim \sup_{\mu, \nu} \sup_{||\psi||_\infty \leq 1/2} pG'(\theta)^2\left|(\mu \textbf{M}_{n,n+1})(\psi)-(\nu \textbf{M}_{n,n+1})(\psi)\right| \\
+    &\leq \sup_{\mu, \nu} pG'(\theta)^2\|\mu \textbf{M}_{n,n+1}-\nu \textbf{M}_{n,n+1}\|_{\mathrm{TV}} \\
+    &\leq pG'(\theta)^2(1-\epsilon)^{\floor{1/(c\log(J))}} ||\hat\pi_{n-1} - \pi_{n-1}||_{\text{TV}} \\
+    &\leq pG'(\theta)^2(1-\epsilon)^{\floor{1/(c\log(J))}} \sup_{||\psi||_{\infty} \leq 1/2} |\hat\pi_{n-1}(\psi) - \pi_{n-1}(\psi)| \\
+    &\lesssim pG'(\theta)^2(1-\epsilon)^{\floor{1/(c\log(J))}}.
+\end{align}
+
+Therefore, we have that
+\begin{align}
+    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_1^1(\theta)||_2^2 \\
+    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta)||_2^2 + \sum_{n=1}^N \E||\nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)||_2^2\\
+    &\lesssim N\frac{CpG'(\theta)^2}{J} + NpG'(\theta)^2(1-\epsilon)^{\floor{1/(c\log(J))}}.
+\end{align}
+\end{proof}

### 2024-06-15 paper/ms.tex
Kevin's edits

@@
 
-\subsection{MOP-$\alpha$ Bias and Variance}
+\subsection{MOP-$\alpha$ Error, Bias and Variance}
+
+We now provide a result showing that MOP-$\alpha$ has a desirable bias-variance tradeoff, achieving desirable rates while trading off between the low-variance but asymptotically biased estimate of \cite{naesseth18} and the high-variance but consistent estimate of \cite{poyiadjis11}.
+As the bias itself is difficult to analyze, we instead analyze the MSE and variance.
+The below result applies for any $\alpha \in [0,1)$, but not $\alpha=1$, as the gradient estimate when $\alpha=1$ has no forgetting properties.
+
 \begin{thm}
     \label{thm:mop-biasvar}
-    When $\alpha\in(0,1)$ and $\theta=\phi$, define $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)(1-\alpha)^{-1}$. 
+    When $\alpha\in (0,1)$ and $\theta=\phi$, define $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)/(1-\alpha)$. 
     There exists an $\epsilon>0$ depending on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karjalainen23} such that the MSE and variance of MOP-$\alpha$ are:
     \vspace*{-1ex}
-    \begin{eqnarray*}
+    \begin{eqnarray}
         \E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)\big\|_2^2 
-        &\lesssim& \min_{k \leq N} Np \, G'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi(\alpha)\right),
+        &\lesssim& \min_{k \leq N} Np \, G'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi(\alpha)\right), \label{eq:mop-mse}
         \\
-        \Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) &\lesssim& \min_{k\leq N} Np \, G'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{16\alpha^{2k}}{(1-\alpha)^2}N\right).
-        \end{eqnarray*}
+        \Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) &\lesssim& \min_{k\leq N} Np \, G'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}N\right). \label{eq:mop-variance}
+        \end{eqnarray}
 \end{thm}
@@
 The MSE bound considers the error between the score and MOP-$(1,k)$, and the error between MOP-$(1,k)$ and MOP-$\alpha$. 
-The latter is $\tilde{O}\big(Np \, G'(\theta)^2(k+\psi(\alpha))\big)$, by routine algebra, and the former can be controlled with a result on the the forgetting of the particle filter from \cite{karjalainen23} and the very same $L^p$ error bound mentioned above.
+The latter can be shown to be $\tilde{O}\big(Np \, G'(\theta)^2(k+\psi(\alpha))\big)$, and the former can be controlled with a result on the the forgetting of the particle filter from \cite{karjalainen23} and the very same $L^p$ error bound mentioned above.
 
-We also show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( Np \, G'(\theta)^2\big/J \big)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p$ and $G'$.
+\paragraph{Interpretation:} The result in Theorem \ref{thm:mop-biasvar} provides theoretical backing for the empirical result observed in Figure \ref{fig:biasvar} of the desirable bias-variance tradeoff that MOP-$\alpha$ enjoys.
+The first term in the MSE bound in Equation \ref{eq:mop-mse} corresponds to the error of the particle approximation, the second mixing error, and the third and fourth the error between the MOP-$(1,k)$ estimate and the MOP-$\alpha$ estimate.
+As $\alpha$ goes to $1$, choosing $k$ appropriately, the first, third and fourth term (corresponding to the variance) increases, while the second term (corresponding to the bias) decreases.
+Likewise, the first term in the variance bound in Equation \ref{eq:mop-variance} corresponds to the error of the particle approximation, while the second corresponds to mixing error.
+As $\alpha$ goes to $1$, the variance increases.
+
+\paragraph{Bias-Variance Tradeoff:} We also show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( Np \, G'(\theta)^2\big/J \big)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p$ and $G'$.
 Combining that result with Theorem \ref{thm:mop-biasvar} shows that MOP-$\alpha$ interpolates between MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$.
 The phase transition arises as even though the particle filter has forgetting properties \cite{karjalainen23}, the resulting derivative estimate of \cite{poyiadjis11} does not, and we require both for the variance reduction.
+
+In contrast, as $Nk^2 \leq N^4$ and $\alpha, k$ can be as large as desired to balance the impact of the last three terms, MOP-$\alpha$ achieves a lower MSE than MOP-$1$ does.
+We also show that MOP-$0$ achieves a MSE of $\tilde{O}\big(NpG'(\theta)^2(J^{-1}+(1-\epsilon)^{\lfloor1/c\log(J)\rfloor})\big)$, where the second term corresponds to uncontrolled mixing error.
+Unlike MOP-$\alpha$, MOP-$0$ has no opportunity to tune $\alpha$ and $k$ to reduce said mixing error, leading to uncontrolled asymptotic bias.
 

### 2024-06-15 paper/appendix/biasvar.tex
Ed edits, mostly cosmetic

--- 2024-06-14 paper/appendix/biasvar.tex
@@
 Now, to illustrate the proof strategy for the general case in an easier context, we first analyze the special case when $\alpha=0.$ We now prove the variance bound for this case, presented below.
+To our knowledge, this result for the variance of the gradient estimate of \cite{naesseth18} is novel.
 
-\begin{thm}[Variance of Gradient Estimate of MOP-$0$]
-The variance of the gradient estimate from MOP-$0$, or the gradient estimate resulting from the algorithm of \cite{naesseth18}, is:
+\begin{thm}[Variance of MOP-$0$ Gradient Estimate]
+The variance of the gradient estimate from MOP-$0$, i.e. the algorithm of \cite{naesseth18}, is:
  \begin{equation}
-     \Var(\nabla_\theta \hat\ell(\theta)) \lesssim \frac{G'(\theta)^2N}{J}.
+     \Var(\nabla_\theta \hat\ell^0(\theta)) \lesssim \frac{Np \, G'(\theta)^2}{J}.
  \end{equation}
@@
 where we use the log-derivative trick, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again.
-The point here was that we establish that the gradient of the log-likelihood estimate is given by the sum of terms over all $N$ and $J$:
+We do this to establish that the gradient of the log-likelihood estimate is given by the sum of terms over all $N$ and $J$:
 \begin{equation}
-\nabla_\theta \hat{\ell}(\theta) = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right).
+\nabla_\theta \hat{\ell}^0(\theta) = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right).
 \end{equation}
-Therefore,
+Therefore, for a given $\theta_i$ in the parameter vector $\theta$,
 \begin{align}
-    \Var\big(\nabla_\theta \hat\ell(\theta)\big) &= \frac{1}{J^2}\Var\left(\sum_{n=1}^N\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    & \hspace{-15mm}= \frac{1}{J^2}\sum_{n=1}^N\Var\left(\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &\hspace{-15mm}+ 2\sum_{m<n}\Cov\left(\frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \frac{1}{J}\sum_{j=1}^{J}\nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
-    &\hspace{-15mm}= \sum_{n=1}^N \Var\big(\nabla_\theta\hat\ell_n(\theta)\big) + 2\sum_{m<n} \Cov\big(\nabla_\theta\hat\ell_m(\theta),\nabla_\theta\hat\ell_n(\theta)\big).
+    \Var\left(\frac{\partial}{\partial_{\theta_i}} \hat\ell^0(\theta)\right) &= \frac{1}{J^2}\Var\left(\sum_{n=1}^N\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    & \hspace{-20mm}= \frac{1}{J^2}\sum_{n=1}^N\Var\left(\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    &\hspace{-20mm}+ 2\sum_{m<n}\Cov\left(\frac{1}{J}\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \frac{1}{J}\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
+    &\hspace{-20mm}= \sum_{n=1}^N \Var\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) + 2\sum_{m<n} \Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta),\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right).
 \end{align}
@@
 Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
-Therefore, by Davydov's inequality, noting that $\nabla_\theta\hat\ell_n(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and without loss of generality labeling $m$ and $n$ such that $\E\big[(\nabla_\theta\hat\ell_m(\theta))^4\big]^{1/4}\leq\E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big]^{1/4}$, we see that
+Therefore, by Davydov's inequality, noting that $\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and without loss of generality labeling $m$ and $n$ such that $\E\big[(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta))^4\big]^{1/4}\leq\E\big[(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta))^4\big]^{1/4}$, we see that
 \begin{align}
-    \Cov\big(\nabla_\theta\hat\ell_m(\theta), \nabla_\theta\hat\ell_n(\theta)\big) 
-    &\leq \alpha(n-m)^{1/2} \, \E\big[(\nabla_\theta\hat\ell_m(\theta))^4\big]^{1/4}
+    \Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta), \frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) 
+    &\leq \alpha(n-m)^{1/2} \, \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta)\right)^4\right]^{1/4}
     \,
-    \E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big]^{1/4}\\
-    &\leq \alpha(n-m)^{1/2}\, \E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big]^{1/2}.
+    \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right]^{1/4}\\
+    &\leq \alpha(n-m)^{1/2}\, \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right]^{1/2}.
 \end{align}
@@
 \begin{equation}
-\E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big] = \E\Big[\big(\nabla_\theta\hat\ell_n(\theta)-\E\big[\nabla_\theta\hat\ell_n(\theta)\big]\big)^4\Big]+\E\big[(\nabla_\theta\hat\ell_n(\theta))^2\big]^2
+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right] = \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)-\E\left[\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right]\right)^4\right]+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^2\right]^2
 \end{equation}
 alongside Lemma 2 of \cite{karjalainen23}, which shows that
-\begin{equation}
-\E\Big[\big(\nabla_\theta\hat\ell_n(\theta)-\E\big[\nabla_\theta\hat\ell_n(\theta)\big]\big)^4\Big] \lesssim \frac{G'(\theta)^4}{J^2}, \hspace{10mm} \E\Big[\big(\nabla_\theta\hat\ell_n(\theta)-\E\big[\nabla_\theta\hat\ell_n(\theta)\big]\big)^2\Big] \lesssim \frac{G'(\theta)^2}{J}.
-\end{equation}
+\begin{eqnarray}
+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)-\E\left[\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right]\right)^4\right] &\lesssim& \frac{G'(\theta)^4}{J^2}, 
+\\
+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)-\E\left[\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right]\right)^2\right] &\lesssim& \frac{G'(\theta)^2}{J}.
+\end{eqnarray}
 It follows that 
 \begin{equation}
-\E\big[(\nabla_\theta\hat\ell_n(\theta))^4\big]^{1/2} \lesssim  \sqrt{\frac{G'(\theta)^4}{J^2}+\left(\frac{G'(\theta)^2}{J}\right)^2} = \frac{G'(\theta)^2}{J},
+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right]^{1/2} \lesssim  \sqrt{\frac{G'(\theta)^4}{J^2}+\left(\frac{G'(\theta)^2}{J}\right)^2} = \frac{G'(\theta)^2}{J},
 \end{equation}
@@
 \end{align}
+It then follows that as $\theta \in \Theta \subseteq \R^p$, 
+\begin{equation}
+    \Var\big(\nabla_{\theta} \hat\ell(\theta)\big) \lesssim \frac{Np \, G'(\theta)^2}{J}.
+\end{equation}
 \end{proof}
@@
 If one divides by $N$, the error is $O(1/\sqrt{NJ})$.
+
+The variance here of $\frac{Np \, G'(\theta)^2}{J}$ can be thought of as a lower bound for the variance of MOP-$\alpha$.
+We will later see that the variance of MOP-$\alpha$ contains this term, and an additional term that corresponds to the memory of the MOP-$\alpha$ gradient estimate.
+
+Although the variance of MOP-$0$ is desirable, it unfortunately has large uncontrollable asymptotic bias.
+Error bounds for the bias and MSE of MOP-$0$ \cite{naesseth18} are also fortunately special cases (where $k=1$, $\alpha=0$) of the bias and MSE analysis we conduct for MOP-$\alpha$.
+We will later see that these showcase the large uncontrollable asymptotic bias of MOP-$0$.
+
+\subsection{MOP-$\alpha$ Variance Bound}
+
+We are now in a position to tackle the MOP-$\alpha$ variance bound.
+Here, we analyze the case for $\alpha \in (0,1)$.
+The proof strategy is as follows.
+Instead of analyzing the variance of the MOP-$\alpha$ gradient estimate proper, we will analyze the variance of a modified estimator where we truncate the weights at $k$ timesteps, so this estimator only looks at most $k$ timesteps back while accumulating the importance weights.
+This enables us to establish strong mixing for the truncated estimator, in order to get a bound on the variance that is $O(NJ^{-1})$.
+We emphasize that this estimator is only a theoretical construct -- we do not actually use this estimator.
+
+Coupled with a bound ensuring that the $k$-truncated estimator provides an estimate that is close to that of MOP-$\alpha$ proper, we get a bound on the variance of MOP-$\alpha$ that comprises of the variance of the $k$-truncated estimator plus the error, for any $k \leq N$.
+It then holds that the final variance bound on MOP-$\alpha$ is given by the minimum over all $k$, and is never larger than $O(\psi(\alpha)+NJ^{-1})$ for some function $\psi$ increasing in $\alpha$.
+
+\begin{thm}[Variance of MOP-$\alpha$ Gradient Estimate]
+    When $\alpha \in (0,1)$, the variance of the gradient estimate from MOP-$\alpha$ is
+    \begin{equation}\Var\big(\nabla_\theta \hat\ell(\theta)\big) \lesssim \min_{k\leq N} \left(\frac{k^2G'(\theta)^2Np}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}Np \, G'(\theta)^2 \right).
+    \end{equation}
+\end{thm}
+\begin{proof}
+
+Using the log-derivative trick and that $w_{n,j}^{F,\theta} = w_{n,j}^{F,\theta,k} = 1$ when $\theta=\phi$,
+\begin{align}
+    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2\\
+    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta}}{{\sum_{j=1}^J w_{n,j}^{F,\theta}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,k}}}\right\rVert_2^2 \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,k}\right\rVert_2^2 \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta}}{w_{n,j}^{F,\theta}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,k}}{w_{n,j}^{F,\theta,k}}\right\rVert_2^2\\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2.
+\end{align}
+
+This lets us bound the cumulative weight discrepancies by
+\begin{align}
+    &  \left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2\\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2 \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta}\right)-\log\left(w_{n,j}^{F,\theta,k}\right)\right)\right\rVert_2^2\\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=n-k}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_2^2 \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^{n-k}\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}p \, G'(\theta)^2\\
+    &\leq p \, G'(\theta)^2\frac{\alpha^k-\alpha^n}{1-\alpha},
+\end{align}
+where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
+We can then bound $\|\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)\|$ as follows:
+\begin{align}
+    &\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_2^2
+    \\ \nonumber
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_2^2
+    \\ \nonumber
+    &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)
+    \\
+    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_2^2
+    \\ 
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\!\log\!\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right) \!-\! \log\!\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)
+    \!-\! \log\!\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) \!+\! \log \! \left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\!\Bigg)\right\rVert_2^2 
+    \\ \nonumber
+    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)\right)\right\lVert_{\infty}
+    \\ 
+    & \qquad\qquad +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\right)\right\rVert_2^2,
+\end{align}
+and each of these two terms is bounded by $\frac{\alpha^k}{1-\alpha}Np \, G'(\theta)^2$. So we now know that
+\begin{equation}\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_2^2 \leq  \frac{2\alpha^k}{1-\alpha}Np \, G'(\theta)^2,\end{equation}
+which is our desired error bound. 
+Now, we bound the variance of $\hat s_k(\theta)$. 
+Recall that we defined
+\begin{eqnarray}
+\hat s_k(\theta) &=& \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right)
+\\
+&=& \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right).
+\end{eqnarray}
+We can then decompose this expression with the log-derivative trick while noting that $w_{n,j}^{F,\theta, k} = 1$ whenever $\theta=\phi$, to see that
+\begin{align}
+    \hat s_k(\theta) 
+    &= \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right) \\
+    &= \sum_{n=1}^N \left(\frac{\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}}{\sum_{j=1}^J w_{n,j}^{F,\theta, k}}-\frac{\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}}{\sum_{j=1}^J w_{n-1,j}^{A,F,\theta, k}}\right) \\
+    &= \sum_{n=1}^N \left(\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}\right)\\
+    &= \frac{1}{J}\sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right).
+\end{align}
+It now follows that we need to bound the variance at a single timestep, namely
+\begin{equation}
+\Var\left(\frac{1}{J}\sum_{j=1}^J\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right).
+\end{equation}
+We use the log-derivative trick yet again to find, noting that $w_{n,j}^{F,\theta, k}=1$ when $\theta=\phi$, that
+\begin{align}
+    \nabla_\theta w_{n,j}^{F,\theta, k} = \frac{\nabla_\theta w_{n,j}^{F,\theta, k}}{w_{n,j}^{F,\theta, k}} = \nabla_\theta \log(w_{n,j}^{F,\theta, k}) &= \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)
+    \\
+    &= \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right).
+\end{align}
+Then,
+\begin{align}
+    &\Var\left(\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right) 
+    \\
+    &\qquad = \Var\left(\nabla_\theta \left(\sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right)- \sum_{i=n-k-1}^{n-1} \alpha^{n-i} \nabla_\theta \log\left(g_{i,j}^{A,F,\theta}\right)\right)\right) \\
+    &\qquad = \Var \left(\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1}\log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right).
+\end{align}
+Note that \begin{equation}\hat s_{n,k}(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\end{equation} is a function bounded by $C\frac{G'(\theta)(1-\alpha^k)}{1-\alpha} =: CG'(\theta)r$ in each coordinate for some constant $C$ (by Assumption \ref{assump:bounded-measurement}) that depends only on $k$ timesteps, and not $n$.
+We can therefore invoke Lemma 2 of \cite{karjalainen23} to bound the $L_2$ error of each of the $k$ (from time $n-k$ to time $n$) functionals from its expectation under the posterior by $Cr\sqrt{p}G'(\theta)J^{-1/2}k$.
+That is, we have that
+\begin{equation}\E\left[\|\hat s_{n,k}(\theta) - \E_{\pi}\hat s_{n,k}(\theta)\|_2^2\right]^{1/2} \leq \frac{Cr\sqrt{p}G'(\theta) k}{\sqrt{J}}.\end{equation}
+
+By the bias-variance decomposition, this in turn bounds the variance at a single timestep by $C^2r^2p \, G'(\theta)^2J^{-1}k^2$. Concretely, we have that
+\begin{equation}\Var\left(\frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \lesssim \frac{r^2k^p2G'(\theta)^2}{J}.\end{equation}
+
+It now remains to bound the variance of $\hat s_k(\theta)$ by considering the covariance of each of the $N$ terms that comprise it. 
+We decompose
+\begin{align}
+    \Var(\nabla_\theta \hat s_k(\theta)) &= \Var\left(\sum_{n=1}^N \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \\
+    &= \sum_{n=1}^N\Var\left(\hat s_{n,k}(\theta)\right)+ 2\sum_{m<n}\Cov\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \\
+    &\lesssim \frac{r^2k^2p \, G'(\theta)^2N}{J} + 2\sum_{m<n}\Cov\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right).
+\end{align}
+
+Similarly to the proof of the MOP-$0$ case, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
+We know from Theorem 3 of \cite{karjalainen23} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
+\begin{equation}
+\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},
+\end{equation}
+i.e.
+the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
+By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strongly mixing, with $\alpha$-mixing coefficients $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c\log(J))}}$.
+Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
+
+We now derive the $\alpha$-mixing coefficients of $(\hat s_{n,k}(\theta))_{n=1}^N$.
+Observe that $\hat s_{n,k}$ is strong mixing at lag $k+1$, as all weights beyond $k$ timesteps are truncated.
+We therefore have that the mixing time of $\hat s_k(\theta)$ is $O(1+k+\log(J))$, and that the $\alpha$-mixing coefficients for $(\hat s_{n,k}(\theta))_{n=1}^N$ are given by $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c(1+k+\log(J)))}}$.
+Therefore, by Davydov's inequality and Lemma 2 of \cite{karjalainen23}, and noting that $\hat s_{n,k}(\theta) \leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, by a similar argument to the MOP-0 case, we find that
+\begin{equation}\Cov\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2p \, G'(\theta)^2}{J}.\end{equation}
+Concretely, noting that $\nabla_\theta\hat s_{n,k}(\theta)\leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and, without loss of generality, assuming $\E[(\nabla_\theta\hat s_{m,k}(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/4}$, we apply Davydov's inequality to see that
+\begin{align}
+    \Cov\big(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)\big) 
+    &\leq \alpha(n-m)^{1/2} \, \E\big[(\nabla_\theta\hat s_{m,k}(\theta))^4\big]^{1/4} \, \E\big[(\nabla_\theta\hat s_{n,k}(\theta))^4\big]^{1/4}\\
+    &\leq \alpha(n-m)^{1/2} \, \E\big[(\nabla_\theta\hat s_{n,k}(\theta))^4\big]^{1/2}.
+\end{align}
+To bound this, we use the fact that 
+\begin{equation}
+\E\big[(\nabla_\theta\hat s_{n,k}(\theta))^4\big] = \E\Big[\big(\nabla_\theta\hat s_{n,k}(\theta)-\E\big[\nabla_\theta\hat s_{n,k}(\theta)]\big)^4\Big]+\E\big[(\nabla_\theta\hat s_{n,k}(\theta))^2\big]^2
+\end{equation}
+alongside Lemma 2 of \cite{karjalainen23}, which alongside the fact that
+\begin{eqnarray}
+\E\Big[\big(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)]\big)^4\Big] &\lesssim& \frac{r^4p^2G'(\theta)^4}{J^2}, 
+\\
+\E\Big[\big(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)]\big)^2\Big] &\lesssim& \frac{r^2p \, G'(\theta)^2}{J},
+\end{eqnarray}
+allows us to find that 
+\begin{equation}
+\E\Big[\big(\nabla_\theta\hat s_{n,k}(\theta)\big)^4\Big]^{1/2} \lesssim  \sqrt{\frac{r^4p^2G'(\theta)^4}{J^2}+\left(\frac{r^2p \, G'(\theta)^2}{J}\right)^2} = \frac{r^2p \, G'(\theta)^2}{J},
+\end{equation}
+and conclude that 
+\begin{equation}\Cov\big(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)\big) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2p \, G'(\theta)^2}{J}.\end{equation}
+Putting it all together, we see that
+\begin{align}
+    \Var\big(\nabla_\theta \hat s_{k}(\theta)\big) &= \sum_{n=1}^N \Var\big(\nabla_\theta\hat s_{n,k}(\theta)\big) + 2\sum_{m<n} \Cov\big(\nabla_\theta\hat s_{m,k}(\theta),\nabla_\theta\hat s_{n,k}(\theta)\big)
+    \\
+    &\leq \frac{Nr^2pk^2G'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\frac{r^2p \, G'(\theta)^2}{J} 
+    \\
+    &\leq \frac{r^2pk^2G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c(1+k+\log(J))}}}\right) 
+    \\
+    &\lesssim \frac{k^2r^2p \, G'(\theta)^2N}{J}.
+\end{align}
+
+We will now use this result to bound $\Var(\nabla_\theta \hat s_k(\theta)$ proper. For random variables $X,Y$ where $|X-Y|$ is bounded almost surely by some $M$, we have that 
+\begin{equation}X-\E X - 2M \leq Y - \E Y \leq X - \E X + 2M\end{equation}
+so,
+\begin{equation}(Y-\E Y)^2 \leq (X-\E X - 2M)^2 + (X-\E X +2M)^2.\end{equation}
+Taking expectations, we get
+\begin{equation}
+\Var(Y) \leq \Var(X) + 8M^2.
+\end{equation}
+It follows from this result and the result we proved earlier, namely that
+\begin{equation}
+\big\|\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)\big\|_2^2 \leq \frac{2\alpha^k}{1-\alpha}Np \, G'(\theta)^2,
+\end{equation}
+that the variance of MOP-$\alpha$ proper is bounded by, for any $k \leq N$,
+that the variance of MOP-$\alpha$ proper is bounded by
+\begin{equation}
+\Var\big(\nabla_\theta \hat\ell(\theta)\big) \lesssim \frac{k^2r^2G'(\theta)^2Np}{J} + \frac{\alpha^{k}}{1-\alpha}Np \, G'(\theta)^2.
+\end{equation}
+As the above holds for any $k \leq N$,
+\begin{equation}
+\Var\big(\nabla_\theta \hat\ell(\theta)\big) \lesssim \min_{k\leq N} \left(\frac{k^2p \, G'(\theta)^2Np}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}Np \, G'(\theta)^2 \right).\end{equation}
+\end{proof}
+
+\subsection{MOP-$\alpha$ MSE Bound}
+
+\begin{thm}[MSE of MOP-$\alpha$]
+    When $\alpha \in (0,1)$, the MSE of MOP-$\alpha$ is given by
+    \begin{equation}
+        \E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)\big\|_2^2 \; \lesssim \; \min_{k \leq N} Np \, G'(\theta)^2\left(\frac{k^2}{J}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\frac{\alpha^k  + \alpha^{k+1} - \alpha}{1-\alpha}\right).
+    \end{equation}
+\end{thm}
+
+\begin{proof}
+
+The broad idea is that the first term can be controlled by a mixing argument, the second term garners $O(Nk)$ error, and the third term is $O\left(\frac{N}{1-\alpha}\right)$.
+It is unclear whether the bias can be reduced further -- the gradient is a sum of $N$ terms, and is therefore $O(N)$ itself.
+
+The first term is the problem term.
+We will first bound the ground truth of conditional scores from particle approximation conditional on correct filtering distribution at time $n-k$, and then bound that from the particle approximation resulting from an arbitrary filtering distribution at time $n-k$.
+That is, where $\hat\pi_n$ is the particle approximation at time $n$ of the posterior $\pi_n$ and we write $\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)$ for the truncated MOP-$1$ conditional score estimate given the correct filtering distribution at time $n-k$, we can first observe that
+\begin{equation}
+    \big\|\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)\big\|_2^2 =  \left\lVert\sum_{n=1}^N \nabla_\theta \hat\ell_n^\alpha(\theta) - \sum_{n=1}^N \nabla_\theta \ell_n(\theta)\right\rVert_2^2 \leq \sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_2^2,
+\end{equation}
+and decompose
+\begin{align} \nonumber
+    & \sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_2^2 \; \leq  \; \sum_{n=1}^N\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)\|_2^2  \; +
+     \\ & \hspace{20mm}
+     \sum_{n=1}^N\|\nabla_\theta\hat s_{n,k}^1(\theta) - \nabla_\theta\hat s_{n,k}^\alpha(\theta)\|_2^2 + \sum_{n=1}^N\|\nabla_\theta\hat s_{n,k}^\alpha(\theta) -  \nabla_\theta\hat\ell_n^\alpha(\theta)\|_2^2.
+\end{align}
+As before, we bound the first term, $\sum_{n=1}^N\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)\|_2^2$, by decomposing it into two terms, 
+\begin{eqnarray}
+\nonumber
+\hspace{-10mm} \sum_{n=1}^N\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)\|_2^2 &\leq&
+\\
+&& \hspace{-40mm}
+\sum_{n=1}^N \|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)\|_2^2 + \sum_{n=1}^N \|\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)\|_2^2.
+\end{eqnarray}
+The first term is a particle approximation dependent on $k$ timesteps, so by Lemma 2 of \cite{karjalainen23}, this is bounded by
+\begin{equation}\E\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)\|_2^2 \leq \frac{Cp \, G'(\theta)^2k^2}{J}.\end{equation}
+
+The second term amounts to bounding the difference between functionals of two different particle measures that mix under the same Markov kernel.
+Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
+We know from Theorem 3 of \cite{karjalainen23} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
+\begin{equation}\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},\end{equation}
+i.e.
+the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
+
+Then, we can bound 
+$\E\|\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)\|_2^2$ by
+\begin{equation}\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}} = \beta_{TV}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},\end{equation}
+implying that
+\begin{eqnarray}
+ && \hspace{-30mm}
+    \E\big\|\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)\big\|_2^2 
+\\
+    &\lesssim& \sup_{\mu, \nu} \sup_{\|\psi\|_\infty \leq 1/2} p \, G'(\theta)^2\big|(\mu \textbf{M}_{n,n+k})(\psi)-(\nu \textbf{M}_{n,n+k})(\psi)\big| 
+\\
+    &\leq& \sup_{\mu, \nu} p \, G'(\theta)^2\big\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\big\|_{\mathrm{TV}} 
+\\
+    &\leq& p \, G'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} \|\hat\pi_{n-k} - \pi_{n-k}\|_{\text{TV}} 
+\\
+    &\leq& p \, G'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} \sup_{\|\psi\|_{\infty} \leq 1/2} \big|\hat\pi_{n-k}(\psi) - \pi_{n-k}(\psi)\big| 
+\\
+    &\lesssim& p \, G'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}}.
+\end{eqnarray}
+Therefore, we have that
+\begin{align}
+    &\E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)\big\|_2^2 \\
+    &\leq \sum_{n=1}^N \E\big\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)\big\|_2^2 + \sum_{n=1}^N \E\big\|\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)\big\|_2^2\\
+    &\lesssim N\frac{Cp \, G'(\theta)^2k^2}{J} + Np \, G'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}}.
+\end{align}
+Now that the first term, $\E\|\nabla_\theta\ell(\theta)\|_2^2$, is taken care of, it remains to bound 
+$\E\|\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)\|_2^2$ 
+and
+$\E\|\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)\|_2^2$.
+We see, by a similar argument to the variance bound, that
+\begin{align}
+    & \hspace{-20mm} \nonumber
+    \E\big\|\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)\big\|_2^2
+    \\ \nonumber
+    &\leq \; \sum_{n=1}^N \E\left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right|
+    + \right.
+    \\ & \hspace{30mm} \left.
+    \left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\right|\Bigg)\right\rVert_2^2.
+\end{align}
+Each of these terms can be bounded using the log-derivative trick and that $w_{n,j}^{F,\theta,1,k} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
+\begin{align}
+    & \hspace{-30mm}
+    \left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_{2}^2
+    \\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2
+    \\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})p \, G'(\theta)^2
+    \\
+    &\leq p \, G'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right),
+\end{align}
+where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
+So, 
+\begin{align}
+    \E\big\|\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)\big\|_\infty
+    \leq 2Np \, G'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) 
+    \leq 2Np \, G'(\theta)^2k.
+\end{align}
+Now, we address the third term, $\E\|\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)\|_\infty$.
+From the variance argument, we know that we can bound $\|\nabla_\theta\hat\ell^\alpha(\theta) - \hat s_k^\alpha(\theta)\|$:
+\begin{align}
+    &\left\lVert \nabla_\theta\hat\ell^\alpha(\theta) - \hat s_k^\alpha(\theta) \right\rVert_2^2
+    \\
+    &= \left\lVert\sum_{n=1}^N \!
+\nabla_\theta \log \!
+\left( \!
+\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right)
+    \!-\! 
+\sum_{n=1}^N\!\nabla_\theta\log\!\left(\!\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \!
+\right\rVert_2^2
+    \\ \nonumber
+    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)- \log\!\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right)\right\lVert_2^2
+    \\ & \hspace{40mm}
+    +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta, \alpha}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha, k}\right)\right)\right\rVert_2^2.
+\end{align}
+Using the log-derivative trick and that $w_{n,j}^{F,\theta,\alpha} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
+\begin{align}
+    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_2^2\\
+    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=1}^{n-k}\alpha^{(n-i)}p \, G'(\theta)^2\\
+    &\leq p \, G'(\theta)^2\frac{\alpha^k-\alpha^n}{1-\alpha},
+\end{align}
+where the second-last line follows from Assumption \ref{assump:bounded-measurement}. Putting it together and taking expectations on both sides, we obtain
+\begin{equation}
+\E\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_2^2 \leq  \frac{2\alpha^k}{1-\alpha}Np \, G'(\theta)^2,
+\end{equation}
+which is our desired error bound. 
+Therefore, our decomposition yields the MSE bound
+\begin{align}
+    & \hspace{-5mm}
+    \E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)\big\|_2^2 
+    \\
+    &\leq \E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)\big\|_2^2 + \E\big\|\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)\big\|_2^2 + \E\big\|\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)\big\|_2^2
+    \\ \nonumber
+    &\lesssim \min_{k \leq N} \left(N\frac{Cp \, G'(\theta)^2k^2}{J} + Np \, G'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} \right.
+    \\
+    & \hspace{40mm} \left.
+    + 2Np \, G'(\theta)^2\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) + 2\frac{\alpha^k}{1-\alpha}Np \, G'(\theta)^2\right) \\
+    &\lesssim \min_{k \leq N} Np \, G'(\theta)^2\left(\frac{k^2}{J}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\frac{\alpha^k  + \alpha^{k+1} - \alpha}{1-\alpha}\right).
+\end{align}
+
+\end{proof}
+
+\begin{cor}[MSE of MOP-$0$]
+    The MSE of MOP-$0$, i.e. the estimator of \cite{naesseth18}, is
+    \begin{equation}
+        \E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^0(\theta)\big\|_2^2 \; \lesssim 
+        \; Np \, G'(\theta)^2\left(\frac{1}{J}+(1-\epsilon)^{\floor{1/(c\log(J))}}\right).
+    \end{equation}
+\end{cor}
+\begin{proof}
+    Observe that MOP-$0$ is equivalent to MOP-$(1,1)$.
+Noting that $k=1$ and that we only need to bound the difference between $\nabla_\theta \ell(\theta)$ and $\nabla_\theta \hat s_1^1(\theta)$, repeating the above arguments almost verbatim allows us to obtain this result.
+
+As before, we bound the first term, $\sum_{n=1}^N\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)\|_2^2$, by decomposing it into two terms, 
+\begin{eqnarray} \nonumber
+\hspace{-10mm} \sum_{n=1}^N\big\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)\big\|_2^2 &\leq 
+\\ &&\hspace{-45mm} 
+\sum_{n=1}^N \big\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta)\big\|_2^2 + \sum_{n=1}^N \big\|\nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)\big\|_2^2.
+\end{eqnarray}
+The first term is a particle approximation dependent on $1$ timesteps, so by Lemma 2 of \cite{karjalainen23}, this is bounded by
+\begin{equation}\E\big\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta)\big\|_2^2 \leq \frac{Cp \, G'(\theta)^2}{J}.\end{equation}
+
+The second term amounts to bounding the difference between functionals of two different particle measures that mix under the same Markov kernel.
+Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
+We know from Theorem 3 of \cite{karjalainen23} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator,
+\begin{equation}
+\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},
+\end{equation}
+i.e.
+the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
+
+Then, we can bound 
+$\E\|\nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)\|_2^2$ by
+\begin{equation}
+\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu \textbf{M}_{n,n+1}-\nu \textbf{M}_{n,n+1}\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}} = \beta_{TV}(\textbf{M}_{n,n+1}) \leq (1-\epsilon)^{\floor{1/(c\log(J))}},
+\end{equation}
+implying that
+\begin{align}
+    & \hspace{-10mm}
+    \E\big\|\nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)\big\|_2^2 
+    \\ & \hspace{30mm}
+    \lesssim \sup_{\mu, \nu} \sup_{\|\psi\|_\infty \leq 1/2} p \, G'(\theta)^2\big|(\mu \textbf{M}_{n,n+1})(\psi)-(\nu \textbf{M}_{n,n+1})(\psi)\big| 
+    \\ & \hspace{30mm}
+    \leq \sup_{\mu, \nu} p \, G'(\theta)^2\big\|\mu \textbf{M}_{n,n+1}-\nu \textbf{M}_{n,n+1}\big\|_{\mathrm{TV}} 
+    \\ & \hspace{30mm}
+    \leq p \, G'(\theta)^2(1-\epsilon)^{\floor{1/(c\log(J))}} \|\hat\pi_{n-1} - \pi_{n-1}\|_{\text{TV}} 
+    \\ & \hspace{30mm}
+    \leq p \, G'(\theta)^2(1-\epsilon)^{\floor{1/(c\log(J))}} \sup_{\|\psi\|_{\infty} \leq 1/2} \big|\hat\pi_{n-1}(\psi) - \pi_{n-1}(\psi)\big| 
+    \\ & \hspace{30mm}
+    \lesssim p \, G'(\theta)^2(1-\epsilon)^{\floor{1/(c\log(J))}}.
+\end{align}
+
+Therefore, we have that
+\begin{align}
+    &\E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_1^1(\theta)\big\|_2^2 \\
+    &\leq \sum_{n=1}^N \E\big\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta)\big\|_2^2 + \sum_{n=1}^N \E\big\|\nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)\big\|_2^2\\
+    &\lesssim N\frac{Cp \, G'(\theta)^2}{J} + Np \, G'(\theta)^2(1-\epsilon)^{\floor{1/(c\log(J))}}.
+\end{align}
+\end{proof}

### 2024-06-16 paper/appendix/biasvar.tex
Ed: some typos, reference tidying, a bit more cosmetics

@@
 When $\alpha=0$, we have:
-\begin{equation}
-    \hat{\lik}^0(\theta) := \prod_{n=1}^N L_n^{A, \theta, 0} = \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta,0}}{\sum_{j=1}^J w_{n,j}^{P,\theta,0}} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
-\end{equation}
+\begin{align}
+    \hat{\lik}^0(\theta) := \prod_{n=1}^N L_n^{A, \theta, 0} &= \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta,0}}{\sum_{j=1}^J w_{n,j}^{P,\theta,0}} 
+    \\&
+    = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
+\end{align}
 Similarly to the proof of Lemma \ref{lem:mop-0-formula}, we define $s_{n,j}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}$, and from the exact same arguments conclude that its gradient when $\theta=\phi$ is therefore
@@
 \begin{align}
-    &\Var\left(\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right) 
+    &\Var\left(\nabla_\theta \big(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\big)\right) 
     \\
@@
 \end{align}
-Note that \begin{equation}\hat s_{n,k}(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\end{equation} is a function bounded by $C\frac{G'(\theta)(1-\alpha^k)}{1-\alpha} =: CG'(\theta)r$ in each coordinate for some constant $C$ (by Assumption \ref{assump:bounded-measurement}) that depends only on $k$ timesteps, and not $n$.
-We can therefore invoke Lemma 2 of \cite{karjalainen23} to bound the $L_2$ error of each of the $k$ (from time $n-k$ to time $n$) functionals from its expectation under the posterior by $Cr\sqrt{p}G'(\theta)J^{-1/2}k$.
+Note that 
+\begin{equation}
+\hat s_{n,k}(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)
+\end{equation} 
+is a function bounded by $C\frac{G'(\theta)(1-\alpha^k)}{1-\alpha} =: CG'(\theta)r$ in each coordinate for some constant $C$ (by Assumption \ref{assump:bounded-measurement}) that depends only on $k$ timesteps, and not $n$.
+Subsequently, we suppress vector and matrix notation for the coordinates of $\theta$ and its derivatives, and their variances and covariances, with inequalities interpreted coordinatewise.
+We invoke Lemma 2 of \cite{karjalainen23} to bound the $L_2$ error of each of the $k$ (from time $n-k$ to time $n$) functionals from its expectation under the posterior by $Cr\sqrt{p}G'(\theta)J^{-1/2}k$.
 That is, we have that
@@
 \begin{align}
-    \Var(\nabla_\theta \hat s_k(\theta)) &= \Var\left(\sum_{n=1}^N \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \\
-    &= \sum_{n=1}^N\Var\left(\hat s_{n,k}(\theta)\right)+ 2\sum_{m<n}\Cov\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \\
-    &\lesssim \frac{r^2k^2p \, G'(\theta)^2N}{J} + 2\sum_{m<n}\Cov\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right).
+    \Var\big(\nabla_\theta \hat s_k(\theta)\big) &= \Var\left(\sum_{n=1}^N \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \\
+    &= \sum_{n=1}^N\Var\big(\hat s_{n,k}(\theta)\big)+ 2\sum_{m<n}\Cov\big(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\big) \\
+    &\lesssim \frac{r^2k^2p \, G'(\theta)^2N}{J} + 2\sum_{m<n}\Cov\big(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\big).
 \end{align}
@@
 Therefore, by Davydov's inequality and Lemma 2 of \cite{karjalainen23}, and noting that $\hat s_{n,k}(\theta) \leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, by a similar argument to the MOP-0 case, we find that
-\begin{equation}\Cov\left(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\right) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2p \, G'(\theta)^2}{J}.\end{equation}
+\begin{equation}
+\Cov\big(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\big) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2p \, G'(\theta)^2}{J}.
+\end{equation}
 Concretely, noting that $\nabla_\theta\hat s_{n,k}(\theta)\leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and, without loss of generality, assuming $\E[(\nabla_\theta\hat s_{m,k}(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/4}$, we apply Davydov's inequality to see that

### 2024-06-16 paper/ms.tex
Ed: some typos, reference tidying, a bit more cosmetics

@@
 
-The cholera transmission model that \cite{king08} developed for Dhaka, Bangladesh, has been used to benchmark the performance of various POMP inference methods \cite{ionides15, wood16, wycoff24}, and we employ it here for the same purpose.
+The cholera transmission model that \cite{king08} developed for Dhaka, Bangladesh, has been used to benchmark the performance of various POMP inference methods \cite{ionides15, fasiolo16, wycoff24}, and we employ it here for the same purpose.
 This model categorizes individuals in a population as susceptible, $S(t)$, infected, $I(t)$, and recovered, $R(t)$ and so is called an SIR compartmental model.
@@
 
-%\bibsplit[2]
-%Use \bibsplit to split the references from the body of the text. Value "[2]" represents the number of reference in the left column (Note: Please avoid single column figures & tables on this page.)
-
-% \bibliographystyle{pnas-new}
-% Bibliography
-\bibliography{bib-ifad}
-
 \arxiv{
@@
 
+%\bibsplit[2]
+%Use \bibsplit to split the references from the body of the text. Value "[2]" represents the number of reference in the left column (Note: Please avoid single column figures & tables on this page.)
+
+% \bibliographystyle{pnas-new}
+% Bibliography
+\bibliography{bib-ifad}
+
 \end{document}

### 2024-06-17 paper/appendix/biasvar.tex
Kevin edits

@@
 is a function bounded by $C\frac{G'(\theta)(1-\alpha^k)}{1-\alpha} =: CG'(\theta)r$ in each coordinate for some constant $C$ (by Assumption \ref{assump:bounded-measurement}) that depends only on $k$ timesteps, and not $n$.
-Subsequently, we suppress vector and matrix notation for the coordinates of $\theta$ and its derivatives, and their variances and covariances, with inequalities interpreted coordinatewise.
+Subsequently, we suppress vector and matrix notation for the coordinates of $\theta$ and its derivatives, and their variances and covariances, with the variance of a vector interpreted as the sum of the variance of its components.
 We invoke Lemma 2 of \cite{karjalainen23} to bound the $L_2$ error of each of the $k$ (from time $n-k$ to time $n$) functionals from its expectation under the posterior by $Cr\sqrt{p}G'(\theta)J^{-1/2}k$.
@@
 By the bias-variance decomposition, this in turn bounds the variance at a single timestep by $C^2r^2p \, G'(\theta)^2J^{-1}k^2$. Concretely, we have that
-\begin{equation}\Var\left(\frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \lesssim \frac{r^2k^p2G'(\theta)^2}{J}.\end{equation}
+\begin{equation}\Var\left(\frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \lesssim \frac{r^2k^2p2G'(\theta)^2}{J}.\end{equation}
 
@@
 
-We will now use this result to bound $\Var(\nabla_\theta \hat s_k(\theta)$ proper. For random variables $X,Y$ where $|X-Y|$ is bounded almost surely by some $M$, we have that 
-\begin{equation}X-\E X - 2M \leq Y - \E Y \leq X - \E X + 2M\end{equation}
+We will now use this result to bound $\Var\big(\nabla_\theta \hat s_k(\theta)\big)$ proper. 
+For random variables $X,Y$ where $|X-Y|$ is bounded almost surely by some $M$, we have that 
+\begin{equation}
+X-\E X - 2M \leq Y - \E Y \leq X - \E X + 2M\end{equation}
 so,
-\begin{equation}(Y-\E Y)^2 \leq (X-\E X - 2M)^2 + (X-\E X +2M)^2.\end{equation}
+\begin{equation}
+(Y-\E Y)^2 \leq (X-\E X - 2M)^2 + (X-\E X +2M)^2.\end{equation}
 Taking expectations, we get

### 2024-06-17 paper/ms.tex
Kevin edits

@@
 Recent advances in AD for particle filters \cite{naesseth18, jonschkowski18, corenflos21, scibior21, singh22} have drawn attention to AD as a tool for inference in POMP models.
-However, existing approaches are either asymptotically biased \cite{naesseth18, jonschkowski18}, have high variance \cite{poyiadjis11, scibior21}, high computational expense \cite{corenflos21, chen24}, or need access to transition densities \cite{poyiadjis11, scibior21, singh22, chen24}.
+However, existing approaches are either asymptotically biased \cite{naesseth18, jonschkowski18}, have high variance \cite{poyiadjis11, scibior21}, are computationally expensive \cite{corenflos21, chen24}, or require access to transition densities \cite{poyiadjis11, scibior21, singh22, chen24}.
 
@@
 In short, \textit{instead of directly differentiating through a basic particle filter, MOP uses AD to differentiate through a series of measurement density ratios and a differentiable simulator}.
-We proceed to generalize the MOP particle filter by adding a discounting factor, $\alpha \in [0,1]$, which interpolates between the  biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$.
-The bias-variance tradeoff induced by $\alpha$ suggests, both in theory and in practice, the use of $\alpha$ values strictly between 0 and 1.
+We proceed to generalize the MOP particle filter by adding a discounting factor, $\alpha \in [0,1]$, which interpolates between the  biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$ (Theorem \ref{thm:mop-functional-forms}).
+The bias-variance tradeoff induced by $\alpha$ suggests, both in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}), the use of $\alpha$ values strictly between 0 and 1.
 
-MOP-$\alpha$ avoids the discontinuity that is problematic for the standard particle filter in a novel way.
+MOP-$\alpha$ avoids the issue of asymptotic bias \cite{corenflos21} arising from having to drop terms arising from resampling when differentiating through the standard particle filter as in \cite{naesseth18} in a novel way.
 This particle filter is smooth by construction, allowing us to bypass the seemingly incompatible paradox of differentiating through discrete Monte Carlo sampling by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
-Yet, MOP-$\alpha$ can nevertheless provide a consistent log-likelihood estimate.
+Yet, MOP-$\alpha$ can nevertheless provide strongly consistent particle and log-likelihood estimates (Theorem \ref{thm:mop-targeting}), and when $\alpha=1$, consistent score estimates (Theorem \ref{thm:mop-grad-consistency}).
 
-We derive a linear convergence rate for stochastic gradient descent (SGD) using MOP-$\alpha$ in the  presence of strong convexity.
-Critically, the estimator is effective within a neighborhood of the maximum but is relatively slow at reaching this neighborhood.
+We derive a linear convergence rate for stochastic gradient descent (SGD) using MOP-$\alpha$ in the  presence of strong convexity (Theorem \ref{thm:mop-convergence}).
+Critically, the estimator converges quickly within a neighborhood of the maximum but struggles to reach this neighborhood in the presence of local minima and saddle points.
 That behavior is complementary to that of iterated filtering algorithms \cite{ionides06-pnas,ionides15} which provides a relatively fast and stable way to identify this neighborhood.
@@
 We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
-Superscripts $x_{n,j}^A$ denote the ancestral trajectory of particle $j$ at time $n$, and $a(\cdot)$ is the ancestor function that maps a particle to its parent, $j \mapsto a(j)$.
+Superscripts $x_{n,j}^A$ of particles denote the ancestral trajectory of particle $j$ at time $n$, $a(\cdot)$ is the ancestor function that maps a particle to its parent, $j \mapsto a(j)$, and $k_j$
+denotes the resample indices drawn for each particle $j$.
 
@@
 
-Directly differentiating through a particle filter, for a fixed value of $\omega$, leads to a biased gradient estimator due to the discreteness of the resampling operation \cite{corenflos21}.
-We avoid this issue by constructing a particle filtering algorithm called {\it Measurement Off-Parameter with discount factor} $\alpha$ (MOP-$\alpha$), defined by the pseudocode in Algorithm~\ref{alg:mop}.
-In short, \textit{instead of directly differentiating through a basic particle filter, we instead use AD to differentiate through a series of measurement density ratios and a differentiable simulator}.
+%Directly differentiating through a particle filter, for a fixed value of $\omega$, leads to a biased gradient estimator due to the discreteness of the resampling operation \cite{corenflos21}.
+Our approach is as follows.
+\textit{Instead of directly differentiating through a basic particle filter, we instead use AD to differentiate through a series of measurement density ratios and a differentiable simulator}.
+This is done through the construction of a novel particle filtering algorithm called {\it Measurement Off-Parameter with discount factor} $\alpha$ (MOP-$\alpha$), defined by the pseudocode in Algorithm~\ref{alg:mop}.
 Note that the plug-and-play property forbids access to transition densities but permits access to the density of the measurements conditional on the value of the latent process.
@@
 
-%%\subsection{Algorithm Outline}
+\subsection{Algorithm Outline}
 
@@
 \end{equation}
-where the weights for each particle are updated recursively by
+where the weights for each particle are updated recursively to correct for the cumulative error incurred by the off-parameter resampling:
 \arxiv{}{\vspace*{-1.5mm}}
@@
 When differentiated, we obtain the estimators of \cite{poyiadjis11, scibior21, naesseth18} as special cases (Theorem \ref{thm:mop-functional-forms}).
-In particular, MOP-$1$ is consistent for the score (Theorem \ref{thm:mop-grad-consistency}), has rates for its bias and variance under different choices of $\alpha$ (Theorem \ref{thm:mop-biasvar}), and enjoys a linear rate of convergence for gradient descent with the resulting gradient estimate (Theorem \ref{thm:mop-convergence}).
+In particular, MOP-$1$ is consistent for the score (Theorem \ref{thm:mop-grad-consistency}), has rates for its bias and variance under different choices of $\alpha$ (Theorem \ref{thm:mop-biasvar}) that illustrate the desirable bias-variance tradeoff observed empirically in Figure \ref{fig:biasvar}, and enjoys a linear rate of convergence for gradient descent with the resulting gradient estimate (Theorem \ref{thm:mop-convergence}).
 
@@
 Choose the learning rate $\eta$ such that $\eta \leq \frac{c(1-\beta)}{2\Gamma}$.
-Then, for sufficiently large $\alpha$ and $J$ to ensure the score estimate is a $1/2$-approximation and the minimum eigenvalue of $H$ is greater than some $c > 0$ for all $m$ with probability at least $1-\delta$, the second stage of IFAD converges linearly to the MLE:
+Then, for sufficiently large $\alpha$ and $J$ to ensure the score estimate is an $\epsilon$-approximation and the minimum eigenvalue of $H$ is greater than some $c > 0$ for all $m$ with probability at least $1-\delta$, the second stage of IFAD converges linearly to the MLE:
 \arxiv{}{\vspace*{-2mm}}
@@
 \arxiv{}{\vspace*{-5mm}}
-We defer the proof, which is similar to that of Theorem 6 in \cite{mahoney16}, to \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}.
-For simplicity, we only prove this for MOP-$1$ and note that the general case follows as the required $1/2$-approximation can still be obtained for sufficiently large $\alpha$.
+We borrow from tools in the field of randomized numerical linear algebra \cite{mahoney16} to solve this problem.
+The proof, which is similar to that of Theorem 6 in \cite{mahoney16}, is deferred to \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}.
+For simplicity, we only prove this for MOP-$1$ and note that the general case follows as the required $\epsilon$-approximation can still be obtained for sufficiently large $\alpha$.
+When $\alpha=1$, the required $J$ is given in the supplementary information in Lemmas \ref{lemma:grad_bound} and \ref{lemma:hess_bound}.
+This result shows that particle estimates for the gradient can enjoy the same linear convergence rate on well-conditioned problems as gradient descent with access to the score.
+As , we note that it is possible.
 
@@
 
-\paragraph{IFAD Successfully Finds the MLE:} Previously, an MLE at a log-likelihood of $-3748.5$ was reported by \cite{king16}.
+\paragraph{IFAD Successfully Finds the MLE:} Previously, an MLE at a log-likelihood of $-3748.6$ was reported by \cite{king16}.
 This MLE was obtained with much computational effort, using many global and local IF2 searches, and with the assistance of likelihood profiling.
 Meanwhile, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.6$, while the best log-likelihood found by \cite{king08} was only $-3793.4$.
-Despite being initialized for a global search, IFAD manages to get much closer to the MLE over the 100 searches than \cite{ionides15} and finds it up to Monte Carlo error, as seen in Table~\ref{table:mle}.
-On this problem, the sequence of local searches and refinement that was previously required for finding the MLE is not necessary with the IFAD algorithm. 
+Despite being initialized for a global search, IFAD manages to get much closer to the MLE over the $100$ searches than \cite{ionides15} and finds it up to $2.5$ standard deviations of Monte Carlo error, as seen in Table~\ref{table:mle}.
+On this problem, the sequence of local searches, refinement, and likelihood profiling that was previously required for finding the MLE is not necessary with the IFAD algorithm. 
 In other words, the improvement in numerical efficiency of IFAD over IF2 is so large that routine application of IFAD (consisting of a collection of Monte Carlo replications from random starting points) generates results outside the reach of a routine application of IF2.
-When interpreting Table~\ref{table:mle}, bear in mind that a difference of order 1 log unit has statistical relevance and therefore scientific value.
+When interpreting Table~\ref{table:mle}, bear in mind that by Wilks' Theorem, any difference of over $1.92$ log units has statistical relevance when testing one parameter at the $0.05$ significance level, and therefore potentially has scientific value.
 
@@
 IFAD therefore successfully combines the best qualities of IFAD and MOP, outperforming either of them alone.
-Monte~Carlo replication is appropriate for IFAD, as for any Monte Carlo algorithm used to solve a challenging numerical problem, but Figure~\ref{fig:boxplot} shows that IFAD can find higher likelihood values, more quickly and more reliably than the previous state-of-the-art.
+Monte~Carlo replication is appropriate for IFAD, or any Monte Carlo algorithm used to solve a challenging numerical problem, but Figure~\ref{fig:boxplot} shows that IFAD can find higher likelihood values, more quickly and more reliably than the previous state-of-the-art.
 In particular, Figure~\ref{fig:boxplot} shows that IFAD with $\alpha=0.97$ can maximize the challenging likelihood of \cite{king08} using a modest number of iterations and Monte Carlo replications, whereas previously it was necessary to carry out an extensive customized search or to live with an incompletely maximized likelihood.
@@
 
+%\bibsplit[2]
+%Use \bibsplit to split the references from the body of the text. Value "[2]" represents the number of reference in the left column (Note: Please avoid single column figures & tables on this page.)
+
+% \bibliographystyle{pnas-new}
+% Bibliography
+\bibliography{bib-ifad}
+
 \arxiv{
@@
 \input{appendix/biasvar}
-}{
-}
 
@@
 
-%\bibsplit[2]
-%Use \bibsplit to split the references from the body of the text. Value "[2]" represents the number of reference in the left column (Note: Please avoid single column figures & tables on this page.)
-
-% \bibliographystyle{pnas-new}
-% Bibliography
-\bibliography{bib-ifad}
+}{
+}
 

### 2024-06-22 paper/ms.tex
push from overleaf

@@
 Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a {\it Measurement Off-Parameter} (MOP) particle filter.
-In short, \textit{instead of directly differentiating through a basic particle filter, MOP uses AD to differentiate through a series of measurement density ratios and a differentiable simulator}.
-We proceed to generalize the MOP particle filter by adding a discounting factor, $\alpha \in [0,1]$, which interpolates between the  biased gradient estimator of \cite{naesseth18} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$ (Theorem \ref{thm:mop-functional-forms}).
+In short, \textit{instead of directly differentiating through a basic particle filter, MOP uses AD to differentiate through a smooth particle filter constructed using measurement density ratios and a differentiable simulator}.
+Our MOP algorithm differs from previous smooth particle filters \cite{svensson18,malik11} by possessing the plug-and-play property.
+Critically, we also avoid the high variance of \cite{svensson18} by generalizing the MOP particle filter to add a discount factor, $\alpha \in [0,1]$.
+This discount factor enables MOP-$\alpha$ to interpolate between the  biased gradient estimator of \cite{naesseth18}, when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21}, when $\alpha=1$ (Theorem \ref{thm:mop-functional-forms}).
 The bias-variance tradeoff induced by $\alpha$ suggests, both in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}), the use of $\alpha$ values strictly between 0 and 1.
@@
 The novel possibility of $0<\alpha<1$ is effective both in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).
+
+As mentioned earlier, our approach has similarities with \cite{svensson18}. 
+\cite{svensson18} construct a deterministic local approximation of the likelihood at the $k$-th optimization step via saving the particles from a run at some $\theta_{k-1}$, performing a reweighting with transition and measurement density ratios to evaluate the likelihood at any sufficiently nearby $\theta$, and then using a $0$-th order optimizer such as \texttt{optim} in \texttt{R} to choose a $\theta_k$ that maximizes the likelihood in that neighborhood.
+However, due to their use of transition densities in the reweighting, their approach is not plug-and-play.
+The variance of their log likelihood estimate is also roughly $O(||\theta - \theta_{k-1}||_2^{2N})$, where $N$ is the horizon.
+Our approach bypasses these two issues.
+The use of the \textit{reparameterization trick} to consider particle paths dependent on $\theta$ (and not $\theta_{k-1}$ or $\phi$) allows us to retain the plug-and-play property.
+On the other hand, our use of AD allows us to only evaluate the likelihood and score estimates when $\theta=\phi$.
+The use of AD bypasses the issue of exponentially increasing variance, but nevertheless can result in problematic polynomial variance.
+Our introduction of $\alpha$ avoids that difficulty.
 
@@
 
-We now provide a result showing that MOP-$\alpha$ has a desirable bias-variance tradeoff, achieving desirable rates while trading off between the low-variance but asymptotically biased estimate of \cite{naesseth18} and the high-variance but consistent estimate of \cite{poyiadjis11}.
+We now provide a result showing that MOP-$\alpha$ has a desirable bias-variance tradeoff when $0<\alpha<1$.
+This is achieved because it combines favorable properties from the low-variance but asymptotically biased estimate of \cite{naesseth18} and the high-variance but consistent estimate of \cite{poyiadjis11}.
 As the bias itself is difficult to analyze, we instead analyze the MSE and variance. 
@@
 
-\paragraph{Interpretation:} The result in Theorem \ref{thm:mop-biasvar} provides theoretical backing for the empirical result observed in Figure \ref{fig:biasvar} of the desirable bias-variance tradeoff that MOP-$\alpha$ enjoys.
+\paragraph{Interpretation:} Theorem \ref{thm:mop-biasvar} provides theoretical understanding of the results observed in Figure \ref{fig:biasvar} that show the empirically favorable bias-variance tradeoff enjoyed by MOP-$\alpha$.
 The first term in the MSE bound in Equation \ref{eq:mop-mse} corresponds to the error of the particle approximation, the second mixing error, and the third and fourth the error between the MOP-$(1,k)$ estimate and the MOP-$\alpha$ estimate.
@@
 
-\paragraph{Bias-Variance Tradeoff:} We also show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( Np \, G'(\theta)^2\big/J \big)$, and note that \cite{poyiadjis11} establish that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p$ and $G'$.
-Combining that result with Theorem \ref{thm:mop-biasvar} shows that MOP-$\alpha$ interpolates between MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$.
+\paragraph{Bias-Variance Tradeoff:} We show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( Np \, G'(\theta)^2\big/J \big)$.
+Previously, \cite{poyiadjis11} established that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p$ and $G'$. 
+Combining these results with Theorem \ref{thm:mop-biasvar} shows that MOP-$\alpha$ interpolates between MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$. 
 The phase transition arises as even though the particle filter has forgetting properties \cite{karjalainen23}, the resulting derivative estimate of \cite{poyiadjis11} does not, and we require both for the variance reduction.
@@
 As the proof of Theorem \ref{thm:mop-biasvar} implies, we can also truncate the weights at a fixed lag, corresponding to the MOP-$(1,k)$ estimate mentioned. 
-The analysis is similar, with slightly better rates but less convenient implementation.
+The analysis is similar, with comparable rates but with a slightly less convenient implementation.
 
@@
 We thank Nicolas Chopin for helpful communications regarding the strong law of large numbers for off-parameter resampled particle filters.
+We also thank Arnaud Doucet for helpful discussion regarding the manuscript, and for making the connection between our approach and that of \cite{svensson18}.
 }{

### 2024-07-10 mop/odmp.tex
mop/odmp.tex for overdispersed discrete-state processes

--- 2023-10-29 mop/mop.tex
@@
 \newcommand\data[1]{#1^*}
-\newcommand\giventh{{\hspace{0.5mm};\hspaxce{0.5mm}}}
+\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
 \newcommand\seq[2]{{#1}\!:\!{#2}}
@@
 
-\title{Automatic differentiation via off-policy particle filters}
+\title{ADPF for overdispersed continuous-time Markov chains}
 \date{\today}
@@
 \begin{abstract}
-Kevin has checked theoretically that the measurement off policy (MOP) particle filter has a fixed-seed derivative matching Scibior \& Wood (SW). 
-Here, we present the MOP algorithm in order to check that it is properly weighted.
-The argument for this remains somewhat heuristic, reasoning about properly weighted samples in a similar way to Jun Liu's 2001 book.
-We also introduce a MOP-$\alpha$ algorithm whose fixed-seed derivative interpolates between the basic ignore-resampling-in-derivative (IRID) PF ($\alpha=0$) and 
-MOP ($\alpha=1$).
+  We propose a combination of MOP and DOP to carry out inference for overdispersed continuous time Markov chains.
+  This algorithm, currently called PPOP (Proposal Partially Off Policy), is described in Section 3.
+  First, we review MOP and DOP.
 \end{abstract}
-
-\section{The Measurement off policy (MOP) particle filter}
-
-DOP has the entire system evolve according to $\phi$, and we differentiate through the likelihood ratios, which are smooth conditional on the resampling being the same (which in turn happens when $\theta \approx \phi$).
-
-POP has the process system evolve according to $\theta$, and the resampling evolve according to $\phi$.
-In essence, we offload the derivative with respect to the process model to the reparametrization trick.
-
-Bias in dropping the resampling gradient increases as process mixes slower.
-ADPF has a long memory, and the variance in the weights add up, and there is significant instability.
-However, by the time the system mixes, ancestral trajectories don't matter as much, so we shouldn't bother with estimating gradients that rely on more than one timestep of the conditional log-likelihood.
-
-Scientifically, this is great that no matter what you do with the gradient it works out well.
-
-Implement MOP-$\alpha$ after the paper is mostly written.
-
-%%%%  MOP PFILTER PSEUDOCODE
-%\begin{algorithm}[t!]
-\begin{algorithm}
-  \caption{\textbf{Measurement off policy (MOP) SMC}:
-    \label{alg:mop-smc}
-  }
-  Initialize filter particles:
-  simulate ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\mydot\giventh{\theta}\right)$ for $j$ in $\seq{1}{J}$\;
-  Initialize relative weights: $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$
-  \;
-  \For{$n$ in $\seq{1}{N}$}{
-    Simulate for prediction:
-    ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\mydot|{X}_{n-1,j}^{F};{\theta}\big)$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{mop:step1}
-    \;
-    Prediction weights: $w_{n,j}^{P,\theta} = w_{n-1,j}^{F,\theta}$ for $j\ \mathrm{in}\ \seq{1}{J}$ 
-    \;
-    Evaluate measurement density:
-    $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for $j$ in $\seq{1}{J}$\;
-    Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-    \;
-    Conditional likelihood under $\phi$: 
-    $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
-    \nllabel{mop:Lphi}
-    \;
-    Normalize weights:
-    $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
-    for $j\ \mathrm{in}\ \seq{1}{J}$
-    \;
-    Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \nllabel{mop:systematic}\;
-    Resample particles: ${X}_{n,j}^{F,\theta}={X}_{n,k_{j}}^{P,\theta}$
-    \;
-    Filter weights corrected for resampling:
-    $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{mop:weight:update}
-    \;
-    Resample filter weights:
-    $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
-    for $j$ in $\seq{1}{J}$ \nllabel{mop:step2}
-    \;
-    After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
-    \;
-  }
-\end{algorithm}
-
-\begin{itemize}
-\item MOP-SMC requires that the algorithm is first run at $\phi$, for which it is a vanilla particle filter.
-$g^{\phi}_{n,j}$ and $\tilde g^{\phi}_{n,j}$ are computed at this first pass.
-Then it is run at $\theta$, with the seed kept fixed.
-
-\item Here, the resampling rule for particle $j$ depends on $j$ (and therefore $\phi$ and $X^{P,\phi}_{n,j}$) but not on $\theta$ or  $X^{P,\theta}_{n,j}$.
-This does not have the standard form for a weighted particle filter, for which we can have general reweighting rules but they are usually a funciton of the location of the particle and the model parameter, $\theta$.
-That may be why it is harder to see that this is a properly weighted filter.
-Nevertheless, inspection of the algorithm shows that each particle is properly reweighted to account for its resampling probability, in step~\ref{mop:weight:update}, so it should be properly weighted.
-
-\item The final estimate of the likelihood is either based on the after-resampling conditional likelihood estimate
-  \begin{equation} \label{mop:likA}
-    L^A(\theta) = \left(\frac{1}{J}\sum_{j=1}^J w^{F,\theta}_{N,j} \right)
-    \prod_{n=1}^N L_n^\phi
-  \end{equation}
-  or the before-resampling estimate,
- \begin{equation}\label{mop:likB}
-   L^B(\theta) = \prod_{n=1}^N \frac{\sum_{j=1}^J w^{P,\theta}_{N,j} \, g^{\theta}_{N,j}}{\sum_{j=1}^J w^{P,\theta}_{N,j}}.
-  \end{equation}
- with $L^B(\theta)$ presumably having slightly lower variance.
-
-\item Weighted samples representing the filter distribution,
-  $f_{X_n|Y_{1:n}}(x_n \given y^*_{1:n} \giventh \theta)$
-  are either
-  $\big\{\big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big),  j\ \mathrm{in}\ \seq{1}{J}\big\}$
-  or
-  $\big\{\big(X^{P,\theta}_{n,j}, g^\theta_{n,j} \, w^{P,\theta}_{n,j}\big),  j\ \mathrm{in}\ \seq{1}{J}\big\}$,
-  meaning that an expectation over $f_{X_n|Y_{1:n}}(x_n \given y^*_{1:n} \giventh \theta)$ is consistently estimated by a corresponding weighted average of the filter or prediction particles.
-
-  \item As long as \texttt{rprocess} is a continuously differentiable function of $\theta$ for fixed seed, and \texttt{dmeasure} is a continuously differentiable function of $\theta$, and $g^{\phi}_{n,j}\neq 0$, we see that MOP is a continuously differentiable function of $\theta$ for fixed seed.
-Since it also provided an unbiased estimate of the likelihood, this justifies exchanging the order of differentiation and integration to ensure that its derivative is an unbiased estimate of the deriative of the likelihood.
-
-  \item Taking the derivative with respect to $\theta$ at $\theta=\phi$, step~\ref{mop:weight:update} looks very much like the stop gradient approach.
-
-  \end{itemize}
-
-Here's an outline of an argument explaining why MOP is properly weighted.
-
-\noindent {\bf Lemma 1}. {\it
-Suppose that $\{(X_j,u_j),j=1,\dots,J\}$ is properly weighted for $f_X$, meaning that, if $X\sim f_X$,
-\[
-E[h(X)] \approx \sum_{j=1}^J \frac{w_j}{\sum_{k=1}^J u_k} h(X_j),
-\]
-for some appropriate formalization of $\approx$.
-Now, let $\{(Y_j,v_j),j=1,\dots,J\}$ be a sample drawn from $\{(X_j,u_j)\}$ where $(X_j,u_j)$ is represented, on average, $\pi_j J$ times.
-This could amount to binomial resampling having $J$ draws each with probability $p_j$, or systematic resampling.
-Suppose
-\[
-(Y_j,v_j) = \big(X_{a(j)},u_{a(j)}/\pi_{a(j)}\big),
-\]
-where $a$ is called the ancestor function. Then, $\{(Y_j,v_j),j=1,\dots,J\}$ is properly weighted for $f_X$.
-}
-
-Notably, Lemma~1 permits $\pi_{1:J}$ to depend on $\{(X_j,u_j)\}$ as long as the resampling is carried out independently of $\{(X_j,u_j)\}$ conditional on $\pi_{1:J}$.
-
-\noindent {\bf Lemma 2}. {\it
-  Suppose that $Z_j \sim f_{Z|X}(\cdot | X_j)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$.
-Then, $\{(Z_j,u_j)\}$ is properly weighted for $f_Z$.
-}
-
-\noindent {\bf Lemma 3}. {\it
-Suppose that $(X^\prime_j,u^\prime_j) = \big(X_j,u_j\, f_{Z|X}(z^*|X_j)\big)$. Then, $\{(X^\prime_j,u^\prime_j)\}$ is properly weighted for $f_{X|Z}(\cdot | z^*)$.
-}
-
-Recursively applying Lemmas~1, 2 and~3, we obtain that 
-%to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
-the MOP filter is properly weighted.
-Specifically, suppose inductively that $\big\{\big(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
-Then, Lemma~2 tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big)\big\}$ is properly weighted for $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
-Lemma~3 tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} \big)\big\}$ is therefore properly weighted for  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-Lemma~1 guarantees that the resampling rule, given by 
-\[
-\big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,j} g^\theta_{n,j}\big/ g^\phi_{n,j}\big),
-\]
-with resampling weights proportional to $g^\phi_{n,j}$, is therefore also properly weighted for $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-
-This has addressed filtering, but not quite yet the likelihood evaluation. For this we use the following lemma.
-
-\noindent {\bf Lemma 4}. {\it
-  $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1_n-1}^*;\theta)$ is properly estimated by either the before-resampling estimate,
-\begin{equation}\label{L1}
-L_n^{B,\theta} =  \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
-\end{equation}
-or by the after-resampling estimate,
-\begin{equation}\label{L2}
-L_n^{A,\theta} = L^\phi_n \frac{\sum_{j=1}^Jw^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}.
-\end{equation}
-where $L^\phi_n$ comes from step~\ref{mop:Lphi}.
-}
-
-Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ is properly weighted for $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
-To see  (\ref{L2}),
-we write the numerator of (\ref{L1}) as
-\[
-L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
-\]
-Using Lemma~1, we see this is properly estimated by
-\[
-L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},
-\]
-from which we obtain (\ref{L2}).
-
-Using Lemma~4, we obtain a likelihood estimate,
-\[
-L^{A,\theta} = \prod_{n=1}^N \left( L^\phi_n \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J w^{P,\theta}_{n,j}}\right).
-\]
-Since $w^{F,\theta}_{n,j}=w^{P,\theta}_{n+1,j}$, this is a telescoping product. The remaining terms are
-$\sum_{j=1}^J w^{P,\theta}_{0,j} = J$ on the denominator and $\sum_{j=1}^J w^{F,\theta}_{N,j}$ on the numerator.
-This derives the MOP estimate in (\ref{mop:likA}).
-
-$L^{B,\theta}$ should generally be preferred, since there is no reason to include the extra variability from resampling when calculating the conditional log likelihood, but it lacks the nice telescoping product.
 
@@
 %%%%  MOP(alpha) PFILTER PSEUDOCODE
-\begin{algorithm}[t!]
+\begin{algorithm}
   \caption{\textbf{Measurement off policy (MOP-$\alpha$) SMC}:
@@
 
-\begin{itemize}
-\item MOP has been rewritten from an earlier draft so that now the only difference between MOP and MOP-$\alpha$ is the discounting exponent $\alpha$ in step~\ref{mop-alpha:discount}.
-\item When $\alpha=1$ we recover MOP. In particular, $L^{A,\theta}_n$ when $\theta=\phi$ and $\theta=1$ recovers Scibior and Wood. 
-\item When $\alpha=0$, $w^{P,\theta}_{n,j}=1$ and so $L^{B,\theta}_n$ corresponds to an equally weighted conditional likelihood, as in a regular particle filter, but with resampling carried out according to $\phi$ not $\theta$.
-  When calculating a derivative of $L_n^{B, \theta}$ when $\alpha=0$ with respect to $\theta$ at $\theta=\phi$, this corresponds to ignoring differentiation through the resampling, i.e., the fixed-seed derivative of the PF algorithm.
-  To see this, we notice that the fixed seed derivative which ``ignores'' resampling is not really ignoring it; in an infinitesimal neighborhood of $\phi$, the resampling is constant and so the derivative is zero.
-  The problem is not that the fixed-seed derivative is missing something, just that we cannot correctly move the derivative through an expectation for a discontinuous function.
-\item Of course, the difference between the before-resampling and after-resampling likelihoods is immaterial. If the functions match, then the gradients match. 
-\item When $\alpha < 1$, the product
-\[
-  L^{A,\alpha}(\theta)=\prod_{n=1}^N L^{A,\theta,\alpha}_n
-\]
-no longer has a telescoping cancellation. 
-\end{itemize}
-
-\section{Unresolved questions}
-
-\begin{itemize}
-\item $L^{A,\theta}_n$ and $L^{B,\theta}_n$ are asymptotically equivalent as $J$ increases, but are not identical.
-  However, both are identical for a particle filter (i.e., MOP when $\theta=\phi$).
-  Presumably, this could allow their fixed-seeed derivatives to be the same at $\theta=\phi$, though it is not obvious if this is true.
-  In that case, we would have an alternative explanation of the telescoping identity of Scibior \& Wood.
-
-\item It seems possible to carry out an extra numerical pass where one recalculates $L^A(\theta)$ taking advantage of the cancellation, which could make the derivative faster and/or more memory efficient, depending on what is the bottleneck.
-This may not be an immediate priority.
-
-  \item The cancellation difference between  $L^{A,\theta}_n$ and $L^{B,\theta}_n$ is moot for $\alpha<1$, so there's not much reason not to use $L^{B,\theta}_n$.
-
-\end{itemize}
+Algorithm~\ref{alg:mop-alpha} just repeats MOP-$\alpha$ from the earlier MOP/DOP draft.
+The final estimate of the likelihood is either based on the after-resampling conditional likelihood estimate
+  \begin{equation} \label{mop:likA}
+    L^A(\theta) = \left(\frac{1}{J}\sum_{j=1}^J w^{F,\theta}_{N,j} \right)
+    \prod_{n=1}^N L_n^\phi
+  \end{equation}
+  or the before-resampling estimate,
+ \begin{equation}\label{mop:likB}
+   L^B(\theta) = \prod_{n=1}^N \frac{\sum_{j=1}^J w^{P,\theta}_{N,j} \, g^{\theta}_{N,j}}{\sum_{j=1}^J w^{P,\theta}_{N,j}}.
+  \end{equation}
+ with $L^B(\theta)$ presumably having slightly lower variance.
 
@@
 
-Here, we consider a doubly off policy particle filter, meaning that both \code{rprocess} and \code{rmeasure} are computed at $\phi$ and particles are reweighted to correspond to $\theta$.
+Algorithm~\ref{alg:dop-alpha} is a doubly off policy particle filter, meaning that both \code{rprocess} and \code{rmeasure} are computed at $\phi$ and particles are reweighted to correspond to $\theta$.
+Something like this is needed to deal with discrete latent state spaces, which are beyond the realm of MOP.
+This algorithm is not plug-and-play.
+It seems to require transition probabilities (or at least their ratios) between observation times.
+In the next section, we show how this can combined with MOP-$\alpha$ to make an algorithm which is practical for over-dispersed continuouse time Markov chains
+
+NOTE: THIS IS STILL PROVISIONAL.
+FOR EXAMPLE, WE PROBABLY NEED dinit SINCE WE CAN'T ASSUME WE CAN DIFFERENTIATE THROUGH TAHT.
+THE PREVIOUS COP VERSION DID NOT DO THAT.
+OR, IN PRACTICE, MAYBE ONE CAN DO A DISCRETE OPTIMIZATION OVER THE IVP, IF IT IS ACTUALLY A DISCRETE PARAMETER?
+ALTERNATIVELY, IT MIGHT BE GOOD TO HAVE A PRIOR OF CONVENIENCE FOR THE IVPS.
+PERHAPS PUTTIG MASS ON AN INITIAL COMPARTMENT VALUE $C_0$ GIVEN BY $\mathrm{round}(\theta_C)-1, \mathrm{round}(\theta_C), \mathrm{round}(\theta_C)+1$ where $\theta_C$ is a continuous parameter for the value of $C_0$.
 
@@
   \caption{\textbf{Doubly off policy (DOP-$\alpha$) SMC}:
-    \label{alg:mop-alpha}
+    \label{alg:dop-alpha}
   }
@@
 
-\begin{itemize}
-\item In order to match the direct derivative for a discrete \code{rprocess} with $\alpha=0$, we would carry out the discounting in step~\ref{dop-alpha:discount} after the weight adjustment in step~\ref{dop-alpha:dproc}.
-  However, if we want to make $\alpha=0$ correspond to a reasonable estimate with no weight memory (i.e., strong mixing) then we should construct the algorithms as stated.
-\item ${X}_{n,j}^{P,\theta}$ in fact depends on $\phi$ not $\theta$ in this algorithm. Renamed accordingly.
-\end{itemize}
+\section{Process partially off policy (PPOP-$\alpha$)}
+
+OR RENAME TO, E.G., COMPARTMENT MODEL AUTOMATIC DIFFERENTIATION (CMAD)
+
+We suppose that, conditionally on a continuous integrated noise process $\Sigma(t)$, the latent process $X(t)$ is a multinomial Markov compartment model.
+We can formalize this definition, but it is intended to correspond to the class of models for which an Euler multinomial approximation is appropriate, where the Euler rate between $t$ and $t+\delta$ depends on $\Sigma(t+\delta)-\Sigma(t)$ and $X(t)$.
+We suppose $K_n$ Euler timesteps between $t_n$ and $t_{n+1}$.
+In practice, the intermediate times $\{t_{n_0}=t_n,t_{n,1},\dots,t_{K_n}=t_{n+1}\}$ may be chosen so that $\delta_{n,k}=t_{n_{k+1}} - t_{n_{k}}\le \delta$ for some $\delta$.
+We write $X_{n,k}=X(t_{n,k})$ and we note that $X_n=X_{n,0}$ and $X_{n+1}=X_{n,K_n}$.
+
+As for MOP and DOP, PPOP in Algorithm~\ref{alg:ppop-alpha} supposes that the algorithm is run first at $\theta=\phi$ and then at arbitrary $\theta$.
+To differentiate it at $\theta=\phi$, we need only one pass, or rather the two passes correspond to the two passes that AD does automatically.
+
+Discounting can be done at each individual Euler timestep, $t_{n,k} $, or just at observation times, $t_n$.
+The finer discretization is slightly smoother and respects the mixing that occurs within the interval $[t_{n-1},t_n]$.
+The extra computation will be negligible in most situations. Therefore, that is what we initially propose.
+
+%%%%  PPOP(alpha) PFILTER PSEUDOCODE
+\begin{algorithm}[t!]
+  \caption{\textbf{Process partially off policy (PPOP-$\alpha$) SMC}:
+    \label{alg:ppop-alpha}
+  }
+  Initialize filter particles:
+  simulate ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\mydot\giventh{\phi}\right)$ for $j$ in $\seq{1}{J}$\;
+  Initialize relative weights: $w^{F,\theta}_{0,j}= \frac{ {f}_{{X}_{0}}\left({X}_{0,j}^{F,\theta}\giventh{\theta}\right)}{ {f}_{{X}_{0}}\left({X}_{0,j}^{F,\theta}\giventh{\phi}\right)}$ for $j$ in $\seq{1}{J}$
+  \;
+  \For{$n$ in $\seq{1}{N}$}{
+    \For{$k$ in $\seq{1}{K_n}$}{
+      Prediction weights with discounting: $w_{n,k,j}^{P,\theta} = \big(w_{n,k-1,j}^{F,\theta}\big)^\alpha$ for $j\ \mathrm{in}\ \seq{1}{J}$
+      \nllabel{ppop-alpha:discount}
+      \;
+      Simulate $\Delta\Sigma^{\theta}_k= \Sigma(t+\delta)-\Sigma(t)$
+      \;
+      Simulate for prediction:
+      ${X}_{n,k,j}^{P,\phi}\sim \mathrm{reulermultinom}\big({X}_{n,k-1,j}^{F}, \rho(\phi,\Delta\Sigma^{\phi}_k),t_{n,k-1},t_{n,k}\big)$
+      %%for $j\ \mathrm{in}\ \seq{1}{J}$
+      \nllabel{ppop-alpha:step1}
+      \;
+      Adjust weights:
+      $\displaystyle w_{n,k,j}^{P,\theta} = w_{n,j}^{P,\theta} \times
+      \frac{
+        \mathrm{deulermultinom}\big({X}_{n,k-1,j}^{F}, {X}_{n,k,j}^{F}, \rho(\theta,\Delta\Sigma^{\theta}_k),t_{n,k-1},t_{n,k}\big)
+  }{
+        \mathrm{deulermultinom}\big({X}_{n,k-1,j}^{F}, {X}_{n,k,j}^{F}, \rho(\phi,\Delta\Sigma^{\phi}_k),t_{n,k-1},t_{n,k}\big)
+}$
+%%      for $j\ \mathrm{in}\ \seq{1}{J}$
+      \nllabel{ppop-alpha:dproc}
+      \;
+    }
+    Evaluate measurement density:
+    $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\phi}\giventh{\theta})$ for $j$ in $\seq{1}{J}$\;
+    Before-resampling conditional likelihood: $\displaystyle L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
+    \;
+    Conditional likelihood under $\phi$: 
+    $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$
+    \nllabel{ppop-alpha:Lphi}
+    \;
+    Normalize weights:
+    $\displaystyle \tilde{g}^{\phi}_{n,j}= \frac{g^{\phi}_{n,j}}{JL_n^{\phi}}$
+    for $j\ \mathrm{in}\ \seq{1}{J}$
+    \;
+    Apply systematic resampling to select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) =\tilde{g}^{\phi}_{n,m}$ \nllabel{ppop-alpha:systematic}\;
+    Resample particles: ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$
+    \;
+    Filter weights corrected for resampling:
+    $\displaystyle w^{FC,\theta}_{n,j}= w^{P,\theta}_{n,j} \times \frac{ g^{\theta}_{n,j}}{ g^{\phi}_{n,j}}$ for $j\ \mathrm{in}\ \seq{1}{J}$ \nllabel{ppop-alpha:weight:update}
+    \;
+    Resample filter weights:
+    $w_{n,j}^{F,\theta}= {w}_{n,k_{j}}^{FC,\theta}$
+    for $j$ in $\seq{1}{J}$ \nllabel{ppop-alpha:step2}
+    \;
+    After-resampling conditional likelihood: $\displaystyle L_n^{A,\theta,\alpha} = L_n^\phi \, \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$
+    \;
+  }
+\end{algorithm}
+
+It remains to define reulermultinom and deulermultinom. This is carried out separately for departured from each compartment, which are supposed to be infinitesially independent given $\Sigma(t)$.
+Suppose compartments are labeled $c=1,\dots,C$, and we have $N_{ab}(t)$ counting total transitions from compartment $a$ to $b$ by time $t$.
+We formally work with the Markov process $X(t)=\{N_{ab}(t), (a,b)\in 1:C \times 1:C\}$.
+together with an initial compartment membership $X(t_0)$, with the compartment membership process, $X_c(t)$, being defined implicitly.
+
+Then, the Euler multinomial approximation, with rate function $\rho_{ab}(x,t,\theta)$ and multiplicative integrated noise process $\Sigma(t)$, with variance depending on $\theta$, has the form
+\[
+\{\Delta N_{cb}(t), b\in 1:C \} \sim \mathrm{multinomial}(X_c,\pi_{c})
+\]
+where $\pi_c=(\pi_{c1},\dots,\pi_{cC})$ is specified in terms of the individual rates, $r_{ca}=\rho_{ca}\Delta\Sigma_{ca}$, and the total rate, $R_c=\sum_{a\neq c}r_{ca}$ as
+\[
+\pi_{ca}=\left\{
+  \begin{array}{lcl}
+    (r_{ca}/R_c) (1-\exp\{-R_c\}) &,& a\neq c \\
+    1-\sum_{b\neq c} pi_{cb} &,& a=c
+  \end{array}
+  \right.
+  \]
+The transitions are independent for $c\in 1:C$ conditional on $\Delta\Sigma(t)$ and $X(t)$.
 

### 2024-07-10 paper/appendix/bayes.tex
Ed: some typos, reference tidying, a bit more cosmetics
Updates from Overleaf

--- 2024-06-13 paper/appendix/bayes.tex
@@
     \centering
-    \includegraphics[width=\textwidth/\real{1.1}]{imgs/pmcmc/nuts.png}
+    \includegraphics[width=\textwidth/\real{1.25}]{imgs/pmcmc/nuts.png}
     \caption{Convergence diagnostics for a No-U-Turn Sampler (NUTS) with uniform priors on a compact set.
@@
     \centering
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Recovery Rate.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Death Rate.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Reciprocal of Mean Immunity Duration.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Trend in Force of Infection.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Noise in Force of Infection.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Measurement Noise.png}
+    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Recovery_Rate.png}
+    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Death_Rate.png}
+    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Reciprocal_of_Mean_Immunity_Duration.png}
+    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Trend_in_Force_of_Infection.png}
+    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Noise_in_Force_of_Infection.png}
+    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Measurement_Noise.png}
     \caption{Posterior estimates from NUTS powered by MOP-$\alpha$ with the informative nonparametric empirical prior obtained from a kernel density estimator on the IF2 parameter cloud.

### 2024-07-10 paper/appendix/biasvar.tex
Updates from Overleaf

@@
 \end{align}
-Similarly to the proof of Lemma \ref{lem:mop-0-formula}, we define $s_{n,j}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}$, and from the exact same arguments conclude that its gradient when $\theta=\phi$ is therefore
+Similarly to the proof of Lemma \ref{lem:mop-0-formula}, we define 
+\begin{equation}
+s_{n,j}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})},
+\end{equation}
+and from the exact same arguments conclude that its gradient when $\theta=\phi$ is therefore 
 \begin{align}
@@
 \end{align}
-where we use the log-derivative trick, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again.
+where we use the derivative of the logarithm, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the derivative of the logarithm where $\theta=\phi$ again.
 We do this to establish that the gradient of the log-likelihood estimate is given by the sum of terms over all $N$ and $J$:
@@
 
-By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strong mixing, with $\alpha$-mixing coefficients $\alpha(k) \leq (1-\epsilon)^{\floor{k/(c\log(J))}}$.
+By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strong mixing, with $\alpha$-mixing coefficients $a(k) \leq (1-\epsilon)^{\floor{k/(c\log(J))}}$.
 Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
@@
     \Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta), \frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) 
-    &\leq \alpha(n-m)^{1/2} \, \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta)\right)^4\right]^{1/4}
+    &\leq a(n-m)^{1/2} \, \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta)\right)^4\right]^{1/4}
     \,
     \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right]^{1/4}\\
-    &\leq \alpha(n-m)^{1/2}\, \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right]^{1/2}.
+    &\leq a(n-m)^{1/2}\, \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right]^{1/2}.
 \end{align}
@@
 \begin{equation}
-\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right] = \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)-\E\left[\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right]\right)^4\right]+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^2\right]^2
+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right] = \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4-\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^2\right]^2\right]+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^2\right]^2
 \end{equation}
@@
 \begin{eqnarray}
-\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)-\E\left[\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right]\right)^4\right] &\lesssim& \frac{G'(\theta)^4}{J^2}, 
+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4-\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^2\right]^2\right] &\lesssim& \frac{G'(\theta)^4}{J^2},
 \\
@@
 \begin{align}
-    \Var\left(\frac{\partial}{\partial_{\theta_i}} \hat\ell(\theta)\right) &= \sum_{n=1}^N \Var\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) + 2\sum_{m<n} \Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta),\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)\\
+    \Var\left(\frac{\partial}{\partial_{\theta_i}} \hat\ell^\alpha(\theta)\right) &= \sum_{n=1}^N \Var\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) + 2\sum_{m<n} \Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta),\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)\\
     &\leq \frac{NG'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\frac{G'(\theta)^2}{J} \\
@@
 \begin{equation}
-    \Var\big(\nabla_{\theta} \hat\ell(\theta)\big) \lesssim \frac{Np \, G'(\theta)^2}{J}.
+    \Var\big(\nabla_{\theta} \hat\ell^\alpha(\theta)\big) \lesssim \frac{Np \, G'(\theta)^2}{J}.
 \end{equation}
@@
 
-Although the variance of MOP-$0$ is desirable, it unfortunately has large uncontrollable asymptotic bias.
+%Although the variance of MOP-$0$ is desirable, it unfortunately has large uncontrollable asymptotic bias.
 Error bounds for the bias and MSE of MOP-$0$ \cite{naesseth18} are also fortunately special cases (where $k=1$, $\alpha=0$) of the bias and MSE analysis we conduct for MOP-$\alpha$.
@@
 Instead of analyzing the variance of the MOP-$\alpha$ gradient estimate proper, we will analyze the variance of a modified estimator where we truncate the weights at $k$ timesteps, so this estimator only looks at most $k$ timesteps back while accumulating the importance weights.
+We write
+\begin{equation}
+    \hat s_k^\alpha(\theta) := \sum_{n=1}^N \log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right)
+\end{equation}
+for the log-likelihood estimate from the $k$-truncated estimator with discount factor $\alpha$. It is the sum of components
+\begin{equation}
+\hat s_{n,k}^\alpha(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)
+\end{equation} 
+over timesteps $N$.
+
 This enables us to establish strong mixing for the truncated estimator, in order to get a bound on the variance that is $O(NJ^{-1})$.
-We emphasize that this estimator is only a theoretical construct -- we do not actually use this estimator.
+We emphasize that this truncated estimator is only a theoretical construct -- we do not actually use the truncated estimator.
+The truncated estimator, as we note in the discussion, possesses similar theoretical guarantees as the MOP-$\alpha$ estimator, but would require the user to specify the number of timesteps to truncate at instead of a discount factor.
 
@@
     When $\alpha \in (0,1)$, the variance of the gradient estimate from MOP-$\alpha$ is
-    \begin{equation}\Var\big(\nabla_\theta \hat\ell(\theta)\big) \lesssim \min_{k\leq N} \left(\frac{k^2G'(\theta)^2Np}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}Np \, G'(\theta)^2 \right).
+    \begin{equation}\Var\big(\nabla_\theta \hat\ell^\alpha(\theta)\big) \lesssim \min_{k\leq N} \left(\frac{k^2G'(\theta)^2Np}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}Np \, G'(\theta)^2 \right).
     \end{equation}
@@
 
-Using the log-derivative trick and that $w_{n,j}^{F,\theta} = w_{n,j}^{F,\theta,k} = 1$ when $\theta=\phi$,
+Using the derivative of the logarithm and that $w_{n,j}^{F,\theta,\alpha} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
 \begin{align}
-    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2\\
-    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta}}{{\sum_{j=1}^J w_{n,j}^{F,\theta}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,k}}}\right\rVert_2^2 \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,k}\right\rVert_2^2 \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta}}{w_{n,j}^{F,\theta}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,k}}{w_{n,j}^{F,\theta,k}}\right\rVert_2^2\\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2.
+    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_2^2\\
+    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}}}\right\rVert_2^2 \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,\alpha}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,\alpha,k}\right\rVert_2^2 \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,\alpha}}{w_{n,j}^{F,\theta,\alpha}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,\alpha,k}}{w_{n,j}^{F,\theta,\alpha,k}}\right\rVert_2^2\\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_2^2.
 \end{align}
@@
 \begin{align}
-    &  \left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2\\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2 \\
-    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta}\right)-\log\left(w_{n,j}^{F,\theta,k}\right)\right)\right\rVert_2^2\\
+    &  \left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_2^2\\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_2^2 \\
+    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta,\alpha}\right)-\log\left(w_{n,j}^{F,\theta,\alpha,k}\right)\right)\right\rVert_2^2\\
     &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=1}^n\alpha^{(n-i)}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=n-k}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_2^2 \\
@@
 where the second-last line follows from Assumption \ref{assump:bounded-measurement}.
-We can then bound $\|\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)\|$ as follows:
+We can then bound $\|\nabla_\theta\hat\ell^\alpha(\theta) - \nabla_\theta \hat s_k^\alpha(\theta)\|$ as follows:
 \begin{align}
-    &\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_2^2
+    &\left\lVert\nabla_\theta\hat\ell^\alpha(\theta) - \nabla_\theta \hat s_k^\alpha(\theta) \right\rVert_2^2
     \\ \nonumber
@@
     \\ 
-    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\!\log\!\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right) \!-\! \log\!\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)
-    \!-\! \log\!\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) \!+\! \log \! \left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\!\Bigg)\right\rVert_2^2 
+    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\!\log\!\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right) \!-\! \log\!\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)
+    \!-\! \log\!\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right) \!+\! \log \! \left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\!\Bigg)\right\rVert_2^2 
     \\ \nonumber
-    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)\right)\right\lVert_{\infty}
+    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right)\right\lVert_{\infty}
     \\ 
-    & \qquad\qquad +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\right)\right\rVert_2^2,
+    & \qquad\qquad +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\right)\right\rVert_2^2,
 \end{align}
 and each of these two terms is bounded by $\frac{\alpha^k}{1-\alpha}Np \, G'(\theta)^2$. So we now know that
-\begin{equation}\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_2^2 \leq  \frac{2\alpha^k}{1-\alpha}Np \, G'(\theta)^2,\end{equation}
+\begin{equation}\left\lVert\nabla_\theta\hat\ell^\alpha(\theta) - \nabla_\theta \hat s_k^\alpha(\theta) \right\rVert_2^2 \leq  \frac{2\alpha^k}{1-\alpha}Np \, G'(\theta)^2,\end{equation}
 which is our desired error bound. 
-Now, we bound the variance of $\hat s_k(\theta)$. 
+Now, we bound the variance of $\nabla_\theta \hat s_k^\alpha(\theta)$. 
 Recall that we defined
 \begin{eqnarray}
-\hat s_k(\theta) &=& \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right)
+\nabla_\theta \hat s_k^\alpha(\theta) &=& \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right)
 \\
-&=& \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right).
+&=& \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, \alpha, k}\right)\right).
 \end{eqnarray}
-We can then decompose this expression with the log-derivative trick while noting that $w_{n,j}^{F,\theta, k} = 1$ whenever $\theta=\phi$, to see that
+We can then decompose this expression with the derivative of the logarithm while noting that $w_{n,j}^{F,\theta,\alpha,k} = 1$ whenever $\theta=\phi$, to see that
 \begin{align}
-    \hat s_k(\theta) 
-    &= \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta, k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, k}\right)\right) \\
-    &= \sum_{n=1}^N \left(\frac{\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}}{\sum_{j=1}^J w_{n,j}^{F,\theta, k}}-\frac{\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}}{\sum_{j=1}^J w_{n-1,j}^{A,F,\theta, k}}\right) \\
-    &= \sum_{n=1}^N \left(\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta, k}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta, k}\right)\\
-    &= \frac{1}{J}\sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right).
+    \nabla_\theta \hat s_k^\alpha(\theta) 
+    &= \sum_{n=1}^N \nabla_\theta \left(\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)-\log\left(\sum_{j=1}^J w_{n-1,j}^{A, F,\theta, \alpha, k}\right)\right) \\
+    &= \sum_{n=1}^N \left(\frac{\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,\alpha,k}}{\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}}-\frac{\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta,\alpha, k}}{\sum_{j=1}^J w_{n-1,j}^{A,F,\theta,\alpha, k}}\right) \\
+    &= \sum_{n=1}^N \left(\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,\alpha,k}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n-1,j}^{A,F,\theta,\alpha, k}\right)\\
+    &= \frac{1}{J}\sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \left(w_{n,j}^{F,\theta,\alpha,k}- w_{n-1,j}^{A,F,\theta,\alpha, k}\right).
 \end{align}
@@
 \begin{equation}
-\Var\left(\frac{1}{J}\sum_{j=1}^J\nabla_\theta \left(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\right)\right).
+\Var\left(\frac{1}{J}\sum_{j=1}^J\nabla_\theta \left(w_{n,j}^{F,\theta,\alpha,k}- w_{n-1,j}^{A,F,\theta,\alpha, k}\right)\right).
 \end{equation}
-We use the log-derivative trick yet again to find, noting that $w_{n,j}^{F,\theta, k}=1$ when $\theta=\phi$, that
+We use the derivative of the logarithm yet again to find, noting that $w_{n,j}^{F,\theta,\alpha,k}=1$ when $\theta=\phi$, that
 \begin{align}
-    \nabla_\theta w_{n,j}^{F,\theta, k} = \frac{\nabla_\theta w_{n,j}^{F,\theta, k}}{w_{n,j}^{F,\theta, k}} = \nabla_\theta \log(w_{n,j}^{F,\theta, k}) &= \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)
+    \nabla_\theta w_{n,j}^{F,\theta,\alpha,k} = \frac{\nabla_\theta w_{n,j}^{F,\theta,\alpha,k}}{w_{n,j}^{F,\theta,\alpha,k}} = \nabla_\theta \log(w_{n,j}^{F,\theta,\alpha,k}) &= \sum_{i=n-k}^n \alpha^{n-i} \nabla_\theta \log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)
     \\
@@
 \begin{align}
-    &\Var\left(\nabla_\theta \big(w_{n,j}^{F,\theta, k}- w_{n-1,j}^{A,F,\theta, k}\big)\right) 
+    &\Var\left(\nabla_\theta \big(w_{n,j}^{F,\theta,\alpha,k}- w_{n-1,j}^{A,F,\theta,\alpha, k}\big)\right) 
     \\
@@
 \begin{equation}
-\hat s_{n,k}(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)
+\nabla_\theta \hat s_{n,k}^\alpha(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)
 \end{equation} 
@@
 That is, we have that
-\begin{equation}\E\left[\|\hat s_{n,k}(\theta) - \E_{\pi}\hat s_{n,k}(\theta)\|_2^2\right]^{1/2} \leq \frac{Cr\sqrt{p}G'(\theta) k}{\sqrt{J}}.\end{equation}
+\begin{equation}\E\left[\|\nabla_\theta \hat s_{n,k}^\alpha(\theta) - \E_{\pi}\nabla_\theta \hat s_{n,k}^\alpha(\theta)\|_2^2\right]^{1/2} \leq \frac{Cr\sqrt{p}G'(\theta) k}{\sqrt{J}}.\end{equation}
 
@@
 
-It now remains to bound the variance of $\hat s_k(\theta)$ by considering the covariance of each of the $N$ terms that comprise it. 
+It now remains to bound the variance of $\nabla_\theta \hat s_k^\alpha(\theta)$ by considering the covariance of each of the $N$ terms that comprise it. 
 We decompose
 \begin{align}
-    \Var\big(\nabla_\theta \hat s_k(\theta)\big) &= \Var\left(\sum_{n=1}^N \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \\
-    &= \sum_{n=1}^N\Var\big(\hat s_{n,k}(\theta)\big)+ 2\sum_{m<n}\Cov\big(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\big) \\
-    &\lesssim \frac{r^2k^2p \, G'(\theta)^2N}{J} + 2\sum_{m<n}\Cov\big(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\big).
+    \Var\big(\nabla_\theta \hat s_k^\alpha(\theta)\big) &= \Var\left(\sum_{n=1}^N \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \nabla_\theta\left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)\right) \\
+    &= \sum_{n=1}^N\Var\big(\nabla_\theta \hat s_{n,k}^\alpha(\theta)\big)+ 2\sum_{m<n}\Cov\big(\nabla_\theta \hat s_{m,k}^\alpha(\theta), \nabla_\theta \hat s_{n,k}^\alpha(\theta)\big) \\
+    &\lesssim \frac{r^2k^2p \, G'(\theta)^2N}{J} + 2\sum_{m<n}\Cov\big(\nabla_\theta \hat s_{m,k}^\alpha(\theta), \nabla_\theta \hat s_{n,k}^\alpha(\theta)\big).
 \end{align}
@@
 the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
-By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strongly mixing, with $\alpha$-mixing coefficients $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c\log(J))}}$.
+By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strongly mixing, with $\alpha$-mixing coefficients $a(l) \leq (1-\epsilon)^{\floor{l/(c\log(J))}}$.
 Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra).
 
-We now derive the $\alpha$-mixing coefficients of $(\hat s_{n,k}(\theta))_{n=1}^N$.
-Observe that $\hat s_{n,k}$ is strong mixing at lag $k+1$, as all weights beyond $k$ timesteps are truncated.
-We therefore have that the mixing time of $\hat s_k(\theta)$ is $O(1+k+\log(J))$, and that the $\alpha$-mixing coefficients for $(\hat s_{n,k}(\theta))_{n=1}^N$ are given by $\alpha(l) \leq (1-\epsilon)^{\floor{l/(c(1+k+\log(J)))}}$.
-Therefore, by Davydov's inequality and Lemma 2 of \cite{karjalainen23}, and noting that $\hat s_{n,k}(\theta) \leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, by a similar argument to the MOP-0 case, we find that
+We now derive the $\alpha$-mixing coefficients of $(\nabla_\theta \hat s_{n,k}^\alpha(\theta))_{n=1}^N$.
+Observe that $\nabla_\theta \hat s_{n,k}^\alpha$ is strong mixing at lag $k+1$, as all weights beyond $k$ timesteps are truncated.
+We therefore have that the mixing time of $\nabla_\theta \hat s_{n,k}^\alpha(\theta)$ is $O(1+k+\log(J))$, and that the $\alpha$-mixing coefficients for $(\nabla_\theta \hat s_{n,k}^\alpha(\theta))_{n=1}^N$ are given by $a(l) \leq (1-\epsilon)^{\floor{l/(c(1+k+\log(J)))}}$.
+Therefore, by Davydov's inequality and Lemma 2 of \cite{karjalainen23}, and noting that $\nabla_\theta \hat s_{n,k}^\alpha(\theta) \leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, by a similar argument to the MOP-0 case, we find that
 \begin{equation}
-\Cov\big(\hat s_{m,k}(\theta), \hat s_{n,k}(\theta)\big) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2p \, G'(\theta)^2}{J}.
+\Cov\big(\nabla_\theta \hat s_{m,k}^\alpha(\theta), \nabla_\theta \hat s_{n,k}^\alpha(\theta)\big) \lesssim (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2p \, G'(\theta)^2}{J}.
 \end{equation}
-Concretely, noting that $\nabla_\theta\hat s_{n,k}(\theta)\leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and, without loss of generality, assuming $\E[(\nabla_\theta\hat s_{m,k}(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat s_{n,k}(\theta))^4]^{1/4}$, we apply Davydov's inequality to see that
+Concretely, noting that $\nabla_\theta\hat s_{n,k}^\alpha(\theta)\leq CrG'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and, without loss of generality, assuming $\E[(\nabla_\theta\hat s_{m,k}^\alpha(\theta))^4]^{1/4}\leq\E[(\nabla_\theta\hat s_{n,k}^\alpha(\theta))^4]^{1/4}$, we apply Davydov's inequality to see that
 \begin{align}
-    \Cov\big(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)\big) 
-    &\leq \alpha(n-m)^{1/2} \, \E\big[(\nabla_\theta\hat s_{m,k}(\theta))^4\big]^{1/4} \, \E\big[(\nabla_\theta\hat s_{n,k}(\theta))^4\big]^{1/4}\\
-    &\leq \alpha(n-m)^{1/2} \, \E\big[(\nabla_\theta\hat s_{n,k}(\theta))^4\big]^{1/2}.
+    \Cov\big(\nabla_\theta\hat s_{m,k}^\alpha(\theta), \nabla_\theta\hat s_{n,k}^\alpha(\theta)\big) 
+    &\leq a(n-m)^{1/2} \, \E\big[(\nabla_\theta\hat s_{m,k}^\alpha(\theta))^4\big]^{1/4} \, \E\big[(\nabla_\theta\hat s_{n,k}^\alpha(\theta))^4\big]^{1/4}\\
+    &\leq a(n-m)^{1/2} \, \E\big[(\nabla_\theta\hat s_{n,k}^\alpha(\theta))^4\big]^{1/2}.
 \end{align}
@@
 \begin{equation}
-\E\big[(\nabla_\theta\hat s_{n,k}(\theta))^4\big] = \E\Big[\big(\nabla_\theta\hat s_{n,k}(\theta)-\E\big[\nabla_\theta\hat s_{n,k}(\theta)]\big)^4\Big]+\E\big[(\nabla_\theta\hat s_{n,k}(\theta))^2\big]^2
+\E\big[(\nabla_\theta\hat s_{n,k}^\alpha(\theta))^4\big] = \E\Big[\big(\nabla_\theta\hat s_{n,k}^\alpha(\theta)-\E\big[\nabla_\theta\hat s_{n,k}^\alpha(\theta)]\big)^4\Big]+\E\big[(\nabla_\theta\hat s_{n,k}^\alpha(\theta))^2\big]^2
 \end{equation}
@@
 \begin{eqnarray}
-\E\Big[\big(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)]\big)^4\Big] &\lesssim& \frac{r^4p^2G'(\theta)^4}{J^2}, 
+\E\Big[\big(\nabla_\theta\hat s_{n,k}^\alpha(\theta)-\E[\nabla_\theta\hat s_{n,k}^\alpha(\theta)]\big)^4\Big] &\lesssim& \frac{r^4p^2G'(\theta)^4}{J^2}, 
 \\
-\E\Big[\big(\nabla_\theta\hat s_{n,k}(\theta)-\E[\nabla_\theta\hat s_{n,k}(\theta)]\big)^2\Big] &\lesssim& \frac{r^2p \, G'(\theta)^2}{J},
+\E\Big[\big(\nabla_\theta\hat s_{n,k}^\alpha(\theta)-\E[\nabla_\theta\hat s_{n,k}^\alpha(\theta)]\big)^2\Big] &\lesssim& \frac{r^2p \, G'(\theta)^2}{J},
 \end{eqnarray}
@@
 \begin{equation}
-\E\Big[\big(\nabla_\theta\hat s_{n,k}(\theta)\big)^4\Big]^{1/2} \lesssim  \sqrt{\frac{r^4p^2G'(\theta)^4}{J^2}+\left(\frac{r^2p \, G'(\theta)^2}{J}\right)^2} = \frac{r^2p \, G'(\theta)^2}{J},
+\E\Big[\big(\nabla_\theta\hat s_{n,k}^\alpha(\theta)\big)^4\Big]^{1/2} \lesssim  \sqrt{\frac{r^4p^2G'(\theta)^4}{J^2}+\left(\frac{r^2p \, G'(\theta)^2}{J}\right)^2} = \frac{r^2p \, G'(\theta)^2}{J},
 \end{equation}
 and conclude that 
-\begin{equation}\Cov\big(\nabla_\theta\hat s_{m,k}(\theta), \nabla_\theta\hat s_{n,k}(\theta)\big) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2p \, G'(\theta)^2}{J}.\end{equation}
+\begin{equation}\Cov\big(\nabla_\theta\hat s_{m,k}^\alpha(\theta), \nabla_\theta\hat s_{n,k}^\alpha(\theta)\big) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c(1+k+\log(J))}}}\frac{r^2p \, G'(\theta)^2}{J}.\end{equation}
 Putting it all together, we see that
 \begin{align}
-    \Var\big(\nabla_\theta \hat s_{k}(\theta)\big) &= \sum_{n=1}^N \Var\big(\nabla_\theta\hat s_{n,k}(\theta)\big) + 2\sum_{m<n} \Cov\big(\nabla_\theta\hat s_{m,k}(\theta),\nabla_\theta\hat s_{n,k}(\theta)\big)
+    \Var\big(\nabla_\theta \hat s_{k}(\theta)\big) &= \sum_{n=1}^N \Var\big(\nabla_\theta\hat s_{n,k}^\alpha(\theta)\big) + 2\sum_{m<n} \Cov\big(\nabla_\theta\hat s_{m,k}^\alpha(\theta),\nabla_\theta\hat s_{n,k}^\alpha(\theta)\big)
     \\
@@
 
-We will now use this result to bound $\Var\big(\nabla_\theta \hat s_k(\theta)\big)$ proper. 
+We will now use this result to bound $\Var\big(\nabla_\theta \hat s_k^\alpha(\theta)\big)$. 
 For random variables $X,Y$ where $|X-Y|$ is bounded almost surely by some $M$, we have that 
@@
 \begin{equation}
-\big\|\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)\big\|_2^2 \leq \frac{2\alpha^k}{1-\alpha}Np \, G'(\theta)^2,
+\big\|\nabla_\theta\hat\ell^\alpha(\theta) - \nabla_\theta \hat s_k^\alpha(\theta)\big\|_2^2 \leq \frac{2\alpha^k}{1-\alpha}Np \, G'(\theta)^2,
 \end{equation}
@@
 \begin{equation}
-\Var\big(\nabla_\theta \hat\ell(\theta)\big) \lesssim \frac{k^2r^2G'(\theta)^2Np}{J} + \frac{\alpha^{k}}{1-\alpha}Np \, G'(\theta)^2.
+\Var\big(\nabla_\theta \hat\ell^\alpha(\theta)\big) \lesssim \frac{k^2r^2G'(\theta)^2Np}{J} + \frac{\alpha^{k}}{1-\alpha}Np \, G'(\theta)^2.
 \end{equation}
@@
 \begin{equation}
-\Var\big(\nabla_\theta \hat\ell(\theta)\big) \lesssim \min_{k\leq N} \left(\frac{k^2p \, G'(\theta)^2Np}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}Np \, G'(\theta)^2 \right).\end{equation}
+\Var\big(\nabla_\theta \hat\ell^\alpha(\theta)\big) \lesssim \min_{k\leq N} \left(\frac{k^2p \, G'(\theta)^2Np}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}Np \, G'(\theta)^2 \right).\end{equation}
 \end{proof}
@@
 
-The broad idea is that the first term can be controlled by a mixing argument, the second term garners $O(Nk)$ error, and the third term is $O\left(\frac{N}{1-\alpha}\right)$.
+The broad idea is to decompose the MSE into three terms as in Equation \ref{eq:mse-decomp}.
+The first term can be controlled by a mixing argument, the second term garners $O(Nk)$ error, and the third term is $O\left(\frac{N}{1-\alpha}\right)$.
 It is unclear whether the bias can be reduced further -- the gradient is a sum of $N$ terms, and is therefore $O(N)$ itself.
@@
      \sum_{n=1}^N\|\nabla_\theta\hat s_{n,k}^1(\theta) - \nabla_\theta\hat s_{n,k}^\alpha(\theta)\|_2^2 + \sum_{n=1}^N\|\nabla_\theta\hat s_{n,k}^\alpha(\theta) -  \nabla_\theta\hat\ell_n^\alpha(\theta)\|_2^2.
+\label{eq:mse-decomp}
 \end{align}
-As before, we bound the first term, $\sum_{n=1}^N\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)\|_2^2$, by decomposing it into two terms, 
+We bound the first term of Equation \ref{eq:mse-decomp} , $\sum_{n=1}^N\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)\|_2^2$, by decomposing it into two terms, 
 \begin{eqnarray}
@@
 \end{eqnarray}
-The first term is a particle approximation dependent on $k$ timesteps, so by Lemma 2 of \cite{karjalainen23}, this is bounded by
+The first term of Equation \ref{eq:mse-decomp} is a particle approximation dependent on $k$ timesteps, so by Lemma 2 of \cite{karjalainen23}, this is bounded by
 \begin{equation}\E\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)\|_2^2 \leq \frac{Cp \, G'(\theta)^2k^2}{J}.\end{equation}
 
-The second term amounts to bounding the difference between functionals of two different particle measures that mix under the same Markov kernel.
+Bounding the second term of Equation \ref{eq:mse-decomp} amounts to bounding the difference between functionals of two different particle measures that mix under the same Markov kernel.
 Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing.
@@
 \end{align}
-Each of these terms can be bounded using the log-derivative trick and that $w_{n,j}^{F,\theta,1,k} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
+Each of these terms can be bounded using the derivative of the logarithm and that $w_{n,j}^{F,\theta,1,k} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
 \begin{align}
@@
 \end{align}
-Now, we address the third term, $\E\|\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)\|_\infty$.
-From the variance argument, we know that we can bound $\|\nabla_\theta\hat\ell^\alpha(\theta) - \hat s_k^\alpha(\theta)\|$:
+Now, we address the third term of Equation \ref{eq:mse-decomp}, $\E\|\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)\|_\infty$.
+From the variance argument, we know that we can bound $\|\nabla_\theta\hat\ell^\alpha(\theta) - \nabla_\theta \hat s_k^\alpha(\theta)\|$:
 \begin{align}
-    &\left\lVert \nabla_\theta\hat\ell^\alpha(\theta) - \hat s_k^\alpha(\theta) \right\rVert_2^2
+    &\left\lVert \nabla_\theta\hat\ell^\alpha(\theta) - \nabla_\theta \hat s_k^\alpha(\theta) \right\rVert_2^2
     \\
@@
 \end{align}
-Using the log-derivative trick and that $w_{n,j}^{F,\theta,\alpha} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
+Using the derivative of the logarithm and that $w_{n,j}^{F,\theta,\alpha} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
 \begin{align}
@@
 \begin{equation}
-\E\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_2^2 \leq  \frac{2\alpha^k}{1-\alpha}Np \, G'(\theta)^2,
+\E\left\lVert\nabla_\theta\hat\ell^\alpha(\theta) - \nabla_\theta \hat s_k^\alpha(\theta) \right\rVert_2^2 \leq  \frac{2\alpha^k}{1-\alpha}Np \, G'(\theta)^2,
 \end{equation}

### 2024-07-10 paper/ms.tex

@@
-\documentclass[11pt]{article} 
+%\documentclass[11pt]{article} 
 %\documentclass[9pt,twocolumn,twoside]{pnas-new}
-%\documentclass[9pt,twocolumn,pnasresearcharticle]{pnas-new}
+\documentclass[9pt,twocolumn,pnasresearcharticle]{pnas-new}
 %\documentclass[9pt,twocolumn,twoside]{pnas-new}
@@
 
-\newcommand\arxiv[2]{#1} % 1st argument for arxiv format
-% \newcommand\arxiv[2]{#2} % 2nd argument for PNAS format
+%\newcommand\arxiv[2]{#1} % 1st argument for arxiv format
+\newcommand\arxiv[2]{#2} % 2nd argument for PNAS format
 % also, swich to \documentclass[11pt]{article} for arxiv format
@@
 Despite their ubiquity, estimation and inference within this broad class of models remains a challenging problem.
-This article concerns the use automatic differentiation (AD) to construct improved algorithms for inference on complex POMP models.
+This article uses automatic differentiation (AD) to construct improved algorithms for inference on complex POMP models.
 
@@
 \cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.
-Unfortunately, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability, while the stop-gradient procedure is introduced without mathematical motivation.
-We use \cite{scibior21} and \cite{poyiadjis11} as our starting point, while developing a new approach which remedies the weaknesses of these papers.
+However, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability, while the stop-gradient procedure is introduced without mathematical motivation.
+We use \cite{scibior21} and \cite{poyiadjis11} as our starting point in developing a new approach remedying these issues.
 
@@
 
-The MOP-$\alpha$ algorithm resamples the particles according to an arbitrary resampling rule that depends on both the target parameter value and a baseline parameter value, and this explains the name {\it measurement off-parameter}.
+The MOP-$\alpha$ algorithm resamples the particles according to an arbitrary resampling rule that depends on both the target parameter value and a baseline parameter value, yielding the name {\it measurement off-parameter}.
 This is intended as an analogy to {\it off-policy} learning in reinforcement learning.
-Specifically, MOP-$\alpha$ evaluates the likelihood at some $\theta \in \Theta$, but instead resamples the particles according to the indices generated by a vanilla particle filter run at a {\it baseline parameter}, $\phi \in \Theta$.
-This ensures that the resampling indices are invariant to $\theta$ when $\omega$ and $\phi$ are fixed, bypassing the issue of Monte Carlo resampling.
+Specifically, MOP-$\alpha$ evaluates the likelihood at some $\theta \in \Theta$, but instead resamples the particles according to the indices generated by a particle filter run at a {\it baseline parameter}, $\phi \in \Theta$.
+This ensures that the resampling indices are invariant in $\theta$ when $\omega$ and $\phi$ are fixed, bypassing the issue of Monte Carlo resampling.
 We show in Theorem~\ref{thm:mop-targeting} that MOP-$\alpha$ exactly targets the filtering distribution when $\theta=\phi$ or $\alpha=1$.
@@
             \hspace*{4mm} Obtain resampled particles ${X}_{n,j}^{F,\theta}={X}_{n,k_{j}}^{P,\theta}$. \newline
-            \hspace*{4mm} Calculate resampled corrected weights
+            \hspace*{4mm} Calculate corrected weights
             $w_{n,j}^{F,\theta}= w^{P,\theta}_{n,k_j} \, g^{\theta}_{n,k_j} \, \big/ \, { g^{\phi}_{n,k_j}}$.\newline
@@
 Algorithm~\ref{alg:mop} reweights the conditional likelihoods by a correction factor accumulated over time to account for the resampling under $\phi$. 
-Writing $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density and $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$, we obtain two suitably weighted estimates of the conditional likelihood under $\theta$,
+Writing $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density and $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$, we obtain two estimates of the conditional likelihood under $\theta$,
 \arxiv{}{\vspace*{-1.5mm}}
@@
      \label{eq:mop-conditional-likelihood}
-     L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j}\, w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}, \hspace{15mm} 
+     L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j}\, w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}, \arxiv{\hspace{15mm}}{\;\;}
      L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
@@
 \end{equation}
-where the weights for each particle are updated recursively to correct for the cumulative error incurred by the off-parameter resampling:
+where we recursively correct the weights for each particle for the cumulative error incurred by the off-parameter resampling:
 \arxiv{}{\vspace*{-1.5mm}}
@@
     w_{n,j}^{P,\theta} = (w_{n-1,j}^{F,\theta})^\alpha, 
-    \hspace{5mm}
+    \arxiv{\hspace{5mm}}{\;}
     w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} \, g^{\theta}_{n,k_j} \, \big/ g^{\phi}_{n,k_j}, 
-    \hspace{5mm}
+    \arxiv{\hspace{5mm}}{\;}
     w^{F,\theta}_{0,j}= 1.
@@
 \end{equation}
-The before-resampling conditional likelihood estimate $L_n^{B,\theta,\alpha}$ is preferable in practice, as it has slightly lower variance than the after-resampling estimate $L_n^{A,\theta,\alpha}$, but the latter is useful in deriving properties of the MOP-$\alpha$ gradient estimate such as that in Theorem \ref{thm:mop-functional-forms}.
+The before-resampling estimate $L_n^{B,\theta,\alpha}$ is preferable in practice, as it has slightly lower variance than the after-resampling estimate $L_n^{A,\theta,\alpha}$, but the latter is useful in deriving properties of the MOP-$\alpha$ gradient estimate such as that in Theorem \ref{thm:mop-functional-forms}.
 
@@
 
-As mentioned earlier, our approach has similarities with \cite{svensson18}. 
-\cite{svensson18} construct a deterministic local approximation of the likelihood at the $k$-th optimization step via saving the particles from a run at some $\theta_{k-1}$, performing a reweighting with transition and measurement density ratios to evaluate the likelihood at any sufficiently nearby $\theta$, and then using a $0$-th order optimizer such as \texttt{optim} in \texttt{R} to choose a $\theta_k$ that maximizes the likelihood in that neighborhood.
+Our approach has similarities with \cite{svensson18}, who construct a deterministic local approximation of the likelihood at the $k$-th optimization step via saving the particles from a run at some $\theta_{k-1}$, perform a reweighting with transition and measurement density ratios to evaluate the likelihood at any sufficiently nearby $\theta$, and then use a $0$-th order optimizer such as \texttt{optim} in \texttt{R} to choose a $\theta_k$ that maximizes the likelihood in that neighborhood.
 However, due to their use of transition densities in the reweighting, their approach is not plug-and-play.
@@
   \centering
-    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{../imgs/095/biasvar.png}
+    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{imgs/095/biasvar.png}
     \caption{Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}.
@@
     \centering
-    \includegraphics[width=\arxiv{14cm}{\textwidth/2}]{../imgs/095/tikzcholera.png}
+    \includegraphics[width=\arxiv{14cm}{\textwidth/2}]{imgs/095/tikzcholera.png}
     \vspace*{-7mm}
@@
 \centering
-\input{../imgs/095/table}
+\input{imgs/095/table}
 \caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone.
@@
 \begin{figure}[htbp!]
-    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/pairs.png}
-    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/qq.png}
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{imgs/095/pairs.png}
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{imgs/095/qq.png}
     \caption{Scatterplots depicting the performance of IFAD against that of IF2.
@@
 \begin{figure}[ht]
-    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/boxplot.png}
-    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{../imgs/095/boxplot_all.png}
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{imgs/095/boxplot.png}
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{imgs/095/boxplot_all.png}
     \caption{\textbf{Left:} Raincloud plot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs, representing the common procedure of running a few searches and choosing the best one.
@@
     \centering
-    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{../imgs/095/optim.png}
+    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{imgs/095/optim.png}
     \caption{Optimization progress of IFAD and IF2.
@@
     \centering
-    \includegraphics[width=\textwidth/\real{1.25}]{../imgs/pmcmc/nuts_eb.png}
+    \includegraphics[width=\arxiv{\textwidth/\real{1.25}}{\textwidth/\real{2.5}}]{imgs/pmcmc/nuts_eb.png}
     \caption{Convergence diagnostics for NUTS with the informative nonparametric empirical Bayes prior from applying KDE to preliminary searches using IF2, with $4$ chains.
@@
 Despite showing that the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} estimate the Fisher score by
-$$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}\left(x_{0:n,j}^{A, F,\theta}, y_{1:N}^* ; \theta\right)$$ and not
-$$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}| X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}; \theta\right).$$
+$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}\left(x_{0:n,j}^{A, F,\theta}, y_{1:N}^* ; \theta\right)$ and not
+$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}| X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}; \theta\right).$
 It is therefore not immediately apparent that these two converge to the same thing. As such we directly show the consistency of the MOP-$1$ gradient estimate below. 
@@
     \vspace*{-1ex}
-    \begin{eqnarray}
+    \arxiv{\begin{eqnarray}
         \E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)\big\|_2^2 
@@
         \Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) &\lesssim& \min_{k\leq N} Np \, G'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}N\right). \label{eq:mop-variance}
-        \end{eqnarray}
+        \end{eqnarray}}{\begin{align}
+        &\E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)\big\|_2^2 \label{eq:mop-mse}\\
+        &\lesssim \min_{k \leq N} Np \, G'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi(\alpha)\right), \nonumber 
+        \\
+        &\Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) \lesssim \min_{k\leq N} Np  G'(\theta)^2\hspace{-0.8ex}\left(\hspace{-0.8ex}\frac{k^2}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}N\hspace{-0.8ex}\right).
+\label{eq:mop-variance}
+        \end{align}}
 \end{thm}
@@
 We thank Nicolas Chopin for helpful communications regarding the strong law of large numbers for off-parameter resampled particle filters.
-We also thank Arnaud Doucet for helpful discussion regarding the manuscript, and for making the connection between our approach and that of \cite{svensson18}.
+We also thank Arnaud Doucet for helpful discussion regarding the manuscript.
 }{

### 2024-07-14 paper/appendix/coverletter.tex
update from overleaf

\documentclass[11pt]{article}

\begin{document}

\noindent Dear PNAS editors,
\newline

Please consider our manuscript, "Accelerated Inference for Partially Observed Markov Processes using Automatic Differentiation", for publication in PNAS.

The contribution of this paper has widespread implications due to the diverse uses of this model class in ecology, engineering, epidemiology, finance, and elsewhere. Our method promises substantially improved inference wherever simulation-based inference is used for the full-information likelihood in complex partially observed Markov processes. That is why we consider publication of our article in PNAS would facilitate a broad range of scientific advances.

Although our algorithm can be implemented by a programmer familiar with current machine learning tools, the technology behind our contribution is rather technical. The justification for our new algorithm requires deep familiarity with the theory of sequential Monte Carlo (SMC) and some experience with the theory and practice of automatic differentiation (AD). Thus, relatively few researchers are well placed to fully appreciate and evaluate our contribution. Leading SMC researchers who have already considered algorithms incorporating AD include Arnaud Doucet, Frank Wood, and George Deligiannidis. Nicolas Chopin is a leading SMC theorist who is also involved in developing SMC computation so he would also be a qualified referee. It would assist the evaluation of our contribution if at least one referee or invited editor could be drawn from this select group of experts.

This manuscript focuses on likelihood-based non-Bayesian inference. We also demonstrate relevance to Bayesian inference, but that is not our primary topic. Therefore, we ask for at least one referee who has experience working outside the Bayesian paradigm.

\vspace{3ex}

\noindent Sincerely,

\noindent Kevin Tan, Ed Ionides and Giles Hooker
\end{document}

### 2024-07-14 paper/ms.tex
update from overleaf

@@
 
+\AtEndDocument{\refstepcounter{thm}\label{finalthm}}
+\AtEndDocument{\refstepcounter{lem}\label{finallem}}
+\AtEndDocument{\refstepcounter{defn}\label{finaldefn}}
+\AtEndDocument{\refstepcounter{prop}\label{finalprop}}
+\AtEndDocument{\refstepcounter{aspt}\label{finalaspt}}
+
 \begin{document}
@@
 \authordeclaration{The authors declare no competing interests.}
-\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: kevtan@wharton.upenn.edu}
+\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: ionides@umich.edu}
 % At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
-\keywords{Sequential Monte Carlo $|$ Automatic Differentiation $|$ Particle Filter $|$ Markov Process $|$ Maximum Likelihood}
+\keywords{Sequential Monte Carlo $|$ Automatic Differentiation $|$ Particle Filter $|$ Markov Process $|$ Likelihood}
 }
@@
 Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques because widely used particle filter algorithms yield an estimated likelihood function that is discontinuous as a function of the model parameters.
-We show how to embed two existing AD particle filter methods in a theoretical framework that provides an extension to a new class of algorithms.
+We embed two existing AD particle filter methods in a theoretical framework that provides an extension to a new class of algorithms.
 This new class permits a bias/variance tradeoff and hence a mean squared error substantially lower than the existing algorithms.
@@
 An attractive feature of basic particle filter algorithms, also known as boostrap filters, is that they do not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
-This plug-and-play property is useful for scientific applications \cite{he10}.
+This \textit{plug-and-play property} is useful for scientific applications \cite{he10}.
 Some plug-and-play methods depend on the construction of low-dimensional summary statistics \cite{wood10,toni09}, sacrificing statistical efficiency for computational convenience.
@@
 However, existing approaches are either asymptotically biased \cite{naesseth18, jonschkowski18}, have high variance \cite{poyiadjis11, scibior21}, are computationally expensive \cite{corenflos21, chen24}, or require access to transition densities \cite{poyiadjis11, scibior21, singh22, chen24}.
-
 Scibior and Wood \cite{scibior21} showed that the estimators derived by Poyiadjis et al.
 \cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.
+
 However, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability, while the stop-gradient procedure is introduced without mathematical motivation.
@@
 
-MOP-$\alpha$ avoids the issue of asymptotic bias \cite{corenflos21} arising from having to drop terms arising from resampling when differentiating through the standard particle filter as in \cite{naesseth18} in a novel way.
-This particle filter is smooth by construction, allowing us to bypass the seemingly incompatible paradox of differentiating through discrete Monte Carlo sampling by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
+One can simply drop terms arising from resampling when differentiating through the particle filter \cite{naesseth18}, but this leads to asymptotic bias \cite{corenflos21}.
+MOP-$\alpha$ is constructed as a smooth particle filter, thus bypassing the seemingly incompatible paradox of differentiating through discrete Monte Carlo sampling by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
 Yet, MOP-$\alpha$ can nevertheless provide strongly consistent particle and log-likelihood estimates (Theorem \ref{thm:mop-targeting}), and when $\alpha=1$, consistent score estimates (Theorem \ref{thm:mop-grad-consistency}).
 
-We derive a linear convergence rate for stochastic gradient descent (SGD) using MOP-$\alpha$ in the  presence of strong convexity (Theorem \ref{thm:mop-convergence}).
+We derive a linear convergence rate for gradient descent using MOP-$\alpha$ in the  presence of strong convexity and smoothness (Theorem \ref{thm:mop-convergence}).
 Critically, the estimator converges quickly within a neighborhood of the maximum but struggles to reach this neighborhood in the presence of local minima and saddle points.
@@
 
+\arxiv{}{\vspace*{-2mm}}
 \section{Problem Setup}
@@
 We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
-Superscripts $x_{n,j}^A$ of particles denote the ancestral trajectory of particle $j$ at time $n$, $a(\cdot)$ is the ancestor function that maps a particle to its parent, $j \mapsto a(j)$, and $k_j$
+Superscripts $x_{n,j}^A$ of particles denote the ancestral trajectory of particle $j$ at time $n$,  and $k_j$
 denotes the resample indices drawn for each particle $j$. 
+%$a(\cdot)$ is the ancestor function that maps a particle to its parent, $j \mapsto a(j)$,
 
@@
 
+\arxiv{}{\vspace*{-2mm}}
 \section{Off-Parameter Particle Filters}
@@
 
+\arxiv{}{\vspace*{-2mm}}
 \subsection{Algorithm Outline}
@@
 
-If $\alpha=1$ in Algorithm~\ref{alg:mop}, the weights, $w^{P,\theta}_{n,j}$ and $w^{F,\theta}_{n,j}$, accumulate as $n$ increases.
-This leads to numerical instability unless $N$ is small.
-For $\alpha<1$, the weights fom previous timesteps are discounted, as in Equation~(\ref{eq:weighting-scheme}).
-Heuristically, $\alpha$ controls a rate of exponential decay of the memory that the filter at time $t_n$, and its resulting gradient estimate, has over the ancestral trajectories prior to $t_n$.
-With $\alpha<1$, Algorithm~\ref{alg:mop} continues to target the filtering distribution and likelihood if $\theta=\phi$.
-However, we expect bias for $\theta\neq\phi$ that shrinks as $\theta$ approaches $\phi$.
+Heuristically, $\alpha$ controls the exponentially decaying memory that the filter at time $t_n$, and its resulting gradient estimate, has over the ancestral trajectories prior to $t_n$.
+If $\alpha=1$ in Algorithm~\ref{alg:mop}, the gradient of the weights, $w^{P,\theta}_{n,j}$ and $w^{F,\theta}_{n,j}$, accumulates as $n$ increases.
+This leads to numerical instability and high variance unless $N$ is small.
+For $\alpha<1$, the weights from previous timesteps are discounted, as in Equation~(\ref{eq:weighting-scheme}).
+Algorithm~\ref{alg:mop} continues to target the filtering distribution and likelihood if $\theta=\phi$, but has an asymptotically biased score estimate.
 This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate.
+%bias for $\theta\neq\phi$ that shrinks as $\theta$ approaches $\phi$
+%
 
@@
 
-Our approach has similarities with \cite{svensson18}, who construct a deterministic local approximation of the likelihood at the $k$-th optimization step via saving the particles from a run at some $\theta_{k-1}$, perform a reweighting with transition and measurement density ratios to evaluate the likelihood at any sufficiently nearby $\theta$, and then use a $0$-th order optimizer such as \texttt{optim} in \texttt{R} to choose a $\theta_k$ that maximizes the likelihood in that neighborhood.
-However, due to their use of transition densities in the reweighting, their approach is not plug-and-play.
-The variance of their log likelihood estimate is also roughly $O(||\theta - \theta_{k-1}||_2^{2N})$, where $N$ is the horizon.
-Our approach bypasses these two issues.
-The use of the \textit{reparameterization trick} to consider particle paths dependent on $\theta$ (and not $\theta_{k-1}$ or $\phi$) allows us to retain the plug-and-play property.
-On the other hand, our use of AD allows us to only evaluate the likelihood and score estimates when $\theta=\phi$.
-The use of AD bypasses the issue of exponentially increasing variance, but nevertheless can result in problematic polynomial variance.
-Our introduction of $\alpha$ avoids that difficulty.
+Our approach has similarities with \cite{svensson18}, who reweight particles from a previous run at $\phi$ to evaluate the likelihood at any sufficiently nearby parameter $\theta$ for $0$-th order optimization.
+However, their approach is not plug-and-play because they use transition densities to reweight particles, and the variance of their log-likelihood estimate is exponentially increasing in the horizon.
+Our approach bypasses these issues -- the first with the \textit{reparameterization trick} to consider particle paths dependent only on $\theta$, and the second as using AD allows us to only evaluate the likelihood and score estimates when $\theta=\phi$.
+%The use of AD bypasses the issue of exponentially increasing variance, but nevertheless can result in problematic polynomial variance.
+% Our introduction of $\alpha$ avoids that difficulty.
+%also roughly $O(||\theta - \theta_{k-1}||_2^{2N})$, where $N$ is
 
+\arxiv{}{\vspace*{-2mm}}
 \subsection{MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, naesseth18}}
@@
     \end{equation}
-    yielding the estimate of \cite{naesseth18} for the bootstrap filter. When $\alpha=1$,
+    yielding \cite{naesseth18}'s estimator with the bootstrap filter. When $\alpha=1$,
     \vspace*{-2.5mm}
@@
 We defer the proof to \arxiv{Appendix~\ref{appendix:functional}}{the supplementary material}. 
-The argument relies on a useful decomposition of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$ that yields a telescoping product in the MOP-$1$ case. 
+The argument relies on a useful decomposition of the after-resampling estimate $L_n^{A,\theta,\alpha}$. 
 Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting $\theta=\phi$ implies that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result. 
-
 This further illustrates how $\alpha$ dictates the memory of the gradient estimate. 
 As \cite{scibior21} remarks, the MOP-$0$ estimator of \cite{naesseth18} only considers single-step quantities, and is ``memoryless'' beyond a single step. 
-This is in contrast to the case when $\alpha=1,$ as that estimate, studied by \cite{poyiadjis11}, only considers the surviving particles at time $N$ and so fully tracks dependencies over time.
+In contrast, the MOP-$1$ estimate of \cite{poyiadjis11, scibior21} only considers the surviving particles at time $N$ and so fully tracks dependencies over time.
 
@@
     \label{fig:biasvar}
+    \arxiv{}{\vspace*{-1ex}}
 \end{figure}
 
+\arxiv{}{\vspace*{-2mm}}
 \subsection{Summary of Theoretical Guarantees}
@@
 
+\arxiv{}{\vspace*{-2mm}}
 \section{Practical Maximum-Likelihood Estimation}
@@
 One then only needs to run one particle filter at $\theta=\phi$, setting the particles at $\phi$ to be copies of the particles at $\theta$ where gradients don't propagate.
-This is done algorithmically through the \texttt{stop\_gradient()} function in \texttt{JAX}, providing a mathematical justification for the use of the stop-gradient operation by \cite{scibior21}.
+This is done algorithmically through the \texttt{stop\_gradient()} function in \texttt{JAX}, coinciding with and justifying the use of the stop-gradient operation in \cite{scibior21}.
 
+\arxiv{}{\vspace*{-2mm}}
 \subsection{Optimization}
 
-This still leaves us with the question of designing an effective procedure for likelihood maximization with MOP-$\alpha$.
-We propose a simple algorithm we call Iterated Filtering with Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad} that runs a few iterations of IF2 to warm-start an iterative (first or second-order) method that uses the MOP-$\alpha$ gradient estimate.
+To design an effective procedure for likelihood maximization with MOP-$\alpha$, we propose a simple algorithm we call Iterated Filtering followed by Automatic Differentiation (IFAD) in Algorithm \ref{alg:ifad}.
+IFAD runs a few iterations of IF2 to warm-start an iterative first or second-order method that uses the MOP-$\alpha$ gradient estimate.
 This leverages IF2's quick empirical convergence to a neighborhood of the MLE, overcoming the tendency of gradient methods to get stuck in saddle points and local minima.
@@
 
+\arxiv{}{\vspace*{-2mm}}
 \subsection{Linear Convergence Rates}
 
-On the other hand, the second stage of IFAD enjoys a linear convergence rate under the usual smoothness and strong convexity assumptions in the theory of convex optimization, as we show below in Theorem \ref{thm:mop-convergence}.
+On the other hand, the second stage of IFAD enjoys a linear convergence rate under the usual smoothness and strong convexity assumptions in convex optimization theory, as we show below in Theorem \ref{thm:mop-convergence}.
 
@@
 For simplicity, we only prove this for MOP-$1$ and note that the general case follows as the required $\epsilon$-approximation can still be obtained for sufficiently large $\alpha$.
-When $\alpha=1$, the required $J$ is given in the supplementary information in Lemmas \ref{lemma:grad_bound} and \ref{lemma:hess_bound}.
+When $\alpha=1$, the required $J$ is given in \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}.
 This result shows that particle estimates for the gradient can enjoy the same linear convergence rate on well-conditioned problems as gradient descent with access to the score.
-As , we note that it is possible.
 
 We therefore see that the second stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the first stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
-This happens fairly often in practice, for example, when sufficient regularity conditions for local asymptotic normality of the MLE hold.
+This happens fairly often in practice, for example, when local asymptotic normality of the MLE holds.
 We conjecture that this applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE, and will explore this in future work.
@@
 	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 cooling schedule $\eta_m$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $m=0.$\newline
-        Run IF2 until initial "convergence" under cooling schedule $\eta_m$, or for a fixed number or iterations, to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
+        Run IF2 until initial "convergence" under cooling schedule $\eta_m$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
 		\textbf{While} procedure not converged: \newline
 		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_m).$ \newline
-		\hspace*{4mm} Obtain $g(\theta_m) = \nabla_{\theta_m} \big(-\hat\loglik(\theta_m)\big)$, $H(\theta_m)$ s.t. $\lambda_{\min}\big(H(\theta_m)\big) \geq c$. \newline
+		\hspace*{4mm} $g(\theta_m) := \nabla_{\theta}(-\hat\loglik(\theta_m))$, $H(\theta_m)$ s.t. $\lambda_{\min}(H(\theta_m)) \geq c$. \newline
 		\hspace*{4mm} Update $\theta_{m+1} := \theta_m - \eta (H(\theta_m))^{-1} g(\theta_m)$, $m:=m+1.$ \newline
@@
 
+\arxiv{}{\vspace*{-2mm}}
 \section{Application to a Cholera Transmission Model}
@@
 \begin{align*}
-    dS&=\big(k \epsilon R^k+\delta(S-P)-\lambda(t) S\big)\, dt+d P-({\sigma S I}/{H})\, dB, \\
-    dI&=\big(\lambda(t) S-(m+\delta+\gamma) I\big)\, dt+({\sigma S I}/{H})\, dB, \\
+    dS&=\big(k \epsilon R^k+\delta(P-S)-\lambda(t) S\big)\, dt+d P-({\sigma S I}/{P})\, dB, \\
+    dI&=\big(\lambda(t) S-(m+\delta+\gamma) I\big)\, dt+({\sigma S I}/{P})\, dB, \\
     dR^1&=\big(\gamma I-(k \epsilon+\delta) R^1\big)\, dt, \hspace{2mm} \dots \\
@@
     \label{fig:tikz-cholera}
+    \arxiv{}{\vspace*{-3mm}}
 \end{figure}
 
+\arxiv{}{\vspace*{-2mm}}
 \subsection{Results}
@@
 We re-implemented IF2 to do so, but we compare our results with the results of \cite{ionides15} (labeled "IF2 2015").
-Our re-implementation ourperforms that of \cite{ionides15}, likely due to a better choice of algorithmic parameters.
+Our re-implementation reaches a better log-likelihood than that of \cite{ionides15}, likely due to a better choice of algorithmic parameters.
 For each method, we performed 100 searches, initialized with 100 initial starting parameter vectors drawn uniformly from the same wide bounding box used in \cite{ionides15}.
@@
 
+\arxiv{}{\vspace*{-1mm}}
 \begin{table}[h!]
@@
 Our implementation of IF2 outperforms that of \cite{ionides15}, but still ultimately underperforms IFAD.
-IFAD manages to find the MLE, matching the highest log-likelihood previously found in the Dhaka cholera model implemented within the \texttt{pomp} package of \cite{king16}.}
+}
+\arxiv{}{\vspace*{-7mm}}
 \label{table:mle}
@@
 \paragraph{IFAD Successfully Finds the MLE:} Previously, an MLE at a log-likelihood of $-3748.6$ was reported by \cite{king16}.
-This MLE was obtained with much computational effort, using many global and local IF2 searches, and with the assistance of likelihood profiling.
+This MLE was obtained with much computational effort, using many global and local IF2 searches and with the assistance of likelihood profiling, 8 years after the model was first proposed by \cite{king08}.
 Meanwhile, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.6$, while the best log-likelihood found by \cite{king08} was only $-3793.4$.
 Despite being initialized for a global search, IFAD manages to get much closer to the MLE over the $100$ searches than \cite{ionides15} and finds it up to $2.5$ standard deviations of Monte Carlo error, as seen in Table~\ref{table:mle}.
-On this problem, the sequence of local searches, refinement, and likelihood profiling that was previously required for finding the MLE is not necessary with the IFAD algorithm. 
+On this problem, the sequence of local searches, refinement, and likelihood profiling that was previously required for finding the MLE is not necessary with IFAD. 
 In other words, the improvement in numerical efficiency of IFAD over IF2 is so large that routine application of IFAD (consisting of a collection of Monte Carlo replications from random starting points) generates results outside the reach of a routine application of IF2.
-When interpreting Table~\ref{table:mle}, bear in mind that by Wilks' Theorem, any difference of over $1.92$ log units has statistical relevance when testing one parameter at the $0.05$ significance level, and therefore potentially has scientific value.
+When interpreting Table~\ref{table:mle}, bear in mind that by Wilks' Theorem, any difference of over $1.92$ log units has statistical relevance when testing one parameter at the $0.05$ significance level, and therefore may have scientific value.
 
@@
 IFAD, in comparison, approaches the MLE quickly due to the IF2 warm-start (as seen in Figure \ref{fig:optim}) and also succeeds at refining the coarse solution found by the warm-start with MOP gradient steps to find the MLE (as seen in Figures~\ref{fig:scatter} and \ref{fig:boxplot}).
-IFAD therefore successfully combines the best qualities of IFAD and MOP, outperforming either of them alone.
+IFAD therefore successfully combines the best qualities of IF2 and MOP, outperforming either of them alone.
 Monte~Carlo replication is appropriate for IFAD, or any Monte Carlo algorithm used to solve a challenging numerical problem, but Figure~\ref{fig:boxplot} shows that IFAD can find higher likelihood values, more quickly and more reliably than the previous state-of-the-art.
@@
 
-For the results presented here, we did not include the many common heuristics used in the machine learning and optimization literature in the gradient descent stage.
+For the results presented here, we did not employ the many common heuristics used in the machine learning and optimization literature in the gradient descent stage.
 We used constant learning rates of $0.01, 0.05$, and $0.2$ for IFAD-$1,0,$ and $0.97$ respectively, and a constant cooling rate of $0.95$ for our IF2 implementation.
 While techniques such as annealing learning rates, gradient normalization, and momentum could further improve the performance of IFAD in other simulations we performed, we chose to report the results of the simplest implementation to serve as a baseline for the method's performance.
-The results for this basic implementation are already sufficient to show the high potential of the approach.
+The results for this basic implementation already showcase the merits of the approach.
 
@@
     \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{imgs/095/qq.png}
+    %\arxiv{}{\vspace*{-1mm}}
     \caption{Scatterplots depicting the performance of IFAD against that of IF2.
 \textbf{Left:} Paired searches from the same starting point.
-Controlling for initial starting point, tuning $\alpha$ allows IFAD-$0.97$ to strictly improve on IF2, \cite{poyiadjis11}, and \cite{naesseth18}, on almost every iteration.
+Tuning $\alpha$ allows IFAD-$0.97$ to strictly improve on IF2, \cite{poyiadjis11}, and \cite{naesseth18}, on almost every initialization.
 \textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches.
-It is clear that on average, IFAD performs best, and manages to find the MLE while no IF2 search successfully gets within 7 log-likelihood units of it.
+IFAD performs best on average, and approaches the MLE while no IF2 search successfully gets within 7 log-likelihood units of it.
 The dotted red line shows the true maximized log-likelihood.
@@
     \label{fig:scatter}
+    \arxiv{}{\vspace*{-4mm}}
 \end{figure}
@@
     \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{imgs/095/boxplot_all.png}
+    %\arxiv{}{\vspace*{-1mm}}
     \caption{\textbf{Left:} Raincloud plot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs, representing the common procedure of running a few searches and choosing the best one.
@@
     \label{fig:boxplot}
+    \arxiv{}{\vspace*{-5mm}}
 \end{figure}
@@
     \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{imgs/095/optim.png}
+    \arxiv{}{\vspace*{-1mm}}
     \caption{Optimization progress of IFAD and IF2.
@@
     \label{fig:optim}
+    \arxiv{}{\vspace*{-4mm}}
 \end{figure}
 
+\arxiv{}{\vspace*{-2mm}}
 \section{Application to Bayesian Inference}
@@
 
-Particle MCMC, as introduced by \cite{andrieu10}, is arguably the most popular method for full-information plug-and-play Bayesian inference for POMP models.
-The plug-and-play property enables its use in simulation-based Bayesian inference for complex scientific models where computing transition densities is not feasible, such as in disease modeling.
-However, the particle Metropolis-Hastings algorithm often experiences slow mixing and requires long burn-in periods.
-For instance, \cite{fasiolo16} reported needing $700,000$ burn-in iterations for effective sampling in a cholera model.
-This computational burden necessitated simplifying the scientific model to reduce computational costs, by adjusting the length of each Euler timestep to be a month instead of a day.
+Particle MCMC \cite{andrieu10} is arguably the most popular method for full-information plug-and-play Bayesian inference for POMP models.
+The plug-and-play property enables its use in simulation-based Bayesian inference for complex scientific models where transition densities are intractable, such as in disease modeling.
+However, particle Metropolis-Hastings often mixes slowly, requiring long burn-in periods.
+For instance, \cite{fasiolo16} reported needing $700000$ burn-in iterations for effective sampling in the \cite{king08} model, even after coarsening each Euler timestep to reduce the computational burden.
 
-We employ a NUTS sampler powered by a MOP-$\alpha$, with a nonparametric empirical prior initialized by the IF2 warm start from the previous section.
-To construct the prior, we perform a kernel density estimate (KDE) on the parameter swarm from the warm start, and use the output of that as an empirical prior.
-The KDE is performed to ensure some degree of mutual contiguity between the densities.
-This yields a prior that is roughly equivalent to a parameter cloud centered at some point in the neighborhood of the MLE, with a standard deviation of $\sqrt{(0.02\cdot0.95^{40})^2 \cdot 600}\approx 6.3\%$ of the MLE.
+To alleviate this, we employ a NUTS sampler powered by MOP-$\alpha$, with a nonparametric empirical prior constructed from density estimation of the parameter swarm from the IF2 warm start having a standard deviation of $6.3\%$ of the MLE.
+%\sqrt{(0.02\cdot0.95^{40})^2 \cdot 600}\approx 6.3\%
 Using a NUTS sampler powered by MOP-$\alpha$ with this prior significantly lowers the mixing time of the Markov chain to as little as $500$ iterations, as seen in Figure \ref{fig:nuts-eb}.
@@
     \includegraphics[width=\arxiv{\textwidth/\real{1.25}}{\textwidth/\real{2.5}}]{imgs/pmcmc/nuts_eb.png}
+    \arxiv{}{\vspace*{-1mm}}
     \caption{Convergence diagnostics for NUTS with the informative nonparametric empirical Bayes prior from applying KDE to preliminary searches using IF2, with $4$ chains.
@@
     \label{fig:nuts-eb}
+    \arxiv{}{\vspace*{-4mm}}
 \end{figure}
 
-For completeness, we note that without the empirical Bayes-style prior from IF2, NUTS with the MOP-$\alpha$ estimate alone tended to diverge.
-In contrast, particle Metropolis-Hastings (with or without the empirical Bayes prior from IF2) fails to effectively explore the parameter space.
-We display the results of the above in Figures \ref{fig:mh} and \ref{fig:nuts} \arxiv{in Appendix~\ref{appendix:bayes}}{within the supplementary information}. 
-The result obtained in this section, of a speedup in the mixing time of particle MCMC by over three orders of magnitude, provides hope that particle Bayesian inference might finally be practical for the complex scientific models commonly encountered in areas like epidemiology.
+Without the IF2 prior, NUTS with the MOP-$\alpha$ estimate alone tended to diverge.
+In contrast, particle Metropolis-Hastings, with or without said prior, fails to effectively explore the parameter space.
+We display these results \arxiv{in Figures \ref{fig:mh} and \ref{fig:nuts} in Appendix~\ref{appendix:bayes}}{within the supplementary information}. 
+This speedup in the mixing time of particle MCMC by over three orders of magnitude provides hope that particle Bayesian inference might be increasingly employed for the complex scientific models commonly encountered in areas like epidemiology.
 
+\arxiv{}{\vspace*{-2mm}}
 \section{Theoretical Analysis of MOP-$\alpha$}
@@
 
+\arxiv{}{\vspace*{-2mm}}
 \begin{enumerate}[label=(A\arabic*),itemsep=-1.2ex] 
@@
 \end{enumerate}
+\arxiv{}{\vspace*{-4mm}}
 
@@
 
+\arxiv{}{\vspace*{-2mm}}
 \subsection{MOP-$\alpha$ Targets the Filtering Distribution}
@@
 
-While the result is presented here as specific to MOP-$\alpha$, we actually prove a more general result in \arxiv{Appendix~\ref{appendix:targeting}}{the supplementary material}.
-That is, we show a strong law of large numbers for triangular arrays of particles with off-parameter resampling, where we resample the particles according to an arbitrary resampling rule not necessarily in proportion to the targeted distribution of interest and employ weights that encode the cumulative discrepancy between the resampling and the target distribution instead of equal weights.
+While the result is presented here as specific to MOP-$\alpha$, we prove a more general result in \arxiv{Appendix~\ref{appendix:targeting}}{the supplementary material}.
+This is a strong law of large numbers for triangular arrays of particles resampled according to an arbitrary resampling rule not necessarily in proportion to the targeted distribution of interest, to weights that encode the cumulative discrepancy between the resampling and the target distribution.
 
@@
 That is, for $\pi_n(\theta)=f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^* ; \theta)$ and any measurable and bounded functional $h$ and for $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\prod_{n=1}^N L_n^{B,\theta,\alpha}$, it holds that
-    \arxiv{}{\vspace*{-2.5mm}}
+    \arxiv{}{\vspace*{-1.5mm}}
     \begin{equation} \nonumber
         \frac{\sum_{j=1}^J h(x_{n,j}^{F, \theta}) \, w_{n,j}^{F,\theta}}{\sum_{j=1}^J w_{n,j}^{F,\theta}} \stackrel{a.s.}{\to} E_{\pi_n(\theta)} \big[h(X_n)\big], \hspace{5mm} \hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).
-    \arxiv{}{\vspace*{-2.5mm}}
+    \arxiv{}{\vspace*{-1.5mm}}
     \end{equation}
@@
 \begin{proof}
-    We provide a proof sketch here, deferring most of the details, and discussion of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$, to \arxiv{Appendix~\ref{appendix:targeting}}{the supplementary information}.
-    When $\theta=\phi$, regardless of the value of $\alpha$, the ratio ${g_{n,j}^\theta}/{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter.
+    We provide a proof sketch here, deferring the details to \arxiv{Appendix~\ref{appendix:targeting}}{the supplementary information}. 
+    When $\theta=\phi$, the ratio ${g_{n,j}^\theta}/{g_{n,j}^\phi}=1$ regardless of $\alpha$, reducing to the vanilla particle filter.
     When $\alpha=1$ and $\theta\neq\phi,$ suppose inductively that $\{(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j})\}_{j=1}^J$ targets $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
@@
     resampling the $k_j$ with probabilities proportional to $g^\phi_{n,j}$, also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-    If the likelihood is estimated with the before-resampling conditional likelihoods $\hat\lik(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, the strong consistency is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+    Estimating the likelihood with the before-resampling estimator $\hat\lik(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, we see that the strong consistency is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
 \end{proof}
 
+\arxiv{}{\vspace*{-2mm}}
 \subsection{MOP-$1$ Is Consistent for the Score}
 
-Despite showing that the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} estimate the Fisher score by
-$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}\left(x_{0:n,j}^{A, F,\theta}, y_{1:N}^* ; \theta\right)$ and not
-$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}| X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}; \theta\right).$
+Although we have shown the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} estimate the Fisher score by
+$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{0:n,j}^{A, F,\theta}, y_{1:N}^* ; \theta)$ and not
+$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}| X_{1:N}}(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}; \theta).$
 It is therefore not immediately apparent that these two converge to the same thing. As such we directly show the consistency of the MOP-$1$ gradient estimate below. 
@@
 \begin{proof}
-    Fix $\omega \in \Omega$, and set $\phi = \theta$, where $\theta$ is the point at which we wish to evaluate the gradient.
+    Fix $\omega \in \Omega$, $\phi = \theta$.
 The sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$ by Assumption \ref{assump:local-bounded-derivative}.
-Again by Assumption \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J^1(\theta)(\omega)|_{\theta=\theta'}$ is also uniformly bounded over all $J$ by $H^*$ for almost every $\omega\in \Omega$ and every $\theta'\in \Theta$.
+Again by \ref{assump:local-bounded-derivative}, the second derivative of $\hat\lik_J^1(\theta)(\omega)|_{\theta=\theta'}$ is also uniformly bounded over all $J$ by $H^*$ for almost every $\omega\in \Omega$ and every $\theta'\in \Theta$.
 So $(\nabla_\theta \hat\lik_J^1(\theta)( \omega))_{J \in \mathbb{N}}$ is uniformly Lipschitz, and therefore uniformly equicontinuous for almost every $\omega \in \Omega$.
@@
     By Arzela-Ascoli, there is a uniformly convergent subsequence.
-But there is only one subsequential limit, as we can treat the gradient estimate at $\theta$ as a bounded functional of the particles by Assumption \ref{assump:local-bounded-derivative}, allowing us to apply Theorem \ref{thm:mop-targeting} to see that the sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$.
+But there is only one subsequential limit, as we can treat the gradient estimate at $\theta$ as a bounded functional of the particles by \ref{assump:local-bounded-derivative}, allowing us to apply Theorem \ref{thm:mop-targeting} to see that the sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$.
 So the whole sequence must converge uniformly to $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega).$
 
-    With uniform convergence for the derivatives established, we can swap the limit and derivative and obtain, in conjunction with the strong consistency $\hat{\lik}_J^1(\theta)(\omega) \stackrel{a.s.}{\to} \lik(\theta)$ in Theorem \ref{thm:mop-targeting}, that for almost every $\omega \in \Omega$,
+    With uniform convergence for the derivatives established, we can swap the limit and derivative and obtain, with the strong consistency $\hat{\lik}_J^1(\theta)(\omega) \stackrel{a.s.}{\to} \lik(\theta)$ in Theorem \ref{thm:mop-targeting}, that for almost every $\omega \in \Omega$,
     $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lik(\theta).$
-    The result then follows by the continuous mapping theorem. 
+    The result follows by the continuous mapping theorem. 
 \end{proof}
 
+\arxiv{}{\vspace*{-2mm}}
 \subsection{MOP-$\alpha$ Error, Bias and Variance}
@@
 
-\begin{thm}
+\begin{thm}[MSE and Variance of MOP-$\alpha$]
     \label{thm:mop-biasvar}
@@
 \end{thm}
+\arxiv{}{\vspace*{-1ex}}
+
 We defer the proof to \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material}, but provide a brief outline here. 
@@
 \paragraph{Interpretation:} Theorem \ref{thm:mop-biasvar} provides theoretical understanding of the results observed in Figure \ref{fig:biasvar} that show the empirically favorable bias-variance tradeoff enjoyed by MOP-$\alpha$.
-The first term in the MSE bound in Equation \ref{eq:mop-mse} corresponds to the error of the particle approximation, the second mixing error, and the third and fourth the error between the MOP-$(1,k)$ estimate and the MOP-$\alpha$ estimate.
+The first term in the MSE bound in Equation \ref{eq:mop-mse} corresponds to the error of the particle approximation, the second mixing error, and the third and fourth the error between the MOP-$(1,k)$ and MOP-$\alpha$ estimates.
 As $\alpha$ goes to $1$, choosing $k$ appropriately, the first, third and fourth term (corresponding to the variance) increases, while the second term (corresponding to the bias) decreases. 
@@
 We also show that MOP-$0$ achieves a MSE of $\tilde{O}\big(NpG'(\theta)^2(J^{-1}+(1-\epsilon)^{\lfloor1/c\log(J)\rfloor})\big)$, where the second term corresponds to uncontrolled mixing error.
-Unlike MOP-$\alpha$, MOP-$0$ has no opportunity to tune $\alpha$ and $k$ to reduce said mixing error, leading to uncontrolled asymptotic bias.
+Unlike MOP-$\alpha$, MOP-$0$ has no opportunity to tune $\alpha$ (or $k$) to reduce said mixing error, leading to uncontrolled asymptotic bias.
 
+\arxiv{}{\vspace*{-2mm}}
 \section{Computational Efficiency}
@@
 
+\arxiv{}{\vspace*{-2mm}}
 \section{Discussion}
@@
 This algorithm may be better suited to large discrete state spaces than either MOP-$\alpha$ or the Baum-Welch algorithm.
-Additionally, our gradient estimate can be used for variational inference to approximate the posterior distribution over latent states \cite{naesseth18} and parameters.
+
+Our gradient estimate can be used for variational inference to approximate the posterior distribution over latent states \cite{naesseth18} and parameters.
 
@@
 \showmatmethods{} % Display the Materials and Methods section
-\acknow{Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.}
+\acknow{
+This research was supported by National Science Foundation grant DMS-1761603. 
+We thank Nicolas Chopin for helpful communications regarding the strong law of large numbers for off-parameter resampled particle filters.
+We also thank Arnaud Doucet for helpful discussions regarding the manuscript.}
 \showacknow{} % Display the acknowledgments section

### 2024-07-14 paper/si.tex
Updates from Overleaf
update from overleaf

@@
 \documentclass[9pt,twoside,lineno]{pnas-new-si}
+
 % Use the lineno option to display guide line numbers if required.
@@
 \externaldocument{ms}
+
 \templatetype{pnassupportinginfo}
-\input{paper/macros}
+\input{macros}
+\usepackage{refcount}
+
+\setcounter{thm}{\getrefnumber{finalthm}}
+\addtocounter{thm}{-1}
+
+\setcounter{lem}{\getrefnumber{finallem}}
+\addtocounter{lem}{-1}
+
+\setcounter{defn}{\getrefnumber{finaldefn}}
+\addtocounter{defn}{-1}
+
+\setcounter{prop}{\getrefnumber{finalprop}}
+\addtocounter{prop}{-1}
+
+\setcounter{aspt}{\getrefnumber{finalaspt}}
+\addtocounter{aspt}{-1}
 
@@
 
-\section{Heading}
+\section{MOP-$\alpha$ Functional Forms}
 
-\ref{alg:mop}
+\label{appendix:functional}
+\input{appendix/functional}
 
-\subsection*{Subhead}
-Type or paste text here. You may break this section up into subheads as needed (e.g., one section on ``Materials'' and one on ``Methods'').
+\section{Optimization Convergence Analysis}
+\label{appendix:convergence}
+\input{appendix/convergence}
 
-\subsection*{Materials}
-Add a materials subsection if you need to.
+\section{Feynman-Kac Models and Monte Carlo Approximations}
+\label{appendix:feynman}
+\input{appendix/feynman}
 
-\subsection*{Methods}
-Add a methods subsection if you need to.
+\section{A Strong Law of Large Numbers for Triangular Arrays of Particles With Off-Parameter Resampling}
+\label{appendix:targeting}
+\input{appendix/targeting}
+
+\section{Consistency of Off-Parameter Resampled Gradient Estimates}
+\label{appendix:consistency}
+\input{appendix/consistency}
+
+\section{Bias-Variance Analysis}
+\label{appendix:biasvar}
+\input{appendix/biasvar}
+
+\section{Figures for Bayesian Inference}
+\label{appendix:bayes}
+\input{appendix/bayes}
 
 %%% Each figure should be on its own page
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{example-image}
-\caption{First figure}
-\end{figure}
 
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{frog}
-\caption{Second figure}
-\end{figure}
-
-\begin{table}\centering
-\caption{This is a table}
-
-\begin{tabular}{lrrr}
-Species & CBS & CV & G3 \\
-\midrule
-1. Acetaldehyde & 0.0 & 0.0 & 0.0 \\
-2. Vinyl alcohol & 9.1 & 9.6 & 13.5 \\
-3. Hydroxyethylidene & 50.8 & 51.2 & 54.0\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-%%% Add this line AFTER all your figures and tables
-\FloatBarrier
-
-\movie{Type legend for the movie here.}
-
-\movie{Type legend for the other movie here. Adding longer text to show what happens, to decide on alignment and/or indentations.}
-
-\movie{A third movie, just for kicks.}
-
-\dataset{dataset_one.txt}{Type or paste legend here.}
-
-\dataset{dataset_two.txt}{Type or paste legend here. Adding longer text to show what happens, to decide on alignment and/or indentations for multi-line or paragraph captions.}
-
-\bibliography{pnas-sample}
+\bibliography{paper/bib-ifad}
 

### 2024-07-14 paper/ms.tex
Sec 3A edits

@@
 MOP-$\alpha$ is constructed as a smooth particle filter, thus bypassing the seemingly incompatible paradox of differentiating through discrete Monte Carlo sampling by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
-Yet, MOP-$\alpha$ can nevertheless provide strongly consistent particle and log-likelihood estimates (Theorem \ref{thm:mop-targeting}), and when $\alpha=1$, consistent score estimates (Theorem \ref{thm:mop-grad-consistency}).
+Yet, MOP-$\alpha$ can nevertheless provide strongly consistent estimates of the filter distribution and the log-likelihood (Theorem \ref{thm:mop-targeting}).
+When $\alpha=1$, it also provides consistent score estimates (Theorem \ref{thm:mop-grad-consistency}).
 
@@
 These improvements also extend to Bayesian inference, as we show in Section \ref{sec:bayes} that we can use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with a nonparametric empirical Bayes-style prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{andrieu10} on the Dhaka cholera model of \cite{king08} from the $700,000$ iterations in \cite{fasiolo16} to just $500$.
-To the best of our knowledge, attaining rapid mixing on this challenging problem has not previously been achievable.
+To the best of our knowledge, attaining rapid mixing on this challenging problem has not previously been achieved.
 
@@
 
-The above densities are defined on a probability space $(\Omega, \Sigma, \prob)$ which is also assumed to enable construction of independent replicates and all other random variables defined in our algorithms.
+The above densities are defined on a probability space $(\Omega, \Sigma, \prob)$ which is also assumed to enable construction of independent replicates and all random variables defined in our algorithms.
 For the algorithmic interpretation of our theory, we identify the random number seed with an element of the sample space $\omega \in \Omega$.
@@
 
-Our approach has similarities with \cite{svensson18}, who reweight particles from a previous run at $\phi$ to evaluate the likelihood at any sufficiently nearby parameter $\theta$ for $0$-th order optimization.
+Our approach has similarities with \cite{svensson18}, in which particles from a previous run at $\phi$ are reweighted to evaluate the likelihood at any sufficiently nearby parameter $\theta$ for $0$-th order optimization.
 However, their approach is not plug-and-play because they use transition densities to reweight particles, and the variance of their log-likelihood estimate is exponentially increasing in the horizon.
@@
 IFAD runs a few iterations of IF2 to warm-start an iterative first or second-order method that uses the MOP-$\alpha$ gradient estimate.
-This leverages IF2's quick empirical convergence to a neighborhood of the MLE, overcoming the tendency of gradient methods to get stuck in saddle points and local minima.
-Conversely, switching to gradient ascent with MOP-$\alpha$ score estimates lets one bypass the difficulty that IF2 has with optimizing the last few units of log-likelihood.
-Combining these two methods in this way lets us enjoy the best of both worlds.
+To see the importance of this innovation, we briefly summarize the empirical results in Section~\ref{sec:cholera}.
 
@@
 In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations.
-Finding the MLE itself with IF2, however, takes much longer, as one has to use a less aggressive cooling rate to do so. 
-For example, \cite{ionides15} use 100 iterations with the Dhaka cholera model of \cite{king08}, while \cite{wheeler23} use 200 with Model 1 in \cite{lee20}. 
-By simply requiring that the first stage of IFAD get to a neighborhood of the MLE and not the MLE itself, we are able to substantially reduce the number of IF2 iterations required.
+However, even slower cooling and large amounts of Monte Carlo replication using IF2 fail to get within five log-likelihood units of the maximum on this problem (Figure~\ref{fig:boxplot}).
+These numerical results are consistent with theoretical results for iterated filtering \cite{doucet15-if}.
+IF2, without assistance from AD, requires additional computationally intensive techniques, such as repeated pruning and profiling, to reach the inferentially critical region within one log unit of the maximum.
+
+Gradient methods can struggle with saddle points, local minima, and nonlinear ridges.
+Without initialization using IF2, Section~\ref{sec:cholera} shows that AD alone can perform poorly.
+Conversely, switching to gradient ascent with MOP-$\alpha$ score estimates lets one bypass the difficulty that IF2 has with optimizing the last few units of log-likelihood. 
+In combination, these two methods avoid their respective weaknesses and provide, for the first time, straightforward and effective likelihood maximization for our benchmark problem.
 
@@
 Choose the learning rate $\eta$ such that $\eta \leq \frac{c(1-\beta)}{2\Gamma}$.
-Then, for sufficiently large $\alpha$ and $J$ to ensure the score estimate is an $\epsilon$-approximation and the minimum eigenvalue of $H$ is greater than some $c > 0$ for all $m$ with probability at least $1-\delta$, the second stage of IFAD converges linearly to the MLE:
+Then, for sufficiently large $\alpha$ and $J$ to ensure the score estimate is an $\epsilon$-approximation and the minimum eigenvalue of $H$ is greater than some $c > 0$ for all $m$ with probability at least $1-\delta$, the second stage of IFAD converges linearly to the MLE.
+
 \arxiv{}{\vspace*{-2mm}}
@@
 \section{Application to a Cholera Transmission Model}
+\label{sec:cholera}
 

### 2024-07-15 paper/ms.tex
Updates from Overleaf
some more edits to 3A and 3B

@@
 The convergence of IF2 (and so the first stage of IFAD) to a neighborhood of the MLE happens rapidly in practice. 
-In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, initial convergence happens within 40 iterations.
-However, even slower cooling and large amounts of Monte Carlo replication using IF2 fail to get within five log-likelihood units of the maximum on this problem (Figure~\ref{fig:boxplot}).
-These numerical results are consistent with theoretical results for iterated filtering \cite{doucet15-if}.
+In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used,  \ed{[initial convergence happens  BE MORE PRECISE] initial convergence to within xxx=40??
+log-likelihood units of the maximum} occurs within 40 iterations.
+However, even when applying much larger amounts of Monte Carlo effort, IF2 fails to get within five log-likelihood units of the maximum on this problem (Figure~\ref{fig:boxplot}).
+These numerical results are consistent with theoretical results for iterated filtering, showing slow convergence close to the maximum \cite{doucet15-if}.
 IF2, without assistance from AD, requires additional computationally intensive techniques, such as repeated pruning and profiling, to reach the inferentially critical region within one log unit of the maximum.
@@
 
-On the other hand, the second stage of IFAD enjoys a linear convergence rate under the usual smoothness and strong convexity assumptions in convex optimization theory, as we show below in Theorem \ref{thm:mop-convergence}.
+The AD stage confers a linear convergence rate for IFAD under the usual smoothness and strong convexity assumptions in convex optimization theory, as we show below in Theorem \ref{thm:mop-convergence}.
 
-\begin{thm}[Linear Convergence of IFAD]
+\begin{thm}[Linear Convergence of IFAD] \label{thm:linear}
 
@@
 \arxiv{}{\vspace*{-5mm}}
-We borrow from tools in the field of randomized numerical linear algebra \cite{mahoney16} to solve this problem.
-The proof, which is similar to that of Theorem 6 in \cite{mahoney16}, is deferred to \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}.
+The proof of Theorem~\ref{thm:linear} uses results from randomized numerical linear algebra, building on Theorem~6 of\cite{mahoney16}. 
+Details are deferred to \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}. 
+\ed{TECHNICALLY, IF WE CLAIM THAT A PROOF FOR ALPHA=1 CAN BE GENERALIZED TO ANY ALPHA, WE ARE CLAIMING A PROOF FOR ALL ALPHA, AND SO WE DO NOT NEED TO SAY WE ARE ONLY PROVING IT FOR ALPHA=1.
+IF WE REALLY ONLY CLAIM TO HAVE SHOWN IT FOR ALPHA=1, THAT SHOULD BE IN THE STATEMENT.
+IF WE CLAIM THE WHOLE RESULT, PERHAPS WE CAN POSTPONE THE DISCUSSION OF THIS 2-STAGE METHOD TO THE SUPPLEMENT?
+AT THIS POINT, ``THE REQUIRED EPSILON-APPROXIMATION'' IS NOT DEFINED FOR THE READER, SO IT IS HARD TO COMPREHEND THE ARGUMENT.}
 For simplicity, we only prove this for MOP-$1$ and note that the general case follows as the required $\epsilon$-approximation can still be obtained for sufficiently large $\alpha$. 
@@
 We therefore see that the second stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the first stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
-This happens fairly often in practice, for example, when local asymptotic normality of the MLE holds.
+Within a neighborhood of the MLE, local quadratic approximations such as local asymptotic normality apply under general conditions \cite{lecam00}.
+The size of this neighborhood, measured in units of log-likelihood, grows as more data become available \cite{ning21}.
+
+%% This happens fairly often in practice, for example, when local asymptotic normality of the MLE holds.
 We conjecture that this applies to the entirety of IFAD, as IF2 converges very quickly to a neighborhood of the MLE, and will explore this in future work.
@@
 We tested IFAD against IF2 on a global search problem for the Dhaka cholera model.
-We re-implemented IF2 to do so, but we compare our results with the results of \cite{ionides15} (labeled "IF2 2015").
+We re-implemented IF2 to do so, but we also compare our results with the results of \cite{ionides15} (labeled "IF2 2015").
 Our re-implementation reaches a better log-likelihood than that of \cite{ionides15}, likely due to a better choice of algorithmic parameters.
@@
 
-To alleviate this, we employ a NUTS sampler powered by MOP-$\alpha$, with a nonparametric empirical prior constructed from density estimation of the parameter swarm from the IF2 warm start having a standard deviation of $6.3\%$ of the MLE.
+To alleviate this, we employ a No U-Turn Sampler (NUTS) \cite{homan14} powered by MOP-$\alpha$, with a nonparametric empirical prior constructed from density estimation of the parameter swarm from the IF2 warm start having a standard deviation of $6.3\%$ of the MLE.
 %\sqrt{(0.02\cdot0.95^{40})^2 \cdot 600}\approx 6.3\%
-Using a NUTS sampler powered by MOP-$\alpha$ with this prior significantly lowers the mixing time of the Markov chain to as little as $500$ iterations, as seen in Figure \ref{fig:nuts-eb}.
+Using a NUTS algorithm powered by MOP-$\alpha$ with this prior significantly lowers the mixing time of the Markov chain to as little as $500$ iterations, as seen in Figure \ref{fig:nuts-eb}.
 
@@
 We display the results for the trend parameter in the Dhaka cholera model of \cite{king08}.
-The NUTS sampler mixes quickly, and the posterior estimates from each chain share roughly the same posterior mode.}
+NUTS mixes quickly, and the posterior estimates from each chain share roughly the same posterior mode.}
     \label{fig:nuts-eb}
@@
 We thank Nicolas Chopin for helpful communications regarding the strong law of large numbers for off-parameter resampled particle filters.
-We also thank Arnaud Doucet for helpful discussions regarding the manuscript.}
+We also thank Jun Chen and Arnaud Doucet for helpful discussions regarding the manuscript.}
 \showacknow{} % Display the acknowledgments section

### 2024-07-17 paper/ms.tex
updates from overleaf

@@
 The convergence of IF2 (and so the first stage of IFAD) to a neighborhood of the MLE happens rapidly in practice. 
-In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used,  \ed{[initial convergence happens  BE MORE PRECISE] initial convergence to within xxx=40??
-log-likelihood units of the maximum} occurs within 40 iterations.
-However, even when applying much larger amounts of Monte Carlo effort, IF2 fails to get within five log-likelihood units of the maximum on this problem (Figure~\ref{fig:boxplot}).
+In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, we require no more than 40 iterations of IF2 to obtain a suitable warm-start of acceptable quality.
+However, even when applying much larger amounts of Monte Carlo effort, IF2 fails to get within seven log-likelihood units of the maximum on this problem (Figure~\ref{fig:boxplot}).
 These numerical results are consistent with theoretical results for iterated filtering, showing slow convergence close to the maximum \cite{doucet15-if}.
@@
 Details are deferred to \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}. 
-\ed{TECHNICALLY, IF WE CLAIM THAT A PROOF FOR ALPHA=1 CAN BE GENERALIZED TO ANY ALPHA, WE ARE CLAIMING A PROOF FOR ALL ALPHA, AND SO WE DO NOT NEED TO SAY WE ARE ONLY PROVING IT FOR ALPHA=1.
-IF WE REALLY ONLY CLAIM TO HAVE SHOWN IT FOR ALPHA=1, THAT SHOULD BE IN THE STATEMENT.
-IF WE CLAIM THE WHOLE RESULT, PERHAPS WE CAN POSTPONE THE DISCUSSION OF THIS 2-STAGE METHOD TO THE SUPPLEMENT?
-AT THIS POINT, ``THE REQUIRED EPSILON-APPROXIMATION'' IS NOT DEFINED FOR THE READER, SO IT IS HARD TO COMPREHEND THE ARGUMENT.}
-For simplicity, we only prove this for MOP-$1$ and note that the general case follows as the required $\epsilon$-approximation can still be obtained for sufficiently large $\alpha$. 
-When $\alpha=1$, the required $J$ is given in \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}. 
 This result shows that particle estimates for the gradient can enjoy the same linear convergence rate on well-conditioned problems as gradient descent with access to the score.

### 2024-07-19 paper/ms.tex
Updates from Overleaf

@@
 We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$.
-The existence of this density is necessary to define the likelihood function, but plug-and-play algorithms do not have access to evaluation of this density.
-Instead, they have access to a corresponding simulator, which we write as $\process_n\left(x_n\mid x_{n-1}; \theta\right)$.
-We set $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ to be the measurement density, which we suppose can be evaluated. We write $y_n^*$ for the actual values of the observations that were observed.
+The existence of this density is necessary to define the likelihood function, but plug-and-play algorithms do not explicitly evaluate this density.
+Instead, they have access to a corresponding simulator, which we, write as $\process_n\left(x_n\mid x_{n-1}; \theta\right)$.
+We set $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ to be the measurement density, which  can be directly evaluated. We write $y_n^*$ for the actual values of the observations that were observed.
 We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
@@
 Theorem~\ref{thm:mop-functional-forms} shows that MOP-$\alpha$ is constructed so this direct derivative obtains the score estimator of  \cite{poyiadjis11, scibior21} when $\alpha=1$ and that of \cite{naesseth18} when $\alpha=0$.
-Setting $\alpha<1$ adds bias but reduces variance, raising opportunities for a favorable tradeoff as illustrated in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).
+Setting $\alpha<1$ adds bias but reduces variance, producing opportunities for a favorable tradeoff as illustrated in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).
 
@@
 The argument relies on a useful decomposition of the after-resampling estimate $L_n^{A,\theta,\alpha}$. 
-Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting $\theta=\phi$ implies that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result. 
+Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting $\theta=\phi$ implies that $w_{n,j}^{P,\theta}$ evaluates to $1$. 
 This further illustrates how $\alpha$ dictates the memory of the gradient estimate. 
@@
 When differentiated, we obtain the estimators of \cite{poyiadjis11, scibior21, naesseth18} as special cases (Theorem \ref{thm:mop-functional-forms}).
-In particular, MOP-$1$ is consistent for the score (Theorem \ref{thm:mop-grad-consistency}), has rates for its bias and variance under different choices of $\alpha$ (Theorem \ref{thm:mop-biasvar}) that illustrate the desirable bias-variance tradeoff observed empirically in Figure \ref{fig:biasvar}, and enjoys a linear rate of convergence for gradient descent with the resulting gradient estimate (Theorem \ref{thm:mop-convergence}).
+In particular, MOP-$1$ is consistent for the score (Theorem \ref{thm:mop-grad-consistency}), has rates for its bias and variance under different choices of $\alpha$ that illustrate the desirable bias-variance tradeoff observed empirically in Figure \ref{fig:biasvar}  (Theorem \ref{thm:mop-biasvar}), and enjoys a linear rate of convergence for gradient descent with the resulting gradient estimate (Theorem \ref{thm:mop-convergence}).
 
@@
 
-The AD stage confers a linear convergence rate for IFAD under the usual smoothness and strong convexity assumptions in convex optimization theory, as we show below in Theorem \ref{thm:mop-convergence}.
+The AD stage confers a linear convergence rate for IFAD, up to a small region near the MLE shrinking in $J$, under the usual smoothness and strong convexity assumptions in convex optimization theory, as we show below in Theorem \ref{thm:mop-convergence}.
 
@@
 \arxiv{}{\vspace*{-5mm}}
-The proof of Theorem~\ref{thm:linear} uses results from randomized numerical linear algebra, building on Theorem~6 of\cite{mahoney16}. 
+The proof of Theorem~\ref{thm:linear} uses results from randomized numerical linear algebra, building on Theorem~6 of \cite{mahoney16}. 
 Details are deferred to \arxiv{Appendix~\ref{appendix:convergence}}{the supplementary information}. 
@@
 Previously, \cite{poyiadjis11} established that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p$ and $G'$. 
-Combining these results with Theorem \ref{thm:mop-biasvar} shows that MOP-$\alpha$ interpolates between MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$. 
-The phase transition arises as even though the particle filter has forgetting properties \cite{karjalainen23}, the resulting derivative estimate of \cite{poyiadjis11} does not, and we require both for the variance reduction.
+Combining these results with Theorem \ref{thm:mop-biasvar} shows that MOP-$\alpha$ interpolates between the variances of MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$.
+The phase transition arises because even though the particle filter has forgetting properties \cite{karjalainen23}, the resulting derivative estimate of \cite{poyiadjis11} does not, and we require both for the variance reduction.
 

### 2024-07-19 paper/ms.tex
Updates from Overleaf

@@
 IFAD runs a few iterations of IF2 to warm-start an iterative first or second-order method that uses the MOP-$\alpha$ gradient estimate.
+We write the algorithm in a fairly general manner, requiring only a Hessian estimate $H(\theta)$ with minimum eigenvalue no smaller than $c$, but in practice we use $H=I_p$ that corresponds to gradient descent.
 To see the importance of this innovation, we briefly summarize the empirical results in Section~\ref{sec:cholera}.
@@
 Meanwhile, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.6$, while the best log-likelihood found by \cite{king08} was only $-3793.4$.
-Despite being initialized for a global search, IFAD manages to get much closer to the MLE over the $100$ searches than \cite{ionides15} and finds it up to $2.5$ standard deviations of Monte Carlo error, as seen in Table~\ref{table:mle}.
+Despite being initialized for a global search, IFAD manages to get much closer to the MLE over the $100$ searches than \cite{ionides15} and finds it up to $2.5$ standard deviations of Monte Carlo log-likelihood error, as seen in Table~\ref{table:mle}.
 On this problem, the sequence of local searches, refinement, and likelihood profiling that was previously required for finding the MLE is not necessary with IFAD. 

### 2024-07-19 paper/ms.tex
Updates from Overleaf

@@
     \arxiv{}{\vspace*{-1mm}}
-    \caption{Convergence diagnostics for NUTS with the informative nonparametric empirical Bayes prior from applying KDE to preliminary searches using IF2, with $4$ chains.
-We display the results for the trend parameter in the Dhaka cholera model of \cite{king08}.
+    \caption{Convergence diagnostics for NUTS when estimating $\beta_{\text{trend}}$ with the informative nonparametric empirical Bayes prior from applying KDE to preliminary searches using IF2, with $4$ chains.
 NUTS mixes quickly, and the posterior estimates from each chain share roughly the same posterior mode.}
@@
 While the result is presented here as specific to MOP-$\alpha$, we prove a more general result in \arxiv{Appendix~\ref{appendix:targeting}}{the supplementary material}.
-This is a strong law of large numbers for triangular arrays of particles resampled according to an arbitrary resampling rule not necessarily in proportion to the targeted distribution of interest, to weights that encode the cumulative discrepancy between the resampling and the target distribution.
+This is a strong law of large numbers for triangular arrays of particles resampled arbitrarily, with resampling probabilities not proportion to the targeted distribution of interest, to weights that correct for the cumulative discrepancy between the resampling and the target distribution.
 

### 2024-07-26 paper/ms.tex
Updates from Overleaf

@@
 \arxiv{}{\vspace*{-2mm}}
-\begin{enumerate}[label=(A\arabic*),itemsep=-1.2ex] 
+\begin{enumerate}[label=(A\arabic*),itemsep=-1.4ex, , font=\small] 
     \item \textbf{Continuity of the Likelihood.} $\ell(\theta)$ has more than two continuous derivatives in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.
@@
 While the result is presented here as specific to MOP-$\alpha$, we prove a more general result in \arxiv{Appendix~\ref{appendix:targeting}}{the supplementary material}.
-This is a strong law of large numbers for triangular arrays of particles resampled arbitrarily, with resampling probabilities not proportion to the targeted distribution of interest, to weights that correct for the cumulative discrepancy between the resampling and the target distribution.
+This is a strong law of large numbers for triangular arrays of particles resampled arbitrarily, with resampling probabilities not proportional to the targeted distribution of interest, reweighted to correct for the cumulative discrepancy between the resampling and the target distribution.
 
@@
 \arxiv{}{\vspace*{-2mm}}
-\subsection{MOP-$1$ Is Consistent for the Score}
+\subsection{MOP-$1$ Gradient Is Consistent for the Score}
 
@@
     \label{thm:mop-biasvar}
-    When $\alpha\in (0,1)$ and $\theta=\phi$, define $\psi(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)/(1-\alpha)$. 
+    When $\alpha\in (0,1)$ and $\theta=\phi$, define $\psi_k(\alpha)=(\alpha^k  + \alpha^{k+1} - \alpha)/(1-\alpha)$. 
     There exists an $\epsilon>0$ depending on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ as in \cite{karjalainen23} such that the MSE and variance of MOP-$\alpha$ are:
@@
         \E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)\big\|_2^2 
-        &\lesssim& \min_{k \leq N} Np \, G'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi(\alpha)\right), \label{eq:mop-mse}
+        &\lesssim& \min_{k \leq N} Np \, G'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi_k(\alpha)\right), \label{eq:mop-mse}
         \\
@@
         &\E\big\|\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)\big\|_2^2 \label{eq:mop-mse}\\
-        &\lesssim \min_{k \leq N} Np \, G'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi(\alpha)\right), \nonumber 
+        &\lesssim \min_{k \leq N} Np \, G'(\theta)^2\left(k^2J^{-1}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\psi_k(\alpha)\right), \nonumber 
         \\
@@
 The MSE bound considers the error between the score and MOP-$(1,k)$, and the error between MOP-$(1,k)$ and MOP-$\alpha$. 
-The latter can be shown to be $\tilde{O}\big(Np \, G'(\theta)^2(k+\psi(\alpha))\big)$, and the former can be controlled with a result on the the forgetting of the particle filter from \cite{karjalainen23} and the very same $L^p$ error bound mentioned above.
+The latter can be shown to be $\tilde{O}\big(Np \, G'(\theta)^2(k+\psi_k(\alpha))\big)$, and the former can be controlled with a result on the the forgetting of the particle filter from \cite{karjalainen23} and the very same $L^p$ error bound mentioned above.
 
@@
 
-In contrast, as $Nk^2 \leq N^4$ and $\alpha, k$ can be as large as desired to balance the impact of the last three terms, MOP-$\alpha$ achieves a lower MSE than MOP-$1$ does.
+In contrast, as $Nk^2 \leq N^4$ and $\alpha, k$ can be as large as desired to balance the impact of the last three terms of Equation \ref{eq:mop-mse}, MOP-$\alpha$ achieves a lower MSE than MOP-$1$ does.
 We also show that MOP-$0$ achieves a MSE of $\tilde{O}\big(NpG'(\theta)^2(J^{-1}+(1-\epsilon)^{\lfloor1/c\log(J)\rfloor})\big)$, where the second term corresponds to uncontrolled mixing error.
@@
 
+Practical likelihood-based data analysis involves many likelihood optimizations \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
+In particular, profile likelihood confidence intervals require a sequence of optimizations.
+Additionally, a careful scientist should consider many model variations, to see whether the conclusions of the study are sensitive to alternative model specifications.
+Coding model variations is relatively simple when using plug-and-play inference methodology.
+However, assessing the scientific potential of these variations requires likelihood optimization.
+If many rounds of laborious searching are required to attain the maximum (as arises in previous methods) that has practical consequences for the rate at which scientists can evaluate hypotheses.
+We have demonstrated, for the first time, a plug-and-play maximum likelihood approach which can directly maximize the likelihood for the complex benchmark model of \cite{king08}, requiring some Monte Carlo replication but not prolonged sequences of explorations.
+We anticipate that this will promote scientific advances in all domains where complex POMP models arise.
+
+\arxiv{}{\vspace*{-2mm}}
+\subsection{Further Work}
+
 If the simulator is discontinuous in $\theta$, MOP-$\alpha$ no longer applies.
@@
 The analysis is similar, with comparable rates but with a slightly less convenient implementation.
-
-Practical likelihood-based data analysis involves many likelihood optimizations \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
-In particular, profile likelihood confidence intervals require a sequence of optimizations.
-Additionally, a careful scientist should consider many model variations, to see whether the conclusions of the study are sensitive to alternative model specifications.
-Coding model variations is relatively simple when using plug-and-play inference methodology.
-However, assessing the scientific potential of these variations requires likelihood optimization.
-If many rounds of laborious searching are required to attain the maximum (as arises in previous methods) that has practical consequences for the rate at which scientists can evaluate hypotheses.
-We have demonstrated, for the first time, a plug-and-play maximum likelihood approach which can directly maximize the likelihood for the complex benchmark model of \cite{king08}, requiring some Monte Carlo replication but not prolonged sequences of explorations.
-We anticipate that this will promote scientific advances in all domains where complex POMP models arise.
 

### 2024-08-12 paper/appendix/coverletter.tex
push from overleaf

@@
 Our method promises substantially improved inference wherever simulation-based inference is used for the full-information likelihood in complex partially observed Markov processes.
-That is why we consider publication of our article in PNAS would facilitate a broad range of scientific advances.
+That is why we believe that publication of our article in PNAS would facilitate a broad range of scientific advances.
 
-Although our algorithm can be implemented by a programmer familiar with current machine learning tools, the technology behind our contribution is rather technical.
+Although our algorithm can be implemented by a competent programmer familiar with current machine learning tools, the theory and motivation behind our contribution is rather technical.
 The justification for our new algorithm requires deep familiarity with the theory of sequential Monte Carlo (SMC) and some experience with the theory and practice of automatic differentiation (AD).

### 2024-08-12 paper/ms.tex
push from overleaf

@@
 
-\title{Accelerated Inference for Partially Observed Markov Processes using Automatic Differentiation}
+\title{Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes}
 
@@
 Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques because widely used particle filter algorithms yield an estimated likelihood function that is discontinuous as a function of the model parameters.
-We embed two existing AD particle filter methods in a theoretical framework that provides an extension to a new class of algorithms.
-This new class permits a bias/variance tradeoff and hence a mean squared error substantially lower than the existing algorithms.
-We develop likelihood maximization algorithms suited to the Monte Carlo properties of the AD gradient estimate.
+To resolve this, we create a theoretical framework that embeds two existing AD particle filter methods within a new class of algorithms.
+This new class permits a bias-variance tradeoff and hence a mean squared error substantially lower than the existing algorithms.
+This allows us to develop likelihood maximization algorithms suited to the Monte Carlo properties of the AD gradient estimate.
 Our algorithms require only a differentiable simulator for the latent dynamic system; by contrast, most previous approaches to AD likelihood maximization for particle filters require access to the system's transition probabilities.
@@
 Commonly, the latent system is modeled as a Markov process, giving rise to a partially observed Markov process (POMP) model, also known as a hidden Markov models or a state space model.
-POMP models arise in fields as diverse as automated control \cite{singh22}, epidemiology \cite{he10, stocks17}, ecology \cite{knape12} and finance \cite{kim08, breto14}.
+POMP models arise in fields as diverse as optimal control \cite{singh22}, epidemiology \cite{he10, stocks17}, ecology \cite{knape12} and finance \cite{kim08, breto14}.
 Despite their ubiquity, estimation and inference within this broad class of models remains a challenging problem.
-This article uses automatic differentiation (AD) to construct improved algorithms for inference on complex POMP models.
+To tackle this, we use automatic differentiation (AD) to construct improved algorithms for inference on complex POMP models.
 
@@
 It provides an unbiased estimate of the likelihood function  \cite{delMoral04}, enabling Bayesian inference \cite{andrieu10,chopin13} and likelihood-based inference \cite{ionides06-pnas,ionides15}.
-Likelihood-based inference is statistically efficient \cite{pawitan01} and is robust to a moderate amount of Monte Carlo error \cite{ionides17,ning21}.
-An attractive feature of basic particle filter algorithms, also known as boostrap filters, is that they do not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
-This \textit{plug-and-play property} is useful for scientific applications \cite{he10}.
+This is desirable, as likelihood-based inference is statistically efficient \cite{pawitan01} and robust to a moderate amount of Monte Carlo error \cite{ionides17,ning21}.
+Furthermore, basic particle filter algorithms, also known as boostrap filters, do not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
+This \textit{plug-and-play} property is useful for scientific applications \cite{he10}.
 Some plug-and-play methods depend on the construction of low-dimensional summary statistics \cite{wood10,toni09}, sacrificing statistical efficiency for computational convenience.
@@
 However, iterated filtering algorithms have shown that this is practical in various scientific investigations, such as \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
-Nevertheless, the Monte Carlo variability arising in iterated filtering applications becomes increasingly problematic as the size of the data and the complexity of the model increases.
-Algorithmic advances are needed to make plug-and-play likelihood maximization for POMP models more numerically efficient.
-We develop plug-and-play AD methods which provide a large computational advantage over the methods used in these previous applications.
+Nevertheless, the Monte Carlo variability arising in iterated filtering applications becomes increasingly problematic as the size of the data and the complexity of the model increase.
+Algorithmic advances are needed to make plug-and-play likelihood maximization for POMPs more numerically efficient.
 
+We examine plug-and-play AD as a tool to do so by providing a likelihood gradient.
 Recent advances in AD for particle filters \cite{naesseth18, jonschkowski18, corenflos21, scibior21, singh22} have drawn attention to AD as a tool for inference in POMP models.
@@
 Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a {\it Measurement Off-Parameter} (MOP) particle filter.
-In short, \textit{instead of directly differentiating through a basic particle filter, MOP uses AD to differentiate through a smooth particle filter constructed using measurement density ratios and a differentiable simulator}.
+In short, \textit{instead of directly differentiating through a basic particle filter, MOP uses AD to differentiate through a series of measurement density ratios and a differentiable simulator}. 
 Our MOP algorithm differs from previous smooth particle filters \cite{svensson18,malik11} by possessing the plug-and-play property.
@@
 This discount factor enables MOP-$\alpha$ to interpolate between the  biased gradient estimator of \cite{naesseth18}, when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21}, when $\alpha=1$ (Theorem \ref{thm:mop-functional-forms}).
-The bias-variance tradeoff induced by $\alpha$ suggests, both in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}), the use of $\alpha$ values strictly between 0 and 1.
+The bias-variance tradeoff induced by $\alpha$ suggests, both in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}), a choice of $\alpha$ strictly in $(0,1)$.
 
-One can simply drop terms arising from resampling when differentiating through the particle filter \cite{naesseth18}, but this leads to asymptotic bias \cite{corenflos21}.
-MOP-$\alpha$ is constructed as a smooth particle filter, thus bypassing the seemingly incompatible paradox of differentiating through discrete Monte Carlo sampling by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
-Yet, MOP-$\alpha$ can nevertheless provide strongly consistent estimates of the filter distribution and the log-likelihood (Theorem \ref{thm:mop-targeting}).
+% One can simply drop terms arising from resampling when differentiating through the particle filter \cite{naesseth18}, but this leads to asymptotic bias \cite{corenflos21}.
+% Instead, MOP-$\alpha$ bypasses the seemingly incompatible paradox of differentiating through discrete Monte Carlo sampling by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
+MOP-$\alpha$ provides strongly consistent estimates of the filtering distribution and the log-likelihood (Theorem \ref{thm:mop-targeting}) when $\alpha=1$ or $\theta=\phi$.
 When $\alpha=1$, it also provides consistent score estimates (Theorem \ref{thm:mop-grad-consistency}).
+This enables us to derive a linear convergence rate for gradient descent using MOP-$\alpha$ in the  presence of strong convexity and smoothness (Theorem \ref{thm:mop-convergence}).
 
-We derive a linear convergence rate for gradient descent using MOP-$\alpha$ in the  presence of strong convexity and smoothness (Theorem \ref{thm:mop-convergence}).
-Critically, the estimator converges quickly within a neighborhood of the maximum but struggles to reach this neighborhood in the presence of local minima and saddle points.
-That behavior is complementary to that of iterated filtering algorithms \cite{ionides06-pnas,ionides15} which provides a relatively fast and stable way to identify this neighborhood.
+In practice, gradient descent with MOP-$\alpha$ converges quickly within a neighborhood of the maximum but struggles to reach this neighborhood in the presence of local minima and saddle points.
+That behavior is complementary to that of iterated filtering algorithms \cite{ionides06-pnas,ionides15} which provide a relatively fast and stable way to identify this neighborhood.
 We therefore build a simple but effective hybrid algorithm called {\it Iterated Filtering with Automatic Differentiation} (IFAD) that warm-starts first-order or second-order gradient methods with a preliminary solution obtained from a few rounds of iterated filtering.
@@
 
-These improvements also extend to Bayesian inference, as we show in Section \ref{sec:bayes} that we can use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with a nonparametric empirical Bayes-style prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{andrieu10} on the Dhaka cholera model of \cite{king08} from the $700,000$ iterations in \cite{fasiolo16} to just $500$.
+These improvements also extend to Bayesian inference.
+In Section \ref{sec:bayes}, we use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with an empirical Bayes-type prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{andrieu10} on the Dhaka cholera model of \cite{king08} from the $700,000$ iterations in \cite{fasiolo16} to just $500$.
 To the best of our knowledge, attaining rapid mixing on this challenging problem has not previously been achieved.
@@
 
-We suppose that the discete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$.
+We suppose that the discrete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$.
 The existence of this density is necessary to define the likelihood function, but plug-and-play algorithms do not explicitly evaluate this density.
@@
 We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
-Superscripts $x_{n,j}^A$ of particles denote the ancestral trajectory of particle $j$ at time $n$,  and $k_j$
-denotes the resample indices drawn for each particle $j$. 
+$x_{n,j}^A$ denotes the ancestral trajectory of particle $j$ at time $n$,  and $k_j$
+denotes the resampling indices for particle $j$. 
 %$a(\cdot)$ is the ancestor function that maps a particle to its parent, $j \mapsto a(j)$,
 
-The above densities are defined on a probability space $(\Omega, \Sigma, \prob)$ which is also assumed to enable construction of independent replicates and all random variables defined in our algorithms.
+The above densities are defined on a probability space $(\Omega, \Sigma, \prob)$ which is assumed to enable construction of independent replicates and all random variables defined in our algorithms.
 For the algorithmic interpretation of our theory, we identify the random number seed with an element of the sample space $\omega \in \Omega$.
-The random seed determines the sequence of pseudo-random numbers as generated by a computer, just as the outcome $\omega\in\Omega$ generates the sequence of random variables.
+The random seed determines the sequence of pseudo-random numbers generated by a computer, just as $\omega\in\Omega$ generates the sequence of random variables.
 
@@
 The simulator can be differentiated using AD as long as the underlying computer code for the simulator is differentiable, and so MOP-$\alpha$ is applicable to a broad class of POMP models.
+To that end, we suppose that the simulator is a differentiable function of $\theta$ for every fixed $\omega$, a condition that requires the latent process to be a continuous random variable. 
+This condition is also known as the {\it reparameterization trick} \cite{corenflos21}.
+Supposing also that the measurement model is differentiable in $\theta$, direct differentiation of MOP-$\alpha$ is available via AD.
 
-The MOP-$\alpha$ algorithm resamples the particles according to an arbitrary resampling rule that depends on both the target parameter value and a baseline parameter value, yielding the name {\it measurement off-parameter}.
+MOP-$\alpha$ resamples the particles according to an arbitrary resampling rule that depends on both the target parameter value and a baseline parameter value, yielding the name {\it measurement off-parameter}.
 This is intended as an analogy to {\it off-policy} learning in reinforcement learning.
@@
 This ensures that the resampling indices are invariant in $\theta$ when $\omega$ and $\phi$ are fixed, bypassing the issue of Monte Carlo resampling.
-We show in Theorem~\ref{thm:mop-targeting} that MOP-$\alpha$ exactly targets the filtering distribution when $\theta=\phi$ or $\alpha=1$.
 
-We suppose that the simulator is a differentiable function of $\theta$ for every fixed $\omega$, a condition that requires the latent process to be a continuous random variable. 
-This condition is also known as the {\it reparameterization trick} \cite{corenflos21}.
-Supposing also that the measurement model is differentiable in $\theta$, direct differentiation of MOP-$\alpha$ is available via AD.
-Theorem~\ref{thm:mop-functional-forms} shows that MOP-$\alpha$ is constructed so this direct derivative obtains the score estimator of  \cite{poyiadjis11, scibior21} when $\alpha=1$ and that of \cite{naesseth18} when $\alpha=0$.
-Setting $\alpha<1$ adds bias but reduces variance, producing opportunities for a favorable tradeoff as illustrated in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).
+% Theorem~\ref{thm:mop-functional-forms} shows that MOP-$\alpha$ is constructed so this direct derivative obtains the score estimator of  \cite{poyiadjis11, scibior21} when $\alpha=1$ and that of \cite{naesseth18} when $\alpha=0$.
+% Setting $\alpha<1$ adds bias but reduces variance, producing opportunities for a favorable tradeoff as illustrated in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).
 
-\begin{algorithm}[H]
+\begin{algorithm}[htbp!]
 	\caption{MOP-$\alpha$}
@@
 
-Algorithm~\ref{alg:mop} constructs two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but with the resampling indices constructed from the first pass, with the baseline parameter, $\phi$.
-%% As the process model follows $\theta$ but the measurement model follows the baseline parameter $\phi$, we therefore call this a ``measurement off-parameter'' particle filter.
-%% DISCUSSED IN THE PREVIOUS SECTION
-The resampling indices, which we write as $k_j \sim \text{Categorical}\big(g^{\phi}_{n,1},...,g^{\phi}_{n,J}\big)$, are a function of $\phi$ for any value of $\theta$.
+Algorithm~\ref{alg:mop} constructs two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but with the resampling indices constructed from the first pass.
+The resampling indices, which we write as $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$, are a function of $\phi$ for any value of $\theta$.
 For the categorical distribution, we use systematic resampling \cite{arulampalam02,king16} which usually has superior performance to multinomial resampling.
-If $\theta$ and $\phi$ coincide, one only needs one particle filter run at $\theta=\phi$, otherwise one needs two runs at the same seed $\omega \in \Omega$.
+In practice, we only ever evaluate MOP-$\alpha$ in the case where $\theta$ and $\phi$ coincide.
+We then only need to run one particle filter at $\theta=\phi$, otherwise one needs two runs at the same seed $\omega \in \Omega$.
 
-Algorithm~\ref{alg:mop} reweights the conditional likelihoods by a correction factor accumulated over time to account for the resampling under $\phi$. 
+In Algorithm~\ref{alg:mop}, the conditional likelihoods are reweighted by a correction factor accumulated over time to account for the resampling under $\phi$. 
 Writing $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density and $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$, we obtain two estimates of the conditional likelihood under $\theta$,
@@
 where we recursively correct the weights for each particle for the cumulative error incurred by the off-parameter resampling:
-\arxiv{}{\vspace*{-1.5mm}}
+\arxiv{}{\vspace*{-2mm}}
 \begin{equation}
@@
 
-When $\alpha=1$, MOP-$1$ maintains complete memory of each particle's ancestral trajectory, and Theorem~\ref{thm:mop-functional-forms} shows that it recovers the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}.
-At the other extreme, when $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{naesseth18}. 
-The novel possibility of $0<\alpha<1$ is effective both in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).
+When $\alpha=1$, MOP-$1$ maintains complete memory of each particle's ancestral trajectory.
+Theorem~\ref{thm:mop-functional-forms} shows that it recovers the consistent but high-variance gradient estimate of \cite{poyiadjis11, scibior21}.
+When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{naesseth18}. 
+The novel possibility of $0<\alpha<1$ effectively balances a bias-variance tradeoff in both theory (Theorem \ref{thm:mop-biasvar}) and practice (Figure \ref{fig:biasvar}).
 
@@
 \cite{scibior21} show that the estimate of \cite{naesseth18} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through the use of a stop-gradient operation in the AD procedure.
-It turns out that both of these, when applied on the bootstrap filter, correspond to special cases of MOP-$\alpha$.
+Both of these, when applied to the bootstrap filter, correspond to special cases of MOP-$\alpha$.
 
@@
 Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting $\theta=\phi$ implies that $w_{n,j}^{P,\theta}$ evaluates to $1$. 
-This further illustrates how $\alpha$ dictates the memory of the gradient estimate. 
-As \cite{scibior21} remarks, the MOP-$0$ estimator of \cite{naesseth18} only considers single-step quantities, and is ``memoryless'' beyond a single step. 
-In contrast, the MOP-$1$ estimate of \cite{poyiadjis11, scibior21} only considers the surviving particles at time $N$ and so fully tracks dependencies over time.
+% This further illustrates how $\alpha$ dictates the memory of the gradient estimate. 
+% As \cite{scibior21} remarks, the MOP-$0$ estimator of \cite{naesseth18} only considers single-step quantities, and is ``memoryless'' beyond a single step. 
+% In contrast, the MOP-$1$ estimate of \cite{poyiadjis11, scibior21} only considers the surviving particles at time $N$ and so fully tracks dependencies over time.
 
@@
 
-\arxiv{}{\vspace*{-2mm}}
-\subsection{Summary of Theoretical Guarantees}
+% \arxiv{}{\vspace*{-2mm}}
+% \subsection{Summary of Theoretical Guarantees}
 
-The construction of MOP-$\alpha$ bypasses the issue of differentiating through a Monte Carlo algorithm with discontinuous resampling by turning it into a problem of differentiating through a simulator and a series of measurement density ratios.
-We defer the theoretical analysis of MOP-$\alpha$ to Section \ref{sec:thms}, but first we highlight a few key results.
-MOP-$\alpha$ estimates the likelihood and conditional distributions of latent variables, as one expects of a particle filter (Theorem \ref{thm:mop-targeting}).
-When differentiated, we obtain the estimators of \cite{poyiadjis11, scibior21, naesseth18} as special cases (Theorem \ref{thm:mop-functional-forms}).
-In particular, MOP-$1$ is consistent for the score (Theorem \ref{thm:mop-grad-consistency}), has rates for its bias and variance under different choices of $\alpha$ that illustrate the desirable bias-variance tradeoff observed empirically in Figure \ref{fig:biasvar}  (Theorem \ref{thm:mop-biasvar}), and enjoys a linear rate of convergence for gradient descent with the resulting gradient estimate (Theorem \ref{thm:mop-convergence}).
+% The construction of MOP-$\alpha$ bypasses the issue of differentiating through a Monte Carlo algorithm with discontinuous resampling by turning it into a problem of differentiating through a simulator and a series of measurement density ratios.
+% We defer the theoretical analysis of MOP-$\alpha$ to Section \ref{sec:thms}, but first we highlight a few key results.
+% MOP-$\alpha$ estimates the likelihood and conditional distributions of latent variables, as one expects of a particle filter (Theorem \ref{thm:mop-targeting}).
+% When differentiated, we obtain the estimators of \cite{poyiadjis11, scibior21, naesseth18} as special cases (Theorem \ref{thm:mop-functional-forms}).
+% In particular, MOP-$1$ is consistent for the score (Theorem \ref{thm:mop-grad-consistency}), has rates for its bias and variance under different choices of $\alpha$ that illustrate the desirable bias-variance tradeoff observed empirically in Figure \ref{fig:biasvar}  (Theorem \ref{thm:mop-biasvar}), and enjoys a linear rate of convergence for gradient descent with the resulting gradient estimate (Theorem \ref{thm:mop-convergence}).
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-6mm}}
 \section{Practical Maximum-Likelihood Estimation}
@@
 IFAD runs a few iterations of IF2 to warm-start an iterative first or second-order method that uses the MOP-$\alpha$ gradient estimate.
-We write the algorithm in a fairly general manner, requiring only a Hessian estimate $H(\theta)$ with minimum eigenvalue no smaller than $c$, but in practice we use $H=I_p$ that corresponds to gradient descent.
+We write the algorithm in a fairly general manner, requiring only a Hessian estimate $H(\theta)$ with minimum eigenvalue no smaller than $c$, but in our experiments we use $H=I_p$, corresponding to gradient descent.
 To see the importance of this innovation, we briefly summarize the empirical results in Section~\ref{sec:cholera}.
@@
 
-Consider the second stage of IFAD (Algorithm \ref{alg:ifad}) where one stops if $\|\nabla_\theta \hat\ell^\alpha(\theta_m)\| \leq (1+\sigma) \epsilon$, where $\sigma \geq \frac{4 \Gamma}{(1-\beta)}$, for some $\beta \in (0,1)$.
+Consider the second stage of IFAD (Algorithm \ref{alg:ifad}), stopping if $\|\nabla_\theta \hat\ell^\alpha(\theta_m)\| \leq (1+\sigma) \epsilon$, where $\sigma \geq \frac{4 \Gamma}{(1-\beta)}$ for some $\beta \in (0,1)$.
 Assume $-\ell$ is strongly convex and smooth, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$.
 Choose the learning rate $\eta$ such that $\eta \leq \frac{c(1-\beta)}{2\Gamma}$.
-Then, for sufficiently large $\alpha$ and $J$ to ensure the score estimate is an $\epsilon$-approximation and the minimum eigenvalue of $H$ is greater than some $c > 0$ for all $m$ with probability at least $1-\delta$, the second stage of IFAD converges linearly to the MLE.
-
+Then, for sufficiently large $\alpha$ and $J$ so $\|\nabla_\theta \hat\ell^\alpha(\theta_m) - \nabla_\theta\ell^\alpha(\theta_m)\|$ and $H(\theta_m) \succ cI_p \succ 0$ for all $m$ with probability at least $1-\delta$,
 \arxiv{}{\vspace*{-2mm}}
@@
 \end{thm}
-\arxiv{}{\vspace*{-5mm}}
+\arxiv{}{\vspace*{-7mm}}
 The proof of Theorem~\ref{thm:linear} uses results from randomized numerical linear algebra, building on Theorem~6 of \cite{mahoney16}. 
@@
 
-We therefore see that the second stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is $\gamma$-strongly convex in a neighborhood of the MLE and (2) the first stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
-Within a neighborhood of the MLE, local quadratic approximations such as local asymptotic normality apply under general conditions \cite{lecam00}.
-The size of this neighborhood, measured in units of log-likelihood, grows as more data become available \cite{ning21}.
+We therefore see that the second stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is well-conditioned in a neighborhood of the MLE and (2) the first stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
+Within a neighborhood of the MLE that grows as more data become available \cite{ning21}, local quadratic approximations such as local asymptotic normality apply under general conditions \cite{lecam00}.
+This suggests that the linear convergence may occur reasonably often in practice.
 
@@
 where the coefficients $(\beta_j)_{j=1}^6$ model seasonality in the force of infection, $\beta_{\text{trend}}$ models the trend in the force of infection, and the $\omega_j$ represent seasonality of a non-human environmental reservoir of disease.
-The measurement model for observed monthly cholera deaths is given by 
+The measurement model for observed monthly cholera deaths is
     $Y_n \sim \gN(M_n, \tau^2M_n^2)$,
-where $M_n=\gamma\int_{t_{n-1}}^{t_n}I(s)\, ds$ is the modeled number of cholera deaths in that month.
+where $M_n=\gamma\int_{t_{n-1}}^{t_n}I(s)\, ds$ is the modeled cholera deaths in that month.
 
@@
 We re-implemented IF2 to do so, but we also compare our results with the results of \cite{ionides15} (labeled "IF2 2015").
-Our re-implementation reaches a better log-likelihood than that of \cite{ionides15}, likely due to a better choice of algorithmic parameters.
-For each method, we performed 100 searches, initialized with 100 initial starting parameter vectors drawn uniformly from the same wide bounding box used in \cite{ionides15}.
-We summarize our findings below.
+Our re-implementation reaches a better log-likelihood than that of \cite{ionides15} due to a better choice of algorithmic parameters.
+For each method, we performed 100 searches, initialized with 100 starting parameter vectors drawn uniformly from the same wide bounding box used in \cite{ionides15}. We summarize our findings below.
 
@@
 \arxiv{}{\vspace*{-1mm}}
-\begin{table}[h!]
+\begin{table}[htbp!]
 \centering
 \input{imgs/095/table}
-\caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone.
-IFAD performs the best among all methods.
-Our implementation of IF2 outperforms that of \cite{ionides15}, but still ultimately underperforms IFAD.
-}
+\caption{Maximum log-likelihood found by IF2, IFAD, and MOP alone. IFAD performs the best among all methods.}
 \arxiv{}{\vspace*{-7mm}}
@@
 This MLE was obtained with much computational effort, using many global and local IF2 searches and with the assistance of likelihood profiling, 8 years after the model was first proposed by \cite{king08}.
-Meanwhile, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.6$, while the best log-likelihood found by \cite{king08} was only $-3793.4$.
+Meanwhile, \cite{ionides15} only achieve a maximum log-likelihood of $-3768.6$, and \cite{king08} only $-3793.4$.
 Despite being initialized for a global search, IFAD manages to get much closer to the MLE over the $100$ searches than \cite{ionides15} and finds it up to $2.5$ standard deviations of Monte Carlo log-likelihood error, as seen in Table~\ref{table:mle}.
 On this problem, the sequence of local searches, refinement, and likelihood profiling that was previously required for finding the MLE is not necessary with IFAD. 
-In other words, the improvement in numerical efficiency of IFAD over IF2 is so large that routine application of IFAD (consisting of a collection of Monte Carlo replications from random starting points) generates results outside the reach of a routine application of IF2.
-When interpreting Table~\ref{table:mle}, bear in mind that by Wilks' Theorem, any difference of over $1.92$ log units has statistical relevance when testing one parameter at the $0.05$ significance level, and therefore may have scientific value.
+% In other words, the improvement in numerical efficiency of IFAD over IF2 is so large that routine application of IFAD (consisting of a collection of Monte Carlo replications from random starting points) generates results outside the reach of a routine application of IF2.
+When interpreting Table~\ref{table:mle}, note that by Wilks' Theorem, any difference of over $1.92$ log units has statistical relevance when testing one parameter at the $0.05$ significance level, and therefore may have scientific value.
 
-\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations, performing IF2 alone ultimately fails to achieve the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter}).
-Conversely, when we tried gradient descent with MOP alone, we encountered many failed searches.
-This is a difficult, nonconvex, and noisy problem with $18$ parameters, and the search gets stuck in local minima and saddle points, failing to approach the MLE.
+\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations, IF2 alone ultimately fails to achieve the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter}).
+Conversely, we encountered many failed searches with gradient descent alone.
+This is a difficult, nonconvex, and noisy problem with $18$ parameters, and gradient descent without a warm-start gets stuck in local minima and saddle points.
 
@@
 %\sqrt{(0.02\cdot0.95^{40})^2 \cdot 600}\approx 6.3\%
-Using a NUTS algorithm powered by MOP-$\alpha$ with this prior significantly lowers the mixing time of the Markov chain to as little as $500$ iterations, as seen in Figure \ref{fig:nuts-eb}.
+Using a NUTS algorithm powered by MOP-$\alpha$ with this prior significantly lowers the mixing time of the Markov chain to as little as $500$ iterations (Figure \ref{fig:nuts-eb}).
 
@@
     \arxiv{}{\vspace*{-1mm}}
-    \caption{Convergence diagnostics for NUTS when estimating $\beta_{\text{trend}}$ with the informative nonparametric empirical Bayes prior from applying KDE to preliminary searches using IF2, with $4$ chains.
+    \caption{Histogram and density estimate of the posterior of $\beta_{\text{trend}}$, as well as convergence diagnostics for NUTS with $4$ chains.
 NUTS mixes quickly, and the posterior estimates from each chain share roughly the same posterior mode.}
@@
 \arxiv{}{\vspace*{-2mm}}
-\begin{enumerate}[label=(A\arabic*),itemsep=-1.4ex, , font=\small] 
-    \item \textbf{Continuity of the Likelihood.} $\ell(\theta)$ has more than two continuous derivatives in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.
-\label{assump:conti-lik}
-    \item \textbf{Bounded Process Model.} There exist $\underbar{M}, \bar{M}$ such that $0 < \underbar{M} \leq f_{X_n|X_{n-1}}(x_n | x_{n-1};\theta) \leq \bar{M} < \infty$.
-\label{assump:bounded-process}
-    \item \textbf{Bounded Measurement Model.} There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G}<\infty$ and there exists $G'(\theta)$ with $\|\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)\|_\infty \leq G'(\theta)< \infty$.
-\label{assump:bounded-measurement}
-    \item \textbf{Bounded Gradient Estimates.} There are functions $G(\theta), H(\theta): \Theta \to [0,\infty)$ uniformly bounded by $G^*, H^*<\infty$, so the MOP-$\alpha$ gradient and Hessian estimates at $\theta=\phi$ are almost surely bounded by $G(\theta)$ and $H(\theta)$ for all $\alpha$.
-\label{assump:local-bounded-derivative}
-    \item \textbf{Differentiability of Density Ratios and Simulator.} The measurement density, \arxiv{\\}{}$f_{Y_n|X_n}(y_n^*|x_n; \theta)$, and simulator have more than two continuous derivatives in $\theta$.
-\label{assump:diff-meas-and-sim}
+\begin{enumerate}[label=(A\arabic*),itemsep=-1ex, , font=\small] 
+    \item \textbf{Continuity of the Likelihood.} $\ell(\theta)$ has more than two continuous derivatives in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.\label{assump:conti-lik}
+    \item \textbf{Bounded Process Model.} There exist $\underbar{M}, \bar{M}$ such that $0 < \underbar{M} \leq f_{X_n|X_{n-1}}(x_n | x_{n-1};\theta) \leq \bar{M} < \infty$.\label{assump:bounded-process}
+    \item \textbf{Bounded Measurement Model.} There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G}<\infty$, and $G'(\theta)$ with $\|\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)\|_\infty \leq G'(\theta)< \infty$.\label{assump:bounded-measurement}
+    \item \textbf{Bounded Gradient Estimates.} There are functions $G(\theta), H(\theta): \Theta \to [0,\infty)$ uniformly bounded by $G^*, H^*<\infty$, so the MOP-$\alpha$ gradient and Hessian estimates at $\theta=\phi$ are almost surely bounded by $G(\theta)$ and $H(\theta)$ for all $\alpha$.\label{assump:local-bounded-derivative}
+    \item \textbf{Differentiable Density Ratios and Simulator.} The measurement density, \arxiv{\\}{}$f_{Y_n|X_n}(y_n^*|x_n; \theta)$, and simulator have more than two continuous derivatives in $\theta$.\label{assump:diff-meas-and-sim}
 \end{enumerate}
-\arxiv{}{\vspace*{-4mm}}
+\arxiv{}{\vspace*{-2mm}}
 
+In particular, assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} imply that the POMP model satisfies a strong mixing condition, and as such are fairly strong but commonplace in the literature \cite{karjalainen23, delMoral04}.
 The fact that the likelihood estimate yielded by MOP-$\alpha$ has more than two continuous derivatives in $\theta$ follows from the construction of the likelihood estimate in equations \ref{eq:mop-conditional-likelihood} and \ref{eq:weighting-scheme}, as well as Assumptions \ref{assump:bounded-measurement} and \ref{assump:diff-meas-and-sim}.
@@
 While the result is presented here as specific to MOP-$\alpha$, we prove a more general result in \arxiv{Appendix~\ref{appendix:targeting}}{the supplementary material}.
-This is a strong law of large numbers for triangular arrays of particles resampled arbitrarily, with resampling probabilities not proportional to the targeted distribution of interest, reweighted to correct for the cumulative discrepancy between the resampling and the target distribution.
+This is a strong law of large numbers for triangular arrays of particles resampled arbitrarily, with resampling probabilities not proportional to the targeted distribution of interest, but reweighted to correct for the cumulative discrepancy between the resampling and the target distribution.
 
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-4mm}}
 \subsection{MOP-$1$ Gradient Is Consistent for the Score}
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-4mm}}
 \subsection{MOP-$\alpha$ Error, Bias and Variance}
@@
 
+\arxiv{}{\vspace*{-1mm}}
 \paragraph{Interpretation:} Theorem \ref{thm:mop-biasvar} provides theoretical understanding of the results observed in Figure \ref{fig:biasvar} that show the empirically favorable bias-variance tradeoff enjoyed by MOP-$\alpha$.
@@
 
+\arxiv{}{\vspace*{-1mm}}
 \paragraph{Bias-Variance Tradeoff:} We show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( Np \, G'(\theta)^2\big/J \big)$.
@@
 
-In contrast, as $Nk^2 \leq N^4$ and $\alpha, k$ can be as large as desired to balance the impact of the last three terms of Equation \ref{eq:mop-mse}, MOP-$\alpha$ achieves a lower MSE than MOP-$1$ does.
+MOP-$\alpha$ achieves a lower MSE than MOP-$1$, as $\alpha, k$ can be as large as desired to balance the impact of the last three terms of Equation \ref{eq:mop-mse}.
 We also show that MOP-$0$ achieves a MSE of $\tilde{O}\big(NpG'(\theta)^2(J^{-1}+(1-\epsilon)^{\lfloor1/c\log(J)\rfloor})\big)$, where the second term corresponds to uncontrolled mixing error.
@@
 Practical likelihood-based data analysis involves many likelihood optimizations \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
-In particular, profile likelihood confidence intervals require a sequence of optimizations.
-Additionally, a careful scientist should consider many model variations, to see whether the conclusions of the study are sensitive to alternative model specifications.
+This occurs with profile likelihood confidence intervals and model selection.
+A careful scientist should consider many model variations, to see whether the conclusions of the study are sensitive to alternative model specifications.
 Coding model variations is relatively simple when using plug-and-play inference methodology.
 However, assessing the scientific potential of these variations requires likelihood optimization.
-If many rounds of laborious searching are required to attain the maximum (as arises in previous methods) that has practical consequences for the rate at which scientists can evaluate hypotheses.
-We have demonstrated, for the first time, a plug-and-play maximum likelihood approach which can directly maximize the likelihood for the complex benchmark model of \cite{king08}, requiring some Monte Carlo replication but not prolonged sequences of explorations.
+Speeding up the process of model fitting has practical consequences for the rate at which scientists can evaluate hypotheses.
+We have demonstrated, for the first time, a plug-and-play maximum likelihood approach which succeeds in finding the MLE from a coarse global search in the complex benchmark model of \cite{king08}.
 We anticipate that this will promote scientific advances in all domains where complex POMP models arise.
@@
 
-Our gradient estimate can be used for variational inference to approximate the posterior distribution over latent states \cite{naesseth18} and parameters.
+Our gradient estimate can be used in variational inference for the posterior over latent states \cite{naesseth18} and parameters.
 

### 2024-08-20 paper/ms.tex
Updates from Overleaf

@@
 % Instead, MOP-$\alpha$ bypasses the seemingly incompatible paradox of differentiating through discrete Monte Carlo sampling by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
-MOP-$\alpha$ provides strongly consistent estimates of the filtering distribution and the log-likelihood (Theorem \ref{thm:mop-targeting}) when $\alpha=1$ or $\theta=\phi$.
-When $\alpha=1$, it also provides consistent score estimates (Theorem \ref{thm:mop-grad-consistency}).
+MOP-$\alpha$ can provide strongly consistent estimates of the filtering distribution and the log-likelihood (Theorem \ref{thm:mop-targeting}).
+When $\alpha=1$, MOP-$\alpha$ also provides consistent score estimates (Theorem \ref{thm:mop-grad-consistency}).
 This enables us to derive a linear convergence rate for gradient descent using MOP-$\alpha$ in the  presence of strong convexity and smoothness (Theorem \ref{thm:mop-convergence}).
@@
 This condition is also known as the {\it reparameterization trick} \cite{corenflos21}.
-Supposing also that the measurement model is differentiable in $\theta$, direct differentiation of MOP-$\alpha$ is available via AD.
+The discrete resampling step is more delicate to address. 
+Our goal involves estimating the derivative of the expectation of the particle filter likelihood, but the discontinuity of the resampling (as a function of $\theta$, for fixed $\omega$) in the standard particle filter means that this is not consistently estimated by the expectation of the derivative.
+MOP-$\alpha$ is constructed to avoid this discontinuity, and its derivative (available via AD) provides consistent estimation of the score (for $\alpha=1)$.
 
-MOP-$\alpha$ resamples the particles according to an arbitrary resampling rule that depends on both the target parameter value and a baseline parameter value, yielding the name {\it measurement off-parameter}.
+MOP-$\alpha$ resamples the particles according to a resampling rule at a baseline parameter value while correcting their weights to target the filtering distribution at another parameter value, yielding the name {\it measurement off-parameter}.
 This is intended as an analogy to {\it off-policy} learning in reinforcement learning.
@@
 This ensures that the resampling indices are invariant in $\theta$ when $\omega$ and $\phi$ are fixed, bypassing the issue of Monte Carlo resampling.
+
+Our approach has similarities with \cite{svensson18}, in which particles from a previous run at $\phi$ are reweighted to evaluate the likelihood at any sufficiently nearby parameter $\theta$ for $0$-th order optimization.
+However, their method is not plug-and-play because they use transition densities to reweight particles, and the variance of their log-likelihood estimate is $O(||\theta - \phi||_2^{2N})$ with an exponential dependence in the horizon.
+Our approach bypasses these issues -- the first with the \textit{reparameterization trick} to consider particle paths dependent only on $\theta$, and the second as using AD allows us to only evaluate the likelihood and score estimates when $\theta=\phi$.
+%The use of AD bypasses the issue of exponentially increasing variance, but nevertheless can result in problematic polynomial variance.
+% Our introduction of $\alpha$ avoids that difficulty.
+%also roughly $O(||\theta - \theta_{k-1}||_2^{2N})$, where $N$ is
 
@@
 
-%\subsection{Discounting Weights for a Bias-Variance Tradeoff}
+\arxiv{}{\vspace*{-2mm}}
+\subsection{Discounting Weights for a Bias-Variance Tradeoff}
 
@@
 This leads to numerical instability and high variance unless $N$ is small.
-For $\alpha<1$, the weights from previous timesteps are discounted, as in Equation~(\ref{eq:weighting-scheme}).
+For $\alpha<1$, the weights from previous timesteps are discounted as in Equation~(\ref{eq:weighting-scheme}), allowing for lower variance.
 Algorithm~\ref{alg:mop} continues to target the filtering distribution and likelihood if $\theta=\phi$, but has an asymptotically biased score estimate.
-This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate.
-%bias for $\theta\neq\phi$ that shrinks as $\theta$ approaches $\phi$
-%
-
-When $\alpha=1$, MOP-$1$ maintains complete memory of each particle's ancestral trajectory.
-Theorem~\ref{thm:mop-functional-forms} shows that it recovers the consistent but high-variance gradient estimate of \cite{poyiadjis11, scibior21}.
-When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{naesseth18}. 
-The novel possibility of $0<\alpha<1$ effectively balances a bias-variance tradeoff in both theory (Theorem \ref{thm:mop-biasvar}) and practice (Figure \ref{fig:biasvar}).
-
-Our approach has similarities with \cite{svensson18}, in which particles from a previous run at $\phi$ are reweighted to evaluate the likelihood at any sufficiently nearby parameter $\theta$ for $0$-th order optimization.
-However, their approach is not plug-and-play because they use transition densities to reweight particles, and the variance of their log-likelihood estimate is exponentially increasing in the horizon.
-Our approach bypasses these issues -- the first with the \textit{reparameterization trick} to consider particle paths dependent only on $\theta$, and the second as using AD allows us to only evaluate the likelihood and score estimates when $\theta=\phi$.
-%The use of AD bypasses the issue of exponentially increasing variance, but nevertheless can result in problematic polynomial variance.
-% Our introduction of $\alpha$ avoids that difficulty.
-%also roughly $O(||\theta - \theta_{k-1}||_2^{2N})$, where $N$ is
+MOP-$1$ and MOP-$0$ correspond to the estimators of \cite{poyiadjis11} and \cite{naesseth18} (Theorem \ref{thm:mop-functional-forms}), while the novel possibility of $0<\alpha<1$ lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate in both theory (Theorem \ref{thm:mop-biasvar}) and practice (Figure \ref{fig:biasvar}).
 
@@
 
-\cite{scibior21} show that the estimate of \cite{naesseth18} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through the use of a stop-gradient operation in the AD procedure.
-Both of these, when applied to the bootstrap filter, correspond to special cases of MOP-$\alpha$.
+When $\alpha=1$, MOP-$1$ maintains complete memory of each particle's ancestral trajectory.
+Theorem~\ref{thm:mop-functional-forms} shows that it recovers the consistent but high-variance score estimate of \cite{poyiadjis11}.
+When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased score estimate of \cite{naesseth18} that \cite{scibior21} show is the gradient of a vanilla particle filter when resampling terms are dropped.
 
@@
 Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting $\theta=\phi$ implies that $w_{n,j}^{P,\theta}$ evaluates to $1$.
+
+The result in Theorem \ref{thm:mop-functional-forms} provides mathematical intuition for why \cite{scibior21}'s use of the stop-gradient procedure in the particle filter recovers the estimate of \cite{poyiadjis11}.
+When $\theta=\phi$, the correction $g_{n,k_j}^\theta/g_{n,k_j}^\phi$ is equivalent to $g_{n,k_j}^\theta / \text{stop\_gradient}(g_{n,k_j}^\theta)$.
+As such, the use of the stop-gradient in \cite{scibior21} can be seen as a special case of the MOP-$\alpha$ weight correction for $\alpha=1, \theta=\phi$.
+
 % This further illustrates how $\alpha$ dictates the memory of the gradient estimate. 
@@
 
-\begin{figure}[ht!]
+\begin{figure}[htbp!]
   \centering
@@
 
-\arxiv{}{\vspace*{-6mm}}
+\arxiv{}{\vspace*{-4mm}}
 \section{Practical Maximum-Likelihood Estimation}
@@
 \textbf{Right:} Raincloud plot of all searches.
-IFAD outperforms all other methods, and the gradient steps improve on the warm-start given by running 40 IF2 iterations.
+IFAD outperforms all other methods.
     The dotted red line shows the true maximized log-likelihood.}
@@
 The dashed orange line depicts the median warm-start given by running 40 IF2 iterations.
-While running 60 more iterations of IF2 improves upon the median warm-start, doing so ultimately underperforms IFAD.
+Running 60 more iterations of IF2 ultimately underperforms IFAD.
 We see that IFAD has better tail control and successfully reaches the MLE.
@@
     \label{fig:optim}
-    \arxiv{}{\vspace*{-4mm}}
+    \arxiv{}{\vspace*{-2mm}}
 \end{figure}
@@
 Particle MCMC \cite{andrieu10} is arguably the most popular method for full-information plug-and-play Bayesian inference for POMP models.
-The plug-and-play property enables its use in simulation-based Bayesian inference for complex scientific models where transition densities are intractable, such as in disease modeling.
 However, particle Metropolis-Hastings often mixes slowly, requiring long burn-in periods.
@@
 
-To alleviate this, we employ a No U-Turn Sampler (NUTS) \cite{homan14} powered by MOP-$\alpha$, with a nonparametric empirical prior constructed from density estimation of the parameter swarm from the IF2 warm start having a standard deviation of $6.3\%$ of the MLE.
+To alleviate this, we employ a No U-Turn Sampler (NUTS) \cite{homan14} powered by MOP-$\alpha$, with a nonparametric empirical prior constructed from density estimation of the parameter swarm from the IF2 warm-start.
+%This prior has a standard deviation of $6.3\%$ of the MLE. 
 %\sqrt{(0.02\cdot0.95^{40})^2 \cdot 600}\approx 6.3\%
-Using a NUTS algorithm powered by MOP-$\alpha$ with this prior significantly lowers the mixing time of the Markov chain to as little as $500$ iterations (Figure \ref{fig:nuts-eb}).
+Using NUTS with MOP-$\alpha$ and this prior significantly lowers the mixing time of the Markov chain to as little as $500$ iterations (Figure \ref{fig:nuts-eb}).
 
@@
     \caption{Histogram and density estimate of the posterior of $\beta_{\text{trend}}$, as well as convergence diagnostics for NUTS with $4$ chains.
-NUTS mixes quickly, and the posterior estimates from each chain share roughly the same posterior mode.}
+NUTS mixes quickly when regularized by an empirical prior with a standard deviation of $6.3\%$ of the MLE, and the posterior estimates from each chain share roughly the same posterior mode.}
     \label{fig:nuts-eb}
@@
 \arxiv{}{\vspace*{-4mm}}
-\subsection{MOP-$1$ Gradient Is Consistent for the Score}
+\subsection{MOP-$1$ Score Estimate Is Consistent}
 
@@
 It is therefore not immediately apparent that these two converge to the same thing. As such we directly show the consistency of the MOP-$1$ gradient estimate below.
-We present an abbreviated proof here, postponing the full argument to \arxiv{Appendix~\ref{appendix:consistency}}{the supplement}.
 
@@
 \end{thm}
-\begin{proof}
-    Fix $\omega \in \Omega$, $\phi = \theta$.
+
+We present a proof outline here, postponing the full argument to \arxiv{Appendix~\ref{appendix:consistency}}{the supplement}
+Assumption \ref{assump:local-bounded-derivative} lets us show uniform equicontinuity of the sequence for almost every $\omega \in \Omega$, allowing us to apply Arzela-Ascoli and Theorem \ref{thm:mop-targeting} to establish uniform convergence for the derivatives as a sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$.
+This is a sufficient condition for swapping the limit and derivative to show that for almost every $\omega \in \Omega$, $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lik(\theta).$
+
+% \begin{proof}
+%     Fix $\omega \in \Omega$, $\phi = \theta$.
 The sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ is uniformly bounded over all $J$ by Assumption \ref{assump:local-bounded-derivative}.
@@
 
-    By Arzela-Ascoli, there is a uniformly convergent subsequence.
+%     By Arzela-Ascoli, there is a uniformly convergent subsequence.
 But there is only one subsequential limit, as we can treat the gradient estimate at $\theta$ as a bounded functional of the particles by \ref{assump:local-bounded-derivative}, allowing us to apply Theorem \ref{thm:mop-targeting} to see that the sequence $(\nabla_\theta \hat\lik_J^1(\theta)(\omega))_{J \in \mathbb{N}}$ converges pointwise for $\theta=\phi$ and almost every $\omega \in \Omega$.
@@
 
-    With uniform convergence for the derivatives established, we can swap the limit and derivative and obtain, with the strong consistency $\hat{\lik}_J^1(\theta)(\omega) \stackrel{a.s.}{\to} \lik(\theta)$ in Theorem \ref{thm:mop-targeting}, that for almost every $\omega \in \Omega$,
-    $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lik(\theta).$
-    The result follows by the continuous mapping theorem. 
-\end{proof}
+%     With uniform convergence for the derivatives established, we can swap the limit and derivative and obtain, with the strong consistency $\hat{\lik}_J^1(\theta)(\omega) \stackrel{a.s.}{\to} \lik(\theta)$ in Theorem \ref{thm:mop-targeting}, that for almost every $\omega \in \Omega$,
+%     $\lim_{J \to \infty} \nabla_\theta \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik_J^1(\theta)(\omega) = \nabla_\theta \lik(\theta).$
+%     The result follows by the continuous mapping theorem. 
+% \end{proof}
 
-\arxiv{}{\vspace*{-4mm}}
+\arxiv{}{\vspace*{-2mm}}
 \subsection{MOP-$\alpha$ Error, Bias and Variance}
@@
 \arxiv{}{\vspace*{-1mm}}
-\paragraph{Interpretation:} Theorem \ref{thm:mop-biasvar} provides theoretical understanding of the results observed in Figure \ref{fig:biasvar} that show the empirically favorable bias-variance tradeoff enjoyed by MOP-$\alpha$.
-The first term in the MSE bound in Equation \ref{eq:mop-mse} corresponds to the error of the particle approximation, the second mixing error, and the third and fourth the error between the MOP-$(1,k)$ and MOP-$\alpha$ estimates.
+\paragraph{Interpretation:} Theorem \ref{thm:mop-biasvar} provides theoretical understanding of the empirically favorable bias-variance tradeoff enjoyed by MOP-$\alpha$ in Figure \ref{fig:biasvar}.
+The first term in the MSE bound in Equation \ref{eq:mop-mse} corresponds to particle approximation error, the second mixing error, and the third and fourth the error between the MOP-$(1,k)$ and MOP-$\alpha$ estimates.
 As $\alpha$ goes to $1$, choosing $k$ appropriately, the first, third and fourth term (corresponding to the variance) increases, while the second term (corresponding to the bias) decreases. 
-Likewise, the first term in the variance bound in Equation \ref{eq:mop-variance} corresponds to the error of the particle approximation, while the second corresponds to mixing error. 
-As $\alpha$ goes to $1$, the variance increases.
+Likewise, the first term in the variance bound in Equation \ref{eq:mop-variance} corresponds to particle approximation error, while the second corresponds to mixing error. 
+The variance increases as $\alpha$ goes to $1$.
 
@@
 In that case, one requires access to the transition densities, or at least their ratios. 
-The discounting parameter, $\alpha$, can be incorporated as for MOP-$\alpha$.
 This algorithm may be better suited to large discrete state spaces than either MOP-$\alpha$ or the Baum-Welch algorithm.

### 2024-08-20 paper/appendix/coverletter.tex
Updates from Overleaf

@@
 
-Please consider our manuscript, "Accelerated Inference for Partially Observed Markov Processes using Automatic Differentiation", for publication in PNAS.
+Please consider our manuscript, ``Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes", for publication in PNAS.
 

### 2024-08-20 paper/ms.tex

@@
 
-However, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability, while the stop-gradient procedure is introduced without mathematical motivation.
+However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability \cite{poyiadjis11}.
+Additionally, while the stop-gradient procedure is undeniably effective, the reason why the correction recovers the \cite{poyiadjis11} estimator remains somewhat unclear.
 We use \cite{scibior21} and \cite{poyiadjis11} as our starting point in developing a new approach remedying these issues.
@@
 The result in Theorem \ref{thm:mop-functional-forms} provides mathematical intuition for why \cite{scibior21}'s use of the stop-gradient procedure in the particle filter recovers the estimate of \cite{poyiadjis11}.
-When $\theta=\phi$, the correction $g_{n,k_j}^\theta/g_{n,k_j}^\phi$ is equivalent to $g_{n,k_j}^\theta / \text{stop\_gradient}(g_{n,k_j}^\theta)$.
-As such, the use of the stop-gradient in \cite{scibior21} can be seen as a special case of the MOP-$\alpha$ weight correction for $\alpha=1, \theta=\phi$.
+When $\theta=\phi$, the correction $g_{n,k_j}^\theta/g_{n,k_j}^\phi$ is equivalent to $g_{n,k_j}^\theta / \text{stop\_gradient}(g_{n,k_j}^\theta)$, and is implemented in this way in practice.
+As such, the use of the stop-gradient in \cite{scibior21} can be seen as the MOP-$\alpha$ weight correction for $\alpha=1, \theta=\phi$.
 
@@
 
-\arxiv{}{\vspace*{-4mm}}
+\arxiv{}{\vspace*{-3mm}}
 \section{Practical Maximum-Likelihood Estimation}
-
+\arxiv{}{\vspace*{-1mm}}
 %One can think of MOP-$\alpha$ as taking a likelihood estimate at $\phi$, reweighting it to instead target the likelihood at $\theta$, and then taking the derivative with respect to $\theta$ of the above estimate.

### 2024-08-26 poster/poster.tex
Updates from Overleaf

% Gemini theme
% https://github.com/anishathalye/gemini
%
% We try to keep this Overleaf template in sync with the canonical source on
% GitHub, but it's recommended that you obtain the template directly from
% GitHub to ensure that you are using the latest version.

\documentclass[final]{beamer}

% ====================
% Packages
% ====================

\newcommand{\off}{\operatorname{off}}
\newcommand{\on}{\operatorname{on}}
\newcommand{\gX}{\mathcal{X}}
\newcommand{\gA}{\mathcal{A}}
\newcommand{\gS}{\mathcal{S}}
\newcommand{\gF}{\mathcal{F}}
\newcommand{\gT}{\mathcal{T}}
\newcommand{\gL}{\mathcal{L}}
\newcommand{\Reg}{\text{Reg}}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[size=custom,width=36,height=24,scale=1]{beamerposter}
\usetheme{gemini}
\usecolortheme{gemini}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{verbatim}
\input{paper/macros}
%\usepackage{enumitem}

\usepackage[skins,theorems]{tcolorbox}
%\tcbset{colback=white}
\tcbset{highlight math style={enhanced,colframe=red,colback=white}}

\usepackage{colortbl} % Add this package to use \cellcolor
\definecolor{lightgreen}{rgb}{0.9, 1, 0.9} % Light green color definition

\pgfplotsset{compat=1.14}

% ====================
% Lengths
% ====================

% If you have N columns, choose \sepwidth and \colwidth such that
% (N+1)*\sepwidth + N*\colwidth = \paperwidth

\setlength{\paperwidth}{36in}
\setlength{\paperheight}{24in}

\newlength{\sepwidth}
\newlength{\colwidth}
\setlength{\sepwidth}{0.025\paperwidth}
\setlength{\colwidth}{0.3\paperwidth}

\newcommand{\separatorcolumn}{\begin{column}{\sepwidth}\end{column}}

\newcommand{\cross}[1][1pt]{\ooalign{%
  \rule[1ex]{1ex}{#1}\cr% Horizontal bar
  \hss\rule{#1}{.7em}\hss\cr}}% Vertical bar

% ====================
% Title
% ====================

\title{Automatic differentiation accelerates inference
for partially-observed Markov processes}

\author{Kevin Tan \inst{*} \and Giles Hooker \inst{*} \and Edward L. Ionides \inst{\cross[.4pt]$\cdot$}}

\institute[shortinst]{\inst{*} Department of Statistics and Data Science, University of Pennsylvania \quad  \inst{\cross[.4pt]} Department of Statistics, University of Michigan}

% ====================
% Footer (optional)
% ====================

\footercontent{
  \hfill
   \hfill
  }
% (can be left out to remove footer)

% ====================
% Logo (optional)
% ====================

% use this to include logos on the left and/or right side of the header:
 %\logoright{\includegraphics[scale=0.24]{logo1.png}}
%\logoleft{\includegraphics[scale=2.5]{sci logo.png}}

% ====================
% Body
% ====================

\begin{document}

\begin{frame}[t]
\begin{columns}[t]
\separatorcolumn

\begin{column}{\colwidth}

  \begin{block}{Introduction}

\textbf{Inference for continuous-time, continuous-state hidden Markov models.}

\begin{tcolorbox}[enhanced,colback=white!100!white,colframe=red!100!red]
\textbf{\textcolor{red}{Partially-Observed Markov Process (POMP)}}: 
\begin{itemize}
    \item Unknown Markov process $\{X_t, t \geq t_0\}$, but we have observations $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$.  
    \item Given measurements $y_n^*$, estimate both the latent state of the system $x_n$ at timesteps $t_0,...,t_N$ and system parameters $\theta \in \Theta$. 
\end{itemize}
\end{tcolorbox}
e.g. given monthly cholera case counts, estimate number of infected and recovered people in the population, and how transmissible the disease is.

  \end{block}

  \begin{block}{Problem}
  Maximum likelihood with only a simulator, cannot evaluate $f_{X_n|X_{n-1};\theta}$.

  \begin{tcolorbox}[enhanced,colback=white!100!white,colframe=red!100!red]
  \begin{center}
      \textbf{Cannot use the EM algorithm!}
  \end{center}
\end{tcolorbox}

  \textbf{What to do now?} Particle filters provide a Monte Carlo approximation of the \textbf{filtering distribution} $f_{X_n|Y_{1:n};\theta}$ and the \textbf{log-likelihood} $\ell(\theta)$.

  \textbf{Gradient descent} using autodiff (AD) on the likelihood estimate? \textcolor{red}{But:}
  \begin{enumerate}
      \item Likelihood estimate not differentiable due to Monte Carlo resampling.
      \item Previous work \cite{corenflos21, naesseth18, scibior21} circumvents this, but computationally expensive, has uncontrolled asymptotic bias, or high variance. 
  \end{enumerate}
  \end{block}

  \begin{alertblock}{Solution: Off-Parameter Resampling}

    \textbf{Reconstruct likelihood} at $\theta$ by cumulatively reweighting particles resampled at a \textbf{baseline parameter} $\phi \in \Theta$. Take gradient at $\theta=\phi$.

    When treating the particles at $\phi$ as constants, you're now differentiating through (1) a differentiable simulator and (2) a series of measurement density ratios, so the resulting likelihood estimate is smooth!

    \textbf{Bias-variance tradeoff:} Discount cumulative reweighting by $\alpha \in [0,1]$.

    \begin{figure}
        \centering
        \includegraphics[width=0.5\linewidth]{}
        \caption{Caption}
        \label{fig:enter-label}
    \end{figure}
  \end{alertblock}
%Only simulation-based method for maximum likelihood available \cite{ionides15} gets close (\~40 nats) to the MLE quickly, but fails to find it ($\geq 7$ nats away).

% \vspace{0.5em}
%  \begin{alertblock}{tl;dr}

%     \begin{itemize}
%         \item Modified an optimistic algorithm for general function approximation algorithm called GOLF \cite{jin2021bellmaneluder, xie2022policy}. 
%         \item Including the offline dataset in parameter estimation achieves provable improvements in regret over pure offline/online learning. 
%         \vspace{0.5ex}
%         \item How? Consider {\it arbitrary} (not necessarily disjoint) partitions of the state-action space $\gX_{\off} \cup \gX_{\on} = \gX$.
%         \begin{itemize}
%             \item Bound regret by the coverage of the behavior policy on $\gX_{\off}$  and a complexity measure for online learning on $\gX_{\on}$.
%             \item Overall regret is characterized by the regret bound on the best possible partition -- even if the algorithm is unaware of it. 
%         \end{itemize}
%         \item Not specific to DISC-GOLF! Yields a general recipe for initializing generic online RL algorithms with offline data without full coverage.
%     \end{itemize}

%   \end{alertblock}

\end{column}

\separatorcolumn

\begin{column}{\colwidth}

  \begin{block}{Complexity Over Partitions}

  \textbf{Observation:} No need to look at the behavior policy's coverage over the whole space for offline learning, just the part it explores well.

  So partition the state-action space $\gX = [H] \times \gS \times \gA$ into $\gX_{\off}, \gX_{\on}$, and consider \textbf{partial offline and online complexity measures defined by the restrictions}:

    \textit{Partial All-Policy Concentrability Coefficient, \textcolor{blue}{restriction in blue}}
        $$c_{\off}(\gF, \gX_{\off}) 
    \coloneqq \max_h \sup_{f \in \gF} \frac{\|(f_h - \gT_h f_{h+1}) \tcbhighmath[colframe=blue]{\mathbbm{1}_{(\cdot, h) \in \gX_{\off}}}\|_{2, d_h^\pi}^2}{\|(f_h - \gT_h f_{h+1})\tcbhighmath[colframe=blue]{\mathbbm{1}_{(\cdot, h) \in \gX_{\off}}}\|_{2, \mu_{h}}^2}.$$
    {\small There exists $\gX_{\off}$ so this is finite, even when the single-policy concentrability coefficient isn't.}

    \textit{Partial Sequential Exploration Coefficient, \textcolor{red}{restriction in red}}
    $$c_{\on}(\gF, \gX_{\on},T) 
    \coloneqq \max_{h} \sup_{f^{(t)}} \sup_{\pi^{(t)}} 
    \left\{\sum_{t=1}^T \frac{ \E_{d_h^{\pi^{(t)}}}[(f_h^{(t)} - \gT_h f_{h+1}^{(t)}) \tcbhighmath[colframe=red]{\mathbbm{1}_{(\cdot, h) \in \gX_{\on}}}]^2}{H^2 \vee \sum_{i=1}^{t-1} \E_{d_h^{\pi^{(i)}}}[(f_h^{(t)} - \gT_h f_{h+1}^{(t)})^2 \tcbhighmath[colframe=red]{\mathbbm{1}_{(\cdot, h) \in \gX_{\on}}}]}\right\}.$$

  \end{block}

    \begin{alertblock}{Punchline}
        Regret of a hybrid algorithm can be decomposed into two terms depending on the offline and online complexity measures for the best possible partition $\gX_{\on}$ and $\gX_{\off}$ -- despite the algorithm not knowing the best partition, if:
        \vspace{-3ex}
        \begin{enumerate}
            \item The regret can be decomposed into e.g. the sum of bonus terms/Bellman errors over all $h, t$.
            \item In-sample error/error over offline dataset is bounded.
            \item The original algorithm has an online-only regret bound.
        \end{enumerate}
    \end{alertblock}

\end{column}

\separatorcolumn

\begin{column}{\colwidth}

    \begin{alertblock}{Theorem: One such example of a regret bound}
\label{thm:regret_bound}
Let $\gX_{\off}, \gX_{\on}$ be an arbitrary partition over $\gX = \gS \times \gA \times [H]$. 
If we warm-start the online exploration of GOLF \cite{jin2021bellmaneluder, xie2022role} by appending the offline data to the experience replay buffer at the beginning, the following regret bound holds with probability at least $1-\delta$: 
\begin{align*}
    \Reg(T)
    \lesssim \inf_{\gX_{\on}, \gX_{\off}} \bigg(\underbrace{\sqrt{\beta H^4 c_{\off}(\gF, \gX_{\off})} T^2/N_{\off}}_{\textcolor{blue}{\text{\small{offline term}}}} + \underbrace{\sqrt{\beta H^4 N_{\on} c_{\on}(\gF, \gX_{\on}, T)}}_{\textcolor{red}{\text{\small{online term}}}}\bigg),
\end{align*} 
where 
$\beta = O(\log (N H \mathcal{N}_{\mathcal{F}}(1/N) / \delta))$ and $N = T + N_{\off}$.
\end{alertblock}

    % \begin{block}{}
    %     \textbf{Tabular:} $$\inf_{\gX_{\on}, \gX_{\off}} \Bigg(\sqrt{H^5SAN_{\on}\left(\frac{N_{\on}}{N_{\off}}\right)\sup_{\pi }\left\lVert\frac{d_h^{\pi}\mathbbm{1}_{\gX_{\off}}}{\mu_h^{\pi}}\right\rVert_{\infty}}  
    % + \sqrt{H^5SA\max_{h \in [H]}|\gX_{\on}|N_{\on}}\Bigg).$$

    % \textbf{Linear:} 
    % $$\inf_{\gX_{\on}, \gX_{\off}} \Bigg(\sqrt{dH^5N_{\on}\left(\frac{N_{\on}}{N_{\off}}\right)\max_h \frac{1}{\lambda_{d_{\off}}(\E_{\mu_h}[ \phi_{\off}\phi_{\off}^\top])}} + \sqrt{d_{\on}dH^5N_{\on}}\Bigg).$$
    % \end{block}

  \begin{exampleblock}{Simulations: Hybrid RL Encourages Exploration}

  \textbf{Does appending the offline dataset to the experience replay buffer encourage sufficient exploration for the portion of the state-action space that does not have good coverage?}

  \textbf{Tabular:} Initialized UCBVI \cite{azar2017minimax} with an offline dataset of $100$ trajectories of length $20$ on a forest management simulator. 
  %Online partition is the state-action pairs with occupancy below $1/SA$.

\begin{figure}[H]
    \centering
    \includegraphics[scale=1]{imgs/095/biasvar.png}
    \caption{Hybrid RL explores more of the online partition (poorly covered $s,a$) to fill in the gaps in the offline dataset than online RL does, unless the behavior policy is optimal.
    }
\end{figure}

    \textbf{Linear:} Initialized LSVI-UCB \cite{jin2020provably} with $200$ trajectories of length $40$ on a scaled-down version of Tetris.

    \begin{figure}[H]
    \centering
    \includegraphics[scale=1]{imgs/095/optim.png}
    \caption{Hybrid RL takes fewer online episodes than online-only RL to achieve a lower concentrability coefficient.}
    \label{fig:linear-coverage}
\end{figure}

  \end{exampleblock}

\begin{block}{Further Work}
    \begin{itemize}
        \item Improving our regret bound to a guarantee based on partial single-policy concentrability would be valuable.
        \item GOLF requires Bellman completeness, but maybe we only need realizability if we can use e.g. OLIVE \cite{du2021bilinear}.
        \item Followup work \cite{tan2024hybridreinforcementlearningbreaks} uses this framework to show provable gains over the online-only regret bound of $\sqrt{d^2H^3T}$ in linear MDPs, 
        $$\text{Reg}(T) \lesssim \inf _{\mathcal{X}_{\text {off}}, \mathcal{X}_{\text {on}}} \sqrt{c_{\text {off }}\left(\mathcal{X}_{\text {off }}\right)^2 d H^3 T^2 / N_{\text {off}}}+\sqrt{d_{\text {on }} d H^3 T},
        \vspace{-0.5ex}$$
        where $d_{\on}$ is the dimension of the online partition.
    \end{itemize}
\end{block}
\vspace{-2ex}

%\vskip-1ex
\begin{block}{}
  \vspace{-0.2em}
    \tiny{\bibliographystyle{plain}
    \bibliography{paper/bib-ifad}}
\end{block}

  %\vskip-1ex
  \vspace{0.2em}

  % \begin{block}{Contact Information}
  % \vskip-0ex
  %   %\nocite{*}
  %   \small{{Email Address}: kevtan@wharton.upenn.edu}
  % \end{block}

\end{column}

\separatorcolumn
\end{columns}
\end{frame}

\end{document}

### 2024-08-27 poster/poster.tex
Updates from Overleaf

@@
 \usepackage[T1]{fontenc}
+\usepackage{boldline}
 \usepackage{lmodern}
@@
 \usecolortheme{gemini}
+\usepackage{natbib}
 \usepackage{graphicx}
@@
 %\tcbset{colback=white}
+
+\definecolor{ballblue}{rgb}{0.13, 0.67, 0.8}
+\definecolor{darkred}{RGB}{139,0,0}
+
 \tcbset{highlight math style={enhanced,colframe=red,colback=white}}
@@
 
+  \vspace{-1ex}
   \begin{block}{Problem}
-  Maximum likelihood with only a simulator, cannot evaluate $f_{X_n|X_{n-1};\theta}$.
-
+  \vspace{-1.6ex}
   \begin{tcolorbox}[enhanced,colback=white!100!white,colframe=red!100!red]
   \begin{center}
-      \textbf{Cannot use the EM algorithm!}
+  Maximum likelihood with only a simulator, cannot evaluate $f_{X_n|X_{n-1};\theta}$. \textbf{Cannot use the EM algorithm!}
   \end{center}
@@
 
-  \textbf{What to do now?} Particle filters provide a Monte Carlo approximation of the \textbf{filtering distribution} $f_{X_n|Y_{1:n};\theta}$ and the \textbf{log-likelihood} $\ell(\theta)$.
+  \textbf{What to do now?} Particle filters provide a Monte Carlo approximation of the \textbf{filtering distribution} $f_{X_n|Y_{1:n};\theta}$ and the \textbf{log-likelihood} $\ell(\theta)$ increasingly accurate in the number of particles $J$.
 
@@
       \item Likelihood estimate not differentiable due to Monte Carlo resampling.
-      \item Previous work \cite{corenflos21, naesseth18, scibior21} circumvents this, but computationally expensive, has uncontrolled asymptotic bias, or high variance. 
+      \item Previous work \cite{corenflos21, naesseth18, scibior21} on this, but computationally expensive \cite{corenflos21}, has uncontrolled asymptotic bias \cite{naesseth18}, or has high variance \cite{poyiadjis11}.
   \end{enumerate}
@@
 
+  \end{alertblock}
+
+  \vspace{-1ex}
     \begin{figure}
         \centering
-        \includegraphics[width=0.5\linewidth]{}
-        \caption{Caption}
+        \includegraphics[width=0.45\linewidth]{imgs/095/mop.png}
+        \includegraphics[width=0.405\linewidth]{imgs/095/biasvar.png}
+        \caption{Dhaka cholera model of \cite{king08}.
+\textbf{Left:} Likelihood slice for the recovery rate parameter, and smooth off-parameter reconstructions for different $\alpha$.
+\textbf{Right:} Bias-variance tradeoff induced by $\alpha$ for estimation of the score for the trend parameter.
+}
         \label{fig:enter-label}
     \end{figure}
-  \end{alertblock}
+
 %Only simulation-based method for maximum likelihood available \cite{ionides15} gets close (\~40 nats) to the MLE quickly, but fails to find it ($\geq 7$ nats away).
@@
 
-  \begin{block}{Complexity Over Partitions}
+  \begin{alertblock}{Guarantees}
 
-  \textbf{Observation:} No need to look at the behavior policy's coverage over the whole space for offline learning, just the part it explores well.
+  \textbf{Correctness:} This procedure targets the filtering distribution, $\pi_n(\theta) = f_{X_n|Y_{1:n}}(x_n|y_{1:n}^*;\theta)$, and is strongly consistent for the likelihood under $\theta$. 
+  We obtain a set of state estimates $(x_{n,j}^{F,\theta})_{j=1}^J$ and a likelihood estimate $\hat \lik_J(\theta)$ so for any $\alpha\in[0,1]$, for any measurable bounded functional $h$:
+  $$\frac{\sum_{j=1}^J h\left(x_{n, j}^{F, \theta}\right) w_{n, j}^{F, \theta}}{\sum_{j=1}^J w_{n, j}^{F, \theta}} \stackrel{\text{a.s.}}{\to} E_{\pi_n(\theta)}\left[h\left(X_n\right)\right], \; \hat{\mathcal{L}}_J^\alpha(\theta) \stackrel{\text{a.s.}}{\to} \mathcal{L}(\theta) \text{ as } J \to \infty.$$
 
-  So partition the state-action space $\gX = [H] \times \gS \times \gA$ into $\gX_{\off}, \gX_{\on}$, and consider \textbf{partial offline and online complexity measures defined by the restrictions}:
+  \textbf{Consistent score estimates:} When $\alpha=1$, $\nabla_\theta \hat\ell_J^1(\theta) \stackrel{\text{a.s.}}{\to} \ell(\theta)$ as $J \to \infty$.
+Else, increasing asymptotic bias as $\alpha \to 0$ in return for lower variance.
 
-    \textit{Partial All-Policy Concentrability Coefficient, \textcolor{blue}{restriction in blue}}
-        $$c_{\off}(\gF, \gX_{\off}) 
-    \coloneqq \max_h \sup_{f \in \gF} \frac{\|(f_h - \gT_h f_{h+1}) \tcbhighmath[colframe=blue]{\mathbbm{1}_{(\cdot, h) \in \gX_{\off}}}\|_{2, d_h^\pi}^2}{\|(f_h - \gT_h f_{h+1})\tcbhighmath[colframe=blue]{\mathbbm{1}_{(\cdot, h) \in \gX_{\off}}}\|_{2, \mu_{h}}^2}.$$
-    {\small There exists $\gX_{\off}$ so this is finite, even when the single-policy concentrability coefficient isn't.}
+  \textbf{Encompasses other estimators:} When $\alpha=0$, $\nabla_\theta \hat\ell_J^0(\theta)$ is equivalent to the low-variance but asymptotically biased estimate of \cite{naesseth18}.
+When $\alpha=1$, $\nabla_\theta \hat\ell_J^1(\theta)$ is equivalent to the high-variance but consistent estimate of \cite{poyiadjis11}.
 
-    \textit{Partial Sequential Exploration Coefficient, \textcolor{red}{restriction in red}}
-    $$c_{\on}(\gF, \gX_{\on},T) 
-    \coloneqq \max_{h} \sup_{f^{(t)}} \sup_{\pi^{(t)}} 
-    \left\{\sum_{t=1}^T \frac{ \E_{d_h^{\pi^{(t)}}}[(f_h^{(t)} - \gT_h f_{h+1}^{(t)}) \tcbhighmath[colframe=red]{\mathbbm{1}_{(\cdot, h) \in \gX_{\on}}}]^2}{H^2 \vee \sum_{i=1}^{t-1} \E_{d_h^{\pi^{(i)}}}[(f_h^{(t)} - \gT_h f_{h+1}^{(t)})^2 \tcbhighmath[colframe=red]{\mathbbm{1}_{(\cdot, h) \in \gX_{\on}}}]}\right\}.$$
+  \end{alertblock}
+
+  \vspace{-2ex}
+  \begin{block}{Rates}
+  \vspace{-6.5mm}
+      \begin{table}[t]
+\centering
+\scalebox{0.85}{
+\begin{tabular}{c  c}
+\toprule & MSE 
+\tabularnewline
+\toprule 
+\vspace{0.2ex}
+$\alpha \in (0,1)$ & $\tcbhighmath[colframe=ballblue]{\min_{k \leq N} N p G^{\prime}(\theta)^2(k^2/J+(1-\epsilon)^{\lfloor k /(c \log (J))\rfloor}+k+\psi_k(\alpha))}$
+\vspace{0.2ex}
+\tabularnewline \hline
+\vspace{0.2ex}
+$\alpha=0$, equivalent to \cite{naesseth18} & $N p G'(\theta)^2(J^{-1} + (1-\epsilon)^{\lfloor 1 /(c \log (J))\rfloor})$
+\vspace{0.2ex}
+\tabularnewline
+\hline
+\vspace{0.2ex}
+$\alpha=1$, equivalent to \cite{poyiadjis11} & $N^4pG'(\theta)^2/J$
+\vspace{0.2ex}
+\tabularnewline
+\toprule
+&  Variance
+\tabularnewline
+\toprule\vspace{0.2ex}
+$\alpha \in (0,1)$ & $\tcbhighmath[colframe=ballblue]{\min _{k \leq N}\left(\frac{k^2 p G^{\prime}(\theta)^2 N p}{(1-\alpha)^2 J}+\frac{\alpha^k}{1-\alpha} N p G^{\prime}(\theta)^2\right)}$
+\vspace{0.2ex}
+\tabularnewline \hline
+\vspace{0.2ex}
+$\alpha=0$, equivalent to \cite{naesseth18} & $N p G'(\theta)^2/J$
+\vspace{0.2ex}
+\tabularnewline
+\hline
+\vspace{0.2ex}
+$\alpha=1$, equivalent to \cite{poyiadjis11} & $N^4pG'(\theta)^2/J$
+\vspace{0.2ex}
+\tabularnewline
+\toprule 
+\end{tabular}}
+\medskip
+\caption{Rates given the number of particles $J$, trajectory length $N$, dimension of parameter space $p$, bound on gradient norm $G'(\theta)$, and model-specific constant $\epsilon>0$.
+We define $\psi_k(\alpha)=\left(\alpha^k+\alpha^{k+1}-\alpha\right) /(1-\alpha)$.
+Our novel contribution of $\alpha \in (0,1)$ has a lower MSE than, and interpolates between the variances of, $\alpha=0$ \cite{naesseth18} and $\alpha=1$ \cite{poyiadjis11}.}
+\label{tab:bounds}
+\vspace{-5mm}
+\end{table}
+  \end{block}
+
+\begin{block}{Practical Maximum Likelihood Estimation}
+    \textbf{Problem:} Difficult, noisy, and nonconvex problem. Gradient descent/L-BFGS/Newton struggles with saddle points, local minima, and ridges.
+
+    \textbf{Solution:} Warm-start gradient descent with IF2 algorithm of \cite{ionides15}.
+IF2 gets to a neighborhood of the MLE quickly, but fails to find it.
+Gradient descent does well in that neighborhood, warm-start achieves best of both worlds.
+
+    \textbf{Theoretical guarantee:}
 
 \end{block}
-
-    \begin{alertblock}{Punchline}
-        Regret of a hybrid algorithm can be decomposed into two terms depending on the offline and online complexity measures for the best possible partition $\gX_{\on}$ and $\gX_{\off}$ -- despite the algorithm not knowing the best partition, if:
-        \vspace{-3ex}
-        \begin{enumerate}
-            \item The regret can be decomposed into e.g. the sum of bonus terms/Bellman errors over all $h, t$.
-            \item In-sample error/error over offline dataset is bounded.
-            \item The original algorithm has an online-only regret bound.
-        \end{enumerate}
-    \end{alertblock}
 
@@
 \begin{column}{\colwidth}
-
-    \begin{alertblock}{Theorem: One such example of a regret bound}
-\label{thm:regret_bound}
-Let $\gX_{\off}, \gX_{\on}$ be an arbitrary partition over $\gX = \gS \times \gA \times [H]$. 
-If we warm-start the online exploration of GOLF \cite{jin2021bellmaneluder, xie2022role} by appending the offline data to the experience replay buffer at the beginning, the following regret bound holds with probability at least $1-\delta$:
-\begin{align*}
-    \Reg(T)
-    \lesssim \inf_{\gX_{\on}, \gX_{\off}} \bigg(\underbrace{\sqrt{\beta H^4 c_{\off}(\gF, \gX_{\off})} T^2/N_{\off}}_{\textcolor{blue}{\text{\small{offline term}}}} + \underbrace{\sqrt{\beta H^4 N_{\on} c_{\on}(\gF, \gX_{\on}, T)}}_{\textcolor{red}{\text{\small{online term}}}}\bigg),
-\end{align*} 
-where 
-$\beta = O(\log (N H \mathcal{N}_{\mathcal{F}}(1/N) / \delta))$ and $N = T + N_{\off}$.
-\end{alertblock}
-
-    % \begin{block}{}
-    %     \textbf{Tabular:} $$\inf_{\gX_{\on}, \gX_{\off}} \Bigg(\sqrt{H^5SAN_{\on}\left(\frac{N_{\on}}{N_{\off}}\right)\sup_{\pi }\left\lVert\frac{d_h^{\pi}\mathbbm{1}_{\gX_{\off}}}{\mu_h^{\pi}}\right\rVert_{\infty}}
-    % + \sqrt{H^5SA\max_{h \in [H]}|\gX_{\on}|N_{\on}}\Bigg).$$
-
-    % \textbf{Linear:} 
-    % $$\inf_{\gX_{\on}, \gX_{\off}} \Bigg(\sqrt{dH^5N_{\on}\left(\frac{N_{\on}}{N_{\off}}\right)\max_h \frac{1}{\lambda_{d_{\off}}(\E_{\mu_h}[ \phi_{\off}\phi_{\off}^\top])}} + \sqrt{d_{\on}dH^5N_{\on}}\Bigg).$$
-    % \end{block}
 
@@
 
-\begin{block}{Further Work}
-    \begin{itemize}
-        \item Improving our regret bound to a guarantee based on partial single-policy concentrability would be valuable.
-        \item GOLF requires Bellman completeness, but maybe we only need realizability if we can use e.g. OLIVE \cite{du2021bilinear}.
-        \item Followup work \cite{tan2024hybridreinforcementlearningbreaks} uses this framework to show provable gains over the online-only regret bound of $\sqrt{d^2H^3T}$ in linear MDPs, 
-        $$\text{Reg}(T) \lesssim \inf _{\mathcal{X}_{\text {off}}, \mathcal{X}_{\text {on}}} \sqrt{c_{\text {off }}\left(\mathcal{X}_{\text {off }}\right)^2 d H^3 T^2 / N_{\text {off}}}+\sqrt{d_{\text {on }} d H^3 T},
-        \vspace{-0.5ex}$$
-        where $d_{\on}$ is the dimension of the online partition.
-    \end{itemize}
-\end{block}
 \vspace{-2ex}

### 2024-10-10 online/online.tex
updates to GitHub

\documentclass{article}
\usepackage[margin=2cm]{geometry}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{positioning}

\setlist{nosep} % to reduce space in lists

\setlength{\parskip}{0.5em}
\setlength{\parindent}{0in}

\input{paper/macros}

\newcommand{\off}{\operatorname{off}}
\newcommand{\on}{\operatorname{on}}

\title{Online Gradient Methods for State and Parameter Estimation in State Space Models}
\author{Kevin Tan, Giles J. Hooker, Edward L. Ionides}
\date{}

\comments=1

\begin{document}
\maketitle
\begin{abstract}

Previous work has focused on obtaining score estimates that balance a bias-variance tradeoff in the batch setting, where one filters through the entire history of observations repeatedly to perform maximum likelihood parameter estimation. We develop an extension for the online setting, where one is interested in inferring both the state of the system as well as performing system identification. We show that the online extension of the MOP-$\alpha$ algorithm of \cite{tan2024acceleratedinferencepartiallyobserved} has a natural interpretation as online gradient descent with momentum, and derive a version of Nesterov's accelerated gradient using this interpretation. Moreover, we show that constructing this estimate using the bootstrap filter bypasses the limitations of the linearly increasing variance of the online high-variance score estimate from \cite{poyiadjis11}. Numerical experiments on robot localization, autonomous driving, and disease modeling illustrate the effectiveness of our approach.
\end{abstract}

\section{Introduction}

$\sum_{n=1}^t \alpha^{t-n}  \nabla_{\theta_t} \log f_{Y_n|X_n}(y_n^*|x_{n,j}^F\theta_t)$ against $\sum_{n=1}^t \alpha^{t-n} \nabla_{\theta_n} \ell_n(\theta_n)$, where the former corresponds to the MOP-$\alpha$ estimate for the score of the whole trajectory seen so far, while the latter corresponds to the SGD with momentum gradient update. The former requires refiltering through the trajectory seen so far to construct the gradient estimate for the log-likelihood at $\theta_t$, while the latter only requires taking the gradient of the log-measurement density observed at time $t$.

Weird observation -- online MOP-$1$ according to Poyiadjis is is MOP-$0$ applied sequentially, in the case of the bootstrap filter. MOP-$1$ sums the log-densities of the surviving particles, while MOP-$0$ sums the log-densities of the particles at each timestep.

\bibliography{paper/bib-ifad}
\bibliographystyle{apalike}

\appendix
\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}

\end{document}

### 2024-10-10 paper/appendix/coverletter.tex
updates to GitHub

@@
 
-\noindent Dear PNAS editors,
+\noindent Dear JRSS-B editors,
 \newline
 
-Please consider our manuscript, ``Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes", for publication in PNAS.
+Please consider our manuscript, ``Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes", for publication in JRSS-B.
 
@@
 Our method promises substantially improved inference wherever simulation-based inference is used for the full-information likelihood in complex partially observed Markov processes.
-That is why we believe that publication of our article in PNAS would facilitate a broad range of scientific advances.
+That is why we believe that publication of our article in JRSS-B would facilitate a broad range of scientific advances.
 
@@
 Nicolas Chopin is a leading SMC theorist who is also involved in developing SMC computation so he would also be a qualified referee.
-It would assist the evaluation of our contribution if at least one referee or invited editor could be drawn from this select group of experts.
+It would assist the evaluation of our contribution if at least one referee could be drawn from this select group of experts.
 
@@
 Therefore, we ask for at least one referee who has experience working outside the Bayesian paradigm.
+Additionally, due to a disagreement that Ed Ionides had with Michael Whitehouse, Nick Whiteley, and Lorenzo Rimella regarding the correctness of a numerical experiment within their paper on Poisson Approximate Likelihoods published at JRSS-B, we similarly ask that none of the referees be drawn from the authors of that paper to preserve the impartiality of reviews.
 

### 2024-10-10 paper/ms.tex
updates to GitHub

@@
 %\documentclass[9pt,twocolumn,twoside]{pnas-new}
-\documentclass[9pt,twocolumn,pnasresearcharticle]{pnas-new}
+\documentclass[numsec,webpdf,modern,medium,namedate]{oup-authoring-template}
 %\documentclass[9pt,twocolumn,twoside]{pnas-new}
 % Use the lineno option to display guide line numbers if required.
+\onecolumn
 
 \usepackage{booktabs}
+\newcommand{\description}{}
+\usepackage{enumitem}
 
@@
 }{
-\templatetype{pnasresearcharticle} % Choose template
-% {pnasresearcharticle} = Template for a two-column research article
-% {pnasmathematics} %= Template for a one-column mathematics article
-% {pnasinvited} %= Template for a PNAS invited submission
-}
+\theoremstyle{thmstyleone}%
+\newtheorem{theorem}{Theorem}%  meant for continuous numbers
+%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
+%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
+\newtheorem{proposition}[theorem]{Proposition}%
+%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
+\theoremstyle{thmstyletwo}%
+\newtheorem{example}{Example}%
+\newtheorem{remark}{Remark}%
+\theoremstyle{thmstylethree}%
+\newtheorem{definition}{Definition}
+\usepackage[doublespacing]{setspace}
+\usepackage[fontsize=12pt]{fontsize}}
 
@@
 
-\title{Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes}
+\arxiv{\title{Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes}
 
-%\arxiv{ \author{K. Tan, G. Hooker and E. L. Ionides\\
-%  \small{Department of Statistics, University of Pennsylvania}\\
-%  \small{Department of Statistics, University of Michigan}}}{}
-
-\arxiv{ \author{K. Tan$^{1}$, G. Hooker$^{1}$ and E. L. Ionides$^{2}$
+\author{K. Tan$^{1}$, G. Hooker$^{1}$ and E. L. Ionides$^{2}$
   \vspace{2mm}\\
   \small{$^{1}$Department of Statistics, University of Pennsylvania}\\
-  \small{$^{2}$Department of Statistics, University of Michigan}}}{}
+  \small{$^{2}$Department of Statistics, University of Michigan}}
 
-\arxiv{}{
-% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
-\author[a]{Kevin Tan}
-\author[a]{Giles Hooker}
-\author[b,1]{Edward L. Ionides}
-%\author[a]{Author Three}
-\affil[a]{University of Pennsylvania}
-\affil[b]{University of Michigan}
-%\affil[c]{Affiliation Three}
-% Please give the surname of the lead author for the running footer
-\leadauthor{Tan}
-% Please add a significance statement to explain the relevance of your work
-\significancestatement{Many scientific models involve highly nonlinear stochastic dynamical systems which can be observed only via noisy and incomplete measurements.
-Prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics.
-We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
-Our new method outperforms previous approaches on a challenging problem in epidemiology.}
-\authorcontributions{K.T. and E.L.I. designed research; K.T. analyzed data; K.T., G.H. and E.L.I. performed research, K.T., G.H., and E.L.I. wrote the manuscript.}
-\authordeclaration{The authors declare no competing interests.}
-\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: ionides@umich.edu}
-% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
-\keywords{Sequential Monte Carlo $|$ Automatic Differentiation $|$ Particle Filter $|$ Markov Process $|$ Likelihood}
+}{
+
+\journaltitle{Journals of the Royal Statistical Society}
+\DOI{DOI HERE}
+\copyrightyear{XXXX}
+\pubyear{XXXX}
+\access{Advance Access Publication Date: Day Month Year}
+\appnotes{Original article}
+
+\firstpage{1}
+
+%\subtitle{Subject Section}
+\title[AD for POMPs]{Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes}
+
+\author[1,$\ast$]{Kevin Tan}
+\author[1]{Giles Hooker}
+\author[2]{Edward L. Ionides}
+%\author[4]{Fifth Author\ORCID{0000-0000-0000-0000}}
+\authormark{Tan et al.}
+
+\address[1]{\orgdiv{Department of Statistics and Data Science}, \orgname{University of Pennsylvania}, \orgaddress{\street{265 South 37th Street, 3rd \& 4th Floors}, \postcode{19104}, \state{Pennsylvania}, \country{United States of America}}}
+\address[2]{\orgdiv{Department of Statistics}, \orgname{University of Michigan}, \orgaddress{\street{323 West Hall, 1085 S University Ave}, \postcode{48109}, \state{Michigan}, \country{United States of America}}}
+
+\corresp[$\ast$]{Address for correspondence. Kevin Tan, University of Pennsylvania, Philadelphia, 19104, USA. \href{Email:kevtan@wharton.upenn.edu}{kevtan@wharton.upenn.edu}}
+
+\received{Date}{0}{Year}
+\revised{Date}{0}{Year}
+\accepted{Date}{0}{Year}
+
 }
 
-\arxiv{\date{Draft compiled on \today}}{
-\dates{This manuscript was compiled on \today}
-\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
-}
-
-\maketitle
-
+\arxiv{
 \begin{abstract}
@@
 Our algorithms require only a differentiable simulator for the latent dynamic system; by contrast, most previous approaches to AD likelihood maximization for particle filters require access to the system's transition probabilities.
-Numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from an iterated filtering algorithm show substantial improvement on current state-of-the-art methods for a challenging scientific benchmark problem.
+Numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from an iterated filtering algorithm shows substantial improvement on current state-of-the-art methods on a challenging scientific benchmark problem.
 \end{abstract}
+}{
 
-\arxiv{}{
-\thispagestyle{firststyle}
-\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
-\firstpage{4}
-% Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting.
-In this example, there are a total of 11 paragraphs on the first page, counting the first level heading as a paragraph.
-The value {12} represents the number of the paragraph starting the second page.
-If a paragraph runs over onto the second page, include a bracket with the paragraph line number starting the second page, followed by the paragraph number in curly brackets, e.g.
-"\firstpage[4]{11}".
+\abstract{Although automatic differentiation (AD) has driven many recent advances in machine learning, partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques because widely used particle filter algorithms yield an estimated likelihood that is discontinuous in the model parameters.
+To resolve this, we create a theoretical framework that embeds two existing AD particle filter methods within a new class of algorithms.
+This new class permits a bias-variance tradeoff and a mean squared error substantially lower than the existing algorithms.
+This allows us to develop likelihood maximization algorithms suited to the Monte Carlo properties of the AD gradient estimate.
+Our algorithms require only a differentiable simulator for the latent dynamic system while most previous approaches require access to the system's transition probabilities.
+Numerical results show that using AD to refine a coarse solution from an iterated filtering algorithm shows substantial improvement on current state-of-the-art methods on a challenging scientific benchmark problem.}
+\keywords{Sequential Monte Carlo, Automatic Differentiation, Particle Filter, Markov Process, Likelihood}
 }
 
-\arxiv{Many}{\dropcap{M}any}
-scientific models involve highly nonlinear stochastic dynamic systems possessing significant random variation in both the process dynamics and the measurements.
+\maketitle
+
+\section{Introduction}
+
+Many scientific models involve highly nonlinear stochastic dynamic systems possessing significant random variation in both the process dynamics and the measurements.
 Commonly, the latent system is modeled as a Markov process, giving rise to a partially observed Markov process (POMP) model, also known as a hidden Markov models or a state space model.
-POMP models arise in fields as diverse as optimal control \cite{singh22}, epidemiology \cite{he10, stocks17}, ecology \cite{knape12} and finance \cite{kim08, breto14}.
+POMP models arise in fields as diverse as optimal control \citep{singh22}, epidemiology \citep{he10, stocks17}, ecology \citep{knape12} and finance \citep{kim08, breto14}.
 Despite their ubiquity, estimation and inference within this broad class of models remains a challenging problem.
@@
 The particle filter, also known as sequential Monte Carlo, serves as the foundation for various inference algorithms for POMP models.
-It provides an unbiased estimate of the likelihood function  \cite{delMoral04}, enabling Bayesian inference \cite{andrieu10,chopin13} and likelihood-based inference \cite{ionides06-pnas,ionides15}.
-This is desirable, as likelihood-based inference is statistically efficient \cite{pawitan01} and robust to a moderate amount of Monte Carlo error \cite{ionides17,ning21}.
+It provides an unbiased estimate of the likelihood function  \citep{delMoral04}, enabling Bayesian inference \citep{andrieu10,chopin13} and likelihood-based inference \citep{ionides06-pnas,ionides15}.
+This is desirable, as likelihood-based inference is statistically efficient \citep{pawitan01} and robust to a moderate amount of Monte Carlo error \citep{ionides17,ning21}.
 Furthermore, basic particle filter algorithms, also known as boostrap filters, do not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
-This \textit{plug-and-play} property is useful for scientific applications \cite{he10}.
-Some plug-and-play methods depend on the construction of low-dimensional summary statistics \cite{wood10,toni09}, sacrificing statistical efficiency for computational convenience.
-Plug-and-play methods are often called likelihood-free \cite{owen15}, and it may be counter-intuitive that likelihood-based inference is possible using likelihood-free methods. 
-However, iterated filtering algorithms have shown that this is practical in various scientific investigations, such as \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
+This \textit{plug-and-play} property is useful for scientific applications \citep{he10}.
+Some plug-and-play methods depend on the construction of low-dimensional summary statistics \citep{wood10,toni09}, sacrificing statistical efficiency for computational convenience.
+Plug-and-play methods are often called likelihood-free \citep{owen15}, and it may be counter-intuitive that likelihood-based inference is possible using likelihood-free methods. 
+However, iterated filtering algorithms have shown that this is practical in various scientific investigations, such as \citep{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
 Nevertheless, the Monte Carlo variability arising in iterated filtering applications becomes increasingly problematic as the size of the data and the complexity of the model increase.
@@
 We examine plug-and-play AD as a tool to do so by providing a likelihood gradient.
-Recent advances in AD for particle filters \cite{naesseth18, jonschkowski18, corenflos21, scibior21, singh22} have drawn attention to AD as a tool for inference in POMP models.
-However, existing approaches are either asymptotically biased \cite{naesseth18, jonschkowski18}, have high variance \cite{poyiadjis11, scibior21}, are computationally expensive \cite{corenflos21, chen24}, or require access to transition densities \cite{poyiadjis11, scibior21, singh22, chen24}.
-Scibior and Wood \cite{scibior21} showed that the estimators derived by Poyiadjis et al.
-\cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.
+Recent advances in AD for particle filters \citep{naesseth18, jonschkowski18, corenflos21, scibior21, singh22} have drawn attention to AD as a tool for inference in POMP models.
+However, existing approaches are either asymptotically biased \citep{naesseth18, jonschkowski18}, have high variance \citep{poyiadjis11, scibior21}, are computationally expensive \citep{corenflos21, chen24}, or require access to transition densities \citep{poyiadjis11, scibior21, singh22, chen24}.
+\cite{scibior21} showed that the estimators derived by \cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.
 
-However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability \cite{poyiadjis11}.
+However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability \citep{poyiadjis11}.
 Additionally, while the stop-gradient procedure is undeniably effective, the reason why the correction recovers the \cite{poyiadjis11} estimator remains somewhat unclear.
@@
 In short, \textit{instead of directly differentiating through a basic particle filter, MOP uses AD to differentiate through a series of measurement density ratios and a differentiable simulator}. 
-Our MOP algorithm differs from previous smooth particle filters \cite{svensson18,malik11} by possessing the plug-and-play property.
+Our MOP algorithm differs from previous smooth particle filters \citep{svensson18,malik11} by possessing the plug-and-play property.
 Critically, we also avoid the high variance of \cite{svensson18} by generalizing the MOP particle filter to add a discount factor, $\alpha \in [0,1]$.
-This discount factor enables MOP-$\alpha$ to interpolate between the  biased gradient estimator of \cite{naesseth18}, when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21}, when $\alpha=1$ (Theorem \ref{thm:mop-functional-forms}).
+This discount factor enables MOP-$\alpha$ to interpolate between the  biased gradient estimator of \cite{naesseth18}, when $\alpha=0$, and the high-variance gradient estimate from \citep{poyiadjis11, scibior21}, when $\alpha=1$ (Theorem \ref{thm:mop-functional-forms}).
 The bias-variance tradeoff induced by $\alpha$ suggests, both in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}), a choice of $\alpha$ strictly in $(0,1)$.
@@
 In practice, gradient descent with MOP-$\alpha$ converges quickly within a neighborhood of the maximum but struggles to reach this neighborhood in the presence of local minima and saddle points.
-That behavior is complementary to that of iterated filtering algorithms \cite{ionides06-pnas,ionides15} which provide a relatively fast and stable way to identify this neighborhood.
+That behavior is complementary to that of iterated filtering algorithms \citep{ionides06-pnas,ionides15} which provide a relatively fast and stable way to identify this neighborhood.
 We therefore build a simple but effective hybrid algorithm called {\it Iterated Filtering with Automatic Differentiation} (IFAD) that warm-starts first-order or second-order gradient methods with a preliminary solution obtained from a few rounds of iterated filtering.
-Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 \cite{ionides06-pnas,ionides11} and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.
+Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 \citep{ionides06-pnas,ionides11} and the Liu-West filter \citep{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.
 
 These improvements also extend to Bayesian inference.
-In Section \ref{sec:bayes}, we use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with an empirical Bayes-type prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{andrieu10} on the Dhaka cholera model of \cite{king08} from the $700,000$ iterations in \cite{fasiolo16} to just $500$.
+In Section \ref{sec:bayes}, we use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \citep{homan14} in conjunction with an empirical Bayes-type prior estimated with IF2 to reduce the burn-in period of particle MCMC \citep{andrieu10} on the Dhaka cholera model of \cite{king08} from the $700,000$ iterations in \cite{fasiolo16} to just $500$.
 To the best of our knowledge, attaining rapid mixing on this challenging problem has not previously been achieved.
@@
 To that end, we suppose that the simulator is a differentiable function of $\theta$ for every fixed $\omega$, a condition that requires the latent process to be a continuous random variable. 
-This condition is also known as the {\it reparameterization trick} \cite{corenflos21}.
+This condition is also known as the {\it reparameterization trick} \citep{corenflos21}.
 The discrete resampling step is more delicate to address. 
@@
 The resampling indices, which we write as $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$, are a function of $\phi$ for any value of $\theta$.
-For the categorical distribution, we use systematic resampling \cite{arulampalam02,king16} which usually has superior performance to multinomial resampling.
+For the categorical distribution, we use systematic resampling \citep{arulampalam02,king16} which usually has superior performance to multinomial resampling.
 In practice, we only ever evaluate MOP-$\alpha$ in the case where $\theta$ and $\phi$ coincide.
@@
 \arxiv{}{\vspace*{-2mm}}
-\subsection{MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, naesseth18}}
+\subsection{MOP-$\alpha$ Encompasses the Estimators of \citep{poyiadjis11, scibior21, naesseth18}}
 
@@
   \centering
-    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{imgs/095/biasvar.png}
-    \caption{Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}.
-We display the MSE of score estimates for the trend in transmission, evaluated at the MLE.}
+    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2}}]{imgs/095/mop.png}
+    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2.2}}]{imgs/095/biasvar.png}
+    \caption{\textbf{Left:} Non‐differentiable likelihood estimates and our smooth off‐parameter reconstructions.
+\textbf{Right:} Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}.
+We display the MSE of score estimates for the trend in transmission at the MLE.}
     \label{fig:biasvar}
@@
 However, even when applying much larger amounts of Monte Carlo effort, IF2 fails to get within seven log-likelihood units of the maximum on this problem (Figure~\ref{fig:boxplot}).
-These numerical results are consistent with theoretical results for iterated filtering, showing slow convergence close to the maximum \cite{doucet15-if}.
+These numerical results are consistent with theoretical results for iterated filtering, showing slow convergence close to the maximum \citep{doucet15-if}.
 IF2, without assistance from AD, requires additional computationally intensive techniques, such as repeated pruning and profiling, to reach the inferentially critical region within one log unit of the maximum.
@@
 We therefore see that the second stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is well-conditioned in a neighborhood of the MLE and (2) the first stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
-Within a neighborhood of the MLE that grows as more data become available \cite{ning21}, local quadratic approximations such as local asymptotic normality apply under general conditions \cite{lecam00}.
+Within a neighborhood of the MLE that grows as more data become available \citep{ning21}, local quadratic approximations such as local asymptotic normality apply under general conditions \citep{lecam00}.
 This suggests that the linear convergence may occur reasonably often in practice.
@@
 
-The cholera transmission model that \cite{king08} developed for Dhaka, Bangladesh, has been used to benchmark the performance of various POMP inference methods \cite{ionides15, fasiolo16, wycoff24}, and we employ it here for the same purpose.
-This model categorizes individuals in a population as susceptible, $S(t)$, infected, $I(t)$, and recovered, $R(t)$ and so is called an SIR compartmental model.
+The cholera transmission model that \cite{king08} developed for Dhaka, Bangladesh, has been used to benchmark the performance of various POMP inference methods \citep{ionides15, fasiolo16, wycoff24}, and we employ it here for the same purpose.
+This model categorizes individuals in a population as susceptible, $S(t)$, infected, $I(t)$, and recovered, $R(t)$ and so is called an SIRS compartmental model.
 In this case, the compartment $R(t)$ is further subdivided into 
@@
 We write $P(t)$ for the total population, and $M_n$ for the cholera deaths in each month.
-As in \cite{king08, ionides15}, the transition dynamics follow a series of stochastic differential equations:
+As in \citep{king08, ionides15}, the transition dynamics follow a series of stochastic differential equations:
 \vspace*{-1mm}
@@
     \centering
-    \includegraphics[width=\arxiv{14cm}{\textwidth/2}]{imgs/095/tikzcholera.png}
+    \includegraphics[width=\arxiv{14cm}{\textwidth}]{imgs/095/tikzcholera.png}
     \vspace*{-7mm}
@@
 
-\paragraph{IFAD Successfully Finds the MLE:} Previously, an MLE at a log-likelihood of $-3748.6$ was reported by \cite{king16}.
+\subsection{IFAD Successfully Finds the MLE:} Previously, an MLE at a log-likelihood of $-3748.6$ was reported by \cite{king16}.
 This MLE was obtained with much computational effort, using many global and local IF2 searches and with the assistance of likelihood profiling, 8 years after the model was first proposed by \cite{king08}.
@@
 
-\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations, IF2 alone ultimately fails to achieve the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter}).
+\subsection{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations, IF2 alone ultimately fails to achieve the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter}).
 Conversely, we encountered many failed searches with gradient descent alone.
@@
 \begin{figure}[htbp!]
-    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{imgs/095/pairs.png}
-    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{imgs/095/qq.png}
+    \centering
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{2.2}}]{imgs/095/pairs.png}
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{2.2}}]{imgs/095/qq.png}
     %\arxiv{}{\vspace*{-1mm}}
@@
 \begin{figure}[ht]
-    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{imgs/095/boxplot.png}
-    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{imgs/095/boxplot_all.png}
+    \centering
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{2.2}}]{imgs/095/boxplot.png}
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{2.2}}]{imgs/095/boxplot_all.png}
     %\arxiv{}{\vspace*{-1mm}}
@@
     \centering
-    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{imgs/095/optim.png}
+    \includegraphics[width=\arxiv{10cm}{\textwidth/2}]{imgs/095/optim.png}
     \arxiv{}{\vspace*{-1mm}}
@@
 
-Particle MCMC \cite{andrieu10} is arguably the most popular method for full-information plug-and-play Bayesian inference for POMP models.
+Particle MCMC \citep{andrieu10} is arguably the most popular method for full-information plug-and-play Bayesian inference for POMP models.
 However, particle Metropolis-Hastings often mixes slowly, requiring long burn-in periods.
@@
 
-To alleviate this, we employ a No U-Turn Sampler (NUTS) \cite{homan14} powered by MOP-$\alpha$, with a nonparametric empirical prior constructed from density estimation of the parameter swarm from the IF2 warm-start.
+To alleviate this, we employ a No U-Turn Sampler (NUTS) \citep{homan14} powered by MOP-$\alpha$, with a nonparametric empirical prior constructed from density estimation of the parameter swarm from the IF2 warm-start.
 %This prior has a standard deviation of $6.3\%$ of the MLE. 
@@
     \centering
-    \includegraphics[width=\arxiv{\textwidth/\real{1.25}}{\textwidth/\real{2.5}}]{imgs/pmcmc/nuts_eb.png}
+    \includegraphics[width=\arxiv{\textwidth/\real{1.25}}{\textwidth}]{imgs/pmcmc/nuts_eb.png}
     \arxiv{}{\vspace*{-1mm}}
@@
 
-In particular, assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} imply that the POMP model satisfies a strong mixing condition, and as such are fairly strong but commonplace in the literature \cite{karjalainen23, delMoral04}.
+In particular, assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} imply that the POMP model satisfies a strong mixing condition, and as such are fairly strong but commonplace in the literature \citep{karjalainen23, delMoral04}.
 The fact that the likelihood estimate yielded by MOP-$\alpha$ has more than two continuous derivatives in $\theta$ follows from the construction of the likelihood estimate in equations \ref{eq:mop-conditional-likelihood} and \ref{eq:weighting-scheme}, as well as Assumptions \ref{assump:bounded-measurement} and \ref{assump:diff-meas-and-sim}.
@@
 
-Although we have shown the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} estimate the Fisher score by
+Although we have shown the MOP-$1$ gradient estimate yields the estimate of \citep{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} estimate the Fisher score by
 $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{0:n,j}^{A, F,\theta}, y_{1:N}^* ; \theta)$ and not
@@
         \\
-        &\Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) \lesssim \min_{k\leq N} Np  G'(\theta)^2\hspace{-0.8ex}\left(\hspace{-0.8ex}\frac{k^2}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}N\hspace{-0.8ex}\right).
+        &\Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) \lesssim \min_{k\leq N} Np  G'(\theta)^2\hspace{-0.8ex}\left(\hspace{-0.8ex}\frac{k^2}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}\hspace{-0.8ex}\right).
 \label{eq:mop-variance}
@@
 \arxiv{}{\vspace*{-1mm}}
-\paragraph{Interpretation:} Theorem \ref{thm:mop-biasvar} provides theoretical understanding of the empirically favorable bias-variance tradeoff enjoyed by MOP-$\alpha$ in Figure \ref{fig:biasvar}.
+\subsection{Interpretation:} Theorem \ref{thm:mop-biasvar} provides theoretical understanding of the empirically favorable bias-variance tradeoff enjoyed by MOP-$\alpha$ in Figure \ref{fig:biasvar}.
 The first term in the MSE bound in Equation \ref{eq:mop-mse} corresponds to particle approximation error, the second mixing error, and the third and fourth the error between the MOP-$(1,k)$ and MOP-$\alpha$ estimates.
@@
 \arxiv{}{\vspace*{-1mm}}
-\paragraph{Bias-Variance Tradeoff:} We show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( Np \, G'(\theta)^2\big/J \big)$.
+\subsection{Bias-Variance Tradeoff:} We show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( Np \, G'(\theta)^2\big/J \big)$.
 Previously, \cite{poyiadjis11} established that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p$ and $G'$. 
 Combining these results with Theorem \ref{thm:mop-biasvar} shows that MOP-$\alpha$ interpolates between the variances of MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$.
-The phase transition arises because even though the particle filter has forgetting properties \cite{karjalainen23}, the resulting derivative estimate of \cite{poyiadjis11} does not, and we require both for the variance reduction.
+The phase transition arises because even though the particle filter has forgetting properties \citep{karjalainen23}, the resulting derivative estimate of \cite{poyiadjis11} does not, and we require both for the variance reduction.
 
@@
 In our simulations, MOP-$\alpha$ took 3.75 times the runtime of the particle filter.
-MOP-$\alpha$ and IFAD therefore share the same $O(NJ)$ time complexity as the particle filter, unlike the $O(NJ^2)$ complexity of \cite{corenflos21} and Algorithm 2 in \cite{poyiadjis11, scibior21}.
+MOP-$\alpha$ and IFAD therefore share the same $O(NJ)$ time complexity as the particle filter, unlike the $O(NJ^2)$ complexity of \cite{corenflos21} and Algorithm 2 in \citep{poyiadjis11, scibior21}.
 
@@
 
-Practical likelihood-based data analysis involves many likelihood optimizations \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
+Practical likelihood-based data analysis involves many likelihood optimizations \citep{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
 This occurs with profile likelihood confidence intervals and model selection.
@@
 
-Our gradient estimate can be used in variational inference for the posterior over latent states \cite{naesseth18} and parameters.
+Our gradient estimate can be used in variational inference for the posterior over latent states \citep{naesseth18} and parameters.
 
@@
 
-\arxiv{
 \section*{Acknowledgments}
-This research was supported by National Science Foundation grant DMS-1761603. 
+This research was supported by National Science Foundation grants DMS-1761603 and DEB-1933497. 
 We thank Nicolas Chopin for helpful communications regarding the strong law of large numbers for off-parameter resampled particle filters.
-We also thank Arnaud Doucet for helpful discussion regarding the manuscript.
-}{
-\showmatmethods{} % Display the Materials and Methods section
-\acknow{
-This research was supported by National Science Foundation grant DMS-1761603. 
-We thank Nicolas Chopin for helpful communications regarding the strong law of large numbers for off-parameter resampled particle filters.
+We also thank Jun Chen and Arnaud Doucet for helpful discussions regarding the manuscript.
+% }{
+% \showmatmethods{} % Display the Materials and Methods section
+% \acknow{
+% This research was supported by National Science Foundation grants DMS-1761603 and DEB-1933497. 
+% We thank Nicolas Chopin for helpful communications regarding the strong law of large numbers for off-parameter resampled particle filters.
 We also thank Jun Chen and Arnaud Doucet for helpful discussions regarding the manuscript.}
-\showacknow{} % Display the acknowledgments section
-}
+% \showacknow{} % Display the acknowledgments section
+% }
 
@@
 
-% \bibliographystyle{pnas-new}
+\bibliographystyle{abbrvnat}
 % Bibliography

### 2024-10-10 paper/oup-authoring-template.tex
updates to GitHub

%%
%% Copyright 2024 OXFORD UNIVERSITY PRESS
%%
%% This file is forl of the 'oup-authoring-template Bundle' for the Royal Statistical Society.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'oup-authoring-template Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
%% with bibliographic references
%%

\documentclass[numsec,webpdf,modern,medium,namedate]{oup-authoring-template}

\onecolumn

%\usepackage{showframe}

\graphicspath{{Fig/}}

% line numbers
%\usepackage[mathlines, switch]{lineno}
%\usepackage[right]{lineno}

\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}
\usepackage[doublespacing]{setspace}
\usepackage[fontsize=12pt]{fontsize}

\begin{document}

\journaltitle{Journals of the Royal Statistical Society}
\DOI{DOI HERE}
\copyrightyear{XXXX}
\pubyear{XXXX}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Original article}

\firstpage{1}

%\subtitle{Subject Section}

\title[Short Article Title]{Article Title}

\author[1,$\ast$]{First Author}
\author[2]{Second Author}
\author[3]{Third Author}
\author[3]{Fourth Author}
\author[4]{Fifth Author\ORCID{0000-0000-0000-0000}}

\authormark{Author Name et al.}

\address[1]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \postcode{Postcode}, \state{State}, \country{Country}}}
\address[2]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \postcode{Postcode}, \state{State}, \country{Country}}}
\address[3]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \postcode{Postcode}, \state{State}, \country{Country}}}
\address[4]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \postcode{Postcode}, \state{State}, \country{Country}}}

\corresp[$\ast$]{Address for correspondence. Name, institution, city, zip code, country. \href{Email:email-id.com}{email-id.com}}

\received{Date}{0}{Year}
\revised{Date}{0}{Year}
\accepted{Date}{0}{Year}

%\editor{Associate Editor: Name}

%\abstract{
%\textbf{Motivation:} .\\
%\textbf{Results:} .\\
%\textbf{Availability:} .\\
%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Journal Name}
%online.}

\abstract{Abstracts must be able to stand alone and so cannot contain citations to
the paper's references, equations, etc. An abstract must consist of a single
paragraph and be concise. Because of online formatting, abstracts must appear
as plain as possible.}
\keywords{keyword1, Keyword2, Keyword3, Keyword4}

% \boxedtext{
% \begin{itemize}
% \item Key boxed text here.
% \item Key boxed text here.
% \item Key boxed text here.
% \end{itemize}}

\maketitle

\section{Introduction}
The introduction introduces the context and summarizes the manuscript. It is importantly to clearly state the contributions
of this piece of work. Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

This is an example of a new parapgraph with a numbered footnote\footnote{\url{https://data.gov.uk/}} and a second footnote marker.\footnote{Example of footnote text.}

\section{This is an example for first level head - section head}\label{sec2}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum (refer Section~\ref{sec5}).

\subsection{This is an example for second level head - subsection head}\label{subsec1}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\subsubsection{This is an example for third level head - subsubsection head}\label{subsubsec1}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. %Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\paragraph{This is an example for fourth level head - paragraph head}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\section{This is an example for first level head}\label{sec3}

\subsection{This is an example for second level head - subsection head}\label{subsec2}

\subsubsection{This is an example for third level head - subsubsection head}\label{subsubsec2}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\paragraph{This is an example for fourth level head - paragraph head}

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\section{Equations}\label{sec4}

Equations in \LaTeX{} can either be inline or set as display equations. For
inline equations use the \verb+$...$+ commands. Eg: the equation
$H\psi = E \psi$ is written via the command \verb+$H \psi = E \psi$+.

For display equations (with auto generated equation numbers)
one can use the equation or eqnarray environments:
\begin{equation}
\|\tilde{X}(k)\|^2 \leq\frac{\sum\limits_{i=1}^{p}\left\|\tilde{Y}_i(k)\right\|^2+\sum\limits_{j=1}^{q}\left\|\tilde{Z}_j(k)\right\|^2 }{p+q},\label{eq1}
\end{equation}
where,
\begin{align}
D_\mu &=  \partial_\mu - ig \frac{\lambda^a}{2} A^a_\mu \nonumber \\
F^a_{\mu\nu} &= \partial_\mu A^a_\nu - \partial_\nu A^a_\mu + g f^{abc} A^b_\mu A^a_\nu.\label{eq2}
\end{align}
Notice the use of \verb+\nonumber+ in the align environment at the end
of each line, except the last, so as not to produce equation numbers on
lines where no equation numbers are required. The \verb+\label{}+ command
should only be used at the last line of an align environment where
\verb+\nonumber+ is not used.
\begin{equation}
Y_\infty = \left( \frac{m}{\textrm{GeV}} \right)^{-3}
    \left[ 1 + \frac{3 \ln(m/\textrm{GeV})}{15}
    + \frac{\ln(c_2/5)}{15} \right].
\end{equation}
The class file also supports the use of \verb+\mathbb{}+, \verb+\mathscr{}+ and
\verb+\mathcal{}+ commands. As such \verb+\mathbb{R}+, \verb+\mathscr{R}+
and \verb+\mathcal{R}+ produces $\mathbb{R}$, $\mathscr{R}$ and $\mathcal{R}$
respectively (refer Subsubsection~\ref{subsubsec3}).

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum. Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum. 
Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.

\section{Tables}\label{sec5}

Tables can be inserted via the normal table and tabular environment. To put
footnotes inside tables one has to Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum. use the additional ``tablenotes" environment
enclosing the tabular environment. The footnote appears just below the table
itself (refer Tables~\ref{tab1} and \ref{tab2}).

\begin{verbatim}
\begin{table}[t]
\begin{center}
\begin{minipage}{<width>}
\caption{<table-caption>\label{<table-label>}}%
\begin{tabular}{@{}llll@{}}
\toprule
column 1 & column 2 & column 3 & column 4\\
\midrule
row 1 & data 1 & data 2          & data 3 \\
row 2 & data 4 & data 5$^{1}$ & data 6 \\
row 3 & data 7 & data 8      & data 9$^{2}$\\
\botrule
\end{tabular}
\begin{tablenotes}%
\item Source: Example for source.
\item[$^{1}$] Example for a 1st table footnote.
\item[$^{2}$] Example for a 2nd table footnote.
\end{tablenotes}
\end{minipage}
\end{center}
\end{table}
\end{verbatim}

Lengthy tables which do not fit within textwidth should be set as rotated tables. For this, we need to use \verb+\begin{sidewaystable}...+ \verb+\end{sidewaystable}+ instead of\break \verb+\begin{table}...+ \verb+\end{table}+ environment.

\begin{table}[!t]
\caption{Caption text\label{tab1}}%
\begin{tabular*}{\columnwidth}{@{\extracolsep\fill}llll@{\extracolsep\fill}}
\toprule
column 1 & column 2  & column 3 & column 4\\
\midrule
row 1    & data 1   & data 2  & data 3  \\
row 2    & data 4   & data 5$^{1}$  & data 6  \\
row 3    & data 7   & data 8  & data 9$^{2}$  \\
\botrule
\end{tabular*}
\begin{tablenotes}%
\item Source: This is an example of table footnote this is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote
\item[$^{1}$] Example for a first table footnote.
\item[$^{2}$] Example for a second table footnote.
\end{tablenotes}
\end{table}

\begin{table*}[t]
\caption{Example of a lengthy table which is set to full textwidth.\label{tab2}}
\tabcolsep=0pt%%
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccccc@{\extracolsep{\fill}}}
\toprule%
& \multicolumn{3}{@{}c@{}}{Element 1$^{1}$} & \multicolumn{3}{@{}c@{}}{Element 2$^{2}$} \\
\cline{2-4}\cline{5-7}%
Project & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
\midrule
Element 3  & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$\\
Element 4  & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$\\
\botrule
\end{tabular*}
\begin{tablenotes}%
\item Note: This is an example of table footnote this is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote
\item[$^{1}$] Example for a first table footnote.
\item[$^{2}$] Example for a second table footnote.\vspace*{6pt}
\end{tablenotes}
\end{table*}

\begin{sidewaystable}%[!p]
\caption{Tables which are too long to fit, should be written using the ``sidewaystable" environment as shown here\label{tab3}}
\tabcolsep=0pt%
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccccc@{\extracolsep{\fill}}}
\toprule%
& \multicolumn{3}{@{}c@{}}{Element 1$^{1}$}& \multicolumn{3}{@{}c@{}}{Element$^{2}$} \\
\cline{2-4}\cline{5-7}%
Projectile & Energy     & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
\midrule
Element 3 & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$ \\
Element 4 & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$ \\
\botrule
\end{tabular*}
\begin{tablenotes}%
\item Note: This is an example of a table footnote this is an example of a table footnote this is an example of a table footnote this is an example of a table footnote this is an example of a table footnote
\item[$^{1}$] This is an example of a table footnote
\end{tablenotes}
\end{sidewaystable}

\section{Figures}\label{sec6}

As per display \LaTeX\ standards one has to use eps images for \verb+latex+ compilation and \verb+pdf/jpg/png+ images for
\verb+pdflatex+ compilation. This is one of the major differences between \verb+latex+
and \verb+pdflatex+. The images should be single-page documents. The command for inserting images
for \verb+latex+ and \verb+pdflatex+ can be generalized. The package used to insert images in \verb+latex/pdflatex+ is the
graphicx package. Figures can be inserted via the normal figure environment as shown in the below example:

\begin{figure}[!t]%
\centering
{\color{black!20}\rule{213pt}{37pt}}
\caption{This is a widefig. This is an example of a long caption this is an example of a long caption  this is an example of a long caption this is an example of a long caption}\label{fig1}
\end{figure}

\begin{figure*}[!t]%
\centering
{\color{black!20}\rule{438pt}{74pt}}
\caption{This is a widefig. This is an example of a long caption this is an example of a long caption  this is an example of a long caption this is an example of a long caption}\label{fig2}
\end{figure*}

\begin{verbatim}
\begin{figure}[t]
        \centering\includegraphics{<eps-file>}
        \caption{<figure-caption>}
        \label{<figure-label>}
\end{figure}
\end{verbatim}

Test text here.

For sample purposes, we have included the width of images in the
optional argument of \verb+\includegraphics+ tag. Please ignore this.
Lengthy figures which do not fit within textwidth should be set in rotated mode. For rotated figures, we need to use \verb+\begin{sidewaysfigure}+ \verb+...+ \verb+\end{sidewaysfigure}+ instead of the \verb+\begin{figure}+ \verb+...+ \verb+\end{figure}+ environment.

\begin{sidewaysfigure}%
\centering
{\color{black!20}\rule{610pt}{102pt}}
\caption{This is an example for a sideways figure. This is an example of a long caption this is an example of a long caption  this is an example of a long caption this is an example of a long caption}\label{fig3}
\end{sidewaysfigure}

\section{Algorithms, Program codes and Listings}\label{sec7}

Packages \verb+algorithm+, \verb+algorithmicx+ and \verb+algpseudocode+ are used for setting algorithms in latex.
For this, one has to use the below format:

\begin{verbatim}
\begin{algorithm}
\caption{<alg-caption>}\label{<alg-label>}
\begin{algorithmic}[1]
. . .
\end{algorithmic}
\end{algorithm}
\end{verbatim}

You may need to refer to the above-listed package documentations for more details before setting an \verb+algorithm+ environment.
To set program codes, one has to use the \verb+program+ package. We need to use the \verb+\begin{program}+ \verb+...+
\verb+\end{program}+ environment to set program codes.

\begin{algorithm}[!t]
\caption{Calculate $y = x^n$}\label{algo1}
\begin{algorithmic}[1]
\Require $n \geq 0 \vee x \neq 0$
\Ensure $y = x^n$
\State $y \Leftarrow 1$
\If{$n < 0$}
        \State $X \Leftarrow 1 / x$
        \State $N \Leftarrow -n$
\Else
        \State $X \Leftarrow x$
        \State $N \Leftarrow n$
\EndIf
\While{$N \neq 0$}
        \If{$N$ is even}
            \State $X \Leftarrow X \times X$
            \State $N \Leftarrow N / 2$
        \Else[$N$ is odd]
            \State $y \Leftarrow y \times X$
            \State $N \Leftarrow N - 1$
        \EndIf
\EndWhile
\end{algorithmic}
\end{algorithm}

Similarly, for \verb+listings+, one has to use the \verb+listings+ package. The \verb+\begin{lstlisting}+ \verb+...+ \verb+\end{lstlisting}+ environment is used to set environments similar to the \verb+verbatim+ environment. Refer to the \verb+lstlisting+ package documentation for more details on this.

\begin{minipage}{\hsize}%
\lstset{language=Pascal}% Set your language (you can change the language for each code-block optionally)
\begin{lstlisting}[frame=single,framexleftmargin=-1pt,framexrightmargin=-17pt,framesep=12pt,linewidth=0.98\textwidth]
for i:=maxint to 0 do
begin
{ do nothing }
end;
Write('Case insensitive ');
Write('Pascal keywords.');
\end{lstlisting}
\end{minipage}

\section{Cross referencing}\label{sec8}

Environments such as figure, table, equation, and align can have a label
declared via the \verb+\label{#label}+ command. For figures and table
environments one should use the \verb+\label{}+ command inside or just
below the \verb+\caption{}+ command.  One can then use the
\verb+\ref{#label}+ command to cross-reference them. As an example, consider
the label declared for Figure \ref{fig1} which is
\verb+\label{fig1}+. To cross-reference it, use the command
\verb+ Figure \ref{fig1}+, for which it comes up as
``Figure~\ref{fig1}".

\subsection{Details on reference citations}\label{subsec3}

With standard numerical .bst files, only numerical citations are possible.
With an author-year .bst file, both numerical and author-year citations are possible.

If author-year citations are selected, \verb+\bibitem+ must have one of the following forms:

{\footnotesize%
\begin{verbatim}
\bibitem[Jones et al.(1990)]{key}...
\bibitem[Jones et al.(1990)Jones,
                Baker, and Williams]{key}...
\bibitem[Jones et al., 1990]{key}...
\bibitem[\protect\citeauthoryear{Jones,
                Baker, and Williams}
                {Jones et al.}{1990}]{key}...
\bibitem[\protect\citeauthoryear{Jones et al.}
                {1990}]{key}...
\bibitem[\protect\astroncite{Jones et al.}
                {1990}]{key}...
\bibitem[\protect\citename{Jones et al., }
                1990]{key}...
\harvarditem[Jones et al.]{Jones, Baker, and
                Williams}{1990}{key}...
\end{verbatim}}

This is either to be made up manually, or to be generated by an
appropriate .bst file with BibTeX. Then,

{%
\begin{verbatim}
                    Author-year mode
                        || Numerical mode
\citet{key} ==>>  Jones et al. (1990)
                        || Jones et al. [21]
\citep{key} ==>> (Jones et al., 1990) || [21]
\end{verbatim}}

\noindent
Multiple citations as normal:

{%
\begin{verbatim}
\citep{key1,key2} ==> (Jones et al., 1990;
                         Smith, 1989)||[21,24]
        or (Jones et al., 1990, 1991)||[21,24]
        or (Jones et al., 1990a,b)   ||[21,24]
\end{verbatim}}

\noindent
\verb+\cite{key}+ is the equivalent of \verb+\citet{key}+ in author-year mode
and  of \verb+\citep{key}+ in numerical mode. Full author lists may be forced with
\verb+\citet*+ or \verb+\citep*+, e.g.

{%
\begin{verbatim}
\citep*{key} ==>> (Jones, Baker, and Mark, 1990)
\end{verbatim}}

\noindent
Optional notes as:

{%
\begin{verbatim}
\citep[chap. 2]{key}     ==>>
        (Jones et al., 1990, chap. 2)
\citep[e.g.,][]{key}     ==>>
        (e.g., Jones et al., 1990)
\citep[see][pg. 34]{key} ==>>
        (see Jones et al., 1990, pg. 34)
\end{verbatim}}

\noindent
(Note: in standard LaTeX, only one note is allowed, after the ref.
Here, one note is like the standard, two make pre- and post-notes.)

{%
\begin{verbatim}
\citealt{key}   ==>> Jones et al. 1990
\citealt*{key}  ==>> Jones, Baker, and
                        Williams 1990
\citealp{key}   ==>> Jones et al., 1990
\citealp*{key}  ==>> Jones, Baker, and
                        Williams, 1990
\end{verbatim}}

\noindent
Additional citation possibilities (both author-year and numerical modes):

{%
\begin{verbatim}
\citeauthor{key}       ==>> Jones et al.
\citeauthor*{key}      ==>> Jones, Baker, and
                                Williams
\citeyear{key}         ==>> 1990
\citeyearpar{key}      ==>> (1990)
\citetext{priv. comm.} ==>> (priv. comm.)
\citenum{key}          ==>> 11 [non-superscripted]
\end{verbatim}}

\noindent
Note: full author lists depend on whether the bib style supports them;
if not, the abbreviated list is printed even when full is requested.

\noindent
For names like della Robbia at the start of a sentence, use

{%
\begin{verbatim}
\Citet{dRob98}      ==>> Della Robbia (1998)
\Citep{dRob98}      ==>> (Della Robbia, 1998)
\Citeauthor{dRob98} ==>> Della Robbia
\end{verbatim}}

\noindent
The following is an example for \verb+\cite{...}+: \cite{rahman2019centroidb}. Another example for \verb+\citep{...}+: \citep{bahdanau2014neural,imboden2018cardiorespiratory,motiian2017unified,murphy2012machine,ji20123d}.

Sample cites here \citet{krizhevsky2012imagenet,horvath2018dna} and \citet{pyrkov2018quantitative}, \citep{wang2018face}, \citep{lecun2015deep,zhang2018fine,ravi2016deep}.

\section{Lists}\label{sec9}

List in \LaTeX{} can be of three types: numbered, bulleted and unnumbered. The ``enumerate'' environment produces a numbered list, the 
``itemize'' environment produces a bulleted list and the ``unlist''
environment produces an unnumbered list.
In each environment, a new entry is added via the \verb+\item+ command.
\begin{enumerate}[1.]
\item This is the 1st item

\item Enumerate creates numbered lists, itemize creates bulleted lists and
unnumerate creates unnumbered lists.
\begin{enumerate}[(a)]
\item Second level numbered list. Enumerate creates numbered lists, itemize creates bulleted lists and
description creates unnumbered lists.

\item Second level numbered list. Enumerate creates numbered lists, itemize creates bulleted lists and
description creates unnumbered lists.
\begin{enumerate}[(ii)]
\item Third level numbered list. Enumerate creates numbered lists, itemize creates bulleted lists and
description creates unnumbered lists.

\item Third level numbered list. Enumerate creates numbered lists, itemize creates bulleted lists and
description creates unnumbered lists.
\end{enumerate}

\item Second level numbered list. Enumerate creates numbered lists, itemize creates bulleted lists and
description creates unnumbered lists.

\item Second level numbered list. Enumerate creates numbered lists, itemize creates bulleted lists and
description creates unnumbered lists.
\end{enumerate}

\item Enumerate creates numbered lists, itemize creates bulleted lists and
description creates unnumbered lists.

\item Numbered lists continue.
\end{enumerate}
Lists in \LaTeX{} can be of three types: enumerate, itemize and description.
In each environment, a new entry is added via the \verb+\item+ command.
\begin{itemize}
\item First level bulleted list. This is the 1st item

\item First level bulleted list. Itemize creates bulleted lists and description creates unnumbered lists.
\begin{itemize}
\item Second level dashed list. Itemize creates bulleted lists and description creates unnumbered lists.

\item Second level dashed list. Itemize creates bulleted lists and description creates unnumbered lists.

\item Second level dashed list. Itemize creates bulleted lists and description creates unnumbered lists.
\end{itemize}

\item First level bulleted list. Itemize creates bulleted lists and description creates unnumbered lists.

\item First level bulleted list. Bullet lists continue.
\end{itemize}

\noindent
Example for unnumbered list items:

\begin{unlist}
\item Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text.

\item Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text.

\item sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text. Sample unnumberd list text.
\end{unlist}

\section{Examples for theorem-like environments}\label{sec10}

For theorem-like environments, we require the \verb+amsthm+ package. There are three types of predefined theorem styles - \verb+thmstyleone+, \verb+thmstyletwo+ and \verb+thmstylethree+   (check your journal's instructions page in case a specific style is required).

\medskip
\noindent\begin{tabular}{|l|p{13pc}|}
\hline
\verb+thmstyleone+ & Numbered, theorem head in bold font and theorem text in italic style \\\hline
\verb+thmstyletwo+ & Numbered, theorem head in roman font and theorem text in italic style \\\hline
\verb+thmstylethree+ & Numbered, theorem head in bold font and theorem text in roman style \\\hline
\end{tabular}

\begin{theorem}[Theorem subhead]\label{thm1}
Example theorem text. Example theorem text. Example theorem text. Example theorem text. Example theorem text.
Example theorem text. Example theorem text. Example theorem text. Example theorem text. Example theorem text.
Example theorem text.
\end{theorem}

Quisque ullamcorper placerat ipsum. Cras nibh. Morbi vel justo vitae lacus tincidunt ultrices. Lorem ipsum dolor sit
amet, consectetuer adipiscing elit. In hac habitasse platea dictumst. Integer tempus convallis augue.

\begin{proposition}
Example proposition text. Example proposition text. Example proposition text. Example proposition text. Example proposition text.
Example proposition text. Example proposition text. Example proposition text. Example proposition text. Example proposition text.
\end{proposition}

Nulla malesuada porttitor diam. Donec felis erat, congue non, volutpat at, tincidunt tristique, libero. Vivamus
viverra fermentum felis. Donec nonummy pellentesque ante.

\begin{example}
Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem.
\end{example}

Nulla malesuada porttitor diam. Donec felis erat, congue non, volutpat at, tincidunt tristique, libero. Vivamus
viverra fermentum felis. Donec nonummy pellentesque ante.

\begin{remark}
Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem.
\end{remark}

Quisque ullamcorper placerat ipsum. Cras nibh. Morbi vel justo vitae lacus tincidunt ultrices. Lorem ipsum dolor sit
amet, consectetuer adipiscing elit. In hac habitasse platea dictumst.

\begin{definition}[Definition sub head]
Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text. Example definition text.
\end{definition}

Apart from the above styles, we have the \verb+\begin{proof}+ \verb+...+ \verb+\end{proof}+ environment - with the proof head in italic style and the body text in roman font with an open square at the end of each proof environment.

\begin{proof}Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text.
\end{proof}

Nam dui ligula, fringilla a, euismod sodales, sollicitudin vel, wisi. Morbi auctor lorem non justo. Nam lacus libero,
pretium at, lobortis vitae, ultricies et, tellus. Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna,
vitae ornare odio metus a mi.

\begin{proof}[Proof of Theorem~{\upshape\ref{thm1}}]
Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text. Example for proof text.
\end{proof}

\noindent
For a quote environment, one has to use\newline \verb+\begin{quote}...\end{quote}+
\begin{quote}
Quoted text example. Aliquam porttitor quam a lacus. Praesent vel arcu ut tortor cursus volutpat. In vitae pede quis diam bibendum placerat. Fusce elementum
convallis neque. Sed dolor orci, scelerisque ac, dapibus nec, ultricies ut, mi. Duis nec dui quis leo sagittis commodo.
\end{quote}
Donec congue. Maecenas urna mi, suscipit in, placerat ut, vestibulum ut, massa. Fusce ultrices nulla et nisl (refer Figure~\ref{fig3}). Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas. Etiam ligula arcu,
elementum a, venenatis quis, sollicitudin sed, metus. Donec nunc pede, tincidunt in, venenatis vitae, faucibus vel (refer Table~\ref{tab3}).

\section{Conclusion}

Some Conclusions here.

%%%%%%%%%%%%%%

\begin{appendices}

\section{Section title of first appendix}\label{sec11}

Nam dui ligula, fringilla a, euismod sodales, sollicitudin vel, wisi. Morbi auctor lorem non justo. Nam lacus libero,
pretium at, lobortis vitae, ultricies et, tellus. Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna,
vitae ornare odio metus a mi. Morbi ac orci et nisl hendrerit mollis. Suspendisse ut massa. Cras nec ante. Pellentesque
a nulla. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Aliquam tincidunt
urna. Nulla ullamcorper vestibulum turpis. Pellentesque cursus luctus mauris.

\subsection{Subsection title of first appendix}\label{subsec4}

Nam dui ligula, fringilla a, euismod sodales, sollicitudin vel, wisi. Morbi auctor lorem non justo. Nam lacus libero,
pretium at, lobortis vitae, ultricies et, tellus. Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna,
vitae ornare odio metus a mi. Morbi ac orci et nisl hendrerit mollis. Suspendisse ut massa. Cras nec ante. Pellentesque
a nulla. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Aliquam tincidunt
urna. Nulla ullamcorper vestibulum turpis. Pellentesque cursus luctus mauris.

\subsubsection{Subsubsection title of first appendix}\label{subsubsec3}

Example for an unnumbered figure:

\begin{figure}[!h]
\centering
{\color{black!20}\rule{85pt}{92pt}}
\end{figure}

Fusce mauris. Vestibulum luctus nibh at lectus. Sed bibendum, nulla a faucibus semper, leo velit ultricies tellus, ac
venenatis arcu wisi vel nisl. Vestibulum diam. Aliquam pellentesque, augue quis sagittis posuere, turpis lacus congue
quam, in hendrerit risus eros eget felis.

\section{Section title of second appendix}\label{sec12}%

Fusce mauris. Vestibulum luctus nibh at lectus. Sed bibendum, nulla a faucibus semper, leo velit ultricies tellus, ac
venenatis arcu wisi vel nisl. Vestibulum diam. Aliquam pellentesque, augue quis sagittis posuere, turpis lacus congue
quam, in hendrerit risus eros eget felis. Maecenas eget erat in sapien mattis porttitor. Vestibulum porttitor. Nulla
facilisi. Sed a turpis eu lacus commodo facilisis. Morbi fringilla, wisi in dignissim interdum, justo lectus sagittis dui, et
vehicula libero dui cursus dui. Mauris tempor ligula sed lacus. Duis cursus enim ut augue. Cras ac magna. Cras nulla.

\begin{figure}[b]
\centering
{\color{black!20}\rule{217pt}{120pt}}
\caption{This is an example for appendix figure\label{fig4}}
\end{figure}

\begin{table}[t]%
\begin{center}
\begin{minipage}{.52\columnwidth}
\caption{This is an example of Appendix table showing food requirements of army, navy and airforce\label{tab4}}%
\begin{tabular}{@{}lcc@{}}%
\toprule
col1 head & col2 head & col3 head \\
\midrule
col1 text & col2 text & col3 text \\
col1 text & col2 text & col3 text \\
col1 text & col2 text & col3 text \\
\botrule
\end{tabular}
\end{minipage}
\end{center}
\end{table}

\subsection{Subsection title of second appendix}\label{subsec5}

Sed commodo posuere pede. Mauris ut est. Ut quis purus. Sed ac odio. Sed vehicula hendrerit sem. Duis non odio.
Morbi ut dui. Sed accumsan risus eget odio. In hac habitasse platea dictumst. Pellentesque non elit. Fusce sed justo
eu urna porta tincidunt. Mauris felis odio, sollicitudin sed, volutpat a, ornare ac, erat. Morbi quis dolor. Donec
pellentesque, erat ac sagittis semper, nunc dui lobortis purus, quis congue purus metus ultricies tellus. Proin et quam.
Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos hymenaeos. Praesent sapien turpis,
fermentum vel, eleifend faucibus, vehicula eu, lacus.

Sed commodo posuere pede. Mauris ut est. Ut quis purus. Sed ac odio. Sed vehicula hendrerit sem. Duis non odio.
Morbi ut dui. Sed accumsan risus eget odio. In hac habitasse platea dictumst. Pellentesque non elit. Fusce sed justo
eu urna porta tincidunt. Mauris felis odio, sollicitudin sed, volutpat a, ornare ac, erat. Morbi quis dolor. Donec
pellentesque, erat ac sagittis semper, nunc dui lobortis purus, quis congue purus metus ultricies tellus. Proin et quam.
Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos hymenaeos. Praesent sapien turpis,
fermentum vel, eleifend faucibus, vehicula eu, lacus.

\subsubsection{Subsubsection title of second appendix}\label{subsubsec4}

Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae,
felis. Curabitur dictum gravida mauris. Nam arcu libero, nonummy eget, consectetuer id, vulputate a, magna. Donec
vehicula augue eu neque.

Example for an equation inside the appendix:
\begin{align}
 p &= \frac{\gamma^{2} - (n_{C} -1)H}{(n_{C} - 1) + H - 2\gamma}, \label{1eq:hybobo:pfromgH} \\
 \theta &= \frac{(\gamma - H)^{2}(\gamma - n_{C} -1)^{2}}{(n_{C} - 1 + H - 2\gamma)^{2}} \label{2eq:hybobo:tfromgH}\; .
\end{align}

\section{Example of another appendix section}\label{sec13}%

Nam dui ligula, fringilla a, euismod sodales, sollicitudin vel, wisi. Morbi auctor lorem non justo. Nam lacus libero,
pretium at, lobortis vitae, ultricies et, tellus. Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna,
vitae ornare odio metus a mi. Morbi ac orci et nisl hendrerit mollis. Suspendisse ut massa. Cras nec ante. Pellentesque
a nulla. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus. Aliquam tincidunt
urna. Nulla ullamcorper vestibulum turpis. Pellentesque cursus luctus mauris
\begin{equation}
\mathcal{L} = i \bar{\psi} \gamma^\mu D_\mu \psi
    - \frac{1}{4} F_{\mu\nu}^a F^{a\mu\nu} - m \bar{\psi} \psi.
\label{eq26}
\end{equation}

Nulla malesuada porttitor diam. Donec felis erat, congue non, volutpat at, tincidunt tristique, libero. Vivamus
viverra fermentum felis. Donec nonummy pellentesque ante. Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. Sed lacinia
nulla vitae enim. Pellentesque tincidunt purus vel magna. Integer non enim. Praesent euismod nunc eu purus. Donec
bibendum quam in tellus. Nullam cursus pulvinar lectus. Donec et mi. Nam vulputate metus eu enim. Vestibulum
pellentesque felis eu massa.

Nulla malesuada porttitor diam. Donec felis erat, congue non, volutpat at, tincidunt tristique, libero. Vivamus
viverra fermentum felis. Donec nonummy pellentesque ante. Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. Sed lacinia
nulla vitae enim. Pellentesque tincidunt purus vel magna. Integer non enim. Praesent euismod nunc eu purus. Donec
bibendum quam in tellus. Nullam cursus pulvinar lectus. Donec et mi. Nam vulputate metus eu enim. Vestibulum
pellentesque felis eu massa.

Nulla malesuada porttitor diam. Donec felis erat, congue non, volutpat at, tincidunt tristique, libero. Vivamus
viverra fermentum felis. Donec nonummy pellentesque ante. Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. Sed lacinia
nulla vitae enim. Pellentesque tincidunt purus vel magna. Integer non enim. Praesent euismod nunc eu purus. Donec
bibendum quam in tellus. Nullam cursus pulvinar lectus. Donec et mi. Nam vulputate metus eu enim. Vestibulum
pellentesque felis eu massa.

Nulla malesuada porttitor diam. Donec felis erat, congue non, volutpat at, tincidunt tristique, libero. Vivamus
viverra fermentum felis. Donec nonummy pellentesque ante. Phasellus adipiscing semper elit. Proin fermentum massa
ac quam. Sed diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas lacinia. Nam ipsum ligula, eleifend
at, accumsan nec, suscipit a, ipsum. Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem. Sed lacinia
nulla vitae enim. Pellentesque tincidunt purus vel magna. Integer non enim. Praesent euismod nunc eu purus. Donec
bibendum quam in tellus. Nullam cursus pulvinar lectus. Donec et mi. Nam vulputate metus eu enim. Vestibulum
pellentesque felis eu massa. Donec
bibendum quam in tellus. Nullam cursus pulvinar lectus. Donec et mi. Nam vulputate metus eu enim. Vestibulum
pellentesque felis eu massa.

%% Example for unnumbered table inside appendix
\begin{table}
\begin{center}
\begin{minipage}{.52\columnwidth}
\caption{}{%
\begin{tabular}{lcc}%
\toprule
col1 head & col2 head & col3 head \\
\midrule
col1 text & col2 text & col3 text \\
col1 text & col2 text & col3 text \\
col1 text & col2 text & col3 text \\
\botrule
\end{tabular}}{}
\end{minipage}
\end{center}
\end{table}

\end{appendices}

\section{Competing interests}
No competing interest is declared.

\section{Author contributions statement}

Must include all authors, identified by initials, for example:
S.R. and D.A. conceived the experiment(s),  S.R. conducted the experiment(s), S.R. and D.A. analysed the results.  S.R. and D.A. wrote and reviewed the manuscript.

\section{Acknowledgments}
The authors thank the anonymous reviewers for their valuable suggestions. This work is supported in part by funds from the National Science Foundation (NSF: \# 1636933 and \# 1920920).

\bibliographystyle{abbrvnat}
\bibliography{reference}

\end{document}

### 2024-10-10 poster/poster.tex
updates to GitHub

@@
 \usepackage[T1]{fontenc}
+%\renewcommand{\thefootnote}{\arabic{footnote}}
 \usepackage{boldline}
+\usepackage{color}
+%\usepackage[symbol]{footmisc}
 \usepackage{lmodern}
@@
 \begin{tcolorbox}[enhanced,colback=white!100!white,colframe=red!100!red]
-\textbf{\textcolor{red}{Partially-Observed Markov Process (POMP)}}: 
-\begin{itemize}
-    \item Unknown Markov process $\{X_t, t \geq t_0\}$, but we have observations $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$.  
-    \item Given measurements $y_n^*$, estimate both the latent state of the system $x_n$ at timesteps $t_0,...,t_N$ and system parameters $\theta \in \Theta$. 
-\end{itemize}
+\textbf{\textcolor{red}{Partially-Observed Markov Process}}:
+
+Unknown Markov process $(X_t)$, observations $y_1^*,...,y_N^*$ at $t_1,..., t_N$. \\
+Estimate latent state of the system $x_{1:N}$ and system parameters $\theta \in \Theta$. 
 \end{tcolorbox}
-e.g. given monthly cholera case counts, estimate number of infected and recovered people in the population, and how transmissible the disease is.
-
+\small{e.g. given monthly cholera case counts, estimate number of infected and recovered people in the population, and how transmissible the disease is.}
   \end{block}
 
-  \vspace{-1ex}
   \begin{block}{Problem}
@@
 
-  \textbf{What to do now?} Particle filters provide a Monte Carlo approximation of the \textbf{filtering distribution} $f_{X_n|Y_{1:n};\theta}$ and the \textbf{log-likelihood} $\ell(\theta)$ increasingly accurate in the number of particles $J$.
+  \textbf{What to do now?} Particle filters give a Monte Carlo approximation of the \textbf{filtering distribution} $f_{X_n|Y_{1:n};\theta}$ and the \textbf{log-likelihood} $\ell(\theta)$ increasingly accurate in the number of particles $J$.
+  %\vspace{1ex}
 
-  \textbf{Gradient descent} using autodiff (AD) on the likelihood estimate? \textcolor{red}{But:}
+  \textbf{Idea:} Gradient descent using autodiff (AD) on the likelihood estimate? \textcolor{red}{But:}
+  \vspace{-2.6ex}
   \begin{enumerate}
       \item Likelihood estimate not differentiable due to Monte Carlo resampling.
-      \item Previous work \cite{corenflos21, naesseth18, scibior21} on this, but computationally expensive \cite{corenflos21}, has uncontrolled asymptotic bias \cite{naesseth18}, or has high variance \cite{poyiadjis11}.
+      \item Previous work \cite{corenflos21, naesseth18, poyiadjis11, scibior21} on this is computationally expensive \cite{corenflos21}, has uncontrolled asymptotic bias \cite{naesseth18}, or has high variance \cite{poyiadjis11, scibior21}.
   \end{enumerate}
+  \vspace{1ex}
   \end{block}
@@
 
-    When treating the particles at $\phi$ as constants, you're now differentiating through (1) a differentiable simulator and (2) a series of measurement density ratios, so the resulting likelihood estimate is smooth!
+    \textbf{Smooth approximation:} Treating the particles at $\phi$ as constants, this is a function of a differentiable simulator and measurement density ratios, so the resulting likelihood estimate is smooth!
 
@@
 
-  \vspace{-1ex}
+  \vspace{1ex}
     \begin{figure}
         \centering
-        \includegraphics[width=0.45\linewidth]{imgs/095/mop.png}
-        \includegraphics[width=0.405\linewidth]{imgs/095/biasvar.png}
-        \caption{Dhaka cholera model of \cite{king08}.
-\textbf{Left:} Likelihood slice for the recovery rate parameter, and smooth off-parameter reconstructions for different $\alpha$.
-\textbf{Right:} Bias-variance tradeoff induced by $\alpha$ for estimation of the score for the trend parameter.
-}
+        \includegraphics[width=0.52\linewidth]{imgs/095/mop.png}
+        \includegraphics[width=0.472\linewidth]{imgs/095/biasvar.png}
+        \caption{\textbf{Left:} Non-differentiable likelihood estimates and our smooth off-parameter reconstructions. 
+        \textbf{Right:} Bias-variance tradeoff in score estimation induced by $\alpha$.}
         \label{fig:enter-label}
@@
   \begin{alertblock}{Guarantees}
+    \begin{enumerate}
+        \item \textbf{Correctness:} Procedure targets the filtering distribution $\pi_n(\theta)$, and is strongly consistent for the likelihood under $\theta$.\footnotemark[1]\footnotetext{\tiny{1.
+We also showed a more general result holds, that of a particle triangular array SLLN corrected for arbitrary off-target resampling.}}
 
-  \textbf{Correctness:} This procedure targets the filtering distribution, $\pi_n(\theta) = f_{X_n|Y_{1:n}}(x_n|y_{1:n}^*;\theta)$, and is strongly consistent for the likelihood under $\theta$. 
-  We obtain a set of state estimates $(x_{n,j}^{F,\theta})_{j=1}^J$ and a likelihood estimate $\hat \lik_J(\theta)$ so for any $\alpha\in[0,1]$, for any measurable bounded functional $h$:
-  $$\frac{\sum_{j=1}^J h\left(x_{n, j}^{F, \theta}\right) w_{n, j}^{F, \theta}}{\sum_{j=1}^J w_{n, j}^{F, \theta}} \stackrel{\text{a.s.}}{\to} E_{\pi_n(\theta)}\left[h\left(X_n\right)\right], \; \hat{\mathcal{L}}_J^\alpha(\theta) \stackrel{\text{a.s.}}{\to} \mathcal{L}(\theta) \text{ as } J \to \infty.$$
+        Obtains state estimates $x_{n,1:J}^{F,\theta}$ and a likelihood estimate $\hat \lik_J^\alpha(\theta)$ so for any $\alpha\in[0,1]$ and measurable bounded functional $h$, as $J \to \infty$:
 
-  \textbf{Consistent score estimates:} When $\alpha=1$, $\nabla_\theta \hat\ell_J^1(\theta) \stackrel{\text{a.s.}}{\to} \ell(\theta)$ as $J \to \infty$.
-Else, increasing asymptotic bias as $\alpha \to 0$ in return for lower variance.
+  $$\frac{\sum_{j=1}^J h\left(x_{n, j}^{F, \theta}\right) w_{n, j}^{F, \theta}}{\sum_{j=1}^J w_{n, j}^{F, \theta}} \stackrel{\text{a.s.}}{\to} \E_{\pi_n(\theta)}[h(X_n)], \qquad \hat{\mathcal{L}}_J^\alpha(\theta) \stackrel{\text{a.s.}}{\to} \mathcal{L}(\theta).$$
+  \vspace{1.5ex}
 
-  \textbf{Encompasses other estimators:} When $\alpha=0$, $\nabla_\theta \hat\ell_J^0(\theta)$ is equivalent to the low-variance but asymptotically biased estimate of \cite{naesseth18}.
-When $\alpha=1$, $\nabla_\theta \hat\ell_J^1(\theta)$ is equivalent to the high-variance but consistent estimate of \cite{poyiadjis11}.
-
+  \item \textbf{Consistent score estimates:} When $\alpha=1$, $$\nabla_\theta \hat\ell_J^1(\theta) \stackrel{\text{a.s.}}{\to} \ell(\theta) \text{ as } J \to \infty.$$ Else, increasing asymptotic bias as $\alpha \to 0$, but lower variance.
+  \vspace{2ex}
+  \item \textbf{Encompasses and interpolates between other estimators:}\\
+  \textbf{$\alpha=0$:} $\nabla_\theta \hat\ell_J^0(\theta)$ is \cite{naesseth18}'s low-variance/asymptotically biased estimate. \\
+  \textbf{$\alpha=1$:} $\nabla_\theta \hat\ell_J^1(\theta)$ is \cite{poyiadjis11}'s high-variance/consistent estimate.
+  \vspace{2ex}
+  \item \textbf{Linear convergence for gradient descent/quasi-Newton/Newton:} \\Early-stopped particle gradient descent with learning rate $\eta$, large $\alpha, J$ converges linearly if log-likelihood strongly convex and smooth:
+  $$\ell\left(\theta^*\right)-\ell\left(\theta_{m+1}\right) \leq\left(1- O(\eta)\right)\left(\ell\left(\theta^*\right)-\ell\left(\theta_m\right)\right).$$
+    \end{enumerate}
   \end{alertblock}
 
-  \vspace{-2ex}
-  \begin{block}{Rates}
-  \vspace{-6.5mm}
+  %\vspace{-2.5ex}
+  \begin{block}{Rates for Score Estimates}
+  \vspace{-1.5ex}
       \begin{table}[t]
 \centering
-\scalebox{0.85}{
+\scalebox{1}{
 \begin{tabular}{c  c}
-\toprule & MSE 
+\toprule \multicolumn{2}{c}{\textbf{MSE}} 
 \tabularnewline
 \toprule 
-\vspace{0.2ex}
-$\alpha \in (0,1)$ & $\tcbhighmath[colframe=ballblue]{\min_{k \leq N} N p G^{\prime}(\theta)^2(k^2/J+(1-\epsilon)^{\lfloor k /(c \log (J))\rfloor}+k+\psi_k(\alpha))}$
+\vspace{1.2ex}
+$\alpha \in (0,1)$ & \qquad$\tcbhighmath[colframe=ballblue]{\min_{k \leq N} N p (k^2/J+
+\begingroup
+\color{ballblue}\underbracket[1pt][1pt]{\color{black}
+(1-\epsilon)^{\lfloor k /(c \log (J))\rfloor}}_{\text{\textcolor{blue}{\tiny{controllable bias}}}}
+\endgroup
++k+\psi_k(\alpha))}$\qquad
 \vspace{0.2ex}
@@
 \vspace{0.2ex}
-$\alpha=0$, equivalent to \cite{naesseth18} & $N p G'(\theta)^2(J^{-1} + (1-\epsilon)^{\lfloor 1 /(c \log (J))\rfloor})$
+$\alpha=0$ \cite{naesseth18} & $N p (1/J + 
+\begingroup
+\color{red}\underbracket[1pt][1pt]{\color{black}(1-\epsilon)^{\lfloor 1 /(c \log (J))\rfloor}}_{\text{\textcolor{red}{\tiny{uncontrollable bias}}}}
+\endgroup
+)$
 \vspace{0.2ex}
@@
 \vspace{0.2ex}
-$\alpha=1$, equivalent to \cite{poyiadjis11} & $N^4pG'(\theta)^2/J$
+$\alpha=1$ \cite{poyiadjis11} & $N^4p/J$
 \vspace{0.2ex}
@@
 \toprule
-&  Variance
+  \multicolumn{2}{c}{\textbf{Variance}}
 \tabularnewline
 \toprule\vspace{0.2ex}
-$\alpha \in (0,1)$ & $\tcbhighmath[colframe=ballblue]{\min _{k \leq N}\left(\frac{k^2 p G^{\prime}(\theta)^2 N p}{(1-\alpha)^2 J}+\frac{\alpha^k}{1-\alpha} N p G^{\prime}(\theta)^2\right)}$
+$\alpha \in (0,1)$ & $\tcbhighmath[colframe=ballblue]{\min _{k \leq N}Np\left(\frac{k^2}{(1-\alpha)^2 J}+\frac{\alpha^k}{1-\alpha} \right)}$
 \vspace{0.2ex}
@@
 \vspace{0.2ex}
-$\alpha=0$, equivalent to \cite{naesseth18} & $N p G'(\theta)^2/J$
+$\alpha=0$ \cite{naesseth18} & $N p /J$
 \vspace{0.2ex}
@@
 \medskip
-\caption{Rates given the number of particles $J$, trajectory length $N$, dimension of parameter space $p$, bound on gradient norm $G'(\theta)$, and model-specific constant $\epsilon>0$.
-We define $\psi_k(\alpha)=\left(\alpha^k+\alpha^{k+1}-\alpha\right) /(1-\alpha)$.
-Our novel contribution of $\alpha \in (0,1)$ has a lower MSE than, and interpolates between the variances of, $\alpha=0$ \cite{naesseth18} and $\alpha=1$ \cite{poyiadjis11}.}
+\caption{Rates given the trajectory length $N$, number of particles $J$, parameter space dimension $p$, model-specific constant $\epsilon>0$, where $\psi_k(\alpha)=\left(\alpha^k+\alpha^{k+1}-\alpha\right) /(1-\alpha)$.
+\textcolor{red}{Our $\alpha \in (0,1)$ has a better rate for the MSE than $\alpha=0$ \cite{naesseth18} and $\alpha=1$ \cite{poyiadjis11}.}}
 \label{tab:bounds}
@@
 
-\begin{block}{Practical Maximum Likelihood Estimation}
-    \textbf{Problem:} Difficult, noisy, and nonconvex problem. Gradient descent/L-BFGS/Newton struggles with saddle points, local minima, and ridges.
-
-    \textbf{Solution:} Warm-start gradient descent with IF2 algorithm of \cite{ionides15}.
-IF2 gets to a neighborhood of the MLE quickly, but fails to find it.
-Gradient descent does well in that neighborhood, warm-start achieves best of both worlds.
-
-    \textbf{Theoretical guarantee:}
-
-\end{block}
-
 \end{column}
@@
 \separatorcolumn
-
 \begin{column}{\colwidth}
 
-  \begin{exampleblock}{Simulations: Hybrid RL Encourages Exploration}
+\vspace{-2ex}
+\begin{figure}[H]
+    \centering{\includegraphics[scale=1]{imgs/095/tikzcholera.png}}
+    \vspace{-2ex}
+    \caption{Diagram of the stochastic SIRS model in \cite{king08}. 
+    }
+\end{figure}
 
-  \textbf{Does appending the offline dataset to the experience replay buffer encourage sufficient exploration for the portion of the state-action space that does not have good coverage?}
+\vspace{.5ex}
+  \begin{exampleblock}{Case Study: Cholera Epidemic in Dhaka, Bangladesh}
 
-  \textbf{Tabular:} Initialized UCBVI \cite{azar2017minimax} with an offline dataset of $100$ trajectories of length $20$ on a forest management simulator. 
-  %Online partition is the state-action pairs with occupancy below $1/SA$.
+  \textbf{Dhaka, 1821-1940:} Stochastic SIRS model from \cite{king08}. Cholera dynamics governed by SDEs. Seasonality, trend in infection modeled with splines.
+
+  \textbf{In practice:} Gradient descent got stuck. Needed a warm-start with IF2 \cite{ionides15} to get to a neighborhood of the MLE quickly and find the MLE.
+  \vspace{0.8ex}
 
@@
     \centering
-    \includegraphics[scale=1]{imgs/095/biasvar.png}
-    \caption{Hybrid RL explores more of the online partition (poorly covered $s,a$) to fill in the gaps in the offline dataset than online RL does, unless the behavior policy is optimal.
+    \includegraphics[scale=1.15]{imgs/095/pairs.png}
+    \includegraphics[scale=0.915]{imgs/095/boxplot_all.png}
+    \caption{\textbf{Left:} Scatterplot depicting paired searches from the same starting point. \textbf{Right:} Raincloud plot of all searches. Our method outperforms the benchmark, IF2. }
+    \label{fig:scatter-raincloud}
+\end{figure}
+%\vspace{1ex}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[scale=0.815]{imgs/095/optim.png}
+    \includegraphics[scale=0.562]{imgs/pmcmc/nuts_eb.png}
+    \caption{\textbf{Left:} Optimization progress of IFAD (ours) and IF2 \cite{ionides15} over 100 trials. \textbf{Right:} Application to Bayesian inference. 
     }
@@
 
-    \textbf{Linear:} Initialized LSVI-UCB \cite{jin2020provably} with $200$ trajectories of length $40$ on a scaled-down version of Tetris.
-
-    \begin{figure}[H]
-    \centering
-    \includegraphics[scale=1]{imgs/095/optim.png}
-    \caption{Hybrid RL takes fewer online episodes than online-only RL to achieve a lower concentrability coefficient.}
-    \label{fig:linear-coverage}
-\end{figure}
-
+    \textbf{Application to Bayesian inference:} Using our score estimate for a No U-Turn Sampler for particle MCMC, with an informative prior from the IF2 warm-start, reduces burn-in time from 700,000 \cite{fasiolo16} to 500 iterations.
   \end{exampleblock}
@@
 \begin{block}{}
-  \vspace{-0.2em}
+  \vspace{-0.5ex}
     \tiny{\bibliographystyle{plain}

### 2024-10-13 online/online.tex
some intro edits

@@
 
+That is, the online MOP-$\alpha$ update is given by
+$$\theta_{t+1} \gets \theta_t + \eta_t \sum_{n=1}^t \alpha^{t-n} \nabla_{\theta_n} \ell_n(\theta_n),$$
+which, defining $m_t = $ can be written
+
+\begin{gathered}
+\theta_{t+1}=\theta_t+\mu_t v_t-\varepsilon_t \nabla f\left(\theta_t\right) \\
+v_{t+1}=\mu_t v_t-\varepsilon_t \nabla f\left(\theta_t\right)
+\end{gathered}
+
 Weird observation -- online MOP-$1$ according to Poyiadjis is is MOP-$0$ applied sequentially, in the case of the bootstrap filter.
 MOP-$1$ sums the log-densities of the surviving particles, while MOP-$0$ sums the log-densities of the particles at each timestep.
+
+Note that the particles now depend on past parameter values -- $x_{n,j}^{F, \theta_{1:n,j}}$
+
+Nesterov may not buy you much in the general adversarial online convex optimization setting, but if we have some stationarity+mixing assumptions on the system, then the losses are effectively the same in some sense!
+Maybe there is some sense of acceleration we can achieve here.
+
+Question -- MOP-0 is inconsistent, but presumably the online analogue that optimizes conditional log-likelihoods yields a consistent estimate.
+What's with this?
+Does taking the online gradient imply we ignore the resampling gradient when thought of as a batch update?
 

### 2024-10-13 paper/ms.tex
some intro edits

@@
 POMP models arise in fields as diverse as optimal control \citep{singh22}, epidemiology \citep{he10, stocks17}, ecology \citep{knape12} and finance \citep{kim08, breto14}.
-Despite their ubiquity, estimation and inference within this broad class of models remains a challenging problem.
-To tackle this, we use automatic differentiation (AD) to construct improved algorithms for inference on complex POMP models.
+Despite their ubiquity, estimation and inference for this broad class of models remains a challenging problem.
+To tackle this, we use automatic differentiation (AD) to construct improved algorithms for complex POMP models.
 
@@
 It provides an unbiased estimate of the likelihood function  \citep{delMoral04}, enabling Bayesian inference \citep{andrieu10,chopin13} and likelihood-based inference \citep{ionides06-pnas,ionides15}.
-This is desirable, as likelihood-based inference is statistically efficient \citep{pawitan01} and robust to a moderate amount of Monte Carlo error \citep{ionides17,ning21}.
-Furthermore, basic particle filter algorithms, also known as boostrap filters, do not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
+Likelihood-based inference is statistically efficient \citep{pawitan01} and robust to a moderate amount of Monte Carlo error \citep{ionides17,ning21}.
+Furthermore, basic particle filters, also known as boostrap filters, do not need to evaluate the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
+
 This \textit{plug-and-play} property is useful for scientific applications \citep{he10}.
@@
 Plug-and-play methods are often called likelihood-free \citep{owen15}, and it may be counter-intuitive that likelihood-based inference is possible using likelihood-free methods. 
-However, iterated filtering algorithms have shown that this is practical in various scientific investigations, such as \citep{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
+However, iterated filtering algorithms have shown that this is practical in various scientific investigations \citep{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
 Nevertheless, the Monte Carlo variability arising in iterated filtering applications becomes increasingly problematic as the size of the data and the complexity of the model increase.
@@
 However, existing approaches are either asymptotically biased \citep{naesseth18, jonschkowski18}, have high variance \citep{poyiadjis11, scibior21}, are computationally expensive \citep{corenflos21, chen24}, or require access to transition densities \citep{poyiadjis11, scibior21, singh22, chen24}.
+
 \cite{scibior21} showed that the estimators derived by \cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.
+However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability, which led the authors to reject this algorithm in favor of an approach without the plug-and-play property and with poorer scalability.
 
-However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability \citep{poyiadjis11}.
-Additionally, while the stop-gradient procedure is undeniably effective, the reason why the correction recovers the \cite{poyiadjis11} estimator remains somewhat unclear.
-We use \cite{scibior21} and \cite{poyiadjis11} as our starting point in developing a new approach remedying these issues.
-
-We start by presenting a new construction of the gradient estimator of \cite{scibior21} and \cite{poyiadjis11}.
-Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a {\it Measurement Off-Parameter} (MOP) particle filter.
-In short, \textit{instead of directly differentiating through a basic particle filter, MOP uses AD to differentiate through a series of measurement density ratios and a differentiable simulator}. 
+We start by presenting a new algorithm which (for reasons which will be discussed later) we call a {\it Measurement Off-Parameter} (MOP) particle filter. 
+In summary, \textit{instead of directly differentiating through a basic particle filter, MOP uses AD to differentiate through a series of measurement density ratios and a differentiable simulator}. 
 Our MOP algorithm differs from previous smooth particle filters \citep{svensson18,malik11} by possessing the plug-and-play property.
 Critically, we also avoid the high variance of \cite{svensson18} by generalizing the MOP particle filter to add a discount factor, $\alpha \in [0,1]$.
-This discount factor enables MOP-$\alpha$ to interpolate between the  biased gradient estimator of \cite{naesseth18}, when $\alpha=0$, and the high-variance gradient estimate from \citep{poyiadjis11, scibior21}, when $\alpha=1$ (Theorem \ref{thm:mop-functional-forms}).
+Remarkably, we show that the derivative of our algorithm coincides with the biased gradient estimator of \cite{naesseth18}, when $\alpha=0$, and the high-variance gradient estimate from \citep{poyiadjis11, scibior21}, when $\alpha=1$ (Theorem \ref{thm:mop-functional-forms}).
 The bias-variance tradeoff induced by $\alpha$ suggests, both in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}), a choice of $\alpha$ strictly in $(0,1)$.
+Choosing $\alpha$ close to one, we obtain a low-bias, low-variance, plug-and-play gradient estimate with computational scaling linear in the number of particles.
 
-% One can simply drop terms arising from resampling when differentiating through the particle filter \cite{naesseth18}, but this leads to asymptotic bias \cite{corenflos21}.
-% Instead, MOP-$\alpha$ bypasses the seemingly incompatible paradox of differentiating through discrete Monte Carlo sampling by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
 MOP-$\alpha$ can provide strongly consistent estimates of the filtering distribution and the log-likelihood (Theorem \ref{thm:mop-targeting}).
 When $\alpha=1$, MOP-$\alpha$ also provides consistent score estimates (Theorem \ref{thm:mop-grad-consistency}).
-This enables us to derive a linear convergence rate for gradient descent using MOP-$\alpha$ in the  presence of strong convexity and smoothness (Theorem \ref{thm:mop-convergence}).
+This lets us derive a linear convergence rate for gradient descent with MOP-$\alpha$ in the strongly convex and smooth setting (Theorem \ref{thm:mop-convergence}).
 
@@
 
-Consider an unobserved Markov process $\{X(t),t  \geq t_0\}$, with discrete-time observations $Y_1,...,Y_N$ realized at values $y_1^*,...,y_N^*$ at times $t_1,..., t_N$.
-The process is parameterized by an unknown parameter $\theta \in \Theta \subseteq \R^p$, where the state $X(t)$ take values in the state space $\gX \subseteq \R^d$, the observations $Y_n$ take values in $\gY,$ and we write $X_n := X(t_n)$.
-
 % By a similar decomposition to that in \cite{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
@@
 
+Consider an unobserved Markov process $\{X(t),t  \geq t_0\}$, with discrete-time observations $Y_1,...,Y_N$ realized at values $y_1^*,...,y_N^*$ at times $t_1,..., t_N$.
+The process is parameterized by an unknown parameter $\theta \in \Theta \subseteq \R^p$, where the state $X(t)$ takes values in the state space $\gX \subseteq \R^d$, the observations $Y_n$ take values in $\gY,$ and we write $X_n := X(t_n)$.
 We suppose that the discrete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$.
@@
 We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
-$x_{n,j}^A$ denotes the ancestral trajectory of particle $j$ at time $n$,  and $k_j$
-denotes the resampling indices for particle $j$. 
+$x_{n,j}^A$ denotes the ancestral trajectory of particle $j$ at time $n$,  and $k_j$ the resampling indices for particle $j$. 
 %$a(\cdot)$ is the ancestor function that maps a particle to its parent, $j \mapsto a(j)$,
@@
 
+One can think of this as reconstructing local smooth approximations of the discontinuous likelihood estimates produced by a particle filter for a fixed number of particles $J$ and fixed seed $\omega$.
+These approximate the likelihood at $\theta$ in a neighborhood around $\phi$, with the intent that differentiating these smooth log-likelihood reconstructions yields a score estimate.
+This is illustrated in the left panel of Figure \ref{fig:biasvar} for various choices of $\alpha$.
+
 Our approach has similarities with \cite{svensson18}, in which particles from a previous run at $\phi$ are reweighted to evaluate the likelihood at any sufficiently nearby parameter $\theta$ for $0$-th order optimization.
@@
     \label{alg:mop}
-	     \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process_n(x_{n+1}|x_n; \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
+	     \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process_n(\cdot|x_n; \theta)$, evaluation parameter $\theta$, baseline parameter $\phi$, seed $\omega$.
 
@@
         \nabla_\theta \hat\ell^0(\theta) 
-        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
+        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
         \vspace*{-2.5mm}
@@
     \end{equation}
-    yielding the estimator of \cite{poyiadjis11, scibior21} for the bootstrap filter.
+    yielding the estimator of \cite{poyiadjis11} for the bootstrap filter.
 \end{thm}
@@
     \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2.2}}]{imgs/095/biasvar.png}
-    \caption{\textbf{Left:} Non‐differentiable likelihood estimates and our smooth off‐parameter reconstructions.
+    \caption{\textbf{Left:} Non‐differentiable likelihood estimates and our smooth off‐parameter approximations that hold locally around the baseline parameter $\phi$.
 \textbf{Right:} Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}.
-We display the MSE of score estimates for the trend in transmission at the MLE.}
+We display the MSE of score estimates for the trend in transmission at the MLE.
+Our novel contribution of $\alpha \in (0,1)$ significantly improves the quality of the likelihood approximation, and therefore the gradient estimates.
+}
     \label{fig:biasvar}
@@
 We tested IFAD against IF2 on a global search problem for the Dhaka cholera model.
-We re-implemented IF2 to do so, but we also compare our results with the results of \cite{ionides15} (labeled "IF2 2015").
+We re-implemented IF2 to do so, but we also compare our results with the results of \cite{ionides15} (labeled ``IF2 2015'').
 Our re-implementation reaches a better log-likelihood than that of \cite{ionides15} due to a better choice of algorithmic parameters.

### 2024-10-17 paper/ms.tex
Updates from Overleaf

@@
 \cite{scibior21} showed that the estimators derived by \cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.
-However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability, which led the authors to reject this algorithm in favor of an approach without the plug-and-play property and with poorer scalability.
+However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability, which led the authors to reject this algorithm in favor of an approach without the plug-and-play property and with inferior scalability.
 
@@
 Within a neighborhood of the MLE that grows as more data become available \citep{ning21}, local quadratic approximations such as local asymptotic normality apply under general conditions \citep{lecam00}.
-This suggests that the linear convergence may occur reasonably often in practice.
+This suggests that the linear convergence may occur often in practice.
 
@@
 	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 cooling schedule $\eta_m$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $m=0.$\newline
-        Run IF2 until initial "convergence" under cooling schedule $\eta_m$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
+        Run IF2 until initial ``convergence'' under cooling schedule $\eta_m$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
 		\textbf{While} procedure not converged: \newline

### 2024-10-18 paper/ms.tex
Updates from Overleaf

@@
 \author{K. Tan$^{1}$, G. Hooker$^{1}$ and E. L. Ionides$^{2}$
-  \vspace{2mm}\\
+  \vspace{3ex}\\
   \small{$^{1}$Department of Statistics, University of Pennsylvania}\\
@@
 However, iterated filtering algorithms have shown that this is practical in various scientific investigations \citep{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
-Nevertheless, the Monte Carlo variability arising in iterated filtering applications becomes increasingly problematic as the size of the data and the complexity of the model increase.
+Nevertheless, the Monte Carlo variability arising in iterated filtering applications is increasingly problematic as the size of the data and the complexity of the model increase.
 Algorithmic advances are needed to make plug-and-play likelihood maximization for POMPs more numerically efficient.
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \section{Problem Setup}
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \section{Off-Parameter Particle Filters}
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \subsection{Algorithm Outline}
@@
 The resampling indices, which we write as $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$, are a function of $\phi$ for any value of $\theta$.
-For the categorical distribution, we use systematic resampling \citep{arulampalam02,king16} which usually has superior performance to multinomial resampling.
+For the categorical distribution, we use systematic resampling \citep{arulampalam02,king16} which outperforms multinomial resampling.
 In practice, we only ever evaluate MOP-$\alpha$ in the case where $\theta$ and $\phi$ coincide.
@@
 where we recursively correct the weights for each particle for the cumulative error incurred by the off-parameter resampling:
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \begin{equation}
@@
     w^{F,\theta}_{0,j}= 1.
-    \arxiv{}{\vspace*{-1.5mm}}
+    \arxiv{}{\vspace*{-3ex}}
 \end{equation}
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \subsection{Discounting Weights for a Bias-Variance Tradeoff}
@@
 For $\alpha<1$, the weights from previous timesteps are discounted as in Equation~(\ref{eq:weighting-scheme}), allowing for lower variance.
-Algorithm~\ref{alg:mop} continues to target the filtering distribution and likelihood if $\theta=\phi$, but has an asymptotically biased score estimate.
-MOP-$1$ and MOP-$0$ correspond to the estimators of \cite{poyiadjis11} and \cite{naesseth18} (Theorem \ref{thm:mop-functional-forms}), while the novel possibility of $0<\alpha<1$ lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate in both theory (Theorem \ref{thm:mop-biasvar}) and practice (Figure \ref{fig:biasvar}).
+Algorithm~\ref{alg:mop} continues to target the filtering distribution and likelihood if $\theta=\phi$, but is asymptotically biased for the score.
+MOP-$1$ and MOP-$0$ correspond to the estimators of \cite{poyiadjis11} and \cite{naesseth18} (Theorem \ref{thm:mop-functional-forms}), while the novel possibility of $0<\alpha<1$ optimizes a bias-variance tradeoff for the score in both theory (Theorem \ref{thm:mop-biasvar}) and practice (Figure \ref{fig:biasvar}).
 
-\arxiv{}{\vspace*{-2mm}}
-\subsection{MOP-$\alpha$ Encompasses the Estimators of \citep{poyiadjis11, scibior21, naesseth18}}
+\arxiv{}{\vspace*{-3ex}}
+\subsection{MOP-$\alpha$ Encompasses the Estimators of \citep{poyiadjis11, naesseth18}}
 
@@
     Writing $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$ and using the after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$, when $\alpha=0$,
-    \vspace*{-2.5mm}
+    \vspace*{-3ex}
     \begin{equation} \nonumber
@@
         = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
-        \vspace*{-2.5mm}
+        \vspace*{-3ex}
     \end{equation}
     yielding \cite{naesseth18}'s estimator with the bootstrap filter. When $\alpha=1$,
-    \vspace*{-2.5mm}
+    \vspace*{-3ex}
     \begin{equation} \nonumber
@@
         = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
-    \vspace*{-2.5mm}
+    \vspace*{-3ex}
     \end{equation}
@@
 \begin{figure}[htbp!]
+\arxiv{}{\vspace*{-2ex}}
   \centering
@@
     \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2.2}}]{imgs/095/biasvar.png}
+    \arxiv{}{\vspace*{-3ex}}
     \caption{\textbf{Left:} Non‐differentiable likelihood estimates and our smooth off‐parameter approximations that hold locally around the baseline parameter $\phi$.
@@
     \label{fig:biasvar}
-    \arxiv{}{\vspace*{-1ex}}
+    \arxiv{}{\vspace*{-4ex}}
 \end{figure}
 
-% \arxiv{}{\vspace*{-2mm}}
+% \arxiv{}{\vspace*{-3ex}}
 % \subsection{Summary of Theoretical Guarantees}
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\begin{algorithm}[h!]
+	\caption{IFAD}
+    \label{alg:ifad}
+	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 and MOP-$\alpha$ cooling schedules $\eta_m$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $m=0.$\newline
+        Run initial IF2 search under cooling schedule $\eta_m$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
+		\textbf{While} procedure not converged: \newline
+		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_m).$ \newline
+		\hspace*{4mm} Set $g(\theta_m) := \nabla_{\theta}(-\hat\loglik(\theta_m))$, and consider any $H(\theta_m)$ s.t. $\lambda_{\min}(H(\theta_m)) \geq c$. \newline
+		\hspace*{4mm} Update $\theta_{m+1} := \theta_m - \eta_m (H(\theta_m))^{-1} g(\theta_m)$, $m:=m+1.$ \newline
+		\textbf{Return} $\hat{\theta} := \theta_m.$
+\end{algorithm}
+
+\arxiv{}{\vspace*{-1ex}}
 \subsection{Optimization}
@@
 The convergence of IF2 (and so the first stage of IFAD) to a neighborhood of the MLE happens rapidly in practice. 
-In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, we require no more than 40 iterations of IF2 to obtain a suitable warm-start of acceptable quality.
+In the case of the Dhaka cholera model of \cite{king08}, when an aggressive geometric cooling multiplier of 0.95 and initial random walk standard deviation of 0.02 is used, we require no more than 40 iterations of IF2 to obtain an acceptable warm-start.
 However, even when applying much larger amounts of Monte Carlo effort, IF2 fails to get within seven log-likelihood units of the maximum on this problem (Figure~\ref{fig:boxplot}).
-These numerical results are consistent with theoretical results for iterated filtering, showing slow convergence close to the maximum \citep{doucet15-if}.
-IF2, without assistance from AD, requires additional computationally intensive techniques, such as repeated pruning and profiling, to reach the inferentially critical region within one log unit of the maximum.
+These observations are consistent with theoretical results for iterated filtering, showing slow convergence close to the maximum \citep{doucet15-if}.
+IF2, without assistance from AD, requires additional computationally intensive techniques, such as repeated pruning and profiling, to reach the inferentially critical region within one log-likelihood unit of the maximum.
 
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-2ex}}
 \subsection{Linear Convergence Rates}
@@
 Consider the second stage of IFAD (Algorithm \ref{alg:ifad}), stopping if $\|\nabla_\theta \hat\ell^\alpha(\theta_m)\| \leq (1+\sigma) \epsilon$, where $\sigma \geq \frac{4 \Gamma}{(1-\beta)}$ for some $\beta \in (0,1)$.
-Assume $-\ell$ is strongly convex and smooth, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$.
-Choose the learning rate $\eta$ such that $\eta \leq \frac{c(1-\beta)}{2\Gamma}$.
-Then, for sufficiently large $\alpha$ and $J$ so $\|\nabla_\theta \hat\ell^\alpha(\theta_m) - \nabla_\theta\ell^\alpha(\theta_m)\|$ and $H(\theta_m) \succ cI_p \succ 0$ for all $m$ with probability at least $1-\delta$,
-\arxiv{}{\vspace*{-2mm}}
+Assume strongly convex and smooth $-\ell$, $\gamma I \preceq \nabla_\theta^2 (-\ell) \preceq \Gamma I$, and assumptions (A1--A5) in Section~\ref{sec:thms}.
+Choose the learning rate $\eta$ so $\eta \leq \frac{c(1-\beta)}{2\Gamma}$. 
+Then, for $\alpha$ sufficiently close to 1 and $J$ sufficiently large (depending on $\epsilon$ and $\delta$), we have that
+$\|\nabla_\theta \hat\ell^\alpha(\theta_m) - \nabla_\theta\ell^\alpha(\theta_m)\| \leq \epsilon$ and $H(\theta_m) \succ cI_p \succ 0$ for all $m$, and that
+\arxiv{}{\vspace*{-3ex}}
 $$
-\ell(\theta^*) - \ell(\theta_{m+1}) \leq \Big(1-\eta\beta\, \frac{8\gamma}{9c}\Big)\big(\ell(\theta^*)-\ell(\theta_m)\big).
+\ell(\theta^*) - \ell(\theta_{m+1}) \leq \Big(1-\eta\beta\, \frac{8\gamma}{9c}\Big)\big(\ell(\theta^*)-\ell(\theta_m)\big) \text{ with probability at least } 1-\delta.
 $$
@@
 \end{thm}
-\arxiv{}{\vspace*{-7mm}}
+\arxiv{}{\vspace*{-10ex}}
 The proof of Theorem~\ref{thm:linear} uses results from randomized numerical linear algebra, building on Theorem~6 of \cite{mahoney16}. 
@@
 
-\begin{algorithm}[H]
-	\caption{IFAD}
-    \label{alg:ifad}
-	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 cooling schedule $\eta_m$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $m=0.$\newline
-        Run IF2 until initial ``convergence'' under cooling schedule $\eta_m$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
-		\textbf{While} procedure not converged: \newline
-		\hspace*{4mm} Run Algorithm \ref{alg:mop} to obtain $\hat\loglik(\theta_m).$ \newline
-		\hspace*{4mm} $g(\theta_m) := \nabla_{\theta}(-\hat\loglik(\theta_m))$, $H(\theta_m)$ s.t. $\lambda_{\min}(H(\theta_m)) \geq c$. \newline
-		\hspace*{4mm} Update $\theta_{m+1} := \theta_m - \eta (H(\theta_m))^{-1} g(\theta_m)$, $m:=m+1.$ \newline
-		\textbf{Return} $\hat{\theta} := \theta_m.$
-\end{algorithm}
-
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \section{Application to a Cholera Transmission Model}
@@
     dI&=\big(\lambda(t) S-(m+\delta+\gamma) I\big)\, dt+({\sigma S I}/{P})\, dB, \\
-    dR^1&=\big(\gamma I-(k \epsilon+\delta) R^1\big)\, dt, \hspace{2mm} \dots \\
+    dR^1&=\big(\gamma I-(k \epsilon+\delta) R^1\big)\, dt, \hspace{3ex} \dots \\
     dR^k&=\big(k \epsilon R^{k-1} -(k \epsilon+\delta) R^k\big)\, dt,
-    \vspace*{-2mm}
+    \vspace*{-3ex}
 \end{align*}
@@
 The force of infection, $\lambda_t$, is modeled by splines $(s_j)_{j=1}^6$
-\vspace*{-2mm}
+\vspace*{-3ex}
 \begin{equation*}    \lambda_t=\exp\hspace{-1mm}\left(\hspace{-.5mm}\beta_{\text{trend}}(t-t_0)+\hspace{-1mm}\sum_{j=1}^{6} \beta_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm}\frac{I}{P} + \exp\hspace{-1mm} \left(\sum_{j=1}^{6} \omega_j s_j(t)\hspace{-1mm}\right)\hspace{-1mm},
-    \vspace*{-2mm}
+    \vspace*{-3ex}
 \end{equation*}
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \subsection{Results}
@@
     \label{fig:optim}
-    \arxiv{}{\vspace*{-2mm}}
+    \arxiv{}{\vspace*{-3ex}}
 \end{figure}
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \section{Application to Bayesian Inference}
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \section{Theoretical Analysis of MOP-$\alpha$}
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \begin{enumerate}[label=(A\arabic*),itemsep=-1ex, , font=\small] 
@@
 \end{enumerate}
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \subsection{MOP-$\alpha$ Targets the Filtering Distribution}
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \subsection{MOP-$\alpha$ Error, Bias and Variance}
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \section{Computational Efficiency}
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \section{Discussion}
@@
 
-\arxiv{}{\vspace*{-2mm}}
+\arxiv{}{\vspace*{-3ex}}
 \subsection{Further Work}

### 2024-10-18 paper/si.tex
Updates from Overleaf

@@
-\documentclass[9pt,twoside,lineno]{pnas-new-si}
+\documentclass[numsec,webpdf,modern,medium,namedate]{oup-authoring-template}
 
@@
 
-\templatetype{pnassupportinginfo}
+%\templatetype{pnassupportinginfo}
 \input{macros}
@@
 
-\title{Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation}
+\title{Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes}
 \author{Kevin Tan, Giles Hooker, Edward L. Ionides}
-\correspondingauthor{Corresponding Author name.\\E-mail: kevtan@wharton.upenn.edu}
+\correspondingauthor{Corresponding Author name.\\E-mail: ionides@umich.edu}
 

### 2024-11-06 paper/si.tex
Updates from Overleaf

@@
 \documentclass[numsec,webpdf,modern,medium,namedate]{oup-authoring-template}
+\onecolumn
 
+\usepackage{booktabs}
+\newcommand{\description}{}
+\usepackage{enumitem}
+\usepackage{tabto}
+
+\theoremstyle{thmstyleone}%
+\newtheorem{theorem}{Theorem}%  meant for continuous numbers
+%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
+%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
+\newtheorem{proposition}[theorem]{Proposition}%
+%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
+\theoremstyle{thmstyletwo}%
+\newtheorem{example}{Example}%
+\newtheorem{remark}{Remark}%
+\theoremstyle{thmstylethree}%
+\newtheorem{definition}{Definition}
+\usepackage[doublespacing]{setspace}
+\usepackage[fontsize=12pt]{fontsize}
+
+\input{macros}
 % Use the lineno option to display guide line numbers if required.
-\usepackage{xr}
+\usepackage{xr, refcount}
 \makeatletter
@@
 
-%\templatetype{pnassupportinginfo}
-\input{macros}
-\usepackage{refcount}
+\journaltitle{Journals of the Royal Statistical Society}
+\DOI{DOI HERE}
+\copyrightyear{XXXX}
+\pubyear{XXXX}
+\access{Advance Access Publication Date: Day Month Year}
+\appnotes{Original article}
 
-\setcounter{thm}{\getrefnumber{finalthm}}
-\addtocounter{thm}{-1}
+\firstpage{1}
 
-\setcounter{lem}{\getrefnumber{finallem}}
-\addtocounter{lem}{-1}
+%\subtitle{Subject Section}
+\title[AD for POMPs]{Web-Based Supporting Materials for Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes}
 
-\setcounter{defn}{\getrefnumber{finaldefn}}
-\addtocounter{defn}{-1}
+\author[1,$\ast$]{Kevin Tan}
+\author[1]{Giles Hooker}
+\author[2]{Edward L. Ionides}
+%\author[4]{Fifth Author\ORCID{0000-0000-0000-0000}}
+\authormark{Tan et al.}
 
-\setcounter{prop}{\getrefnumber{finalprop}}
-\addtocounter{prop}{-1}
+\address[1]{\orgdiv{Department of Statistics and Data Science}, \orgname{University of Pennsylvania}, \orgaddress{\street{265 South 37th Street, 3rd \& 4th Floors}, \postcode{19104}, \state{Pennsylvania}, \country{United States of America}}}
+\address[2]{\orgdiv{Department of Statistics}, \orgname{University of Michigan}, \orgaddress{\street{323 West Hall, 1085 S University Ave}, \postcode{48109}, \state{Michigan}, \country{United States of America}}}
 
-\setcounter{aspt}{\getrefnumber{finalaspt}}
-\addtocounter{aspt}{-1}
+\corresp[$\ast$]{Address for correspondence. Kevin Tan, University of Pennsylvania, Philadelphia, 19104, USA. \href{Email:kevtan@wharton.upenn.edu}{kevtan@wharton.upenn.edu}}
+
+% \received{Date}{0}{Year}
+% \revised{Date}{0}{Year}
+% \accepted{Date}{0}{Year}
+\AtBeginDocument{\setcounter{thm}{\getrefnumber{finalthm}}}
+\AtBeginDocument{\setcounter{lem}{\getrefnumber{finallem}}}
+\AtBeginDocument{\setcounter{defn}{\getrefnumber{finaldefn}}}
+\AtBeginDocument{\setcounter{prop}{\getrefnumber{finalprop}}}
+\AtBeginDocument{\setcounter{aspt}{\getrefnumber{finalaspt}}}
+\AtBeginDocument{\setcounter{figure}{\getrefnumber{finalfig}}}
 
@@
 
-\title{Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes}
-\author{Kevin Tan, Giles Hooker, Edward L. Ionides}
-\correspondingauthor{Corresponding Author name.\\E-mail: ionides@umich.edu}
+%\addtocounter{thm}{-1}
 
-%% Comment out or remove this line before generating final copy for submission; this will also remove the warning re: "Consecutive odd pages found".
-%\instructionspage
+%\addtocounter{lem}{-1}
+
+%\addtocounter{defn}{-1}
+
+%\addtocounter{prop}{-1}
+
+%\addtocounter{aspt}{-1}
+
+\abstract{Although automatic differentiation (AD) has driven many recent advances in machine learning, partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques because widely used particle filter algorithms yield an estimated likelihood that is discontinuous in the model parameters.
+To resolve this, we create a theoretical framework that embeds two existing AD particle filter methods within a new class of algorithms.
+This new class permits a bias-variance tradeoff and a mean squared error substantially lower than the existing algorithms.
+This allows us to develop likelihood maximization algorithms suited to the Monte Carlo properties of the AD gradient estimate.
+Our algorithms require only a differentiable simulator for the latent dynamic system while most previous approaches require access to the system's transition probabilities.
+Numerical results show that using AD to refine a coarse solution from an iterated filtering algorithm shows substantial improvement on current state-of-the-art methods on a challenging scientific benchmark problem.}
 
 \maketitle
-
-%% Adds the main heading for the SI text. Comment out this line if you do not have any supporting information text.
-\SItext
 
@@
 
+\bibliographystyle{apalike}
 \bibliography{paper/bib-ifad}

### 2024-11-26 paper/appendix/bayes.tex
Updates from Overleaf

@@
 \end{figure}
+\FloatBarrier
+\newpage
 
@@
     \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Measurement_Noise.png}
-    \caption{Posterior estimates from NUTS powered by MOP-$\alpha$ with the informative nonparametric empirical prior obtained from a kernel density estimator on the IF2 parameter cloud.
+    \caption{Posterior estimates (on the logarithmic scale for everything but the trend in force of infection) from NUTS powered by MOP-$\alpha$ with the informative nonparametric empirical prior obtained from a kernel density estimator on the IF2 parameter cloud.
 The posterior estimates from each chain are largely in agreement.}
@@
 \end{figure}
+
+\FloatBarrier
+\newpage

### 2024-11-26 paper/ms.tex
Updates from Overleaf

@@
 %%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
-\theoremstyle{thmstyletwo}%
+\theoremstyle{thmstyletwo}%w
 \newtheorem{example}{Example}%
@@
 \AtEndDocument{\refstepcounter{aspt}\label{finalaspt}}
+\AtEndDocument{\refstepcounter{figure}\label{finalfig}}
 
@@
 However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability, which led the authors to reject this algorithm in favor of an approach without the plug-and-play property and with inferior scalability.
+\textit{Our approach is the most significant methodological advance since \cite{ionides15}, in nearly a decade, for plug-and-play statistical inference on POMP models.}
 
@@
 
-\arxiv{}{\vspace*{-3ex}}
 \section{Problem Setup}
@@
 
-\arxiv{}{\vspace*{-3ex}}
 \section{Off-Parameter Particle Filters}
@@
 
-Our approach has similarities with \cite{svensson18}, in which particles from a previous run at $\phi$ are reweighted to evaluate the likelihood at any sufficiently nearby parameter $\theta$ for $0$-th order optimization.
-However, their method is not plug-and-play because they use transition densities to reweight particles, and the variance of their log-likelihood estimate is $O(||\theta - \phi||_2^{2N})$ with an exponential dependence in the horizon.
-Our approach bypasses these issues -- the first with the \textit{reparameterization trick} to consider particle paths dependent only on $\theta$, and the second as using AD allows us to only evaluate the likelihood and score estimates when $\theta=\phi$.
-%The use of AD bypasses the issue of exponentially increasing variance, but nevertheless can result in problematic polynomial variance.
-% Our introduction of $\alpha$ avoids that difficulty.
-%also roughly $O(||\theta - \theta_{k-1}||_2^{2N})$, where $N$ is
-
-% Theorem~\ref{thm:mop-functional-forms} shows that MOP-$\alpha$ is constructed so this direct derivative obtains the score estimator of  \cite{poyiadjis11, scibior21} when $\alpha=1$ and that of \cite{naesseth18} when $\alpha=0$.
-% Setting $\alpha<1$ adds bias but reduces variance, producing opportunities for a favorable tradeoff as illustrated in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).
+\begin{figure}[htbp!]
+  \centering
+    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2}}]{imgs/095/mop.png}
+    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2.2}}]{imgs/095/biasvar.png}
+    \arxiv{}{\vspace*{-3ex}}
+    \caption{\textbf{Left:} Non‐differentiable likelihood estimates and our smooth off‐parameter approximations that hold locally around the baseline parameter $\phi$.
+\textbf{Right:} Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}.
+We display the MSE of score estimates for the trend in transmission at the MLE.
+Our novel contribution of $\alpha \in (0,1)$ significantly improves the quality of the likelihood approximation, and therefore the gradient estimates.
+}
+    \label{fig:biasvar}
+    \arxiv{}{\vspace*{-4ex}}
+\end{figure}
 
@@
 
-\arxiv{}{\vspace*{-3ex}}
+Our approach has similarities with \cite{svensson18}, in which particles from a previous run at $\phi$ are reweighted to evaluate the likelihood at any sufficiently nearby parameter $\theta$ for $0$-th order optimization.
+However, their method is not plug-and-play because they use transition densities to reweight particles, and the variance of their log-likelihood estimate is $O(||\theta - \phi||_2^{2N})$ with an exponential dependence in the horizon.
+Our approach bypasses these issues -- the first with the \textit{reparameterization trick} to consider particle paths dependent only on $\theta$, and the second as using AD allows us to only evaluate the likelihood and score estimates when $\theta=\phi$.
+%The use of AD bypasses the issue of exponentially increasing variance, but nevertheless can result in problematic polynomial variance.
+% Our introduction of $\alpha$ avoids that difficulty.
+%also roughly $O(||\theta - \theta_{k-1}||_2^{2N})$, where $N$ is
+
+% Theorem~\ref{thm:mop-functional-forms} shows that MOP-$\alpha$ is constructed so this direct derivative obtains the score estimator of  \cite{poyiadjis11, scibior21} when $\alpha=1$ and that of \cite{naesseth18} when $\alpha=0$.
+% Setting $\alpha<1$ adds bias but reduces variance, producing opportunities for a favorable tradeoff as illustrated in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}).
+
 \subsection{Algorithm Outline}
@@
 
-\arxiv{}{\vspace*{-3ex}}
 \subsection{Discounting Weights for a Bias-Variance Tradeoff}
@@
 
-\arxiv{}{\vspace*{-3ex}}
 \subsection{MOP-$\alpha$ Encompasses the Estimators of \citep{poyiadjis11, naesseth18}}
@@
 
-\begin{figure}[htbp!]
-\arxiv{}{\vspace*{-2ex}}
-  \centering
-    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2}}]{imgs/095/mop.png}
-    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2.2}}]{imgs/095/biasvar.png}
-    \arxiv{}{\vspace*{-3ex}}
-    \caption{\textbf{Left:} Non‐differentiable likelihood estimates and our smooth off‐parameter approximations that hold locally around the baseline parameter $\phi$.
-\textbf{Right:} Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}.
-We display the MSE of score estimates for the trend in transmission at the MLE.
-Our novel contribution of $\alpha \in (0,1)$ significantly improves the quality of the likelihood approximation, and therefore the gradient estimates.
-}
-    \label{fig:biasvar}
-    \arxiv{}{\vspace*{-4ex}}
-\end{figure}
-
 % \arxiv{}{\vspace*{-3ex}}
@@
 
-\arxiv{}{\vspace*{-3mm}}
 \section{Practical Maximum-Likelihood Estimation}
-\arxiv{}{\vspace*{-1mm}}
+
 %One can think of MOP-$\alpha$ as taking a likelihood estimate at $\phi$, reweighting it to instead target the likelihood at $\theta$, and then taking the derivative with respect to $\theta$ of the above estimate.
@@
 
-\begin{algorithm}[h!]
+\begin{algorithm}[htbp!]
 	\caption{IFAD}
@@
 
-\arxiv{}{\vspace*{-1ex}}
 \subsection{Optimization}
@@
 
-\arxiv{}{\vspace*{-2ex}}
 \subsection{Linear Convergence Rates}
@@
 
-\arxiv{}{\vspace*{-3ex}}
 \section{Application to a Cholera Transmission Model}
@@
 
-\arxiv{}{\vspace*{-3ex}}
 \subsection{Results}
@@
     \centering
+    \vspace{1mm}
     \includegraphics[width=\arxiv{8cm}{\textwidth/\real{2.2}}]{imgs/095/pairs.png}
@@
     \centering
+    \vspace{1mm}
     \includegraphics[width=\arxiv{10cm}{\textwidth/2}]{imgs/095/optim.png}
@@
 
-\arxiv{}{\vspace*{-3ex}}
 \section{Application to Bayesian Inference}
@@
     \centering
-    \includegraphics[width=\arxiv{\textwidth/\real{1.25}}{\textwidth}]{imgs/pmcmc/nuts_eb.png}
-    \arxiv{}{\vspace*{-1mm}}
+    \vspace{1mm}
+    \includegraphics[width=\arxiv{\textwidth/\real{1.25}}{\textwidth}/\real{1.25}]{imgs/pmcmc/nuts_eb.png}
     \caption{Histogram and density estimate of the posterior of $\beta_{\text{trend}}$, as well as convergence diagnostics for NUTS with $4$ chains.
@@
 
-\arxiv{}{\vspace*{-3ex}}
 \section{Theoretical Analysis of MOP-$\alpha$}
@@
 
-\arxiv{}{\vspace*{-3ex}}
 \subsection{MOP-$\alpha$ Targets the Filtering Distribution}
@@
     resampling the $k_j$ with probabilities proportional to $g^\phi_{n,j}$, also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
-    Estimating the likelihood with the before-resampling estimator $\hat\lik(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, we see that the strong consistency is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
+    Estimating the likelihood with the before-resampling estimator $\hat\lik(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, we see that the strong consistency is a direct consequence of our result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
 \end{proof}
 
-\arxiv{}{\vspace*{-4mm}}
 \subsection{MOP-$1$ Score Estimate Is Consistent}
@@
 
-\arxiv{}{\vspace*{-3ex}}
 \subsection{MOP-$\alpha$ Error, Bias and Variance}
@@
         \\
-        &\Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) \lesssim \min_{k\leq N} Np  G'(\theta)^2\hspace{-0.8ex}\left(\hspace{-0.8ex}\frac{k^2}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}\hspace{-0.8ex}\right).
-\label{eq:mop-variance}
+        &\Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) \lesssim \min_{k\leq N} Np  G'(\theta)^2\left(\frac{k^2}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}\right). \label{eq:mop-variance}
         \end{align}}
@@
 
-\arxiv{}{\vspace*{-1mm}}
 \subsection{Interpretation:} Theorem \ref{thm:mop-biasvar} provides theoretical understanding of the empirically favorable bias-variance tradeoff enjoyed by MOP-$\alpha$ in Figure \ref{fig:biasvar}.
@@
 
-\arxiv{}{\vspace*{-1mm}}
 \subsection{Bias-Variance Tradeoff:} We show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( Np \, G'(\theta)^2\big/J \big)$.
@@
 
-\arxiv{}{\vspace*{-3ex}}
 \section{Computational Efficiency}
@@
 This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter (with a simulator written in C++) in the \texttt{pomp} package of \cite{king16}.
+This has since spurred the development of a new open source Python library named \texttt{pypomp}\footnote{The package can be found at \href{https://pypi.org/project/pypomp/}{https://pypi.org/project/pypomp/}.
+The source code for this paper and original implementation can be found at \href{https://github.com/hetankevin/diffPomp}{https://github.com/hetankevin/diffPomp}.}, which incorporates the improvements in methodology and computation described within this paper.
 
@@
 
-\arxiv{}{\vspace*{-3ex}}
 \section{Discussion}
@@
 
-\arxiv{}{\vspace*{-3ex}}
 \subsection{Further Work}

### 2024-11-26 paper/ms-kevtan.tex
changed plots

--- 2024-10-17 paper/ms.tex
@@
 %\documentclass[9pt,twocolumn,twoside]{pnas-new}
-\documentclass[numsec,webpdf,modern,medium,namedate]{oup-authoring-template}
+\documentclass[9pt,twocolumn,pnasresearcharticle]{pnas-new}
 %\documentclass[9pt,twocolumn,twoside]{pnas-new}
 % Use the lineno option to display guide line numbers if required.
-\onecolumn
 
 \usepackage{booktabs}
-\newcommand{\description}{}
-\usepackage{enumitem}
 
@@
 }{
-\theoremstyle{thmstyleone}%
-\newtheorem{theorem}{Theorem}%  meant for continuous numbers
-%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
-%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
-\newtheorem{proposition}[theorem]{Proposition}%
-%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
-\theoremstyle{thmstyletwo}%
-\newtheorem{example}{Example}%
-\newtheorem{remark}{Remark}%
-\theoremstyle{thmstylethree}%
-\newtheorem{definition}{Definition}
-\usepackage[doublespacing]{setspace}
-\usepackage[fontsize=12pt]{fontsize}}
+\templatetype{pnasresearcharticle} % Choose template
+% {pnasresearcharticle} = Template for a two-column research article
+% {pnasmathematics} %= Template for a one-column mathematics article
+% {pnasinvited} %= Template for a PNAS invited submission
+}
 
@@
 
-\arxiv{\title{Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes}
+\title{Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes}
 
-\author{K. Tan$^{1}$, G. Hooker$^{1}$ and E. L. Ionides$^{2}$
+%\arxiv{ \author{K. Tan, G. Hooker and E. L. Ionides\\
+%  \small{Department of Statistics, University of Pennsylvania}\\
+%  \small{Department of Statistics, University of Michigan}}}{}
+
+\arxiv{ \author{K. Tan$^{1}$, G. Hooker$^{1}$ and E. L. Ionides$^{2}$
   \vspace{2mm}\\
   \small{$^{1}$Department of Statistics, University of Pennsylvania}\\
-  \small{$^{2}$Department of Statistics, University of Michigan}}
+  \small{$^{2}$Department of Statistics, University of Michigan}}}{}
 
-}{
-
-\journaltitle{Journals of the Royal Statistical Society}
-\DOI{DOI HERE}
-\copyrightyear{XXXX}
-\pubyear{XXXX}
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Original article}
-
-\firstpage{1}
-
-%\subtitle{Subject Section}
-\title[AD for POMPs]{Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes}
-
-\author[1,$\ast$]{Kevin Tan}
-\author[1]{Giles Hooker}
-\author[2]{Edward L. Ionides}
-%\author[4]{Fifth Author\ORCID{0000-0000-0000-0000}}
-\authormark{Tan et al.}
-
-\address[1]{\orgdiv{Department of Statistics and Data Science}, \orgname{University of Pennsylvania}, \orgaddress{\street{265 South 37th Street, 3rd \& 4th Floors}, \postcode{19104}, \state{Pennsylvania}, \country{United States of America}}}
-\address[2]{\orgdiv{Department of Statistics}, \orgname{University of Michigan}, \orgaddress{\street{323 West Hall, 1085 S University Ave}, \postcode{48109}, \state{Michigan}, \country{United States of America}}}
-
-\corresp[$\ast$]{Address for correspondence. Kevin Tan, University of Pennsylvania, Philadelphia, 19104, USA. \href{Email:kevtan@wharton.upenn.edu}{kevtan@wharton.upenn.edu}}
-
-\received{Date}{0}{Year}
-\revised{Date}{0}{Year}
-\accepted{Date}{0}{Year}
-
+\arxiv{}{
+% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
+\author[a]{Kevin Tan}
+\author[a]{Giles Hooker}
+\author[b,1]{Edward L. Ionides}
+%\author[a]{Author Three}
+\affil[a]{University of Pennsylvania}
+\affil[b]{University of Michigan}
+%\affil[c]{Affiliation Three}
+% Please give the surname of the lead author for the running footer
+\leadauthor{Tan}
+% Please add a significance statement to explain the relevance of your work
+\significancestatement{Many scientific models involve highly nonlinear stochastic dynamical systems which can be observed only via noisy and incomplete measurements.
+Prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics.
+We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation.
+Our new method outperforms previous approaches on a challenging problem in epidemiology.}
+\authorcontributions{K.T. and E.L.I. designed research; K.T. analyzed data; K.T., G.H. and E.L.I. performed research, K.T., G.H., and E.L.I. wrote the manuscript.}
+\authordeclaration{The authors declare no competing interests.}
+\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: ionides@umich.edu}
+% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
+\keywords{Sequential Monte Carlo $|$ Automatic Differentiation $|$ Particle Filter $|$ Markov Process $|$ Likelihood}
 }
 
-\arxiv{
+\arxiv{\date{Draft compiled on \today}}{
+\dates{This manuscript was compiled on \today}
+\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
+}
+
+\maketitle
+
 \begin{abstract}
@@
 Our algorithms require only a differentiable simulator for the latent dynamic system; by contrast, most previous approaches to AD likelihood maximization for particle filters require access to the system's transition probabilities.
-Numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from an iterated filtering algorithm shows substantial improvement on current state-of-the-art methods on a challenging scientific benchmark problem.
+Numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from an iterated filtering algorithm show substantial improvement on current state-of-the-art methods for a challenging scientific benchmark problem.
 \end{abstract}
-}{
 
-\abstract{Although automatic differentiation (AD) has driven many recent advances in machine learning, partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques because widely used particle filter algorithms yield an estimated likelihood that is discontinuous in the model parameters.
-To resolve this, we create a theoretical framework that embeds two existing AD particle filter methods within a new class of algorithms.
-This new class permits a bias-variance tradeoff and a mean squared error substantially lower than the existing algorithms.
-This allows us to develop likelihood maximization algorithms suited to the Monte Carlo properties of the AD gradient estimate.
-Our algorithms require only a differentiable simulator for the latent dynamic system while most previous approaches require access to the system's transition probabilities.
-Numerical results show that using AD to refine a coarse solution from an iterated filtering algorithm shows substantial improvement on current state-of-the-art methods on a challenging scientific benchmark problem.}
-\keywords{Sequential Monte Carlo, Automatic Differentiation, Particle Filter, Markov Process, Likelihood}
+\arxiv{}{
+\thispagestyle{firststyle}
+\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
+\firstpage{4}
+% Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting.
+In this example, there are a total of 11 paragraphs on the first page, counting the first level heading as a paragraph.
+The value {12} represents the number of the paragraph starting the second page.
+If a paragraph runs over onto the second page, include a bracket with the paragraph line number starting the second page, followed by the paragraph number in curly brackets, e.g.
+"\firstpage[4]{11}".
 }
 
-\maketitle
-
-\section{Introduction}
-
-Many scientific models involve highly nonlinear stochastic dynamic systems possessing significant random variation in both the process dynamics and the measurements.
+\arxiv{Many}{\dropcap{M}any}
+scientific models involve highly nonlinear stochastic dynamic systems possessing significant random variation in both the process dynamics and the measurements.
 Commonly, the latent system is modeled as a Markov process, giving rise to a partially observed Markov process (POMP) model, also known as a hidden Markov models or a state space model.
-POMP models arise in fields as diverse as optimal control \citep{singh22}, epidemiology \citep{he10, stocks17}, ecology \citep{knape12} and finance \citep{kim08, breto14}.
-Despite their ubiquity, estimation and inference for this broad class of models remains a challenging problem.
-To tackle this, we use automatic differentiation (AD) to construct improved algorithms for complex POMP models.
+POMP models arise in fields as diverse as optimal control \cite{singh22}, epidemiology \cite{he10, stocks17}, ecology \cite{knape12} and finance \cite{kim08, breto14}.
+Despite their ubiquity, estimation and inference within this broad class of models remains a challenging problem.
+To tackle this, we use automatic differentiation (AD) to construct improved algorithms for inference on complex POMP models.
 
 The particle filter, also known as sequential Monte Carlo, serves as the foundation for various inference algorithms for POMP models.
-It provides an unbiased estimate of the likelihood function  \citep{delMoral04}, enabling Bayesian inference \citep{andrieu10,chopin13} and likelihood-based inference \citep{ionides06-pnas,ionides15}.
-Likelihood-based inference is statistically efficient \citep{pawitan01} and robust to a moderate amount of Monte Carlo error \citep{ionides17,ning21}.
-Furthermore, basic particle filters, also known as boostrap filters, do not need to evaluate the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
-
-This \textit{plug-and-play} property is useful for scientific applications \citep{he10}.
-Some plug-and-play methods depend on the construction of low-dimensional summary statistics \citep{wood10,toni09}, sacrificing statistical efficiency for computational convenience.
-Plug-and-play methods are often called likelihood-free \citep{owen15}, and it may be counter-intuitive that likelihood-based inference is possible using likelihood-free methods. 
-However, iterated filtering algorithms have shown that this is practical in various scientific investigations \citep{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
+It provides an unbiased estimate of the likelihood function  \cite{delMoral04}, enabling Bayesian inference \cite{andrieu10,chopin13} and likelihood-based inference \cite{ionides06-pnas,ionides15}.
+This is desirable, as likelihood-based inference is statistically efficient \cite{pawitan01} and robust to a moderate amount of Monte Carlo error \cite{ionides17,ning21}.
+Furthermore, basic particle filter algorithms, also known as boostrap filters, do not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
+This \textit{plug-and-play} property is useful for scientific applications \cite{he10}.
+Some plug-and-play methods depend on the construction of low-dimensional summary statistics \cite{wood10,toni09}, sacrificing statistical efficiency for computational convenience.
+Plug-and-play methods are often called likelihood-free \cite{owen15}, and it may be counter-intuitive that likelihood-based inference is possible using likelihood-free methods. 
+However, iterated filtering algorithms have shown that this is practical in various scientific investigations, such as \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
 Nevertheless, the Monte Carlo variability arising in iterated filtering applications becomes increasingly problematic as the size of the data and the complexity of the model increase.
@@
 We examine plug-and-play AD as a tool to do so by providing a likelihood gradient.
-Recent advances in AD for particle filters \citep{naesseth18, jonschkowski18, corenflos21, scibior21, singh22} have drawn attention to AD as a tool for inference in POMP models.
-However, existing approaches are either asymptotically biased \citep{naesseth18, jonschkowski18}, have high variance \citep{poyiadjis11, scibior21}, are computationally expensive \citep{corenflos21, chen24}, or require access to transition densities \citep{poyiadjis11, scibior21, singh22, chen24}.
+Recent advances in AD for particle filters \cite{naesseth18, jonschkowski18, corenflos21, scibior21, singh22} have drawn attention to AD as a tool for inference in POMP models.
+However, existing approaches are either asymptotically biased \cite{naesseth18, jonschkowski18}, have high variance \cite{poyiadjis11, scibior21}, are computationally expensive \cite{corenflos21, chen24}, or require access to transition densities \cite{poyiadjis11, scibior21, singh22, chen24}.
+Scibior and Wood \cite{scibior21} showed that the estimators derived by Poyiadjis et al.
+\cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.
 
-\cite{scibior21} showed that the estimators derived by \cite{poyiadjis11} can be attained with standard AD software using an algorithmic procedure, called a {\it stop-gradient}, which allows selected expressions to be evaluated but not differentiated.
-However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability, which led the authors to reject this algorithm in favor of an approach without the plug-and-play property and with inferior scalability.
+However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability \cite{poyiadjis11}.
+Additionally, while the stop-gradient procedure is undeniably effective, the reason why the correction recovers the \cite{poyiadjis11} estimator remains somewhat unclear.
+We use \cite{scibior21} and \cite{poyiadjis11} as our starting point in developing a new approach remedying these issues.
 
-We start by presenting a new algorithm which (for reasons which will be discussed later) we call a {\it Measurement Off-Parameter} (MOP) particle filter. 
-In summary, \textit{instead of directly differentiating through a basic particle filter, MOP uses AD to differentiate through a series of measurement density ratios and a differentiable simulator}. 
-Our MOP algorithm differs from previous smooth particle filters \citep{svensson18,malik11} by possessing the plug-and-play property.
+We start by presenting a new construction of the gradient estimator of \cite{scibior21} and \cite{poyiadjis11}.
+Specifically, we show how this estimator can be derived as the direct derivative of a suitably weighted particle filter, which (for reasons which will be discussed later) we call a {\it Measurement Off-Parameter} (MOP) particle filter.
+In short, \textit{instead of directly differentiating through a basic particle filter, MOP uses AD to differentiate through a series of measurement density ratios and a differentiable simulator}. 
+Our MOP algorithm differs from previous smooth particle filters \cite{svensson18,malik11} by possessing the plug-and-play property.
 Critically, we also avoid the high variance of \cite{svensson18} by generalizing the MOP particle filter to add a discount factor, $\alpha \in [0,1]$.
-Remarkably, we show that the derivative of our algorithm coincides with the biased gradient estimator of \cite{naesseth18}, when $\alpha=0$, and the high-variance gradient estimate from \citep{poyiadjis11, scibior21}, when $\alpha=1$ (Theorem \ref{thm:mop-functional-forms}).
+This discount factor enables MOP-$\alpha$ to interpolate between the  biased gradient estimator of \cite{naesseth18}, when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21}, when $\alpha=1$ (Theorem \ref{thm:mop-functional-forms}).
 The bias-variance tradeoff induced by $\alpha$ suggests, both in theory (Theorem \ref{thm:mop-biasvar}) and in practice (Figure \ref{fig:biasvar}), a choice of $\alpha$ strictly in $(0,1)$.
-Choosing $\alpha$ close to one, we obtain a low-bias, low-variance, plug-and-play gradient estimate with computational scaling linear in the number of particles.
 
+% One can simply drop terms arising from resampling when differentiating through the particle filter \cite{naesseth18}, but this leads to asymptotic bias \cite{corenflos21}.
+% Instead, MOP-$\alpha$ bypasses the seemingly incompatible paradox of differentiating through discrete Monte Carlo sampling by instead using AD to differentiate through a series of measurement density ratios and a differentiable simulator.
 MOP-$\alpha$ can provide strongly consistent estimates of the filtering distribution and the log-likelihood (Theorem \ref{thm:mop-targeting}).
 When $\alpha=1$, MOP-$\alpha$ also provides consistent score estimates (Theorem \ref{thm:mop-grad-consistency}).
-This lets us derive a linear convergence rate for gradient descent with MOP-$\alpha$ in the strongly convex and smooth setting (Theorem \ref{thm:mop-convergence}).
+This enables us to derive a linear convergence rate for gradient descent using MOP-$\alpha$ in the  presence of strong convexity and smoothness (Theorem \ref{thm:mop-convergence}).
 
 In practice, gradient descent with MOP-$\alpha$ converges quickly within a neighborhood of the maximum but struggles to reach this neighborhood in the presence of local minima and saddle points.
-That behavior is complementary to that of iterated filtering algorithms \citep{ionides06-pnas,ionides15} which provide a relatively fast and stable way to identify this neighborhood.
+That behavior is complementary to that of iterated filtering algorithms \cite{ionides06-pnas,ionides15} which provide a relatively fast and stable way to identify this neighborhood.
 We therefore build a simple but effective hybrid algorithm called {\it Iterated Filtering with Automatic Differentiation} (IFAD) that warm-starts first-order or second-order gradient methods with a preliminary solution obtained from a few rounds of iterated filtering.
-Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 \citep{ionides06-pnas,ionides11} and the Liu-West filter \citep{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.
+Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 \cite{ionides06-pnas,ionides11} and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.
 
 These improvements also extend to Bayesian inference.
-In Section \ref{sec:bayes}, we use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \citep{homan14} in conjunction with an empirical Bayes-type prior estimated with IF2 to reduce the burn-in period of particle MCMC \citep{andrieu10} on the Dhaka cholera model of \cite{king08} from the $700,000$ iterations in \cite{fasiolo16} to just $500$.
+In Section \ref{sec:bayes}, we use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with an empirical Bayes-type prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{andrieu10} on the Dhaka cholera model of \cite{king08} from the $700,000$ iterations in \cite{fasiolo16} to just $500$.
 To the best of our knowledge, attaining rapid mixing on this challenging problem has not previously been achieved.
@@
 \section{Problem Setup}
+
+Consider an unobserved Markov process $\{X(t),t  \geq t_0\}$, with discrete-time observations $Y_1,...,Y_N$ realized at values $y_1^*,...,y_N^*$ at times $t_1,..., t_N$.
+The process is parameterized by an unknown parameter $\theta \in \Theta \subseteq \R^p$, where the state $X(t)$ take values in the state space $\gX \subseteq \R^d$, the observations $Y_n$ take values in $\gY,$ and we write $X_n := X(t_n)$.
 
@@
 
-Consider an unobserved Markov process $\{X(t),t  \geq t_0\}$, with discrete-time observations $Y_1,...,Y_N$ realized at values $y_1^*,...,y_N^*$ at times $t_1,..., t_N$.
-The process is parameterized by an unknown parameter $\theta \in \Theta \subseteq \R^p$, where the state $X(t)$ takes values in the state space $\gX \subseteq \R^d$, the observations $Y_n$ take values in $\gY,$ and we write $X_n := X(t_n)$.
 We suppose that the discrete-time latent process model has a density $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$.
@@
 We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.
-$x_{n,j}^A$ denotes the ancestral trajectory of particle $j$ at time $n$,  and $k_j$ the resampling indices for particle $j$. 
+$x_{n,j}^A$ denotes the ancestral trajectory of particle $j$ at time $n$,  and $k_j$
+denotes the resampling indices for particle $j$. 
 %$a(\cdot)$ is the ancestor function that maps a particle to its parent, $j \mapsto a(j)$,
@@
 To that end, we suppose that the simulator is a differentiable function of $\theta$ for every fixed $\omega$, a condition that requires the latent process to be a continuous random variable. 
-This condition is also known as the {\it reparameterization trick} \citep{corenflos21}.
+This condition is also known as the {\it reparameterization trick} \cite{corenflos21}.
 The discrete resampling step is more delicate to address. 
@@
 This ensures that the resampling indices are invariant in $\theta$ when $\omega$ and $\phi$ are fixed, bypassing the issue of Monte Carlo resampling.
-
-One can think of this as reconstructing local smooth approximations of the discontinuous likelihood estimates produced by a particle filter for a fixed number of particles $J$ and fixed seed $\omega$.
-These approximate the likelihood at $\theta$ in a neighborhood around $\phi$, with the intent that differentiating these smooth log-likelihood reconstructions yields a score estimate.
-This is illustrated in the left panel of Figure \ref{fig:biasvar} for various choices of $\alpha$.
 
@@
     \label{alg:mop}
-	     \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process_n(\cdot|x_n; \theta)$, evaluation parameter $\theta$, baseline parameter $\phi$, seed $\omega$.
+	     \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process_n(x_{n+1}|x_n; \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$.
 
@@
 The resampling indices, which we write as $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$, are a function of $\phi$ for any value of $\theta$.
-For the categorical distribution, we use systematic resampling \citep{arulampalam02,king16} which usually has superior performance to multinomial resampling.
+For the categorical distribution, we use systematic resampling \cite{arulampalam02,king16} which usually has superior performance to multinomial resampling.
 In practice, we only ever evaluate MOP-$\alpha$ in the case where $\theta$ and $\phi$ coincide.
@@
 \arxiv{}{\vspace*{-2mm}}
-\subsection{MOP-$\alpha$ Encompasses the Estimators of \citep{poyiadjis11, scibior21, naesseth18}}
+\subsection{MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, naesseth18}}
 
@@
         \nabla_\theta \hat\ell^0(\theta) 
-        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
+        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
         \vspace*{-2.5mm}
@@
     \end{equation}
-    yielding the estimator of \cite{poyiadjis11} for the bootstrap filter.
+    yielding the estimator of \cite{poyiadjis11, scibior21} for the bootstrap filter.
 \end{thm}
@@
   \centering
-    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2}}]{imgs/095/mop.png}
-    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2.2}}]{imgs/095/biasvar.png}
-    \caption{\textbf{Left:} Non‐differentiable likelihood estimates and our smooth off‐parameter approximations that hold locally around the baseline parameter $\phi$.
-\textbf{Right:} Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}.
-We display the MSE of score estimates for the trend in transmission at the MLE.
-Our novel contribution of $\alpha \in (0,1)$ significantly improves the quality of the likelihood approximation, and therefore the gradient estimates.
-}
+    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{imgs/095/biasvar.png}
+    \caption{Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}.
+We display the MSE of score estimates for the trend in transmission, evaluated at the MLE.}
     \label{fig:biasvar}
@@
 However, even when applying much larger amounts of Monte Carlo effort, IF2 fails to get within seven log-likelihood units of the maximum on this problem (Figure~\ref{fig:boxplot}).
-These numerical results are consistent with theoretical results for iterated filtering, showing slow convergence close to the maximum \citep{doucet15-if}.
+These numerical results are consistent with theoretical results for iterated filtering, showing slow convergence close to the maximum \cite{doucet15-if}.
 IF2, without assistance from AD, requires additional computationally intensive techniques, such as repeated pruning and profiling, to reach the inferentially critical region within one log unit of the maximum.
@@
 We therefore see that the second stage of IFAD converges linearly to the MLE if (1) the log-likelihood surface is well-conditioned in a neighborhood of the MLE and (2) the first stage of IFAD successfully reaches a (high-probability) basin of attraction of the MLE.
-Within a neighborhood of the MLE that grows as more data become available \citep{ning21}, local quadratic approximations such as local asymptotic normality apply under general conditions \citep{lecam00}.
-This suggests that the linear convergence may occur often in practice.
+Within a neighborhood of the MLE that grows as more data become available \cite{ning21}, local quadratic approximations such as local asymptotic normality apply under general conditions \cite{lecam00}.
+This suggests that the linear convergence may occur reasonably often in practice.
 
@@
 	    \textbf{Input:} Number of particles $J$, timesteps $N$, IF2 cooling schedule $\eta_m$, MOP-$\alpha$ discounting parameter $\alpha$, $\theta_0$, $m=0.$\newline
-        Run IF2 until initial ``convergence'' under cooling schedule $\eta_m$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
+        Run IF2 until initial "convergence" under cooling schedule $\eta_m$ to obtain $\{\Theta_j, j=1,...,J\}$, set $\theta_m := \frac{1}{J}\sum_{j=1}^J \Theta_j.$\newline
 		\textbf{While} procedure not converged: \newline
@@
 
-The cholera transmission model that \cite{king08} developed for Dhaka, Bangladesh, has been used to benchmark the performance of various POMP inference methods \citep{ionides15, fasiolo16, wycoff24}, and we employ it here for the same purpose.
-This model categorizes individuals in a population as susceptible, $S(t)$, infected, $I(t)$, and recovered, $R(t)$ and so is called an SIRS compartmental model.
+The cholera transmission model that \cite{king08} developed for Dhaka, Bangladesh, has been used to benchmark the performance of various POMP inference methods \cite{ionides15, fasiolo16, wycoff24}, and we employ it here for the same purpose.
+This model categorizes individuals in a population as susceptible, $S(t)$, infected, $I(t)$, and recovered, $R(t)$ and so is called an SIR compartmental model.
 In this case, the compartment $R(t)$ is further subdivided into 
@@
 We write $P(t)$ for the total population, and $M_n$ for the cholera deaths in each month.
-As in \citep{king08, ionides15}, the transition dynamics follow a series of stochastic differential equations:
+As in \cite{king08, ionides15}, the transition dynamics follow a series of stochastic differential equations:
 \vspace*{-1mm}
@@
     \centering
-    \includegraphics[width=\arxiv{14cm}{\textwidth}]{imgs/095/tikzcholera.png}
+    \includegraphics[width=\arxiv{14cm}{\textwidth/2}]{imgs/095/tikzcholera.png}
     \vspace*{-7mm}
@@
 We tested IFAD against IF2 on a global search problem for the Dhaka cholera model.
-We re-implemented IF2 to do so, but we also compare our results with the results of \cite{ionides15} (labeled ``IF2 2015'').
+We re-implemented IF2 to do so, but we also compare our results with the results of \cite{ionides15} (labeled "IF2 2015").
 Our re-implementation reaches a better log-likelihood than that of \cite{ionides15} due to a better choice of algorithmic parameters.
@@
 
-\subsection{IFAD Successfully Finds the MLE:} Previously, an MLE at a log-likelihood of $-3748.6$ was reported by \cite{king16}.
+\paragraph{IFAD Successfully Finds the MLE:} Previously, an MLE at a log-likelihood of $-3748.6$ was reported by \cite{king16}.
 This MLE was obtained with much computational effort, using many global and local IF2 searches and with the assistance of likelihood profiling, 8 years after the model was first proposed by \cite{king08}.
@@
 
-\subsection{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations, IF2 alone ultimately fails to achieve the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter}).
+\paragraph{IFAD Outperforms Both IF2 and MOP Alone:} While IF2 quickly approaches a neighborhood of the MLE within only 40 iterations, IF2 alone ultimately fails to achieve the last few log-likelihood units, as no IF2 search comes within 7 log-likelihood units of the MLE (as seen in Figure \ref{fig:scatter}).
 Conversely, we encountered many failed searches with gradient descent alone.
@@
 \begin{figure}[htbp!]
-    \centering
-    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{2.2}}]{imgs/095/pairs.png}
-    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{2.2}}]{imgs/095/qq.png}
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{imgs/095/pairs.png}
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{imgs/095/qq.png}
     %\arxiv{}{\vspace*{-1mm}}
@@
 \begin{figure}[ht]
-    \centering
-    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{2.2}}]{imgs/095/boxplot.png}
-    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{2.2}}]{imgs/095/boxplot_all.png}
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{imgs/095/boxplot.png}
+    \includegraphics[width=\arxiv{8cm}{\textwidth/\real{4.2}}]{imgs/095/boxplot_all.png}
     %\arxiv{}{\vspace*{-1mm}}
@@
     \centering
-    \includegraphics[width=\arxiv{10cm}{\textwidth/2}]{imgs/095/optim.png}
+    \includegraphics[width=\arxiv{10cm}{\textwidth/3}]{imgs/095/optim.png}
     \arxiv{}{\vspace*{-1mm}}
@@
 
-Particle MCMC \citep{andrieu10} is arguably the most popular method for full-information plug-and-play Bayesian inference for POMP models.
+Particle MCMC \cite{andrieu10} is arguably the most popular method for full-information plug-and-play Bayesian inference for POMP models.
 However, particle Metropolis-Hastings often mixes slowly, requiring long burn-in periods.
@@
 
-To alleviate this, we employ a No U-Turn Sampler (NUTS) \citep{homan14} powered by MOP-$\alpha$, with a nonparametric empirical prior constructed from density estimation of the parameter swarm from the IF2 warm-start.
+To alleviate this, we employ a No U-Turn Sampler (NUTS) \cite{homan14} powered by MOP-$\alpha$, with a nonparametric empirical prior constructed from density estimation of the parameter swarm from the IF2 warm-start.
 %This prior has a standard deviation of $6.3\%$ of the MLE. 
@@
     \centering
-    \includegraphics[width=\arxiv{\textwidth/\real{1.25}}{\textwidth}]{imgs/pmcmc/nuts_eb.png}
+    \includegraphics[width=\arxiv{\textwidth/\real{1.25}}{\textwidth/\real{2.5}}]{imgs/pmcmc/nuts_eb.png}
     \arxiv{}{\vspace*{-1mm}}
@@
 
-In particular, assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} imply that the POMP model satisfies a strong mixing condition, and as such are fairly strong but commonplace in the literature \citep{karjalainen23, delMoral04}.
+In particular, assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} imply that the POMP model satisfies a strong mixing condition, and as such are fairly strong but commonplace in the literature \cite{karjalainen23, delMoral04}.
 The fact that the likelihood estimate yielded by MOP-$\alpha$ has more than two continuous derivatives in $\theta$ follows from the construction of the likelihood estimate in equations \ref{eq:mop-conditional-likelihood} and \ref{eq:weighting-scheme}, as well as Assumptions \ref{assump:bounded-measurement} and \ref{assump:diff-meas-and-sim}.
@@
 
-Although we have shown the MOP-$1$ gradient estimate yields the estimate of \citep{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} estimate the Fisher score by
+Although we have shown the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} estimate the Fisher score by
 $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{0:n,j}^{A, F,\theta}, y_{1:N}^* ; \theta)$ and not
@@
         \\
-        &\Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) \lesssim \min_{k\leq N} Np  G'(\theta)^2\hspace{-0.8ex}\left(\hspace{-0.8ex}\frac{k^2}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}\hspace{-0.8ex}\right).
+        &\Var\big(\nabla_\theta \hat\ell^{\alpha}(\theta)\big) \lesssim \min_{k\leq N} Np  G'(\theta)^2\hspace{-0.8ex}\left(\hspace{-0.8ex}\frac{k^2}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}N\hspace{-0.8ex}\right).
 \label{eq:mop-variance}
@@
 \arxiv{}{\vspace*{-1mm}}
-\subsection{Interpretation:} Theorem \ref{thm:mop-biasvar} provides theoretical understanding of the empirically favorable bias-variance tradeoff enjoyed by MOP-$\alpha$ in Figure \ref{fig:biasvar}.
+\paragraph{Interpretation:} Theorem \ref{thm:mop-biasvar} provides theoretical understanding of the empirically favorable bias-variance tradeoff enjoyed by MOP-$\alpha$ in Figure \ref{fig:biasvar}.
 The first term in the MSE bound in Equation \ref{eq:mop-mse} corresponds to particle approximation error, the second mixing error, and the third and fourth the error between the MOP-$(1,k)$ and MOP-$\alpha$ estimates.
@@
 \arxiv{}{\vspace*{-1mm}}
-\subsection{Bias-Variance Tradeoff:} We show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( Np \, G'(\theta)^2\big/J \big)$.
+\paragraph{Bias-Variance Tradeoff:} We show in \arxiv{Appendix~\ref{appendix:biasvar}}{the supplementary material} that the variance of MOP-$0$ is $\tilde{O}\big( Np \, G'(\theta)^2\big/J \big)$.
 Previously, \cite{poyiadjis11} established that the variance of MOP-$1$ is $\tilde{O}(N^4/J)$, ignoring factors of $p$ and $G'$. 
 Combining these results with Theorem \ref{thm:mop-biasvar} shows that MOP-$\alpha$ interpolates between the variances of MOP-$0$ and MOP-$1$, as $N \leq Nk^2 \leq N^4$, with a phase transition as soon as $\alpha<1$.
-The phase transition arises because even though the particle filter has forgetting properties \citep{karjalainen23}, the resulting derivative estimate of \cite{poyiadjis11} does not, and we require both for the variance reduction.
+The phase transition arises because even though the particle filter has forgetting properties \cite{karjalainen23}, the resulting derivative estimate of \cite{poyiadjis11} does not, and we require both for the variance reduction.
 
@@
 In our simulations, MOP-$\alpha$ took 3.75 times the runtime of the particle filter.
-MOP-$\alpha$ and IFAD therefore share the same $O(NJ)$ time complexity as the particle filter, unlike the $O(NJ^2)$ complexity of \cite{corenflos21} and Algorithm 2 in \citep{poyiadjis11, scibior21}.
+MOP-$\alpha$ and IFAD therefore share the same $O(NJ)$ time complexity as the particle filter, unlike the $O(NJ^2)$ complexity of \cite{corenflos21} and Algorithm 2 in \cite{poyiadjis11, scibior21}.
 
@@
 
-Practical likelihood-based data analysis involves many likelihood optimizations \citep{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
+Practical likelihood-based data analysis involves many likelihood optimizations \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
 This occurs with profile likelihood confidence intervals and model selection.
@@
 
-Our gradient estimate can be used in variational inference for the posterior over latent states \citep{naesseth18} and parameters.
+Our gradient estimate can be used in variational inference for the posterior over latent states \cite{naesseth18} and parameters.
 
@@
 
+\arxiv{
 \section*{Acknowledgments}
+This research was supported by National Science Foundation grant DMS-1761603. 
+We thank Nicolas Chopin for helpful communications regarding the strong law of large numbers for off-parameter resampled particle filters.
+We also thank Arnaud Doucet for helpful discussion regarding the manuscript.
+}{
+\showmatmethods{} % Display the Materials and Methods section
+\acknow{
 This research was supported by National Science Foundation grants DMS-1761603 and DEB-1933497. 
 We thank Nicolas Chopin for helpful communications regarding the strong law of large numbers for off-parameter resampled particle filters.
-We also thank Jun Chen and Arnaud Doucet for helpful discussions regarding the manuscript.
-% }{
-% \showmatmethods{} % Display the Materials and Methods section
-% \acknow{
-% This research was supported by National Science Foundation grants DMS-1761603 and DEB-1933497. 
-% We thank Nicolas Chopin for helpful communications regarding the strong law of large numbers for off-parameter resampled particle filters.
 We also thank Jun Chen and Arnaud Doucet for helpful discussions regarding the manuscript.}
-% \showacknow{} % Display the acknowledgments section
-% }
+\showacknow{} % Display the acknowledgments section
+}
 
@@
 
-\bibliographystyle{abbrvnat}
+% \bibliographystyle{pnas-new}
 % Bibliography

### 2024-11-26 archive/filtering.py

@@ -1,0 +1,295 @@
+
+
+import jax
+import itertools
+import numpy as onp
+import jax.numpy as np
+import ipywidgets as widgets
+
+from jax.numpy.linalg import inv, pinv
+from scipy.linalg import solve_discrete_are as dare
+from jax import jit, grad
+from functools import partial
+from IPython import display
+from toolz.dicttoolz import valmap, itemmap
+from itertools import chain
+
+from tqdm.notebook import tqdm
+from tensorflow_probability.substrates import jax as tfp
+tfd = tfp.distributions
+tfb = tfp.bijectors
+tfpk = tfp.math.psd_kernels
+
+import matplotlib.pyplot as plt
+plt.style.use('matplotlibrc')
+
+from resampling import *
+from pomps import rinit, rprocess, dmeasure, rinits, rprocesses, dmeasures
+
+def resampler(counts, particlesP, norm_weights):
+    J = norm_weights.shape[-1]
+    counts = resample(norm_weights)
+    particlesF = particlesP[counts]
+    norm_weights = norm_weights[counts] - jax.lax.stop_gradient(norm_weights[counts]) - np.log(J)
+    return counts, particlesF, norm_weights
+
+def no_resampler(counts, particlesP, norm_weights):
+    return counts, particlesP, norm_weights
+
+def resampler_thetas(counts, particlesP, norm_weights, thetas):
+    J = norm_weights.shape[-1]
+    counts = resample(norm_weights)
+    particlesF = particlesP[counts]
+    norm_weights = norm_weights[counts] - jax.lax.stop_gradient(norm_weights[counts]) - np.log(J)
+    thetasF = thetas[counts]
+    return counts, particlesF, norm_weights, thetasF
+
+def no_resampler_thetas(counts, particlesP, norm_weights, thetas):
+    return counts, particlesP, norm_weights, thetas
+    
+'''
+# Resampling condition
+if np.exp(np.max(norm_weights))/np.exp(np.min(norm_weights)) > thresh:
+    resamples += 1 #tracker
+    # Systematic resampling
+    counts = resample(norm_weights, J)
+    particlesF = particlesP[counts]
+    weights = norm_weights[counts] - jax.lax.stop_gradient(norm_weights[counts]) - np.log(J)
+else:
+    particlesF = particlesP
+    weights = norm_weights
+'''
+
+
+def pfilter_helper(t, inputs):
+    particlesF, theta, covars, loglik, norm_weights, counts, ys, thresh, key = inputs
+    J = len(particlesF)
+    
[…]

### 2024-11-26 paper/appendix/consistency.tex

@@
     Explicitly, for almost every $\omega \in \Omega$ and every $\epsilon>0$, there exists some $\delta(\omega)>0$ such that for every $||\theta - \theta'||_{\infty}<\delta$ and every $J \in \mathbb{N}$ we have that
-    $$||\eta_n^J(h_\theta)(\omega)-\eta_n^J(h_{\theta'})(\omega)||_\infty < \epsilon.$$
+    \begin{equation}
+    \big||\eta_n^J(h_\theta)(\omega)-\eta_n^J(h_{\theta'})(\omega)\big||_\infty < \epsilon.
+    \end{equation}
     Then, by Arzela-Ascoli, there exists a uniformly convergent subsequence.
@@
 The sequence therefore converges uniformly to its limit $\lim_{J \to \infty} \eta_n^J(h_\theta)(\omega).$ Therefore, with uniform convergence for the derivatives established, we can swap the limit and derivative, and obtain that for almost every $\omega \in \Omega$,
-    $$\lim_{J \to \infty} \eta_n^J(h_\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \eta_n^J(h_\theta)(\omega).$$
+    \begin{equation}
+    \lim_{J \to \infty} \eta_n^J(h_\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \eta_n^J(h_\theta)(\omega).
+    \end{equation}
 Again from Theorem \ref{thm:off-parameter-targeting}, we know that
-    $$\eta_n^J(h_\theta)(\omega) \to \eta_n(h_\theta) = h^*_\theta$$ for almost every $\omega\in\Omega$.
+    \begin{equation}\eta_n^J(h_\theta)(\omega) \to \eta_n(h_\theta) = h^*_\theta\end{equation} for almost every $\omega\in\Omega$.
     We then have that for almost every $\omega \in \Omega$, 
-    $$\lim_{J \to \infty} \eta_n^J(h_\theta)(\omega) = \nabla_\theta \eta_n(h_\theta) = \nabla_\theta h^*_\theta,$$
+    \begin{equation}
+    \lim_{J \to \infty} \eta_n^J(h_\theta)(\omega) = \nabla_\theta \eta_n(h_\theta) = \nabla_\theta h^*_\theta,
+    \end{equation}
     as we wanted. 
@@
 %     Explicitly, for almost every $\omega \in \Omega$ and every $\epsilon>0$, there exists some $\delta(\omega)>0$ such that for every $||\theta - \theta'||_{\infty}<\delta$ and every $J \in \mathbb{N}$ we have that
-%     $$||\nabla_\theta \hat\lik^1_J(\theta)(\omega)-\nabla_\theta \hat\lik^1_J(\theta')(\omega)||_\infty < \epsilon.$$
+%     \begin{equation}||\nabla_\theta \hat\lik^1_J(\theta)(\omega)-\nabla_\theta \hat\lik^1_J(\theta')(\omega)||_\infty < \epsilon.\end{equation}
 
@@
 The sequence therefore converges uniformly to its limit $\lim_{J \to \infty} \nabla_\theta \hat\lik^1_J(\theta)(\omega).$ Therefore, with uniform convergence for the derivatives established, we can swap the limit and derivative, and obtain that for almost every $\omega \in \Omega$,
-%     $$\lim_{J \to \infty} \nabla_\theta \hat\lik^1_J(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik^1_J(\theta)(\omega).$$
+%     \begin{equation}\lim_{J \to \infty} \nabla_\theta \hat\lik^1_J(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik^1_J(\theta)(\omega).\end{equation}
 
 %     From Lemma \ref{lem:lik-proper-weight}, we know that
-%     $$\hat{\lik}^1_J(\theta)(\omega) \to \lik(\theta)$$ for almost every $\omega\in\Omega$.
+%     \begin{equation}\hat{\lik}^1_J(\theta)(\omega) \to \lik(\theta)\end{equation} for almost every $\omega\in\Omega$.
 
 %     We then have that for almost every $\omega \in \Omega$, 
-%     $$\lim_{J \to \infty} \nabla_\theta \hat\lik^1_J(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik^1_J(\theta)(\omega) = \nabla_\theta \lik(\theta),$$
+%     \begin{equation}\lim_{J \to \infty} \nabla_\theta \hat\lik^1_J(\theta)(\omega) = \nabla_\theta \lim_{J \to \infty} \hat\lik^1_J(\theta)(\omega) = \nabla_\theta \lik(\theta),\end{equation}
 %     as we wanted. The result then follows by the continuous mapping theorem.

### 2024-11-26 paper/ms.tex
Updates from Overleaf

@@
   \centering
-    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2}}]{imgs/095/mop.png}
+    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2.2}}]{imgs/095/mop.png}
     \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2.2}}]{imgs/095/biasvar.png}
     \arxiv{}{\vspace*{-3ex}}
-    \caption{\textbf{Left:} Non‐differentiable likelihood estimates and our smooth off‐parameter approximations that hold locally around the baseline parameter $\phi$.
-\textbf{Right:} Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}.
+    \caption{\textbf{A:} Non‐differentiable likelihood estimates and our smooth off‐parameter approximations that hold locally around the baseline parameter $\phi$.
+\textbf{B:} Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}.
 We display the MSE of score estimates for the trend in transmission at the MLE.
@@
     \caption{Scatterplots depicting the performance of IFAD against that of IF2.
-\textbf{Left:} Paired searches from the same starting point.
+\textbf{A:} Paired searches from the same starting point.
 Tuning $\alpha$ allows IFAD-$0.97$ to strictly improve on IF2, \cite{poyiadjis11}, and \cite{naesseth18}, on almost every initialization.
-\textbf{Right:} Q-Q plot of ranked IFAD searches against ranked IF2 searches.
+\textbf{B:} Q-Q plot of ranked IFAD searches against ranked IF2 searches.
 IFAD performs best on average, and approaches the MLE while no IF2 search successfully gets within 7 log-likelihood units of it.
@@
     %\arxiv{}{\vspace*{-1mm}}
-    \caption{\textbf{Left:} Raincloud plot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs, representing the common procedure of running a few searches and choosing the best one.
-\textbf{Right:} Raincloud plot of all searches.
+    \caption{\textbf{A:} Raincloud plot depicting the performance of IFAD and IF2 where we plot the results of the best run out of every ten runs, representing the common procedure of running a few searches and choosing the best one.
+\textbf{B:} Raincloud plot of all searches.
 IFAD outperforms all other methods.

### 2024-11-27 paper/ms.tex
almost ready for jrssb

@@
   \centering
+
+    \arxiv{}{\vspace*{0.15ex}}
     \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2.2}}]{imgs/095/mop.png}
-    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2.2}}]{imgs/095/biasvar.png}
-    \arxiv{}{\vspace*{-3ex}}
-    \caption{\textbf{A:} Non‐differentiable likelihood estimates and our smooth off‐parameter approximations that hold locally around the baseline parameter $\phi$.
-\textbf{B:} Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}.
-We display the MSE of score estimates for the trend in transmission at the MLE.
-Our novel contribution of $\alpha \in (0,1)$ significantly improves the quality of the likelihood approximation, and therefore the gradient estimates.
-}
+    \includegraphics[width=\arxiv{10cm}{\textwidth/\real{2.21}}]{imgs/095/biasvar.png}
+    \arxiv{}{\vspace*{-2ex}}
+    \caption{\textbf{A:} Non‐differentiable likelihood estimates and smooth off‐parameter approximations that hold locally around the baseline parameter $\phi$.
+\textbf{B:} Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$.
+We display the MSE of score estimates for the trend in transmission at the MLE for the Dhaka cholera model of \cite{king08}.
+Our novel contribution of $\alpha \in (0,1)$ significantly improves the quality of the likelihood approximation and the score estimates.}
     \label{fig:biasvar}
-    \arxiv{}{\vspace*{-4ex}}
+    \arxiv{}{\vspace*{-2ex}}
 \end{figure}
@@
 We defer the proof to \arxiv{Appendix~\ref{appendix:functional}}{the supplementary material}. 
-The argument relies on a useful decomposition of the after-resampling estimate $L_n^{A,\theta,\alpha}$. 
-Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting $\theta=\phi$ implies that $w_{n,j}^{P,\theta}$ evaluates to $1$.
+The argument relies on a useful decomposition of the after-resampling estimate $L_n^{A,\theta,\alpha}$, repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting $\theta=\phi$ implies that $w_{n,j}^{P,\theta}$ evaluates to $1$.
 
@@
 
-\begin{figure}
+\begin{figure}[htbp!]
     \centering
@@
     \vspace*{-7mm}
-    \caption{A compartment flow diagram for the Dhaka cholera model from \cite{king08}.}
+    \caption{A compartment flow diagram for the \cite{king08} Dhaka cholera model.}
     \label{fig:tikz-cholera}
@@
 
-\arxiv{}{\vspace*{-1mm}}
+%\arxiv{}{\vspace*{-1mm}}
 \begin{table}[htbp!]
@@
     \centering
+    \arxiv{}{\vspace*{0.1ex}}
     \includegraphics[width=\arxiv{8cm}{\textwidth/\real{2.2}}]{imgs/095/boxplot.png}

### 2024-11-27 paper/appendix/bayes.tex
changed plots
Updates from Overleaf

--- 2024-07-10 paper/appendix/bayes.tex
@@
 \end{figure}
+\FloatBarrier
+\newpage
 
 \begin{figure}[H]
-    \centering
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Recovery_Rate.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Death_Rate.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Reciprocal_of_Mean_Immunity_Duration.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Trend_in_Force_of_Infection.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Noise_in_Force_of_Infection.png}
-    \includegraphics[scale=0.33]{imgs/pmcmc/nuts_eb/Measurement_Noise.png}
+    \centering\
+    \includegraphics[scale=0.30]{imgs/pmcmc/nuts_eb/Recovery_Rate.png}
+    \includegraphics[scale=0.30]{imgs/pmcmc/nuts_eb/Death_Rate.png}
+    \includegraphics[scale=0.30]{imgs/pmcmc/nuts_eb/Reciprocal_of_Mean_Immunity_Duration.png}
+    \includegraphics[scale=0.30]{imgs/pmcmc/nuts_eb/Trend_in_Force_of_Infection.png}
+    \includegraphics[scale=0.30]{imgs/pmcmc/nuts_eb/Noise_in_Force_of_Infection.png}
+    \includegraphics[scale=0.30]{imgs/pmcmc/nuts_eb/Measurement_Noise.png}
     \caption{Posterior estimates from NUTS powered by MOP-$\alpha$ with the informative nonparametric empirical prior obtained from a kernel density estimator on the IF2 parameter cloud.

### 2024-11-27 paper/appendix/biasvar.tex
Updates from Overleaf

@@
 To our knowledge, this result for the variance of the gradient estimate of \cite{naesseth18} is novel.
-
 \begin{thm}[Variance of MOP-$0$ Gradient Estimate]
@@
 \end{thm}
-
 \begin{proof}
@@
 We can then bound $\|\nabla_\theta\hat\ell^\alpha(\theta) - \nabla_\theta \hat s_k^\alpha(\theta)\|$ as follows:
+{\small
 \begin{align}
@@
 \end{align}
+}
 and each of these two terms is bounded by $\frac{\alpha^k}{1-\alpha}Np \, G'(\theta)^2$. So we now know that
@@
 \begin{equation}
-\Var\big(\nabla_\theta \hat\ell^\alpha(\theta)\big) \lesssim \min_{k\leq N} \left(\frac{k^2p \, G'(\theta)^2Np}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}Np \, G'(\theta)^2 \right).\end{equation}
+\Var\big(\nabla_\theta \hat\ell^\alpha(\theta)\big) \lesssim \min_{k\leq N} \left(\frac{k^2 \, G'(\theta)^2Np}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}Np \, G'(\theta)^2 \right).\end{equation}
 \end{proof}
@@
 \\
-&& \hspace{-40mm}
+&& \hspace{-65mm}
 \sum_{n=1}^N \|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)\|_2^2 + \sum_{n=1}^N \|\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)\|_2^2.
@@
 the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.
-
 Then, we can bound 
@@
 \begin{eqnarray} \nonumber
-\hspace{-10mm} \sum_{n=1}^N\big\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)\big\|_2^2 &\leq 
-\\ &&\hspace{-45mm} 
+\hspace{-5mm} \sum_{n=1}^N\big\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)\big\|_2^2 &\leq 
+\\ &&\hspace{-70mm} 
 \sum_{n=1}^N \big\|\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta)\big\|_2^2 + \sum_{n=1}^N \big\|\nabla_\theta\hat s_{n,1|\hat\pi_{n-1}=\pi_{n-1}}^1(\theta) - \nabla_\theta\hat s_{n,1}^1(\theta)\big\|_2^2.

### 2024-11-27 paper/appendix/convergence.tex

@@
 
+%\ed{TECHNICALLY, IF WE CLAIM THAT A PROOF FOR ALPHA=1 CAN BE GENERALIZED TO ANY ALPHA, WE ARE CLAIMING A PROOF FOR ALL ALPHA, AND SO WE DO NOT NEED TO SAY WE ARE ONLY PROVING IT FOR ALPHA=1.
+IF WE REALLY ONLY CLAIM TO HAVE SHOWN IT FOR ALPHA=1, THAT SHOULD BE IN THE STATEMENT.
+IF WE CLAIM THE WHOLE RESULT, PERHAPS WE CAN POSTPONE THE DISCUSSION OF THIS 2-STAGE METHOD TO THE SUPPLEMENT?
+AT THIS POINT, ``THE REQUIRED EPSILON-APPROXIMATION'' IS NOT DEFINED FOR THE READER, SO IT IS HARD TO COMPREHEND THE ARGUMENT.}
+For simplicity, we only provide the argument for MOP-$1$ and note that the general case follows as the required $\epsilon$-approximation can still be obtained for sufficiently large $\alpha$. 
+When $\alpha=1$, the required $J$ is given in the below lemmas.
+
 The convergence analysis in Theorem \ref{thm:mop-convergence} is limited to the case where $-\ell$ is $\gamma$-strongly convex.
-Though it is true that in a neighborhood of the optimum local asymptotic normality holds and the log-likelihood is strongly convex in this neighborhood, in practice likelihood surfaces for POMPs are often highly nonconvex globally.
+Under regularity conditions, we expect the log-likelihood to be approximately quadratic in a neighborhood of its maximum, though in practice the likelihood surfaces for POMP models are often highly nonconvex globally.
 The convergence to an optimum, local or global, must therefore be sensitive to initialization.
@@
 
-We will seek to use the concentration inequality of \cite{delMoral11} to bound the deviation of the gradient estimate from the gradient of the negative log-likelihood in the sup norm with a union bound.
+We use the concentration inequality of \cite{delMoral11} to bound the deviation of the gradient estimate from the gradient of the negative log-likelihood in the sup norm with a union bound.
 Fix $\theta = \phi$.
@@
 
-Now we apply the Hoeffding-type concentration inequality from Del Moral and Rio \cite{delMoral11} and a union bound over each $\varphi_n^i(x_{n,j}^{F,\theta})$, totaling $N$ timesteps and $p$ parameters, to find that
+Now we apply the Hoeffding-type concentration inequality from \cite{delMoral11} and a union bound over each $\varphi_n^i(x_{n,j}^{F,\theta})$, totaling $N$ timesteps and $p$ parameters, to find that
 \begin{align}
@@
 Define $\theta_\eta = \theta_m + \eta p_m$, where $p_m=-(H(\theta_m))^{-1}g(\theta_m)$. 
-As in Roosta-Khorasani and Mahoney \cite{mahoney16}, we want to show there is some iteration-independent $\tilde{\eta}>0$ such that the Armijo condition
+As in \cite{mahoney16}, we want to show there is some iteration-independent $\tilde{\eta}>0$ such that the Armijo condition
 \begin{equation}
@@
 holds for any $0< \eta < \tilde{\eta}$ and some $\beta \in (0,1)$.
-By an argument found in the beginning of the proof of Theorem 6 in \cite{mahoney16}, we have that choosing $J$ such that $\|\nabla_\theta\hat{\ell}(\theta_m) - \nabla_\theta \ell(\theta_m)\| \leq \epsilon$ and $\lambda_{\min}(H(\theta_m)) \geq c>0$ for each $m$, yields
+By an argument found in the beginning of the proof of Theorem 6 of \cite{mahoney16}, we have that choosing $J$ such that $\|\nabla_\theta\hat{\ell}(\theta_m) - \nabla_\theta \ell(\theta_m)\| \leq \epsilon$ and $\lambda_{\min}(H(\theta_m)) \geq c>0$ for each $m$, yields
 \begin{align}

### 2024-11-27 paper/appendix/coverletter.tex
almost ready for jrssb

@@
 The contribution of this paper has widespread implications due to the diverse uses of this model class in ecology, engineering, epidemiology, finance, and elsewhere.
-Our method promises substantially improved inference wherever simulation-based inference is used for the full-information likelihood in complex partially observed Markov processes.
+Our method promises substantially improved inference wherever simulation-based full-information inference is used for complex partially observed Markov processes.
 That is why we believe that publication of our article in JRSS-B would facilitate a broad range of scientific advances.
@@
 Nicolas Chopin is a leading SMC theorist who is also involved in developing SMC computation so he would also be a qualified referee.
-It would assist the evaluation of our contribution if at least one referee could be drawn from this select group of experts.
 
@@
 Therefore, we ask for at least one referee who has experience working outside the Bayesian paradigm.
-Additionally, due to a disagreement that Ed Ionides had with Michael Whitehouse, Nick Whiteley, and Lorenzo Rimella regarding the correctness of a numerical experiment within their paper on Poisson Approximate Likelihoods published at JRSS-B, we similarly ask that none of the referees be drawn from the authors of that paper to preserve the impartiality of reviews.
-
+Additionally, due to a recent disagreement that Ed Ionides had with Michael Whitehouse, Nick Whiteley, and Lorenzo Rimella regarding the correctness of a numerical experiment within their paper on Poisson Approximate Likelihoods published at JRSS-B, we similarly ask that none of the referees be drawn from the authors of that paper to preserve the impartiality of reviews.
 \vspace{3ex}
@@
 
-\noindent Kevin Tan, Ed Ionides and Giles Hooker
+\noindent Kevin Tan, Giles Hooker and Ed Ionides
 \end{document}

### 2024-11-27 paper/appendix/feynman.tex
another day, another dollar

@@
-In this section, we introduce the Feynman-Kac convention of \cite{delMoral04} that has since become commonplace \cite{karjalainen23} for the analysis of the particle filter.
+In this section, we introduce the Feynman-Kac convention of \cite{delMoral04} that has since become commonplace \citep{karjalainen23} for the analysis of the particle filter.
 The mathematical formalization and notation introduced here will be adopted in the remainder of the analysis, in order to prove Theorems \ref{thm:mop-targeting}, \ref{thm:mop-grad-consistency}, and \ref{thm:mop-biasvar}.
@@
 
-\paragraph{Markov kernels and the process model:} A Markov kernel $M$ with source $\gX_1$ and target $\gX_2$ is a map $M: \gX_1 \times \mathcal{B}(\gX_2) \to [0,1]$ such that for every set $A \in \mathcal{B}(\gX_2)$ and every point $x \in \gX_1$, the map $x \mapsto M(x, A)$ is a measurable function of $x$, and the map $A \mapsto M(x,A)$ is a probability measure on $\gX_2$.
+\paragraph{\bf Markov kernels and the process model:} A Markov kernel $M$ with source $\gX_1$ and target $\gX_2$ is a map $M: \gX_1 \times \mathcal{B}(\gX_2) \to [0,1]$ such that for every set $A \in \mathcal{B}(\gX_2)$ and every point $x \in \gX_1$, the map $x \mapsto M(x, A)$ is a measurable function of $x$, and the map $A \mapsto M(x,A)$ is a probability measure on $\gX_2$.
 The quantity $M(x,A)$ can be thought of as the probability of transitioning to the set $A$ given that we are at the point $x$.
@@
 
-\paragraph{Markov kernels and measures:} For any measure $\eta$, any Markov kernel $M$ on $\gX$, any point $x \in \gX$ and any measurable subset $A \subseteq \gX$, let 
+\paragraph{\bf Markov kernels and measures:} For any measure $\eta$, any Markov kernel $M$ on $\gX$, any point $x \in \gX$ and any measurable subset $A \subseteq \gX$, let 
 \begin{align}
@@
 
-\paragraph{Compositions of Markov kernels:} The composition of a Markov kernel $M_1$ with another Markov kernel $M_2$ is another Markov kernel, given by 
+\paragraph{\bf Compositions of Markov kernels:} The composition of a Markov kernel $M_1$ with another Markov kernel $M_2$ is another Markov kernel, given by 
 \begin{equation}
@@
 
-\paragraph{Total variation distance:} The total variation distance between two measures $\mu$ and $\nu$ on $\gX$ is
+\paragraph{\bf Total variation distance:} The total variation distance between two measures $\mu$ and $\nu$ on $\gX$ is
 \begin{equation}
@@
 
-\paragraph{Dobrushin contraction:} The Dobrushin contraction coefficient $\beta_{\text{TV}}$ of a Markov kernel $M$ is given by
+\paragraph{\bf Dobrushin contraction:} The Dobrushin contraction coefficient $\beta_{\text{TV}}$ of a Markov kernel $M$ is given by
 \begin{equation}
-\beta_{\mathrm{TV}}(M)=\sup _{x, y \in \gX}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}.    
+\beta_{\mathrm{TV}}(M)=\sup _{x, y \in \gX}\big\|M(x, \cdot)-M(y, \cdot)\big\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}.
 \end{equation}
 
-\paragraph{Potential functions and the measurement model:} A potential function $G : \gX \to [0,\infty)$ is a non-negative function of an element of the state space $x \in \gX$.
+\paragraph{\bf Potential functions and the measurement model:} A potential function $G : \gX \to [0,\infty)$ is a non-negative function of an element of the state space $x \in \gX$. 
 In our case, this corresponds to the measurement model, and in our previous notation is written as $g_{n,j} = f_{Y_n|X_n}(y_n^*|x_{n,j}^F) = G_n(x_{n,j}^F)$, where in a slight abuse of notation we suppress the dependence on $\theta$ for notational simplicity.
@@
 
-\paragraph{Feynman-Kac models:} A Feynman-Kac model on $\gX$ is a tuple $(\pi_0, (M_n)_{n=1}^N, (G_n)_{n=1}^N)$ of an initial probability measure on the state space $\pi_0$, a sequence of transition kernels $(M_n)_{n=1}^N$, and a sequence of potential functions $(G_n)_{n=1}^N$.
+\paragraph{\bf Feynman-Kac models:} A Feynman-Kac model on $\gX$ is a tuple $(\pi_0, (M_n)_{n=1}^N, (G_n)_{n=1}^N)$ of an initial probability measure on the state space $\pi_0$, a sequence of transition kernels $(M_n)_{n=1}^N$, and a sequence of potential functions $(G_n)_{n=1}^N$.
 In the notation used in the main text, this corresponds to the starting distribution $f_{X_0}$, the sequence of transition densities $f_{X_{n}|X_{n-1}}$, and the measurement densities $f_{Y_n|X_n}$.
@@
 \end{itemize}
-\paragraph{The particle filter:} The particle filter then yields a Monte Carlo approximation to the above Feynman-Kac model, via a sequence of mixture Dirac measures.
+\paragraph{\bf The particle filter:} The particle filter then yields a Monte Carlo approximation to the above Feynman-Kac model, via a sequence of mixture Dirac measures.
 When one resamples at every timestep, the prediction measure at timestep $n$ is then given by
@@
 
-\paragraph{Forgetting of the particle filter:} 
+\paragraph{\bf Forgetting of the particle filter:} 
 The above formalization yields a result from \cite{karjalainen23} on the forgetting of the particle filter that we require for our analysis of the bias, variance, and error of MOP-$\alpha$.  

### 2024-11-27 paper/appendix/functional.tex
update from overleaf

@@
     \end{equation}
-    yielding the estimator of \cite{poyiadjis11, scibior21} with the bootstrap filter.
+    yielding the bootstrap filter estimator of \cite{poyiadjis11} and \cite{scibior21}.
 \end{lem}
@@
 \end{equation}
-Substituting the above identity into the log-likelihood decomposition obtained earlier \ed{USE EQUATION NUMBERS TO MAKE THIS ARGUMENT MORE CONCRETE} yields
+Substituting the above identity into the log-likelihood decomposition obtained earlier in Equation \ref{eq:log-derivative-identity} yields
 \begin{equation}

### 2024-11-27 paper/ms.tex

@@
 We write $P(t)$ for the total population, and $M_n$ for the cholera deaths in each month.
-As in \citep{king08, ionides15}, the transition dynamics follow a series of stochastic differential equations:
+As in \cite{king08}, the transition dynamics follow a series of stochastic differential equations:
 \vspace*{-1mm}
@@
 
-Although we have shown the MOP-$1$ gradient estimate yields the estimate of \citep{poyiadjis11, scibior21} when applied to the bootstrap filter, \cite{poyiadjis11, scibior21} estimate the Fisher score by
+Although we have shown the MOP-$1$ gradient estimate yields the estimate of \cite{poyiadjis11} and \cite{scibior21} when applied to the bootstrap filter, those authors estimate the Fisher score by 
 $\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{0:n,j}^{A, F,\theta}, y_{1:N}^* ; \theta)$ and not
@@
 
-Our implementation of the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \cite{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, even with a simulator written in Python.
+Our implementation of the particle filter, simulator, and MOP-$\alpha$ in \texttt{JAX} \citep{jax} enabled us to take advantage of just-in-time compilation and GPU acceleration, even with a simulator written in Python.
 This led to a 16x speedup (379ms vs 6.29s on a Intel i9-13900K CPU and NVIDIA RTX3090 GPU) over the CPU-only implementation of the particle filter (with a simulator written in C++) in the \texttt{pomp} package of \cite{king16}.

### 2024-12-07 paper/ms.tex
810 presentation

@@
     \includegraphics[width=\arxiv{\textwidth/\real{1.25}}{\textwidth}/\real{1.25}]{imgs/pmcmc/nuts_eb.png}
-    \caption{Histogram and density estimate of the posterior of $\beta_{\text{trend}}$, as well as convergence diagnostics for NUTS with $4$ chains.
+    \caption{Convergence diagnostics for NUTS with $4$ chains, where we plot $1000$ samples from the posterior of $\beta_{\text{trend}}$ within the Dhaka cholera model of \cite{king08}.
+\textbf{A:} Histogram of posterior samples.
+\textbf{B:} density estimate of the posterior.
+\textbf{C:} Autocorrelation plot for MCMC samples.
+\textbf{D:} Trace plot for MCMC samples, pre and post burn-in.
 NUTS mixes quickly when regularized by an empirical prior with a standard deviation of $6.3\%$ of the MLE, and the posterior estimates from each chain share roughly the same posterior mode.}

### 2024-12-07 presentation/810class14.tex
810 presentation

--- 2024-03-06 presentation/presentation.tex
@@
 \title[AD for POMPs]{Automatic Differentiation Accelerates Inference for Partially Observed Stochastic Processes}
-\subtitle{\small{Inference for Partially Observed Structured Dynamic Systems}}
-\author[Kevin Tan]{Kevin Tan \\ \small{University of Pennsylvania, Department of Statistics and Data Science} \\\vspace{0.5cm} \small{Joint work with Giles Hooker and Edward Ionides}}
-\date[January 4, 2024]{IMS Asia Pacific Rim Meeting \\ January 4, 2024}
+\subtitle{\url{https://arxiv.org/abs/2407.03085}}
+\author[Tan, Hooker \& Ionides]{Kevin Tan$^1$, Giles Hooker$^1$ and Ed Ionides$^2$ \\ {\small $^1$University of Pennsylvania, $^2$University of Michigan}} 
+\date[STATS 810]{STATS 810 \\December 9, 2024}
 
@@
 \usepackage{multicol,lipsum,xcolor}
-\usepackage{hyperref}       % hyperlinks
+%\usepackage{hyperref}       % hyperlinks
 \usepackage{url}            % simple URL typesetting
@@
 
-\include{mop/macros}
+\input{mop/macros}
 
@@
             \\ 
-            -- Graphic from \cite{berg2019pfilterimage}.}
+            -- Graphic from \cite{berg19}.}
         \end{minipage}
@@
             \\$f(x,y) = x^2+xy+2$, $f'(x,y) = (2x+y,x).$
-            -- Figure from \href{https://avinashselvam.medium.com/automatic-differentiation-explained-9f02c74e9a90}{Medium}.}
+            -- Figure from             %\url{https://avinashselvam.medium.com/automatic%-differentiation-explained-9f02c74e9a90}  
+            \url{https://avinashselvam.medium.com}
+            }
         \end{minipage}
@@
     \begin{itemize}
-        \item \pause IF2 (\cite{ionides15}), particle MCMC (\cite{doucet2010pmcmc}), only full-information methods that can deal with this.
+        \item \pause IF2 \citep{ionides15}, particle MCMC \citep{andrieu10}, only full-information methods that can deal with this.
     \end{itemize}
@@
     \begin{itemize}
-        \item Original iterated filtering algorithm: (noisy) approximation of score (\cite{ionides06-pnas}). 
+        \item Original iterated filtering algorithm: (noisy) approximation of score \citep{ionides06-pnas}. 
         %\item \pause Performed a single "gradient step" at the end of each filtering iteration.
@@
          \hline
-         \cite{blei2018vsmc} & Backprop through vanilla PF \\ 
+         \cite{naesseth18} & Backprop through vanilla PF \\ 
          & Asymptotic bias \\
@@
         \begin{itemize}
-            \item \pause Gradient estimates of \cite{blei2018vsmc} (MOP-0), \cite{poyiadjis11} and \cite{scibior2021dpf} (MOP-1) are special cases.
+            \item \pause Gradient estimates of \cite{naesseth18} (MOP-0), \cite{poyiadjis11} and \cite{scibior21} (MOP-1) are special cases.
             \item \pause Does not need transition densities, only a differentiable simulator.
@@
     \item \pause\textbf{Exponentially-weighted moving average} where $\alpha$ controls the amount of discounting.
-    \item \pause Bias small if $y_{n+1:N}^*$ not informative for $x_n$ given $y_{0:n}^*$ (\cite{corenflos21}).
+    \item \pause Bias small if $y_{n+1:N}^*$ not informative for $x_n$ given $y_{0:n}^*$ \citep{corenflos21}.
 \end{itemize}
-
-\end{frame}
-
-\begin{frame}{Bias-Variance Tradeoff with $\alpha$}
-\begin{figure}
-    \centering
-    \includegraphics[scale=0.5]{imgs/mlegrad.png}
-    \caption{Gradient estimates at MLE for trend in \cite{king08} cholera model. When $\alpha=1$, estimate close to unbiased with high variance. When $\alpha=0$, lower variance but biased.}
-    \label{fig:bias-variance}
-\end{figure}
 
@@
     $$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{0:N}}(y_{1:N}^*|x_{1:N,j}^{A, F,\theta}),$$
-    and when $\alpha=0,$ is the gradient estimator of \cite{blei2018vsmc},
+    and when $\alpha=0,$ is the gradient estimator of \cite{naesseth18},
     $$
@@
     $$
-    i.e. differentiating through a vanilla particle filter (\cite{scibior21}).
+    i.e. differentiating through a vanilla particle filter \citep{scibior21}.
 \end{prop}
@@
     \begin{itemize}
-        \item Similar to \cite{mahoney16}, uses concentration inequalities from \cite{delmoral2011ci}.
+        \item Similar to \cite{mahoney16}, uses concentration inequalities from \cite{delmoral11}.
         \item Gradient stage of IFAD-1 converges linearly to the MLE if:
@@
     \begin{itemize}
-        \item \pause Gradient takes 3.75x time of \texttt{pfilter()}, in line with cheap gradient principle (\cite{kakade2019provably}).
+        \item \pause Gradient takes 3.75x time of \texttt{pfilter()}, in line with cheap gradient principle \citep{kakade2019provably}.
     \end{itemize}
     \item \pause Runtime \textbf{linear, and not quadratic}, in number of particles.
-    \item \pause Re-implementation in \texttt{JAX} (\cite{jax2018github}) led to 16x speedup v.s. \texttt{pomp} package of \cite{king08}.
+    \item \pause Re-implementation in \texttt{JAX} \citep{jax} led to 16x speedup v.s. \texttt{pomp} package of \cite{king08}.
 \end{itemize}
@@
         \item \pause Extensions to panel and spatiotemporal data, e.g. with a block particle filter in the lens of \cite{ionides22} and \cite{ning23}.
-        \item \pause Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual}.
+        \item \pause Python counterpart to the popular \texttt{pomp} R package by \citet{king16}.
     \end{itemize}
@@
 \bibliographystyle{apalike}
-\bibliography{mop/bib-ifad, mop/bib-ref}
+\bibliography{paper/bib-ifad}
 
@@
         \item 
-        \pause \cite{blei2018vsmc}, backpropagate through vanilla particle filter.
+        \pause \cite{naesseth18}, backpropagate through vanilla particle filter.
         \item 
@@
         \begin{itemize}
-            \item \pause Either one drops resampling terms and accepts asymptotic bias (\cite{blei2018vsmc}), or has variance quadratic in horizon (\cite{poyiadjis11}, \cite{scibior21})
+            \item \pause Either one drops resampling terms and accepts asymptotic bias (\cite{naesseth18}), or has variance quadratic in horizon (\cite{poyiadjis11}, \cite{scibior21})
         \end{itemize}

### 2024-12-11 presentation/810class14.tex
submitted to JRSSB

@@
 \begin{frame}{What is this reweighting?}
-    Ordinarily, the likelihood estimate for $\phi$ is 
-    $$\hat\lik(\phi) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J \left(\underbrace{f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}, \phi)}_{\textcolor{blue}{\text{Measurement Model}}}
+    The properly weighted estimate at $\theta$ with proposal from $\phi$ is 
+    $$\hat\lik(\theta) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J \left(\underbrace{f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}, \theta)}_{\textcolor{blue}{\text{Measurement Model}}}
     \cdot 
-    \underbrace{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi} | x_{n-1,j}^{F,\phi}, \phi)}_{\textcolor{red}{\text{Process model}}}\right).$$
+    \frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi} | x_{n-1,j}^{F,\phi}, \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi} | x_{n-1,j}^{F,\phi}, \phi)}
+    \right).$$
     \pause
-    So to estimate the likelihood for $\theta$, we can correct 
+    Which we write as
         $$
         \hat\lik(\theta) 
-        = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}, \phi) \cdot 
-        f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi} | x_{n-1,j}^{F,\phi}, \phi) 
+        = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}, \phi) 
         \cdot \underbrace{s_{n,j} \cdot r_{n,j}}_{\text{Correction Term}},
@@
     -- \textbf{Right:} MOP alone drastically underperforms all other methods, failing to get close to the MLE. \\
-    -- IF2 warm-start necessary in challenging nonconvex and noisy problems.}
+    -- IF2 warm start necessary in challenging nonconvex and noisy problems.}
     \label{fig:boxplot-search}
@@
     \includegraphics[scale=0.35]{imgs/095/hist.png}
-    \caption{Histogram comparison of IFAD, IF2, and MOP alone. Right is better.}
+    \caption{Histogram comparison of IFAD (with various $\alpha$) and IF2. Right is better.}
     \label{fig:hist-all}
@@
         \item \pause Gradient steps perform better at fine-grained refinement.
-        \item \pause Performing gradient steps alone without a warm-start leads to the search getting stuck in local minima and saddle points.
+        \item \pause Performing gradient steps alone without a warm start leads to the search getting stuck in local minima and saddle points.
         \item \pause IFAD combines the best of IF2 and MOP, approaching the MLE quickly and successfully performing refinement -- even on a very difficult global search problem.
